### 2016-10-31 analysis/0_field_work_proposal.Rmd
add fieldwork proposal

---
title: "Field work"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    orientation: rows
---

```{r setup, include=FALSE}
library('flexdashboard')
```

```{r pacakges}
library('rgdal')
library("leaflet") 
library("sp")
library("raster")

library('dplyr')

library('ggplot2')

```

```{r readData}
# mymachine <- '/Users/ajpeluLap/' 
mymachine <- '/Users/ajpelu/' 
di <- paste0(mymachine, 'Dropbox/phd/phd_repos/qpyr_dendro/')

# Data 
ot <- read.csv(paste0(di,'data_raw/geoinfo/other_projects.csv'), 
               header = TRUE, sep = ',')
mig <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'migrame_transectos', verbose = FALSE)

# Spatial info
ots <- SpatialPointsDataFrame(coords = ot[,c('x','y')], data = ot,
                               proj4string = CRS("+proj=utm +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))

mig <- spTransform(mig, CRS("+init=epsg:4236"))
ots <- spTransform(ots, CRS("+init=epsg:4236"))

# Select only transect of forest (migrame)
mig_forest <- mig[mig$tipo == "ROBLEDAL",]
mig_forest_ca <- mig_forest[mig_forest$localidad == "CANAR",]
mig_forest_sj <- mig_forest[mig_forest$localidad == "SANJUAN",]

ots_phe <- ots[ots$notes == 'Fenologia_Obsnev',]
ots_gea <- ots[ots$notes == 'GeaIzquierdo2014',]

```

```{r}
mymap <- leaflet() %>% 
  # BaseMaps 
  addWMSTiles('http://www.ideandalucia.es/wms/ortofoto2009?',
              layers = 'oca10_2008-09',
              options = WMSTileOptions(format = "image/png", transparent = FALSE),
              attribution = '<a href="http://www.juntadeandalucia.es/institutodeestadisticaycartografia" target="_blank">Instituto de Estadística y Cartografía de Andalucía</a>',
              group = 'Ortophotography 2009') %>%
   addWMSTiles('http://www.ideandalucia.es/services/toporaster10/wms?',
              layers = 'toporaster10',
              options = WMSTileOptions(format = "image/png", transparent = FALSE),
              attribution = '<a href="http://www.juntadeandalucia.es/institutodeestadisticaycartografia" target="_blank">Instituto de Estadística y Cartografía de Andalucía</a>',
              group = 'Topographical') %>%

  # Layers control
  addLayersControl(position = 'bottomright',
                   # baseGroups = c("Topographical","Ortophotography 2009","PNOA"),
                   baseGroups = c("Ortophotography 2009", "Topographical"),
                   overlayGroups = c("Migrame", "OBSNEV Fen", "Gea Izquierdo 2014"),
                   options = layersControlOptions(collapsed = FALSE))

```

San Juan
=======================================================================

```{r}
myext <- extent(mig_forest_sj)

mymap %>% 
  fitBounds(myext@xmin, myext@ymin, myext@xmax, myext@ymax) %>% 
  addPolylines(data=mig_forest_sj, color='blue', group='Migrame', popup=mig_forest_sj$nombre) %>% 
  addCircles(ots_phe, 
             lng=coordinates(ots_phe)[,'x'], 
             lat=coordinates(ots_phe)[,'y'], 
             group='OBSNEV Fen', radius=2, color = 'red', popup=ots_phe$name) %>% 
  addCircles(ots_gea, 
             lng=coordinates(ots_gea)[,'x'], 
             lat=coordinates(ots_gea)[,'y'], 
             group='Gea Izquierdo 2014', radius=2, color = 'green', popup=ots_gea$name)

```

Caniar
=======================================================================

```{r}
myext <- extent(mig_forest_ca)

mymap %>% 
  fitBounds(myext@xmin, myext@ymin, myext@xmax, myext@ymax) %>% 
  addPolylines(data=mig_forest_ca, color='blue', group='Migrame', popup=mig_forest_ca$nombre) %>% 
  addCircles(ots_phe, 
             lng=coordinates(ots_phe)[,'x'], 
             lat=coordinates(ots_phe)[,'y'], 
             group='OBSNEV Fen', radius=2, color = 'red', popup=ots_phe$name) %>% 
  addCircles(ots_gea, 
             lng=coordinates(ots_gea)[,'x'], 
             lat=coordinates(ots_gea)[,'y'], 
             group='Gea Izquierdo 2014', radius=2, color = 'green', popup=ots_gea$name)

```

### 2016-11-30 analysis/geo.Rmd
updating

---
title: "Get topographic info of gps points"
author: "AJ Perez-Luque (@ajpelu)"
date: "2016 Nov"
output:  
  md_document:
    variant: markdown_github
---

```{r packages, warning=FALSE, message=FALSE}}
library("rgdal")
library("sp")
library("raster")
library("dplyr")
```

# Read data 
```{r wd, echo=FALSE}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# Connect with NAS and copy the mde file located at 
# '/cartografia/Informacion_Ambiental/BASES_DE_REFERENCIA_TERRITORIAL/MDE_SN_AreaInfluecia_10m/mde_sn_area_influencia.asc' 
#---------------------------------
```

# Extract data
```{r} 
ditemp <- paste('/Users/', machine, '/Documents/', sep='')
setwd(ditemp)

# read DEM 
demsn <- raster("mde_sn_area_influencia.asc")

setwd(di)

## Projection 
# Get projection
projection(demsn)

# Set projection
crs(demsn) <- "+proj=utm +zone=30 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"

# Read points data

gps_point <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'dendro_garmin', verbose = FALSE)

# Read pixels data 
eviresi <- read.csv(file=paste(di, "/data/evi_resilience.csv", sep=""), header = TRUE, sep = ',')

## convert to spatial df 
eviresi_sp <- SpatialPointsDataFrame(coords = eviresi[,c('lng','lat')], data = eviresi,
                               proj4string = CRS("+init=epsg:4326"))
## reproject 
eviresi_sp <- spTransform(eviresi_sp, crs(demsn))

evi_elev <- extract(demsn, eviresi_sp, method='simple', sp=TRUE)

elev <- evi_elev@data %>% 
  dplyr::select(iv_malla_modi_id, mde_sn_area_influencia) %>%
  transmute(iv_malla_modi_id, elev = mde_sn_area_influencia)

# Export data
write.csv(elev, file=paste(di, "/data/elev.csv", sep=""), row.names = FALSE)
```
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2016-11-30 analysis/geo2.Rmd
updating

--- 2016-10-31 analysis/0_field_work_proposal.Rmd
@@
 ---
-title: "Field work"
+title: "Field work 2016"
 output: 
@@
 library("raster")
-
 library('dplyr')
-
 library('ggplot2')
-
 ```

### 2016-12-02 analysis/get_elevation_points.Rmd
get elevation points

--- 2016-11-30 analysis/geo.Rmd
@@
 ```{r wd, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
+# machine <- 'ajpelu'
+machine <- 'ajpeluLap'
 di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 # Read points data 
-
 gps_point <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'dendro_garmin', verbose = FALSE)
 
-# Read pixels data 
-eviresi <- read.csv(file=paste(di, "/data/evi_resilience.csv", sep=""), header = TRUE, sep = ',')
+## reproject 
+gps_point_re <- spTransform(gps_point, crs(demsn))
 
-## convert to spatial df 
-eviresi_sp <- SpatialPointsDataFrame(coords = eviresi[,c('lng','lat')], data = eviresi,
-                               proj4string = CRS("+init=epsg:4326"))
-## reproject 
-eviresi_sp <- spTransform(eviresi_sp, crs(demsn))
+gps_elev <- extract(demsn, gps_point_re, method='simple', sp=TRUE)
 
-evi_elev <- extract(demsn, eviresi_sp, method='simple', sp=TRUE)
-
-elev <- evi_elev@data %>% 
-  dplyr::select(iv_malla_modi_id, mde_sn_area_influencia) %>%
-  transmute(iv_malla_modi_id, elev = mde_sn_area_influencia)
+elev <- gps_elev@data %>% 
+  dplyr::select(lat, lon, elev, name, date, elev_mde = mde_sn_area_influencia)
 
@@
 ```
-title: "Untitled"
-output: html_document
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-## R Markdown
-
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
-
-```{r cars}
-summary(cars)
-```
-
-## Including Plots
-
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2016-12-07 analysis/geo2.Rmd
add geospatial analysis add notes dendro

@@
 ```{r readData}
-# mymachine <- '/Users/ajpeluLap/' 
-mymachine <- '/Users/ajpelu/' 
+mymachine <- '/Users/ajpeluLap/' 
+# mymachine <- '/Users/ajpelu/' 
 di <- paste0(mymachine, 'Dropbox/phd/phd_repos/qpyr_dendro/')
@@
 
+# field woek points 
+fw <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'dendro_garmin', verbose = FALSE)
+
 # Spatial info
@@
 
+# Reproject 
 mig <- spTransform(mig, CRS("+init=epsg:4236"))
 ots <- spTransform(ots, CRS("+init=epsg:4236"))
+fw <- spTransform(fw, CRS("+init=epsg:4236"))
 
@@
                    baseGroups = c("Ortophotography 2009", "Topographical"),
-                   overlayGroups = c("Migrame", "OBSNEV Fen", "Gea Izquierdo 2014"),
+                   overlayGroups = c("Field Work 2016", "Migrame", "OBSNEV Fen", "Gea Izquierdo 2014"),
                    options = layersControlOptions(collapsed = FALSE))
-
 ```
@@
              lat=coordinates(ots_gea)[,'y'], 
-             group='Gea Izquierdo 2014', radius=2, color = 'green', popup=ots_gea$name)
+             group='Gea Izquierdo 2014', radius=2, color = 'green', popup=ots_gea$name) %>% 
+  addCircles(fw, 
+             lng=coordinates(fw)[,'coords.x1'], 
+             lat=coordinates(fw)[,'coords.x2'], 
+             group='Field Work 2016', radius=2, color = 'yellow', popup=fw$name)
 
@@
              lat=coordinates(ots_gea)[,'y'], 
-             group='Gea Izquierdo 2014', radius=2, color = 'green', popup=ots_gea$name)
-
+             group='Gea Izquierdo 2014', radius=2, color = 'green', popup=ots_gea$name) %>% 
+  addCircles(fw, 
+             lng=coordinates(fw)[,'coords.x1'], 
+             lat=coordinates(fw)[,'coords.x2'], 
+             group='Field Work 2016', radius=2, color = 'yellow', popup=fw$name)
 ```

### 2016-12-07 man/notes_dendro.Rmd
add geospatial analysis add notes dendro

---
title: "Dendrocronology notes"
author: "AJ Perez-Luque (@ajpelu)"
date: "2016 Nov"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: references.bib
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Ring Width measurement

## Notes from [@GeaIzquierdo2012Tree]

* Two cores were taken at 1.30 m from each selected tree
* The stem diamenter and tree height were measured
* Cores were air dried, fixed in woodend mounts, sanded and RW measured 
* Ring-width series were visually cross-dated and the dating was checked using COFECHA `$TODO$` Check COFECHA
* Ring-width inidces were obtained by removing the age trend using 30-year individual splines and autocorrelation before computing the final indices using the ration between the observed and expected annual mean growth

# Features of vascular system of *Q. pyrenaica*

* Ring-porous tress are assumed to be particularly sensitive to climate warming, because they operate with narrow hydraulic safety margins (see Choat et al. )[@PerezdeLis2016]

## References

### 2016-12-07 man/notes_dendro.Rmd
update notes

@@
 
+* Vessel size chronologies were related to previous summer rainfall (from González-González et al. 2014)[@PerezdeLis2016]
+
+### Spring tree phenology [@PerezdeLis2016]
+
+| Phases                                 | DOY       | Months                | Sites <sup>1</sup> | Year <sup>1</sup> |
+|----------------------------------------|-----------|-----------------------|--------------------|-------------------|
+| Cambial resumption (cambial divisions) | 58 - 94   | late Feb - early Apr  | ***                | -                 |
+| Earlywood vessel (EV) enlargement      | 71 - 182  | mid March - early Jul | ***                | ***               |
+| EV maturation                          | 100 - 162 | early April - mid Jun | ***                | ***               |
+| Leaf expansion                         | 126 - 177 | early May - late June | ***                | ***               |
+
+* <sup>1</sup>: Differences between sites and years (2012, 2013)
+
+* Budburst dates were highly correlated to those of wood formation particularly to the onset of EV maturation
+* Full leaf expansion and the end of EV enlargement were simultaneous. EV formation and leaf unfoulding showed to be coupled
+
+* `$TODO$`: Interesting methodology of thermal drivers of phenology. To be used with phenological data of JMM
+
+#### Linkage between leaf phenology and earlywood formation [@PerezdeLis2016]
+* High correlations were found between budburst and stem maduration phases.
+It may be due to the fact that expanding leaves are major sources of auxin indole-3-acetic acid (IAA), a primary hormone promoting vascular differentiaion.
+* Leaf unfolding in ring-porous species occurs when first-formed vessels are already expanding, evidencing that auxin supply from growing leaves is not required for triggering cell division and vascular differentiation.
+Endogenous auxin levels in the dormant cambium, probably provided in winter by either dormant or swelling buds may induce xylogenesis due to higher cell sensitivity to auxin in ring porous species.
+* Ring-porous trees yearly restore their photosynthetic apparatus and hydraulic architecture using exclusively carbon reserves from the previous season.
+This is particularly true for *Q.
+pyrenaica* whose leaves were not fully expanded until the end of earlywood formation.
+
+#### Spring phenology
+* Phenological sensitivity to temperature is specific-dependent (see Vitasse et al. 2009 in [@PerezdeLis2016])
+
+* A delayed leaf appearance is assumed to be an adaptive response to avoid damage by late frost events (see Vitasse et al. 2014 in [@PerezdeLis2016]) as could be the case of Q. pyrenaica.
+
+#### Phenology-dependent hydraulic features of EV 
+* Climate signals in multi-decadal tree rings series indicate that vessel traits in ring-porous species are mostly influenced by climatic conditions in two physiological distinc periods: during the previous year and at the onset of cambial activity.
+`$TODO$`: Relation with lag effects of the drought
+
+* **Previous summer rainfall** has been stated as the most relevant climatic factor determining vessel size rather than temperature in winter and spring (see González-González et al.
+2014; 2015 in [@PerezdeLis2016]).
+Moisture availability was reported to be the most limiting factor driving radial growth in Iberian *Q.
+pyrenaica* populations (see Gea-Izquierdo and Cañellas 2014).
+:arrow_right: Vessel size could be more dependent on environmental signals operating at time windows that are different from those of vessel enlargement, such as precipitation during the previous growing season.
+* Formation of narrower vessels is considered an adaptive response of trees to water stress in ring-porous species.
+It could be somewhat triggered by declining carbon gain after severe droughts, due to reliance of ring-porous oaks on the reserves of the previous year for earlywood formation and canopy development
+
 ## References

### 2016-12-21 analysis/explore_tree.Rmd
add explore tree analysis

---
title: "First analysis"
author: "AJ Perez-Luque (@ajpelu)"
date: "2016 Nov"
output:  
  md_document:
    variant: markdown_github
---

```{r packages, warning=FALSE, message=FALSE}}
library("dplyr")
library("ggplot2")
```

# Read data 
```{r wd, echo=FALSE}
# machine <- 'ajpelu'
machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# tree data
tree <- read.csv(file=paste0(di, 'data_raw/tree_data.csv'), header=TRUE, sep=',')

# gps data 
elev <- read.csv(file=paste0(di, "/data/elev.csv"), header=TRUE, sep = ',')
```

# Prepare data
```{r}
qpyr <- tree %>% 
  inner_join(elev, by='name') %>% 
  filter(name != 'Fresno')

low_ca<- c('B01', 'B02', 'B03', 'B04', 'B05', 'B06', 'B07', 'B08', 'B09', 'B10',
         'B26', 'B27', 'B28', 'B29', 'B30')
low_sj <- c('A01', 'A02', 'A03', 'A04', 'A05', 'A06', 'A07', 'A08', 'A09', 'A10')
# high_sj<- c('A11', 'A12', 'A13', 'A14', 'A15', 'A16', 'A17', 'A18', 'A19', 'A20')
# high_ca <- c('B11', 'B12', 'B13', 'B14', 'B15', 'B16', 'B17', 'B18', 'B19', 'B20', 
#           'B21', 'B22', 'B23', 'B24', 'B25')

qpyr <- qpyr %>% 
  mutate(stand = ifelse(loc == 'SJ',
       ifelse(name %in% low_sj, 'low', 'high'), 
       ifelse(name %in% low_ca, 'low', 'high')))

qpyr %>% 
  ggplot(aes(y=height, x=perimeter)) + 
  geom_point(size=2) + geom_smooth(method = "lm", se=FALSE) +
  facet_grid(loc~stand) + theme_bw()

# TODO: Tape sensu Korner --> how many wood does tree invert into growht in height by basimetric area?? LINK to BAI (explore the same pattern with BAI) 
qpyr <- qpyr %>% 
  mutate(diam = 2*pi*(sqrt(perimeter/pi)),
         taper = diam / height)

qpyr %>% 
  ggplot(aes(y=taper, x=elev)) + 
  geom_point(size=2) + facet_wrap(~loc) + geom_smooth(method = "lm", se=FALSE)

+
  facet_grid(loc~stand) + theme_bw()

```

### 2016-12-21 man/notes_competition_index.Rmd
update competition index

---
title: "Competition notes"
author: "AJ Perez-Luque (@ajpelu)"
date: "2016 Dic"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: references.bib
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Competition index
He estado mirando los papers de [@SanchezSalguero2015;@Linares2009;@Linares2010] donde utilizan un *Competition Index (CI)* basado en distancia y dbh. Está bastante relacionado con growth. Otros papers analizan diferentes índices de competencia (ver por ejemplo [@Contreras2011;@Maleki2015]) y concluyen que los índices espacialmente explicitos tienen mejor correlación con crecimiento, aunque tienen la desventaja de que son mas tediosos.

## References

### 2017-02-20 analysis/analysis_competence.Rmd
add data reprojected

---
title: "Competence distance"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Feb"
output:  
  md_document:
    variant: markdown_github
---

```{r, warning=FALSE, message=FALSE}}
#library("tidyverse")
library("dplyr")
library("ggplot2")
```

# Read data 
```{r, echo=FALSE}
# machine <- 'ajpelu'
machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

compete <- read.csv(file=paste0(di, '/data/tree_compete.csv'), header=TRUE, sep=',')
```

```{r}
# prepare data

get.coords <- function(a, d, x0, y0) {
a <- ifelse(a <= 90, 90 - a, 450 - a)
data.frame(x = x0 + d * cos(a / 180 * pi),
y = y0+ d * sin(a / 180 * pi))
}

compete <- compete %>% 
  mutate(xcord = get.coords(a=rumbo_degree, 
                           d=distance_cm/100,
                           x0 = 0, y0 = 0)$x,
         ycord = get.coords(a=rumbo_degree, 
                           d=distance_cm/100,
                           x0 = 0, y0 = 0)$y)

mis <- get.coords(a=compete$rumbo_degree, 
                  d=compete$distance_cm, 
                  x0 = 0, y0 = 0)

```

```{r}
# see stackoverflow questions 6862742
gg_circle <- function(radius, xcenter, ycenter, color='black', fill=NA, ...){
  x <- xcenter + radius*cos(seq(0,pi, length.out = 100))
  ymax <- ycenter + radius*sin(seq(0, pi, length.out = 100))
  ymin <- ycenter + radius*sin(seq(0, -pi, length.out = 100))
  annotate("ribbon", x=x, ymin=ymin, ymax=ymax, color=color, fill=fill, ...)
}

compete %>%  
  ggplot(aes(x=xcord, y=ycord)) + 
  facet_wrap(~id_core) + 
  geom_point(aes(size=diam_mm), shape=21, fill='transparent') + 
  xlim(-11,11) + ylim(-11,11) + 
  geom_point(aes(x=0, y=0), color='green') +
  theme_bw() +
  gg_circle(radius = 10, xcenter = 0, ycenter = 0) +
  theme(panel.grid.major=element_blank(), 
        panel.grid.minor=element_blank())

```

# Extract data
```{r} 
ditemp <- paste('/Users/', machine, '/Documents/', sep='')
setwd(ditemp)

# read DEM 
demsn <- raster("mde_sn_area_influencia.asc")

setwd(di)

## Projection 
# Get projection
projection(demsn)

# Set projection
crs(demsn) <- "+proj=utm +zone=30 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"

# Read points data 
gps_point <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'dendro_garmin', verbose = FALSE)

## reproject 
gps_point_re <- spTransform(gps_point, crs(demsn))

gps_elev <- extract(demsn, gps_point_re, method='simple', sp=TRUE)

elev <- gps_elev@data %>% 
  dplyr::select(lat, lon, elev, name, date, elev_mde = mde_sn_area_influencia)

# Export data
write.csv(elev, file=paste(di, "/data/elev.csv", sep=""), row.names = FALSE)
```

### 2017-02-20 analysis/geo2.Rmd
add data reprojected

@@
 library("raster")
-library('dplyr')
-library('ggplot2')
+library("tidyverse")
 ```
@@
 
-# field woek points 
+# field wok points 
 fw <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'dendro_garmin', verbose = FALSE)
+
+mygpx <- paste0(di, 'data_raw/geoinfo/points_dendro_rep.GPX')
+my_points <- readOGR(mygpx, layer = 'waypoints', verbose = TRUE)
 
@@
              lat=coordinates(fw)[,'coords.x2'], 
-             group='Field Work 2016', radius=2, color = 'yellow', popup=fw$name)
+             group='Field Work 2016', radius=2, color = 'yellow', popup=fw$name) %>% 
+   addCircles(my_points, 
+             lng=coordinates(my_points)[,'coords.x1'], 
+             lat=coordinates(my_points)[,'coords.x2'], 
+             group='Field Work rep ', radius=2, color = 'purple', popup=my_points$name)
 
@@
              lat=coordinates(fw)[,'coords.x2'], 
-             group='Field Work 2016', radius=2, color = 'yellow', popup=fw$name)
+             group='Field Work 2016', radius=2, color = 'yellow', popup=fw$name) %>% 
+     addCircles(my_points, 
+             lng=coordinates(my_points)[,'coords.x1'], 
+             lat=coordinates(my_points)[,'coords.x2'], 
+             group='Field Work rep ', radius=2, color = 'purple', popup=my_points$name)
+
 ```

### 2017-03-01 analysis/analysis_competence.Rmd
updating

@@
 ```{r, warning=FALSE, message=FALSE}}
-#library("tidyverse")
-library("dplyr")
-library("ggplot2")
+library("tidyverse")
 ```
@@
 
+# density data 
 compete <- read.csv(file=paste0(di, '/data/tree_compete.csv'), header=TRUE, sep=',')
@@
 ```{r}
-# prepare data
+radius_plot <- 10 #
+sf_plot <- pi * radius_plot^2
 
-get.coords <- function(a, d, x0, y0) {
-a <- ifelse(a <= 90, 90 - a, 450 - a)
-data.frame(x = x0 + d * cos(a / 180 * pi),
-y = y0+ d * sin(a / 180 * pi))
+den_tree <- compete %>% 
+  group_by(id_core) %>% 
+  summarise(abun = n(), 
+            den = abun*10000 / sf_plot ) # stand density (tree / ha)
+
+diameter = perimeter/pi
+
+ba <- compete %>% 
+  mutate(dn = (diam_mm/1000) / pi, 
+         dn2 = dn * dn) %>% 
+  group_by(id_core) %>% 
+  summarise(
+    ba = (pi/4) * sum(dn2),
+    ba_mean = ba / n())
+
+get_coords_trees <- function(angle, distance, x0, y0) {
+  angle <- ifelse(angle <= 90, 90 - angle, 450 - angle)
+  data.frame(x = x0 + distance * cos(angle / 180 * pi),
+             y = y0+ distance * sin(angle / 180 * pi))
 }
@@
 compete <- compete %>% 
-  mutate(xcord = get.coords(a=rumbo_degree, 
-                           d=distance_cm/100,
+  mutate(xcord = get_coords_trees(angle= rumbo_degree, 
+                           distance = distance_cm/100,
                            x0 = 0, y0 = 0)$x,
-         ycord = get.coords(a=rumbo_degree, 
-                           d=distance_cm/100,
+
+         ycord = get_coords_trees(angle = rumbo_degree, 
+                           distance = distance_cm/100,
                            x0 = 0, y0 = 0)$y)
-
-mis <- get.coords(a=compete$rumbo_degree, 
-                  d=compete$distance_cm, 
-                  x0 = 0, y0 = 0)
 
@@
         panel.grid.minor=element_blank())
+
+radi <- 10 
+(pi*radi*radi)/10000
+
+```
+
+```{r}
+
+den <- compete %>% group_by(id_core) %>% 
+  summarise(ntree = n(),
+            den = ntree*10000 / (pi*100))
 

### 2017-03-01 analysis/get_elevation_points.Rmd
updating

@@
 library("dplyr")
+library("maptools")
 ```
@@
 ```{r wd, echo=FALSE}
-# machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 # read DEM 
-demsn <- raster("mde_sn_area_influencia.asc")
+# demsn <- raster("mde_sn_area_influencia.asc")
+# dme5t ign 
+dem1027 <- raster("1027.asc")
+dem1042 <- raster("1042.asc")
 
@@
 ## Projection 
-# Get projection
-projection(demsn)
-
-# Set projection
-crs(demsn) <- "+proj=utm +zone=30 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
+# Set projection(demsn)
+crs(demsn) <- "+init=epsg:23030"
 
 # Read points data 
-gps_point <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'dendro_garmin', verbose = FALSE)
+gps_point_ca <- readOGR(paste0(di, 'data_raw/geoinfo/dendro_ca.GPX'), layer='waypoints', verbose = FALSE) 
+gps_point_sj <- readOGR(paste0(di, 'data_raw/geoinfo/dendro_sj.GPX'), layer='waypoints', verbose = FALSE)
 
-## reproject 
-gps_point_re <- spTransform(gps_point, crs(demsn))
+# Join point 
+gps_point <- spRbind(gps_point_ca,gps_point_sj)
 
-gps_elev <- extract(demsn, gps_point_re, method='simple', sp=TRUE)
+# Get elev 
+gps_elev <- extract(demsn, gps_point, method='simple', sp=TRUE)
 
 elev <- gps_elev@data %>%
-  dplyr::select(lat, lon, elev, name, date, elev_mde = mde_sn_area_influencia)
+  dplyr::select(ele, name, elev_mde = mde_sn_area_influencia)
 

### 2017-03-01 analysis/spatial_focal_tree.Rmd
change point apparence

--- 2017-02-20 analysis/geo2.Rmd
@@
 
-```{r pacakges}
+```{r pacakges, warning=FALSE, message=FALSE}
 library('rgdal')
@@
 ```{r readData}
-mymachine <- '/Users/ajpeluLap/' 
-# mymachine <- '/Users/ajpelu/' 
+# mymachine <- '/Users/ajpeluLap/' 
+mymachine <- '/Users/ajpelu/' 
 di <- paste0(mymachine, 'Dropbox/phd/phd_repos/qpyr_dendro/')
+
+# Read spatial data 
+# Centroid of focal tree 
+## filename 
+mygpx_ca <- paste0(di, 'data_raw/geoinfo/dendro_ca.GPX')
+mygpx_sj <- paste0(di, 'data_raw/geoinfo/dendro_sj.GPX')
+
+## reading  
+field_work_ca <- readOGR(mygpx_ca, layer="waypoints", verbose = FALSE)
+field_work_sj <- readOGR(mygpx_sj, layer="waypoints", verbose = FALSE)
+```
+
+```{r prepareBaseMap}
+nombre_capas_base <- c("Ortophotography 2009", "Topographical")
+nombre_capas_overlay <- c("Field Work 2016")
+
+mymap <- leaflet() %>% 
+  # BaseMaps 
+  addWMSTiles('http://www.ideandalucia.es/wms/ortofoto2009?',
+              layers = 'oca10_2008-09',
+              options = WMSTileOptions(format = "image/png", transparent = FALSE),
+              attribution = '<a href="http://www.juntadeandalucia.es/institutodeestadisticaycartografia" target="_blank">Instituto de Estadística y Cartografía de Andalucía</a>',
+              group = 'Ortophotography 2009') %>%
+   addWMSTiles('http://www.ideandalucia.es/services/toporaster10/wms?',
+              layers = 'toporaster10',
+              options = WMSTileOptions(format = "image/png", transparent = FALSE),
+              attribution = '<a href="http://www.juntadeandalucia.es/institutodeestadisticaycartografia" target="_blank">Instituto de Estadística y Cartografía de Andalucía</a>',
+              group = 'Topographical') %>%
+
+  # Layers control
+  addLayersControl(position = 'bottomright',
+                   baseGroups = nombre_capas_base,
+                   overlayGroups = nombre_capas_overlay,
+                   options = layersControlOptions(collapsed = FALSE))
+```
+
+San Juan
+=======================================================================
+
+```{r}
+# Spatial object 
+fw <- field_work_sj
+
+# Set spatial extension 
+myext <- extent(fw)
+
+mymap %>% 
+  fitBounds(myext@xmin, myext@ymin, myext@xmax, myext@ymax) %>% 
+  addCircles(fw, 
+             lng=coordinates(fw)[,'coords.x1'], 
+             lat=coordinates(fw)[,'coords.x2'], 
+             group='Field Work 2016', radius=5, 
+             color = 'red', popup=fw$name, 
+             stroke = TRUE, fillOpacity = 0.5)
+
+```
+
+Caniar
+=======================================================================
+
+```{r}
+# Spatial object 
+fw <- field_work_ca
+
+# Set spatial extension 
+myext <- extent(fw)
+
+mymap %>% 
+  fitBounds(myext@xmin, myext@ymin, myext@xmax, myext@ymax) %>% 
+  addCircles(fw, 
+             lng=coordinates(fw)[,'coords.x1'], 
+             lat=coordinates(fw)[,'coords.x2'], 
+             group='Field Work 2016', radius=5, 
+             color = 'red', popup=fw$name, 
+             stroke = TRUE, fillOpacity = 0.5)
+
+```
+
+```{r eval=FALSE, echo=FALSE}
 
@@
 ots_gea <- ots[ots$notes == 'GeaIzquierdo2014',]
-
-```
-
-```{r}
-mymap <- leaflet() %>% 
-  # BaseMaps 
-  addWMSTiles('http://www.ideandalucia.es/wms/ortofoto2009?',
-              layers = 'oca10_2008-09',
-              options = WMSTileOptions(format = "image/png", transparent = FALSE),
-              attribution = '<a href="http://www.juntadeandalucia.es/institutodeestadisticaycartografia" target="_blank">Instituto de Estadística y Cartografía de Andalucía</a>',
-              group = 'Ortophotography 2009') %>%
-   addWMSTiles('http://www.ideandalucia.es/services/toporaster10/wms?',
-              layers = 'toporaster10',
-              options = WMSTileOptions(format = "image/png", transparent = FALSE),
-              attribution = '<a href="http://www.juntadeandalucia.es/institutodeestadisticaycartografia" target="_blank">Instituto de Estadística y Cartografía de Andalucía</a>',
-              group = 'Topographical') %>%
-
-  # Layers control
-  addLayersControl(position = 'bottomright',
-                   # baseGroups = c("Topographical","Ortophotography 2009","PNOA"),
-                   baseGroups = c("Ortophotography 2009", "Topographical"),
-                   overlayGroups = c("Field Work 2016", "Migrame", "OBSNEV Fen", "Gea Izquierdo 2014"),
-                   options = layersControlOptions(collapsed = FALSE))
-```
-
-San Juan
-=======================================================================
-
-```{r}
-myext <- extent(mig_forest_sj)
-
-mymap %>% 
-  fitBounds(myext@xmin, myext@ymin, myext@xmax, myext@ymax) %>% 
-  addPolylines(data=mig_forest_sj, color='blue', group='Migrame', popup=mig_forest_sj$nombre) %>% 
-  addCircles(ots_phe, 
-             lng=coordinates(ots_phe)[,'x'], 
-             lat=coordinates(ots_phe)[,'y'], 
-             group='OBSNEV Fen', radius=2, color = 'red', popup=ots_phe$name) %>% 
-  addCircles(ots_gea, 
-             lng=coordinates(ots_gea)[,'x'], 
-             lat=coordinates(ots_gea)[,'y'], 
-             group='Gea Izquierdo 2014', radius=2, color = 'green', popup=ots_gea$name) %>% 
-  addCircles(fw, 
-             lng=coordinates(fw)[,'coords.x1'], 
-             lat=coordinates(fw)[,'coords.x2'], 
-             group='Field Work 2016', radius=2, color = 'yellow', popup=fw$name) %>% 
-   addCircles(my_points, 
-             lng=coordinates(my_points)[,'coords.x1'], 
-             lat=coordinates(my_points)[,'coords.x2'], 
-             group='Field Work rep ', radius=2, color = 'purple', popup=my_points$name)
-
-```
-
-Caniar
-=======================================================================
-
-```{r}
 myext <- extent(mig_forest_ca)

### 2017-03-01 analysis/analysis_competence.Rmd
update computation of index

@@
 ```{r, echo=FALSE}
-# machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 # density data 
-compete <- read.csv(file=paste0(di, '/data/tree_compete.csv'), header=TRUE, sep=',')
+compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
 ```
@@
 ```{r}
-radius_plot <- 10 #
-sf_plot <- pi * radius_plot^2
+# Compute diameter normal (in meters)
+compete <- compete %>% 
+  mutate(dn = (perim_mm/1000) /pi) 
+```
+
+```{r}
+focal_trees <- unique(compete$id_core)
+
+output<- vector("list", length(focal_trees))
+
+for (i in 1:length(focal_trees)){
+
+# Create df   
+df <- compete %>% 
+  filter(id_core == focal_trees[[i]])
+
+# Fixed Variables 
+radius_plot <- 10 # radius plot (10 m)
+sf_plot <- pi * radius_plot^2 # surface plot (m2)
+dnft <- filter(df, sp== 'Focal')$dn # normal diameter focal tree    
+# ----------
+
+# Competence indices 
+## Number of competitors (r = 10)
+n_competitors <- filter(df, sp != 'Focal') %>% count()
+
+## Number of competitors with dn_j > dn_i (focal tree)
+n_competitors_hihger <- df %>%
+  filter(sp != 'Focal') %>%
+  filter(dn > dnft) %>% 
+  count()
+
+## Plot density (tree/plot)
+plot_density <- n_competitors / sf_plot
+
+## Stand density (tree / ha) 
+stand_density <- (n_competitors * 10000) / sf_plot
+
+## Sum sizes 
+sum_sizes <- df %>% 
+  filter(sp != 'Focal') %>%
+  summarise(sum_sizes=sum(dn))
+
+## Size ratio 
+size_ratio <- dnft / (dnft + sum_sizes$sum_sizes)
+
+## Distance nearest tree
+dist_nn <- df %>% 
+  filter(sp != 'Focal') %>% 
+  slice(which.min(distance_cm))
+
+dist_nn <- (dist_nn$distance_cm / 100)
+
+## Various 
+### Basal area
+### Size ratio proportional to distance 
+### Size difference proportional to distance 
+### Negative exponential size ratio
+### Negative exponential weighted size ratio
+### Lorimer 
+### Crowding 
+df_more_indices <- df %>%
+  filter(sp != 'Focal') %>%
+  mutate(
+    basal_area = ((pi/4) * dn * dn), 
+    size_ratio_distance = (dn / dnft) * (1 / ((distance_cm/100) + 1)),
+    size_diff_distance = (dn - dnft) / ((distance_cm/100) + 1),
+    neg_exp_size_ratio = (dn / dnft) * (1 / exp((distance_cm/100) + 1)),
+    neg_exp_w_size_ratio = (dn / dnft) * exp((-(distance_cm/100) + 1) / (dn + dnft)),
+    lorimer = (dn / dnft) / (sqrt((distance_cm/100) / radius_plot)),
+    crowding = dn / (distance_cm/100))
+
+df_indices <- df_more_indices %>% 
+  summarise(ba = sum(basal_area),
+            srd = sum(size_ratio_distance),
+            sdd = sum(size_diff_distance),
+            nesr = sum(neg_exp_size_ratio),
+            newsr = sum(neg_exp_w_size_ratio),
+            lorimer = sum(lorimer),
+            crowding = sum(crowding)) %>% 
+  mutate(nc = n_competitors,
+         nch = n_competitors_hihger,
+         pd = plot_density,
+         std = stand_density,
+         ss = sum_sizes,
+         sr = size_ratio,
+         dnn = dist_nn, 
+         id_plot = focal_trees[[i]])
+
+output[[i]] <- df_indices 
+}
+
+df_indices_plot <- bind_rows(output)
+```
+
+```{r}
+# Some constante variables 
+radius_plot <- 10 # radius plot (10 m)
+sf_plot <- pi * radius_plot^2 # surface plot (m2)
 
 den_tree <- compete %>% 
+  filter(sp != 'Focal') %>% 
   group_by(id_core) %>% 
-  summarise(abun = n(), 
-            den = abun*10000 / sf_plot ) # stand density (tree / ha)
+  summarise(ntree = n(), 
+            stand_density = ntree*10000 / sf_plot ) # stand density (tree / ha)
 

### 2017-03-03 analysis/analysis_competence.Rmd
update

@@
 ```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
+#machine <- 'ajpelu'
+machine <- 'ajpeluLap'
 di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 
+## Compute competence index
+
+`TODO` documentar
+
 ```{r}
-focal_trees <- unique(compete$id_core)
+focal_trees <- compete %>% filter(id_core != 'Fresno') 
+focal_trees <- unique(focal_trees$id_core)
 
@@
 df_indices_plot <- bind_rows(output)
+
+rm(n_competitors, n_competitors_hihger, plot_density, stand_density, sum_sizes, size_ratio, dist_nn, df, df_indices, output, df_more_indices, dnft)
 ```
 
+## Plot
+
 ```{r}
-# Some constante variables 
-radius_plot <- 10 # radius plot (10 m)
-sf_plot <- pi * radius_plot^2 # surface plot (m2)
-
-den_tree <- compete %>% 
-  filter(sp != 'Focal') %>% 
-  group_by(id_core) %>% 
-  summarise(ntree = n(), 
-            stand_density = ntree*10000 / sf_plot ) # stand density (tree / ha)
-
-diameter = perimeter/pi
-
-ba <- compete %>% 
-  mutate(dn = (diam_mm/1000) / pi, 
-         dn2 = dn * dn) %>% 
-  group_by(id_core) %>% 
-  summarise(
-    ba = (pi/4) * sum(dn2),
-    ba_mean = ba / n())
-
 get_coords_trees <- function(angle, distance, x0, y0) {
@@
                            x0 = 0, y0 = 0)$x,
-
          ycord = get_coords_trees(angle = rumbo_degree, 
@@
                            x0 = 0, y0 = 0)$y)
-
 ```
@@
 compete %>%  
+  filter(sp != 'Focal') %>%
   ggplot(aes(x=xcord, y=ycord)) + 
   facet_wrap(~id_core) + 
-  geom_point(aes(size=diam_mm), shape=21, fill='transparent') + 
+  geom_point(aes(size=perim_mm), shape=21, fill='transparent') + 
   xlim(-11,11) + ylim(-11,11) + 
@@
         panel.grid.minor=element_blank()) 
-
-radi <- 10 
-(pi*radi*radi)/10000
-
 ```
@@
 
-den <- compete %>% group_by(id_core) %>% 
-  summarise(ntree = n(),
-            den = ntree*10000 / (pi*100))
-
 ```
-
-# Extract data
-```{r} 
-ditemp <- paste('/Users/', machine, '/Documents/', sep='')
-setwd(ditemp)
-
-# read DEM 
-demsn <- raster("mde_sn_area_influencia.asc")
-
-setwd(di)
-
-## Projection 
-# Get projection
-projection(demsn)
-
-# Set projection
-crs(demsn) <- "+proj=utm +zone=30 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
-
-# Read points data 
-gps_point <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'dendro_garmin', verbose = FALSE)
-
-## reproject 
-gps_point_re <- spTransform(gps_point, crs(demsn))
-
-gps_elev <- extract(demsn, gps_point_re, method='simple', sp=TRUE)
-
-elev <- gps_elev@data %>% 
-  dplyr::select(lat, lon, elev, name, date, elev_mde = mde_sn_area_influencia)
-
-# Export data
-write.csv(elev, file=paste(di, "/data/elev.csv", sep=""), row.names = FALSE)
-```

### 2017-03-03 analysis/test-bai.Rmd
update

---
title: "dendro"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Feb"
output:  
  md_document:
    variant: markdown_github
---

```{r, warning=FALSE, message=FALSE}}
library("tidyverse")
library("dplR")
```

# Read data 
```{r, echo=FALSE}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# sj 
sj <- read.rwl(fname=paste0(di, '/data/SNA_sanjuan.rwl'), format="tucson")
```

```{r}
compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')

compete <- compete %>% 
  mutate(dn = (perim_mm/1000) /pi)

dd <- compete %>% 
  filter(sp=='Focal') %>% 
  filter(loc=='SJ') %>% 
  filter(id_core!='Fresno')
```

```{r}

# Function to compute BAI

# rwl_file
myrwl <- sj

# Extract rwl serie 
myrwl_i <- myrwl[[1]]

# Remove NA 
myrwl_io <- na.omit(myrwl_i)

bai <- (pi/4)*(dbh^2 - ((d - 2*w)^2))

sjD <- sj[["SNA0101"]]
dbh <- 1170 / pi

baiD <- bai.out(sj[["SNA0101"]], dbh)

```

### 2017-03-30 man/field_data.Rmd
update field data

---
title: "Field data work"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: references.bib
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Muestreo
## Spatial Coverage

Dos localidades de muestreo en Sierra Nevada: Robledal de Cañar y Robledal de San Juan. En cada una de las zonas dos zonas (high and low elevation). $TODO$ :red_circle: `INCLUIR LINK A MAPA CON ZONZAS Y TABLA CON NOMBRES DE MUESTREO`

## Metodologia muestreo

* Número de árboles: En Cáñar muestreamos 30 árboles (15 y 15) mientras que en San Juan 20 (10 y 10) árboles
* De cada árbol se midió la altura y el perímetro a la altura del pecho (para sacar dbh)
* Los árboles fueron seleccionados de forma aleatoria
* Tomamos al menos 2 cores perpendiculares por árbol a una altura de 1.3 m. Para ello utilizamos una barrena de Pressler
* Los cores se etiquetaron y se guardaron en pajitas hasta su manipulacion en la laboratorio

## Temporal coverage 
* 13-14 Noviembre 2016

## Collectors 
* Guillermo Gea-Izquierdo 
* Regino Zamora-Rodríguez
* Antonio J. Pérez-Luque
* Francisco J. Bonet-García (13 Nov)

# Laboratorio:

* Cada core se montó sobre soportes de madera convenientemente etiquetados
* Cada core se pegó utilizando cola soluble al agua y fixo (scott) 
* Ojo con montar bien las muestras (plano transversal, ver @Fritts1976)
* Una vez montadas, son lijadas (empezando por lija gruesa hasta lija menos gruesa)

## Medición: 
* Primero datación visual cruzada (sincronización // cross-dating)
* La medición de anillos se realizó utilizando un dispositivo LinTab 6 (RINNTECH) con una precision de 0.01 mm

Se utilizó un LINTAB conectado a TSAP ??  
LinTab Measuring System

## References

### 2017-03-30 man/field_data.Rmd
Update field work

@@
 ## Medición:
+
 * Primero datación visual cruzada (sincronización // cross-dating)
-* La medición de anillos se realizó utilizando un dispositivo LinTab 6 (RINNTECH) con una precision de 0.01 mm
-
-Se utilizó un LINTAB conectado a TSAP ??  
-LinTab Measuring System
+* La medición de anillos se realizó utilizando un estereomicroscópio acoplado a un dispositivo LINTAB (Rinntech, Heidelberg, Germany) con una precision de 0.01 mm; y utilizando el software [TSAP-Win](http://www.rinntech.de/content/view/17/48/lang,english/index.html)
+* La sincronización se realizó primero de forma visual y luego se verificó utilizando los valores de los estadísticos: 
+    * Gleichläufigkeit (Glk)
+    * t-value 
+    * Cross-datin index (CDI)
+* Posteriormente la sincronización (cross-dating) fue evaluada utilizando COFECHA [@Holmes1983]
 

### 2017-03-31 man/field_data.Rmd
update field work and add data

--- 2017-03-30 man/field_data.Rmd
@@
 ## Medición:
+
 * Primero datación visual cruzada (sincronización // cross-dating)
-* La medición de anillos se realizó utilizando un dispositivo LinTab 6 (RINNTECH) con una precision de 0.01 mm
+* La medición de anillos se realizó utilizando un estereomicroscópio acoplado a un dispositivo LINTAB (Rinntech, Heidelberg, Germany) con una precision de 0.01 mm (1/100 mm); y utilizando el software [TSAP-Win](http://www.rinntech.de/content/view/17/48/lang,english/index.html)
+* La sincronización se realizó primero de forma visual y luego se verificó utilizando los valores de los estadísticos: 
+    * Gleichläufigkeit (Glk)
+    * t-value 
+    * Cross-datin index (CDI)
+* Posteriormente la sincronización (cross-dating) fue evaluada utilizando COFECHA [@Holmes1983]
 
-Se utilizó un LINTAB conectado a TSAP ??  
-LinTab Measuring System
+## Output
+De la medición y posterior sincronización se han generado varios archivos, en concreto de tres tipos para cada localidad:
+
+* `sn_qpyr_LOC2016.fh`. Archivo en formato Heidelberg, generado por el software TSAP-Win con los datos de las series de cada core, de cada árbol, y las cronos medias. 
+* `sn_LOC.rwl`. Archivo en formato Tucson con las mediciones de cada core 
+* `sn_LOC_cofecha.OUT`. Archivo de salida de la comprobación de la sincronización realizada con COFECHA [@Holmes; @GrissinoMayer2001]
+
+donde `LOC` puede ser: `sanjuan` o `canar`. Se ha generado dos carpetas, una para cada localidad:
+
+* `./data_raw/dendro_ring/sn_canar/` 
+* `./data_raw/dendro_ring/sn_sanjuan/`
+
+En el caso de Cáñar, tuvimos algunos desajustes en la sincronización, por que también existe otro archivo `.fh` llamado `SN_QPYR_Canar_2016_antonio_viejas.fh` que contiene los datos antes de las correcciones realizadas por Guillermo.
+
+Finalmente, existe una copia de seguridad de todos los archivos
+
+Para mas información sobre formatos de datos en dendrocronología ver [material suplementario](http://www.treeringsociety.org/resources/SOM/Brewer_Murphy_SupplementaryMaterial.pdf) de [@Brewer2011].
 

### 2017-04-07 analysis/test-bai.Rmd
update

@@
 
+# Compute diameter (mm)
 compete <- compete %>% 
-  mutate(dn = (perim_mm/1000) /pi)
+  mutate(dn_mm = (perim_mm / pi))
 
-dd <- compete %>% 
+# Get only focal trees, and only selected variables 
+ft <- compete %>% 
   filter(sp=='Focal') %>% 
-  filter(loc=='SJ') %>% 
-  filter(id_core!='Fresno')
+  filter(id_focal!='Fresno') %>% 
+  select(id_focal, loc, dn_mm, height_cm)
+
+# Select only SJ trees and 
+ft_sj <- ft %>% filter(loc=='SJ')
+
+# get Create a dataframe with core ID
+cores <- data.frame(id_cores=colnames(sj))
+
+# Extract replicate and Tree ID from core ID 
+cores <- cores %>% 
+  mutate(id_focal = as.factor(stringr::str_sub(id_cores, 3,5)),
+         id_replica = stringr::str_sub(id_cores, 6,8))
+
+# Create df with diameter and height for each core ID 
+diam_cores <- cores %>% inner_join(ft_sj, by='id_focal') %>% 
+  select(id_cores, dn_mm)
+
+```
+
+```{r}
+# BAI dplR
+baisj <- bai.out(rwl = sj, diam = diam_cores)
+
 ```
@@
 
+# Year
+n.vec <- seq_len(nrow(myrwl))
+
+# Series
+seq_len(ncol(myrwl))
+
+# Select a ring-width series
+dat <- myrwl[[1]] 
+# Remove NA
+dat2 <- na.omit(dat)
+
+    for (i in seq_len(ncol(rwl))) {
+        dat <- rwl[[i]]
+        dat2 <- na.omit(dat)
+        if (is.null(diam)) 
+            d <- sum(dat2) * 2
+        else d <- diam.vec[i]
+        r0 <- d/2 - c(0, cumsum(rev(dat2)))
+        bai <- -pi * rev(diff(r0 * r0))
+        na <- attributes(dat2)$na.action
+        no.na <- n.vec[!n.vec %in% na]
+        out[no.na, i] <- bai
+    }
+    out
+
 bai <- (pi/4)*(dbh^2 - ((d - 2*w)^2))
@@
 
+rwl
+
+# Test bai.out 
+s1 <- sj[1]
+b1 <- bai.out(s1)/100
+
+ggplot(s1, aes(y=SNA0101, x=row.names(s1))) + 
+  geom_point() + 
+  geom_line(b1, aes(y=SNA0101, x=row.names(b1)))
+
+ggplot(s1, aes(y=SNA0101, x=row.names(b1), group=1)) + geom_line()
+
 ```

### 2017-04-10 analysis/test-bai.Rmd
split sj datasets

@@
 sj <- read.rwl(fname=paste0(di, '/data/SNA_sanjuan.rwl'), format="tucson")
+```
+
+Algunos árboles tienen 3 cores. Sin embargo, no los tres cores llegan hasta la corteza, por lo tanto, no podemos utilizar el diametro para estimar bai. Vamos a realizar lo siguiente:
+
+* Crear dos datasets: 
+    * Dataset con series de datos que llegan hasta corteza (`sj_cor`)
+    * Dataset con series de datos que no llegan hasta corteza (`sj_sin`): se trata de cores que no llegan hasta la corteza.
+
+* Para el cálculo del BAI, en el dataset `sj` utilizamos el diámetro medido en campo. En el caso del dataset `sj_sin` utilizamos la suma de todos los diámetros (:red_circle: `$TODO$ ASK to Guillermo`)
+
+```{r}
+# Get summary ring-width series
+sj_summ <- summary(sj)
+
+# Get names cores with last year different to 2016 
+id_cores_no_bark <- sj_summ %>% 
+  filter(last != 2016) %>% select(series) %>% mutate(series = factor(series))
+
+id_cores_no_bark <- unname(unlist(id_cores_no_bark))
+
+# Subbet datasets 
+sj_cor <- sj[which(! colnames(sj) %in% (id_cores_no_bark))]
+sj_sin <- sj[which(colnames(sj) %in% (id_cores_no_bark))]
 ```

### 2017-04-10 analysis/test-bai.Rmd
update piovesan test bai

@@
     variant: markdown_github
+bibliography: references.bib 
+csl: ecology.csl
 ---
 
-```{r, warning=FALSE, message=FALSE}}
+```{r, warning=FALSE, message=FALSE}
 library("tidyverse")
@@
 sj <- read.rwl(fname=paste0(di, '/data/SNA_sanjuan.rwl'), format="tucson")
+sjt <- read.rwl(fname=paste0(di, '/data/sj_test.rwl'), format="tucson")
 ```
@@
 ```
+
+Lectura y preparación de datos de diámetro
 
@@
 
+# remove diamm of cores without bark 
+diam_cores_sj <- diam_cores %>% filter(! id_cores %in% id_cores_no_bark)
 ```
 
+## Cómputo del BAI
+
+He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]
+
 ```{r} 
-# BAI dplR
-baisj <- bai.out(rwl = sj, diam = diam_cores)
+## BAI Function Piovesan
+
+i <- "SNA0101"
+
+bai_piovesan <- function(rwdf, diam_df){
+  # rwdf: a ringwidth dataframe. row: years; col: id_cores
+  # diam_df: a dataframe with diameter and id_core (two columns)
+
+  # Output 
+  bai <- rwdf
+
+  # length years in Ring-width Data frame 
+  vec_ly <- seq_len(nrow(rwdf))
+
+  # ID cores 
+  cores <- colnames(rwdf)
+
+  for (i in cores){ 
+    # 1 # Select core 
+    id_core <- i
+
+    # 2 # Select serie 
+    rs <- rwdf[[id_core]]
+
+    # 3 # Select diameter 
+    dbh <- subset(diam_df, id_cores == id_core)
+    dbh <- dbh[,2]
+
+    # 4 # Prepare data to compute bai 
+    #### Omit NA 
+    rsomit <- na.omit(rs)
+
+    #### Revert w 
+    rs_rev <- rev(rsomit)
+
+    # Cumsum rever w
+    # cum_rs_rev <- cumsum(rs_rev)
+    cum_rs_rev <- c(0, cumsum(rs_rev))
+
+    # Compute d_{t} 
+    d <- dbh - 2*cum_rs_rev
+
+    # Remove last element 
+    d <- d[-length(d)]
+
+    # Bai
+    b <- rev((pi/4) * ((d)^2 - (d - 2*rs_rev)^2))
+
+    # Store BAI and NA 
+    ### Which elements are NA? 
+    na <- attributes(rsomit)$na.action
+    ### Which ones no? 
+    no_na <- vec_ly[!vec_ly %in% na]
+    ### Put bai in those element with no NA
+    bai[no_na, i] <- b 
+  }
+  bai 
+}
+
+# Test 
+b_test_piovesan <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
+b_test_dplR <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
 
@@
 
-```{r}
+```{r, eval=FALSE, echo=FALSE}
+# BAI dplR
+baisj <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
 
-# Function to compute BAI
+baisjt <- bai.out(rwl = sjt)
+
+# Function to compute BAI (Piovesan et al)
 
 # rwl_file
-myrwl <- sj
+myrwl <- sj_cor
+# Diameter df  
+diam_df <- diam_cores_sj
+
+# Core
+id_core <- 'SNA0101'
+
+# Ring-width serie
+myrwl_serie <- myrwl[[id_core]]
+
+# Diameter for core id_core 
+dbh <- subset(diam_df, id_cores == id_core)
+dbh <- dbh[,2]
+
+myrwl_serie2 <- na.omit(myrwl_serie)
+cum
+
+r0 <- dbh/2 - c(0,cumsum(rev(myrwl_serie2)))
+
+bai <- (pi/4)*(dbh^2 - ((dbh - 2*myrwl_serie)^2))
+
+s <- myrwl_serie[94:96]
+
+d <- sum(s)*2
+
+rev(s)
 
@@
 
+diam_df_i <- diam_cores_sj[1]
+
 # Remove NA 
 myrwl_io <- na.omit(myrwl_i)
+
+bai <- 
+diam.vec <- diam[, 2]
+
+########## test Piovesan
+# rwl_file
+myrwl <- baisj
+
+#id_core
+
+# Extract rwl serie 
+myrwl_i <- myrwl[[1]]
+
+dbh <- 372.4226
 
@@
 
-    for (i in seq_len(ncol(rwl))) {
-        dat <- rwl[[i]]
-        dat2 <- na.omit(dat)
-        if (is.null(diam)) 
-            d <- sum(dat2) * 2
-        else d <- diam.vec[i]
-        r0 <- d/2 - c(0, cumsum(rev(dat2)))
-        bai <- -pi * rev(diff(r0 * r0))
-        na <- attributes(dat2)$na.action
-        no.na <- n.vec[!n.vec %in% na]
-        out[no.na, i] <- bai
-    }
-    out
-
-bai <- (pi/4)*(dbh^2 - ((d - 2*w)^2))
-
-sjD <- sj[["SNA0101"]]
-dbh <- 1170 / pi
-
-baiD <- bai.out(sj[["SNA0101"]], dbh)
-
-rwl
-
-# Test bai.out 
-s1 <- sj[1]
-b1 <- bai.out(s1)/100
-
 ggplot(s1, aes(y=SNA0101, x=row.names(s1))) + 

### 2017-04-10 analysis/test-bai.Rmd
update bai

@@
 
+```{r}
+set.seed(1234)
+serie <- sample(colnames(b_test_dplR),1)
+
+require(reshape2)
+
+bp <- melt(t(b_test_piovesan))
+names(bp) <- c('cores', 'years', 'bai_p')
+
+bd <- melt(t(b_test_dplR))
+names(bd) <- c('cores', 'years', 'bai_d')
+
+diff <- bp %>% 
+  inner_join(bd, by=c('cores','years')) %>% 
+  mutate(diff = bai_p - bai_d)
+
+ggplot(x, aes(y=bai, x=years, colour=test)) + 
+  geom_point() + 
+  facet_wrap(~cores)
+
 ```{r, eval=FALSE, echo=FALSE}

### 2017-04-11 analysis/test-bai.Rmd
update script

@@
 library("dplR")
+source(paste0(getwd(),"/script/R/bai_piovesan.R"))
 ```
@@
 # sj 
-sj <- read.rwl(fname=paste0(di, '/data/SNA_sanjuan.rwl'), format="tucson")
-sjt <- read.rwl(fname=paste0(di, '/data/sj_test.rwl'), format="tucson")
+sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
+
+# canar 
+ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
 ```
+
+## Canar Data
 
@@
 ```{r} 
-## BAI Function Piovesan
+# Compute BAI for SJ locality 
+baisj <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
+```
 
-i <- "SNA0101"
+```{r}
+rwl.report(sj_cor)
 
-bai_piovesan <- function(rwdf, diam_df){
-  # rwdf: a ringwidth dataframe. row: years; col: id_cores
-  # diam_df: a dataframe with diameter and id_core (two columns)
+bai_p <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
+bai_d <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
 
-  # Output 
-  bai <- rwdf
+diam_cores_sjsin <- data.frame(id_cores = colnames(sj_sin),
+                               dn_mm = colSums(sj_sin, na.rm = TRUE))
+rownames(diam_cores_sjsin) <- NULL
 
-  # length years in Ring-width Data frame 
-  vec_ly <- seq_len(nrow(rwdf))
-
-  # ID cores 
-  cores <- colnames(rwdf)
-
-  for (i in cores){ 
-    # 1 # Select core 
-    id_core <- i
-
-    # 2 # Select serie 
-    rs <- rwdf[[id_core]]
-
-    # 3 # Select diameter 
-    dbh <- subset(diam_df, id_cores == id_core)
-    dbh <- dbh[,2]
-
-    # 4 # Prepare data to compute bai 
-    #### Omit NA 
-    rsomit <- na.omit(rs)
-
-    #### Revert w 
-    rs_rev <- rev(rsomit)
-
-    # Cumsum rever w
-    # cum_rs_rev <- cumsum(rs_rev)
-    cum_rs_rev <- c(0, cumsum(rs_rev))
-
-    # Compute d_{t} 
-    d <- dbh - 2*cum_rs_rev
-
-    # Remove last element 
-    d <- d[-length(d)]
-
-    # Bai
-    b <- rev((pi/4) * ((d)^2 - (d - 2*rs_rev)^2))
-
-    # Store BAI and NA 
-    ### Which elements are NA? 
-    na <- attributes(rsomit)$na.action
-    ### Which ones no? 
-    no_na <- vec_ly[!vec_ly %in% na]
-    ### Put bai in those element with no NA
-    bai[no_na, i] <- b 
-  }
-  bai 
-}
-
-# Test 
-b_test_piovesan <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
-b_test_dplR <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
+bai_psin <- bai_piovesan(rwdf = sj_sin, diam_df = diam_cores_sjsin)
+bai_dsinsin <- bai.out(rwl = sj_sin)
+bai_dsin <- bai.out(rwl = sj_sin, diam = diam_cores_sjsin)
 
@@
 ```
+
+```{r}
+
+x <- c(1,2,2,3,1,3,4,5)
+
+xx <- rollmean(x=x, k=3, align = 'center', na.pad = T)
+
+```

### 2017-04-11 script/R/bai_piovesan.R

@@ -1,0 +1,60 @@
+## BAI Function Piovesan 
+
+## See Piovesan et al. 2008 http://dx.doi.org/10.1111/j.1365-2486.2008.01570.x 
+## BAI = (pi/4)*[ (dbh_{t})^2 - [dbh_{t} - 2*w_{t}]^2 ] 
+## r_{t} = r_{t-1} + w_{t} 
+
+bai_piovesan <- function(rwdf, diam_df){
+  # rwdf: a ringwidth dataframe. row: years; col: id_cores
+  # diam_df: a dataframe with diameter and id_core (two columns) 
+  
+  # Output 
+  bai <- rwdf 
+  
+  # length years in Ring-width Data frame 
+  vec_ly <- seq_len(nrow(rwdf))
+  
+  # ID cores 
+  cores <- colnames(rwdf)
+  
+  for (i in cores){ 
+    # 1 # Select core 
+    id_core <- i 
+    
+    # 2 # Select ring width serie 
+    rws <- rwdf[[id_core]]
+    
+    # 3 # Select diameter 
+    dbh <- subset(diam_df, id_cores == id_core)
+    dbh <- dbh[,2]
+    
+    # 4 # Prepare data to compute bai 
+    #### Omit NA 
+    rwsomit <- na.omit(rws)
+    
+    #### Revert w 
+    rws_rev <- rev(rwsomit)
+    
+    # Cumsum rever w
+    # cum_rs_rev <- cumsum(rs_rev)
+    cum_rws_rev <- c(0, cumsum(rws_rev))
+    
+    # Compute d_{t} 
+    d <- dbh - 2*cum_rws_rev
+    
+    # Remove last element 
+    d <- d[-length(d)]
+    
+    # Bai
+    b <- rev((pi/4) * ((d)^2 - (d - 2*rws_rev)^2))
+    
+    # Store BAI and NA 
+    ### Which elements are NA? 
+    na <- attributes(rwsomit)$na.action
+    ### Which ones no? 
+    no_na <- vec_ly[!vec_ly %in% na]
+    ### Put bai in those element with no NA
+    bai[no_na, i] <- b 
+  }
+  bai 
+}

### 2017-04-11 man/field_data.Rmd
update biblio
update
update notas datos de campo

@@
 
-# Muestreo
-## Spatial Coverage
+# Dendrocronología
+
+## Muestreo
+
+### Spatial Coverage
 
@@
 
-## Metodologia muestreo
+### Metodología
 
@@
 
-## Temporal coverage 
+### Temporal coverage 
 * 13-14 Noviembre 2016
 
-## Collectors 
+### Collectors 
 * Guillermo Gea-Izquierdo 
@@
 
-# Laboratorio:
+## Laboratorio:
 
@@
 
-## Medición:
+### Medición:
 
@@
 
-## Output
+### Output
 De la medición y posterior sincronización se han generado varios archivos, en concreto de tres tipos para cada localidad:
@@
 * `sn_LOC.rwl`. Archivo en formato Tucson con las mediciones de cada core 
-* `sn_LOC_cofecha.OUT`. Archivo de salida de la comprobación de la sincronización realizada con COFECHA [@Holmes; @GrissinoMayer2001]
+* `sn_LOC_cofecha.OUT`. Archivo de salida de la comprobación de la sincronización realizada con COFECHA [@Holmes1983 @GrissinoMayer2001]
 
@@
 
-Finalmente, existe una copia de seguridad de todos los archivos
+Finalmente, existe una copia de seguridad de todos los archivos en el archivo [`./data_raw/dendro_ring/copia_seg/dendro.zip`] y en mi carpeta drive backups
 
@@
 
-## References
+# References

### 2017-04-11 analysis/test-bai.Rmd
update source

@@
 library("dplR")
-source(paste0(getwd(),"/script/R/bai_piovesan.R"))
+source('./script/R/bai_piovesan.R')
 ```
@@
 # Read data
+
 ```{r, echo=FALSE}
@@
 
-## Canar Data
+## San Juan Data
 

### 2017-04-12 analysis/test_gc.R
code snippet test GC disturbance

@@ -1,0 +1,21 @@
+# vector 
+install.packages('rowr')
+
+x <- 1:10
+
+y <- c(1,2,1,3,1,2,4,3,2,1)
+
+library(zoo)
+
+rollapply(y, 2, mean)
+
+y <- c(1,2,3,2,2)
+
+
+rollApply(y, sum, window = 2, align = 'left')
+rollApply(y, sum, window = 2, align='right')
+
+
+
+
+

### 2017-04-21 man/notes_dendro_analysis.Rmd
update dendro analysis + bib

--- 2017-04-11 man/field_data.Rmd
@@
 ---
-title: "Field data work"
-author: "AJ Perez-Luque (@ajpelu)"
+title: "Notes Dendro Analysis"
+author: "AJ Perez-Luque (@ajpelu); G. Gea-Izquierdo"
 date: "2017"
@@
 
-# Dendrocronología
+# Notas análisis dendro
+
+## Sincronización 
+Sincronizamos visualmene cada una de las series de radios (rw) y posteriormente verificamos la sincronización con  COFECHA [@Holmes1983]
+
+* [] :red_circle: Link al ouptut y análisis de los datos de COFECHA
+* [] :red_circle: Tabla 1 con estadísticos (ver otros ejemplos)
+
+## Transformación RW a BAI
+* Vamos a transformar los datos de RW a BAI. 
+* Existe una tendencia de decrecimiento en las medidas de RW con el incremento del tamaño del árbol o con el incremento de la edad (ver [@Biondi2008, @Dorado2017])
+
+> ...
+one of the main elements of dendrochronological standardization is removing the biological trend, i.e.
+the progressive decline of ring width along a cross-sectional radius that is caused by the corresponding increase in stem size and tree age over time...
+[@Biondi2008]
+
+> ... Individual tree-ring series were coverted to Basal Area Increment (BAI) to avoid age-related trends in non-juvenile ring-width measurement (see [@Dorado2017])
+
+* El objetivo de esta transformación es remover las tendencias relacionadas con la edad. Podría tratarse de una forma de estandarización. 
+* Teóricamente y bajo unas condiciones climáticas medias y en ausencia de perturbaciones, BAI tiene una tendencia creciente en edades tempranas (hasta 40-50 años) y posteriormente se estabiliza por un largo periodod hasta que al final decrece antes de morir (ver @GeaIzquierdo2014 p.229).
+Por tanto, una vez transformado a BAI, las tendencias observadas estarían relacionadas con cambios ambientales a largo plazo.
+
+Por tanto, tendencias en BAI (ver @GeaIzquierdo2014): 
+* Tendencias negativas: expresan un declive en la productividad 
+* Tendencias positivas: efecto positivo del clima o de la fertilización atmosférica
 

### 2017-04-21 man/notes_dendro_analysis.Rmd
update analysis dendro

@@
 
-* [] :red_circle: Link al ouptut y análisis de los datos de COFECHA
-* [] :red_circle: Tabla 1 con estadísticos (ver otros ejemplos)
+* [ ] :red_circle: Link al ouptut y análisis de los datos de COFECHA
+* [ ] :red_circle: Tabla 1 con estadísticos (ver otros ejemplos)
 
@@
 * Vamos a transformar los datos de RW a BAI. 
-* Existe una tendencia de decrecimiento en las medidas de RW con el incremento del tamaño del árbol o con el incremento de la edad (ver [@Biondi2008, @Dorado2017])
+* Existe una tendencia de decrecimiento en las medidas de RW con el incremento del tamaño del árbol o con el incremento de la edad (ver @Biondi2008, @Dorado2017)
 
@@
 
-> ... Individual tree-ring series were coverted to Basal Area Increment (BAI) to avoid age-related trends in non-juvenile ring-width measurement (see [@Dorado2017])
+> ... Individual tree-ring series were coverted to Basal Area Increment (BAI) to avoid age-related trends in non-juvenile ring-width measurement [@Dorado2017]
 
 * El objetivo de esta transformación es remover las tendencias relacionadas con la edad. Podría tratarse de una forma de estandarización. 
-* Teóricamente y bajo unas condiciones climáticas medias y en ausencia de perturbaciones, BAI tiene una tendencia creciente en edades tempranas (hasta 40-50 años) y posteriormente se estabiliza por un largo periodod hasta que al final decrece antes de morir (ver @GeaIzquierdo2014 p.229).
+* Teóricamente y bajo unas condiciones climáticas medias y en ausencia de perturbaciones, BAI tiene una tendencia creciente en edades tempranas (hasta 40-50 años) y posteriormente se estabiliza por un largo periodod hasta que al final decrece antes de morir (ver p.
+229 en @GeaIzquierdo2014).
 Por tanto, una vez transformado a BAI, las tendencias observadas estarían relacionadas con cambios ambientales a largo plazo.
 
-Por tanto, tendencias en BAI (ver @GeaIzquierdo2014): 
+Por tanto, tendencias en BAI (ver en @GeaIzquierdo2014):
+
 * Tendencias negativas: expresan un declive en la productividad 
@@
 
-## Muestreo
-
-### Spatial Coverage
-
-Dos localidades de muestreo en Sierra Nevada: Robledal de Cañar y Robledal de San Juan.
-En cada una de las zonas dos zonas (high and low elevation).
-$TODO$ :red_circle: `INCLUIR LINK A MAPA CON ZONZAS Y TABLA CON NOMBRES DE MUESTREO`
-
-### Metodología
-
-* Número de árboles: En Cáñar muestreamos 30 árboles (15 y 15) mientras que en San Juan 20 (10 y 10) árboles
-* De cada árbol se midió la altura y el perímetro a la altura del pecho (para sacar dbh)
-* Los árboles fueron seleccionados de forma aleatoria
-* Tomamos al menos 2 cores perpendiculares por árbol a una altura de 1.3 m. Para ello utilizamos una barrena de Pressler
-* Los cores se etiquetaron y se guardaron en pajitas hasta su manipulacion en la laboratorio
-
-### Temporal coverage 
-* 13-14 Noviembre 2016
-
-### Collectors 
-* Guillermo Gea-Izquierdo 
-* Regino Zamora-Rodríguez
-* Antonio J. Pérez-Luque
-* Francisco J. Bonet-García (13 Nov)
-
-## Laboratorio:
-
-* Cada core se montó sobre soportes de madera convenientemente etiquetados
-* Cada core se pegó utilizando cola soluble al agua y fixo (scott) 
-* Ojo con montar bien las muestras (plano transversal, ver @Fritts1976)
-* Una vez montadas, son lijadas (empezando por lija gruesa hasta lija menos gruesa)
-
-### Medición:
-
-* Primero datación visual cruzada (sincronización // cross-dating)
-* La medición de anillos se realizó utilizando un estereomicroscópio acoplado a un dispositivo LINTAB (Rinntech, Heidelberg, Germany) con una precision de 0.01 mm (1/100 mm); y utilizando el software [TSAP-Win](http://www.rinntech.de/content/view/17/48/lang,english/index.html)
-* La sincronización se realizó primero de forma visual y luego se verificó utilizando los valores de los estadísticos: 
-    * Gleichläufigkeit (Glk)
-    * t-value 
-    * Cross-datin index (CDI)
-* Posteriormente la sincronización (cross-dating) fue evaluada utilizando COFECHA [@Holmes1983]
-
-### Output
-De la medición y posterior sincronización se han generado varios archivos, en concreto de tres tipos para cada localidad:
-
-* `sn_qpyr_LOC2016.fh`. Archivo en formato Heidelberg, generado por el software TSAP-Win con los datos de las series de cada core, de cada árbol, y las cronos medias. 
-* `sn_LOC.rwl`. Archivo en formato Tucson con las mediciones de cada core 
-* `sn_LOC_cofecha.OUT`. Archivo de salida de la comprobación de la sincronización realizada con COFECHA [@Holmes1983 @GrissinoMayer2001]
-
-donde `LOC` puede ser: `sanjuan` o `canar`. Se ha generado dos carpetas, una para cada localidad:
-
-* `./data_raw/dendro_ring/sn_canar/` 
-* `./data_raw/dendro_ring/sn_sanjuan/`
-
-En el caso de Cáñar, tuvimos algunos desajustes en la sincronización, por que también existe otro archivo `.fh` llamado `SN_QPYR_Canar_2016_antonio_viejas.fh` que contiene los datos antes de las correcciones realizadas por Guillermo.
-
-Finalmente, existe una copia de seguridad de todos los archivos en el archivo [`./data_raw/dendro_ring/copia_seg/dendro.zip`] y en mi carpeta drive backups
-
-Para mas información sobre formatos de datos en dendrocronología ver [material suplementario](http://www.treeringsociety.org/resources/SOM/Brewer_Murphy_SupplementaryMaterial.pdf) de [@Brewer2011].
-
 # References

### 2017-04-21 man/notes_dendro_analysis.Rmd
* add BAItree function * update notes analysis dendro

@@
 ## Transformación RW a BAI
+### Justificación 
 * Vamos a transformar los datos de RW a BAI. 
@@
 
+### Procedimiento
+Existen varias formas de hacerlos:
+
+* Usando la función `dplR::bai.out()`
+* Aproximación la de @Piovesan2008:
+
+    * Función `BAItree()` escrita por G. Gea. Enlace aquí [./script/R/gea/BAItree.R](/script/R/gea/BAItree.R)
+    * Función `bai_piovesan()` escrita por A.J. Pérez
+
 # References

### 2017-04-21 script/R/gea/BAItree.R

@@ -1,0 +1,83 @@
+####Ejemplo calculo BAI por arbol
+###17042017
+
+
+####### BAI  #####
+###MACRO PARA CALCULARSE BAI POR PARCELA
+### CUIDADO CON SERIES QUE NO EMPIEZAN EN LA MISMA FECHA QUE EL FINAL (PUEDEN METER SESGO): por eso uso BAItree
+### BAI calculado por árbol para evitar sesgo de series que no lleguen hasta el final ("flotantes")
+##
+
+###daps = archivo donde est´án los diametros que se han metido en campo
+### rwls = archivo .rwl que me importo con library(dplR) funci´ón read.rwl
+
+BAItree=function(daps, rwls)
+{
+  BAIdata=NULL
+  #
+  for (i in 1:(ncol(rwls)-1))
+  {
+    Var=as.data.frame(cbind(as.numeric((rwls[,i])), as.numeric(rwls$year)))
+    Var=subset(Var, !is.na(Var[,1]))
+    colnames(Var)=c("Growth", "year")
+    Var$age1=seq(1, nrow(Var), 1)
+    parcela2=as.numeric(substr(colnames(rwls[i]), 3,4))
+    arbol=as.numeric(substr(colnames(rwls[i]),5,7))
+    checkage=subset(daps, Parcela==parcela2 & Identidad..==arbol)
+    Var$dapcc=as.numeric(c(checkage[13]))
+    Var$dapsc=(Var$dapcc-(2*(0.02*Var$dapcc+0.39)))*10   ###pongo la de la encina de siempre, lo paso todo a mm
+    Var$tree=arbol
+    #
+    BAIdata=rbind(BAIdata, Var)
+  }
+  
+  BAIdatatree=aggregate(BAIdata, by=list(BAIdata$year, BAIdata$tree), mean)
+  #
+  BAIdatatree=BAIdatatree[order(BAIdatatree$tree, BAIdatatree$year),-c(4, 8)]
+  colnames(BAIdatatree)=c("year", "tree", "Growth", "age1", "dapcc", "dapsc")
+  #  
+  arboles=aggregate(BAIdatatree[,2], by=list(arbol=BAIdatatree$tree), mean)  
+  arboles=arboles[,1]
[…]

### 2017-04-23 analysis/test-bai.Rmd
update source function
update

@@
 library("dplR")
-source('./script/R/bai_piovesan.R')
+library("knitr")
+source(paste0(getwd(), '/script/R/bai_piovesan.R'))
+
 ```
@@
 ```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
-di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
+# machine <- 'ajpelu'
+machine <- 'ajpeluLap'
+di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
 
@@
 ```
+
+## Aggregate RW by tree
+
+```{r}
+str(sj)
+
+name colnames(sj)[1]
+
+i <- '01'
+
+# Name of the tree to be selected  
+id_tree <- paste0('SNA',i)
+
+# Name to store the mean value 
+name_avg <- paste0('a_',i)
+
+# Get year (rownames)
+df <- sj %>% mutate(year=rownames(.))
+
+# Get name of the trees 
+# Name of each serie: i.e.:  SNA0101 (SNA|id_tree|id_core) 
+name_trees <- unique(stringr::str_sub(colnames(sj), start=1, end=5))
+
+# name <- name_trees[1]
+name <- name_trees[i]
+
+aver <- df %>% 
+  gather(variable, value, starts_with(name)) %>%  
+  group_by(year) %>%  
+  mutate(avg = mean(value, na.rm=TRUE)) %>% 
+  spread(variable, value)
+
+i<- 'SNA01'
+
+output <- list()
+
+for (i in seq_along(name_trees)) {
+  name <- name_trees[i]
+
+  name_avg <- paste0('a_',stringr::str_sub(name, star=4, end=5))
+
+  aux <- sj %>% select(starts_with(name)) %>% rowMeans() %>% as.data.frame()
+  names(aux) <- paste0('a_',stringr::str_sub(name, star=4, end=5))
+
+  output <- cbind(output, aux)
+  }
+
+name_tr
+
+names_trees
+
+name_rwsei
+code_site <- 'SNA' 
+id_tree <- stringr::str_sub('SNA0102', 4,5)
+id_replicate <-
+
+mm <- sj %>% select(SNA0101, SNA0102) %>% mutate(SNA01 = rowMeans(.))
+
+mutate(year=rownames_to_column(.)) %>%
 

### 2017-04-23 man/notes_dendro_analysis.Rmd
update

--- 2017-04-21 man/notes_dendro_analysis.Rmd
@@
 
+### Procedimiento
+* Aproximaciones y funciones en R:
+
+    * `dplR::bai.out()`: Utiliza sumatorio de rw si no se le proporcionan los diámetros. No elimina valor de corteza.
+    * Aproximación de @Piovesan2008.
+Podemos utilizar varias funciones:  `BAItree()` escrita por G.
+Gea (código aquí: [./script/R/gea/BAItree.R](/script/R/gea/BAItree.R)); o la función `bai_piovesan()` escrita por A.J.
+Pérez (código aquí: [./script/R/bai_piovesan.R](/script/R/bai_piovesan.R))
+
+* Calcular el BAI por cada árbol haciendo la media de crecimiento (RW) a nivel de árbol.
+
+* :red_circle: Estimación corteza en BAItree como lo ha calculado (lo de la encina)
+* :red_circle: Dudas funcion BAItree Guillermo.
+
+--------------------------
+
+### Planteamiento del análisis
+
+* 2 sitios y dos cotas por sitio 
+* Comparación formal entre crecimientos entre sitios 
+* 4 cronos separadas y dos cronos promedio
+
 # References

### 2017-04-27 script/R/rwl_byTree.R
Create function to compute rwl by tree

@@ -1,0 +1,41 @@
+# Function rwl by tree
+
+# Compute rwl mean by tree
+# This function needs a dataframe with the ring width measurements (a dataframe of the class 'rwl' from dplR pkg) 
+# It creates a new dataframe with the average value of the ring width (RW) by tree; and also a varible with the year. 
+# Tree names coming from colnames of the dataframe (see from class::rwl). The colnames of this rwl have the 
+# following structure: SNA0101 (SNA|id_tree|id_core). 
+ 
+rwl_byTree <- function(rwldf){ 
+  require(dplyr)
+  
+  # Create output with rownmaes (year) as a variable 
+  output_avg <- rwldf %>% 
+    mutate(year = as.numeric(rownames(.))) %>% 
+    dplyr::select(year)
+  
+  # Get name of each tree (each serie: i.e.:  SNA0101 (SNA|id_tree|id_core)) 
+  tree_names <- unique(stringr::str_sub(names(rwldf), start=1, end=5))
+  
+  # Loop 
+  for (i in tree_names) {
+    # Create a variable name for the average by tree 
+    name_avg <- paste0('rw_',stringr::str_sub(i, star=3, end=5))
+    
+    # Create aux dataframe with rwl mean values and the year
+    aux <- rwldf %>% 
+      dplyr::select(starts_with(i)) %>% 
+      mutate_(.dots = setNames(list(~rowMeans(., na.rm=TRUE),
+                                    ~as.numeric(rownames(.))),
+                               nm=c(name_avg, "year"))) %>% 
+      dplyr::select(-starts_with(i))
+    
+    # Inner join with output by year  
+    output_avg <- output_avg %>% inner_join(aux, by='year')
+  }
+  
+  # Change NaN by NA
+  output_avg[is.na(output_avg)] <- NA
+  
+  output_avg
+}

### 2017-04-27 analysis/test-bai.Rmd
update analysis

--- 2017-04-11 analysis/test-bai.Rmd
@@
 library("tidyverse")
+library("stringr")
 library("dplR")
-source('./script/R/bai_piovesan.R')
+library("knitr")
+source(paste0(getwd(), '/script/R/bai_piovesan.R'))
+source(paste0(getwd(), '/script/R/rw_byTree.R'))
 ```
@@
 # machine <- 'ajpeluLap'
-di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
+di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
 
@@
 ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
+
+# Read diameters data
+compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+```
+
+## Aggregate RW by tree
+
+* Lectura y preparación de datos de diámetro
+```{r} 
+# Prepare Diameter data
+
+# Compute diameter (mm)
+compete <- compete %>% 
+  mutate(dn_mm = (perim_mm / pi))
+
+# Get only focal trees, and only selected variables 
+ft <- compete %>% 
+  filter(sp=='Focal') %>% 
+  filter(id_focal!='Fresno') %>% 
+  select(id_focal, loc, dn_mm, height_cm)
+
+# Set levels of eleveation 
+sj_lowcode  <- paste0('A', str_pad(1:10, 2, pad='0'))
+sj_highcode <- paste0('A', 11:20)
+ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')),
+            paste0('B', 26:30))
+ca_highcode <- paste0('B', 11:25)
+
+ft <- ft %>% 
+  mutate(elev = ifelse(id_focal %in% sj_lowcode, 'Low',
+                       ifelse(id_focal %in% sj_highcode, 'High',
+                              ifelse(id_focal %in% ca_lowcode, 'Low', 'High')))) %>%
+  mutate(site = paste0(loc, '_', elev))
+```
+
+* Agregar valores medios de RW 
+:red_circle: DOCUMENTAR 
+```{r}
+sj_tree <- rw_byTree(sj)
+ca_tree <- rw_byTree(ca)
+```
+
+* Crear diferentes dataset por sitio 
+```{r}
+# function to subset averaged RW 
+subsetRW <- function(rwdf, codes){ 
+  require(dplyr)
+  aux_codes <- paste0('rw_', codes)
+  mysubset <- rwdf %>% select(one_of(aux_codes), year)
+  mysubset
+  }
+
+sjL <- subsetRW(sj_tree, sj_lowcode)
+sjH <- subsetRW(sj_tree, sj_highcode)
+caL <- subsetRW(ca_tree, ca_lowcode)
+caH <- subsetRW(ca_tree, ca_highcode)
+
+ft %>% group_by(site) %>% do(data = (.)) %>% 
+             select(data) %>% 
+             lapply(function(x) {(x)}) %>% .[[1]]
+
 ```

### 2017-04-27 script/R/rw_byTree.R

@@ -1,21 +1,21 @@
-# Function rwl by tree
+# Name: rw_byTree
 
-# Compute rwl mean by tree
+# Compute rw mean by tree
 # This function needs a dataframe with the ring width measurements (a dataframe of the class 'rwl' from dplR pkg) 
 # It creates a new dataframe with the average value of the ring width (RW) by tree; and also a varible with the year. 
 # Tree names coming from colnames of the dataframe (see from class::rwl). The colnames of this rwl have the 
 # following structure: SNA0101 (SNA|id_tree|id_core). 
  
-rwl_byTree <- function(rwldf){ 
+rw_byTree <- function(rwdf){ 
   require(dplyr)
   
   # Create output with rownmaes (year) as a variable 
-  output_avg <- rwldf %>% 
+  output_avg <- rwdf %>% 
     mutate(year = as.numeric(rownames(.))) %>% 
     dplyr::select(year)
   
   # Get name of each tree (each serie: i.e.:  SNA0101 (SNA|id_tree|id_core)) 
-  tree_names <- unique(stringr::str_sub(names(rwldf), start=1, end=5))
+  tree_names <- unique(stringr::str_sub(names(rwdf), start=1, end=5))
   
   # Loop 
   for (i in tree_names) {
@@ -23,7 +23,7 @@ rwl_byTree <- function(rwldf){
     name_avg <- paste0('rw_',stringr::str_sub(i, star=3, end=5))
     
     # Create aux dataframe with rwl mean values and the year
-    aux <- rwldf %>% 
+    aux <- rwdf %>% 
       dplyr::select(starts_with(i)) %>% 
       mutate_(.dots = setNames(list(~rowMeans(., na.rm=TRUE),
                                     ~as.numeric(rownames(.))),

### 2017-05-08 man/notes_dendro_analysis.Rmd
update links cofecha

@@
 ## Sincronización 
-Sincronizamos visualmene cada una de las series de radios (rw) y posteriormente verificamos la sincronización con  COFECHA [@Holmes1983]
+Sincronizamos visualmene cada una de las series de radios (rw) y posteriormente verificamos la sincronización con  COFECHA [@Holmes1983].
+Se puede obtener COFECHA en este [enlace](http://web.utk.edu/~grissino/software.htm).
+Resultados COFECHA:
 
-* [ ] :red_circle: Link al ouptut y análisis de los datos de COFECHA
-* [ ] :red_circle: Tabla 1 con estadísticos (ver otros ejemplos)
+* San Juan: [`./data_raw/dendro_ring/sn_sanjuan_cofecha.OUT`](/data_raw/dendro_ring/sn_sanjuan_cofecha.OUT) 
+* Cañar: [`./data_raw/dendro_ring/sn_sanjuan_cofecha.OUT`](/data_raw/dendro_ring/sn_sanjuan_cofecha.OUT)
+
+* [ ] :red_circle: Tabla con estadísticos (ver otros ejemplos): Ha de te
 

### 2017-05-08 man/notes_dendro_analysis.Rmd
update dendro analysis

@@
 
-* San Juan: [`./data_raw/dendro_ring/sn_sanjuan_cofecha.OUT`](/data_raw/dendro_ring/sn_sanjuan_cofecha.OUT) 
-* Cañar: [`./data_raw/dendro_ring/sn_sanjuan_cofecha.OUT`](/data_raw/dendro_ring/sn_sanjuan_cofecha.OUT)
+* San Juan: [`./data_raw/dendro_ring/sn_sanjuan/sn_sanjuan_cofecha.OUT`](/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan_cofecha.OUT) 
+* Cañar: [`./data_raw/dendro_ring/sn_canar/sn_canar_cofecha.OUT`](/data_raw/dendro_ring/sn_canar/sn_canar_cofecha.OUT)
 

### 2017-05-09 analysis/get_topo_values.Rmd
Get topo values upload csv with topo values

---
title: "Topographical values"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 May"
output:  
  md_document:
    variant: markdown_github
---

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("raster")
library("stringr")
library('rgdal')
library("sp")
```

# Read and prepare data 
```{r, eval=FALSE}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# Read spatial data 
# Centroid of focal tree 
## filename 
mygpx_ca <- paste0(di, 'data_raw/geoinfo/dendro_ca.GPX')
mygpx_sj <- paste0(di, 'data_raw/geoinfo/dendro_sj.GPX')

## reading  
field_work_ca <- readOGR(mygpx_ca, layer="waypoints", verbose = FALSE)
field_work_sj <- readOGR(mygpx_sj, layer="waypoints", verbose = FALSE)
```

# Topographic data
```{r, eval=FALSE}
# Get Elevation and derivate data 
# We use DEM fron SN

# Connect with NAS and copy the mde file located at 
# '/cartografia/Informacion_Ambiental/BASES_DE_REFERENCIA_TERRITORIAL/MDE_SN_AreaInfluecia_10m/mde_sn_area_influencia.asc' 
ditemp <- paste('/Users/', machine, '/Documents/', sep='')
setwd(ditemp)

# read DEM 
demsn <- raster("mde_sn_area_influencia.asc")

setwd(di)

## Projection 
# Get projection
projection(demsn)

# Set projection
crs(demsn) <- "+proj=utm +zone=30 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"

# Get slope and aspect 
slopesn10 <- terrain(demsn, opt="slope", unit = 'degrees')
aspectsn10 <- terrain(demsn, opt="aspect", unit = 'degrees')

# Reproject spatial info of trees 
field_work_ca_re <- spTransform(field_work_ca, crs(demsn))
field_work_sj_re <- spTransform(field_work_sj, crs(demsn))

# Extract elev 
ca_elev <- extract(demsn, field_work_ca_re, method = 'simple', sp=TRUE)
sj_elev <- extract(demsn, field_work_sj_re, method = 'simple', sp=TRUE)

# Extract Slope
ca_slope <- extract(slopesn10, field_work_ca_re, method = 'simple', sp=TRUE)
sj_slope <- extract(slopesn10, field_work_sj_re, method = 'simple', sp=TRUE)

# Extract Aspect 
ca_aspect <- extract(aspectsn10, field_work_ca_re, method = 'simple', sp=TRUE)
sj_aspect <- extract(aspectsn10, field_work_sj_re, method = 'simple', sp=TRUE)

## Reorganize and join 
ca_elev_s <- as.data.frame(ca_elev) %>% 
  dplyr::select(name, mde = mde_sn_area_influencia)

ca_slope_s <- as.data.frame(ca_slope) %>% 
  dplyr::select(name,slope)

ca_aspect_s <- as.data.frame(ca_aspect) %>% 
  dplyr::select(name,aspect)

ca_topo <- ca_elev_s %>% left_join(ca_slope_s, by='name') %>% left_join(ca_aspect_s, by= 'name')

## Reorganize and join 
sj_elev_s <- as.data.frame(sj_elev) %>% 
  dplyr::select(name, mde = mde_sn_area_influencia)

sj_slope_s <- as.data.frame(sj_slope) %>% 
  dplyr::select(name,slope)

sj_aspect_s <- as.data.frame(sj_aspect) %>% 
  dplyr::select(name,aspect)

sj_topo <- sj_elev_s %>% left_join(sj_slope_s, by='name') %>% left_join(sj_aspect_s, by= 'name')

# Join all sites and convert aspect into categorical 
topo <- sj_topo %>% rbind(ca_topo) %>% 
  mutate(aspectF = cut(aspect, 
                       breaks= c(0,22.5, 67.5, 112.5, 157.5, 202.5, 247.5, 292.5, 337.5, 359.5),
                       labels = c("N", "NE", "E", "SE", "S", "SW", "W", "NW","N")))

# Export data
write.csv(topo, file=paste(di, "/data/topo/topo.csv", sep=""), row.names = FALSE)
```

### 2017-05-09 analysis/analysis_stand_features.Rmd
add stand features

--- 2017-03-03 analysis/analysis_competence.Rmd
@@
 ---
-title: "Competence distance"
+title: "Analysis Stand Features "
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 Feb"
+date: "2017 May"
 output:  
@@
 library("tidyverse")
+library("raster")
+library("stringr")
+library('rgdal')
+library("sp")
 ```
 
-# Read data 
+# Read and prepare data 
 ```{r, echo=FALSE}
-#machine <- 'ajpelu'
-machine <- 'ajpeluLap'
-di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
+di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
 
-# density data 
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+# Read tree data
+tree <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+
+# Read spatial data 
+# Centroid of focal tree 
+## filename 
+mygpx_ca <- paste0(di, 'data_raw/geoinfo/dendro_ca.GPX')
+mygpx_sj <- paste0(di, 'data_raw/geoinfo/dendro_sj.GPX')
+
+## reading  
+field_work_ca <- readOGR(mygpx_ca, layer="waypoints", verbose = FALSE)
+field_work_sj <- readOGR(mygpx_sj, layer="waypoints", verbose = FALSE)
 ```
 
+* Two datasets: focal tree and competence 
 ```{r}
-# Compute diameter normal (in meters)
-compete <- compete %>% 
-  mutate(dn = (perim_mm/1000) /pi) 
+# Compute diameter (mm)
+tree <- tree %>% 
+  mutate(dn_mm = (perim_mm / pi))
+
+# Set levels of eleveation 
+sj_lowcode  <- paste0('A', str_pad(1:10, 2, pad='0'))
+sj_highcode <- paste0('A', 11:20)
+ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')),
+            paste0('B', 26:30))
+ca_highcode <- paste0('B', 11:25)
+
+tree <- tree %>% 
+  mutate(elevF = ifelse(id_focal %in% sj_lowcode, 'Low',
+                       ifelse(id_focal %in% sj_highcode, 'High',
+                              ifelse(id_focal %in% ca_lowcode, 'Low', 'High')))) %>%
+  mutate(site = paste0(loc, '_', elevF))
+
+# Get only focal trees  
+ft <- tree %>% 
+  filter(sp=='Focal') %>% 
+  filter(id_focal!='Fresno')
+
+# Get only no focal trees 
+nft <- tree %>% 
+  filter(sp!='Focal') 
+```
+
+# Spatial Info
+```{r, spatialData} 
+## Get coordinates of spatial data  
+sp_ca <- as.data.frame(field_work_ca) %>% 
+  dplyr::select(ele, name, lat = coords.x2, long = coords.x1)
+
+sp_sj <- as.data.frame(field_work_sj) %>% 
+  dplyr::select(ele, name, lat = coords.x2, long = coords.x1)
+
+## Add code site and get centroid (see # http://rspatial.org/analysis/rst/8-pointpat.html)
+sp_ca <- sp_ca %>% 
+  mutate(loc = 'CA', 
+         elevF = ifelse(name %in% ca_lowcode, 'Low', 'High'),
+         site = paste0(loc, '_', elevF))
+
+coord_ca <- sp_ca %>% 
+  group_by(site) %>%
+  summarise(lat_m = mean(lat),
+            long_m = mean(long))
+
+# plot(sp_ca$long, sp_ca$lat, pch=19, col='gray')
+# points(coord_ca$long_m, coord_ca$lat_m, pch=19, col='blue')
+
+sp_sj <- sp_sj %>% 
+  mutate(loc = 'SJ', 
+         elevF = ifelse(name %in% sj_lowcode, 'Low', 'High'),
+         site = paste0(loc, '_', elevF))        
+coord_sj <- sp_sj %>% 
+  group_by(site) %>%
+  summarise(lat_m = mean(lat),
+            long_m = mean(long))
+
+coords_sites <- coord_sj %>% rbind(coord_ca)
+
+#plot(sp_sj$long, sp_sj$lat, pch=19, col='gray')
+#points(coord_sj$long_m, coord_sj$lat_m, pch=19, col='blue')
+
+```
+
+# Topographic data
+```{r}
+# Read topo data 
+topo <- read.csv(file=paste(di, "/data/topo/topo.csv", sep=""), header=TRUE, sep=',')
+
+topo <- topo %>% 
+  mutate(loc = ifelse(str_detect(name, "A"), 'SJ', 'CA'),
+         elevF = ifelse(name %in% sj_lowcode, 'Low',
+                        ifelse(name %in% sj_highcode, 'High',
+                              ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
+  mutate(site = paste0(loc, '_', elevF))
+
+topo_summary <- topo %>%
+  group_by(site) %>%
+  summarise(mde_m = mean(mde),
+            mde_sd = sd(mde),
+            mde_min = min(mde),
+            mde_max = max(mde),
+            slope_m = mean(slope),
+            slope_sd = sd(slope))
+
 ```

### 2017-05-09 script/R/get_coords_trees.R
function to get coordinates of trees inside a plot

@@ -1,0 +1,12 @@
+## Coordinates of Trees relative to center of plot 
+# arguments: 
+# * angle: azimut between focal tree and measured tree (with 'brujula')
+# * distance: distance to center of plot (focal tree)
+# * x0, y0: coordinates of plot center (focal tree)
+
+
+get_coords_trees <- function(angle, distance, x0, y0) {
+  angle <- ifelse(angle <= 90, 90 - angle, 450 - angle)
+  data.frame(x = x0 + distance * cos(angle / 180 * pi),
+             y = y0+ distance * sin(angle / 180 * pi))
+}

### 2017-05-10 analysis/analysis_competence.Rmd
update competence analysis

@@
 library("tidyverse")
+library("stringr")
 ```
@@
 
-## Compute competence index
+## Compute competence indexes
 
-`TODO` documentar
+* Get unique identifier of focal tree 
+* Competence indexes: 
+    *
 
 ```{r}
-focal_trees <- compete %>% filter(id_core != 'Fresno') 
-focal_trees <- unique(focal_trees$id_core)
+# Get name of focal trees
+focal_trees <- compete %>% filter(id_focal != 'Fresno') 
+focal_trees <- unique(focal_trees$id_focal)
 
+# Empty object to store results 
 output<- vector("list", length(focal_trees))
@@
 
-# Create df   
-df <- compete %>% 
-  filter(id_core == focal_trees[[i]])
+  # Create df with all trees inside plot (10 m radius)
+  df <- compete %>% filter(id_focal == focal_trees[[i]])
 
-# Fixed Variables 
+  # Compute fixed variables: 
   radius_plot <- 10 # radius plot (10 m)
   sf_plot <- pi * radius_plot^2 # surface plot (m2)
-dnft <- filter(df, sp== 'Focal')$dn # normal diameter focal tree    
+  dnft <- filter(df, sp== 'Focal')$dn # normal diameter of focal tree    
   # ----------
@@
   # Competence indices 
-## Number of competitors (r = 10)
+  ## Number of competitors inside plot wihtout focal tree (r = 10)
   n_competitors <- filter(df, sp != 'Focal') %>% count()
@@
 
-## Plot density (tree/plot)
+  ## plot density (tree/plot)
   plot_density <- n_competitors / sf_plot
@@
 
-dist_nn <- (dist_nn$distance_cm / 100)
+  dist_nn <- (dist_nn$distance_cm / 100) #in meters
 
@@
   ### Crowding
+
   df_more_indices <- df %>%
     filter(sp != 'Focal') %>%
-  mutate(
-    basal_area = ((pi/4) * dn * dn), 
+    mutate(basal_area = ((pi/4) * dn * dn), 
            size_ratio_distance = (dn / dnft) * (1 / ((distance_cm/100) + 1)),
@@
            dnn = dist_nn,
-         id_plot = focal_trees[[i]])
+           id_focal = focal_trees[[i]])
 
   output[[i]] <- df_indices
+
 }
@@
 rm(n_competitors, n_competitors_hihger, plot_density, stand_density, sum_sizes, size_ratio, dist_nn, df, df_indices, output, df_more_indices, dnft)
+
+# Export data
+
 ```
@@
 ```{r}
-get_coords_trees <- function(angle, distance, x0, y0) {
-  angle <- ifelse(angle <= 90, 90 - angle, 450 - angle)
-  data.frame(x = x0 + distance * cos(angle / 180 * pi),
-             y = y0+ distance * sin(angle / 180 * pi))
-}
+source(paste0(getwd(), '/script/R/get_coords_trees.R'))
 
+# Add variables with coordinates of each tree. Use get_coords_trees() function 
 compete <- compete %>% 
@@
 
+## Plot by site 
 ```{r}
+# Set levels of eleveation 
+sj_lowcode  <- paste0('A', str_pad(1:10, 2, pad='0'))
+sj_highcode <- paste0('A', 11:20)
+ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')),
+            paste0('B', 26:30))
+ca_highcode <- paste0('B', 11:25)
+
+compete <- compete %>% 
+  mutate(elev = ifelse(id_focal %in% sj_lowcode, 'Low',
+                       ifelse(id_focal %in% sj_highcode, 'High',
+                              ifelse(id_focal %in% ca_lowcode, 'Low', 'High')))) %>%
+  mutate(site = paste0(loc, '_', elev))
+```
+
 # see stackoverflow questions 6862742
@@
 }
+```
 
@@
   ggplot(aes(x=xcord, y=ycord)) + 
-  facet_wrap(~id_core) + 
+  facet_wrap(~id_focal) + 
   geom_point(aes(size=perim_mm), shape=21, fill='transparent') + 

### 2017-05-10 analysis/analysis_competence.Rmd
Create csv with competence indexes

@@
 ```{r, echo=FALSE}
-#machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 ```{r}
-# Get name of focal trees
+# 1 Get name of focal trees
 focal_trees <- compete %>% filter(id_focal != 'Fresno') 
@@
 
-# Empty object to store results 
+# 2 Compute Fixed variables 
+radius_plot <- 10 # radius plot (10 m)
+sf_plot <- pi * radius_plot^2 # surface plot (m2)
+
+# 3 Empty object to store results 
 output<- vector("list", length(focal_trees))
 
+# 4 Loop 
 for (i in 1:length(focal_trees)){
@@
 
-  # Compute fixed variables: 
-  radius_plot <- 10 # radius plot (10 m)
-  sf_plot <- pi * radius_plot^2 # surface plot (m2)
-  dnft <- filter(df, sp== 'Focal')$dn # normal diameter of focal tree    
-  # ----------
+  # Focal tree
+  focal_t <- filter(df, sp== 'Focal') 
+  ## Normal diameter of focal tree
+  dnft <- focal_t$dn
 
@@
   ## Number of competitors inside plot wihtout focal tree (r = 10)
-  n_competitors <- filter(df, sp != 'Focal') %>% count()
+  n_competitors <- filter(df, sp != 'Focal') %>% count() %>% transmute(n_competitors = n)
 
@@
     filter(dn > dnft) %>% 
-    count()
+    count() %>% 
+    transmute(n_competitors_hihger = n)
 
   ## plot density (tree/plot)
-  plot_density <- n_competitors / sf_plot
+  plot_density <- n_competitors$n_competitors / sf_plot
 
   ## Stand density (tree / ha) 
-  stand_density <- (n_competitors * 10000) / sf_plot
+  stand_density <- (n_competitors$n_competitors * 10000) / sf_plot
 
@@
 
-df_indices_plot <- bind_rows(output)
+# df_indices_plot <- bind_rows(output)
+# As the output is a nested list I used this solution (http://stackoverflow.com/questions/27930883/converting-elements-in-a-nested-list-to-dataframe) to convert to a dataframe
+
+df_indices_plot <- bind_rows(lapply(output, as.data.frame.list, stringsAsFactors=FALSE))
 
@@
 # Export data 
-
+write.csv(df_indices_plot, file=paste(di, "data/competence/competence_indexes.csv", sep=""), row.names = FALSE)
 ```

### 2017-05-10 analysis/analysis_competence.Rmd
update

@@
 ---
-title: "Competence distance"
+title: "Analysis of Competence Index"
 author: "AJ Perez-Luque (@ajpelu)"
@@
 
-```{r, warning=FALSE, message=FALSE}}
+```{r, warning=FALSE, message=FALSE}
 library("tidyverse")
@@
 
-## Compute competence indexes
+## Compute competition indices
+
+The following competition indices were computed: 
+### Basal Area
+
+$$\sum_{i=1}^{n}pi\cdot (dbh_{i} / 2)^2$$
 
@@
 ```{r}
-source(paste0(getwd(), '/script/R/get_coords_trees.R'))
+source(paste0(di, '/script/R/get_coords_trees.R'))
 
@@
 ## Plot by site
+
 ```{r}
@@
 
+#### San Juan 
+```{r, echo=FALSE}
 # see stackoverflow questions 6862742
@@
 }
-```
 
@@
   filter(sp != 'Focal') %>%
+  filter(loc == 'SJ') %>% 
   ggplot(aes(x=xcord, y=ycord)) + 
   facet_wrap(~id_focal) + 
-  geom_point(aes(size=perim_mm), shape=21, fill='transparent') + 
+  geom_point(aes(size=dn), shape=21, fill='transparent') +
   xlim(-11,11) + ylim(-11,11) + 
@@
   theme(panel.grid.major=element_blank(), 
-        panel.grid.minor=element_blank()) 
+        panel.grid.minor=element_blank(), 
+        strip.background = element_rect(colour = "black", fill ='transparent')) 
 ```
 
-```{r}
-
+#### Canar 
+```{r, echo=FALSE, fig.height=8, fig.width=12}
+compete %>%  
+  filter(sp != 'Focal') %>%
+  filter(loc == 'CA') %>% 
+  ggplot(aes(x=xcord, y=ycord)) + 
+  facet_wrap(~id_focal) + 
+  geom_point(aes(size=dn), shape=21, fill='transparent') +
+  xlim(-11,11) + ylim(-11,11) + 
+  geom_point(aes(x=0, y=0), color='green') +
+  theme_bw() +
+  gg_circle(radius = 10, xcenter = 0, ycenter = 0) +
+  theme(panel.grid.major=element_blank(), 
+        panel.grid.minor=element_blank(), 
+        strip.background = element_rect(colour = "black", fill ='transparent')) 
 ```

### 2017-05-10 analysis/analysis_competence.Rmd
update latex and fig

@@
 
-$$\sum_{i=1}^{n}pi\cdot (dbh_{i} / 2)^2$$
+$$ \sum_{i=1}^{n}pi\cdot (dbh_{i} / 2)^2 $$
 
@@
 #### San Juan 
-```{r, echo=FALSE}
+```{r, echo=FALSE, fig.height=8, fig.width=10}
 # see stackoverflow questions 6862742
@@
 #### Canar 
-```{r, echo=FALSE, fig.height=8, fig.width=12}
+```{r, echo=FALSE, fig.height=9, fig.width=12}
 compete %>%  

### 2017-05-10 analysis/analysis_competence.Rmd
update references and include equations

@@
     variant: markdown_github
+bibliography: references.bib 
+csl: ecology.csl
 ---
@@
 
-The following competition indices were computed: 
-### Basal Area
+The following competition indices were computed. See @GeaIzquierdo2009 for more info.
 
-$$ \sum_{i=1}^{n}pi\cdot (dbh_{i} / 2)^2 $$
+### Distance-Independet Indices
 
-* Get unique identifier of focal tree 
-* Competence indexes: 
-    *
+* Basal Area (**ba**)
+
+$$ BA = \sum_{i=1}^{n}pi\cdot (dbh_{i} / 2)^2 $$
+
+* Stand Density (**std**)
+
+$$ std = n_{trees} / \mathrm{plot\, area(ha)} $$
+
+* Plot Density (**pd**)
+
+$$ pd = n_{trees} / \mathrm{plot\, area}$$
+
+* Number of competitors (**n_competitors**): number of competitors within *r* meters (10 m)
+
+* Number of competitors higher (**n_competitors_higher**): number of competitors within *r* meters (10 m) such that $ dbh_j > dbh_i $
+
+* Sum of size (**sum_sizes**) of trees within *r* meters (10 m): $ \sum dbh_{j} $
+
+* Size ratio (**sr**)
+
+$$ sr = dbh_i / \sum_{i=1}^{n}dbh_i $$
+
+### Distance-Dependet Indices
+
+* Distance to nearest tree (**dnn**)
+
+* Crowding
+
+$$ \mathrm{crowding} = \sum_{i=1}^{n}(dbh_i / dist_{ij}) $$
+
+* Lorimer
+
+$$ \mathrm{lorimer} = \sum_{i=1}^{n}\left [ (dbh_i / dbh_i) / \sqrt{dist_{ij}/r} \right ] $$
+
+* Negative Exponential size ratio (**nesr**)
+
+$$ \mathrm{nesr} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[ 1 / \exp(dist_{ij} + 1) \right ] $$
+
+* Negative Exponential Weighted size ratio (**newsr**)
+
+$$ \mathrm{newsr} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \exp\left[ (-dist_{ij} + 1) / (dbh_i + dbh_j) \right ] $$
+
+* Size ratio proportional to distance (**srd**)
+
+$$ \mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ] $$
+
+* Size difference proportional to distance (**sdd**)
+
+$$ \mathrm{sdd} = \sum_{i=1}^{n} \left[ (dbh_j  - dbh_i) / (dist_{ij} + 1) \right ] $$
 
@@
   ## Number of competitors with dn_j > dn_i (focal tree)
-  n_competitors_hihger <- df %>%
+  n_competitors_higher <- df %>%
     filter(sp != 'Focal') %>%
@@
     count() %>% 
-    transmute(n_competitors_hihger = n)
+    transmute(n_competitors_higher = n)
 
@@
               crowding = sum(crowding)) %>% 
-    mutate(nc = n_competitors,
-           nch = n_competitors_hihger,
+    mutate(n_competitors,
+           n_competitors_higher,
            pd = plot_density,
            std = stand_density,
-           ss = sum_sizes,
+           sum_sizes,
            sr = size_ratio,
@@
 
-rm(n_competitors, n_competitors_hihger, plot_density, stand_density, sum_sizes, size_ratio, dist_nn, df, df_indices, output, df_more_indices, dnft)
+rm(n_competitors, n_competitors_higher, plot_density, stand_density, sum_sizes, size_ratio, dist_nn, df, df_indices, output, df_more_indices, dnft)
 
@@
 ```
+
+## References

### 2017-05-10 analysis/analysis_competence.Rmd
update name variables

@@
   md_document:
+    toc: true
     variant: markdown_github
@@
 # Export data 
-write.csv(df_indices_plot, file=paste(di, "data/competence/competence_indexes.csv", sep=""), row.names = FALSE)
+write.csv(df_indices_plot, file=paste(di, "data/competence/competence_indices.csv", sep=""), row.names = FALSE)
 ```

### 2017-05-10 analysis/analysis_stand_features.Rmd
update stand features

@@
 
-## Compute competence index
+# Competition data
+```{r}
+# Read competition
+compe <- read.csv(file=paste(di, "/data/competence/competence_indices.csv", sep=""), header=TRUE, sep=',')
 
-`TODO` documentar
-
-```{r}
-focal_trees <- compete %>% filter(id_core != 'Fresno') 
-focal_trees <- unique(focal_trees$id_core)
-
-output<- vector("list", length(focal_trees))
-
-for (i in 1:length(focal_trees)){
-
-# Create df   
-df <- compete %>% 
-  filter(id_core == focal_trees[[i]])
-
-# Fixed Variables 
-radius_plot <- 10 # radius plot (10 m)
-sf_plot <- pi * radius_plot^2 # surface plot (m2)
-dnft <- filter(df, sp== 'Focal')$dn # normal diameter focal tree    
-# ----------
-
-# Competence indices 
-## Number of competitors (r = 10)
-n_competitors <- filter(df, sp != 'Focal') %>% count()
-
-## Number of competitors with dn_j > dn_i (focal tree)
-n_competitors_hihger <- df %>%
-  filter(sp != 'Focal') %>%
-  filter(dn > dnft) %>% 
-  count()
-
-## Plot density (tree/plot)
-plot_density <- n_competitors / sf_plot
-
-## Stand density (tree / ha) 
-stand_density <- (n_competitors * 10000) / sf_plot
-
-## Sum sizes 
-sum_sizes <- df %>% 
-  filter(sp != 'Focal') %>%
-  summarise(sum_sizes=sum(dn))
-
-## Size ratio 
-size_ratio <- dnft / (dnft + sum_sizes$sum_sizes)
-
-## Distance nearest tree
-dist_nn <- df %>% 
-  filter(sp != 'Focal') %>% 
-  slice(which.min(distance_cm))
-
-dist_nn <- (dist_nn$distance_cm / 100)
-
-## Various 
-### Basal area
-### Size ratio proportional to distance 
-### Size difference proportional to distance 
-### Negative exponential size ratio
-### Negative exponential weighted size ratio
-### Lorimer 
-### Crowding 
-df_more_indices <- df %>%
-  filter(sp != 'Focal') %>%
-  mutate(
-    basal_area = ((pi/4) * dn * dn), 
-    size_ratio_distance = (dn / dnft) * (1 / ((distance_cm/100) + 1)),
-    size_diff_distance = (dn - dnft) / ((distance_cm/100) + 1),
-    neg_exp_size_ratio = (dn / dnft) * (1 / exp((distance_cm/100) + 1)),
-    neg_exp_w_size_ratio = (dn / dnft) * exp((-(distance_cm/100) + 1) / (dn + dnft)),
-    lorimer = (dn / dnft) / (sqrt((distance_cm/100) / radius_plot)),
-    crowding = dn / (distance_cm/100))
-
-df_indices <- df_more_indices %>% 
-  summarise(ba = sum(basal_area),
-            srd = sum(size_ratio_distance),
-            sdd = sum(size_diff_distance),
-            nesr = sum(neg_exp_size_ratio),
-            newsr = sum(neg_exp_w_size_ratio),
-            lorimer = sum(lorimer),
-            crowding = sum(crowding)) %>% 
-  mutate(nc = n_competitors,
-         nch = n_competitors_hihger,
-         pd = plot_density,
-         std = stand_density,
-         ss = sum_sizes,
-         sr = size_ratio,
-         dnn = dist_nn, 
-         id_plot = focal_trees[[i]])
-
-output[[i]] <- df_indices 
-}
-
-df_indices_plot <- bind_rows(output)
-
-rm(n_competitors, n_competitors_hihger, plot_density, stand_density, sum_sizes, size_ratio, dist_nn, df, df_indices, output, df_more_indices, dnft)
 ```
 
-## Plot
-
-```{r}
-get_coords_trees <- function(angle, distance, x0, y0) {
-  angle <- ifelse(angle <= 90, 90 - angle, 450 - angle)
-  data.frame(x = x0 + distance * cos(angle / 180 * pi),
-             y = y0+ distance * sin(angle / 180 * pi))
-}
-
-compete <- compete %>% 
-  mutate(xcord = get_coords_trees(angle= rumbo_degree, 
-                           distance = distance_cm/100,
-                           x0 = 0, y0 = 0)$x,
-         ycord = get_coords_trees(angle = rumbo_degree, 
-                           distance = distance_cm/100,
-                           x0 = 0, y0 = 0)$y)
-```
-
-```{r}
-# see stackoverflow questions 6862742
-gg_circle <- function(radius, xcenter, ycenter, color='black', fill=NA, ...){
-  x <- xcenter + radius*cos(seq(0,pi, length.out = 100))
-  ymax <- ycenter + radius*sin(seq(0, pi, length.out = 100))
-  ymin <- ycenter + radius*sin(seq(0, -pi, length.out = 100))
-  annotate("ribbon", x=x, ymin=ymin, ymax=ymax, color=color, fill=fill, ...)
-}
-
-compete %>%  
-  filter(sp != 'Focal') %>%
-  ggplot(aes(x=xcord, y=ycord)) + 
-  facet_wrap(~id_core) + 
-  geom_point(aes(size=perim_mm), shape=21, fill='transparent') + 
-  xlim(-11,11) + ylim(-11,11) + 
-  geom_point(aes(x=0, y=0), color='green') +
-  theme_bw() +
-  gg_circle(radius = 10, xcenter = 0, ycenter = 0) +
-  theme(panel.grid.major=element_blank(), 
-        panel.grid.minor=element_blank()) 
-```
-
-```{r}
-
-```
+Conclusiones_Taller_EME_SN_borrador_para_devolucion.doc

### 2017-05-11 analysis/analysis_stand_features.Rmd
update aov analysis

@@
 library("sp")
+library("knitr")
+library("broom")
+library("multcomp")
+library("pander")
+library("devtools")
+source_gist("https://gist.github.com/ajpelu/194e721077ec045a2b864088908e7aca", filename = 'table_glht.r') # multicomp table
 ```
@@
 
+# General Variables 
+```{r}
+general_var <- ft %>% group_by(loc, site) %>% count()
+
+general_var %>% kable()
+```
+
 # Spatial Info
@@
 
+coords_sites %>% kable()
 ```
@@
 # Topographic data
+## Data summary 
 ```{r}
@@
                               ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
-  mutate(site = paste0(loc, '_', elevF))
+  mutate(site = paste0(loc, '_', elevF)) %>% 
+  mutate(site = as.factor(site),
+         loc = as.factor(loc),
+         elevF = as.factor(elevF))
 
@@
 
+# Another way to obtain summary values 
+variables <- c('mde','slope','aspect')
+auxdf <- data.frame()
+
+for (i in variables){ 
+aux <- topo %>% 
+  dplyr::group_by(site) %>% 
+  summarise_each_(funs(mean, sd, se=sd(.)/sqrt(n())), i) %>% mutate(variable=i)
+
+auxdf <- rbind(auxdf, aux) }
+```
+
+## ANOVAs 
+### Elevation
+```{r}
+mivariable <- 'mde'
+
+# Model 
+myformula <- as.formula(paste0(mivariable, " ~ site"))
+mymodel <- aov(myformula, data=topo)
+
+auxdf %>% 
+  dplyr::filter(variable==mivariable) %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+## Summary model 
+tm <- broom::tidy(mymodel)
+
+# See http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1) 
+pander(tm, round=5,caption = "ANOVA table", missing = '', 
+       emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+
+## Multiple comparison 
+tuk <- glht(mymodel, linfct = mcp(site = "Tukey"))
+
+# Convert comparisons into letters 
+df_letter <- fortify(cld(tuk)) %>%
+  transmute(site = as.factor(lhs),
+         tukey = letters) %>%
+  mutate(variable = mivariable)
+
+aux_name <- paste0('df_tuk_', mivariable)
+assign(aux_name, df_letter)
+
+# Get table 
+mymult <- as.data.frame(table_glht(tuk))
+pander(mymult, round=4,caption = "Post hoc comparison (Tukey, alpha = 0.05)", missing = '', 
+       emphasize.strong.cells = 
+               which(mymult < 0.05 & mymult == mymult[,4], arr.ind = T))
+```
+
+### Slope
+```{r}
+mivariable <- 'slope'
+
+# Model 
+myformula <- as.formula(paste0(mivariable, " ~ site"))
+mymodel <- aov(myformula, data=topo)
+
+auxdf %>% 
+  dplyr::filter(variable==mivariable) %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+## Summary model 
+tm <- broom::tidy(mymodel)
+
+# See http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1) 
+pander(tm, round=5,caption = "ANOVA table", missing = '', 
+       emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+
+## Multiple comparison 
+tuk <- glht(mymodel, linfct = mcp(site = "Tukey"))
+
+# Convert comparisons into letters 
+df_letter <- fortify(cld(tuk)) %>%
+  transmute(site = as.factor(lhs),
+         tukey = letters) %>%
+  mutate(variable = mivariable)
+
+aux_name <- paste0('df_tuk_', mivariable)
+assign(aux_name, df_letter)
+
+# Get table 
+mymult <- as.data.frame(table_glht(tuk))
+a <- pander(mymult, round=4,caption = "Post hoc comparison (Tukey, alpha = 0.05)", missing = '', 
+       emphasize.strong.cells = 
+               which(mymult < 0.05 & mymult == mymult[,4], arr.ind = T))
 ```
@@
 
+compe <- compe %>% 
+  mutate(name = id_focal) %>% 
+  mutate(loc = ifelse(str_detect(name, "A"), 'SJ', 'CA'),
+         elevF = ifelse(name %in% sj_lowcode, 'Low',
+                        ifelse(name %in% sj_highcode, 'High',
+                              ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
+  mutate(site = paste0(loc, '_', elevF)) %>% 
+  mutate(site = as.factor(site),
+         loc = as.factor(loc),
+         elevF = as.factor(elevF))
+
 ```
 
-Conclusiones_Taller_EME_SN_borrador_para_devolucion.doc
+```{r}
+xxx <- function(df, mivariable, mygroup){ 
+  require('dplyr')
+  require('multcomp')
+  require('broom') 
+  # mivariable: variabe of interest
+  # mygroup: factor variable
+
+  # Model formulation
+  myformula <- as.formula(paste0(mivariable, " ~ ", mygroup))
+
+  # AOV 
+  mymodel <- aov(myformula, data=df)
+
+  # Summary AOV (broom style)
+  tm <- broom::tidy(mymodel)
+
+  # Multiple comparison 
+  tuk <- glht(mymodel, linfct = mcp(mygroup = "Tukey"))
+
+  # Convert comparisons into letters 
+  df_letter <- fortify(cld(tuk)) %>% 
+    transmute(site = as.factor(lhs),
+         tukey = letters) %>%
+    mutate(variable = mivariable)
+
+  aux_name <- paste0('df_tuk_', mivariable)
+  assign(aux_name, df_letter)
+
+  mymult <- as.data.frame(table_glht(tuk))
+
+  mymodel
+  tm
+  aux_name
+  mymult
+}
+
+ args <- list("Tukey")
+           names(args) <- X
+           cmp <- do.call(mcp, args)
+           glht(fm1, linfct = cmp)
+
+df <- topo
+mivariable <- 'slope'
+mygroup <- 'site'
+
+ala <- xxx(df=topo, mivariable='slope', mygroup='site' )
+
+auxdf %>% 
+  dplyr::filter(variable==mivariable) %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+# See http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1) 
+pander(tm, round=5,caption = "ANOVA table", missing = '', 
+       emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+
+## Multiple comparison 
+tuk <- glht(mymodel, linfct = mcp(site = "Tukey"))
+
+# Convert comparisons into letters 
+df_letter <- fortify(cld(tuk)) %>%
+  transmute(site = as.factor(lhs),
+         tukey = letters) %>%
+  mutate(variable = mivariable)
+
+aux_name <- paste0('df_tuk_', mivariable)
+assign(aux_name, df_letter)
+
+# Get table 
+mymult <- as.data.frame(table_glht(tuk))
+a <- pander(mymult, round=4,caption = "Post hoc comparison (Tukey, alpha = 0.05)", missing = '', 
+       emphasize.strong.cells = 
+               which(mymult < 0.05 & mymult == mymult[,4], arr.ind = T))
+
+```

### 2017-05-15 analysis/analysis_stand_features.Rmd
update function

@@
 ```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
+# machine <- 'ajpelu'
+machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 
-general_var %>% kable()
+general_var %>% kable
 ```
@@
 ```{r}
-xxx <- function(df, mivariable, mygroup){ 
+compara <- function(df, mivariable){ 
   require('dplyr')
@@
   # mivariable: variabe of interest
-  # mygroup: factor variable
+
+  output <- list()
 
   # Model formulation
-  myformula <- as.formula(paste0(mivariable, " ~ ", mygroup))
+  myformula <- as.formula(paste0(mivariable, " ~ site"))
 
@@
   # Multiple comparison 
-  tuk <- glht(mymodel, linfct = mcp(mygroup = "Tukey"))
+  tuk <- glht(mymodel, linfct = mcp(site = "Tukey"))
 
@@
 
-  mymodel
-  tm
-  aux_name
-  mymult
+  output$mymodel <- mymodel
+  output$tm <- tm
+  output$aux_name <- df_letter
+  output$mymult <- mymult
+
+  return(output)
+}  
+```
+
+```{r}
+
+# Select only variables to compare 
+compe_sel <- compe %>% dplyr::select(ba:dnn,site)
+# Get vector with variables 
+variables <- compe %>% dplyr::select(ba:dnn) %>% names()
+
+for (i in variables){
+
+  # apply comparison
+  out_compara <- compara(df=compe_sel, mivariable = i)
+
+  out_name  <- paste0('aov_', i)
+  assign(out_name, out_compara)
+
   }
 
- args <- list("Tukey")
-           names(args) <- X
-           cmp <- do.call(mcp, args)
-           glht(fm1, linfct = cmp)
-
-df <- topo
-mivariable <- 'slope'
-mygroup <- 'site'
-
-ala <- xxx(df=topo, mivariable='slope', mygroup='site' )
+ala <- xxx(df=topo, mivariable='slope')
 
@@
 mymult <- as.data.frame(table_glht(tuk))
-a <- pander(mymult, round=4,caption = "Post hoc comparison (Tukey, alpha = 0.05)", missing = '', 
+
+for (i in variables){
+
+  mymult <-
+
+  aux <- 
+pander(aov_dnn$mymult, round=4,caption = "Post hoc comparison (Tukey, alpha = 0.05)", missing = '', 
      emphasize.strong.cells = 
-               which(mymult < 0.05 & mymult == mymult[,4], arr.ind = T))
+             which(aov_ba$mymult < 0.05 & aov_ba$mymult == aov_ba$mymult[,4], arr.ind = T))
+
+aov_ba$mymult
 

### 2017-05-15 analysis/analysis_stand_features.Rmd
analyze stand features and create output

@@
 ```{r, echo=FALSE}
-# machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 
+#### Function to comparison (aov)
 ```{r}
@@
 
+  # Return also summary values 
+  summ_values <- df %>% 
+    dplyr::select_(mivariable, 'site') %>%
+    group_by(site) %>% 
+    summarise_each(funs(mean, sd, se=sd(.)/sqrt(n())))
+
+  # Join letters and summary values
+  output$summ_values <- summ_values
+
+  summ_comparison <- summ_values %>% left_join(df_letter, by='site')
+
+  output$summ_comparison <- summ_comparison
+
   return(output)
 }
+
 ```
+
+## Comparison
 
@@
 compe_sel <- compe %>% dplyr::select(ba:dnn,site)
+
 # Get vector with variables 
@@
 
-ala <- xxx(df=topo, mivariable='slope')
+# Loop to export into txt files (see ./out/anovas_competition ... )
+for (i in variables){
 
-auxdf %>% 
-  dplyr::filter(variable==mivariable) %>% 
+  out <- get(paste0('aov_', i))
+
+  sink(paste0(di,'out/anovas_competition/aov_', i, '.txt'))
+  cat("MODEL \n")
+  print(out$mymodel) 
+  cat("\n")
+
+  cat("MODEL pretty \n")
+  print(out$tm) 
+  cat("\n")
+
+  cat("POST HOC \n")
+  print(out$mymult) 
+  cat("\n")
+
+  cat("SUMMARY VALUES \n")
+  print(out$summ_comparison) 
+  cat("\n")
+
+  sink()
+}
+```
+
+## Distance-Independet Indices 
+### Basal Area
+```{r}
+mivariable <- 'ba'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
   pander(round=4, caption=paste0('Mean values (', mivariable,')'))
 
-# See http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1) 
-pander(tm, round=5,caption = "ANOVA table", missing = '', 
-       emphasize.strong.cells = 
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
                which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
 
-## Multiple comparison 
-tuk <- glht(mymodel, linfct = mcp(site = "Tukey"))
+### Stand Density
+```{r}
+mivariable <- 'std'
+s <- get(paste0('aov_', mivariable))
 
-# Convert comparisons into letters 
-df_letter <- fortify(cld(tuk)) %>%
-  transmute(site = as.factor(lhs),
-         tukey = letters) %>%
-  mutate(variable = mivariable)
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
 
-aux_name <- paste0('df_tuk_', mivariable)
-assign(aux_name, df_letter)
+tm <- s$tm
 
-# Get table 
-mymult <- as.data.frame(table_glht(tuk))
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
 
-for (i in variables){
+### Plot Density
+```{r}
+mivariable <- 'pd'
+s <- get(paste0('aov_', mivariable))
 
-  mymult <-
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
 
-  aux <- 
-pander(aov_dnn$mymult, round=4,caption = "Post hoc comparison (Tukey, alpha = 0.05)", missing = '', 
-     emphasize.strong.cells = 
-             which(aov_ba$mymult < 0.05 & aov_ba$mymult == aov_ba$mymult[,4], arr.ind = T))
+tm <- s$tm
 
-aov_ba$mymult
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
 
+### Number of competitors within *r* meters (10 m)
+```{r}
+mivariable <- 'n_competitors'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
 ```
+
+### Number of competitors within *r* meters (10 m) such that $ dbh_j > dbh_i $
+```{r}
+mivariable <- 'n_competitors_higher'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Sum of size of trees within *r* meters (10 m)
+
+```{r}
+mivariable <- 'sum_sizes'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+## Size ratio
+
+```{r}
+mivariable <- 'sr'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+## Distance-Dependet Indices
+
+### Distance to nearest tree       
+```{r}
+mivariable <- 'dnn'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Crowding       
+```{r}
+mivariable <- 'crowding'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Lorimer
+
+```{r}
+mivariable <- 'lorimer'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Negative Exponential size ratio
+
+```{r}
+mivariable <- 'nesr'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Negative Exponential Weighted size ratio
+```{r}
+mivariable <- 'newsr'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Size ratio proportional to distance
+
+```{r}
+mivariable <- 'srd'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Size difference proportional to distance
+
+```{r}
+mivariable <- 'sdd'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```

### 2017-05-15 analysis/analysis_stand_features.Rmd
update files

@@
 
-```{r, warning=FALSE, message=FALSE}
+```{r, warning=FALSE, message=FALSE, echo=FALSE}
 library("tidyverse")
@@
 
-# Read and prepare data 
+# Prepare Data
 ```{r, echo=FALSE}
@@
 * Two datasets: focal tree and competence
+
 ```{r}
@@
 # General Variables 
+Numbers of focal trees by site
+
 ```{r}
@@
 # Spatial Info
+Coordinates of the centroid for each site
+
 ```{r, spatialData} 
@@
 # Competition data
-```{r}
+
+* Read data 
+```{r, echo=FALSE}
 # Read competition
@@
 
+* Create a custom function to compare between sites (aov & post hoc)
+
+```{r, echo=FALSE}
 #### Function to comparison (aov)
-```{r}
 compara <- function(df, mivariable){ 
@@
 
+* Export data into text files (see `/out/anovas_competition/`)
+
+```{r, echo=FALSE}
 ## Comparison 
-
-```{r}
-
 # Select only variables to compare 
@@
 ### Basal Area
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'ba'

### 2017-05-15 analysis/analysis_stand_features.Rmd
update function; add min max; loop for topo variables; close output

@@
 
-# Topographic data
-## Data summary 
-```{r}
-# Read topo data 
-topo <- read.csv(file=paste(di, "/data/topo/topo.csv", sep=""), header=TRUE, sep=',')
-
-topo <- topo %>% 
-  mutate(loc = ifelse(str_detect(name, "A"), 'SJ', 'CA'),
-         elevF = ifelse(name %in% sj_lowcode, 'Low',
-                        ifelse(name %in% sj_highcode, 'High',
-                              ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
-  mutate(site = paste0(loc, '_', elevF)) %>% 
-  mutate(site = as.factor(site),
-         loc = as.factor(loc),
-         elevF = as.factor(elevF))
-
-topo_summary <- topo %>%
-  group_by(site) %>%
-  summarise(mde_m = mean(mde),
-            mde_sd = sd(mde),
-            mde_min = min(mde),
-            mde_max = max(mde),
-            slope_m = mean(slope),
-            slope_sd = sd(slope))
-
-# Another way to obtain summary values 
-variables <- c('mde','slope','aspect')
-auxdf <- data.frame()
-
-for (i in variables){ 
-aux <- topo %>% 
-  dplyr::group_by(site) %>% 
-  summarise_each_(funs(mean, sd, se=sd(.)/sqrt(n())), i) %>% mutate(variable=i)
-
-auxdf <- rbind(auxdf, aux) }
-```
-
-## ANOVAs 
-### Elevation
-```{r}
-mivariable <- 'mde'
-
-# Model 
-myformula <- as.formula(paste0(mivariable, " ~ site"))
-mymodel <- aov(myformula, data=topo)
-
-auxdf %>% 
-  dplyr::filter(variable==mivariable) %>% 
-  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
-
-## Summary model 
-tm <- broom::tidy(mymodel)
-
-# See http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1) 
-pander(tm, round=5,caption = "ANOVA table", missing = '', 
-       emphasize.strong.cells = 
-               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
-
-## Multiple comparison 
-tuk <- glht(mymodel, linfct = mcp(site = "Tukey"))
-
-# Convert comparisons into letters 
-df_letter <- fortify(cld(tuk)) %>%
-  transmute(site = as.factor(lhs),
-         tukey = letters) %>%
-  mutate(variable = mivariable)
-
-aux_name <- paste0('df_tuk_', mivariable)
-assign(aux_name, df_letter)
-
-# Get table 
-mymult <- as.data.frame(table_glht(tuk))
-pander(mymult, round=4,caption = "Post hoc comparison (Tukey, alpha = 0.05)", missing = '', 
-       emphasize.strong.cells = 
-               which(mymult < 0.05 & mymult == mymult[,4], arr.ind = T))
-```
-
-### Slope
-```{r}
-mivariable <- 'slope'
-
-# Model 
-myformula <- as.formula(paste0(mivariable, " ~ site"))
-mymodel <- aov(myformula, data=topo)
-
-auxdf %>% 
-  dplyr::filter(variable==mivariable) %>% 
-  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
-
-## Summary model 
-tm <- broom::tidy(mymodel)
-
-# See http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1) 
-pander(tm, round=5,caption = "ANOVA table", missing = '', 
-       emphasize.strong.cells = 
-               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
-
-## Multiple comparison 
-tuk <- glht(mymodel, linfct = mcp(site = "Tukey"))
-
-# Convert comparisons into letters 
-df_letter <- fortify(cld(tuk)) %>%
-  transmute(site = as.factor(lhs),
-         tukey = letters) %>%
-  mutate(variable = mivariable)
-
-aux_name <- paste0('df_tuk_', mivariable)
-assign(aux_name, df_letter)
-
-# Get table 
-mymult <- as.data.frame(table_glht(tuk))
-a <- pander(mymult, round=4,caption = "Post hoc comparison (Tukey, alpha = 0.05)", missing = '', 
-       emphasize.strong.cells = 
-               which(mymult < 0.05 & mymult == mymult[,4], arr.ind = T))
-```
-
 # Competition data
@@
     group_by(site) %>% 
-    summarise_each(funs(mean, sd, se=sd(.)/sqrt(n())))
+    summarise_each(funs(mean, sd, se=sd(.)/sqrt(n()), min, max))
 
@@
 
-  sink(paste0(di,'out/anovas_competition/aov_', i, '.txt'))
+  file_out <- file(paste0(di,'out/anovas_competition/aov_', i, '.txt'))
+  sink(file_out)
   cat("MODEL \n")
@@
   sink()
+  close(file_out)
 }
@@
 ### Stand Density
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'std'
@@
 ### Plot Density
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'pd'
@@
 ### Number of competitors within *r* meters (10 m)
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'n_competitors'
@@
 ### Number of competitors within *r* meters (10 m) such that $ dbh_j > dbh_i $
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'n_competitors_higher'
@@
 
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'sum_sizes'
@@
 
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'sr'
@@
 ### Distance to nearest tree       
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'dnn'
@@
 ### Crowding       
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'crowding'
@@
 
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'lorimer'
@@
 
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'nesr'
@@
 ### Negative Exponential Weighted size ratio
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'newsr'
@@
 
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'srd'
@@
 
-```{r}
+```{r, echo=FALSE}
 mivariable <- 'sdd'
@@
 ```
+
+# Topographic data
+
+* Read data 
+```{r}
+# Read topo data 
+topo <- read.csv(file=paste(di, "/data/topo/topo.csv", sep=""), header=TRUE, sep=',')
+
+topo <- topo %>% 
+  mutate(loc = ifelse(str_detect(name, "A"), 'SJ', 'CA'),
+         elevF = ifelse(name %in% sj_lowcode, 'Low',
+                        ifelse(name %in% sj_highcode, 'High',
+                              ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
+  mutate(site = paste0(loc, '_', elevF)) %>% 
+  mutate(site = as.factor(site),
+         loc = as.factor(loc),
+         elevF = as.factor(elevF))
+
+topo_summary <- topo %>%
+  group_by(site) %>%
+  summarise(mde_m = mean(mde),
+            mde_sd = sd(mde),
+            mde_min = min(mde),
+            mde_max = max(mde),
+            slope_m = mean(slope),
+            slope_sd = sd(slope))
+
+# Another way to obtain summary values 
+variables <- c('mde','slope','aspect')
+auxdf <- data.frame()
+
+for (i in variables){ 
+aux <- topo %>% 
+  dplyr::group_by(site) %>% 
+  summarise_each_(funs(mean, sd, se=sd(.)/sqrt(n())), i) %>% mutate(variable=i)
+
+auxdf <- rbind(auxdf, aux) }
+```
+
+* Compare data and export data into text files (see `/out/anovas_topo/`)
+
+```{r, echo=FALSE}
+## Comparison 
+# Select only variables to compare
+
+topo_sel <- topo %>% dplyr::select(mde, slope, site)
+
+# Get vector with variables 
+variables <- c('mde','slope')
+
+for (i in variables){
+
+  # apply comparison
+  out_compara <- compara(df=topo_sel, mivariable = i)
+
+  out_name  <- paste0('aov_', i)
+  assign(out_name, out_compara)
+
+}
+
+# Loop to export into txt files (see ./out/anovas_topo ... )
+for (i in variables){
+
+  out <- get(paste0('aov_', i))
+
+  file_out <- file(paste0(di,'out/anovas_topo/aov_', i, '.txt'))
+  sink(file_out)
+  cat("MODEL \n")
+  print(out$mymodel) 
+  cat("\n")
+
+  cat("MODEL pretty \n")
+  print(out$tm) 
+  cat("\n")
+
+  cat("POST HOC \n")
+  print(out$mymult) 
+  cat("\n")
+
+  cat("SUMMARY VALUES \n")
+  print(out$summ_comparison) 
+  cat("\n")
+
+  sink()
+  close(file_out)
+}
+```
+
+### Elevation
+
+```{r, echo=FALSE}
+mivariable <- 'mde'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Slope
+
+```{r, echo=FALSE}
+mivariable <- 'slope'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### Min & max mde

### 2017-05-15 analysis/analysis_stand_features.Rmd
updating

@@
   cat("SUMMARY VALUES \n")
-  print(out$summ_comparison) 
+  print(as.data.frame(out$summ_comparison))
   cat("\n")
@@
 
-topo_summary <- topo %>%
-  group_by(site) %>%
-  summarise(mde_m = mean(mde),
-            mde_sd = sd(mde),
-            mde_min = min(mde),
-            mde_max = max(mde),
-            slope_m = mean(slope),
-            slope_sd = sd(slope))
-
-# Another way to obtain summary values 
-variables <- c('mde','slope','aspect')
-auxdf <- data.frame()
-
-for (i in variables){ 
-aux <- topo %>% 
-  dplyr::group_by(site) %>% 
-  summarise_each_(funs(mean, sd, se=sd(.)/sqrt(n())), i) %>% mutate(variable=i)
-
-auxdf <- rbind(auxdf, aux) }
+# 
+# topo_summary <- topo %>%
+#   group_by(site) %>%
+#   summarise(mde_m = mean(mde),
+#             mde_sd = sd(mde),
+#             mde_min = min(mde),
+#             mde_max = max(mde),
+#             slope_m = mean(slope),
+#             slope_sd = sd(slope))
+# 
+# 
+# 
+# # Another way to obtain summary values 
+# variables <- c('mde','slope','aspect')
+# auxdf <- data.frame() 
+# 
+# for (i in variables){ 
+# aux <- topo %>% 
+#   dplyr::group_by(site) %>% 
+#   summarise_each_(funs(mean, sd, se=sd(.)/sqrt(n())), i) %>% mutate(variable=i) 
+# 
+# auxdf <- rbind(auxdf, aux) }
 ```
@@
 ```
-
-### Min & max mde

### 2017-05-15 analysis/analysis_stand_features.Rmd
close sink et al

@@
   out <- get(paste0('aov_', i))
-
-  file_out <- file(paste0(di,'out/anovas_competition/aov_', i, '.txt'))
+  file_out <- file(paste0(di,'out/anovas_competition/aov_', i, '.txt'), "w")
   sink(file_out)
@@
 
-  sink()
-  close(file_out)
+  while (sink.number()>0) sink()
+  # close(file_out)
 }
+
+while (sink.number()>0) sink()
 ```
@@
 
-  file_out <- file(paste0(di,'out/anovas_topo/aov_', i, '.txt'))
+  file_out <- file(paste0(di,'out/anovas_topo/aov_', i, '.txt'), "w")
   sink(file_out)
+
   cat("MODEL \n")
@@
 
-  sink()
-  close(file_out)
+  while (sink.number()>0) sink()
+  # close(file_out)
+
 }
+
+while (sink.number()>0) sink()
 ```

### 2017-05-15 analysis/analysis_stand_features.Rmd
update focal tree

@@
   filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno')
+  filter(id_focal!='Fresno') %>% 
+  mutate(site = as.factor(site))
 
@@
 ```
+
+# Focal tree summary
+
+```{r}
+## Comparison 
+# Select only variables to compare 
+ft_sel <- ft %>%
+  mutate(dn = dn_mm / 1000, 
+         h = height_cm / 100) %>% 
+  dplyr::select(dn, h, site)
+
+# Get vector with variables 
+variables <- c('dn','h')
+
+for (i in variables){
+
+  # apply comparison
+  out_compara <- compara(df=ft_sel, mivariable = i)
+
+  out_name  <- paste0('aov_', i)
+  assign(out_name, out_compara)
+
+}
+
+# Loop to export into txt files (see ./out/anovas_ft ... )
+for (i in variables){
+
+  out <- get(paste0('aov_', i))
+
+  file_out <- file(paste0(di,'out/anovas_ft/aov_', i, '.txt'), "w")
+  sink(file_out)
+
+  cat("MODEL \n")
+  print(out$mymodel) 
+  cat("\n")
+
+  cat("MODEL pretty \n")
+  print(out$tm) 
+  cat("\n")
+
+  cat("POST HOC \n")
+  print(out$mymult) 
+  cat("\n")
+
+  cat("SUMMARY VALUES \n")
+  print(as.data.frame(out$summ_comparison))
+  cat("\n")
+
+  while (sink.number()>0) sink()
+  # close(file_out)
+
+}
+
+while (sink.number()>0) sink()
+```
+
+### dn Focal tree
+
+```{r, echo=FALSE}
+mivariable <- 'dn'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```
+
+### height Focal tree
+
+```{r, echo=FALSE}
+mivariable <- 'h'
+s <- get(paste0('aov_', mivariable))
+
+s$summ_comparison %>% 
+  pander(round=4, caption=paste0('Mean values (', mivariable,')'))
+
+tm <- s$tm
+
+pander(tm, round=5,
+       caption=paste0('ANOVA table (', mivariable,')'), 
+       missing = '', emphasize.strong.cells = 
+               which(tm < 0.1 & tm == tm$p.value, arr.ind = T))
+```

### 2017-05-15 analysis/analysis_stand_features.Rmd
update pdf

@@
 date: "2017 May"
-output:  
-  md_document:
-    toc: true
-    variant: markdown_github
+output: pdf_document
 ---
+
+```{r global_options, include=FALSE}
+knitr::opts_chunk$set(warning=FALSE, message=FALSE)
+```
 

### 2017-05-16 analysis/analysis_stand_features.Rmd
update md
change format table

--- 2017-05-15 analysis/analysis_stand_features.Rmd
@@
 ---
+
+```{r global_options, include=FALSE}
+knitr::opts_chunk$set(warning=FALSE, message=FALSE)
+```
 
@@
 
-  summ_comparison <- summ_values %>% left_join(df_letter, by='site')
+  summ_comparison <- summ_values %>% left_join(df_letter, by='site') %>% as.data.frame()
 
@@
 ```
+
+```{r}
+# Get name of variables 
+variables <- compe %>% dplyr::select(ba:dnn) %>% names() %>% rbind()
+# Add topo and focal tree variables 
+nombres <- c(variables, 'mde','slope', 'dn', 'h')
+
+out <- c()
+
+for (i in nombres){ 
+  # name of aov object
+  aov_name <- paste0('aov_',i)
+  # get aov_object 
+  aux <- get(aov_name)
+  # get summ comparison dataframe 
+  aux2 <- aux$summ_comparison
+  out <- rbind(out, aux2)
+}
+
+out2 <- out %>% dplyr::select(site, mean, sd, tukey, mivar = variable)
+
+out2
+
+# See this solution 
+# http://stackoverflow.com/questions/39066811/long-to-wide-with-dplyr
+out3 <- gather(out2, variable, value, mean, sd, tukey) %>% 
+  unite(var, variable, site) %>% 
+  spread(var, value)
+
+out3 <- out3 %>% dplyr::select(mivar, mean_CA_High, sd_CA_High, tukey_CA_High,
+                 mean_CA_Low, sd_CA_Low, tukey_CA_Low,
+                 mean_SJ_High, sd_SJ_High, tukey_SJ_High,
+                 mean_SJ_Low, sd_SJ_Low, tukey_SJ_Low)
+
+```

### 2017-05-16 analysis/analysis_stand_features.Rmd
update analysis; +create csv with tables

--- 2017-05-15 analysis/analysis_stand_features.Rmd
@@
 ```
+
+```{r}
+# Join all data and export as table
+
+# Get name of variables 
+variables <- compe %>% dplyr::select(ba:dnn) %>% names() %>% rbind()
+# Add topo and focal tree variables 
+nombres <- c(variables, 'mde','slope', 'dn', 'h')
+
+out <- c()
+
+for (i in nombres){ 
+  # name of aov object
+  aov_name <- paste0('aov_',i)
+  # get aov_object 
+  aux <- get(aov_name)
+  # get summ comparison dataframe 
+  aux2 <- aux$summ_comparison
+  out <- rbind(out, aux2)
+}
+
+out2 <- out %>% dplyr::select(site, mean, sd, tukey, mivar = variable)
+
+# See this solution 
+# http://stackoverflow.com/questions/39066811/long-to-wide-with-dplyr
+# http://garrettgman.github.io/tidying/
+out3 <- gather(out2, variable, value, mean, sd, tukey) %>% 
+  unite(var, variable, site) %>% 
+  spread(var, value, convert=TRUE)
+
+out4 <- out3 %>% 
+  transmute(variables = mivar,
+            CA_High_m = round(mean_CA_High, 3),
+            CA_High_sd = round(sd_CA_High, 3),
+            CA_High_l = tukey_CA_High,
+            CA_Low_m = round(mean_CA_Low, 3),
+            CA_Low_sd = round(sd_CA_Low, 3),
+            CA_Low_l = tukey_CA_Low,
+            SJ_High_m = round(mean_SJ_High, 3),
+            SJ_High_sd = round(sd_SJ_High, 3),
+            SJ_High_l = tukey_SJ_High,
+            SJ_Low_m = round(mean_SJ_Low, 3),
+            SJ_Low_sd = round(sd_SJ_Low, 3),
+            SJ_Low_l = tukey_SJ_Low)
+
+# Export data 
+write.csv(out4, file=paste(di, "data/proto_tables/site_features.csv", sep=""), row.names = FALSE)
+
+## General Variables and coordinates
+
+general_var %>% 
+  left_join(coords_sites, by='site') %>% 
+  write.csv(file=paste(di, "data/proto_tables/site_general.csv", sep=""), row.names = FALSE)
+```

### 2017-05-16 analysis/test-bai.Rmd
update

@@
 library("knitr")
+library("detrendeR")
 source(paste0(getwd(), '/script/R/bai_piovesan.R'))
@@
   filter(id_focal!='Fresno') %>% 
-  select(id_focal, loc, dn_mm, height_cm)
+  dplyr::select(id_focal, loc, dn_mm, height_cm)
 
@@
 ft <- ft %>% 
-  mutate(elev = ifelse(id_focal %in% sj_lowcode, 'Low',
+  mutate(elev = as.factor(
+    ifelse(id_focal %in% sj_lowcode, 'Low',
                        ifelse(id_focal %in% sj_highcode, 'High',
-                              ifelse(id_focal %in% ca_lowcode, 'Low', 'High')))) %>%
-  mutate(site = paste0(loc, '_', elev))
+                              ifelse(id_focal %in% ca_lowcode, 'Low', 'High'))))) %>%
+  mutate(site = as.factor(paste0(loc, '_', elev)))
+```
+
+### General statistics by site
+
+```{r}
+# Create subset to compare between sites 
+caL <- ca[,c("SNB0101","SNB0102","SNB0201","SNB0202","SNB0301","SNB0302","SNB0401","SNB0402","SNB0501","SNB0502g",
+             "SNB0601","SNB0602","SNB0701","SNB0702g","SNB0801g","SNB0802g","SNB0901g","SNB0902","SNB1001","SNB1002",
+             "SNB2601","SNB2602","SNB2701","SNB2702g","SNB2801","SNB2802","SNB2901","SNB2902","SNB3001","SNB3002")]
+
+caH <- ca[, c("SNB1101","SNB1102","SNB1201","SNB1202","SNB1301","SNB1302","SNB1401","SNB1402","SNB1501","SNB1502",
+              "SNB1601","SNB1602","SNB1701","SNB1702","SNB1801","SNB1802","SNB1901","SNB1902","SNB2001","SNB2002g",
+              "SNB2101","SNB2102","SNB2201g","SNB2202g","SNB2301g","SNB2302","SNB2401","SNB2402","SNB2501","SNB2502")]
+
+sjL <- sj[, c("SNA0101","SNA0102","SNA0201","SNA0202","SNA0301","SNA0302","SNA0401","SNA0402","SNA0501","SNA0502",
+              "SNA0601","SNA0602","SNA0603","SNA0701","SNA0702","SNA0801","SNA0802","SNA0901","SNA0902","SNA0903",
+              "SNA1001","SNA1002","SNA1003")]
+sjH <- sj[, c("SNA1101","SNA1102","SNA1103","SNA1201","SNA1202","SNA1203","SNA1301","SNA1302","SNA1303","SNA1401",
+              "SNA1402","SNA1501","SNA1502","SNA1601","SNA1602","SNA1701","SNA1702","SNA1703","SNA1801","SNA1802",
+              "SNA1901","SNA1902","SNA2001","SNA2003","SNA2002")]
+```
+
+### Dendrochronological summary 
+```{r}
+objects_rwl <- c('caL','caH','sjL','sjH')
+
+out <- c()
+
+for (i in objects_rwl){
+
+  aux <- get(i)
+  aux_stats <- rwi.stats(aux)
+
+  aux_stats$site <- i
+
+  out <- rbind(out, aux_stats)
+  }
+
+d <- rwl.report(ca)
+
 ```

### 2017-05-16 analysis/test-bai.Rmd
update analysis; update computo de BAI;

@@
 library("knitr")
-library("detrendeR")
+# library("detrendeR")
+library("pander")
 source(paste0(getwd(), '/script/R/bai_piovesan.R'))
@@
 
-# Read data
+# Read y Prepare data 
+* Leer datos `rwl` de SJ y CA
+* Leer datos de diametros de los focal tree
 
 ```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
+# machine <- 'ajpelu'
+machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 
-## Aggregate RW by tree
+* Lectura y preparación de datos de diámetro
 
-* Lectura y preparación de datos de diámetro
 ```{r} 
@@
 
-### General statistics by site
+* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
 
@@
 
-### Dendrochronological summary 
+# Summary dendrochronology statistics
+
+* Por cada loc (ca/sj) o site (ca_high, ca_low, sj_high, sj_low) calculamos los siguientes datos: 
+    * age: mean + sd (average of all series)
+    * year: min, max of all series
+    * year range
+    * autocorrelation order 1 (ar1): mean + sd 
+    * seires intercorrelation (icor): mean + sd
+
+## By site 
 ```{r}
@@
   aux <- get(i)
-  aux_stats <- rwi.stats(aux)
+  # Stats of singles RWL
+  aux_stats <- rwl.stats(aux)
+  # Add site 
+  aux_stats$site <- as.factor(i)
 
-  aux_stats$site <- i
+  # Compute interseries correlation 
+  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
+  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
 
@@
 
-d <- rwl.report(ca)
+rwl_by_site <- out %>% group_by(site) %>% 
+  summarize(age_mean = round(mean(year),2), 
+            age_sd = round(sd(year), 2),
+            year_min = min(first),
+            year_max = max(last), 
+            year_range = year_max - year_min + 1, 
+            ar1_mean = round(mean(ar1), 3),
+            ar1_sd = round(sd(ar1), 3),
+            icor_mean = round(mean(interser_cor_val), 3),
+            icor_sd = round(sd(interser_cor_val), 3)) %>% 
+  as.data.frame()
 
+rwl_by_site %>% 
+  write.csv(file=paste(di, "data/proto_tables/site_dendro.csv", sep=""), row.names = FALSE)
+
+pander(rwl_by_site, caption='Dendrochronological summary by sites') 
 ```
 
-* Agregar valores medios de RW 
-:red_circle: DOCUMENTAR 
+## By Loc
+```{r}
+objects_rwl <- c('ca','sj')
+
+out <- c()
+
+for (i in objects_rwl){
+
+  aux <- get(i)
+  # Stats of singles RWL
+  aux_stats <- rwl.stats(aux)
+  # Add site 
+  aux_stats$loc <- as.factor(i)
+
+  # Compute interseries correlation 
+  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
+  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
+
+  out <- rbind(out, aux_stats)
+  }
+
+rwl_by_loc <- out %>% group_by(loc) %>% 
+  summarize(age_mean = round(mean(year),2), 
+            age_sd = round(sd(year), 2),
+            year_min = min(first),
+            year_max = max(last), 
+            year_range = year_max - year_min + 1, 
+            ar1_mean = round(mean(ar1), 3),
+            ar1_sd = round(sd(ar1), 3),
+            icor_mean = round(mean(interser_cor_val), 3),
+            icor_sd = round(sd(interser_cor_val), 3)) %>% 
+  as.data.frame()
+
+rwl_by_loc %>% 
+  write.csv(file=paste(di, "data/proto_tables/loc_dendro.csv", sep=""), row.names = FALSE)
+
+pander(rwl_by_loc, caption='Dendrochronological summary by sites') 
+```
+
+# Aggregate RW by tree 
+* Agregar valores medios de RW por loc (obtenemos sj_tree / ca_tree) o por site (sjL_tree, sjH_tree / caL_tree, caH_tree) 
+* ver fun
+
 ```{r}
@@
 ca_tree <- rw_byTree(ca)
+
+sjL_tree <- rw_byTree(sjL)
+sjH_tree <- rw_byTree(sjH)
+caL_tree <- rw_byTree(caL)
+caH_tree <- rw_byTree(caH)
 ```
@@
 ```{r}
-# function to subset averaged RW 
-subsetRW <- function(rwdf, codes){ 
-  require(dplyr)
-  aux_codes <- paste0('rw_', codes)
-  mysubset <- rwdf %>% select(one_of(aux_codes), year)
-  mysubset
-  }
-
-sjL <- subsetRW(sj_tree, sj_lowcode)
-sjH <- subsetRW(sj_tree, sj_highcode)
-caL <- subsetRW(ca_tree, ca_lowcode)
-caH <- subsetRW(ca_tree, ca_highcode)
+# # function to subset averaged RW 
+# subsetRW <- function(rwdf, codes){ 
+#   require(dplyr)
+#   aux_codes <- paste0('rw_', codes)
+#   mysubset <- rwdf %>% dplyr::select(one_of(aux_codes), year)
+#   mysubset
+#   }
 
@@
              select(data) %>% 
-             lapply(function(x) {(x)}) %>% .[[1]]
+             lapply(function(x) {(x)}) %>% .[1]
+
+%>% .[[2]]
 
 ```
+
+---> VAS por AQUI
 
@@
 
-rev(s)
-
-# Extract rwl serie 
-myrwl_i <- myrwl[[1]]
-
-diam_df_i <- diam_cores_sj[1]
-
-# Remove NA 
-myrwl_io <- na.omit(myrwl_i)
-
-bai <- 
-diam.vec <- diam[, 2]
-
 ########## test Piovesan

### 2017-05-17 analysis/test-bai.Rmd
update

@@
 
+```{r}
+
+rownames(sj_tree) <- sj_tree$year
+sj_tree <- sj_tree[,-1]
+carwi <- detrend(rwl = ca_tree, method = "ModNegExp")
+ca_prwi <- data.frame(rw = rowMeans(carwi, na.rm=TRUE))
+ca_prwi <- ca_prwi %>% mutate(year = rownames(.),
+                        site = as.factor('ca'),
+                        rw_log = log(rw))
+
+xx <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
+
+plot(xx$rw~rownames(xx))
+
+sjcrn <- chron(sjrwi, prefix = "rw_")
+plot(sjcrn, add.spline=TRUE, nyrs=20)
+plot()
+
+sj_tree <- rw_byTree(sj)
+ca_tree <- rw_byTree(ca)
+
+rownames(ca_tree) <- ca_tree$year
+ca_tree <- ca_tree[,-1]
+ca_p <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
+
+ca_p <- ca_p %>% mutate(year = rownames(.),
+                        site = as.factor('ca'),
+                        rw_log = log(rw))
+
+rownames(sj_tree) <- sj_tree$year
+sj_tree <- sj_tree[,-1]
+sj_p <- data.frame(rw = rowMeans(sj_tree, na.rm=TRUE))
+
+sj_p <- sj_p %>% mutate(year = rownames(.),
+                        site = as.factor('sj'),
+                        rw_log = log(rw))
+
+xx <- rbind(ca_p, sj_p)
+
+xx %>% ggplot(aes(x=as.Date(year, format='%Y'), y=rw_log)) + 
+  geom_line(aes(group=1)) + 
+  scale_x_date(date_labels = "%Y") + 
+  facet_wrap(~site, ncol=1, scales = 'free_y') + 
+  theme_bw() + 
+  geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
+  geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
+  xlab('Year') + ylab('Ring width (1/100 mm)')
+
+xx %>% filter(site =='ca') %>% 
+  ggplot(aes(x=as.Date(year, format='%Y'), y=rw)) + geom_point(colour='blue') +
+  geom_line(aes(group=1), colour='blue') + 
+  scale_x_date(date_labels = "%Y") + 
+  theme_bw() + 
+  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
+  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
+  xlab('Year') + ylab('Ring width (1/100 mm)')
+
+ca_prwi %>%
+  ggplot(aes(x = as.Date(year, format = '%Y'), y = rw)) + geom_point() +
+  geom_line(aes(group = 1)) +
+  scale_x_date(date_labels = "%Y") +
+  theme_bw() +
+  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
+  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
+  xlab('Year') + ylab('Ring width Index')
+
 ## San Juan Data

### 2017-05-19 script/R/rw_byTree.R
update function: * add argument snc: structure of name core * change the way getting treenames using snc * change the way the mean by tree is named * covert year into rownames

@@ -6,8 +6,14 @@
 # Tree names coming from colnames of the dataframe (see from class::rwl). The colnames of this rwl have the 
 # following structure: SNA0101 (SNA|id_tree|id_core). 
  
-rw_byTree <- function(rwdf){ 
+rw_byTree <- function(rwdf, snc){ 
   require(dplyr)
+  
+  # snc: structure of name core: vector with structure of name of the core, example SJ0102; 
+  # i.e.: snc=c(2,2,2)
+  s_site <- snc[1]
+  s_tree <- snc[2]
+  s_core <- snc[3]
   
   # Create output with rownmaes (year) as a variable 
   output_avg <- rwdf %>% 
@@ -15,12 +21,12 @@ rw_byTree <- function(rwdf){
     dplyr::select(year)
   
   # Get name of each tree (each serie: i.e.:  SNA0101 (SNA|id_tree|id_core)) 
-  tree_names <- unique(stringr::str_sub(names(rwdf), start=1, end=5))
+  tree_names <- unique(stringr::str_sub(names(rwdf), start=1, end=s_site+s_tree))
   
   # Loop 
   for (i in tree_names) {
     # Create a variable name for the average by tree 
-    name_avg <- paste0('rw_',stringr::str_sub(i, star=3, end=5))
+    name_avg <- paste0('mT_',stringr::str_sub(i, star=s_site+1, end=s_site+s_tree))
     
     # Create aux dataframe with rwl mean values and the year
     aux <- rwdf %>% 
@@ -37,5 +43,11 @@ rw_byTree <- function(rwdf){
   # Change NaN by NA
   output_avg[is.na(output_avg)] <- NA
   
+  # Convert year to rownames 
+  row.names(output_avg) <- output_avg$year
+  output_avg <- subset(output_avg, select=-year)
+  
   output_avg
+  
+  
 }

### 2017-05-19 script/R/rw_byTree.R
add name of loc

@@ -6,8 +6,9 @@
 # Tree names coming from colnames of the dataframe (see from class::rwl). The colnames of this rwl have the 
 # following structure: SNA0101 (SNA|id_tree|id_core). 
  
-rw_byTree <- function(rwdf, snc){ 
+rw_byTree <- function(rwdf, snc, loc){ 
   require(dplyr)
+  # loc is used for the name of the trees 
   
   # snc: structure of name core: vector with structure of name of the core, example SJ0102; 
   # i.e.: snc=c(2,2,2)
@@ -26,7 +27,7 @@ rw_byTree <- function(rwdf, snc){
   # Loop 
   for (i in tree_names) {
     # Create a variable name for the average by tree 
-    name_avg <- paste0('mT_',stringr::str_sub(i, star=s_site+1, end=s_site+s_tree))
+    name_avg <- paste0(loc, stringr::str_sub(i, star=s_site+1, end=s_site+s_tree))
     
     # Create aux dataframe with rwl mean values and the year
     aux <- rwdf %>% 

### 2017-05-19 analysis/test-bai.Rmd
update

@@
 library("knitr")
-# library("detrendeR")
+library("detrendeR")
 library("pander")
@@
 
+* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
+    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
+
+```{r}
+# Replace SNA by SJ and SNB by CA
+names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
+names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")
+
+# Remove g in name of some cores of CA. 
+names(ca) <- stringr::str_replace(names(ca), "g", "")
+```
+
+```{r}
+# Create subset to compare between sites 
+caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
+             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
+             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
+caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
+              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
+              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
+
+sjL <- sj[, c("SJ0101","SJ0102","SJ0201","SJ0202","SJ0301","SJ0302","SJ0401","SJ0402","SJ0501","SJ0502",
+              "SJ0601","SJ0602","SJ0603","SJ0701","SJ0702","SJ0801","SJ0802","SJ0901","SJ0902","SJ0903",
+              "SJ1001","SJ1002","SJ1003")]
+sjH <- sj[, c("SJ1101","SJ1102","SJ1103","SJ1201","SJ1202","SJ1203","SJ1301","SJ1302","SJ1303","SJ1401",
+              "SJ1402","SJ1501","SJ1502","SJ1601","SJ1602","SJ1701","SJ1702","SJ1703","SJ1801","SJ1802",
+              "SJ1901","SJ1902","SJ2001","SJ2003","SJ2002")]
+```
+
 * Lectura y preparación de datos de diámetro
@@
 
+# Change name focal according to loc
+compete <- compete %>% 
+  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))
+
 # Get only focal trees, and only selected variables 
@@
   filter(id_focal!='Fresno') %>% 
-  dplyr::select(id_focal, loc, dn_mm, height_cm)
+  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm)
 
 # Set levels of eleveation 
-sj_lowcode  <- paste0('A', str_pad(1:10, 2, pad='0'))
-sj_highcode <- paste0('A', 11:20)
-ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')),
-            paste0('B', 26:30))
-ca_highcode <- paste0('B', 11:25)
+sj_lowcode  <- paste0('SJ', str_pad(1:10, 2, pad='0'))
+sj_highcode <- paste0('SJ', 11:20)
+ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
+            paste0('CA', 26:30))
+ca_highcode <- paste0('CA', 11:25)
 
@@
   mutate(elev = as.factor(
-    ifelse(id_focal %in% sj_lowcode, 'Low',
-                       ifelse(id_focal %in% sj_highcode, 'High',
-                              ifelse(id_focal %in% ca_lowcode, 'Low', 'High'))))) %>%
+    ifelse(id_focalLoc %in% sj_lowcode, 'Low',
+                       ifelse(id_focalLoc %in% sj_highcode, 'High',
+                              ifelse(id_focalLoc %in% ca_lowcode, 'Low', 'High'))))) %>%
   mutate(site = as.factor(paste0(loc, '_', elev)))
@@
 * Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
-
-```{r}
-# Create subset to compare between sites 
-caL <- ca[,c("SNB0101","SNB0102","SNB0201","SNB0202","SNB0301","SNB0302","SNB0401","SNB0402","SNB0501","SNB0502g",
-             "SNB0601","SNB0602","SNB0701","SNB0702g","SNB0801g","SNB0802g","SNB0901g","SNB0902","SNB1001","SNB1002",
-             "SNB2601","SNB2602","SNB2701","SNB2702g","SNB2801","SNB2802","SNB2901","SNB2902","SNB3001","SNB3002")]
-
-caH <- ca[, c("SNB1101","SNB1102","SNB1201","SNB1202","SNB1301","SNB1302","SNB1401","SNB1402","SNB1501","SNB1502",
-              "SNB1601","SNB1602","SNB1701","SNB1702","SNB1801","SNB1802","SNB1901","SNB1902","SNB2001","SNB2002g",
-              "SNB2101","SNB2102","SNB2201g","SNB2202g","SNB2301g","SNB2302","SNB2401","SNB2402","SNB2501","SNB2502")]
-
-sjL <- sj[, c("SNA0101","SNA0102","SNA0201","SNA0202","SNA0301","SNA0302","SNA0401","SNA0402","SNA0501","SNA0502",
-              "SNA0601","SNA0602","SNA0603","SNA0701","SNA0702","SNA0801","SNA0802","SNA0901","SNA0902","SNA0903",
-              "SNA1001","SNA1002","SNA1003")]
-sjH <- sj[, c("SNA1101","SNA1102","SNA1103","SNA1201","SNA1202","SNA1203","SNA1301","SNA1302","SNA1303","SNA1401",
-              "SNA1402","SNA1501","SNA1502","SNA1601","SNA1602","SNA1701","SNA1702","SNA1703","SNA1801","SNA1802",
-              "SNA1901","SNA1902","SNA2001","SNA2003","SNA2002")]
-```
+    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
 
@@
 # Aggregate RW by tree
+
 * Agregar valores medios de RW por loc (obtenemos sj_tree / ca_tree) o por site (sjL_tree, sjH_tree / caL_tree, caH_tree) 
-* ver fun
+* ver fun rw_byTree o utilizar treeMean (dplR)
 
 ```{r}
-sj_tree <- rw_byTree(sj)
-ca_tree <- rw_byTree(ca)
+# Remember snc = structure of core name SJ0101 (site | tree | core)
+sj_tree <- rw_byTree(sj, snc =c(2,2,2))
+ca_tree <- rw_byTree(ca, snc =c(2,2,2))
 
-sjL_tree <- rw_byTree(sjL)
-sjH_tree <- rw_byTree(sjH)
-caL_tree <- rw_byTree(caL)
-caH_tree <- rw_byTree(caH)
+sjL_tree <- rw_byTree(sjL, snc =c(2,2,2))
+sjH_tree <- rw_byTree(sjH, snc =c(2,2,2))
+caL_tree <- rw_byTree(caL, snc =c(2,2,2))
+caH_tree <- rw_byTree(caH, snc =c(2,2,2))
+
+# Si decido usar treeMean debería ser (solo un ejemplo)
+# ids_sj <- read.ids(sj, stc = c(2, 2, 2))
+# sj_meantree <- treeMean(sj, ids = ids_sj,  na.rm=TRUE)
 ```

### 2017-05-19 script/R/rw_byTree.R

@@ -6,9 +6,9 @@
 # Tree names coming from colnames of the dataframe (see from class::rwl). The colnames of this rwl have the 
 # following structure: SNA0101 (SNA|id_tree|id_core). 
  
-rw_byTree <- function(rwdf, snc, loc){ 
+rw_byTree <- function(rwdf, snc){ 
   require(dplyr)
-  # loc is used for the name of the trees 
+  # locname is used for the name of the trees 
   
   # snc: structure of name core: vector with structure of name of the core, example SJ0102; 
   # i.e.: snc=c(2,2,2)
@@ -27,7 +27,7 @@ rw_byTree <- function(rwdf, snc, loc){
   # Loop 
   for (i in tree_names) {
     # Create a variable name for the average by tree 
-    name_avg <- paste0(loc, stringr::str_sub(i, star=s_site+1, end=s_site+s_tree))
+    name_avg <- paste0('SJ', stringr::str_sub(i, star=s_site+1, end=s_site+s_tree))
     
     # Create aux dataframe with rwl mean values and the year
     aux <- rwdf %>% 

### 2017-05-19 script/R/rw_byTree.R
add locname at the end of the function

@@ -6,10 +6,10 @@
 # Tree names coming from colnames of the dataframe (see from class::rwl). The colnames of this rwl have the 
 # following structure: SNA0101 (SNA|id_tree|id_core). 
  
-rw_byTree <- function(rwdf, snc){ 
+rw_byTree <- function(rwdf, snc, locname){ 
   require(dplyr)
   # locname is used for the name of the trees 
-  
+
   # snc: structure of name core: vector with structure of name of the core, example SJ0102; 
   # i.e.: snc=c(2,2,2)
   s_site <- snc[1]
@@ -27,7 +27,7 @@ rw_byTree <- function(rwdf, snc){
   # Loop 
   for (i in tree_names) {
     # Create a variable name for the average by tree 
-    name_avg <- paste0('SJ', stringr::str_sub(i, star=s_site+1, end=s_site+s_tree))
+    name_avg <- paste0(stringr::str_sub(i, star=s_site+1, end=s_site+s_tree))
     
     # Create aux dataframe with rwl mean values and the year
     aux <- rwdf %>% 
@@ -46,8 +46,11 @@ rw_byTree <- function(rwdf, snc){
   
   # Convert year to rownames 
   row.names(output_avg) <- output_avg$year
+  
   output_avg <- subset(output_avg, select=-year)
   
+  # Add locname 
+  names(output_avg) <- paste0(locname, names(output_avg))
   output_avg
   
   

### 2017-05-19 man/notes_dendro_analysis.Rmd
update analysis notes
update analysis; update computo de BAI;
update notes analysis

@@
 * Cañar: [`./data_raw/dendro_ring/sn_canar/sn_canar_cofecha.OUT`](/data_raw/dendro_ring/sn_canar/sn_canar_cofecha.OUT)
-
-* [ ] :red_circle: Tabla con estadísticos (ver otros ejemplos): Ha de te
 
@@
 * Calcular el BAI por cada árbol haciendo la media de crecimiento (RW) a nivel de árbol.
+Para el cálculo de la media de crecimiento por árbol utilizamos la siguiente función [./script/R/rw_byTree.R](/script/R/rw_byTree.R)) creada específicamente para ello.
+También podríamos hacerlo con la función `treeMean`
 
@@
 * :red_circle: Dudas funcion BAItree Guillermo.
+
+## Chronologias
+Para cada sitio y/o loc construimos una cronología utilizando las series de BAI. Podemos utilizar varias aproximaciones: 
+ * media aritmética 
+ * Tukey’s Biweight Robust Mean (ver pág 123 @Cook1990)
+ * :red_circle: ¿cual usar de las dos? ¿Ask to Guillermo?
 

### 2017-05-19 script/R/bai_piovesan.R
change way to compute bai

@@ -6,7 +6,9 @@
 
 bai_piovesan <- function(rwdf, diam_df){
   # rwdf: a ringwidth dataframe. row: years; col: id_cores
-  # diam_df: a dataframe with diameter and id_core (two columns) 
+  # diam_df: a dataframe with two columns: 
+  #   * diameter = diameter 
+  #   * id = id_core or id_tree or whatever 
   
   # Output 
   bai <- rwdf 
@@ -25,8 +27,8 @@ bai_piovesan <- function(rwdf, diam_df){
     rws <- rwdf[[id_core]]
     
     # 3 # Select diameter 
-    dbh <- subset(diam_df, id_cores == id_core)
-    dbh <- dbh[,2]
+    dbh <- subset(diam_df, id == id_core)
+    dbh <- dbh$diameter
     
     # 4 # Prepare data to compute bai 
     #### Omit NA 
@@ -56,5 +58,6 @@ bai_piovesan <- function(rwdf, diam_df){
     ### Put bai in those element with no NA
     bai[no_na, i] <- b 
   }
+
   bai 
 }

### 2017-05-19 analysis/test-bai.Rmd
add new analysis bai

@@
 # Remember snc = structure of core name SJ0101 (site | tree | core)
-sj_tree <- rw_byTree(sj, snc =c(2,2,2))
-ca_tree <- rw_byTree(ca, snc =c(2,2,2))
+sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
+ca_tree <- rw_byTree(ca, snc =c(2,2,2), locname = 'CA')
 
-sjL_tree <- rw_byTree(sjL, snc =c(2,2,2))
-sjH_tree <- rw_byTree(sjH, snc =c(2,2,2))
-caL_tree <- rw_byTree(caL, snc =c(2,2,2))
-caH_tree <- rw_byTree(caH, snc =c(2,2,2))
+sjL_tree <- rw_byTree(sjL, snc =c(2,2,2), locname = 'SJ')
+sjH_tree <- rw_byTree(sjH, snc =c(2,2,2), locname = 'SJ')
+caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
+caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')
 
@@
 
-* Crear diferentes dataset por sitio 
+* Crear diferentes dataset de diametro por sitio 
 ```{r}
@@
 
-ft %>% group_by(site) %>% do(data = (.)) %>% 
-             select(data) %>% 
-             lapply(function(x) {(x)}) %>% .[1]
+# diam <- ft %>% group_by(site) %>% do(data = (.)) %>% 
+#              dplyr::select(data) %>% 
+#              lapply(function(x) {(x)})
 
-%>% .[[2]]
+diam <- ft %>%
+  mutate(diameter = dn_mm, 
+         id = id_focalLoc) %>%
+  dplyr::select(id, diameter, site) %>% 
+  split(.$site)
+
+d_caH <- diam$CA_High[,c('id','diameter')]
+d_caL <- diam$CA_Low[,c('id','diameter')]
+d_sjH <- diam$SJ_High[,c('id','diameter')]
+d_sjL <- diam$SJ_Low[,c('id','diameter')]
+
+diamloc <- ft %>% 
+  mutate(diameter = dn_mm, 
+         id = id_focalLoc) %>%
+  dplyr::select(id, diameter, loc) %>% split(.$loc)
+
+d_ca <- diamloc$CA[,c('id','diameter')]
+d_sj <- diamloc$SJ[,c('id','diameter')]
+```
+
+# Cómputo del BAI por loc y/o site
+
+* He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]. Es similar a `bai.out`
+
+```{r}
+bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
+bai_ca <- bai_piovesan(rwdf = ca_tree, diam_df = d_ca)
+
+bai_sjH <- bai_piovesan(rwdf = sjH_tree, diam_df = d_sjH)
+bai_sjL <- bai_piovesan(rwdf = sjL_tree, diam_df = d_sjL)
+bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
+bai_caL <- bai_piovesan(rwdf = caL_tree, diam_df = d_caL)
+
+# bai_test <- round(bai.out(rwl = sj_tree, diam = d_sj), 4)
+# test bai.out and bai_piovesan  
+# unique(round(bai_sj$SJ03,4) == round(bai_test$SJ03,4))
+
+# Set class to bai object 
+# Esto es para que funcionen algunas otras funciones de dplR 
+bais <- c('bai_sj', 'bai_ca', 'bai_sjH', 'bai_sjL', 'bai_caH', 'bai_caL')
+
+for (i in bais){
+  aux <- get(i)
+
+  class(aux) <- c('rwl', 'data.frame')
+
+  assign(i, aux)
+}
+
+```
+
+# Cronologías by site / loc
+
+```{r}
+
+w <- summary(bai_sj)
+mean(w$year)
+mean(w$stdev)
+
+plot(bai_sj, plot.type="spag")
+
+cro_sjb <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
+ plot(cro_sjb, add.spline=TRUE, nyrs=30)
+
+cro_sj <- chron(bai_sj, prefix = "SJ", biweight = FALSE)
+
+cc <- cbind(cro_sj, cro_sjb)
+
+su
 
@@
 
-```{r}
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-carwi <- detrend(rwl = ca_tree, method = "ModNegExp")
-ca_prwi <- data.frame(rw = rowMeans(carwi, na.rm=TRUE))
-ca_prwi <- ca_prwi %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-xx <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-plot(xx$rw~rownames(xx))
-
-sjcrn <- chron(sjrwi, prefix = "rw_")
-plot(sjcrn, add.spline=TRUE, nyrs=20)
-plot()
-
-sj_tree <- rw_byTree(sj)
-ca_tree <- rw_byTree(ca)
-
-rownames(ca_tree) <- ca_tree$year
-ca_tree <- ca_tree[,-1]
-ca_p <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-ca_p <- ca_p %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-sj_p <- data.frame(rw = rowMeans(sj_tree, na.rm=TRUE))
-
-sj_p <- sj_p %>% mutate(year = rownames(.),
-                        site = as.factor('sj'),
-                        rw_log = log(rw))
-
-xx <- rbind(ca_p, sj_p)
-
-xx %>% ggplot(aes(x=as.Date(year, format='%Y'), y=rw_log)) + 
-  geom_line(aes(group=1)) + 
-  scale_x_date(date_labels = "%Y") + 
-  facet_wrap(~site, ncol=1, scales = 'free_y') + 
-  theme_bw() + 
-  geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-xx %>% filter(site =='ca') %>% 
-  ggplot(aes(x=as.Date(year, format='%Y'), y=rw)) + geom_point(colour='blue') +
-  geom_line(aes(group=1), colour='blue') + 
-  scale_x_date(date_labels = "%Y") + 
-  theme_bw() + 
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-ca_prwi %>%
-  ggplot(aes(x = as.Date(year, format = '%Y'), y = rw)) + geom_point() +
-  geom_line(aes(group = 1)) +
-  scale_x_date(date_labels = "%Y") +
-  theme_bw() +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width Index')
+```
 
@@
 diam_cores_sj <- diam_cores %>% filter(! id_cores %in% id_cores_no_bark)
-```
-
-## Cómputo del BAI
-
-He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]
-
-```{r} 
-# Compute BAI for SJ locality 
-baisj <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
 ```
@@
 ```
+
+```{r}
+# tes plots
+
+rownames(sj_tree) <- sj_tree$year
+sj_tree <- sj_tree[,-1]
+carwi <- detrend(rwl = ca_tree, method = "ModNegExp")
+ca_prwi <- data.frame(rw = rowMeans(carwi, na.rm=TRUE))
+ca_prwi <- ca_prwi %>% mutate(year = rownames(.),
+                        site = as.factor('ca'),
+                        rw_log = log(rw))
+
+xx <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
+
+plot(xx$rw~rownames(xx))
+
+sjcrn <- chron(sjrwi, prefix = "rw_")
+plot(sjcrn, add.spline=TRUE, nyrs=20)
+plot()
+
+sj_tree <- rw_byTree(sj)
+ca_tree <- rw_byTree(ca)
+
+rownames(ca_tree) <- ca_tree$year
+ca_tree <- ca_tree[,-1]
+ca_p <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
+
+ca_p <- ca_p %>% mutate(year = rownames(.),
+                        site = as.factor('ca'),
+                        rw_log = log(rw))
+
+rownames(sj_tree) <- sj_tree$year
+sj_tree <- sj_tree[,-1]
+sj_p <- data.frame(rw = rowMeans(sj_tree, na.rm=TRUE))
+
+sj_p <- sj_p %>% mutate(year = rownames(.),
+                        site = as.factor('sj'),
+                        rw_log = log(rw))
+
+xx <- rbind(ca_p, sj_p)
+
+xx %>% ggplot(aes(x=as.Date(year, format='%Y'), y=rw_log)) + 
+  geom_line(aes(group=1)) + 
+  scale_x_date(date_labels = "%Y") + 
+  facet_wrap(~site, ncol=1, scales = 'free_y') + 
+  theme_bw() + 
+  geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
+  geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
+  xlab('Year') + ylab('Ring width (1/100 mm)')
+
+xx %>% filter(site =='ca') %>% 
+  ggplot(aes(x=as.Date(year, format='%Y'), y=rw)) + geom_point(colour='blue') +
+  geom_line(aes(group=1), colour='blue') + 
+  scale_x_date(date_labels = "%Y") + 
+  theme_bw() + 
+  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
+  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
+  xlab('Year') + ylab('Ring width (1/100 mm)')
+
+ca_prwi %>%
+  ggplot(aes(x = as.Date(year, format = '%Y'), y = rw)) + geom_point() +
+  geom_line(aes(group = 1)) +
+  scale_x_date(date_labels = "%Y") +
+  theme_bw() +
+  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
+  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
+  xlab('Year') + ylab('Ring width Index')
+
+```

### 2017-05-25 analysis/test-bai.Rmd
update chron

@@
 library("knitr")
-library("detrendeR")
+# library("detrendeR")
 library("pander")
@@
 ```{r}
+# Robust mean biweith or not
+cro_sjb <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
+cro_sjnotb <- chron(bai_sj, prefix = "SJ", biweight = FALSE)
 
-w <- summary(bai_sj)
-mean(w$year)
-mean(w$stdev)
+cro_b <- cro_sjb %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('biweight')) %>% 
+  select(bai_mean = SJstd, samp.depth, year, treatment)
 
-plot(bai_sj, plot.type="spag")
+cro_notb <- cro_sjnotb %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('biweight')) %>% 
+  select(bai_mean = SJstd, samp.depth, year, treatment)
 
-cro_sjb <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
+# Compute chronologies of BAI
+
+chrono_bai <- function(bai_data, prefix){
+
+    if (length(prefix) == 0) {
+        prefix.str <- ""
+    }
+    else {
+        prefix.str <- as.character(prefix)[1]
+        if (is.na(prefix.str) || Encoding(prefix.str) == "bytes" || 
+            nchar(prefix.str) > 3) {
+            stop("'prefix' must be a character string with less than 4 characters")
+        }
+    }
+    n_trees <- rowSums(!is.na(bai_data))
+
+    out <- bai_data
+
+    out <- data.frame(samps)
+    out
+}
+
+### Compute Chrono
+
+b <- bai_sj
+
+chrono_bai <- function(bai_object, alpha){ 
+  require(dplyr)
+
+  b <- bai_object 
+  alpha <- as.numeric(alpha)
+
+  # Prepare data 
+  b_aux <- b %>% 
+  ##  Create a variable of year 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  ## melt data 
+  gather(series, bai, -year) %>% 
+  # remove NA values 
+  filter(bai != "")
+
+  # Compute mean, sd, se and sample deepth (n_trees)
+  out <- b_aux %>% 
+  group_by(year) %>% 
+  summarise(bai_mean = mean(bai), 
+            bai_sd = sd(bai), 
+            bai_se = bai_sd / sqrt(n()), 
+            n_trees = n()) %>% 
+  as.data.frame()
+
+  # Loess 
+  l <- loess(b_aux$bai ~ b_aux$year, span = alpha)
+  years_v <- seq(min(b_aux$year), max(b_aux$year), 1)
+
+  l_pred_aux <- predict(l, years_v, se=TRUE)
+  l_pred <- as.data.frame(cbind(loess_fit = l_pred_aux$fit, 
+                loess_se = l_pred_aux$se.fit,
+                year = years_v))
+
+  out <- out %>% inner_join(l_pred, by='year')
+
+  out
+}
+
+c_sj <- chrono_bai(bai_sj, alpha=0.75)
+
+c_sj
+
+c_sj %>% ggplot(aes(x=year, y=bai_mean)) + 
+  theme_bw() + 
+  geom_ribbon(aes(x= year, 
+                  ymin = loess_fit - 1.96*loess_se,
+                  ymax = loess_fit + 1.96*loess_se), na.rm = FALSE, 
+              fill = "grey") +
+  geom_line() + geom_point() + 
+  geom_line() + geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+
+bb2 %>% ggplot(aes(x=year, y=bai_mean)) + 
+  theme_bw() + 
+  geom_ribbon(aes(x= year, 
+                  ymin = bai_mean - bai_se,
+                  ymax = bai_mean + bai_se), na.rm = FALSE, 
+              fill = "grey") +
+  geom_line() + geom_point() + 
+  geom_smooth(method= 'loess', span = 0.30, se=FALSE)
+
+# IC LOESS
+
+  geom_smooth(method= 'loess', span = 0.30, se=FALSE) + theme_bw() + 
+  geom_bar(aes(x=year, stat = (samp.depth*10), colour=treatment))
+
+# cc <- chrono_bai(bai_sj, prefix = "SJ")
+
+```
+
+```{r}
+
+# SJ
+
+cro_sj <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
+cro_sj %>% plot(add.spline=TRUE, nyrs=30)
+
+cro_ca <- chron(bai_ca, prefix = "CA", biweight = TRUE)
+cro_ca %>% plot(add.spline=TRUE, nyrs=30)
+
+cro_ca2 <- cro_ca %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('CA')) %>% 
+  select(bai_mean = CAstd, samp.depth, year, treatment)
+
+cro_sj2 <- cro_sj %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('SJ')) %>% 
+  select(bai_mean = SJstd, samp.depth, year, treatment)
+
+cro2 <- rbind(cro_ca2, cro_sj2)
+
+cro2 %>% ggplot(aes(x=year, y=bai_mean, colour=treatment)) + 
+  geom_line() + geom_smooth(method= 'loess', span = 0.30, se=FALSE) + theme_bw() + 
+  geom_bar(aes(x=year, stat = (samp.depth*10), colour=treatment))
+
+  geom_point(aes(x=year, y = samp.depth*100))
+
+cro_ca2 %>% ggplot(aes(x=year, y=bai_mean)) + 
+  geom_line(colour='gray') + geom_smooth(method= 'loess', span = 0.80, se=FALSE) + theme_bw() 
++
+  geom
+
+plot()
+
+str(cro_ca)
+
+names(cro_ca)
+
+cro_ca <- chron(bai_ca, prefix = "CA", biweight = TRUE)
 plot(cro_sjb, add.spline=TRUE, nyrs=30)

### 2017-05-26 analysis/test-bai.Rmd
update correlation coefficient

@@
 
-chrono_bai <- function(bai_data, prefix){
-
-    if (length(prefix) == 0) {
-        prefix.str <- ""
-    }
-    else {
-        prefix.str <- as.character(prefix)[1]
-        if (is.na(prefix.str) || Encoding(prefix.str) == "bytes" || 
-            nchar(prefix.str) > 3) {
-            stop("'prefix' must be a character string with less than 4 characters")
-        }
-    }
-    n_trees <- rowSums(!is.na(bai_data))
-
-    out <- bai_data
-
-    out <- data.frame(samps)
-    out
-}
-
 ### Compute Chrono
@@
 
-chrono_bai <- function(bai_object, alpha){ 
+chrono_bai <- function(bai_object){ 
   require(dplyr)
@@
   b <- bai_object 
-  alpha <- as.numeric(alpha)
+  # alpha <- as.numeric(alpha)
 
@@
 
+  out 
+}
+
   # Loess 
@@
 
-c_sj <- chrono_bai(bai_sj, alpha=0.75)
+cro_sj <- chrono_bai(bai_sj)
+cro_sj$level <- 'SJ'
+cro_ca <- chrono_bai(bai_ca)
+cro_ca$level <- 'CA'
 
+# San JUan 
+cro_sjH <- chrono_bai(bai_sjH)
+cro_sjH$site <- 'SJ_High'
+
+cro_sjL <- chrono_bai(bai_sjL)
+cro_sjL$site <- 'SJ_Low'
+
+cro_sj<- rbind(cro_sjH, cro_sjL)
+cro_sj$loc <- 'Norte (San Juan)'
+
+cro %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
+  theme(panel.grid = element_blank())
+
+# Canar 
+cro_caH <- chrono_bai(bai_caH)
+cro_caH$site <- 'CA_High'
+te <- chrono_bai(bai_caH)
+plot.crn(te)
+plot(te)
+plot(te, add.spline=TRUE, nyrs=30)
+
+cro_caL <- chrono_bai(bai_caL)
+cro_caL$site <- 'CA_Low'
+
+cro_ca <- rbind(cro_caH,cro_caL)
+cro_ca$loc <- 'Sur (Canar)'
+
+cro_ca %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
+  theme(panel.grid = element_blank())
+
+cro_sites <- rbind(cro_ca, cro_sj)
+
+cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_line() + geom_point() + 
+  theme(panel.grid = element_blank()) + facet_wrap(~loc, nrow = 2) + 
+  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) + 
+  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+
+ + 
+geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+
+  geom_ribbon(aes(x= year, 
+                  ymin = loess_fit - 1.96*loess_se,
+                  ymax = loess_fit + 1.96*loess_se), na.rm = FALSE, 
+              fill = "grey") +
 c_sj
@@
 
+chrono_bai <- function(bai_data, prefix){
+
+    if (length(prefix) == 0) {
+        prefix.str <- ""
+    }
+    else {
+        prefix.str <- as.character(prefix)[1]
+        if (is.na(prefix.str) || Encoding(prefix.str) == "bytes" || 
+            nchar(prefix.str) > 3) {
+            stop("'prefix' must be a character string with less than 4 characters")
+        }
+    }
+    n_trees <- rowSums(!is.na(bai_data))
+
+    out <- bai_data
+
+    out <- data.frame(samps)
+    out
+}
+
 ```
+
+```{r}
+# centred moved average smooth
+# https://cran.r-project.org/web/packages/smooth/vignettes/sma.html
+
+correla_cronos <- function(cronologia, win_vector, crononame){
+
+  cro_object <- cronologia
+
+  # Set bai and years variables 
+  bai <- cro_object[,'bai_mean']
+  yearss <- cro_object[,'year']
+
+  # object to store 
+  out <- c()
+
+  for (i in seq_along(win_vector)){ 
+    # moving window size 
+    f <- rep(1/i, i)
+
+    # moving average smooth 
+    bai_smooth <- stats::filter(bai, f, sides=1)
+
+    # out 
+    out_aux <- as.data.frame(cbind(yearss, bai_smooth))
+
+    out_aux$size_wind <- i
+    out <- rbind(out, out_aux)
+  }
+  out$crononame <- crononame 
+  out
+}
+
+correla_cro_caH <- correla_cronos(cro_caH, win_vector=seq(1,40, by=1), crononame = 'cro_caH')
+correla_cro_caL <- correla_cronos(cro_caL, win_vector=seq(1,40, by=1), crononame = 'cro_caL')
+correla_cro_sjH <- correla_cronos(cro_sjH, win_vector=seq(1,40, by=1), crononame = 'cro_sjH')
+correla_cro_sjL <- correla_cronos(cro_sjL, win_vector=seq(1,40, by=1), crononame = 'cro_sjL')
+
+correla_croSJ <- rbind(correla_cro_sjH, correla_cro_sjL)
+
+# SJ 
+out_final <- c()
+
+for (j in 1:40){
+
+  aux <- correla_croSJ %>% 
+    filter(size_wind == j) %>%
+    select(-size_wind) %>% 
+    spread(crononame, bai_smooth) %>% 
+    as.data.frame()
+
+  rownames(aux) <- aux$yearss 
+  aux1 <- aux[,-1]
+
+  output_correla <- interseries.cor(aux1)
+
+  out <- as.data.frame(cbind(res.cor = output_correla$res.cor[1], 
+                           p_val = output_correla$p.val[1]))
+
+  out$size <- j
+
+  out_final <- rbind(out_final, out)
+
+  }
+
+out_final$site <- 'SJ'
+
+correla_final_sj <- out_final
+
+# CA
+correla_croCA <- rbind(correla_cro_caH, correla_cro_caL)
+out_final <- c()
+
+for (j in 1:40){
+
+  aux <- correla_croCA %>% 
+    filter(size_wind == j) %>%
+    select(-size_wind) %>% 
+    spread(crononame, bai_smooth) %>% 
+    as.data.frame()
+
+  rownames(aux) <- aux$yearss 
+  aux1 <- aux[,-1]
+
+  output_correla <- interseries.cor(aux1)
+
+  out <- as.data.frame(cbind(res.cor = output_correla$res.cor[1], 
+                           p_val = output_correla$p.val[1]))
+
+  out$size <- j
+
+  out_final <- rbind(out_final, out)
+
+  }
+
+out_final$site <- 'CA'
+
+correla_final_ca <- out_final
+
+correla_finales <- rbind(correla_final_ca, correla_final_sj)
+correla_finales %>% 
+  ggplot(aes(x=size, y=res.cor, colour=site)) + geom_line()+
+  theme_bw() + 
+  ylab('Correlation coefficient') + xlab('years(smooth)') + geom_smooth(method = 'loess', se=FALSE)
+
+c11 <- c1[26:nrow(c1),]
+
+cor(c11$cro_sjH, c11$cro_sjL)
+
+ggplot(c1, aes(x=yearss, y=cro_sjH)) + geom_line() + 
+  geom_line(aes(x=yearss, y=cro_sjL), colour='blue')
+
+cronologia <- cro_caH
+cro_object <- get()
+cro_object <- get(cronologia)
+
+plot(cro_caH$year, cro_caH$bai_mean, type='l')
+lines(cro_caH$year, y_lag, col="blue")
+
+  y_lag <- filter(y, f20, sides=1)
+lines(x, y_lag, col="red")
+
+# average of current sample and 19 previous samples (red)
+f20 <- rep(1/20, 20)
+f20
+#>  [1] 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05 0.05
+#> [18] 0.05 0.05 0.05
 

### 2017-05-26 analysis/test-bai.Rmd
compute chronology using biweight or aritmetic mean

@@
 ```{r, echo=FALSE}
-# machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+#machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 
+### Mean vs. Tukey’s Biweight Robust Mean
+
+* Computo la cronologia usando la función `chron`. 
+* Para SJ hag una comparación de la cronologia construida con biweight (*treatment = biweigth*) o construida con la media aritmética (*treatment = No biweigth*) (calculada con la función `chron`)
+* Idem para CA, pero en CA si separo entre CA high y CA low
+
 ```{r}
@@
   mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('No biweight')) %>% 
+  select(bai_mean = SJstd, samp.depth, year, treatment)
+
+aux_cro <- rbind(cro_b, cro_notb)
+
+aux_cro %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
+  geom_point() + geom_line() + 
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
+  theme_bw() + theme(panel.grid.major = element_blank()) +
+  ggtitle('SJ chronology built using Biweight or Mean')
+```
+
+```{r}
+# Robust mean biweith or not
+cro_cab_h <- chron(bai_caH, prefix = "CA", biweight = TRUE)
+cro_canotb_h <- chron(bai_caH, prefix = "CA", biweight = FALSE)
+
+cro_cab_h_aux <- cro_cab_h %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
   mutate(treatment = as.factor('biweight')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
+  select(bai_mean = CAstd, samp.depth, year, treatment)
+
+cro_canotb_h_aux <- cro_canotb_h  %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('No biweight')) %>% 
+  select(bai_mean = CAstd, samp.depth, year, treatment)
+
+aux_croca_h <- rbind(cro_cab_h_aux, cro_canotb_h_aux)
+aux_croca_h$site <- 'CA High'
+
+cro_cab_l <- chron(bai_caL, prefix = "CA", biweight = TRUE)
+cro_canotb_l <- chron(bai_caL, prefix = "CA", biweight = FALSE)
+
+cro_cab_l_aux <- cro_cab_l %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('biweight')) %>% 
+  select(bai_mean = CAstd, samp.depth, year, treatment)
+
+cro_canotb_l_aux <- cro_canotb_l  %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('No biweight')) %>% 
+  select(bai_mean = CAstd, samp.depth, year, treatment)
+
+aux_croca_l <- rbind(cro_cab_l_aux, cro_canotb_l_aux)
+aux_croca_l$site <- 'CA Low'
+
+aux_croca <- rbind(aux_croca_h, aux_croca_l)
+
+aux_croca %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
+  geom_point() + geom_line() + 
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
+  theme_bw() + theme(panel.grid.major = element_blank()) + 
+  facet_wrap(~site, nrow = 2) +
+    ggtitle('CA chronology built using Biweight or Mean')
+```
 

### 2017-05-26 script/R/chrono_bai.R
create chrono bai function

@@ -1,0 +1,31 @@
+## Function to create a chronology 
+
+## The function require a bai_object (mean bai of each year by tree). 
+## It computes the mean, sd and se of BAI by row (ie by year)
+## It also computes the sample deepth, i.e. number of trees used to build the chrono. 
+
+chrono_bai <- function(bai_object){ 
+  require(dplyr) 
+  
+  b <- bai_object 
+  
+  # Prepare data 
+  b_aux <- b %>% 
+    ##  Create a variable of year 
+    mutate(year = as.numeric(rownames(.))) %>% 
+    ## melt data 
+    gather(series, bai, -year) %>% 
+    # remove NA values 
+    filter(bai != "") 
+  
+  # Compute mean, sd, se and sample deepth (n_trees)
+  out <- b_aux %>% 
+    group_by(year) %>% 
+    summarise(bai_mean = mean(bai), 
+              bai_sd = sd(bai), 
+              bai_se = bai_sd / sqrt(n()), 
+              n_trees = n()) %>% 
+    as.data.frame()
+  
+  out 
+} 

### 2017-05-26 script/R/correla_cronos.R
function to compute correlation cronos

@@ -1,0 +1,34 @@
+## Function to get correlation between cronos 
+
+## The function require:
+## * cronologia: chronology computed with chrono_bai 
+## * win_vector: a vector with the window size (i.e.: c(1,2,3))
+## * cronomame: a character with the name of the chrono (i.e: 'cro_caH')
+
+correla_cronos <- function(cronologia, win_vector, crononame){
+  
+  cro_object <- cronologia 
+  
+  # Set bai and years variables 
+  bai <- cro_object[,'bai_mean']
+  yearss <- cro_object[,'year']
+  
+  # object to store 
+  out <- c() 
+  
+  for (i in seq_along(win_vector)){ 
+    # moving window size 
+    f <- rep(1/i, i)
+    
+    # moving average smooth 
+    bai_smooth <- stats::filter(bai, f, sides=1)
+    
+    # out 
+    out_aux <- as.data.frame(cbind(yearss, bai_smooth))
+    
+    out_aux$size_wind <- i
+    out <- rbind(out, out_aux)
+  }
+  out$crononame <- crononame 
+  out
+} 

### 2017-05-26 script/R/correla_cronos.R
update fnction

@@ -5,7 +5,7 @@
 ## * win_vector: a vector with the window size (i.e.: c(1,2,3))
 ## * cronomame: a character with the name of the chrono (i.e: 'cro_caH')
 
-correla_cronos <- function(cronologia, win_vector, crononame){
+suaviza_cronos <- function(cronologia, win_vector, crononame){
   
   cro_object <- cronologia 
   
@@ -21,7 +21,7 @@ correla_cronos <- function(cronologia, win_vector, crononame){
     f <- rep(1/i, i)
     
     # moving average smooth 
-    bai_smooth <- stats::filter(bai, f, sides=1)
+    bai_smooth <- stats::filter(bai, f, sides=2)
     
     # out 
     out_aux <- as.data.frame(cbind(yearss, bai_smooth))

### 2017-05-26 script/R/suaviza_cronos.R
update description

@@ -1,4 +1,4 @@
-## Function to get correlation between cronos 
+## Function to smooth cronos using centred moving averages with different window sizes 
 
 ## The function require:
 ## * cronologia: chronology computed with chrono_bai 

### 2017-05-26 man/notes_dendro_analysis.Rmd
update analysis and comments

@@
  * Tukey’s Biweight Robust Mean (ver pág 123 @Cook1990)
- * :red_circle: ¿cual usar de las dos? ¿Ask to Guillermo?
+
+La media aritmética es la que usa Guillermo para BAI, si queremos hacer cronologías directamente con RW, pues utilizaríamos la biweight.
+(G.
+Gea *com.
+per*).
+He encontrado algunas diferencias sobre todo en las cronologias de Cañar :red_circle: ASK to GUILLERMO.
+
+Puedo calcularlas con la función `chron()` pero esto no me permite obtener sd, se del bai por año.
+Por ello me creo una función llamada `chrono_bai` (código aquí: [./script/R/chrono_bai.R](/script/R/gea/chrono_bai.R))
+
+### Similitud chronologias
+Testamos si utilizar una crono por localidad (CA / SJ) o una por site (CA High, CA Low, SJ High, SJ Low). Para ello varias aproximaciones:
+
+* Crear una crono por cada sitio 
+* Comparación visual (grafica)
+* Comparación similitud series (ver la aproximación de @Dorado2017):
+
+    * Cada cronologia es suavizada usando centred moving averages con diferentes windows sizes.
+Para ello utilizamos la función `suaviza_cronos` (código aquí: [./script/R/suaviza_cronos.R](/script/R/gea/suaviza_cronos.R))
+    * Posteriormente se calcula la correlación entre chronos para cada suavizado. 
+    * Con boosptrap se obtienen niveles de significación :red_circle: ASK to Guillermo and Isabel.
 

### 2017-05-26 analysis/analysis_chronologies.Rmd
change name

--- 2017-05-25 analysis/test-bai.Rmd
@@
 source(paste0(getwd(), '/script/R/rw_byTree.R'))
+source(paste0(getwd(), '/script/R/chrono_bai.R'))
+source(paste0(getwd(), '/script/R/suaviza_cronos.R'))
 ```
@@
 machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
+#machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
   mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
+  mutate(treatment = as.factor('No biweight')) %>% 
   select(bai_mean = SJstd, samp.depth, year, treatment)
 
-# Compute chronologies of BAI
+aux_cro <- rbind(cro_b, cro_notb)
 
-chrono_bai <- function(bai_data, prefix){
+aux_cro %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
+  geom_point() + geom_line() + 
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
+  theme_bw() + theme(panel.grid.major = element_blank()) +
+  ggtitle('SJ chronology built using Biweight or Mean')
+```
 
-    if (length(prefix) == 0) {
-        prefix.str <- ""
-    }
-    else {
-        prefix.str <- as.character(prefix)[1]
-        if (is.na(prefix.str) || Encoding(prefix.str) == "bytes" || 
-            nchar(prefix.str) > 3) {
-            stop("'prefix' must be a character string with less than 4 characters")
-        }
-    }
-    n_trees <- rowSums(!is.na(bai_data))
+```{r}
+# Robust mean biweith or not
+cro_cab_h <- chron(bai_caH, prefix = "CA", biweight = TRUE)
+cro_canotb_h <- chron(bai_caH, prefix = "CA", biweight = FALSE)
 
-    out <- bai_data
+cro_cab_h_aux <- cro_cab_h %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('biweight')) %>% 
+  select(bai_mean = CAstd, samp.depth, year, treatment)
 
-    out <- data.frame(samps)
-    out
+cro_canotb_h_aux <- cro_canotb_h  %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('No biweight')) %>% 
+  select(bai_mean = CAstd, samp.depth, year, treatment)
+
+aux_croca_h <- rbind(cro_cab_h_aux, cro_canotb_h_aux)
+aux_croca_h$site <- 'CA High'
+
+cro_cab_l <- chron(bai_caL, prefix = "CA", biweight = TRUE)
+cro_canotb_l <- chron(bai_caL, prefix = "CA", biweight = FALSE)
+
+cro_cab_l_aux <- cro_cab_l %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('biweight')) %>% 
+  select(bai_mean = CAstd, samp.depth, year, treatment)
+
+cro_canotb_l_aux <- cro_canotb_l  %>% 
+  mutate(year = as.numeric(rownames(.))) %>% 
+  mutate(treatment = as.factor('No biweight')) %>% 
+  select(bai_mean = CAstd, samp.depth, year, treatment)
+
+aux_croca_l <- rbind(cro_cab_l_aux, cro_canotb_l_aux)
+aux_croca_l$site <- 'CA Low'
+
+aux_croca <- rbind(aux_croca_h, aux_croca_l)
+
+aux_croca %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
+  geom_point() + geom_line() + 
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
+  theme_bw() + theme(panel.grid.major = element_blank()) + 
+  facet_wrap(~site, nrow = 2) +
+    ggtitle('CA chronology built using Biweight or Mean')
+```
+
+## Compute Chrono 
+### San Juan chronologies by site
+
+```{r}
+# San Juan 
+cro_sjH <- chrono_bai(bai_sjH)
+cro_sjH$site <- 'SJ_High'
+
+cro_sjL <- chrono_bai(bai_sjL)
+cro_sjL$site <- 'SJ_Low'
+
+cro_sj<- rbind(cro_sjH, cro_sjL)
+cro_sj$loc <- 'Norte (San Juan)'
+
+cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
+  theme(panel.grid = element_blank()) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+```
+
+### Cañar chronologies by site
+
+```{r}
+# Canar 
+cro_caH <- chrono_bai(bai_caH)
+cro_caH$site <- 'CA_High'
+
+cro_caL <- chrono_bai(bai_caL)
+cro_caL$site <- 'CA_Low'
+
+cro_ca <- rbind(cro_caH,cro_caL)
+cro_ca$loc <- 'Sur (Canar)'
+
+cro_ca %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
+  theme(panel.grid = element_blank()) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+```
+
+```{r}
+# Combine and plot 
+cro_sites <- rbind(cro_ca, cro_sj)
+
+cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_line() + geom_point() + 
+  theme(panel.grid = element_blank()) + facet_wrap(~loc, nrow = 2) + 
+  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+```
+
+## Compute similarity between chronos 
+Use @Dorado2017 approach.
+
+* Para cada chrono se realiza un suavizado, utilizando diferente tamaño de ventana. 
+* Luego calculamos la correlación entre series :red_circle: ask to Isabel and guillermo
+
+```{r}
+# centred moved average smooth
+suaviza_caH <- suaviza_cronos(cro_caH, win_vector=seq(1,40, by=1), crononame = 'cro_caH')
+suaviza_caL <- suaviza_cronos(cro_caL, win_vector=seq(1,40, by=1), crononame = 'cro_caL')
+suaviza_sjH <- suaviza_cronos(cro_sjH, win_vector=seq(1,40, by=1), crononame = 'cro_sjH')
+suaviza_sjL <- suaviza_cronos(cro_sjL, win_vector=seq(1,40, by=1), crononame = 'cro_sjL')
+
+```
+
+```{r}
+# compute corelations between series 
+# SJ 
+suaviza_sj <- rbind(suaviza_sjH, suaviza_sjL)
+
+out_final <- c()
+
+for (j in 1:40){
+
+  aux <- suaviza_sj  %>% 
+    filter(size_wind == j) %>%
+    select(-size_wind) %>% 
+    spread(crononame, bai_smooth) %>% 
+    as.data.frame()
+
+  rownames(aux) <- aux$yearss 
+  aux1 <- aux[,-1]
+
+  output_correla <- interseries.cor(aux1)
+
+  out <- as.data.frame(cbind(res.cor = output_correla$res.cor[1], 
+                           p_val = output_correla$p.val[1]))
+
+  out$size <- j
+
+  out_final <- rbind(out_final, out)
+
   }
 
-### Compute Chrono
+out_final$site <- 'SJ'
+correla_sj <- out_final
 
-b <- bai_sj
+# CA
+suaviza_ca <- rbind(suaviza_caH, suaviza_caL) 
+out_final <- c()
 
-chrono_bai <- function(bai_object, alpha){ 
-  require(dplyr)
+for (j in 1:40){
 
-  b <- bai_object 
-  alpha <- as.numeric(alpha)
-
-  # Prepare data 
-  b_aux <- b %>% 
-  ##  Create a variable of year 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  ## melt data 
-  gather(series, bai, -year) %>% 
-  # remove NA values 
-  filter(bai != "")
-
-  # Compute mean, sd, se and sample deepth (n_trees)
-  out <- b_aux %>% 
-  group_by(year) %>% 
-  summarise(bai_mean = mean(bai), 
-            bai_sd = sd(bai), 
-            bai_se = bai_sd / sqrt(n()), 
-            n_trees = n()) %>% 
+  aux <- suaviza_ca %>% 
+    filter(size_wind == j) %>%
+    select(-size_wind) %>% 
+    spread(crononame, bai_smooth) %>% 
     as.data.frame()
 
+  rownames(aux) <- aux$yearss 
+  aux1 <- aux[,-1]
+
+  output_correla <- interseries.cor(aux1)
+
+  out <- as.data.frame(cbind(res.cor = output_correla$res.cor[1], 
+                           p_val = output_correla$p.val[1]))
+
+  out$size <- j
+
+  out_final <- rbind(out_final, out)
+
+  }
+
+out_final$site <- 'CA'
+
+correla_ca <- out_final
+
+# Combine datasets and plot coefficient correlation 
+correla <- rbind(correla_ca, correla_sj)
+
+correla %>% 
+  ggplot(aes(x=size, y=res.cor, colour=site)) + geom_line()+
+  theme_bw() + ylim(c(0,1)) +
+  ylab('Correlation coefficient') + 
+  xlab('Suavizado (centred moving average window size)') + 
+  geom_smooth(method = 'loess', se=FALSE) + 
+  theme(panel.grid = element_blank())
+```
+
+---> VAS por AQUI
+
+```{r, echo=FALSE, eval=FALSE}
+
+# CODE SNIPPETS 
   # Loess 
@@
 }
-
-c_sj <- chrono_bai(bai_sj, alpha=0.75)
-
-c_sj
-
-c_sj %>% ggplot(aes(x=year, y=bai_mean)) + 
-  theme_bw() + 
-  geom_ribbon(aes(x= year, 
-                  ymin = loess_fit - 1.96*loess_se,
-                  ymax = loess_fit + 1.96*loess_se), na.rm = FALSE, 
-              fill = "grey") +
-  geom_line() + geom_point() + 
-  geom_line() + geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-
-bb2 %>% ggplot(aes(x=year, y=bai_mean)) + 
-  theme_bw() + 
-  geom_ribbon(aes(x= year, 
-                  ymin = bai_mean - bai_se,
-                  ymax = bai_mean + bai_se), na.rm = FALSE, 
-              fill = "grey") +
-  geom_line() + geom_point() + 
-  geom_smooth(method= 'loess', span = 0.30, se=FALSE)
-
-# IC LOESS
-
-  geom_smooth(method= 'loess', span = 0.30, se=FALSE) + theme_bw() + 
-  geom_bar(aes(x=year, stat = (samp.depth*10), colour=treatment))
-
-# cc <- chrono_bai(bai_sj, prefix = "SJ")
-
-```
-
-```{r}
 
@@
   geom_bar(aes(x=year, stat = (samp.depth*10), colour=treatment))
-
-  geom_point(aes(x=year, y = samp.depth*100))
-
-cro_ca2 %>% ggplot(aes(x=year, y=bai_mean)) + 
-  geom_line(colour='gray') + geom_smooth(method= 'loess', span = 0.80, se=FALSE) + theme_bw() 
-+
-  geom
-
-plot()
-
-str(cro_ca)
-
-names(cro_ca)
-
-cro_ca <- chron(bai_ca, prefix = "CA", biweight = TRUE)
-plot(cro_sjb, add.spline=TRUE, nyrs=30)
-
-cro_sj <- chron(bai_sj, prefix = "SJ", biweight = FALSE)
-
-cc <- cbind(cro_sj, cro_sjb)
-
-su
-
-```
-
----> VAS por AQUI
-
-```
 

### 2017-05-26 analysis/analysis_chronologies.Rmd
update file and render

@@
 library("pander")
-source(paste0(getwd(), '/script/R/bai_piovesan.R'))
-source(paste0(getwd(), '/script/R/rw_byTree.R'))
-source(paste0(getwd(), '/script/R/chrono_bai.R'))
-source(paste0(getwd(), '/script/R/suaviza_cronos.R'))
 ```
@@
 compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+```
+
+```{r}
+source(paste0(di, 'script/R/bai_piovesan.R'))
+source(paste0(di, 'script/R/rw_byTree.R'))
+source(paste0(di, 'script/R/chrono_bai.R'))
+source(paste0(di, 'script/R/suaviza_cronos.R'))
 ```
@@
 
-```{r}
 # Get summary ring-width series
@@
 sj_sin <- sj[which(colnames(sj) %in% (id_cores_no_bark))]
-```
 
@@
 
-```{r}
 compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
@@
 diam_cores_sj <- diam_cores %>% filter(! id_cores %in% id_cores_no_bark)
-```
 
-```{r}
 rwl.report(sj_cor)
@@
 
-```
-
-```{r}
 set.seed(1234)
@@
 
-```{r, eval=FALSE, echo=FALSE}
 # BAI dplR
@@
 
-```
-
-```{r}
-
 x <- c(1,2,2,3,1,3,4,5)
@@
 
-```
-
-```{r}
 # tes plots

### 2017-05-29 script/R/gea/pgc_rw.R
upload disturbance script of GG

@@ -1,0 +1,312 @@
+########Plots growth changes Montejo RW
+########
+######
+### este es un ejemplo que usé para el artículo Dorado-Liñán et al. (2017) Dendrochronlogia
+### Hay 3 especies (misma función uso): Q. pyrenaica, Q. petraea, Fagus sylvatica en el norte de Madrid (Montejo)
+
+####A LA VUELTA DE LUGO SEGUIR CALCULANDO EL % DE ÁRBOLES QUE PASAN LOS LÍMITES 25-50%, LUEGO SEGUIR CON BAI
+####
+
+library(dplR)
+#
+##Primero me construyo una matriz con todos los crecimientos por columnas
+#
+AGC2=function(RWL)
+{
+GC=as.data.frame(NULL)
+#
+for (n in 1:ncol(RWL))
+{
+	series <- as.data.frame(RWL[, n])
+	series$year=as.numeric(row.names(RWL))
+	series=subset(series, !is.na(series[1]))
+	tree=as.numeric(substr(colnames(RWL)[n],3,4))
+	tree2=as.numeric(substr(colnames(RWL)[n], 5,5))
+	if(!is.na(tree2)){tree3=as.numeric(substr(colnames(RWL[n]),3,5))} else {tree3=tree}
+	series$tree=as.numeric(tree3)
+	GC=rbind(GC, series)
+}
+names(GC)[1]="RW"
+GC2=aggregate(GC[,1], by=list(year=GC$year,tree=GC$tree), mean)
+names(GC2)[3]="RW"
+GC2=GC2[order(GC2$tree, GC2$year),]
+GC2$tree2=1
+for (i in 2:nrow(GC2))
+{
+if(GC2$year[i]<GC2$year[i-1]) GC2$tree2[i]=1+GC2$tree2[i-1]
+else GC2$tree2[i]=GC2$tree2[i-1]
+}
+GC2<<-GC2
+}
+#
+#
+AGC2haya=function(RWL)
+{
+  GC=as.data.frame(NULL)
+  #
+  for (n in 1:ncol(RWL))
+  {
+    series <- as.data.frame(RWL[, n])
+    series$year=as.numeric(row.names(RWL))
+    series=subset(series, !is.na(series[1]))
+    #tree=as.numeric(substr(colnames(RWL)[n],3,4))
[…]

### 2017-05-30 analysis/analysis_chronologies.Rmd
update biblio and compute CI by bootstrap

@@
 library("pander")
+library('boot')
 ```
@@
 
-* Para cada chrono se realiza un suavizado, utilizando diferente tamaño de ventana. 
-* Luego calculamos la correlación entre series :red_circle: ask to Isabel and guillermo
+* Para cada chrono se realiza un suavizado, utilizando diferente tamaño de ventana (1 a 40 years). Así evaluamos la sincronía entre dos series, tanto a nivel interanual como a largo plazo. 
+* Luego calculamos la correlación entre series para cada tamaño de ventana.
+Usamos correlaciones de Pearson o Sperman entre las series (es decir entre las zonas comunes de las cronologías).
+Ojo con la autocorrelación.
+Para ello puedo utilizar `interseries.cor()` del pkg `dplR`; o utilizar `cor.test()` (`stats`).
+El primero está basado en el segundo, así que utilizo el segundo.
+* Obtener IC utilizando bootstrap. Varias aproximaciones:
+
+    * Utilizar el pkg `boot` y conseguir IC. 
+    * Aproximación de @Mudelsee2014.
+En su paper habla de correlacion entre dos series y utiliza un programa en Fortran-90.
+He contactado con el autor y no tiene pensado hacerlo en R.
+He intentado utilizar Fortran-90, pero no tengo ni idea.
+Asi que desisto de utilizar esta computación de los IC.
 
@@
 
+### Bootstrap IC
+
 ```{r}
-# compute corelations between series 
-# SJ 
+# Metodo correlation
+tipos_correla <- c('spearman', 'pearson')
+
+for (m in tipos_correla){
+
+  metodo_correla <- m
+
 suaviza_sj <- rbind(suaviza_sjH, suaviza_sjL)
@@
     spread(crononame, bai_smooth) %>% 
+    na.omit() %>% # to compare the series in the years they are overlapping 
     as.data.frame()
 
+  # Put years as rownames and delete yearss variable
   rownames(aux) <- aux$yearss 
@@
 
-  output_correla <- interseries.cor(aux1)
+  # Names the series 
+  names(aux1) <- c('H', 'L')
 
-  out <- as.data.frame(cbind(res.cor = output_correla$res.cor[1], 
-                           p_val = output_correla$p.val[1]))
+  # Get real value of correlation between smoothed series 
+  pearson <- cor.test(aux1$H, aux1$L, method = metodo_correla)$estimate
 
+  ## Bootstratp to ger CI 
+  correlationTest <- function(x,index){
+    ct <- cor.test(x$H[index], x$L[index], method = metodo_correla)
+    return(ct$estimate)
+    }
+
+  ### boostrap   
+  boot_cor <- boot(aux1, correlationTest, R=1000)
+  ### CI
+  bci <- boot.ci(boot_cor, conf = 0.95, type = 'bca')
+
+  # Output results 
+  estimate <- bci$t0 # Pearson coefficient
+  ci_lower <- bci$bca[4] # Lower CI
+  ci_upper <- bci$bca[5] # Upper CI
+
+  out <- as.data.frame(cbind(estimate, 
+                             ci_lower, 
+                             ci_upper))
+  # Smoothing year (window size)
   out$size <- j
@@
 out_final$site <- 'CA'
-
 correla_ca <- out_final
@@
 correla <- rbind(correla_ca, correla_sj) 
+correla$metodo_correla <- metodo_correla
+assign(paste0('correla_', metodo_correla), correla)
+}
+
+```
+
+### Plot similarity among chronologies
+
+* Correlation among chronologies for the common period of each site in different time-domains after pre-filtering the time-series with increasing size of the moving-average window.
+
+```{r, fig.width=12, fig.height=8}
+correla <- rbind(correla_pearson, correla_spearman)
+correla$series <- ifelse(correla$site == 'CA', 
+                         'CAHigh-CALow',
+                         'SJHigh-SJLow')
 
 correla %>% 
-  ggplot(aes(x=size, y=res.cor, colour=site)) + geom_line()+
-  theme_bw() + ylim(c(0,1)) +
+  ggplot(aes(x=size, y=estimate, colour=series)) + geom_line()+
+  theme_bw() + 
   ylab('Correlation coefficient') +
-  xlab('Suavizado (centred moving average window size)') + 
-  geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank())
+  xlab('Smoothing (centred moving average window size)') + 
+  # geom_smooth(method = 'loess', se=FALSE) + 
+  theme(panel.grid = element_blank()) +
+  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=series), linetype = 2)  +
+  geom_line(aes(x=size, y=ci_upper, colour=series), linetype = 2) + 
+  facet_wrap(~metodo_correla, ncol=2) +
+  theme(strip.background = element_blank())
 ```
-
----> VAS por AQUI
 

### 2017-05-30 man/material_methods.Rmd
update material methods

---
title: "material and methods"
author: "AJ Perez-Luque (@ajpelu); G. Gea-Izquierdo"
date: "2017"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: references.bib
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Materials and methods

## Dendrochronological methods
... dendro
For each focal tree we measured diameter at breast height (DBH) and total height. 
A total of xx trees were sampled. 
...

We built chronologies for each site (two)

Site chronologies were built by averaging all tree BAI measurement of the same site. To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` (:red_circle: cite boot).

### 2017-05-31 man/material_methods.Rmd
update bib
update

@@
 To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` (:red_circle: cite boot).
+Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
+
+```{r}
+py <- pointer.rgc(s033, nb.yrs=4, rgc.thresh.pos=60, rgc.thresh.neg=40, series.thresh=75)
+
+rgc.plot(py,start.yr=1950,end.yr=NULL,sd.disp=FALSE
+```
+
+## References

### 2017-05-31 man/ms_discussion.Rmd
update

---
title: "discussion"
author: "AJ Perez-Luque (@ajpelu); G. Gea-Izquierdo"
date: "2017"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: references.bib
csl: ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Discussions notes

## Dos cronos en Cañar 
* @GeaIzquierdo2014 muestrea solo en CA-Low (QUPY10) y obtiene resultados similares a CA-Low. 
* En tan poco espacio hay diferencias enormes entre las dos cronos
* *Las tendencias de BAI de Cáñar baja altitud son diferentes a las del norte, las cuales se parecen mucho a las que suelen encontrarse en la mayoría de sitios. Puedes mirar por ejemplo las tendencias de las líneas negras (de más de 100 años, sobre todo) en la Fig 2 del artículo de Ecosystems… fíjate que QUPY3 se parece más a QUPY10… y QUPY3 era la única dehesa (en parte) y uno de los sitios más secos (por otro lado, aunque frío). En fin, todo esto es algo especulativo, pero sirva como exploración de tus cronos, y para mostrar que me parece muy interesante la diferencia entre Cáñar alto y bajo (y nuevo).*

## References

### 2017-05-31 script/R/correlaBootIC.R
add new function to compute correlation

@@ -1,0 +1,81 @@
+# Function to compute correlation between two series and compute IC by boot 
+# - df_suaviza: df with smoothed series (the two series together, by rbind), and the following attributes: 
+#  * yearss
+#  * bai_smooth: the value of bai smoothed
+#  * size_window: the size of window used to smoothed the serie (centred moving average)
+#  * cronomame: the name of the chrono 
+# - tipos_correla: types of correlation. ie: tc <- c('spearman', 'pearson')
+# - years_suavizado: the length of years of smoothed used (i.e. 40)
+# - name_comparison: name to the two series comparison. i.e.: 'sjL-sjH'
+# BootIC are computed using the following parameters: 
+# * replicates = 1000 
+# * ic = 0.95 
+# * type of interval = "bca"
+
+
+correlaBootIC <- function(df_suaviza, tipos_correla, years_suavizado, 
+                          name_comparison){
+  
+  ys <- as.numeric(years_suavizado)
+  nc <- as.character(name_comparison)
+  
+  out_correla <- c() 
+  for (m in tipos_correla){ 
+    # Choose correlation approach 
+    metodo_correla <- m
+    
+    out_final <- c() 
+    
+    for (j in 1:ys){
+      aux <- df_suaviza %>% 
+        filter(size_wind == j) %>%
+        select(-size_wind) %>% 
+        spread(crononame, bai_smooth) %>% 
+        na.omit() %>% # to compare the series in the years they are overlapping 
+        as.data.frame() 
+      
+      # Put years as rownames and delete yearss variable
+      rownames(aux) <- aux$yearss 
+      aux1 <- aux[,-1]
+      
+      # Names the series 
+      names(aux1) <- c('a', 'b')
+      
+      # Get real value of correlation between smoothed series 
+      coeff_correlation <- cor.test(aux1$a, aux1$b, method = metodo_correla)$estimate 
+      
+      ## Bootstratp to ger CI 
+      ### Function to apply 
+      correlationTest <- function(x,index){
+        ct <- cor.test(x$a[index], x$b[index], method = metodo_correla)
+        return(ct$estimate)
+        }
+      
+      ### boostrap   
+      boot_cor <- boot(aux1, correlationTest, R=1000)
+      ### CI
+      bci <- boot.ci(boot_cor, conf = 0.95, type = 'bca')
+      
+      # Output results 
+      estimate <- bci$t0 # Estimate coefficient
[…]

### 2017-05-31 analysis/analysis_chronologies.Rmd
update
update messages
update boot CI

@@
 ---
+
+```{r global_options, include=FALSE}
+knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
+```
 
@@
 source(paste0(di, 'script/R/suaviza_cronos.R'))
+source(paste0(di, 'script/R/correlaBootIC.R'))
 ```
@@
 
-## Compute similarity between chronos 
-Use @Dorado2017 approach.
+# Correlation between chronos
+
+To compute similarity between chronos we use @Dorado2017 approach.
 
@@
 suaviza_sjL <- suaviza_cronos(cro_sjL, win_vector=seq(1,40, by=1), crononame = 'cro_sjL')
-
 ```
 
-### Bootstrap IC
+## between sites similarity (sjH *vs* sjL; caH *vs* caL)
 
-```{r}
+This note is for the plot:
+
+* Correlation among chronologies for the common period of each site in different time-domains after pre-filtering the time-series with increasing size of the moving-average window.
+
+```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
 # Metodo correlation
-tipos_correla <- c('spearman', 'pearson')
+tc <- c('spearman', 'pearson')
 
-for (m in tipos_correla){
-
-  metodo_correla <- m
-
+# SJ 
 suaviza_sj <- rbind(suaviza_sjH, suaviza_sjL)
-
-out_final <- c()
-
-for (j in 1:40){
-
-  aux <- suaviza_sj  %>% 
-    filter(size_wind == j) %>%
-    select(-size_wind) %>% 
-    spread(crononame, bai_smooth) %>% 
-    na.omit() %>% # to compare the series in the years they are overlapping 
-    as.data.frame()
-
-  # Put years as rownames and delete yearss variable
-  rownames(aux) <- aux$yearss 
-  aux1 <- aux[,-1]
-
-  # Names the series 
-  names(aux1) <- c('H', 'L')
-
-  # Get real value of correlation between smoothed series 
-  pearson <- cor.test(aux1$H, aux1$L, method = metodo_correla)$estimate
-
-  ## Bootstratp to ger CI 
-  correlationTest <- function(x,index){
-    ct <- cor.test(x$H[index], x$L[index], method = metodo_correla)
-    return(ct$estimate)
-    }
-
-  ### boostrap   
-  boot_cor <- boot(aux1, correlationTest, R=1000)
-  ### CI
-  bci <- boot.ci(boot_cor, conf = 0.95, type = 'bca')
-
-  # Output results 
-  estimate <- bci$t0 # Pearson coefficient
-  ci_lower <- bci$bca[4] # Lower CI
-  ci_upper <- bci$bca[5] # Upper CI
-
-  out <- as.data.frame(cbind(estimate, 
-                             ci_lower, 
-                             ci_upper))
-  # Smoothing year (window size)
-  out$size <- j
-
-  out_final <- rbind(out_final, out)
-
-}
-
-out_final$site <- 'SJ'
-correla_sj <- out_final
+correla_sj <- correlaBootIC(suaviza_sj, tc, years_suavizado = 40, 
+                            name_comparison = 'sjH-sjL')
 
@@
 suaviza_ca <- rbind(suaviza_caH, suaviza_caL)
-out_final <- c()
+correla_ca <- correlaBootIC(suaviza_ca, tc, years_suavizado = 40, 
+                            name_comparison = 'caH-caL')
 
-for (j in 1:40){
-
-  aux <- suaviza_ca  %>% 
-    filter(size_wind == j) %>%
-    select(-size_wind) %>% 
-    spread(crononame, bai_smooth) %>% 
-    na.omit() %>% # to compare the series in the years they are overlapping 
-    as.data.frame()
-
-  # Put years as rownames and delete yearss variable
-  rownames(aux) <- aux$yearss 
-  aux1 <- aux[,-1]
-
-  # Names the series 
-  names(aux1) <- c('H', 'L')
-
-  # Get real value of correlation between smoothed series 
-  pearson <- cor.test(aux1$H, aux1$L, method = metodo_correla)$estimate
-
-  ## Bootstratp to ger CI 
-  correlationTest <- function(x,index){
-    ct <- cor.test(x$H[index], x$L[index], method = metodo_correla)
-    return(ct$estimate)
-    }
-
-  ### boostrap   
-  boot_cor <- boot(aux1, correlationTest, R=1000)
-  ### CI
-  bci <- boot.ci(boot_cor, conf = 0.95, type = 'bca')
-
-  # Output results 
-  estimate <- bci$t0 # Pearson coefficient
-  ci_lower <- bci$bca[4] # Lower CI
-  ci_upper <- bci$bca[5] # Upper CI
-
-  out <- as.data.frame(cbind(estimate, 
-                             ci_lower, 
-                             ci_upper))
-  # Smoothing year (window size)
-  out$size <- j
-
-  out_final <- rbind(out_final, out)
-
-}
-
-out_final$site <- 'CA'
-correla_ca <- out_final
-
-# Combine datasets and plot coefficient correlation 
 correla <- rbind(correla_ca, correla_sj)
-correla$metodo_correla <- metodo_correla
-assign(paste0('correla_', metodo_correla), correla)
-}
-
-```
-
-### Plot similarity among chronologies
-
-* Correlation among chronologies for the common period of each site in different time-domains after pre-filtering the time-series with increasing size of the moving-average window.
-
-```{r, fig.width=12, fig.height=8}
-correla <- rbind(correla_pearson, correla_spearman)
-correla$series <- ifelse(correla$site == 'CA', 
-                         'CAHigh-CALow',
-                         'SJHigh-SJLow')
 
 correla %>% 
-  ggplot(aes(x=size, y=estimate, colour=series)) + geom_line()+
-  theme_bw() + 
-  ylab('Correlation coefficient') +
+  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
+  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
   xlab('Smoothing (centred moving average window size)') + 
@@
   geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=series), linetype = 2)  +
-  geom_line(aes(x=size, y=ci_upper, colour=series), linetype = 2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
+  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
+  facet_wrap(~metodo_correla, ncol=2) +
+  theme(strip.background = element_blank())
+
+```
+
+## between final sites similarity (sj *vs* caH; sj *vs* caL; caH *vs* caL)
+
+```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
+# Metodo correlation
+tc <- c('spearman', 'pearson')
+
+# Compute the suaviza_sj desde el principio, sin diferenciar h and l. Para ello la crono la computo de nuevo 
+cro_sjloc <- chrono_bai(bai_sj)
+cro_sjloc$site <- 'SJ'
+suaviza_sjloc <- suaviza_cronos(cro_sjloc, win_vector=seq(1,40, by=1), crononame = 'cro_sj')
+
+# sjcaH 
+suaviza_sjcaH <- rbind(suaviza_sjloc, suaviza_caH)
+correla_sjcaH <- correlaBootIC(suaviza_sjcaH, tc, years_suavizado = 40, 
+                            name_comparison = 'sj-caH')
+
+# sjcaL 
+suaviza_sjcaL <- rbind(suaviza_sjloc, suaviza_caL)
+correla_sjcaL <- correlaBootIC(suaviza_sjcaL, tc, years_suavizado = 40, 
+                            name_comparison = 'sj-caL')
+
+# Join all correlations 
+correla <- rbind(correla_ca, correla_sjcaL, correla_sjcaH)
+
+correla %>% 
+  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
+  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
+  xlab('Smoothing (centred moving average window size)') + 
+  # geom_smooth(method = 'loess', se=FALSE) + 
+  theme(panel.grid = element_blank()) +
+  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
+  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
   facet_wrap(~metodo_correla, ncol=2) +
@@
 ```
+
+---> here
 

### 2017-05-31 man/ms_material_methods.Rmd
update ms_mm
update methods

--- 2017-05-31 man/material_methods.Rmd
@@
 
+## Tree sampling 
+Samplig was carried during autumn of 2016.
+Trees were sampled at two locations located in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Table 1).
+Both sites were oak monospecific and representatives of two of the three the population's cluster identified for this specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..).
+In each site between 15 and 20 dominant trees were sampled.
+Two cores of 5 mm of diameter were taken per tree at 1.3 m using an increment borer.
+Diameter at breat height (DBH) and total height were recorded for each tree.
+Increment cores were air dried, glued onto wooden mounts and sanded.
+Annual radial growth (ring width, RW) were measured with a LINTAB measuring device (:red_circle: Rinntech 2003) coupled to a stereomicroscope, with an accuracy of 0.01 mm.
+Individual ring series were visually and statistically cross-dated with TSAP software (:red_circle: Frank Rinn, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
+Validation of the croos-dating was done using COFECHA software [@Holmes1983].
+
 ## Dendrochronological methods
@@
 
-```{r}
-py <- pointer.rgc(s033, nb.yrs=4, rgc.thresh.pos=60, rgc.thresh.neg=40, series.thresh=75)
-
-rgc.plot(py,start.yr=1950,end.yr=NULL,sd.disp=FALSE
-```
-
 ## References

### 2017-06-01 script/R/computeGC.R
UPDATE FUNCTION

@@ -1,0 +1,53 @@
+# Function ton compute Percentage of Growth Change using the approach of Nowacki and Abrams 1997. Ecol. Monograph. 67(2):225
+# Arguments: 
+# * dfrwl: a RWL dataframe
+# * ws: window size: the 
+# * prefijo: name of the series site, i.e.: 'CA' 
+
+
+computeGC <- function(dfrwl, ws, prefijo){ 
+  require(dplyr)
+  require(stringr)
+  
+  # window_size 
+  ws <- as.numeric(ws) 
+  # n_series (or n_series by tree)
+  name_series <- colnames(dfrwl)
+  
+  out_gc <- c() 
+  
+  for (ns in name_series){ 
+    # selected series 
+    ss <- dfrwl %>% 
+      dplyr::select_(.dots = ns) %>% 
+      mutate(year = as.numeric(row.names(.))) %>% 
+      na.omit() %>% 
+      as.data.frame()
+    
+    # Range years 
+    range_years <- (min(ss$year)+ws):(max(ss$year)-ws)
+    
+    aux_serie <- c()
+    
+    for (y in range_years){
+      # subsequent and precedent subset 
+      m2 <- ss %>% filter(year > y & year <= (y+ws)) %>% data.frame()
+      m1 <- ss %>% filter(year <= y & year >= (y-ws+1)) %>% data.frame()
+      # mean values
+      mean_m2 <- mean(m2[,1])
+      mean_m1 <- mean(m1[,1])
+      # GC 
+      gc = ((mean_m2 - mean_m1) / mean_m1) * 100 
+      
+      # out 
+      gc_aux <- data.frame(gc = gc, year = y, name_serie = ns)
+      gc_aux$tree <- as.numeric(stringr::str_replace(gc_aux$name_serie, prefijo, ""))
+      
+      aux_serie <- rbind(aux_serie, gc_aux)
+    }
+    
+    # Out 
+    out_gc <- rbind(out_gc, aux_serie)
+  }
+  return(out_gc)
+}

### 2017-06-01 analysis/analysis_disturbance_chronos.Rmd
generate disturbance chronos

--- 2017-05-31 analysis/analysis_chronologies.Rmd
@@
 ---
-title: "dendro"
+title: "disturbance chronologies"
 author: "AJ Perez-Luque (@ajpelu)"
@@
 
-# Read y Prepare data 
+# Disturbance chronologies
+
+* Vamos a calcular las disturbance chronologies siguiendo la aproximación de [@Nowacki1997](varios ejemplos se pueden ver en [@GeaIzquierdo2014, @Dorado2017])
+
+* Utilizaremos tres chronologias: SJ, CAH y CAL (ver [./analysis/analysis_chronologies.md]('./analysis/analysis_chronologies.md))
+
+## Prepare data
+
 * Leer datos `rwl` de SJ y CA
@@
 ```{r}
+source(paste0(di, 'script/R/rw_byTree.R'))
 source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/rw_byTree.R'))
+
 source(paste0(di, 'script/R/chrono_bai.R'))
@@
 
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
+* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low
+
     * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
@@
 
-sjL <- sj[, c("SJ0101","SJ0102","SJ0201","SJ0202","SJ0301","SJ0302","SJ0401","SJ0402","SJ0501","SJ0502",
-              "SJ0601","SJ0602","SJ0603","SJ0701","SJ0702","SJ0801","SJ0802","SJ0901","SJ0902","SJ0903",
-              "SJ1001","SJ1002","SJ1003")]
-sjH <- sj[, c("SJ1101","SJ1102","SJ1103","SJ1201","SJ1202","SJ1203","SJ1301","SJ1302","SJ1303","SJ1401",
-              "SJ1402","SJ1501","SJ1502","SJ1601","SJ1602","SJ1701","SJ1702","SJ1703","SJ1801","SJ1802",
-              "SJ1901","SJ1902","SJ2001","SJ2003","SJ2002")]
+# sjL <- sj[, c("SJ0101","SJ0102","SJ0201","SJ0202","SJ0301","SJ0302","SJ0401","SJ0402","SJ0501","SJ0502",
+#               "SJ0601","SJ0602","SJ0603","SJ0701","SJ0702","SJ0801","SJ0802","SJ0901","SJ0902","SJ0903",
+#               "SJ1001","SJ1002","SJ1003")]
+# sjH <- sj[, c("SJ1101","SJ1102","SJ1103","SJ1201","SJ1202","SJ1203","SJ1301","SJ1302","SJ1303","SJ1401",
+#               "SJ1402","SJ1501","SJ1502","SJ1601","SJ1602","SJ1701","SJ1702","SJ1703","SJ1801","SJ1802",
+#               "SJ1901","SJ1902","SJ2001","SJ2003","SJ2002")]
 ```
@@
 # Set levels of eleveation 
-sj_lowcode  <- paste0('SJ', str_pad(1:10, 2, pad='0'))
-sj_highcode <- paste0('SJ', 11:20)
 ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
@@
 ft <- ft %>% 
-  mutate(elev = as.factor(
-    ifelse(id_focalLoc %in% sj_lowcode, 'Low',
-                       ifelse(id_focalLoc %in% sj_highcode, 'High',
-                              ifelse(id_focalLoc %in% ca_lowcode, 'Low', 'High'))))) %>%
-  mutate(site = as.factor(paste0(loc, '_', elev)))
+  mutate(site = as.factor(
+    ifelse(id_focalLoc %in% ca_lowcode, 'CAL', 
+           ifelse(id_focalLoc %in% ca_highcode, 'CAH', 'SJ'))))
 ```
-
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
-    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
 
@@
 
-* Por cada loc (ca/sj) o site (ca_high, ca_low, sj_high, sj_low) calculamos los siguientes datos: 
+* Por cada site (CAH, CAL, SJ) calculamos los siguientes datos: 
     * age: mean + sd (average of all series)
@@
 ```{r}
-objects_rwl <- c('caL','caH','sjL','sjH')
+objects_rwl <- c('caL','caH','sj')
 
@@
 rwl_by_site %>% 
-  write.csv(file=paste(di, "data/proto_tables/site_dendro.csv", sep=""), row.names = FALSE)
+  write.csv(file=paste(di, "data/proto_tables/site_3_dendro.csv", sep=""), row.names = FALSE)
 
-pander(rwl_by_site, caption='Dendrochronological summary by sites') 
-```
-
-## By Loc
-```{r}
-objects_rwl <- c('ca','sj')
-
-out <- c()
-
-for (i in objects_rwl){
-
-  aux <- get(i)
-  # Stats of singles RWL
-  aux_stats <- rwl.stats(aux)
-  # Add site 
-  aux_stats$loc <- as.factor(i)
-
-  # Compute interseries correlation 
-  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
-  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
-
-  out <- rbind(out, aux_stats)
-  }
-
-rwl_by_loc <- out %>% group_by(loc) %>% 
-  summarize(age_mean = round(mean(year),2), 
-            age_sd = round(sd(year), 2),
-            year_min = min(first),
-            year_max = max(last), 
-            year_range = year_max - year_min + 1, 
-            ar1_mean = round(mean(ar1), 3),
-            ar1_sd = round(sd(ar1), 3),
-            icor_mean = round(mean(interser_cor_val), 3),
-            icor_sd = round(sd(interser_cor_val), 3)) %>% 
-  as.data.frame()
-
-rwl_by_loc %>% 
-  write.csv(file=paste(di, "data/proto_tables/loc_dendro.csv", sep=""), row.names = FALSE)
-
-pander(rwl_by_loc, caption='Dendrochronological summary by sites') 
+pander(rwl_by_site, caption='Dendrochronological summary by sites (3)') 
 ```
@@
 
-* Agregar valores medios de RW por loc (obtenemos sj_tree / ca_tree) o por site (sjL_tree, sjH_tree / caL_tree, caH_tree) 
+* Agregar valores medios de RW por site (obtenemos sj_tree / caL_tree, caH_tree) 
 * ver fun rw_byTree o utilizar treeMean (dplR)
@@
 sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
-ca_tree <- rw_byTree(ca, snc =c(2,2,2), locname = 'CA')
-
-sjL_tree <- rw_byTree(sjL, snc =c(2,2,2), locname = 'SJ')
-sjH_tree <- rw_byTree(sjH, snc =c(2,2,2), locname = 'SJ')
 caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
@@
 ```{r}
-# # function to subset averaged RW 
-# subsetRW <- function(rwdf, codes){ 
-#   require(dplyr)
-#   aux_codes <- paste0('rw_', codes)
-#   mysubset <- rwdf %>% dplyr::select(one_of(aux_codes), year)
-#   mysubset
-#   }
-
-# diam <- ft %>% group_by(site) %>% do(data = (.)) %>% 
-#              dplyr::select(data) %>% 
-#              lapply(function(x) {(x)})
-
 diam <- ft %>%
@@
 
-d_caH <- diam$CA_High[,c('id','diameter')]
-d_caL <- diam$CA_Low[,c('id','diameter')]
-d_sjH <- diam$SJ_High[,c('id','diameter')]
-d_sjL <- diam$SJ_Low[,c('id','diameter')]
-
-diamloc <- ft %>% 
-  mutate(diameter = dn_mm, 
-         id = id_focalLoc) %>%
-  dplyr::select(id, diameter, loc) %>% split(.$loc)
-
-d_ca <- diamloc$CA[,c('id','diameter')]
-d_sj <- diamloc$SJ[,c('id','diameter')]
+d_caH <- diam$CAH[,c('id','diameter')]
+d_caL <- diam$CAL[,c('id','diameter')]
+d_sj <- diam$SJ[,c('id','diameter')]
 ```
 
-# Cómputo del BAI por loc y/o site
+# Cómputo del BAI por site
 
@@
 bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
-bai_ca <- bai_piovesan(rwdf = ca_tree, diam_df = d_ca)
-
-bai_sjH <- bai_piovesan(rwdf = sjH_tree, diam_df = d_sjH)
-bai_sjL <- bai_piovesan(rwdf = sjL_tree, diam_df = d_sjL)
 bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
@@
 
-# bai_test <- round(bai.out(rwl = sj_tree, diam = d_sj), 4)
-# test bai.out and bai_piovesan  
-# unique(round(bai_sj$SJ03,4) == round(bai_test$SJ03,4))
-
 # Set class to bai object 
 # Esto es para que funcionen algunas otras funciones de dplR 
-bais <- c('bai_sj', 'bai_ca', 'bai_sjH', 'bai_sjL', 'bai_caH', 'bai_caL')
+bais <- c('bai_sj', 'bai_caH', 'bai_caL')
 
@@
 }
-
 ```

### 2017-06-01 script/R/computeGC.R
update function include ngc

@@ -37,10 +37,12 @@ computeGC <- function(dfrwl, ws, prefijo){
       mean_m2 <- mean(m2[,1])
       mean_m1 <- mean(m1[,1])
       # GC 
-      gc = ((mean_m2 - mean_m1) / mean_m1) * 100 
+      pgc = ((mean_m2 - mean_m1) / mean_m1) * 100 
+      ngc = ((mean_m1 - mean_m2) / mean_m2) * 100 
       
       # out 
-      gc_aux <- data.frame(gc = gc, year = y, name_serie = ns)
+      gc_aux <- data.frame(pgc = pgc, ngc =ngc,
+                           year = y, name_serie = ns)
       gc_aux$tree <- as.numeric(stringr::str_replace(gc_aux$name_serie, prefijo, ""))
       
       aux_serie <- rbind(aux_serie, gc_aux)

### 2017-06-01 script/R/computeGC.R
update function

@@ -43,6 +43,8 @@ computeGC <- function(dfrwl, ws, prefijo){
       # out 
       gc_aux <- data.frame(pgc = pgc, ngc =ngc,
                            year = y, name_serie = ns)
+      
+      gc_aux$gc <- ifelse(gc_aux$pgc > 0, gc_aux$pgc, (-1)*gc_aux$ngc)
       gc_aux$tree <- as.numeric(stringr::str_replace(gc_aux$name_serie, prefijo, ""))
       
       aux_serie <- rbind(aux_serie, gc_aux)

### 2017-06-01 analysis/analysis_disturbance_chronos.Rmd
update chronos

@@
 source(paste0(di, 'script/R/bai_piovesan.R'))
+source(paste0(di, 'script/R/computeGC.R'))
 
@@
 ```
+
+```{r}
+
+gc_caL <- computeGC(caL_tree, ws=10, prefijo = 'CA')
+gc_caH <- computeGC(caH_tree, ws=10, prefijo = 'CA')
+gc_sj <- computeGC(sj_tree, ws=10, prefijo = 'SJ')
+
+	GCper=100*((mean(m2[,1])-mean(m1[,1]))/mean(m1[,1]))  ##PGC
+
+	GCperneg=100*((mean(m1[,1])-mean(m2[,1]))/mean(m2[,1])) ###NGC
+	GCfin=rbind(GCfin, c(i, j, GCper, GCperneg))
+	names(GCfin)=c("tree2", "year", "GCper", "GCperneg")
+
+	GCper = 100
+
+GCfin2$GCfinal <- ifelse(GCfin2$GCper>0, GCfin2$GCper, -1*GCfin2$GCperneg) ###pongo para los positivos el PGC, para los negativos el NGC
+
+GCfin2<<-GCfin2
+
+xx <- getGC(caL_tree, ws=10, prefijo='CA')
+
+GCfin2 <- data.frame(GCper = GCper, GCperneg = GCperneg)
+
+AGC2=function(RWL){
+
+  GC=as.data.frame(NULL)
+
+  #
+  for (n in 1:ncol(RWL)){
+    series <- as.data.frame(RWL[, n])
+    series$year=as.numeric(row.names(RWL))
+    series=subset(series, !is.na(series[1]))
+    tree=as.numeric(substr(colnames(RWL)[n],3,4))
+    tree2=as.numeric(substr(colnames(RWL)[n], 5,5))
+    if(!is.na(tree2)){
+      tree3=as.numeric(substr(colnames(RWL[n]),3,5))}
+    else {tree3=tree}
+    series$tree=as.numeric(tree3)
+
+    GC=rbind(GC, series)
+    }
+  names(GC)[1]="RW"
+  GC2=aggregate(GC[,1], by=list(year=GC$year,tree=GC$tree), mean)
+  names(GC2)[3]="RW"
+  GC2=GC2[order(GC2$tree, GC2$year),]
+
+  GC2$tree2=1
+  for (i in 2:nrow(GC2)){
+    if(GC2$year[i]<GC2$year[i-1]) GC2$tree2[i]=1+GC2$tree2[i-1] else GC2$tree2[i]=GC2$tree2[i-1]}
+  GC2<<-GC2
+  }
+
+AGC3=function(RWL, n)
+{
+GCfin2=data.frame(NULL)
+
+for (i in 1:max(RWL[,4])) {
+
+  series=subset(RWL, tree2==i)
+
+  GCfin=data.frame(NULL)
+
+  for (j in (min(series$year)+n):(max(series$year)-n)){
+
+    M2=subset(series, series$year>=j & series$year<(j+n-1))
+    M1=subset(series, series$year<(j-1) & series$year>=(j-n))
+
+	GCper=100*((mean(M2[,3])-mean(M1[,3]))/mean(M1[,3]))  ##PGC
+	GCperneg=100*((mean(M1[,3])-mean(M2[,3]))/mean(M2[,3])) ###NGC
+
+	GCfin=rbind(GCfin, c(i, j, GCper, GCperneg))
+	names(GCfin)=c("tree2", "year", "GCper", "GCperneg")
+}
+GCfin2=rbind(GCfin2, GCfin)
+}
+GCfin2$GCfinal=
+  ifelse(GCfin2$GCper>0, GCfin2$GCper, -1*GCfin2$GCperneg) ###pongo para los positivos el PGC, para los negativos el NGC
+GCfin2<<-GCfin2
+}
+
+rr <- AGC2(sj)
+rr2 <- AGC3(rr, 10)
+
+RWL <- rr
+n <- 10
+1:max(RWL[,4])
+i <- 1 
+series=subset(RWL, tree2==i)
+
+rangej <- (min(series$year)+n):(max(series$year)-n)
+j <- rangej[1]
+
+M2=subset(series, series$year>=j & series$year<(j+n-1))
+M2b =subset(series, series$year<(j+n-1))
+
+rr
+
+head(rr)
+
+range <- (min(rr$year) + 10):(max(rr$year) - 10)
+
+series <- subset(rr, tree2 == 1)
+M2 = subset(series, series$year>=j & series$year<(j+n-1))
+
+j<- 1931
+n <- 10
 

### 2017-06-02 script/R/computeGC.R
update median computation

@@ -36,15 +36,28 @@ computeGC <- function(dfrwl, ws, prefijo){
       # mean values
       mean_m2 <- mean(m2[,1])
       mean_m1 <- mean(m1[,1])
+      
+      # median 
+      median_m2 <- median(m2[,1])
+      median_m1 <- median(m1[,1])
+      
       # GC 
-      pgc = ((mean_m2 - mean_m1) / mean_m1) * 100 
-      ngc = ((mean_m1 - mean_m2) / mean_m2) * 100 
+      pgc_m = ((mean_m2 - mean_m1) / mean_m1) * 100 
+      ngc_m = ((mean_m1 - mean_m2) / mean_m2) * 100 
+      
+      # 
+      # GC 
+      pgc_md = ((median_m2 - median_m1) / median_m1) * 100 
+      ngc_md = ((median_m1 - median_m2) / median_m2) * 100 
       
       # out 
-      gc_aux <- data.frame(pgc = pgc, ngc =ngc,
+      gc_aux <- data.frame(pgc = pgc_m, ngc = ngc_m,
+                           pgc_md = pgc_md, ngc_md = ngc_md,
                            year = y, name_serie = ns)
       
-      gc_aux$gc <- ifelse(gc_aux$pgc > 0, gc_aux$pgc, (-1)*gc_aux$ngc)
+      gc_aux$gc_mean <- ifelse(gc_aux$pgc > 0, gc_aux$pgc, (-1)*gc_aux$ngc)
+      gc_aux$gc_median <- ifelse(gc_aux$pgc_md > 0, gc_aux$pgc_md, (-1)*gc_aux$ngc_md)
+      
       gc_aux$tree <- as.numeric(stringr::str_replace(gc_aux$name_serie, prefijo, ""))
       
       aux_serie <- rbind(aux_serie, gc_aux)

### 2017-06-02 analysis/analysis_disturbance_chronos.Rmd
update disturbance chronos

@@
 ```{r}
-
 gc_caL <- computeGC(caL_tree, ws=10, prefijo = 'CA')
@@
 gc_sj <- computeGC(sj_tree, ws=10, prefijo = 'SJ')
+
+# mean by year 
+# dataframe for plot 
+gc_todos <- c()
+
+datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
+
+for (d in datasetes){ 
+  # get df
+  df <- get(d)
+
+  aux_df <- df %>%
+    group_by(year) %>% 
+    summarise(ntrees = length(tree),
+              gc_m = mean(gc_mean),
+              gc_m_sd = sd(gc_mean),
+              gc_m_se = sd(gc_mean) / sqrt(n()), 
+              gc_md = mean(gc_median),
+              gc_md_sd = sd(gc_median),
+              gc_md_se = sd(gc_median) / sqrt(n())) %>%  
+    mutate(site=paste(d))
+
+  gc_todos <- rbind(gc_todos, aux_df)
+}
+
+gc_todos %>% ggplot(aes(x=year, y=gc_m)) + 
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  # geom_point() +
+  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray') +
+  facet_wrap(~site, nrow = 3, scales = 'free_y',
+             labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low',
+                                 'gc_sj' = 'San Juan'))) +
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
+  ylab('Mean GC') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
+
+## Númber of trees 
+GC_trees <- function(gcdf, treshold){
+
+  th <- as.numeric(treshold)
+
+  # Get total tree by year 
+  n_tree <- gcdf %>% group_by(year) %>% summarise(total_tree = length(ntrees))
+
+  # Get n tree with GC positve upper treshold 
+  n_pos <- gcdf %>% filter(gc_m > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees))
+
+  # Get n tree with GC negative lower treshold 
+  n_neg <- gcdf %>% filter(gc_m < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees))
+
+  # Join and compute percentage 
+  outgc <- n_tree %>% 
+    left_join(n_pos, by='year') %>% 
+    left_join(n_neg, by='year') %>% 
+    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
+         per_neg = round((n_neg / total_tree) * 100, 2))
+
+  return(outgc)
+}
+
+treesth <- GC_trees(gcdf = gc_todos, treshold = 50)
+
+t_caL <- gc_todos %>% filter(site == 'gc_caL') %>% 
+  GC_trees(treshold = 50) %>% data.frame()
+t_caL$site <- 'gc_caL'
+
+t_caH <- gc_todos %>% filter(site == 'gc_caH') %>% 
+  GC_trees(treshold = 50) %>% data.frame()
+t_caH$site <- 'gc_caH'
+
+t_sj <- gc_todos %>% filter(site == 'gc_sj') %>% 
+  GC_trees(treshold = 50) %>% data.frame()
+t_sj$site <- 'gc_sj'
+
+per_t <- rbind(t_caH, t_caL, t_sj)
+
+per_t %>% ggplot(aes(x=year, y=gc_m)) + 
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  # geom_point() +
+  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray') +
+  facet_wrap(~site, nrow = 3, scales = 'free_y',
+             labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low',
+                                 'gc_sj' = 'San Juan'))) +
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
+  ylab('Mean GC') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
+
+# EOBS data climate
+
+xx %>% group_by(year, gc_mean) %>% count(tree) %>% filter(gc_mean > 50)
+
+p %>% filter(site=='sj') %>% 
+  ggplot(aes(x=year, y=mgc_md)) + 
+  geom_bar(stat='identity') + theme_bw() + 
+  # facet_wrap(~site, nrow = 3, scales = 'free_y') +
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red')
+
+gc_caH %>% 
+  group_by(year) %>% 
+  summarise(nt = length(tree),
+            mgc = mean(gc)) %>% 
+  ggplot(aes(x=year, y=mgc)) + geom_bar(stat='identity') + theme_bw() + 
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red')
+
+gc_sj %>% 
+  group_by(year) %>% 
+  summarise(nt = length(tree),
+            mgc = mean(gc)) %>% 
+  ggplot(aes(x=year, y=mgc)) + geom_bar(stat='identity') + theme_bw() + 
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red')
+
+# + ylim(c(-300,300))
 

### 2017-06-05 analysis/analysis_disturbance_chronos.Rmd
updating

@@
 
-# sjL <- sj[, c("SJ0101","SJ0102","SJ0201","SJ0202","SJ0301","SJ0302","SJ0401","SJ0402","SJ0501","SJ0502",
-#               "SJ0601","SJ0602","SJ0603","SJ0701","SJ0702","SJ0801","SJ0802","SJ0901","SJ0902","SJ0903",
-#               "SJ1001","SJ1002","SJ1003")]
-# sjH <- sj[, c("SJ1101","SJ1102","SJ1103","SJ1201","SJ1202","SJ1203","SJ1301","SJ1302","SJ1303","SJ1401",
-#               "SJ1402","SJ1501","SJ1502","SJ1601","SJ1602","SJ1701","SJ1702","SJ1703","SJ1801","SJ1802",
-#               "SJ1901","SJ1902","SJ2001","SJ2003","SJ2002")]
 ```
@@
 
+# Disturbance chronologies
+
+* calcular chronologies con media y mediana. 
+* ver [./analysis/analysis_disturbance_chronos.md](/analysis/analysis_disturbance_chronos.md))
+
 ```{r}
@@
     group_by(year) %>% 
-    summarise(ntrees = length(tree),
+    summarise(ntrees = length(ntrees),
               gc_m = mean(gc_mean),
@@
                                         fill='white'))
+```
 
 ## Númber of trees 
-GC_trees <- function(gcdf, treshold){
+```{r}
+
+computeGCtrees <- function(gcdf, treshold){
 
@@
   # Get n tree with GC positve upper treshold 
-  n_pos <- gcdf %>% filter(gc_m > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees))
+  n_pos <- gcdf %>% dplyr::filter(gc_mean > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees))
 
   # Get n tree with GC negative lower treshold 
-  n_neg <- gcdf %>% filter(gc_m < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees))
+  n_neg <- gcdf %>% dplyr::filter(gc_mean < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees))
 
@@
 
-treesth <- GC_trees(gcdf = gc_todos, treshold = 50)
+gc_trees<- c()
+
+datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
+
+for (d in datasetes){ 
+  # get df
+  df <- get(d)
+
+  au <- df %>% computeGCtrees(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
+  gc_trees <- rbind(gc_trees, au)
+}
+
+xx <- gc_trees %>% 
+  mutate(per_negn = (per_neg)*(-1))
+
+gc_trees %>% ggplot() +
+  geom_bar(aes(x=year, y=per_pos), fill='blue', stat = "identity", position = "dodge") +
+  geom_bar(aes(x=year, y=(-1)*per_neg), fill='blue',  stat = "identity", position = "dodge") + 
+  facet_wrap(~site, nrow = 3) +
+  theme_bw() + ylab('%trees GC > 50 %') +
+  theme(panel.grid = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
+
+gc_trees %>% ggplot() + 
+  geom_point(aes(x=year, y=per_pos), shape=17) + 
+  geom_point(aes(x=year, y=per_neg), shape=6) +
+  facet_wrap(~site, nrow = 3)
+
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  # geom_point() +
+  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray') +
+  facet_wrap(~site, nrow = 3, scales = 'free_y',
+             labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low',
+                                 'gc_sj' = 'San Juan'))) +
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
+  ylab('Mean GC') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
+
+computeGCtrees
 

### 2017-06-05 script/R/computeGC.R

@@ -43,22 +43,22 @@ computeGC <- function(dfrwl, ws, prefijo){
       
       # GC 
       pgc_m = ((mean_m2 - mean_m1) / mean_m1) * 100 
-      ngc_m = ((mean_m1 - mean_m2) / mean_m2) * 100 
+      ngc_m = ((mean_m2 - mean_m1) / mean_m2) * 100 
       
       # 
       # GC 
       pgc_md = ((median_m2 - median_m1) / median_m1) * 100 
-      ngc_md = ((median_m1 - median_m2) / median_m2) * 100 
+      ngc_md = ((median_m2 - median_m1) / median_m2) * 100 
       
       # out 
       gc_aux <- data.frame(pgc = pgc_m, ngc = ngc_m,
                            pgc_md = pgc_md, ngc_md = ngc_md,
                            year = y, name_serie = ns)
       
-      gc_aux$gc_mean <- ifelse(gc_aux$pgc > 0, gc_aux$pgc, (-1)*gc_aux$ngc)
-      gc_aux$gc_median <- ifelse(gc_aux$pgc_md > 0, gc_aux$pgc_md, (-1)*gc_aux$ngc_md)
+      gc_aux$gc_mean <- ifelse(gc_aux$pgc > 0, gc_aux$pgc, gc_aux$ngc)
+      gc_aux$gc_median <- ifelse(gc_aux$pgc_md > 0, gc_aux$pgc_md, gc_aux$ngc_md)
       
-      gc_aux$tree <- as.numeric(stringr::str_replace(gc_aux$name_serie, prefijo, ""))
+      gc_aux$ntrees <- as.numeric(stringr::str_replace(gc_aux$name_serie, prefijo, ""))
       
       aux_serie <- rbind(aux_serie, gc_aux)
     }

### 2017-06-05 man/notes_dendro_analysis.Rmd
update

@@
 
+## Disturbance chronologies
+
+Se han construido cronologías de perturbaciones siguiendo la aproximación de aproximación de [@Nowacki1997](varios ejemplos se pueden ver en [@GeaIzquierdo2014, @Dorado2017]).
+Utilizamos la función `computeGC` (código aquí: [./script/R/computeGC.R](/script/R/gea/computeGC.R)).
+En concreto hemos llevado a cabo lo siguiente:
+
+* Computamos medias (o medianas) de RW series (por árbol) en una ventana de 10 años.
+G.Gea utiliza medias, aunque en otros trabajos utilizan medianas (las medianas son estimadores mas robustos de la tendencia central que la media [@Camarero2011]).
+Las medianas son menos sensibles a los valores extremos (son mas conservadores.
+G.
+Gea *com.
+per.*).
+No obstante la función calcula ambos (medias y medianas)
+* La formula es:
+
+    * PGC: [(M2 - M1)/M1]*100 
+    * NGC: [(M2 - M1)/M2]*100
+
+- Puedes probar los pointer years y lo de Lloret, efectivamente la idea era comparar los años de sequía de las cronos con tus análisis en imágenes.
+Lo bueno es que en las cronos puedes comparar con años secos antes de 2005, ya sabes.
+
+- Me resulta raro que no se parezcan en nada en la alta frecuencia (lag 0) ninguna de las cronos, pero bueno, si te sale así es cuestión de pensar bien si hay información útil en ese gráfico.
+La baja frecuencia en SJ y CL es opuesta, no sé, habría que discutir si tiene sentido presentar esa figura.
+
+- Para los GC: yo siempre he usado medias porque creo recordar que es lo que había en el artículo original.
+Hay gente que usa medianas… usa el que creas más conveniente, la técnica tiene sus debilidades (por ejemplo, puede haber “liberaciones” en cronos muy sensibles durante períodos muy húmedos, por ejemplo, y a efectos de crecimiento parecen lo mismo que una liberación por competencia).
+Las medianas son menos sensibles a valores extremos, en este sentido pueden ser más conservadoras, pero bueno, presentando intervalos de confianza y/o mezclando gráficos de medias con gráficos de % de pies que presentan liberaciones/supresiones se debería llegar a los mismos resultados.
+
+- Las NGC: lo suyo es (M2-M1)/M2, es decir, poner en el denominador siempre el período teóricamente más pequeño.
+De algún modo así se aumenta la probabilidad de que salga significativo, pero bueno.
+Creo recordar que en los que yo usé de referencia usaban esa fórmula.
+Son pequeños matices, y como hemos hablado, la técnica tiene sus limitaciones, así que de una u otra manera, cuando son claras deberían salir igual (picos en esa época).
+Pero vamos, yo he usado la última que dices tú.
+
+- se contabilizan ese año los que superen el umbral, efectivamente, y para los PGC se computan sólo los positivos, y para el NGC sólo los negativos.
+Es decir, independientemente (entiendo, así pensándolo rápido que hace un tiempo que no hago un cálculo de éstos) por un lado se calculan los PGC y los NGC, y luego se componen (los cálculos independientes) en el gráfico.
+- El gráfico de SJ me parece coherente con el gráfico de crecimiento medio.
+Se ve la liberación en los años 40-50 que debió ser de la última corta para leñas.
+Queda bonito y tiene sentido, lo que no sé es cuántos árboles tienes antes y después del pico, pero es coherente con la historia que estuvimos discutiendo.
+En cuanto a Cáñar, habrá que pensar bien qué son esas supresiones: podrían ser épocas de estrés climático, o épocas en las que había más competencia (entonces deberían salir luego liberaciones… pero todo esto es especulativo).
+En cualquier caso, la de SJ me parece interesante, bonita y coherente con lo que estuvimos discutiendo.
+
+Un abrazo, buena semana.
+
 --------------------------

### 2017-06-06 analysis/analysis_disturbance_chronos.Rmd
update disturbance chronologies

--- 2017-06-01 analysis/analysis_disturbance_chronos.Rmd
@@
 library('boot')
+library('gtable')
+library('grid')
 ```
@@
 source(paste0(di, 'script/R/computeGC.R'))
-
-source(paste0(di, 'script/R/chrono_bai.R'))
-source(paste0(di, 'script/R/suaviza_cronos.R'))
-source(paste0(di, 'script/R/correlaBootIC.R'))
 ```
@@
 
-# Cómputo del BAI por site
+# Disturbance chronologies
 
-* He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]. Es similar a `bai.out`
+* calcular chronologies con media y mediana. 
+* ver [./analysis/analysis_disturbance_chronos.md](/analysis/analysis_disturbance_chronos.md))
 
 ```{r}
-bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
-bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
-bai_caL <- bai_piovesan(rwdf = caL_tree, diam_df = d_caL)
-
-# Set class to bai object 
-# Esto es para que funcionen algunas otras funciones de dplR 
-bais <- c('bai_sj', 'bai_caH', 'bai_caL')
-
-for (i in bais){
-  aux <- get(i)
-
-  class(aux) <- c('rwl', 'data.frame')
-
-  assign(i, aux)
-}
-```
-
-```{r}
-
 gc_caL <- computeGC(caL_tree, ws=10, prefijo = 'CA')
@@
 gc_sj <- computeGC(sj_tree, ws=10, prefijo = 'SJ')
+```
 
-	GCper=100*((mean(m2[,1])-mean(m1[,1]))/mean(m1[,1]))  ##PGC
+### Plot 
+```{r}
+# Plot mean value by year 
+# dataframe for plot 
+gc_todos <- c()
 
-	GCperneg=100*((mean(m1[,1])-mean(m2[,1]))/mean(m2[,1])) ###NGC
-	GCfin=rbind(GCfin, c(i, j, GCper, GCperneg))
-	names(GCfin)=c("tree2", "year", "GCper", "GCperneg")
+datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
 
-	GCper = 100
+for (d in datasetes){ 
+  # get df
+  df <- get(d)
 
-GCfin2$GCfinal <- ifelse(GCfin2$GCper>0, GCfin2$GCper, -1*GCfin2$GCperneg) ###pongo para los positivos el PGC, para los negativos el NGC
+  aux_df <- df %>%
+    group_by(year) %>% 
+    summarise(ntrees = length(ntrees),
+              gc_m = mean(gc_mean),
+              gc_m_sd = sd(gc_mean),
+              gc_m_se = sd(gc_mean) / sqrt(n()), 
+              gc_md = mean(gc_median),
+              gc_md_sd = sd(gc_median),
+              gc_md_se = sd(gc_median) / sqrt(n())) %>%  
+    mutate(site=paste(d))
 
-GCfin2<<-GCfin2
-
-xx <- getGC(caL_tree, ws=10, prefijo='CA')
-
-GCfin2 <- data.frame(GCper = GCper, GCperneg = GCperneg)
-
-AGC2=function(RWL){
-
-  GC=as.data.frame(NULL)
-
-  #
-  for (n in 1:ncol(RWL)){
-    series <- as.data.frame(RWL[, n])
-    series$year=as.numeric(row.names(RWL))
-    series=subset(series, !is.na(series[1]))
-    tree=as.numeric(substr(colnames(RWL)[n],3,4))
-    tree2=as.numeric(substr(colnames(RWL)[n], 5,5))
-    if(!is.na(tree2)){
-      tree3=as.numeric(substr(colnames(RWL[n]),3,5))}
-    else {tree3=tree}
-    series$tree=as.numeric(tree3)
-
-    GC=rbind(GC, series)
-    }
-  names(GC)[1]="RW"
-  GC2=aggregate(GC[,1], by=list(year=GC$year,tree=GC$tree), mean)
-  names(GC2)[3]="RW"
-  GC2=GC2[order(GC2$tree, GC2$year),]
-
-  GC2$tree2=1
-  for (i in 2:nrow(GC2)){
-    if(GC2$year[i]<GC2$year[i-1]) GC2$tree2[i]=1+GC2$tree2[i-1] else GC2$tree2[i]=GC2$tree2[i-1]}
-  GC2<<-GC2
+  gc_todos <- rbind(gc_todos, aux_df)
 }
 
-AGC3=function(RWL, n)
-{
-GCfin2=data.frame(NULL)
+gc_todos %>% ggplot(aes(x=year, y=gc_m)) + 
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  # geom_point() +
+  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray') +
+  facet_wrap(~site, nrow = 3, scales = 'free_y',
+             labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low',
+                                 'gc_sj' = 'San Juan'))) +
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
+  ylab('Mean GC') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
+```
 
-for (i in 1:max(RWL[,4])) {
+# Plot (correcting the scale)
+```{r}
 
-  series=subset(RWL, tree2==i)
+# Create a transformation scale 
+# https://stackoverflow.com/questions/37446064/i-need-ggplot-scale-x-log10-to-give-me-both-negative-and-positive-numbers-as-o
 
-  GCfin=data.frame(NULL)
+pos_neg_log <- scales::trans_new('signed_log', 
+                                 transform = function(x) sign(x)*log(abs(x)),
+                                 inverse=function(x) sign(x)*exp(abs(x)))
 
-  for (j in (min(series$year)+n):(max(series$year)-n)){
+# solution for arrange
+# http://felixfan.github.io/stacking-plots-same-x/
 
-    M2=subset(series, series$year>=j & series$year<(j+n-1))
-    M1=subset(series, series$year<(j-1) & series$year>=(j-n))
+a <- gc_todos %>% 
+  filter(site != 'gc_sj') %>% 
+  ggplot(aes(x=year, y=gc_m)) + 
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray', width=.3, size=.2) +
+  # scale_y_continuous(trans = pos_neg_log) +
+  facet_wrap(~site, nrow = 2,
+             labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low'))) +
+                                # 'gc_sj' = 'San Juan'))) + 
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
+  ylab('Mean GC') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'),
+        axis.title.x = element_blank(),
+        axis.text.x = element_blank())
 
-	GCper=100*((mean(M2[,3])-mean(M1[,3]))/mean(M1[,3]))  ##PGC
-	GCperneg=100*((mean(M1[,3])-mean(M2[,3]))/mean(M2[,3])) ###NGC
+b <- gc_todos %>% 
+  filter(site == 'gc_sj') %>% 
+  ggplot(aes(x=year, y=gc_m)) + 
+  xlim(layer_scales(a)$x$range$range[1],layer_scales(a)$x$range$range[2]) +
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray', width=.3, size=.2) +
+  scale_y_continuous(trans = pos_neg_log, 
+                     breaks=c(-1000, -100, 0, 100, 500)) +
+  facet_wrap(~site, nrow = 2,
+             labeller=as_labeller(c('gc_sj' = 'San Juan'))) + 
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
+  ylab('') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
 
-	GCfin=rbind(GCfin, c(i, j, GCper, GCperneg))
-	names(GCfin)=c("tree2", "year", "GCper", "GCperneg")
-}
-GCfin2=rbind(GCfin2, GCfin)
-}
-GCfin2$GCfinal=
-  ifelse(GCfin2$GCper>0, GCfin2$GCper, -1*GCfin2$GCperneg) ###pongo para los positivos el PGC, para los negativos el NGC
-GCfin2<<-GCfin2
+grid.newpage()
+grid.draw(rbind(ggplotGrob(a), ggplotGrob(b), size='last'))
+```
+
+## Númber of trees 
+```{r}
+
+computeGCtrees <- function(gcdf, treshold){
+
+  th <- as.numeric(treshold)
+
+  # Get total tree by year 
+  n_tree <- gcdf %>% group_by(year) %>% summarise(total_tree = length(ntrees))
+
+  # Get n tree with GC positve upper treshold 
+  n_pos <- gcdf %>% dplyr::filter(gc_mean > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees))
+
+  # Get n tree with GC negative lower treshold 
+  n_neg <- gcdf %>% dplyr::filter(gc_mean < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees))
+
+  # Join and compute percentage 
+  outgc <- n_tree %>% 
+    left_join(n_pos, by='year') %>% 
+    left_join(n_neg, by='year') %>% 
+    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
+         per_neg = round((n_neg / total_tree) * 100, 2))
+
+  return(outgc)
 }
 
-rr <- AGC2(sj)
-rr2 <- AGC3(rr, 10)
+gc_trees<- c()
 
-RWL <- rr
-n <- 10
-1:max(RWL[,4])
-i <- 1 
-series=subset(RWL, tree2==i)
+datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
 
-rangej <- (min(series$year)+n):(max(series$year)-n)
-j <- rangej[1]
+for (d in datasetes){ 
+  # get df
+  df <- get(d)
 
-M2=subset(series, series$year>=j & series$year<(j+n-1))
-M2b =subset(series, series$year<(j+n-1))
+  au <- df %>% computeGCtrees(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
+  gc_trees <- rbind(gc_trees, au)
+}
 
-rr
+# secondary axis
+# https://whatalnk.github.io/r-tips/ggplot2-secondary-y-axis.nb.html
 
-head(rr)
-
-range <- (min(rr$year) + 10):(max(rr$year) - 10)
-
-series <- subset(rr, tree2 == 1)
-M2 = subset(series, series$year>=j & series$year<(j+n-1))
-
-j<- 1931
-n <- 10
-
-# Cronologías by site / loc
-
-### Mean vs. Tukey’s Biweight Robust Mean
-
-* Computo la cronologia usando la función `chron`. 
-* Para SJ hag una comparación de la cronologia construida con biweight (*treatment = biweigth*) o construida con la media aritmética (*treatment = No biweigth*) (calculada con la función `chron`)
-* Idem para CA, pero en CA si separo entre CA high y CA low
-
-```{r}
-# Robust mean biweith or not
-cro_sjb <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
-cro_sjnotb <- chron(bai_sj, prefix = "SJ", biweight = FALSE)
-
-cro_b <- cro_sjb %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
-
-cro_notb <- cro_sjnotb %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
-
-aux_cro <- rbind(cro_b, cro_notb)
-
-aux_cro %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
-  geom_point() + geom_line() + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
-  theme_bw() + theme(panel.grid.major = element_blank()) +
-  ggtitle('SJ chronology built using Biweight or Mean')
-```
-
-```{r}
-# Robust mean biweith or not
-cro_cab_h <- chron(bai_caH, prefix = "CA", biweight = TRUE)
-cro_canotb_h <- chron(bai_caH, prefix = "CA", biweight = FALSE)
-
-cro_cab_h_aux <- cro_cab_h %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_canotb_h_aux <- cro_canotb_h  %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
-
-aux_croca_h <- rbind(cro_cab_h_aux, cro_canotb_h_aux)
-aux_croca_h$site <- 'CA High'
-
-cro_cab_l <- chron(bai_caL, prefix = "CA", biweight = TRUE)
-cro_canotb_l <- chron(bai_caL, prefix = "CA", biweight = FALSE)
-
-cro_cab_l_aux <- cro_cab_l %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_canotb_l_aux <- cro_canotb_l  %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
-
-aux_croca_l <- rbind(cro_cab_l_aux, cro_canotb_l_aux)
-aux_croca_l$site <- 'CA Low'
-
-aux_croca <- rbind(aux_croca_h, aux_croca_l)
-
-aux_croca %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
-  geom_point() + geom_line() + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
-  theme_bw() + theme(panel.grid.major = element_blank()) + 
-  facet_wrap(~site, nrow = 2) +
-    ggtitle('CA chronology built using Biweight or Mean')
-```
-
-## Compute Chrono 
-### San Juan chronologies by site
-
-```{r}
-# San Juan 
-cro_sjH <- chrono_bai(bai_sjH)
-cro_sjH$site <- 'SJ_High'
-
-cro_sjL <- chrono_bai(bai_sjL)
-cro_sjL$site <- 'SJ_Low'
-
-cro_sj<- rbind(cro_sjH, cro_sjL)
-cro_sj$loc <- 'Norte (San Juan)'
-
-cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
-  theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
-```
-
-### Cañar chronologies by site
-
-```{r}
-# Canar 
-cro_caH <- chrono_bai(bai_caH)
-cro_caH$site <- 'CA_High'
-
-cro_caL <- chrono_bai(bai_caL)
-cro_caL$site <- 'CA_Low'
-
-cro_ca <- rbind(cro_caH,cro_caL)
-cro_ca$loc <- 'Sur (Canar)'
-
-cro_ca %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
-  theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
-```
-
-```{r}
-# Combine and plot 
-cro_sites <- rbind(cro_ca, cro_sj)
-
-cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + 
-  theme(panel.grid = element_blank()) + facet_wrap(~loc, nrow = 2) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-```
-
-# Correlation between chronos
-
-To compute similarity between chronos we use @Dorado2017 approach.
-
-* Para cada chrono se realiza un suavizado, utilizando diferente tamaño de ventana (1 a 40 years). Así evaluamos la sincronía entre dos series, tanto a nivel interanual como a largo plazo. 
-* Luego calculamos la correlación entre series para cada tamaño de ventana.
-Usamos correlaciones de Pearson o Sperman entre las series (es decir entre las zonas comunes de las cronologías).
-Ojo con la autocorrelación.
-Para ello puedo utilizar `interseries.cor()` del pkg `dplR`; o utilizar `cor.test()` (`stats`).
-El primero está basado en el segundo, así que utilizo el segundo.
-* Obtener IC utilizando bootstrap. Varias aproximaciones:
-
-    * Utilizar el pkg `boot` y conseguir IC. 
-    * Aproximación de @Mudelsee2014.
-En su paper habla de correlacion entre dos series y utiliza un programa en Fortran-90.
-He contactado con el autor y no tiene pensado hacerlo en R.
-He intentado utilizar Fortran-90, pero no tengo ni idea.
-Asi que desisto de utilizar esta computación de los IC.
-
-```{r}
-# centred moved average smooth
-suaviza_caH <- suaviza_cronos(cro_caH, win_vector=seq(1,40, by=1), crononame = 'cro_caH')
-suaviza_caL <- suaviza_cronos(cro_caL, win_vector=seq(1,40, by=1), crononame = 'cro_caL')
-suaviza_sjH <- suaviza_cronos(cro_sjH, win_vector=seq(1,40, by=1), crononame = 'cro_sjH')
-suaviza_sjL <- suaviza_cronos(cro_sjL, win_vector=seq(1,40, by=1), crononame = 'cro_sjL')
-```
-
-## between sites similarity (sjH *vs* sjL; caH *vs* caL)
-
-This note is for the plot:
-
-* Correlation among chronologies for the common period of each site in different time-domains after pre-filtering the time-series with increasing size of the moving-average window.
-
-```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-# Metodo correlation
-tc <- c('spearman', 'pearson')
-
-# SJ 
-suaviza_sj <- rbind(suaviza_sjH, suaviza_sjL)
-correla_sj <- correlaBootIC(suaviza_sj, tc, years_suavizado = 40, 
-                            name_comparison = 'sjH-sjL')
-
-# CA 
-suaviza_ca <- rbind(suaviza_caH, suaviza_caL)
-correla_ca <- correlaBootIC(suaviza_ca, tc, years_suavizado = 40, 
-                            name_comparison = 'caH-caL')
-
-correla <- rbind(correla_ca, correla_sj)
-
-correla %>% 
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  facet_wrap(~metodo_correla, ncol=2) +
-  theme(strip.background = element_blank())
+# Plot 
+gc_trees %>% ggplot() +
+  geom_bar(aes(x=year, y=per_pos), fill='blue', colour='blue', stat = "identity", position = "dodge") + 
+  geom_bar(aes(x=year, y=(-1)*per_neg), fill='red', colour='red', stat = "identity", position = "dodge") +
+  facet_wrap(~site, nrow = 3, labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low',
+                                 'gc_sj' = 'San Juan'))) + 
+  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
+  geom_line(aes(x=year, y=total_tree), colour = 'gray') + 
+  theme_bw() + ylab('%trees GC > 50 %') +
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
 
 ```
-
-## between final sites similarity (sj *vs* caH; sj *vs* caL; caH *vs* caL)
-
-```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-# Metodo correlation
-tc <- c('spearman', 'pearson')
-
-# Compute the suaviza_sj desde el principio, sin diferenciar h and l. Para ello la crono la computo de nuevo 
-cro_sjloc <- chrono_bai(bai_sj)
-cro_sjloc$site <- 'SJ'
-suaviza_sjloc <- suaviza_cronos(cro_sjloc, win_vector=seq(1,40, by=1), crononame = 'cro_sj')
-
-# sjcaH 
-suaviza_sjcaH <- rbind(suaviza_sjloc, suaviza_caH)
-correla_sjcaH <- correlaBootIC(suaviza_sjcaH, tc, years_suavizado = 40, 
-                            name_comparison = 'sj-caH')
-
-# sjcaL 
-suaviza_sjcaL <- rbind(suaviza_sjloc, suaviza_caL)
-correla_sjcaL <- correlaBootIC(suaviza_sjcaL, tc, years_suavizado = 40, 
-                            name_comparison = 'sj-caL')
-
-# Join all correlations 
-correla <- rbind(correla_ca, correla_sjcaL, correla_sjcaH)
-
-correla %>% 
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  facet_wrap(~metodo_correla, ncol=2) +
-  theme(strip.background = element_blank())
-```
-
----> here
-
-```{r, echo=FALSE, eval=FALSE}
-
-# CODE SNIPPETS 
-  # Loess 
-  l <- loess(b_aux$bai ~ b_aux$year, span = alpha)
-  years_v <- seq(min(b_aux$year), max(b_aux$year), 1)
-
-  l_pred_aux <- predict(l, years_v, se=TRUE)
-  l_pred <- as.data.frame(cbind(loess_fit = l_pred_aux$fit, 
-                loess_se = l_pred_aux$se.fit,
-                year = years_v))
-
-  out <- out %>% inner_join(l_pred, by='year')
-
-  out
-}
-
-# SJ
-
-cro_sj <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
-cro_sj %>% plot(add.spline=TRUE, nyrs=30)
-
-cro_ca <- chron(bai_ca, prefix = "CA", biweight = TRUE)
-cro_ca %>% plot(add.spline=TRUE, nyrs=30)
-
-cro_ca2 <- cro_ca %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('CA')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_sj2 <- cro_sj %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('SJ')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
-
-cro2 <- rbind(cro_ca2, cro_sj2)
-
-cro2 %>% ggplot(aes(x=year, y=bai_mean, colour=treatment)) + 
-  geom_line() + geom_smooth(method= 'loess', span = 0.30, se=FALSE) + theme_bw() + 
-  geom_bar(aes(x=year, stat = (samp.depth*10), colour=treatment))
-
-## San Juan Data
-
-Algunos árboles tienen 3 cores. Sin embargo, no los tres cores llegan hasta la corteza, por lo tanto, no podemos utilizar el diametro para estimar bai. Vamos a realizar lo siguiente:
-
-* Crear dos datasets: 
-    * Dataset con series de datos que llegan hasta corteza (`sj_cor`)
-    * Dataset con series de datos que no llegan hasta corteza (`sj_sin`): se trata de cores que no llegan hasta la corteza.
-
-* Para el cálculo del BAI, en el dataset `sj` utilizamos el diámetro medido en campo. En el caso del dataset `sj_sin` utilizamos la suma de todos los diámetros (:red_circle: `$TODO$ ASK to Guillermo`)
-
-# Get summary ring-width series
-sj_summ <- summary(sj)
-
-# Get names cores with last year different to 2016 
-id_cores_no_bark <- sj_summ %>% 
-  filter(last != 2016) %>% select(series) %>% mutate(series = factor(series))
-
-id_cores_no_bark <- unname(unlist(id_cores_no_bark))
-
-# Subbet datasets 
-sj_cor <- sj[which(! colnames(sj) %in% (id_cores_no_bark))]
-sj_sin <- sj[which(colnames(sj) %in% (id_cores_no_bark))]
-
-Lectura y preparación de datos de diámetro
-
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  select(id_focal, loc, dn_mm, height_cm)
-
-# Select only SJ trees and 
-ft_sj <- ft %>% filter(loc=='SJ')
-
-# get Create a dataframe with core ID
-cores <- data.frame(id_cores=colnames(sj))
-
-# Extract replicate and Tree ID from core ID 
-cores <- cores %>% 
-  mutate(id_focal = as.factor(stringr::str_sub(id_cores, 3,5)),
-         id_replica = stringr::str_sub(id_cores, 6,8))
-
-# Create df with diameter and height for each core ID 
-diam_cores <- cores %>% inner_join(ft_sj, by='id_focal') %>% 
-  select(id_cores, dn_mm)
-
-# remove diamm of cores without bark 
-diam_cores_sj <- diam_cores %>% filter(! id_cores %in% id_cores_no_bark)
-
-rwl.report(sj_cor)
-
-bai_p <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
-bai_d <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
-
-diam_cores_sjsin <- data.frame(id_cores = colnames(sj_sin),
-                               dn_mm = colSums(sj_sin, na.rm = TRUE))
-rownames(diam_cores_sjsin) <- NULL
-
-bai_psin <- bai_piovesan(rwdf = sj_sin, diam_df = diam_cores_sjsin)
-bai_dsinsin <- bai.out(rwl = sj_sin)
-bai_dsin <- bai.out(rwl = sj_sin, diam = diam_cores_sjsin)
-
-set.seed(1234)
-serie <- sample(colnames(b_test_dplR),1)
-
-require(reshape2)
-
-bp <- melt(t(b_test_piovesan))
-names(bp) <- c('cores', 'years', 'bai_p')
-
-bd <- melt(t(b_test_dplR))
-names(bd) <- c('cores', 'years', 'bai_d')
-
-diff <- bp %>% 
-  inner_join(bd, by=c('cores','years')) %>% 
-  mutate(diff = bai_p - bai_d)
-
-ggplot(x, aes(y=bai, x=years, colour=test)) + 
-  geom_point() + 
-  facet_wrap(~cores)
-
-# BAI dplR
-baisj <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
-
-baisjt <- bai.out(rwl = sjt)
-
-# Function to compute BAI (Piovesan et al)
-
-# rwl_file
-myrwl <- sj_cor
-# Diameter df  
-diam_df <- diam_cores_sj
-
-# Core
-id_core <- 'SNA0101'
-
-# Ring-width serie
-myrwl_serie <- myrwl[[id_core]]
-
-# Diameter for core id_core 
-dbh <- subset(diam_df, id_cores == id_core)
-dbh <- dbh[,2]
-
-myrwl_serie2 <- na.omit(myrwl_serie)
-cum
-
-r0 <- dbh/2 - c(0,cumsum(rev(myrwl_serie2)))
-
-bai <- (pi/4)*(dbh^2 - ((dbh - 2*myrwl_serie)^2))
-
-s <- myrwl_serie[94:96]
-
-d <- sum(s)*2
-
-########## test Piovesan
-# rwl_file
-myrwl <- baisj
-
-#id_core
-
-# Extract rwl serie 
-myrwl_i <- myrwl[[1]]
-
-dbh <- 372.4226
-
-# Year
-n.vec <- seq_len(nrow(myrwl))
-
-# Series
-seq_len(ncol(myrwl))
-
-# Select a ring-width series
-dat <- myrwl[[1]] 
-# Remove NA
-dat2 <- na.omit(dat)
-
-ggplot(s1, aes(y=SNA0101, x=row.names(s1))) + 
-  geom_point() + 
-  geom_line(b1, aes(y=SNA0101, x=row.names(b1)))
-
-ggplot(s1, aes(y=SNA0101, x=row.names(b1), group=1)) + geom_line()
-
-x <- c(1,2,2,3,1,3,4,5)
-
-xx <- rollmean(x=x, k=3, align = 'center', na.pad = T)
-
-# tes plots
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-carwi <- detrend(rwl = ca_tree, method = "ModNegExp")
-ca_prwi <- data.frame(rw = rowMeans(carwi, na.rm=TRUE))
-ca_prwi <- ca_prwi %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-xx <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-plot(xx$rw~rownames(xx))
-
-sjcrn <- chron(sjrwi, prefix = "rw_")
-plot(sjcrn, add.spline=TRUE, nyrs=20)
-plot()
-
-sj_tree <- rw_byTree(sj)
-ca_tree <- rw_byTree(ca)
-
-rownames(ca_tree) <- ca_tree$year
-ca_tree <- ca_tree[,-1]
-ca_p <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-ca_p <- ca_p %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-sj_p <- data.frame(rw = rowMeans(sj_tree, na.rm=TRUE))
-
-sj_p <- sj_p %>% mutate(year = rownames(.),
-                        site = as.factor('sj'),
-                        rw_log = log(rw))
-
-xx <- rbind(ca_p, sj_p)
-
-xx %>% ggplot(aes(x=as.Date(year, format='%Y'), y=rw_log)) + 
-  geom_line(aes(group=1)) + 
-  scale_x_date(date_labels = "%Y") + 
-  facet_wrap(~site, ncol=1, scales = 'free_y') + 
-  theme_bw() + 
-  geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-xx %>% filter(site =='ca') %>% 
-  ggplot(aes(x=as.Date(year, format='%Y'), y=rw)) + geom_point(colour='blue') +
-  geom_line(aes(group=1), colour='blue') + 
-  scale_x_date(date_labels = "%Y") + 
-  theme_bw() + 
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-ca_prwi %>%
-  ggplot(aes(x = as.Date(year, format = '%Y'), y = rw)) + geom_point() +
-  geom_line(aes(group = 1)) +
-  scale_x_date(date_labels = "%Y") +
-  theme_bw() +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width Index')
-
-```

### 2017-06-06 analysis/analysis_disturbance_chronos.Rmd
update titles
update

@@
 
-### Plot 
+### Plot (mean) 
 ```{r}
@@
   geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
-  ylab('Mean GC') + 
+  ylab('% GC (mean)') + 
   theme(panel.grid.minor = element_blank(), 
@@
 
-# Plot (correcting the scale)
+### Plot (median) 
 ```{r}
+gc_todos %>% ggplot(aes(x=year, y=gc_md)) + 
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  # geom_point() +
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray') +
+  facet_wrap(~site, nrow = 3, scales = 'free_y',
+             labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low',
+                                 'gc_sj' = 'San Juan'))) +
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
+  ylab('% GC (median)') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
+```
 
+### Plot (correcting the scale) (mean)
+```{r}
 # Create a transformation scale 
@@
   geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
-  ylab('Mean GC') + 
+  ylab('% GC (mean)') + 
   theme(panel.grid.minor = element_blank(), 
@@
 
-## Númber of trees 
+### Plot (correcting the scale) (median)
+```{r}
+c <- gc_todos %>% 
+  filter(site != 'gc_sj') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
+  # scale_y_continuous(trans = pos_neg_log) +
+  facet_wrap(~site, nrow = 2,
+             labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low'))) +
+                                # 'gc_sj' = 'San Juan'))) + 
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
+  ylab('% GC (median)') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'),
+        axis.title.x = element_blank(),
+        axis.text.x = element_blank())
+
+d <- gc_todos %>% 
+  filter(site == 'gc_sj') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  xlim(layer_scales(a)$x$range$range[1],layer_scales(a)$x$range$range[2]) +
+  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
+  scale_y_continuous(trans = pos_neg_log, 
+                     breaks=c(-1000, -100, 0, 100, 500)) +
+  facet_wrap(~site, nrow = 2,
+             labeller=as_labeller(c('gc_sj' = 'San Juan'))) + 
+  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
+  ylab('') + 
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
+
+grid.newpage()
+grid.draw(rbind(ggplotGrob(c), ggplotGrob(d), size='last'))
+```
+
+## Number of trees with %GC > 50% (mean)
 ```{r}
@@
 ```
+
+## Number of trees with %GC > 50% (median)
+```{r}
+
+computeGCtrees <- function(gcdf, treshold){
+
+  th <- as.numeric(treshold)
+
+  # Get total tree by year 
+  n_tree <- gcdf %>% group_by(year) %>% summarise(total_tree = length(ntrees))
+
+  # Get n tree with GC positve upper treshold 
+  n_pos <- gcdf %>% dplyr::filter(gc_median > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees))
+
+  # Get n tree with GC negative lower treshold 
+  n_neg <- gcdf %>% dplyr::filter(gc_median < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees))
+
+  # Join and compute percentage 
+  outgc <- n_tree %>% 
+    left_join(n_pos, by='year') %>% 
+    left_join(n_neg, by='year') %>% 
+    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
+         per_neg = round((n_neg / total_tree) * 100, 2))
+
+  return(outgc)
+}
+
+gc_trees<- c()
+
+datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
+
+for (d in datasetes){ 
+  # get df
+  df <- get(d)
+
+  au <- df %>% computeGCtrees(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
+  gc_trees <- rbind(gc_trees, au)
+}
+
+# secondary axis
+# https://whatalnk.github.io/r-tips/ggplot2-secondary-y-axis.nb.html
+
+# Plot 
+gc_trees %>% ggplot() +
+  geom_bar(aes(x=year, y=per_pos), fill='blue', colour='blue', stat = "identity", position = "dodge") + 
+  geom_bar(aes(x=year, y=(-1)*per_neg), fill='red', colour='red', stat = "identity", position = "dodge") +
+  facet_wrap(~site, nrow = 3, labeller=as_labeller(c('gc_caH' = 'Canar High', 
+                                 'gc_caL' = 'Canar Low',
+                                 'gc_sj' = 'San Juan'))) + 
+  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
+  geom_line(aes(x=year, y=total_tree), colour = 'gray') + 
+  theme_bw() + ylab('% trees GC > 50 % (median)') +
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'))
+
+```
+
+# References

### 2017-06-07 analysis/analysis_chronologies.Rmd
update analysis

@@
 ```
-
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
-    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
 
@@
 
-aux_cro %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
+aux_cro %>% ggplot(aes(x = year, y = bai_mean / 100, colour=treatment)) + 
   geom_point() + geom_line() + 
@@
 
-cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+plot_cro_sj <- cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
   theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                    ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + geom_point(size=1) + 
+  # scale_colour_manual(values = c("red", "blue")) + 
   theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1920, 2020, by=10))
+
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sj_HL.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sj, width=12, height = 4) 
+plot_cro_sj 
 ```
@@
 
-cro_ca %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+plot_cro_ca <- cro_ca %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
   theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=1) + 
+  # scale_colour_manual(values = c("red", "blue")) + 
   theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep='')))+
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_ca_HL.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_ca, width=12, height = 4) 
+plot_cro_ca
 ```
@@
 
-cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+plot_cro_sites <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
   theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + 
-  theme(panel.grid = element_blank()) + facet_wrap(~loc, nrow = 2) + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=1) +
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + facet_wrap(~loc, nrow = 2) + 
   scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
 #  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sites.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sites, width=12, height = 8)
+plot_cro_sites
+```
+
+```{r}
+plot_cro_sites_collapsed <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  # geom_point(size=1) +
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
+#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sites_collapsed.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sites_collapsed, width=12, height = 4)
+plot_cro_sites_collapsed
+```
+
+```{r}
+plot_cro_sites_collapsed50  <- cro_sites %>% 
+  filter(year > 1949) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  # geom_point(size=1) +
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1950, 2020, by=5)) 
+#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sites_collapsed50.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sites_collapsed50, width=12, height = 4) 
+plot_cro_sites_collapsed50
+```
+
+```{r}
+plot_cro_sites_collapsed90  <- cro_sites %>% 
+  filter(year > 1989) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  # geom_point(size=1) +
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1990, 2016, by=2)) 
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sites_collapsed90.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sites_collapsed90, width=12, height = 4) 
+plot_cro_sites_collapsed90
 ```
@@
   theme(strip.background = element_blank())
+```
 
+```{r}
+correla_sites4 <- correla %>% filter(metodo_correla == 'pearson') %>%
+  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
+  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
+  xlab('Smoothing (centred moving average window size)') + 
+  theme(panel.grid = element_blank()) +
+  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
+  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
+  theme(legend.position = c(.8,.7))
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/correla_boot_sitesHL.pdf') 
+ggsave(filename=fileplot, plot=correla_sites4, width=5, height = 5) 
+correla_sites4
 ```
@@
 
----> here
+```{r}
+correla_sitesSJCALH  <- correla %>% 
+  filter(metodo_correla == 'pearson') %>%
+  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
+  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
+  xlab('Smoothing (centred moving average window size)') + 
+  # geom_smooth(method = 'loess', se=FALSE) + 
+  theme(panel.grid = element_blank()) +
+  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
+  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
+  theme(legend.position = c(.8,.9), legend.title = element_blank())
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/correla_boot_sitesSJCALH.pdf') 
+ggsave(filename=fileplot, plot=correla_sitesSJCALH, width=5, height = 5) 
+correla_sitesSJCALH 
+```
+
+# Chronologies by site (SJ, CAL, CAH)
+
+```{r}
+# Compute la crono de SJ de nuevo y llamar al site SJ
+cro_sjf <- chrono_bai(bai_sj)
+cro_sjf$site <- 'SJ'
+```
+
+```{r}
+# Combine and plot 
+cro_sites <- rbind(cro_caL, cro_caH, cro_sjf)
+
+plot_cro_sitesSJCALH <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sitesSJCALH.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH, width=12, height = 5)
+plot_cro_sitesSJCALH
+```
+
+```{r}
+plot_cro_sitesSJCALH50 <- cro_sites %>% 
+  filter(year > 1949) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sitesSJCALH50.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH50, width=12, height = 5)
+plot_cro_sitesSJCALH50
+```
+
+```{r}
+plot_cro_sitesSJCALH90 <- cro_sites %>% 
+  filter(year > 1989) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
+```
+
+```{r}
+fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sitesSJCALH90.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH90, width=12, height = 5)
+plot_cro_sitesSJCALH90
+```
 

### 2017-06-07 man/notes_dendro_analysis.Rmd
update
update analysis

@@
 
+#### Salidas gráficas
+Ver `./out/fig/chronos/`. Hacemos una comparación de cronos (BAi medios +- se), enfocado en sitio High and Low:
+
+  * `crono_compara_sj_HL.pdf`
+  * `crono_compara_ca_HL.pdf`
+  * Todos los sitios juntos:
+
+    * Facet: `crono_compara_sites.pdf`
+    * Collapsed: `crono_compara_sites_collapsed.pdf`
+    * Collapsed (>1950): `crono_compara_sites_collapsed50.pdf`
+    * Collapsed (>1990): `crono_compara_sites_collapsed90.pdf`
+
+Una vez analizadas las similitudes entre cronologías (ver apartado siguiente), decidimos quedarnos con 3 cronos: SJ, CAL y CAH.
+Y el plot de las cronos, con su error estandar es: `crono_compara_sitesSJCALH.pdf`
+
 ### Similitud chronologias
@@
     * Con boosptrap se obtienen niveles de significación :red_circle: ASK to Guillermo and Isabel.
+
+#### Salidas gráficas
+Ver `./out/fig/chronos/`
+
+* correlación High-Low por sitio: `correla_boot_sitesHL.pdf`
+* correlación SJ, CAL, CAH: `correla_boot_sitesSJCALH.pdf`
 
@@
     * NGC: [(M2 - M1)/M2]*100
+
+* Aquí también hay varias aproximaciones. Yo sigo las indicaciones de G. Gea y utilizamos ésta opción.
 

### 2017-06-08 man/notes_dendro_analysis.Rmd
update notes

@@
 
-Se han construido cronologías de perturbaciones siguiendo la aproximación de aproximación de [@Nowacki1997](varios ejemplos se pueden ver en [@GeaIzquierdo2014, @Dorado2017]).
+Se han construido cronologías de perturbaciones siguiendo la aproximación de aproximación del método *Percent Increase*  [@Nowacki1997](varios ejemplos se pueden ver en [@GeaIzquierdo2014, @Dorado2017]).
 Utilizamos la función `computeGC` (código aquí: [./script/R/computeGC.R](/script/R/gea/computeGC.R)).
@@
 
+siendo *M1* la media (o mediana) del crecimiento (media de RW) en los 10 años precedentes (incluye el año target); mientras que *M2* es la media del crecimiento en los 10 años subsecuentes (excluyendo el año target).
+
 * Aquí también hay varias aproximaciones. Yo sigo las indicaciones de G. Gea y utilizamos ésta opción. 
+* En la aproximación de [@Nowacki1997], además se establecen dos tresholds:
+
+    * 25 % GC implica una liberación moderada
+    * 50 % GC implica una liberación intensa (major)
+
+* Algunas ventajas del método de [@Nowacki1997] (según [@Altman2014]):
+
+    * Se puede aplicar aun con pequeños tamaños de muestra
+    * No es necesaria información sobre la autoecología de la especie
+
+En una revisión sobre métodos para detectar disturbance events, [@Rubio2014] encontró que el método mas común es el de *runneing mean*, es decir, la comparación la tasa de crecimiento de grupos de anillos adyacentes.
 

### 2017-06-08 man/notes_dendro_analysis.Rmd
update bib and notes

@@
 Para cada sitio y/o loc construimos una cronología utilizando las series de BAI. Podemos utilizar varias aproximaciones:
+
  * media aritmética 
@@
 
-#### Salidas gráficas
-Ver `./out/fig/chronos/`. Hacemos una comparación de cronos (BAi medios +- se), enfocado en sitio High and Low:
+#### Salidas gráficas (ver `./out/fig/chronos/`)
+Hacemos una comparación de cronos (BAi medios ± se), enfocado en cada sitio entre High and Low:
 
@@
 
-#### Salidas gráficas
-Ver `./out/fig/chronos/`
+#### Salidas gráficas (ver `./out/fig/chronos/`)
 
@@
 
-Se han construido cronologías de perturbaciones siguiendo la aproximación de aproximación del método *Percent Increase*  [@Nowacki1997](varios ejemplos se pueden ver en [@GeaIzquierdo2014, @Dorado2017]).
-Utilizamos la función `computeGC` (código aquí: [./script/R/computeGC.R](/script/R/gea/computeGC.R)).
-En concreto hemos llevado a cabo lo siguiente:
+Se han construido cronologías de perturbaciones siguiendo la aproximación de [@Nowacki1997]: método *Percent Increase* (varios ejemplos se pueden ver en @GeaIzquierdo2014, @Dorado2017).
+Con éste método de Percent incre
+
+Utilizamos la función `computeGC` (código aquí: [./script/R/computeGC.R](/script/R/gea/computeGC.R)). En concreto hemos llevado a cabo lo siguiente:
 

### 2017-06-09 man/field_data.Rmd
update bib file

@@
     variant: markdown_github
-bibliography: references.bib
-csl: ecology.csl
+bibliography: ../references.bib
+csl: ../ecology.csl
 ---

### 2017-06-09 man/notes_competition_index.Rmd
udd nota sobre libro
update bib file

@@
 
+* Revisar el libro de Burkhart and Tomé (2012). Modeling Forest Trees
+and Stands. En este libro hay un capítulo sobre índices de competencia a nivel de árbol
+
 ## References

### 2017-06-13 script/xml_Parser.R
upload script parser

@@ -1,0 +1,286 @@
+library("XML")
+library('xml2')
+library(tidyverse)
+
+# machine <- 'ajpelu'
+machine <- 'ajpeluLap'
+di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
+
+# XML 
+x <- paste0(di, 'data_raw/meteo/5514_base_aerea_monthly.xml')
+
+# Create a tree with the elements of the XML document
+doc <- xmlTreeParse(x, getDTD = FALSE, encoding = 'ISO-8859-1')
+
+r <- xmlRoot(doc)
+
+# How many nodes 
+xmlSize(r[[1]])
+
+# Get metadata of estacion
+xmlRoot(doc)[['Estacion']]
+
+# Name station 
+md_station <- as.data.frame(t(xmlSApply(r, xmlAttrs)))
+rownames(md_station) <- NULL 
+
+# Get variables 
+md_variables <- as.data.frame(t(xmlSApply(r[['Estacion']][['Sensores']], xmlAttrs)))
+rownames(md_variables) <- NULL
+
+
+# Get all nodes of Estacion 
+valores <- r[['Estacion']]
+
+# Remove two first nodes
+valores <- valores[-(1:2)]
+
+valores[[800]][c('Medida')]
+
+Dates 
+d <- as.data.frame(t(xmlAttrs(valores[[1]])))
+
+
+dd = sapply(xmlChildren(valores),
+            function(x)
+              xmlValue(x[['Valores']][['Medida']]))
+
+
+tt = xmlSApply(r, xmlGetAttr, 'Fecha')
+
+
+r <- xmlRoot(doc[[]])
+ttt <- r[['Estacion']][[800]]
+
+getSibling(ttt, after= FALSE)
+
+cc = xmlSApply(r, "[[", 800)
+mg = as.character(sapply(cc, xmlGetAttr, 'Medida'))
+
+mg = as.numeric(xmlSApply(r, 
+                          function(node)
+                            xmlGetAttr(node[[800]], 'Medida')))
+
+
+valores[[800]]
+length(valores)
+xmlSize(valores[[3]])
+
+# Get dates 
+xx <- xmlSApply(r[['Estacion']], xmlValue)
+
+
+variables_names <- md_variables$ID
+
[…]

### 2017-06-13 script/xml_Parser.R
update script parser

@@ -1,16 +1,16 @@
 library("XML")
-library('xml2')
-library(tidyverse)
+#library('xml2')
+#library(tidyverse)
 
-# machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
 
 # XML 
 x <- paste0(di, 'data_raw/meteo/5514_base_aerea_monthly.xml')
 
 # Create a tree with the elements of the XML document
-doc <- xmlTreeParse(x, getDTD = FALSE, encoding = 'ISO-8859-1')
+doc <- xmlTreeParse(x, getDTD = FALSE)
 
 r <- xmlRoot(doc)
 
@@ -35,252 +35,35 @@ valores <- r[['Estacion']]
 # Remove two first nodes
 valores <- valores[-(1:2)]
 
-valores[[800]][c('Medida')]
 
-Dates 
-d <- as.data.frame(t(xmlAttrs(valores[[1]])))
 
 
-dd = sapply(xmlChildren(valores),
-            function(x)
-              xmlValue(x[['Valores']][['Medida']]))
+aux_final <- c()
 
+for (n in 1:xmlSize(valores)){ 
 
-tt = xmlSApply(r, xmlGetAttr, 'Fecha')
+  # Get date 
+  mydate <- xmlGetAttr(valores[[n]], "Fecha")
+ 
+  # Get subnodes by date  
+  subn <- xmlSize(valores[[n]])
 
+  aux_valores <- c() 
+    for (j in 1:subn){
+      vb <- as.character(xmlAttrs(valores[[n]][[j]]))
+      # vb <- as.character(xmlAttrs(valores[[800]][[1]]))
+      value <- xmlValue(valores[[n]][[j]])
+      # value <- xmlValue(valores[[800]][[1]])
+      
+      aux <- data.frame(vb, value, mydate)
+      
+      aux_valores <- rbind(aux_valores, aux)
+    }
+    
+  aux_final <- rbind(aux_final, aux_valores)
+}
 
-r <- xmlRoot(doc[[]])
-ttt <- r[['Estacion']][[800]]
-
-getSibling(ttt, after= FALSE)
-
[…]

### 2017-06-14 script/xml_Parser.R
update parser

@@ -1,18 +1,46 @@
 library("XML")
-#library('xml2')
-#library(tidyverse)
+library('tidyverse')
 
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
+# WD 
+# machine <- 'ajpelu'
+machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
 
-# XML 
+# read xmls 
+
+f <- list.files(paste0(di, 'data_raw/meteo/'))
+
+for (m in seq_along(f)){
+  
+  # get name of file 
+  name_f <- str_replace(f[1], pattern = '.xml', '')
+  
+  # get file 
+  xml_file <- paste0(di, 'data_raw/meteo/', f[1]) 
+  
+  # Parse file
+  # Create a tree with the elements of the XML document
+  doc <- xmlTreeParse(xml_file, getDTD = FALSE)
+  
+  # rootXML
+  r <- xmlRoot(doc)
+  
+  # Get Name station 
+  md_station <- as.data.frame(t(xmlSApply(r, xmlAttrs)))
+  rownames(md_station) <- NULL 
+  
+  # Get name variables 
+  md_variables <- as.data.frame(t(xmlSApply(r[['Estacion']][['Sensores']], xmlAttrs)))
+  rownames(md_variables) <- NULL
+  
+  
+}
+
 x <- paste0(di, 'data_raw/meteo/5514_base_aerea_monthly.xml')
 
-# Create a tree with the elements of the XML document
-doc <- xmlTreeParse(x, getDTD = FALSE)
 
-r <- xmlRoot(doc)
+
+
 
 # How many nodes 
 xmlSize(r[[1]])

### 2017-06-14 script/R/parseaXMLclima.R
update anlaysis chrono and parser climate data

@@ -1,0 +1,70 @@
+## Function to parser REDIAM XML Climate data 
+# @ajpelu 2017 
+# A custom function to parser the climate data 
+# nameXML: '5514_base_aerea_dialy.xml' 
+# path: paste0(di, 'data_raw/meteo/') 
+
+parseaXMLclima <- function(nameXML, path){ 
+  require('XML')
+  require('stringr')
+  require('lubridate')
+  
+  # get name of file 
+  name_f <- str_replace(nameXML, pattern = '.xml', '')
+  
+  # get file 
+  xml_f <- paste0(path, nameXML)
+  
+  # Parse file
+  # Create a tree with the elements of the XML document
+  doc <- xmlTreeParse(xml_f, getDTD = FALSE)
+  
+  # rootXML
+  r <- xmlRoot(doc)
+  
+  # Get Name station 
+  md_station <- as.data.frame(t(xmlSApply(r, xmlAttrs)))
+  rownames(md_station) <- NULL 
+  
+  # Get name variables 
+  md_variables <- as.data.frame(t(xmlSApply(r[['Estacion']][['Sensores']], xmlAttrs)))
+  rownames(md_variables) <- NULL
+  
+  # Get all nodes of Estacion 
+  valores <- r[['Estacion']]
+  
+  # Remove two first nodes
+  valores <- valores[-(1:2)]
+  
+  # Create dataframe 
+  formateados <- c()
+  
+  for (n in 1:xmlSize(valores)){ 
+    # Get date 
+    mydate <- xmlGetAttr(valores[[n]], "Fecha")
+    
+    # Get subnodes by date  
+    subn <- xmlSize(valores[[n]])
+    
+    aux_valores <- c() 
+    for (j in 1:subn){
+      vb <- as.factor(as.character(xmlAttrs(valores[[n]][[j]]))) 
+      value <- as.numeric(xmlValue(valores[[n]][[j]])) 
+      
+      aux <- data.frame(vb, value)
+      
+      aux_valores <- rbind(aux_valores, aux)
+    }
+    
+    aux_valores$date <- dmy(mydate)
+    
+    formateados <- rbind(formateados, aux_valores)
+  }
+  
+  out <- list() 
+  out$station <- md_station
+  out$variables <- md_variables
+  out$values <- formateados
+  return(out) 
+} 
+

### 2017-06-14 script/xml_Parser.R

@@ -1,97 +1,56 @@
-library("XML")
+# Script to parser XML data from CLIMA-REDIAM
+# @ajpelu 2017
+
+library('XML')
 library('tidyverse')
+library('stringr')
 
 # WD 
-# machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
 
 # read xmls 
+f <- list.files(paste0(di, 'data_raw/meteo/'), pattern = '*.xml')
 
-f <- list.files(paste0(di, 'data_raw/meteo/'))
 
-for (m in seq_along(f)){
-  
-  # get name of file 
-  name_f <- str_replace(f[1], pattern = '.xml', '')
-  
-  # get file 
-  xml_file <- paste0(di, 'data_raw/meteo/', f[1]) 
-  
-  # Parse file
-  # Create a tree with the elements of the XML document
-  doc <- xmlTreeParse(xml_file, getDTD = FALSE)
-  
-  # rootXML
-  r <- xmlRoot(doc)
-  
-  # Get Name station 
-  md_station <- as.data.frame(t(xmlSApply(r, xmlAttrs)))
-  rownames(md_station) <- NULL 
-  
-  # Get name variables 
-  md_variables <- as.data.frame(t(xmlSApply(r[['Estacion']][['Sensores']], xmlAttrs)))
-  rownames(md_variables) <- NULL
-  
-  
-}
+# Load function 
+source(paste0(di, 'script/R/parseaXMLclima.R'))
 
-x <- paste0(di, 'data_raw/meteo/5514_base_aerea_monthly.xml')
+# Set path 
+mypath <- paste0(di, 'data_raw/meteo/') 
+
+# # Loop to process all data 
+# for (i in 1:length(f)){ 
+#   d <- parseaXMLclima(f[i], path=mypath)
+#   name_d <- str_replace(f[i], pattern = '.xml', '')
+#   assign(name_d, d)
+# }
+
+d_base <- parseaXMLclima("5514_base_aerea_dialy.xml", mypath) 
+m_base <- parseaXMLclima("5514_base_aerea_monthly.xml", mypath) 
+m_cartuja <- parseaXMLclima("5515_cartuja_monthly.xml", mypath) 
+m_soportujar <- parseaXMLclima("6246_soportujaar_monthly.xml", mypath) 
[…]

### 2017-06-19 analysis/analysis_climate.Rmd
update climate

--- 2017-05-10 analysis/analysis_competence.Rmd
@@
 ---
-title: "Analysis of Competence Index"
+title: "Climate REDIAM"
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 Feb"
+date: "2017"
 output:  
@@
 library("stringr")
+library("lubridate")
 ```
 
-# Read data 
+## Compute competition indices
+
 ```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
+# machine <- 'ajpelu'
+machine <- 'ajpeluLap'
 di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
 
-# density data 
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+# climate data
+
+cli <- read.csv(file=paste0(di, '/data_raw/meteo/5514_base_aerea_dialy.csv'), header=TRUE, sep=',')
 ```
+
+# Minimun temperature
+```{r}
+
+cli %>% 
+  filter(vb == 'TD4') %>% 
+  mutate(y = lubridate::year(date)) %>% 
+  filter(value > 34) %>% 
+  group_by(y) %>% 
+  summarise(ndays = n()) %>% 
+  ggplot(aes(x=y, y=ndays)) +
+  geom_point() + 
+  geom_smooth(method = 'loess', span = 0.75, se=FALSE) +
+  theme_bw()
+
++ 
+  geom_vline(xintercept = 1995)
+
+  group_by(y) %>% 
+  slice(which.max(value))
+
+df %>% group_by(A) %>% slice(which.min(x)
 
@@
 ```
-
-## Compute competition indices
 
@@
 
-## Plot
+## Plot by site
 
@@
 ```
-
-## Plot by site
 

### 2017-06-19 analysis/analysis_disturbance_chronos.Rmd
include trader analysis

@@
 
+# TRADER approach 
+Vamos a utilizar la aproximación del pkg `trader` [Altman2014]
+
+## Nowacki & Abrams 1997
+Mismo análisis de PG [@Nowacki1997], pero con trader.
+
+```{r}
+## PG approach 
+library('TRADER')
+
+temp <- getwd()
+
+setwd(paste0(di, 'analysis/trader_chronos/caL/'))
+growthAveragingALL(caL_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
+                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
+                   prefix = 'CA', gfun = median, length = 5, storedev = pdf)
+
+setwd(paste0(di, 'analysis/trader_chronos/caH/'))
+growthAveragingALL(caH_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
+                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
+                   prefix = 'CA', gfun = median, length = 5, storedev = pdf)
+
+setwd(paste0(di, 'analysis/trader_chronos/sj/'))
+growthAveragingALL(sj_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
+                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
+                   prefix = 'SJ', gfun = median, length = 5, storedev = pdf) 
+```
+
+## Black & Abrams 2003
+Análisis de boundary-line [@Black2003]
+```{r}
+setwd(paste0(di, 'analysis/trader_chronos/boundary_caL/'))
+
+boundaryLineALL(caL_tree, releases = NULL, m1 = 10, m2 = 10,
+                boundary = NULL, buffer = 10,
+                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
+                prefix = 'CA', drawing = TRUE, gfun = median, length = 5, 
+                notop = 10, notop2 = 10, storedev = pdf)
+
+setwd(paste0(di, 'analysis/trader_chronos/boundary_caH/'))
+boundaryLineALL(caH_tree, releases = NULL, m1 = 10, m2 = 10,
+                boundary = NULL, buffer = 10,
+                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
+                prefix = 'CA', drawing = TRUE, gfun = median, length = 5, 
+                notop = 10, notop2 = 10, storedev = pdf)
+
+setwd(paste0(di, 'analysis/trader_chronos/boundary_sj/'))
+boundaryLineALL(sj_tree, releases = NULL, m1 = 10, m2 = 10,
+                boundary = NULL, buffer = 10,
+                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
+                prefix = 'SJ', drawing = TRUE, gfun = median, length = 5, 
+                notop = 10, notop2 = 10, storedev = pdf)
+
+```
+
+## Fraver & White 2005
+Análisis de absolute increase [@Fraver2005]
+```{r}
+setwd(paste0(di, 'analysis/trader_chronos/absoluteI_caL/'))
+absoluteIncreaseALL(caL_tree, abs = NULL, m1 = 10, m2 = 10,
+                buffer = 10, prefix = 'CA', drawing = TRUE, gfun = median,
+                length = 5, storedev = pdf)
+
+setwd(paste0(di, 'analysis/trader_chronos/absoluteI_caH/'))
+absoluteIncreaseALL(caH_tree, abs = NULL, m1 = 10, m2 = 10,
+                buffer = 10, prefix = 'CA', drawing = TRUE, gfun = median,
+                length = 5, storedev = pdf)
+
+setwd(paste0(di, 'analysis/trader_chronos/absoluteI_sj/'))
+absoluteIncreaseALL(sj_tree, abs = NULL, m1 = 10, m2 = 10,
+                buffer = 10, prefix = 'SJ', drawing = TRUE, gfun = median,
+                length = 5, storedev = pdf) 
+```
+
 # References

### 2017-06-19 man/notes_dendro_analysis.Rmd
update bib file
include trader analysis

@@
 Se han construido cronologías de perturbaciones siguiendo la aproximación de [@Nowacki1997]: método *Percent Increase* (varios ejemplos se pueden ver en @GeaIzquierdo2014, @Dorado2017).
-Con éste método de Percent incre
+Con éste método, utilizando una ventana temporal de longitud suficiente (~10 años) se filtran (se ignoran) las respuestas a cambios short-term en temperatura y precipitación [@Fraver2005].
 
-Utilizamos la función `computeGC` (código aquí: [./script/R/computeGC.R](/script/R/gea/computeGC.R)). En concreto hemos llevado a cabo lo siguiente:
+Para su cómputo hemos utilizado la función `computeGC` (código aquí: [./script/R/computeGC.R](/script/R/gea/computeGC.R)). En concreto hemos llevado a cabo lo siguiente:
 
 * Computamos medias (o medianas) de RW series (por árbol) en una ventana de 10 años.
-G.Gea utiliza medias, aunque en otros trabajos utilizan medianas (las medianas son estimadores mas robustos de la tendencia central que la media [@Camarero2011]).
+G.Gea utiliza medias, aunque en otros trabajos utilizan medianas (las medianas son estimadores mas robustos de la tendencia central que la media [@Rubino2004; @Camarero2011]).
 Las medianas son menos sensibles a los valores extremos (son mas conservadores.
@@
 
-* Algunas ventajas del método de [@Nowacki1997] (según [@Altman2014]):
+* Ventajas del método de [@Nowacki1997]:
 
-    * Se puede aplicar aun con pequeños tamaños de muestra
-    * No es necesaria información sobre la autoecología de la especie
+    * Se puede aplicar aun con pequeños tamaños de muestra [@Altman2014]
+    * No es necesaria información sobre la autoecología de la especie [@Altman2014]
 
-En una revisión sobre métodos para detectar disturbance events, [@Rubio2014] encontró que el método mas común es el de *runneing mean*, es decir, la comparación la tasa de crecimiento de grupos de anillos adyacentes.
+* Algunos inconvenientes:
+
+    * La generalización del promedio de crecimiento radial puede conducir tanto a la detección de falsas liberaciones como a la exclusión de liberaciones verdaderas [@Altman2014]
+    * Excesivamente sensitivo a bajas tasas de crecimiento (acepta false positive releases) asi como excesivamente estricto a altas tasas de crecimiento (produciendo false negative releases) [@Fraver2005]
+    * Subjetividad en la elección del umbral minimo para considerar una perturbación. Cambios en el porcentaje elegido implican cambios en la detección de perturbaciones [@Rubino2004]
+    * La variación en la longitud de la ventana temporal, tiene un gran impacto en la detección de liberaciones [@Rubino2004].
+
+En una revisión sobre métodos para detectar disturbance events, [@Rubino2004] encontró que el método mas común es el de *running mean*, es decir, la comparación la tasa de crecimiento de grupos de anillos adyacentes.
+Comparando diferentes tipos de métodos para detectar perturbaciones (5 tipos de métodos: *static releases*; *detrending o estandarización*; *tasa de crecimiento medio*; *running means*; y *event response*), vieron que el método de *running means* es el que produce resultados que mejor concuerdan con perturbaciones antrópicas bien documentadas.
+
+Existe una aproximación llamada *boundary-line method* [@Black2003] que escala el porcentaje de cambio de [@Nowacki1997] utilizando la tasa de crecimiento previa a la perturbación.
+Se considera liberación menor y mayor aquellas que caen dentro del 20 - 49.9 % y del 50 - 100 % de la boundary-line respectivamente.
+La ventaja del método es la estandarización, ya que escalar el porcentaje de crecimiento respecto a la tase de crecimiento previa considera las relaciones entre la edad, el tamaño y el canopy class, que influencian en algun modo la tasa de crecimiento radial.
+Una de las desventajas es la gran cantidad de medidas de anillos necesarias para su computo, que está en torno a 50000 para una determinada especie [@Altman2014].
 
@@
 
-Un abrazo, buena semana.
-
 --------------------------

### 2017-06-19 analysis/analysis_disturbance_chronos.Rmd
update chronos y disturbance effects

@@
                 length = 5, storedev = pdf)
+
+setwd(temp)
+```
+
+```{r}
+
+# SJ 
+pg <- read.csv(paste0(di, 'analysis/trader_chronos/sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'PG')
+bl <- read.csv(paste0(di, 'analysis/trader_chronos/boundary_sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'BL')
+ai <- read.csv(paste0(di, 'analysis/trader_chronos/absoluteI_sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'AI')
+
+dist_sj <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'sj')
+
+# caL 
+pg <- read.csv(paste0(di, 'analysis/trader_chronos/caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'PG')
+bl <- read.csv(paste0(di, 'analysis/trader_chronos/boundary_caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'BL')
+ai <- read.csv(paste0(di, 'analysis/trader_chronos/absoluteI_caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'AI')
+
+dist_caL <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'caL')
+
+# caH 
+pg <- read.csv(paste0(di, 'analysis/trader_chronos/caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'PG')
+bl <- read.csv(paste0(di, 'analysis/trader_chronos/boundary_caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'BL')
+ai <- read.csv(paste0(di, 'analysis/trader_chronos/absoluteI_caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
+  mutate(metodo = 'AI')
+
+dist_caH <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'caH')
+
+disturbances <- dist_caH %>% bind_rows(dist_caL, dist_sj)
+```
+
+```{r}
+disturbances %>% ggplot(aes(x=AllReleasesYear, y=AllReleasesFreq)) + 
+  geom_point(aes(shape=metodo, size=AllReleasesFreq)) + scale_shape(solid = FALSE) +
+  facet_wrap(~site, nrow = 3) +
+  theme_bw()
+```
+
+```{r}
+disturbances %>% ggplot(aes(x=AllReleasesYear, y=metodo)) + 
+  geom_point(aes(size=AllReleasesFreq), shape=21, colour = "black")+
+  facet_wrap(~site, nrow = 3) +
+  theme_bw()
 ```

### 2017-06-19 man/notes_dendro_analysis.Rmd
update chronos y disturbance effects

@@
 
+He llevado a cabo una comparación de métodos con el pkg `trader` [@Altman2014] (siendo `XXX` cada sitio, i.e.: `sj`, `caL`, `caH`):
+
+* Método de percentage growth [@Nowacki1997]: `/analysis/trader_chronos/XXX/`
+* Método de boundary-line [@Black2003]: `/analysis/trader_chronos/boundary_XXX/`
+* Método de absolute increment [@Fraver2005]: `/analysis/trader_chronos/absoluteI_XXX/`
+
+Posteriormente he analizado los años en los que ocurren liberaciones detectadas con los diferentes métodos y en diferentes sitios.
+
+### notas sueltas de correos
 - Puedes probar los pointer years y lo de Lloret, efectivamente la idea era comparar los años de sequía de las cronos con tus análisis en imágenes.

### 2017-06-19 man/notes_dendro_analysis.Rmd
updating notes of analysis

@@
 
+# Resilience Analysis
+
+### [@Gazol2017]
+* Para cada año seco y en cada sitio se calculó la resiliencia del crecimiento de los árboles utilizando tres indices propuestos por [@Lloret2011], en los cuales, la respuesta del crecimiento del árbol a las condiciones secas se descompone en:
+    * Resistance (Rt): quantifies the difference between tree growth in the dry year and the years before it.
+Rt = Gd / Gprev.
+It quantifies the capacity of trees to buffer the drought stress and continue growth during drought
+    * Recovery (Rc):  quantifies the difference in growth between the dry year and a subsequent period. Rc = Gpost / Gd
+    * Resilience (Rs): quantifies the difference in tree growth before and after the dry year.
+Rs = Gpost / Gprev.
+It measures the capacity of trees to recover the growth rates observed before the drought stress
+
+* Gd =  crecimiento durante el periodo de sequía (mean tree-ring width)
+* Gprev y Gpost = crecimiento medio durante los tres años antes y despues del periodo de sequía
+
+* Cada año seco lo consideraron como un evento de sequía
+* Utilizaron una ventana de 3 años, porque obtuvieron resultados similares 2, 3 y 4 años (ver correlaciones del material suplementario de ese trabajo)
+
+### [@Cavin2017]
+* Utilizan los cuatro índices de [@Lloret2011] (los anteriores mas la Relative Resilience)
+* Evaluan tree growth usando BAI 
+* Determinan el año seco, utilizando una aproximación de pointer year. 
+* Relative Resilience (Gpost - Gd) / Gprev. Se trata de la resiliencia ponderada por la severidad del evento de sequía
+
+### [@Sanguesa2015]
+* Seleccionan tres años de sequía basándose en Spei de agosto de 4 meses (ver material suplemtario de su trabajo). Valores mas bajos de los últimos 25 años.  
+* Definen dos métricas para la respuesta del BAI a la sequía: Drougth Sensitivity (DS) y Drought Recovery (DR)
+
+    * Drougth Sensitivity (DS): [(BAId - BAId-3) / (BAId-3)]*100 
+    * Drought Recovery (DR): [(BAId+3 - BAId-3) / (BAId-3)]*100
+
+* Selección de ventana de 3 años, basado en
+
 ### notas sueltas de correos

### 2017-06-19 analysis/analysis_resilience_bai.Rmd
compute resilience BAI

--- 2017-06-06 analysis/analysis_disturbance_chronos.Rmd
@@
 ---
-title: "disturbance chronologies"
+title: "Growth Resilience"
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 Feb"
+date: "2017 June"
 output:  
@@
 
-# Disturbance chronologies
+# Resilience
 
-* Vamos a calcular las disturbance chronologies siguiendo la aproximación de [@Nowacki1997](varios ejemplos se pueden ver en [@GeaIzquierdo2014, @Dorado2017])
-
-* Utilizaremos tres chronologias: SJ, CAH y CAL (ver [./analysis/analysis_chronologies.md]('./analysis/analysis_chronologies.md))
+* Calcularemos las métricas resiliencia de [@Lloret2011] sobre el crecimiento. 
+* Vamos a calcularlas sobre el BAI de cada árbol. 
+* Utilizaremos tres sitios: SJ, CAH y CAL (ver [./analysis/analysis_chronologies.md]('./analysis/analysis_chronologies.md))
 
@@
 ```{r, echo=FALSE}
-machine <- 'ajpelu'
-#machine <- 'ajpeluLap'
+#machine <- 'ajpelu'
+machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/computeGC.R'))
 ```
@@
 * Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low
-
-    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
 
@@
 
-# Summary dendrochronology statistics
-
-* Por cada site (CAH, CAL, SJ) calculamos los siguientes datos: 
-    * age: mean + sd (average of all series)
-    * year: min, max of all series
-    * year range
-    * autocorrelation order 1 (ar1): mean + sd 
-    * seires intercorrelation (icor): mean + sd
-
-## By site 
-```{r}
-objects_rwl <- c('caL','caH','sj')
-
-out <- c()
-
-for (i in objects_rwl){
-
-  aux <- get(i)
-  # Stats of singles RWL
-  aux_stats <- rwl.stats(aux)
-  # Add site 
-  aux_stats$site <- as.factor(i)
-
-  # Compute interseries correlation 
-  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
-  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
-
-  out <- rbind(out, aux_stats)
-  }
-
-rwl_by_site <- out %>% group_by(site) %>% 
-  summarize(age_mean = round(mean(year),2), 
-            age_sd = round(sd(year), 2),
-            year_min = min(first),
-            year_max = max(last), 
-            year_range = year_max - year_min + 1, 
-            ar1_mean = round(mean(ar1), 3),
-            ar1_sd = round(sd(ar1), 3),
-            icor_mean = round(mean(interser_cor_val), 3),
-            icor_sd = round(sd(interser_cor_val), 3)) %>% 
-  as.data.frame()
-
-rwl_by_site %>% 
-  write.csv(file=paste(di, "data/proto_tables/site_3_dendro.csv", sep=""), row.names = FALSE)
-
-pander(rwl_by_site, caption='Dendrochronological summary by sites (3)') 
-```
-
 # Aggregate RW by tree
@@
 caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')
-
-# Si decido usar treeMean debería ser (solo un ejemplo)
-# ids_sj <- read.ids(sj, stc = c(2, 2, 2))
-# sj_meantree <- treeMean(sj, ids = ids_sj,  na.rm=TRUE)
 ```
@@
 
-# Disturbance chronologies
+# Cómputo del BAI por site
 
-* calcular chronologies con media y mediana. 
-* ver [./analysis/analysis_disturbance_chronos.md](/analysis/analysis_disturbance_chronos.md))
+* He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]. Es similar a `bai.out`
 
 ```{r}
-gc_caL <- computeGC(caL_tree, ws=10, prefijo = 'CA')
-gc_caH <- computeGC(caH_tree, ws=10, prefijo = 'CA')
-gc_sj <- computeGC(sj_tree, ws=10, prefijo = 'SJ')
-```
+bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
+bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
+bai_caL <- bai_piovesan(rwdf = caL_tree, diam_df = d_caL)
 
-### Plot (mean) 
-```{r}
-# Plot mean value by year 
-# dataframe for plot 
-gc_todos <- c()
+# Set class to bai object 
+# Esto es para que funcionen algunas otras funciones de dplR 
+bais <- c('bai_sj', 'bai_caH', 'bai_caL')
 
-datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
+for (i in bais){
+  aux <- get(i)
 
-for (d in datasetes){ 
-  # get df
-  df <- get(d)
+  class(aux) <- c('rwl', 'data.frame')
 
-  aux_df <- df %>%
-    group_by(year) %>% 
-    summarise(ntrees = length(ntrees),
-              gc_m = mean(gc_mean),
-              gc_m_sd = sd(gc_mean),
-              gc_m_se = sd(gc_mean) / sqrt(n()), 
-              gc_md = mean(gc_median),
-              gc_md_sd = sd(gc_median),
-              gc_md_se = sd(gc_median) / sqrt(n())) %>%  
-    mutate(site=paste(d))
-
-  gc_todos <- rbind(gc_todos, aux_df)
+  assign(i, aux)
 }
-
-gc_todos %>% ggplot(aes(x=year, y=gc_m)) + 
-  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
-  # geom_point() +
-  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray') +
-  facet_wrap(~site, nrow = 3, scales = 'free_y',
-             labeller=as_labeller(c('gc_caH' = 'Canar High', 
-                                 'gc_caL' = 'Canar Low',
-                                 'gc_sj' = 'San Juan'))) +
-  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
-  ylab('% GC (mean)') + 
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'))
-```
-
-### Plot (median) 
-```{r}
-gc_todos %>% ggplot(aes(x=year, y=gc_md)) + 
-  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
-  # geom_point() +
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray') +
-  facet_wrap(~site, nrow = 3, scales = 'free_y',
-             labeller=as_labeller(c('gc_caH' = 'Canar High', 
-                                 'gc_caL' = 'Canar Low',
-                                 'gc_sj' = 'San Juan'))) +
-  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
-  ylab('% GC (median)') + 
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'))
-```
-
-### Plot (correcting the scale) (mean)
-```{r}
-# Create a transformation scale 
-# https://stackoverflow.com/questions/37446064/i-need-ggplot-scale-x-log10-to-give-me-both-negative-and-positive-numbers-as-o
-
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse=function(x) sign(x)*exp(abs(x)))
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-a <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_m)) + 
-  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
-  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray', width=.3, size=.2) +
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site, nrow = 2,
-             labeller=as_labeller(c('gc_caH' = 'Canar High', 
-                                 'gc_caL' = 'Canar Low'))) +
-                                # 'gc_sj' = 'San Juan'))) + 
-  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
-  ylab('% GC (mean)') + 
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'),
-        axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-b <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_m)) + 
-  xlim(layer_scales(a)$x$range$range[1],layer_scales(a)$x$range$range[2]) +
-  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
-  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray', width=.3, size=.2) +
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -100, 0, 100, 500)) +
-  facet_wrap(~site, nrow = 2,
-             labeller=as_labeller(c('gc_sj' = 'San Juan'))) + 
-  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
-  ylab('') + 
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'))
-
-grid.newpage()
-grid.draw(rbind(ggplotGrob(a), ggplotGrob(b), size='last'))
-```
-
-### Plot (correcting the scale) (median)
-```{r}
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site, nrow = 2,
-             labeller=as_labeller(c('gc_caH' = 'Canar High', 
-                                 'gc_caL' = 'Canar Low'))) +
-                                # 'gc_sj' = 'San Juan'))) + 
-  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
-  ylab('% GC (median)') + 
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'),
-        axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(a)$x$range$range[1],layer_scales(a)$x$range$range[2]) +
-  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -100, 0, 100, 500)) +
-  facet_wrap(~site, nrow = 2,
-             labeller=as_labeller(c('gc_sj' = 'San Juan'))) + 
-  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
-  ylab('') + 
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'))
-
-grid.newpage()
-grid.draw(rbind(ggplotGrob(c), ggplotGrob(d), size='last'))
-```
-
-## Number of trees with %GC > 50% (mean)
-```{r}
-
-computeGCtrees <- function(gcdf, treshold){
-
-  th <- as.numeric(treshold)
-
-  # Get total tree by year 
-  n_tree <- gcdf %>% group_by(year) %>% summarise(total_tree = length(ntrees))
-
-  # Get n tree with GC positve upper treshold 
-  n_pos <- gcdf %>% dplyr::filter(gc_mean > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees))
-
-  # Get n tree with GC negative lower treshold 
-  n_neg <- gcdf %>% dplyr::filter(gc_mean < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees))
-
-  # Join and compute percentage 
-  outgc <- n_tree %>% 
-    left_join(n_pos, by='year') %>% 
-    left_join(n_neg, by='year') %>% 
-    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
-         per_neg = round((n_neg / total_tree) * 100, 2))
-
-  return(outgc)
-}
-
-gc_trees<- c()
-
-datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
-
-for (d in datasetes){ 
-  # get df
-  df <- get(d)
-
-  au <- df %>% computeGCtrees(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
-  gc_trees <- rbind(gc_trees, au)
-}
-
-# secondary axis
-# https://whatalnk.github.io/r-tips/ggplot2-secondary-y-axis.nb.html
-
-# Plot 
-gc_trees %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill='blue', colour='blue', stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill='red', colour='red', stat = "identity", position = "dodge") +
-  facet_wrap(~site, nrow = 3, labeller=as_labeller(c('gc_caH' = 'Canar High', 
-                                 'gc_caL' = 'Canar Low',
-                                 'gc_sj' = 'San Juan'))) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'gray') + 
-  theme_bw() + ylab('%trees GC > 50 %') +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'))
 
@@
 
-## Number of trees with %GC > 50% (median)
+## Resilience function
+
 ```{r}
 
-computeGCtrees <- function(gcdf, treshold){
+event_year <- c(1995, 2005, 2012)
 
-  th <- as.numeric(treshold)
+window <- c(2,3) 
+  i <- 1995
+  window <- 3
+  bai_object <- bai_sj
 
-  # Get total tree by year 
-  n_tree <- gcdf %>% group_by(year) %>% summarise(total_tree = length(ntrees))
+baiResilience <- function(bai_object, event_years, window){
 
-  # Get n tree with GC positve upper treshold 
-  n_pos <- gcdf %>% dplyr::filter(gc_median > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees))
+  out_growth <- c() 
+  out_res <- c() 
+  out <- list()
 
-  # Get n tree with GC negative lower treshold 
-  n_neg <- gcdf %>% dplyr::filter(gc_median < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees))
+  for (i in event_years){
+    # Create df previous, during and post event
+    df <- bai_object %>% 
+      mutate(year = as.numeric(row.names(.))) %>% as.data.frame()
 
-  # Join and compute percentage 
-  outgc <- n_tree %>% 
-    left_join(n_pos, by='year') %>% 
-    left_join(n_neg, by='year') %>% 
-    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
-         per_neg = round((n_neg / total_tree) * 100, 2))
+    df_pre <- df %>% filter(year < i & year >= (i - window)) %>% mutate(disturb = 'pre')
+    df_event <- df %>% filter(year == i) %>% mutate(disturb = 'dr')
+    df_post <- df %>% filter(year > i & year <= (i + window)) %>% mutate(disturb = 'post')
 
-  return(outgc)
+    # Melt data 
+    dfmelt <- bind_rows(df_pre, df_event, df_post) %>% 
+      gather(tree, value, -disturb, -year)
+
+    # Compute mean of Growth (previous, during and posterior)
+    growth_df <- dfmelt %>% group_by(disturb, tree) %>% 
+      summarise(mean_period = mean(value),
+                sd_period = sd(value),
+                se_period = sd_period/sqrt(length(value))) %>% 
+      mutate(disturb_year = i)
+
+    out_growth <- rbind(out_growth, growth_df)
+
+    # Compute resilience by trees 
+    aux_resilience <- c() 
+    trees <- unique(growth_df$tree)
+
+    for (t in trees){
+      # Filter by tree
+      df_tree <- growth_df %>% filter(tree == t)
+
+      # Compute resilience metrics by tree
+      aux_df <- df_tree %>% select(-c(tree, sd_period, se_period)) %>% 
+        spread(key=disturb, value=mean_period) %>% 
+        mutate(rt = dr / pre,
+               rc = post / dr,
+               rs = post / pre,
+               rrs = (post - dr) / pre,tree = t) %>% 
+        select(rt, rc, rs, rrs, disturb_year, tree)
+
+      aux_resilience <- rbind(aux_resilience, aux_df)
     }
 
-gc_trees<- c()
-
-datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
-
-for (d in datasetes){ 
-  # get df
-  df <- get(d)
-
-  au <- df %>% computeGCtrees(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
-  gc_trees <- rbind(gc_trees, au)
+    out_res <- rbind(out_res, aux_resilience)
   }
 
-# secondary axis
-# https://whatalnk.github.io/r-tips/ggplot2-secondary-y-axis.nb.html
+  out$growth <- out_growth
+  out$resilience <- out_res
 
-# Plot 
-gc_trees %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill='blue', colour='blue', stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill='red', colour='red', stat = "identity", position = "dodge") +
-  facet_wrap(~site, nrow = 3, labeller=as_labeller(c('gc_caH' = 'Canar High', 
-                                 'gc_caL' = 'Canar Low',
-                                 'gc_sj' = 'San Juan'))) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'gray') + 
-  theme_bw() + ylab('% trees GC > 50 % (median)') +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'))
+  return(out)
+}
+
+dyears <- c(1995, 2005, 2012)
+res_sj3 <- baiResilience(bai_sj, event_years = dyears, window = 3)
+res_sj2 <- baiResilience(bai_sj, event_years = dyears, window = 2)
+
+resile2 <- res_sj2$resilience
+resile3 <- res_sj3$resilience
+
+summary(lm(resile2$rs~resile3$rs))
 

### 2017-06-20 script/R/baiResilience.R
function to compute bai Resilience

@@ -1,0 +1,70 @@
+## Compute Resilience of BAI 
+
+## Resilience metrics: See Lloret et al. 2011 http://dx.doi.org/10.1111/j.1600-0706.2011.19372.x
+## bai_object: an object of bai (dplR)
+## event_years: vector with drought years, i.e.: c(1995, 2005)
+## window: size of the window to compute previous and post event segments 
+
+## A list with  to df 
+## $growth: summary statistics by tree of BAI for each period (prev, dr, post)
+## $resilience: resilience metrics for each drougth period 
+
+baiResilience <- function(bai_object, event_years, window){ 
+  
+  out_growth <- c() 
+  out_res <- c() 
+  out <- list()
+  
+  for (i in event_years){
+    # Create df previous, during and post event
+    df <- bai_object %>% 
+      mutate(year = as.numeric(row.names(.))) %>% as.data.frame()
+    
+    df_pre <- df %>% filter(year < i & year >= (i - window)) %>% mutate(disturb = 'pre')
+    df_event <- df %>% filter(year == i) %>% mutate(disturb = 'dr')
+    df_post <- df %>% filter(year > i & year <= (i + window)) %>% mutate(disturb = 'post')
+    
+    # Melt data 
+    dfmelt <- bind_rows(df_pre, df_event, df_post) %>% 
+      gather(tree, value, -disturb, -year)
+    
+    # Compute mean of Growth (previous, during and posterior)
+    growth_df <- dfmelt %>% group_by(disturb, tree) %>% 
+      summarise(mean_period = mean(value),
+                sd_period = sd(value),
+                se_period = sd_period/sqrt(length(value))) %>% 
+      mutate(disturb_year = i) %>% as.data.frame() 
+    
+    out_growth <- rbind(out_growth, growth_df)
+    
+    # Compute resilience by trees 
+    aux_resilience <- c() 
+    trees <- unique(growth_df$tree)
+    
+    for (t in trees){
+      # Filter by tree
+      df_tree <- growth_df %>% filter(tree == t) 
+      
+      # Compute resilience metrics by tree
+      aux_df <- df_tree %>% select(-c(tree, sd_period, se_period)) %>% 
+        spread(key=disturb, value=mean_period) %>% 
+        mutate(rt = dr / pre,
+               rc = post / dr,
+               rs = post / pre,
+               rrs = ((post - dr) / pre),
+               tree = t) %>% 
[…]

### 2017-06-20 analysis/analysis_resilience_bai.Rmd
updating reslience BAI analysis

@@
 library("tidyverse")
+library("dplR")
 library("stringr")
-library("dplR")
 library("knitr")
-# library("detrendeR")
-library("pander")
-library('boot')
 library('gtable')
 library('grid')
+library('gridExtra')
 ```
@@
 ```{r, echo=FALSE}
-#machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 source(paste0(di, 'script/R/bai_piovesan.R'))
+source(paste0(di, 'script/R/baiResilience.R'))
 ```
@@
 
-# Cómputo del BAI por site
+## Cómputo del BAI por site
 
@@
 
-## Resilience function
+## Resilience
+
+* Computar métricas de resiliencia BAI para los tres sitios. 
+* Computar tres eventos climáticos: 1995, 2005, 2012
+* Computar dos ventanas temporales: 2 y 3
 
 ```{r}
+# Drought years 
+dyears <- c(1995, 2005, 2012)
 
-event_year <- c(1995, 2005, 2012)
+# SJ 
+res_3_sj <- baiResilience(bai_sj, event_years = dyears, window = 3)
+res_2_sj <- baiResilience(bai_sj, event_years = dyears, window = 2)
 
-window <- c(2,3) 
-  i <- 1995
-  window <- 3
-  bai_object <- bai_sj
+# caL 
+res_3_caL <- baiResilience(bai_caL, event_years = dyears, window = 3)
+res_2_caL <- baiResilience(bai_caL, event_years = dyears, window = 2)
 
-baiResilience <- function(bai_object, event_years, window){
+# caH 
+res_3_caH <- baiResilience(bai_caH, event_years = dyears, window = 3)
+res_2_caH <- baiResilience(bai_caH, event_years = dyears, window = 2)
+```
 
-  out_growth <- c() 
-  out_res <- c() 
-  out <- list()
+* Computar correlaciones ventanas temporales
 
-  for (i in event_years){
-    # Create df previous, during and post event
-    df <- bai_object %>% 
-      mutate(year = as.numeric(row.names(.))) %>% as.data.frame()
+```{r}
+# Vector with objects name
+obj <- c('res_2_sj', 'res_3_sj',
+         'res_2_caL', 'res_3_caL',
+         'res_2_caH', 'res_3_caH')
 
-    df_pre <- df %>% filter(year < i & year >= (i - window)) %>% mutate(disturb = 'pre')
-    df_event <- df %>% filter(year == i) %>% mutate(disturb = 'dr')
-    df_post <- df %>% filter(year > i & year <= (i + window)) %>% mutate(disturb = 'post')
+variables <- c('rt', 'rs', 'rc', 'rrs')
 
-    # Melt data 
-    dfmelt <- bind_rows(df_pre, df_event, df_post) %>% 
-      gather(tree, value, -disturb, -year)
+correla_ws <- c()
 
-    # Compute mean of Growth (previous, during and posterior)
-    growth_df <- dfmelt %>% group_by(disturb, tree) %>% 
-      summarise(mean_period = mean(value),
-                sd_period = sd(value),
-                se_period = sd_period/sqrt(length(value))) %>% 
-      mutate(disturb_year = i)
+for (i in obj){ 
+  x <- get(i)
+  xres <- x$resilience
+  out <- xres %>% 
+    mutate(ws = paste0('ws_', as.character(str_extract(i, "([0-9])"))),
+           site = str_replace(i, "res_[0-9]_", '')) %>%
+    select(-disturb_year, -tree)
 
-    out_growth <- rbind(out_growth, growth_df)
+  correla_ws <- bind_rows(correla_ws, out)
 
-    # Compute resilience by trees 
-    aux_resilience <- c() 
-    trees <- unique(growth_df$tree)
-
-    for (t in trees){
-      # Filter by tree
-      df_tree <- growth_df %>% filter(tree == t)
-
-      # Compute resilience metrics by tree
-      aux_df <- df_tree %>% select(-c(tree, sd_period, se_period)) %>% 
-        spread(key=disturb, value=mean_period) %>% 
-        mutate(rt = dr / pre,
-               rc = post / dr,
-               rs = post / pre,
-               rrs = (post - dr) / pre,tree = t) %>% 
-        select(rt, rc, rs, rrs, disturb_year, tree)
-
-      aux_resilience <- rbind(aux_resilience, aux_df)
 }
 
-    out_res <- rbind(out_res, aux_resilience)
-  }
+# Split by window size 
+correla <- correla_ws %>% split(.$ws)
 
-  out$growth <- out_growth
-  out$resilience <- out_res
+# Change names 
+names(correla[["ws_2"]])[1:4] <- paste0(names(correla[["ws_2"]])[1:4], '2')
+names(correla[["ws_3"]])[1:4] <- paste0(names(correla[["ws_3"]])[1:4], '3')
 
-  return(out)
-}
+cor2 <- correla[["ws_2"]] %>% select(-ws) %>% mutate(ind = row_number())
+cor3 <- correla[["ws_3"]] %>% select(-ws) %>% mutate(ind = row_number())
 
-dyears <- c(1995, 2005, 2012)
-res_sj3 <- baiResilience(bai_sj, event_years = dyears, window = 3)
-res_sj2 <- baiResilience(bai_sj, event_years = dyears, window = 2)
+correlations <- inner_join(cor2, cor3, by='ind')
 
-resile2 <- res_sj2$resilience
-resile3 <- res_sj3$resilience
+# Resistance
+model <- lm(rt2~rt3, data=correlations)
+p_rt <- correlations %>% ggplot(aes(rt2, rt3)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3)))
 
-summary(lm(resile2$rs~resile3$rs))
+# Recovery
+model <- lm(rc2~rc3, data=correlations)
+p_rc <- correlations %>% ggplot(aes(rc2, rc3)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3)))
 
+# Resilience
+model <- lm(rs2~rs3, data=correlations)
+p_rs <- correlations %>% ggplot(aes(rs2, rs3)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3)))
+
+# Relative Resilience
+model <- lm(rrs2~rrs3, data=correlations)
+p_rrs <- correlations %>% ggplot(aes(rrs2, rrs3)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3)))
+
+grid.arrange(p_rc, p_rt, p_rs, p_rrs, ncol=2) 
 ```

### 2017-06-20 man/notes_history.Rmd
updating reslience BAI analysis

---
title: "Notas Historicas"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../references.bib
csl: ../ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Minería

## Mina La Probadora
* Antigua mina de cobre explotada desde finales del siglo XIX (1890) hasta 1957.
* Situada en el cortijo de su mismo nombre 
* Actualmente ruinas de antigua fundición 
* En su época de explotación, los materiales eran llevados a lomos de caballerías hasta la estación del tranvía de Sierra Nevada, que dispuso de un vagón especial para el transporte del mineral hasta Granada.

* Óxidos de Hierro
* Mineralización de cobre

* Hay una galería (la mas grande) que esta entibada

# Tranvía
* En 1941 había una propuesta de ampliación del tranvía de Sierra Nevada, desde el Maitena hasta la Estrella. Ver plano de José Pérez-Pozuelo (Colección Miguel Giménez Yanguas) en Luces de Sulayr

# Links

## Minería 
* http://www.mtiblog.com/2012/01/mina-la-probadora-guejar-sierra-granada.html 
* Plano Topográfico-Minero de una parte de Sierra Nevada [Material cartográfico]: Término de Güéjar Sierra, Provincia de Granada. 1858 http://www.ign.es/fondoscartograficos/ 
* http://www.foro-minerales.com/forum/viewtopic.php?p=69374 
* La Probadora: http://mining-andaluz.de/?page_id=4025

## References

### 2017-06-20 analysis/analysis_resilience_bai.Rmd
update notes; add csv with correlations

@@
 # SJ 
+res_4_sj <- baiResilience(bai_sj, event_years = dyears, window = 4)
 res_3_sj <- baiResilience(bai_sj, event_years = dyears, window = 3)
@@
 # caL
+res_4_caL <- baiResilience(bai_caL, event_years = dyears, window = 4)
 res_3_caL <- baiResilience(bai_caL, event_years = dyears, window = 3)
@@
 # caH
+res_4_caH <- baiResilience(bai_caH, event_years = dyears, window = 4)
 res_3_caH <- baiResilience(bai_caH, event_years = dyears, window = 3)
@@
 # Vector with objects name
-obj <- c('res_2_sj', 'res_3_sj',
-         'res_2_caL', 'res_3_caL',
-         'res_2_caH', 'res_3_caH')
-
-variables <- c('rt', 'rs', 'rc', 'rrs')
+obj <- c('res_2_sj', 'res_3_sj', 'res_4_sj',
+         'res_2_caL', 'res_3_caL', 'res_4_caL',
+         'res_2_caH', 'res_3_caH', 'res_4_caH')
 
@@
 names(correla[["ws_3"]])[1:4] <- paste0(names(correla[["ws_3"]])[1:4], '3')
+names(correla[["ws_4"]])[1:4] <- paste0(names(correla[["ws_4"]])[1:4], '4')
 
@@
 cor3 <- correla[["ws_3"]] %>% select(-ws) %>% mutate(ind = row_number())
+cor4 <- correla[["ws_4"]] %>% select(-ws) %>% mutate(ind = row_number())
 
-correlations <- inner_join(cor2, cor3, by='ind')
+correlations <- inner_join(cor2, cor3, by='ind') %>% inner_join(cor4, by='ind')
 
@@
 model <- lm(rt2~rt3, data=correlations)
-p_rt <- correlations %>% ggplot(aes(rt2, rt3)) + 
+p_rt23 <- correlations %>% ggplot(aes(rt2, rt3)) + 
   geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3)))
+  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) + 
+  theme(legend.position = c(.2, .75))
+
+model <- lm(rt2~rt4, data=correlations)
+p_rt24 <- correlations %>% ggplot(aes(rt2, rt4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+
+model <- lm(rt3~rt4, data=correlations)
+p_rt34 <- correlations %>% ggplot(aes(rt3, rt4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+
+grid.arrange(p_rt23, p_rt24, p_rt34,ncol=3)
 
@@
 model <- lm(rc2~rc3, data=correlations)
-p_rc <- correlations %>% ggplot(aes(rc2, rc3)) + 
+p_rc23 <- correlations %>% ggplot(aes(rc2, rc3)) + 
   geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3)))
+  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) + 
+  theme(legend.position = c(.2, .75))
+
+model <- lm(rc2~rc4, data=correlations)
+p_rc24 <- correlations %>% ggplot(aes(rc2, rc4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+
+model <- lm(rc3~rc4, data=correlations)
+p_rc34 <- correlations %>% ggplot(aes(rc3, rc4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+
+grid.arrange(p_rc23, p_rc24, p_rc34,ncol=3)
 
@@
 model <- lm(rs2~rs3, data=correlations)
-p_rs <- correlations %>% ggplot(aes(rs2, rs3)) + 
+p_rs23 <- correlations %>% ggplot(aes(rs2, rs3)) + 
   geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3)))
+  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) + 
+  theme(legend.position = c(.2, .75))
+
+model <- lm(rs2~rs4, data=correlations)
+p_rs24 <- correlations %>% ggplot(aes(rs2, rs4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+
+model <- lm(rs3~rs4, data=correlations)
+p_rs34 <- correlations %>% ggplot(aes(rs3, rs4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+
+grid.arrange(p_rs23, p_rs24, p_rs34,ncol=3)
 
@@
 model <- lm(rrs2~rrs3, data=correlations)
-p_rrs <- correlations %>% ggplot(aes(rrs2, rrs3)) + 
+p_rrs23 <- correlations %>% ggplot(aes(rrs2, rrs3)) + 
   geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3)))
+  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) + 
+  theme(legend.position = c(.2, .75))
 
-grid.arrange(p_rc, p_rt, p_rs, p_rrs, ncol=2) 
+model <- lm(rrs2~rrs4, data=correlations)
+p_rrs24 <- correlations %>% ggplot(aes(rrs2, rrs4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+
+model <- lm(rrs3~rrs4, data=correlations)
+p_rrs34 <- correlations %>% ggplot(aes(rrs3, rrs4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+
+grid.arrange(p_rrs23, p_rrs24, p_rrs34,ncol=3)
+
 ```

### 2017-06-20 man/notes_dendro_analysis.Rmd
update notes; add csv with correlations

@@
 
+* Utilizamos BAI por árbol 
+* Computamos índices de resiliencia de [@Lloret2011]: Resistance (rt), Recovery (rc), Resilience (rs) y Relative Resilience (rrs). 
+* Computamos para 3 eventos de sequía (:red_circle: Mirar esto y decidir que años): 1995, 2005 y 2012
+* Utilizo la la función `baiResilience` (código aquí: [./script/R/baiResilience.R](/script/R/baiResilience.R)).
+* Utilizamos diferentes tamaños ventana para los crecimientos previo y posterior (2, 3 y 4) (ver aproximación de [@Gazol2017]) y computamos correlaciones entre resultados de cada métrica.
+Los resultados de las correlaciones se guardan [aquí](/out/correla_resilience/correla_window_size.csv)
+
 ### [@Gazol2017]

### 2017-06-20 man/notes_dendro_analysis.Rmd
update links

@@
 Puedo calcularlas con la función `chron()` pero esto no me permite obtener sd, se del bai por año.
-Por ello me creo una función llamada `chrono_bai` (código aquí: [./script/R/chrono_bai.R](/script/R/gea/chrono_bai.R))
+Por ello me creo una función llamada `chrono_bai` (código aquí: [./script/R/chrono_bai.R](/script/R/chrono_bai.R))
 
@@
     * Cada cronologia es suavizada usando centred moving averages con diferentes windows sizes.
-Para ello utilizamos la función `suaviza_cronos` (código aquí: [./script/R/suaviza_cronos.R](/script/R/gea/suaviza_cronos.R))
+Para ello utilizamos la función `suaviza_cronos` (código aquí: [./script/R/suaviza_cronos.R](/script/R/suaviza_cronos.R))
     * Posteriormente se calcula la correlación entre chronos para cada suavizado. 
@@
 
-Para su cómputo hemos utilizado la función `computeGC` (código aquí: [./script/R/computeGC.R](/script/R/gea/computeGC.R)). En concreto hemos llevado a cabo lo siguiente:
+Para su cómputo hemos utilizado la función `computeGC` (código aquí: [./script/R/computeGC.R](/script/R/computeGC.R)). En concreto hemos llevado a cabo lo siguiente:
 

### 2017-06-20 analysis/analysis_resilience_bai.Rmd
update resilience

@@
 library('gridExtra')
+library('pander')
 ```
@@
 
-* Computar correlaciones ventanas temporales
-
+### Computar correlaciones ventanas temporales 
 ```{r}
@@
 # Resistance
+aux_coefs <- c()
+
 model <- lm(rt2~rt3, data=correlations)
@@
   theme(legend.position = c(.2, .75))
+aux <- as.data.frame(cbind('rt','2-3', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rt','2-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rt','3-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = c(.2, .75))
+aux <- as.data.frame(cbind('rc','2-3', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rc','2-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rc','3-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = c(.2, .75))
+aux <- as.data.frame(cbind('rs','2-3', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rs','2-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rs','3-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = c(.2, .75))
+aux <- as.data.frame(cbind('rrs','2-3', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rrs','2-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
@@
   theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rrs','3-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
 
 grid.arrange(p_rrs23, p_rrs24, p_rrs34,ncol=3)
+
+names(aux_coefs) <- c('var', 'window_size', 'r2')
+
+write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), row.names = F)
+```
+
+Nos quedamos con 3 años de ventana temporal.
+
+# Plots Crecimiento
+```{r}
+gsj <- res_3_sj$growth %>% mutate(site='SJ')
+gcaL <- res_3_caL$growth %>% mutate(site='caL')
+gcaH <- res_3_caH$growth %>% mutate(site='caH')
+
+g <- bind_rows(gsj, gcaL, gcaH)
+
+g %>% mutate(disturb = recode(disturb, 
+                              dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  ggplot(aes(y=mean_period/100, x=disturb)) + 
+  geom_boxplot() + facet_grid(site~disturb_year, scales='free_y') + 
+  theme_bw()
+```
+
+```{r}
+g %>% 
+  mutate(disturb = recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  group_by(disturb, disturb_year, site) %>% 
+  summarise(mean = mean(mean_period),
+            sd = sd(mean_period),
+            se = sd/sqrt(length(mean_period))) %>%
+  ggplot(aes(y=mean/100, x=disturb)) + 
+  geom_errorbar(aes(ymin=mean/100 - se/100, 
+                    ymax=mean/100 + se/100),
+                width = 0.2) +
+  geom_point(size=2, shape=21, fill='white') + 
+  facet_grid(site~disturb_year, scales='free_y') + 
+  theme_bw() + ylab('BAI (cm2/year)') + 
+  theme(panel.grid = element_blank())
+```
+
+```{r}
+g %>% 
+  mutate(disturb = recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  group_by(site, disturb_year, disturb) %>% 
+  summarise(mean = mean(mean_period/100),
+            sd = sd(mean_period/100),
+            se = sd/sqrt(length(mean_period/100))) %>% pander() 
+```
+
+# Plots Resiliencia
+```{r}
+rsj <- res_3_sj$resilience %>% mutate(site='SJ')
+rcaL<- res_3_caL$resilience %>% mutate(site='caL')
+rcaH <- res_3_caH$resilience %>% mutate(site='caH')
+
+re <- bind_rows(rsj, rcaL, rcaH)
 

### 2017-06-20 man/notes_dendro_analysis.Rmd
update resilience

@@
 * Utilizamos diferentes tamaños ventana para los crecimientos previo y posterior (2, 3 y 4) (ver aproximación de [@Gazol2017]) y computamos correlaciones entre resultados de cada métrica.
-Los resultados de las correlaciones se guardan [aquí](/out/correla_resilience/correla_window_size.csv)
+Los resultados de las correlaciones se guardan [aquí](/out/correla_resilience/correla_window_size.csv).
+* Observamos que no existen diferencias en los resultados en función del tamaño de ventana que utilizamos (altas correlaciones) y decidimos utilizar 3 años: no solapamiento entre 2005 y 2012; se ha utilizado en otros trabajos.
 

### 2017-06-21 man/notes_dendro_analysis.Rmd
update notes

@@
 
-* Utilizamos BAI por árbol 
-* Computamos índices de resiliencia de [@Lloret2011]: Resistance (rt), Recovery (rc), Resilience (rs) y Relative Resilience (rrs). 
-* Computamos para 3 eventos de sequía (:red_circle: Mirar esto y decidir que años): 1995, 2005 y 2012
-* Utilizo la la función `baiResilience` (código aquí: [./script/R/baiResilience.R](/script/R/baiResilience.R)).
-* Utilizamos diferentes tamaños ventana para los crecimientos previo y posterior (2, 3 y 4) (ver aproximación de [@Gazol2017]) y computamos correlaciones entre resultados de cada métrica.
-Los resultados de las correlaciones se guardan [aquí](/out/correla_resilience/correla_window_size.csv).
-* Observamos que no existen diferencias en los resultados en función del tamaño de ventana que utilizamos (altas correlaciones) y decidimos utilizar 3 años: no solapamiento entre 2005 y 2012; se ha utilizado en otros trabajos.
-
+## Revision otros trabajos 
 ### [@Gazol2017]
 * Para cada año seco y en cada sitio se calculó la resiliencia del crecimiento de los árboles utilizando tres indices propuestos por [@Lloret2011], en los cuales, la respuesta del crecimiento del árbol a las condiciones secas se descompone en:
+
     * Resistance (Rt): quantifies the difference between tree growth in the dry year and the years before it.
@@
 
-* Selección de ventana de 3 años, basado en
+## Aproximación utilizada
+
+* Utilizamos BAI por árbol y computamos índices de resiliencia de [@Lloret2011]: Resistance (rt), Recovery (rc), Resilience (rs) y Relative Resilience (rrs). 
+* 3 eventos de sequía (:red_circle: Mirar esto y decidir que años): 1995, 2005 y 2012
+* Utilizo la la función `baiResilience` (código aquí: [./script/R/baiResilience.R](/script/R/baiResilience.R)).
+* Utilizamos diferentes tamaños ventana para los crecimientos previo y posterior (2, 3 y 4) (ver aproximación de [@Gazol2017]) y computamos correlaciones entre resultados de cada métrica.
+Los resultados de las correlaciones se guardan [aquí](/out/correla_resilience/correla_window_size.csv).
+* Observamos que no existen diferencias en los resultados en función del tamaño de ventana que utilizamos (altas correlaciones) y decidimos utilizar 3 años: no solapamiento entre 2005 y 2012; se ha utilizado en otros trabajos.
+En el archivo [./out/correla_resilience/correla_window_size.csv](/out/correla_resilience/correla_window_size.csv) están los resultados de las correlaciones utilizando diferentes tamaños de ventana.
+
+* Seleccionamos 3 años como ventana temporal. 
+* Generamos dos conjuntos de datos para su ploteo y análisis:
+
+    * Crecimientos: dataframe con valores de BAI por árbol (*n* = 50) para cada evento de sequía (*disturb_year*, *n*=3: 1995, 2005, 2012) y para cada periodo (*disturb*; *n*=3, *prev*, *dr* y *post*) (OJO para *prev* y *post* se trata del crecimiento medio para cada árbol de 3 años, por eso aparecen datos de sd y se a nivel de árbol).
+El archivo se guarda en: [./data/resilience/crecimientos_drought.csv](data/resilience/crecimientos_drought.csv)
+    * Resiliencia: dataframe con los valores de las métricas de resiliencia por árbol (*n* = 50) para cada evento de sequía (*disturb_year*, *n*=3: 1995, 2005, 2012).
+Cada metrica se almacena en una variable.
+Se incluye también otra variable de sitio (también implicita en tree).
+El archivo se guarda en [./data/resilience/resilience_bai.csv](data/resilience/resilience_bai.csv)
 

### 2017-06-21 man/notes_dendro_analysis.Rmd
fix link error

@@
     * Crecimientos: dataframe con valores de BAI por árbol (*n* = 50) para cada evento de sequía (*disturb_year*, *n*=3: 1995, 2005, 2012) y para cada periodo (*disturb*; *n*=3, *prev*, *dr* y *post*) (OJO para *prev* y *post* se trata del crecimiento medio para cada árbol de 3 años, por eso aparecen datos de sd y se a nivel de árbol).
-El archivo se guarda en: [./data/resilience/crecimientos_drought.csv](data/resilience/crecimientos_drought.csv)
+El archivo se guarda en: [./data/resilience/crecimientos_drought.csv](/data/resilience/crecimientos_drought.csv)
     * Resiliencia: dataframe con los valores de las métricas de resiliencia por árbol (*n* = 50) para cada evento de sequía (*disturb_year*, *n*=3: 1995, 2005, 2012).
@@
 Se incluye también otra variable de sitio (también implicita en tree).
-El archivo se guarda en [./data/resilience/resilience_bai.csv](data/resilience/resilience_bai.csv)
+El archivo se guarda en [./data/resilience/resilience_bai.csv](/data/resilience/resilience_bai.csv)
 

### 2017-06-21 analysis/analysis_resilience_bai.Rmd
Resilience anovas

@@
 library('pander')
+library('broom')
+library('effects')
 ```
@@
 
+```{r}
+aux_coefs %>% pander()
+```
+
 Nos quedamos con 3 años de ventana temporal.
@@
 # Plots Crecimiento
+## Boxplot with outliers 
 ```{r}
@@
 
+# Export csv 
+write.csv(g, file=paste0(di, 'data/resilience/crecimientos_drought.csv'), row.names = FALSE)
+
 g %>% mutate(disturb = recode(disturb, 
@@
   ggplot(aes(y=mean_period/100, x=disturb)) + 
-  geom_boxplot() + facet_grid(site~disturb_year, scales='free_y') + 
+  geom_boxplot() + 
+  facet_grid(site~disturb_year, scales='free_y') + 
   theme_bw()
@@
 
+## Mean + se 
 ```{r}
@@
 ```{r}
+pgrowht <- g %>% 
+  mutate(disturb = recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  group_by(disturb, disturb_year, site) %>% 
+  summarise(mean = mean(mean_period),
+            sd = sd(mean_period),
+            se = sd/sqrt(length(mean_period))) %>%
+  ggplot(aes(y=mean/100, x=disturb, colour=site)) + 
+  geom_errorbar(aes(ymin=mean/100 - se/100, 
+                    ymax=mean/100 + se/100),
+                width = 0.15) +
+  geom_point(size=3, aes(shape=site), fill='white') + 
+  geom_line(aes(group=site))+
+  facet_wrap(~disturb_year, scales='free_y') + 
+  theme_bw() + ylab('BAI (cm2/year)') + 
+  theme(panel.grid = element_blank())
+
+pgrowht
+ggsave(plot=pgrowht, width=8, height = 4,
+       filename=paste0(di, 'out/fig/resilience/bai_events.pdf')) 
+```
+
+```{r}
 g %>% 
@@
 
-# Plots Resiliencia
+# Anovas Resiliencia
+
 ```{r}
+# Prepara data 
 rsj <- res_3_sj$resilience %>% mutate(site='SJ')
@@
 re <- bind_rows(rsj, rcaL, rcaH)
+re$disturb_year <- as.factor(re$disturb_year)
+re$site <- as.factor(re$site)
 
+# Export csv 
+write.csv(re, file=paste0(di, 'data/resilience/resilience_bai.csv'), row.names = FALSE)
+```
+
+## Custom functions
+
+```{r}
+# Custom Function to compute ANOVAS
+aovas <- function(df, vars, resp_var){ 
+  require('dplyr')
+  require('broom')
+
+  # Create subset 
+  dfsel <- df %>% dplyr::select_(.dots=c(vars, resp_var))
+
+  # Model 
+  myformula <- as.formula(paste0(resp_var,  " ~ ",
+                                 paste(vars, collapse = '*')))
+
+  mymodel <- aov(myformula, data=dfsel)
+
+  # Output model Summary http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1)
+  model_coeff <- broom::tidy(mymodel)
+  model_summary <- broom::glance(mymodel)
+
+  out <- c() 
+  out$model_coeff <- model_coeff
+  out$model_summary <- model_summary
+  out$mymodel <- mymodel
+
+  return(out)
+}
+
+# Post-Hoc comparison
+phc <- function(mymodel, resp_var){
+  require(lsmeans)
+
+  # Disturb Event 
+  ph_event <- lsmeans(mymodel, pairwise ~ disturb_year, adjust = "tukey")
+
+  # differences letters 
+  cld_event <- cld(ph_event, alpha   = 0.01, 
+                   Letters = letters, 
+                   adjust  = "tukey")
+
+  # Site  
+  ph_site <- lsmeans(mymodel, pairwise ~ site, adjust = "tukey")
+  cld_site <- cld(ph_site, alpha   = 0.01, 
+                 Letters = letters, 
+                 adjust  = "tukey")
+
+  # interaction 
+  ph_i <- lsmeans(mymodel, pairwise ~ disturb_year:site, adjust = "tukey")
+
+  # Objets for plot
+  aux_ph_site <- as.data.frame(summary(ph_site$lsmeans)) 
+  aux_ph_site <- aux_ph_site %>% mutate(var = resp_var)
+  aux_ph_event <- as.data.frame(summary(ph_event$lsmeans)) 
+  aux_ph_event <- aux_ph_event %>% mutate(var = resp_var)
+  aux_ph_i <- as.data.frame(summary(ph_i$lsmeans)) 
+  aux_ph_i <- aux_ph_i %>% mutate(var = resp_var)
+
+  # Return objects
+  cat('\n### Event ###\n')
+  print(ph_event)
+  print(cld_event)
+  cat('\n### Clu pop ###\n')
+  print(ph_site)
+  print(cld_site)
+  cat('\n### Event:Clu pop ###\n')
+  print(ph_i)
+  return(list(aux_ph_site, aux_ph_event, aux_ph_i))
+}
+```
+
+```{r}
+vars <- c('disturb_year','site')
+```
+
+## Recovery 
+```{r}
+# Variable
+resp_var <- 'rc'
+
+# AOV
+aov_rc <- aovas(re, vars=vars, resp_var = resp_var)
+```
+
+```{r}
+mc <- aov_rc$model_coeff
+
+pander(mc, round=5,
+       caption = paste0("ANOVA table: ", resp_var), missing = '', 
+       emphasize.strong.cells = 
+         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
+```
+
+```{r}
+gm <- aov_rc$model_summary
+
+gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
+colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
+rownames(gm) <- "Statistic"
+pander(t(gm)) 
+```
+
+### Post hoc comparison
+
+```{r}
+# Post hoc Define model
+mymodel <- aov_rc$mymodel
+postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
+```
+
+### Plots
+
+```{r}
+#### ~ Site
+ps <- plot(effect("site",mymodel))
+#### ~ Disturb Year
+pd <- plot(effect('disturb_year', mymodel))
+#### Disturb Year:Site
+picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(3,1))
+```
+
+```{r}
+ps
+```
+
+```{r}
+pd
+```
+
+```{r}
+picollapse
+```
+
+```{r}
+pi
+```
+
+## Resistance 
+```{r}
+# Variable
+resp_var <- 'rt'
+
+# AOV
+aov_rt <- aovas(re, vars=vars, resp_var = resp_var)
+```
+
+```{r}
+mc <- aov_rt$model_coeff
+
+pander(mc, round=5,
+       caption = paste0("ANOVA table: ", resp_var), missing = '', 
+       emphasize.strong.cells = 
+         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
+```
+
+```{r}
+gm <- aov_rt$model_summary
+
+gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
+colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
+rownames(gm) <- "Statistic"
+pander(t(gm)) 
+```
+
+### Post hoc comparison
+
+```{r}
+# Post hoc Define model
+mymodel <- aov_rt$mymodel
+postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
+```
+
+### Plots
+
+```{r}
+#### ~ Site
+ps <- plot(effect("site",mymodel))
+#### ~ Disturb Year
+pd <- plot(effect('disturb_year', mymodel))
+#### Disturb Year:Site
+picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(3,1))
+```
+
+```{r}
+ps
+```
+
+```{r}
+pd
+```
+
+```{r}
+picollapse
+```
+
+```{r}
+pi
+```
+
+## Relative Resilience 
+```{r}
+# Variable
+resp_var <- 'rrs'
+
+# AOV
+aov_rrs <- aovas(re, vars=vars, resp_var = resp_var)
+```
+
+```{r}
+mc <- aov_rrs$model_coeff
+
+pander(mc, round=5,
+       caption = paste0("ANOVA table: ", resp_var), missing = '', 
+       emphasize.strong.cells = 
+         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
+```
+
+```{r}
+gm <- aov_rrs$model_summary
+
+gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
+colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
+rownames(gm) <- "Statistic"
+pander(t(gm)) 
+```
+
+### Post hoc comparison
+
+```{r}
+# Post hoc Define model
+mymodel <- aov_rrs$mymodel
+postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
+```
+
+### Plots
+
+```{r}
+#### ~ Site
+ps <- plot(effect("site",mymodel))
+#### ~ Disturb Year
+pd <- plot(effect('disturb_year', mymodel))
+#### Disturb Year:Site
+picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(3,1))
+```
+
+```{r}
+ps
+```
+
+```{r}
+pd
+```
+
+```{r}
+picollapse
+```
+
+```{r}
+pi
+```
+
+## Resilience 
+```{r}
+# Variable
+resp_var <- 'rs'
+
+# AOV
+aov_rs <- aovas(re, vars=vars, resp_var = resp_var)
+```
+
+```{r}
+mc <- aov_rs$model_coeff
+
+pander(mc, round=5,
+       caption = paste0("ANOVA table: ", resp_var), missing = '', 
+       emphasize.strong.cells = 
+         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
+```
+
+```{r}
+gm <- aov_rs$model_summary
+
+gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
+colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
+rownames(gm) <- "Statistic"
+pander(t(gm)) 
+```
+
+### Post hoc comparison
+
+```{r}
+# Post hoc Define model
+mymodel <- aov_rs$mymodel
+postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
+```
+
+### Plots
+
+```{r}
+#### ~ Site
+ps <- plot(effect("site",mymodel))
+#### ~ Disturb Year
+pd <- plot(effect('disturb_year', mymodel))
+#### Disturb Year:Site
+picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(3,1))
+```
+
+```{r}
+ps
+```
+
+```{r}
+pd
+```
+
+```{r}
+picollapse
+```
+
+```{r}
+pi
 ```

### 2017-06-21 analysis/analysis_resilience_bai.Rmd
update references analysis

@@
     variant: markdown_github
-bibliography: references.bib 
-csl: ecology.csl
+bibliography: ../references.bib 
+csl: ../ecology.csl
 ---

### 2017-06-21 analysis/analysis_resilience_bai.Rmd
update only three years

@@
 ```{r}
+# Only 2005 and 2012
+re <- re %>% filter(disturb_year != 1995) %>% as.data.frame()
 vars <- c('disturb_year','site')
@@
 mymodel <- aov_rt$mymodel
-postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
+postH_rt <- phc(mymodel = mymodel, resp_var = resp_var)
 ```
@@
 mymodel <- aov_rrs$mymodel
-postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
+postH_rrs <- phc(mymodel = mymodel, resp_var = resp_var)
 ```
@@
 mymodel <- aov_rs$mymodel
-postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
+postH_rs <- phc(mymodel = mymodel, resp_var = resp_var)
 ```

### 2017-06-21 analysis/analysis_resilience_bai.Rmd
update plots two events

@@
   ph_i <- lsmeans(mymodel, pairwise ~ disturb_year:site, adjust = "tukey")
+  cld_i <- cld(ph_i, alpha   = 0.01, 
+                 Letters = letters, 
+                 adjust  = "tukey")
 
@@
   print(ph_i)
-  return(list(aux_ph_site, aux_ph_event, aux_ph_i))
+  return(list(aux_ph_site, aux_ph_event, aux_ph_i, cld_site, cld_event, cld_i))
 }
@@
 
+```{r}
+# Interactions plot all  
+means_site <- postH_rc[[4]] %>% mutate(var = 'rc') %>% 
+  bind_rows(postH_rt[[4]] %>% mutate(var = 'rt')) %>% 
+  bind_rows(postH_rs[[4]] %>% mutate(var = 'rs')) %>% 
+  bind_rows(postH_rrs[[4]] %>% mutate(var = 'rrs')) %>% 
+  rename(letras = .group)
+
+means_disturb <- postH_rc[[5]] %>% mutate(var = 'rc') %>% 
+  bind_rows(postH_rt[[5]] %>% mutate(var = 'rt')) %>% 
+  bind_rows(postH_rs[[5]] %>% mutate(var = 'rs')) %>% 
+  bind_rows(postH_rrs[[5]] %>% mutate(var = 'rrs')) %>% 
+  rename(letras = .group)
+
+means_distub_site <- postH_rc[[6]] %>% mutate(var = 'rc') %>% 
+  bind_rows(postH_rt[[6]] %>% mutate(var = 'rt')) %>% 
+  bind_rows(postH_rs[[6]] %>% mutate(var = 'rs')) %>% 
+  bind_rows(postH_rrs[[6]] %>% mutate(var = 'rrs')) %>% 
+  rename(letras = .group)
+
+dodge <- position_dodge(width = 0.3)
+micolor <- '#455883'
+```
+
+```{r}
+means_site %>%  
+  ggplot(aes(x=site, y=lsmean)) + 
+  geom_errorbar(aes(ymin=lsmean - SE, ymax=lsmean + SE), 
+                color=micolor, size=.5, width=.15, position = dodge) +
+  geom_point(colour=micolor, 
+             size=3, position = dodge) +
+  theme_bw() + xlab('') + ylab('') + 
+  facet_wrap(~var, scales='free_y', ncol = 1) +
+  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15)
+
+```
+
+```{r}
+means_disturb %>%  
+  ggplot(aes(x=disturb_year, y=lsmean)) + 
+  geom_errorbar(aes(ymin=lsmean - SE, ymax=lsmean + SE), 
+                color=micolor, size=.5, width=.15, position = dodge) +
+  geom_point(colour=micolor, 
+             size=3, position = dodge) +
+  theme_bw() + xlab('') + ylab('') + 
+  facet_wrap(~var, scales='free_y', ncol = 1) +
+  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15)
+```
+
+```{r}
+means_distub_site %>%  
+  ggplot(aes(x=site, y=lsmean, group=disturb_year, colour=disturb_year)) + 
+  geom_errorbar(aes(ymin=lsmean - SE, ymax=lsmean + SE), 
+                size=.5, width=.15) + 
+  geom_point(aes(shape=disturb_year), size=3) + 
+  geom_line() +
+  theme_bw() + xlab('') + ylab('') + 
+  facet_wrap(~var, scales='free_y', ncol = 1) +
+  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15)
+```
+
 # References

### 2017-06-22 analysis/analysis_resilience_bai.Rmd
update resilience analisys bai

@@
 ```{r}
-means_site %>%  
+plot_ms <- means_site %>%  
   ggplot(aes(x=site, y=lsmean)) + 
@@
   facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15)
-
+  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15) +
+  theme(strip.background = element_rect(colour = "black", fill = "white"))
 ```
@@
 ```{r}
-means_disturb %>%  
+plot_md <- means_disturb %>%  
   ggplot(aes(x=disturb_year, y=lsmean)) + 
@@
   facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15)
+  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15) +
+  theme(strip.background = element_rect(colour = "black", fill = "white"))
 ```
@@
 ```{r}
-means_distub_site %>%  
+plot_mds <- means_distub_site %>%  
   ggplot(aes(x=site, y=lsmean, group=disturb_year, colour=disturb_year)) + 
@@
   facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15)
+  geom_text(aes(y=lsmean+SE, label=letras), nudge_x = 0.15)+
+  theme(strip.background = element_rect(colour = "black", fill = "white"),
+        legend.position = c(0.8, 0.93),
+        legend.background = element_blank()) +
+  scale_colour_manual(values = c(micolor, "red"))
+
+```
+
+```{r}
+pdf(paste0(di, 'out/fig/resilience/interaction_plots.pdf'), width=9, height = 9)
+grid.arrange(plot_md, plot_ms, plot_mds, ncol=3)
+dev.off()
+```
+
+```{r}
+aovas_coeff <- aov_rc$model_coeff %>% mutate(var = 'rc') %>% 
+  bind_rows(aov_rt$model_coeff %>% mutate(var = 'rt')) %>% 
+  bind_rows(aov_rs$model_coeff %>% mutate(var = 'rs')) %>% 
+  bind_rows(aov_rrs$model_coeff%>% mutate(var = 'rrs'))
+
+write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/anovas_statistics.csv'), row.names = F)
+
+aovas_coeff %>% pander()
+```
+
+```{r}
+aovas_model_summary <- aov_rc$model_summary %>% mutate(var = 'rc') %>% 
+  bind_rows(aov_rt$model_summary %>% mutate(var = 'rt')) %>% 
+  bind_rows(aov_rs$model_summary %>% mutate(var = 'rs')) %>% 
+  bind_rows(aov_rrs$model_summary%>% mutate(var = 'rrs'))
+
+write.csv(aovas_model_summary, 
+          file=paste0(di, '/out/anovas_resilience/anovas_summary_modelos.csv'), row.names = F)
+
+gm <- apply(aovas_model_summary, 1, formatC, digits = 2, format = "f") 
+rownames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e", "variable"),"$")
+colnames(gm) <- c("rc", "rt", "rs", "rrs")
+
+pander(gm)
 ```

### 2017-06-22 man/notes_dendro_analysis.Rmd
Resilience anovas
update resilience analisys bai

@@
 
+### Gráficos exploratorios
+
+* Plot crecimiento medio (BAI) previo, durante y posterior a cada evento de sequía en los tres sitios. Ver `/out/fig/resilience/bai_events.pdf`.
+
+### Análisis ANOVAS
+* Modelos: var_res ~ site + disturb_year + site:disturb_year
+* Utilizamos solo 2005 y 2012 
+* Summary ANOVAS:
+
+    * estadísticos y demás en este archivo [./out/anovas_resilience/anovas_statistics.csv](/out/anovas_resilience/anovas_statistics.csv)
+    * Summary modelo: [./out/anovas_resilience/anovas_summary_modelos.csv](/out/anovas_resilience/anovas_summary_modelos.csv).
+
+* Interaction plots Anovas: comparación valores medios ± SE por evento de sequía, sitio e interacción:  Ver `/out/fig/resilience/interaction_plots.pdf`.
+
 ### notas sueltas de correos

### 2017-06-22 analysis/analysis_resilience_bai.Rmd
update plot interactions

@@
 
+```
+
+```{r}
 dodge <- position_dodge(width = 0.3)
 micolor <- '#455883'
+mierrorbarSE <- aes(ymin=lsmean - SE, ymax=lsmean + SE)
+mierrorbar <- aes(ymin=lower.CL, ymax=upper.CL)
 ```
@@
   ggplot(aes(x=site, y=lsmean)) + 
-  geom_errorbar(aes(ymin=lsmean - SE, ymax=lsmean + SE), 
-                color=micolor, size=.5, width=.15, position = dodge) +
   geom_point(colour=micolor, 
@@
   theme(strip.background = element_rect(colour = "black", fill = "white"))
+
+plot_msSE <- plot_ms + geom_errorbar(mierrorbarSE,color=micolor, 
+                                     size=.5, width=.15, position = dodge)
+
+plot_msCI <- plot_ms + geom_errorbar(mierrorbar,color=micolor, 
+                                     size=.5, width=.15, position = dodge)
 ```
@@
   ggplot(aes(x=disturb_year, y=lsmean)) + 
-  geom_errorbar(aes(ymin=lsmean - SE, ymax=lsmean + SE), 
-                color=micolor, size=.5, width=.15, position = dodge) +
   geom_point(colour=micolor, 
@@
   theme(strip.background = element_rect(colour = "black", fill = "white"))
+
+plot_mdSE <- plot_md + geom_errorbar(mierrorbarSE,color=micolor, 
+                                     size=.5, width=.15, position = dodge)
+
+plot_mdCI <- plot_md + geom_errorbar(mierrorbar,color=micolor, 
+                                     size=.5, width=.15, position = dodge)
 ```
@@
   ggplot(aes(x=site, y=lsmean, group=disturb_year, colour=disturb_year)) + 
-  geom_errorbar(aes(ymin=lsmean - SE, ymax=lsmean + SE), 
-                size=.5, width=.15) + 
   geom_point(aes(shape=disturb_year), size=3) + 
@@
 
+plot_mdsSE <- plot_mds + geom_errorbar(mierrorbarSE, size=.5, width=.15)
+plot_mdsCI <- plot_mds + geom_errorbar(mierrorbar, size=.5, width=.15)
 ```
@@
 ```{r}
-pdf(paste0(di, 'out/fig/resilience/interaction_plots.pdf'), width=9, height = 9)
-grid.arrange(plot_md, plot_ms, plot_mds, ncol=3)
+pdf(paste0(di, 'out/fig/resilience/interaction_plotsSE.pdf'), width=9, height = 9)
+grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
 dev.off()
+
+pdf(paste0(di, 'out/fig/resilience/interaction_plotsCI.pdf'), width=9, height = 9)
+grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
+dev.off()
+
 ```

### 2017-06-22 man/notes_dendro_analysis.Rmd
update notes

@@
 
-* Interaction plots Anovas: comparación valores medios ± SE por evento de sequía, sitio e interacción:  Ver `/out/fig/resilience/interaction_plots.pdf`.
+* Interaction plots Anovas: comparación valores medios ± SE por evento de sequía, sitio e interacción:  Ver `/out/fig/resilience/interaction_plotsCI.pdf` o `/out/fig/resilience/interaction_plotsSE.pdf`.
+La diferencia esta en utilizar CI o SE.
 

### 2017-08-16 man/ms_material_methods.Rmd
update bib file
update material and methods

@@
 
-## Tree sampling 
+## Study sites
+
+Aqui meter descripción de SN y de la distribución del Qp en Sierra Nevada.
+
+## Field sampling and dendrochronological methods
+
+### Tree sampling 
 Samplig was carried during autumn of 2016.
 Trees were sampled at two locations located in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Table 1).
-Both sites were oak monospecific and representatives of two of the three the population's cluster identified for this specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..).
-In each site between 15 and 20 dominant trees were sampled.
+Both sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..).
+In each site between 15 and 20 dominant trees were randomly selected.
 Two cores of 5 mm of diameter were taken per tree at 1.3 m using an increment borer.
@@
 
+## Statistical analysis
+
 ## References

### 2017-08-17 man/ms_0_intro.Rmd
update structure ms

--- 2017-05-31 man/material_methods.Rmd
@@
 ---
-title: "material and methods"
+title: "ms_0_intro"
 author: "AJ Perez-Luque (@ajpelu); G. Gea-Izquierdo"
@@
 
-# Materials and methods
-
-## Dendrochronological methods
-... dendro
-For each focal tree we measured diameter at breast height (DBH) and total height. 
-A total of xx trees were sampled. 
-...
-
-We built chronologies for each site (two)
-
-Site chronologies were built by averaging all tree BAI measurement of the same site.
-To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
-
-```{r}
-py <- pointer.rgc(s033, nb.yrs=4, rgc.thresh.pos=60, rgc.thresh.neg=40, series.thresh=75)
-
-rgc.plot(py,start.yr=1950,end.yr=NULL,sd.disp=FALSE
-```
+# Intro
 

### 2017-08-17 man/ms_2_results.Rmd
update structure ms

--- 2017-05-31 man/material_methods.Rmd
@@
 ---
-title: "material and methods"
+title: "ms_2_results"
 author: "AJ Perez-Luque (@ajpelu); G. Gea-Izquierdo"
@@
 
-# Materials and methods
-
-## Dendrochronological methods
-... dendro
-For each focal tree we measured diameter at breast height (DBH) and total height. 
-A total of xx trees were sampled. 
-...
-
-We built chronologies for each site (two)
-
-Site chronologies were built by averaging all tree BAI measurement of the same site.
-To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
-
-```{r}
-py <- pointer.rgc(s033, nb.yrs=4, rgc.thresh.pos=60, rgc.thresh.neg=40, series.thresh=75)
-
-rgc.plot(py,start.yr=1950,end.yr=NULL,sd.disp=FALSE
-```
+# Results
 

### 2017-08-17 man/ms_0_intro.Rmd
add aims of study

@@
 
+* The vegetation response to water deficit was assessed using remote sensing information and dendrocronologial data.
+
+:red_circle: Algo de las sequías
+
+## Aims
+`$IMPROVE$`
+
+In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both greenees and growth of *Q. pyrenaica* forests in Sierra Nevada. Specifically,
+
+The aims of this work were:
+
+* To evaluate how extreme drought events affected the greennes and growth of *Q. pyreancia* forests in ther rear edge,
+
+* to analyze the resilience of these forests to sucessive extrme drought events, 
+* and to explore differences in the resilience metrics between populations located in contrasting slopes within the rear edge of the distribution of this species.
+
 ## References

### 2017-08-17 man/ms_1_material_methods.Rmd
update structure ms
add content about site and species

--- 2017-08-16 man/ms_material_methods.Rmd
@@
 bibliography: ../references.bib
-csl: ../ecology.csl
+csl: ../geb.csl
 ---
@@
 
-## Study sites
+## Species and study site
 
-Aqui meter descripción de SN y de la distribución del Qp en Sierra Nevada.
+The Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula [@Franco1990] reaching its southern limit in north of Morocco.
+In the Iberian Peninsula these forests live under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002] living on siliceous soils, or soils poor in basic ions [@Vilches2014].
+*Q.
+pyrenaica* requires between 650 and 1200 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm (Martínez-Parras and Molero-Mesa 1982, García and Jiménez 2009), with summer rainfall being a key factor in the distribution of the species (Gavilán et al.
+2007, Río et al.
+2007).
+
+This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with altitudes of between 860 m and 3482 m a.s.l.
+The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
+There are eight oak patches (2400 Has) identified (:red_circle: FIGURE) in this mountain range, ranging between 1100 and 2000 *m a.s.l.* and generally associated to major river valleys.
+Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation (Brewer et al.
+2002, Olalde et al.
+2002, Rodríguez-Sánchez et al.
+2010) and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change (Hampe and Petit 2005).
+
+The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests (Melendo and Valle 2000, Vivero et al.
+2000), undergoing intensive anthropic use in the last few decades (Camacho-Olmedo et al.
+2002, Valbuena-Carabaña et al.
+2010).
+In fact, the status of conservation of this species for southern Spain is "Vulnerable" (Vivero et al.
+2000).
+The relict presence of this species in Sierra Nevada is related both to its genetic resilience as well as to its high intraspecific genetic diversity (Valbuena-Carabaña and Gil 2013).
+However, they are also expected to suffer the impact of climate change, due to their climate requirements (wet summers).
+Thus, simulations of the climate change effects on this habitat forecast a reduction in suitable habitats for Sierra Nevada (Benito et al.
+2011).
+
+:red_circle: La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
+Ver MIGRAME dataset
 
@@
 
+Resilience
+
 ## Statistical analysis

### 2017-08-18 man/ms_0_intro.Rmd
UPDATE

--- 2017-08-17 man/ms_0_intro.Rmd
@@
 
+* The vegetation response to water deficit was assessed using remote sensing information and dendrocronologial data.
+
+Ideas a expresar
+
+### Remote sensing
+
+* Uso de Remote sensing para estudiar la sequía --> Leer @Zhang2013 y escribir algo.
+Drought monitoring using remote-sensing approach was originally applied to agriculture.
+Several remote-sensing derived indices have been used to study the drought effects on vegetation ...
+* Leer también a @AghaKouchak2015
+
+### Sequías 
+* :red_circle: Algo de las sequías de 2005 y 2014
+
+* Extreme sequías y ecología
+
+### Ideas a incluir
+
+* Descripción de las sequía de 2005 y 2014
+
+La sequía es una característica del clima mediterráneo [@Lionello2012], pero en los últimos años se ha venido observando un aumento de la frecuencia y la severidad de los periodos de sequía (:red_circle: CITA; @VicenteSerrano2014 para la P.
+Iberica).
+
+The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southeren sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013].
+These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
+
+an increase in the drought severity in the Iberian Peninsula has been observed in the last decades [@VicenteSerrano2014].
+
+* Extreme sequias y ecologia 
+* Como la dendro ayuda a evaluar las sequías (ver Gazol 2017)
+
+### Aims
+`$IMPROVE$`
+
+In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both greenees and growth of *Q. pyrenaica* forests in Sierra Nevada. Specifically,
+
+The aims of this work were:
+
+* To quantify how two extreme drought events influenced the greennes and radial growth of *Q. pyreancia* forests in their rear edge,
+
+* to analyze the resilience of these forests to sucessive extrme drought events,
+
+* and to explore differences in the resilience metrics between populations located in contrasting slopes within the rear edge of the distribution of this species.
+
 ## References

### 2017-08-18 man/ms_1_material_methods.Rmd
remove blanks
UPDATE

@@
 
+## NDVI
+
 ## Species and study site
@@
 
-This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with altitudes of between 860 m and 3482 m a.s.l.
+This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m a.s.l.
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
@@
 2010) and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change (Hampe and Petit 2005).
+
+----- 
+:red_circle: `duda aqui` Varias referencias hablan de los años 2005 y 2012 como  extremadamente secos.
+Pero habría que hacer alguna referencia y/o análisis.
+Tengo dudas de si hemos de analizar (e incluir) que efectivamente los años 2005 y 2012 fueron caracterizados por un extrema sequía, por lo que habría que  incluyendo referencia a apéndice
+-----
 
@@
 Ver MIGRAME dataset
+
+## Datos de sequía.
+
+* Meter aquí algunos datos de sequñia, similar a lo planteado por Gazol
+
+### Greenness data
 

### 2017-08-18 man/ms_0_intro.Rmd
update

@@
 
+Althought several works have reported these two years as some of the worst drought events, we characterised the drought at several spatio-temporal scales in the study area.
+From a long-term perspective, we compare the accumulated monthly precipitation at a meteorological station (Granada, Base Aérea) during the hydrological years 2004-2005 and 2011-2012 with the average of accumulated monthly precipitation for the period 1950-2015.
+
 * Extreme sequias y ecologia 

### 2017-08-21 man/ms_0_intro.Rmd
update mat and methods

@@
 
+### 
+Uso de NDVI como estimador de la NPP:
+
+* The NDVI properties have allowed the use of this information for estimating the Net Primary Production (NPP) (Goward and Dye, 1987; Running et al., 2004; Hasenauer et al., 2012).
+Different studies have already found a strong relationship between NPP and radial growth (e.g., Granier et al., 2008; Babst et al., 2013, 2014a, 2014b; Vicente-Serrano et al., 2015), albeit with significant differences, particularly those related to species, sites and environmental conditions.
+
+* Ver Gilaber et al. 2017 http://www.mdpi.com/2072-4292/9/3/193
+
 ### Aims

### 2017-08-21 man/ms_1_material_methods.Rmd
update mat and methods

@@
 # Materials and methods
-
-## NDVI
 
@@
 Tengo dudas de si hemos de analizar (e incluir) que efectivamente los años 2005 y 2012 fueron caracterizados por un extrema sequía, por lo que habría que  incluyendo referencia a apéndice
+
+* O quizá un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
 -----
@@
 
-* Meter aquí algunos datos de sequñia, similar a lo planteado por Gazol
+* Meter aquí algunos datos de sequia, similar a lo planteado por Gazol
 
 ### Greenness data
+
+To characterize the vegetation greeness we used the Enhanced Vegetation Index (EVI) derived from MOD13Q1 product obtained by the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor [@Didan2015].
+EVI product consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 231 m x 231 m.
+
+EVI junto con NDVI ha sido utilizado para cuantificar la productividad primaria ...
+No obstante we used EVI instead of NDVI because it is more sensitive to change in high-biomass areas; EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002, @Krapivin2015].
+
+NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI.
+
+Data were obtained using a Google Earth Engine script (:red_circle: cite gists) for the 2000 - 2015 period.
+The EVI data are geometrically and atmospherically corrected and include information about the quality ass....
+:red_circle:
+
+These data are geometrically and atmospherically corrected, and include an index of data quality (reliability, which range from 0 – good quality data – to 4 – raw data or absent for different reasons) based on the environmental conditions in which the data was recorded
+
+We first used the Quality Assesment (QA band) information of this product to filter out those values affected by high content of aerosols, clouds, shadows, snow or water; and then a quality assessment was carried out to filter the ...
+[@Reyes2015]
+
+:red_circle: reescribir esto de la calidad.
+
+After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2015.
 
@@
 
-Resilience
+## Resilience
+
+To evaluate the effects of the drought events on greeennes and tree growth we used the four different resilience metrics [@Lloret2011]
 
@@
 
+* Evaluación de la resiliencia en cada variable 
+* Long-term trends in RW
+* Anomalías en EVI??
+
 ## References

### 2017-08-22 man/ms_1_material_methods.Rmd
update resilience notes

@@
 
-To evaluate the effects of the drought events on greeennes and tree growth we used the four different resilience metrics [@Lloret2011]
+To evaluate the effects of the disturbance events on greeennes and tree growth we used four resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance, recovery (*Rc*) and relative resilience (*RRs*).
+We computed the values of these indices for tree growth and greenness during each drought event.
+We considered 2005 and 2012 as a single drought event.
+
+The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and befor the disturbance:
+
+Resistance (*Rt*) = Drought / Predrought
+
+The Recovery index (*Rc*) is the ability to recover relative to the damage experienced during disturbance, and it is estimated as the ratio between performance after and during disturbance:
+
+Recovery (*Rc*) = Postdrought / Drought
+
+The Resilience index (*Rs*) is the capacity to reach pre-disturbance performance levels, and it is estimated as the ratio between the performance after and before disturbance: Resilience (*Rs*)
+
+Re
+
+The indices For each target variable (tree growth and greenness) the indices evaluate
 

### 2017-08-22 man/ms_1_material_methods.Rmd
update mat and methods: resilience and filter MODIS

--- 2017-08-21 man/ms_1_material_methods.Rmd
@@
 
-* Meter aquí algunos datos de sequia, similar a lo planteado por Gazol
+* :red_circle: Meter aquí algunos datos de sequia, similar a lo planteado por Gazol
 
@@
 
-To characterize the vegetation greeness we used the Enhanced Vegetation Index (EVI) derived from MOD13Q1 product obtained by the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor [@Didan2015].
+To characterize the vegetation greeness of *Quercus pyrenaica* we used the Enhanced Vegetation Index (EVI) derived from MOD13Q1 product obtained by the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor [@Didan2015].
 EVI product consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 231 m x 231 m.
@@
 Data were obtained using a Google Earth Engine script (:red_circle: cite gists) for the 2000 - 2015 period.
+We selected the pixels covering the distribution of Quercus pyrenaica forests in Sierra Nevada (*n* = 928 pixels).
 The EVI data are geometrically and atmospherically corrected and include information about the quality ass....
@@
 After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2015.
+(:red_circle: Hemos seleccionado EVI medio, además de por los consejos que me ha dicho Domingo, porque he comprobado que existe una correlación entre el evi medio y el evi estacional, sobre todo el de verano.
+Ver esto: [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md).
+Además presenta alta correlaciones significativas con el EVI de verano: 0.88; de primavera: 0.76 y anual: 0.81)
+
+> Procedimiento de Filtrado de datos (ver [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md))
+
+* Información contenida en banda QA.
+
+    * Nos quedamos con pixeles marcados como Good Data (57.89 %)
+    * Filter out los marcados como Snow/Ice y/o Cloudy (2.57 + 7.08 = 9.65 %)
+    * Pixeles marcados como Marginal Data (32.33 %) (ver siguiente paso)
+
+* Explorar distribución temporal y analizar banda QA Detailed y llevar a cabo un filtrado siguiendo las especificaciones de @Reyes2015.
+
+    * Vemos los composites marcados con Aerosoles, Adjacent cluods, y Shadow.  
+    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds.
+On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]
+
+* Finalmente nos hemos quedado con las siguientes cifras.
+De un total de 360064 images composites for the study zone were downloaded (928 x 20 x 1 + 928 x 23 x 16 = 360064), tras el filtrado, nos quedamos con 286825 (79.65 %)
 
@@
 
-To evaluate the effects of the drought events on greeennes and tree growth we used the four different resilience metrics [@Lloret2011]
+To evaluate the effects of the disturbance events on greeennes and tree growth we used four resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), recovery (*Rc*) and relative resilience (*RRs*).
+
+The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and befor the disturbance:
+
+Resistance (*Rt*) = Drought / Predrought
+
+The Recovery index (*Rc*) is the ability to recover from disturbance relative to its severity, and it is estimated as the ratio between performance after and during disturbance:
+
+Recovery (*Rc*) = Postdrought / Drought
+
+The Resilience index (*Rs*) is the capacity to reach pre-disturbance performance levels, and it is estimated as the ratio between the performance after and before disturbance:
+
+Resilience (*Rs*) = Postdrought / Predrought
+
+The Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance, and it is estimated as:
+
+Relative Resilience (*RRs*) = (Postdrought - Drought) / Predrought
+
+We computed the values of these indices for tree growth and greenness during each drought event.
+We considered 2005 and 2012 as singles drought events.
+The Predrought and Postdrought values of each target variable (i.e.: tree growth or EVI) we computed as the mean value during a period of three years before and after the disturbance events respectively.
+(:red_circle: Incluir algo de computación de ventana temporal 2,3,4 años: hacerlo tanto en dendro como en EVI)
 
@@
 
-* Evaluación de la resiliencia en cada variable 
-* Long-term trends in RW
-* Anomalías en EVI??
+* Explore anomalies EVI 
+* Explore long and short term trends in RW :red_circle: ver correo Guillermo
+* ANOVA analysis EVI events and populations
 

### 2017-08-22 man/ms_1_material_methods.Rmd
update mat and methods

@@
 To characterize the vegetation greeness of *Quercus pyrenaica* we used the Enhanced Vegetation Index (EVI) derived from MOD13Q1 product obtained by the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor [@Didan2015].
+EVI and NDVI (Normalized Difference Vegetation Index) are the most common greenness vegetation indices.
+We used EVI instead of NDVI (Normalized Difference Vegetation Index) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002, @Krapivin2015, @Cabello2012].
+
 EVI product consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 231 m x 231 m.
-
-EVI junto con NDVI ha sido utilizado para cuantificar la productividad primaria ...
-No obstante we used EVI instead of NDVI because it is more sensitive to change in high-biomass areas; EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002, @Krapivin2015].
-
-NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI.
-
-Data were obtained using a Google Earth Engine script (:red_circle: cite gists) for the 2000 - 2015 period.
+Data were obtained using a Google Earth Engine script (:red_circle: cite gists) for the 2000 - 2016 period.
 We selected the pixels covering the distribution of Quercus pyrenaica forests in Sierra Nevada (*n* = 928 pixels).
@@
 :red_circle:
+
+`$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI.
 

### 2017-08-24 man/ms_2_results.Rmd
update

@@
 
+Del análisis de las anomalías observamos:
+
+* En 2005 y en 2012 las anomalías fueron negativas. 
+* 2005 las anomalías fueron mucho mas negativas que para 2012. 
+*
+
+* Analizar magnitud de las anomalías para 2005 y 2012 (comparación entre ellas y entre NyS):
+
+Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.
+
+* 2005 fue el año en el que las anomalías fueron mas negativas, siendo de las anomalías de mas magnitud negativa en las poblaciones del norte.
+
+* Si atendemos a las sa (standardized anomalíes) y aplicamos el criterio de Gao, podemos decir que en 2005 se observó un bronwing en los bosques de Q.
+pyrenaica, sobre todo en las situadas en el northern slopes.
+
+* Asimismo, las anomalías 2012 fueron negativas pero
+
+tanto el las poblaciones del N como en las del S.
+
+*
+
 ## References

### 2017-08-25 man/ms_0_intro.Rmd
update intro and references

@@
 
+Major drought episodes in the IP were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014]
+
 La sequía es una característica del clima mediterráneo [@Lionello2012], pero en los últimos años se ha venido observando un aumento de la frecuencia y la severidad de los periodos de sequía (:red_circle: CITA;  @VicenteSerrano2014 para la P.
@@
 
+frecuencia de las sequías severas (ver Dai 2011)
+
+Algunos eventos extremos han aumentado su frecuencia en los últimos años
+
 Althought several works have reported these two years as some of the worst drought events, we characterised the drought at several spatio-temporal scales in the study area.
@@
 * Como la dendro ayuda a evaluar las sequías (ver Gazol 2017)
+
+Extreme climate events (as droughts, heatwaves) severely affect forests and grasslands throuhg changes in plant physiology, phenology and carbon allocation [@Ummenhofer2017].
+
+Droughts are most likely to have the largest and most long-lasting impacts globally due to large indirect and lagged impacts and long recovery especially for forest ecosystems (ver  18 en @Ummenhofer2017)
 

### 2017-08-25 man/ms_1_material_methods.Rmd
update mat and methods. Standardized anomalies

@@
 
+To explore the effect of drought events on greenness we used the EVI standardized anomaly (EVI~sa)
+
+calculated pixel-by-pixel. For each pixel we averaged all the EVI valid values within a year, and then the standardized anomaly was computed as:
+
+$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
+
+(anomaly divided by the standard deviation) are calculated pixel-by-pixel
+
+The SAs, which are also referred to as normalized anomalies, are calculated by dividing the anomalies by the standard deviation.
+The SAs generally provide more information about the magnitude of the anomalies, because the influences of dispersion have been removed.
+
+En el caso de las anomalías estandarizadas, utilizamos la aproximación de Gao et al.
+(2016) [doi:10.1038/srep26958](https://dx.doi.org/10.1038/srep26958), donde las anomalias las dividimos por la desviación estandar de los valores de EVI medio para el periodo de referencia.
+Las anomalías estandarizadas generalmente proporcionan mas información referente a la magnitud de la anomalía ya que las potenciales influencias de la dispersión de los datos han sido eliminadas.
+Por tanto las anomalías estandarizadas se calculan como: sa = (evi_year - evi_ref) / sd_ref.
+
 ## Field sampling and dendrochronological methods

### 2017-08-25 man/ms_1_material_methods.Rmd
update methods: standardized anomalies

--- 2017-08-22 man/ms_1_material_methods.Rmd
@@
 
-:red_circle: reescribir esto de la calidad.
+:red_circle: reescribir esto de la calidad. Ver Samanta et al 2012 y como describe el proceso de calidad
 
@@
 De un total de 360064 images composites for the study zone were downloaded (928 x 20 x 1 + 928 x 23 x 16 = 360064), tras el filtrado, nos quedamos con 286825 (79.65 %)
+
+To explore the effect of drought events on greenness we calculated the EVI standardized anomaly (EVI~sa) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012, @Gao2016].
+For each pixel we averaged all the EVI valid values within a year (:red_filter: see quality filter), and then the standardized anomaly was computed as:
+
+$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
+where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for the year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for the year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
 

### 2017-08-28 man/ms_2_results.Rmd
update writing

@@
 # Results
+
+## Vegetation Greenness
+Standardized a
+
+Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005.
+Sin embargo vemos valores menos negativos para 2005.
+🔴 Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years 🔴
+
+Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 ± 0.029)que para 2012 (-0.418 ± 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
+
+Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (Fig 1a :red_circle: EVI profile comparison), particu
 

### 2017-08-29 man/ms_2_results.Rmd
UPDATE
update results resilience

@@
 ## Vegetation Greenness
+
+:red_circle: referencia a las tendencias en EVI (lo hemos vuelto a calcular y además en el trabajo de ontologías también nos sale)
+
 Standardized a
@@
 * 2005 las anomalías fueron mucho mas negativas que para 2012.
-*
 
@@
 
-tanto el las poblaciones del N como en las del S.
+## Resiliencia EVI
 
-*
+* Resistencia
+
+    * Menor resistencia al evento de 2005 que al de 2012 (2005=0.86, 2012=0.94, p <0.0001)
+    * Poblaciones del norte menos resistentes que las del sur (N=0.883; S=0.921; p < 0.0001)
+    * La resistencia de las poblaciones varia segun el evento de sequía y la población, así observamos que para la sequía de 2005 la resistencia fue diferente entre el N y el S.
+Sin embargo las poblaciones resistieron igual (no diferencias) para el evento de 2012 (p = 0.0921).
+La resistencia fue mucho menor para las poblaciones del N, en el evento de 2005 (p < 0.0001).
+Las poblaciones del sur fueron mas resistentes que las del N en el pero de los eventos de sequía.
+
+* Recovery
+
+    * El recovery fue mayor tras el evento de 2005 que tras el evento de 2012 (2005 = 1.12; 2012 = 1.05; p<0.0001)
+    * Mayor recovery de las poblaciones del N que las del S (N=1.107; S=1.069; p<0.001)
+    * El recovery en la poblaciones del S no fue diferente entre 2005 y 2012
 

### 2017-08-29 man/ms_2_results.Rmd
update results

@@
 
+* Resilience
+
+    * Mayores valores de resiliencia para 2005 que para 2012 (2005 = 0.996; 2012= 0.959; p < 0.0001)
+    * Resilienca fue mayor para las poblaciones del sur (S= 0.983; N=0.973; p < 0.0001)
+    * Interacción: Para 2005, no existieron diferencias en la resiliencia de las poblaciones N y S (Las poblaciones del N y S fueron resilientes de una forma similar al evento de 2005).
+Sin embargo, para 2012 si encontramos valores de resiliencia mas grandes para las poblaciones del sur que para las del norte (p<0.0001)
+
 ## References

### 2017-08-31 man/ms_1_material_methods.Rmd
update writing
update anovas methods

@@
 
+We tested for significant differences between drought events (2005 and 2012) and oak population (northern and southern slopes) for each of the resilience indices.
+Robust two-way ANOVAs were used beacuse original and log-transformed data both did not match the assumptions of normality and homogeneity of variance [@Wilcox2012].
+Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to mean value in all cases [@Wilcox2012].
+When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field, @Wilcox2012].
+Post-hoc differences were assessed pairwise using a similar boostrap test.
+All the robust ANOVA and post-hoc tests were carried out using the WRS2 [@Mair2017] and rcompanion [@Mangiafico2017] R packages.
+The level of significance was set at 0.05 and adjusted for multiple comparisons.
+
 ## References

### 2017-08-31 man/ms_1_material_methods.Rmd
fix typo biblio

@@
 Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to mean value in all cases [@Wilcox2012].
-When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field, @Wilcox2012].
+When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012, @Wilcox2012].
 Post-hoc differences were assessed pairwise using a similar boostrap test.

### 2017-09-01 man/ms_2_results.Rmd
update results

--- 2017-08-28 man/ms_2_results.Rmd
@@
 
-tanto el las poblaciones del N como en las del S.
+## Resiliencia EVI
 
-*
+* Resistencia
+
+    * Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001]
+    * Menor resistencia de las poblaciones del Norte a los eventos de sequía que las del Sur [N: 0.883 (0.877-0.889); S: 0.921 (0.918 - 0.925); p <0.0001]
+    * La resistencia varió en función de la sequía y de la población.
+Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
+
+* Recovery
+
+    * La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
+    * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
+    * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001].
+En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
+
+* Resilience
+
+    * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
+    * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152).
+En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
+
+* Relative Resilience
+
+  * Los robledales mostraron mayor resiliencia relativa a la sequía de 2005 que a la de 2012 [2005: 0.099 (0.095-0.105); 2012: 0.053 (0.050 - 0.056); p <0.0001]  
+  * Las poblaciones del sur mostraron menor resiliencia relativa que las del norte [N: 0.086 (0.082-0.092); S: 0.063 (0.060 - 0.066); p <0.0001], debido sobre todo a la diferencia en la resiliencia relativa para las poblaciones del norte entre los dos eventos de sequía (mucha mayor resiliencia relativa en 2012 que en 2005 para las poblaciones del N)
+  * Las poblaciones del sur no mostraron diferencias en cuanto a la resiliencia relativa entre los dos eventos de sequía (padj = 0.152)
 

### 2017-09-05 analysis/analysis_resilience_bai.Rmd
upload anovas

@@
 library('effects')
+library('devtools')
+# devtools::install_github("ajpelu/auxiliar")
+library('car')
+library('auxiliar')
+library('WRS2')
+library('MASS')
+library('rcompanion')
 ```
@@
   mutate(disturb = recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  group_by(site, disturb_year, disturb) %>% 
-  summarise(mean = mean(mean_period/100),
+  dplyr::group_by(site, disturb_year, disturb) %>% 
+  dplyr::summarise(mean = mean(mean_period/100), 
                    sd = sd(mean_period/100),
-            se = sd/sqrt(length(mean_period/100))) %>% pander() 
+                   se = sd/sqrt(length(mean_period/100))) %>% as.data.frame() %>% pander() 
 ```
@@
 ```
+
+## Asumptions
+
+* Explorar si se cumplen los supuestos de normalidad y homocedasticidad. Tenemos que comprobar que cada uno de los grupos son normales (1995,2005,2012; site: SJ, CaH, CaL; e interactions)
+
+```{r}
+shapirosNormal <- function(df, resp_var, factor_vars) {
+  rv <- enquo(resp_var)
+  out <- df %>%
+    group_by_(.dots=factor_vars) %>%
+    summarise(statistic = round(shapiro.test(!!rv)$statistic,5),
+              p_value = round(shapiro.test(!!rv)$p.value,5)) %>%
+    data.frame()
+
+  return(out)
+}
+
+homogetest <- function(resp_var, factores, df){
+
+  out_factores <- c()
+
+  for (f in factores){
+    hv <- c()
+    myformula <- as.formula(paste0(resp_var, "~", f))
+    #tests
+    fk <- fligner.test(myformula, data = df)
+    lv <- leveneTest(myformula, data = df)
+    # out
+    hv$fk_stat <- fk$statistic
+    hv$fk_pvalue <- fk$p.value
+    hv$lev_stat <- lv$`F value`[1]
+    hv$lev_pvalue <- lv$`Pr(>F)`[1]
+    hv$factor <- f
+    hv <- as.data.frame(hv)
+    row.names(hv) <- NULL
+
+    out_factores <- rbind(out_factores, hv)}
+  return(out_factores)
+
+}
+```
+## Normalidad
+
+```{r}
+# See auxiliar::shapirosNormal
+### Resilience
+
+nrsA <- shapirosNormal(re, resp_var = rs, 'disturb_year')
+nrsA$var <- 'rs'
+nrsB <- shapirosNormal(re, resp_var = rs, 'site')
+nrsB$var <- 'rs'
+nrsAB <- shapirosNormal(re, resp_var = rs, c('disturb_year','site'))
+nrsAB$var <- 'rs'
+
+### Recovery
+nrcA <- shapirosNormal(re, resp_var = rc, 'disturb_year')
+nrcA$var <- 'rc'
+nrcB <- shapirosNormal(re, resp_var = rc, 'site') 
+nrcB$var <- 'rc'
+nrcAB <- shapirosNormal(re, resp_var = rc, c('disturb_year','site'))
+nrcAB$var <- 'rc'
+
+### Resistance
+nrtA <- shapirosNormal(re, resp_var = rt, 'disturb_year')
+nrtA$var <- 'rt'
+nrtB <- shapirosNormal(re, resp_var = rt, 'site') 
+nrtB$var <- 'rt'
+nrtAB <- shapirosNormal(re, resp_var = rt, c('disturb_year','site'))
+nrtAB$var <- 'rt'
+
+### Relative Resilience 
+nrrsA <- shapirosNormal(re, resp_var = rrs, 'disturb_year')
+nrrsA$var <- 'rrs'
+nrrsB <- shapirosNormal(re, resp_var = rrs, 'site') 
+nrrsB$var <- 'rrs'
+nrrsAB <- shapirosNormal(re, resp_var = rrs, c('disturb_year','site'))
+nrrsAB$var <- 'rrs'
+```
+
+```{r}
+normtestAB <- rbind(nrcAB, nrtAB, nrsAB, nrrsAB) 
+normtestAB %>% pander()
+write.csv(normtestAB, 
+          file=paste0(di, '/out/anovas_resilience/normo_disturb_year_site.csv'), row.names = F)
+```
+
+```{r}
+rm(nrcA, nrcB, nrcAB,
+   nrsA, nrsB, nrsAB,
+   nrrsA, nrrsB, nrrsAB,
+   nrtA, nrtB, nrtAB)
+```
+
+* No se cumplen los requisitos de normalidad
+
+## Heterocedasticidad
+
+```{r}
+## See auxiliar::homogetest
+factores <- c('disturb_year', 'site', 'interaction(disturb_year, site)') 
+responses <- c('rs', 'rc', 'rt', 'rrs')
+homo <- c()
+
+for (i in responses){ 
+  ht <- homogetest(resp_var = i, factores = factores, df = re)
+  ht <- ht %>% mutate(response = i)
+  homo <- rbind(homo, ht)
+}
+
+homo %>% pander()
+write.csv(homo, 
+          file=paste0(di, '/out/anovas_resilience/homocedasticidad.csv'), row.names = F)
+```
+
+* Tampoco se cumplen los requisitos de homogeneidad de varianzas entre grupos
+
+* Probamos a transformar los datos con log y reanalizar los supuestos de homocedasticidad
+
+```{r}
+factores <- c('disturb_year', 'site', 'interaction(disturb_year, site)') 
+responses <- c('logrs', 'logrc', 'logrt', 'logrrs')
+homo_log <- c()
+
+re <- re %>% 
+  mutate(
+    logrs = log(rs),
+    logrc = log(rc),
+    logrt = log(rc),
+    logrrs = log(rrs)
+)
+
+for (i in responses){ 
+  ht <- homogetest(resp_var = i, factores = factores, df = re)
+  ht <- ht %>% mutate(response = i)
+  homo_log <- rbind(homo_log, ht)
+}
+
+homo_log %>% pander()
+write.csv(homo_log, 
+          file=paste0(di, '/out/anovas_resilience/homocedasticidad_log.csv'), row.names = F)
+```
+
+* Tampoco se cumplen
 
@@
   # Disturb Event 
-  ph_event <- lsmeans(mymodel, pairwise ~ disturb_year, adjust = "tukey")
+  ph_event <- lsmeans(mymodel, pairwise ~ disturb_year, adjust = "bon")
 
@@
                    Letters = letters, 
-                   adjust  = "tukey")
+                   adjust  = "bon")
 
   # Site  
-  ph_site <- lsmeans(mymodel, pairwise ~ site, adjust = "tukey")
+  ph_site <- lsmeans(mymodel, pairwise ~ site, adjust = "bon")
   cld_site <- cld(ph_site, alpha   = 0.01, 
                  Letters = letters, 
-                 adjust  = "tukey")
+                 adjust  = "bon")
 
   # interaction 
-  ph_i <- lsmeans(mymodel, pairwise ~ disturb_year:site, adjust = "tukey")
+  ph_i <- lsmeans(mymodel, pairwise ~ disturb_year:site, adjust = "bon")
   cld_i <- cld(ph_i, alpha   = 0.01, 
                  Letters = letters, 
-                 adjust  = "tukey")
+                 adjust  = "bon")
 
@@
 
+**OJO SOLO 2005 y 2012**
+
 ```{r}
@@
 aov_rc <- aovas(re, vars=vars, resp_var = resp_var)
-```
 
-```{r}
 mc <- aov_rc$model_coeff
@@
          which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-```
 
-```{r}
 gm <- aov_rc$model_summary
@@
 
-### Post hoc comparison
-
 ```{r}
@@
 
-### Plots
-
-```{r}
+```{r, echo=FALSE}
 #### ~ Site
@@
 picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(3,1))
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
+
 ```
 
-```{r}
+```{r, fig.width = 4, fig.height=4}
 ps
@@
 
-```{r}
+```{r, fig.width = 4, fig.height=4}
 pd
@@
 
-```{r}
+```{r, fig.width = 4, fig.height=4}
 picollapse
@@
 
-```{r}
+```{r, fig.width = 6}
 pi
@@
 ## Resistance 
-```{r}
+```{r, echo =FALSE}
 # Variable
 resp_var <- 'rt'
+vars <- c('disturb_year','site')
 
@@
 aov_rt <- aovas(re, vars=vars, resp_var = resp_var)
-```
 
-```{r}
 mc <- aov_rt$model_coeff
@@
          which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-```
 
-```{r}
 gm <- aov_rt$model_summary
@@
 ## Relative Resilience 
-```{r}
+```{r, echo=FALSE}
 # Variable
@@
 aov_rrs <- aovas(re, vars=vars, resp_var = resp_var)
-```
 
-```{r}
 mc <- aov_rrs$model_coeff
@@
          which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-```
 
-```{r}
 gm <- aov_rrs$model_summary
@@
 ## Resilience 
-```{r}
+```{r, echo=FALSE}
 # Variable
@@
 aov_rs <- aovas(re, vars=vars, resp_var = resp_var)
-```
 
-```{r}
 mc <- aov_rs$model_coeff
@@
          which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-```
 
-```{r}
 gm <- aov_rs$model_summary
@@
 
-```{r}
+```{r, echo=FALSE}
 # Interactions plot all  
@@
 
-```{r}
+```{r, echo=FALSE}
 dodge <- position_dodge(width = 0.3)
@@
 
-```{r}
+```{r, echo=FALSE}
 plot_ms <- means_site %>%  
@@
 
-```{r}
+```{r, echo=FALSE}
 plot_md <- means_disturb %>%  
@@
 
+# ROBUST ANOVA
+
+* Ver Wilcox (2005, 2012)
+* Vamos a realizar un Robust factorial ANOVA. En concreto:
+
+  * Two-way robust factorial ANOVA on M-estimator
+  * pkg WRS2
+
+```{r}
+robustANOVA <- function(df, resp_var, factores, 
+         alpha, nboot, treshold) {
+  # alpha: alpha ci for huber m-estimation 
+  # nboot: numer of iterations 
+  # treshoold for letter (posthoc)
+  # See http://rcompanion.org/rcompanion/d_08a.html
+
+  df <- re
+  alpha = 0.95
+  nboot = 3000
+  treshold = 0.01
+
+  # Create interaction 
+  df$interaction <- interaction(df$disturb_year, df$site)
+
+  # Formulas 
+  formulaFull <- as.formula(paste0(resp_var,  " ~ ",
+                                 paste(factores, collapse = '+')))
+
+  formula_A <- as.formula(paste0(resp_var,  " ~ ", factores[1]))
+  formula_B <- as.formula(paste0(resp_var,  " ~ ", factores[2]))
+  formula_AB <- as.formula(paste0(resp_var,  " ~ interaction"))
+
+  # Produce Huber M-estimators and confidence intervals by group
+  mest <- groupwiseHuber(formulaFull, data = df, ci.type = 'wald', conf.level = alpha)
+  mest_a <- groupwiseHuber(formula_A, data = df, ci.type = 'wald', conf.level = alpha)
+  mest_b <- groupwiseHuber(formula_B, data = df, ci.type = 'wald', conf.level = alpha)
+
+  # Two-way robust analysis 
+  x <- pbad2way(formulaFull, data = df, est = "mom", nboot = nboot)
+
+  out_ra <- data.frame(
+    term = c(x$varnames[2], 
+             x$varnames[3], 
+             paste0(x$varnames[2], ':', x$varnames[3])),
+    p_value = c(x$A.p.value, x$B.p.value, x$AB.p.value))
+
+  # post-hoc 
+  ## factor A
+  pha <- pairwiseRobustTest(formula_A, data = df, est = "mom", 
+                            nboot = nboot, method="bonferroni")
+
+  ## factor B
+  phb <- pairwiseRobustTest(formula_B, data = df, est = "mom", 
+                            nboot = nboot, method="bonferroni")
+  ## interaction effect (AB)
+  phab <- pairwiseRobustTest(formula_AB, data = df, est = "mom", 
+                            nboot = nboot, method="bonferroni")
+  ## letters
+  letters_pha <- cldList(comparison = pha$Comparison,
+        p.value    = pha$p.adjust,
+        threshold  = treshold)
+  letters_phb <- cldList(comparison = phb$Comparison,
+        p.value    = phb$p.adjust,
+        threshold  = treshold)
+
+  letters_phab <- cldList(comparison = phab$Comparison,
+        p.value    = phab$p.adjust,
+        threshold  = treshold)
+
+  ph <- rbind(pha, phb, phab)
+
+  phRWS2 <- mcp2a(formulaFull, data=df, est = "mom", nboot = nboot)
+
+  out <- list()  
+  out$mest <- mest # Huber M-estimators and Confidence Intervals
+  out$mest_a <- mest_a
+  out$mest_b <- mest_b
+  out$ra <- out_ra # Output for Two-way robust analysis (M-estimators)
+  out$letters_pha <- if(exists(letters_pha), letters_pha,  # Letters comparison posthoc
+  out$letters_phb <- letters_pha # Letters comparison posthoc
+  out$letters_phab <- letters_pha # Letters comparison posthoc
+  out$ph <- ph # posthoc comparison usinng pairwiseRobustTest
+
+  print(out_ra)
+  print(phRWS2)
+
+  return(out)
+}
+```
+
+```{r}
+factores = c('disturb_year', 'site', 'disturb_year:site')
+
+```
+
+```{r}
+rars <- robustANOVA(df=re, resp_var='rs', factores=factores,
+              alpha = 0.95, nboot = 3000, treshold = 0.01)
+```
+
+```{r}
+rarc <- robustANOVA(df=evires, resp_var='rc', factores=factores,
+              alpha = 0.95, nboot = 3000, treshold = 0.01)
+```
+
+```{r}
+rart <- robustANOVA(df=evires, resp_var='rt', factores=factores,
+              alpha = 0.95, nboot = 3000, treshold = 0.01)
+```
+
+```{r}
+rarrs <- robustANOVA(df=evires, resp_var='rrs', factores=factores,
+              alpha = 0.95, nboot = 3000, treshold = 0.01)
+```
+
 # References

### 2017-09-06 script/R/baiResilience.R
fix dplyr::select

@@ -46,14 +46,14 @@ baiResilience <- function(bai_object, event_years, window){
       df_tree <- growth_df %>% filter(tree == t) 
       
       # Compute resilience metrics by tree
-      aux_df <- df_tree %>% select(-c(tree, sd_period, se_period)) %>% 
+      aux_df <- df_tree %>% dplyr::select(-c(tree, sd_period, se_period)) %>% 
         spread(key=disturb, value=mean_period) %>% 
         mutate(rt = dr / pre,
                rc = post / dr,
                rs = post / pre,
                rrs = ((post - dr) / pre),
                tree = t) %>% 
-        select(rt, rc, rs, rrs, disturb_year, tree) %>% as.data.frame() 
+        dplyr::select(rt, rc, rs, rrs, disturb_year, tree) %>% as.data.frame() 
       
       aux_resilience <- rbind(aux_resilience, aux_df)
     }

### 2017-09-06 analysis/analysis_resilience_bai.Rmd
update resilience bai

@@
 ```{r global_options, include=FALSE}
-knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
+knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, error=TRUE, message=FALSE)
 ```
@@
 ```
-
 # Resilience
@@
            site = str_replace(i, "res_[0-9]_", '')) %>%
-    select(-disturb_year, -tree)
+    dplyr::select(-disturb_year, -tree)
 
@@
 
-cor2 <- correla[["ws_2"]] %>% select(-ws) %>% mutate(ind = row_number())
-cor3 <- correla[["ws_3"]] %>% select(-ws) %>% mutate(ind = row_number())
-cor4 <- correla[["ws_4"]] %>% select(-ws) %>% mutate(ind = row_number())
+cor2 <- correla[["ws_2"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
+cor3 <- correla[["ws_3"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
+cor4 <- correla[["ws_4"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
 
@@
 
-g %>% mutate(disturb = recode(disturb, 
+g %>% mutate(disturb = dplyr::recode(disturb, 
                               dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
@@
 g %>% 
-  mutate(disturb = recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
   group_by(disturb, disturb_year, site) %>% 
@@
 pgrowht <- g %>% 
-  mutate(disturb = recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
   group_by(disturb, disturb_year, site) %>% 
@@
 g %>% 
-  mutate(disturb = recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
   dplyr::group_by(site, disturb_year, disturb) %>% 
@@
 vars <- c('disturb_year','site')
+re$disturb_year <- factor(re$disturb_year)
 ```
@@
 
-  df <- re
-  alpha = 0.95
-  nboot = 3000
-  treshold = 0.01
-
   # Create interaction 
@@
   x <- pbad2way(formulaFull, data = df, est = "mom", nboot = nboot)
+
+  pbad2way(rs ~ disturb_year + site + disturb_year:site, data = df, 
+           est = "mom", nboot = nboot)
 
@@
                             nboot = nboot, method="bonferroni")
-  ## letters
-  letters_pha <- cldList(comparison = pha$Comparison,
-        p.value    = pha$p.adjust,
-        threshold  = treshold)
-  letters_phb <- cldList(comparison = phb$Comparison,
-        p.value    = phb$p.adjust,
-        threshold  = treshold)
-
-  letters_phab <- cldList(comparison = phab$Comparison,
-        p.value    = phab$p.adjust,
-        threshold  = treshold)
 
@@
   out$ra <- out_ra # Output for Two-way robust analysis (M-estimators)
-  out$letters_pha <- if(exists(letters_pha), letters_pha,  # Letters comparison posthoc
-  out$letters_phb <- letters_pha # Letters comparison posthoc
-  out$letters_phab <- letters_pha # Letters comparison posthoc
   out$ph <- ph # posthoc comparison usinng pairwiseRobustTest 
+  out$pha <- pha
+  out$phb <- phb
+  out$phab <- phab
 
@@
 }
+
+  # if (exists('letters_phb')) {
+  # letters_phb <- letters_phb} else {
+  # myerror <- evaluate('cldList(comparison = phb$Comparison, 
+  #                     p.value = phb$p.adjust, threshold  = treshold)')
+  # letters_phb <- as.character(myerror[[2]]$message)}
+  # 
+  # if (exists('letters_pha')) {
+  # letters_pha <- letters_pha} else {
+  # myerror <- evaluate('cldList(comparison = pha$Comparison, 
+  #                     p.value = pha$p.adjust, threshold  = treshold)')
+  # letters_pha <- as.character(myerror[[2]]$message)}
+  # 
+  # if (exists('letters_phab')) {
+  # letters_phab <- letters_phab} else {
+  # myerror <- evaluate('cldList(comparison = phab$Comparison, 
+  #                     p.value = phab$p.adjust, threshold  = treshold)')
+  # letters_phab <- as.character(myerror[[2]]$message)}
+
 ```
@@
 factores = c('disturb_year', 'site', 'disturb_year:site')
-
-```
-
+```
+
+## Resilience 
 ```{r}
@@
 
-```{r}
-rarc <- robustANOVA(df=evires, resp_var='rc', factores=factores,
+#### Rs Letters 
+```{r}
+x <-rars
+
+cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01)
+cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01)
+cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01)
+```
+
+## Recovery 
+```{r}
+rarc <- robustANOVA(df=re, resp_var='rc', factores=factores,
               alpha = 0.95, nboot = 3000, treshold = 0.01)
@@
 
-```{r}
-rart <- robustANOVA(df=evires, resp_var='rt', factores=factores,
+#### Rc Letters 
+```{r}
+x <-rarc
+
+cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01)
+cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01)
+cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01)
+```
+
+## Resistance 
+```{r}
+rart <- robustANOVA(df=re, resp_var='rt', factores=factores,
               alpha = 0.95, nboot = 3000, treshold = 0.01)
@@
 
-```{r}
-rarrs <- robustANOVA(df=evires, resp_var='rrs', factores=factores,
+#### Rt Letters 
+```{r}
+x <-rart
+
+cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01)
+cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01)
+cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01)
+```
+
+## Relative Resilience
+```{r}
+rarrs <- robustANOVA(df=re, resp_var='rrs', factores=factores,
               alpha = 0.95, nboot = 3000, treshold = 0.01)
@@
 
+#### Rt Letters 
+```{r}
+x <-rarrs
+
+cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01)
+cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01)
+cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01)
+```
+
+### Estimadores de huber 
+```{r} 
+rars$mest$var <- 'rs'
+rarc$mest$var <- 'rc'
+rart$mest$var <- 'rt'
+rarrs$mest$var <- 'rrs'
+
+mhuber <- rbind(rarc$mest, rart$mest, rars$mest, rarrs$mest)
+mhuber %>% pander()
+```
+
+```{r}
+rars$mest_a$var <- 'rs'
+rarc$mest_a$var <- 'rc'
+rart$mest_a$var <- 'rt'
+rarrs$mest_a$var <- 'rrs'
+
+mhuber_a <- rbind(rarc$mest_a, rart$mest_a, rars$mest_a, rarrs$mest_a)
+mhuber_a %>% pander()
+```
+
+```{r}
+rars$mest_b$var <- 'rs'
+rarc$mest_b$var <- 'rc'
+rart$mest_b$var <- 'rt'
+rarrs$mest_b$var <- 'rrs'
+
+mhuber_b <- rbind(rarc$mest_b, rart$mest_b, rars$mest_b, rarrs$mest_b)
+mhuber_b %>% pander()
+```
+
+### Pairwise comparison
+```{r}
+
+rars$ph$var <- 'rs'
+rarc$ph$var <- 'rc'
+rart$ph$var <- 'rt'
+rarrs$ph$var <- 'rrs'
+
+pairwise <- rbind(rarc$ph, rart$ph, rars$ph, rarrs$ph) 
+pairwise %>% pander()
+```
+
+## Interaction plot 
+### Response ~ (x=Drought)
+
+```{r}
+mhuber<- mhuber %>% 
+  mutate(var_sorted = case_when(var == "rc" ~ "1_rc",
+                                var == "rt" ~ "0_rt",
+                                var == "rs" ~ "2_rs",
+                                var == "rrs" ~ "3_rrs"))
+
+pd <- position_dodge(.2)
+
+robust_plot_evi_drought <- ggplot(mhuber, aes(x=disturb_year, y=M.Huber, color = site, group=site, fill=site)) + 
+  geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), 
+                width=.1, size=0.7, position=pd) + 
+  geom_line(aes(group=site,color=site, linetype=site), position=pd) + 
+  geom_point(shape=21, size=3.5, position=pd) +
+  facet_wrap(~var_sorted, nrow = 2, scales = 'free_y',
+             labeller=as_labeller(c('0_rt' = 'Resistance', 
+                                 '1_rc' = 'Recovery',
+                                 '2_rs' = 'Resilience',
+                                 '3_rrs' = 'Relative Resilience'))) +
+  scale_color_manual(values=c('black','blue','red')) + 
+  scale_fill_manual(values=c('black','blue','red')) + theme_bw() +
+  scale_linetype_manual(values=c("solid", "dashed", 'dashed')) +
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'),
+        legend.position="bottom") +
+  ylab('') + xlab('Drought event')
+
+robust_plot_evi_drought
+
+pdf(paste0(di, 'out/fig/resilience/robust_plot_evi_drought.pdf'), width=9, height = 9)
+robust_plot_evi_drought 
+dev.off()
+```
+
+### Response ~ (x=site)
+```{r}
+pd <- position_dodge(.2)
+
+robust_plot_evi_site <- ggplot(mhuber, aes(x=site, y=M.Huber, color = disturb_year, group=disturb_year, fill=disturb_year)) + 
+  geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), 
+                width=.1, size=0.7, position=pd) + 
+  geom_line(aes(group=disturb_year,color=disturb_year, linetype=disturb_year), position=pd) + 
+  geom_point(shape=21, size=3.5, position=pd) +
+  facet_wrap(~var_sorted, nrow = 2, scales = 'free_y',
+             labeller=as_labeller(c('0_rt' = 'Resistance', 
+                                 '1_rc' = 'Recovery',
+                                 '2_rs' = 'Resilience',
+                                 '3_rrs' = 'Relative Resilience'))) +
+  scale_color_manual(values=c('black','blue')) + 
+  scale_fill_manual(values=c('black','blue')) + theme_bw() +
+  scale_linetype_manual(values=c("solid", "dashed")) +
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white'),
+        legend.position="bottom") +
+  ylab('') + xlab('Drought event')
+
+robust_plot_evi_site
+
+pdf(paste0(di, 'out/fig/resilience/robust_plot_evi_site.pdf'), width=9, height = 9)
+robust_plot_evi_site 
+dev.off()
+```
+
+```{r}
+# Export data 
+write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/robust_mhuber.csv'), row.names = F)
+write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_a.csv'), row.names = F)
+write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_b.csv'), row.names = F)
+
+write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/robust_pairwise.csv'), row.names = F)
+```
+
 # References

### 2017-09-06 analysis/analysis_chronologies.Rmd
update anlaysis chrono and parser climate data
update and get crono by site

@@
 machine <- 'ajpelu'
-#machine <- 'ajpeluLap'
+# machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
   mutate(treatment = as.factor('biweight')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
+  dplyr::select(bai_mean = SJstd, samp.depth, year, treatment)
 
@@
   mutate(treatment = as.factor('No biweight')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
+  dplyr::select(bai_mean = SJstd, samp.depth, year, treatment)
 
@@
   mutate(treatment = as.factor('biweight')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
+  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
 
@@
   mutate(treatment = as.factor('No biweight')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
+  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
 
@@
 
+## Export data of bai (chro)
+write.csv(cro_sites, file=paste(di, "data/cronos_medias/cronos_sites.csv", sep=""), row.names = FALSE)
+
 plot_cro_sitesSJCALH <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 

### 2017-09-06 script/R/correlaBootIC.R

@@ -29,7 +29,7 @@ correlaBootIC <- function(df_suaviza, tipos_correla, years_suavizado,
     for (j in 1:ys){
       aux <- df_suaviza %>% 
         filter(size_wind == j) %>%
-        select(-size_wind) %>% 
+        dplyr::select(-size_wind) %>% 
         spread(crononame, bai_smooth) %>% 
         na.omit() %>% # to compare the series in the years they are overlapping 
         as.data.frame() 

### 2017-09-06 analysis/fig_ts_bai_evi.Rmd
update fig

--- 2017-09-06 analysis/analysis_resilience_bai.Rmd
@@
 ---
-title: "Growth Resilience"
+title: "Plot_BAI_EVI"
 author: "AJ Perez-Luque (@ajpelu)"
@@
 library("tidyverse")
-library("dplR")
 library("stringr")
 library("knitr")
-library('gtable')
 library('grid')
 library('gridExtra')
-library('pander')
-library('broom')
-library('effects')
-library('devtools')
-# devtools::install_github("ajpelu/auxiliar")
-library('car')
-library('auxiliar')
-library('WRS2')
-library('MASS')
-library('rcompanion')
-```
-# Resilience
-
-* Calcularemos las métricas resiliencia de [@Lloret2011] sobre el crecimiento. 
-* Vamos a calcularlas sobre el BAI de cada árbol. 
-* Utilizaremos tres sitios: SJ, CAH y CAL (ver [./analysis/analysis_chronologies.md]('./analysis/analysis_chronologies.md))
-
-## Prepare data
-
-* Leer datos `rwl` de SJ y CA
-* Leer datos de diametros de los focal tree
+```
 
@@
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
-
-# sj 
-sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
-
-# canar 
-ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
-
+```
+
+```{r}
 # Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+cro <- read.csv(file=paste0(di, '/data/cronos_medias/cronos_sites.csv'), header=TRUE, sep=',')
+
+# evi data from 
+# https://github.com/ajpelu/qpyr_modis_resilience/blob/master/data/evi_mean.csv
+
+# Read data
+evi <- read.csv(file=paste(di, "/data/evi/evi_mean.csv", sep=""), header = TRUE, sep = ',')
+
+evi <- evi %>% 
+  mutate(
+    clu_pop = as.factor(case_when(
+      pop == 1 ~ "Camarate",
+      pop %in% c(2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
+      pop == 9 ~ 'out')),
+    clu_pop2 = as.factor(case_when(
+      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
+      pop == 9 ~ 'out'))) %>% 
+  filter(clu_pop != 'out')
+
+evi_year <- evi %>% group_by(year, clu_pop2) %>% 
+  summarise(mean = mean(evi, na.rm=TRUE),
+            sd = sd(evi),
+            se = sd /sqrt(length(sd))) %>% as.data.frame()
+```
+
+```{r}
+p <- cro %>% 
+  filter(year > 1949) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
+
+str(cro)
+
+cro <- cro %>% 
+  dplyr::rename(mean = bai_mean, sd = bai_sd, se = bai_se) %>%
+  dplyr::select(-n_trees) %>% 
+  mutate(clu_pop2 = as.factor(case_when(
+      site =='SJ' ~ 'Northern slope',
+      site =='CA_Low' ~ 'Southern slope',
+      site =='CA_High' ~ 'Southern slope')),
+      variable = 'bai')
+
+cro %>% 
+  filter(year > 1949) %>% 
+  ggplot(aes(x=year, y=mean)) + 
+  geom_line() + 
+  facet_wrap(~clu_pop2) + 
+  theme_bw() + ylab('BAI') + 
+  geom_point(size=.8)
+
+psj <- cro %>% 
+  filter(year > 1949, site=='SJ')
+
+  ggplot(aes(x=year, y=bai_mean/100)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) +
+    ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
+
+psj + geom_line(evi_year, aes(x=year, y=mean*100))
+
++ 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
+
+p + geom_line(evi_year, aes(x=year, y=mean*100))
+
+head(evi)
+min(evi_year$mean)
+max(evi_year$mean)
+
+```
+
+```{r}
+#fileplot <- paste0(di, '/out/fig/chronos/crono_compara_sitesSJCALH50.pdf') 
+#ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH50, width=12, height = 5)
+#plot_cro_sitesSJCALH50
 ```

### 2017-09-06 man/ms_1_material_methods.Rmd
update

@@
 We considered 2005 and 2012 as singles drought events.
-The Predrought and Postdrought values of each target variable (i.e.: tree growth or EVI) we computed as the mean value during a period of three years before and after the disturbance events respectively.
-(:red_circle: Incluir algo de computación de ventana temporal 2,3,4 años: hacerlo tanto en dendro como en EVI)
+The predrought and postdrought values of each target variable (i.e.: tree growth or EVI) we computed as the mean value during a period of three years before and after the disturbance events respectively.
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (:red_circle: incluir tabla de coeficientes y/o gráfica??
+como suplement, see Gazol 2017)
 

### 2017-09-14 man/ms_2_results.Rmd
update results

@@
 
+## Resiliencia BAI
+
+* Resistance: 
+  * No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012.
+Valores de resistencia menor al evento de 2005.
+[2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.03]
+  * Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001).
+Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].
+  * La interacción también fue significativa.
+De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ???
+
+* Recovery 
+    * En general mayor recuperación para el evento de 2012 que para el evento de 2005 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
+    *
+
+    La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
+    * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
+    * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001].
+En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
+
 ## References

### 2017-09-15 analysis/analysis_resilience_bai.Rmd
update
change letters and update tables

@@
 
-cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01)
-cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01)
-cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01)
+letraArs <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rs')
+letraArs
+
+letraBrs <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rs')
+letraBrs # Manual (IMPROVE IT)
+letraBrs <- data.frame(Group = c('SJ', 'caH', 'caL'),
+                     Letter = c('a','a','a'),
+                     MonoLetter = c('a','a','a')) %>% mutate(var ='rs')
+letraBrs
+
+letraABrs  <- cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01) %>% mutate(var ='rs')
+letraABrs
+
 ```
@@
 
-cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01)
-cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01)
-cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01)
+letraArc <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rc')
+letraArc
+letraBrc <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rc')
+letraBrc
+letraABrc <- cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01) %>% mutate(var ='rc')
+letraABrc
+
 ```
@@
 
-cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01)
-cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01)
-cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01)
+letraArt <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rt')
+letraArt #Manual (IMPROVE IT)
+letraArt <- data.frame(Group = as.factor(c('25', '212')),
+                     Letter = as.factor(c('a','a')),
+                     MonoLetter = as.factor(c('a','a'))) %>% mutate(var ='rt')
+letraBrt <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rt')
+letraBrt
+letraABrt <- cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01) %>% mutate(var ='rt')
+letraABrt 
 ```
@@
 
-cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01)
-cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01)
-cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01)
+letraArrs <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rrs')
+letraArrs
+letraBrrs <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rrs')
+letraBrrs
+letraABrrs <- cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01) %>% mutate(var ='rrs')
+letraABrrs 
+```
+
+```{r}
+
+letrasA <- rbind(letraArs, letraArc, letraArt, letraArrs) %>% 
+  mutate(disturb_year = 
+           case_when(Group == "25" ~ "2005",
+                     Group == "212" ~ "2012")) %>% 
+  dplyr::select(-Group)
+
+letrasB <- letrasB %>% rename(site = Group)
+
+letrasAB <- rbind(letraABrs, letraABrc, letraABrt, letraABrrs) %>% 
+  separate(Group, into=c('disturb_year', 'site')) %>% 
+  mutate(disturb_year = 
+           case_when(disturb_year == "25" ~ "2005",
+                     disturb_year == "212" ~ "2012"))
 ```
@@
 mhuber <- rbind(rarc$mest, rart$mest, rars$mest, rarrs$mest)
-mhuber %>% pander()
+
+# add letras 
+mhuber <- mhuber %>% inner_join(letrasAB, by=c('var','disturb_year', 'site'))
+
+mhuber_agg <- mhuber %>% 
+  mutate(M.Huber = round(M.Huber, 4),
+         lower.ci = round(lower.ci, 4),
+         upper.ci = round(upper.ci, 4)) %>% 
+  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  dplyr::select(-MonoLetter) %>% 
+  dplyr::select(var, disturb_year:ci, Letter) %>% as.data.frame()
+
+mhuber_agg %>% pander() 
 ```
@@
 mhuber_a <- rbind(rarc$mest_a, rart$mest_a, rars$mest_a, rarrs$mest_a)
-mhuber_a %>% pander()
+mhuber_a <- mhuber_a %>% inner_join(letrasA, by=c('var','disturb_year'))
+
+mhuber_agg_a <- mhuber_a %>% 
+  mutate(M.Huber = round(M.Huber, 4),
+         lower.ci = round(lower.ci, 4),
+         upper.ci = round(upper.ci, 4)) %>% 
+  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  dplyr::select(-MonoLetter) %>% 
+  dplyr::select(var, disturb_year:ci, Letter) %>% as.data.frame()
+
+mhuber_agg_a %>% pander()
+
 ```
@@
 mhuber_b <- rbind(rarc$mest_b, rart$mest_b, rars$mest_b, rarrs$mest_b)
-mhuber_b %>% pander()
+mhuber_b <- mhuber_b %>% inner_join(letrasB, by=c('var','site'))
+
+mhuber_agg_b <- mhuber_b %>% 
+  mutate(M.Huber = round(M.Huber, 4),
+         lower.ci = round(lower.ci, 4),
+         upper.ci = round(upper.ci, 4)) %>% 
+  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  dplyr::select(-MonoLetter) %>% 
+  dplyr::select(var, site:ci, Letter) %>% as.data.frame()
+
+mhuber_agg_b %>% pander() 
 ```
@@
 write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/robust_mhuber.csv'), row.names = F)
+write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg.csv'), row.names = F)
 write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_a.csv'), row.names = F)
+write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg_a.csv'), row.names = F)
 write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_b.csv'), row.names = F)
+write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg_b.csv'), row.names = F)
 

### 2017-09-15 analysis/analysis_resilience_bai.Rmd
update tables mhuber

@@
                      MonoLetter = as.factor(c('a','a'))) %>% mutate(var ='rt')
+letraArt 
 letraBrt <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rt')
@@
 ```{r}
-
 letrasA <- rbind(letraArs, letraArc, letraArt, letraArrs) %>% 
@@
 
-letrasB <- letrasB %>% rename(site = Group)
+letrasB <- rbind(letraBrs, letraBrc, letraBrt, letraBrrs) %>% rename(site = Group)
 

### 2017-09-15 analysis/tables_resiliences.Rmd
update tables

---
title: "tabla_mean_values Mhuber"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017"
output:  
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, error=TRUE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("pander")
```

```{r, echo=FALSE}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
```

# Read data from EVI

```{r}
# Estos datos vienen del repo de modis_resilience (/analysis/explore_Resilience.Rmd)
mhuber_evi <- read.csv(file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber.csv'), header = TRUE, sep=",")
mhuber_eviA <- read.csv(file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_a.csv'), header = TRUE, sep=",")
mhuber_eviB <- read.csv(file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_b.csv'), header = TRUE, sep=",")
```

```{r}
# Estos datos vienen del repo de modis_resilience (/analysis/explore_Resilience.Rmd)
mhuber_bai <- read.csv(file=paste0(di, '/out/anovas_resilience/robust_mhuber.csv'), header = TRUE, sep=",")
mhuber_baiA <- read.csv(file=paste0(di, '/out/anovas_resilience/robust_mhuber_a.csv'), header = TRUE, sep=",")
mhuber_baiB <- read.csv(file=paste0(di, '/out/anovas_resilience/robust_mhuber_b.csv'), header = TRUE, sep=",")
```

### Tablas 
```{r}
evi_a <- mhuber_eviA %>% 
  mutate(M.Huber = round(M.Huber, 4),
         lower.ci = round(lower.ci, 4),
         upper.ci = round(upper.ci, 4)) %>% 
  unite("ci", c("lower.ci", "upper.ci"), sep=", ") %>% 
  mutate(ci = paste0('(', ci, ')')) %>% 
  unite("value", c("M.Huber", "ci"), sep= " ") %>%
  dplyr::select(-MonoLetter, -n) %>% 
  unite("value", c("value", "Letter"), sep= "^") %>% 
  dplyr::select(var, disturb_year, value) %>% as.data.frame()

evi_a %>% pander() 
```

### 2017-09-15 analysis/tables_resiliences.Rmd
create tables

@@
 
-### Tablas 
+### Tablas Disturb year  
 ```{r}
@@
   mutate(ci = paste0('(', ci, ')')) %>% 
-  unite("value", c("M.Huber", "ci"), sep= " ") %>%
+  unite("value_EVI", c("M.Huber", "ci"), sep= " ") %>%
   dplyr::select(-MonoLetter, -n) %>% 
-  unite("value", c("value", "Letter"), sep= "^") %>% 
-  dplyr::select(var, disturb_year, value) %>% as.data.frame()
+  dplyr::select(var, disturb_year, value_EVI, letter_EVI=Letter) %>% as.data.frame()
 
 evi_a %>% pander()
+
+bai_a <- mhuber_baiA %>% 
+  mutate(M.Huber = round(M.Huber, 4),
+         lower.ci = round(lower.ci, 4),
+         upper.ci = round(upper.ci, 4)) %>% 
+  unite("ci", c("lower.ci", "upper.ci"), sep=", ") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  unite("value_BAI", c("M.Huber", "ci"), sep= " ") %>%
+  dplyr::select(-MonoLetter, -n) %>% 
+  dplyr::select(var, disturb_year, value_BAI, letter_BAI=Letter) %>% as.data.frame()
+
+bai_a %>% pander()
+
+mhuberA <- evi_a %>% inner_join(bai_a, by=c("var", "disturb_year"))
+
+mhuberA %>% pander()
+
+mhuberA
+
 ```
+
+# Tablas site
+```{r}
+evi_b <- mhuber_eviB %>% 
+  mutate(M.Huber = round(M.Huber, 4),
+         lower.ci = round(lower.ci, 4),
+         upper.ci = round(upper.ci, 4)) %>% 
+  unite("ci", c("lower.ci", "upper.ci"), sep=", ") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  unite("value_EVI", c("M.Huber", "ci"), sep= " ") %>%
+  dplyr::select(-MonoLetter, -n) %>% 
+  dplyr::select(var, site, value_EVI, letter_EVI=Letter) %>% as.data.frame()
+
+evi_b %>% pander()
+
+bai_b <- mhuber_baiB %>% 
+  mutate(M.Huber = round(M.Huber, 4),
+         lower.ci = round(lower.ci, 4),
+         upper.ci = round(upper.ci, 4)) %>% 
+  unite("ci", c("lower.ci", "upper.ci"), sep=", ") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  unite("value_BAI", c("M.Huber", "ci"), sep= " ") %>%
+  dplyr::select(-MonoLetter, -n) %>% 
+  dplyr::select(var, site, value_BAI, letter_BAI=Letter) %>% as.data.frame()
+
+bai_b %>% pander()
+
+```
+
+# Tablas Interaction
+```{r}
+evi <- mhuber_evi %>% 
+  mutate(M.Huber = round(M.Huber, 4),
+         lower.ci = round(lower.ci, 4),
+         upper.ci = round(upper.ci, 4)) %>% 
+  unite("ci", c("lower.ci", "upper.ci"), sep=", ") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  unite("value_EVI", c("M.Huber", "ci"), sep= " ") %>%
+  dplyr::select(-MonoLetter, -n) %>% 
+  dplyr::select(var,  disturb_year, site, value_EVI, letter_EVI=Letter) %>% as.data.frame()
+
+evi %>% pander()
+
+bai <- mhuber_bai %>% 
+  mutate(M.Huber = round(M.Huber, 4),
+         lower.ci = round(lower.ci, 4),
+         upper.ci = round(upper.ci, 4)) %>% 
+  unite("ci", c("lower.ci", "upper.ci"), sep=", ") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  unite("value_BAI", c("M.Huber", "ci"), sep= " ") %>%
+  dplyr::select(-MonoLetter, -n) %>% 
+  dplyr::select(var, disturb_year, site, value_BAI, letter_BAI=Letter) %>% as.data.frame()
+
+bai %>% pander() 
+```

### 2017-09-15 analysis/tables_resiliences.Rmd
update tablas

@@
 date: "2017"
-output:  
-  md_document:
-    variant: markdown_github
+output: pdf_document
 bibliography: ../references.bib 
@@
 
-### Tablas Disturb year  
-```{r}
+## Tablas Disturb year  
+```{r, echo=FALSE}
 evi_a <- mhuber_eviA %>% 
@@
 mhuberA %>% pander() 
-
-mhuberA
-
 ```
 
-# Tablas site
-```{r}
+## Tablas site
+```{r, echo=FALSE}
 evi_b <- mhuber_eviB %>% 
@@
 
-# Tablas Interaction
-```{r}
+## Tablas Interaction
+```{r, echo=FALSE}
 evi <- mhuber_evi %>% 

### 2017-09-15 man/ms_0_intro.Rmd
update ms and add compile document

@@
 title: "ms_0_intro"
-author: "AJ Perez-Luque (@ajpelu); G. Gea-Izquierdo"
-date: "2017"
 output:  
@@
 
-```{r setup, include=FALSE}
+```{r, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)

### 2017-09-15 man/ms_1_material_methods.Rmd
update ms and add compile document

@@
 title: "material and methods"
-author: "AJ Perez-Luque (@ajpelu); G. Gea-Izquierdo"
-date: "2017"
 output:  
@@
 
------ 
 :red_circle: `duda aqui` Varias referencias hablan de los años 2005 y 2012 como  extremadamente secos.
@@
 * O quizá un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
------
 

### 2017-09-15 man/ms_3_discussion.Rmd
update ms and add compile document

--- 2017-05-31 man/ms_discussion.Rmd
@@
     variant: markdown_github
-bibliography: references.bib
-csl: ecology.csl
+bibliography: ../references.bib
+csl: ../ecology.csl
 ---
 
-```{r setup, include=FALSE}
+```{r, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)

### 2017-09-15 man/report.Rmd
update ms and add compile document

---
title: "ms_borrador"
author: "AJ Perez-Luque; G. Gea-Izquierdo; R. Zamora; FJ. Bonet"
date: "2017"
output:  word_document
bibliography: ../references.bib
csl: ../ecology.csl
---

```{r child = 'ms_0_intro.Rmd'}
```

```{r child = 'ms_1_material_methods.Rmd'}
```

```{r child = 'ms_2_results.Rmd'}
```

### 2017-09-15 man/ms_2_results.Rmd
update ms r

@@
 title: "ms_2_results"
-author: "AJ Perez-Luque (@ajpelu); G. Gea-Izquierdo"
-date: "2017"
 output:  
@@
 
-:red_circle: referencia a las tendencias en EVI (lo hemos vuelto a calcular y además en el trabajo de ontologías también nos sale)
+referencia a las tendencias en EVI (lo hemos vuelto a calcular y además en el trabajo de ontologías también nos sale)
 
@@
 Sin embargo vemos valores menos negativos para 2005.
-🔴 Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years 🔴
+Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years
 
-Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 ± 0.029)que para 2012 (-0.418 ± 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
+Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029)que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
 
-Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (Fig 1a :red_circle: EVI profile comparison), particu
+Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (:red_circle: Fig 1a EVI profile comparison), particu
 
@@
     * En general mayor recuperación para el evento de 2012 que para el evento de 2005 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
-    *
-
-    La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
-    * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
-    * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001].
-En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
 

### 2017-09-15 man/ms_1_material_methods.Rmd
fix typo

@@
 
-This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m a.s.l.
+This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*.
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
-There are eight oak patches (2400 Has) identified (:red_circle: FIGURE) in this mountain range, ranging between 1100 and 2000 *m a.s.l.* and generally associated to major river valleys.
+There are eight oak patches (2400 Has) identified (:red_circle: FIGURE) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys.
 Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation (Brewer et al.

### 2017-09-15 man/ms_1_material_methods.Rmd
update methods

@@
 bibliography: ../references.bib
+templat
 csl: ../geb.csl
@@
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
-There are eight oak patches (2400 Has) identified (:red_circle: FIGURE) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys.
+There are eight oak patches (2400 Has) identified (**Figure 1**) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys.
 Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation (Brewer et al.
@@
 
-:red_circle: La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset
-
 ## Datos de sequía.
@@
 
-To characterize the vegetation greeness of *Quercus pyrenaica* we used the Enhanced Vegetation Index (EVI) derived from MOD13Q1 product obtained by the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor [@Didan2015].
-EVI and NDVI (Normalized Difference Vegetation Index) are the most common greenness vegetation indices.
-We used EVI instead of NDVI (Normalized Difference Vegetation Index) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002, @Krapivin2015, @Cabello2012].
+To characterize the vegetation greenness of *Q.
+pyrenaica* we used the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
+EVI and NDVI (*Normalized Difference Vegetation Index*) are the most common greenness vegetation indices.
+We used EVI instead of NDVI because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].
 
-EVI product consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 231 m x 231 m.
-Data were obtained using a Google Earth Engine script (:red_circle: cite gists) for the 2000 - 2016 period.
-We selected the pixels covering the distribution of Quercus pyrenaica forests in Sierra Nevada (*n* = 928 pixels).
+EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 231 m x 231 m.
+We selected the pixels covering the distribution of *Q.
+pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
+MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016.
+
+OJITO---------- reescribir esto
+
+A quality filter was applied to select only valid values of EVI.
+
+First we used information from the Quality Assesment band (QA band) to filter out those values affected by high content of aerosols, clouds, shadows, snow or water;
+
+snow, ice, cloud
+
+and then a quality assessment was carried out to filter the ... [@Reyes2015]
+
+    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds.
+On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]
+
+EVI data were
+
+The presence of clouds (adjacent clouds, mixed clouds and shadows) 'obscures' the surface in a radiometric sense, thus corrupting inferred EVI values.
+In addition, two types of aerosol loadings typically corrupt EVI—climatology and high aerosols.
+Use of aerosol climatology
+
+Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2).
+Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked.
+Besides, VI usefulness indices between 0 to 11 essentially include all EVI data.
+Thus, these two conditions serve as additional checks.
+
 The EVI data are geometrically and atmospherically corrected and include information about the quality ass....
-:red_circle:
 
@@
 :red_circle: reescribir esto de la calidad. Ver Samanta et al 2012 y como describe el proceso de calidad
+
+OJITO---------- reescribir esto de arriba
 
@@
 Samplig was carried during autumn of 2016.
-Trees were sampled at two locations located in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Table 1).
-Both sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..).
+Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1).
+Two elevations were sampled for the southern site (CA-Low and CA-High).
+All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..).
 In each site between 15 and 20 dominant trees were randomly selected.
-Two cores of 5 mm of diameter were taken per tree at 1.3 m using an increment borer.
-Diameter at breat height (DBH) and total height were recorded for each tree.
+Two cores of 5 mm of diameter were taken from each at breast heith (1.3 m) using an increment borer.
+For each tree, diameter at breat height (DBH) and total height were recorded.
 Increment cores were air dried, glued onto wooden mounts and sanded.
-Annual radial growth (ring width, RW) were measured with a LINTAB measuring device (:red_circle: Rinntech 2003) coupled to a stereomicroscope, with an accuracy of 0.01 mm.
-Individual ring series were visually and statistically cross-dated with TSAP software (:red_circle: Frank Rinn, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
+Annual radial growth (ring width, RW) were measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.01 mm.
+Individual ring series were visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
 Validation of the croos-dating was done using COFECHA software [@Holmes1983].
@@
 ## Dendrochronological methods
-... dendro
-For each focal tree we measured diameter at breast height (DBH) and total height. 
-A total of xx trees were sampled. 
-...
-
-We built chronologies for each site (two)
+* incluir citas a Frits; Cooks ... 
+* site cronologies
+* BAI 
+* GC Nowacki??
 

### 2017-09-17 man/ms_1_material_methods.Rmd
update

@@
 
-EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 231 m x 231 m.
+EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
 We selected the pixels covering the distribution of *Q.
@@
 MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016.
+A data filtering was applied to select EVI valid values.
+The filtering was done using quality flags and VI Usefulness Indices accompanying the EVI data.
+We filter out those values affected by high content of aerosols, clouds, shadows, snow or water.
 
-OJITO---------- reescribir esto
-
-A quality filter was applied to select only valid values of EVI.
-
-First we used information from the Quality Assesment band (QA band) to filter out those values affected by high content of aerosols, clouds, shadows, snow or water;
-
-snow, ice, cloud
+Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2).
+Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked.
+Besides, VI usefulness indices between 0 to 11 essentially include all EVI data.
+Thus, these two conditions serve as additional checks.
 
@@
 
-EVI data were
-
 The presence of clouds (adjacent clouds, mixed clouds and shadows) 'obscures' the surface in a radiometric sense, thus corrupting inferred EVI values.
@@
 Use of aerosol climatology
-
-Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2).
-Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked.
-Besides, VI usefulness indices between 0 to 11 essentially include all EVI data.
-Thus, these two conditions serve as additional checks.
 

### 2017-11-06 man/ms_0_intro.Rmd
update

@@
 ### Sequías
+
+Although summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
+
+In the Iberian Peninsula, major drought episodies were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
+The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013].
+These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
+
+### Ideas a incluir
+
 * :red_circle: Algo de las sequías de 2005 y 2014
@@
 
-### Ideas a incluir
-
-* Descripción de las sequía de 2005 y 2014
-
-Major drought episodes in the IP were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014]
-
-La sequía es una característica del clima mediterráneo [@Lionello2012], pero en los últimos años se ha venido observando un aumento de la frecuencia y la severidad de los periodos de sequía (:red_circle: CITA;  @VicenteSerrano2014 para la P.
-Iberica).
-
-The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southeren sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013].
-These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
-
-an increase in the drought severity in the Iberian Peninsula has been observed in the last decades [@VicenteSerrano2014].
+> notas sequia
+> 
+> * the drought events have been longest and most severe in the period 1991-2010 for mediterranean are of Southern Europe  
+> * Seasonally, drought frequency is projected to increase everywhere in Europe for both scenarios in spring and summer, especially over southern Europe, and less intensely in autumn [@Spinoni2017a]
+> Aumento de la frecuencia de sequías en spring and summer desde 1950 hasta 2014 [@Spinoni2017b]
+> Both for frequency and severity, the evolution towards drier conditions is more relevant in the last three decades over Mediterranean area in summer, 
+> an increase in the drought severity in the Iberian Peninsula has been observed in the last decades [@VicenteSerrano2014].
 
 frecuencia de las sequías severas (ver Dai 2011)
-
-Algunos eventos extremos han aumentado su frecuencia en los últimos años
 

### 2017-11-06 pres/state_ms.Rmd
update

--- 2016-11-30 analysis/geo.Rmd
@@
 ---
-title: "Get topographic info of gps points"
-author: "AJ Perez-Luque (@ajpelu)"
-date: "2016 Nov"
+title: "Resiliencia de *Quercus pyreancia* a dos eventos de sequía"
+author: "Perez-Luque AJ; Gea-Izquierdo G; Zamora R; Bonet FJ"
 output:
-  md_document:
-    variant: markdown_github
----
-
-```{r packages, warning=FALSE, message=FALSE}}
-library("rgdal")
-library("sp")
-library("raster")
-library("dplyr")
-```
-
-# Read data 
-```{r wd, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
-di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
-
-# Connect with NAS and copy the mde file located at 
-# '/cartografia/Informacion_Ambiental/BASES_DE_REFERENCIA_TERRITORIAL/MDE_SN_AreaInfluecia_10m/mde_sn_area_influencia.asc' 
-#---------------------------------
-```
-
-# Extract data
-```{r} 
-ditemp <- paste('/Users/', machine, '/Documents/', sep='')
-setwd(ditemp)
-
-# read DEM 
-demsn <- raster("mde_sn_area_influencia.asc")
-
-setwd(di)
-
-## Projection 
-# Get projection
-projection(demsn)
-
-# Set projection
-crs(demsn) <- "+proj=utm +zone=30 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
-
-# Read points data
-
-gps_point <- readOGR(paste0(di, 'data_raw/geoinfo/'), layer = 'dendro_garmin', verbose = FALSE)
-
-# Read pixels data 
-eviresi <- read.csv(file=paste(di, "/data/evi_resilience.csv", sep=""), header = TRUE, sep = ',')
-
-## convert to spatial df 
-eviresi_sp <- SpatialPointsDataFrame(coords = eviresi[,c('lng','lat')], data = eviresi,
-                               proj4string = CRS("+init=epsg:4326"))
-## reproject 
-eviresi_sp <- spTransform(eviresi_sp, crs(demsn))
-
-evi_elev <- extract(demsn, eviresi_sp, method='simple', sp=TRUE)
-
-elev <- evi_elev@data %>% 
-  dplyr::select(iv_malla_modi_id, mde_sn_area_influencia) %>%
-  transmute(iv_malla_modi_id, elev = mde_sn_area_influencia)
-
-# Export data
-write.csv(elev, file=paste(di, "/data/elev.csv", sep=""), row.names = FALSE)
-```
-title: "Untitled"
-output: html_document
+  rmdshower::shower_presentation:
+    self_contained: false
+    katex: true
+    ratio: 16x10
 ---
@@
 
-## R Markdown
+## Introduccion
 

### 2017-11-07 analysis/drought/explore_drought_spei_regional.Rmd
regional drought analysis from old repo

---
title: "SPEI drought SN"
author: "AJ Perez-Luque (@ajpelu)"
date: "2016 March"
output:  
    md_document:
      variant: markdown_github

---
# Drought data

* Obtain data of drought index SPEI from [SPEI Global Drought Monitor](http://sac.csic.es/spei/map/maps.html)
* Query data from Sierra Nevada [36.75, -3.75], [37.25, -3.25]

```{r metadata, echo=FALSE}
# Set working directory 
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

```

```{r packages, warning=FALSE, echo=FALSE, message=FALSE}
# Load packages 
library("dplyr")
library("stringr")
library("lubridate")
library("reshape2")
library("ggplot2")
source(paste0(di,"/script/R/exportpdf.R"))
```

## Prepare data
```{r}

# Get file names
myfiles <- list.files(path = paste(di, "/data_raw/spei_all", sep=""), pattern = "\\.csv$")

mydf <- data.frame()

# Loop to read files and prepare data
for (j in myfiles){ 
  aux <- read.csv(file=paste(di, "/data_raw/spei_all/", j, sep= ""),
              header = TRUE,
              sep = ',')

  # Remove csv and get name 
  name_aux <- str_replace(j, ".csv", "")

  # Get lat long
  latlong <- str_replace(name_aux, "SPEI_", "")

  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]

  aux$lat <- mylat
  aux$long <- mylong

  # Split time 
  # Get year 
  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))

  # Get month 
  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")

  # Variable month in number 
  aux$months <- match(aux$month, month.abb)

  # assign(name_aux, aux)

  mydf <- rbind(mydf, aux)
}
```

Select data for SPEI_1, SPEI_3, SPEI_6, SPEI_12, SPEI_24, and SPEI_48 and compute the mean of the 6 cells (0.5º size)

```{r}
spei <- mydf %>% 
  select(SPEI_1,SPEI_3,SPEI_6,SPEI_12,SPEI_24, SPEI_48, lat, long, year, months)

# Get mean values of the all cells (4 celss ) 
# and create a date format variable 
spei_mean <- spei %>% 
  select(-lat, -long) %>% 
  group_by(year,months) %>% 
  summarise(spei1 = mean(SPEI_1),
            spei3 = mean(SPEI_3),
            spei6 = mean(SPEI_6),
            spei12 = mean(SPEI_12),
            spei24 = mean(SPEI_24),
            spei48 = mean(SPEI_48)) %>%
  mutate(date = lubridate::ymd(paste(year, months, "01")))

# Melting data and compute a sing auxiliar column to colour bars 
spei_mean_melt <- melt(spei_mean, id.vars = c("year", "months", "date"))
spei_mean_melt$sign <- ifelse(spei_mean_melt$value > 0, "pos", "neg")

spei_mean_melt2000 <- spei_mean_melt %>% 
  filter(year >= 2000)
```

### Plot 
```{r}
label_variables <- c('spei1' = '1-month',
                     'spei3' = '3-month',
                     'spei6' = '6-month',
                     'spei12' = '12-month',
                     'spei24' = '24-month',
                     'spei48' = '48-month')

d <- ggplot(spei_mean_melt, aes(x=date, y=value, fill=sign)) +  
  geom_bar(stat = "identity") + 
  facet_wrap(~variable, ncol=1, labeller = as_labeller(label_variables)) + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
  theme_bw()  + 
  labs(title= 'Drought evolution for Sierra Nevada',
    x= '',
    y='SPEI') +
  theme(strip.background = element_rect(fill = "white"),
        legend.position = "none")

d

# >2000
d2000 <- ggplot(spei_mean_melt2000, aes(x=date, y=value, fill=sign)) +  
  geom_bar(stat = "identity") + 
  facet_wrap(~variable, ncol=1, labeller = as_labeller(label_variables)) + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
  theme_bw()  + 
  labs(title= 'Drought evolution for Sierra Nevada (From year 2000)',
    x= '',
    y='SPEI') +
  theme(strip.background = element_rect(fill = "white"),
        legend.position = "none")

d2000
```

# Explore 3, 6 months scales
```{r}
spei_mean_melt1970 <- spei_mean_melt %>% 
  filter(year >= 1970) %>%
  filter(variable %in% c('spei3', 'spei6'))

d1970 <- ggplot(spei_mean_melt1970, aes(x=date, y=value, fill=sign)) +  
  geom_bar(stat = "identity") + 
  facet_wrap(~variable, ncol=1, labeller = as_labeller(label_variables)) + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
  theme_bw()  + 
  labs(title= 'Drought evolution for Sierra Nevada (From year 1995)',
    x= '',
    y='SPEI') +
  theme(strip.background = element_rect(fill = "white"),
        legend.position = "none") +
  scale_x_date(date_breaks = "2 year", date_labels = "%Y")

d1970
```

```{r}
### Export all pdfs
di_export <- paste0(di, "/out/fig/drought")

pdf(file=paste0(di_export, "/spei_sn_all.pdf"), height = 10, width =9)
d 
dev.off()

pdf(file=paste0(di_export, "/spei_sn_all_2000.pdf"), height = 10, width =9)
d2000
dev.off()

pdf(file=paste0(di_export, "/spei_sn_all_1970.pdf"), height = 5, width =9)
d1970
dev.off()
```

### 2017-11-07 pres/state_ms.Rmd
clear pres
remove error checkout

@@
 output: 
-  rmdshower::shower_presentation:
-    self_contained: false
-    katex: true
-    ratio: 16x10
+  beamer_presentation
+bibliography: ../references.bib
+csl: ../nature.csl
 ---
@@
 
-## Introduccion
+# Introduccion
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+- Item 1 
+    - sub
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+---
 
-```{r cars}
-summary(cars)
-```
+# Sequía
+
+- Aumento en la **severidad** y **frecuencia** de los eventos de sequías en las últimas décadas, especialmente para el sur de Europa [@VicenteSerrano2014; @Spinoni2015; @Stagge2017]
+- Eventos extremos P. Ibérica: 1981, 1995, 2000, **2005**, **2012**  [@GarciaHerrera2007; @Gouveia2015; @Trigo2013]
+
+## Análisis adicionales
+
+- SPEI (Standardised Precipitation-Evapotranspiration Index) para datos
+
+---
+
+Sequía (análisis adicionales): SPEI
+
+- Escala regional: exploramos la sequía a escala regional utilizando datos del [SPEI Global Drought Monitor](http://sac.csic.es/spei/database.html) para Sierra Nevada (spatial resolution of 0.5º) Ver [esto](https://github.com/ajpelu/qpyr_resilience/blob/master/analysis/drought/explore_drought_spei_regional.md):
 
@@
 
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
 Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2017-11-07 analysis/drought/explore_drought_station.Rmd

---
title: "SPEI by stations"
author: "AJ Perez-Luque (@ajpelu)"
date: "2016 Sep"
output:  
  md_document:
    toc: true
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Set working directory 
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

#---------------------------------
```

```{r, warning=FALSE, message=FALSE}
library('plyr')
library('dplyr')
library('lubridate')
library('SPEI')
library('ggplot2')
library('zoo')
library('plotly')
library('RCurl')
source(paste0(di, '/script/R/hydro_year.R'))
library("rgdal")
library("sp")
library("raster")
library("leaflet")
```

## SPEI Sierra Nevada 
* Compute SPEI of Sierra Nevada for three stations from [RIA network](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController): [Cadiar](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=7), [Padul](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=10) and [Jerez del Marquesado](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=6)
* Temporal serie: 2000 - 2016

```{r, eval=FALSE}
# Map of station

geoinfo_station <- read.table(file=paste0(di, '/data_raw/meteo/metadata_station_ria.txt'), header=TRUE, sep=';')

m <- leaflet(geoinfo_station) %>%  
    addProviderTiles("Esri.WorldImagery", 
                   group='Satellite') %>% 
  addMarkers(~long, ~lat, popup = ~as.character(name))

m
```

* Read raw data and prepare data: 
    * Get month and year
    * Compute hydrological year
    * Filter (up to 2015)
    * Compute the precipitation, evapotranspiration and mean temperature (average by month)

```{r}
cadiar <- read.delim(file=paste0(di, '/data_raw/meteo/Cadiar.txt'), header=T, sep='')
padul <- read.delim(file=paste0(di, '/data_raw/meteo/Padul.txt'), header=T, sep='')
jerez <- read.delim(file=paste0(di, '/data_raw/meteo/Jerez del Marquesado.txt'), header=T, sep='')

# Prepare data 
ca <- cadiar %>% 
  filter(complete.cases(.)) %>% 
  mutate(date = as.Date(FECHA, format="%d-%m-%y"), 
         year = lubridate::year(date), 
         month = lubridate::month(date),
         hyear = hydro_year(date), 
         hyear_f = paste0(hyear-1,'-', hyear)) %>%
  filter(year < 2016) %>% 
  dplyr::select(date, year, month, doy = DIA, tmed = Gr07TMed, prec = Gr07Precip, et0 = Gr07ETo, hyear, hyear_f) %>% 
  # Group by year and month 
  group_by(year, month, hyear, hyear_f) %>% 
  dplyr::summarise(prec_ac = sum(prec),
                   et0_ac = sum(et0),
                   tmed_avg = mean(tmed))

je <- jerez %>% 
  filter(complete.cases(.)) %>% 
  mutate(date = as.Date(FECHA, format="%d-%m-%y"), 
         year = lubridate::year(date), 
         month = lubridate::month(date),
         hyear = hydro_year(date), 
         hyear_f = paste0(hyear-1,'-', hyear)) %>%
  filter(year < 2016) %>% 
  dplyr::select(date, year, month, doy = DIA, tmed = Gr06TMed, prec = Gr06Precip, et0 = Gr06ETo, hyear, hyear_f) %>% 
  # Group by year and month 
  group_by(year, month, hyear, hyear_f) %>% 
  dplyr::summarise(prec_ac = sum(prec),
            et0_ac = sum(et0),
            tmed_avg = mean(tmed))

pa <- padul %>% 
  filter(complete.cases(.)) %>% 
  mutate(date = as.Date(FECHA, format="%d-%m-%y"), 
         year = lubridate::year(date), 
         month = lubridate::month(date),
         hyear = hydro_year(date), 
         hyear_f = paste0(hyear-1,'-', hyear)) %>%
  filter(year < 2016) %>% 
  dplyr::select(date, year, month, doy = DIA, tmed = Gr10TMed, prec = Gr10Precip, et0 = Gr10ETo, hyear, hyear_f) %>% 
  # Group by year and month 
  group_by(year, month, hyear, hyear_f) %>% 
  dplyr::summarise(prec_ac = sum(prec),
            et0_ac = sum(et0),
            tmed_avg = mean(tmed))

```

## SPEI index 
Now compute the SPEI index. We first created a function to compute the spei (using SPEI package) in our dataframes.

```{r}
# Compute the spei index

compute_spei <- function(x, rango_mensual, inicio){
  # x = dataframe 
  # rango_mensual = vector with temporal range of spei

  # dataframe for output 
  outdf <- data.frame()

  for (i in rango_mensual){ 
    # Compute spei
    my_spei <- spei(x$prec_ac - x$et0_ac, i)

    # Create the ts 
    ts_aux <- ts(my_spei$fitted, start=inicio,frequency=12)

    # Convert to dataframe 
    tsdf <- data.frame(Y=as.matrix(ts_aux), fecha=as.Date(as.yearmon(time(ts_aux))))

    # Rename 
    names(tsdf) <- c('value', 'fecha')

    tsdf <- tsdf %>% 
       mutate(signo = ifelse(value >= 0, 'pos', 'neg'),
              spei = paste0('spei_', i))

    # Rbind 
    outdf <- rbind(outdf, tsdf)

  }

  outdf <- outdf %>% 
    mutate(spei_ord = plyr::mapvalues(spei,
                                      c('spei_1', 'spei_3', 'spei_6', 'spei_12', 'spei_24'),
                                      c('spei_01', 'spei_03', 'spei_06', 'spei_12', 'spei_24')))

} 
```

### Plot the SPEI indices for all sites

```{r}
# SPEI Cadiar
spei_ca <- compute_spei(ca, rango_mensual = c(1,3,6,12,24), inicio = c(2000,9))

df <- spei_ca
site <- 'Cadiar'

# PLOT 
ggplot(df, aes(x=fecha, y=value, fill=signo)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
  facet_wrap(~spei_ord, ncol=1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  theme_bw() + 
  ylab('SPEI') + 
  xlab('year') + 
  theme(legend.position = "none", 
        strip.background = element_blank()) +
  ggtitle(site)
```

```{r}
# SPEI Padul
spei_pa <- compute_spei(pa, rango_mensual = c(1,3,6,12,24), inicio = c(2000,9))

df <- spei_pa
site <- 'Padul'

# PLOT 
ggplot(df, aes(x=fecha, y=value, fill=signo)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
  facet_wrap(~spei_ord, ncol=1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  theme_bw() + 
  ylab('SPEI') + 
  xlab('year') + 
  theme(legend.position = "none", 
        strip.background = element_blank()) +
  ggtitle(site)
```

```{r}
# SPEI Jerez 
spei_je <- compute_spei(je, rango_mensual = c(1,3,6,12,24), inicio = c(2000,9))

df <- spei_je
site <- 'Jerez Marquesado'

# PLOT 
ggplot(df, aes(x=fecha, y=value, fill=signo)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
  facet_wrap(~spei_ord, ncol=1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  theme_bw() + 
  ylab('SPEI') + 
  xlab('year') + 
  theme(legend.position = "none", 
        strip.background = element_blank()) +
  ggtitle(site)
```

## Precipitation curves

We plot the cummulated precipitation (from september to september next year) (See Trigo et al. 2013).

```{r}

mysites <- c('ca','je','pa')
all_sites <- data.frame()

for (i in mysites){

  df <- get(i)
  aux <- df %>% 
    dplyr::mutate(hmonth = ifelse(month <= 8, month + 4, month -8)) %>% 
    group_by(hyear_f, hmonth) %>% 
    dplyr::summarise(prec_ac = sum(prec_ac)) %>%
    mutate(csum = cumsum(prec_ac)) %>% 
    mutate(loc = i) %>% 
    as.data.frame()

  all_sites <- rbind(all_sites, aux)
}

all_sites <-  all_sites %>% 
  mutate(station= plyr::mapvalues(loc, c('ca', 'je', 'pa'), c("Cadiar", "Jerez", "Padul")))

all_sites %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot() + facet_wrap(~station) + 
  geom_smooth(aes(group=1), se=TRUE, level=0.95) + 
  geom_line(data=subset(all_sites, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1)) +
  geom_line(data=subset(all_sites, hyear_f == '2011-2012'),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') + xlab('') +
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug"))
```

The boxplots show the accumulated monthly precipitation (average from 2000 to 2015; 2005 and 2012 hidro. years are excluded). The red and black lines corresponding to 2004-2005 and 2011-2012 hydrological years respectively.

### Interactively by year

```{r}
g <- ggplot(data=all_sites, aes(x=as.factor(hmonth), y=csum, group=hyear_f, colour=hyear_f)) + 
  geom_line() + facet_wrap(~station) + theme_bw() +
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  xlab('months (hydrological year; 1= sep)') +
  ylab('Precipitation (mm) acumulated')

p <- ggplotly(g)
print(p)
```

## Long term series REDIAM

### Data from stations 5514 and 5515

We requested official data to REDIAM from the following stations: 
* 5514 Granada Base Aérea 
* 5515 Granada Cartuja

as long as these two stations are the longest term series for southern Spain.

```{r}
ltdf <- read.table(file=paste0(di, '/data_raw/meteo/long_term_data_rediam_data.txt'), header=T, sep=';', dec=',')

ltdf <- ltdf %>% 
  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
         month = lubridate::month(fecha),
         hmonth = ifelse(month <= 8, month + 4, month -8), 
         hyear = hydro_year(fecha),
         hyear_f = paste0(hyear-1,'-', hyear))
```

#### 5514 station

It's the longest term series.

```{r}
# We used only the 5514 station (the most long term)
# Filter by station and prec
# Filter from 1950 
st5514 <- ltdf %>% 
  filter(VARIABLE == 'PM1') %>% 
  filter(cod == '5514') %>% 
  filter(hyear >= 1950) %>% 
  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)

st5514_pro <- st5514 %>% 
  group_by(hyear_f, hmonth) %>% 
  dplyr::summarise(pre = sum(pre)) %>%
  dplyr::mutate(csum = cumsum(pre))

st5514_pro  %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot() +  
  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') + xlab('') + 
  ggtitle('Granada Base Aerea, 1950-2015') + 
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug"))
```

Fig. Acumulated monthly precipitation during the hydrological year 2004-2005 (blue line) and 2011-2012 (red line). The boxplot representing the averaged from 1940-2015 period.

##### Cumulated precipitation from 2000 to 2015 for 5514 station

```{r}
# Cumulated precipitation from 2000 to 2015 for 5514 station 
st5514_2000  <- ltdf %>% 
  filter(VARIABLE == 'PM1') %>% 
  filter(cod == '5514') %>% 
  filter(hyear >= 1999) %>% 
  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR) %>% 
  group_by(hyear_f, hmonth) %>% 
  dplyr::summarise(pre = sum(pre)) %>%
  dplyr::mutate(csum = cumsum(pre))

st5514_2000 %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot() +  
  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
  geom_line(data=subset(st5514_2000, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
  geom_line(data=subset(st5514_2000, hyear_f == '2011-2012'),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') +  xlab('') + 
  ggtitle('Granada Base Aerea, 2000-2015') + 
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug"))

```

#### Interactive plots by year

```{r}
ba_ip <- ggplot(data=st5514_pro, 
       aes(x=as.factor(hmonth), y=csum, group=hyear_f, colour=as.factor(hyear_f))) +
  geom_line() + 
  theme_bw() + 
  ylab('Cumulative precipitation (mm)') + 
  xlab('') + ggtitle('Granada Base Aerea, 1950-2015') + 
  theme(legend.title = element_blank()) +
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug"))

ba_ipp <- ggplotly(ba_ip)
print(ba_ipp)
```

```{r, echo=FALSE, eval=FALSE, warning=FALSE}
## This code is for the data of REDIAM (not official)

# Data from REDIAM 
r <- read.table(file=paste0(di, '/data_raw/meteo/meteo_base_aerea_rediam.csv'), header=T, sep=';', dec=',')

r <- r %>%
  dplyr::select(year = AGNO, month = MES, pre=PREFINAL, tmed = TEMPMED) %>%
  mutate(fecha=as.Date(paste(year, month, '01', sep='-'), format='%Y-%m-%d'),
         hyear = hydro_year(fecha),
         hyear_f = paste0(hyear-1,'-', hyear))

rba <- r %>% 
  mutate(hmonth = ifelse(month <= 8, month + 4, month -8)) %>%
  filter(year >= 1951) %>% 
  filter(!(year == 1951 & hmonth >= 5)) %>% 
  group_by(hyear_f, hmonth) %>% 
  summarise(pre = sum(pre)) %>%
  mutate(csum = cumsum(pre))

rba %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot() +  
  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
  geom_line(data=subset(rba, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
  geom_line(data=subset(rba, hyear_f == '2011-2012'),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') + 
  xlab('months (hydrological year; 1= sep)') 
```

```{r, eval=FALSE, warning=FALSE, message=FALSE}
### NOT EVALUATED -----------------------------

###  Long term serie for Granada Base Aérea (From Linaria)
raw <- read.csv(file=paste0(di, '/data_raw/meteo/meteo_data_base_aerea.csv'), header=T, sep=';')

raw <- raw %>% 
  dplyr::select(-codigo) %>% 
  mutate(miyear = lubridate::year(fecha),
         mimonth = lubridate::month(fecha),
         hyear = hydro_year(fecha),
         hyear_f = paste0(hyear-1,'-', hyear))

# Group by month-year and summarize 
pre <- raw %>%
  filter(codigo.1 == 'PI1') %>% 
  group_by(miyear, mimonth, hyear, hyear_f) %>% 
  dplyr::summarize(prec_ac = sum(valor))

tmed <- raw %>%
  filter(codigo.1 == 'TI1') %>% 
  group_by(miyear, mimonth, hyear, hyear_f) %>% 
  dplyr::summarize(tmed = mean(valor))

ba <- pre %>% 
  inner_join(tmed, by=c('miyear', 'mimonth', 'hyear', 'hyear_f'))

# Compute et0 
# Latitud: 37° 8' 13'' N - Longitud: 3° 37' 53'' O 
lat_ba <- 37.13694
ba$et0_ac <-thornthwaite(ba$tmed,lat_ba)

# Export data
write.csv(ba, file=paste0(di, '/data_raw/meteo/base_aerea_aggregated.csv'), row.names = FALSE)

spei_ba <- compute_spei(ba, rango_mensual = c(6,12,24), inicio = c(1951,1))

df <- spei_ba
site <- 'Base Aerea'

# PLOT 
ggplot(df, aes(x=fecha, y=value, fill=signo)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
  facet_wrap(~spei_ord, ncol=1) +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") + 
  theme_bw() + 
  ylab('SPEI') + 
  xlab('year') + 
  theme(legend.position = "none", 
        strip.background = element_blank()) +
  ggtitle(site) + 
  geom_vline(xintercept=as.numeric(as.Date("2005-01-01"))) +
  geom_vline(xintercept=as.numeric(as.Date("2005-12-31")))

pre_acu <- pre %>% 
  mutate(hmonth = ifelse(mimonth <= 8, mimonth + 4, mimonth -8)) %>%
  filter(!(miyear == 1951 & hmonth >= 5)) %>% 
  mutate(hyear_corrected = hyear -1)

cumulate_preci <- pre_acu %>% 
  group_by(hyear_corrected, hmonth) %>% 
  summarise(prec_ac = sum(prec_ac)) %>%
  mutate(csum = cumsum(prec_ac))

cumulate_preci %>% 
  filter(!(hyear_corrected %in% c(2004, 1998))) %>% 
  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot() + 
  geom_smooth(aes(group=1), se=TRUE, level=0.95) + 
  geom_line(data=subset(cumulate_preci, hyear_corrected == 2004),
            aes(x=as.factor(hmonth), y=csum, group=1)) +
  geom_line(data=subset(cumulate_preci, hyear_corrected == 1998),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + ylab('Precipitation (mm) acumulated') + 
  xlab('months (hydrological year; 1= sep)') + 
  annotate("text", label = "2004-2005", x = 2, y = 500, size = 6, colour = "black") + 
  annotate("text", label = "1998-1999", x = 2, y = 550, size = 6, colour = "red") + 
  annotate("text", label = "mean 1950-2009", x = 3, y = 600, size = 6, colour = "blue")
```

### 2017-11-07 analysis/drought/explore_drought_spei_regional.Rmd
update regional drought

@@
   theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada (From year 1995)',
+  labs(title= 'Drought evolution for Sierra Nevada (From year 1970)',
     x= '',

### 2017-11-07 pres/state_ms.Rmd
update

@@
 
-- SPEI (Standardised Precipitation-Evapotranspiration Index) para datos
+- SPEI (Standardised Precipitation-Evapotranspiration Index) para datos regionales procedentes de SPEI Global Drought Monitor.
+
+    - spatial resolution of 0.5º
 
@@
 
-- Escala regional: exploramos la sequía a escala regional utilizando datos del [SPEI Global Drought Monitor](http://sac.csic.es/spei/database.html) para Sierra Nevada (spatial resolution of 0.5º) Ver [esto](https://github.com/ajpelu/qpyr_resilience/blob/master/analysis/drought/explore_drought_spei_regional.md):
+- Escala **regional**:
+    - Datos de [SPEI Global Drought Monitor](http://sac.csic.es/spei/database.html) para Sierra Nevada (spatial resolution of 0.5º) Ver [esto](https://github.com/ajpelu/qpyr_resilience/blob/master/analysis/drought/explore_drought_spei_regional.md)
+
+---
+
+![picture](../analysis/drought/explore_drought_spei_regional_files/figure-markdown_github/unnamed-chunk-4-1.png)
+
+---
+
+<img src="./out/fig/drought/spei_sn_all_1970.pdf" alt="some text"  width="4200" height="4200">
+
+---
 

### 2017-11-08 analysis/drought/explore_drought_spei_regional.Rmd
set chunk name; change axis text size

@@
 ### Plot 
-```{r}
+```{r plotspei}
 label_variables <- c('spei1' = '1-month',
@@
 d
+```
 
+```{r plot_spei2000}
 # >2000
@@
 # Explore 3, 6 months scales
-```{r}
+```{r plot_spei1970}
 spei_mean_melt1970 <- spei_mean_melt %>% 
@@
   theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
+        legend.position = "none",
+        axis.text=element_text(size=7)) +
   scale_x_date(date_breaks = "2 year", date_labels = "%Y")

### 2017-11-08 analysis/drought/explore_drought_station.Rmd
add map

@@
     variant: markdown_github
+always_allow_html: yes
 ---
@@
 
-```{r, eval=FALSE}
+```{r}
 # Map of station

### 2017-11-08 analysis/drought/drought_dashboard.Rmd
add drought dashboard

---
title: "Exploring drought in southern Spain"
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    orientation: rows
---

```{r setup, include=FALSE}
library('flexdashboard')
```

```{r pacakges}
library('RCurl')
library('plyr')
library('dplyr')
library('lubridate')
library('SPEI')
library('ggplot2')
library('zoo')
library('plotly')
source("https://gist.githubusercontent.com/ajpelu/bb0d3617b5a026ab1de67bcf84fa577a/raw/4ca37e4d5b49275146ffb7fe9a58549a6778bf5b/hydrological_year.R%2520")
library("leaflet") 
```

```{r readData}
cadiar <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Cadiar.txt?token=AGHS0tWFx0TZR9fAsC6-6I53z0x6K5PHks5aC_Q8wA%3D%3D"), sep='', header=T)
padul <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Padul.txt?token=AGHS0oMFoEA4Iwvx2yhUGXHbkCpDmPlOks5aC_RbwA%3D%3D"), sep='', header=T)
jerez <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Jerez%20del%20Marquesado.txt?token=AGHS0omrawIULbdaFa1lHovFNDq_HAyXks5aC_RxwA%3D%3D"), sep='', header=T)
# ba <- read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_resilience/master/data_raw/meteo/base_aerea_aggregated.csv"), header=T, sep=',')

ltdf <- read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/long_term_data_rediam_data.txt?token=AGHS0isI6zUuuXN0rnR96VFc71bpahgOks5aC_SKwA%3D%3D"), header=T, sep=';', dec=',')
```

```{r}
# Compute the spei index 
compute_spei <- function(x, rango_mensual, inicio){
  # x = dataframe 
  # rango_mensual = vector with temporal range of spei

  # dataframe for output 
  outdf <- data.frame()

  for (i in rango_mensual){ 
    # Compute spei
    my_spei <- spei(x$prec_ac - x$et0_ac, i)

    # Create the ts 
    ts_aux <- ts(my_spei$fitted, start=inicio,frequency=12)

    # Convert to dataframe 
    tsdf <- data.frame(Y=as.matrix(ts_aux), fecha=as.Date(as.yearmon(time(ts_aux))))

    # Rename 
    names(tsdf) <- c('value', 'fecha')

    tsdf <- tsdf %>% 
       mutate(signo = ifelse(value >= 0, 'pos', 'neg'),
              spei = paste0('spei_', i))

    # Rbind 
    outdf <- rbind(outdf, tsdf)

  }

  outdf <- outdf %>% 
    mutate(spei_ord = plyr::mapvalues(spei,
                                      c('spei_1', 'spei_3', 'spei_6', 'spei_12', 'spei_24'),
                                      c('spei_01', 'spei_03', 'spei_06', 'spei_12', 'spei_24')))

} 
```

Short-Term Serie 
=======================================================================

Row {.tabset .tabset-fade}
-------------------------------

```{r}
# Prepare data 
# Prepare data 
ca <- cadiar %>% 
  filter(complete.cases(.)) %>% 
  mutate(date = as.Date(FECHA, format="%d-%m-%y"), 
         year = lubridate::year(date), 
         month = lubridate::month(date),
         hyear = hydro_year(date), 
         hyear_f = paste0(hyear-1,'-', hyear)) %>%
  filter(year < 2016) %>% 
  dplyr::select(date, year, month, doy = DIA, tmed = Gr07TMed, prec = Gr07Precip, et0 = Gr07ETo, hyear, hyear_f) %>% 
  # Group by year and month 
  group_by(year, month, hyear, hyear_f) %>% 
  dplyr::summarise(prec_ac = sum(prec),
                   et0_ac = sum(et0),
                   tmed_avg = mean(tmed))

je <- jerez %>% 
  filter(complete.cases(.)) %>% 
  mutate(date = as.Date(FECHA, format="%d-%m-%y"), 
         year = lubridate::year(date), 
         month = lubridate::month(date),
         hyear = hydro_year(date), 
         hyear_f = paste0(hyear-1,'-', hyear)) %>%
  filter(year < 2016) %>% 
  dplyr::select(date, year, month, doy = DIA, tmed = Gr06TMed, prec = Gr06Precip, et0 = Gr06ETo, hyear, hyear_f) %>% 
  # Group by year and month 
  group_by(year, month, hyear, hyear_f) %>% 
  dplyr::summarise(prec_ac = sum(prec),
            et0_ac = sum(et0),
            tmed_avg = mean(tmed))

pa <- padul %>% 
  filter(complete.cases(.)) %>% 
  mutate(date = as.Date(FECHA, format="%d-%m-%y"), 
         year = lubridate::year(date), 
         month = lubridate::month(date),
         hyear = hydro_year(date), 
         hyear_f = paste0(hyear-1,'-', hyear)) %>%
  filter(year < 2016) %>% 
  dplyr::select(date, year, month, doy = DIA, tmed = Gr10TMed, prec = Gr10Precip, et0 = Gr10ETo, hyear, hyear_f) %>% 
  # Group by year and month 
  group_by(year, month, hyear, hyear_f) %>% 
  dplyr::summarise(prec_ac = sum(prec),
            et0_ac = sum(et0),
            tmed_avg = mean(tmed)) 
```

### Cadiar

```{r}
# SPEI Cadiar
spei_ca <- compute_spei(ca, rango_mensual = c(1,3,6,12,24), inicio = c(2000,9))

df <- spei_ca %>%
  filter(!(spei %in% c('spei_1', 'spei_24')))
site <- 'Cadiar'

# PLOT 
ggplot(df, aes(x=fecha, y=value, fill=signo)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
  facet_wrap(~spei_ord, ncol=1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  theme_bw() + 
  ylab('SPEI') + 
  xlab('year') + 
  theme(legend.position = "none", 
        strip.background = element_blank()) +
  ggtitle(site)
```

### Padul

```{r}
# SPEI Padul
spei_pa <- compute_spei(pa, rango_mensual = c(1,3,6,12,24), inicio = c(2000,9))

df <- spei_pa %>% 
  filter(!(spei %in% c('spei_1', 'spei_24')))
site <- 'Padul'

# PLOT 
ggplot(df, aes(x=fecha, y=value, fill=signo)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
  facet_wrap(~spei_ord, ncol=1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  theme_bw() + 
  ylab('SPEI') + 
  xlab('year') + 
  theme(legend.position = "none", 
        strip.background = element_blank()) +
  ggtitle(site)
```

### Jerez Marquesado

```{r}

# SPEI Jerez 
spei_je <- compute_spei(je, rango_mensual = c(1,3,6,12,24), inicio = c(2000,9))

df <- spei_je %>% 
  filter(!(spei %in% c('spei_1', 'spei_24')))
site <- 'Jerez Marquesado'

# PLOT 
ggplot(df, aes(x=fecha, y=value, fill=signo)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
  facet_wrap(~spei_ord, ncol=1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") + 
  theme_bw() + 
  ylab('SPEI') + 
  xlab('year') + 
  theme(legend.position = "none", 
        strip.background = element_blank()) +
  ggtitle(site)
```

```{r}
mysites <- c('ca','je','pa')
all_sites <- data.frame()

for (i in mysites){

  df <- get(i)
  aux <- df %>% 
    dplyr::mutate(hmonth = ifelse(month <= 8, month + 4, month -8)) %>% 
    group_by(hyear_f, hmonth) %>% 
    dplyr::summarise(prec_ac = sum(prec_ac)) %>%
    mutate(csum = cumsum(prec_ac)) %>% 
    mutate(loc = i) %>% 
    as.data.frame()

  all_sites <- rbind(all_sites, aux)
}

all_sites <-  all_sites %>% 
  mutate(station= plyr::mapvalues(loc, c('ca', 'je', 'pa'), c("Cadiar", "Jerez", "Padul")))
```

### Precipitation (comparison)
```{r}
df_aux <- all_sites %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))

plot_pre_all_sites <- ggplot(df_aux, aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot() + facet_wrap(~station) + 
  geom_smooth(aes(group=1), se=TRUE, level=0.95) + 
  geom_line(data=subset(all_sites, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
  geom_line(data=subset(all_sites, hyear_f == '2011-2012'),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') + xlab('') +
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
  annotate("text", label = "2004-2005", x = 2, y = 500, size = 3, colour = "blue") + 
  annotate("text", label = "2011-2012", x = 2, y = 550, size = 3, colour = "red") + 
  annotate("text", label = "mean 2000-2015", x = 3, y = 600, size = 3, colour = "black")

ggplotly(plot_pre_all_sites)
```

### Precipitation (year)

```{r}
g <- ggplot(data=all_sites, aes(x=as.factor(hmonth), y=csum, group=hyear_f, colour=hyear_f)) + 
  geom_line() + facet_wrap(~station) + theme_bw() +
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') + xlab('') +
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug"))

ggplotly(g)
```

Long Term Serie 
=======================================================================

Row {.tabset .tabset-fade}
-------------------------------

```{r} 
ltdf <- ltdf %>% 
  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
         month = lubridate::month(fecha),
         hmonth = ifelse(month <= 8, month + 4, month -8), 
         hyear = hydro_year(fecha),
         hyear_f = paste0(hyear-1,'-', hyear))

```

### Precipitation acc. (1950-2015)

```{r}
# We used only the 5514 station (the most long term)
# Filter by station and prec
# Filter from 1950 
st5514 <- ltdf %>% 
  filter(VARIABLE == 'PM1') %>% 
  filter(cod == '5514') %>% 
  filter(hyear >= 1950) %>% 
  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)

st5514_pro <- st5514 %>% 
  group_by(hyear_f, hmonth) %>% 
  dplyr::summarise(pre = sum(pre)) %>%
  dplyr::mutate(csum = cumsum(pre))

cum_5514 <- st5514_pro  %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot() +  
  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') + xlab('') + 
  ggtitle('Granada Base Aerea, 1950-2015') + 
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
  annotate("text", label = "2004-2005", x = 2, y = 500, size = 3, colour = "blue") + 
  annotate("text", label = "2011-2012", x = 2, y = 550, size = 3, colour = "red") + 
  annotate("text", label = "mean 1950-2015", x = 3, y = 600, size = 3, colour = "black")

ggplotly(cum_5514)
```

### Precipitation acc. (2000-2015)

```{r}
# Cumulated precipitation from 2000 to 2015 for 5514 station 
st5514_2000  <- ltdf %>% 
  filter(VARIABLE == 'PM1') %>% 
  filter(cod == '5514') %>% 
  filter(hyear >= 1999) %>% 
  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR) %>% 
  group_by(hyear_f, hmonth) %>% 
  dplyr::summarise(pre = sum(pre)) %>%
  dplyr::mutate(csum = cumsum(pre))

cum_5514_2000 <- st5514_2000 %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot() +  
  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
  geom_line(data=subset(st5514_2000, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
  geom_line(data=subset(st5514_2000, hyear_f == '2011-2012'),
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red') +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black")) + 
  ylab('Precipitation (mm) acumulated') +  xlab('') + 
  ggtitle('Granada Base Aerea, 2000-2015') + 
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
  annotate("text", label = "2004-2005", x = 2, y = 500, size = 3, colour = "blue") + 
  annotate("text", label = "2011-2012", x = 2, y = 550, size = 3, colour = "red") + 
  annotate("text", label = "mean 2000-2015", x = 3, y = 600, size = 3, colour = "black")

ggplotly(cum_5514_2000)
```

### Interactive plots by year

```{r}
ba_ip <- ggplot(data=st5514_pro, 
       aes(x=as.factor(hmonth), y=csum, group=hyear_f, colour=as.factor(hyear_f))) +
  geom_line() + 
  theme_bw() + 
  ylab('Cumulative precipitation (mm)') + 
  xlab('') + ggtitle('Granada Base Aerea, 1950-2015') + 
  theme(legend.title = element_blank()) +
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug"))

ggplotly(ba_ip)
```

```{r, SPEI, eval=FALSE, echo=FALSE}

### TODO HACER SPEI datos Rediam 
spei_ba <- compute_spei(ba, rango_mensual = c(1,3,6,12,24), inicio = c(1951,1))

df <- spei_ba %>%
  filter(!(spei %in% c('spei_1', 'spei_3')))
site <- 'Meteorological station Granada (Base Aerea)'

ggplot(df, aes(x=fecha, y=value, fill=signo)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) + 
  facet_wrap(~spei_ord, ncol=1) +
  scale_x_date(date_breaks = "10 year", date_labels = "%Y") + 
  theme_bw() + 
  ylab('Standardised Precipitation-Evapotranspiration Index') + 
  xlab('year') + 
  theme(legend.position = "none", 
        strip.background = element_blank(),
        plot.title = element_text(size = 8)) +
  ggtitle(site)  
```

Map 
=======================================================================

```{r}
geoinfo_station <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_resilience/master/data_raw/meteo/metadata_station_ria.txt"), sep=';', header=T)

m <- leaflet(geoinfo_station) %>%  
    addProviderTiles("Esri.WorldImagery", 
                   group='Satellite') %>% 
  addMarkers(~long, ~lat, popup = ~as.character(name))

m
```

### 2017-11-08 analysis/drought/explore_drought_station.Rmd
add drought station

--- 2017-11-07 analysis/drought/explore_drought_station.Rmd
@@
     variant: markdown_github
+
 ---
@@
 
-```{r, eval=FALSE}
+```{r eval=FALSE}
 # Map of station
@@
 
-```{r}
+```{r speiCadiar}
 # SPEI Cadiar
@@
 
-```{r}
+```{r speiPadul}
 # SPEI Padul
@@
 
-```{r}
+```{r speiJerez}
 # SPEI Jerez 
@@
 
-```{r}
+```{r prec_curvesRIA}
 
@@
 
-```{r}
+```{r prec_curvesBaseAerea}
 # We used only the 5514 station (the most long term)
@@
   scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug"))
+                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
+  annotate("text", label = "2004-2005", x = 2, y = 500, size = 6, colour = "blue") + 
+  annotate("text", label = "2011-2012", x = 2, y = 550, size = 6, colour = "red") + 
+  annotate("text", label = "mean 1950-2015", x = 3, y = 600, size = 6, colour = "black")
 ```
@@
 
-```{r}
+```{r prec_curvesBaseAerea2000}
 # Cumulated precipitation from 2000 to 2015 for 5514 station 
@@
   scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug"))
+                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
+  annotate("text", label = "2004-2005", x = 2, y = 500, size = 6, colour = "blue") + 
+  annotate("text", label = "2011-2012", x = 2, y = 550, size = 6, colour = "red") + 
+  annotate("text", label = "mean 2000-2015", x = 3, y = 600, size = 6, colour = "black")
 
@@
   ylab('Precipitation (mm) acumulated') + 
-  xlab('months (hydrological year; 1= sep)') 
+  xlab('months (hydrological year; 1= sep)') +
+  annotate("text", label = "2004-2005", x = 2, y = 500, size = 6, colour = "blue") + 
+  annotate("text", label = "2011-2012", x = 2, y = 550, size = 6, colour = "red") + 
+  annotate("text", label = "mean 1950-2009", x = 3, y = 600, size = 6, colour = "black")
 ```
@@
   geom_vline(xintercept=as.numeric(as.Date("2005-01-01"))) +
-  geom_vline(xintercept=as.numeric(as.Date("2005-12-31")))
+  geom_vline(xintercept=as.numeric(as.Date("2005-12-31"))) +
+  scale_x_date(date_breaks = "2 year", date_labels = "%Y")
 

### 2017-11-08 analysis/drought/drought_dashboard.Rmd
remove token

@@
 ```{r readData}
-cadiar <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Cadiar.txt?token=AGHS0tWFx0TZR9fAsC6-6I53z0x6K5PHks5aC_Q8wA%3D%3D"), sep='', header=T)
-padul <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Padul.txt?token=AGHS0oMFoEA4Iwvx2yhUGXHbkCpDmPlOks5aC_RbwA%3D%3D"), sep='', header=T)
-jerez <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Jerez%20del%20Marquesado.txt?token=AGHS0omrawIULbdaFa1lHovFNDq_HAyXks5aC_RxwA%3D%3D"), sep='', header=T)
+cadiar <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Cadiar.txt"), sep='', header=T)
+padul <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Padul.txt"), sep='', header=T)
+jerez <-read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/Jerez%20del%20Marquesado.txt"), sep='', header=T)
 # ba <- read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_resilience/master/data_raw/meteo/base_aerea_aggregated.csv"), header=T, sep=',')
 
-ltdf <- read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/long_term_data_rediam_data.txt?token=AGHS0isI6zUuuXN0rnR96VFc71bpahgOks5aC_SKwA%3D%3D"), header=T, sep=';', dec=',')
+ltdf <- read.csv(text=getURL("https://raw.githubusercontent.com/ajpelu/qpyr_dendro/master/data_raw/meteo/long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
 ```

### 2017-11-08 man/notes_drought.Rmd
update drought notes

---
title: "notes Drought"
author: "AJ Perez-Luque (@ajpelu)"
date: "2016 Aug"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../references.bib
csl: ../ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Drought Characterization

* Three most severe drought events (**1950–2012**) [@Trigo2013, @Gouveia2015]:
    * **1998-1999**
    * **2004-2005**
    * **2011-2012**

* The two drought episodes of 2004/2005 and 2011/2012 were extreme in both its magnitude and spatial extent [@Gouveia2014]

<center>![](/out/images/trigo_2012.jpg)</center>

<br>

Figure. Accumulated monthly precipitation averaged over the IP during the hydrological year 2011–12 (black line) and during the other two most severe drought events in the period 1950–2012 (colored lines). Light, dark, and medium gray shaded areas show the ensemble spread from the corresponding evolutions of all hydrological years between 2002–03 and 2011–12, 1950–51 and 1959–60, and the overlapping between them, respectively. The gray thick line indicates the 1950 –2010 mean evolution, with boxes and whiskers representing the ±0.5 sigma level and the 10th–90th percentile interval, respectively.

## Summer of 2005

* 2004/2005 hydrological year is considered one of the worst drooughts ever recorded in the Iberian Peninsula, particularly in the central and southern sectors [@GarciaHerrera2007].

* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].

* The hydrological year from October 2004 to September 2005 was the driest on record at several locations throughout Iberia [@GarciaHerrera2007].

## Summer of 2012

<center>![](/man/images/trigo_2012.jpg)</center>

## Long-term perspective

* @GarciaHerrera2007 analyzed three longest dialy series available (1941-2005). 
The following figure shows the daily accumulated precipitation anomaly compared to the 1961–90 normal for Granada station. The colored lines display the six greater droughts ever recorded in each station (1941–2005 period) during the corresponding October (year - 1) to June (year) hydrological year.

<center>![](/man/images/garcia_herrera_granada_zoom.jpg)</center>

* From a long-term perspective, the decades of 40s and the early 50s along with the 80s and 90s are described as particularly dry throughout the whole Iberian territory (Vicente-Serrano 2006a,b)

* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].

# Drought characterization

Althought drought events are characteristic of the Mediterranean area (`TODO`: `CITA`), an increase in the drought severity in the Iberian Peninsula has been observed in the last decades [@VicenteSerrano2014]. The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southeren sector [@GarciaHerrera2007 @Gouveia2015 @Trigo2013]. These events were extreme in both its magnitude and spatial extent [@Gouveia2014].

Althought several works have reported these two years as some of the worst drought events, we characterised the drought at several spatio-temporal scales in the study area. From a long-term perspective, we compare the accumulated monthly precipitation at a meteorological station (Granada, Base Aérea) during the hydrological years 2004-2005 and 2011-2012 with the average of accumulated monthly precipitation for the period 1950-2015.

Figure. Accumulated monthly precipitation averaged over the IP during the hydrological year 2011–12 (black line) and during the other two most severe drought events in the period 1950–2012 (colored lines). Light, dark, and medium gray shaded areas show the ensemble spread from the corresponding evolutions of all hydrological years between 2002–03 and 2011–12, 1950–51 and 1959–60, and the overlapping between them, respectively. The gray thick line indicates the 1950 –2010 mean evolution, with boxes and whiskers representing the ±0.5 sigma level and the 10th–90th percentile interval, respectively.

# References

### 2017-11-08 man/ms_0_intro.Rmd
add more ideas to intro

@@
 
-Ideas a expresar
-
-### Remote sensing
-
-* Uso de Remote sensing para estudiar la sequía --> Leer @Zhang2013 y escribir algo.
-Drought monitoring using remote-sensing approach was originally applied to agriculture.
-Several remote-sensing derived indices have been used to study the drought effects on vegetation ...
-* Leer también a @AghaKouchak2015
-
 ### Sequías
@@
 Although summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
+Además las proyecciones apuntan a que este incremento en la frecuencia continuará [@Spinoni2017a].
 
@@
 These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
-
-### Ideas a incluir
-
-* :red_circle: Algo de las sequías de 2005 y 2014
-
-* Extreme sequías y ecología
 
@@
 
-frecuencia de las sequías severas (ver Dai 2011)
-
 Althought several works have reported these two years as some of the worst drought events, we characterised the drought at several spatio-temporal scales in the study area.
@@
 
-* Extreme sequias y ecologia 
+### Extreme sequias y ecologia
+
+Extreme climate events (e.g.
+droughts) severely affect forests and grasslands throuhg changes in plant physiology, phenology and carbon allocation [@Ummenhofer2017] (incluir citas para los efectos de la sequía sobre la vegetación).
+
+No solo efectos aislados, sino también en conjunción con otros factores, sobre todo en el mediterráneo donde convergen muchos de los factores que pueden interaccionar:
+
+* La sequía es un factor crucial a tener en cuenta, ya que es además de los efectos que puede tener de forma aislada, se ha visto que además presenta muchas interacciones con otros factores, siendo por tanto un factor crucial (`$IMPROVE$` ver [@DoblasMiranda2017]):
+* Some interactions alter the effects of a single factor, as drought enhances or decreases the effects of atmospheric components on plant ecophysiology
+* Drought and land use changes, among others, alter water resources and lead to land degradation, vegetation regeneration decline, and expansion of forest diseases.
+* Climate change, and especially drought, emerges as a crucial factor in most of the reviewed interactions and therefore it should be considered when it comes to designing and applying international management policies
+* Drought should be considered when designing and applying management policies.
+
+Además la sequía se espera que tengan ...Droughts are most likely to have the largest and most long-lasting impacts globally due to large indirect and lagged impacts and long recovery especially for forest ecosystems (ver  18 en @Ummenhofer2017).
+Así por ejemplo se ha visto que todos los biomas presentan una vulnerabilidad similar ...
+En una revision sobre la vulnerabilidad del sistema de trasnporte en plantas al embolismo inducido por sequía, ha mostrado que una convergencia de la vulnerabilidad de los bosques a la sequía, mostrando que todos los biomas son igualmente vulnerables a los fallos hidráulicos independientemente del régimen de precipitación (Choat et al.
+2012, doi:10.1038/nature11688)
+
+-- Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
+(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar)
+
+Una vez dicho esto...
+
 * Como la dendro ayuda a evaluar las sequías (ver Gazol 2017) 
+* Remote sensing:
 
-Extreme climate events (as droughts, heatwaves) severely affect forests and grasslands throuhg changes in plant physiology, phenology and carbon allocation [@Ummenhofer2017].
-
-Droughts are most likely to have the largest and most long-lasting impacts globally due to large indirect and lagged impacts and long recovery especially for forest ecosystems (ver  18 en @Ummenhofer2017)
+    * Uso de Remote sensing para estudiar la sequía --> Leer @Zhang2013 y escribir algo.
+Drought monitoring using remote-sensing approach was originally applied to agriculture.
+Several remote-sensing derived indices have been used to study the drought effects on vegetation ...
+    * Leer también a @AghaKouchak2015
 

### 2017-11-08 man/ms_0_intro.Rmd
update pres and intro

@@
 
+* ojo hablar de que la sequía es un proceso natural, pero que se está viendo aumentado en los últimos años:
+
+Drought is a natural phenomenon that occurs when water availability is significantly below normal levels over a long period and the supply cannot meet the existing demand.
+
+> ver intro Ecosphere:
+> Drought affects ecological systems across every climatic zone worldwide and is exacerbated by climate change and increasing anthropogenic water demands (Mishra and Singh 2010).
+Characterized by below-normal precipitation (Dai 2011), meteorological drought results from complex interactions between the atmosphere and hydrologic processes within the biosphere.
+Unlike aridity, which is a permanent feature of climate (Wilhite 1992), drought is a temporary extreme event (Palmer 1965, Mishra and Singh 2010) that can persist for extended time periods (months to years; Mishra and Singh 2010).
+Drought can cause significant changes in ecosystem productivity and water dynamics, and it is one of the most economically and ecologically disruptive extreme events affecting millions of people globally (Dai 2011).
+
 Extreme climate events (e.g.
@@
 
+Climate change projections indicate that extreme events will become more common in the future (IPCC 2013), making it important that we understand how ecosystems respond to these events and the potential feedbacks to radiative forcing.
+
 ### 

### 2017-11-08 pres/state_ms.Rmd
update pres and intro

--- 2017-11-07 pres/state_ms.Rmd
@@
 
-- Item 1 
-    - sub
-
 ---
@@
 
-- SPEI (Standardised Precipitation-Evapotranspiration Index) para datos
+- SPEI (Standardised Precipitation-Evapotranspiration Index) para datos regionales y locales
+- Precipitación acumulada del año hidrológico en curso 
+- Long term (>1950) y short-term (>2000)
 
@@
 
-- Escala regional: exploramos la sequía a escala regional utilizando datos del [SPEI Global Drought Monitor](http://sac.csic.es/spei/database.html) para Sierra Nevada (spatial resolution of 0.5º) Ver [esto](https://github.com/ajpelu/qpyr_resilience/blob/master/analysis/drought/explore_drought_spei_regional.md):
+- Escala **regional**:
+    - Datos de [\textcolor{blue}{SPEI Global Drought Monitor}](http://sac.csic.es/spei/database.html) para Sierra Nevada (spatial resolution of 0.5º) Ver [\textcolor{blue}{esto}](https://github.com/ajpelu/qpyr_resilience/blob/master/analysis/drought/explore_drought_spei_regional.md)
 
-## Including Plots
+- Escala **local**:
+    - Calcular datos SPEI para tres estaciones de la [\textcolor{blue}{red RIA}](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController) durante el periodo 2000 - 2016: [\textcolor{blue}{Cadiar}](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=7), [\textcolor{blue}{Padul}](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=10) y [\textcolor{blue}{Jerez del Marquesado}](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=6)
+    - Curvas de precipitación acumulada RIA y datos base Aérea (obtenidos de la REDIAM) (desde 1950) 
+    - Ver [\textcolor{blue}{flexdashboard drought}](https://rawgit.com/ajpelu/qpyr_dendro/master/analysis/drought/drought_dashboard.html)
 
-You can also embed plots, for example:
+---
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+![](../analysis/drought/explore_drought_spei_regional_files/figure-markdown_github/plot_spei1970-1.png)
+
+---
+
+- Para la caracterización de los dos periodos de sequía ¿Incluir alguna referencia a estos análisis (quizá como *supplementary materials*) o solamente referencias bibliofrágicas?
+
+---
+
+Aquí deberíamos incluir los objetivos del paper y las preguntas
+
+---
+
+# References
+
+---

### 2017-11-13 man/notes_EVI.Rmd
add notes EVI; add info from get_modis_raw_data.md qpyr_modis_resilience

---
title: "notes EVI"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 nov"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../references.bib
csl: ../ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Notas sobre EVI

## MODIS Vegetation Indices

***Aim***: To obtain vegetation indices for Quercus pyrenaica forests in Sierra Nevada from 2000 to 2016.

Technical info: 
* MODIS IV:

 * Collection `MODIS 006`
 * Temporal range: 2000 - 2016 
 * Pixel approach

* Distribution of Pyrenan oak forest in Sierra Nevada:

 * Derived from Ecosystems map of Sierra Nevada 
 * Datum: `epsg:4326`
 * See this [link](https://rawgit.com/ajpelu/qpyr_distribution/master/analysis/distribution_map_sn.html)

* Pixels covering Pyrenan oak forest in Sierra Nevada: `./github/ajpelu/qpyr_distribution/data_raw/geoinfo/iv_malla_modis_qp_centroid.shp` 
 * *n* pixels = 927 pixels 
 * See this [link](https://rawgit.com/ajpelu/qpyr_distribution/master/analysis/distribution_map_sn.html)

The script to download MODIS data is at: `./script/GEE/get_iv_modis_qp.js`

# References

### 2017-11-13 analysis/prepare_modis_qa.Rmd
add quality asessment qa modis

---
title: "Quality Assessment of MODIS IV data"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 March"
output:  
  md_document:
    toc: true
    variant: markdown_github
---

# Filter data 
```{r wd, echo=FALSE}
#---------------------------------
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
#---------------------------------
```

```{r packages, warning=FALSE, message=FALSE}
library("tidyverse")
library("knitr")
library("binaryLogic")
library("corrplot")
source(paste0(di,"/script/R/getComposite.R"))
```

```{r}
# Read data
rawdata <- read.csv(file=paste(di, "/data_raw/evi/iv_qp_raw_qa_2017_aug.csv", sep= ""), header = TRUE, sep = ',')
```

## Prepare Raw Data
* Raw data come from GEE script (see [`/script/GEE/get_iv_modis.qp.js`](/script/GEE/get_iv_modis_qp.js)

* Date:
    + Get date of the image (from hdf title, system.index) 
    + Store as date (date format) and create new variable for year

* Select and rename variables of interest

```{r}
raw <- rawdata %>% 
  mutate(
    # Date
    date = as.Date(substr(system.index,1,10), format = "%Y_%m_%d"),
    # date2 = as.Date(substr(system.index,1,10), format = "%Y_%m_%d"),
    year = lubridate::year(date),

    # GEE index
    gee_index = stringr::str_replace(
      substr(system.index,22, nchar(as.character(system.index))),
    pattern = '_', "")) %>% 
    # test = ifelse(date == date2, 0, 1))
  dplyr::select(doy = DayOfYear, evi = EVI, ndvi = NDVI, summQA = SummaryQA, iv_malla_modi_id,
                pop, date, year, gee_index, lat, long, 
                qa_quality, qa_use, qa_aerosol, qa_adj_cloud, 
                qa_atmos, qa_mix_cloud, qa_landwater, qa_snow, qa_shadow)
```

## Some metadata of the time series 
* Temporal range of the time series

```{r}
# Get temporal range of the data
# Start date
unique(min(as.Date(raw$date)))

# End date 
unique(max(as.Date(raw$date)))
```

* Number of images per year 
```{r}
# See n of images per year and per pixel 
n_images_pixel <- raw %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>%
  summarise(n = n(),
            n_pixel = n()/length(unique(iv_malla_modi_id)))

kable(n_images_pixel)
```

* Summary table of layer summaryQA

```{r}
# Table of QA
raw %>% group_by(summQA) %>% 
  summarise(npixels = n()) %>% 
  mutate(freq = round((npixels / sum(npixels)*100),2),
         QA = plyr::mapvalues(summQA, c(0, 1, 2, 3), c("Good Data", "Marginal data", "Snow/Ice", "Cloudy"))) %>% 
  kable() 
```

### QA 
* Explore temporal distribution of QA

```{r, message=FALSE, warning=FALSE}
qadate <- raw %>% 
  group_by(date, summQA) %>% 
  count(date, summQA) %>% 
  mutate(freq = round((n / sum(n))*100, 2),
         QA = plyr::mapvalues(summQA, c(0, 1, 2, 3), 
                              c("Good Data", "Marginal data", "Snow/Ice", "Cloudy"))) 
qadate %>% 
  ggplot(aes(x=as.Date(date), y=freq, fill=QA)) + 
  geom_bar(stat = 'identity', position='fill') + 
#  facet_wrap(~QA, ncol=1) + 
  theme_bw() +
  scale_x_date(date_labels = "%Y %d")
```

## QA detailed 
* Only explore the marginal data

```{r} 
decodeQA <- function(x, nb){
  bit <- intToBits(x)
  paste(tail(rev(as.integer(bit)), nb), collapse="")}

decodeQAv <- Vectorize(decodeQA)

marginal <- raw %>% filter(summQA == 1) 
```

### Quality 
```{r}
marginal %>% 
  mutate(qa_quality_dec = decodeQAv(qa_quality, nb=2)) %>% 
  group_by(qa_quality_dec) %>% 
  summarise(n = n()) %>% 
  mutate(freq = round(n / sum (n)*100,2)) %>% 
  kable()
```

### Quality usefulness
* All data
```{r}
# See https://lpdaac.usgs.gov/sites/default/files/public/product_documentation/mod13_user_guide.pdf 
raw %>% 
  mutate(qa_use_dec = decodeQAv(qa_use, nb=4)) %>% 
  group_by(qa_use_dec) %>% 
  summarise(n = n()) %>% 
  mutate(freq = round(n / sum (n)*100,2)) %>% 
  kable()
```

* Marginal data 
```{r}
marginal %>% 
  mutate(qa_use_dec = decodeQAv(qa_use, nb=4)) %>% 
  group_by(qa_use_dec) %>% 
  summarise(n = n()) %>% 
  mutate(freq = round(n / sum (n)*100,2)) %>% 
  kable() 
```

### Aerosol 
```{r}
marginal %>% 
  mutate(qa_aerosol_dec = decodeQAv(qa_aerosol, nb=2)) %>% 
  group_by(qa_aerosol_dec) %>% 
  summarise(n = n()) %>% 
  mutate(freq = round(n / sum (n)*100,2)) %>% 
  kable()
```

* Only the 6.96 % of the pixels marked as 'Marginal data' contain high concentration of Aerosols.

### Clouds 
```{r}
# Adjacents
marginal %>% 
  mutate(qa_adj_cloud_dec = decodeQAv(qa_adj_cloud, nb=1)) %>% 
  group_by(qa_adj_cloud_dec) %>% 
  summarise(n = n()) %>% 
  mutate(freq = round(n / sum (n)*100,2)) %>% 
  kable()

# Mixed
marginal %>% 
  mutate(qa_mix_cloud_dec = decodeQAv(qa_mix_cloud, nb=1)) %>% 
  group_by(qa_mix_cloud_dec) %>% 
  summarise(n = n()) %>% 
  mutate(freq = round(n / sum (n)*100,2)) %>% 
  kable()
```

* All the pixel marked as 'Marginal data' do not contain mixed cloud.
* We have to consider the Adjacent clouds

### Snow 
```{r}
marginal %>% 
  mutate(qa_snow_dec = decodeQAv(qa_snow, nb=1)) %>% 
  group_by(qa_snow_dec) %>% 
  summarise(n = n()) %>% 
  mutate(freq = round(n / sum (n)*100,2)) %>% 
  kable()
```

* All the pixel marked as 'Marginal data' do not contain snow or ice.

### Shadow
```{r}
marginal %>% 
  mutate(qa_shadow_dec = decodeQAv(qa_shadow, nb=1)) %>% 
  group_by(qa_shadow_dec) %>% 
  summarise(n = n()) %>% 
  mutate(freq = round(n / sum (n)*100,2)) %>% 
  kable()
```

* We have to consider the shadow

### Filter 
We applied customized filter based on Reyes-Díez et al. 2015.

* Creamos una variable nueva, llamada `filtered`. Aquellos pixeles con *Good Data* le asignamos valor `90`. Los *Snow/Ice* y los *Cloudy* le asignamos valor `99`. Para los *Marginal Data* miramos la calidad. En concreto previamente hemos analizado la cantidad de aerosoles, la de nubes y la de sombras (ver mas arriba). La forma de proceder es la siguiente: 
    + Si tiene alta concentración de aerosoles, marcamos el pixel con 1. 
    + Si tiene nubes adjacentes, marcamos el pixel con 1.
    + Si tiene sombra, marcamos el pixel con 1. 
    + Al final, creamos una variable llamada `f_sum` con la suma de esos tres filtros. Si en alguno de los tres filtros se da la condición 1, es decir `f_sum > = 1`, entonces, le asignamos el valor `99` en la variable filtered.

```{r}
rawfilter <- raw %>% 
  mutate(f_aerosol = ifelse(decodeQAv(qa_aerosol, nb=2) == '11', 'Aerosol (high)', 'OK'),
         f_cloud_a = ifelse(decodeQAv(qa_adj_cloud, nb=1) == '1', 'Clouds: adjacent', 'OK'),
         f_shadow = ifelse(decodeQAv(qa_shadow, nb=1) == '1', 'Shadow','OK'),
         # as integer
         f_aerosoli = ifelse(f_aerosol == 'OK', 0, 1),
         f_cloud_ai = ifelse(f_cloud_a == 'OK', 0, 2),
         f_shadowi = ifelse(f_shadow == 'OK', 0, 4))

rawfilter <- rawfilter %>% 
  mutate(f_sum = f_aerosoli + f_cloud_ai + f_shadowi,
         filtered = ifelse(summQA == 0, 90, 
                           ifelse(summQA %in% c(2,3), 99, f_sum)),

         filtered_qa = plyr::mapvalues(filtered, 
                                       c(0, 1, 2, 3, 4, 5, 6, 7, 90, 99, NA),
                                       c("Others", "Aerosol",
                                         "Clouds adj", "Aerosol + Clouds adj",
                                         "Shadow", "Aerosol + Shadow", 
                                         "Clouds adj + Shadow", 
                                         "Aerosol + Clouds adj + Shadow",
                                         "Good Data", "Snow / Ice or Cloudy", "NA")),
         mes = lubridate::month(date))

rawfilter %>% group_by(filtered_qa) %>% 
  summarise(npixels = n()) %>% 
  mutate(freq = round((npixels / sum(npixels)*100),2)) %>% 
  kable() 
```

* According to Reyes-Díez et al. (2015) we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds. On the other hand, the use of EVI mean is highly stable under the use of any filter (see Reyes-Díez et al. 2015).

```{r}
rawfilter <- rawfilter %>% 
  mutate(filtered_2 = ifelse(filtered %in% c(0,1,2,3), 90, filtered),
         filtered_2_qa = plyr::mapvalues(filtered_2, 
                                       c(4, 5, 6, 7, 90, 99, NA),
                                       c("Shadow", "Aerosol + Shadow", 
                                         "Clouds adj + Shadow", 
                                         "Aerosol + Clouds adj + Shadow",
                                         "Good Data", "Snow / Ice or Cloudy", "NA")))

rawfilter %>% group_by(filtered_2_qa) %>% 
  summarise(npixels = n()) %>% 
  mutate(freq = round((npixels / sum(npixels)*100),2)) %>% 
  kable()

```

We have to explore the temporal distribution of the filtered data (especially for Shadow)

```{r shadow_temporal_evolution}
sombras <- c("Shadow", "Aerosol + Shadow","Clouds adj + Shadow", "Aerosol + Clouds adj + Shadow") 
sombras_n <- c(4, 5, 6, 7)

# Shadow 
rawfilter %>% filter(filtered_2 %in% sombras_n) %>% group_by(mes) %>% 
  summarise(n=n()) %>% 
  mutate(freq = round((n  / sum(n )*100),2)) %>% 
  ggplot(aes(x=as.factor(mes), y=n)) + 
  geom_bar(stat='identity') +
  geom_text(aes(label=freq, size=3, vjust=-.25)) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  xlab('month') + 
  ylab('# Images with shadow')
```

Vemos que la mayoría de las imagenes con sombra están en los meses de invierno: mes 1, 2, 3, y 12 (69.32 %), esto es: el 69.32 % de las imágenes que están marcadas con sombra en QA Detailed, se distribuyen temporalmente entre los meses 1,2,3 y 12, que coincide con el periodo donde el Quercus pyrenaica no presenta actividad fotosintética (incluir cita de estación de crecimiento, y de cuando tira la hoja), por lo tanto los valores con sombra los debemos eliminar a la hora de calcular la media de EVI, ya que se encuentran la mayoría en los meses de periodo en los que el roble no tiene actividad fotosintética.

Now for Ice / Cloudy 
```{r snow_temporal_evolution}
# "Snow / Ice or Cloudy" (99)

# Shadow 
rawfilter %>% filter(filtered_2 == 99) %>% group_by(mes) %>% 
  summarise(n=n()) %>% 
  mutate(freq = round((n  / sum(n )*100),2)) %>% 
  ggplot(aes(x=as.factor(mes), y=n)) + 
  geom_bar(stat='identity') +
  geom_text(aes(label=freq, size=3, vjust=-.25)) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  xlab('month') + 
  ylab('# Images with Snow / Ice or Cloudy')
```

We select only data with high Good Quality, because shadow, ice or cloudy are mainly locate at winter months.

```{r good_temporal_evolution}
df <- rawfilter %>% 
  filter(filtered_2 == 90) %>% 
  select(doy, evi, ndvi, iv_malla_modi_id, pop, date, year, lat, long, summQA, qa_use, filtered, filtered_2, mes)

df %>% group_by(mes) %>% 
  summarise(n=n()) %>% 
  mutate(freq = round((n/sum(n )*100),2)) %>% 
  ggplot(aes(x=as.factor(mes), y=n)) + 
  geom_bar(stat='identity') +
  geom_text(aes(label=freq, size=3, vjust=-.25)) + 
  theme_bw() + 
  theme(legend.position = 'none') + 
  xlab('month') + 
  ylab('# Images with Good Quality')
```

```{r n_composites_year}
ncomposites_year <- df %>% group_by(iv_malla_modi_id, year) %>% summarise(n=n())

ncomposites_year %>% ggplot(aes(x=n)) + 
  geom_histogram(stat='count') +
  xlab('# composites by year and pixel') + 
  ylab('# ')+
  theme_bw()
```

En este último gráfico, exploramos el número de pixeles que tienen todos las imágenes de un año completo. Aunque no se si tiene mucho sentido este gráfico.

We do not apply a filter treshold, because we selected only the Good Data.

# Prepare data of processed IV

### Get the composite of the images and the season
* See [Testa et al. 2014](https://www.researchgate.net/publication/262566793_Correcting_MODIS_16-day_composite_NDVI_time-series_with_actual_acquisition_dates)
* Use a [custom function](/script/R/getComposite.R)

```{r}
# Get leap years 
years <- unique(df$year)
ly <- years[lubridate::leap_year(years)]

# Two functions 
rd <- df %>% 
  mutate(m= lubridate::month(date),
         d= lubridate::day(date)) %>% 
  mutate(composite = ifelse(year %in% ly, 
                            getComposite_leap(m,d),
                            getComposite_nonleap(m,d))) %>% 
  mutate(season = ifelse(composite < 6, 'winter',
                   ifelse(composite < 12, 'spring',
                    ifelse(composite < 18, 'summer','autumn'))))
```

### Scale factor of the NDVI and EVI data 
```{r}
# Apply scale factor https://lpdaac.usgs.gov/dataset_discovery/modis/modis_products_table/mod13q1 
rd <- rd %>% 
  mutate(evi = evi * 0.0001,
         ndvi = ndvi * 0.0001) 
```

We created several datasets:

### ***iv_composite*** 
iv by pixel and by composite (output as `./data/evi/iv_composite.csv`). It contains the following fields:

* `iv_malla_modi_id`: the identifier of the modis cell
* `year`
* `evi` or `ndvi`: the value of the EVI (or NDVI) for the composite 
* `date`: date of adquisition of the image
* `composite`: number of composite (23 by year)
* `long`: longitude coordinates
* `lat`: latitute coordinates 
* `pop`: numeric code of the *Q. pyrenaica* population
* `season`: 
    + `0` annual value
    + `1` spring value
    + `2` summer value 
    + `3` autumn value 
    + `4` winter value 
* `seasonF`: the season coded as factor 
* `summQA`: the summary Quality Assessment (see MODIS)
* `qa_use`: the usseful index (see QA MODIS)

### ***evi_mean***
evi mean by pixel and year (output as `./data/evi/evi_mean.csv`). It contains the following fields:

* `iv_malla_modi_id`: the identifier of the modis cell
* `year`
* `evi`: mean value of evi for pixel *i* at year *j*
* `long`: longitude coordinates
* `lat`: latitute coordinates 
* `pop`: numeric code of the *Q. pyrenaica* population
* `n_composites`: number of composites used to computed the annual mean value of EVI (according to previously filtering)

### ***evi_seasonal*** and ***ndvi_seasonal***
* annual and seasonal evi by pixel (output as `./data/evi/evi_atributes_all.csv`)
* annual and seasonal ndvi by pixel (output as `./data/evi/ndvi_atributes_all.csv`)

These two dataframes have the following fields:

* `iv_malla_modi_id`: the identifier of the modis cell
* `year`
* `evi` or `ndvi`: the value of the EVI (or NDVI) (cumulative value for each season)
* `season`: the season of cumulative evi:
    + `0` annual value
    + `1` spring value
    + `2` summer value 
    + `3` autumn value 
    + `4` winter value 
* `seasonF`: the season coded as factor 
* `long`: longitude coordinates
* `lat`: latitute coordinates 
* `pop`: numeric code of the *Q. pyrenaica* population

#### Create dataframe with composites
```{r}
iv_composite <- rd %>% 
  dplyr::select(iv_malla_modi_id, evi, ndvi, pop, date, year, long, lat, composite, seasonF = season, summQA, qa_use)
```

#### Create EVI mean dataset
```{r}
evimean <- rd %>% 
  group_by(iv_malla_modi_id, year) %>%
  summarise(evi = mean(evi[evi >=0], na.rm=T),
            n_composites = length(year))

# Add coordinates and pob and other useful info 
aux_rd <- rd %>% dplyr::select(iv_malla_modi_id, long, lat, pop) %>%
  group_by(iv_malla_modi_id) %>% unique()

# Join dataframes 
evimean <- evimean %>% dplyr::inner_join(aux_rd, by="iv_malla_modi_id") 
```

####  Create seasonal dataframes of EVI and NDVI (integrate, EVI and NDVI sum)
* EVI
```{r}
# Create annual evi by pixel 
evi_annual <- rd %>% 
  group_by(iv_malla_modi_id, year) %>%
  summarise(evi = sum(evi[evi >=0],  na.rm=T)) %>%
  mutate(seasonF='annual', 
         season = 0)

# Create seasonal evi by pixel 
evi_season <- rd %>% 
  group_by(iv_malla_modi_id, year, season) %>%
  summarise(evi = sum(evi[evi >=0], na.rm=T)) %>% 
  mutate(seasonF = season) %>% 
  mutate(season = ifelse(season == 'autumn', 3, 
                   ifelse(season == 'winter', 4, 
                    ifelse(season == 'spring', 1,2))))

evidf <- rbind(evi_annual, evi_season)

# Add coordinates and pob 
aux_rd <- rd %>% dplyr::select(iv_malla_modi_id, long, lat, pop) %>%
  group_by(iv_malla_modi_id) %>% unique()

# Join dataframes 
evidf <- evidf %>% dplyr::inner_join(aux_rd, by="iv_malla_modi_id") 
```

## Correlaciones entre el EVI medio y los EVI estacionales para los robledales

```{r correlation_evi}
co <- evidf %>% dplyr::select(iv_malla_modi_id, year, evi, seasonF) %>% 
  spread(seasonF, evi) %>% as.data.frame()

co2 <- evimean %>% 
  dplyr::select(iv_malla_modi_id, year, evimean = evi) %>% 
  inner_join(co, by =c("iv_malla_modi_id","year")) %>% as.data.frame()

co3 <- co2 %>% dplyr::select(-iv_malla_modi_id, -year)

# Ver https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html 
cor.mtest <- function(mat, conf.level = 0.95){
  mat <- as.matrix(mat)
    n <- ncol(mat)
    p.mat <- lowCI.mat <- uppCI.mat <- matrix(NA, n, n)
    diag(p.mat) <- 0
    diag(lowCI.mat) <- diag(uppCI.mat) <- 1
    for(i in 1:(n-1)){
        for(j in (i+1):n){
            tmp <- cor.test(mat[,i], mat[,j], conf.level = conf.level)
            p.mat[i,j] <- p.mat[j,i] <- tmp$p.value
            lowCI.mat[i,j] <- lowCI.mat[j,i] <- tmp$conf.int[1]
            uppCI.mat[i,j] <- uppCI.mat[j,i] <- tmp$conf.int[2]
        }
    }
    return(list(p.mat, lowCI.mat, uppCI.mat))
}

m <- cor(co3, use='complete.obs')
res1 <- cor.mtest(co3,0.99)

corrplot(m, type='upper',diag=TRUE, 
         tl.cex=.9, tl.col='black', tl.pos='lt',
         p.mat=res1[[1]], insig='p-value', sig.level=0.001)

corrplot(m, method= 'number', type='lower',
           diag=TRUE, col='grey40',
           tl.pos='n', cl.pos='n',
           add=TRUE)

```

* NDVI 
```{r}
# Create annual ndvi by pixel 
ndvi_annual <- rd %>% 
  group_by(iv_malla_modi_id, year) %>%
  summarise(ndvi = sum(ndvi[ndvi >=0], na.rm=T)) %>%
  mutate(seasonF='annual', 
         season = 0)

# Create seasonal ndvi by pixel 
ndvi_season <- rd %>% 
  group_by(iv_malla_modi_id, year, season) %>%
  summarise(ndvi = sum(ndvi[ndvi >=0], na.rm=T)) %>% 
  mutate(seasonF = season) %>% 
  mutate(season = ifelse(season == 'autumn', 3, 
                   ifelse(season == 'winter', 4, 
                    ifelse(season == 'spring', 1,2))))

ndvidf <- rbind(ndvi_annual, ndvi_season)

# Join dataframes 
ndvidf <- ndvidf %>% dplyr::inner_join(aux_rd, by="iv_malla_modi_id")
```

### Export dataframes
```{r}
# Export dataframes 
write.csv(evimean, file=paste(di, "/data/evi/evi_mean.csv", sep=""), row.names = FALSE)
write.csv(evidf, file=paste(di, "/data/evi/evi_atributes_all.csv", sep=""), row.names = FALSE)
write.csv(ndvidf, file=paste(di, "/data/evi/ndvi_atributes_all.csv", sep=""), row.names = FALSE)
write.csv(iv_composite, file=paste(di, "/data/evi/iv_composite.csv", sep=""), row.names = FALSE)
```

```{r, echo=FALSE, eval=FALSE}
# * Now we convert data of EVI, NDVI according to our ***filter criteria***: 
#  * we remove the year value of a pixel if the pixel has bad quality in more than 4 composites in a year (4/23 = 17.39%; we retained 82.6 %) 
#  * we remove a pixel of the series if the pixel has bad quality for more 3 years (3/16 = 18.75 %; we retained 81.25 %)

df <- rawfilter %>% 
  filter(year != 2016) %>% 
  select(doy, evi, ndvi, iv_malla_modi_id, pop, date, year, lat, long, summQA, qa_use, filtered, filtered_2) %>% 
  mutate(evi_f = ifelse(filtered_2 == 90, evi, NA),
         ndvi_f = ifelse(filtered_2 == 90, ndvi,NA), 
         mes = lubridate::month(date))

# Create counters 
pixeles <- unique(df$iv_malla_modi_id)
years <- unique(df$year)

# Dataframe for output
filter_pixel_year <- data.frame()

for (i in pixeles){ 
  # Create a subset of pixel of interest   
  df_pixel <- df %>% filter(iv_malla_modi_id == i)

  # Create empty dataframe 
  filter_pixel_year_aux <- data.frame()

  for (j in years){ 
    # empty dataframe 
    out_aux <- data.frame()

    # Create a dataframe for year "j"
    df_pixel_year <- df_pixel %>% filter(year == j)

    # Get numbers of composites by year 
    n_composites <- df_pixel_year %>% count()

    # How many composites are marked as "bad data" 
    bad_composites <- sum(is.na(df_pixel_year$evi_f))

    # Treshold composites
    # For year 2000 is 3; for other year is 4 
    tc <- ifelse(j == 2000, 4, 5)

    # Output 
    status <- ifelse(bad_composites > tc, 'Bad year', 'Ok') 
    out_aux <- cbind(i, y, status)

    # status <- ifelse(bad_composites > tc, 'Bad year', 'Ok') 
    # out_aux$iv_malla_modi_id <- i 
    # out_aux$status <- ifelse(bad_composites > tc, 'Bad year', 'Ok') 
    #   
    #   <- cbind(iv_malla_modi_id = as.numeric(as.character(i)), 
    #              year=as.numeric(as.character(j)), 
    #              status)

    # Rbinds
    filter_pixel_year_aux <- rbind(filter_pixel_year_aux, out_aux)
    }

  # Rbinds
  filter_pixel_year <- rbind(filter_pixel_year, filter_pixel_year_aux)
}

filter_pixel_year <- filter_pixel_year %>% 
  mutate(year = as.numeric(as.character(y)), 
         iv_malla_modi_id = as.numeric(as.character(i))) %>% 
  select(year, iv_malla_modi_id, status)

filter_pixel_year %>% 
  group_by(iv_malla_modi_id, status) %>% 
  summarise(n = n())

```

### 2017-11-13 pres/state_ms.Rmd
update presentation

@@
 ---
-title: "Resiliencia de *Quercus pyreancia* a dos eventos de sequía"
+title: "Resiliencia de *Quercus pyrenaica* a dos eventos de sequía"
 author: "Perez-Luque AJ; Gea-Izquierdo G; Zamora R; Bonet FJ"
 output: 
-  beamer_presentation
+  beamer_presentation: 
+    template: NULL 
 bibliography: ../references.bib
@@
 
+* Utilizar RS information and tree ring to evaluate el efecto de la sequía
+* Resiliencia ..
+
 ---
+
+# Species study
+
+Aquí mapa de Sierra Nevada y mapa de robledales (8 manchas)
+
+---
+
+# Greenness data
+
+Para caracterizar el verdor de la vegetación utilizamos *Enhanced Vegetation Index* (**EVI**) [@Huete2002; @Krapivin2015; @Cabello2012]:
+
+  * Mas sensible que NDVI en áreas con mucha biomasa
+  * La influencia de las condiciones atmosféricas en el cálculo del índice de vegetación es menor en EVI que en NDVI
+  * EVI corrige las señales de fondo del dosel
+
+## Dataset 
+* 2000-2016: Una imagen cada 16 días (23 imágenes por año) 
+* 250 x 250 m 
+* Pixels cubriendo la distribución de los robledales en Sierra Nevada (*n* = 928 pixels) 
+* MODIS MOD13Q1 Collection 6
+
+---
+
+mapa pixeles sobre Sierra Nevada 
+`$TODO$`
+
+---
+
+## Filtrado de datos EVI
+
+Quality Assesment (QA band) & VI Usefulness Indices
+
+928 x 20 x 1 + 928 x 23 x 16 = 360064
+
+* 208437 (57.89 %) Good Data 
+* Filter out: snow/ice (9268) + cloudy (25504) + NA (44) = 35176 (9.77 %)
+* Marginal Data (32.33 %)
+
+    * Explorar distribución temporal y analizar banda QA Detailed [@Reyes2015] 
+    * En zonas montañosas poner especial atención a datos de sombras [@Reyes2015]
+
+Mas info del [\textcolor{blue}{filtrado de datos}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/prepare_modis_qa.md)
+
+Tras el filtrado, nos quedamos con **286825 (79.65 %)**
+
+---
+
+## EVI anual medio
+
+* Además del filtrado, EVI medio es estable frente a la pérdida de datos [@Reyes2015]
+
+![](../analysis/prepare_modis_qa_files/figure-markdown_github/correlation_evi-1.png){height=200px}
+--- 
+# Anomalías estandarizadas EVI
+
+* Computo pixel a pixel 
+* Proporcionan mas información sobre la magnitud de la anomalía
+
+$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
 

### 2017-11-13 man/ms_1_material_methods.Rmd
update

@@
 
-To explore the effect of drought events on greenness we calculated the EVI standardized anomaly (EVI~sa) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012, @Gao2016].
+To explore the effect of drought events on greenness we calculated the EVI standardized anomaly ($EVI~sa$) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012, @Gao2016].
 For each pixel we averaged all the EVI valid values within a year (:red_filter: see quality filter), and then the standardized anomaly was computed as:

### 2017-11-13 analysis/get_shapes_field_work.Rmd
update

---
title: "Field work points shapes"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 May"
output:  
  md_document:
    variant: markdown_github
---

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("raster")
library("stringr")
library('rgdal')
library("sp")
library('rgeos')
```

# Read and prepare data 
```{r, eval=FALSE}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# Read spatial data 
# Centroid of focal tree 
## filename 
mygpx_ca <- paste0(di, 'data_raw/geoinfo/dendro_ca.GPX')
mygpx_sj <- paste0(di, 'data_raw/geoinfo/dendro_sj.GPX')

## reading  
field_work_ca <- readOGR(mygpx_ca, layer="waypoints", verbose = FALSE)
field_work_sj <- readOGR(mygpx_sj, layer="waypoints", verbose = FALSE)

# join 
fw <- rbind(field_work_ca, field_work_sj)
```

```{r}
low_ca<- c('B01', 'B02', 'B03', 'B04', 'B05', 'B06', 'B07', 'B08', 'B09', 'B10',
         'B26', 'B27', 'B28', 'B29', 'B30')
low_sj <- c('A01', 'A02', 'A03', 'A04', 'A05', 'A06', 'A07', 'A08', 'A09', 'A10')
high_sj<- c('A11', 'A12', 'A13', 'A14', 'A15', 'A16', 'A17', 'A18', 'A19', 'A20')
high_ca <- c('B11', 'B12', 'B13', 'B14', 'B15', 'B16', 'B17', 'B18', 'B19', 'B20',
           'B21', 'B22', 'B23', 'B24', 'B25')

# Create variable site 
fw@data$site <- ifelse(fw@data$name %in% low_sj, 'SJLow', 
                       ifelse(fw@data$name %in% high_sj, 'SJHigh',
                              ifelse(fw@data$name %in% low_ca, 'CALow', 'CAHigh')))
# Variable Loc
ca <- c(low_ca, high_ca)
fw@data$loc <- ifelse(fw@data$name %in% ca, "CA", "SJ")

# Centroids
cCALow <- SpatialPointsDataFrame(gCentroid(fw[fw$site == 'CALow',], byid=FALSE),
                                 data.frame(site='CALow'), match.ID = FALSE)

cCAHigh <- SpatialPointsDataFrame(gCentroid(fw[fw$site == 'CAHigh',], byid=FALSE),
                                 data.frame(site='CAHigh'), match.ID = FALSE)

cSJLow <- SpatialPointsDataFrame(gCentroid(fw[fw$site == 'SJLow',], byid=FALSE),
                                 data.frame(site='SJLow'), match.ID = FALSE)

cSJHigh <- SpatialPointsDataFrame(gCentroid(fw[fw$site == 'SJHigh',], byid=FALSE),
                                 data.frame(site='SJHigh'), match.ID = FALSE)

centroides <- rbind(cSJHigh, cSJLow,
                    cCAHigh, cCALow)

# Export shapefiles
exportdsn <- paste0(di, 'data_raw/geoinfo/shapes/')

writeOGR(obj=fw, dsn=exportdsn, layer="field_work_4326", driver="ESRI Shapefile", overwrite_layer = TRUE)
writeOGR(obj=centroides, dsn=exportdsn, layer="centroides_4326", driver="ESRI Shapefile", overwrite_layer = TRUE)
```

### 2017-11-13 pres/state_ms.Rmd
update
update pres

@@
 
-Aquí mapa de Sierra Nevada y mapa de robledales (8 manchas)
+![](../out/fig/geomap/mapa_situacion.jpg){height=300px}
 
@@
 * MODIS MOD13Q1 Collection 6
-
----
-
-mapa pixeles sobre Sierra Nevada 
-`$TODO$`
 
@@
 * Computo pixel a pixel 
-* Proporcionan mas información sobre la magnitud de la anomalía
+* Proporcionan mas información sobre la magnitud de la anomalía [@Samanta2012, @Gao2016]
 
 $$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
+
+---
+
+# FieldWork
+
+* 2 localidades: SJ (norte) y CA (sur)
+* 2 elevaciones por sitio (High-Low)
+
+* Focal trees:
+
+    * 10 - 15  árboles dominantes por cada sitio
+    * 2 cores 5mm 
+    * dbh, altura
+
+* Competencia:
+
+    * Recuento todos los árboles (dbh > 7.5) dentro de un radio de 10 metros 
+    * dbh, altura, distancia árbol focal, rumbo 
+    * Índices de competencia dependientes e independientes de la distancia [@GeaIzquierdo2009]
+
+---
+
+# Dendrochronological methods
 

### 2017-11-20 analysis/analysis_competence.Rmd
update indices
update name competences

@@
 
-# Read data 
+## Compute competition indices
+
 ```{r, echo=FALSE}
@@
 #### San Juan 
-```{r, echo=FALSE, fig.height=8, fig.width=10}
+```{r competence_sj, echo=FALSE, fig.height=8, fig.width=10}
 # see stackoverflow questions 6862742
@@
 #### Canar 
-```{r, echo=FALSE, fig.height=9, fig.width=12}
+```{r competence_ca, echo=FALSE, fig.height=9, fig.width=12}
 compete %>%  

### 2017-11-20 pres/state_ms.Rmd
add slides

@@
 ---
-title: "Resiliencia de *Quercus pyrenaica* a dos eventos de sequía"
+title: |
+  Resiliencia de *Quercus pyrenaica* a dos eventos de sequía
 author: "Perez-Luque AJ; Gea-Izquierdo G; Zamora R; Bonet FJ"
@@
     template: NULL 
+    fig_caption: FALSE
 bibliography: ../references.bib
@@
     * Índices de competencia dependientes e independientes de la distancia [@GeaIzquierdo2009] 
+    * [\textcolor{blue}{Resultados}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/analysis_competence.md)
 
 --- 
+![](../analysis/analysis_competence_files/figure-markdown_github/competence_sj-1.png){height=300px}
+--- 
+![](../analysis/analysis_competence_files/figure-markdown_github/competence_ca-1.png){height=300px}
+---
+# Resilience metrics 
+Resistance quantifies the severity of the impact of the disturbance in the year it occurred 
+Recovery index (*Rc*) is the ability to recover from disturbance relative to its severity. Recovery  = Postdrought / Drought
+Resilience index (*Rs*) is the capacity to reach pre-disturbance performance levels. Resilience = Postdrought / Predrought
+Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance. Relative Resilience = (Postdrought - Drought) / Predrought
 
-# Dendrochronological methods
+---
 

### 2017-11-20 analysis/explore_correlations_windows_size.Rmd
create correla window plots

---
title: "Correlation Resilience Window"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 March"
output:  
  md_document:
    toc: true
    variant: markdown_github
---

# Filter data 
```{r wd, echo=FALSE}
#---------------------------------
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
#---------------------------------
```

```{r packages, warning=FALSE, message=FALSE}
library("tidyverse")
library("ggcorrplot")
library("gridExtra")
```

```{r}

# Read correlation tables of BAI from: ../analysis/analysis_resilience_bai.md 
# sj 
corBai <- read.csv(file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), header = TRUE)
```

```{r}
# Genera correlations matrix

generaMatrix <- function(cordf, variable){ 
  m <- cordf %>% filter(var == variable) %>% 
    dplyr::select(-var, -window_size) %>% 
    mutate(w1 = c(2, 2, 3), 
           w2 = c(3,4,4)) 
  mf <- m %>% 
    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)

  df <- m %>% rbind(mf) %>% 
    add_row(r2 = 1, w1=2, w2=2) %>% 
    add_row(r2 = 1, w1=3, w2=3) %>% 
    add_row(r2 = 1, w1=4, w2=4)

    df <- df %>% 
    mutate(w1 = case_when(
      w1 == 2 ~ 'ws2',
      w1 == 3 ~ 'ws3',
      w1 == 4 ~ 'ws4'),
      w2 = case_when(
      w2 == 2 ~ 'ws2',
      w2 == 3 ~ 'ws3',
      w2 == 4 ~ 'ws4'))

      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))

return(cor_df)
}

cor_rt <- generaMatrix(corBai, 'rt')
cor_rc <- generaMatrix(corBai, 'rc')
cor_rs <- generaMatrix(corBai, 'rs')
cor_rrs <- generaMatrix(corBai, 'rrs')
```

```{r}
g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
g_rrs <- ggcorrplot(cor_rrs, type='upper', lab=TRUE, title='BAI RRs', show.legend = TRUE)

```

```{r correla_window_bai}
grid.arrange(g_rt, g_rc, g_rs, g_rrs)
```

### 2017-11-20 analysis/evi/compute_resilience_evi.Rmd
add script and rmd to compute evi resilience from old repo qpyr_modis_resilience

---
title: "Compute Resilience EVI"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Aug"
output:  
  md_document:
    variant: markdown_github
---
```{r, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.align='center', message = FALSE, warning = FALSE) 
```

```{r wd, echo=FALSE}
#---------------------------------
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
```

```{r packages, warning=FALSE, message=FALSE}
library('tidyverse')
library('stringr')
library('grid')
library('gridExtra')
library('pander')
source(paste0(di, '/script/R/eviResilience.R'))
```

```{r}
# Read data
iv <- read.csv(file=paste(di, "/data/evi_mean.csv", sep=""), header = TRUE, sep = ',')

# Prepare data
# NOTE remove pop 9 
evi <- iv %>% filter(pop != 9) %>% 
  dplyr::select(-n_composites, -long, -lat, -pop)
```

## Compute resilience metrics

* Para los disturbance events de 2005 y 2012, vamos a computar las métricas de resiliencia. Lo hacemos aplicando diferentes ventanas temporales (2, 3 y 4).

* Luego analizamos la correlación para cada indicador (rs, rc, rt y rrs) entre las diferentes ventanas. Si están muy correlacionados podemos decir que el resultado no varía mucho entre ellos.

```{r}
dyears <- c(2005, 2012)

# Compute resilience for 2, 3 and 4 years 
res2 <- eviResilience(evi, event_years = dyears, window = 2)
res3 <- eviResilience(evi, event_years = dyears, window = 3)
res4 <- eviResilience(evi, event_years = dyears, window = 4)
```

```{r}
# Vector with objects name
obj <- c('res2', 'res3', 'res4')

correla_ws <- c()

for (i in obj){ 
  x <- get(i)
  xres <- x$resilience
  out <- xres %>% 
    mutate(ws = paste0('ws_', as.character(str_extract(i, "([0-9])")))) %>% 
    select(-disturb_year, -iv_malla_modi_id)

  correla_ws <- bind_rows(correla_ws, out)
}

# Split by window size 
correla <- correla_ws %>% split(.$ws)

# Change names 
names(correla[["ws_2"]])[1:4] <- paste0(names(correla[["ws_2"]])[1:4], '2')
names(correla[["ws_3"]])[1:4] <- paste0(names(correla[["ws_3"]])[1:4], '3')
names(correla[["ws_4"]])[1:4] <- paste0(names(correla[["ws_4"]])[1:4], '4')

cor2 <- correla[["ws_2"]] %>% select(-ws) %>% mutate(ind = row_number())
cor3 <- correla[["ws_3"]] %>% select(-ws) %>% mutate(ind = row_number())
cor4 <- correla[["ws_4"]] %>% select(-ws) %>% mutate(ind = row_number())

correlations <- inner_join(cor2, cor3, by='ind') %>% inner_join(cor4, by='ind')
```

### Correlations window size 
#### Resistance
```{r}
# Resistance
aux_coefs <- c()

model <- lm(rt2~rt3, data=correlations)
p_rt23 <- correlations %>% ggplot(aes(rt2, rt3)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) + 
  theme(legend.position = c(.2, .75))
aux <- as.data.frame(cbind('rt','2-3', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

model <- lm(rt2~rt4, data=correlations)
p_rt24 <- correlations %>% ggplot(aes(rt2, rt4)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
  theme(legend.position = 'none')
aux <- as.data.frame(cbind('rt','2-4', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

model <- lm(rt3~rt4, data=correlations)
p_rt34 <- correlations %>% ggplot(aes(rt3, rt4)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
  theme(legend.position = 'none')
aux <- as.data.frame(cbind('rt','3-4', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

grid.arrange(p_rt23, p_rt24, p_rt34,ncol=3) 
```

#### Recovery
```{r}
# Recovery 
model <- lm(rc2~rc3, data=correlations)
p_rc23 <- correlations %>% ggplot(aes(rc2, rc3)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) + 
  theme(legend.position = c(.2, .75))
aux <- as.data.frame(cbind('rc','2-3', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

model <- lm(rc2~rc4, data=correlations)
p_rc24 <- correlations %>% ggplot(aes(rc2, rc4)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
  theme(legend.position = 'none')
aux <- as.data.frame(cbind('rc','2-4', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

model <- lm(rc3~rc4, data=correlations)
p_rc34 <- correlations %>% ggplot(aes(rc3, rc4)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
  theme(legend.position = 'none')
aux <- as.data.frame(cbind('rc','3-4', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

grid.arrange(p_rc23, p_rc24, p_rc34,ncol=3) 
```

#### Resilience 
```{r}
# Resilience
model <- lm(rs2~rs3, data=correlations)
p_rs23 <- correlations %>% ggplot(aes(rs2, rs3)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) + 
  theme(legend.position = c(.2, .75))
aux <- as.data.frame(cbind('rs','2-3', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

model <- lm(rs2~rs4, data=correlations)
p_rs24 <- correlations %>% ggplot(aes(rs2, rs4)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
  theme(legend.position = 'none')
aux <- as.data.frame(cbind('rs','2-4', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

model <- lm(rs3~rs4, data=correlations)
p_rs34 <- correlations %>% ggplot(aes(rs3, rs4)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
  theme(legend.position = 'none')
aux <- as.data.frame(cbind('rs','3-4', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

grid.arrange(p_rs23, p_rs24, p_rs34,ncol=3) 
```

#### Relative Resilience
```{r}
# Relative Resilience
model <- lm(rrs2~rrs3, data=correlations)
p_rrs23 <- correlations %>% ggplot(aes(rrs2, rrs3)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) + 
  theme(legend.position = c(.2, .75))
aux <- as.data.frame(cbind('rrs','2-3', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

model <- lm(rrs2~rrs4, data=correlations)
p_rrs24 <- correlations %>% ggplot(aes(rrs2, rrs4)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
  theme(legend.position = 'none')
aux <- as.data.frame(cbind('rrs','2-4', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

model <- lm(rrs3~rrs4, data=correlations)
p_rrs34 <- correlations %>% ggplot(aes(rrs3, rrs4)) + 
  geom_point() + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
  theme(legend.position = 'none')
aux <- as.data.frame(cbind('rrs','3-4', as.numeric(summary(model)$r.squared)))
aux_coefs <- rbind(aux_coefs, aux)

grid.arrange(p_rrs23, p_rrs24, p_rrs34,ncol=3)

names(aux_coefs) <- c('var', 'window_size', 'r2')

write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), row.names = F)
```

```{r}
aux_coefs %>% pander()
```

** NOS QUEDAMOS CON 3 AÑOS COMO VENTANA TEMPORAL**

## Export and save 
```{r}
# Export csv 
write.csv(res3$evi, file=paste0(di, '/data/resilience/evi_period.csv'), row.names = FALSE)
write.csv(res3$resilience, file=paste0(di, '/data/resilience/resiliences.csv'), row.names = FALSE)
```

### 2017-11-20 analysis/explore_correlations_windows_size.Rmd
update correlations window size

@@
 corBai <- read.csv(file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), header = TRUE)
+corEvi <- read.csv(file=paste0(di, '/out/correla_resilience/correla_window_size_evi.csv'), header = TRUE)
 ```
@@
 }
+```
 
+# BAI
+
+```{r}
 cor_rt <- generaMatrix(corBai, 'rt')
@@
 ```
+
+# EVI
+
+```{r}
+cor_rte <- generaMatrix(corEvi, 'rt')
+cor_rce <- generaMatrix(corEvi, 'rc')
+cor_rse <- generaMatrix(corEvi, 'rs')
+cor_rrse <- generaMatrix(corEvi, 'rrs')
+```
+
+```{r}
+g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
+g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
+g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = FALSE)
+g_rrse <- ggcorrplot(cor_rrse, type='upper', lab=TRUE, title='EVI RRs', show.legend = TRUE)
+
+```
+
+```{r correla_window_evi}
+grid.arrange(g_rte, g_rce, g_rse, g_rrse)
+```

### 2017-11-20 analysis/evi/compute_MannKendall_evimean.Rmd
add Mann Kendall analysis of evi mean from  old repo qpyr_modis_resilience

---
title: "Compute EVI mean trends"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Aug"
output:  
  md_document:
    variant: markdown_github
---
```{r, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.align='center', fig.width = 14, message = FALSE, warning = FALSE) 
```

```{r wd, echo=FALSE}
#---------------------------------
machine <- 'ajpelu'
#machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro', sep='')
#---------------------------------
```

```{r packages, warning=FALSE, message=FALSE}
library("trend")
library('tidyverse')
source(paste0(di, '/script/R/exploreMKTS.R'))
library("rgdal")
library("sp")
library("raster")
library("rasterVis")
```

```{r}
# Read data
iv <- read.csv(file=paste(di, "/data/evi/evi_mean.csv", sep=""), header = TRUE, sep = ',')
```

```{r}
## ojo filtramos por pop (quitmos 9 pop)
ivtrend <- iv %>% filter(pop != 9) %>% dplyr::select(iv_malla_modi_id, year, evi)
```

```{r}

pixel <- unique(ivtrend$iv_malla_modi_id)

out <- data.frame()

for (i in pixel){ 
  m_aux <- c()

  # dataframe by pixel
  iv_pixel <- ivtrend %>% filter(iv_malla_modi_id == i)

  # MK trend test
  m <- mk.test(iv_pixel$evi, alternative = "two.sided", continuity = TRUE)

  # Sen slope 
  s <- sens.slope(iv_pixel$evi, conf.level = 0.95)

  p_value = round(m$p.value, 5)

  m_aux <- as.data.frame(
    cbind.data.frame(iv_malla_modi_id = i, 
          p_value = round(m$p.value, 6),
          statistic = round(m$statistic, 3),
          t(round(m$estimates, 5)), 
          senslope = round(s$estimates, 5),
          sen_lowci = round(s$conf.int[1],5),
          sen_uppci = round(s$conf.int[2],5),
          sen_levelci = as.character(attributes(s$conf.int))))

  row.names(m_aux) <- NULL

  out <- rbind.data.frame(out, m_aux)
}

iv_att <- iv %>% dplyr::select(-year, -evi, -n_composites) %>% unique()

mk_evi  <- out %>% inner_join(iv_att, by = 'iv_malla_modi_id') 
```

## Explore results Trends

```{r}
mk_per <- exploreMKTS(mk_evi, alpha=0.05) 
mk_per$summary %>% kable()

```

```{r trend_sig_pixels}
mk_evi <- mk_evi %>% mutate(sig = ifelse(p_value < 0.05, 'sig', 'nosig'))

mk_evi %>% ggplot(aes(x=tau)) + 
  geom_histogram(stat='bin', bindwidth=.1, fill='blue') +
  facet_wrap(~sig, labeller = as_labeller(c('nosig' = 'Non significant',
                                            'sig' = 'Significant'))) +
  theme_bw() + ylab('npixels') + 
  theme(strip.background = element_rect(colour='black', 
                                        fill='white'))
```

# Maps 
```{r}
# Create a spatial dataframe 
mk_evi_sp <- SpatialPointsDataFrame(coords = mk_evi[,c('long','lat')], 
                                  data = mk_evi, proj4string = CRS("+init=epsg:4326"))

variables <- c('tau', 'senslope', 'p_value')

# Transform 
aux_spatial <- spTransform(mk_evi_sp, CRS("+init=epsg:23030"))

for (i in variables){

  # raster auxiliar 
  aux_rast <- raster(aux_spatial, resolution=250)

  # raster_tau
  mr <- rasterize(aux_spatial, aux_rast, i, fun=mean)
  names(mr) <- paste0(i)

  # reprojected raster
  mr_re <- projectRaster(mr, crs=crs("+init=epsg:4326"))
  names(mr_re) <- paste0(i)

  # assign 
  name_raster <- paste0('r_', i)
  assign(name_raster, mr)

  name_raster_re <- paste0('r_re_', i)
  assign(name_raster_re, mr_re)

  writeRaster(mr, file=paste(di, "/data/raster/r_trendEVI_" , i, ".asc", sep=""), overwrite=TRUE)
  writeRaster(mr_re, file=paste(di, "/data/raster/r_trendEVI_re_", i, ".tiff", sep=""), overwrite=TRUE)

}

# Create stack of raster 
stack_trends <- stack(r_tau, r_senslope, r_p_value)
stack_trends_re <- stack(r_re_tau, r_re_senslope, r_re_p_value)
```

```{r}
# Export stack 
temp <- getwd()
setwd(paste(di, "/data/raster/", sep=""))
writeRaster(stack_trends, filename = 'r_trend_stack', overwrite =TRUE) 
writeRaster(stack_trends_re, filename = 'r_trend_re_stack', overwrite =TRUE) 
setwd(temp)
```

```{r rater_map_tau_EVI}
# Select a palette http://colorbrewer2.org/
mypal <- brewer.pal(11, "RdYlGn")
# Specify the color palette
myTheme=rasterTheme(region=mypal)

lp <- levelplot(stack_trends_re, 
          margin=FALSE,
          layer='tau',
          par.settings=myTheme, 
          #par.settings=RdBuTheme, 
          names.attr="tau",
          pretty=TRUE,
          layout=c(1,1),
          main='tau', xlab=NULL, ylab=NULL)

print(lp)

tiff(filename=paste0(di, '/out/raster_maps/trends_evi_tau_.tiff'), 
     height = 1600, width = 2500, res=150, pointsize = 20, bg='transparent')
print(lp)
dev.off() 
```

```{r raster_map_sen_evi}
lp <- levelplot(stack_trends_re, 
          margin=FALSE,
          layer='senslope',
          #par.settings=myTheme, 
          par.settings=RdBuTheme, 
          names.attr="senslope",
          pretty=TRUE,
          layout=c(1,1),
          main='Sen Slope', xlab=NULL, ylab=NULL)

print(lp)

tiff(filename=paste0(di, '/out/raster_maps/trends_evi_senslope_.tiff'), 
     height = 1600, width = 2500, res=150, pointsize = 20, bg='transparent')
print(lp)
dev.off() 
```

### 2017-11-20 analysis/evi/compute_MannKendall_evimean.Rmd
modify size

@@
 require(knitr)
-opts_chunk$set(fig.align='center', fig.width = 14, message = FALSE, warning = FALSE) 
+opts_chunk$set(message = FALSE, warning = FALSE) 
 ```

### 2017-11-20 analysis/evi/explore_evi_mean.Rmd
add analysis of EVI mean from old repo qpyr_modis_resilience

---
title: "Compute EVI mean trajectories"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Aug"
output:  
  md_document:
    variant: markdown_github
---
```{r, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(message = FALSE, warning = FALSE) 
```

```{r wd, echo=FALSE}
#---------------------------------
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro', sep='')
#---------------------------------
```

```{r packages, warning=FALSE, message=FALSE}
# library("plyr")
# library('dplyr')
library("scales")
library('tidyverse')
```

# Trajectories EVI mean

```{r}
# Read data
evi <- read.csv(file=paste(di, "/data/evi/evi_mean.csv", sep=""), header = TRUE, sep = ',')

evi <- evi %>% 
  mutate(
    clu_pop = as.factor(case_when(
      pop == 1 ~ "Camarate",
      pop %in% c(2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out')),
    clu_pop2 = as.factor(case_when(
      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out'))) %>% 
  filter(clu_pop != 'out')
```

## global

```{r trajectories_evi}
traj <- evi %>% dplyr::group_by(pop, year) %>% 
  dplyr::summarise(mean = mean(evi),
            sd = sd(evi))

traj_mean <- traj %>% 
  group_by(year) %>% 
  dplyr::summarise(meanOfmean = mean(mean), 
            sdOfmean = sd(mean),
            seOfmean = sdOfmean/sqrt(length(mean)),
            meanOfsd = mean(sd),
            sdOfsd = sd(sd),
            seOfsd = sdOfsd /sqrt(length(sd))) %>% 
  as.data.frame()

traj_plot <- traj_mean %>% 
  ggplot(aes(x=meanOfmean, y=meanOfsd, label=year)) +
  geom_errorbar(aes(ymin=meanOfsd - seOfsd, ymax=meanOfsd + seOfsd)) + 
  geom_errorbarh(aes(xmin=meanOfmean - seOfmean, xmax=meanOfmean + seOfmean)) + 
  geom_path(colour='gray') +
  geom_point(size=3, shape=21, fill='white') + 
  geom_text(hjust = 0.001, nudge_x = 0.001) + 
  geom_vline(aes(xintercept = mean(meanOfmean)), colour='red') +
  geom_hline(aes(yintercept = mean(meanOfsd)), colour ='red')+
  theme_bw() + xlab('mean') + ylab('variance') + 
  theme(strip.background = element_rect(fill = "white"), 
        legend.position="none")

traj_plot

pdf(file=paste0(di, "/out/trajectories/plot_trajectories_evi.pdf"), height = 6, width =6)
traj_plot
dev.off()
```

## by cluster pop

```{r trajectories_evi_clu}
traj_mean_pop <- traj %>% 
  mutate(clu_pop = as.factor(ifelse(pop %in% c(1,2,3,4,5), 'N', 'S'))) %>% 
  group_by(clu_pop,year) %>% 
  dplyr::summarise(meanOfmean = mean(mean), 
            sdOfmean = sd(mean),
            seOfmean = sdOfmean/sqrt(length(mean)),
            meanOfsd = mean(sd),
            sdOfsd = sd(sd),
            seOfsd = sdOfsd /sqrt(length(sd))) %>%
  as.data.frame()

line_traj_mean_pop <- traj_mean_pop %>% 
  group_by(clu_pop) %>% 
  dplyr::summarise(
    meanOfmean = mean(meanOfmean), 
    meanOfsd = mean(meanOfsd)
  )

traj_plot_pop <- ggplot(traj_mean_pop,
  aes(x=meanOfmean, y=meanOfsd, label=year)) +
  geom_errorbar(aes(ymin=meanOfsd - seOfsd, ymax=meanOfsd + seOfsd)) + 
  geom_errorbarh(aes(xmin=meanOfmean - seOfmean, xmax=meanOfmean + seOfmean)) + 
  geom_path(colour='gray') +
  geom_point(size=3, shape=21, fill='white') +
  geom_text(hjust = 0.001, nudge_x = 0.001) + 
  facet_wrap(~clu_pop) +
  theme_bw() + xlab('mean') + ylab('variance') + 
  theme(strip.background = element_rect(fill = "white"), 
        legend.position="none")

traj_plot_pop <- 
  traj_plot_pop + 
  geom_vline(aes(xintercept = meanOfmean), line_traj_mean_pop,  colour='red') +
  geom_hline(aes(yintercept = meanOfsd), line_traj_mean_pop, colour ='red')

traj_plot_pop
pdf(file=paste0(di, "/out/trajectories/plot_trajectories_evi_clu.pdf"), height = 6, width =10)
traj_plot_pop
dev.off()
```

# EVI profiles

```{r}
# Read data
iv <- read.csv(file=paste(di, "/data/evi/iv_composite.csv", sep=""), header = TRUE, sep = ',')
```

## Prepare data

```{r}
evi_profile_dat <- iv %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=T),
            sd = sd(evi, na.rm=T),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d'))

ndvi_profile_dat <- iv %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(ndvi, na.rm=TRUE),
            sd = sd(ndvi, na.rm=TRUE),
            se = sd/sqrt(length(ndvi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d'))

```

## Plot EVI profile

```{r evi_profile}
micolor <- '#455883'

evi_profile <- ggplot(evi_profile_dat, aes(cd, y=mean)) + 
  geom_errorbar(aes(ymin = mean - 10*se, ymax= mean + 10*se), width=4, colour=micolor, size=.8) + 
  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') +
  geom_line(colour=micolor, size=.8) + 
  geom_point(size=3,colour=micolor) +
  geom_point(size=1.5, colour='white')+
  scale_x_date(labels = function(x) format(x, "%b"),
               breaks = date_breaks('month')) + 
  ylab('EVI') + xlab('Date') + 
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank()) +
  theme_classic()
#format(x, "%d-%b")

print(evi_profile)

pdf(file=paste0(di, "/out/evi_profiles/plot_profile_evi.pdf"), height = 8, width = 12)
evi_profile
dev.off()

```

## Plot NDVI profile

```{r ndvi_profile}
ndvi_profile <- ggplot(ndvi_profile_dat, aes(cd, y=mean)) + 
  geom_errorbar(aes(ymin = mean - 10*se, ymax= mean + 10*se), width=4, colour=micolor, size=.8) + 
  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') +
  geom_line(colour=micolor, size=.8) + 
  geom_point(size=3,colour=micolor) +
  geom_point(size=1.5, colour='white')+
  scale_x_date(labels = function(x) format(x, "%b"),
               breaks = date_breaks('month')) + 
  ylab('NDVI') + xlab('Date') + 
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank()) +
  theme_classic()
#format(x, "%d-%b")

print(ndvi_profile)

pdf(file=paste0(di, "/out/evi_profiles/plot_profile_ndvi.pdf"), height = 8, width = 12)
ndvi_profile
dev.off()

```

## Comparing 2005 and 2012 with reference

```{r}
drought_years <- c(2005, 2012)

evi_profile_dat_ref <- iv %>% 
  filter(!year %in% drought_years) %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
            sd = sd(evi, na.rm=TRUE),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
  mutate(period = 'reference')

evi_profile_dat_2005 <- iv %>% 
  filter(year == 2005) %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
            sd = sd(evi, na.rm=TRUE),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
  mutate(period = '2005')

evi_profile_dat_2012 <- iv %>% 
  filter(year == 2012) %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
            sd = sd(evi, na.rm=TRUE),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
  mutate(period = '2012')

# Join the three dataframes
evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
```

```{r evi_profile_compara}
# colref <- '#455883'
# colref2 <- '#10253F'
# col2005 <- '#00BA38'
# col2012 <- '#F8766D'

col2012 <- '#0700fe'
col2005 <- '#19e00b'
colref <- '#7b7b7b'

profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
  geom_line(size=.9) + 
  geom_point(size=3, fill='white', shape=21) +
  scale_x_date(labels = function(x) format(x, "%b"),
               breaks = date_breaks('month')) + 
  ylab('EVI') + xlab('Date') + 
  theme_bw() +
  theme_classic() + 
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank()) +
  scale_color_manual(values=c(col2005, col2012, colref))

#format(x, "%d-%b")

profile_compara

pdf(file=paste0(di, "/out/evi_profiles/plot_evi_profiles_compara.pdf"), height = 8, width = 12)
profile_compara
dev.off()
```

### 2017-11-20 analysis/evi/compute_anomalies.Rmd
add compute evi anomialies from old repo qpyr_modis_resilience

---
title: "Compute EVI anomalies"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Aug"
output:  
  md_document:
    variant: markdown_github
---
```{r, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(message = FALSE, warning = FALSE) 
```

```{r wd, echo=FALSE}
#---------------------------------
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro', sep='')
#---------------------------------
```

```{r packages, warning=FALSE, message=FALSE}
library('tidyverse')
```

# Read data 
Two dataframes: 
* EVI mean dataframe
* IV composite

(for another previous analysis see this [repo](https://github.com/ajpelu/qpyr_resilience)) and

```{r}
# Read data
# iv <- read.csv(file=paste(di, "/data/iv_composite.csv", sep=""), header = TRUE, sep = ',')
evi <- read.csv(file=paste(di, "/data/evi/evi_mean.csv", sep=""), header = TRUE, sep = ',')
```

# Compute anomalies

# By evi_mean
```{r, eval=FALSE}
# Set counters
pixels <- unique(evi$iv_malla_modi_id)
# composites <- unique(evi$seasonF)
years <- unique(evi$year)

# Set dataframe to compute anomaly
df <- evi

# Create emtpy datafraje to store output 
anomalos <- data.frame()

for (i in pixels){
  df_aux <- df[df$iv_malla_modi_id == i,]

    # Create empty df to store anomlay of composite j of all year 
    aux_anomaly <- data.frame()

    for (y in years){
      # get mean evi for the reference period
      iv_ref <- df_aux %>% 
        filter(year != y) %>% 
        summarise(mean(evi, na.rm=T))

      # Get sd ref for standardized anomaly
      sd_ref <- df_aux %>% 
        filter(year != y) %>% 
        summarise(sd(evi, na.rm=T))

      # get evi for the year
      iv_year <- df_aux[df_aux$year == y, 'evi']
      # To solve the problems with the year without composites (i.e 2000)
      iv_year <- ifelse(length(iv_year) == 0, 0, iv_year)

      # Compute anomaly
      anomaly <- (iv_year - iv_ref)
      names(anomaly) <- 'a'

      # Compute normalized anomaly 
      anomaly_nor <- ((iv_year - iv_ref) / (iv_year + iv_ref))*100
      names(anomaly_nor) <- 'nora'

      # Compute standardized anomaly
      anomaly_std <- (iv_year - iv_ref) / sd_ref
      names(anomaly_std) <- 'sa'

      # Create dataframe 
      aux <- cbind(y, anomaly, anomaly_nor, anomaly_std)
      aux_anomaly <- rbind(aux_anomaly, aux)
      }

  # Add name of the pixel 
  aux_anomaly <- aux_anomaly %>% mutate(iv_malla_modi_id = i)

  # Output 
  anomalos <- rbind(anomalos, aux_anomaly)
}

# Add pob, lat, long, etc 
evi_aux <- evi %>% dplyr::select(iv_malla_modi_id, long, lat, pop) %>%
  group_by(iv_malla_modi_id) %>% unique()

# Join dataframes 
anomalias_evimean <- anomalos %>% dplyr::inner_join(evi_aux, by="iv_malla_modi_id")

write.csv(anomalias_evimean, file=paste(di, "/data/anomalies/anomalias_evimean.csv", sep=""), row.names = FALSE)
```

## Notas

Las anomalias se calculan para cada pixel del siguiente modo: a = evi_year - evi_ref; siendo evi_year el valor del EVI medio para el año en cuestión y *evi_ref* la media de todos los valores de EVI medio en todo el periodo de referencia.

En el caso de las anomalías estandarizadas, utilizamos la aproximación de Gao et al. (2016) [doi:10.1038/srep26958](https://dx.doi.org/10.1038/srep26958), donde las anomalias las dividimos por la desviación estandar de los valores de EVI medio para el periodo de referencia. Las anomalías estandarizadas generalmente proporcionan mas información referente a la magnitud de la anomalía ya que las potenciales influencias de la dispersión de los datos han sido eliminadas. Por tanto las anomalías estandarizadas se calculan como: sa = (evi_year - evi_ref) / sd_ref. 
Ver esto http://iopscience.iop.org/article/10.1088/1748-9326/7/2/024018/meta

Finalmente, computamos las anomalias normalizadas, en las cuales escalamos los valores de las anomalías entre -100 y 100, de acuerdo a la formula: nora = (evi_year - evi_ref) / (evi_year + evi_ref)*100. Mas información ver https://www.researchgate.net/post/how_can_I_normalize_the_range_of_NDVI_anomaly

Por otro lado, podemos utilizar el criterio de Samanta et al. 2010 (http://onlinelibrary.wiley.com/doi/10.1029/2009GL042154/abstract y  http://dx.doi.org/10.1111/j.1469-8137.2010.03516.x) --> Ojo leer bien.

## By composite (OJO esto aun no lo he computado: 23 / aug)

```{r, eval=FALSE}
# Set counters
pixels <- unique(iv$iv_malla_modi_id)
composites <- unique(iv$composite)
years <- unique(iv$year)

# Set dataframe to compute anomaly
df <- iv

# Create emtpy datafraje to store output 
anomalos <- data.frame()

for (i in pixels){
  df_aux <- df[df$iv_malla_modi_id == i,]

  # Create empty auxiliar dataframe to store results
  aux_composite <- data.frame()

  for (j in composites) {
    # Create df_auxiliar by composite
    df_by_composite <- df_aux[df_aux$composite == j,]

    # Create empty df to store anomlay of composite j of all year 
    aux_anomaly <- data.frame()

    for (y in years){
      # get mean evi for the reference period
      iv_ref <- df_by_composite %>% 
        filter(year != y) %>% 
        summarise(mean(evi, na.rm=T))

      # get evi for the year
      iv_year <- df_by_composite[df_by_composite$year == y, 'evi']
      # To solve the problems with the year without composites (i.e 2000)
      iv_year <- ifelse(length(iv_year) == 0, 0, iv_year)

      # Compute standardized anomaly
      anomaly_std <- ((iv_year - iv_ref) / (iv_year + iv_ref))*100
      names(anomaly_std) <- 'anomaly_std'

      # Compute anomaly
      anomaly <- (iv_year - iv_ref)
      names(anomaly) <- 'anomaly'

      # Create dataframe 
      aux <- cbind(y, anomaly, anomaly_std)
      aux_anomaly <- rbind(aux_anomaly, aux)
      }

    aux_compos <- aux_anomaly %>% mutate(composite = j)

    aux_composite <- rbind(aux_composite, aux_compos)

  }

  # Remove the composites 1,2 an 3 for year 2000 
  aux_composite <- aux_composite[!(aux_composite$y == 2000 & aux_composite$composite %in% c(1:3)), ]

  # Add name of the pixel 
  aux_composite <- aux_composite %>% mutate(iv_malla_modi_id = i)

  # Output 
  anomalos <- rbind(anomalos, aux_composite)
}

# Add pob, lat, long, etc 
iv_aux <- iv %>% dplyr::select(iv_malla_modi_id, long, lat, pop) %>%
  group_by(iv_malla_modi_id) %>% unique()

# Join dataframes 
anomalias_composite <- anomalos %>% dplyr::inner_join(iv_aux, by="iv_malla_modi_id")

write.csv(anomalias_composite, file=paste(di, "/data/anomalies/anomalias_composite.csv", sep=""), row.names = FALSE)
```

### 2017-11-20 analysis/evi/explore_anomalies.Rmd
explore anomalies from old repo qpyr_modis_resilience

---
title: "Explore anomalies"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Aug"
output:  
  md_document:
    variant: markdown_github
---
```{r, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(message = FALSE, warning = FALSE) 
```

```{r wd, echo=FALSE}
#---------------------------------
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro', sep='')
#---------------------------------
```

```{r packages, warning=FALSE, message=FALSE}
library("rgdal")
library("sp")
library("raster")
library("rasterVis")
source(paste0(di,"/script/R/exportpdf.R")) # function to export raster levelplots maps as pdf
library("dplyr")
library("ggplot2")
library("reshape2")
library("purrr")
# library("pander")
library("knitr")
library("effects")
library("pander")
library("broom")
```

## Read data and Prepare Data

* Crear variable factorial para agrupar a las poblaciones 
* No incluiremos la poblacion 9

```{r}
# Read data 
anomalias_evimean <- read.csv(file=paste(di, "/data/anomalies/anomalias_evimean.csv", sep=""), header = TRUE, sep = ',')
# anomalias_composite <- read.csv(file=paste(di, "/data/anomalies/anomalias_composite.csv", sep=""), header = TRUE, sep = ',')

anomalias_evimean <- anomalias_evimean %>% 
  mutate(
    clu_pop = as.factor(case_when(
      pop == 1 ~ "Camarate",
      pop %in% c(2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out')),
    clu_pop2 = as.factor(case_when(
      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out'))) %>% 
  filter(clu_pop != 'out')
```

```{r, echo=FALSE}
# colours 
# color_neg <- '#d73027'
color_neg <- '#a63603'
# darkblue
#color_pos <- '#1a9850'
color_pos <- '#006d2c'
```

## Anomalies by cluster populations

* Datos medios de las anomalias (*a*), anomalias normalizadas (*nora*) y de las anomalias estandarizadas(*sa*) por cluster

* Consideramos al Camarate (pop 1, dentro del Northern Slopes)

```{r}
# Anomalies 
avg_a_clu <- anomalias_evimean %>% 
  group_by(clu_pop2, y) %>% 
  dplyr::summarise(mean=mean(a, na.rm=T),
            sd = sd(a, na.rm=T),
            se = sd/sqrt(length(a))) %>% 
 mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))

# Normalized Anomalies 
avg_nora_clu <- anomalias_evimean %>% 
  group_by(clu_pop2, y) %>% 
  dplyr::summarise(mean=mean(nora, na.rm=T),
            sd = sd(nora, na.rm=T),
            se = sd/sqrt(length(nora))) %>% 
 mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))

# Standardized Anomalies 
avg_sa_clu <- anomalias_evimean %>% 
  group_by(clu_pop2, y) %>% 
  dplyr::summarise(mean=mean(sa, na.rm=T),
            sd = sd(sa, na.rm=T),
            se = sd/sqrt(length(sa))) %>% 
 mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
```

### EVI Anomaly by cluster pop

```{r plot_anoma_EVI_clu_a}
myylab <- 'EVI Anomaly'

plot_a_clu <- avg_a_clu %>%  
  ggplot(aes(x=y, y=mean, fill=signo)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
  facet_wrap(~clu_pop2, nrow = 2) + 
  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  ylab(myylab) + xlab('') +
  theme_bw() +
  theme(text = element_text(size=16),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_clu_a.pdf"), height = 7, width = 7)
plot_a_clu
dev.off()
```

### EVI Normalized Anomaly by cluster pop

```{r plot_anoma_EVI_clu_nora}
myylab <- 'EVI Normalized Anomaly'

plot_nora_clu <- avg_nora_clu %>%  
  ggplot(aes(x=y, y=mean, fill=signo)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
  facet_wrap(~clu_pop2, nrow = 2) + 
  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  ylab(myylab) + xlab('') +
  theme_bw() +
  theme(text = element_text(size=16),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_clu_nora.pdf"), height = 7, width = 7)
plot_nora_clu 
dev.off()
```

### EVI Standardized Anomaly by cluster pop

```{r plot_anoma_EVI_clu_sa}
myylab <- 'EVI Standardized Anomaly'

plot_sa_clu  <- avg_sa_clu %>%  
  ggplot(aes(x=y, y=mean, fill=signo)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
  facet_wrap(~clu_pop2, nrow = 2) + 
  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  ylab(myylab) + xlab('') +
  theme_bw() +
  theme(text = element_text(size=16),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_clu_sa.pdf"), height = 7, width = 7)
plot_sa_clu
dev.off()
```

## Anomalies by populations

* Datos medios de las anomalias (*a*), anomalias normalizadas (*nora*) y de las anomalias estandarizadas(*sa*) por poblaciones

```{r}
# Anomalies 
avg_a_pop <- anomalias_evimean %>% 
  group_by(pop, y) %>% 
  dplyr::summarise(mean=mean(a, na.rm=T),
            sd = sd(a, na.rm=T),
            se = sd/sqrt(length(a))) %>% 
 mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))

# Normalized Anomalies 
avg_nora_pop <- anomalias_evimean %>% 
  group_by(pop, y) %>% 
  dplyr::summarise(mean=mean(nora, na.rm=T),
            sd = sd(nora, na.rm=T),
            se = sd/sqrt(length(nora))) %>% 
 mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))

# Standardized Anomalies 
avg_sa_pop <- anomalias_evimean %>% 
  group_by(pop, y) %>% 
  dplyr::summarise(mean=mean(sa, na.rm=T),
            sd = sd(sa, na.rm=T),
            se = sd/sqrt(length(sa))) %>% 
 mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
```

```{r}
label_pop <- c('1' = 'Lugros', '2' = 'Guejar-Sierra', '3' = 'Monachil',
                      '4' = 'Dilar', '5' = 'Durcal', '6' = 'Caniar', 
                      '7' = 'Poqueira', '8' = 'Trevelez', '9' = 'Cadiar')
```

### EVI Anomaly by pop

```{r plot_anoma_EVI_pop_a}
myylab <- 'EVI Anomaly'

plot_a_pop <- avg_a_pop %>%  
  ggplot(aes(x=y, y=mean, fill=signo)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
  facet_wrap(~pop, nrow = 3, labeller = as_labeller(label_pop)) + 
  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  ylab(myylab) + xlab('') +
  theme_bw() +
  theme(text = element_text(size=16), legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_a.pdf"), height = 14, width = 14)
plot_a_pop
dev.off()
```

### EVI Normalized Anomaly by pop

```{r plot_anoma_EVI_pop_nora}
myylab <- 'EVI Normalized Anomaly'

plot_nora_pop <- avg_nora_pop %>%  
  ggplot(aes(x=y, y=mean, fill=signo)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
  facet_wrap(~pop, nrow = 3, labeller = as_labeller(label_pop)) +
  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  ylab(myylab) + xlab('') +
  theme_bw() +
  theme(text = element_text(size=16),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_nora.pdf"), height = 14, width = 14)
plot_nora_pop 
dev.off()
```

### EVI Standardized Anomaly by pop

```{r plot_anoma_EVI_pop_sa}
myylab <- 'EVI Standardized Anomaly'

plot_sa_pop  <- avg_sa_pop %>%  
  ggplot(aes(x=y, y=mean, fill=signo)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
  facet_wrap(~pop, nrow = 3, labeller = as_labeller(label_pop)) + 
  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  ylab(myylab) + xlab('') +
  theme_bw() +
  theme(text = element_text(size=16),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_sa.pdf"), height = 14, width = 14)
plot_sa_pop
dev.off()
```

# Maps

* Create raster maps (two projections: `epsg:4326` and `epsg:23030`)

```{r, echo=FALSE}
# Create a spatial dataframe 
anom_sp <- SpatialPointsDataFrame(coords = anomalias_evimean[,c('long','lat')], 
                                  data = anomalias_evimean, proj4string = CRS("+init=epsg:4326"))

# Transform 
anom_sp <- spTransform(anom_sp, CRS("+init=epsg:23030"))

myyear <- c(2005,2012)

for (j in myyear){

  # Filter by year 
  aux_spatial_year <- anom_sp[anom_sp$y ==j,]

  # raster auxiliar 
  aux_rast <- raster(aux_spatial_year, resolution=250)

  # raster_layers 
  r_a <- rasterize(aux_spatial_year, aux_rast, 'a', fun=mean)
  names(r_a) <- paste0('a_', j)

  r_nora <- rasterize(aux_spatial_year, aux_rast, 'nora', fun=mean)
  names(r_nora) <- paste0('nora_', j)

  r_sa <- rasterize(aux_spatial_year, aux_rast, 'sa', fun=mean)
  names(r_sa) <- paste0('sa_', j)

  # reprojected raster
  r_a_re <- projectRaster(r_a, crs=crs("+init=epsg:4326"))
  names(r_a_re) <- paste0('a_',j)

  r_nora_re <- projectRaster(r_nora, crs=crs("+init=epsg:4326"))
  names(r_nora_re) <- paste0('nora_',j)

  r_sa_re <- projectRaster(r_sa, crs=crs("+init=epsg:4326"))
  names(r_sa_re) <- paste0('sa_',j)

  # assign 
  name_raster_a <- paste0('r_anomalia_a', j) 
  name_raster_a_re <- paste0('r_anomalia_a', j, "_re") 
  assign(name_raster_a, r_a)
  assign(name_raster_a_re, r_a_re)
  writeRaster(r_a, file=paste(di, "/data/raster/r_anomalia_a_" , j, ".asc", sep=""), overwrite=TRUE)
  writeRaster(r_a_re, file=paste(di, "/data/raster/r_anomalia_a_" , j, "_re", ".tiff", sep=""), overwrite=TRUE)

  name_raster_nora <- paste0('r_anomalia_nora', j) 
  name_raster_nora_re <- paste0('r_anomalia_nora', j, "_re") 
  assign(name_raster_nora, r_nora)
  assign(name_raster_nora_re, r_nora_re)
  writeRaster(r_nora, file=paste(di, "/data/raster/r_anomalia_nora_" , j, ".asc", sep=""), overwrite=TRUE)
  writeRaster(r_nora_re, file=paste(di, "/data/raster/r_anomalia_nora_", j, "_re", ".tiff", sep=""), overwrite=TRUE)

  name_raster_sa <- paste0('r_anomalia_sa', j) 
  name_raster_sa_re <- paste0('r_anomalia_sa', j, "_re") 
  assign(name_raster_sa, r_sa)
  assign(name_raster_sa_re, r_sa_re)
  writeRaster(r_sa, file=paste(di, "/data/raster/r_anomalia_sa_" , j, ".asc", sep=""), overwrite=TRUE)
  writeRaster(r_sa_re, file=paste(di, "/data/raster/r_anomalia_sa_" , j, "_re", ".tiff", sep=""), overwrite=TRUE)

}

```

```{r}
# Create stack of raster 
stack_anomalias <- stack(r_anomalia_a2005, r_anomalia_a2012, r_anomalia_nora2005, r_anomalia_nora2012, r_anomalia_sa2005, r_anomalia_sa2012)

stack_anomalias_re <- stack(r_anomalia_a2005_re, r_anomalia_a2012_re, r_anomalia_nora2005_re, r_anomalia_nora2012_re, r_anomalia_sa2005_re, r_anomalia_sa2012_re)

# Export stack 
temp <- getwd()
setwd(paste(di, "/data/raster/", sep=""))
writeRaster(stack_anomalias, filename = 'r_all_nomalias_stack', overwrite =TRUE) 
writeRaster(stack_anomalias_re, filename = 'r_all_anomalias_re_stack', overwrite =TRUE) 
setwd(temp)
```

## Spatial exploration of the anomaly

### Anomaly  
```{r map_anomaly_2005_2012_a}
# Select a palette http://colorbrewer2.org/
mypal <- brewer.pal(11, "RdYlGn")
# Specify the color palette
myTheme=rasterTheme(region=mypal)

lp <- levelplot(stack_anomalias, 
          margin=FALSE,
          layer=c('a_2005', 'a_2012'),
          #par.settings=myTheme, 
          par.settings=RdBuTheme, 
          names.attr=c("2005", "2012"),
          pretty=TRUE,
          layout=c(2,1),
          main='EVI Anomaly', xlab=NULL, ylab=NULL)

print(lp)

tiff(filename=paste0(di, '/out/raster_maps/anomaly_2005_2012_a.tiff'), 
     height = 1600, width = 2500, res=150, pointsize = 20, bg='transparent')
print(lp)
dev.off()

exportpdf(mypdf=paste0(di, '/out/raster_maps/anomaly_2005_2012_a.pdf'), lp) 
```

### Normalized Anomaly  
```{r map_anomaly_2005_2012_nora}
# Select a palette http://colorbrewer2.org/
mypal <- brewer.pal(11, "RdYlGn")
# Specify the color palette
myTheme=rasterTheme(region=mypal)

lp <- levelplot(stack_anomalias, 
          margin=FALSE,
          layer=c('nora_2005', 'nora_2012'),
          #par.settings=myTheme, 
          par.settings=RdBuTheme, 
          names.attr=c("2005", "2012"),
          pretty=TRUE,
          layout=c(2,1),
          main='EVI Normalized Anomaly', xlab=NULL, ylab=NULL)

print(lp)

tiff(filename=paste0(di, '/out/raster_maps/anomaly_2005_2012_nora.tiff'), 
     height = 1600, width = 2500, res=150, pointsize = 20, bg='transparent')
print(lp)
dev.off()

exportpdf(mypdf=paste0(di, '/out/raster_maps/anomaly_2005_2012_nora.pdf'), lp) 
```

### Standardized Anomaly  
```{r map_anomaly_2005_2012_sa}
# Select a palette http://colorbrewer2.org/
mypal <- brewer.pal(11, "RdYlGn")
# Specify the color palette
myTheme=rasterTheme(region=mypal)

lp <- levelplot(stack_anomalias, 
          margin=FALSE,
          layer=c('sa_2005', 'sa_2012'),
          #par.settings=myTheme, 
          par.settings=RdBuTheme, 
          names.attr=c("2005", "2012"),
          pretty=TRUE,
          layout=c(2,1),
          main='EVI Standardized Anomaly', xlab=NULL, ylab=NULL)

print(lp)

tiff(filename=paste0(di, '/out/raster_maps/anomaly_2005_2012_sa.tiff'), 
     height = 1600, width = 2500, res=150, pointsize = 20, bg='transparent')
print(lp)
dev.off()

exportpdf(mypdf=paste0(di, '/out/raster_maps/anomaly_2005_2012_sa.pdf'), lp) 
```

### Analisis de diferencias year y clu para las anomalias

* Preparo un dataframe. Solamente selecciono los años 2005 y 2012 (distrub year)

```{r}
compara_anoma <- anomalias_evimean %>% 
  filter(y %in% c(2005,2012)) %>% 
  mutate(y = as.factor(y)) %>% 
  dplyr::rename(disturb_year = y, 
         site = clu_pop2)
```

* Incluyo dos funciones customizadas para analizar las anovas y posthoc comparison

```{r}
# Custom Function to compute ANOVAS
aovas <- function(df, vars, resp_var){ 
  require('dplyr')
  require('broom')

  # Create subset 
  dfsel <- df %>% dplyr::select_(.dots=c(vars, resp_var))

  # Model 
  myformula <- as.formula(paste0(resp_var,  " ~ ",
                                 paste(vars, collapse = '*')))

  mymodel <- aov(myformula, data=dfsel)

  # Output model Summary http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1)
  model_coeff <- broom::tidy(mymodel)
  model_summary <- broom::glance(mymodel)

  out <- c() 
  out$model_coeff <- model_coeff
  out$model_summary <- model_summary
  out$mymodel <- mymodel

  return(out)
}

# Post-Hoc comparison
phc <- function(mymodel, resp_var){
  require(lsmeans)

  # Disturb Event 
  ph_event <- lsmeans(mymodel, pairwise ~ disturb_year, adjust = "bon")

  # differences letters 
  cld_event <- cld(ph_event, alpha   = 0.01, 
                   Letters = letters, 
                   adjust  = "bon")

  # Site  
  ph_site <- lsmeans(mymodel, pairwise ~ site, adjust = "bon")
  cld_site <- cld(ph_site, alpha   = 0.01, 
                 Letters = letters, 
                 adjust  = "bon")

  # interaction 
  ph_i <- lsmeans(mymodel, pairwise ~ disturb_year:site, adjust = "bon")
  cld_i <- cld(ph_i, alpha   = 0.01, 
                 Letters = letters, 
                 adjust  = "bon")

  # Objets for plot
  aux_ph_site <- as.data.frame(summary(ph_site$lsmeans)) 
  aux_ph_site <- aux_ph_site %>% mutate(var = resp_var)
  aux_ph_event <- as.data.frame(summary(ph_event$lsmeans)) 
  aux_ph_event <- aux_ph_event %>% mutate(var = resp_var)
  aux_ph_i <- as.data.frame(summary(ph_i$lsmeans)) 
  aux_ph_i <- aux_ph_i %>% mutate(var = resp_var)

  # Return objects
  cat('\n### Event ###\n')
  print(ph_event)
  print(cld_event)
  cat('\n### Clu pop ###\n')
  print(ph_site)
  print(cld_site)
  cat('\n### Event:Clu pop ###\n')
  print(ph_i)
  return(list(aux_ph_site, aux_ph_event, aux_ph_i, cld_site, cld_event, cld_i))
}
```

#### Compara Anomaly (a)
```{r}
resp_var <- 'a'
vars <- c('disturb_year','site')

# AOV
aov_a <- aovas(compara_anoma, vars=vars, resp_var = resp_var)

mc <- aov_a$model_coeff

pander(mc, round=5,
       caption = paste0("ANOVA table: ", resp_var), missing = '', 
       emphasize.strong.cells = 
         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))

gm <- aov_a$model_summary

gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
rownames(gm) <- "Statistic"
pander(t(gm))

# Post hoc Define model
mymodel <- aov_a$mymodel
postH_a <- phc(mymodel = mymodel, resp_var = resp_var)
```

```{r}
#### ~ Site
ps <- plot(effect("site",mymodel))
#### ~ Disturb Year
pd <- plot(effect('disturb_year', mymodel))
#### Disturb Year:Site
picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
pi <- plot(effect("disturb_year:site",mymodel), layout=c(3,1))

```

```{r aov_anomaly_site}
ps
```

```{r aov_anomaly_disturb}
pd
```

```{r aov_anomaly_inter_same_plot}
picollapse
```

```{r aov_anomaly_inter}
pi
```

#### Compara Normalized Anomaly (nora)
```{r}
resp_var <- 'nora'
vars <- c('disturb_year','site')

# AOV
aov_nora <- aovas(compara_anoma, vars=vars, resp_var = resp_var)

mc <- aov_nora$model_coeff

pander(mc, round=5,
       caption = paste0("ANOVA table: ", resp_var), missing = '', 
       emphasize.strong.cells = 
         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))

gm <- aov_nora$model_summary

gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
rownames(gm) <- "Statistic"
pander(t(gm))

# Post hoc Define model
mymodel <- aov_nora$mymodel
postH_nora <- phc(mymodel = mymodel, resp_var = resp_var)
```

```{r}
#### ~ Site
ps <- plot(effect("site",mymodel))
#### ~ Disturb Year
pd <- plot(effect('disturb_year', mymodel))
#### Disturb Year:Site
picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
pi <- plot(effect("disturb_year:site",mymodel), layout=c(3,1))

```

```{r aov_nora_site}
ps
```

```{r aov_nora_disturb}
pd
```

```{r aov_nora_inter_same_plot}
picollapse
```

```{r aov_nora_inter}
pi
```

#### Compara Standardized Anomaly (sa)
```{r}
resp_var <- 'sa'
vars <- c('disturb_year','site')

# AOV
aov_sa <- aovas(compara_anoma, vars=vars, resp_var = resp_var)

mc <- aov_sa$model_coeff

pander(mc, round=5,
       caption = paste0("ANOVA table: ", resp_var), missing = '', 
       emphasize.strong.cells = 
         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))

gm <- aov_sa$model_summary

gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
rownames(gm) <- "Statistic"
pander(t(gm))

# Post hoc Define model
mymodel <- aov_sa$mymodel
postH_sa <- phc(mymodel = mymodel, resp_var = resp_var)
```

```{r}
#### ~ Site
ps <- plot(effect("site",mymodel))
#### ~ Disturb Year
pd <- plot(effect('disturb_year', mymodel))
#### Disturb Year:Site
picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
pi <- plot(effect("disturb_year:site",mymodel), layout=c(3,1))

```

```{r aov_sa_site}
ps
```

```{r aov_sa_disturb}
pd
```

```{r aov_sa_inter_same_plot}
picollapse
```

```{r aov_sa_inter}
pi
```

#### Plot de trayectoria de las anomalías 
* Idea de plot de la media y la varianza (Trayectoria). Relacionado con las métricas de estabilidad (com. personal. Paula Escribano).

* Referencias intersantes para las trayectorias:

    * Zimmermann, N. E., R. A. Washington-Allen, R. D. Ramsey, M. E. Schaepman, L. Mathys, B. Koetz, M. Kneubuehler, and T. C. Edwards. 2007. Modern remote sensing for environmental monitoring of landscape states and trajectories. Pages 65-91 in F. Kienast, O. Wildi, and S. Ghosh, editors. A changing world: challenges for landscape research. Springer, Dordrecht, The Netherlands.
    * Washington-Allen, R. A., R. D. Ramsey, N. E. West, and B. E. Norton. 2008. Quantification of the ecological resilience of drylands using digital remote sensing. Ecology and Society 13(1): 33.  http://www.ecologyandsociety.org/vol13/iss1/art33/

```{r trajectories_anomalies}
traj <- anomalias_evimean %>% group_by(pop, y) %>% 
  dplyr::summarise(mean = mean(a),
            sd = sd(a))

traj_mean <- traj %>% 
  group_by(y) %>% 
  dplyr::summarise(meanOfmean = mean(mean), 
            sdOfmean = sd(mean),
            seOfmean = sdOfmean/sqrt(length(mean)),
            meanOfsd = mean(sd),
            sdOfsd = sd(sd),
            seOfsd = sdOfsd /sqrt(length(sd))) %>% 
  as.data.frame()

traj_plot <- traj_mean%>% 
  ggplot(aes(x=meanOfmean, y=meanOfsd, label=y)) +
  geom_errorbar(aes(ymin=meanOfsd - seOfsd, ymax=meanOfsd + seOfsd)) + 
  geom_errorbarh(aes(xmin=meanOfmean - seOfmean, xmax=meanOfmean + seOfmean)) + 
  geom_path(colour='gray') +
  geom_point(size=3, shape=21, fill='white') + 
  geom_text(hjust = 0.001, nudge_x = 0.001) + 
  geom_vline(aes(xintercept = mean(meanOfmean)), colour='red') +
  geom_hline(aes(yintercept = mean(meanOfsd)), colour ='red')+
  theme_bw() + xlab('mean') + ylab('variance') + 
  theme(strip.background = element_rect(fill = "white"), 
        legend.position="none")

traj_plot
pdf(file=paste0(di, "/out/trajectories/plot_trajectories_anomalies.pdf"), height = 6, width =6)
traj_plot
dev.off()
```

```{r}
traj_mean_pop <- traj %>% 
  mutate(clu_pop = as.factor(ifelse(pop %in% c(1,2,3,4,5), 'N', 'S'))) %>% 
  group_by(clu_pop,y) %>% 
  dplyr::summarise(meanOfmean = mean(mean), 
            sdOfmean = sd(mean),
            seOfmean = sdOfmean/sqrt(length(mean)),
            meanOfsd = mean(sd),
            sdOfsd = sd(sd),
            seOfsd = sdOfsd /sqrt(length(sd))) %>%
  as.data.frame()

line_traj_mean_pop <- traj_mean_pop %>% 
  group_by(clu_pop) %>% 
  dplyr::summarise(
    meanOfmean = mean(meanOfmean), 
    meanOfsd = mean(meanOfsd)
  )

traj_plot_pop <- ggplot(traj_mean_pop,
  aes(x=meanOfmean, y=meanOfsd, label=y)) +
  geom_errorbar(aes(ymin=meanOfsd - seOfsd, ymax=meanOfsd + seOfsd)) + 
  geom_errorbarh(aes(xmin=meanOfmean - seOfmean, xmax=meanOfmean + seOfmean)) + 
  geom_path(colour='gray') +
  geom_point(size=3, shape=21, fill='white') +
  geom_text(hjust = 0.001, nudge_x = 0.001) + 
  facet_wrap(~clu_pop) +
  theme_bw() + xlab('mean') + ylab('variance') + 
  theme(strip.background = element_rect(fill = "white"), 
        legend.position="none")

traj_plot_pop <- 
  traj_plot_pop + 
  geom_vline(aes(xintercept = meanOfmean), line_traj_mean_pop,  colour='red') +
  geom_hline(aes(yintercept = meanOfsd), line_traj_mean_pop, colour ='red')

traj_plot_pop
pdf(file=paste0(di, "/out/trajectories/plot_trajectories_anomalies_clu.pdf"), height = 6, width =10)
traj_plot_pop
dev.off()
```

# Notas que podemos obtener de explorar las anomalías

## Del análisis del plot de trayectorias

* 2005 supuso una disminución del greenness (el promedio de las anomalías es muy bajo respecto a los otros años), y esta disminución es bastante homogénea para todos los pixeles.

* Aunque 2012 presentó anomalías negativas (disminuición del greenness), su magnitud promedio (es decir el promedio para todos los pixeles) fue inferior al promedio observado para los años 2006, 2007 y 2000. Nótese que los años 2006 y 2007 son los posteriores a la gran sequía de 2005.

* Un análisis detallado entre las poblaciones del N y del S, refleja un patron simliar respecto a 2005, esto es, es el año en el que el promedio de las anomalías fue el mas bajo. Con 2012 ocurre algo similar. Llama la atención, que para 2005 sin embargo el patron fue mas heterogéno en las poblaciones del sur, esto es, mucha varibilidad en las anomalías de las poblaciones situadas en el sur.

## Explorando las anomalías

* Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005. Sin embargo vemos valores menos negativos para 2005. :red_circle: Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years :red_circle:

    * Las anomalías (sa) fueron significativamente menores en 2005 (`r round(postH_sa[[5]][1,2],3)` ± `r round(postH_sa[[5]][1,3],3)`)que para 2012 (`r round(postH_sa[[5]][2,2],3)` ± `r round(postH_sa[[5]][2,3],3)`), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)

* Si utilizamos el criterio de Samanta et al. 2010, de las anomalias estandarizadas, vemos que solamente ocurre bronwning en el evento de sequía de 2005, ya que para el resto de la serie todos los valores de las anomalías están dentro -1 y 1. Además observamos que esto ocurre en ambas cluster de poblaciones (N y S), siendo mas intensa la reducción en las poblaciones del N.

* Explorando las anomalias entre N y S, vemos que en 2005 la reducción fue mayor para las poblaciones del N. Este se ve muy bien sobre todo en los mapas, en las que vemos que para 2005, existe una mayor concentración de valores negativos de anomalías en las poblaciones de N. Asimismo, vemos que las poblaciones del S, muestran valores heterogéneos de anomalías en 2005. Esto concuerda con lo observado en el análisis de trayectoria.

    * También observamos anomalías (sa) significativamente menores en para las poblaciones del N (`r round(postH_sa[[4]][1,2],3)` ± `r round(postH_sa[[4]][1,3],3)`) que para las del S (`r round(postH_sa[[4]][2,2],3)` ± `r round(postH_sa[[4]][2,3],3)`), (LSMEANS, t.ratio = -7.056; p_value < 0.0001)

### 2017-11-20 analysis/evi/explore_anomalies.Rmd
update explore anomalies evi. adding png

@@
 
+plot_a_clu
 pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_clu_a.pdf"), height = 7, width = 7)
@@
 
+plot_nora_clu 
 pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_clu_nora.pdf"), height = 7, width = 7)
@@
 
+plot_sa_clu
 pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_clu_sa.pdf"), height = 7, width = 7)
@@
 
+plot_a_pop
 pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_a.pdf"), height = 14, width = 14)
@@
 
+plot_nora_pop 
 pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_nora.pdf"), height = 14, width = 14)
@@
 
+plot_sa_pop
 pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_sa.pdf"), height = 14, width = 14)

### 2017-11-20 pres/state_ms.Rmd
update pres

--- 2017-11-13 pres/state_ms.Rmd
@@
     * Índices de competencia dependientes e independientes de la distancia [@GeaIzquierdo2009] 
+    * [\textcolor{blue}{Resultados}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/analysis_competence.md)
 
@@
 
-# Dendrochronological methods
+![](../analysis/analysis_competence_files/figure-markdown_github/competence_sj-1.png)
+
+---
+
+![](../analysis/analysis_competence_files/figure-markdown_github/competence_ca-1.png)
+
+---
+
+# Resilience metrics
+
+To evaluate the effects of the disturbance events on greeennes and tree growth we used four resilience indices [@Lloret2011]
+
+* **Resistance (*Rt*)** quantifies the severity of the impact of the disturbance in the year it occurred. $$Resistance = Drought / Predrought$$
+* **Recovery index (*Rc*)** is the ability to recover from disturbance relative to its severity. 
+$$Recovery  = Postdrought / Drought$$
+* **Resilience index (*Rs*)** is the capacity to reach pre-disturbance performance levels. 
+$$Resilience = Postdrought / Predrought$$
+* **Relative Resilience (*RRs*)** is the resilience weighted by the severity of the disturbance. 
+$$Relative Resilience = (Postdrought - Drought) / Predrought$$
+
+---
+
+* Se calcularon los valores de cada índice de resilience para las variables EVI medio y Tree growth en cada evento de sequía 
+* Consideramos 2005 y 2012 como dos eventos de sequía
+* Los valores *PreDrought* y *PostDrought* de cada variable (EVI medio y tree growth) se calcularon como la media durante un periodo de tres años antes y despues respectivamente del evento de sequía.
+Probamos con periodos de 2,3 y 4 años y obtuvimos resultados similares (Elegimos 3 años)
+
+    * Ver [\textcolor{blue}{gráficas}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/analysis_resilience_bai.md) para BAI
+    * Ver [\textcolor{blue}{gráficas}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/evi/compute_resilience_evi.md) para EVI
+
+---
+
+![](../analysis/explore_correlations_windows_size_files/figure-markdown_github/correla_window_bai-1.png)
+
+---
+
+![](../analysis/explore_correlations_windows_size_files/figure-markdown_github/correla_window_evi-1.png)
+
+---
+
+# Analyses
+
+---
+
+# R - Explore EVI mean
+
+![](../analysis/evi/explore_evi_mean_files/figure-markdown_github/evi_profile_compara-1.png)
+
+---
+
+# R - Plot de trayectorias: general y por cluster pops (ojito con esto)
+
+![](../analysis/evi/explore_evi_mean_files/figure-markdown_github/trajectories_evi-1.png)
+
+---
+
+# R - Plot de trayectorias: general y por cluster pops (ojito con esto)
+
+![](../analysis/evi/explore_evi_mean_files/figure-markdown_github/trajectories_evi_clu-1.png)
+
+---
+
+# R - Mann Kendall trends in EVI mean
+
+![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
+
+---
+
+# R - Mann Kendall trends in EVI mean
+
+![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/trend_sig_pixels-1.png){width=150%}
+
+---
+
+# R- Mann Kendall trends in EVI mean
+
+![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/raster_map_sen_evi-1.png){width=150%}
+
+---
+
+# R - EVI anomalies
+
+![](../analysis/evi/explore_anomalies_files/figure-markdown_github/plot_anoma_EVI_clu_sa-1.png){width=90%}
+
+---
+
+# Notas anomalias para incluir
+
+https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/evi/explore_anomalies.md
+
+---
+
+# Mapa de anomalias estandarizadas
+
+![](../analysis/evi/explore_anomalies_files/figure-markdown_github/map_anomaly_2005_2012_sa-1.png){width=90%}
+
+---
 

### 2017-11-21 analysis/evi/explore_resilience.Rmd
update resilience from old repo qpyr_modis_resilience

--- 2017-09-15 analysis/analysis_resilience_bai.Rmd
@@
 ---
-title: "Growth Resilience"
+title: "Explore resilience evi"
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 June"
+date: "2017 Aug"
 output:
@@
     variant: markdown_github
-bibliography: ../references.bib 
-csl: ../ecology.csl
 ---
-
-```{r global_options, include=FALSE}
-knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, error=TRUE, message=FALSE)
-```
-
-```{r, warning=FALSE, message=FALSE}
-library("tidyverse")
-library("dplR")
-library("stringr")
-library("knitr")
-library('gtable')
-library('grid')
+```{r, echo=FALSE, message=FALSE}
+require(knitr)
+opts_chunk$set(fig.align='center', message = FALSE, warning = FALSE) 
+```
+
+```{r wd, echo=FALSE}
+#---------------------------------
+machine <- 'ajpelu'
+#machine <- 'ajpeluLap'
+di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro', sep='')
+#---------------------------------
+```
+
+```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
+library('dplyr')
+library('tidyverse')
+library('broom')
+library('pander')
+library('lsmeans')
+library('effects')
 library('gridExtra')
-library('pander')
-library('broom')
-library('effects')
+library('WRS2')
+library('car')
+library('MASS')
+library('rcompanion')
 library('devtools')
 # devtools::install_github("ajpelu/auxiliar")
-library('car')
-library('auxiliar')
-library('WRS2')
-library('MASS')
-library('rcompanion')
-```
-# Resilience
-
-* Calcularemos las métricas resiliencia de [@Lloret2011] sobre el crecimiento. 
-* Vamos a calcularlas sobre el BAI de cada árbol. 
-* Utilizaremos tres sitios: SJ, CAH y CAL (ver [./analysis/analysis_chronologies.md]('./analysis/analysis_chronologies.md))
-
-## Prepare data
-
-* Leer datos `rwl` de SJ y CA
-* Leer datos de diametros de los focal tree
-
-```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
-
-# sj 
-sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
-
-# canar 
-ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
-
-# Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-```
-
-```{r}
-source(paste0(di, 'script/R/rw_byTree.R'))
-source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/baiResilience.R'))
-```
-
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low
-
-```{r}
-# Replace SNA by SJ and SNB by CA
-names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
-names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")
-
-# Remove g in name of some cores of CA. 
-names(ca) <- stringr::str_replace(names(ca), "g", "")
-```
-
-```{r}
-# Create subset to compare between sites 
-caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
-             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
-             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
-caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
-              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
-              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
-
-```
-
-* Lectura y preparación de datos de diámetro
-
-```{r} 
-# Prepare Diameter data
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Change name focal according to loc
-compete <- compete %>% 
-  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm)
-
-# Set levels of eleveation 
-ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
-            paste0('CA', 26:30))
-ca_highcode <- paste0('CA', 11:25)
-
-ft <- ft %>% 
-  mutate(site = as.factor(
-    ifelse(id_focalLoc %in% ca_lowcode, 'CAL', 
-           ifelse(id_focalLoc %in% ca_highcode, 'CAH', 'SJ'))))
-```
-
-# Aggregate RW by tree
-
-* Agregar valores medios de RW por site (obtenemos sj_tree / caL_tree, caH_tree) 
-* ver fun rw_byTree o utilizar treeMean (dplR)
-
-```{r}
-# Remember snc = structure of core name SJ0101 (site | tree | core)
-sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
-caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
-caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')
-```
-
-* Crear diferentes dataset de diametro por sitio 
-```{r}
-diam <- ft %>%
-  mutate(diameter = dn_mm, 
-         id = id_focalLoc) %>%
-  dplyr::select(id, diameter, site) %>% 
-  split(.$site)
-
-d_caH <- diam$CAH[,c('id','diameter')]
-d_caL <- diam$CAL[,c('id','diameter')]
-d_sj <- diam$SJ[,c('id','diameter')]
-```
-
-## Cómputo del BAI por site
-
-* He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]. Es similar a `bai.out`
-
-```{r}
-bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
-bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
-bai_caL <- bai_piovesan(rwdf = caL_tree, diam_df = d_caL)
-
-# Set class to bai object 
-# Esto es para que funcionen algunas otras funciones de dplR 
-bais <- c('bai_sj', 'bai_caH', 'bai_caL')
-
-for (i in bais){
-  aux <- get(i)
-
-  class(aux) <- c('rwl', 'data.frame')
-
-  assign(i, aux)
+```
+
+# Prepare data
+
+```{r}
+# Read data
+raw_evires <- read.csv(file=paste(di, "/data/resilience/resilience_evi.csv", sep=""), header = TRUE, sep = ',')
+
+# add data of pop
+anomalias <- read.csv(file=paste(di, "/data/anomalies/anomalias_evimean.csv", sep=""), header = TRUE, sep = ',')
+
+attr_iv_malla_modis_id <- anomalias %>% dplyr::select(iv_malla_modi_id,long,lat,pop) %>% unique()
+
+raw_evires <- raw_evires %>% inner_join(attr_iv_malla_modis_id, by='iv_malla_modi_id')
+
+# filter by pop and add new variable
+
+evires <- raw_evires %>% 
+  mutate(
+    clu_pop = as.factor(case_when(
+      pop == 1 ~ "Camarate",
+      pop %in% c(2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
+      pop == 9 ~ 'out')),
+    clu_pop2 = as.factor(case_when(
+      pop %in% c(1,2,3,4,5) ~ 'N',
+      pop %in% c(6,7,8) ~ 'S',
+      pop == 9 ~ 'out'))) %>% 
+  filter(clu_pop != 'out')
+
+# Change name of clu_pop2 and disturb_year para los analisis anovas
+evires <- evires %>% dplyr::rename(site = clu_pop2) %>% 
+  mutate(disturb_year = as.factor(disturb_year))
+
+```
+
+# ANOVAS
+
+```{r, echo=FALSE}
+# Custom Function to compute ANOVAS
+aovas <- function(df, vars, resp_var){ 
+  require('dplyr')
+  require('broom')
+
+  # Create subset 
+  dfsel <- df %>% dplyr::select_(.dots=c(vars, resp_var))
+
+  # Model 
+  myformula <- as.formula(paste0(resp_var,  " ~ ",
+                                 paste(vars, collapse = '*')))
+
+  mymodel <- aov(myformula, data=dfsel)
+
+  # Output model Summary http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1)
+  model_coeff <- broom::tidy(mymodel)
+  model_summary <- broom::glance(mymodel)
+
+  out <- c() 
+  out$model_coeff <- model_coeff
+  out$model_summary <- model_summary
+  out$mymodel <- mymodel
+
+  return(out)
 }
 
+# Post-Hoc comparison
+phc <- function(mymodel, resp_var){
+  require(lsmeans)
+
+  # Disturb Event 
+  ph_event <- lsmeans(mymodel, pairwise ~ disturb_year, adjust = "bon")
+
+  # differences letters 
+  cld_event <- cld(ph_event, alpha   = 0.01, 
+                   Letters = letters, 
+                   adjust  = "bon")
+
+  # Site  
+  ph_site <- lsmeans(mymodel, pairwise ~ site, adjust = "bon")
+  cld_site <- cld(ph_site, alpha   = 0.01, 
+                 Letters = letters, 
+                 adjust  = "bon")
+
+  # interaction 
+  ph_i <- lsmeans(mymodel, pairwise ~ disturb_year:site, adjust = "bon")
+  cld_i <- cld(ph_i, alpha   = 0.01, 
+                 Letters = letters, 
+                 adjust  = "bon")
+
+  # Objets for plot
+  aux_ph_site <- as.data.frame(summary(ph_site$lsmeans)) 
+  aux_ph_site <- aux_ph_site %>% mutate(var = resp_var)
+  aux_ph_event <- as.data.frame(summary(ph_event$lsmeans)) 
+  aux_ph_event <- aux_ph_event %>% mutate(var = resp_var)
+  aux_ph_i <- as.data.frame(summary(ph_i$lsmeans)) 
+  aux_ph_i <- aux_ph_i %>% mutate(var = resp_var)
+
+  # Return objects
+  cat('\n### Event ###\n')
+  print(ph_event)
+  print(cld_event)
+  cat('\n### Clu pop ###\n')
+  print(ph_site)
+  print(cld_site)
+  cat('\n### Event:Clu pop ###\n')
+  print(ph_i)
+  return(list(aux_ph_site, aux_ph_event, aux_ph_i, cld_site, cld_event, cld_i))
+}
+```
+
+## Recovery
+
+```{r, echo=FALSE}
+resp_var <- 'rc'
+vars <- c('disturb_year','site')
+
+# AOV
+aov_rc <- aovas(evires, vars=vars, resp_var = resp_var)
+
+mc <- aov_rc$model_coeff
+
+pander(mc, round=5,
+       caption = paste0("ANOVA table: ", resp_var), missing = '', 
+       emphasize.strong.cells = 
+         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
+
+gm <- aov_rc$model_summary
+
+gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
+colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
+rownames(gm) <- "Statistic"
+pander(t(gm)) 
+```
+
+```{r}
+# Post hoc Define model
+mymodel <- aov_rc$mymodel
+postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
+```
+
+```{r, echo=FALSE}
+#### ~ Site
+ps <- plot(effect("site",mymodel))
+#### ~ Disturb Year
+pd <- plot(effect('disturb_year', mymodel))
+#### Disturb Year:Site
+picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
+
+```
+
+```{r aov_rc_site, fig.width = 4, fig.height=4}
+ps
+```
+
+```{r aov_rc_disturb, fig.width = 4, fig.height=4}
+pd
+```
+
+```{r aov_rc_inter_collapse, fig.width = 4, fig.height=4}
+picollapse
+```
+
+```{r aov_rc_inter, fig.width = 6}
+pi
+```
+
+## Resistance 
+```{r, echo=FALSE}
+# Variable
+resp_var <- 'rt'
+
+vars <- c('disturb_year','site')
+
+# AOV
+aov_rt <- aovas(evires, vars=vars, resp_var = resp_var)
+
+mc <- aov_rt$model_coeff
+
+pander(mc, round=5,
+       caption = paste0("ANOVA table: ", resp_var), missing = '', 
+       emphasize.strong.cells = 
+         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
+
+gm <- aov_rt$model_summary
+
+gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
+colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
+rownames(gm) <- "Statistic"
+pander(t(gm)) 
+```
+
+```{r}
+# Post hoc Define model
+mymodel <- aov_rt$mymodel
+postH_rt <- phc(mymodel = mymodel, resp_var = resp_var)
+```
+
+```{r, echo=FALSE}
+#### ~ Site
+ps <- plot(effect("site",mymodel))
+#### ~ Disturb Year
+pd <- plot(effect('disturb_year', mymodel))
+#### Disturb Year:Site
+picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
+
+```
+
+```{r aov_rt_site, fig.width = 4, fig.height=4}
+ps
+```
+
+```{r aov_rt_disturb, fig.width = 4, fig.height=4}
+pd
+```
+
+```{r aov_rt_inter_collapse, fig.width = 4, fig.height=4}
+picollapse
+```
+
+```{r aov_rt_inter, fig.width = 6}
+pi
 ```
@@
 ## Resilience 
-
-* Computar métricas de resiliencia BAI para los tres sitios. 
-* Computar tres eventos climáticos: 1995, 2005, 2012
-* Computar dos ventanas temporales: 2 y 3
-
-```{r}
-# Drought years 
-dyears <- c(1995, 2005, 2012)
-
-# SJ 
-res_4_sj <- baiResilience(bai_sj, event_years = dyears, window = 4)
-res_3_sj <- baiResilience(bai_sj, event_years = dyears, window = 3)
-res_2_sj <- baiResilience(bai_sj, event_years = dyears, window = 2)
-
-# caL
-res_4_caL <- baiResilience(bai_caL, event_years = dyears, window = 4)
-res_3_caL <- baiResilience(bai_caL, event_years = dyears, window = 3)
-res_2_caL <- baiResilience(bai_caL, event_years = dyears, window = 2)
-
-# caH
-res_4_caH <- baiResilience(bai_caH, event_years = dyears, window = 4)
-res_3_caH <- baiResilience(bai_caH, event_years = dyears, window = 3)
-res_2_caH <- baiResilience(bai_caH, event_years = dyears, window = 2)
-```
-
-### Computar correlaciones ventanas temporales 
-```{r}
-# Vector with objects name
-obj <- c('res_2_sj', 'res_3_sj', 'res_4_sj',
-         'res_2_caL', 'res_3_caL', 'res_4_caL',
-         'res_2_caH', 'res_3_caH', 'res_4_caH')
-
-correla_ws <- c()
-
-for (i in obj){ 
-  x <- get(i)
-  xres <- x$resilience
-  out <- xres %>% 
-    mutate(ws = paste0('ws_', as.character(str_extract(i, "([0-9])"))),
-           site = str_replace(i, "res_[0-9]_", '')) %>%
-    dplyr::select(-disturb_year, -tree)
-
-  correla_ws <- bind_rows(correla_ws, out)
-
-}
-
-# Split by window size 
-correla <- correla_ws %>% split(.$ws)
-
-# Change names 
-names(correla[["ws_2"]])[1:4] <- paste0(names(correla[["ws_2"]])[1:4], '2')
-names(correla[["ws_3"]])[1:4] <- paste0(names(correla[["ws_3"]])[1:4], '3')
-names(correla[["ws_4"]])[1:4] <- paste0(names(correla[["ws_4"]])[1:4], '4')
-
-cor2 <- correla[["ws_2"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-cor3 <- correla[["ws_3"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-cor4 <- correla[["ws_4"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-
-correlations <- inner_join(cor2, cor3, by='ind') %>% inner_join(cor4, by='ind')
-
-# Resistance
-aux_coefs <- c()
-
-model <- lm(rt2~rt3, data=correlations)
-p_rt23 <- correlations %>% ggplot(aes(rt2, rt3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rt','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rt2~rt4, data=correlations)
-p_rt24 <- correlations %>% ggplot(aes(rt2, rt4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rt','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rt3~rt4, data=correlations)
-p_rt34 <- correlations %>% ggplot(aes(rt3, rt4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rt','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rt23, p_rt24, p_rt34,ncol=3)
-
-# Recovery 
-model <- lm(rc2~rc3, data=correlations)
-p_rc23 <- correlations %>% ggplot(aes(rc2, rc3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rc','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rc2~rc4, data=correlations)
-p_rc24 <- correlations %>% ggplot(aes(rc2, rc4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rc','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rc3~rc4, data=correlations)
-p_rc34 <- correlations %>% ggplot(aes(rc3, rc4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rc','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rc23, p_rc24, p_rc34,ncol=3)
-
-# Resilience
-model <- lm(rs2~rs3, data=correlations)
-p_rs23 <- correlations %>% ggplot(aes(rs2, rs3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rs','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rs2~rs4, data=correlations)
-p_rs24 <- correlations %>% ggplot(aes(rs2, rs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rs','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rs3~rs4, data=correlations)
-p_rs34 <- correlations %>% ggplot(aes(rs3, rs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rs','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rs23, p_rs24, p_rs34,ncol=3)
-
-# Relative Resilience
-model <- lm(rrs2~rrs3, data=correlations)
-p_rrs23 <- correlations %>% ggplot(aes(rrs2, rrs3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rrs','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rrs2~rrs4, data=correlations)
-p_rrs24 <- correlations %>% ggplot(aes(rrs2, rrs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rrs','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rrs3~rrs4, data=correlations)
-p_rrs34 <- correlations %>% ggplot(aes(rrs3, rrs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rrs','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rrs23, p_rrs24, p_rrs34,ncol=3)
-
-names(aux_coefs) <- c('var', 'window_size', 'r2')
-
-write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), row.names = F)
-```
-
-```{r}
-aux_coefs %>% pander()
-```
-
-Nos quedamos con 3 años de ventana temporal.
-
-# Plots Crecimiento
-## Boxplot with outliers 
-```{r}
-gsj <- res_3_sj$growth %>% mutate(site='SJ')
-gcaL <- res_3_caL$growth %>% mutate(site='caL')
-gcaH <- res_3_caH$growth %>% mutate(site='caH')
-
-g <- bind_rows(gsj, gcaL, gcaH)
-
-# Export csv 
-write.csv(g, file=paste0(di, 'data/resilience/crecimientos_drought.csv'), row.names = FALSE)
-
-g %>% mutate(disturb = dplyr::recode(disturb, 
-                              dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  ggplot(aes(y=mean_period/100, x=disturb)) + 
-  geom_boxplot() + 
-  facet_grid(site~disturb_year, scales='free_y') + 
-  theme_bw()
-```
-
-## Mean + se 
-```{r}
-g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  group_by(disturb, disturb_year, site) %>% 
-  summarise(mean = mean(mean_period),
-            sd = sd(mean_period),
-            se = sd/sqrt(length(mean_period))) %>%
-  ggplot(aes(y=mean/100, x=disturb)) + 
-  geom_errorbar(aes(ymin=mean/100 - se/100, 
-                    ymax=mean/100 + se/100),
-                width = 0.2) +
-  geom_point(size=2, shape=21, fill='white') + 
-  facet_grid(site~disturb_year, scales='free_y') + 
-  theme_bw() + ylab('BAI (cm2/year)') + 
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-pgrowht <- g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  group_by(disturb, disturb_year, site) %>% 
-  summarise(mean = mean(mean_period),
-            sd = sd(mean_period),
-            se = sd/sqrt(length(mean_period))) %>%
-  ggplot(aes(y=mean/100, x=disturb, colour=site)) + 
-  geom_errorbar(aes(ymin=mean/100 - se/100, 
-                    ymax=mean/100 + se/100),
-                width = 0.15) +
-  geom_point(size=3, aes(shape=site), fill='white') + 
-  geom_line(aes(group=site))+
-  facet_wrap(~disturb_year, scales='free_y') + 
-  theme_bw() + ylab('BAI (cm2/year)') + 
-  theme(panel.grid = element_blank())
-
-pgrowht
-ggsave(plot=pgrowht, width=8, height = 4,
-       filename=paste0(di, 'out/fig/resilience/bai_events.pdf')) 
-```
-
-```{r}
-g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  dplyr::group_by(site, disturb_year, disturb) %>% 
-  dplyr::summarise(mean = mean(mean_period/100), 
-                   sd = sd(mean_period/100),
-                   se = sd/sqrt(length(mean_period/100))) %>% as.data.frame() %>% pander() 
-```
-
-# Anovas Resiliencia
-
-```{r}
-# Prepara data 
-rsj <- res_3_sj$resilience %>% mutate(site='SJ')
-rcaL<- res_3_caL$resilience %>% mutate(site='caL')
-rcaH <- res_3_caH$resilience %>% mutate(site='caH')
-
-re <- bind_rows(rsj, rcaL, rcaH)
-re$disturb_year <- as.factor(re$disturb_year)
-re$site <- as.factor(re$site)
-
-# Export csv 
-write.csv(re, file=paste0(di, 'data/resilience/resilience_bai.csv'), row.names = FALSE)
-```
-
-## Asumptions
-
-* Explorar si se cumplen los supuestos de normalidad y homocedasticidad. Tenemos que comprobar que cada uno de los grupos son normales (1995,2005,2012; site: SJ, CaH, CaL; e interactions)
+```{r, echo=FALSE}
+# Variable
+resp_var <- 'rs'
+
+vars <- c('disturb_year','site')
+
+# AOV
+aov_rs <- aovas(evires, vars=vars, resp_var = resp_var)
+
+mc <- aov_rs$model_coeff
+
+pander(mc, round=5,
+       caption = paste0("ANOVA table: ", resp_var), missing = '', 
+       emphasize.strong.cells = 
+         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
+
+gm <- aov_rs$model_summary
+
+gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
+colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
+rownames(gm) <- "Statistic"
+pander(t(gm)) 
+```
+
+```{r}
+# Post hoc Define model
+mymodel <- aov_rs$mymodel
+postH_rs <- phc(mymodel = mymodel, resp_var = resp_var)
+```
+
+```{r, echo=FALSE}
+#### ~ Site
+ps <- plot(effect("site",mymodel))
+#### ~ Disturb Year
+pd <- plot(effect('disturb_year', mymodel))
+#### Disturb Year:Site
+picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
+
+```
+
+```{r aov_rs_site, fig.width = 4, fig.height=4}
+ps
+```
+
+```{r aov_rs_disturb, fig.width = 4, fig.height=4}
+pd
+```
+
+```{r aov_rs_inter_collapse, fig.width = 4, fig.height=4}
+picollapse
+```
+
+```{r aov_rs_inter, fig.width = 6}
+pi
+```
+
+## Relative Resilience
+```{r, echo=FALSE}
+# Variable
+resp_var <- 'rrs'
+
+vars <- c('disturb_year','site')
+
+# AOV
+aov_rrs <- aovas(evires, vars=vars, resp_var = resp_var)
+
+mc <- aov_rrs$model_coeff
+
+pander(mc, round=5,
+       caption = paste0("ANOVA table: ", resp_var), missing = '', 
+       emphasize.strong.cells = 
+         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
+
+gm <- aov_rrs$model_summary
+
+gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
+colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
+rownames(gm) <- "Statistic"
+pander(t(gm)) 
+```
+
+```{r}
+# Post hoc Define model
+mymodel <- aov_rrs$mymodel
+postH_rrs <- phc(mymodel = mymodel, resp_var = resp_var)
+```
+
+```{r, echo=FALSE}
+#### ~ Site
+ps <- plot(effect("site",mymodel))
+#### ~ Disturb Year
+pd <- plot(effect('disturb_year', mymodel))
+#### Disturb Year:Site
+picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
+pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
+
+```
+
+```{r aov_rrs_site, fig.width = 4, fig.height=4}
+ps
+```
+
+```{r aov_rrs_disturb, fig.width = 4, fig.height=4}
+pd
+```
+
+```{r aov_rrs_inter_collapse, fig.width = 4, fig.height=4}
+picollapse
+```
+
+```{r aov_rrs_inter, fig.width = 6}
+pi
+```
+
+```{r, echo=FALSE}
+means_site <- postH_rc[[4]] %>% mutate(var = 'rc') %>% 
+  bind_rows(postH_rt[[4]] %>% mutate(var = 'rt')) %>% 
+  bind_rows(postH_rs[[4]] %>% mutate(var = 'rs')) %>% 
+  bind_rows(postH_rrs[[4]] %>% mutate(var = 'rrs')) %>% 
+  dplyr::rename(letras = .group)
+
+means_disturb <- postH_rc[[5]] %>% mutate(var = 'rc') %>% 
+  bind_rows(postH_rt[[5]] %>% mutate(var = 'rt')) %>% 
+  bind_rows(postH_rs[[5]] %>% mutate(var = 'rs')) %>% 
+  bind_rows(postH_rrs[[5]] %>% mutate(var = 'rrs')) %>% 
+  dplyr::rename(letras = .group)
+
+means_distub_site <- postH_rc[[6]] %>% mutate(var = 'rc') %>% 
+  bind_rows(postH_rt[[6]] %>% mutate(var = 'rt')) %>% 
+  bind_rows(postH_rs[[6]] %>% mutate(var = 'rs')) %>% 
+  bind_rows(postH_rrs[[6]] %>% mutate(var = 'rrs')) %>% 
+  dplyr::rename(letras = .group)
+```
+
+```{r, echo=FALSE}
+dodge <- position_dodge(width = 0.3)
+micolor <- '#455883'
+mierrorbarSE <- aes(ymin=lsmean - SE, ymax=lsmean + SE)
+mierrorbar <- aes(ymin=lower.CL, ymax=upper.CL)
+```
+
+```{r, echo=FALSE}
+plot_ms <- means_site %>%  
+  ggplot(aes(x=site, y=lsmean)) + 
+  geom_point(colour=micolor, 
+             size=3, position = dodge) +
+  theme_bw() + xlab('') + ylab('') + 
+  facet_wrap(~var, scales='free_y', ncol = 1) +
+  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15) +
+  theme(strip.background = element_rect(colour = "black", fill = "white"))
+
+plot_msSE <- plot_ms + geom_errorbar(mierrorbarSE,color=micolor, 
+                                     size=.5, width=.15, position = dodge)
+
+plot_msCI <- plot_ms + geom_errorbar(mierrorbar,color=micolor, 
+                                     size=.5, width=.15, position = dodge)
+```
+
+```{r, echo=FALSE}
+plot_md <- means_disturb %>%  
+  ggplot(aes(x=disturb_year, y=lsmean)) + 
+  geom_point(colour=micolor, 
+             size=3, position = dodge) +
+  theme_bw() + xlab('') + ylab('') + 
+  facet_wrap(~var, scales='free_y', ncol = 1) +
+  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15) +
+  theme(strip.background = element_rect(colour = "black", fill = "white"))
+
+plot_mdSE <- plot_md + geom_errorbar(mierrorbarSE,color=micolor, 
+                                     size=.5, width=.15, position = dodge)
+
+plot_mdCI <- plot_md + geom_errorbar(mierrorbar,color=micolor, 
+                                     size=.5, width=.15, position = dodge)
+```
+
+```{r, echo=FALSE}
+plot_mds <- means_distub_site %>%  
+  ggplot(aes(x=site, y=lsmean, group=disturb_year, colour=disturb_year)) + 
+  geom_point(aes(shape=disturb_year), size=3) + 
+  geom_line() +
+  theme_bw() + xlab('') + ylab('') + 
+  facet_wrap(~var, scales='free_y', ncol = 1) +
+  geom_text(aes(y=lsmean+SE, label=letras), nudge_x = 0.15)+
+  theme(strip.background = element_rect(colour = "black", fill = "white"),
+        legend.position = c(0.8, 0.93),
+        legend.background = element_blank()) +
+  scale_colour_manual(values = c(micolor, "red"))
+
+plot_mdsSE <- plot_mds + geom_errorbar(mierrorbarSE, size=.5, width=.15)
+plot_mdsCI <- plot_mds + geom_errorbar(mierrorbar, size=.5, width=.15)
+```
+
+### mean + sd
+```{r plot_resiliences_evi_sd,fig.width = 12, fig.height=12}
+grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
+```
+
+### mean + ci
+
+```{r plot_resiliences_evi_ci, fig.width = 12, fig.height=12}
+grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
+```
+
+```{r, echo=FALSE}
+pdf(paste0(di, '/out/mean_resilience/evi/interaction_plotsSE.pdf'), width=9, height = 9)
+grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
+dev.off()
+
+pdf(paste0(di, '/out/mean_resilience/evi/interaction_plotsCI.pdf'), width=9, height = 9)
+grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
+dev.off()
+```
+
+```{r, echo=FALSE}
+aovas_coeff <- aov_rc$model_coeff %>% mutate(var = 'rc') %>% 
+  bind_rows(aov_rt$model_coeff %>% mutate(var = 'rt')) %>% 
+  bind_rows(aov_rs$model_coeff %>% mutate(var = 'rs')) %>% 
+  bind_rows(aov_rrs$model_coeff%>% mutate(var = 'rrs'))
+
+write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/evi/anovas_statistics.csv'), row.names = F)
+
+aovas_coeff %>% pander()
+```
+
+```{r, echo=FALSE}
+aovas_model_summary <- aov_rc$model_summary %>% mutate(var = 'rc') %>% 
+  bind_rows(aov_rt$model_summary %>% mutate(var = 'rt')) %>% 
+  bind_rows(aov_rs$model_summary %>% mutate(var = 'rs')) %>% 
+  bind_rows(aov_rrs$model_summary%>% mutate(var = 'rrs'))
+
+write.csv(aovas_model_summary, 
+          file=paste0(di, '/out/anovas_resilience/evi/anovas_summary_modelos.csv'), row.names = F)
+
+gm <- apply(aovas_model_summary, 1, formatC, digits = 2, format = "f") 
+rownames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e", "variable"),"$")
+colnames(gm) <- c("rc", "rt", "rs", "rrs")
+
+pander(gm)
+```
+
+# Asumptions
 
@@
   out <- df %>%
-    group_by_(.dots=factor_vars) %>%
-    summarise(statistic = round(shapiro.test(!!rv)$statistic,5),
+    dplyr::group_by_(.dots=factor_vars) %>%
+    dplyr::summarise(statistic = round(shapiro.test(!!rv)$statistic,5),
               p_value = round(shapiro.test(!!rv)$p.value,5)) %>%
@@
 ```
+
+* Explorar si se cumplen los supuestos de normalidad y homocedasticidad. Tenemos que comprobar que cada uno de los grupos son normales (2005 vs 2012; N vs S; e interactions)
+
 ## Normalidad
@@
 ### Resilience 
-
-nrsA <- shapirosNormal(re, resp_var = rs, 'disturb_year')
+nrsA<- shapirosNormal(evires, resp_var = rs, 'disturb_year')
 nrsA$var <- 'rs'
-nrsB <- shapirosNormal(re, resp_var = rs, 'site')
+nrsB <- shapirosNormal(evires, resp_var = rs, 'site') 
 nrsB$var <- 'rs'
-nrsAB <- shapirosNormal(re, resp_var = rs, c('disturb_year','site'))
+nrsAB <- shapirosNormal(evires, resp_var = rs, c('disturb_year','site'))
 nrsAB$var <- 'rs'
@@
 ### Recovery
-nrcA <- shapirosNormal(re, resp_var = rc, 'disturb_year')
+nrcA <- shapirosNormal(evires, resp_var = rc, 'disturb_year')
 nrcA$var <- 'rc'
-nrcB <- shapirosNormal(re, resp_var = rc, 'site') 
+nrcB <- shapirosNormal(evires, resp_var = rc, 'site') 
 nrcB$var <- 'rc'
-nrcAB <- shapirosNormal(re, resp_var = rc, c('disturb_year','site'))
+nrcAB <- shapirosNormal(evires, resp_var = rc, c('disturb_year','site'))
 nrcAB$var <- 'rc'
@@
 ### Resistance
-nrtA <- shapirosNormal(re, resp_var = rt, 'disturb_year')
+nrtA <- shapirosNormal(evires, resp_var = rt, 'disturb_year')
 nrtA$var <- 'rt'
-nrtB <- shapirosNormal(re, resp_var = rt, 'site') 
+nrtB <- shapirosNormal(evires, resp_var = rt, 'site') 
 nrtB$var <- 'rt'
-nrtAB <- shapirosNormal(re, resp_var = rt, c('disturb_year','site'))
+nrtAB <- shapirosNormal(evires, resp_var = rt, c('disturb_year','site'))
 nrtAB$var <- 'rt'
@@
 ### Relative Resilience 
-nrrsA <- shapirosNormal(re, resp_var = rrs, 'disturb_year')
+nrrsA <- shapirosNormal(evires, resp_var = rrs, 'disturb_year')
 nrrsA$var <- 'rrs'
-nrrsB <- shapirosNormal(re, resp_var = rrs, 'site') 
+nrrsB <- shapirosNormal(evires, resp_var = rrs, 'site') 
 nrrsB$var <- 'rrs'
-nrrsAB <- shapirosNormal(re, resp_var = rrs, c('disturb_year','site'))
+nrrsAB <- shapirosNormal(evires, resp_var = rrs, c('disturb_year','site'))
 nrrsAB$var <- 'rrs'
@@
 ```{r}
+normtestA <- rbind(nrcA, nrtA, nrsA, nrrsA) 
+normtestA %>% pander()
+write.csv(normtestA, 
+          file=paste0(di, '/out/anovas_resilience/evi/normo_disturb_year.csv'), row.names = F)
+```
+
+```{r}
+normtestB <- rbind(nrcB, nrtB, nrsB, nrrsB) 
+normtestB %>% pander()
+write.csv(normtestB, 
+          file=paste0(di, '/out/anovas_resilience/evi/normo_site.csv'), row.names = F)
+```
+
+```{r}
 normtestAB <- rbind(nrcAB, nrtAB, nrsAB, nrrsAB) 
-normtestAB %>% pander()
+normtestAB%>% pander()
 write.csv(normtestAB, 
-          file=paste0(di, '/out/anovas_resilience/normo_disturb_year_site.csv'), row.names = F)
-```
-
-```{r}
-rm(nrcA, nrcB, nrcAB,
-   nrsA, nrsB, nrsAB,
-   nrrsA, nrrsB, nrrsAB,
-   nrtA, nrtB, nrtAB)
+          file=paste0(di, '/out/anovas_resilience/evi/normo_disturb_year_site.csv'), row.names = F)
+```
+
+```{r}
+# rm(nrcA, nrcB, nrcAB, 
+#    nrsA, nrsB, nrsAB,
+#    nrrsA, nrrsB, nrrsAB,
+#    nrtA, nrtB, nrtAB)
 ```
@@
 ## See auxiliar::homogetest
+
 factores <- c('disturb_year', 'site', 'interaction(disturb_year, site)') 
@@
 for (i in responses){ 
-  ht <- homogetest(resp_var = i, factores = factores, df = re)
+  ht <- homogetest(resp_var = i, factores = factores, df = evires)
   ht <- ht %>% mutate(response = i)
@@
 write.csv(homo, 
-          file=paste0(di, '/out/anovas_resilience/homocedasticidad.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/evi/homocedasticidad.csv'), row.names = F)
 ```
@@
 * Tampoco se cumplen los requisitos de homogeneidad de varianzas entre grupos
+
+# Transformación datos
+
+## Log
 
@@
 
-re <- re %>% 
+evires <- evires %>% 
   mutate(
@@
 write.csv(homo_log, 
-          file=paste0(di, '/out/anovas_resilience/homocedasticidad_log.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/evi/homocedasticidad_log.csv'), row.names = F)
 ```
@@
 
-## Custom functions
-
-```{r}
-# Custom Function to compute ANOVAS
-aovas <- function(df, vars, resp_var){ 
-  require('dplyr')
-  require('broom')
-
-  # Create subset 
-  dfsel <- df %>% dplyr::select_(.dots=c(vars, resp_var))
-
-  # Model 
-  myformula <- as.formula(paste0(resp_var,  " ~ ",
-                                 paste(vars, collapse = '*')))
-
-  mymodel <- aov(myformula, data=dfsel)
-
-  # Output model Summary http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1)
-  model_coeff <- broom::tidy(mymodel)
-  model_summary <- broom::glance(mymodel)
-
-  out <- c() 
-  out$model_coeff <- model_coeff
-  out$model_summary <- model_summary
-  out$mymodel <- mymodel
-
-  return(out)
+## Log + 1
+
+```{r}
+factores <- c('disturb_year', 'site', 'interaction(disturb_year, site)') 
+responses <- c('log1rs', 'log1rc', 'log1rt', 'log1rrs')
+homo_log1 <- c()
+
+evires <- evires %>% 
+  mutate(
+    log1rs = log(rs + 1),
+    log1rc = log(rc + 1),
+    log1rt = log(rc + 1),
+    log1rrs = log(rrs +1 )
+)
+
+for (i in responses){ 
+  ht <- homogetest(resp_var = i, factores = factores, df = evires)
+  ht <- ht %>% mutate(response = i)
+  homo_log1 <- rbind(homo_log1, ht)
 }
 
-# Post-Hoc comparison
-phc <- function(mymodel, resp_var){
-  require(lsmeans)
-
-  # Disturb Event 
-  ph_event <- lsmeans(mymodel, pairwise ~ disturb_year, adjust = "bon")
-
-  # differences letters 
-  cld_event <- cld(ph_event, alpha   = 0.01, 
-                   Letters = letters, 
-                   adjust  = "bon")
-
-  # Site  
-  ph_site <- lsmeans(mymodel, pairwise ~ site, adjust = "bon")
-  cld_site <- cld(ph_site, alpha   = 0.01, 
-                 Letters = letters, 
-                 adjust  = "bon")
-
-  # interaction 
-  ph_i <- lsmeans(mymodel, pairwise ~ disturb_year:site, adjust = "bon")
-  cld_i <- cld(ph_i, alpha   = 0.01, 
-                 Letters = letters, 
-                 adjust  = "bon")
-
-  # Objets for plot
-  aux_ph_site <- as.data.frame(summary(ph_site$lsmeans)) 
-  aux_ph_site <- aux_ph_site %>% mutate(var = resp_var)
-  aux_ph_event <- as.data.frame(summary(ph_event$lsmeans)) 
-  aux_ph_event <- aux_ph_event %>% mutate(var = resp_var)
-  aux_ph_i <- as.data.frame(summary(ph_i$lsmeans)) 
-  aux_ph_i <- aux_ph_i %>% mutate(var = resp_var)
-
-  # Return objects
-  cat('\n### Event ###\n')
-  print(ph_event)
-  print(cld_event)
-  cat('\n### Clu pop ###\n')
-  print(ph_site)
-  print(cld_site)
-  cat('\n### Event:Clu pop ###\n')
-  print(ph_i)
-  return(list(aux_ph_site, aux_ph_event, aux_ph_i, cld_site, cld_event, cld_i))
-}
-```
-
-**OJO SOLO 2005 y 2012**
-
-```{r}
-# Only 2005 and 2012
-re <- re %>% filter(disturb_year != 1995) %>% as.data.frame()
-vars <- c('disturb_year','site')
-re$disturb_year <- factor(re$disturb_year)
-```
-
-## Recovery 
-```{r}
-# Variable
-resp_var <- 'rc'
-
-# AOV
-aov_rc <- aovas(re, vars=vars, resp_var = resp_var)
-
-mc <- aov_rc$model_coeff
-
-pander(mc, round=5,
-       caption = paste0("ANOVA table: ", resp_var), missing = '', 
-       emphasize.strong.cells = 
-         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-
-gm <- aov_rc$model_summary
-
-gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
-colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
-rownames(gm) <- "Statistic"
-pander(t(gm)) 
-```
-
-```{r}
-# Post hoc Define model
-mymodel <- aov_rc$mymodel
-postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
-```
-
-```{r, echo=FALSE}
-#### ~ Site
-ps <- plot(effect("site",mymodel))
-#### ~ Disturb Year
-pd <- plot(effect('disturb_year', mymodel))
-#### Disturb Year:Site
-picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
-
-```
-
-```{r, fig.width = 4, fig.height=4}
-ps
-```
-
-```{r, fig.width = 4, fig.height=4}
-pd
-```
-
-```{r, fig.width = 4, fig.height=4}
-picollapse
-```
-
-```{r, fig.width = 6}
-pi
-```
-
-## Resistance 
-```{r, echo =FALSE}
-# Variable
-resp_var <- 'rt'
-vars <- c('disturb_year','site')
-
-# AOV
-aov_rt <- aovas(re, vars=vars, resp_var = resp_var)
-
-mc <- aov_rt$model_coeff
-
-pander(mc, round=5,
-       caption = paste0("ANOVA table: ", resp_var), missing = '', 
-       emphasize.strong.cells = 
-         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-
-gm <- aov_rt$model_summary
-
-gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
-colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
-rownames(gm) <- "Statistic"
-pander(t(gm)) 
-```
-
-```{r}
-# Post hoc Define model
-mymodel <- aov_rt$mymodel
-postH_rt <- phc(mymodel = mymodel, resp_var = resp_var)
-```
-
-```{r, echo=FALSE}
-#### ~ Site
-ps <- plot(effect("site",mymodel))
-#### ~ Disturb Year
-pd <- plot(effect('disturb_year', mymodel))
-#### Disturb Year:Site
-picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
-
-```
-
-```{r, fig.width = 4, fig.height=4}
-ps
-```
-
-```{r, fig.width = 4, fig.height=4}
-pd
-```
-
-```{r, fig.width = 4, fig.height=4}
-picollapse
-```
-
-```{r, fig.width = 6}
-pi
-```
-
-## Relative Resilience 
-```{r, echo=FALSE}
-# Variable
-resp_var <- 'rrs'
-
-# AOV
-aov_rrs <- aovas(re, vars=vars, resp_var = resp_var)
-
-mc <- aov_rrs$model_coeff
-
-pander(mc, round=5,
-       caption = paste0("ANOVA table: ", resp_var), missing = '', 
-       emphasize.strong.cells = 
-         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-
-gm <- aov_rrs$model_summary
-
-gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
-colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
-rownames(gm) <- "Statistic"
-pander(t(gm)) 
-```
-
-```{r}
-# Post hoc Define model
-mymodel <- aov_rrs$mymodel
-postH_rrs <- phc(mymodel = mymodel, resp_var = resp_var)
-```
-
-```{r, echo=FALSE}
-#### ~ Site
-ps <- plot(effect("site",mymodel))
-#### ~ Disturb Year
-pd <- plot(effect('disturb_year', mymodel))
-#### Disturb Year:Site
-picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
-
-```
-
-```{r, fig.width = 4, fig.height=4}
-ps
-```
-
-```{r, fig.width = 4, fig.height=4}
-pd
-```
-
-```{r, fig.width = 4, fig.height=4}
-picollapse
-```
-
-```{r, fig.width = 6}
-pi
-```
-
-## Resilience 
-```{r, echo=FALSE}
-# Variable
-resp_var <- 'rs'
-
-# AOV
-aov_rs <- aovas(re, vars=vars, resp_var = resp_var)
-
-mc <- aov_rs$model_coeff
-
-pander(mc, round=5,
-       caption = paste0("ANOVA table: ", resp_var), missing = '', 
-       emphasize.strong.cells = 
-         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-
-gm <- aov_rs$model_summary
-
-gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
-colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
-rownames(gm) <- "Statistic"
-pander(t(gm)) 
-```
-
-```{r}
-# Post hoc Define model
-mymodel <- aov_rs$mymodel
-postH_rs <- phc(mymodel = mymodel, resp_var = resp_var)
-```
-
-```{r, echo=FALSE}
-#### ~ Site
-ps <- plot(effect("site",mymodel))
-#### ~ Disturb Year
-pd <- plot(effect('disturb_year', mymodel))
-#### Disturb Year:Site
-picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
-
-```
-
-```{r, fig.width = 4, fig.height=4}
-ps
-```
-
-```{r, fig.width = 4, fig.height=4}
-pd
-```
-
-```{r, fig.width = 4, fig.height=4}
-picollapse
-```
-
-```{r, fig.width = 6}
-pi
-```
-
-```{r, echo=FALSE}
-# Interactions plot all  
-means_site <- postH_rc[[4]] %>% mutate(var = 'rc') %>% 
-  bind_rows(postH_rt[[4]] %>% mutate(var = 'rt')) %>% 
-  bind_rows(postH_rs[[4]] %>% mutate(var = 'rs')) %>% 
-  bind_rows(postH_rrs[[4]] %>% mutate(var = 'rrs')) %>% 
-  rename(letras = .group)
-
-means_disturb <- postH_rc[[5]] %>% mutate(var = 'rc') %>% 
-  bind_rows(postH_rt[[5]] %>% mutate(var = 'rt')) %>% 
-  bind_rows(postH_rs[[5]] %>% mutate(var = 'rs')) %>% 
-  bind_rows(postH_rrs[[5]] %>% mutate(var = 'rrs')) %>% 
-  rename(letras = .group)
-
-means_distub_site <- postH_rc[[6]] %>% mutate(var = 'rc') %>% 
-  bind_rows(postH_rt[[6]] %>% mutate(var = 'rt')) %>% 
-  bind_rows(postH_rs[[6]] %>% mutate(var = 'rs')) %>% 
-  bind_rows(postH_rrs[[6]] %>% mutate(var = 'rrs')) %>% 
-  rename(letras = .group)
-
-```
-
-```{r, echo=FALSE}
-dodge <- position_dodge(width = 0.3)
-micolor <- '#455883'
-mierrorbarSE <- aes(ymin=lsmean - SE, ymax=lsmean + SE)
-mierrorbar <- aes(ymin=lower.CL, ymax=upper.CL)
-```
-
-```{r, echo=FALSE}
-plot_ms <- means_site %>%  
-  ggplot(aes(x=site, y=lsmean)) + 
-  geom_point(colour=micolor, 
-             size=3, position = dodge) +
-  theme_bw() + xlab('') + ylab('') + 
-  facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15) +
-  theme(strip.background = element_rect(colour = "black", fill = "white"))
-
-plot_msSE <- plot_ms + geom_errorbar(mierrorbarSE,color=micolor, 
-                                     size=.5, width=.15, position = dodge)
-
-plot_msCI <- plot_ms + geom_errorbar(mierrorbar,color=micolor, 
-                                     size=.5, width=.15, position = dodge)
-```
-
-```{r, echo=FALSE}
-plot_md <- means_disturb %>%  
-  ggplot(aes(x=disturb_year, y=lsmean)) + 
-  geom_point(colour=micolor, 
-             size=3, position = dodge) +
-  theme_bw() + xlab('') + ylab('') + 
-  facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15) +
-  theme(strip.background = element_rect(colour = "black", fill = "white"))
-
-plot_mdSE <- plot_md + geom_errorbar(mierrorbarSE,color=micolor, 
-                                     size=.5, width=.15, position = dodge)
-
-plot_mdCI <- plot_md + geom_errorbar(mierrorbar,color=micolor, 
-                                     size=.5, width=.15, position = dodge)
-```
-
-```{r}
-plot_mds <- means_distub_site %>%  
-  ggplot(aes(x=site, y=lsmean, group=disturb_year, colour=disturb_year)) + 
-  geom_point(aes(shape=disturb_year), size=3) + 
-  geom_line() +
-  theme_bw() + xlab('') + ylab('') + 
-  facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean+SE, label=letras), nudge_x = 0.15)+
-  theme(strip.background = element_rect(colour = "black", fill = "white"),
-        legend.position = c(0.8, 0.93),
-        legend.background = element_blank()) +
-  scale_colour_manual(values = c(micolor, "red"))
-
-plot_mdsSE <- plot_mds + geom_errorbar(mierrorbarSE, size=.5, width=.15)
-plot_mdsCI <- plot_mds + geom_errorbar(mierrorbar, size=.5, width=.15)
-```
-
-```{r}
-pdf(paste0(di, 'out/fig/resilience/interaction_plotsSE.pdf'), width=9, height = 9)
-grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
-dev.off()
-
-pdf(paste0(di, 'out/fig/resilience/interaction_plotsCI.pdf'), width=9, height = 9)
-grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
-dev.off()
-
-```
-
-```{r}
-aovas_coeff <- aov_rc$model_coeff %>% mutate(var = 'rc') %>% 
-  bind_rows(aov_rt$model_coeff %>% mutate(var = 'rt')) %>% 
-  bind_rows(aov_rs$model_coeff %>% mutate(var = 'rs')) %>% 
-  bind_rows(aov_rrs$model_coeff%>% mutate(var = 'rrs'))
-
-write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/anovas_statistics.csv'), row.names = F)
-
-aovas_coeff %>% pander()
-```
-
-```{r}
-aovas_model_summary <- aov_rc$model_summary %>% mutate(var = 'rc') %>% 
-  bind_rows(aov_rt$model_summary %>% mutate(var = 'rt')) %>% 
-  bind_rows(aov_rs$model_summary %>% mutate(var = 'rs')) %>% 
-  bind_rows(aov_rrs$model_summary%>% mutate(var = 'rrs'))
-
-write.csv(aovas_model_summary, 
-          file=paste0(di, '/out/anovas_resilience/anovas_summary_modelos.csv'), row.names = F)
-
-gm <- apply(aovas_model_summary, 1, formatC, digits = 2, format = "f") 
-rownames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e", "variable"),"$")
-colnames(gm) <- c("rc", "rt", "rs", "rrs")
-
-pander(gm)
-```
+homo_log1 %>% pander()
+write.csv(homo_log, 
+          file=paste0(di, '/out/anovas_resilience/evi/homocedasticidad_log_plus_1.csv'), row.names = F)
+```
+
+* Tampoco se cumplen
+
+## Buscar mejor transformación de Box-Cox
+
+* Buscamos el mejor lambda para cada variable para estudiar posibles transformaciones
+
+#### Lambda Resilience 
+```{r}
+m  <- lm(rs ~ disturb_year*site, evires)
+b <- boxcox(m)
+b$x[which.max(b$y)]
+```
+
+#### Lambda Resistance
+```{r}
+m  <- lm(rt ~ disturb_year*site, evires)
+b <- boxcox(m)
+b$x[which.max(b$y)]
+```
+
+#### Lambda Recovery
+```{r}
+m  <- lm(rc ~ disturb_year*site, evires)
+b <- boxcox(m)
+b$x[which.max(b$y)]
+```
+
+#### Lambda Relative Resilience
+```{r, eval=FALSE}
+m  <- lm(rrs ~ disturb_year*site, evires)
+b <- boxcox(m)
+b$x[which.max(b$y)]
+```
+
+Obtengo diferentes lambdas, lo cual complica las transformaciones. Por lo que opto por ROBUST ANOVA
 
@@
   x <- pbad2way(formulaFull, data = df, est = "mom", nboot = nboot)
-
-  pbad2way(rs ~ disturb_year + site + disturb_year:site, data = df, 
-           est = "mom", nboot = nboot)
 
@@
 }
-
-  # if (exists('letters_phb')) {
-  # letters_phb <- letters_phb} else {
-  # myerror <- evaluate('cldList(comparison = phb$Comparison, 
-  #                     p.value = phb$p.adjust, threshold  = treshold)')
-  # letters_phb <- as.character(myerror[[2]]$message)}
-  # 
-  # if (exists('letters_pha')) {
-  # letters_pha <- letters_pha} else {
-  # myerror <- evaluate('cldList(comparison = pha$Comparison, 
-  #                     p.value = pha$p.adjust, threshold  = treshold)')
-  # letters_pha <- as.character(myerror[[2]]$message)}
-  # 
-  # if (exists('letters_phab')) {
-  # letters_phab <- letters_phab} else {
-  # myerror <- evaluate('cldList(comparison = phab$Comparison, 
-  #                     p.value = phab$p.adjust, threshold  = treshold)')
-  # letters_phab <- as.character(myerror[[2]]$message)}
-
 ```
@@
 
-## Resilience 
-```{r}
-rars <- robustANOVA(df=re, resp_var='rs', factores=factores,
+```{r}
+rars <- robustANOVA(df=evires, resp_var='rs', factores=factores,
               alpha = 0.95, nboot = 3000, treshold = 0.01)
@@
 letraBrs <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rs')
-letraBrs # Manual (IMPROVE IT)
-letraBrs <- data.frame(Group = c('SJ', 'caH', 'caL'),
-                     Letter = c('a','a','a'),
-                     MonoLetter = c('a','a','a')) %>% mutate(var ='rs')
 letraBrs
@@
 
-## Recovery 
-```{r}
-rarc <- robustANOVA(df=re, resp_var='rc', factores=factores,
+```{r}
+rarc <- robustANOVA(df=evires, resp_var='rc', factores=factores,
               alpha = 0.95, nboot = 3000, treshold = 0.01)
@@
 
-## Resistance 
-```{r}
-rart <- robustANOVA(df=re, resp_var='rt', factores=factores,
+```{r}
+rart <- robustANOVA(df=evires, resp_var='rt', factores=factores,
               alpha = 0.95, nboot = 3000, treshold = 0.01)
@@
 letraArt <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rt')
-letraArt #Manual (IMPROVE IT)
-letraArt <- data.frame(Group = as.factor(c('25', '212')),
-                     Letter = as.factor(c('a','a')),
-                     MonoLetter = as.factor(c('a','a'))) %>% mutate(var ='rt')
 letraArt 
@@
 
-## Relative Resilience
-```{r}
-rarrs <- robustANOVA(df=re, resp_var='rrs', factores=factores,
+```{r}
+rarrs <- robustANOVA(df=evires, resp_var='rrs', factores=factores,
               alpha = 0.95, nboot = 3000, treshold = 0.01)
@@
 
-#### Rt Letters 
 ```{r}
@@
 mhuber_agg %>% pander()
+
 ```
@@
 mhuber_agg_a %>% pander() 
-
 ```
@@
 
-```{r}
+```{r robust_plot_evi_drought}
 mhuber<- mhuber %>% 
@@
                                  '3_rrs' = 'Relative Resilience'))) +
-  scale_color_manual(values=c('black','blue','red')) + 
-  scale_fill_manual(values=c('black','blue','red')) + theme_bw() +
-  scale_linetype_manual(values=c("solid", "dashed", 'dashed')) +
+  scale_color_manual(values=c('black','blue')) + 
+  scale_fill_manual(values=c('black','blue')) + theme_bw() +
+  scale_linetype_manual(values=c("solid", "dashed")) +
   theme(panel.grid.minor = element_blank(), 
@@
 
-pdf(paste0(di, 'out/fig/resilience/robust_plot_evi_drought.pdf'), width=9, height = 9)
+pdf(paste0(di, '/out/resilience_robust/evi/robust_plot_evi_drought.pdf'), width=9, height = 9)
 robust_plot_evi_drought 
@@
 ### Response ~ (x=site)
-```{r}
+```{r robust_plot_evi_site}
 pd <- position_dodge(.2)
@@
 
-pdf(paste0(di, 'out/fig/resilience/robust_plot_evi_site.pdf'), width=9, height = 9)
+pdf(paste0(di, '/out/resilience_robust/evi/robust_plot_evi_site.pdf'), width=9, height = 9)
 robust_plot_evi_site 
@@
 # Export data 
-write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/robust_mhuber.csv'), row.names = F)
-write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg.csv'), row.names = F)
-write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_a.csv'), row.names = F)
-write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg_a.csv'), row.names = F)
-write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_b.csv'), row.names = F)
-write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg_b.csv'), row.names = F)
-
-write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/robust_pairwise.csv'), row.names = F)
-```
-
-# References
+write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber.csv'), row.names = F)
+write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_agg.csv'), row.names = F)
+write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_a.csv'), row.names = F)
+write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_agg_a.csv'), row.names = F)
+write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_b.csv'), row.names = F)
+write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_agg_b.csv'), row.names = F)
+
+write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_pairwise.csv'), row.names = F)
+```

### 2017-11-21 analysis/evi/explore_anomalies.Rmd
update anomalies EVI

@@
 
-```{r}
+```{r trajectories_anomalies_pop}
 traj_mean_pop <- traj %>% 
@@
 
+```{r trajectories_anomalies_sa}
+trajsa <- anomalias_evimean %>% group_by(pop, y) %>% 
+  dplyr::summarise(mean = mean(sa),
+            sd = sd(sa))
+
+traj_meansa <- trajsa %>% 
+  group_by(y) %>% 
+  dplyr::summarise(meanOfmean = mean(mean), 
+            sdOfmean = sd(mean),
+            seOfmean = sdOfmean/sqrt(length(mean)),
+            meanOfsd = mean(sd),
+            sdOfsd = sd(sd),
+            seOfsd = sdOfsd /sqrt(length(sd))) %>% 
+  as.data.frame()
+
+traj_plotsa <- traj_meansa %>% 
+  ggplot(aes(x=meanOfmean, y=meanOfsd, label=y)) +
+  geom_errorbar(aes(ymin=meanOfsd - seOfsd, ymax=meanOfsd + seOfsd)) + 
+  geom_errorbarh(aes(xmin=meanOfmean - seOfmean, xmax=meanOfmean + seOfmean)) + 
+  geom_path(colour='gray') +
+  geom_point(size=3, shape=21, fill='white') + 
+  geom_text(hjust = 0.001, nudge_x = 0.001) + 
+  geom_vline(aes(xintercept = mean(meanOfmean)), colour='red') +
+  geom_hline(aes(yintercept = mean(meanOfsd)), colour ='red')+
+  theme_bw() + xlab('mean') + ylab('variance') + 
+  theme(strip.background = element_rect(fill = "white"), 
+        legend.position="none")
+
+traj_plotsa
+pdf(file=paste0(di, "/out/trajectories/plot_trajectories_anomalies_sa.pdf"), height = 6, width =6)
+traj_plotsa
+dev.off()
+```
+
+```{r trajectories_anomalies_pop_sa}
+traj_mean_popsa <- trajsa %>% 
+  mutate(clu_pop = as.factor(ifelse(pop %in% c(1,2,3,4,5), 'N', 'S'))) %>% 
+  group_by(clu_pop,y) %>% 
+  dplyr::summarise(meanOfmean = mean(mean), 
+            sdOfmean = sd(mean),
+            seOfmean = sdOfmean/sqrt(length(mean)),
+            meanOfsd = mean(sd),
+            sdOfsd = sd(sd),
+            seOfsd = sdOfsd /sqrt(length(sd))) %>%
+  as.data.frame()
+
+line_traj_mean_popsa <- traj_mean_popsa %>% 
+  group_by(clu_pop) %>% 
+  dplyr::summarise(
+    meanOfmean = mean(meanOfmean), 
+    meanOfsd = mean(meanOfsd)
+  )
+
+traj_plot_popsa <- ggplot(traj_mean_popsa,
+  aes(x=meanOfmean, y=meanOfsd, label=y)) +
+  geom_errorbar(aes(ymin=meanOfsd - seOfsd, ymax=meanOfsd + seOfsd)) + 
+  geom_errorbarh(aes(xmin=meanOfmean - seOfmean, xmax=meanOfmean + seOfmean)) + 
+  geom_path(colour='gray') +
+  geom_point(size=3, shape=21, fill='white') +
+  geom_text(hjust = 0.001, nudge_x = 0.001) + 
+  facet_wrap(~clu_pop) +
+  theme_bw() + xlab('mean') + ylab('variance') + 
+  theme(strip.background = element_rect(fill = "white"), 
+        legend.position="none")
+
+traj_plot_popsa <- 
+  traj_plot_popsa + 
+  geom_vline(aes(xintercept = meanOfmean), line_traj_mean_popsa,  colour='red') +
+  geom_hline(aes(yintercept = meanOfsd), line_traj_mean_popsa, colour ='red')
+
+traj_plot_popsa
+pdf(file=paste0(di, "/out/trajectories/plot_trajectories_anomalies_clusa.pdf"), height = 6, width =10)
+traj_plot_popsa
+dev.off()
+```
+
 # Notas que podemos obtener de explorar las anomalías
 
-## Del análisis del plot de trayectorias
+## Del análisis del plot de trayectorias:
+
+### Std anomalies
+
+* 2005 fue una gran disminución del greenness (promedio de las sa es muy bajo respecto a los otros años) aunque es heterogenea (alta varianza)
+
+* La disminución en 2012 fue menor, y no se puede considerar un bronwing (sa < - 1), pero homogénea. 2000, 2006 y 2007 presentaron menos anomalías que 2012.
+
+* Al explorar los patrones N y S, observamos lo mismo, aunque mayor variabilidad para el evento de 2005 en las poblaciones del sur.
+
+### Anomalias
 

### 2017-11-21 analysis/evi/explore_anomalies.Rmd
remove errors // add percentage browing

@@
 plot_a_pop
-pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_a.pdf"), height = 14, width = 14)
+pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_pop_a.pdf"), height = 14, width = 14)
 plot_a_pop
@@
 plot_nora_pop 
-pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_nora.pdf"), height = 14, width = 14)
+pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_pop_nora.pdf"), height = 14, width = 14)
 plot_nora_pop 
@@
 plot_sa_pop
-pdf(file=paste0(di, "/out/anomalies/plot_anoma_EVI_pop_sa.pdf"), height = 14, width = 14)
+pdf(file=paste0(di, "/out/anomalies/evi/plot_anoma_EVI_pop_sa.pdf"), height = 14, width = 14)
 plot_sa_pop
@@
 
+## Analisis espacial bronwing vs. greenning
+
+```{r}
+sa2005 <- anomalias_evimean %>% filter(y == 2005) %>% 
+  mutate(type = ifelse(sa < -1, "browning", 
+                       ifelse(sa > 1, "greening", "no change"))) %>% 
+  group_by(clu_pop2, type) %>% 
+  dplyr::summarise(count_clu = n()) %>% 
+  group_by(clu_pop2) %>% 
+  mutate(per = round(count_clu / sum(count_clu)*100, 2),
+         y = 2005)
+
+sa2012 <- anomalias_evimean %>% filter(y == 2012) %>% 
+  mutate(type = ifelse(sa < -1, "browning", 
+                       ifelse(sa > 1, "greening", "no change"))) %>% 
+  group_by(clu_pop2, type) %>% 
+  dplyr::summarise(count_clu = n()) %>% 
+  group_by(clu_pop2) %>% 
+  mutate(per = round(count_clu / sum(count_clu)*100, 2),
+         y = 2012)
+
+per_sa <- sa2005 %>% rbind(sa2012) %>% as.data.frame()
+
+per_sa %>% kable()
+
+write.csv(per_sa, file=paste0(di, '/out/anomalies/evi/percen_browing_greening.csv'), row.names = F)
+
+```
+
+```{r per_evi_browing_greening}
+# How many pixels 
+npixels <- length(unique(anomalias_evimean$iv_malla_modi_id))
+# 912
+
+saper <- anomalias_evimean %>% filter(y %in% c(2005, 2012))
+
+saper %>% ggplot(aes(sa)) + facet_grid(y~clu_pop2) + geom_histogram()
+
+```
+
 # Notas que podemos obtener de explorar las anomalías

### 2017-11-21 analysis/evi/explore_anomalies.Rmd
include new analysis evi anomalies

@@
 
+```{r percentage_greening}
+
+per_sa <- per_sa %>% add_row(clu_pop2 = 'Northern slope',
+                             type = 'greening',
+                             count_clu = 0, per = 0.00, y = 2005)
+
+per_pixels_green <- per_sa %>% 
+  ggplot(aes(x=type, y=per, fill=as.factor(y))) + geom_bar(stat='identity', position = "dodge") + facet_wrap(~clu_pop2) + 
+  theme_bw() + 
+  theme(strip.background = element_rect(colour = "black", fill = "white"))
+
+per_pixels_green 
+pdf(file=paste0(di, "/out/anomalies/evi/per_pixels_green.pdf"), height = 6, width =10)
+traj_plot_popsa
+dev.off()
+
+```
+
 ```{r per_evi_browing_greening}

### 2017-11-21 pres/state_ms.Rmd
update theme and links

@@
     fig_caption: FALSE
+    theme: metropolis
+fontsize: 10pt
+classoption: compress
 bibliography: ../references.bib
 csl: ../nature.csl
+header-includes:
+  \setbeamercolor{frametitle}{bg=gray}
+  \hypersetup{colorlinks,citecolor=orange,filecolor=red,linkcolor=brown,urlcolor=blue}
 ---
@@
 - Escala **regional**:
-    - Datos de [\textcolor{blue}{SPEI Global Drought Monitor}](http://sac.csic.es/spei/database.html) para Sierra Nevada (spatial resolution of 0.5º) Ver [\textcolor{blue}{esto}](https://github.com/ajpelu/qpyr_resilience/blob/master/analysis/drought/explore_drought_spei_regional.md)
+    - Datos de [SPEI Global Drought Monitor](http://sac.csic.es/spei/database.html) para Sierra Nevada (spatial resolution of 0.5º) Ver [esto](https://github.com/ajpelu/qpyr_resilience/blob/master/analysis/drought/explore_drought_spei_regional.md)
 
 - Escala **local**:
-    - Calcular datos SPEI para tres estaciones de la [\textcolor{blue}{red RIA}](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController) durante el periodo 2000 - 2016: [\textcolor{blue}{Cadiar}](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=7), [\textcolor{blue}{Padul}](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=10) y [\textcolor{blue}{Jerez del Marquesado}](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=6)
+    - Calcular datos SPEI para tres estaciones de la [red RIA](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController) durante el periodo 2000 - 2016: [Cadiar](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=7), [Padul](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=10) y [Jerez del Marquesado](http://www.juntadeandalucia.es/agriculturaypesca/ifapa/ria/servlet/FrontController?action=Static&url=coordenadas.jsp&c_provincia=18&c_estacion=6)
     - Curvas de precipitación acumulada RIA y datos base Aérea (obtenidos de la REDIAM) (desde 1950) 
-    - Ver [\textcolor{blue}{flexdashboard drought}](https://rawgit.com/ajpelu/qpyr_dendro/master/analysis/drought/drought_dashboard.html)
+    - Ver [flexdashboard drought](https://rawgit.com/ajpelu/qpyr_dendro/master/analysis/drought/drought_dashboard.html)
 
@@
 
-Mas info del [\textcolor{blue}{filtrado de datos}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/prepare_modis_qa.md)
+Mas info del [filtrado de datos](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/prepare_modis_qa.md)
 
@@
     * Índices de competencia dependientes e independientes de la distancia [@GeaIzquierdo2009] 
-    * [\textcolor{blue}{Resultados}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/analysis_competence.md)
+    * [Resultados](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/analysis_competence.md)
 
@@
 
-    * Ver [\textcolor{blue}{gráficas}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/analysis_resilience_bai.md) para BAI
-    * Ver [\textcolor{blue}{gráficas}](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/evi/compute_resilience_evi.md) para EVI
+    * Ver [gráficas](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/analysis_resilience_bai.md) para BAI
+    * Ver [gráficas](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/evi/compute_resilience_evi.md) para EVI
 
@@
 
-# R - Explore EVI mean
+# Explore EVI mean
 
 ![](../analysis/evi/explore_evi_mean_files/figure-markdown_github/evi_profile_compara-1.png)
+
+---
+
+# EVI anomalies
+
+![](../analysis/evi/explore_anomalies_files/figure-markdown_github/plot_anoma_EVI_clu_sa-1.png){width=90%}
+
+---
+
+# Notas anomalias
+
+* EVI sa menores en 2005 *** (lowest) que en 2012, sobre todo para poblaciones del norte
+* Criterio Samanta et al. > + 1 greening | < -1 bronwing 
+* [Más info anomalias](https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/evi/explore_anomalies.md ) EVI
+* Heterogeneidad de anomalias estandarizadas
+
+---
+
+# Mapa de anomalias estandarizadas
+
+![](../analysis/evi/explore_anomalies_files/figure-markdown_github/map_anomaly_2005_2012_sa-1.png){width=90%}
+
+---
+
+# Trayectorias anomalias estandarizadas
+
+![](../analysis/evi/explore_anomalies_files/figure-markdown_github/trajectories_anomalies_sa-1.png){width=90%}
+
+---
+
+# Browing
+
+![](../analysis/evi/explore_anomalies_files/figure-markdown_github/percentage_greening-1.png){width=90%}
+
+---
+
+```{r, echo=FALSE}
+library(knitr)
+df <- read.csv(file='../out/anomalies/evi/percen_browing_greening.csv', header=TRUE)
+
+kable(df)
+```
 
@@
 
+* EVI sa menores en 2005 *** (lowest) que en 2012, sobre todo para poblaciones del norte
+* Criterio Samanta et al. > + 1 greening | < -1 bronwing
+
 https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/evi/explore_anomalies.md

### 2017-11-21 analysis/evi/explore_anomalies.Rmd
add percentage to plot

--- 2017-11-21 analysis/evi/explore_anomalies.Rmd
@@
 
+```{r percentage_greening}
+
+per_sa <- per_sa %>% add_row(clu_pop2 = 'Northern slope',
+                             type = 'greening',
+                             count_clu = 0, per = 0.00, y = 2005)
+library(scales)
+
+per_pixels_green <- per_sa %>% 
+  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
+  geom_bar(stat='identity', position = "dodge") + 
+  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), vjust=-0.25) + 
+  facet_wrap(~clu_pop2) + 
+  theme_bw() + 
+  theme(strip.background = element_rect(colour = "black", fill = "white"))
+
+per_pixels_green 
+pdf(file=paste0(di, "/out/anomalies/evi/per_pixels_green.pdf"), height = 6, width =10)
+traj_plot_popsa
+dev.off()
+
+```
+
 ```{r per_evi_browing_greening}

### 2017-11-21 man/ms_2_results.Rmd
update

@@
 
-referencia a las tendencias en EVI (lo hemos vuelto a calcular y además en el trabajo de ontologías también nos sale)
+Aproximadamete el x% de los robledales mostraron un incremento en los valores de productividad (greenees) ... (relacionar con tendencias EVI y ontologias).
 
-Standardized a
+Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (Fig.
+R1 Profile EVI).
+The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Fig.
+R2 y R3).
 
-Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005.
+Esta disminución para 2005 sin embargo fue heterogénea (ver plot de trajectorias)
+
+En 2005 la mayoría de los pixeles mostraron browning (99.36 % y 79.37 % para las poblaciones del norte y del sur respectivamente), mientras que en 2012 la mayoría de los pixeles no mostraron cambio
+
+de los pixels que cubren las poblaciones del norte mostraron browning durante la sequía de 2005,
+
+La mayor parte del área Tanto en las poblaciones del norte como en las del sur se observó que l
+
+La mayoría de los pixels que cubren los robledales de las poblaciones del norte y las del sur mostraron browning para la sequía del 2005, sin embargo en 2012 la mayoría de los pixeles se clasificaron como no changes en las poblaciones del norte (89.60 %) y en las del sur (70.07 %) (Tabla supplementaria??)
+* Si atendemos a las sa (standardized anomalíes) y aplicamos el criterio de Gao, podemos decir que en 2005 se observó un bronwing en los bosques de Q.
+pyrenaica, sobre todo en las situadas en el northern slopes.
+
+> Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005.
 Sin embargo vemos valores menos negativos para 2005.
@@
 
-Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029)que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
+> Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029)que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
 
-Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (:red_circle: Fig 1a EVI profile comparison), particu
+> * En 2005 y en 2012 las anomalías fueron negativas.
+> Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.
 
-Del análisis de las anomalías observamos:
+#### Temporal trend EVI values 
+El 78.95 % de los pixeles mostraron una tendencia positiva en cuanto al EVI medio anual (siendo significativa para el 31.67 % de los pixeles).
+Esta tendencia positiva fué sobre todo mayor en algunas de las poblaciones del suroeste.
+La
 
-* En 2005 y en 2012 las anomalías fueron negativas. 
-* 2005 las anomalías fueron mucho mas negativas que para 2012.
+--> referencia a las tendencias en EVI (lo hemos vuelto a calcular y además en el trabajo de ontologías también nos sale)
 
-* Analizar magnitud de las anomalías para 2005 y 2012 (comparación entre ellas y entre NyS):
-
-Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.
-
-* 2005 fue el año en el que las anomalías fueron mas negativas, siendo de las anomalías de mas magnitud negativa en las poblaciones del norte.
-
-* Si atendemos a las sa (standardized anomalíes) y aplicamos el criterio de Gao, podemos decir que en 2005 se observó un bronwing en los bosques de Q.
-pyrenaica, sobre todo en las situadas en el northern slopes.
-
-* Asimismo, las anomalías 2012 fueron negativas pero
+En 2005 y 2012 se observó un browning para los valores medios de EVI. En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.
 

### 2017-11-21 pres/state_ms.Rmd
update

@@
 
-# R - Plot de trayectorias: general y por cluster pops (ojito con esto)
+# Patrones generales del EVI medio anual
 
-![](../analysis/evi/explore_evi_mean_files/figure-markdown_github/trajectories_evi-1.png)
+* El 78.95 % de los pixeles mostraron una tendencia positiva en cuanto al EVI medio anual (siendo significativa para el 31.67 % de los pixeles).
+
+* Esta tendencia positiva fué sobre todo mayor en algunas de las poblaciones del suroeste
+
+* Concuerda con datos de ontología (TODO comentar)
 
@@
 
-# R - Plot de trayectorias: general y por cluster pops (ojito con esto)
-
-![](../analysis/evi/explore_evi_mean_files/figure-markdown_github/trajectories_evi_clu-1.png)
+![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/trend_sig_pixels-1.png){width=150%}
 
 ---
-
-# R - Mann Kendall trends in EVI mean
 
@@
 
-# R - Mann Kendall trends in EVI mean
+Trayectorias EVI medio anual
 
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/trend_sig_pixels-1.png){width=150%}
-
----
-
-# R- Mann Kendall trends in EVI mean
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/raster_map_sen_evi-1.png){width=150%}
+![](../analysis/evi/explore_evi_mean_files/figure-markdown_github/trajectories_evi-1.png)
 

### 2017-11-22 man/ms_2_results.Rmd
update
updating

@@
 
+!!! Vas por aqui pollo
+
+Pyrenean oak forests showed significantly lower resistance to 2005 drought event than to 2012 one [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001].
+
+* Menor resistencia de los robledales a la sequía de 2005 que a la de 2012
+
+Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001], siendo diferencial el comportamiento en la sequía de 2005, donde se observó menor resistencia de las poblaciones del norte.
+
 ## Resiliencia EVI
@@
 Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
+
+La sequía de 2005 redujo el EVI medio hasta el 85.8%, mientras que la de 2012 la redujo hasta el 94.3%.
+
+La sequía de 2005 redujo el evi medio hasta el 81.9 % en las poblaciones del N, y hasta el 90.2 % en las del S
+
+... drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)
 

### 2017-11-24 analysis/evi/explore_resilience.Rmd
update
update robust anovas

@@
 
+  set.seed(123)
+
   # Create interaction 
@@
     p_value = c(x$A.p.value, x$B.p.value, x$AB.p.value))
+
+  out_raTrimmed <- t2way(formulaFull, data = df)
 
@@
 
+  phtrimmed <- mcp2atm(formulaFull, data=df)
+
   out <- list()  
@@
   out$ra <- out_ra # Output for Two-way robust analysis (M-estimators)
+  out$raTrimmed <- out_raTrimmed
   out$ph <- ph # posthoc comparison usinng pairwiseRobustTest 
@@
 
+  print(cat('\n Robust M-Anova \n'))
   print(out_ra)
+  print(cat('\n Robust Trimmed \n'))
+  print(out_raTrimmed)
+  print(cat('\n post hoc Mhuber \n'))
   print(phRWS2)
+  print(cat('\n post hoc Trimmed means \n'))
+  print(phtrimmed)
 

### 2017-11-24 analysis/evi/explore_resilience.Rmd
add pdf

@@
 date: "2017 Aug"
-output: 
-  md_document:
-    variant: markdown_github
+output: pdf_document
 ---

### 2017-11-24 analysis/evi/explore_resilience.Rmd
add anovas table

--- 2017-11-24 analysis/evi/explore_resilience.Rmd
@@
 
+  out_ratr_df <- data.frame(var = c(out_raTrimmed$varnames[2], 
+                                    out_raTrimmed$varnames[3], 
+                                    paste0(out_raTrimmed$varnames[2], ':', out_raTrimmed$varnames[3])),
+              statistic = c(out_raTrimmed$Qa, out_raTrimmed$Qb, out_raTrimmed$Qab),
+              pvalue = c(out_raTrimmed$A.p.value, out_raTrimmed$B.p.value, out_raTrimmed$AB.p.value))
+
   # post-hoc 
@@
   out$raTrimmed <- out_raTrimmed
+  out$out_ratr_df <- out_ratr_df
   out$ph <- ph # posthoc comparison usinng pairwiseRobustTest 
@@
 ```{r}
+
+rars$out_ratr_df$var <- 'rs'
+rarc$out_ratr_df$var <- 'rc'
+rart$out_ratr_df$var <- 'rt'
+rarrs$out_ratr_df$var <- 'rrs'
+
+trimmedanovas <- rbind(rarc$out_ratr_df, rart$out_ratr_df, rars$out_ratr_df, rarrs$out_ratr_df) 
+trimmedanovas %>% pander()
+
+```
+
+```{r}
 # Export data 
@@
 write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_pairwise.csv'), row.names = F)
-```
+write.csv(trimmedanovas, file=paste0(di, '/out/anovas_resilience/huber_evi/trimmed_anovas.csv', row.names = F)
+```

### 2017-11-24 analysis/evi/explore_resilience.Rmd
update

@@
 
-  out_ratr_df <- data.frame(var = c(out_raTrimmed$varnames[2], 
+  out_ratr_df <- data.frame(fact = c(out_raTrimmed$varnames[2], 
                                     out_raTrimmed$varnames[3], 
@@
 write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_pairwise.csv'), row.names = F)
-write.csv(trimmedanovas, file=paste0(di, '/out/anovas_resilience/huber_evi/trimmed_anovas.csv', row.names = F)
-```
+write.csv(trimmedanovas, file=paste0(di, '/out/anovas_resilience/huber_evi/trimmed_anovas.csv'), row.names = F)
+```

### 2017-12-01 analysis/plots_correlations_resiliences.Rmd
add plots of resiliences

---
title: "Resilience Correlations beetwen drought events"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017"
output: 
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, error=TRUE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("pander")
library('gridExtra')
```

```{r, echo=FALSE}
machine <- 'ajpelu'
#machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
```

# Read data from EVI

```{r}
r.bai <- read.csv(file=paste0(di, '/data/resilience/resilience_bai.csv'), header=T, sep = ",")
r.evi <- read.csv(file=paste0(di, '/data/resilience/resilience_evi.csv'), header=T, sep = ",")

# add data of pop
anomalias <- read.csv(file=paste(di, "/data/anomalies/anomalias_evimean.csv", sep=""), header = TRUE, sep = ',')

attr_iv_malla_modis_id <- anomalias %>% dplyr::select(iv_malla_modi_id,long,lat,pop) %>% unique()
r.evi <- r.evi %>% inner_join(attr_iv_malla_modis_id, by='iv_malla_modi_id')

r.evi <- r.evi %>% 
  mutate(
    clu_pop = as.factor(case_when(
      pop == 1 ~ "Camarate",
      pop %in% c(2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out')),
    clu_pop2 = as.factor(case_when(
      pop %in% c(1,2,3,4,5) ~ 'N',
      pop %in% c(6,7,8) ~ 'S',
      pop == 9 ~ 'out'))) %>% 
  filter(clu_pop != 'out')

# Change name of clu_pop2 and disturb_year 
r.evi <- r.evi %>% dplyr::rename(site = clu_pop2) %>% 
  mutate(disturb_year = as.factor(disturb_year))

```

# BAI

```{r}
df2005 <- r.bai %>% 
  filter(disturb_year == 2005) %>% 
  rename(rs.2005 = rs, 
         rc.2005 = rc,
         rt.2005 = rt, 
         rrs.2005 = rrs) %>% 
  dplyr::select(-disturb_year)

df2012 <- r.bai %>% 
  filter(disturb_year == 2012) %>% 
  rename(rs.2012 = rs, 
         rc.2012 = rc,
         rt.2012 = rt, 
         rrs.2012 = rrs) %>% 
  dplyr::select(-disturb_year) %>% 
  dplyr::select(-site)

d <- inner_join(df2005, df2012, by = 'tree')
```

# Plots BAI

```{r bai_rs}
# Resilience
bai_rs <- d %>% ggplot(aes(x=rs.2005, y=rs.2012, fill=site)) + 
  stat_ellipse(aes(fill=site), geom = 'polygon', alpha=1/6, type='norm', level=.95) +
  geom_point(aes(colour=site)) + 
  xlab('Resilience to 2005 drouht event') + 
  ylab('Resilience to 2012 drouht event') + 
  xlim(0,1.4) + ylim(0.4,1.8) + 
  theme_bw()+
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) 
```

```{r bai_rt}
# Resistance
bai_rt <- d %>% ggplot(aes(x=rt.2005, y=rt.2012, fill=site)) + 
  stat_ellipse(aes(fill=site), geom = 'polygon', alpha=1/6, type='norm', level=.95) +
  geom_point(aes(colour=site)) + 
  xlab('Resistance to 2005 drouht event') + 
  ylab('Resistance to 2012 drouht event') + 
  xlim(0,1.3) + ylim(0,1.3) + 
  theme_bw()+
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1)

```

```{r bai_rc}
# Recovery
bai_rc <- d %>% ggplot(aes(x=rc.2005, y=rc.2012, fill=site)) + 
  stat_ellipse(aes(fill=site), geom = 'polygon', alpha=1/6, type='norm', level=.95) +
  geom_point(aes(colour=site)) + 
  xlab('Recovery to 2005 drouht event') + 
  ylab('Recovery to 2012 drouht event') + 
  theme_bw()+
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) 
```

# EVI
```{r}
e2005 <- r.evi %>% 
  filter(disturb_year == 2005) %>% 
  rename(rs.2005 = rs, 
         rc.2005 = rc,
         rt.2005 = rt, 
         rrs.2005 = rrs) %>% 
  dplyr::select(-disturb_year)

e2012 <- r.evi %>% 
  filter(disturb_year == 2012) %>% 
  rename(rs.2012 = rs, 
         rc.2012 = rc,
         rt.2012 = rt, 
         rrs.2012 = rrs) %>% 
  dplyr::select(-disturb_year) %>% 
  dplyr::select(-c(long:site))

evi <- inner_join(e2005, e2012, by = 'iv_malla_modi_id')
```

# Resilience
```{r evi_rs}
evi_rs <- evi %>% ggplot(aes(x=rs.2005, y=rs.2012, fill=site)) + 
  stat_ellipse(aes(fill=site), geom = 'polygon', alpha=1/6, type='norm', level=.95) +
  geom_point(aes(colour=site)) + 
  xlab('Resilience to 2005 drouht event') + 
  ylab('Resilience to 2012 drouht event') + 
  theme_bw()+
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) 
```

```{r}
# Resistance
evi_rt <- evi %>% ggplot(aes(x=rt.2005, y=rt.2012, fill=site)) + 
  stat_ellipse(aes(fill=site), geom = 'polygon', alpha=1/6, type='norm', level=.95) +
  geom_point(aes(colour=site)) + 
  xlab('Resistance to 2005 drouht event') + 
  ylab('Resistance to 2012 drouht event') + 
  theme_bw()+
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) 
```

```{r}
# Recovery
evi_rc <- evi %>% ggplot(aes(x=rc.2005, y=rc.2012, fill=site)) + 
  stat_ellipse(aes(fill=site), geom = 'polygon', alpha=1/6, type='norm', level=.95) +
  geom_point(aes(colour=site)) + 
  xlab('Recovery to 2005 drouht event') + 
  ylab('Recovery to 2012 drouht event') + 
  theme_bw()+
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = 1) 
```

```{r plot_resiliences_evi_ci, fig.width = 12, fig.height=12}
grid.arrange(bai_rc, bai_rt, bai_rs, ncol=3)
```

```{r, echo=FALSE}
pdf(paste0(di, '/out/mean_resilience/evi/interaction_plotsSE.pdf'), width=9, height = 9)
grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
dev.off()

grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)

### 2017-12-11 pres/state_ms.Rmd
update
add new slides

@@
 
-# Explore EVI mean
+# EVI profile comparison
 
@@
 
-# R - EVI anomalies
+# EVI Resilience metrics
 
-![](../analysis/evi/explore_anomalies_files/figure-markdown_github/plot_anoma_EVI_clu_sa-1.png){width=90%}
+![](../analysis/evi/explore_resilience_files/figure-markdown_github/robust_plot_evi_drought-1.png)
 
@@
 
-# Notas anomalias para incluir
+```{r, echo=FALSE}
+library(knitr)
+df <- read.csv(file='../out/anovas_resilience/huber_evi/trimmed_anovas.csv', header=TRUE)
 
-* EVI sa menores en 2005 *** (lowest) que en 2012, sobre todo para poblaciones del norte
-* Criterio Samanta et al. > + 1 greening | < -1 bronwing
-
-https://github.com/ajpelu/qpyr_dendro/blob/master/analysis/evi/explore_anomalies.md
+kable(df)
+```
 
@@
 
-# Mapa de anomalias estandarizadas
+```{r, echo=FALSE}
+library(knitr)
+df <- read.csv(file='../out/anovas_resilience/huber_evi/robust_mhuber_agg.csv', header=TRUE)
 
-![](../analysis/evi/explore_anomalies_files/figure-markdown_github/map_anomaly_2005_2012_sa-1.png){width=90%}
+kable(df)
+```
+
+---
+
+```{r, echo=FALSE}
+library(knitr)
+df <- read.csv(file='../out/anovas_resilience/huber_evi/robust_mhuber_agg_a.csv', header=TRUE)
+
+kable(df)
+```
+
+---
+
+```{r, echo=FALSE}
+library(knitr)
+df <- read.csv(file='../out/anovas_resilience/huber_evi/robust_mhuber_agg_b.csv', header=TRUE)
+
+kable(df)
+```
 

### 2017-12-11 man/ms_2_results.Rmd
update results

@@
 
-Aproximadamete el x% de los robledales mostraron un incremento en los valores de productividad (greenees) ... (relacionar con tendencias EVI y ontologias).
-
 Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (Fig.
@@
 
-En 2005 la mayoría de los pixeles mostraron browning (99.36 % y 79.37 % para las poblaciones del norte y del sur respectivamente), mientras que en 2012 la mayoría de los pixeles no mostraron cambio
+Durante 2005 la mayoría de los pixeles mostraron browning (99.36 % y 79.37 % para las poblaciones del norte y del sur respectivamente), sin embargo en 2012 la mayoría de los pixeles se clasificaron como no changes en las poblaciones del norte (89.60 %) y en las del sur (70.07 %) (Tabla supplementaria??)
 
-de los pixels que cubren las poblaciones del norte mostraron browning durante la sequía de 2005,
-
-La mayor parte del área Tanto en las poblaciones del norte como en las del sur se observó que l
-
-La mayoría de los pixels que cubren los robledales de las poblaciones del norte y las del sur mostraron browning para la sequía del 2005, sin embargo en 2012 la mayoría de los pixeles se clasificaron como no changes en las poblaciones del norte (89.60 %) y en las del sur (70.07 %) (Tabla supplementaria??)
 * Si atendemos a las sa (standardized anomalíes) y aplicamos el criterio de Gao, podemos decir que en 2005 se observó un bronwing en los bosques de Q.
@@
 
-En 2005 y 2012 se observó un browning para los valores medios de EVI. En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.
+Aproximadamete el x% de los robledales mostraron un incremento en los valores de productividad (greenees) ... (relacionar con tendencias EVI y ontologias).
 
-!!! Vas por aqui pollo
+Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI.
+En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.
 
-Pyrenean oak forests showed significantly lower resistance to 2005 drought event than to 2012 one [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001].
+#### Resilience metrics
+
+Pyrenean oak forests showed significantly lower resistance to 2005 drought event than to 2012 one [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); table R1; Figura R4a, F = 799.86, p <0.0001].
+The 2005 drought reduced the greenness of oak to 85.8 % while the 2012 reduced 94.3 %.
+Southern populations showed significantly higher values of resistance to drought than northern ones, except for 2012 where non-significant differences were recorded (table R1, Figure R4a).
+
+> Resistencia
+> 
+> * Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001]
+> * Menor resistencia de las poblaciones del Norte a los eventos de sequía que las del Sur [N: 0.883 (0.877-0.889); S: 0.921 (0.918 - 0.925); p <0.0001]
+> * La resistencia varió en función de la sequía y de la población.
+Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
+
+table R1. Robust anovas con F-Values 
+Figura R4. Interaction plot resilience metrics
 
@@
 ## Resiliencia EVI
-
-* Resistencia
-
-    * Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001]
-    * Menor resistencia de las poblaciones del Norte a los eventos de sequía que las del Sur [N: 0.883 (0.877-0.889); S: 0.921 (0.918 - 0.925); p <0.0001]
-    * La resistencia varió en función de la sequía y de la población.
-Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
 

### 2017-12-11 man/ms_2_results.Rmd
update results

@@
 Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
+> * notas:  La sequía de 2005 redujo el EVI medio hasta el 85.8%, mientras que la de 2012 la redujo hasta el 94.3%.
+La sequía de 2005 redujo el evi medio hasta el 81.9 % en las poblaciones del N, y hasta el 90.2 % en las del S.
+...
+drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)
+
+The oak forests recovered their greenness more rapidly after the 2005 drought than after 2012.
+In the period after 2005 drought, greeness achieved was 112 % (Rc = 1.12) and after 2012 105.7 % (Rc = 1.057).
+A similar recovery after the 2005 and 2012 drought event was observed for southern populations (p = 0.2453; Figure R4), whilst the northern populations showed a significantly greater recovery after the 2005 drought than after the 2012 drought.
+
+> Recovery 
+> 
+> * La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
+> * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
+> * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001].
+En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
+
+-----> POR AQUI POLLITO
+
+Resilience values indicate that greeness was rather similar
+
+    * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
+    * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152).
+En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
 
@@
 
-* Menor resistencia de los robledales a la sequía de 2005 que a la de 2012
-
-Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001], siendo diferencial el comportamiento en la sequía de 2005, donde se observó menor resistencia de las poblaciones del norte.
-
 ## Resiliencia EVI
-
-La sequía de 2005 redujo el EVI medio hasta el 85.8%, mientras que la de 2012 la redujo hasta el 94.3%.
-
-La sequía de 2005 redujo el evi medio hasta el 81.9 % en las poblaciones del N, y hasta el 90.2 % en las del S
-
-... drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)
-
-* Recovery
-
-    * La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
-    * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
-    * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001].
-En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
 

### 2017-12-12 man/ms_2_results.Rmd
update resilience results

@@
 
+Resilience values were significantly higher for the 2012 drought event than for 2005, although both values were close to 1 indicating that greenness level was rather similar after each disturbance event.
+The southern populations showed higher resilience values than the northern ones, although they were not significantly different for 2005 drought (p = 0.036).
+
+> Resilience 
+> * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
+> * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152).
+En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
+
 -----> POR AQUI POLLITO
-
-Resilience values indicate that greeness was rather similar
-
-    * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
-    * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152).
-En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
 
@@
 
-## Resiliencia EVI
-
-* Resilience
-
-    * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
-    * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152).
-En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
-
-* Relative Resilience
+> Relative Resilience (De esto no he dicho nada)
 

### 2017-12-12 pres/state_ms.Rmd
update state

@@
 knitr::opts_chunk$set(echo = TRUE)
+library(tidyverse)
+library(here)
+library(pander)
+library(kableExtra)
 ```
@@
 ```{r, echo=FALSE}
-library(knitr)
-df <- read.csv(file='../out/anovas_resilience/huber_evi/trimmed_anovas.csv', header=TRUE)
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
 
-kable(df)
+dff <- df %>% gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+kable(dff, format = "latex", booktabs = T,
+      col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+      digits=c(0,2,3,2,3,2,3,2,3)) %>% 
+  kable_styling(font_size = 8) %>% 
+  add_header_above(c("", "rc" = 2, "rrs" = 2, "rs" = 2, "rt" = 2))
+
+# pandoc.table(dff,
+#              round=c(0,2,3,2,3,2,3,2,3),
+#              digits=c(0,5,4,5,4,5,4,5,4)
+#              )
 ```

### 2017-12-12 man/ms_2_results.Rmd
include table

@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r}
+library(tidyverse)
+library(here)
+library(pander)
+library(kableExtra)
+```
+
+```{r setup, echo=FALSE}
+show_text <- FALSE
 ```
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 > Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005.
@@
 > * En 2005 y en 2012 las anomalías fueron negativas.
-> Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.
+> Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.")
+```
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 > Resistencia
@@
 drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)
+")
+```
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 > Recovery 
@@
 En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
+")
+```
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 > Resilience 
@@
 En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
+")
+```
+
+```{r, echo=FALSE}
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df %>% gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+kable(dff, format = "latex", booktabs = T,
+      col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+      digits=c(0,2,3,2,3,2,3,2,3)) %>% 
+  kable_styling(font_size = 8) %>% 
+  add_header_above(c("", "rc" = 2, "rrs" = 2, "rs" = 2, "rt" = 2))
+```
 

### 2017-12-12 man/ms_tables.Rmd
add tables

--- 2017-12-11 man/ms_2_results.Rmd
@@
 ---
-title: "ms_2_results"
+title: "ms_tables"
 output:  
@@
 
-# Results
+```{r, message=FALSE, echo=FALSE}
+library(tidyverse)
+library(here)
+library(pander)
+library(kableExtra)
+library(knitr)
+library(dplyr)
+```
 
-## Vegetation Greenness
+### Table R4
 
-Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (Fig.
-R1 Profile EVI).
-The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Fig.
-R2 y R3).
+```{r, echo=FALSE}
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
 
-Esta disminución para 2005 sin embargo fue heterogénea (ver plot de trajectorias)
+dff <- df %>% gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
 
-Durante 2005 la mayoría de los pixeles mostraron browning (99.36 % y 79.37 % para las poblaciones del norte y del sur respectivamente), sin embargo en 2012 la mayoría de los pixeles se clasificaron como no changes en las poblaciones del norte (89.60 %) y en las del sur (70.07 %) (Tabla supplementaria??)
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
 
-* Si atendemos a las sa (standardized anomalíes) y aplicamos el criterio de Gao, podemos decir que en 2005 se observó un bronwing en los bosques de Q.
-pyrenaica, sobre todo en las situadas en el northern slopes.
+colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rc\np", "rs\nF", "rc\np", "rt\nF", "rc\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
 
-> Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005.
-Sin embargo vemos valores menos negativos para 2005.
-Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years
-
-> Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029)que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
-
-> * En 2005 y en 2012 las anomalías fueron negativas.
-> Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.
-
-#### Temporal trend EVI values 
-El 78.95 % de los pixeles mostraron una tendencia positiva en cuanto al EVI medio anual (siendo significativa para el 31.67 % de los pixeles).
-Esta tendencia positiva fué sobre todo mayor en algunas de las poblaciones del suroeste.
-La
-
---> referencia a las tendencias en EVI (lo hemos vuelto a calcular y además en el trabajo de ontologías también nos sale)
-
-Aproximadamete el x% de los robledales mostraron un incremento en los valores de productividad (greenees) ... (relacionar con tendencias EVI y ontologias).
-
-Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI.
-En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.
-
-#### Resilience metrics
-
-Pyrenean oak forests showed significantly lower resistance to 2005 drought event than to 2012 one [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); table R1; Figura R4a, F = 799.86, p <0.0001].
-The 2005 drought reduced the greenness of oak to 85.8 % while the 2012 reduced 94.3 %.
-Southern populations showed significantly higher values of resistance to drought than northern ones, except for 2012 where non-significant differences were recorded (table R1, Figure R4a).
-
-> Resistencia
-> 
-> * Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001]
-> * Menor resistencia de las poblaciones del Norte a los eventos de sequía que las del Sur [N: 0.883 (0.877-0.889); S: 0.921 (0.918 - 0.925); p <0.0001]
-> * La resistencia varió en función de la sequía y de la población.
-Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
-> * notas:  La sequía de 2005 redujo el EVI medio hasta el 85.8%, mientras que la de 2012 la redujo hasta el 94.3%.
-La sequía de 2005 redujo el evi medio hasta el 81.9 % en las poblaciones del N, y hasta el 90.2 % en las del S.
-...
-drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)
-
-The oak forests recovered their greenness more rapidly after the 2005 drought than after 2012.
-In the period after 2005 drought, greeness achieved was 112 % (Rc = 1.12) and after 2012 105.7 % (Rc = 1.057).
-A similar recovery after the 2005 and 2012 drought event was observed for southern populations (p = 0.2453; Figure R4), whilst the northern populations showed a significantly greater recovery after the 2005 drought than after the 2012 drought.
-
-> Recovery 
-> 
-> * La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
-> * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
-> * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001].
-En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
-
------> POR AQUI POLLITO
-
-Resilience values indicate that greeness was rather similar
-
-    * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
-    * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152).
-En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
-
-table R1. Robust anovas con F-Values 
-Figura R4. Interaction plot resilience metrics
-
-## Resiliencia EVI
-
-* Resilience
-
-    * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
-    * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152).
-En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
-
-* Relative Resilience
-
-  * Los robledales mostraron mayor resiliencia relativa a la sequía de 2005 que a la de 2012 [2005: 0.099 (0.095-0.105); 2012: 0.053 (0.050 - 0.056); p <0.0001]  
-  * Las poblaciones del sur mostraron menor resiliencia relativa que las del norte [N: 0.086 (0.082-0.092); S: 0.063 (0.060 - 0.066); p <0.0001], debido sobre todo a la diferencia en la resiliencia relativa para las poblaciones del norte entre los dos eventos de sequía (mucha mayor resiliencia relativa en 2012 que en 2005 para las poblaciones del N)
-  * Las poblaciones del sur no mostraron diferencias en cuanto a la resiliencia relativa entre los dos eventos de sequía (padj = 0.152)
-
-## Resiliencia BAI
-
-* Resistance: 
-  * No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012.
-Valores de resistencia menor al evento de 2005.
-[2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.03]
-  * Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001).
-Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].
-  * La interacción también fue significativa.
-De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ???
-
-* Recovery 
-    * En general mayor recuperación para el evento de 2012 que para el evento de 2005 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
-
-## References
+```

### 2017-12-12 man/ms_figures.Rmd
create figures

---
title: "ms_figures"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../references.bib
csl: ../ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")

library("grid")
library("gridExtra")
```

### Figure 2

```{r, echo=FALSE}
iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')

drought_years <- c(2005, 2012)

evi_profile_dat_ref <- iv %>% 
  filter(!year %in% drought_years) %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
            sd = sd(evi, na.rm=TRUE),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
  mutate(period = 'reference')

evi_profile_dat_2005 <- iv %>% 
  filter(year == 2005) %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
            sd = sd(evi, na.rm=TRUE),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
  mutate(period = '2005')

evi_profile_dat_2012 <- iv %>% 
  filter(year == 2012) %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
            sd = sd(evi, na.rm=TRUE),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
  mutate(period = '2012')

# Join the three dataframes
evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)

# colref <- '#455883'
# colref2 <- '#10253F'
# col2005 <- '#00BA38'
# col2012 <- '#F8766D'

col2012 <- '#0700fe'
col2005 <- '#19e00b'
colref <- '#7b7b7b'

profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
  geom_line(size=.9) + 
  geom_point(size=3, fill='white', shape=21) +
  scale_x_date(labels = function(x) format(x, "%b"),
               breaks = date_breaks('month')) + 
  ylab('EVI') + xlab('Date') + 
  theme_bw() +
  theme_classic() + 
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        legend.position = c(0.1,.9)) +
  scale_color_manual(values=c(col2005, col2012, colref))

# profile_compara 
```

```{r, echo=FALSE}
# From explore_anomalies.Rmd 
# Read data 
anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')

anomalias_evimean <- anomalias_evimean %>% 
  mutate(
    clu_pop = as.factor(case_when(
      pop == 1 ~ "Camarate",
      pop %in% c(2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out')),
    clu_pop2 = as.factor(case_when(
      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out'))) %>% 
  filter(clu_pop != 'out')

# colours 
color_neg <- '#a63603'
color_pos <- '#006d2c'

# Standardized Anomalies 
avg_sa_clu <- anomalias_evimean %>% 
  group_by(clu_pop2, y) %>% 
  dplyr::summarise(mean=mean(sa, na.rm=T),
            sd = sd(sa, na.rm=T),
            se = sd/sqrt(length(sa))) %>% 
 mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))

myylab <- 'EVI Standardized Anomaly'

plot_sa_clu  <- avg_sa_clu %>%  
  ggplot(aes(x=y, y=mean, fill=signo)) + 
  geom_bar(stat = "identity") + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
  facet_wrap(~clu_pop2, nrow = 2) + 
  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  ylab(myylab) + xlab('') +
  theme_bw() +
  theme(text = element_text(size=16),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "white"))

# plot_sa_clu
```

```{r, fig.height=9, echo=FALSE} 
grid.arrange(profile_compara, plot_sa_clu)
```

### Figure R2 = 3 
```{r, echo=FALSE}
mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)

# NO plot rrs 
mhuber_filter <- mhuber %>% filter(var != 'rrs')

pd <- position_dodge(.2)

robust_plot_evi_drought <-
  ggplot(mhuber_filter, 
         aes(x = as.factor(disturb_year), y = M.Huber, 
             color = site, 
             group = site, 
             fill = site)) +
  geom_errorbar(
    aes(ymin = lower.ci, ymax = upper.ci),
    width = .1, size = 0.7, position = pd) +
  geom_line(aes(group = site, color = site, linetype = site), position = pd) +
  geom_point(shape = 21, size = 2.5, position = pd) +
  geom_text_repel(aes(label=Letter), 
                  nudge_x = 0.3, size = 5,
                  segment.size = 0) +
  facet_wrap(
    ~var_sorted, nrow = 1, scales = "free_y",
    labeller = as_labeller(c(
      "0_rt" = "Resistance",
      "1_rc" = "Recovery",
      "2_rs" = "Resilience"))) +
  scale_color_manual(values = c("black", "blue")) +
  scale_fill_manual(values = c("black", "blue")) + theme_bw() +
  scale_linetype_manual(values = c("solid", "dashed")) +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(
      colour = "black",
      fill = "white"),
    legend.position = "bottom") +
  ylab("") + xlab("Drought event")

robust_plot_evi_drought 
```

### 2017-12-12 man/ms_2_results.Rmd
update results

@@
 library(kableExtra)
+library(knitr)
 ```
@@
 
-```{r, echo=FALSE}
-df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
-names(df) <- c('factores', 'F', 'p', 'variables')
+table R1. Robust anovas con F-Values 
+Figura R2. Interaction plot resilience metrics
 
-dff <- df %>% gather(temp, numeros, F, p) %>% 
-  unite(temp1, variables, temp) %>% 
-  spread(temp1, numeros) %>% 
-  mutate(factores = recode(factores, 
-                           "disturb_year" = "Disturb", 
-                           "disturb_year:site" = "Disturb X Site", 
-                           "site" = "Site")) %>% 
-  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
-
-kable(dff, format = "latex", booktabs = T,
-      col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
-      digits=c(0,2,3,2,3,2,3,2,3)) %>% 
-  kable_styling(font_size = 8) %>% 
-  add_header_above(c("", "rc" = 2, "rrs" = 2, "rs" = 2, "rt" = 2))
-```
-
------> POR AQUI POLLITO
-
-table R1. Robust anovas con F-Values 
-Figura R4. Interaction plot resilience metrics
-
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 > Relative Resilience (De esto no he dicho nada)
@@
   * Las poblaciones del sur no mostraron diferencias en cuanto a la resiliencia relativa entre los dos eventos de sequía (padj = 0.152)
+")
+```
+
+-----> POR AQUI POLLITO
 

### 2017-12-15 analysis/bai/analysis_resilience_bai.Rmd
update anovas resiliences

--- 2017-09-15 analysis/analysis_resilience_bai.Rmd
@@
 ---
-title: "Growth Resilience"
+title: "Growth Resilience (reanalisis)"
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 June"
-output:
-  md_document:
-    variant: markdown_github
+date: "2017 Dec"
+output: pdf_document 
 bibliography: ../references.bib 
@@
 ```{r global_options, include=FALSE}
-knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, error=TRUE, message=FALSE)
-```
-
-```{r, warning=FALSE, message=FALSE}
+knitr::opts_chunk$set(fig.width=12, fig.height=8, warning=FALSE, error=TRUE, message=FALSE)
+```
+
+```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
 library("tidyverse")
 library("dplR")
-library("stringr")
 library("knitr")
@@
 library('gridExtra')
+library('stringr')
 library('pander')
@@
 
-```{r, echo=FALSE}
+```{r}
 machine <- 'ajpelu'
 # machine <- 'ajpeluLap'
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
+di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
 
@@
 
-```{r}
+```{r, echo=FALSE}
 # Replace SNA by SJ and SNB by CA
@@
 
-```{r}
+```{r, echo=FALSE}
 # Create subset to compare between sites 
@@
 * Computar tres eventos climáticos: 1995, 2005, 2012
-* Computar dos ventanas temporales: 2 y 3
+* Computar ventanas temporales: 2, 3 y 4
 
@@
 ### Computar correlaciones ventanas temporales 
-```{r}
+```{r, echo = FALSE}
 # Vector with objects name
@@
 correlations <- inner_join(cor2, cor3, by='ind') %>% inner_join(cor4, by='ind')
-
+```
+
+#### Resistance
+
+```{r correla_temporal_window_rt, echo = FALSE}
 # Resistance
@@
 grid.arrange(p_rt23, p_rt24, p_rt34,ncol=3) 
-
+```
+
+#### Recovery
+
+```{r correla_temporal_window_rc, echo = FALSE}
 # Recovery 
@@
 grid.arrange(p_rc23, p_rc24, p_rc34,ncol=3) 
-
+```
+
+#### Resilience
+
+```{r correla_temporal_window_rs, echo = FALSE}
 # Resilience
@@
 grid.arrange(p_rs23, p_rs24, p_rs34,ncol=3) 
-
+```
+
+#### Relative Resilience
+
+```{r correla_temporal_window_rrs, echo = FALSE}
 # Relative Resilience
@@
 grid.arrange(p_rrs23, p_rrs24, p_rrs34,ncol=3) 
-
+```
+
+#### Coeficientes de correlacion entre ventanas temporales
+
+```{r echo=FALSE}
+# Export correlations window size 
 names(aux_coefs) <- c('var', 'window_size', 'r2')
@@
 ## Boxplot with outliers 
-```{r}
+```{r growth_disturbances_boxplot, echo = FALSE}
 gsj <- res_3_sj$growth %>% mutate(site='SJ')
@@
 ## Mean + se 
-```{r}
+```{r growth_disturbances_mean_se, echo = FALSE}
 g %>% 
@@
 
-```{r}
+```{r growth_disturbances_mean_se_colours, echo = FALSE}
 pgrowht <- g %>% 
@@
   theme_bw() + ylab('BAI (cm2/year)') + 
+  scale_shape_manual(values=c(15,19,19)) + 
   theme(panel.grid = element_blank())
@@
 ```
+
+## tables
 
@@
 
-```{r}
+```{r, echo = FALSE}
 shapirosNormal <- function(df, resp_var, factor_vars) {
@@
 
-```{r}
+```{r, echo = FALSE}
 # See auxiliar::shapirosNormal
@@
 ```{r}
+normtestA <- rbind(nrcA, nrtA, nrsA, nrrsA) 
+normtestA %>% pander()
+write.csv(normtestA, 
+          file=paste0(di, '/out/anovas_resilience/bai/normo_disturb_year.csv'), row.names = F)
+```
+
+```{r}
+normtestB <- rbind(nrcB, nrtB, nrsB, nrrsB) 
+normtestB %>% pander()
+write.csv(normtestB, 
+          file=paste0(di, '/out/anovas_resilience/bai/normo_site.csv'), row.names = F)
+```
+
+```{r}
 normtestAB <- rbind(nrcAB, nrtAB, nrsAB, nrrsAB) 
@@
 write.csv(normtestAB, 
-          file=paste0(di, '/out/anovas_resilience/normo_disturb_year_site.csv'), row.names = F)
-```
-
-```{r}
-rm(nrcA, nrcB, nrcAB,
-   nrsA, nrsB, nrsAB,
-   nrrsA, nrrsB, nrrsAB,
-   nrtA, nrtB, nrtAB)
+          file=paste0(di, '/out/anovas_resilience/bai/normo_disturb_year_site.csv'), row.names = F)
+```
+
+```{r}
+# rm(nrcA, nrcB, nrcAB,
+#    nrsA, nrsB, nrsAB,
+#    nrrsA, nrrsB, nrrsAB,
+#    nrtA, nrtB, nrtAB)
 ```
@@
 
-```{r}
+```{r, echo = FALSE}
 ## See auxiliar::homogetest
@@
 write.csv(homo, 
-          file=paste0(di, '/out/anovas_resilience/homocedasticidad.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/bai/homocedasticidad.csv'), row.names = F)
 ```
@@
 
+# Transformación datos
+
+## Log
+
 * Probamos a transformar los datos con log y reanalizar los supuestos de homocedasticidad
 
-```{r}
+```{r, echo = FALSE}
 factores <- c('disturb_year', 'site', 'interaction(disturb_year, site)') 
@@
 write.csv(homo_log, 
-          file=paste0(di, '/out/anovas_resilience/homocedasticidad_log.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/bai/homocedasticidad_log.csv'), row.names = F)
 ```
@@
 
-## Custom functions
-
-```{r}
+# ANOVAS
+
+* Utilizamos una custom function
+
+```{r, echo = FALSE}
 # Custom Function to compute ANOVAS
@@
 ## Recovery 
-```{r}
+```{r, echo=FALSE}
 # Variable
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rc_site, fig.width = 4, fig.height=4}
 ps
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rc_disturb, fig.width = 4, fig.height=4}
 pd
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rc_inter_collapse, fig.width = 4, fig.height=4}
 picollapse
@@
 
-```{r, fig.width = 6}
+```{r aov_rc_inter, fig.width = 6}
 pi
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rt_site, fig.width = 4, fig.height=4}
 ps
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rt_disturb, fig.width = 4, fig.height=4}
 pd
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rt_inter_collapse, fig.width = 4, fig.height=4}
 picollapse
@@
 
-```{r, fig.width = 6}
+```{r aov_rt_inter, fig.width = 6}
 pi
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rrs_site, fig.width = 4, fig.height=4}
 ps
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rrs_disturb, fig.width = 4, fig.height=4}
 pd
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rrs_inter_collapse, fig.width = 4, fig.height=4}
 picollapse
@@
 
-```{r, fig.width = 6}
+```{r aov_rrs_inter, fig.width = 6}
 pi
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rs_site, fig.width = 4, fig.height=4}
 ps
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rs_disturb, fig.width = 4, fig.height=4}
 pd
@@
 
-```{r, fig.width = 4, fig.height=4}
+```{r aov_rs_inter_collapse, fig.width = 4, fig.height=4}
 picollapse
@@
 
-```{r, fig.width = 6}
+```{r aov_rs_inter, fig.width = 6}
 pi
@@
 
-```{r}
-pdf(paste0(di, 'out/fig/resilience/interaction_plotsSE.pdf'), width=9, height = 9)
+### mean + sd
+```{r plot_resiliences_bai_sd,fig.width = 12, fig.height=12}
+grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
+```
+
+### mean + ci
+```{r plot_resiliences_bai_ci, fig.width = 12, fig.height=12}
+grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
+```
+
+```{r, echo=FALSE}
+pdf(paste0(di, '/out/mean_resilience/bai/interaction_plotsSE.pdf'), width=9, height = 9)
 grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
@@
 
-pdf(paste0(di, 'out/fig/resilience/interaction_plotsCI.pdf'), width=9, height = 9)
+pdf(paste0(di, '/out/mean_resilience/bai/interaction_plotsCI.pdf'), width=9, height = 9)
 grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
 dev.off()
-
-```
-
-```{r}
+```
+
+```{r, echo=FALSE}
 aovas_coeff <- aov_rc$model_coeff %>% mutate(var = 'rc') %>% 
@@
 
-write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/anovas_statistics.csv'), row.names = F)
+write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/bai/anovas_statistics.csv'), row.names = F)
 
@@
 write.csv(aovas_model_summary, 
-          file=paste0(di, '/out/anovas_resilience/anovas_summary_modelos.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/bai/anovas_summary_modelos.csv'), row.names = F)
 
@@
     p_value = c(x$A.p.value, x$B.p.value, x$AB.p.value))
+
+  out_raTrimmed <- t2way(formulaFull, data = df)
+
+  out_ratr_df <- data.frame(fact = c(out_raTrimmed$varnames[2], 
+                                    out_raTrimmed$varnames[3], 
+                                    paste0(out_raTrimmed$varnames[2], ':', out_raTrimmed$varnames[3])),
+              statistic = c(out_raTrimmed$Qa, out_raTrimmed$Qb, out_raTrimmed$Qab),
+              pvalue = c(out_raTrimmed$A.p.value, out_raTrimmed$B.p.value, out_raTrimmed$AB.p.value))
 
@@
   out$mest_b <- mest_b
+  out$raTrimmed <- out_raTrimmed
+  out$out_ratr_df <- out_ratr_df
   out$ra <- out_ra # Output for Two-way robust analysis (M-estimators)
@@
 
-#### Rt Letters 
+#### RRs Letters 
 ```{r}
@@
 
-```{r}
+```{r robust_plot_bai_drought}
 mhuber<- mhuber %>% 
@@
 
-robust_plot_evi_drought <- ggplot(mhuber, aes(x=disturb_year, y=M.Huber, color = site, group=site, fill=site)) + 
+robust_plot_bai_drought <- ggplot(mhuber, aes(x=disturb_year, y=M.Huber, color = site, group=site, fill=site)) + 
   geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), 
@@
 
-robust_plot_evi_drought
-
-pdf(paste0(di, 'out/fig/resilience/robust_plot_evi_drought.pdf'), width=9, height = 9)
-robust_plot_evi_drought 
+robust_plot_bai_drought
+
+pdf(paste0(di, 'out/resilience_robust/bai/robust_plot_bai_drought.pdf'), width=9, height = 9)
+robust_plot_bai_drought 
 dev.off()
@@
 ### Response ~ (x=site)
-```{r}
+```{r robust_plot_bai_site}
 pd <- position_dodge(.2)
 
-robust_plot_evi_site <- ggplot(mhuber, aes(x=site, y=M.Huber, color = disturb_year, group=disturb_year, fill=disturb_year)) + 
+robust_plot_bai_site <- ggplot(mhuber, aes(x=site, y=M.Huber, color = disturb_year, group=disturb_year, fill=disturb_year)) + 
   geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), 
@@
 
-robust_plot_evi_site
-
-pdf(paste0(di, 'out/fig/resilience/robust_plot_evi_site.pdf'), width=9, height = 9)
-robust_plot_evi_site 
+robust_plot_bai_site
+
+pdf(paste0(di, 'out/resilience_robust/bai/robust_plot_bai_site.pdf'), width=9, height = 9)
+robust_plot_bai_site 
 dev.off()
@@
 ```{r}
+
+rars$out_ratr_df$var <- 'rs'
+rarc$out_ratr_df$var <- 'rc'
+rart$out_ratr_df$var <- 'rt'
+rarrs$out_ratr_df$var <- 'rrs'
+
+trimmedanovas <- rbind(rarc$out_ratr_df, rart$out_ratr_df, rars$out_ratr_df, rarrs$out_ratr_df) 
+trimmedanovas %>% pander()
+
+```
+
+```{r}
 # Export data 
-write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/robust_mhuber.csv'), row.names = F)
-write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg.csv'), row.names = F)
-write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_a.csv'), row.names = F)
-write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg_a.csv'), row.names = F)
-write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_b.csv'), row.names = F)
-write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg_b.csv'), row.names = F)
-
-write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/robust_pairwise.csv'), row.names = F)
+write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber.csv'), row.names = F)
+write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg.csv'), row.names = F)
+write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_a.csv'), row.names = F)
+write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg_a.csv'), row.names = F)
+write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_b.csv'), row.names = F)
+write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg_b.csv'), row.names = F)
+
+write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_pairwise.csv'), row.names = F)
+
+write.csv(trimmedanovas, file=paste0(di, '/out/anovas_resilience/huber_bai/trimmed_anovas.csv'), row.names = F)
 ```

### 2017-12-18 man/ms_2_results.Rmd
update results

@@
 
-#### Resilience metrics
+#### Resilience metrics EVI
 
@@
 
------> POR AQUI POLLITO
-
 ## Resiliencia BAI
 
-* Resistance: 
+Although no significant differences were observed in the resistance of the oak growth to both drought events (F = 6.0189; p = 0.019; table R2; Figura R5), the 2005 drought reduced growth more than 2012 did (Rt = 0.721 and 0.819 respectively).
+Northern population showed resistance values lower than southern ones, especially for 2005 drought event where the growth was reduced to 44.5 % of the preceding period.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+> Resistance:
+
 * No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012.
 Valores de resistencia menor al evento de 2005.
-[2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.03]
+[2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.025; trimmed-p = 0.019]
+
 * Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001).
 Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].
+
 * La interacción también fue significativa.
 De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ???
+")
+```
 
-* Recovery 
-    * En general mayor recuperación para el evento de 2012 que para el evento de 2005 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
+Oak forests recovered their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1.
+Norhtern populations showed signigicantly higher values of recovery than southern ones, which even not reach recovery values of Rc = 1.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+> Recovery 
+> * Mayor recuperación del crecimiento de los robledales tras la sequía de 2005 que tras la de 2012 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
+> * Las poblacionees del sur (caH y caL; no diferencias entre ellas) mostraron una menor recuperación que las poblaciones del norte, es decir, las del norte se recuperaron mas rápidamente que las del sur que no llegaron incluso a valores de Rc = 1.
+") 
+```
+
+Significantly higher values of resilience of growth were observerd for the 2012 drought event than for the 2005 one.
+We found no differences of resilience values for growth between northern and southern populations (F = 1.31; p = 0.404).
+For 2005 drought event the resilience of northern population was lower than southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity.
+
+notas para discusión
+
+En ambos casos (evi y bai) la sequía de 2005 y de 2012 redujeron el greenness y el crecimiento.
+Además en ambos casos la sequía de 2005 redujo mas la variable, sin embargo, para el greenness si existen diferencias entre ambos eventos de sequía, mientras que para el crecimiento no.
+Por tanto podemos decir que se ha reducido el greennes pero para el crecimiento esto está mas amortiguado.
+
+Al explorar la resistencia por sitio y drought se observa una curiosidad.
+La resistencia de los robledales de CaLow se mantiene casi igual.
+La de la cara norte (SJ) sufre mucho en 2005 y luego se recupera un poco.
+La de caH presenta menor resistencia en 2012 que en 2005 (y eso llama la atención)
 

### 2017-12-18 analysis/bai/analysis_resilience_bai.Rmd
resilience bai render md

@@
 date: "2017 Dec"
-output: pdf_document 
+output: 
+  md_document:
+    variant: markdown_github
 bibliography: ../references.bib 

### 2017-12-18 man/ms_tables.Rmd
update tables

@@
 
-colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rc\np", "rs\nF", "rc\np", "rt\nF", "rc\np")
+colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rrs\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
 pander(dff,
@@
 ```
+
+### Table R4 sin rrs
+
+```{r, echo=FALSE}
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df %>% 
+  filter(variables != "rrs") %>% 
+  gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
+
+colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
+
+```

### 2017-12-18 man/ms_tables.Rmd
updating

@@
 
-### Table R4
+```{r, echo=FALSE, include = FALSE}
 
-```{r, echo=FALSE}
+### Table R1_evi
 df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
@@
 
-### Table R4 sin rrs
-
-```{r, echo=FALSE}
+```{r, echo=FALSE, include = FALSE}
+### Table R1_evi sin rrs 
 df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
@@
 
+dff_evi <- dff %>% mutate(target = 'EVI')
+
 colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
@@
 ```
+
+```{r, echo=FALSE, include = FALSE}
+
+### Table R1_bai
+df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df %>% gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
+
+colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rrs\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
+
+```
+
+```{r, echo=FALSE, include = FALSE}
+### Table R1_bai sin rrs 
+df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df %>% 
+  filter(variables != "rrs") %>% 
+  gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+dff_bai <- dff %>% mutate(target = 'BAI')
+
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
+
+colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
+
+```
+
+### Table R1
+
+```{r, echo=FALSE}
+
+dffs <- dff_bai %>% bind_rows(dff_evi)
+
+# write.csv(dffs, here("/out/anovas_resilience", "trimmed_anovas_joined.csv"), row.names = FALSE)
+
+# kable(dffs, format = "latex", booktabs = T,
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "variable"),
+#       digits=c(0,2,3,2,3,2,3,2,3,0)) %>% 
+#   kable_styling(font_size = 8) %>% 
+#   add_header_above(c("", "rc" = 2, "rs" = 2, "rt" = 2))
+
+colnames(dffs) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np", "variable")
+pander(dffs,
+             round=c(0,2,3,2,3,2,3,2,3,0),
+             digits=c(0,5,4,5,4,5,4,5,4,0),
+             keep.line.breaks = TRUE
+             )
+```
+
+# Table RS1 
+* Estimadores huber resiliences EVI
+
+# Table RS2
+
+* Estimadores huber resiliences Bai

### 2017-12-18 man/ms_tables_portrait.Rmd
updating

---
title: "ms_tables_portrait"
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 7.5
    highlight: null
    reference_docx: template_tables_portrait.docx
bibliography: ../references.bib
csl: ../ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
library(dplyr)
```

# Table RS1

* Estimadores huber resiliences EVI

```{r, echo = FALSE, warning = FALSE}
mraw <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)

m <- mraw %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>% 
 unite("mh", c("M.Huber", "ci"), sep = "\n")

# Datos agrupados para todos los sitios 
mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)

m_a <- mraw_a %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)

m_b <- mraw_b %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
  mutate(disturb_year = 2099)

mm <- m %>% 
  bind_rows(m_a) %>%
  bind_rows(m_b) %>% 
  gather(temp, numeros, mh) %>% 
  unite(temp1, c(variables, disturb_year), temp) %>% 
  spread(temp1, numeros, fill = "")

mm <- mm %>% 
  dplyr::select(site, 
                rt_2005_mh, rc_2005_mh, rs_2005_mh,
                rt_2012_mh, rc_2012_mh, rs_2012_mh,
                rt_2099_mh, rc_2099_mh, rs_2099_mh)

colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
pander(mm, keep.line.breaks = TRUE, split.table = Inf)
```

# Table RS2

* Estimadores huber resiliences Bai

```{r, echo = FALSE, warning = FALSE}
mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)

m <- mraw %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>% 
 unite("mh", c("M.Huber", "ci"), sep = "\n")

# Datos agrupados para todos los sitios 
mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)

m_a <- mraw_a %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)

m_b <- mraw_b %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
  mutate(disturb_year = 2099)

mm <- m %>% 
  bind_rows(m_a) %>%
  bind_rows(m_b) %>% 
  gather(temp, numeros, mh) %>% 
  unite(temp1, c(variables, disturb_year), temp) %>% 
  spread(temp1, numeros, fill = "")

mm <- mm %>% 
  dplyr::select(site, 
                rt_2005_mh, rc_2005_mh, rs_2005_mh,
                rt_2012_mh, rc_2012_mh, rs_2012_mh,
                rt_2099_mh, rc_2099_mh, rs_2099_mh)

colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
pander(mm, keep.line.breaks = TRUE, split.table = Inf)
```

```{r, echo = FALSE, warning = FALSE, eval = FALSE}

# Este código es para print la tabla anterior en formato pdf latex (ver state_ms)c
### OJO REVISAR 
# ------------------------------

mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)

m <- mraw %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter))

# Datos agrupados para todos los sitios 
mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)

m_a <- mraw_a %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)

m_b <- mraw_b %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  mutate(disturb_year = 2099)

mm <- m %>% 
  bind_rows(m_a) %>%
  bind_rows(m_b) %>% 
  gather(temp, numeros, c(M.Huber, ci)) %>% 
  unite(temp1, c(variables, disturb_year), temp) %>% 
  spread(temp1, numeros, fill = "")

mm <- mm %>% 
  dplyr::select(site, 
                rc_2005_M.Huber, rc_2005_ci, rt_2005_M.Huber, rt_2005_ci, rs_2005_M.Huber, rs_2005_ci,
                rc_2012_M.Huber, rc_2012_ci, rt_2012_M.Huber, rt_2012_ci, rs_2012_M.Huber, rs_2012_ci,
                rc_2099_M.Huber, rc_2099_ci, rt_2099_M.Huber, rt_2099_ci, rs_2099_M.Huber, rs_2099_ci)

mmt <- mm %>%
   rownames_to_column %>% 
   gather(var, value, -rowname) %>% 
   spread(rowname, value)

colnames(mmt) <- c("var", "SJ", "caH", "caL", "z_all") 
mmt <- mmt[-19,] 
mmt <- mmt %>% 
  mutate(vars = c(rep("Rc",6), rep("Rt",6), rep("Rs",6)),
         year = c(rep(
           c(rep("2005", 2), 
             rep("2012", 2),
             rep(" ", 2)), 3))) %>% 
  dplyr::select(vars, year, SJ, caH, caL, var)

kable(mmt, 
      format = "latex", booktabs = T) %>% 
  kable_styling(font_size = 6) %>% 
  add_header_above(c("", "", "Northern", "Southern" = 2, "")) 
```

### 2017-12-18 pres/state_ms.Rmd
update presentation
updating

--- 2017-12-11 pres/state_ms.Rmd
@@
 ```{r, echo=FALSE}
-library(knitr)
-df <- read.csv(file='../out/anovas_resilience/huber_evi/trimmed_anovas.csv', header=TRUE)
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
 
-kable(df)
+dff <- df  %>%  
+  filter(variables != "rrs") %>% 
+  gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+kable(dff, format = "latex", booktabs = T,
+      col.names = c("", "F", "p", "F", "p", "F", "p"),
+      digits=c(0,2,3,2,3,2,3,2,3)) %>% 
+  kable_styling(font_size = 8) %>% 
+  add_header_above(c("", "rc" = 2, "rs" = 2, "rt" = 2))
+
 ```
@@
 
+# BAI Resilience metrics
+
+![](../analysis/bai/analysis_resilience_bai_files/figure-markdown_github/robust_plot_bai_drought-1.png)
+
+---
+
+```{r, echo=FALSE}
+df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df  %>%  
+  filter(variables != "rrs") %>% 
+  gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+kable(dff, format = "latex", booktabs = T,
+      col.names = c("", "F", "p", "F", "p", "F", "p"),
+      digits=c(0,2,3,2,3,2,3,2,3)) %>% 
+  kable_styling(font_size = 8) %>% 
+  add_header_above(c("", "rc" = 2, "rs" = 2, "rt" = 2))
+
+```
+
+---
+
+```{r, echo=FALSE}
+dffs <- read.csv(here("/out/anovas_resilience", "trimmed_anovas_joined.csv"), header = TRUE)
+
+kable(dffs, format = "latex", booktabs = T,
+      col.names = c("", "F", "p", "F", "p", "F", "p", "variable"),
+      digits=c(0,2,3,2,3,2,3,2,3,0)) %>% 
+  kable_styling(font_size = 8) %>% 
+  add_header_above(c("", "rc" = 2, "rs" = 2, "rt" = 2))
+
+```
+
+---
+
+# OJO CON ESTA TABLA (revisar Tabla RS2)
+
+```{r, echo = FALSE, warning = FALSE}
+mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
+
+m <- mraw %>% 
+  filter(var != 'rrs') %>% 
+  rename(variables = var)  %>% 
+  dplyr::select(c(-n, -Letter))
+
+# Datos agrupados para todos los sitios 
+mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
+
+m_a <- mraw_a %>% 
+  filter(var != 'rrs') %>% 
+  rename(variables = var)  %>% 
+  dplyr::select(c(-n, -Letter)) %>%  
+  mutate(site = "z_all")
+
+# Datos agrupados para todos los sitios 
+mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
+
+m_b <- mraw_b %>% 
+  filter(var != 'rrs') %>% 
+  rename(variables = var)  %>% 
+  dplyr::select(c(-n, -Letter)) %>%  
+  mutate(disturb_year = 2099)
+
+mm <- m %>% 
+  bind_rows(m_a) %>%
+  bind_rows(m_b) %>% 
+  gather(temp, numeros, c(M.Huber, ci)) %>% 
+  unite(temp1, c(variables, disturb_year), temp) %>% 
+  spread(temp1, numeros, fill = "")
+
+mm <- mm %>% 
+  dplyr::select(site, 
+                rc_2005_M.Huber, rc_2005_ci, rt_2005_M.Huber, rt_2005_ci, rs_2005_M.Huber, rs_2005_ci,
+                rc_2012_M.Huber, rc_2012_ci, rt_2012_M.Huber, rt_2012_ci, rs_2012_M.Huber, rs_2012_ci,
+                rc_2099_M.Huber, rc_2099_ci, rt_2099_M.Huber, rt_2099_ci, rs_2099_M.Huber, rs_2099_ci)
+
+mmt <- mm %>%
+   rownames_to_column %>% 
+   gather(var, value, -rowname) %>% 
+   spread(rowname, value)
+
+colnames(mmt) <- c("var", "SJ", "caH", "caL", "z_all") 
+mmt <- mmt[-19,] 
+mmt <- mmt %>% 
+  mutate(vars = c(rep("Rc",6), rep("Rt",6), rep("Rs",6)),
+         year = c(rep(
+           c(rep("2005", 2), 
+             rep("2012", 2),
+             rep(" ", 2)), 3))) %>% 
+  dplyr::select(vars, year, SJ, caH, caL, var)
+
+kable(mmt, 
+      format = "latex", booktabs = T) %>% 
+  kable_styling(font_size = 7) %>% 
+  add_header_above(c("", "", "Northern", "Southern" = 2, "")) 
+```
+
+---
+
 # References

### 2017-12-20 analysis/analysis_chronologies.Rmd
update chronologies

--- 2017-05-31 analysis/analysis_chronologies.Rmd
@@
 ---
-title: "dendro"
+title: "chronologies"
 author: "AJ Perez-Luque (@ajpelu)"
@@
 
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
-    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
-
 # Summary dendrochronology statistics
@@
 rwl_by_site %>% 
-  write.csv(file=paste(di, "data/proto_tables/site_dendro.csv", sep=""), row.names = FALSE)
+  write.csv(file=paste(di, "data/dendro_summary/site_dendro.csv", sep=""), row.names = FALSE)
 
@@
 rwl_by_loc %>% 
-  write.csv(file=paste(di, "data/proto_tables/loc_dendro.csv", sep=""), row.names = FALSE)
+  write.csv(file=paste(di, "data/dendro_summary/loc_dendro.csv", sep=""), row.names = FALSE)
 
@@
 
-```{r}
+```{r chro_sj_compare_method}
 # Robust mean biweith or not
@@
 
-```{r}
+```{r chro_ca_compare_method}
 # Robust mean biweith or not
@@
 
-```{r}
+```{r chrono_sj_two_sites}
 # San Juan 
@@
 
-cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+plot_cro_sj <- cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
   theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                    ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + geom_point(size=1) + 
+  # scale_colour_manual(values = c("red", "blue")) + 
   theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1920, 2020, by=10))
+
+plot_cro_sj
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_sj_HL.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sj, width=12, height = 4) 
 ```
@@
 
-```{r}
+```{r chrono_ca_two_sites}
 # Canar 
@@
 
-cro_ca %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+plot_cro_ca <- cro_ca %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
   theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + scale_colour_manual(values = c("red", "blue")) + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=1) + 
+  # scale_colour_manual(values = c("red", "blue")) + 
   theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep='')))+
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
+
+plot_cro_ca
 ```
@@
 ```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_ca_HL.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_ca, width=12, height = 4) 
+```
+
+```{r chronos_4_sites}
 # Combine and plot 
@@
 
-cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+plot_cro_sites <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
   theme_bw() + ylab('BAI') + 
-  geom_line() + geom_point() + 
-  theme(panel.grid = element_blank()) + facet_wrap(~loc, nrow = 2) + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=1) +
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + facet_wrap(~loc, nrow = 2) + 
   scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) 
+  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
 #  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_sites.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sites, width=12, height = 8)
+```
+
+```{r chronos_4_sites_collapsed}
+plot_cro_sites_collapsed <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  # geom_point(size=1) +
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
+#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sites_collapsed, width=12, height = 4)
+plot_cro_sites_collapsed
+```
+
+```{r chronos_4_sites_collapsed50}
+plot_cro_sites_collapsed50  <- cro_sites %>% 
+  filter(year > 1949) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  # geom_point(size=1) +
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1950, 2020, by=5)) 
+#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+
+plot_cro_sites_collapsed50
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed50.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sites_collapsed50, width=12, height = 4) 
+plot_cro_sites_collapsed50
+```
+
+```{r chronos_4_sites_collapsed90}
+plot_cro_sites_collapsed90  <- cro_sites %>% 
+  filter(year > 1989) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  # geom_point(size=1) +
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1990, 2016, by=2))
+
+plot_cro_sites_collapsed90
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed90.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sites_collapsed90, width=12, height = 4) 
+plot_cro_sites_collapsed90
 ```
@@
 ```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
+# set seed
+set.seed(333)
+
 # Metodo correlation
@@
 correla <- rbind(correla_ca, correla_sj)
+```
 
+```{r correla_chronos_sites, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
 correla %>% 
@@
   theme(strip.background = element_blank())
+```
 
+```{r correla_chronos_sites_pearson}
+correla_sites_pearson <- correla %>% filter(metodo_correla == 'pearson') %>%
+  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
+  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
+  xlab('Smoothing (centred moving average window size)') + 
+  theme(panel.grid = element_blank()) +
+  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
+  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
+  theme(legend.position = c(.8,.7))
+correla_sites_pearson
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/correla_boot_sitesHL.pdf') 
+ggsave(filename=fileplot, plot=correla_sites_pearson, width=5, height = 5) 
 ```
@@
 tc <- c('spearman', 'pearson')
+
+set.seed(333)
 
@@
 
----> here
+```{r correla_sitesSJCALH}
+correla_sitesSJCALH  <- correla %>% 
+  filter(metodo_correla == 'pearson') %>%
+  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
+  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
+  xlab('Smoothing (centred moving average window size)') + 
+  # geom_smooth(method = 'loess', se=FALSE) + 
+  theme(panel.grid = element_blank()) +
+  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
+  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
+  theme(legend.position = c(.8,.9), legend.title = element_blank())
+
+correla_sitesSJCALH
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/correla_boot_sitesSJCALH.pdf') 
+ggsave(filename=fileplot, plot=correla_sitesSJCALH, width=5, height = 5) 
+```
+
+# Chronologies by site (SJ, CAL, CAH)
+
+```{r}
+# Compute la crono de SJ de nuevo y llamar al site SJ
+cro_sjf <- chrono_bai(bai_sj)
+cro_sjf$site <- 'SJ'
+```
+
+```{r crono_compara_sitesSJCALH}
+# Combine and plot 
+cro_sites <- rbind(cro_caL, cro_caH, cro_sjf)
+
+## Export data of bai (chro)
+write.csv(cro_sites, file=paste(di, "data/cronos_medias/cronos_sites.csv", sep=""), row.names = FALSE)
+
+plot_cro_sitesSJCALH <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
+
+plot_cro_sitesSJCALH
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH, width=12, height = 5)
+plot_cro_sitesSJCALH
+```
+
+```{r crono_compara_sitesSJCALH50}
+plot_cro_sitesSJCALH50 <- cro_sites %>% 
+  filter(year > 1949) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
+plot_cro_sitesSJCALH50
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH50.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH50, width=12, height = 5)
+```
+
+```{r crono_compara_sitesSJCALH90}
+plot_cro_sitesSJCALH90 <- cro_sites %>% 
+  filter(year > 1989) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
+plot_cro_sitesSJCALH90
+```
+
+```{r}
+fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH90.pdf') 
+ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH90, width=12, height = 5)
+```
 

### 2017-12-20 analysis/chronos/analysis_chronologies.Rmd
change dir

--- 2017-12-20 analysis/analysis_chronologies.Rmd
@@
     variant: markdown_github
-bibliography: references.bib 
-csl: ecology.csl
+bibliography: ../references.bib
+csl: ../ecology.csl
 ---

### 2017-12-20 reports/report_dendro.Rmd
add report dendro

--- 2017-09-17 man/ms_1_material_methods.Rmd
@@
 ---
-title: "material and methods"
+title: "report_dendro"
 output:  
@@
 
-# Materials and methods
+```{r setup, echo=FALSE}
+show_text <- FALSE
+```
 
-## Species and study site
+# Cronologías
 
-The Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula [@Franco1990] reaching its southern limit in north of Morocco.
-In the Iberian Peninsula these forests live under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002] living on siliceous soils, or soils poor in basic ions [@Vilches2014].
-*Q.
-pyrenaica* requires between 650 and 1200 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm (Martínez-Parras and Molero-Mesa 1982, García and Jiménez 2009), with summer rainfall being a key factor in the distribution of the species (Gavilán et al.
-2007, Río et al.
-2007).
+* Para cada sitio construimos una cronología utilizando las series de BAI. 
+* Hemos utilizado la media aritmética
 
-This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*.
-The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
-There are eight oak patches (2400 Has) identified (**Figure 1**) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys.
-Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation (Brewer et al.
-2002, Olalde et al.
-2002, Rodríguez-Sánchez et al.
-2010) and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change (Hampe and Petit 2005).
+## Cronologías San Juan (northern face)
 
-:red_circle: `duda aqui` Varias referencias hablan de los años 2005 y 2012 como  extremadamente secos.
-Pero habría que hacer alguna referencia y/o análisis.
-Tengo dudas de si hemos de analizar (e incluir) que efectivamente los años 2005 y 2012 fueron caracterizados por un extrema sequía, por lo que habría que  incluyendo referencia a apéndice
-
-* O quizá un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
-
-The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests (Melendo and Valle 2000, Vivero et al.
-2000), undergoing intensive anthropic use in the last few decades (Camacho-Olmedo et al.
-2002, Valbuena-Carabaña et al.
-2010).
-In fact, the status of conservation of this species for southern Spain is "Vulnerable" (Vivero et al.
-2000).
-The relict presence of this species in Sierra Nevada is related both to its genetic resilience as well as to its high intraspecific genetic diversity (Valbuena-Carabaña and Gil 2013).
-However, they are also expected to suffer the impact of climate change, due to their climate requirements (wet summers).
-Thus, simulations of the climate change effects on this habitat forecast a reduction in suitable habitats for Sierra Nevada (Benito et al.
-2011).
-
-## Datos de sequía.
-
-* :red_circle: Meter aquí algunos datos de sequia, similar a lo planteado por Gazol
-
-### Greenness data
-
-To characterize the vegetation greenness of *Q.
-pyrenaica* we used the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
-EVI and NDVI (*Normalized Difference Vegetation Index*) are the most common greenness vegetation indices.
-We used EVI instead of NDVI because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].
-
-EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
-We selected the pixels covering the distribution of *Q.
-pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
-MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016.
-A data filtering was applied to select EVI valid values.
-The filtering was done using quality flags and VI Usefulness Indices accompanying the EVI data.
-We filter out those values affected by high content of aerosols, clouds, shadows, snow or water.
-
-Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2).
-Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked.
-Besides, VI usefulness indices between 0 to 11 essentially include all EVI data.
-Thus, these two conditions serve as additional checks.
-
-and then a quality assessment was carried out to filter the ... [@Reyes2015]
-
-    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds.
-On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]
-
-The presence of clouds (adjacent clouds, mixed clouds and shadows) 'obscures' the surface in a radiometric sense, thus corrupting inferred EVI values.
-In addition, two types of aerosol loadings typically corrupt EVI—climatology and high aerosols.
-Use of aerosol climatology
-
-The EVI data are geometrically and atmospherically corrected and include information about the quality ass....
-
-`$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI.
-
-These data are geometrically and atmospherically corrected, and include an index of data quality (reliability, which range from 0 – good quality data – to 4 – raw data or absent for different reasons) based on the environmental conditions in which the data was recorded
-
-We first used the Quality Assesment (QA band) information of this product to filter out those values affected by high content of aerosols, clouds, shadows, snow or water; and then a quality assessment was carried out to filter the ...
-[@Reyes2015]
-
-:red_circle: reescribir esto de la calidad. Ver Samanta et al 2012 y como describe el proceso de calidad
-
-OJITO---------- reescribir esto de arriba
-
-After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2015.
-(:red_circle: Hemos seleccionado EVI medio, además de por los consejos que me ha dicho Domingo, porque he comprobado que existe una correlación entre el evi medio y el evi estacional, sobre todo el de verano.
-Ver esto: [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md).
-Además presenta alta correlaciones significativas con el EVI de verano: 0.88; de primavera: 0.76 y anual: 0.81)
-
-> Procedimiento de Filtrado de datos (ver [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md))
-
-* Información contenida en banda QA.
-
-    * Nos quedamos con pixeles marcados como Good Data (57.89 %)
-    * Filter out los marcados como Snow/Ice y/o Cloudy (2.57 + 7.08 = 9.65 %)
-    * Pixeles marcados como Marginal Data (32.33 %) (ver siguiente paso)
-
-* Explorar distribución temporal y analizar banda QA Detailed y llevar a cabo un filtrado siguiendo las especificaciones de @Reyes2015.
-
-    * Vemos los composites marcados con Aerosoles, Adjacent cluods, y Shadow.  
-    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds.
-On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]
-
-* Finalmente nos hemos quedado con las siguientes cifras.
-De un total de 360064 images composites for the study zone were downloaded (928 x 20 x 1 + 928 x 23 x 16 = 360064), tras el filtrado, nos quedamos con 286825 (79.65 %)
-
-To explore the effect of drought events on greenness we calculated the EVI standardized anomaly (EVI~sa) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012, @Gao2016].
-For each pixel we averaged all the EVI valid values within a year (:red_filter: see quality filter), and then the standardized anomaly was computed as:
-
-$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
-where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for the year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for the year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
+![]()
 
@@
 For each tree, diameter at breat height (DBH) and total height were recorded.
-Increment cores were air dried, glued onto wooden mounts and sanded.
+
+### Dendroecological analyses 
+Tree cores were air dried, glued onto wooden mounts and sanded.
 Annual radial growth (ring width, RW) were measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.01 mm.
@@
 
-## Dendrochronological methods
-* incluir citas a Frits; Cooks ... 
-* site cronologies
-* BAI 
-* GC Nowacki??
+The growth trends were analyzed at different time scales.
+The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
+BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributatle to increasing stem circumference [@Biondi2008].
+We used measured dbh and raw nondetrened ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
+We calculated a mean BAI series for each individual tree.
+To study the rsponse of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies (growth index, GIres) calculated from biweigth means of ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976]
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+ * BAI represents a more accurate indicator of growth than ring-width, since it reduces the variation caused by adding volume to a circular stem 
+ * BAI is a more meaningful indicator of growth than ring width because it removes variation in growth attributale to increasing circumference. 
+ * BAI is more closely related to biomass increment 
+ * BAI removes variation in growth attributatle to increasing stem circumference and captures changes in growth better than linear measures as tree-ring width [@Biondi2008] 
+")
+```
 
@@
 Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
+
+* incluir citas a Frits; Cooks ...
+
+* BAI 
+* site cronologies
+* GC Nowacki??
 

### 2017-12-20 man/ms_1_material_methods.Rmd
update
material methods

@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r setup, echo=FALSE}
+show_text <- FALSE
 ```
@@
 
+Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1)
+
 ## Field sampling and dendrochronological methods

### 2017-12-20 analysis/chronos/analysis_chronologies.Rmd
update chronos

@@
 
+plot_cro_sites
 ```
@@
 #  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+plot_cro_sites_collapsed
 ```

### 2017-12-20 reports/report_dendro.Rmd
update report dendro

@@
 bibliography: ../references.bib
-templat
 csl: ../geb.csl
@@
 
-# Cronologías
+# BAI
 
-* Para cada sitio construimos una cronología utilizando las series de BAI. 
+## Cronologias
+
+* Para cada sitio construimos una cronología utilizando las series de BAI de cada árbol  
 * Hemos utilizado la media aritmética
 
-## Cronologías San Juan (northern face)
+En los siguientes gráficos se muestran las cronologías medias para cada sitio (mean ± se)
 
-![]()
+#### Figura 1. Cronologías por sitio
+![](../analysis/chronos/analysis_chronologies_files/figure-markdown_github/chronos_4_sites-1.png)
 
-## Field sampling and dendrochronological methods
+### Comparación cronologías 
+Las cronologías de la cara norte (SJ) (SJ-High y SJ-Low)s son muy similares.
+Además en cada sitio de SJ, muestreamos solamente 15 árboles, frente a los 20 de cad sitio de Cáñar (sur).
+Por tanto consideramos incluir solamente dos sitios en cara sur (CA-Low y CA-High) y un solo sitio en cara norte.
+Adicionalmente a la comparación visual de chronos se realizó el siguiente análisis:
 
-### Tree sampling 
-Samplig was carried during autumn of 2016.
-Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1).
-Two elevations were sampled for the southern site (CA-Low and CA-High).
-All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..).
-In each site between 15 and 20 dominant trees were randomly selected.
-Two cores of 5 mm of diameter were taken from each at breast heith (1.3 m) using an increment borer.
-For each tree, diameter at breat height (DBH) and total height were recorded.
-
-### Dendroecological analyses 
-Tree cores were air dried, glued onto wooden mounts and sanded.
-Annual radial growth (ring width, RW) were measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.01 mm.
-Individual ring series were visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
-Validation of the croos-dating was done using COFECHA software [@Holmes1983].
-
-The growth trends were analyzed at different time scales.
-The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
-BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributatle to increasing stem circumference [@Biondi2008].
-We used measured dbh and raw nondetrened ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
-We calculated a mean BAI series for each individual tree.
-To study the rsponse of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies (growth index, GIres) calculated from biweigth means of ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976]
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
- * BAI represents a more accurate indicator of growth than ring-width, since it reduces the variation caused by adding volume to a circular stem 
- * BAI is a more meaningful indicator of growth than ring width because it removes variation in growth attributale to increasing circumference. 
- * BAI is more closely related to biomass increment 
- * BAI removes variation in growth attributatle to increasing stem circumference and captures changes in growth better than linear measures as tree-ring width [@Biondi2008] 
-")
-```
-
-Site chronologies were built by averaging all tree BAI measurement of the same site.
-To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
-
-* incluir citas a Frits; Cooks ...
-
-* BAI 
-* site cronologies
-* GC Nowacki??
-
-## Resilience
-
-To evaluate the effects of the disturbance events on greeennes and tree growth we used four resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), recovery (*Rc*) and relative resilience (*RRs*).
-
-The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and befor the disturbance:
-
-Resistance (*Rt*) = Drought / Predrought
-
-The Recovery index (*Rc*) is the ability to recover from disturbance relative to its severity, and it is estimated as the ratio between performance after and during disturbance:
-
-Recovery (*Rc*) = Postdrought / Drought
-
-The Resilience index (*Rs*) is the capacity to reach pre-disturbance performance levels, and it is estimated as the ratio between the performance after and before disturbance:
-
-Resilience (*Rs*) = Postdrought / Predrought
-
-The Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance, and it is estimated as:
-
-Relative Resilience (*RRs*) = (Postdrought - Drought) / Predrought
-
-We computed the values of these indices for tree growth and greenness during each drought event.
-We considered 2005 and 2012 as singles drought events.
-The predrought and postdrought values of each target variable (i.e.: tree growth or EVI) we computed as the mean value during a period of three years before and after the disturbance events respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (:red_circle: incluir tabla de coeficientes y/o gráfica??
-como suplement, see Gazol 2017)
-
-## Statistical analysis
-
-* Explore anomalies EVI 
-* Explore long and short term trends in RW :red_circle: ver correo Guillermo
-* ANOVA analysis EVI events and populations
-
-We tested for significant differences between drought events (2005 and 2012) and oak population (northern and southern slopes) for each of the resilience indices.
-Robust two-way ANOVAs were used beacuse original and log-transformed data both did not match the assumptions of normality and homogeneity of variance [@Wilcox2012].
-Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to mean value in all cases [@Wilcox2012].
-When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012, @Wilcox2012].
-Post-hoc differences were assessed pairwise using a similar boostrap test.
-All the robust ANOVA and post-hoc tests were carried out using the WRS2 [@Mair2017] and rcompanion [@Mangiafico2017] R packages.
-The level of significance was set at 0.05 and adjusted for multiple comparisons.
-
-## References
+* Se estudió la similitud de las series de cada sitio dentro de cada localidad, es decir, CA-High vs. CA-Low  y SJ-High vs. SJ-Low. 
+* Cada cronología se suavizó utilizando centred moving averages con diferentes tamaño de ventana (entre 1 y 40)
+* Posteriormente se calculó la correlación entre las cronologías de cada sitio suavizadas con una misma ventana temporal, por ejemplo: SJ-High suavidaza con window size 5 años vs.
+SJ-Low suavidaza con window size de 5 años.
+* Además se obtuvieron los intervalos de confianza de dichos coeficientes de correlación mediante boostrap.

### 2017-12-21 reports/report_dendro.Rmd
update report dendro

@@
 * Además se obtuvieron los intervalos de confianza de dichos coeficientes de correlación mediante boostrap.
+
+En el siguiente gráfico se muestran los resultados de dicha comparación.
+
+![](../analysis/chronos/analysis_chronologies_files/figure-markdown_github/correla_chronos_sites_pearson-1.png)
+
+Parece claro que en SJ debemos hacer solo una chrono.
+La duda aquí es la siguiente, en Material y methods decimos que hemos muestreado dos sitios dentro de cada localidad y que posteriormente nos quedamos con dos en el sur y uno en el norte basados en este análisis anterior, o directamente decimos que hemos muestreado solo tres sitios.
+Esto también repercutirá en algunas tablas (la de característica de stand, age, etc.)
+
+### Cronologías por sitio
+
+![](../analysis/chronos/analysis_chronologies_files/figure-markdown_github/crono_compara_sitesSJCALH-1.png)
+
+![](../analysis/chronos/analysis_chronologies_files/figure-markdown_github/crono_compara_sitesSJCALH-1.png)
+
+Algunas cosas que observamos:
+
+*

### 2017-12-21 analysis/chronos/analysis_splines.Rmd
splines

--- 2017-12-20 analysis/chronos/analysis_chronologies.Rmd
@@
 ---
-title: "chronologies"
+title: "chronos_splines"
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 Feb"
+date: "2017 Dec"
 output:  
@@
 library("pander")
-library('boot')
 ```
@@
 * Leer datos `rwl` de SJ y CA
-* Leer datos de diametros de los focal tree
 
 ```{r, echo=FALSE}
-# machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
@@
 ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
-
-# Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-```
+```
+
+# Computar los rwi
+
+```{r}
+  # rwdf: a ringwidth dataframe. row: years; col: id_cores
+# nsmooth: es el tamaño de la wavelength (0.67*long serie; ej.: 30 years)
+
+computeSpline <- function (rwdf, nsmooth){
+
+  rwl_rwi <- c()
+
+  # ID cores 
+  cores <- colnames(rwdf)
+
+  for (i in cores) {
+    # 1 # Select core
+    id_core <- i
+
+    # 1 # Select ring width serie 
+    rws <- rwdf[[id_core]]
+
+    # 2 # Prepare data to detrend
+    names(rws) <- rownames(rwdf)
+
+    serie_suavizada <- detrend.series(y = rws, method = "Spline",
+                                      y.name = id_core, nyrs = nsmooth) 
+    serie_suavizada <- as.data.frame(serie_suavizada) 
+    colnames(serie_suavizada) <- id_core
+
+    # 4 # Join another detrened series
+    rwl_rwi <- cbind(rwl_rwi, serie_suavizada)
+  }
+  return(rwl_rwi)
+}
+
+rwi_sj <- computeSpline(sj, nsmooth = 30)
+
+rwdf <- sj 
+cores <- colnames(rwdf)
+id_core <- cores[1]
 

### 2017-12-21 man/ms_1_material_methods.Rmd
update report dendro
splines

@@
 
+Incluir datos de muestreo de competencia
+
 ### Dendroecological analyses 
@@
 We calculated a mean BAI series for each individual tree.
-To study the rsponse of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies (growth index, GIres) calculated from biweigth means of ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976]
+To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies (growth index, GIres) calculated from biweight means of ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976].
+
+```{r}
+cat("
+To obtain ring-width series of each individual and species (species chronology) the cross-dated ring-width series were standardized and detrended using the dplR library (Bunn 2010) in the R statistical language (R Development Core Team 2015).
+First, we fitted a spline through the ring-width series of each measured radius.
+We used a spline where the frequency response is 0.5 at a wavelength of 0.67 multiplied by the length of the series in years (Bunn 2010).
+Then, the ring-width data were divided by the fitted lines to obtain residual ring-width indices (RWI) that were subjected to autoregressive modelling (pre-whitening) and then averaged for each year using a biweight robust estimation of the mean (Fritts 1976).
+Calculating RWIs using subtraction instead of division rendered similar results as those here presented.
+Thus, we obtained mean residual chronologies for each species of prewhitened growth indices.
+    ")
+
+```
 

### 2017-12-21 man/notes_dendro_analysis.Rmd
splines

@@
 
-#### Salidas gráficas (ver `./out/fig/chronos/`)
+#### Salidas gráficas (ver `./out/chronos/`)
 Hacemos una comparación de cronos (BAi medios ± se), enfocado en cada sitio entre High and Low:
@@
 
-#### Salidas gráficas (ver `./out/fig/chronos/`)
+#### Salidas gráficas (ver `./out/chronos/`)
 
@@
 Posteriormente he analizado los años en los que ocurren liberaciones detectadas con los diferentes métodos y en diferentes sitios.
+
+## RING-WIDTH INDICES 
+Para estudiar la respuesta del crecimiento a la variabilidad climática inter-anual (short-term response) se utilizan chronologias residuales ...
+
+Se trata de una estandarización que se calcula dividiendo los datos de los crecimientos de los anillos (raw growth measurements) entre el crecimiento esperado para cada año [@Fritts1976].
+De esta forma se elimina la tendencia en crecimiento y además se escala la varianza para que sea aproximadamente igual para toda la serie.
+
+$$I_{t}=\frac{R_{t}}{G_{t}}$$
+
+siendo $I_{t}$ el indice, $R_{t}$ el valor del crecimiento (ring-width) y $G_{t}$ el valor del crecimiento estimado (esperado) para el año $t$.
+La estandarización se calcula por división y no por substracción, porque las ring-width series son heterocedasticas.
+
+Nota.
+Al usar tree-ring indices, que son estacionarios, tienen una media definida y una varianza homogénea, entonces podemos promediar los valores de los indices de varios árboles de un mismo sitio y formar una función media.
+
+### Estimación del crecimiento esperado
+
+El crecimiento esperado se computa mediante métodos *detrending* que lo que hacen es estimar y eliminar la variabilidad de baja-frecuencia debida a efectos biológicos o de competencia.
+Existen varios métodos de dentrending: negativo exponencial, splines.
+
+Los ***splines*** es un tipo de suavizado y se incluyen dentro de los filtros low-pass.
+Es decir, cada valor de RW se promedia utilizando un moving averages (media móvil) pero ponderada con unos pesos.
+@Cook1981 fueron los primeros en utilizar esta aproximación, sobre todo para eliminar la varianza debida a competencia de los árboles por la luz y nutrientes y a las perturbaciones del sitio.
+En definitiva una forma de eliminar la varianza no climática [@Fritts1976].
+
+Los splines son como una concatenación de segmentos polinómicos (cúbicos) que se unen en unos puntos (knots)  (*sensu* @Cook1981: "the smoothing spline is a series of piecewise cubic polynomials with a knot at each data point abscissa").
+
+El grado de suavizado del filtro low-pass (spline) depende de la función de frecuency-response.
+50 % frecuency-response cutoff es la frecuencia a la cual el 50 % de la amplitdud de una señal es retenida (o removida), y se utiliza para definr el grado de suavizado realizado por un filtro digital.
+
+El tamaño de ventana utilizado es otro factor importante. @Cook1981 habla de 2/3 la longitud de la serie en años (0.67 wavelength) (ver también detrend.series en pkg dplR)
+
+### Procedimiento
+* Utilizar un spline para cada serie donde la frecuencia respuesta es 0.5 (50 % cutoff) a una wavelength de 0.67 la longitud de la serie de datos.
+He preguntado a Guillermo y me comentó que 30 años estaba bien.
+Así que utilizo 30 años.
+
+* Seguidamente dividir los datos brutos (rw) entre los datos ajustados (obtenidos con splines) y obtendremos una ***residual ring-width indices**  
+VAS POR AQUI EN LA EXPLICACIÓN
 

### 2017-12-21 reports/report_dendro.Rmd
splines

@@
 
-# BAI
+# Long-term growth response
+
+## BAI 
+Computamos el BAI
 
@@
 ### Comparación cronologías 
-Las cronologías de la cara norte (SJ) (SJ-High y SJ-Low)s son muy similares.
+Las cronologías de la cara norte (SJ) (SJ-High y SJ-Low) son muy similares.
 Además en cada sitio de SJ, muestreamos solamente 15 árboles, frente a los 20 de cad sitio de Cáñar (sur).
@@
 
-![](../analysis/chronos/analysis_chronologies_files/figure-markdown_github/crono_compara_sitesSJCALH-1.png)
-
 Algunas cosas que observamos:
 
-*
+...
+
+# Short-term response
+
+* Usamos pre-withened residual chronologies (growth index, GI res) calculated from biweight means of ratios between raw growth measurements and cubic splines with a frequency cutoff at 30 years [@Fritts1976]

### 2017-12-24 script/R/computeSpline.R
add function computesplines based on GGea script

@@ -1,0 +1,44 @@
+## Compute RWI with splines 
+# rwdf: a ring-width dataframe. row: years; col: id_cores
+# nsmooth: size of the wavelength (0.67*long serie; ej.: 30 years) 
+
+# Return: a dataframe with RWI value
+
+computeSpline <- function (rwdf, nsmooth) {
+  
+  require(dplyr)
+  require(dplR)
+  
+  # rwl_rwi <- data.frame()
+  rwl_rwi <- data.frame(year = as.numeric(row.names(rwdf)))
+  
+  # ID cores 
+  cores <- colnames(rwdf)
+  
+  for (i in cores) {
+    # 1 # Select core (serie)
+    id_core <- i
+    
+    # 2 # Select ring width serie 
+    rws <- rwdf[[id_core]]
+    
+    # 3 # Detrended serie
+    names(rws) <- rownames(rwdf)
+    
+    serie_suavizada <- detrend.series(y = rws, method = "Spline",
+                                      y.name = id_core, nyrs = nsmooth, 
+                                      make.plot = FALSE) 
+    serie_suavizada <- as.data.frame(serie_suavizada) 
+    colnames(serie_suavizada) <- id_core 
+    serie_suavizada$year <- as.numeric(row.names(serie_suavizada))
+    
+    # 4 # Join another detrended series
+    rwl_rwi <- rwl_rwi %>% inner_join(serie_suavizada, by = "year") %>% as.data.frame()
+  }
+  
+  row.names(rwl_rwi) <- rwl_rwi$year 
+  rwl_rwi <- rwl_rwi %>% dplyr::select(-year) %>% as.data.frame()
+  
+  return(rwl_rwi)
+}     
+

### 2017-12-26 analysis/chronos/analysis_splines.Rmd
add rwi computations

@@
 
-# Computar los rwi
-
-```{r}
-  # rwdf: a ringwidth dataframe. row: years; col: id_cores
-# nsmooth: es el tamaño de la wavelength (0.67*long serie; ej.: 30 years)
-
-computeSpline <- function (rwdf, nsmooth){
-
-  rwl_rwi <- c()
-
-  # ID cores 
-  cores <- colnames(rwdf)
-
-  for (i in cores) {
-    # 1 # Select core
-    id_core <- i
-
-    # 1 # Select ring width serie 
-    rws <- rwdf[[id_core]]
-
-    # 2 # Prepare data to detrend
-    names(rws) <- rownames(rwdf)
-
-    serie_suavizada <- detrend.series(y = rws, method = "Spline",
-                                      y.name = id_core, nyrs = nsmooth) 
-    serie_suavizada <- as.data.frame(serie_suavizada) 
-    colnames(serie_suavizada) <- id_core
-
-    # 4 # Join another detrened series
-    rwl_rwi <- cbind(rwl_rwi, serie_suavizada)
-  }
-  return(rwl_rwi)
-}
-
-rwi_sj <- computeSpline(sj, nsmooth = 30)
-
-rwdf <- sj 
-cores <- colnames(rwdf)
-id_core <- cores[1]
-
-```{r}
-source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/rw_byTree.R'))
-source(paste0(di, 'script/R/chrono_bai.R'))
-source(paste0(di, 'script/R/suaviza_cronos.R'))
-source(paste0(di, 'script/R/correlaBootIC.R'))
-```
-
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
-    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
+Computar RWI para tres sitios CA_High, CA_Low, SJ.
+
+# Preparar datos para computar RWI 
+```{r}
+source(paste0(di, 'script/R/computeSpline.R'))
+```
 
@@
 ```
+
+* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ
 
@@
               "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
-
-sjL <- sj[, c("SJ0101","SJ0102","SJ0201","SJ0202","SJ0301","SJ0302","SJ0401","SJ0402","SJ0501","SJ0502",
-              "SJ0601","SJ0602","SJ0603","SJ0701","SJ0702","SJ0801","SJ0802","SJ0901","SJ0902","SJ0903",
-              "SJ1001","SJ1002","SJ1003")]
-sjH <- sj[, c("SJ1101","SJ1102","SJ1103","SJ1201","SJ1202","SJ1203","SJ1301","SJ1302","SJ1303","SJ1401",
-              "SJ1402","SJ1501","SJ1502","SJ1601","SJ1602","SJ1701","SJ1702","SJ1703","SJ1801","SJ1802",
-              "SJ1901","SJ1902","SJ2001","SJ2003","SJ2002")]
-```
-
-* Lectura y preparación de datos de diámetro
-
-```{r} 
-# Prepare Diameter data
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Change name focal according to loc
-compete <- compete %>% 
-  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm)
-
-# Set levels of eleveation 
-sj_lowcode  <- paste0('SJ', str_pad(1:10, 2, pad='0'))
-sj_highcode <- paste0('SJ', 11:20)
-ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
-            paste0('CA', 26:30))
-ca_highcode <- paste0('CA', 11:25)
-
-ft <- ft %>% 
-  mutate(elev = as.factor(
-    ifelse(id_focalLoc %in% sj_lowcode, 'Low',
-                       ifelse(id_focalLoc %in% sj_highcode, 'High',
-                              ifelse(id_focalLoc %in% ca_lowcode, 'Low', 'High'))))) %>%
-  mutate(site = as.factor(paste0(loc, '_', elev)))
-```
+```
+
+## Compute RWI
+
+```{r}
+rwi_sj <- computeSpline(rwdf = sj, nsmooth = 30)
+rwi_caL <- computeSpline(rwdf = caL, nsmooth = 30)
+rwi_caH <- computeSpline(rwdf = caH, nsmooth = 30)
+```
+
+```{r}
+
+objects_rwi <- c('rwi_caL','rwi_caH','rwi_sj')
+
+out <- c()
+
+for (i in objects_rwi){
+
+  aux <- get(i)
+
+  # Stats of RWI 
+  aux_stats <- rwi.stats(aux)
+
+  # Add site 
+  aux_stats$site <- as.factor(i)
+
+  out <- rbind(out, aux_stats)
+  }
+
+d <- chron(rwi_sj, biweight = TRUE, prewhiten = TRUE)
+dd <- rwi.stats(rwi_sj)
+
+crn.plot(d)
 
@@
 ```{r}
+
+d <- rwi.stats()
+
 objects_rwl <- c('caL','caH','sjL','sjH')

### 2017-12-26 analysis/chronos/analysis_splines.Rmd
update analysis of splines

@@
               "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
+
+# remove the rows with NA across all columns 
+caL <- caL[rowSums(is.na(caL))!=ncol(caL), ]
+
 ```
@@
 rwi_caH <- computeSpline(rwdf = caH, nsmooth = 30)
+
+# Export data 
+write.csv(rwi_sj, file=paste(di, "data/rwi/rwi_sj.csv", sep=""), row.names = TRUE)
+write.csv(rwi_caL, file=paste(di, "data/rwi/rwi_caL.csv", sep=""), row.names = TRUE)
+write.csv(rwi_caH, file=paste(di, "data/rwi/rwi_caH.csv", sep=""), row.names = TRUE)
 ```
+
+### Summary dendrochronology statistics
+
+* Por cada site (ca_high, ca_low, sj) calculamos los estadísticos de la serie rwi. En concreto:
+
+:red_circle: DUDAS cual utilizar como reportes
+
+## By site
 
@@
   aux <- get(i)
+  name_aux <- sub(".*\\_", '', i)
+
+  # Get ids 
+  aux_ids <- read.ids(aux, stc = c(2,2,2))
 
   # Stats of RWI 
-  aux_stats <- rwi.stats(aux)
+  aux_stats <- rwi.stats(aux, ids = aux_ids)
 
   # Add site 
-  aux_stats$site <- as.factor(i)
+  aux_stats$site <- as.factor(name_aux)
 
@@
 
-d <- chron(rwi_sj, biweight = TRUE, prewhiten = TRUE)
-dd <- rwi.stats(rwi_sj)
+rwi_statistics <- out
 
-crn.plot(d)
+rwi_statistics %>% 
+  write.csv(file=paste(di, "data/dendro_summary/site3_dendro_rwi.csv", sep=""), row.names = FALSE)
 
-# Summary dendrochronology statistics
+pander(rwi_statistics, caption='Dendrochronological summary by sites') 
+```
 
-* Por cada loc (ca/sj) o site (ca_high, ca_low, sj_high, sj_low) calculamos los siguientes datos: 
-    * age: mean + sd (average of all series)
-    * year: min, max of all series
-    * year range
-    * autocorrelation order 1 (ar1): mean + sd 
-    * seires intercorrelation (icor): mean + sd
+## Compute residual chronologies
 
-## By site 
+A partir de cada serie (RWI) vamos a computar chronologias por sitio.
+
+* Cada serie RWI se somete a un modelo autorregresivo (pre-withening) y posteriormente se calcula el promedio para cada año mediante una estimación robusta de la media (biweight).
+
+* La calidad estadistica de cada serie se chequeó via la EPS (expressed population signal) [@Wigley1984]. Se calculó para cada serie la EPS utilizando una ventana temporal de 30 años. 
+* Posteriormente, a threshold value of EPS >0.85 was used to determine the cutoff year of the time span that could be considered reliable
+
 ```{r}
 
-d <- rwi.stats()
-
-objects_rwl <- c('caL','caH','sjL','sjH')
+objects_rwi <- c('rwi_caL','rwi_caH','rwi_sj')
 
 out <- c() 
+out_eps <- c()
 
-for (i in objects_rwl){
+for (i in objects_rwi){
 
   aux <- get(i)
-  # Stats of singles RWL
-  aux_stats <- rwl.stats(aux)
-  # Add site 
-  aux_stats$site <- as.factor(i)
 
-  # Compute interseries correlation 
-  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
-  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
+  name_aux <- sub(".*\\_", '', i)
 
-  out <- rbind(out, aux_stats)
+  # Compute chrono 
+  crono_aux <- chron(aux, prefix = "", biweight = TRUE, prewhiten = TRUE)
+
+  # Add year and site 
+  crono_aux$year <- as.numeric(row.names(crono_aux))
+  crono_aux$site <- as.factor(name_aux)
+
+  # name export 
+  nameexport <- paste0("cro_", name_aux)
+
+  assign(nameexport, crono_aux)
+
+  # eps 
+  # Get ids 
+  aux_ids <- read.ids(aux, stc = c(2,2,2))
+
+  # eps cutoff
+  aux_stats <- rwi.stats.running(aux, ids = aux_ids, window.length = 30)
+
+  aux_eps <- data.frame(year = c(min(crono_aux$year), 
+                                aux_stats$mid.year, 
+                                max(crono_aux$year)),
+                       eps = c(NA, aux_stats$eps, NA), 
+                       site = as.factor(name_aux))
+
+  out <- rbind(out, crono_aux)
+  out_eps <- rbind(out_eps, aux_eps)
   }
 
-rwl_by_site <- out %>% group_by(site) %>% 
-  summarize(age_mean = round(mean(year),2), 
-            age_sd = round(sd(year), 2),
-            year_min = min(first),
-            year_max = max(last), 
-            year_range = year_max - year_min + 1, 
-            ar1_mean = round(mean(ar1), 3),
-            ar1_sd = round(sd(ar1), 3),
-            icor_mean = round(mean(interser_cor_val), 3),
-            icor_sd = round(sd(interser_cor_val), 3)) %>% 
-  as.data.frame()
+# cronos 
+cro_rwi <- out
 
-rwl_by_site %>% 
-  write.csv(file=paste(di, "data/dendro_summary/site_dendro.csv", sep=""), row.names = FALSE)
-
-pander(rwl_by_site, caption='Dendrochronological summary by sites') 
+# Export data 
+write.csv(cro_rwi, file=paste(di, "data/cronos_medias/cronos_sites_rwi.csv", sep=""), row.names = FALSE)
 ```
 
-## By Loc
+# Obtener el año a partir del cual podemos considerar reliable la chrono. 
+## eps
+
 ```{r}
-objects_rwl <- c('ca','sj')
+# eps umbral 
+eps_cut <- 0.85
+year_cutoffs <- 
+  out_eps %>% 
+  group_by(site) %>% 
+  # filter(eps < eps_cut) %>% 
+  slice(which.max(eps < eps_cut)) %>% as.data.frame()
 
-out <- c()
-
-for (i in objects_rwl){
-
-  aux <- get(i)
-  # Stats of singles RWL
-  aux_stats <- rwl.stats(aux)
-  # Add site 
-  aux_stats$loc <- as.factor(i)
-
-  # Compute interseries correlation 
-  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
-  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
-
-  out <- rbind(out, aux_stats)
-  }
-
-rwl_by_loc <- out %>% group_by(loc) %>% 
-  summarize(age_mean = round(mean(year),2), 
-            age_sd = round(sd(year), 2),
-            year_min = min(first),
-            year_max = max(last), 
-            year_range = year_max - year_min + 1, 
-            ar1_mean = round(mean(ar1), 3),
-            ar1_sd = round(sd(ar1), 3),
-            icor_mean = round(mean(interser_cor_val), 3),
-            icor_sd = round(sd(interser_cor_val), 3)) %>% 
-  as.data.frame()
-
-rwl_by_loc %>% 
-  write.csv(file=paste(di, "data/dendro_summary/loc_dendro.csv", sep=""), row.names = FALSE)
-
-pander(rwl_by_loc, caption='Dendrochronological summary by sites') 
+pander(year_cutoffs) 
 ```
 
-# Aggregate RW by tree
+Ojo el año es el mid year del periodo considerado.
+Por ejemplo para SJ, todos los valores de EPS son mayores del valor del umbral, y coincide que el menor valor de eps es el del primer periodo.
+Ojo a eso.
 
-* Agregar valores medios de RW por loc (obtenemos sj_tree / ca_tree) o por site (sjL_tree, sjH_tree / caL_tree, caH_tree) 
-* ver fun rw_byTree o utilizar treeMean (dplR)
+## sample depth 
+```{r}
+# ssamp.depth
+sample_treshold <- 5
+
+year_sampledepth <- 
+  cro_rwi %>% 
+  group_by(site) %>% 
+  # filter(eps < eps_cut) %>% 
+  slice(which.max(samp.depth > sample_treshold)) %>% as.data.frame()
+
+pander(year_sampledepth) 
+```
+
+### Plots of RWI
 
 ```{r}
-# Remember snc = structure of core name SJ0101 (site | tree | core)
-sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
-ca_tree <- rw_byTree(ca, snc =c(2,2,2), locname = 'CA')
+cro_rwi <- cro_rwi %>% 
+  mutate(site_sorted = case_when(
+    site == "sj" ~ "0_sj", 
+    site == "caH" ~ "1_caH",
+    site == "caL" ~ "2_caL"))
 
-sjL_tree <- rw_byTree(sjL, snc =c(2,2,2), locname = 'SJ')
-sjH_tree <- rw_byTree(sjH, snc =c(2,2,2), locname = 'SJ')
-caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
-caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')
+year_cutoffs <- year_cutoffs %>% 
+  mutate(site_sorted = case_when(
+    site == "sj" ~ "0_sj", 
+    site == "caH" ~ "1_caH",
+    site == "caL" ~ "2_caL"))
 
-# Si decido usar treeMean debería ser (solo un ejemplo)
-# ids_sj <- read.ids(sj, stc = c(2, 2, 2))
-# sj_meantree <- treeMean(sj, ids = ids_sj,  na.rm=TRUE)
+year_sampledepth <- year_sampledepth %>% 
+  mutate(site_sorted = case_when(
+    site == "sj" ~ "0_sj", 
+    site == "caH" ~ "1_caH",
+    site == "caL" ~ "2_caL")
 ```
 
-* Crear diferentes dataset de diametro por sitio 
-```{r}
-# # function to subset averaged RW 
-# subsetRW <- function(rwdf, codes){ 
-#   require(dplyr)
-#   aux_codes <- paste0('rw_', codes)
-#   mysubset <- rwdf %>% dplyr::select(one_of(aux_codes), year)
-#   mysubset
-#   }
+Plots de las cronos, añadiendo líneas para los años de sequía 1995, 2005 y 2012 (líneas negras) y líneas para los años con valor eps > 0.85 (ver ejemplo paper Dorado-liñan https://link.springer.com/article/10.1007/s00484-017-1410-5).
+También añado una con el sample depth > 5 (línea azul).
+Escoger uno de los dos criterios.
+Preguntar a G.
+Gea
 
-# diam <- ft %>% group_by(site) %>% do(data = (.)) %>% 
-#              dplyr::select(data) %>% 
-#              lapply(function(x) {(x)})
+```{r, rwi_crono}
+label_site <- c("0_sj" = "Northern (SJ)",
+                "1_caH" = "Southern High (CaH)",
+                "2_caL" = "Southern Low (CaL)")
 
-diam <- ft %>%
-  mutate(diameter = dn_mm, 
-         id = id_focalLoc) %>%
-  dplyr::select(id, diameter, site) %>% 
-  split(.$site)
+# 
+plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
+  theme_bw() + 
+  theme(panel.grid = element_blank(), 
+        strip.background = element_rect(fill = "white"),
+        strip.text = element_text(face="bold", size=12), 
+        axis.text = element_text(face="bold", size=10), 
+        axis.title = element_text(face="bold", size=12)) + 
+  geom_hline(yintercept = 1, colour = "grey") + 
+  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
+  facet_wrap(~site_sorted, scales = "fixed", ncol = 1,
+             labeller = as_labeller(label_site)) + 
+  ylab("Ring width index") + xlab("Year") + 
+  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
+  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
+  geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
+  scale_x_continuous(limits = c(1819, 2016), 
+                     breaks= seq(1820, 2020, by=20))
 
-d_caH <- diam$CA_High[,c('id','diameter')]
-d_caL <- diam$CA_Low[,c('id','diameter')]
-d_sjH <- diam$SJ_High[,c('id','diameter')]
-d_sjL <- diam$SJ_Low[,c('id','diameter')]
-
-diamloc <- ft %>% 
-  mutate(diameter = dn_mm, 
-         id = id_focalLoc) %>%
-  dplyr::select(id, diameter, loc) %>% split(.$loc)
-
-d_ca <- diamloc$CA[,c('id','diameter')]
-d_sj <- diamloc$SJ[,c('id','diameter')]
+plot_rwi_crono   
+ggsave(plot=plot_rwi_crono, width=8, height = 6,
+       filename=paste0(di, 'out/chronos_rwi/rwi_crono_3sites.pdf')) 
 ```
 
-# Cómputo del BAI por loc y/o site
+```{r, rwi_crono_1950}
+plot_rwi_crono50 <- cro_rwi %>% 
+  filter(year > 1949) %>% 
+  ggplot(aes(x=year, y=res)) +
+  theme_bw() + 
+  theme(panel.grid = element_blank( ), 
+        strip.background = element_rect(fill = "white"),
+        strip.text = element_text(face="bold", size=12), 
+        axis.text = element_text(face="bold", size=10), 
+        axis.title = element_text(face="bold", size=12)) + 
+  geom_hline(yintercept = 1, colour = "grey") + 
+  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
+  facet_wrap(~site_sorted, scales = "fixed", ncol = 1,
+             labeller = as_labeller(label_site)) + 
+  ylab("Ring width index") + xlab("Year") + 
+  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") +
+  scale_x_continuous(limits = c(1950, 2016), 
+                     breaks = seq(1950, 2016, by=10))
 
-* He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]. Es similar a `bai.out`
-
-```{r}
-bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
-bai_ca <- bai_piovesan(rwdf = ca_tree, diam_df = d_ca)
-
-bai_sjH <- bai_piovesan(rwdf = sjH_tree, diam_df = d_sjH)
-bai_sjL <- bai_piovesan(rwdf = sjL_tree, diam_df = d_sjL)
-bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
-bai_caL <- bai_piovesan(rwdf = caL_tree, diam_df = d_caL)
-
-# bai_test <- round(bai.out(rwl = sj_tree, diam = d_sj), 4)
-# test bai.out and bai_piovesan  
-# unique(round(bai_sj$SJ03,4) == round(bai_test$SJ03,4))
-
-# Set class to bai object 
-# Esto es para que funcionen algunas otras funciones de dplR 
-bais <- c('bai_sj', 'bai_ca', 'bai_sjH', 'bai_sjL', 'bai_caH', 'bai_caL')
-
-for (i in bais){
-  aux <- get(i)
-
-  class(aux) <- c('rwl', 'data.frame')
-
-  assign(i, aux)
-}
+plot_rwi_crono50   
+ggsave(plot=plot_rwi_crono50, width=8, height = 6,
+       filename=paste0(di, 'out/chronos_rwi/rwi_crono_3sites50.pdf'))
 
@@
 
-# Cronologías by site / loc
+VAS POR AQUI
 
-### Mean vs. Tukey’s Biweight Robust Mean
+# CLIMATE RELATIONSHIPS 
+```{r}
 
-* Computo la cronologia usando la función `chron`. 
-* Para SJ hag una comparación de la cronologia construida con biweight (*treatment = biweigth*) o construida con la media aritmética (*treatment = No biweigth*) (calculada con la función `chron`)
-* Idem para CA, pero en CA si separo entre CA high y CA low
+library(ncdf4)
 
-```{r chro_sj_compare_method}
-# Robust mean biweith or not
-cro_sjb <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
-cro_sjnotb <- chron(bai_sj, prefix = "SJ", biweight = FALSE)
+nc <- nc_open( "/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc")
+print(nc)
 
-cro_b <- cro_sjb %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = SJstd, samp.depth, year, treatment)
+lon <- ncvar_get(nc,"longitude")
 
-cro_notb <- cro_sjnotb %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = SJstd, samp.depth, year, treatment)
+http://geog.uoregon.edu/bartlein/courses/geog490/week04-netCDF.html#open-the-netcdf-file
 
-aux_cro <- rbind(cro_b, cro_notb)
+eobs
 
-aux_cro %>% ggplot(aes(x = year, y = bai_mean / 100, colour=treatment)) + 
-  geom_point() + geom_line() + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
-  theme_bw() + theme(panel.grid.major = element_blank()) +
-  ggtitle('SJ chronology built using Biweight or Mean')
+untar("/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc", exdir = "/Users/ajpelu/Downloads/climaobs")
+
+# Precipitation dataset URL
+ds <- "/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc"
+
+di <- dataInventory("/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc")
+  "http://opendap.knmi.nl/knmi/thredds/dodsC/e-obs_0.25regular/rr_0.25deg_reg_v16.0.nc"
+
+http://meteo.unican.es/work/loadeR/wiki/html/OaccessingLoadingStationData.html
+download.file("http://meteo.unican.es/work/loadeR/data/VALUE_ECA_86_v2.tar.gz", destfile ="/Users/ajpelu/Downloads/climaobs/VALUE_ECA_86_v2.tar.gz")
+
+value <- "/Users/ajpelu/Downloads/climaobs/VALUE_ECA_86_v2"
+
+di <-dataInventory(value)
+
+?loa
+
+precip  <- loadGridData(dataset = ds,
+			   var = "rr",
+			   lonLim = -3.428,
+			   latLim = 36.957,
+			   years = 1950:2016,
+			   season = 1:12)
+
 ```
 
-```{r chro_ca_compare_method}
-# Robust mean biweith or not
-cro_cab_h <- chron(bai_caH, prefix = "CA", biweight = TRUE)
-cro_canotb_h <- chron(bai_caH, prefix = "CA", biweight = FALSE)
+Resilience RWI ???
 
-cro_cab_h_aux <- cro_cab_h %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_canotb_h_aux <- cro_canotb_h  %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-aux_croca_h <- rbind(cro_cab_h_aux, cro_canotb_h_aux)
-aux_croca_h$site <- 'CA High'
-
-cro_cab_l <- chron(bai_caL, prefix = "CA", biweight = TRUE)
-cro_canotb_l <- chron(bai_caL, prefix = "CA", biweight = FALSE)
-
-cro_cab_l_aux <- cro_cab_l %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_canotb_l_aux <- cro_canotb_l  %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-aux_croca_l <- rbind(cro_cab_l_aux, cro_canotb_l_aux)
-aux_croca_l$site <- 'CA Low'
-
-aux_croca <- rbind(aux_croca_h, aux_croca_l)
-
-aux_croca %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
-  geom_point() + geom_line() + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
-  theme_bw() + theme(panel.grid.major = element_blank()) + 
-  facet_wrap(~site, nrow = 2) +
-    ggtitle('CA chronology built using Biweight or Mean')
-```
-
-## Compute Chrono 
-### San Juan chronologies by site
-
-```{r chrono_sj_two_sites}
-# San Juan 
-cro_sjH <- chrono_bai(bai_sjH)
-cro_sjH$site <- 'SJ_High'
-
-cro_sjL <- chrono_bai(bai_sjL)
-cro_sjL$site <- 'SJ_Low'
-
-cro_sj<- rbind(cro_sjH, cro_sjL)
-cro_sj$loc <- 'Norte (San Juan)'
-
-plot_cro_sj <- cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                    ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + geom_point(size=1) + 
-  # scale_colour_manual(values = c("red", "blue")) + 
-  theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1920, 2020, by=10))
-
-plot_cro_sj
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sj_HL.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sj, width=12, height = 4) 
-```
-
-### Cañar chronologies by site
-
-```{r chrono_ca_two_sites}
-# Canar 
-cro_caH <- chrono_bai(bai_caH)
-cro_caH$site <- 'CA_High'
-
-cro_caL <- chrono_bai(bai_caL)
-cro_caL$site <- 'CA_Low'
-
-cro_ca <- rbind(cro_caH,cro_caL)
-cro_ca$loc <- 'Sur (Canar)'
-
-plot_cro_ca <- cro_ca %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=1) + 
-  # scale_colour_manual(values = c("red", "blue")) + 
-  theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep='')))+
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-
-plot_cro_ca
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_ca_HL.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_ca, width=12, height = 4) 
-```
-
-```{r chronos_4_sites}
-# Combine and plot 
-cro_sites <- rbind(cro_ca, cro_sj)
-
-plot_cro_sites <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + facet_wrap(~loc, nrow = 2) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-
-plot_cro_sites
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites, width=12, height = 8)
-```
-
-```{r chronos_4_sites_collapsed}
-plot_cro_sites_collapsed <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  # geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-plot_cro_sites_collapsed
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed, width=12, height = 4)
-plot_cro_sites_collapsed
-```
-
-```{r chronos_4_sites_collapsed50}
-plot_cro_sites_collapsed50  <- cro_sites %>% 
-  filter(year > 1949) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  # geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1950, 2020, by=5)) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-
-plot_cro_sites_collapsed50
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed50.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed50, width=12, height = 4) 
-plot_cro_sites_collapsed50
-```
-
-```{r chronos_4_sites_collapsed90}
-plot_cro_sites_collapsed90  <- cro_sites %>% 
-  filter(year > 1989) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  # geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1990, 2016, by=2))
-
-plot_cro_sites_collapsed90
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed90.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed90, width=12, height = 4) 
-plot_cro_sites_collapsed90
-```
-
-# Correlation between chronos
-
-To compute similarity between chronos we use @Dorado2017 approach.
-
-* Para cada chrono se realiza un suavizado, utilizando diferente tamaño de ventana (1 a 40 years). Así evaluamos la sincronía entre dos series, tanto a nivel interanual como a largo plazo. 
-* Luego calculamos la correlación entre series para cada tamaño de ventana.
-Usamos correlaciones de Pearson o Sperman entre las series (es decir entre las zonas comunes de las cronologías).
-Ojo con la autocorrelación.
-Para ello puedo utilizar `interseries.cor()` del pkg `dplR`; o utilizar `cor.test()` (`stats`).
-El primero está basado en el segundo, así que utilizo el segundo.
-* Obtener IC utilizando bootstrap. Varias aproximaciones:
-
-    * Utilizar el pkg `boot` y conseguir IC. 
-    * Aproximación de @Mudelsee2014.
-En su paper habla de correlacion entre dos series y utiliza un programa en Fortran-90.
-He contactado con el autor y no tiene pensado hacerlo en R.
-He intentado utilizar Fortran-90, pero no tengo ni idea.
-Asi que desisto de utilizar esta computación de los IC.
-
-```{r}
-# centred moved average smooth
-suaviza_caH <- suaviza_cronos(cro_caH, win_vector=seq(1,40, by=1), crononame = 'cro_caH')
-suaviza_caL <- suaviza_cronos(cro_caL, win_vector=seq(1,40, by=1), crononame = 'cro_caL')
-suaviza_sjH <- suaviza_cronos(cro_sjH, win_vector=seq(1,40, by=1), crononame = 'cro_sjH')
-suaviza_sjL <- suaviza_cronos(cro_sjL, win_vector=seq(1,40, by=1), crononame = 'cro_sjL')
-```
-
-## between sites similarity (sjH *vs* sjL; caH *vs* caL)
-
-This note is for the plot:
-
-* Correlation among chronologies for the common period of each site in different time-domains after pre-filtering the time-series with increasing size of the moving-average window.
-
-```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-# set seed
-set.seed(333)
-
-# Metodo correlation
-tc <- c('spearman', 'pearson')
-
-# SJ 
-suaviza_sj <- rbind(suaviza_sjH, suaviza_sjL)
-correla_sj <- correlaBootIC(suaviza_sj, tc, years_suavizado = 40, 
-                            name_comparison = 'sjH-sjL')
-
-# CA 
-suaviza_ca <- rbind(suaviza_caH, suaviza_caL)
-correla_ca <- correlaBootIC(suaviza_ca, tc, years_suavizado = 40, 
-                            name_comparison = 'caH-caL')
-
-correla <- rbind(correla_ca, correla_sj)
-```
-
-```{r correla_chronos_sites, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-correla %>% 
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  facet_wrap(~metodo_correla, ncol=2) +
-  theme(strip.background = element_blank())
-```
-
-```{r correla_chronos_sites_pearson}
-correla_sites_pearson <- correla %>% filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  theme(legend.position = c(.8,.7))
-correla_sites_pearson
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/correla_boot_sitesHL.pdf') 
-ggsave(filename=fileplot, plot=correla_sites_pearson, width=5, height = 5) 
-```
-
-## between final sites similarity (sj *vs* caH; sj *vs* caL; caH *vs* caL)
-
-```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-# Metodo correlation
-tc <- c('spearman', 'pearson')
-
-set.seed(333)
-
-# Compute the suaviza_sj desde el principio, sin diferenciar h and l. Para ello la crono la computo de nuevo 
-cro_sjloc <- chrono_bai(bai_sj)
-cro_sjloc$site <- 'SJ'
-suaviza_sjloc <- suaviza_cronos(cro_sjloc, win_vector=seq(1,40, by=1), crononame = 'cro_sj')
-
-# sjcaH 
-suaviza_sjcaH <- rbind(suaviza_sjloc, suaviza_caH)
-correla_sjcaH <- correlaBootIC(suaviza_sjcaH, tc, years_suavizado = 40, 
-                            name_comparison = 'sj-caH')
-
-# sjcaL 
-suaviza_sjcaL <- rbind(suaviza_sjloc, suaviza_caL)
-correla_sjcaL <- correlaBootIC(suaviza_sjcaL, tc, years_suavizado = 40, 
-                            name_comparison = 'sj-caL')
-
-# Join all correlations 
-correla <- rbind(correla_ca, correla_sjcaL, correla_sjcaH)
-
-correla %>% 
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  facet_wrap(~metodo_correla, ncol=2) +
-  theme(strip.background = element_blank())
-```
-
-```{r correla_sitesSJCALH}
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/correla_boot_sitesSJCALH.pdf') 
-ggsave(filename=fileplot, plot=correla_sitesSJCALH, width=5, height = 5) 
-```
-
-# Chronologies by site (SJ, CAL, CAH)
-
-```{r}
-# Compute la crono de SJ de nuevo y llamar al site SJ
-cro_sjf <- chrono_bai(bai_sj)
-cro_sjf$site <- 'SJ'
-```
-
-```{r crono_compara_sitesSJCALH}
-# Combine and plot 
-cro_sites <- rbind(cro_caL, cro_caH, cro_sjf)
-
-## Export data of bai (chro)
-write.csv(cro_sites, file=paste(di, "data/cronos_medias/cronos_sites.csv", sep=""), row.names = FALSE)
-
-plot_cro_sitesSJCALH <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-
-plot_cro_sitesSJCALH
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH, width=12, height = 5)
-plot_cro_sitesSJCALH
-```
-
-```{r crono_compara_sitesSJCALH50}
-plot_cro_sitesSJCALH50 <- cro_sites %>% 
-  filter(year > 1949) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
-plot_cro_sitesSJCALH50
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH50.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH50, width=12, height = 5)
-```
-
-```{r crono_compara_sitesSJCALH90}
-plot_cro_sitesSJCALH90 <- cro_sites %>% 
-  filter(year > 1989) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-plot_cro_sitesSJCALH90
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH90.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH90, width=12, height = 5)
-```
-
-```{r, echo=FALSE, eval=FALSE}
-
-# CODE SNIPPETS 
-  # Loess 
-  l <- loess(b_aux$bai ~ b_aux$year, span = alpha)
-  years_v <- seq(min(b_aux$year), max(b_aux$year), 1)
-
-  l_pred_aux <- predict(l, years_v, se=TRUE)
-  l_pred <- as.data.frame(cbind(loess_fit = l_pred_aux$fit, 
-                loess_se = l_pred_aux$se.fit,
-                year = years_v))
-
-  out <- out %>% inner_join(l_pred, by='year')
-
-  out
-}
-
-# SJ
-
-cro_sj <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
-cro_sj %>% plot(add.spline=TRUE, nyrs=30)
-
-cro_ca <- chron(bai_ca, prefix = "CA", biweight = TRUE)
-cro_ca %>% plot(add.spline=TRUE, nyrs=30)
-
-cro_ca2 <- cro_ca %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('CA')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_sj2 <- cro_sj %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('SJ')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
-
-cro2 <- rbind(cro_ca2, cro_sj2)
-
-cro2 %>% ggplot(aes(x=year, y=bai_mean, colour=treatment)) + 
-  geom_line() + geom_smooth(method= 'loess', span = 0.30, se=FALSE) + theme_bw() + 
-  geom_bar(aes(x=year, stat = (samp.depth*10), colour=treatment))
-
-## San Juan Data
-
-Algunos árboles tienen 3 cores. Sin embargo, no los tres cores llegan hasta la corteza, por lo tanto, no podemos utilizar el diametro para estimar bai. Vamos a realizar lo siguiente:
-
-* Crear dos datasets: 
-    * Dataset con series de datos que llegan hasta corteza (`sj_cor`)
-    * Dataset con series de datos que no llegan hasta corteza (`sj_sin`): se trata de cores que no llegan hasta la corteza.
-
-* Para el cálculo del BAI, en el dataset `sj` utilizamos el diámetro medido en campo. En el caso del dataset `sj_sin` utilizamos la suma de todos los diámetros (:red_circle: `$TODO$ ASK to Guillermo`)
-
-# Get summary ring-width series
-sj_summ <- summary(sj)
-
-# Get names cores with last year different to 2016 
-id_cores_no_bark <- sj_summ %>% 
-  filter(last != 2016) %>% select(series) %>% mutate(series = factor(series))
-
-id_cores_no_bark <- unname(unlist(id_cores_no_bark))
-
-# Subbet datasets 
-sj_cor <- sj[which(! colnames(sj) %in% (id_cores_no_bark))]
-sj_sin <- sj[which(colnames(sj) %in% (id_cores_no_bark))]
-
-Lectura y preparación de datos de diámetro
-
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  select(id_focal, loc, dn_mm, height_cm)
-
-# Select only SJ trees and 
-ft_sj <- ft %>% filter(loc=='SJ')
-
-# get Create a dataframe with core ID
-cores <- data.frame(id_cores=colnames(sj))
-
-# Extract replicate and Tree ID from core ID 
-cores <- cores %>% 
-  mutate(id_focal = as.factor(stringr::str_sub(id_cores, 3,5)),
-         id_replica = stringr::str_sub(id_cores, 6,8))
-
-# Create df with diameter and height for each core ID 
-diam_cores <- cores %>% inner_join(ft_sj, by='id_focal') %>% 
-  select(id_cores, dn_mm)
-
-# remove diamm of cores without bark 
-diam_cores_sj <- diam_cores %>% filter(! id_cores %in% id_cores_no_bark)
-
-rwl.report(sj_cor)
-
-bai_p <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
-bai_d <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
-
-diam_cores_sjsin <- data.frame(id_cores = colnames(sj_sin),
-                               dn_mm = colSums(sj_sin, na.rm = TRUE))
-rownames(diam_cores_sjsin) <- NULL
-
-bai_psin <- bai_piovesan(rwdf = sj_sin, diam_df = diam_cores_sjsin)
-bai_dsinsin <- bai.out(rwl = sj_sin)
-bai_dsin <- bai.out(rwl = sj_sin, diam = diam_cores_sjsin)
-
-set.seed(1234)
-serie <- sample(colnames(b_test_dplR),1)
-
-require(reshape2)
-
-bp <- melt(t(b_test_piovesan))
-names(bp) <- c('cores', 'years', 'bai_p')
-
-bd <- melt(t(b_test_dplR))
-names(bd) <- c('cores', 'years', 'bai_d')
-
-diff <- bp %>% 
-  inner_join(bd, by=c('cores','years')) %>% 
-  mutate(diff = bai_p - bai_d)
-
-ggplot(x, aes(y=bai, x=years, colour=test)) + 
-  geom_point() + 
-  facet_wrap(~cores)
-
-# BAI dplR
-baisj <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
-
-baisjt <- bai.out(rwl = sjt)
-
-# Function to compute BAI (Piovesan et al)
-
-# rwl_file
-myrwl <- sj_cor
-# Diameter df  
-diam_df <- diam_cores_sj
-
-# Core
-id_core <- 'SNA0101'
-
-# Ring-width serie
-myrwl_serie <- myrwl[[id_core]]
-
-# Diameter for core id_core 
-dbh <- subset(diam_df, id_cores == id_core)
-dbh <- dbh[,2]
-
-myrwl_serie2 <- na.omit(myrwl_serie)
-cum
-
-r0 <- dbh/2 - c(0,cumsum(rev(myrwl_serie2)))
-
-bai <- (pi/4)*(dbh^2 - ((dbh - 2*myrwl_serie)^2))
-
-s <- myrwl_serie[94:96]
-
-d <- sum(s)*2
-
-########## test Piovesan
-# rwl_file
-myrwl <- baisj
-
-#id_core
-
-# Extract rwl serie 
-myrwl_i <- myrwl[[1]]
-
-dbh <- 372.4226
-
-# Year
-n.vec <- seq_len(nrow(myrwl))
-
-# Series
-seq_len(ncol(myrwl))
-
-# Select a ring-width series
-dat <- myrwl[[1]] 
-# Remove NA
-dat2 <- na.omit(dat)
-
-ggplot(s1, aes(y=SNA0101, x=row.names(s1))) + 
-  geom_point() + 
-  geom_line(b1, aes(y=SNA0101, x=row.names(b1)))
-
-ggplot(s1, aes(y=SNA0101, x=row.names(b1), group=1)) + geom_line()
-
-x <- c(1,2,2,3,1,3,4,5)
-
-xx <- rollmean(x=x, k=3, align = 'center', na.pad = T)
-
-# tes plots
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-carwi <- detrend(rwl = ca_tree, method = "ModNegExp")
-ca_prwi <- data.frame(rw = rowMeans(carwi, na.rm=TRUE))
-ca_prwi <- ca_prwi %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-xx <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-plot(xx$rw~rownames(xx))
-
-sjcrn <- chron(sjrwi, prefix = "rw_")
-plot(sjcrn, add.spline=TRUE, nyrs=20)
-plot()
-
-sj_tree <- rw_byTree(sj)
-ca_tree <- rw_byTree(ca)
-
-rownames(ca_tree) <- ca_tree$year
-ca_tree <- ca_tree[,-1]
-ca_p <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-ca_p <- ca_p %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-sj_p <- data.frame(rw = rowMeans(sj_tree, na.rm=TRUE))
-
-sj_p <- sj_p %>% mutate(year = rownames(.),
-                        site = as.factor('sj'),
-                        rw_log = log(rw))
-
-xx <- rbind(ca_p, sj_p)
-
-xx %>% ggplot(aes(x=as.Date(year, format='%Y'), y=rw_log)) + 
-  geom_line(aes(group=1)) + 
-  scale_x_date(date_labels = "%Y") + 
-  facet_wrap(~site, ncol=1, scales = 'free_y') + 
-  theme_bw() + 
-  geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-xx %>% filter(site =='ca') %>% 
-  ggplot(aes(x=as.Date(year, format='%Y'), y=rw)) + geom_point(colour='blue') +
-  geom_line(aes(group=1), colour='blue') + 
-  scale_x_date(date_labels = "%Y") + 
-  theme_bw() + 
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-ca_prwi %>%
-  ggplot(aes(x = as.Date(year, format = '%Y'), y = rw)) + geom_point() +
-  geom_line(aes(group = 1)) +
-  scale_x_date(date_labels = "%Y") +
-  theme_bw() +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width Index')
-
-```
+# RESILIENCE RWI

### 2017-12-27 analysis/explore_correlations_windows_size.Rmd
update correlations window size

@@
 corEvi <- read.csv(file=paste0(di, '/out/correla_resilience/correla_window_size_evi.csv'), header = TRUE)
+corRwi <- read.csv(file=paste0(di, '/out/correla_resilience/correla_window_size_rwi.csv'), header = TRUE)
 ```
@@
 ```
+
+# RWI
+
+```{r}
+cor_rte <- generaMatrix(corRwi, 'rt')
+cor_rce <- generaMatrix(corRwi, 'rc')
+cor_rse <- generaMatrix(corRwi, 'rs')
+cor_rrse <- generaMatrix(corRwi, 'rrs')
+```
+
+```{r}
+g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='RWI Rt', show.legend = FALSE)
+g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='RWI Rc', show.legend = FALSE)
+g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='RWI Rs', show.legend = FALSE)
+g_rrse <- ggcorrplot(cor_rrse, type='upper', lab=TRUE, title='RWI RRs', show.legend = TRUE)
+
+```
+
+```{r correla_window_rwi}
+grid.arrange(g_rte, g_rce, g_rse, g_rrse)
+```

### 2017-12-27 script/R/rwiResilience.R
function to compute RWI resilience metrics

@@ -1,0 +1,79 @@
+## Compute Resilience of RWI
+
+## Resilience metrics: See Lloret et al. 2011 http://dx.doi.org/10.1111/j.1600-0706.2011.19372.x
+## rwi_tree: an object of rwi by tree
+## event_years: vector with drought years, i.e.: c(1995, 2005)
+## window: size of the window to compute previous and post event segments
+
+## A list with  to df
+## $rwi: summary statistics by tree of RWI for each period (prev, dr, post)
+## $resilience: resilience metrics for each drougth period
+
+rwiResilience <- function(rwi_tree, event_years, window) {
+  out_rwi <- c()
+  out_res <- c()
+  out <- list()
+
+  require(dplyr)
+
+  for (i in event_years) {
+    # Create df previous, during and post event
+    df <- rwi_tree %>%
+      mutate(year = as.numeric(row.names(.))) %>%
+      as.data.frame()
+
+    df_pre <- df %>% filter(year < i & year >= (i - window)) %>% mutate(disturb = "pre")
+    df_event <- df %>% filter(year == i) %>% mutate(disturb = "dr")
+    df_post <- df %>% filter(year > i & year <= (i + window)) %>% mutate(disturb = "post")
+
+    # Melt data
+    dfmelt <- bind_rows(df_pre, df_event, df_post) %>%
+      gather(tree, value, -disturb, -year)
+
+    # Compute mean of RWI (previous, during and posterior)
+    rwi_df <- dfmelt %>%
+      group_by(disturb, tree) %>%
+      summarise(
+        mean_period = mean(value),
+        sd_period = sd(value),
+        se_period = sd_period / sqrt(length(value))
+      ) %>%
+      mutate(disturb_year = i) %>%
+      as.data.frame()
+
+    out_rwi <- rbind(out_rwi, rwi_df)
+
+
+    # Compute resilience by trees
+    aux_resilience <- c()
+    trees <- unique(rwi_df$tree)
+
+    for (t in trees) {
+      # Filter by tree
+      df_tree <- rwi_df %>% filter(tree == t)
+
+      # Compute resilience metrics by tree
+      aux_df <- df_tree %>%
+        dplyr::select(-c(tree, sd_period, se_period)) %>%
+        spread(key = disturb, value = mean_period) %>%
+        mutate(
+          rt = dr / pre,
+          rc = post / dr,
+          rs = post / pre,
[…]

### 2017-12-27 analysis/chronos/analysis_splines.Rmd
add resilience analysis to splines RWI

@@
 source(paste0(di, 'script/R/computeSpline.R'))
+source(paste0(di, 'script/R/rw_byTree.R'))
+source(paste0(di, 'script/R/rwiResilience.R'))
 ```
@@
     site == "caH" ~ "1_caH",
-    site == "caL" ~ "2_caL")
+    site == "caL" ~ "2_caL"))
 ```
@@
 
+# Resilience
+
+* Computar métricas de resiliencia RWI para los tres sitios. 
+* Computar tres eventos climáticos: 1995, 2005, 2012
+* Computar ventanas temporales: 2, 3 y 4
+
+## Aggregate RWI by tree
+
+* Agregar valores medios de RWI por site (obtenemos sj_tree / caL_tree, caH_tree) 
+* ver fun rw_byTree (utilizada para bai_resilience.Rmd)
+
+```{r}
+# Remember snc = structure of core name SJ0101 (site | tree | core)
+rwi_sj_tree <- rw_byTree(rwi_sj, snc =c(2,2,2), locname = 'SJ')
+rwi_caL_tree <- rw_byTree(rwi_caL, snc =c(2,2,2), locname = 'CA')
+rwi_caH_tree <- rw_byTree(rwi_caH, snc =c(2,2,2), locname = 'CA')
+```
+
+```{r}
+# Drought years 
+dyears <- c(1995, 2005, 2012)
+
+# SJ 
+res_4_sj <- rwiResilience(bai_sj, event_years = dyears, window = 4)
+res_3_sj <- rwiResilience(bai_sj, event_years = dyears, window = 3)
+res_2_sj <- rwiResilience(bai_sj, event_years = dyears, window = 2)
+
+# caL
+res_4_caL <- rwiResilience(bai_caL, event_years = dyears, window = 4)
+res_3_caL <- rwiResilience(bai_caL, event_years = dyears, window = 3)
+res_2_caL <- rwiResilience(bai_caL, event_years = dyears, window = 2)
+
+# caH
+res_4_caH <- rwiResilience(bai_caH, event_years = dyears, window = 4)
+res_3_caH <- rwiResilience(bai_caH, event_years = dyears, window = 3)
+res_2_caH <- rwiResilience(bai_caH, event_years = dyears, window = 2)
+```
+
+```{r}
+
+dyears <- c(1995, 2005, 2012)
+event_years <- dyears
+window = 4
+
+rwiResilience <- function(rwi_tree, event_years)
+
+  out_rwi <- c() 
+  out_res <- c() 
+  out <- list()
+
+  require(dplyr)
+
+  for (i in event_years){
+    # Create df previous, during and post event
+    df <- rwi_tree %>% 
+      mutate(year = as.numeric(row.names(.))) %>% as.data.frame()
+
+    df_pre <- df %>% filter(year < i & year >= (i - window)) %>% mutate(disturb = 'pre')
+    df_event <- df %>% filter(year == i) %>% mutate(disturb = 'dr')
+    df_post <- df %>% filter(year > i & year <= (i + window)) %>% mutate(disturb = 'post')
+
+    # Melt data 
+    dfmelt <- bind_rows(df_pre, df_event, df_post) %>% 
+      gather(tree, value, -disturb, -year)
+
+    # Compute mean of RWI (previous, during and posterior)
+    rwi_df <- dfmelt %>% group_by(disturb, tree) %>% 
+      summarise(mean_period = mean(value),
+                sd_period = sd(value),
+                se_period = sd_period/sqrt(length(value))) %>% 
+      mutate(disturb_year = i) %>% as.data.frame()
+
+    out_rwi <- rbind(out_rwi, rwi_df)
+
+    # Compute resilience by trees 
+    aux_resilience <- c() 
+    trees <- unique(rwi_df$tree)
+
+    for (t in trees){
+      # Filter by tree
+      df_tree <- rwi_df %>% filter(tree == t)
+
+      # Compute resilience metrics by tree
+      aux_df <- df_tree %>% dplyr::select(-c(tree, sd_period, se_period)) %>% 
+        spread(key=disturb, value=mean_period) %>% 
+        mutate(rt = dr / pre,
+               rc = post / dr,
+               rs = post / pre,
+               rrs = ((post - dr) / pre),
+               tree = t) %>% 
+        dplyr::select(rt, rc, rs, rrs, disturb_year, tree) %>% as.data.frame()
+
+      aux_resilience <- rbind(aux_resilience, aux_df)
+    }
+
+    out_res <- rbind(out_res, aux_resilience)
+  }
+
+  out$rwi <- out_rwi
+  out$resilience <- out_res
+
+    return(out)
+}
+
+i <- 1995
+rwi_tree <- rwi_sj_tree
+
+```
+
 VAS POR AQUI
@@
 ```{r}
+
+## Download data from E-OBS
+# http://www.ecad.eu/download/ensembles/download.php#datafiles
+
+library(loadeR)
+
+# Define dir and list nc files 
+mydir <- '/Users/ajpelu/Downloads/climaobs'
+
+list.files(mydir, pattern = "\\.nc$")
+
+di <- dataInventory("/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc")
 

### 2017-12-27 analysis/chronos/analysis_splines.Rmd
update analysis splines

--- 2017-12-26 analysis/chronos/analysis_splines.Rmd
@@
 library("pander")
+library("gridExtra")
 ```
@@
 
-VAS POR AQUI
+# Resilience
 
-# CLIMATE RELATIONSHIPS 
+* Computar métricas de resiliencia RWI para los tres sitios. 
+* Computar tres eventos climáticos: 1995, 2005, 2012
+* Computar ventanas temporales: 2, 3 y 4
+
+## Aggregate RWI by tree
+
+* Agregar valores medios de RWI por site (obtenemos sj_tree / caL_tree, caH_tree) 
+* ver fun rw_byTree (utilizada para bai_resilience.Rmd)
+
 ```{r}
-
-library(ncdf4)
-
-nc <- nc_open( "/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc")
-print(nc)
-
-lon <- ncvar_get(nc,"longitude")
-
-http://geog.uoregon.edu/bartlein/courses/geog490/week04-netCDF.html#open-the-netcdf-file
-
-eobs
-
-untar("/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc", exdir = "/Users/ajpelu/Downloads/climaobs")
-
-# Precipitation dataset URL
-ds <- "/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc"
-
-di <- dataInventory("/Users/ajpelu/Downloads/climaobs/tg_0.25deg_reg_v16.0.nc")
-  "http://opendap.knmi.nl/knmi/thredds/dodsC/e-obs_0.25regular/rr_0.25deg_reg_v16.0.nc"
-
-http://meteo.unican.es/work/loadeR/wiki/html/OaccessingLoadingStationData.html
-download.file("http://meteo.unican.es/work/loadeR/data/VALUE_ECA_86_v2.tar.gz", destfile ="/Users/ajpelu/Downloads/climaobs/VALUE_ECA_86_v2.tar.gz")
-
-value <- "/Users/ajpelu/Downloads/climaobs/VALUE_ECA_86_v2"
-
-di <-dataInventory(value)
-
-?loa
-
-precip  <- loadGridData(dataset = ds,
-			   var = "rr",
-			   lonLim = -3.428,
-			   latLim = 36.957,
-			   years = 1950:2016,
-			   season = 1:12)
-
+# Remember snc = structure of core name SJ0101 (site | tree | core)
+rwi_sj_tree <- rw_byTree(rwi_sj, snc =c(2,2,2), locname = 'SJ')
+rwi_caL_tree <- rw_byTree(rwi_caL, snc =c(2,2,2), locname = 'CA')
+rwi_caH_tree <- rw_byTree(rwi_caH, snc =c(2,2,2), locname = 'CA')
 ```
 
-Resilience RWI ???
+```{r}
+# Drought years 
+dyears <- c(1995, 2005, 2012)
 
-# RESILIENCE RWI
+# SJ 
+res_4_sj <- rwiResilience(rwi_sj_tree, event_years = dyears, window = 4)
+res_3_sj <- rwiResilience(rwi_sj_tree, event_years = dyears, window = 3)
+res_2_sj <- rwiResilience(rwi_sj_tree, event_years = dyears, window = 2)
+
+# caL
+res_4_caL <- rwiResilience(rwi_caL_tree, event_years = dyears, window = 4)
+res_3_caL <- rwiResilience(rwi_caL_tree, event_years = dyears, window = 3)
+res_2_caL <- rwiResilience(rwi_caL_tree, event_years = dyears, window = 2)
+
+# caH
+res_4_caH <- rwiResilience(rwi_caH_tree , event_years = dyears, window = 4)
+res_3_caH <- rwiResilience(rwi_caH_tree , event_years = dyears, window = 3)
+res_2_caH <- rwiResilience(rwi_caH_tree , event_years = dyears, window = 2)
+```
+
+### Computar correlaciones ventanas temporales 
+```{r, echo = FALSE}
+# Vector with objects name
+obj <- c('res_2_sj', 'res_3_sj', 'res_4_sj',
+         'res_2_caL', 'res_3_caL', 'res_4_caL',
+         'res_2_caH', 'res_3_caH', 'res_4_caH')
+
+correla_ws <- c()
+
+for (i in obj){ 
+  x <- get(i)
+  xres <- x$resilience
+  out <- xres %>% 
+    mutate(ws = paste0('ws_', as.character(str_extract(i, "([0-9])"))),
+           site = str_replace(i, "res_[0-9]_", '')) %>%
+    dplyr::select(-disturb_year, -tree)
+
+  correla_ws <- bind_rows(correla_ws, out)
+
+}
+
+# Split by window size 
+correla <- correla_ws %>% split(.$ws)
+
+# Change names 
+names(correla[["ws_2"]])[1:4] <- paste0(names(correla[["ws_2"]])[1:4], '2')
+names(correla[["ws_3"]])[1:4] <- paste0(names(correla[["ws_3"]])[1:4], '3')
+names(correla[["ws_4"]])[1:4] <- paste0(names(correla[["ws_4"]])[1:4], '4')
+
+cor2 <- correla[["ws_2"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
+cor3 <- correla[["ws_3"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
+cor4 <- correla[["ws_4"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
+
+correlations <- inner_join(cor2, cor3, by='ind') %>% inner_join(cor4, by='ind')
+```
+
+#### Resistance
+
+```{r correla_temporal_window_rt, echo = FALSE}
+# Resistance
+aux_coefs <- c()
+
+model <- lm(rt2~rt3, data=correlations)
+p_rt23 <- correlations %>% ggplot(aes(rt2, rt3)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) + 
+  theme(legend.position = c(.2, .75))
+aux <- as.data.frame(cbind('rt','2-3', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+model <- lm(rt2~rt4, data=correlations)
+p_rt24 <- correlations %>% ggplot(aes(rt2, rt4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rt','2-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+model <- lm(rt3~rt4, data=correlations)
+p_rt34 <- correlations %>% ggplot(aes(rt3, rt4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rt','3-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+grid.arrange(p_rt23, p_rt24, p_rt34,ncol=3) 
+```
+
+#### Recovery
+
+```{r correla_temporal_window_rc, echo = FALSE}
+# Recovery 
+model <- lm(rc2~rc3, data=correlations)
+p_rc23 <- correlations %>% ggplot(aes(rc2, rc3)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) + 
+  theme(legend.position = c(.2, .75))
+aux <- as.data.frame(cbind('rc','2-3', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+model <- lm(rc2~rc4, data=correlations)
+p_rc24 <- correlations %>% ggplot(aes(rc2, rc4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rc','2-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+model <- lm(rc3~rc4, data=correlations)
+p_rc34 <- correlations %>% ggplot(aes(rc3, rc4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rc','3-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+grid.arrange(p_rc23, p_rc24, p_rc34,ncol=3) 
+```
+
+#### Resilience
+
+```{r correla_temporal_window_rs, echo = FALSE}
+# Resilience
+model <- lm(rs2~rs3, data=correlations)
+p_rs23 <- correlations %>% ggplot(aes(rs2, rs3)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) + 
+  theme(legend.position = c(.2, .75))
+aux <- as.data.frame(cbind('rs','2-3', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+model <- lm(rs2~rs4, data=correlations)
+p_rs24 <- correlations %>% ggplot(aes(rs2, rs4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rs','2-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+model <- lm(rs3~rs4, data=correlations)
+p_rs34 <- correlations %>% ggplot(aes(rs3, rs4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rs','3-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+grid.arrange(p_rs23, p_rs24, p_rs34,ncol=3) 
+```
+
+#### Relative Resilience
+
+```{r correla_temporal_window_rrs, echo = FALSE}
+# Relative Resilience
+model <- lm(rrs2~rrs3, data=correlations)
+p_rrs23 <- correlations %>% ggplot(aes(rrs2, rrs3)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) + 
+  theme(legend.position = c(.2, .75))
+aux <- as.data.frame(cbind('rrs','2-3', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+model <- lm(rrs2~rrs4, data=correlations)
+p_rrs24 <- correlations %>% ggplot(aes(rrs2, rrs4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rrs','2-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+model <- lm(rrs3~rrs4, data=correlations)
+p_rrs34 <- correlations %>% ggplot(aes(rrs3, rrs4)) + 
+  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
+  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
+  theme(legend.position = 'none')
+aux <- as.data.frame(cbind('rrs','3-4', as.numeric(summary(model)$r.squared)))
+aux_coefs <- rbind(aux_coefs, aux)
+
+grid.arrange(p_rrs23, p_rrs24, p_rrs34,ncol=3) 
+```
+
+#### Coeficientes de correlacion entre ventanas temporales
+
+```{r echo=FALSE}
+# Export correlations window size 
+names(aux_coefs) <- c('var', 'window_size', 'r2')
+
+write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size_rwi.csv'), row.names = F)
+```
+
+```{r}
+aux_coefs %>% pander()
+```
+
+Nos quedamos con 3 años de ventana temporal.
+
+# Plots Crecimiento
+## Boxplot with outliers 
+```{r rwi_disturbances_boxplot, echo = FALSE}
+gsj <- res_3_sj$rwi %>% mutate(site='SJ')
+gcaL <- res_3_caL$rwi %>% mutate(site='caL')
+gcaH <- res_3_caH$rwi %>% mutate(site='caH')
+
+g <- bind_rows(gsj, gcaL, gcaH)
+
+# Export csv 
+write.csv(g, file=paste0(di, 'data/resilience/rwi_drought.csv'), row.names = FALSE)
+
+g %>% mutate(disturb = dplyr::recode(disturb, 
+                              dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  ggplot(aes(y=mean_period, x=disturb)) + 
+  geom_boxplot() + 
+  facet_grid(site~disturb_year, scales='free_y') + 
+  theme_bw()
+```
+
+## Mean + se 
+```{r rwi_disturbances_mean_se, echo = FALSE}
+g %>% 
+  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  group_by(disturb, disturb_year, site) %>% 
+  summarise(mean = mean(mean_period),
+            sd = sd(mean_period),
+            se = sd/sqrt(length(mean_period))) %>%
+  ggplot(aes(y=mean, x=disturb)) + 
+  geom_errorbar(aes(ymin=mean - se, 
+                    ymax=mean + se),
+                width = 0.2) +
+  geom_point(size=2, shape=21, fill='white') + 
+  facet_grid(site~disturb_year, scales='free_y') + 
+  theme_bw() + ylab('Ring width index') + 
+  theme(panel.grid = element_blank())
+```
+
+```{r rwi_disturbances_mean_se_colours, echo = FALSE}
+pgrowht <- g %>% 
+  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  group_by(disturb, disturb_year, site) %>% 
+  summarise(mean = mean(mean_period),
+            sd = sd(mean_period),
+            se = sd/sqrt(length(mean_period))) %>%
+  ggplot(aes(y=mean, x=disturb, colour=site)) + 
+  geom_errorbar(aes(ymin=mean - se, 
+                    ymax=mean + se),
+                width = 0.15) +
+  geom_point(size=3, aes(shape=site), fill='white') + 
+  geom_line(aes(group=site))+
+  facet_wrap(~disturb_year, scales='free_y') + 
+  theme_bw() + ylab('Ring width index') +  
+  scale_shape_manual(values=c(15,19,19)) + 
+  theme(panel.grid = element_blank())
+
+pgrowht
+ggsave(plot=pgrowht, width=8, height = 4,
+       filename=paste0(di, 'out/mean_resilience/rwi/rwi_mean.pdf')) 
+```
+
+## tables
+
+```{r}
+g %>% 
+  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
+  dplyr::group_by(site, disturb_year, disturb) %>% 
+  dplyr::summarise(mean = mean(mean_period/100), 
+                   sd = sd(mean_period/100),
+                   se = sd/sqrt(length(mean_period/100))) %>% as.data.frame() %>% pander() 
+```
+
+# Prepare data of Resiliencia for ANOVAS
+```{r}
+# Prepara data 
+rsj <- res_3_sj$resilience %>% mutate(site='SJ')
+rcaL<- res_3_caL$resilience %>% mutate(site='caL')
+rcaH <- res_3_caH$resilience %>% mutate(site='caH')
+
+re <- bind_rows(rsj, rcaL, rcaH)
+re$disturb_year <- as.factor(re$disturb_year)
+re$site <- as.factor(re$site)
+
+# Export csv 
+write.csv(re, file=paste0(di, 'data/resilience/resilience_rwi.csv'), row.names = FALSE)
+```

### 2017-12-27 man/notes_dendro_analysis.Rmd
update analysis splines

@@
 * Seguidamente dividir los datos brutos (rw) entre los datos ajustados (obtenidos con splines) y obtendremos una ***residual ring-width indices**
-VAS POR AQUI EN LA EXPLICACIÓN
+
+### Chronologias short-term
+
+Una vez obtenidos los RWI, vamos a construir una chronología.
+Podemos hacerlo con los valores estándar o con los valores residuales.
+En ambos casos, se procede a obtener los valores medios de rwi para cada año.
+Para ello utilizamos una media robusta, en concreto, Tukey's biweight robust mean, que es un método insensible a los outliers [@Riitters1990].
+
+En el caso de la chornologia residual, previamente a la agregación de los valores anuales (biweight) se aplica un modelo autorregresivo (prewithening = TRUE) para minimizar el impacto de la autocorrelación.
+Así obtendremos una serie estacionara de eventos independientes con media y varianza constante.
 

### 2017-12-27 analysis/rwi/analysis_resilience_rwi.Rmd
resilience RWI

--- 2017-12-15 analysis/bai/analysis_resilience_bai.Rmd
@@
 ---
-title: "Growth Resilience (reanalisis)"
+title: "Anovas RWI Resilience"
 author: "AJ Perez-Luque (@ajpelu)"
 date: "2017 Dec"
-output: pdf_document 
+output: 
+  md_document:
+    variant: markdown_github
 bibliography: ../references.bib 
@@
 ```
-# Resilience
-
-* Calcularemos las métricas resiliencia de [@Lloret2011] sobre el crecimiento. 
-* Vamos a calcularlas sobre el BAI de cada árbol. 
-* Utilizaremos tres sitios: SJ, CAH y CAL (ver [./analysis/analysis_chronologies.md]('./analysis/analysis_chronologies.md))
-
-## Prepare data
-
-* Leer datos `rwl` de SJ y CA
-* Leer datos de diametros de los focal tree
+
+# Resilience ANOVAs
+
+* Partimos de los datos obtenidos de resiliencia para los RWI ([./analysis/chronos/analysis_splines.md]('./analysis/chronos/analysis_splines.md)) )
 
@@
 
-# sj 
-sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
-
-# canar 
-ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
-
 # Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-```
-
-```{r}
-source(paste0(di, 'script/R/rw_byTree.R'))
-source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/baiResilience.R'))
-```
-
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low
-
-```{r, echo=FALSE}
-# Replace SNA by SJ and SNB by CA
-names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
-names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")
-
-# Remove g in name of some cores of CA. 
-names(ca) <- stringr::str_replace(names(ca), "g", "")
-```
-
-```{r, echo=FALSE}
-# Create subset to compare between sites 
-caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
-             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
-             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
-caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
-              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
-              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
-
-```
-
-* Lectura y preparación de datos de diámetro
-
-```{r} 
-# Prepare Diameter data
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Change name focal according to loc
-compete <- compete %>% 
-  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm)
-
-# Set levels of eleveation 
-ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
-            paste0('CA', 26:30))
-ca_highcode <- paste0('CA', 11:25)
-
-ft <- ft %>% 
-  mutate(site = as.factor(
-    ifelse(id_focalLoc %in% ca_lowcode, 'CAL', 
-           ifelse(id_focalLoc %in% ca_highcode, 'CAH', 'SJ'))))
-```
-
-# Aggregate RW by tree
-
-* Agregar valores medios de RW por site (obtenemos sj_tree / caL_tree, caH_tree) 
-* ver fun rw_byTree o utilizar treeMean (dplR)
-
-```{r}
-# Remember snc = structure of core name SJ0101 (site | tree | core)
-sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
-caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
-caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')
-```
-
-* Crear diferentes dataset de diametro por sitio 
-```{r}
-diam <- ft %>%
-  mutate(diameter = dn_mm, 
-         id = id_focalLoc) %>%
-  dplyr::select(id, diameter, site) %>% 
-  split(.$site)
-
-d_caH <- diam$CAH[,c('id','diameter')]
-d_caL <- diam$CAL[,c('id','diameter')]
-d_sj <- diam$SJ[,c('id','diameter')]
-```
-
-## Cómputo del BAI por site
-
-* He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]. Es similar a `bai.out`
-
-```{r}
-bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
-bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
-bai_caL <- bai_piovesan(rwdf = caL_tree, diam_df = d_caL)
-
-# Set class to bai object 
-# Esto es para que funcionen algunas otras funciones de dplR 
-bais <- c('bai_sj', 'bai_caH', 'bai_caL')
-
-for (i in bais){
-  aux <- get(i)
-
-  class(aux) <- c('rwl', 'data.frame')
-
-  assign(i, aux)
-}
-
-```
-
-## Resilience
-
-* Computar métricas de resiliencia BAI para los tres sitios. 
-* Computar tres eventos climáticos: 1995, 2005, 2012
-* Computar ventanas temporales: 2, 3 y 4
-
-```{r}
-# Drought years 
-dyears <- c(1995, 2005, 2012)
-
-# SJ 
-res_4_sj <- baiResilience(bai_sj, event_years = dyears, window = 4)
-res_3_sj <- baiResilience(bai_sj, event_years = dyears, window = 3)
-res_2_sj <- baiResilience(bai_sj, event_years = dyears, window = 2)
-
-# caL
-res_4_caL <- baiResilience(bai_caL, event_years = dyears, window = 4)
-res_3_caL <- baiResilience(bai_caL, event_years = dyears, window = 3)
-res_2_caL <- baiResilience(bai_caL, event_years = dyears, window = 2)
-
-# caH
-res_4_caH <- baiResilience(bai_caH, event_years = dyears, window = 4)
-res_3_caH <- baiResilience(bai_caH, event_years = dyears, window = 3)
-res_2_caH <- baiResilience(bai_caH, event_years = dyears, window = 2)
-```
-
-### Computar correlaciones ventanas temporales 
-```{r, echo = FALSE}
-# Vector with objects name
-obj <- c('res_2_sj', 'res_3_sj', 'res_4_sj',
-         'res_2_caL', 'res_3_caL', 'res_4_caL',
-         'res_2_caH', 'res_3_caH', 'res_4_caH')
-
-correla_ws <- c()
-
-for (i in obj){ 
-  x <- get(i)
-  xres <- x$resilience
-  out <- xres %>% 
-    mutate(ws = paste0('ws_', as.character(str_extract(i, "([0-9])"))),
-           site = str_replace(i, "res_[0-9]_", '')) %>%
-    dplyr::select(-disturb_year, -tree)
-
-  correla_ws <- bind_rows(correla_ws, out)
-
-}
-
-# Split by window size 
-correla <- correla_ws %>% split(.$ws)
-
-# Change names 
-names(correla[["ws_2"]])[1:4] <- paste0(names(correla[["ws_2"]])[1:4], '2')
-names(correla[["ws_3"]])[1:4] <- paste0(names(correla[["ws_3"]])[1:4], '3')
-names(correla[["ws_4"]])[1:4] <- paste0(names(correla[["ws_4"]])[1:4], '4')
-
-cor2 <- correla[["ws_2"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-cor3 <- correla[["ws_3"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-cor4 <- correla[["ws_4"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-
-correlations <- inner_join(cor2, cor3, by='ind') %>% inner_join(cor4, by='ind')
-```
-
-#### Resistance
-
-```{r correla_temporal_window_rt, echo = FALSE}
-# Resistance
-aux_coefs <- c()
-
-model <- lm(rt2~rt3, data=correlations)
-p_rt23 <- correlations %>% ggplot(aes(rt2, rt3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rt','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rt2~rt4, data=correlations)
-p_rt24 <- correlations %>% ggplot(aes(rt2, rt4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rt','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rt3~rt4, data=correlations)
-p_rt34 <- correlations %>% ggplot(aes(rt3, rt4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rt','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rt23, p_rt24, p_rt34,ncol=3) 
-```
-
-#### Recovery
-
-```{r correla_temporal_window_rc, echo = FALSE}
-# Recovery 
-model <- lm(rc2~rc3, data=correlations)
-p_rc23 <- correlations %>% ggplot(aes(rc2, rc3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rc','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rc2~rc4, data=correlations)
-p_rc24 <- correlations %>% ggplot(aes(rc2, rc4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rc','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rc3~rc4, data=correlations)
-p_rc34 <- correlations %>% ggplot(aes(rc3, rc4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rc','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rc23, p_rc24, p_rc34,ncol=3) 
-```
-
-#### Resilience
-
-```{r correla_temporal_window_rs, echo = FALSE}
-# Resilience
-model <- lm(rs2~rs3, data=correlations)
-p_rs23 <- correlations %>% ggplot(aes(rs2, rs3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rs','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rs2~rs4, data=correlations)
-p_rs24 <- correlations %>% ggplot(aes(rs2, rs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rs','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rs3~rs4, data=correlations)
-p_rs34 <- correlations %>% ggplot(aes(rs3, rs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rs','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rs23, p_rs24, p_rs34,ncol=3) 
-```
-
-#### Relative Resilience
-
-```{r correla_temporal_window_rrs, echo = FALSE}
-# Relative Resilience
-model <- lm(rrs2~rrs3, data=correlations)
-p_rrs23 <- correlations %>% ggplot(aes(rrs2, rrs3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rrs','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rrs2~rrs4, data=correlations)
-p_rrs24 <- correlations %>% ggplot(aes(rrs2, rrs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rrs','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rrs3~rrs4, data=correlations)
-p_rrs34 <- correlations %>% ggplot(aes(rrs3, rrs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rrs','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rrs23, p_rrs24, p_rrs34,ncol=3) 
-```
-
-#### Coeficientes de correlacion entre ventanas temporales
-
-```{r echo=FALSE}
-# Export correlations window size 
-names(aux_coefs) <- c('var', 'window_size', 'r2')
-
-write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), row.names = F)
-```
-
-```{r}
-aux_coefs %>% pander()
-```
-
-Nos quedamos con 3 años de ventana temporal.
-
-# Plots Crecimiento
-## Boxplot with outliers 
-```{r growth_disturbances_boxplot, echo = FALSE}
-gsj <- res_3_sj$growth %>% mutate(site='SJ')
-gcaL <- res_3_caL$growth %>% mutate(site='caL')
-gcaH <- res_3_caH$growth %>% mutate(site='caH')
-
-g <- bind_rows(gsj, gcaL, gcaH)
-
-# Export csv 
-write.csv(g, file=paste0(di, 'data/resilience/crecimientos_drought.csv'), row.names = FALSE)
-
-g %>% mutate(disturb = dplyr::recode(disturb, 
-                              dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  ggplot(aes(y=mean_period/100, x=disturb)) + 
-  geom_boxplot() + 
-  facet_grid(site~disturb_year, scales='free_y') + 
-  theme_bw()
-```
-
-## Mean + se 
-```{r growth_disturbances_mean_se, echo = FALSE}
-g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  group_by(disturb, disturb_year, site) %>% 
-  summarise(mean = mean(mean_period),
-            sd = sd(mean_period),
-            se = sd/sqrt(length(mean_period))) %>%
-  ggplot(aes(y=mean/100, x=disturb)) + 
-  geom_errorbar(aes(ymin=mean/100 - se/100, 
-                    ymax=mean/100 + se/100),
-                width = 0.2) +
-  geom_point(size=2, shape=21, fill='white') + 
-  facet_grid(site~disturb_year, scales='free_y') + 
-  theme_bw() + ylab('BAI (cm2/year)') + 
-  theme(panel.grid = element_blank())
-```
-
-```{r growth_disturbances_mean_se_colours, echo = FALSE}
-pgrowht <- g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  group_by(disturb, disturb_year, site) %>% 
-  summarise(mean = mean(mean_period),
-            sd = sd(mean_period),
-            se = sd/sqrt(length(mean_period))) %>%
-  ggplot(aes(y=mean/100, x=disturb, colour=site)) + 
-  geom_errorbar(aes(ymin=mean/100 - se/100, 
-                    ymax=mean/100 + se/100),
-                width = 0.15) +
-  geom_point(size=3, aes(shape=site), fill='white') + 
-  geom_line(aes(group=site))+
-  facet_wrap(~disturb_year, scales='free_y') + 
-  theme_bw() + ylab('BAI (cm2/year)') + 
-  scale_shape_manual(values=c(15,19,19)) + 
-  theme(panel.grid = element_blank())
-
-pgrowht
-ggsave(plot=pgrowht, width=8, height = 4,
-       filename=paste0(di, 'out/fig/resilience/bai_events.pdf')) 
-```
-
-## tables
-
-```{r}
-g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  dplyr::group_by(site, disturb_year, disturb) %>% 
-  dplyr::summarise(mean = mean(mean_period/100), 
-                   sd = sd(mean_period/100),
-                   se = sd/sqrt(length(mean_period/100))) %>% as.data.frame() %>% pander() 
-```
-
-# Anovas Resiliencia
-
-```{r}
-# Prepara data 
-rsj <- res_3_sj$resilience %>% mutate(site='SJ')
-rcaL<- res_3_caL$resilience %>% mutate(site='caL')
-rcaH <- res_3_caH$resilience %>% mutate(site='caH')
-
-re <- bind_rows(rsj, rcaL, rcaH)
-re$disturb_year <- as.factor(re$disturb_year)
-re$site <- as.factor(re$site)
-
-# Export csv 
-write.csv(re, file=paste0(di, 'data/resilience/resilience_bai.csv'), row.names = FALSE)
+re <- read.csv(file=paste0(di, 'data/resilience/resilience_rwi.csv'), header=TRUE, sep=',')
 ```
@@
 write.csv(normtestA, 
-          file=paste0(di, '/out/anovas_resilience/bai/normo_disturb_year.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/rwi/normo_disturb_year.csv'), row.names = F)
 ```
@@
 write.csv(normtestB, 
-          file=paste0(di, '/out/anovas_resilience/bai/normo_site.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/rwi/normo_site.csv'), row.names = F)
 ```
@@
 write.csv(normtestAB, 
-          file=paste0(di, '/out/anovas_resilience/bai/normo_disturb_year_site.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/rwi/normo_disturb_year_site.csv'), row.names = F)
 ```
@@
 ## See auxiliar::homogetest
+re$disturb_year <- as.factor(as.character(re$disturb_year))
 factores <- c('disturb_year', 'site', 'interaction(disturb_year, site)') 
@@
 write.csv(homo, 
-          file=paste0(di, '/out/anovas_resilience/bai/homocedasticidad.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/rwi/homocedasticidad.csv'), row.names = F)
 ```
@@
 write.csv(homo_log, 
-          file=paste0(di, '/out/anovas_resilience/bai/homocedasticidad_log.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/rwi/homocedasticidad_log.csv'), row.names = F)
 ```
@@
 ### mean + sd
-```{r plot_resiliences_bai_sd,fig.width = 12, fig.height=12}
+```{r plot_resiliences_rwi_sd,fig.width = 12, fig.height=12}
 grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
@@
 ### mean + ci
-```{r plot_resiliences_bai_ci, fig.width = 12, fig.height=12}
+```{r plot_resiliences_rwi_ci, fig.width = 12, fig.height=12}
 grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
@@
 ```{r, echo=FALSE}
-pdf(paste0(di, '/out/mean_resilience/bai/interaction_plotsSE.pdf'), width=9, height = 9)
+pdf(paste0(di, '/out/mean_resilience/rwi/interaction_plotsSE.pdf'), width=9, height = 9)
 grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
@@
 
-pdf(paste0(di, '/out/mean_resilience/bai/interaction_plotsCI.pdf'), width=9, height = 9)
+pdf(paste0(di, '/out/mean_resilience/rwi/interaction_plotsCI.pdf'), width=9, height = 9)
 grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
@@
 
-write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/bai/anovas_statistics.csv'), row.names = F)
+write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/rwi/anovas_statistics.csv'), row.names = F)
 
@@
 write.csv(aovas_model_summary, 
-          file=paste0(di, '/out/anovas_resilience/bai/anovas_summary_modelos.csv'), row.names = F)
+          file=paste0(di, '/out/anovas_resilience/rwi/anovas_summary_modelos.csv'), row.names = F)
 
@@
 
-```{r robust_plot_bai_drought}
+```{r robust_plot_rwi_drought}
 mhuber<- mhuber %>% 
@@
 
-robust_plot_bai_drought <- ggplot(mhuber, aes(x=disturb_year, y=M.Huber, color = site, group=site, fill=site)) + 
+robust_plot_rwi_drought <- ggplot(mhuber, aes(x=disturb_year, y=M.Huber, color = site, group=site, fill=site)) + 
   geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), 
@@
 
-robust_plot_bai_drought
-
-pdf(paste0(di, 'out/resilience_robust/bai/robust_plot_bai_drought.pdf'), width=9, height = 9)
-robust_plot_bai_drought 
+robust_plot_rwi_drought
+
+pdf(paste0(di, 'out/resilience_robust/rwi/robust_plot_rwi_drought.pdf'), width=9, height = 9)
+robust_plot_rwi_drought
 dev.off()
@@
 ### Response ~ (x=site)
-```{r robust_plot_bai_site}
+```{r robust_plot_rwi_site}
 pd <- position_dodge(.2)
 
-robust_plot_bai_site <- ggplot(mhuber, aes(x=site, y=M.Huber, color = disturb_year, group=disturb_year, fill=disturb_year)) + 
+robust_plot_rwi_site <- ggplot(mhuber, aes(x=site, y=M.Huber, color = disturb_year, group=disturb_year, fill=disturb_year)) + 
   geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), 
@@
 
-robust_plot_bai_site
-
-pdf(paste0(di, 'out/resilience_robust/bai/robust_plot_bai_site.pdf'), width=9, height = 9)
-robust_plot_bai_site 
+robust_plot_rwi_site
+
+pdf(paste0(di, 'out/resilience_robust/rwi/robust_plot_rwi_site.pdf'), width=9, height = 9)
+robust_plot_rwi_site 
 dev.off()
@@
 # Export data 
-write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber.csv'), row.names = F)
-write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg.csv'), row.names = F)
-write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_a.csv'), row.names = F)
-write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg_a.csv'), row.names = F)
-write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_b.csv'), row.names = F)
-write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg_b.csv'), row.names = F)
-
-write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_pairwise.csv'), row.names = F)
-
-write.csv(trimmedanovas, file=paste0(di, '/out/anovas_resilience/huber_bai/trimmed_anovas.csv'), row.names = F)
+write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/huber_rwi/robust_mhuber.csv'), row.names = F)
+write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/huber_rwi/robust_mhuber_agg.csv'), row.names = F)
+write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/huber_rwi/robust_mhuber_a.csv'), row.names = F)
+write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/huber_rwi/robust_mhuber_agg_a.csv'), row.names = F)
+write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/huber_rwi/robust_mhuber_b.csv'), row.names = F)
+write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/huber_rwi/robust_mhuber_agg_b.csv'), row.names = F)
+
+write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/huber_rwi/robust_pairwise.csv'), row.names = F)
+
+write.csv(trimmedanovas, file=paste0(di, '/out/anovas_resilience/huber_rwi/trimmed_anovas.csv'), row.names = F)
 ```

### 2017-12-27 data_raw/eobs/download_and_process_EOBS.R
add Script and processes data of EOBS

@@ -1,0 +1,155 @@
+# Load pkgs
+
+library('loadeR')
+library('transformeR')
+library('rgdal')
+library('raster')
+library('latticeExtra')
+
+# tutorials 
+# http://www.meteo.unican.es/climate4R
+# https://github.com/SantanderMetGroup/loadeR/wiki
+# https://github.com/SantanderMetGroup/transformeR/wiki
+# https://github.com/SantanderMetGroup/visualizeR/wiki 
+
+# download data from EOBS (dec 2017)
+# http://www.ecad.eu/download/ensembles/download.php#datafiles 
+
+dir <- "C:/Users/alpelu/eobs"
+
+
+# Load qp distribution
+# Read spatial data 
+qp<- rgdal::readOGR(dsn=paste0(dir,'/geo_data'),
+                    layer = 'q_pyr_sn_4326', verbose = FALSE, encoding = "UTF-8")
+
+
+
+# list files nc
+list.files(dir, pattern = "\\.nc$")
+
+# Doing inventory 
+di <- dataInventory("C:/Users/alpelu/eobs/rr_0.25deg_reg_v16.0.nc")
+
+# Define limits location: 
+
+# Load data
+preraw <- loadGridData(dataset = "C:/Users/alpelu/eobs/rr_0.25deg_reg_v16.0.nc", 
+                    var = "rr", 
+                    lonLim = c(-3.6, -3.2), 
+                    latLim= c(36.75,37.25), 
+                    aggr.m = "sum")
+
+pre_cli_raw <- climatology(pre_raw)
+
+
+
+spatialPlot(pre_cli,
+            backdrop.theme = "none",
+            scales = list(draw = TRUE)) + layer(sp.polygons(qp))
+
+
+pre <- loadGridData(dataset = "C:/Users/alpelu/eobs/rr_0.25deg_reg_v16.0.nc", 
+                    var = "rr", 
+                    lonLim = c(-3.625,-3.375), 
+                    latLim= c(36.875,37.125), 
+                    aggr.m = "sum")
+pre_cli <- climatology(pre)
+spatialPlot(pre_cli,
+            backdrop.theme = "none",
+            scales = list(draw = TRUE)) + layer(sp.polygons(qp))
+
+# See coordinates
+pre_cli$xyCoords
+
+
+# Create multigrids with variables
+
+# Doing inventory 
[…]

### 2017-12-27 analysis/resilience-plots.Rmd
add resilience plots

--- 2017-12-27 analysis/rwi/analysis_resilience_rwi.Rmd
@@
 ---
-title: "Anovas RWI Resilience"
+title: "Resilience plots"
 author: "AJ Perez-Luque (@ajpelu)"
@@
 library('effects')
-library('devtools')
-# devtools::install_github("ajpelu/auxiliar")
-library('car')
-library('auxiliar')
-library('WRS2')
-library('MASS')
-library('rcompanion')
+```
+
+# Cargar los datos
+```{r}
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
+di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
+```
+
+```{r}
+re_bai <- read.csv(file=paste0(di, 'data/resilience/resilience_bai.csv'), header=TRUE, sep=',')
+re_evi <- read.csv(file=paste0(di, 'data/resilience/resilience_evi.csv'), header=TRUE, sep=',')
+```
+
+```{r}
+
+bai <- re_bai %>% dplyr::select(-tree, -rrs) %>% mutate(mivariable = "BAI")
+evi <- re_evi %>% dplyr::select(-iv_malla_modi_id, -rrs) %>% mutate(mivariable = "EVI")
+
 ```

### 2017-12-27 analysis/prepare_eobs_data.Rmd
add eobs data

---
title: "EOBS Climate data"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Dec"
output:
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, warning=FALSE, error=TRUE, message=FALSE)
```

```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
library("tidyverse")
```

# Cargar los datos
```{r}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
```

```{r}
eobs <- read.csv(file=paste0(di, 'data_raw/eobs/eobs_data_robles.csv'), header=TRUE, sep=',')

eobs <- eobs %>% 
  mutate(date = as.Date(str_sub(startDate, 1, 11), format="%Y-%m-%d"), 
         year = lubridate::year(date), 
         month = lubridate::month(date))

eobs %>% filter(variable == 'rr') %>% ggplot(aes(x=date, y=value, colour=loc)) + geom_line()

```

### 2017-12-27 analysis/prepare_eobs_data.Rmd
try treeclim data

@@
 library("tidyverse")
+library("treeclim")
 ```
@@
 eobs <- read.csv(file=paste0(di, 'data_raw/eobs/eobs_data_robles.csv'), header=TRUE, sep=',')
+cro <- read.csv(file=paste(di, "data/cronos_medias/cronos_sites_rwi.csv", sep=""), header=TRUE, sep=',')
+
+croN <- cro %>% filter(site == 'sj') %>% dplyr::select(-site) %>% 
+  filter(year > 1949)
+rownames(croN) <- croN$year
 
@@
 
+cl <- eobs %>% 
+  filter(loc == 'N') %>% 
+  filter(variable %in% c('rr', 'tg')) %>% 
+  dplyr::select(value, variable, year, month) %>% 
+  spread(variable, value)
+
+cl <- cl %>% rename(prec = rr, temp = tg) %>% 
+  filter(year < 2017)
+
+test <- dcc(chrono = croN, climate = cl)
+
+croN 
+# http://czang.org/treeclim.html
+
 eobs %>% filter(variable == 'rr') %>% ggplot(aes(x=date, y=value, colour=loc)) + geom_line()

### 2017-12-28 analysis/prepare_eobs_data.Rmd
prepare eobs data

--- 2017-12-27 analysis/prepare_eobs_data.Rmd
@@
 library("tidyverse")
+library("stringr")
 ```
 
-# Cargar los datos
 ```{r}
@@
 
+Utilizamos como fuente de datos el dataset E-OBS [@Besselaar2011].
+Se trata de un dataset de datos climáticos de alta resolución de toda Europa.
+Se pueden descargar para varias resoluciones.
+Hemos descargado los datos de 0.25º de resolución, de temperatura media, máxima y mínima; y precipitación.
+Son datos diarios a partir de estaciones y tienen desde 1950.
+Ver mas info en http://www.ecad.eu/download/ensembles/download.php
+
+Una vez descargados, hemos utilizado varios paquetes desarrollados por el [grupo de Meteorología Santander](http://www.meteo.unican.es/climate4r), en concreto:
+
+* loadeR: https://github.com/SantanderMetGroup/loadeR 
+* transformeR: https://github.com/SantanderMetGroup/transformeR/ 
+* visualizeR: https://github.com/SantanderMetGroup/visualizeR/
+
+Para mas informacion ver [@Cofino2018, @Frias2018].
+
+Tuve problemas con JAVA y tuve que hacer la descarga y el post-procesado para seleccionar los datos en windows.
+En el script [`./data_raw/eobs/download_and_process_EOBS.R`](/data_raw/eobs/download_and_process_EOBS.R) está el procedimiento.
+Ojo no están los datos originales (ocupan 17 Gb.
+Ir a la web http://www.ecad.eu/download/ensembles/download.php para descargarlos).
+Una vez seleccionadas las celdas, obtuve un conjunto de datos llamado `eobs_data_robles.csv` que es el que vamos a utilizar.
+
+# Prepare data
+
 ```{r}
@@
 
-eobs <- eobs %>% 
+eobs_wide <- eobs %>% 
   mutate(date = as.Date(str_sub(startDate, 1, 11), format="%Y-%m-%d"), 
          year = lubridate::year(date), 
-         month = lubridate::month(date))
+         month = lubridate::month(date),
+         var_clima = case_when(
+           variable == "rr" ~ "prec",
+           variable == "tg" ~ "tmean",
+           variable == "tn" ~ "tmin",
+           variable == "tx" ~ "tmax")
+         ) %>% 
+  dplyr::select(value, loc, date, year, month, var_clima) %>% 
+  spread(var_clima, value)
 
-eobs %>% filter(variable == 'rr') %>% ggplot(aes(x=date, y=value, colour=loc)) + geom_line()
+write.csv(eobs_wide, file=paste0(di, "data/eobs/eobs_formatted.csv"), row.names = FALSE) 
+```
 
-```
+# References

### 2017-12-28 analysis/rwi/climate_rwi_relate.Rmd
update climate eobs relationships

---
title: "Climate-RWI relations"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Dec"
output:
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, warning=FALSE, error=TRUE, message=FALSE)
```

```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
library("tidyverse")
library("stringr")
library("treeclim")
```

```{r}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
```

# Read and prepare data 
```{r}
eobs <- read.csv(file=paste0(di, "data/eobs/eobs_formatted.csv"), header=TRUE, sep=',')
crono <- read.csv(file=paste(di, "data/cronos_medias/cronos_sites_rwi.csv", sep=""), header=TRUE, sep=',')
```

* Utilizamos la crono residual no la estandar :red_circle: DUDA

```{r}
cro_sj <- crono %>% filter(site == 'sj') %>% dplyr::select(-site, -std) %>% column_to_rownames(var = "year") %>% na.omit()
cro_caL <- crono %>% filter(site == 'caL') %>% dplyr::select(-site, -std) %>% column_to_rownames(var = "year") %>% na.omit()
cro_caH <- crono %>% filter(site == 'caH') %>% dplyr::select(-site, -std) %>% column_to_rownames(var = "year") %>% na.omit()
```

# Create subsets data for eobs

```{r}
eobsN <- eobs %>% filter(year < 2017) %>% filter(loc == 'N')

prec_N <- eobsN %>% dplyr::select(year, month, prec) 
tmean_N <- eobsN %>% dplyr::select(year, month, tmean) 
tmin_N <- eobsN %>% dplyr::select(year, month, tmin) 
tmax_N <- eobsN %>% dplyr::select(year, month, tmax)

eobsS <- eobs %>% filter(year < 2017) %>% filter(loc == 'S')

prec_S <- eobsS %>% dplyr::select(year, month, prec) 
tmean_S <- eobsS %>% dplyr::select(year, month, tmean) 
tmin_S <- eobsS %>% dplyr::select(year, month, tmin) 
tmax_S <- eobsS %>% dplyr::select(year, month, tmax) 
```

# Precipitation
from previous June to

```{r}
rprec_SJ <- dcc(chrono = cro_sj, climate = prec_N, 
         method = "correlation",  boot = "std",
        .range(-3:9) + .sum(-9:8))

rprec_caL <- dcc(chrono = cro_caL, climate = prec_S, 
         method = "correlation",  boot = "std",
        .range(-3:9) + .sum(-9:8))

rprec_caH <- dcc(chrono = cro_caH, climate = prec_S, 
         method = "correlation",  boot = "std",
        .range(-3:9) + .sum(-9:8))

rprecs <- c('rprec_SJ', 'rprec_caH', 'rprec_caL')
rprec_all <- c()

for (i in rprecs){

  # Get object dcc
  aux <- get(i)

  # Get coeffs 
  aux_coef <- aux$coef

  # remove rownames
  row.names(aux_coef) <- NULL

  # Change name for aggregate variables 
  aux_coef <- aux_coef %>% 
    mutate(month = case_when(
      month == "sep...AUG" ~ "Hydrol", 
      TRUE ~ as.character(.$month)),
      sig = case_when(
      significant == "TRUE" ~ "*", 
      significant == "FALSE" ~ ""))

  # Add loc variable
  name_site <- str_sub(i, start=7)
  aux_coef$site <- name_site

  # rbind
  rprec_all <- rbind(rprec_all, aux_coef)
  } 
```

```{r}

rprec_all

rprec_all %>% ggplot(aes(x=id, y=coef, group = site)) +
  geom_bar(aes(fill=site),
           stat="identity", position = "dodge") +
  scale_x_continuous(breaks = rprec_all$id, labels = rprec_all$month) +
  theme_bw() + 
  theme(axis.title.x = element_blank()) +
  geom_text(aes(label = sig), 
            position = position_dodge(0.9), 
            vjust = 0.1, 
            size = 10) +
  ylab('Correlation')

```{r}

eobs_norht_no_prec <- eobs_norht %>% dplyr::select(-prec)

mc <- dcc(chrono = cro_sj, 
          climate = eobs_norht, 
          var_names = c("prec", "tmax", "tmean", "tmin"), 
          method = "correlation", boot = "stationary")

sea_sj <- seascorr(chrono = cro_sj, 
             climate = eobs_norht, 
             var_names = c("prec", "tmax", "tmean", "tmin"))

plot(sea_sj)

dc_sj

croN <- cro %>% filter(site == 'sj') %>% dplyr::select(-site) %>% 
  filter(year > 1949)
rownames(croN) <- croN$year

cl <- eobs %>% 
  filter(loc == 'N') %>% 
  filter(variable %in% c('rr', 'tg')) %>% 
  dplyr::select(value, variable, year, month) %>% 
  spread(variable, value)

cl <- cl %>% rename(prec = rr, temp = tg) %>% 
  filter(year < 2017)

test <- dcc(chrono = croN, climate = cl)

croN 
# http://czang.org/treeclim.html

```

The relationships between monthly climatic variables (mean maximum and minimum temperatures and total precipitation) and the chronologies of the ring-width indices were calculated using bootstrapped correlation functions in the treeclim package in R (Zang & Biondi, 2015).

### 2017-12-29 analysis/figuras_bai_population.Rmd
figures bai and population

---
title: "Figura Chronos_population"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Dec"
output:
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, warning=FALSE, error=TRUE, message=FALSE)
```

```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
library("tidyverse")
library("stringr")
library("patchwork")
```

```{r}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
```

# Read and prepare data 
```{r}
crono <- read.csv(file=paste(di, "data/cronos_medias/cronos_sites.csv", sep=""), header=TRUE, sep=',')
pop <- read.csv(file=paste0(di, "data_raw/population/demography.csv"), header = TRUE)
```

```{r}
year_select <- c(1842, 1857, 1860, 1877, 1887, 1897, 1900, 1910, 1920, 1930, 1940, 1950, 1960, 1970, 1981, 1991, 1996, 2000, 2005, 2010, 2015)
pop <- pop %>% filter(year %in% year_select)
```

# Pop SJ 
```{r population_bai_sj}
pop_sj <- crono %>% filter(site == 'SJ') %>%  
  ggplot(aes(x=year, y=bai_mean/100)) +
  theme_bw() + ylab('BAI') + 
  theme(panel.grid = element_blank(),
        legend.position = c(0.1,.8),
        strip.background = element_blank()) +
  geom_bar(data = pop, aes(x=year, y=gu / 1500), stat = 'identity', fill="#ABABAB") + 
  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
                  ymax = (bai_mean + bai_se)/100,
                  fill=site), alpha=.6, colour=NA) +
  geom_line(aes(color = site), size=1) + 
  scale_y_continuous(sec.axis = sec_axis(~ .*1500, name = '# inhabitants', breaks = c(0,1500, 3000, 4500))) +
  scale_colour_manual(values = '#7570b3') +
  scale_fill_manual(values = '#7570b3')
pop_sj
```

```{r}
fileplot <- paste0(di, '/out/bai_population/bai_pop_sj.pdf') 
ggsave(filename=fileplot, plot=pop_sj, width=12, height = 5)
pop_sj
dev.off()
```

# Pop CA 
```{r population_bai_ca}

pop_ca <- crono %>% filter(site != 'SJ') %>%  
  ggplot(aes(x=year, y=bai_mean/100)) +
  theme_bw() + ylab('BAI') + 
  theme(panel.grid = element_blank(),
        legend.position = c(0.1,.8),
        strip.background = element_blank()) +
  geom_bar(data = pop, aes(x=year, y=ca / 300), stat = 'identity', fill="#ABABAB") + 
  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
                  ymax = (bai_mean + bai_se)/100,
                  fill=site), alpha=.6, colour=NA) +
  geom_line(aes(color = site), size=1) + 
  scale_y_continuous(sec.axis = sec_axis(~ .*300, name = '# inhabitants', breaks = c(0,500, 1000))) +
  scale_colour_manual(values = c('#e7298a','#1b9e77')) +
  scale_fill_manual(values = c('#e7298a','#1b9e77')) 
```

```{r}
fileplot <- paste0(di, '/out/bai_population/bai_pop_ca.pdf') 
ggsave(filename=fileplot, plot=pop_ca, width=12, height = 5)
pop_ca
dev.off()
```

### 2018-01-03 man/ms_1_material_methods.Rmd
update

@@
 Two cores of 5 mm of diameter were taken from each at breast heith (1.3 m) using an increment borer.
-For each tree, diameter at breat height (DBH) and total height were recorded.
-
-Incluir datos de muestreo de competencia
+Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively.
+We assessed stand competition of the target tree by recording all neighboring living trees with dbh > 7.5 cm in a circle of 10 m radi.
+The species, dbh and total height were annotated.
+We also measured distance and azimuth to the target tree.
+(:red_circle: indices competencia??)
 
@@
 We used measured dbh and raw nondetrened ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
-We calculated a mean BAI series for each individual tree.
-To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies (growth index, GIres) calculated from biweight means of ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976].
+We calculated a mean BAI serie for each individual tree.
+(:red_circle: equation BAI).
+...
+Site chronologies were built by averaging all tree BAI measurement of the same site.
+
+To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976](:red:circle: (Cook and Kairiukstis 1990)..
+We standardized and detrended the tree-ring width series using the `dplR` [@Bunn2008, @Bunn2010] package.
+Residual site chronologies were obtained computing the bi-weight robust mean of all prewhitened growth indices for the trees of the same site.
+
+Statistical descriptive parameters (Fritts, 2001), including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology considering the common 1950–2013 period.
+
+* incluir tree clim [@Zang2015],  EPS
+
+... The diameters at the breast height (DBH) measured in the field were used to convert each individual ring-width series to basal area increment (BAI) using dplR (Bunn 2008).
+
+The chronologies were computed as bi-weight robust mean to reduce the bias caused by extreme values and an autoregressive model was applied to remove the autocorrelation of the series (Cook and Kairiukstis 1990).
+The statistical quality of each chronology was checked via expressed population signal (EPS; Wigley et al.
+1984).
+A threshold value of EPS >0.85 was used to determine the cutoff year of the time span that could be considered reliable.
+The common period 1950–2012 was used for further analyses in order to avoid the period of intensive logging in TJN (see also Fig.
+2)....
+
+### Climate and growth
+
+Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008].
+Monthly precipitation and mean, minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period.
+We selected grid cells covering each sampled sites.
+
+treeclim aqui
+
+The SPEI drought index was calculated by applying the R Package “SPEI” [Vicente-Serrano et al., 2010].
+The index was calculated for different seasonal windows from 1 to 12 months and at different starting months (from January to December) for the period 1950–2013.
 
@@
 
-Site chronologies were built by averaging all tree BAI measurement of the same site.
  To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated.

### 2018-01-04 analysis/rwi/climate_rwi_relate.Rmd
add spei correlations

@@
 library("treeclim")
+# devtools::install_github("thomasp85/patchwork")
+library("patchwork")
 ```
@@
 eobsN <- eobs %>% filter(year < 2017) %>% filter(loc == 'N') 
-
 prec_N <- eobsN %>% dplyr::select(year, month, prec) 
@@
 eobsS <- eobs %>% filter(year < 2017) %>% filter(loc == 'S') 
-
 prec_S <- eobsS %>% dplyr::select(year, month, prec) 
@@
 
-# Precipitation
-from previous June to
+```{r}
+eobs <- read.csv(file=paste0(di, "data/eobs/eobs_formatted.csv"), header=TRUE, sep=',')
+
+# sanjuan: 37.133, -3.382 = spei_-3.25_37.25.csv 
+# canar: 36.957, -3.4284 = spei_-3.25_36.75.csv
+
+spei_sj <- read.csv(file=paste0(di, "data_raw/spei/single_location/SPEI_37.25_-3.25.csv"), header=TRUE)
+spei_ca <- read.csv(file=paste0(di, "data_raw/spei/single_location/SPEI_36.75_-3.25.csv"), header=TRUE)
+
+# Prepare data 
+spei_N <- spei_sj %>% 
+  mutate(date = as.Date(paste0('01', DATA), format =  '%d%b%Y'),
+         year = lubridate::year(date),
+         month = lubridate::month(date)) %>% 
+  dplyr::select(year, month, spei6 = SPEI_6) %>% 
+  filter(year > 1950) %>% 
+  filter(year < 2017)
+
+spei_S <- spei_ca %>% 
+  mutate(date = as.Date(paste0('01', DATA), format =  '%d%b%Y'),
+         year = lubridate::year(date),
+         month = lubridate::month(date)) %>% 
+  dplyr::select(year, month, spei6 = SPEI_6) %>% 
+  filter(year > 1950) %>% 
+  filter(year < 2017)
+```
+
+## Some useful functions
 
 ```{r}
+# for join dccs 
+join_dccs <- function(dfs){ 
+  out <- c() 
+  for (i in dfs){
+    # Get object dcc
+    aux <- get(i)
+    # Get coeffs 
+    aux_coef <- aux$coef
+    # remove rownames
+    row.names(aux_coef) <- NULL
+
+    # # Add loc variable
+    # name_site <- sub(".*_", "", i)
+    # aux_coef$site <- name_site
+
+    # Add variable sig (asterisk) and site 
+    aux_coef <- aux_coef %>% 
+      mutate(site = sub(".*_", "", i), 
+            sig = case_when(
+              significant == "TRUE" ~ "*", 
+              significant == "FALSE" ~ ""))
+
+    # rbind
+    out <- rbind(out, aux_coef)
+
+  } 
+  return(out)
+}
+
+## PLOTs functions 
+theme_cor <- function(){
+  theme_bw() %+replace%
+    theme(axis.title.x = element_blank(),
+          panel.grid = element_blank(),
+          axis.text.x = element_text(angle = 90))
+}
+# http://joeystanley.com/blog/custom-themes-in-ggplot2
+
+custom_gg <- list(theme_cor(), 
+                  geom_hline(yintercept = 0, colour="gray"),
+                  ylab('Correlation'), 
+                  geom_bar(aes(fill=site), stat="identity", position = "dodge"),
+                  geom_text(aes(label = sig, 
+                                x = id, 
+                                y = (coef + ifelse(coef >= 0, 0.02, -0.02))), 
+                            position = position_dodge(.9), size = 10),
+                  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3")))
+                  # scale_fill_brewer(palette = 'Paired', type='qual'))
+
+# "#A6CEE3" azul clarito
+# "#B2DF8A" verde
+
+```
+
+# Precipitation
+
+* Correlación entre Marzo del mes anterior y hasta Sep del mes en curso
+* Correlación con datos de Pp acumulada
+
+```{r}
+set.seed(3333)
 rprec_SJ <- dcc(chrono = cro_sj, climate = prec_N, 
@@
 
+# Joins dccs 
 rprecs <- c('rprec_SJ', 'rprec_caH', 'rprec_caL')
-rprec_all <- c()
+rprec_all <- join_dccs(rprecs)
 
-for (i in rprecs){
+# Add custom names for aggregation variables
 
-  # Get object dcc
-  aux <- get(i)
-
-  # Get coeffs 
-  aux_coef <- aux$coef
-
-  # remove rownames
-  row.names(aux_coef) <- NULL
-
-  # Change name for aggregate variables 
-  aux_coef <- aux_coef %>% 
-    mutate(month = case_when(
+rprec_all <- rprec_all %>% 
+  mutate(month_name = case_when(
       month == "sep...AUG" ~ "Hydrol", 
-      TRUE ~ as.character(.$month)),
-      sig = case_when(
-      significant == "TRUE" ~ "*", 
-      significant == "FALSE" ~ ""))
-
-  # Add loc variable
-  name_site <- str_sub(i, start=7)
-  aux_coef$site <- name_site
-
-  # rbind
-  rprec_all <- rbind(rprec_all, aux_coef)
-  } 
+      TRUE ~ as.character(.$month)))
 ```
 
-```{r}
+## Plot precipitation 
+```{r rwi_clim_prec}
+p_prec <- rprec_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rprec_all$id, labels = rprec_all$month_name) + 
+  theme(legend.position = c(.1, .8)) +
+  annotate("text", label = "Prec", x = 1, y= -0.4) 
+p_prec
 
-rprec_all
+pdf(paste0(di, '/out/climate_rwi/prec.pdf'), width=9, height = 5)
+p_prec
+dev.off()
+```
 
-rprec_all %>% ggplot(aes(x=id, y=coef, group = site)) +
-  geom_bar(aes(fill=site),
-           stat="identity", position = "dodge") +
-  scale_x_continuous(breaks = rprec_all$id, labels = rprec_all$month) +
-  theme_bw() + 
-  theme(axis.title.x = element_blank()) +
-  geom_text(aes(label = sig), 
-            position = position_dodge(0.9), 
-            vjust = 0.1, 
-            size = 10) +
-  ylab('Correlation')
+# SPEI
+
+* Correlación entre Marzo del mes anterior y hasta Sep del mes en curso
+* Correlación con datos de spei6 hydrol, winter, summer, spring, autumn,
 
 ```{r}
+set.seed(3333)
+rspei_SJ <- dcc(chrono = cro_sj, climate = spei_N, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
 
-eobs_norht_no_prec <- eobs_norht %>% dplyr::select(-prec)
+rspei_caL <- dcc(chrono = cro_caL, climate = spei_N, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
 
-mc <- dcc(chrono = cro_sj, 
-          climate = eobs_norht, 
-          var_names = c("prec", "tmax", "tmean", "tmin"), 
-          method = "correlation", boot = "stationary")
+rspei_caH <- dcc(chrono = cro_caH, climate = spei_N, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
 
-sea_sj <- seascorr(chrono = cro_sj, 
-             climate = eobs_norht, 
-             var_names = c("prec", "tmax", "tmean", "tmin"))
+# Joins dccs 
+rspeis <- c('rspei_SJ', 'rspei_caH', 'rspei_caL')
+rspei_all <- join_dccs(rspeis)
 
-plot(sea_sj)
+# Add custom names for aggregation variables of  and / or SPEI
+customNames <- function(x){ 
+  out <- x %>% 
+    mutate(month_name = case_when(
+      month == "sep...AUG" ~ "Hydrol",
+      month == "dec...FEB" ~ "Winter",
+      month == "MAR...MAY" ~ "Spring",
+      month == "JUN...AUG" ~ "Summer",
+      month == "SEP...NOV" ~ "Autumn",
+      TRUE ~ as.character(.$month)))
 
-dc_sj
+  return(out)
+}
 
-croN <- cro %>% filter(site == 'sj') %>% dplyr::select(-site) %>% 
-  filter(year > 1949)
-rownames(croN) <- croN$year
+rspei_all <- customNames(rspei_all) 
+```
 
-cl <- eobs %>% 
-  filter(loc == 'N') %>% 
-  filter(variable %in% c('rr', 'tg')) %>% 
-  dplyr::select(value, variable, year, month) %>% 
-  spread(variable, value)
+## Plot SPEI
+```{r rwi_clim_spei}
+p_spei <- rspei_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rspei_all$id, labels = rspei_all$month_name) + 
+  theme(legend.position = c(.1, .8)) +
+  annotate("text", label = "SPEI 6", x = 1, y= -0.4) 
+p_spei
 
-cl <- cl %>% rename(prec = rr, temp = tg) %>% 
-  filter(year < 2017)
+pdf(paste0(di, '/out/climate_rwi/spei.pdf'), width=9, height = 5)
+p_spei
+dev.off()
+```
 
-test <- dcc(chrono = croN, climate = cl)
+# Tmean
 
-croN 
-# http://czang.org/treeclim.html
+* Correlación entre Marzo del mes anterior y hasta Sep del mes en curso
+* Correlación con tmean winter, summer, spring, autumn, tmean hydrol
 
+```{r}
+set.seed(3333)
+rtmean_SJ <- dcc(chrono = cro_sj, climate = tmean_N, 
+         method = "correlation",  boot = "std", 
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rtmean_caH <- dcc(chrono = cro_caH, climate = tmean_S, 
+         method = "correlation",  boot = "std", 
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rtmean_caL <- dcc(chrono = cro_caL, climate = tmean_S, 
+         method = "correlation",  boot = "std", 
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+# Joins dccs 
+rtmeans <- c('rtmean_SJ', 'rtmean_caH', 'rtmean_caL')
+rtmean_all <- join_dccs(rtmeans)
+
+# Add custom names for aggregation variables of Temperatures
+customNames <- function(x){ 
+  out <- x %>% 
+    mutate(month_name = case_when(
+      month == "sep...AUG" ~ "Hydrol",
+      month == "dec...FEB" ~ "Winter",
+      month == "MAR...MAY" ~ "Spring",
+      month == "JUN...AUG" ~ "Summer",
+      month == "SEP...NOV" ~ "Autumn",
+      TRUE ~ as.character(.$month)))
+
+  return(out)
+  }
+
+rtmean_all <- customNames(rtmean_all) 
+```
+
+## Plot Tmean 
+```{r rwi_clim_tmean}
+p_tmean <- rtmean_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rtmean_all$id, labels = rtmean_all$month_name) + 
+  theme(legend.position ="top") + 
+  annotate("text", label = "Tmean", x = 1, y= -0.4) + ylim(-.4, .4)
+p_tmean
+
+pdf(paste0(di, '/out/climate_rwi/tmean.pdf'), width=9, height = 5)
+p_tmean
+dev.off()
+```
+
+# Tmin
+
+* Correlación entre Marzo del mes anterior y hasta Sep del mes en curso
+* Correlación con tmin winter, summer, spring, autumn, tmean hydrol
+
+```{r}
+set.seed(3333)
+rtmin_SJ <- dcc(chrono = cro_sj, climate = tmin_N, 
+         method = "correlation",  boot = "std", 
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rtmin_caH <- dcc(chrono = cro_caH, climate = tmin_S, 
+         method = "correlation",  boot = "std", 
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rtmin_caL <- dcc(chrono = cro_caL, climate = tmin_S, 
+         method = "correlation",  boot = "std", 
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+# Join dccs 
+rtmins <- c('rtmin_SJ', 'rtmin_caH', 'rtmin_caL')
+rtmin_all <- join_dccs(rtmins)
+
+# Rename variable 
+rtmin_all <- customNames(rtmin_all) 
+```
+
+## Plot Tmin 
+```{r rwi_cli_tmin}
+p_tmin <- rtmin_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rtmin_all$id, labels = rtmin_all$month_name) + 
+  theme(legend.position = "none") +
+  annotate("text", label = "Tmin", x = 1, y= -0.4) + ylim(-.4, .4)
+p_tmin
+
+pdf(paste0(di, '/out/climate_rwi/tmin.pdf'), width=9, height = 5)
+p_tmin
+dev.off()
+```
+
+# Tmax
+
+* Correlación entre Marzo del mes anterior y hasta Sep del mes en curso
+* Correlación con tmax winter, summer, spring, autumn, hydrol, may-jul, june-july
+
+```{r}
+set.seed(3333)
+rtmax_SJ <- dcc(chrono = cro_sj, climate = tmax_N, 
+         method = "correlation",  boot = "std", 
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11) + .mean(5:7) + .mean(6:7))
+
+rtmax_caH <- dcc(chrono = cro_caH, climate = tmax_S, 
+         method = "correlation",  boot = "std", 
+         .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11) + .mean(5:7) + .mean(6:7))
+
+rtmax_caL <- dcc(chrono = cro_caL, climate = tmax_S, 
+         method = "correlation",  boot = "std", 
+         .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11) + .mean(5:7) + .mean(6:7))
+
+# Join dccs 
+rtmaxs <- c('rtmax_SJ', 'rtmax_caH', 'rtmax_caL')
+rtmax_all <- join_dccs(rtmaxs)
+
+# Rename variable 
+rtmax_all <- customNames(rtmax_all) 
+```
+
+## Plot Tmax 
+```{r rwi_cli_tmax}
+p_tmax <- rtmax_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rtmin_all$id, labels = rtmin_all$month_name) +
+  theme(legend.position = "none") +
+  annotate("text", label = "Tmax", x = 1, y= -0.4) + ylim(-.4, .4)
+p_tmax
+
+pdf(paste0(di, '/out/climate_rwi/tmax.pdf'), width=9, height = 5)
+p_tmax
+dev.off()
+```
+
+```{r rwi_cli_temperatures}
+p_temp <- p_tmean + p_tmax + p_tmin + plot_layout(ncol=1)
+pdf(paste0(di, '/out/climate_rwi/temp.pdf'), width=9, height = 12)
+p_temp
+dev.off()
+```
+
+```{r rwi_cli_prec_spei}
+p_prec_spei <- p_prec + p_spei + plot_layout(ncol=1)
+pdf(paste0(di, '/out/climate_rwi/prec_spei.pdf'), width=9, height = 8)
+p_prec_spei
+dev.off()
+```
+
+```{r rwi_cli_all}
+p <- p_tmean + p_tmax + p_tmin + p_prec + plot_layout(ncol=2)
+pdf(paste0(di, '/out/climate_rwi/all.pdf'), width=13, height = 8)
+p
+dev.off()
+```
+
+# Correlations RUNNING
+
+Para explorar la variabilidad temporal (estabilidad) en la relación entre clima y crecimiento, utilizamos una correlación con window size de 30 años
+
+## Prec 
+```{r moving_precSJ}
+set.seed(3333)
+mprec_SJ <- dcc(chrono = cro_sj, climate = prec_N, selection =.sum(-9:8) + .range(3,12), 
+         dynamic = "moving", win_size = 10, win_offset = 5, sb = FALSE, method = "cor", boot = 'std')
+plot(mprec_SJ)
+```
+
+```{r moving_preccaH}
+set.seed(3333)
+mprec_caH <- dcc(chrono = cro_caH, climate = prec_S, selection =.sum(-9:8) + .range(3,12), 
+         dynamic = "moving", win_size = 10, win_offset = 5, sb = FALSE, method = "cor", boot = 'std')
+plot(mprec_caH)
+```
+
+```{r moving_preccaL}
+set.seed(3333)
+mprec_caL <- dcc(chrono = cro_caL, climate = prec_S, selection =.sum(-9:8) + .range(3,12), 
+         dynamic = "moving", win_size = 10, win_offset = 5, sb = FALSE, method = "cor", boot = 'std')
+plot(mprec_caL)
 ```

### 2018-01-04 man/ms_1_material_methods.Rmd
update methods

@@
 The growth trends were analyzed at different time scales.
+To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976].
+Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008, @Bunn2010] package.
+Residual site chronologies were obtained computing the bi-weight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990].
+The statistical quality of each chronology was checked via expressed population signal [@Wigley1984].
+A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.
+
 The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
@@
 We used measured dbh and raw nondetrened ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
+We used the following equation:
+
+$$\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$$
+where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
 We calculated a mean BAI serie for each individual tree.
-(:red_circle: equation BAI).
-...
-Site chronologies were built by averaging all tree BAI measurement of the same site.
-
-To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976](:red:circle: (Cook and Kairiukstis 1990)..
-We standardized and detrended the tree-ring width series using the `dplR` [@Bunn2008, @Bunn2010] package.
-Residual site chronologies were obtained computing the bi-weight robust mean of all prewhitened growth indices for the trees of the same site.
-
-Statistical descriptive parameters (Fritts, 2001), including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology considering the common 1950–2013 period.
-
-* incluir tree clim [@Zang2015],  EPS
-
-... The diameters at the breast height (DBH) measured in the field were used to convert each individual ring-width series to basal area increment (BAI) using dplR (Bunn 2008).
-
-The chronologies were computed as bi-weight robust mean to reduce the bias caused by extreme values and an autoregressive model was applied to remove the autocorrelation of the series (Cook and Kairiukstis 1990).
-The statistical quality of each chronology was checked via expressed population signal (EPS; Wigley et al.
-1984).
-A threshold value of EPS >0.85 was used to determine the cutoff year of the time span that could be considered reliable.
-The common period 1950–2012 was used for further analyses in order to avoid the period of intensive logging in TJN (see also Fig.
-2)....
+Site BAI chronologies were obtained by averaging individual tree BAI time series.
 
@@
 We selected grid cells covering each sampled sites.
+We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010, @Begueria2014].
+SPEI values for the period 1961-2014 were obtained with a spatial resolution of 1.1 km from the Drought indices dataset for Spain database (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017].
+We select a temporal scale of 6 months.
 
-treeclim aqui
+The relationships between residual site chronologies and the climatic variables were assessed by a bootstrapped Pearson's correlation estimate using the `treeclim` package [@Zang2015].
+The bootstrapped confidence intervals were used to estimate the significance (p < 0.05) of the correlation coefficients.
 
-The SPEI drought index was calculated by applying the R Package “SPEI” [Vicente-Serrano et al., 2010].
-The index was calculated for different seasonal windows from 1 to 12 months and at different starting months (from January to December) for the period 1950–2013.
+ ```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+* Para obtener los datos de SPEI, hemos ido bajado los datos desde http://sac.csic.es/spei/map/maps.html 
+* sanjuan: 37.133, -3.382 = spei_-3.25_37.25.csv 
+* canar: 36.957, -3.4284 = spei_-3.25_36.75.csv
 
-```{r}
+OJO ver http://monitordesequia.csic.es/
+")
+```
+
+Statistical descriptive parameters (Fritts, 2001), including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology considering the common 1950–2013 period.
+
+The chronologies were computed as bi-weight robust mean to reduce the bias caused by extreme values and an autoregressive model was applied to remove the autocorrelation of the series (Cook and Kairiukstis 1990).
+
+```{r echo=FALSE, results='asis', eval=show_text}
 cat("
@@
 Thus, we obtained mean residual chronologies for each species of prewhitened growth indices.
-    ")
 
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
  * BAI represents a more accurate indicator of growth than ring-width, since it reduces the variation caused by adding volume to a circular stem 
@@
 
-* incluir citas a Frits; Cooks ...
-
-* BAI 
-* site cronologies
 * GC Nowacki??
@@
 
+All analyses were carried out in R software (ver.
+2.11.1), using library dplR to standardize data, library nlme to fit the mixed models and vegan to run the PCA.
+All tests were at α = 0.05, except when indicated.
+
 ## References

### 2018-01-04 analysis/rwi/climate_rwi_relate.Rmd
update tree-climate

@@
 library("patchwork")
+library("readr")
 ```
@@
 
+# Spei data: 0.5 º - 1950
+
 ```{r}
-eobs <- read.csv(file=paste0(di, "data/eobs/eobs_formatted.csv"), header=TRUE, sep=',')
-
 # sanjuan: 37.133, -3.382 = spei_-3.25_37.25.csv 
@@
   filter(year < 2017)
+```
+
+# Spei data: 1.1 km º - 1961 (4 for months )
+```{r}
+# http://monitordesequia.csic.es/ 
+# sanjuan: 37.13, -3.37 = DATA_37.13_-3.37.csv
+# canar Low: 36.956, -3.424 = DATA_36.96_-3.42.csv
+# canar High: 36.966, -3.421 = DATA_36.97_-3.42.csv
+
+spei_sjD <- read.csv(file=paste0(di, "data_raw/spei/monitor_sequia/DATA_37.13_-3.37.csv"), header=TRUE)
+spei_caHD <- read.csv(file=paste0(di,  "data_raw/spei/monitor_sequia/DATA_36.97_-3.42.csv"), header=TRUE)  
+spei_caLD <- read.csv(file=paste0(di,  "data_raw/spei/monitor_sequia/DATA_36.96_-3.42.csv"), header=TRUE)
+
+# Prepare data 
+sequia_sj <- spei_sjD %>% 
+  mutate(date = parse_date(DATA, format='%d-%B-%Y', locale = locale('es')),
+         year = lubridate::year(date),
+         month = lubridate::month(date)) %>% 
+  dplyr::select(year, month, spei6 = spei_6) %>% 
+  filter(year > 1961) %>% 
+  group_by(year, month) %>% 
+  summarise(spei6 = mean(spei6)) %>% as.data.frame()
+
+sequia_caH <- spei_caHD %>% 
+  mutate(date = parse_date(DATA, format='%d-%B-%Y', locale = locale('es')),
+         year = lubridate::year(date),
+         month = lubridate::month(date)) %>% 
+  dplyr::select(year, month, spei6 = spei_6) %>% 
+  filter(year > 1961) %>% 
+  group_by(year, month) %>% 
+  summarise(spei6 = mean(spei6)) %>% as.data.frame()
+
+sequia_caL <- spei_caLD %>% 
+  mutate(date = parse_date(DATA, format='%d-%B-%Y', locale = locale('es')),
+         year = lubridate::year(date),
+         month = lubridate::month(date)) %>% 
+  dplyr::select(year, month, spei6 = spei_6) %>% 
+  filter(year > 1961) %>% 
+  group_by(year, month) %>% 
+  summarise(spei6 = mean(spei6)) %>% as.data.frame()
+
 ```
@@
 
+# SPEI High-resol
+
+* Correlación entre Marzo del mes anterior y hasta Sep del mes en curso
+* Correlación con datos de spei6 hydrol, winter, summer, spring, autumn,
+
+```{r}
+set.seed(3333)
+rsequia_SJ <- dcc(chrono = cro_sj, climate = sequia_sj, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rsequia_caL <- dcc(chrono = cro_caL, climate = sequia_caL, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rsequia_caH <- dcc(chrono = cro_caH, climate = sequia_caH, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+# Joins dccs 
+rsequias <- c('rsequia_SJ', 'rsequia_caH', 'rsequia_caL')
+rsequia_all <- join_dccs(rsequias)
+
+rsequia_all <- customNames(rsequia_all) 
+```
+
+## Plot SPEI HR
+```{r rwi_clim_sequia}
+p_sequia <- rsequia_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rsequia_all$id, labels = rsequia_all$month_name) + 
+  theme(legend.position = c(.1, .8)) +
+  annotate("text", label = "SPEI 6 - HR", x = 1, y= -0.4) 
+p_sequia
+
+pdf(paste0(di, '/out/climate_rwi/sequia.pdf'), width=9, height = 5)
+p_sequia
+dev.off()
+```
+
 # Tmean
@@
 
-```{r rwi_cli_prec_spei}
-p_prec_spei <- p_prec + p_spei + plot_layout(ncol=1)
+```{r rwi_cli_prec_spei_sequia}
+p_prec_spei <- p_prec + p_spei + p_sequia + plot_layout(ncol=1)
 pdf(paste0(di, '/out/climate_rwi/prec_spei.pdf'), width=9, height = 8)
 p_prec_spei
+dev.off()
+```
+
+```{r rwi_cli_prec_sequia}
+p_prec_sequia <- p_prec + p_sequia + plot_layout(ncol=1)
+pdf(paste0(di, '/out/climate_rwi/prec_sequia.pdf'), width=9, height = 8)
+p_prec_sequia
 dev.off()

### 2018-01-10 analysis/disturbance/analysis_disturbance_chronos.Rmd
update disturbances

---
title: "disturbance chronologies"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Feb"
output:
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("stringr")
library("dplR")
library("knitr")
# library("detrendeR")
library("pander")
library('boot')
library('gtable')
library('grid')
library('patchwork')
```

# Disturbance chronologies

* Vamos a calcular las disturbance chronologies siguiendo la aproximación de [@Nowacki1997](varios ejemplos se pueden ver en [@GeaIzquierdo2014, @Dorado2017])

* Utilizaremos tres chronologias: SJ, CAH y CAL (ver [./analysis/chronos/analysis_chronologies.md]('./analysis/chronos/analysis_chronologies.md))

## Prepare data

* Leer datos `rwl` de SJ y CA
* Leer datos de diametros de los focal tree

```{r, echo=FALSE}
machine <- 'ajpelu'
#machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# sj 
sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")

# canar 
ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")

# Read diameters data
compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
```

```{r}
source(paste0(di, 'script/R/rw_byTree.R'))
source(paste0(di, 'script/R/bai_piovesan.R'))
source(paste0(di, 'script/R/computeGC.R'))
```

* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low

* :red_circle: ojo he cambiado los nombres de las series en SJ y en CA

```{r}
# Replace SNA by SJ and SNB by CA
names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")

# Remove g in name of some cores of CA. 
names(ca) <- stringr::str_replace(names(ca), "g", "")
```

```{r}
# Create subset to compare between sites 
caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]

```

* Lectura y preparación de datos de diámetro

```{r} 
# Prepare Diameter data

# Compute diameter (mm)
compete <- compete %>% 
  mutate(dn_mm = (perim_mm / pi))

# Change name focal according to loc
compete <- compete %>% 
  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))

# Get only focal trees, and only selected variables 
ft <- compete %>% 
  filter(sp=='Focal') %>% 
  filter(id_focal!='Fresno') %>% 
  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm)

# Set levels of eleveation 
ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
            paste0('CA', 26:30))
ca_highcode <- paste0('CA', 11:25)

ft <- ft %>% 
  mutate(site = as.factor(
    ifelse(id_focalLoc %in% ca_lowcode, 'CAL', 
           ifelse(id_focalLoc %in% ca_highcode, 'CAH', 'SJ'))))
```

# Summary dendrochronology statistics

* Por cada site (CAH, CAL, SJ) calculamos los siguientes datos: 
    * age: mean + sd (average of all series)
    * year: min, max of all series
    * year range
    * autocorrelation order 1 (ar1): mean + sd 
    * seires intercorrelation (icor): mean + sd

## By site 
```{r}
objects_rwl <- c('caL','caH','sj')

out <- c()

for (i in objects_rwl){

  aux <- get(i)
  # Stats of singles RWL
  aux_stats <- rwl.stats(aux)
  # Add site 
  aux_stats$site <- as.factor(i)

  # Compute interseries correlation 
  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val

  out <- rbind(out, aux_stats)
  }

rwl_by_site <- out %>% group_by(site) %>% 
  summarize(age_mean = round(mean(year),2), 
            age_sd = round(sd(year), 2),
            year_min = min(first),
            year_max = max(last), 
            year_range = year_max - year_min + 1, 
            ar1_mean = round(mean(ar1), 3),
            ar1_sd = round(sd(ar1), 3),
            icor_mean = round(mean(interser_cor_val), 3),
            icor_sd = round(sd(interser_cor_val), 3)) %>% 
  as.data.frame()

rwl_by_site %>% 
  write.csv(file=paste(di, "data/proto_tables/site_3_dendro.csv", sep=""), row.names = FALSE)

pander(rwl_by_site, caption='Dendrochronological summary by sites (3)') 
```

# Aggregate RW by tree

* Agregar valores medios de RW por site (obtenemos sj_tree / caL_tree, caH_tree) 
* ver fun rw_byTree o utilizar treeMean (dplR)

```{r}
# Remember snc = structure of core name SJ0101 (site | tree | core)
sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')

# Si decido usar treeMean debería ser (solo un ejemplo)
# ids_sj <- read.ids(sj, stc = c(2, 2, 2))
# sj_meantree <- treeMean(sj, ids = ids_sj,  na.rm=TRUE)
```

* Crear diferentes dataset de diametro por sitio 
```{r}
diam <- ft %>%
  mutate(diameter = dn_mm, 
         id = id_focalLoc) %>%
  dplyr::select(id, diameter, site) %>% 
  split(.$site)

d_caH <- diam$CAH[,c('id','diameter')]
d_caL <- diam$CAL[,c('id','diameter')]
d_sj <- diam$SJ[,c('id','diameter')]
```

# Disturbance chronologies

* calcular chronologies con media y mediana.

```{r}
gc_caL <- computeGC(caL_tree, ws=10, prefijo = 'CA')
gc_caH <- computeGC(caH_tree, ws=10, prefijo = 'CA')
gc_sj <- computeGC(sj_tree, ws=10, prefijo = 'SJ')
```

*  Create a dataset with all values

```{r}
# dataframe for plot 
gc_todos <- c()

datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')

for (d in datasetes){ 
  # get df
  df <- get(d)

  aux_df <- df %>%
    group_by(year) %>% 
    dplyr::summarise(ntrees = length(ntrees),
              gc_m = mean(gc_mean, na.rm=TRUE),
              gc_m_sd = sd(gc_mean, na.rm=TRUE),
              gc_m_se = sd(gc_mean) / sqrt(n()), 
              gc_md = mean(gc_median, na.rm=TRUE),
              gc_md_sd = sd(gc_median, na.rm=TRUE),
              gc_md_se = sd(gc_median) / sqrt(n())) %>%  
    mutate(site=paste(d))

  gc_todos <- rbind(gc_todos, aux_df)
}

gc_todos <- gc_todos %>% 
  mutate(site_sorted = case_when(
    site == "gc_sj" ~ "0_sj", 
    site == "gc_caH" ~ "1_caH",
    site == "gc_caL" ~ "2_caL"))

gc_todos %>%   
  write.csv(file=paste(di, "data/disturbance/disturbance_gc.csv", sep=""), row.names = FALSE)
```

# Plots 
### Plots: auxiliar fx 
```{r}
# plot themes  
label_site <- c("0_sj" = "Northern (SJ)",
                "1_caH" = "Southern High (CaH)",
                "2_caL" = "Southern Low (CaL)")

theme_gc <- function(){
  theme_bw() %+replace% 
    theme(panel.grid.minor = element_blank(),
          strip.background = element_rect(colour='black', fill='white'))
}

custom_gc <- list(
  theme_gc(), 
  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue'),
  geom_hline(yintercept=c(50, -50), linetype=2, colour='red'), 
  facet_wrap(~site_sorted, nrow = 3, scales = 'free_y',
             labeller = as_labeller(label_site)),
  geom_bar(stat='identity', fill='gray', colour='gray')
  )
```

## Mean 
```{r meanGC}
p_gc_mean <- gc_todos %>% 
  ggplot(aes(x=year, y=gc_m)) +
  custom_gc + ylab('% GC (mean)') + 
  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour = ' gray')

p_gc_mean

pdf(paste0(di, 'out/disturbance/gc_mean.pdf'), width=9, height = 9)
p_gc_mean 
dev.off()

```

## Median
```{r mediaGC}
p_gc_median <- gc_todos %>% 
  ggplot(aes(x=year, y=gc_md)) +
  custom_gc + ylab('% GC (median)') + 
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour = ' gray')

p_gc_median

pdf(paste0(di, 'out/disturbance/gc_median.pdf'), width=9, height = 9)
p_gc_median 
dev.off()
```

### Plot (correcting the scale) (mean)
```{r meanGC_corregido}
# Create a transformation scale 
# https://stackoverflow.com/questions/37446064/i-need-ggplot-scale-x-log10-to-give-me-both-negative-and-positive-numbers-as-o

pos_neg_log <- scales::trans_new('signed_log', 
                                 transform = function(x) sign(x)*log(abs(x)),
                                 inverse = function(x) sign(x)*exp(abs(x)))

# solution for arrange
# http://felixfan.github.io/stacking-plots-same-x/

custom_corrected_gc <- list(
  geom_bar(stat='identity', fill='gray', colour='gray'),
  theme_bw(), 
  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4), 
  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4),
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black',fill='white'))
  )

custom_ylab <- '% GC (mean)'

a <- gc_todos %>% 
  filter(site != 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_m)) + 
  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), 
                colour='gray', width=.3, size=.2) +
  # scale_y_continuous(trans = pos_neg_log) +
  facet_wrap(~site_sorted, nrow = 2,
             labeller=as_labeller(c("1_caH" = "Southern High (CaH)", "2_caL" = "Southern Low (CaL)"))) + 
    ylab(custom_ylab) + custom_corrected_gc

b <- gc_todos %>% 
  filter(site == 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_m)) + 
  xlim(layer_scales(a)$x$range$range[1],layer_scales(a)$x$range$range[2]) +
  geom_hline(yintercept = 1, colour = 'black', lwd = 0.3) + 
  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray', width=.3, size=.2) +
  facet_wrap(~site_sorted, nrow = 2,
             labeller=as_labeller(c("0_sj" = "Northern (SJ)"))) + 
  custom_corrected_gc + ylab('') + 
  scale_y_continuous(trans = pos_neg_log, 
                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

# grid.newpage()
# mean_corregido <- grid.draw(rbind(ggplotGrob(b), ggplotGrob(a), size='last'))

plot_means_corregido <- b + a + plot_layout(ncol=1, heights = c(1, 2))

pdf(paste0(di, 'out/disturbance/gc_mean_corregido.pdf'), width=9, height = 9)
plot_means_corregido 
dev.off()
```

```{r medianGC_corregido}
custom_ylab <- '% GC (median)'

c <- gc_todos %>% 
  filter(site != 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_md)) + 
  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
  # scale_y_continuous(trans = pos_neg_log) +
  facet_wrap(~site_sorted, nrow = 2,
             labeller=as_labeller(c("1_caH" = "Southern High (CaH)", "2_caL" = "Southern Low (CaL)"))) + 
    ylab(custom_ylab) + custom_corrected_gc

d <- gc_todos %>% 
  filter(site == 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_md)) + 
  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
  geom_hline(yintercept = 1, colour = 'black', lwd = 0.3) + 
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
  facet_wrap(~site_sorted, nrow = 2,
             labeller=as_labeller(c("0_sj" = "Northern (SJ)"))) + 
  custom_corrected_gc + ylab('') + 
  scale_y_continuous(trans = pos_neg_log, 
                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
plot_medians_corregido

pdf(paste0(di, 'out/disturbance/gc_median_corregido.pdf'), width=9, height = 9)
plot_medians_corregido
dev.off()
```

# Number trees with %GC > 25 %

## Number of trees with %GC > 25% (mean)
```{r}
computeGCtrees <- function(gcdf, treshold){

  th <- as.numeric(treshold)

  # Get total tree by year 
  n_tree <- gcdf %>% group_by(year) %>% dplyr::summarise(total_tree = length(ntrees))

  # Get n tree with GC positve upper treshold 
  n_pos <- gcdf %>% dplyr::filter(gc_mean > th) %>% group_by(year) %>% dplyr::summarise(n_pos = length(ntrees))

  # Get n tree with GC negative lower treshold 
  n_neg <- gcdf %>% dplyr::filter(gc_mean < (-1*th)) %>% group_by(year) %>% dplyr::summarise(n_neg = length(ntrees))

  # Join and compute percentage 
  outgc <- n_tree %>% 
    left_join(n_pos, by='year') %>% 
    left_join(n_neg, by='year') %>% 
    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
         per_neg = round((n_neg / total_tree) * 100, 2))

  return(outgc)
}
```

```{r}
gc_trees<- c()

datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')

for (d in datasetes){ 
  # get df
  df <- get(d)

  au <- df %>% computeGCtrees(treshold = 25) %>% data.frame() %>% mutate(site = paste(d))
  gc_trees <- rbind(gc_trees, au)
}

gc_trees  <- gc_trees  %>% 
  mutate(site_sorted = case_when(
    site == "gc_sj" ~ "0_sj", 
    site == "gc_caH" ~ "1_caH",
    site == "gc_caL" ~ "2_caL"))

gc_trees %>%   
  write.csv(file=paste(di, "data/disturbance/disturbance_trees_gc_treshold.csv", sep=""), row.names = FALSE)

```

## Plots (mean)

```{r trees_GC25_mean}
# secondary axis
# https://whatalnk.github.io/r-tips/ggplot2-secondary-y-axis.nb.html

# Plot 
pgc25_mean <- gc_trees %>% ggplot() +
  geom_bar(aes(x=year, y=per_pos), fill='blue', colour='blue', stat = "identity", position = "dodge") + 
  geom_bar(aes(x=year, y=(-1)*per_neg), fill='red', colour='red', stat = "identity", position = "dodge") +
  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) +
  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
  geom_line(aes(x=year, y=total_tree), colour = 'gray') + 
  theme_bw() + ylab('% trees GC > 25 % (mean)') +
  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white'))

pgc25_mean

pdf(paste0(di, 'out/disturbance/gc_tree25_mean.pdf'), width=9, height = 9)
pgc25_mean
dev.off()
```

## Number of trees with %GC > 25% (median)
```{r}
computeGCtreesMedian <- function(gcdf, treshold){

  th <- as.numeric(treshold)

  # Get total tree by year 
  n_tree <- gcdf %>% group_by(year) %>% dplyr::summarise(total_tree = length(ntrees))

  # Get n tree with GC positve upper treshold 
  n_pos <- gcdf %>% dplyr::filter(gc_median > th) %>% group_by(year) %>% dplyr::summarise(n_pos = length(ntrees))

  # Get n tree with GC negative lower treshold 
  n_neg <- gcdf %>% dplyr::filter(gc_median < (-1*th)) %>% group_by(year) %>% dplyr::summarise(n_neg = length(ntrees))

  # Join and compute percentage 
  outgc <- n_tree %>% 
    left_join(n_pos, by='year') %>% 
    left_join(n_neg, by='year') %>% 
    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
         per_neg = round((n_neg / total_tree) * 100, 2))

  return(outgc)
}

gc_treesMed <- c()

datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')

for (d in datasetes){ 
  # get df
  df <- get(d)

  au <- df %>% computeGCtreesMedian(treshold = 25) %>% data.frame() %>% mutate(site = paste(d))
  gc_treesMed <- rbind(gc_treesMed, au)
}

gc_treesMed  <- gc_treesMed  %>% 
  mutate(site_sorted = case_when(
    site == "gc_sj" ~ "0_sj", 
    site == "gc_caH" ~ "1_caH",
    site == "gc_caL" ~ "2_caL"))

gc_treesMed %>%   
  write.csv(file=paste(di, "data/disturbance/disturbance_trees_gc_treshold_median.csv", sep=""), row.names = FALSE)
```

```{r trees_GC25_median}
# secondary axis
# https://whatalnk.github.io/r-tips/ggplot2-secondary-y-axis.nb.html

# Plot 
pgc25_median <- gc_treesMed %>% ggplot() +
  geom_bar(aes(x=year, y=per_pos), fill='blue', colour='blue', stat = "identity", position = "dodge") + 
  geom_bar(aes(x=year, y=(-1)*per_neg), fill='red', colour='red', stat = "identity", position = "dodge") +
  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
  geom_line(aes(x=year, y=total_tree), colour = 'gray') + 
  theme_bw() + ylab('% trees GC > 25 % (median)') +
  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white'))

pgc25_median

pdf(paste0(di, 'out/disturbance/gc_tree25_median.pdf'), width=9, height = 9)
pgc25_median
dev.off()  
```

# TRADER approach 
Vamos a utilizar la aproximación del pkg `trader` [Altman2014]

## Nowacki & Abrams 1997
Mismo análisis de PG [@Nowacki1997], pero con trader.

```{r}
## PG approach 
library('TRADER')

temp <- getwd()

setwd(paste0(di, 'analysis/disturbance/trader_chronos/caL/'))
growthAveragingALL(caL_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
                   prefix = 'CA', gfun = median, length = 5, storedev = pdf)

setwd(paste0(di, 'analysis/disturbance/trader_chronos/caH/'))
growthAveragingALL(caH_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
                   prefix = 'CA', gfun = median, length = 5, storedev = pdf)

setwd(paste0(di, 'analysis/disturbance/trader_chronos/sj/'))
growthAveragingALL(sj_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
                   prefix = 'SJ', gfun = median, length = 5, storedev = pdf) 
```

## Black & Abrams 2003
Análisis de boundary-line [@Black2003]
```{r}
setwd(paste0(di, 'analysis/disturbance/trader_chronos/boundary_caL/'))

boundaryLineALL(caL_tree, releases = NULL, m1 = 10, m2 = 10,
                boundary = NULL, buffer = 10,
                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
                prefix = 'CA', drawing = TRUE, gfun = median, length = 5, 
                notop = 10, notop2 = 10, storedev = pdf)

setwd(paste0(di, 'analysis/disturbance/trader_chronos/boundary_caH/'))
boundaryLineALL(caH_tree, releases = NULL, m1 = 10, m2 = 10,
                boundary = NULL, buffer = 10,
                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
                prefix = 'CA', drawing = TRUE, gfun = median, length = 5, 
                notop = 10, notop2 = 10, storedev = pdf)

setwd(paste0(di, 'analysis/disturbance/trader_chronos/boundary_sj/'))
boundaryLineALL(sj_tree, releases = NULL, m1 = 10, m2 = 10,
                boundary = NULL, buffer = 10,
                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
                prefix = 'SJ', drawing = TRUE, gfun = median, length = 5, 
                notop = 10, notop2 = 10, storedev = pdf)

```

## Fraver & White 2005
Análisis de absolute increase [@Fraver2005]
```{r}
setwd(paste0(di, 'analysis/disturbance/trader_chronos/absoluteI_caL/'))
absoluteIncreaseALL(caL_tree, abs = NULL, m1 = 10, m2 = 10,
                buffer = 10, prefix = 'CA', drawing = TRUE, gfun = median,
                length = 5, storedev = pdf)

setwd(paste0(di, 'analysis/disturbance/trader_chronos/absoluteI_caH/'))
absoluteIncreaseALL(caH_tree, abs = NULL, m1 = 10, m2 = 10,
                buffer = 10, prefix = 'CA', drawing = TRUE, gfun = median,
                length = 5, storedev = pdf)

setwd(paste0(di, 'analysis/disturbance/trader_chronos/absoluteI_sj/'))
absoluteIncreaseALL(sj_tree, abs = NULL, m1 = 10, m2 = 10,
                buffer = 10, prefix = 'SJ', drawing = TRUE, gfun = median,
                length = 5, storedev = pdf)

setwd(temp)
```

```{r}

# SJ 
pg <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'PG')
bl <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/boundary_sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'BL')
ai <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/absoluteI_sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'AI')

dist_sj <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'sj')

# caL 
pg <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'PG')
bl <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/boundary_caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'BL')
ai <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/absoluteI_caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'AI')

dist_caL <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'caL')

# caH 
pg <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'PG')
bl <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/boundary_caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'BL')
ai <- read.csv(paste0(di, 'analysis/disturbance/trader_chronos/absoluteI_caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'AI')

dist_caH <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'caH')

disturbances <- dist_caH %>% bind_rows(dist_caL, dist_sj) %>% 
    mutate(site_sorted = case_when(
    site == "sj" ~ "0_sj", 
    site == "caH" ~ "1_caH",
    site == "caL" ~ "2_caL"))

```

```{r plot_other_methods}
trader_compare <- disturbances %>% ggplot(aes(x=AllReleasesYear, y=AllReleasesFreq)) + 
  geom_point(aes(shape=metodo, size=AllReleasesFreq)) + scale_shape(solid = FALSE) +
  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) +
  theme_bw()

trader_compare

pdf(paste0(di, 'out/disturbance/trader_compare.pdf'), width=10, height = 9)
trader_compare
dev.off()
```

```{r plot_other_methods_separate}
trader_compare_separe <- disturbances %>% ggplot(aes(x=AllReleasesYear, y=AllReleasesFreq, fill=metodo)) + 
  geom_bar(aes(fill=metodo), stat="identity", position = "dodge") +
  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + theme_bw()

trader_compare_separe

pdf(paste0(di, 'out/disturbance/trader_compare_separados.pdf'), width=10, height = 9)
trader_compare_separe
dev.off()
```

# References

### 2018-01-10 analysis/disturbance/explore_disturbance_chronos.Rmd
update disturbances

---
title: "disturbance chronologies"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 April"
output:  
  md_document:
    variant: markdown_github
bibliography: references.bib 
csl: ecology.csl
---

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("dplR")
library("knitr")
library("rowr")
library("zoo")
```

# Read data

```{r, echo=FALSE}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# sj 
sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")

# canar 
ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
```

# Disturbance chronologies 
* Vamos a computar las disturbances chronologies 
...

```{r}
# Get summary ring-width series
sj_summ <- summary(sj)

# Get names cores with last year different to 2016 
id_cores_no_bark <- sj_summ %>% 
  filter(last != 2016) %>% select(series) %>% mutate(series = factor(series))

id_cores_no_bark <- unname(unlist(id_cores_no_bark))

# Subbet datasets 
sj_cor <- sj[which(! colnames(sj) %in% (id_cores_no_bark))]
sj_sin <- sj[which(colnames(sj) %in% (id_cores_no_bark))]
```

```{r}
rws <- data.frame(year = as.numeric(as.character(rownames(sj_cor))),
                  rw =sj_cor[,'SNA1102'])

# Some notes about lag, windows, etc. 
# https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html

rws <- rws %>% 
  mutate(g2= rollapply(rw, width = 10, sum, align = 'left', partial=TRUE), 
         g1_nolag = rollapply(rw, width = 11, sum, align = 'right', partial=TRUE),
         rw_lag = lag(rw, 1),
         g1_previous10 = rollapply(rw_lag, width = 10, sum, align = 'right', partial=TRUE, na.rm=T),
         g1_mio = g1_nolag - rw,
         gc = ((g2 - g1_previous10) / g1_previous10)*100,
         gc_mio = ((g2 - g1_mio) / g1_mio)*100)

rws %>% 
  ggplot(aes(x=year, y=gc)) + geom_bar(stat='identity') +
  geom_hline(yintercept = c(-25,25), linetype = "dashed", colour="red") + 
  theme_bw()
```

### 2018-01-10 man/ms_1_material_methods.Rmd
update disturbances

@@
 
+una frase conectora 
+We also explored the climate-growth relationships and the disturbance  ...
+
 ### Climate and growth
@@
 We selected grid cells covering each sampled sites.
+Data were downloaded and preprocess using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018, @Frias2018].
 We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010, @Begueria2014].
@@
 The bootstrapped confidence intervals were used to estimate the significance (p < 0.05) of the correlation coefficients.
+
+### Disturbance analyses
+
+Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (releases) or decreases (supressions) in radial growth [@Nowacki1997].
+...
+This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
+By using a temporal window of reasonable length, this method filters out the response to short-term changes in temperature and precipitation [@Nowacki1997, @Fraver2005].
+
+Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:
+
+$$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
+where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median.
+We used medians since they are mores robust estimator of central tendency than means @Rubino2004 @Camarero2011].
+Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
+A
+
+To separate growth periods produced by disturbance events from those by climate, we considered a treshold of 25 % of GC for consi
+
+A minimum $GC$ of 25% was established to depict canopy disturbances. Furthermore, more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
+
+This approach filters out
+
+We used the percentage Growth-change filter (GC).
+
+, calculated annually using 10-year windows
+
+Disturbance chronologies were built using tree-ring width to identify historical abrupt positive (releases) or negative (suppressions) changes in growth in each species (Nowacki and Abrams, 1997).
+This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
 

### 2018-01-10 man/ms_2_results.Rmd
update
update disturbances

@@
 
-```{r setup, echo=FALSE}
+```{r, echo=FALSE}
 show_text <- FALSE
@@
 
+# Climate and tree
+
+Tree-growth exhibited a signigicant response to water availability more than to the temperatures.
+Cumulative precipitation of the hydrological year (September of the previous year to August of the growth year) influenced positively the tree-growth for the both northern and southern populations (FIG ).
+Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
+Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth, specially for northern population.
+Spring Maximum temperature was the most significant limitation of the tree-growth for the southern population while maximum temperatures of the previous August did for the northern one.
+Conversely, mean, minimum and maximum temperatures of current September positively influenced the tree-growth in the northern population.
+
 notas para discusión

### 2018-01-10 man/ms_2_results.Rmd
update results

@@
 
+# Disturbance
+
+En las poblaciones del norte osbervamos dos eventos de PGC por encima del 25 e incluso el 50 %.
+El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
+Estos periodos se alternan con periodos de NGC: previo 1940, 1960 - 1991, y quizá aunque no lo vemos en al actualidad (??
+duda.
+Esto es especular un poco).
+Esto responde creo que a:
+
+(i) actividad minera y tratamientos forestales.
+La actividad minera en la zona fue intensa (aunque a veces intermitente, entre 1890 y 1957 (cuando se clausuró); y previamente desde 1858 tras los trabajos de Amalio Maestre).
+Hay costancia de la existencia de hornos de fundición de mineral en alguna de las minas (La Probadora).
+(ii) actividades forestales:
+* Se conoce la existencia de un proyecto redactado en 1943 de repoblación forestal de la cabecera del río Genil (he pedido los mapas).
+* Ver actividades forestales (bd proyectos de gestión)
+
+En el trabajo de [@GeaIzquierdo2014] se observa una liberación posterior a 1950 (QPUY9) (entre 1950 y 1960 aprox.) y luego una supresión hasta mediados de 1990.
+Esto coincide con lo que hemos observado en nuestra chrono.
+
+Respecto a las poblaciones del sur, destaca en la caH una primera liberación en torno a 1829-1830 y 1837-1840; y luego predominan los eventos de supresión que se alternan con algunos de liberación (muy débiles) hastas 1950.
+Posteriormente se alternan liberaciones-supresiones débiles.
+Destaca una liberación en torno a 2000-2003 (relación con clima???
+posterior a sequía 1999???)
+
+En caL, observamos un periodod de limitación hasta 1885 aprox.
+Luego se alternan al igual que antes periodos de liberación con otos de supresión con señales más fuertes que en caL, y al igual que antes destaca con mas intensidad si cabe, una liberación entre 2000 y 2003 (antrópico o climática ??)
+
+Duda: ocurre en los dos sitios caL y caH, entonces, es aumento de crecimiento debido a actuación forestal? o debido a clima??
+
+[@GeaIzquierdo2014] encontró un pico de disturbance en torno a 1860 en QUPY10 (cañar) sugiriendo que la baja densidad en esos sitios fue presumible el resultado de un thinning (aclareo).
+El patrón de supresión que observa para este sitio va en la linea de lo que nosotros encontramos (supresiones que alcanzan a veces el 25 % y otras no)
+
+En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles.
+En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles
+
 notas para discusión

### 2018-01-16 analysis/rwi/climate_rwi_relate.Rmd
UPDATE

@@
 * Correlación entre Marzo del mes anterior y hasta Sep del mes en curso
-* Correlación con tmax winter, summer, spring, autumn, hydrol, may-jul, june-july
+* Correlación con tmax winter, summer, spring, autumn, hydrol
 
@@
          method = "correlation",  boot = "std", 
-        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11) + .mean(5:7) + .mean(6:7))
+        .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
 
@@
          method = "correlation",  boot = "std", 
-         .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11) + .mean(5:7) + .mean(6:7))
+         .range(-3:9) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
 

### 2018-01-16 man/ms_0_intro.Rmd
updating
UPDATE

@@
 
-* To quantify how two extreme drought events influenced the greennes and radial growth of *Q. pyreancia* forests in their rear edge,
+* To quantify how two extreme drought events influenced the greenness and radial growth of *Q. pyreancia* forests in their rear edge,
 
-* to analyze the resilience of these forests to sucessive extrme drought events,
+* to analyze the resilience of these forests to successive extreme drought events,
 
@@
 
+Do northern and southern populations of Pyrenean oak forests differ in their resistance, resilience or recovery to extreme drought events?
+
 ## References

### 2018-01-16 man/ms_1_material_methods.Rmd
UPDATE

@@
 bibliography: ../references.bib
-templat
-csl: ../geb.csl
+csl: ../ecology.csl
 ---
@@
 ### Tree sampling 
-Samplig was carried during autumn of 2016.
+Sampling was carried during autumn of 2016.
 Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1).
-Two elevations were sampled for the southern site (CA-Low and CA-High).
+For the southern site two elevations were sampled: CA-Low and CA-High.
 All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..).
 In each site between 15 and 20 dominant trees were randomly selected.
-Two cores of 5 mm of diameter were taken from each at breast heith (1.3 m) using an increment borer.
+Two cores of 5 mm of diameter were taken from each at breast heigth (1.3 m) using an increment borer.
 Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively.
@@
 Individual ring series were visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
-Validation of the croos-dating was done using COFECHA software [@Holmes1983].
+Validation of the cross-dating was done using COFECHA software [@Holmes1983].
 
 The growth trends were analyzed at different time scales.
-To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-withened residual chronologies calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frecuency cutoff at 30 years [@Fritts1976].
-Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008, @Bunn2010] package.
-Residual site chronologies were obtained computing the bi-weight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990].
+To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-whitened residual chronologies calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976].
+Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008; @Bunn2010] package.
+Residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990].
 The statistical quality of each chronology was checked via expressed population signal [@Wigley1984].
@@
 The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
-BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributatle to increasing stem circumference [@Biondi2008].
+BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference [@Biondi2008].
 We used measured dbh and raw nondetrened ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
@@
 Data were downloaded and preprocess using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018, @Frias2018].
-We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010, @Begueria2014].
+We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014].
 SPEI values for the period 1961-2014 were obtained with a spatial resolution of 1.1 km from the Drought indices dataset for Spain database (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017].
@@
 where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median.
-We used medians since they are mores robust estimator of central tendency than means @Rubino2004 @Camarero2011].
+We used medians since they are mores robust estimator of central tendency than means [@Rubino2004; @Camarero2011].
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
@@
 
-To separate growth periods produced by disturbance events from those by climate, we considered a treshold of 25 % of GC for consi
+To separate growth periods produced by disturbance events from those by climate, we considered a threshold of 25 % of GC for consi
 

### 2018-01-16 man/ms_3_discussion.Rmd
UPDATE

@@
 
+## Reducción crecimiento y sequías
+
+Se observó una gran reducción de crecimiento en 1994-1995, 1999 y 2005 para Pinus nigra en el SE de Andalusia [@SanchezSalguero2013] coincidiendo con años de sequía para esa zona.
+La máxima reducción en crecimiento fue observada para 2005, siendo significativamente mayor que para el resto de años [@SanchezSalguero2013].
+
+## Relaciones crecimiento-clima en otras especies
+
 ## References

### 2018-01-16 man/notes_dendro_analysis.Rmd
UPDATE

@@
 
+# Dendro statistics
+
+see Camarero et al 2013. Ann Forest sciences 
+* Mean and standard deviation (SD) of raw tree-ring width data
+* First-order autocorrelation of raw data (AC1), which measures the year-to-year persistence in ring width
+* Mean sensitivity (MS) of the residual ring-width series, which measures the year-to-year variability in the width index
+* Mean between-trees correlation within each site (r bt), which quantifies the similarity in residual width indices among trees
+* Expressed population signal (EPS) of residual series, which assesses the representativeness of the sample size compared with theoretical infinite population
+
 # Resilience Analysis

### 2018-01-17 man/ms_1_material_methods.Rmd
update

@@
 *Q.
-pyrenaica* requires between 650 and 1200 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm (Martínez-Parras and Molero-Mesa 1982, García and Jiménez 2009), with summer rainfall being a key factor in the distribution of the species (Gavilán et al.
-2007, Río et al.
-2007).
+pyrenaica* requires between 650 and 1200 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species [@Gavilan2007; @delRio2007].
 
@@
 There are eight oak patches (2400 Has) identified (**Figure 1**) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys.
-Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation (Brewer et al.
-2002, Olalde et al.
-2002, Rodríguez-Sánchez et al.
-2010) and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change (Hampe and Petit 2005).
+Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation [@Brewer2002; @Olalde2002; RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].
+
+The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000, @Vivero2000], undergoing intensive anthropic use in the last few decades [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
+In fact, the status of conservation of this species for southern Spain is "Vulnerable" [@Vivero2000].
+The relict presence of this species in Sierra Nevada is related both to its genetic resilience as well as to its high intraspecific genetic diversity [@Valbuena2010].
+However, they are also expected to suffer the impact of climate change, due to their climate requirements (wet summers).
+Thus, simulations of the climate change effects on this habitat forecast a reduction in suitable habitats for Sierra Nevada [@Benito2011].
+
+## Drought episodes
+
+* :red_circle: Meter aquí algunos datos de sequia, similar a lo planteado por Gazol
 
@@
 * O quizá un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
-
-The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests (Melendo and Valle 2000, Vivero et al.
-2000), undergoing intensive anthropic use in the last few decades (Camacho-Olmedo et al.
-2002, Valbuena-Carabaña et al.
-2010).
-In fact, the status of conservation of this species for southern Spain is "Vulnerable" (Vivero et al.
-2000).
-The relict presence of this species in Sierra Nevada is related both to its genetic resilience as well as to its high intraspecific genetic diversity (Valbuena-Carabaña and Gil 2013).
-However, they are also expected to suffer the impact of climate change, due to their climate requirements (wet summers).
-Thus, simulations of the climate change effects on this habitat forecast a reduction in suitable habitats for Sierra Nevada (Benito et al.
-2011).
-
-## Datos de sequía.
-
-* :red_circle: Meter aquí algunos datos de sequia, similar a lo planteado por Gazol
 

### 2018-01-17 man/ms_1_material_methods.Rmd
update methods

@@
 
-### Greenness data
+## Greenness data
 
@@
 
+MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016.
 EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
@@
 pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
-MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016.
 A data filtering was applied to select EVI valid values.
 The filtering was done using quality flags and VI Usefulness Indices accompanying the EVI data.
-We filter out those values affected by high content of aerosols, clouds, shadows, snow or water.
+We filter out those values affected by high content of aerosols, clouds, snow and shadows, following the filtering recomendations of EVI data for mountain regions [@Reyes2015].
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2).
@@
 
-`$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI.
-
 These data are geometrically and atmospherically corrected, and include an index of data quality (reliability, which range from 0 – good quality data – to 4 – raw data or absent for different reasons) based on the environmental conditions in which the data was recorded
@@
 
-OJITO---------- reescribir esto de arriba
+")
+```
 
-After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2015.
+After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2016.
+EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 (:red_circle: Hemos seleccionado EVI medio, además de por los consejos que me ha dicho Domingo, porque he comprobado que existe una correlación entre el evi medio y el evi estacional, sobre todo el de verano.
@@
 
-To explore the effect of drought events on greenness we calculated the EVI standardized anomaly ($EVI~sa$) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012, @Gao2016].
-For each pixel we averaged all the EVI valid values within a year (:red_filter: see quality filter), and then the standardized anomaly was computed as:
+")
+```
+
+To explore the effect of drought events on greenness we calculated the EVI standardized anomaly ($EVI~sa$) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016].
+For each pixel we annually averaged all the EVI valid values, and then the standardized anomaly was computed as:
 
@@
 where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for the year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for the year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
-
-Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1)
-
+Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) (incluir cita).
 ## Field sampling and dendrochronological methods
@@
 For the southern site two elevations were sampled: CA-Low and CA-High.
-All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range (:red_circle: mejorar; citar Pérz-Luque et al..).
+All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range [@PerezLuque2015].
 In each site between 15 and 20 dominant trees were randomly selected.
@@
 
+Statistical descriptive parameters [@Fritts1976] including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology.
+
 una frase conectora 
@@
 We selected grid cells covering each sampled sites.
-Data were downloaded and preprocess using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018, @Frias2018].
+Data were downloaded and preprocess using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018].
 We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014].
@@
 ### Disturbance analyses 
+Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (releases) or decreases (supressions) in radial growth [@Nowacki1997].
+By using a temporal window of reasonable length, this approach filters out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].
 
-Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (releases) or decreases (supressions) in radial growth [@Nowacki1997].
-...
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
-By using a temporal window of reasonable length, this method filters out the response to short-term changes in temperature and precipitation [@Nowacki1997, @Fraver2005].
+")
+```
 
@@
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
-A
+A minimum $GC$ of 25% was established to depict canopy disturbances, and we considered a stand-wise disturbance when more than 50% of the individual trees display growth changes above 25 %.
 
-To separate growth periods produced by disturbance events from those by climate, we considered a threshold of 25 % of GC for consi
-
-A minimum $GC$ of 25% was established to depict canopy disturbances. Furthermore, more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
-
-This approach filters out
-
-We used the percentage Growth-change filter (GC).
-
-, calculated annually using 10-year windows
-
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 Disturbance chronologies were built using tree-ring width to identify historical abrupt positive (releases) or negative (suppressions) changes in growth in each species (Nowacki and Abrams, 1997).
 This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
+")
+```
 
@@
 ```
-
-Statistical descriptive parameters (Fritts, 2001), including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology considering the common 1950–2013 period.
-
-The chronologies were computed as bi-weight robust mean to reduce the bias caused by extreme values and an autoregressive model was applied to remove the autocorrelation of the series (Cook and Kairiukstis 1990).
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
  To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated.
 Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
-
-* GC Nowacki??
+")
+```
 
@@
 
-To evaluate the effects of the disturbance events on greeennes and tree growth we used four resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), recovery (*Rc*) and relative resilience (*RRs*).
-
-The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and befor the disturbance:
+To evaluate the effects of the drougth events on greeennes and tree growth (BAI) we used three resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
+The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred.
+It is estimated as the ratio between the performance during and befor the disturbance:
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 The Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance, and it is estimated as:
@@
 Relative Resilience (*RRs*) = (Postdrought - Drought) / Predrought
+")
+```
 
-We computed the values of these indices for tree growth and greenness during each drought event.
+We computed the values of these indices for tree growth (BAI) and greenness during each drought event.
 We considered 2005 and 2012 as singles drought events.
-The predrought and postdrought values of each target variable (i.e.: tree growth or EVI) we computed as the mean value during a period of three years before and after the disturbance events respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (:red_circle: incluir tabla de coeficientes y/o gráfica??
-como suplement, see Gazol 2017)
+The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (DUDA: incluir tabla de coeficientes y/o gráfica??
+como suplement, see Gazol 2017).
 
 ## Statistical analysis
-
-* Explore anomalies EVI 
-* Explore long and short term trends in RW :red_circle: ver correo Guillermo
-* ANOVA analysis EVI events and populations
 
@@
 Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to mean value in all cases [@Wilcox2012].
-When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012, @Wilcox2012].
+When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012].
 Post-hoc differences were assessed pairwise using a similar boostrap test.
-All the robust ANOVA and post-hoc tests were carried out using the WRS2 [@Mair2017] and rcompanion [@Mangiafico2017] R packages.
+All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages.
 The level of significance was set at 0.05 and adjusted for multiple comparisons.
 
-All analyses were carried out in R software (ver.
-2.11.1), using library dplR to standardize data, library nlme to fit the mixed models and vegan to run the PCA.
+All analyses were carried out in R software [@R2017] and  using library dplR to standardize data, library nlme to fit the mixed models and vegan to run the PCA.
 All tests were at α = 0.05, except when indicated.

### 2018-01-18 man/field_data.Rmd
update ms

--- 2017-04-11 man/field_data.Rmd
@@
     variant: markdown_github
-bibliography: references.bib
-csl: ecology.csl
+bibliography: ../refs/references.bib
+csl: ../refs/ecology.csl
 ---

### 2018-01-18 man/ms_0_intro.Rmd
update methods
update ms

@@
     variant: markdown_github
-bibliography: ../references.bib
-csl: ../ecology.csl
+bibliography: ../refs/references.bib
+csl: ../refs/ecology.csl
 ---
@@
 
+`$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI.
+
 ### Aims

### 2018-01-18 man/ms_1_material_methods.Rmd
update methods
update ms

@@
 The level of significance was set at 0.05 and adjusted for multiple comparisons.
-
-All analyses were carried out in R software [@R2017] and  using library dplR to standardize data, library nlme to fit the mixed models and vegan to run the PCA.
-All tests were at α = 0.05, except when indicated.
-
-## References
+All analyses were carried out in R software [@R2017] using several packages (see Supplementary S)

### 2018-01-18 man/ms_2_results.Rmd
UPDATE
update ms

@@
 
-Tree-growth exhibited a signigicant response to water availability more than to the temperatures.
+Tree-growth exhibited a significant response to water availability more than to the temperatures.
 Cumulative precipitation of the hydrological year (September of the previous year to August of the growth year) influenced positively the tree-growth for the both northern and southern populations (FIG ).

### 2018-01-18 man/ms_figures.Rmd
update
update ms

@@
 ```
+
+### Figure 1
+
+![](../out/fig/geomap/mapa_situacion.jpg)
 

### 2018-01-18 man/ms_tables_portrait.Rmd
update ms

@@
     reference_docx: template_tables_portrait.docx
-bibliography: ../references.bib
-csl: ../ecology.csl
+bibliography: ../refs/references.bib
+csl: ../refs/ecology.csl
 ---

### 2018-01-18 man/notes_EVI.Rmd
add notas sobre EVI
update ms

@@
 
+# Filtrado de datos
+
+Para el filtrado de datos ver el siguiente link
+[`./analysis/prepare_modis_qa.md`](/analysis/prepare_modis_qa.md)
+
 # References

### 2018-01-18 man/report.Rmd
update ms r
add template
include tables rmd child
add figures
updating
update ms

@@
 date: "2017"
-output:  word_document
-bibliography: ../references.bib
-csl: ../ecology.csl
+output:
+  word_document:
+    fig_caption: yes
+    fig_height: 6
+    fig_width: 7.5
+    highlight: null
+    reference_docx: template_ms.docx
+bibliography: ../refs/references.bib
+csl: ../refs/ecology.csl
 ---
@@
 ```
+
+```{r child = 'ms_3_discussion.Rmd'}
+```
+
+```{r child = 'ms_tables.Rmd'}
+```
+
+```{r child = 'ms_figures.Rmd'}
+```

### 2018-01-18 script/genera_all_pkgs.R

@@ -1,0 +1,17 @@
+# Script to generate citations of all pkgs
+
+# Load libraries 
+library("here")
+library("grateful")
+
+# Get all pkgs
+paqui <- scan_packages(all.pkgs = TRUE, include.Rmd = TRUE)
+
+# Get cites of all pkgs 
+citas_paqui <- get_citations(paqui)
+
+# Create rmd 
+create_rmd(citas_paqui, filename = here("man", "supp_0.rmd"))
+
+
+

### 2018-01-18 man/notes/notes_drought.Rmd
move notes to new folder

--- 2017-11-08 man/notes_drought.Rmd
@@
 
-<center>![](/out/images/trigo_2012.jpg)</center>
-
-<br>
+![](../out/images/trigo_2012.jpg)
 
@@
 
-<center>![](/man/images/trigo_2012.jpg)</center>
+![](../out/images/trigo_2012.jpg)
 
@@
 
-<center>![](/man/images/garcia_herrera_granada_zoom.jpg)</center>
+![](../out/images/garcia_herrera_granada_zoom.jpg)
 

### 2018-01-18 man/ms_1_material_methods.Rmd
update text

@@
 where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for the year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for the year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
-Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) (incluir cita).
+Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2010; @Samanta2012].
+
 ## Field sampling and dendrochronological methods

### 2018-01-18 man/ms_2_results.Rmd
update text

@@
 
-```{r}
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
 library(tidyverse)
@@
 
-Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (Fig.
-R1 Profile EVI).
-The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Fig.
-R2 y R3).
+Trends analysis showed that 78.9 % of the pixels of *Q.
+pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
+The strongest trends were observed in southwestern populations (Figura S2).
+Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (Figure 2a).
+The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b).
+Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought (Figure 2b).
 
-Esta disminución para 2005 sin embargo fue heterogénea (ver plot de trajectorias)
+According to the standardized anomalies, *Q.
+pyrenaica* forests suffered a browning during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), however no changes were observed for the 2012 drought (Figure S3).
 
-Durante 2005 la mayoría de los pixeles mostraron browning (99.36 % y 79.37 % para las poblaciones del norte y del sur respectivamente), sin embargo en 2012 la mayoría de los pixeles se clasificaron como no changes en las poblaciones del norte (89.60 %) y en las del sur (70.07 %) (Tabla supplementaria??)
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+* Disminución del EVI --> Espacialización ver plots de trajectorias
+* Por ejemplo la disminución para 2005 fue heterogénea (
+
+* Durante 2005 la mayoría de los pixeles mostraron browning (99.36 % y 79.37 % para las poblaciones del norte y del sur respectivamente), sin embargo en 2012 la mayoría de los pixeles se clasificaron como no changes en las poblaciones del norte (89.60 %) y en las del sur (70.07 %) (Tabla supplementaria??)
 
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
 > Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005.
@@
 > * En 2005 y en 2012 las anomalías fueron negativas.
-> Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.")
-```
+> Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.
 
@@
 En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.
+")
+```
 

### 2018-01-18 man/ms_3_discussion.Rmd
update ms
update text

@@
 # Discussions notes
+
+## Greenness
+
+* Aproximadamete el x% de los robledales mostraron un incremento en los valores de productividad (greenees) ... (relacionar con tendencias EVI y ontologias).
+* tendencias ontologias
+
+* Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI.
+En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.
 

### 2018-01-18 man/ms_supp.Rmd
create supplementary rmd

---
title: "ms_supplementary"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")

library("grid")
library("gridExtra")
library("patchwork")

```

# Supplementary

## Supplementary 1

Listado de pkgs usados

## Suplementary 2

## Figure S2

![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}

Figure S2. Spatial distribution of the EVI-annual trends for the period 2000-2016. Trends ($\tau$) with the most positive values (green) indicate positive trend.

## Figure S3

```{r}
df <- read.csv(file='../out/anomalies/evi/percen_browing_greening.csv', header=TRUE)

df <- df %>% add_row(clu_pop2 = 'Northern slope',
                             type = 'greening',
                             count_clu = 0, per = 0.00, y = 2005)

per_pixels_green <- df %>% 
  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
  geom_bar(stat='identity', position = "dodge") + 
  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), vjust= - 0.75) + 
  facet_wrap(~clu_pop2) + 
  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
  theme(strip.background = element_rect(colour = "black", fill = "white")
        ) + theme_minimal() + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B"))

per_pixels_green
```

Figure S3. Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.

### 2018-01-18 man/ms_figures.Rmd
update figures

@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r, message=FALSE, echo=FALSE}
+```{r}
 library("tidyverse")
@@
 library("gridExtra")
+library("patchwork")
+```
+
+```{r, echo=FALSE}
+show_text <- FALSE
+```
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+# Lista de figuras
+
+## Figure 1 Location map 
+- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
+Ver MIGRAME dataset.
+
+    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf 
+    - Ojo con los nombres SJ (Que hacemos?)
+
+## Figure 2. Plot combinado con EVI profile (a) y anomalías (b)
+")
 ```
@@
 
+Figure 1.
+Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain range, where three cluster of oak populations have been identified [@PerezLuque2015] (showed in different colour) (b).
+A grid of with the MODIS pixels for each population is shown (see material and methods).
+Detailed location of the sampling sites: northern (San Juan, SJ) (c) and southern ones (Cáñar: CALow and CAHigh)(d).
+Colour Orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
+
 ### Figure 2
 
-```{r, echo=FALSE}
+```{r}
 iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
@@
         panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9)) +
+        legend.position = c(0.1,.9),
+        plot.title = element_text(hjust = -0.05, size = 16, 
+                                    margin = margin(t=-5, b=1))) +
   scale_color_manual(values=c(col2005, col2012, colref))
@@
 
-```{r, echo=FALSE}
+```{r}
 # From explore_anomalies.Rmd 
@@
         panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"))
+        strip.background = element_rect(fill = "white"),
+        plot.title = element_text(hjust = -0.05, size = 16, 
+                                    margin = margin(t=-5, b=1)))
 
@@
 
-```{r, fig.height=9, echo=FALSE} 
-grid.arrange(profile_compara, plot_sa_clu)
+```{r, fig.height=9} 
+# grid.arrange(profile_compara, plot_sa_clu)
 ```
 
+```{r, fig.height=7}
+a_profile_compara <- profile_compara + labs(title = "a")
+b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
+a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
+```
+
+Figure 2.
+Profile of montly EVI values for *Q.
+pyrenaica* forests during the 2005 (*green*) and 2012 (*blue*) drought events and for the reference period (*gray*) (a).
+EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations (b).
+
+Comparison of EVI profile for the reference period (*grey*) and the 2005 (*green*) and 2012 (*blue*) drought events
+
 ### Figure R2 = 3 
-```{r, echo=FALSE}
+```{r}
 mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)

### 2018-01-18 pres/state_ms.Rmd
update state

@@
 classoption: compress
-bibliography: ../references.bib
-csl: ../nature.csl
+bibliography: ../refs/references.bib
+csl: ../refs/nature.csl
 header-includes:

### 2018-01-18 analysis/rwi/climate_rwi_relate.Rmd
update figures

@@
     variant: markdown_github
-bibliography: ../references.bib 
-csl: ../ecology.csl
+bibliography: ../refs/references.bib 
+csl: ../refs/ecology.csl
 ---

### 2018-01-18 man/ms_2_results.Rmd
update figures

@@
 
-#### Resilience metrics EVI
+## Responses to drougth: greenness
 
-Pyrenean oak forests showed significantly lower resistance to 2005 drought event than to 2012 one [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); table R1; Figura R4a, F = 799.86, p <0.0001].
-The 2005 drought reduced the greenness of oak to 85.8 % while the 2012 reduced 94.3 %.
-Southern populations showed significantly higher values of resistance to drought than northern ones, except for 2012 where non-significant differences were recorded (table R1, Figure R4a).
+*Q.
+pyrenaica* forest showed significantly lower resistance to 2005 drought event than to 2012 one [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); table 2; Figure 3a, F = 799.86, p <0.0001].
+The 2005 drought reduced the greenness of oak to 85.8 % while the 2012 reduced 94.3 % (tabla rs1 quizás, duda en esto).
+Southern populations showed significantly higher values of resistance to drought than northern ones, except for 2012 where non-significant differences were recorded (Figure 3a).
 
@@
 In the period after 2005 drought, greeness achieved was 112 % (Rc = 1.12) and after 2012 105.7 % (Rc = 1.057).
-A similar recovery after the 2005 and 2012 drought event was observed for southern populations (p = 0.2453; Figure R4), whilst the northern populations showed a significantly greater recovery after the 2005 drought than after the 2012 drought.
+A similar recovery after the 2005 and 2012 drought event was observed for southern populations (p = 0.2453; Figure 3a), whilst the northern populations showed a significantly greater recovery after the 2005 drought than after the 2012 drought.
 
@@
 
-table R1. Robust anovas con F-Values 
 Figura R2. Interaction plot resilience metrics
@@
 
-## Resiliencia BAI
+## Responses to drougth: tree-growth (BAI)
 
-Although no significant differences were observed in the resistance of the oak growth to both drought events (F = 6.0189; p = 0.019; table R2; Figura R5), the 2005 drought reduced growth more than 2012 did (Rt = 0.721 and 0.819 respectively).
+Although no significant differences were observed in the resistance of the oak growth to both drought events (F = 6.0189; p = 0.019; table 2; Figure 3b), the 2005 drought reduced growth more than 2012 did (Rt = 0.721 and 0.819 respectively).
 Northern population showed resistance values lower than southern ones, especially for 2005 drought event where the growth was reduced to 44.5 % of the preceding period.

### 2018-01-18 man/ms_figures.Rmd
update figures

@@
 
-![](../out/fig/geomap/mapa_situacion.jpg)
+![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
 
@@
 Figure 2.
-Profile of montly EVI values for *Q.
-pyrenaica* forests during the 2005 (*green*) and 2012 (*blue*) drought events and for the reference period (*gray*) (a).
+Comparison of EVI profile for the reference period (*gray*) and during the 2005 (*green*) and 2012 (*blue*) drought events (a).
 EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations (b).
 
-Comparison of EVI profile for the reference period (*grey*) and the 2005 (*green*) and 2012 (*blue*) drought events
+### Figure 3
 
-### Figure R2 = 3 
 ```{r}
 mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
+mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
 
 # NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs')
+mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
+mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
 
 pd <- position_dodge(.2)
+line_lwd <- .8
 
-robust_plot_evi_drought <-
-  ggplot(mhuber_filter, 
-         aes(x = as.factor(disturb_year), y = M.Huber, 
-             color = site, 
-             group = site, 
-             fill = site)) +
+aes_disturb <- aes(x = disturb_year, y = M.Huber, color = site, group = site, fill = site)
+aes_site <- aes(x=site, y=M.Huber, color = disturb_year, group=disturb_year, fill=disturb_year)
+
+theme_res_site  <- list(
   geom_errorbar( 
-    aes(ymin = lower.ci, ymax = upper.ci),
-    width = .1, size = 0.7, position = pd) +
-  geom_line(aes(group = site, color = site, linetype = site), position = pd) +
-  geom_point(shape = 21, size = 2.5, position = pd) +
+    aes(ymin = lower.ci, ymax = upper.ci, linetype=disturb_year),
+    width=.2, position=pd, lwd=line_lwd),
+  geom_line(aes(group = disturb_year, color = disturb_year, linetype = disturb_year), position = pd, lwd=line_lwd), 
+  geom_point(shape = 21, size = 2.5, position = pd), 
   geom_text_repel(aes(label=Letter), 
-                  nudge_x = 0.3, size = 5,
-                  segment.size = 0) +
+                  nudge_x = 0.3, direction="both", 
+                  size = 5, show.legend = FALSE,
+                  segment.size = 0.7, 
+                  segment.alpha = 0.5),
   facet_wrap(
-    ~var_sorted, nrow = 1, scales = "free_y",
+    ~var_sorted, nrow = 1, scales="free_y",
     labeller = as_labeller(c(
@@
       "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))) +
-  scale_color_manual(values = c("black", "blue")) +
-  scale_fill_manual(values = c("black", "blue")) + theme_bw() +
-  scale_linetype_manual(values = c("solid", "dashed")) +
+      "2_rs" = "Resilience"))),
+  theme_bw(), 
   theme(
@@
       fill = "white"),
-    legend.position = "bottom") +
-  ylab("") + xlab("Drought event")
+    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
+  ylab("Response value"), 
+  xlab("")
+)
 
-robust_plot_evi_drought 
+theme_res_disturbyear  <- list(
+  geom_errorbar( 
+    aes(ymin = lower.ci, ymax = upper.ci, linetype=site),
+    width=.2, position=pd, lwd=line_lwd),
+  geom_line(aes(group = site, color = site, linetype = site), position = pd, lwd=line_lwd), 
+  geom_point(shape = 21, size = 2.5, position = pd), 
+  geom_text_repel(aes(label=Letter), 
+                  nudge_x = 0.3, direction="both", 
+                  size = 5, show.legend = FALSE,
+                  segment.size = 0.7, 
+                  segment.alpha = 0.5),
+  facet_wrap(
+    ~var_sorted, nrow = 1, scales="free_y",
+    labeller = as_labeller(c(
+      "0_rt" = "Resistance",
+      "1_rc" = "Recovery",
+      "2_rs" = "Resilience"))),
+  theme_bw(), 
+  theme(
+    panel.grid.minor = element_blank(),
+    strip.background = element_rect(
+      colour = "black",
+      fill = "white"),
+    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
+  ylab("Response value"), 
+  xlab("")
+)
+
+theme_res_evi <-  list(
+  scale_color_manual(values = c("black", "blue")),
+  scale_shape_manual(values=c(19, 22)),
+  scale_fill_manual(values = c("black", "blue")),
+  scale_linetype_manual(values = c("solid", "solid"))
+)
+
+theme_res_bai <- list(
+  scale_color_manual(values=c('black','blue','blue')), 
+  scale_shape_manual(values=c(19, 22, 15)),
+  scale_fill_manual(values=c('black','white','blue')),  
+  scale_linetype_manual(values=c("solid", "solid", 'solid'))
+)
+
 ```
+
+```{r}
+res_evi_drought <-
+  ggplot(mhuber_filter, aes_disturb) +
+  theme_res_disturbyear + 
+  theme_res_evi +
+  theme(legend.position='top') + 
+  ggtitle("a) EVI")
+
+res_bai_drought <-
+  ggplot(mhuberbai_filter, aes_disturb) + 
+  theme_res_disturbyear + 
+  theme_res_bai + 
+  theme(legend.position='bottom') + 
+  xlab('Drought event') + 
+  ggtitle("b) BAI")
+```
+
+```{r, fig.height=7}
+res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
+```
+
+Figure 3.
+Response of northern (*black*) and southern (*blue*) populations of *Q.
+pyrenaica* forests to drougth in terms of resistance, recovery and resilience of greenness (EVI; a) and tree growth (BAI; b) for the years 2005 and 2012.
+
+### Figure 4
+```{r}
+res_evi_site <-
+  ggplot(mhuber_filter, aes_site) +
+  theme_res_site + 
+  theme_res_evi + theme(legend.position='top') + 
+  ggtitle("a) EVI")
+
+res_bai_site <-
+  ggplot(mhuberbai_filter, aes_site) + 
+  theme_res_site + 
+  theme_res_bai + 
+  theme(legend.position='bottom') + 
+  xlab('Drought event') + 
+  ggtitle("b) BAI")
+```
+
+```{r, fig.height=7}
+res_evi_site  + res_bai_site + plot_layout(ncol=1)
+```

### 2018-01-18 man/report.Rmd
update figures

@@
     fig_height: 6
-    fig_width: 7.5
+    fig_width: 6
     highlight: null
@@
 ```
+
+```{r child = 'ms_supp.Rmd'}
+```

### 2018-01-22 analysis/dendro_summary.Rmd
add summary dendro

---
title: "Summary Dendro"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Dec"
output:  
  md_document:
    variant: markdown_github
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("stringr")
library("dplR")
library("knitr")
library("pander")

```

# Read y Prepare data 
* Leer datos `rwl` de SJ y CA

```{r, echo=FALSE}
machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# sj 
sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")

# canar 
ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
```

Computar RWI para tres sitios CA_High, CA_Low, SJ.

# Preparar datos para computar RWI 
```{r}
source(paste0(di, 'script/R/computeSpline.R'))
source(paste0(di, 'script/R/rw_byTree.R'))
```

```{r}
# Replace SNA by SJ and SNB by CA
names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")

# Remove g in name of some cores of CA. 
names(ca) <- stringr::str_replace(names(ca), "g", "")
```

* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ

```{r}
# Create subset to compare between sites 
caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]

# remove the rows with NA across all columns 
caL <- caL[rowSums(is.na(caL))!=ncol(caL), ]
```

### Summary Dendro

* Summary stats of RWL

```{r}
rwl_stat_sj <- rwl.stats(sj)
rwl_stat_caH <- rwl.stats(caH)
rwl_stat_caL <- rwl.stats(caL)

rwl_stat_sites <- rbind(rwl_stat_sj, rwl_stat_caH, rwl_stat_caL) %>% 
  mutate(
  site = as.factor(case_when(
    series %in% names(sj) ~ "sj", 
    series %in% names(caH) ~ "caH", 
    series %in% names(caL) ~ "caL")),
  tree = as.numeric(substr(series, 3, 4)),
  radi = as.numeric(substr(series, 5, 6))
  )

# average age 
age_sites <- rwl_stat_sites %>% 
  group_by(site) %>% 
  dplyr::summarise(age_mean = mean(year), 
         age_std = sd(year)) %>% as.data.frame()

pander(age_sites) 
write.csv(age_sites, file=paste(di, "data/dendro_summary/site3_avg_ages.csv", sep=""), row.names = FALSE)

rwl_sites_summ <- rwl_stat_sites %>% 
  group_by(site) %>% 
  dplyr::summarise(n_trees = length(unique(tree)),
                   n_radii =n(), 
                   year_first = min(first), 
                   year_last = max(last), 
                   year_length = max(year),
                   rw_mean = mean(mean), 
                   rw_std = mean(stdev),
                   sens1 = mean(sens1), 
                   sens2 = mean(sens2), 
                   ar1 = mean(ar1)) %>% 
  as.data.frame()

pander(rwl_sites_summ)
```

* sample deepth

```{r}
# sample depth 
sampleDepth <- function(x, site){ 
  x$samps <- rowSums(!is.na(x))
  x$year <- row.names(x)
  x$site <- site
  out <- x[,c("samps", "year", "site")]
  return(out)
}

d_sj <- sampleDepth(sj, site='sj')
d_caL <- sampleDepth(caL, site='caL')
d_caH <- sampleDepth(caH, site='caH')

d_sites <- rbind(d_sj, d_caH, d_caL)

d <- d_sites %>% 
  filter(samps >= 5) %>% 
  group_by(site) %>% 
  slice(which.min(year)) %>% 
  mutate(
    year = as.numeric(year),
    length_year = (2016 - year + 1)) %>%  
  as.data.frame()

pander(d) 
```

* Combining tables

```{r}
df <- d %>% 
  dplyr::select(-samps, -year) %>% 
  rename(year_length5 = length_year) %>% 
  left_join(rwl_sites_summ, by='site') %>% 
  dplyr::select(
    site, n_trees, n_radii, 
    first_year = year_first, last_year = year_last, 
    length_year = year_length, length_year5 = year_length5, 
    rw_mean, rw_std, ms2 = sens2, ar1)

pander(df)
```

* Summary stats of Residual chronos

```{r}
rwi_stats <- read.csv(file=paste(di, "data/dendro_summary/site3_dendro_rwi.csv", sep=""),  header=TRUE, sep=',')
```

# Dendro table 
```{r}
dendro_summary <- rwi_stats %>% 
  dplyr::select(rbar.wt, eps, site) %>% 
  inner_join(df, by='site') %>% 
  dplyr::select(site:ar1, rbar.wt, eps)

write.csv(dendro_summary, file=paste(di, "data/dendro_summary/site3_dendro_summary.csv", sep=""), row.names = FALSE)

pander(dendro_summary, caption='Dendrochronological summary by sites') 
```

### 2018-01-22 analysis/analysis_stand_features.Rmd
create table
stand analysis features updated 3 sites

@@
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 May"
+date: "2018 Jan"
 output:   
@@
 
+### - old code 4 sites 
 # Set levels of eleveation 
-sj_lowcode  <- paste0('A', str_pad(1:10, 2, pad='0'))
-sj_highcode <- paste0('A', 11:20)
-ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')),
-            paste0('B', 26:30))
+# sj_lowcode  <- paste0('A', str_pad(1:10, 2, pad='0'))
+# sj_highcode <- paste0('A', 11:20)
+# ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')),
+#             paste0('B', 26:30))
+# ca_highcode <- paste0('B', 11:25)
+# 
+# tree <- tree %>% 
+#   mutate(elevF = ifelse(id_focal %in% sj_lowcode, 'Low',
+#                        ifelse(id_focal %in% sj_highcode, 'High',
+#                               ifelse(id_focal %in% ca_lowcode, 'Low', 'High')))) %>%
+#   mutate(site = paste0(loc, '_', elevF))
+# 
+#
+
+sj_code  <- paste0('A', str_pad(1:20, 2, pad='0'))
+ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')), paste0('B', 26:30))
 ca_highcode <- paste0('B', 11:25)
 
-tree <- tree %>% 
-  mutate(elevF = ifelse(id_focal %in% sj_lowcode, 'Low',
-                       ifelse(id_focal %in% sj_highcode, 'High',
-                              ifelse(id_focal %in% ca_lowcode, 'Low', 'High')))) %>%
-  mutate(site = paste0(loc, '_', elevF))
+tree <- tree %>% mutate(
+  site = as.factor(case_when(
+    id_focal %in% sj_code ~ "sj", 
+    id_focal %in% ca_highcode ~ "caH", 
+    id_focal %in% ca_lowcode ~ "caL")))
 
@@
   filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  mutate(site = as.factor(site))
+  filter(id_focal!='Fresno')
 
@@
 ```{r}
-general_var <- ft %>% group_by(loc, site) %>% count()
+general_var <- ft %>% group_by(site) %>% count()
 
@@
 
-## Add code site and get centroid (see # http://rspatial.org/analysis/rst/8-pointpat.html)
-sp_ca <- sp_ca %>% 
-  mutate(loc = 'CA', 
-         elevF = ifelse(name %in% ca_lowcode, 'Low', 'High'),
-         site = paste0(loc, '_', elevF))
+sp_sites <- rbind(sp_sj, sp_ca) %>% 
+  mutate(site = as.factor(case_when(
+    name %in% sj_code ~ "sj", 
+    name %in% ca_highcode ~ "caH", 
+    name %in% ca_lowcode ~ "caL")))
 
-coord_ca <- sp_ca %>% 
+coord_sites <- sp_sites %>% 
   group_by(site) %>%
-  summarise(lat_m = mean(lat),
-            long_m = mean(long))
+  dplyr::summarise(lat_m = mean(lat), long_m = mean(long))
 
-# plot(sp_ca$long, sp_ca$lat, pch=19, col='gray')
-# points(coord_ca$long_m, coord_ca$lat_m, pch=19, col='blue')
+coord_sites %>% kable()
 
-sp_sj <- sp_sj %>% 
-  mutate(loc = 'SJ', 
-         elevF = ifelse(name %in% sj_lowcode, 'Low', 'High'),
-         site = paste0(loc, '_', elevF))        
-coord_sj <- sp_sj %>% 
-  group_by(site) %>%
-  summarise(lat_m = mean(lat),
-            long_m = mean(long))
-
-coords_sites <- coord_sj %>% rbind(coord_ca)
-
-#plot(sp_sj$long, sp_sj$lat, pch=19, col='gray')
-#points(coord_sj$long_m, coord_sj$lat_m, pch=19, col='blue')
-
-coords_sites %>% kable()
+# old code 
+# 
+# 
+# 
+# ## Add code site and get centroid (see # http://rspatial.org/analysis/rst/8-pointpat.html)
+# sp_ca <- sp_ca %>% 
+#   mutate(loc = 'CA', 
+#          
+#          
+#          elevF = ifelse(name %in% ca_lowcode, 'Low', 'High'),
+#          site = paste0(loc, '_', elevF)) 
+# 
+# coord_ca <- sp_ca %>% 
+#   group_by(site) %>%
+#   summarise(lat_m = mean(lat),
+#             long_m = mean(long))
+# 
+# # plot(sp_ca$long, sp_ca$lat, pch=19, col='gray')
+# # points(coord_ca$long_m, coord_ca$lat_m, pch=19, col='blue')
+#   
+# sp_sj <- sp_sj %>% 
+#   mutate(loc = 'SJ', 
+#          elevF = ifelse(name %in% sj_lowcode, 'Low', 'High'),
+#          site = paste0(loc, '_', elevF))        
+# coord_sj <- sp_sj %>% 
+#   group_by(site) %>%
+#   summarise(lat_m = mean(lat),
+#             long_m = mean(long))
+# 
+# coords_sites <- coord_sj %>% rbind(coord_ca)
+# 
+# #plot(sp_sj$long, sp_sj$lat, pch=19, col='gray')
+# #points(coord_sj$long_m, coord_sj$lat_m, pch=19, col='blue')
+# 
+# coords_sites %>% kable()
 ```
@@
 compe <- compe %>% 
-  mutate(name = id_focal) %>% 
-  mutate(loc = ifelse(str_detect(name, "A"), 'SJ', 'CA'),
-         elevF = ifelse(name %in% sj_lowcode, 'Low',
-                        ifelse(name %in% sj_highcode, 'High',
-                              ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
-  mutate(site = paste0(loc, '_', elevF)) %>% 
-  mutate(site = as.factor(site),
-         loc = as.factor(loc),
-         elevF = as.factor(elevF))
+  mutate(site = as.factor(case_when(
+    id_focal %in% sj_code ~ "sj", 
+    id_focal %in% ca_highcode ~ "caH", 
+    id_focal %in% ca_lowcode ~ "caL")))
+
+# old code 
+# compe <- compe %>% 
+#   mutate(name = id_focal) %>% 
+#   mutate(loc = ifelse(str_detect(name, "A"), 'SJ', 'CA'),
+#          elevF = ifelse(name %in% sj_lowcode, 'Low',
+#                         ifelse(name %in% sj_highcode, 'High',
+#                               ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
+#   mutate(site = paste0(loc, '_', elevF)) %>% 
+#   mutate(site = as.factor(site),
+#          loc = as.factor(loc),
+#          elevF = as.factor(elevF))
 
@@
     group_by(site) %>% 
-    summarise_each(funs(mean, sd, se=sd(.)/sqrt(n()), min, max))
+    summarise_all(funs(mean, sd, se=sd(.)/sqrt(n()), min, max))
 
@@
 
+# topo <- topo %>% 
+#   mutate(loc = ifelse(str_detect(name, "A"), 'SJ', 'CA'),
+#          elevF = ifelse(name %in% sj_lowcode, 'Low',
+#                         ifelse(name %in% sj_highcode, 'High',
+#                               ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
+#   mutate(site = paste0(loc, '_', elevF)) %>% 
+#   mutate(site = as.factor(site),
+#          loc = as.factor(loc),
+#          elevF = as.factor(elevF))
+
 topo <- topo %>% 
-  mutate(loc = ifelse(str_detect(name, "A"), 'SJ', 'CA'),
-         elevF = ifelse(name %in% sj_lowcode, 'Low',
-                        ifelse(name %in% sj_highcode, 'High',
-                              ifelse(name %in% ca_lowcode, 'Low', 'High')))) %>%
-  mutate(site = paste0(loc, '_', elevF)) %>% 
-  mutate(site = as.factor(site),
-         loc = as.factor(loc),
-         elevF = as.factor(elevF))
+  mutate(site = as.factor(case_when(
+    name %in% sj_code ~ "sj", 
+    name %in% ca_highcode ~ "caH", 
+    name %in% ca_lowcode ~ "caL")))
 
-# 
-# topo_summary <- topo %>%
-#   group_by(site) %>%
-#   summarise(mde_m = mean(mde),
-#             mde_sd = sd(mde),
-#             mde_min = min(mde),
-#             mde_max = max(mde),
-#             slope_m = mean(slope),
-#             slope_sd = sd(slope))
-# 
-# 
-# 
-# # Another way to obtain summary values 
-# variables <- c('mde','slope','aspect')
-# auxdf <- data.frame() 
-# 
-# for (i in variables){ 
-# aux <- topo %>% 
-#   dplyr::group_by(site) %>% 
-#   summarise_each_(funs(mean, sd, se=sd(.)/sqrt(n())), i) %>% mutate(variable=i) 
-# 
-# auxdf <- rbind(auxdf, aux) }
 ```
@@
 ft_sel <- ft %>%
-  mutate(dn = dn_mm / 1000, 
+  mutate(dn = dn_mm / 10, 
          h = height_cm / 100) %>% 
@@
 
+# out4 <- out3 %>% 
+#   transmute(variables = mivar,
+#             CA_High_m = round(mean_CA_High, 3),
+#             CA_High_sd = round(sd_CA_High, 3),
+#             CA_High_l = tukey_CA_High,
+#             CA_Low_m = round(mean_CA_Low, 3),
+#             CA_Low_sd = round(sd_CA_Low, 3),
+#             CA_Low_l = tukey_CA_Low,
+#             SJ_High_m = round(mean_SJ_High, 3),
+#             SJ_High_sd = round(sd_SJ_High, 3),
+#             SJ_High_l = tukey_SJ_High,
+#             SJ_Low_m = round(mean_SJ_Low, 3),
+#             SJ_Low_sd = round(sd_SJ_Low, 3),
+#             SJ_Low_l = tukey_SJ_Low)
+
 out4 <- out3 %>% 
   transmute(variables = mivar,
-            CA_High_m = round(mean_CA_High, 3),
-            CA_High_sd = round(sd_CA_High, 3),
-            CA_High_l = tukey_CA_High,
-            CA_Low_m = round(mean_CA_Low, 3),
-            CA_Low_sd = round(sd_CA_Low, 3),
-            CA_Low_l = tukey_CA_Low,
-            SJ_High_m = round(mean_SJ_High, 3),
-            SJ_High_sd = round(sd_SJ_High, 3),
-            SJ_High_l = tukey_SJ_High,
-            SJ_Low_m = round(mean_SJ_Low, 3),
-            SJ_Low_sd = round(sd_SJ_Low, 3),
-            SJ_Low_l = tukey_SJ_Low)
+            caH_m = round(mean_caH, 3),
+            caH_sd = round(sd_caH, 3),
+            caH_l = tukey_caH,
+            caL_m = round(mean_caL, 3),
+            caL_sd = round(sd_caL, 3),
+            caL_l = tukey_caL,
+            sj_m = round(mean_sj, 3),
+            sj_sd = round(sd_sj, 3),
+            sj_l = tukey_sj)
 
 # Export data 
-write.csv(out4, file=paste(di, "data/proto_tables/site_features.csv", sep=""), row.names = FALSE)
+# write.csv(out4, file=paste(di, "data/proto_tables/site_features.csv", sep=""), row.names = FALSE)
+write.csv(out4, file=paste(di, "data/competence/resumen_medias_indices_competencia.csv", sep=""), row.names = FALSE)
 
-## General Variables and coordinates
+pander(out4, caption='Summary Values')
+```
 
-general_var %>% 
-  left_join(coords_sites, by='site') %>% 
-  write.csv(file=paste(di, "data/proto_tables/site_general.csv", sep=""), row.names = FALSE)
+# General topo of the sites
+
+```{r}
+topo_site <- topo %>% group_by(site) %>% summarise(
+  elev_mean = mean(mde), 
+  elev_min = min(mde), 
+  elev_max = max(mde), 
+  slope_mean = mean(slope)
+)
+
+site_general <- general_var %>% 
+  left_join(coord_sites, by='site') %>% 
+  left_join(topo_site, by='site')
+
+write.csv(site_general, file=paste(di, "data/sites_summary/site_general.csv", sep=""), row.names = FALSE)
 ```

### 2018-01-22 man/ms_tables.Rmd
update figures
update tables

@@
 
+```{r, echo=FALSE}
+show_text <- FALSE
+```
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+# Lista de tablas
+
+# Table 1 
+* topograficos, y dendro summary
+
+# Table 2 ANOVAS resilience BAI and EVI
+
+# Table RS1 
+* Estimadores huber resiliences EVI
+
+# Table RS2
+
+* Estimadores huber resiliences Bai
+
+")
+```
+
+## Table 1. Dendro summary statistics 
+```{r, echo=FALSE, include = FALSE}
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+
+```
+
+## Table 2. Robust ANOVAS (F-value)
 ```{r, echo=FALSE, include = FALSE}
@@
 
-### Table R1
+## Table 2
 
 ```{r, echo=FALSE}
-
+# table 2 (r1)
 dffs <- dff_bai %>% bind_rows(dff_evi)
@@
 ```
-
-# Table RS1 
-* Estimadores huber resiliences EVI
-
-# Table RS2
-
-* Estimadores huber resiliences Bai

### 2018-01-22 analysis/analysis_stand_features.Rmd
update

--- 2018-01-22 analysis/analysis_stand_features.Rmd
@@
 date: "2018 Jan"
-output:   
-  md_document:
-    toc: true
-    variant: markdown_github
+output:  pdf_document
+bibliography: references.bib 
+csl: ecology.csl
 ---
@@
 ```{r}
+ft_summary <- ft_sel %>% group_by(site) %>% 
+  dplyr::summarise(h_mean = mean(h),
+                   h_sd = sd(h),
+                   dn_mean = mean(dn),
+                   dn_sd = sd(dn)) %>% 
+  as.data.frame()
+
+compe_summary <- compe %>%  
+  mutate(ba_ha = (ba * 10000)/(pi*100)) %>% 
+  group_by(site) %>% 
+  dplyr::select(ba_ha, std, site) %>% 
+  dplyr::summarise(ba_mean = mean(ba_ha), 
+                   ba_sd = sd(ba_ha),
+                   stand_density_mean = mean(std), 
+                   stand_density_sd = sd(std)) %>% 
+  as.data.frame()
+
+# DBH and H of all trees (no focal tree)
+
+all <- nft %>% mutate(dn = dn_mm / 10, 
+                      h = height_cm / 100) %>% 
+  dplyr::select(dn, h, site) %>% 
+  bind_rows(ft_sel) %>% 
+  group_by(site) %>% 
+  dplyr::summarise(h_mean_all = mean(h),
+                   h_sd_all = sd(h),
+                   dn_mean_all = mean(dn),
+                   dn_sd_all = sd(dn)) %>% 
+  as.data.frame()
+
+stand_biotic <- ft_summary %>% 
+  inner_join(compe_summary, by = 'site') %>% 
+  inner_join(all, by='site')
+
+# Export data 
+write.csv(stand_biotic, file=paste(di, "data/competence/biotic_stand_features.csv", sep=""), row.names = FALSE)
+
+pander(stand_biotic)
+```
+
+```{r}
 # Join all data and export as table
@@
   left_join(coord_sites, by='site') %>% 
-  left_join(topo_site, by='site')
+  left_join(topo_site, by='site') %>% 
+  as.data.frame()
+
+pander(site_general)
 

### 2018-01-22 man/ms_tables_portrait.Rmd
update

@@
 
+```{r, echo=FALSE}
+show_text <- FALSE
+```
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+# Lista de tablas
+
+# Table 1 
+* Site features
+
+# Table 2 
+* Dendro summary
+
+# Table 3 ANOVAS resilience BAI and EVI
+
 # Table RS1 
+* Estimadores huber resiliences EVI
+
+# Table RS2
+
+* Estimadores huber resiliences Bai
+
+")
+```
+
+## Table 1. Characteristics of the sampled plots
+
+```{r, echo=FALSE}
+# https://rdrr.io/cran/pander/man/panderOptions.html
+
+topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
+ages <- read.csv(here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
+dendro_summ <- read.csv(here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
+biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)
+
+t1raw <- topo %>% inner_join(ages, by='site') %>% 
+  inner_join(dendro_summ, by='site') %>% 
+  inner_join(biotic, by='site')
+
+t1raw <- t1raw %>% mutate(name = case_when(
+  site == 'caH' ~ 'Cáñar-High', 
+  site == 'caL' ~ 'Cáñar-Low',
+  site == 'sj' ~ 'San Juan'))
+
+t1 <- t1raw %>% dplyr::select(
+  site, lat_m, long_m, elev_mean, slope_mean, 
+  n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, 
+  dn_mean_all, dn_sd_all, h_mean_all, h_sd_all,
+  ba_mean, ba_sd, stand_density_mean, stand_density_sd) %>% 
+  mutate(pare = ")") %>% 
+  mutate_at(c(2,3,5,18,19), round, 2) %>% 
+  mutate_at(4, round, 0) %>% 
+  mutate_at(c(8:17, 20:21), round, 1) %>% 
+  unite(dn, c(dn_mean, dn_sd), sep=" (") %>% 
+  unite(h, c(h_mean, h_sd), sep=" (") %>% 
+  unite(age, c(age_mean, age_std), sep=" (") %>% 
+  unite(dn_all, c(dn_mean_all, dn_sd_all), sep=" (") %>% 
+  unite(h_all, c(h_mean_all, h_sd_all), sep=" (") %>% 
+  unite(ba, c(ba_mean, ba_sd), sep=" (") %>% 
+  unite(stand_density, c(stand_density_mean, stand_density_sd), sep=" (") %>% 
+  mutate(
+    dn = paste0(dn, pare),
+    h = paste0(h, pare), 
+    age = paste0(age, pare),
+    dn_all = paste0(dn_all, pare),
+    h_all = paste0(h_all, pare),
+    ba = paste0(ba, pare),
+    stand_density = paste0(stand_density, pare)) %>% 
+  dplyr::select(-pare)
+
+colnames (t1) <- c("Name", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
+                   "# trees", '# cores', "Dbh (cm)", "Height (m)", "Age (years)", 
+                   "Dbh all (cm)", "Height all (m)", "BA (m2/ha)", "Density (trees/ha)")
+
+pander(t1, split.table = Inf)
+```
+
+## Table 2. Dendro summary
+
+```{r, echo=FALSE}
+t2 <- t1raw %>% dplyr::select(
+  site, first_year, last_year, 
+  length_year, length_year5, 
+  n_trees, n_radii,
+  rw_mean, rw_std, ms2, ar1, rbar.wt, eps)  %>% 
+  mutate(pare = ")") %>% 
+  mutate_at(c(8:13), round, 3) %>% 
+  unite(length_year, c(length_year,length_year5), sep=" (") %>%
+  unite(RW, c(rw_mean,rw_std), sep=" (") %>%
+  mutate(
+    length_year = paste0(length_year, pare),
+    RW = paste0(RW, pare)) %>% 
+  dplyr::select(-pare)
+
+colnames(t2) <- c("Name", "First year", "Last year", "Length (years)",  
+                   "# trees", '# cores', "RW (mm)", "MS", "AR(1)", "Rbt", "EPS")
+
+pander(t2,
+       split.table = Inf)
+
+```
+
+## Table 3. Robust ANOVAS (F-value)
+```{r, echo=FALSE, include = FALSE}
+### Table R1_evi
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df %>% gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
+
+colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rrs\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
+
+```
+
+```{r, echo=FALSE, include = FALSE}
+### Table R1_evi sin rrs 
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df %>% 
+  filter(variables != "rrs") %>% 
+  gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
+
+dff_evi <- dff %>% mutate(target = 'EVI')
+
+colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
+
+```
+
+```{r, echo=FALSE, include = FALSE}
+
+### Table R1_bai
+df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df %>% gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
+
+colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rrs\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
+
+```
+
+```{r, echo=FALSE, include = FALSE}
+### Table R1_bai sin rrs 
+df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
+
+dff <- df %>% 
+  filter(variables != "rrs") %>% 
+  gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+
+dff_bai <- dff %>% mutate(target = 'BAI')
+
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
+
+colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
+
+```
+
+```{r, echo=FALSE}
+# table 2 (r1)
+dffs <- dff_bai %>% bind_rows(dff_evi)
+
+# write.csv(dffs, here("/out/anovas_resilience", "trimmed_anovas_joined.csv"), row.names = FALSE)
+
+# kable(dffs, format = "latex", booktabs = T,
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "variable"),
+#       digits=c(0,2,3,2,3,2,3,2,3,0)) %>% 
+#   kable_styling(font_size = 8) %>% 
+#   add_header_above(c("", "rc" = 2, "rs" = 2, "rt" = 2))
+
+colnames(dffs) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np", "variable")
+pander(dffs,
+             round=c(0,2,3,2,3,2,3,2,3,0),
+             digits=c(0,5,4,5,4,5,4,5,4,0),
+             keep.line.breaks = TRUE
+             )
+```
+
+## Table RS1
 

### 2018-01-22 man/ms_tables/t_dendro_summary.Rmd
add tables separately

---
output:  
  word_document:
    reference_docx: "../../man/template_tables_portrait.docx"
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
library(dplyr)
```

**Table 2.** Characteristics of the mean tree ring chronologies. Values of the length year in parenthesis indicate years replicated more than five series. RW = mean annual ring width (standard deviation in parenthesis). MS = mean sensitivity. AR(1) = mean autocorrelation of raw series. Rbt = mean correlation between series. EPS = mean expressed population signal. EPS and Rbt are calculated for the mean residual chronologies of growth indices.

```{r, echo=FALSE}
topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
ages <- read.csv(here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
dendro_summ <- read.csv(here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)

t1raw <- topo %>% inner_join(ages, by='site') %>% 
  inner_join(dendro_summ, by='site') %>% 
  inner_join(biotic, by='site')

t1raw <- t1raw %>% mutate(name = case_when(
  site == 'caH' ~ 'Cáñar-High', 
  site == 'caL' ~ 'Cáñar-Low',
  site == 'sj' ~ 'San Juan'))

t2 <- t1raw %>% dplyr::select(
  site, first_year, last_year, 
  length_year, length_year5, 
  n_trees, n_radii,
  rw_mean, rw_std, ms2, ar1, rbar.wt, eps)  %>% 
  mutate(pare = ")") %>% 
  mutate_at(c(8:13), round, 3) %>% 
  unite(length_year, c(length_year,length_year5), sep=" (") %>%
  unite(RW, c(rw_mean,rw_std), sep=" (") %>%
  mutate(
    length_year = paste0(length_year, pare),
    RW = paste0(RW, pare)) %>% 
  dplyr::select(-pare)

colnames(t2) <- c("Site", "First year", "Last year", "Length (years)",  
                   "# trees", '# cores', "RW (mm)", "MS", "AR(1)", "Rbt", "EPS")

pander(t2,
       split.table = Inf)

```

### 2018-01-22 man/ms_tables/t_huber_res_bai.Rmd
add tables separately

--- 2017-12-18 man/ms_tables_portrait.Rmd
@@
 ---
-title: "ms_tables_portrait"
 output:  
   word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 7.5
-    highlight: null
-    reference_docx: template_tables_portrait.docx
-bibliography: ../references.bib
-csl: ../ecology.csl
+    reference_docx: "../../man/template_tables_portrait.docx"
 ---
@@
 
-# Table RS1
-
-* Estimadores huber resiliences EVI
-
-```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)
-
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
-
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
-
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
-```
-
-# Table RS2
-
-* Estimadores huber resiliences Bai
+**Table S2.** Estimadores huber resiliences BAI
 

### 2018-01-22 man/ms_tables/t_huber_res_evi.Rmd
add tables separately

--- 2017-12-18 man/ms_tables_portrait.Rmd
@@
 
-# Table RS1
-
-* Estimadores huber resiliences EVI
+**Table S1.** Estimadores huber resiliences EVI
 
@@
 # Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
+mraw_a <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
 
@@
 # Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)
+mraw_b <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)
 
@@
 ```
-
-# Table RS2
-
-* Estimadores huber resiliences Bai
-
-```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
-
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
-
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
-
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
-```
-
-```{r, echo = FALSE, warning = FALSE, eval = FALSE}
-
-# Este código es para print la tabla anterior en formato pdf latex (ver state_ms)c
-### OJO REVISAR 
-# ------------------------------
-
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
-
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter))
-
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, c(M.Huber, ci)) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rc_2005_M.Huber, rc_2005_ci, rt_2005_M.Huber, rt_2005_ci, rs_2005_M.Huber, rs_2005_ci,
-                rc_2012_M.Huber, rc_2012_ci, rt_2012_M.Huber, rt_2012_ci, rs_2012_M.Huber, rs_2012_ci,
-                rc_2099_M.Huber, rc_2099_ci, rt_2099_M.Huber, rt_2099_ci, rs_2099_M.Huber, rs_2099_ci)
-
-mmt <- mm %>%
-   rownames_to_column %>% 
-   gather(var, value, -rowname) %>% 
-   spread(rowname, value)
-
-colnames(mmt) <- c("var", "SJ", "caH", "caL", "z_all") 
-mmt <- mmt[-19,] 
-mmt <- mmt %>% 
-  mutate(vars = c(rep("Rc",6), rep("Rt",6), rep("Rs",6)),
-         year = c(rep(
-           c(rep("2005", 2), 
-             rep("2012", 2),
-             rep(" ", 2)), 3))) %>% 
-  dplyr::select(vars, year, SJ, caH, caL, var)
-
-kable(mmt, 
-      format = "latex", booktabs = T) %>% 
-  kable_styling(font_size = 6) %>% 
-  add_header_above(c("", "", "Northern", "Southern" = 2, "")) 
-```

### 2018-01-22 man/ms_tables/t_robust_anovas.Rmd
add tables separately

--- 2017-12-18 man/ms_tables_portrait.Rmd
@@
 
-# Table RS1
+**Table 3.** Robust ANOVAS
 
-* Estimadores huber resiliences EVI
+```{r, echo=FALSE, include = FALSE}
+### Table R1_evi
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
 
-```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)
+dff <- df %>% gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
 
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
+colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rrs\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
 
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
-
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
 ```
 
-# Table RS2
+```{r, echo=FALSE, include = FALSE}
+### Table R1_evi sin rrs 
+df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
 
-* Estimadores huber resiliences Bai
+dff <- df %>% 
+  filter(variables != "rrs") %>% 
+  gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
 
-```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
 
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
+dff_evi <- dff %>% mutate(target = 'EVI')
 
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
+colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
 
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
-
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
 ```
 
-```{r, echo = FALSE, warning = FALSE, eval = FALSE}
+```{r, echo=FALSE, include = FALSE}
 
-# Este código es para print la tabla anterior en formato pdf latex (ver state_ms)c
-### OJO REVISAR 
-# ------------------------------
+### Table R1_bai
+df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
 
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
+dff <- df %>% gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
 
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter))
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
 
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
+colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rrs\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
 
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  mutate(site = "z_all")
+```
 
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
+```{r, echo=FALSE, include = FALSE}
+### Table R1_bai sin rrs 
+df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
+names(df) <- c('factores', 'F', 'p', 'variables')
 
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  mutate(disturb_year = 2099)
+dff <- df %>% 
+  filter(variables != "rrs") %>% 
+  gather(temp, numeros, F, p) %>% 
+  unite(temp1, variables, temp) %>% 
+  spread(temp1, numeros) %>% 
+  mutate(factores = recode(factores, 
+                           "disturb_year" = "Disturb", 
+                           "disturb_year:site" = "Disturb X Site", 
+                           "site" = "Site")) %>% 
+  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
 
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, c(M.Huber, ci)) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
+dff_bai <- dff %>% mutate(target = 'BAI')
 
-mm <- mm %>% 
-  dplyr::select(site, 
-                rc_2005_M.Huber, rc_2005_ci, rt_2005_M.Huber, rt_2005_ci, rs_2005_M.Huber, rs_2005_ci,
-                rc_2012_M.Huber, rc_2012_ci, rt_2012_M.Huber, rt_2012_ci, rs_2012_M.Huber, rs_2012_ci,
-                rc_2099_M.Huber, rc_2099_ci, rt_2099_M.Huber, rt_2099_ci, rs_2099_M.Huber, rs_2099_ci)
+# knitr::kable(dff, 
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
+#       digits=c(0,2,3,2,3,2,3,2,3))
 
-mmt <- mm %>%
-   rownames_to_column %>% 
-   gather(var, value, -rowname) %>% 
-   spread(rowname, value)
+colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dff,
+             round=c(0,2,3,2,3,2,3,2,3),
+             digits=c(0,5,4,5,4,5,4,5,4),
+             keep.line.breaks = TRUE
+             )
+dff
 
-colnames(mmt) <- c("var", "SJ", "caH", "caL", "z_all") 
-mmt <- mmt[-19,] 
-mmt <- mmt %>% 
-  mutate(vars = c(rep("Rc",6), rep("Rt",6), rep("Rs",6)),
-         year = c(rep(
-           c(rep("2005", 2), 
-             rep("2012", 2),
-             rep(" ", 2)), 3))) %>% 
-  dplyr::select(vars, year, SJ, caH, caL, var)
+```
 
-kable(mmt, 
-      format = "latex", booktabs = T) %>% 
-  kable_styling(font_size = 6) %>% 
-  add_header_above(c("", "", "Northern", "Southern" = 2, "")) 
+```{r, echo=FALSE}
+# table 2 (r1)
+dffs <- dff_bai %>% bind_rows(dff_evi) %>% dplyr::select(8,1:7)
+
+# write.csv(dffs, here("/out/anovas_resilience", "trimmed_anovas_joined.csv"), row.names = FALSE)
+
+# kable(dffs, format = "latex", booktabs = T,
+#       col.names = c("", "F", "p", "F", "p", "F", "p", "variable"),
+#       digits=c(0,2,3,2,3,2,3,2,3,0)) %>% 
+#   kable_styling(font_size = 8) %>% 
+#   add_header_above(c("", "rc" = 2, "rs" = 2, "rt" = 2))
+
+dffs <- dffs %>% 
+  mutate_at(c(3,5,7), round, 2) %>%
+  mutate_at(c(4,6,8), round, 3)
+
+colnames(dffs) <- c("variable", "factores", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
+pander(dffs)
 ```

### 2018-01-22 man/ms_tables/t_sampled_plots.Rmd
add tables separately

--- 2017-12-18 man/ms_tables_portrait.Rmd
@@
 
-# Table RS1
+**Table 1.** Characteristics of sampled plot.
+Lat = latitude; Long = longitude.
+Dbh and height of all trees, Basal Area (BA) and Density are computed for all trees within a 10-m radius of focal trees (see methods).
+Values shown here correspond to site averages.
+Standard deviations are shown in parentheses.
 
-* Estimadores huber resiliences EVI
+```{r, echo=FALSE}
+topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
+ages <- read.csv(here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
+dendro_summ <- read.csv(here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
+biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)
 
-```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)
+t1raw <- topo %>% inner_join(ages, by='site') %>% 
+  inner_join(dendro_summ, by='site') %>% 
+  inner_join(biotic, by='site')
 
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
+t1raw <- t1raw %>% mutate(name = case_when(
+  site == 'caH' ~ 'Cáñar-High', 
+  site == 'caL' ~ 'Cáñar-Low',
+  site == 'sj' ~ 'San Juan'))
 
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
+t1 <- t1raw %>% dplyr::select(
+  site, lat_m, long_m, elev_mean, slope_mean, 
+  n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, 
+  dn_mean_all, dn_sd_all, h_mean_all, h_sd_all,
+  ba_mean, ba_sd, stand_density_mean, stand_density_sd) %>% 
+  mutate(pare = ")") %>% 
+  mutate_at(c(2,3,5,18,19), round, 2) %>% 
+  mutate_at(4, round, 0) %>% 
+  mutate_at(c(8:17, 20:21), round, 1) %>% 
+  unite(dn, c(dn_mean, dn_sd), sep=" (") %>% 
+  unite(h, c(h_mean, h_sd), sep=" (") %>% 
+  unite(age, c(age_mean, age_std), sep=" (") %>% 
+  unite(dn_all, c(dn_mean_all, dn_sd_all), sep=" (") %>% 
+  unite(h_all, c(h_mean_all, h_sd_all), sep=" (") %>% 
+  unite(ba, c(ba_mean, ba_sd), sep=" (") %>% 
+  unite(stand_density, c(stand_density_mean, stand_density_sd), sep=" (") %>% 
+  mutate(
+    dn = paste0(dn, pare),
+    h = paste0(h, pare), 
+    age = paste0(age, pare),
+    dn_all = paste0(dn_all, pare),
+    h_all = paste0(h_all, pare),
+    ba = paste0(ba, pare),
+    stand_density = paste0(stand_density, pare)) %>% 
+  dplyr::select(-pare)
 
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
-  mutate(site = "z_all")
+colnames (t1) <- c("Site", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
+                   "# trees", '# cores', "Dbh (cm)", "Height (m)", "Age (years)", 
+                   "Dbh all (cm)", "Height all (m)", "BA (m2/ha)", "Density (trees/ha)")
 
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)
+pander(t1, split.table = Inf)
 
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
-
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
 ```
-
-# Table RS2
-
-* Estimadores huber resiliences Bai
-
-```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
-
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
-
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
-
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
-```
-
-```{r, echo = FALSE, warning = FALSE, eval = FALSE}
-
-# Este código es para print la tabla anterior en formato pdf latex (ver state_ms)c
-### OJO REVISAR 
-# ------------------------------
-
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
-
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter))
-
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, c(M.Huber, ci)) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rc_2005_M.Huber, rc_2005_ci, rt_2005_M.Huber, rt_2005_ci, rs_2005_M.Huber, rs_2005_ci,
-                rc_2012_M.Huber, rc_2012_ci, rt_2012_M.Huber, rt_2012_ci, rs_2012_M.Huber, rs_2012_ci,
-                rc_2099_M.Huber, rc_2099_ci, rt_2099_M.Huber, rt_2099_ci, rs_2099_M.Huber, rs_2099_ci)
-
-mmt <- mm %>%
-   rownames_to_column %>% 
-   gather(var, value, -rowname) %>% 
-   spread(rowname, value)
-
-colnames(mmt) <- c("var", "SJ", "caH", "caL", "z_all") 
-mmt <- mmt[-19,] 
-mmt <- mmt %>% 
-  mutate(vars = c(rep("Rc",6), rep("Rt",6), rep("Rs",6)),
-         year = c(rep(
-           c(rep("2005", 2), 
-             rep("2012", 2),
-             rep(" ", 2)), 3))) %>% 
-  dplyr::select(vars, year, SJ, caH, caL, var)
-
-kable(mmt, 
-      format = "latex", booktabs = T) %>% 
-  kable_styling(font_size = 6) %>% 
-  add_header_above(c("", "", "Northern", "Southern" = 2, "")) 
-```

### 2018-01-23 man/ms_tables/t_robust_anovas.Rmd
formatted robust table

@@
 
-**Table 3.** Robust ANOVAS
+**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the factor drought events (2005 and 2012) and site.
 
@@
   mutate(factores = recode(factores, 
-                           "disturb_year" = "Disturb", 
-                           "disturb_year:site" = "Disturb X Site", 
+                           "disturb_year" = "Drougth event", 
+                           "disturb_year:site" = "Drougth event X Site", 
                            "site" = "Site")) %>% 
-  arrange(match(factores, c("Disturb", "Site", "Disturb X Site")))
+  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site")))
 

### 2018-01-23 man/ms_tables/t_huber_res_bai.Rmd
update tables

@@
 
-**Table S2.** Estimadores huber resiliences BAI
+**Table S2.** Robust measures of central tendency of resilience indices for tree-growth (BAI) groupped by drought events, site and interaction.
+Measures of central tendency are M-estimators based on Huber's Psi (see material and methods).
+95 % confidence intervals using 3000 bootstrap are included in parentheses.
 

### 2018-01-23 man/ms_tables/t_huber_res_evi.Rmd
update tables

@@
 
-**Table S1.** Estimadores huber resiliences EVI
+**Table S1.** Robust measures of central tendency of resilience indices for greenness (EVI) groupped by drought events, site and interaction.
+Measures of central tendency are M-estimators based on Huber's Psi (see material and methods).
+95 % confidence intervals using 3000 bootstrap are included in parentheses.
 

### 2018-01-23 man/ms_tables/t_robust_anovas.Rmd
update tables

@@
 ### Table R1_evi
-df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
-names(df) <- c('factores', 'F', 'p', 'variables')
+evi <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
+names(evi) <- c('factores', 'F', 'p', 'variables')
 
-dff <- df %>% gather(temp, numeros, F, p) %>% 
+dff_evi <- evi %>% gather(temp, numeros, F, p) %>% 
   unite(temp1, variables, temp) %>% 
@@
                            "site" = "Site")) %>% 
-  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site")))
+  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site"))) %>% 
+  mutate_at(c(2,4,6,8), round, 2) %>% 
+  mutate_at(c(3,5,7,9), round, 3) %>% 
+  mutate(variable = 'evi')
 
-colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rrs\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
-pander(dff,
-             round=c(0,2,3,2,3,2,3,2,3),
-             digits=c(0,5,4,5,4,5,4,5,4),
-             keep.line.breaks = TRUE
-             )
-dff
+bai <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
+names(bai) <- c('factores', 'F', 'p', 'variables')
 
-```
-
-```{r, echo=FALSE, include = FALSE}
-### Table R1_evi sin rrs 
-df <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
-names(df) <- c('factores', 'F', 'p', 'variables')
-
-dff <- df %>% 
-  filter(variables != "rrs") %>% 
-  gather(temp, numeros, F, p) %>% 
+dff_bai <- bai %>% gather(temp, numeros, F, p) %>% 
   unite(temp1, variables, temp) %>% 
@@
                            "site" = "Site")) %>% 
-  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site")))
+  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site"))) %>% 
+  mutate_at(c(2,4,6,8), round, 2) %>% 
+  mutate_at(c(3,5,7,9), round, 3) %>% 
+  mutate(variable = 'bai')
 
-# knitr::kable(dff, 
-#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
-#       digits=c(0,2,3,2,3,2,3,2,3))
-
-dff_evi <- dff %>% mutate(target = 'EVI')
-
-colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
-pander(dff,
-             round=c(0,2,3,2,3,2,3,2,3),
-             digits=c(0,5,4,5,4,5,4,5,4),
-             keep.line.breaks = TRUE
-             )
-dff
-
-```
-
-```{r, echo=FALSE, include = FALSE}
-
-### Table R1_bai
-df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
-names(df) <- c('factores', 'F', 'p', 'variables')
-
-dff <- df %>% gather(temp, numeros, F, p) %>% 
-  unite(temp1, variables, temp) %>% 
-  spread(temp1, numeros) %>% 
-  mutate(factores = recode(factores, 
-                           "disturb_year" = "Drougth event", 
-                           "disturb_year:site" = "Drougth event X Site", 
-                           "site" = "Site")) %>% 
-  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site")))
-
-# knitr::kable(dff, 
-#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
-#       digits=c(0,2,3,2,3,2,3,2,3))
-
-colnames(dff) <- c("", "rc\nF", "rc\np", "rrs\nF", "rrs\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
-pander(dff,
-             round=c(0,2,3,2,3,2,3,2,3),
-             digits=c(0,5,4,5,4,5,4,5,4),
-             keep.line.breaks = TRUE
-             )
-dff
-
-```
-
-```{r, echo=FALSE, include = FALSE}
-### Table R1_bai sin rrs 
-df <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
-names(df) <- c('factores', 'F', 'p', 'variables')
-
-dff <- df %>% 
-  filter(variables != "rrs") %>% 
-  gather(temp, numeros, F, p) %>% 
-  unite(temp1, variables, temp) %>% 
-  spread(temp1, numeros) %>% 
-  mutate(factores = recode(factores, 
-                           "disturb_year" = "Drougth event", 
-                           "disturb_year:site" = "Drougth event X Site", 
-                           "site" = "Site")) %>% 
-  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site")))
-
-dff_bai <- dff %>% mutate(target = 'BAI')
-
-# knitr::kable(dff, 
-#       col.names = c("", "F", "p", "F", "p", "F", "p", "F", "p"),
-#       digits=c(0,2,3,2,3,2,3,2,3))
-
-colnames(dff) <- c("", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
-pander(dff,
-             round=c(0,2,3,2,3,2,3,2,3),
-             digits=c(0,5,4,5,4,5,4,5,4),
-             keep.line.breaks = TRUE
-             )
-dff
-
+dff <- dff_evi %>% 
+  bind_rows(dff_bai) %>% 
+  dplyr::select(-rrs_F, -rrs_p) %>% 
+  dplyr::select(variable, factores, rt_F, rt_p, rc_F, rc_p, rs_F, rs_p)
 ```
@@
 ```{r, echo=FALSE}
-# table 2 (r1)
-dffs <- dff_bai %>% bind_rows(dff_evi) %>% dplyr::select(8,1:7)
-
-# write.csv(dffs, here("/out/anovas_resilience", "trimmed_anovas_joined.csv"), row.names = FALSE)
-
-# kable(dffs, format = "latex", booktabs = T,
-#       col.names = c("", "F", "p", "F", "p", "F", "p", "variable"),
-#       digits=c(0,2,3,2,3,2,3,2,3,0)) %>% 
-#   kable_styling(font_size = 8) %>% 
-#   add_header_above(c("", "rc" = 2, "rs" = 2, "rt" = 2))
-
-dffs <- dffs %>% 
-  mutate_at(c(3,5,7), round, 2) %>%
-  mutate_at(c(4,6,8), round, 3)
-
-colnames(dffs) <- c("variable", "factores", "rc\nF", "rc\np", "rs\nF", "rs\np", "rt\nF", "rt\np")
-pander(dffs)
+colnames(dff) <- c("variable", "factores", "rt\nF", "rt\np", "rc\nF", "rc\np", "rs\nF", "rs\np")
+pander(dff)
 ```

### 2018-01-23 man/ms_figures.Rmd
update figures and tables

--- 2018-01-18 man/ms_figures.Rmd
@@
   geom_line(size=.9) + 
-  geom_point(size=3, fill='white', shape=21) +
+  geom_point(size=1.5, fill='white', shape=21) +
   scale_x_date(labels = function(x) format(x, "%b"),
@@
 Figure 2.
-Profile of montly EVI values for *Q.
-pyrenaica* forests during the 2005 (*green*) and 2012 (*blue*) drought events and for the reference period (*gray*) (a).
+Comparison of EVI profile for the reference period (*gray*) and during the 2005 (*green*) and 2012 (*blue*) drought events (a).
 EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations (b).
+Error bars show standard error.
 
-Comparison of EVI profile for the reference period (*grey*) and the 2005 (*green*) and 2012 (*blue*) drought events
+### Figure 3
 
-### Figure R2 = 3 
 ```{r}
 mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
+mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
 
 # NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs')
+mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
+mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
 
 pd <- position_dodge(.2)
+line_lwd <- .8
 
-robust_plot_evi_drought <-
-  ggplot(mhuber_filter, 
-         aes(x = as.factor(disturb_year), y = M.Huber, 
-             color = site, 
-             group = site, 
-             fill = site)) +
-  geom_errorbar(
-    aes(ymin = lower.ci, ymax = upper.ci),
-    width = .1, size = 0.7, position = pd) +
-  geom_line(aes(group = site, color = site, linetype = site), position = pd) +
-  geom_point(shape = 21, size = 2.5, position = pd) +
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = 0.3, size = 5,
-                  segment.size = 0) +
+aes_disturb <- aes(x = disturb_year, y = M.Huber, 
+                   color = site, group = site, 
+                   fill = site, linetype=site,
+                   shape = site, 
+                   ymin = lower.ci, ymax = upper.ci)
+aes_site <- aes(x = site, y = M.Huber, 
+                color = disturb_year, group=disturb_year, 
+                fill=disturb_year, linetype=disturb_year,
+                shape = disturb_year, 
+                ymin = lower.ci, ymax = upper.ci)
+
+theme_res <- list(
+  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
+  geom_line(position = pd, lwd=line_lwd),
+  geom_point(size = 2.5, position = pd), 
   facet_wrap(
-    ~var_sorted, nrow = 1, scales = "free_y",
+    ~var_sorted, nrow = 1, scales="free_y",
     labeller = as_labeller(c(
@@
       "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))) +
-  scale_color_manual(values = c("black", "blue")) +
-  scale_fill_manual(values = c("black", "blue")) + theme_bw() +
-  scale_linetype_manual(values = c("solid", "dashed")) +
+      "2_rs" = "Resilience"))),
+  geom_text_repel(aes(label=Letter), 
+                  nudge_x = - 0.5,
+                  nudge_y = 0.01,
+                  direction="both", 
+                  size = 3.5, show.legend = FALSE,
+                  segment.size = 0.5, 
+                  segment.alpha = 0.5),
+  theme_bw(), 
   theme(
@@
       fill = "white"),
-    legend.position = "bottom") +
-  ylab("") + xlab("Drought event")
+    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
+  ylab("Response value"), 
+  xlab("")
+)
 
-robust_plot_evi_drought 
+theme_evi <-  list(
+  scale_color_manual(values = c("black", "blue")),
+  scale_shape_manual(values=c(19, 22)),
+  scale_fill_manual(values = c("black", "blue")),
+  scale_linetype_manual(values = c("solid", "solid"))
+)
+
+theme_bai <- list(
+  scale_color_manual(values=c('black','blue','blue')), 
+  scale_shape_manual(values=c(19, 22, 15)),
+  scale_fill_manual(values=c('black','white','blue')),  
+  scale_linetype_manual(values=c("solid", "solid", 'solid'))
+)
 ```
+
+```{r}
+# EVI_drought 
+res_evi_drought <-
+  ggplot(mhuber_filter, aes_disturb) + 
+  theme_res + theme_evi + theme(legend.position='top') + 
+  ggtitle("a) EVI")
+
+# BAI_drought 
+res_bai_drought <-
+  ggplot(mhuberbai_filter, aes_disturb) + 
+  theme_res + theme_bai + theme(legend.position='bottom') + 
+  xlab('Drought event') + 
+  ggtitle("b) BAI")
+
+```
+
+```{r, fig.height=7}
+res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
+```
+
+Figure 3.
+Response of northern (*black*) and southern (*blue*) populations of *Q.
+pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; a) and tree growth (BAI; b) for the years 2005 and 2012.
+Different letters above bars indicate significant post hoc differences between groups (see material and methods).
+
+### Figure 4
+```{r}
+theme_de <- list(
+  scale_color_manual(values=c('black','red')), 
+  scale_shape_manual(values=c(19, 15)),
+  scale_fill_manual(values=c('black','red')),  
+  scale_linetype_manual(values=c("solid", "solid")),
+  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
+
+)
+
+# EVI_site
+res_evi_site <-
+  ggplot(mhuber_filter, aes_site) + 
+  theme_res + theme_de + theme(legend.position='top') + 
+  ggtitle("a) EVI")
+
+# BAI_site
+res_bai_site <-
+  ggplot(mhuberbai_filter, aes_site) + 
+  theme_res + theme_de + theme(legend.position='bottom') + 
+  xlab('Site') + 
+  ggtitle("b) BAI") 
+```
+
+```{r}
+res_evi_site  + res_bai_site + plot_layout(ncol=1)
+```
+
+Figure 4.
+Response of *Q.
+pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree growth (BAI; b) by site.
+Different letters above bars indicate significant post hoc differences between groups (see material and methods).

### 2018-01-23 analysis/analysis_KW_stand_features.Rmd
include KW

---
title: "Comparison ANOVAS stand features"
author: "AJ Perez-Luque (@ajpelu)"
date: "2018 Jan"
output:  
  md_document:
    variant: markdown_github
bibliography: references.bib 
csl: ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library("tidyverse")
library("raster")
library("stringr")
library('rgdal')
library("sp")
library("knitr")
library("broom")
library("multcomp")
library("pander")
library("devtools")
source_gist("https://gist.github.com/ajpelu/194e721077ec045a2b864088908e7aca", filename = 'table_glht.r') # multicomp table
```

# Prepare Data
```{r, echo=FALSE}

machine <- 'ajpelu'
# machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# Read tree data
tree <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
```

```{r}
# Compute diameter (mm)
tree <- tree %>% 
  mutate(dn_mm = (perim_mm / pi))

sj_code  <- paste0('A', str_pad(1:20, 2, pad='0'))
ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')), paste0('B', 26:30))
ca_highcode <- paste0('B', 11:25)

tree <- tree %>% mutate(
  site = as.factor(case_when(
    id_focal %in% sj_code ~ "sj", 
    id_focal %in% ca_highcode ~ "caH", 
    id_focal %in% ca_lowcode ~ "caL")))

# Get only focal trees  
ft <- tree %>% 
  filter(sp=='Focal') %>% 
  filter(id_focal!='Fresno')

# Get only no focal trees 
nft <- tree %>% 
  filter(sp!='Focal')

# Select only variables to compare (ft)
ft_sel <- ft %>%
  mutate(dn = dn_mm / 10, 
         h = height_cm / 100) %>% 
  dplyr::select(dn, h, site)

# Select only variable to compare (all trees)
all <- nft %>% mutate(dn = dn_mm / 10, 
                      h = height_cm / 100) %>% 
  dplyr::select(dn, h, site) %>% 
  bind_rows(ft_sel)

```

# Competition data

* Read data 
```{r, echo=FALSE}
# Read competition
compe <- read.csv(file=paste(di, "/data/competence/competence_indices.csv", sep=""), header=TRUE, sep=',')

compe <- compe %>% 
  mutate(site = as.factor(case_when(
    id_focal %in% sj_code ~ "sj", 
    id_focal %in% ca_highcode ~ "caH", 
    id_focal %in% ca_lowcode ~ "caL")))

compe_sel <- compe %>% dplyr::select(ba, std, site)
```

# Kruskal-Wallis comparison

```{r}
comparaKW <- function(df, mivariable){ 
  require(PMCMR)
  require(multcompView)

  output <- list()

  # Model formulation
  myformula <- as.formula(paste0(mivariable, " ~ site"))

  # Kruskal Wallis  
  kt <- kruskal.test(myformula, data = df)

  # Summary AOV (broom style)
  tm <- broom::tidy(kt)
  tm$mi_variable <- mivariable

  # Dunn's test 
  object <- posthoc.kruskal.dunn.test(myformula, data=df, p.adjust.method = "bonferroni")

  # Get dataframe with letters and pvalues (#from summary.PMCMR)
  pval <- as.numeric(object$p.value)
  stat <- as.numeric(object$statistic)
  grp1 <- as.numeric(c(col(object$p.value)))
  cnam <- colnames(object$p.value)
  grp2 <- as.numeric(c(row(object$p.value)))
  rnam <- rownames(object$p.value)
  H0 <- paste(cnam[grp1], " = ", rnam[grp2])
  OK <- !is.na(pval)
  xdf <- data.frame(H0 = H0[OK], statistic = stat[OK], p.value = format.pval(pval[OK], 5))

  # Get letters (using multcompView) See viggnete PMCMR 
  dt_letters  <- multcompView::multcompLetters(get.pvalues(object), threshold = 0.05)

  output$KW <- tm
  output$post_hoc <- xdf
  output$letters <- dt_letters
  return(output)
  }
```

```{r, warning=FALSE}
# Focal Tree 
kw_ft_dn <- comparaKW(ft_sel, 'dn')
kw_ft_h <- comparaKW(ft_sel, 'h')

# All trees 
kw_all_dn <- comparaKW(all, 'dn') 
kw_all_h <- comparaKW(all, 'h')

# Competence 
kw_ba <- comparaKW(compe_sel, 'ba')
kw_std <- comparaKW(compe_sel, 'std')
```

# Export data to text files (see `/out/kw_stand_features/`)

```{r}

variables <- c('ft_dn', 'ft_h', 'all_dn', 'all_h', 'ba', 'std')

for (i in variables){

  out <- get(paste0('kw_', i))
  file_out <- file(paste0(di,'out/kw_stand_features/kw_', i, '.txt'), "w")
  sink(file_out)
  cat(paste0("VARIABLE = ",i," \n"))
  cat("\n")
  cat("KRUSKAL WALLIS \n")
  print(out$KW) 
  cat("\n")

  cat("POST HOC \n")
  print(out$post_hoc) 
  cat("\n")

  cat("Letters \n")
  print(out$letters)
  cat("\n")

  while (sink.number()>0) sink()
  # close(file_out)
}

while (sink.number()>0) sink()
```

```{r}

variables <- c('ft_dn', 'ft_h', 'all_dn', 'all_h', 'ba', 'std') 
output_letras <- c()

for (i in variables){

  out <- get(paste0('kw_', i))
  l <- as.data.frame(out$letters$Letters)
  l$site <- row.names(l)
  colnames(l) <- c('letter', 'site')
  l$mivariable <- i

  output_letras <- rbind(output_letras, l)
}

# Export data 
write.csv(output_letras, file=paste(di, "out/kw_stand_features/kw_letras.csv", sep=""), row.names = FALSE)
```

### 2018-01-23 man/ms_1_material_methods.Rmd
update material and meethods

@@
 Sampling was carried during autumn of 2016.
-Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern site) and Cáñar (CA; southern site) (Figure 1; Table 1).
+Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern aspect) and Cáñar (CA; southern aspect) (Figure 1; Table 1).
 For the southern site two elevations were sampled: CA-Low and CA-High.
-All the sites were oak monospecific and representatives of two of the three the population's cluster identified for the specie in this mountain range [@PerezLuque2015].
+All the sites were oak monospecific and representatives of two of the three population clusters identified for the species in this mountain range [@PerezLuque2015].
 In each site between 15 and 20 dominant trees were randomly selected.
-Two cores of 5 mm of diameter were taken from each at breast heigth (1.3 m) using an increment borer.
+Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer.
 Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively.
-We assessed stand competition of the target tree by recording all neighboring living trees with dbh > 7.5 cm in a circle of 10 m radi.
-The species, dbh and total height were annotated.
-We also measured distance and azimuth to the target tree.
-(:red_circle: indices competencia??)
+We assessed stand competition of target trees by recording distance, azimuth, dbh, species and total height of all neighboring living trees with dbh > 7.5 cm in a circular plot of 10 m radius.
+We used non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] for search differences between sites for height, DBH and competence indices.
+When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values.
+These analysis were performed uisng the `PMCMR` package [@Pohlert2014].
 
@@
 Tree cores were air dried, glued onto wooden mounts and sanded.
-Annual radial growth (ring width, RW) were measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.01 mm.
-Individual ring series were visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
+Annual radial growth (ring width, RW) was measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.001 mm.
+Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
 Validation of the cross-dating was done using COFECHA software [@Holmes1983].
@@
 The growth trends were analyzed at different time scales.
-To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-whitened residual chronologies calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976].
+To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-whitened residual chronologies (RWI) calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976].
 Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008; @Bunn2010] package.
 Residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990].
-The statistical quality of each chronology was checked via expressed population signal [@Wigley1984].
+The statistical quality of each chronology was checked via the expressed population signal [@Wigley1984].
 A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.
@@
 BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference [@Biondi2008].
-We used measured dbh and raw nondetrened ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
+We used measured dbh and raw ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
 We used the following equation:
@@
 We calculated a mean BAI serie for each individual tree.
-Site BAI chronologies were obtained by averaging individual tree BAI time series.
+Mean site BAI chronologies were obtained by averaging individual tree BAI time series.
 
+Both RWI and BAI chronologies were compared to different climatic indices to analyse the tree low and the high-frequency response to climate.
 Statistical descriptive parameters [@Fritts1976] including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology.
-
-una frase conectora 
-We also explored the climate-growth relationships and the disturbance  ...
 
@@
 We selected grid cells covering each sampled sites.
-Data were downloaded and preprocess using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018].
+Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018].
 We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014].
 SPEI values for the period 1961-2014 were obtained with a spatial resolution of 1.1 km from the Drought indices dataset for Spain database (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017].
-We select a temporal scale of 6 months.
+We selected a temporal scale for the SPEI of 6 months.
 
@@
 ### Disturbance analyses 
-Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (releases) or decreases (supressions) in radial growth [@Nowacki1997].
-By using a temporal window of reasonable length, this approach filters out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].
+Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release event from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g.
+logging) in the past.
+By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].
 
@@
 where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median.
-We used medians since they are mores robust estimator of central tendency than means [@Rubino2004; @Camarero2011].
+We used medians since they are more robust estimator of central tendency than means [@Rubino2004; @Camarero2011].
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
-A minimum $GC$ of 25% was established to depict canopy disturbances, and we considered a stand-wise disturbance when more than 50% of the individual trees display growth changes above 25 %.
+To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 25 % (DUDA) of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
 
@@
 
-To evaluate the effects of the drougth events on greeennes and tree growth (BAI) we used three resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
+To evaluate the effects of drougth events on greeennes and tree growth (BAI) we used three resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
 The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred.
-It is estimated as the ratio between the performance during and befor the disturbance:
+It is estimated as the ratio between the performance during and before the disturbance (drought) event:
 
@@
 The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (DUDA: incluir tabla de coeficientes y/o gráfica??
-como suplement, see Gazol 2017).
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3) and this length was used in other studies [Gazol2017].
 
@@
 
-We tested for significant differences between drought events (2005 and 2012) and oak population (northern and southern slopes) for each of the resilience indices.
-Robust two-way ANOVAs were used beacuse original and log-transformed data both did not match the assumptions of normality and homogeneity of variance [@Wilcox2012].
-Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to mean value in all cases [@Wilcox2012].
+We used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and oak population (northern and southern slopes) for each of the three resilience indices studied.
+These test were used beacuse original and log-transformed data both did not match the assumptions of normality and homogeneity of variance [@Wilcox2012].
+Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012].
 When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012].
@@
 All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages.
-The level of significance was set at 0.05 and adjusted for multiple comparisons.
-All analyses were carried out in R software [@R2017] using several packages (see Supplementary S)
+The level of significance was set to 0.05 and adjusted for multiple comparisons.
+All analyses were carried out in R software [@R2017] using several packages (see Supplementary S?)

### 2018-01-23 man/ms_2_results.Rmd
update results

@@
 
-Trends analysis showed that 78.9 % of the pixels of *Q.
-pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
-The strongest trends were observed in southwestern populations (Figura S2).
-Vegetation greenness of *Quercus pyrenaica* forests were lower during the 2005 and 2012 year than the greenness observed for the reference period (Figure 2a).
+Trends analysis showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
+The strongest trends were observed in southwestern populations (Figure S1).
+Vegetation greenness of *Q.
+pyrenaica* forests were lower during both the 2005 and 2012 years than the greenness observed for the reference period (Figure 2a).
 The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b).
@@
 According to the standardized anomalies, *Q.
-pyrenaica* forests suffered a browning during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), however no changes were observed for the 2012 drought (Figure S3).
+pyrenaica* forests suffered a browning during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), however no changes were observed for the 2012 drought (Figure S2).
 
@@
 *Q.
-pyrenaica* forest showed significantly lower resistance to 2005 drought event than to 2012 one [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); table 2; Figure 3a, F = 799.86, p <0.0001].
-The 2005 drought reduced the greenness of oak to 85.8 % while the 2012 reduced 94.3 % (tabla rs1 quizás, duda en esto).
-Southern populations showed significantly higher values of resistance to drought than northern ones, except for 2012 where non-significant differences were recorded (Figure 3a).
+pyrenaica* forest showed significantly lower resistance to 2005 drought event than to 2012 one (Table 3; Figure 3a).
+The 2005 drought reduced the greenness of oak to 85.8 % while the 2012 reduced 94.3 % (Table S1).
+Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1.
+Figures 3a, 4a).
 
@@
 
-The oak forests recovered their greenness more rapidly after the 2005 drought than after 2012.
-In the period after 2005 drought, greeness achieved was 112 % (Rc = 1.12) and after 2012 105.7 % (Rc = 1.057).
-A similar recovery after the 2005 and 2012 drought event was observed for southern populations (p = 0.2453; Figure 3a), whilst the northern populations showed a significantly greater recovery after the 2005 drought than after the 2012 drought.
+The oak forests recovered their greenness significantly more rapidly after the 2005 drought than after 2012 (Tables 3, S1).
+In the period after 2005 drought, greeness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
+A similar recovery after the 2005 and 2012 drought event was observed for southern populations (p = 0.2453; Figure 4a), whilst the northern populations showed a significantly greater recovery after the 2005 drought than after the 2012 drought (Figure 4a).
 
@@
 
-Resilience values were significantly higher for the 2012 drought event than for 2005, although both values were close to 1 indicating that greenness level was rather similar after each disturbance event.
-The southern populations showed higher resilience values than the northern ones, although they were not significantly different for 2005 drought (p = 0.036).
+Resilience values were significantly higher for the 2012 drought event than for 2005 (Tables 3, S1), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
+The southern populations showed higher resilience values than the northern ones, although they were not significantly different for 2005 drought (p = 0.036; Figures 3a, 4a).
 
@@
 ```
-
-Figura R2. Interaction plot resilience metrics
 
@@
 
-Although no significant differences were observed in the resistance of the oak growth to both drought events (F = 6.0189; p = 0.019; table 2; Figure 3b), the 2005 drought reduced growth more than 2012 did (Rt = 0.721 and 0.819 respectively).
-Northern population showed resistance values lower than southern ones, especially for 2005 drought event where the growth was reduced to 44.5 % of the preceding period.
+Although no significant differences were observed in the resistance of the oak growth to both drought events (Table 3; Figure 3b), the 2005 drought reduced growth more than 2012 did (Rt = 0.721 and 0.819 respectively) (Table S2).
+Northern population showed resistance values lower than southern ones, especially for 2005 drought event where the growth was reduced to 44.5 % of the preceding period (Figures 3b, 4b).
 
@@
 
-* No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012.
+* No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012 (F = 6.0189; p = 0.019).
 Valores de resistencia menor al evento de 2005.
@@
 
-Oak forests recovered their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1.
-Norhtern populations showed signigicantly higher values of recovery than southern ones, which even not reach recovery values of Rc = 1.
+Oak forests recovered their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1 (Table S2).
+Norhtern populations showed signigicantly higher values of recovery than southern ones (Tables 3, S2), which even not reach recovery values of Rc = 1 (Figure 3b, 4b).
 
@@
 
-Significantly higher values of resilience of growth were observerd for the 2012 drought event than for the 2005 one.
-We found no differences of resilience values for growth between northern and southern populations (F = 1.31; p = 0.404).
-For 2005 drought event the resilience of northern population was lower than southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity.
+Significantly higher values of resilience of growth were observerd for the 2012 drought event than for the 2005 one (Table S2; Figure 4b).
+We found no differences of resilience values for growth between northern and southern populations (Table 3).
+For 2005 drought event the resilience of northern population was lower than southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity (Figure 3b).
+
+# Chronology
+
+The trees of the southern population were older and larger than northern ones, particularly for the highest elevation site (Ca-High) which had bigger trees than the other sites (Tables 1, 2).
+The basal area increment (BAI) were different between sites (Figure 5).
+El sitio mas elevado (Ca-High) mostró mayores valores de BAI que los otros sitios, mostrando además una tendencia positiva a partir de 1970.
+
+Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5).
+La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona.
+Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.
+
+and higher values of tree-growth (Table 2).
+
+El sitio caH mostró además los mayores valores de AR(1) y de MS
+
+El sitio del norte, mostró mayores valores de MS (Table 2). ¿esto sugiere que este sitio tiene una alta limitación climática?
+
+Different growth trends were observed for each site (Figure 5). Ca-High (the highest elevation site) had higher values of tree growth than the other sites.
 

### 2018-01-23 man/ms_figures.Rmd
update results

@@
 Different letters above bars indicate significant post hoc differences between groups (see material and methods).
+
+### Figure 5
+
+![](../analysis/chronos/analysis_chronologies_files/figure-markdown_github/crono_compara_sitesSJCALH-1.png)
+
+Figure 5.
+Basal Area Increment (BAI) chronologies of *Q.
+pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA_Low; *pink*) and high-elevation (CA_High, *purple*) sites.
+Shading areas coorespond to standard error of the mean.

### 2018-01-23 man/ms_tables/t_sampled_plots.Rmd
update results

@@
 Standard deviations are shown in parentheses.
+Different letters indicate statistically significant differences between sites (Kruskal-Wallis test followed by Dunn’s test, p<0.05).
 
@@
 biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)
+letritas <- read.csv(here("/out/kw_stand_features", "kw_letras.csv"), header = TRUE)
+
+letritas <- letritas %>% 
+  mutate(letter = as.character(letter)) %>% 
+  mutate(mivar = paste0(mivariable, '_l')) %>%
+  dplyr::select(-mivariable) %>% 
+  spread(key=mivar, letter)
 
@@
   inner_join(dendro_summ, by='site') %>% 
-  inner_join(biotic, by='site')
+  inner_join(biotic, by='site') %>% 
+  inner_join(letritas, by='site')
 
@@
   dn_mean_all, dn_sd_all, h_mean_all, h_sd_all,
-  ba_mean, ba_sd, stand_density_mean, stand_density_sd) %>% 
+  ba_mean, ba_sd, stand_density_mean, stand_density_sd, 
+  all_dn_l, all_h_l, ba_l, ft_dn_l, ft_h_l, std_l
+  ) %>% 
   mutate(pare = ")") %>% 
@@
   mutate(
-    dn = paste0(dn, pare),
-    h = paste0(h, pare), 
+    dn = paste(paste0(dn, pare), ft_dn_l, sep=" "),
+    h = paste(paste0(h, pare), ft_h_l, sep = ' '),
     age = paste0(age, pare),
-    dn_all = paste0(dn_all, pare),
-    h_all = paste0(h_all, pare),
-    ba = paste0(ba, pare),
-    stand_density = paste0(stand_density, pare)) %>% 
-  dplyr::select(-pare)
+    dn_all = paste(paste0(dn_all, pare), all_dn_l, sep = ' '),
+    h_all = paste(paste0(h_all, pare), all_h_l, sep = ' '),
+    ba = paste(paste0(ba, pare), ba_l, sep = ' '),
+    stand_density = paste(paste0(stand_density, pare), std_l, sep = ' ')) %>%
+  dplyr::select(-pare, -all_dn_l, -all_h_l, -ba_l, -ft_dn_l, -ft_h_l, -std_l)
 

### 2018-01-24 analysis/analysis_stand_features.Rmd
update distance-dependet indices

@@
   group_by(site) %>% 
-  dplyr::select(ba_ha, std, site) %>% 
+  dplyr::select(ba_ha, std, srd, site) %>% 
   dplyr::summarise(ba_mean = mean(ba_ha), 
@@
                    stand_density_mean = mean(std), 
-                   stand_density_sd = sd(std)) %>% 
+                   stand_density_sd = sd(std), 
+                   srd_mean = mean(srd),
+                   srd_sd = sd(srd)) %>% 
   as.data.frame()

### 2018-01-24 man/ms_tables/t_sampled_plots.Rmd
update sampled plots

@@
 Lat = latitude; Long = longitude.
-Dbh and height of all trees, Basal Area (BA) and Density are computed for all trees within a 10-m radius of focal trees (see methods).
+Dbh and height of all trees, Basal Area (BA), Density and SRD (Size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see methods).
 Values shown here correspond to site averages.
@@
   dn_mean_all, dn_sd_all, h_mean_all, h_sd_all,
-  ba_mean, ba_sd, stand_density_mean, stand_density_sd, 
-  all_dn_l, all_h_l, ba_l, ft_dn_l, ft_h_l, std_l
+  ba_mean, ba_sd, stand_density_mean, stand_density_sd, srd_mean, srd_sd,
+  all_dn_l, all_h_l, ba_l, ft_dn_l, ft_h_l, std_l, srd_l
   ) %>% 
   mutate(pare = ")") %>% 
-  mutate_at(c(2,3,5,18,19), round, 2) %>% 
+  mutate_at(c(2,3,5,18,19, 22:23), round, 2) %>% 
   mutate_at(4, round, 0) %>% 
   mutate_at(c(8:17, 20:21), round, 1) %>% 
+  unite(n_trees, c(n_trees, n_radii), sep=" (") %>% 
   unite(dn, c(dn_mean, dn_sd), sep=" (") %>% 
@@
   unite(stand_density, c(stand_density_mean, stand_density_sd), sep=" (") %>% 
+  unite(srd, c(srd_mean, srd_sd), sep=" (") %>% 
   mutate(
+    n_trees = paste0(n_trees, pare),
     dn = paste(paste0(dn, pare), ft_dn_l, sep=" "),
@@
     ba = paste(paste0(ba, pare), ba_l, sep = ' '),
-    stand_density = paste(paste0(stand_density, pare), std_l, sep = ' ')) %>%
-  dplyr::select(-pare, -all_dn_l, -all_h_l, -ba_l, -ft_dn_l, -ft_h_l, -std_l)
+    stand_density = paste(paste0(stand_density, pare), std_l, sep = ' '),
+    srd = paste(paste0(srd, pare), srd_l, sep = ' ')) %>%
+  dplyr::select(-pare, -all_dn_l, -all_h_l, -ba_l, -ft_dn_l, -ft_h_l, -std_l, -srd_l)
 
 colnames (t1) <- c("Site", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
-                   "# trees", '# cores', "Dbh (cm)", "Height (m)", "Age (years)", 
-                   "Dbh all (cm)", "Height all (m)", "BA (m2/ha)", "Density (trees/ha)")
+                   "# trees (# cores)", "Dbh (cm)", "Height (m)", "Age (years)", 
+                   "Dbh all (cm)", "Height all (m)", "BA (m2/ha)", "Density (trees/ha)", "SRD")
 

### 2018-01-24 analysis/analysis_KW_stand_features.Rmd
update kw

@@
 
-compe_sel <- compe %>% dplyr::select(ba, std, site)
+# Seleccionamos: 
+# * distance independet: BA, Stand Densisty (std) 
+# * distance dependent: Site ratio proportional to distance (srd); as in Gea2014 Foreco 320:70 & Lorimer
+
+compe_sel <- compe %>% dplyr::select(ba, std, srd, pd, site)
 ```
@@
 kw_std <- comparaKW(compe_sel, 'std')
+kw_srd <- comparaKW(compe_sel, 'srd')
+kw_pd <- comparaKW(compe_sel, 'pd')
 ```
@@
 
-variables <- c('ft_dn', 'ft_h', 'all_dn', 'all_h', 'ba', 'std')
+variables <- c('ft_dn', 'ft_h', 'all_dn', 'all_h', 'ba', 'std', 'srd', 'pd')
 
@@
 
-variables <- c('ft_dn', 'ft_h', 'all_dn', 'all_h', 'ba', 'std') 
+variables <- c('ft_dn', 'ft_h', 'all_dn', 'all_h', 'ba', 'std', 'srd', 'pd')
 output_letras <- c()

### 2018-01-24 analysis/rwi/climate_rwi_relate.Rmd
update

@@
 
+# SELECTED MONTHS
+
+* Tras comentar con GGea, decidimos, poner desde sep-1 hasta oct1 
+* quitar tmean
+
+```{r}
+set.seed(3333)
+rprec_SJ <- dcc(chrono = cro_sj, climate = prec_N, 
+         method = "correlation",  boot = "std",
+        .range(-9:10) + .sum(-9:8))
+
+rprec_caL <- dcc(chrono = cro_caL, climate = prec_S, 
+         method = "correlation",  boot = "std",
+        .range(-9:10) + .sum(-9:8))
+
+rprec_caH <- dcc(chrono = cro_caH, climate = prec_S, 
+         method = "correlation",  boot = "std",
+        .range(-9:10) + .sum(-9:8))
+
+# Joins dccs 
+rprecs <- c('rprec_SJ', 'rprec_caH', 'rprec_caL')
+rprec_all <- join_dccs(rprecs)
+
+# Add custom names for aggregation variables
+
+rprec_all <- rprec_all %>% 
+  mutate(month_name = case_when(
+      month == "sep...AUG" ~ "Hydrol", 
+      TRUE ~ as.character(.$month)))
+```
+
+## Plot precipitation 
+```{r rwi_clim_prec_selec}
+p_prec <- rprec_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rprec_all$id, labels = rprec_all$month_name) + 
+  theme(legend.position = "none") + 
+  annotate("text", label = "Prec", x = 1, y= -0.4) 
+p_prec
+
+pdf(paste0(di, '/out/climate_rwi/prec_selec.pdf'), width=9, height = 5)
+p_prec
+dev.off()
+```
+
+# SPEI High-resol
+
+```{r}
+set.seed(3333)
+rsequia_SJ <- dcc(chrono = cro_sj, climate = sequia_sj, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rsequia_caL <- dcc(chrono = cro_caL, climate = sequia_caL, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rsequia_caH <- dcc(chrono = cro_caH, climate = sequia_caH, 
+         method = "correlation",  boot = "std", var_names = 'spei6',
+        .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+# Joins dccs 
+rsequias <- c('rsequia_SJ', 'rsequia_caH', 'rsequia_caL')
+rsequia_all <- join_dccs(rsequias)
+
+rsequia_all <- customNames(rsequia_all) 
+```
+
+## Plot SPEI HR
+```{r rwi_clim_sequia_selec}
+p_sequia <- rsequia_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rsequia_all$id, labels = rsequia_all$month_name) + 
+  theme(legend.position = "none") + 
+  annotate("text", label = "SPEI 6 - HR", x = 1.5, y= -0.4) 
+p_sequia
+
+pdf(paste0(di, '/out/climate_rwi/sequia_selec.pdf'), width=9, height = 5)
+p_sequia
+dev.off()
+```
+
+# Tmin
+
+```{r}
+set.seed(3333)
+rtmin_SJ <- dcc(chrono = cro_sj, climate = tmin_N, 
+         method = "correlation",  boot = "std", 
+        .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rtmin_caH <- dcc(chrono = cro_caH, climate = tmin_S, 
+         method = "correlation",  boot = "std", 
+        .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rtmin_caL <- dcc(chrono = cro_caL, climate = tmin_S, 
+         method = "correlation",  boot = "std", 
+        .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+# Join dccs 
+rtmins <- c('rtmin_SJ', 'rtmin_caH', 'rtmin_caL')
+rtmin_all <- join_dccs(rtmins)
+
+# Rename variable 
+rtmin_all <- customNames(rtmin_all) 
+```
+
+## Plot Tmin 
+```{r rwi_cli_tmin_selec}
+p_tmin <- rtmin_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rtmin_all$id, labels = rtmin_all$month_name) + 
+  theme(legend.position = c(.1, .8)) +
+  annotate("text", label = "Tmin", x = 1, y= -0.4) + ylim(-.4, .4)
+p_tmin
+
+pdf(paste0(di, '/out/climate_rwi/tmin_selec.pdf'), width=9, height = 5)
+p_tmin
+dev.off()
+```
+
+# Tmax
+
+```{r}
+set.seed(3333)
+rtmax_SJ <- dcc(chrono = cro_sj, climate = tmax_N, 
+         method = "correlation",  boot = "std", 
+        .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rtmax_caH <- dcc(chrono = cro_caH, climate = tmax_S, 
+         method = "correlation",  boot = "std", 
+         .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+rtmax_caL <- dcc(chrono = cro_caL, climate = tmax_S, 
+         method = "correlation",  boot = "std", 
+         .range(-9:10) + .mean(-9:8) + .mean(-12:2) + .mean(3:5) + .mean(6:8) + .mean(9:11))
+
+# Join dccs 
+rtmaxs <- c('rtmax_SJ', 'rtmax_caH', 'rtmax_caL')
+rtmax_all <- join_dccs(rtmaxs)
+
+# Rename variable 
+rtmax_all <- customNames(rtmax_all) 
+```
+
+## Plot Tmax 
+```{r rwi_cli_tmax_selec}
+p_tmax <- rtmax_all %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  custom_gg +
+  scale_x_continuous(breaks = rtmin_all$id, labels = rtmin_all$month_name) +
+  theme(legend.position = "none") +
+  annotate("text", label = "Tmax", x = 1, y= -0.4) + ylim(-.4, .4)
+p_tmax
+
+pdf(paste0(di, '/out/climate_rwi/tmax_selec.pdf'), width=9, height = 5)
+p_tmax
+dev.off()
+```
+
+```{r rwi_cli_relation}
+p <- grid.arrange(p_tmin, p_tmax, p_prec, p_sequia, ncol=2)
+pdf(paste0(di, '/out/climate_rwi/rwi_cli_relation.pdf'), width=9, height = 12)
+p
+dev.off()
+```
+
 # Correlations RUNNING

### 2018-01-24 man/ms_1_material_methods.Rmd
update

@@
 We assessed stand competition of target trees by recording distance, azimuth, dbh, species and total height of all neighboring living trees with dbh > 7.5 cm in a circular plot of 10 m radius.
+We computed two distance independent indices: density in $\mathrm{trees \cdot ha^{-1}}$, and basal area (BA) in $\mathrm{m^{2} \cdot ha^{-1}}$; and one distance dependent indices: size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
 We used non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] for search differences between sites for height, DBH and competence indices.
@@
 Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008].
-Monthly precipitation and mean, minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period.
+Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period.
 We selected grid cells covering each sampled sites.

### 2018-01-24 man/ms_2_results.Rmd
update

@@
 
-The trees of the southern population were older and larger than northern ones, particularly for the highest elevation site (Ca-High) which had bigger trees than the other sites (Tables 1, 2).
+The trees of the southern population were older and larger than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2).
+Competence indices were similar among sites (Table 1).
 The basal area increment (BAI) were different between sites (Figure 5).
 El sitio mas elevado (Ca-High) mostró mayores valores de BAI que los otros sitios, mostrando además una tendencia positiva a partir de 1970.
+Se observan diferencias en cuanto al crecimiento entre los sitios de las poblaciones del sur.
+Ca_Low se parece en crecimiento mas a SJ que a Ca_High.
 

### 2018-01-24 man/ms_3_discussion.Rmd
update

@@
 
+## Dendro
+
+### Diferencias crecimiento entre sitios. 
+Mayor crecimiento en los robles situados en CaHigh.
+Sorprende que en el rear edge de la distribución de la especie, en la cara sur dentro de este rear edge, y en la parte mas alta (estas parcelas están en el treeline de la especie en SN) es donde encontramos mayores crecimientos.
+Parece que los robles estuvieran mejor en esta zona.
+Algunas reflexiones
+
+* Quizá están creciendo donde les están dejando (menor impacto antrópico)
+* ¿Menor estrés hídrico en zonas mas altas? 
+* ¿Diferencias entre suelos?
+(ver artículo [@CoboDiaz2017] que compara suelos de Cáñar).
+En este estudio sobre microbiota del suelo, analizó tres sitios: por encima del treline (XZF), low altitudinal oak forest (LAF) and high altitudinal oak forest (HAF).
+LAF está mucho mas bajo, pero HAF está cerca de las parcelas caHigh y encontró lo siguiente:
+
+  * tipo de suelo: LAF es Sandy-loam, HAF es loam
+  * HAF tuvo los valores mayores de disponibilidad hídrica (% available water tabla s1 en [@CoboDiaz2017]) 
+  * Los tres sitios eran pobres en materia orgánica, pero el HAF dobló los valores de los otros dos sitios
+  * No diferencias en contenido de N, C/N ratio similares en HAF, y LAF
+
+Sorprende que en Cáñar haya diferencias tan grandes de crecimiento estándo tan cerca. ¿Puede estar relacionado con intensidad de uso antrópico? De hecho la Ca-Low tiene crecimientos similares a SJ.
+
 ## Relaciones crecimiento-clima en otras especies

### 2018-01-24 man/ms_figures.Rmd
add figure 5

@@
 Shading areas coorespond to standard error of the mean.
+
+```{r, warning=FALSE}
+cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
+
+cronos_bai <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+  theme_bw() + ylab('BAI') + 
+  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
+                  ymax = (bai_mean + bai_se)/100,
+                  fill=site), alpha=.2, colour=NA) +
+  geom_line() + 
+  geom_point(size=.8) + 
+  theme(panel.grid = element_blank(),
+        legend.position = c(0.1,.8),
+        strip.background = element_blank()) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
+  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
+  # geom_line(aes(x=year, y=n_trees+40, colour=site))
+
+samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
+  theme_bw() + ylab('# trees') + 
+  geom_line() + 
+  theme(panel.grid = element_blank(),
+        legend.position = "none", 
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(), 
+        plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")) + 
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
+
+  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
+  #       axis.title.y = element_text(vjust =0.25))
+
+plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.1, 0.9))
+
+plot_bai_s
+```

### 2018-01-24 analysis/chronos/analysis_splines.Rmd
update figures

@@
 # Export data 
-write.csv(cro_rwi, file=paste(di, "data/cronos_medias/cronos_sites_rwi.csv", sep=""), row.names = FALSE)
+# write.csv(cro_rwi, file=paste(di, "data/cronos_medias/cronos_sites_rwi.csv", sep=""), row.names = FALSE)
 ```
@@
     site == "caL" ~ "2_caL"))
+
+# Export data 
+write.csv(cro_rwi, file=paste(di, "data/cronos_medias/cronos_sites_rwi.csv", sep=""), row.names = FALSE)
+write.csv(year_cutoffs, file=paste(di, "data/cronos_medias/cronos_sites_rwi_year_cutoffs.csv", sep=""), row.names = FALSE)
+write.csv(year_sampledepth, file=paste(di, "data/cronos_medias/cronos_sites_rwi_year_sampledepth.csv", sep=""), row.names = FALSE)
 ```

### 2018-01-24 man/ms_figures.Rmd
update figures

@@
 
-![](../analysis/chronos/analysis_chronologies_files/figure-markdown_github/crono_compara_sitesSJCALH-1.png)
-
-Figure 5.
-Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA_Low; *pink*) and high-elevation (CA_High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-
 ```{r, warning=FALSE}
@@
 ```
+
+Figure 5.
+Basal Area Increment (BAI) chronologies of *Q.
+pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA_Low; *pink*) and high-elevation (CA_High, *purple*) sites.
+Shading areas coorespond to standard error of the mean.
+Number of series are displayed in the upper plot.
+
+# Figure 6
+
+```{r}
+cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
+year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
+year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
+
+label_site <- c("0_sj" = "Northern (SJ)",
+                "1_caH" = "Southern High (CaH)",
+                "2_caL" = "Southern Low (CaL)")
+
+# 
+plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
+  theme_bw() + 
+  theme(panel.grid = element_blank(), 
+        strip.background = element_rect(fill = "white"),
+        strip.text = element_text(face="bold", size=12), 
+        axis.text = element_text(face="bold", size=10), 
+        axis.title = element_text(face="bold", size=12)) + 
+  geom_hline(yintercept = 1, colour = "grey") + 
+  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
+  facet_wrap(~site_sorted, scales = "fixed", ncol = 1,
+             labeller = as_labeller(label_site)) + 
+  ylab("Ring width index") + xlab("Year") + 
+  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
+  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
+  geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
+  scale_x_continuous(limits = c(1819, 2016), 
+                     breaks= seq(1820, 2020, by=20))
+
+plot_rwi_crono
+
+```

### 2018-01-25 analysis/rwi/climate_rwi_relate.Rmd
update rwi climate

@@
 ```{r}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
+# machine <- 'ajpelu'
+machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
@@
                                 y = (coef + ifelse(coef >= 0, 0.02, -0.02))), 
-                            position = position_dodge(.9), size = 10),
+                            position = position_dodge(.9), size = 8),
                   scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3")))
@@
       TRUE ~ as.character(.$month)))
+
+write.csv(rprec_all, file=paste(di, "data/rwi_climate/rprec_selected.csv", sep=""), row.names = TRUE)
 ```
@@
 rsequia_all <- customNames(rsequia_all) 
+write.csv(rsequia_all, file=paste(di, "data/rwi_climate/rsequia_selected.csv", sep=""), row.names = TRUE)
 ```
@@
 rtmin_all <- customNames(rtmin_all) 
+write.csv(rtmin_all, file=paste(di, "data/rwi_climate/rtmin_selected.csv", sep=""), row.names = TRUE)
 ```
@@
 rtmax_all <- customNames(rtmax_all) 
+write.csv(rtmax_all, file=paste(di, "data/rwi_climate/rtmax_selected.csv", sep=""), row.names = TRUE)
 ```

### 2018-01-25 analysis/rwi/climate_rwi_relate.Rmd
update rwi climate with selected monts

@@
     variant: markdown_github
-bibliography: ../refs/references.bib 
-csl: ../refs/ecology.csl
+bibliography: ../../refs/references.bib 
+csl: ../../refs/ecology.csl
 ---
@@
 ```{r}
-# machine <- 'ajpelu'
-machine <- 'ajpeluLap'
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
 di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
@@
       TRUE ~ as.character(.$month)))
+
+write.csv(rprec_all, here("/data/rwi_climate", "rprec_all.csv"), row.names = FALSE)
 ```
@@
 rspei_all <- customNames(rspei_all)
+
+write.csv(rspei_all, here("/data/rwi_climate", "rspei_all.csv"), row.names = FALSE)
 ```
@@
 rsequia_all <- customNames(rsequia_all)
+
+write.csv(rsequia_all, here("/data/rwi_climate", "rsequia_all.csv"), row.names = FALSE)
 ```
@@
 rtmean_all <- customNames(rtmean_all) 
+write.csv(rtmean_all, here("/data/rwi_climate", "rtmean_all.csv"), row.names = FALSE)
 ```
@@
 rtmin_all <- customNames(rtmin_all) 
+write.csv(rtmin_all, here("/data/rwi_climate", "rtmin_all.csv"), row.names = FALSE)
 ```
@@
 rtmax_all <- customNames(rtmax_all) 
+write.csv(rtmax_all, here("/data/rwi_climate", "rtmax_all.csv"), row.names = FALSE)
 ```

### 2018-01-25 analysis/disturbance/analysis_disturbance_chronos.Rmd
update figures

@@
 
+```{r}
+gc_trees50<- c()
+
+datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
+
+for (d in datasetes){ 
+  # get df
+  df <- get(d)
+
+  au <- df %>% computeGCtrees(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
+  gc_trees50 <- rbind(gc_trees50, au)
+}
+
+gc_trees50  <- gc_trees50  %>% 
+  mutate(site_sorted = case_when(
+    site == "gc_sj" ~ "0_sj", 
+    site == "gc_caH" ~ "1_caH",
+    site == "gc_caL" ~ "2_caL"))
+
+gc_trees %>%   
+  write.csv(file=paste(di, "data/disturbance/disturbance_trees_gc_treshold50.csv", sep=""), row.names = FALSE)
+```
+
 ## Plots (mean)
@@
   write.csv(file=paste(di, "data/disturbance/disturbance_trees_gc_treshold_median.csv", sep=""), row.names = FALSE)
+```
+
+```{r}
+gc_treesMed50 <- c()
+
+datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')
+
+for (d in datasetes){ 
+  # get df
+  df <- get(d)
+
+  au <- df %>% computeGCtreesMedian(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
+  gc_treesMed50 <- rbind(gc_treesMed50, au)
+}
+
+gc_treesMed50  <- gc_treesMed50  %>% 
+  mutate(site_sorted = case_when(
+    site == "gc_sj" ~ "0_sj", 
+    site == "gc_caH" ~ "1_caH",
+    site == "gc_caL" ~ "2_caL"))
+
+gc_treesMed50 %>%   
+  write.csv(file=paste(di, "data/disturbance/disturbance_trees_gc_treshold_median50.csv", sep=""), row.names = FALSE)
 ```

### 2018-01-25 man/ms_figures.Rmd
update rwi climate with selected monts
update figures

@@
 ---
-title: "ms_figures"
 output:
-  md_document:
-    toc: true
-    variant: markdown_github
+  word_document:
+    fig_caption: yes
+    fig_height: 6
+    fig_width: 6
+    highlight: null
+    reference_docx: template_ms.docx
 bibliography: ../refs/references.bib
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
 ```
@@
 
-Figure 1.
+**Figure 1**.
 Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain range, where three cluster of oak populations have been identified [@PerezLuque2015] (showed in different colour) (b).
@@
 
-Figure 2.
-Comparison of EVI profile for the reference period (*gray*) and during the 2005 (*green*) and 2012 (*blue*) drought events (a).
+**Figure 2.** Comparison of EVI profile for the reference period (*gray*) and during the 2005 (*green*) and 2012 (*blue*) drought events (a).
 EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations (b).
@@
 
-Figure 3.
-Response of northern (*black*) and southern (*blue*) populations of *Q.
+**Figure 3.** Response of northern (*black*) and southern (*blue*) populations of *Q.
 pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; a) and tree growth (BAI; b) for the years 2005 and 2012.
@@
 
-Figure 4.
-Response of *Q.
+**Figure 4.** Response of *Q.
 pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree growth (BAI; b) by site.
@@
 
-```{r, warning=FALSE}
+```{r, warning=FALSE, out.extra='angle=90'}
 cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
@@
   theme(panel.grid = element_blank(),
+        axis.text.x=element_text(size=6),
         legend.position = c(0.1,.8),
@@
 
-Figure 5.
-Basal Area Increment (BAI) chronologies of *Q.
+**Figure 5.** Basal Area Increment (BAI) chronologies of *Q.
 pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA_Low; *pink*) and high-elevation (CA_High, *purple*) sites.
@@
 ```
+
+**Figure 6.** Residual Tree-ring chronologies obtained for the *Q.
+pyrenaica* sites.
+Dashed red lines indicate the start of the reliable period (EPS > 0.85).
+Dotted black lines showing the three most recent severe drought years (1995, 2005 and 2012).
+
+# Figure 7
+```{r}
+rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
+rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
+rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
+rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
+
+rtmax <- rtmax %>% mutate(v_name = 'T max')
+rtmin <- rtmin %>% mutate(v_name = 'T min')
+rsequia <- rsequia %>% mutate(v_name ='SPEI')
+rprec <- rprec %>% mutate(v_name = 'Prec')
+
+df <- rbind(rtmax, rtmin, rsequia, rprec)
+
+# cutre solucion para el problema de PREC 
+exdf <- data.frame(
+  id = c(16:19),
+  varname = rep("prec", 4), 
+  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
+  coef = rep(0, 4), 
+  significant = rep('FALSE', 4),
+  ci_lower = rep(0, 4),
+  ci_upper = rep(0, 4), 
+  sig = rep("", 4), 
+  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
+  v_name = rep("Prec", 4))
+
+exdf <- rbind(exdf, exdf, exdf)
+exdf$X <- c(46:57)
+exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
+
+ddf <- rbind(df, exdf)
+
+label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
+               "T max" = "(c) T max", "T min" = "(d) T min")
+
+clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
+  geom_hline(yintercept = 0, colour="gray") +
+  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
+  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
+  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
+            position = position_dodge(.9), size = 6) + 
+  theme_bw() + 
+  theme(axis.title.x = element_blank(),
+        panel.grid.major = element_blank(),
+        panel.grid.minor.y = element_blank(),
+        axis.text.x = element_text(angle = 90, vjust = 0.3),
+        legend.position = "bottom",
+        strip.background = element_rect(fill=NA, colour=NA),
+        strip.text.x = element_text(size = 10, face = "bold")) + 
+  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
+  scale_y_continuous(limits = c(-0.4, 0.8), 
+                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
+  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3")) + 
+  ylab('Correlation')
+
+clima_ring 
+```
+
+**Figure 7.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
+pyrenaica* and monthly climatic data (precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures) for northern site (*green* bars), low-elevation southern site (*light blue* bars) and high-elevation shouthern (*dark blue* bars) site.
+Asteriks indicate significant ($P < 0.05$) correlation coefficients.
+
+# Figure 8
+
+```{r}
+gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
+
+# customize plot (corrected scale)
+pos_neg_log <- scales::trans_new('signed_log', 
+                                 transform = function(x) sign(x)*log(abs(x)),
+                                 inverse = function(x) sign(x)*exp(abs(x)))
+
+custom_ylab <- '% GC'
+micolor <- c("#838B8B")
+
+# solution for arrange
+# http://felixfan.github.io/stacking-plots-same-x/
+
+custom_corrected_gc <- list(
+  geom_bar(stat='identity', fill=micolor, colour=micolor),
+  theme_bw(), 
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black',fill='white'))
+  )
+
+c <- gc_todos %>% 
+  filter(site != 'gc_sj') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
+  # scale_y_continuous(trans = pos_neg_log) +
+  facet_wrap(~site_sorted, nrow = 2,
+             labeller=as_labeller(c("1_caH" = "Southern High (CaH)", "2_caL" = "Southern Low (CaL)"))) + 
+    ylab(custom_ylab) + custom_corrected_gc
+
+d <- gc_todos %>% 
+  filter(site == 'gc_sj') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
+  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
+  facet_wrap(~site_sorted, nrow = 2,
+             labeller=as_labeller(c("0_sj" = "Northern (SJ)"))) + 
+  custom_corrected_gc + ylab('') + 
+  scale_y_continuous(trans = pos_neg_log, 
+                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
+  theme(axis.title.x = element_blank(),
+        axis.text.x = element_blank())
+
+plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
+plot_medians_corregido
+
+```
+
+**Figure 8.** Comparison of median growth change ($GC$) following [@Nowacki1997] for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
+
+# Figure 9
+
+```{r}
+gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
+gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
+
+colorpositivo <- c("#104E8B")
+colornegativo <- c("#CD3333")
+
+pgc_median <- gc_treesMed %>% ggplot() +
+  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
+  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
+  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
+  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
+  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
+  theme_bw() + ylab('% trees') +
+  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black', 
+                                        fill='white')) 
+pgc_median
+
+```
+
+**Figure 9**. Percentage of *Q. pyrenaica* trees affected by GC > 25 % by site. *Black* line shows number of trees (rigth-axis).

### 2018-01-26 man/ms_1_material_methods.Rmd
update

@@
 The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
-BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference [@Biondi2008].
+BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008].
 We used measured dbh and raw ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
@@
 ### Disturbance analyses 
-Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release event from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g.
+Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g.
 logging) in the past.
@@
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
-To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 25 % (DUDA) of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
+To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % (DUDA) of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
 
@@
 The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3) and this length was used in other studies [Gazol2017].
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3) and this length was used in other studies [@Gazol2017].
 

### 2018-01-26 man/ms_supp.Rmd
update

@@
 library("scales")
-
 library("grid")
@@
 library("patchwork")
-
+library("ggcorrplot")
 ```
@@
 
-## Supplementary 1
+## Figures
 
-Listado de pkgs usados
-
-## Suplementary 2
-
-## Figure S2
+**Figure S1**.
+Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
+Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
 
@@
 
-Figure S2. Spatial distribution of the EVI-annual trends for the period 2000-2016. Trends ($\tau$) with the most positive values (green) indicate positive trend.
-
-## Figure S3
+**Figure S2.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
 
@@
 
-Figure S3. Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
+**Figure S3.** Correlation
+
+```{r}
+
+corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
+corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
+corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
+
+generaMatrix <- function(cordf, variable){ 
+  m <- cordf %>% filter(var == variable) %>% 
+    dplyr::select(-var, -window_size) %>% 
+    mutate(w1 = c(2, 2, 3), 
+           w2 = c(3,4,4)) 
+  mf <- m %>% 
+    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
+
+  df <- m %>% rbind(mf) %>% 
+    add_row(r2 = 1, w1=2, w2=2) %>% 
+    add_row(r2 = 1, w1=3, w2=3) %>% 
+    add_row(r2 = 1, w1=4, w2=4)
+
+    df <- df %>% 
+    mutate(w1 = case_when(
+      w1 == 2 ~ 'ws2',
+      w1 == 3 ~ 'ws3',
+      w1 == 4 ~ 'ws4'),
+      w2 = case_when(
+      w2 == 2 ~ 'ws2',
+      w2 == 3 ~ 'ws3',
+      w2 == 4 ~ 'ws4'))
+
+      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
+
+return(cor_df)
+}
+
+cor_rt <- generaMatrix(corBai, 'rt')
+cor_rc <- generaMatrix(corBai, 'rc')
+cor_rs <- generaMatrix(corBai, 'rs')
+
+cor_rte <- generaMatrix(corEvi, 'rt')
+cor_rce <- generaMatrix(corEvi, 'rc')
+cor_rse <- generaMatrix(corEvi, 'rs')
+
+g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
+g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
+g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
+g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
+g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
+g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
+
+(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
+```

### 2018-01-29 man/ms_1_material_methods.Rmd
update text

@@
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
-There are eight oak patches (2400 Has) identified (**Figure 1**) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys.
-Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation [@Brewer2002; @Olalde2002; RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].
+There are eight oak patches (2400 Has) identified (Figure 1) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys.
+Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].
 
-The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000, @Vivero2000], undergoing intensive anthropic use in the last few decades [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
+The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], undergoing intensive anthropic use in the last few decades [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
 In fact, the status of conservation of this species for southern Spain is "Vulnerable" [@Vivero2000].
@@
 
-* :red_circle: Meter aquí algunos datos de sequia, similar a lo planteado por Gazol
+The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years particularly for the southern of the Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and the were charecterized by extreme drogught (Appendix S1).
+Aunque sabemos que existen otros años de sequía extrema (e.g.
+1995, 1981) [@VicenteSerrano2014], no se han elegido por una limitación metodológica, esto es, nuestro objetivo es analizar el efecto de la sequía y la resiliencia a los eventos de sequía utilizando información de satélite y de dendrocronología, y la información de satélite utilizada (ver mas adelante) solamente está disponible desde el año 2000 en adelante.
 
-:red_circle: `duda aqui` Varias referencias hablan de los años 2005 y 2012 como  extremadamente secos.
-Pero habría que hacer alguna referencia y/o análisis.
-Tengo dudas de si hemos de analizar (e incluir) que efectivamente los años 2005 y 2012 fueron caracterizados por un extrema sequía, por lo que habría que  incluyendo referencia a apéndice
-
-* O quizá un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+* Ver ejemplo Gazol et al 2017 Plan Ecology
+* un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
+")
+```
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+SPEI values between 1 and −1 are considered normal, whereas values <−1 indicate drought and values <−1.5 indicate severe drought47. We defined a drought as a period where SPEI12 decreased to ≤−1.5.
+")
+```
+
 The relationships between residual site chronologies and the climatic variables were assessed by a bootstrapped Pearson's correlation estimate using the `treeclim` package [@Zang2015].

### 2018-01-29 man/ms_0_intro.Rmd
updating texts

@@
 
+```{r setup, echo=FALSE}
+show_text <- FALSE
+```
+
 # Intro
+
+## 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 
@@
 `$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI. 
+")
+```
+
+Objetivos del trabajo
 
 ### Aims
-`$IMPROVE$`
-
 In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both greenees and growth of *Q. pyrenaica* forests in Sierra Nevada. Specifically,
@@
 
-* and to explore differences in the resilience metrics between populations located in contrasting slopes within the rear edge of the distribution of this species.
-
-Do northern and southern populations of Pyrenean oak forests differ in their resistance, resilience or recovery to extreme drought events?
-
-## References
+* and to explore differences in the resilience metrics between populations located in contrasting slopes within the rear edge of the distribution of this species.Do northern and southern populations of Pyrenean oak forests differ in their resistance, resilience or recovery to extreme drought events?

### 2018-01-29 man/ms_1_material_methods.Rmd
updating texts

@@
 
-The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years particularly for the southern of the Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and the were charecterized by extreme drogught (Appendix S1).
+The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years particularly for the southern of the Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and the were charecterized by extreme drogught (Figure S1).
 Aunque sabemos que existen otros años de sequía extrema (e.g.
@@
 EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
+We explored the temporal trends of EVI at a pixel scale.
+We computed the nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945), using the `trend` package [@Thorsten2017].
+We set 0.05 the alpha level for the test.
 
@@
 We assessed stand competition of target trees by recording distance, azimuth, dbh, species and total height of all neighboring living trees with dbh > 7.5 cm in a circular plot of 10 m radius.
-We computed two distance independent indices: density in $\mathrm{trees \cdot ha^{-1}}$, and basal area (BA) in $\mathrm{m^{2} \cdot ha^{-1}}$; and one distance dependent indices: size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
+We computed two distance independent indices: *density* in $\mathrm{trees \cdot ha^{-1}}$, and *basal area* (BA) in $\mathrm{m^{2} \cdot ha^{-1}}$; and one distance dependent indices: size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
 We used non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] for search differences between sites for height, DBH and competence indices.

### 2018-01-29 man/ms_2_results.Rmd
updating texts

@@
 Trends analysis showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
-The strongest trends were observed in southwestern populations (Figure S1).
+The strongest trends were observed in southwestern populations (Figure S2).
 Vegetation greenness of *Q.
@@
 According to the standardized anomalies, *Q.
-pyrenaica* forests suffered a browning during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), however no changes were observed for the 2012 drought (Figure S2).
+pyrenaica* forests suffered a browning during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), however no changes were observed for the 2012 drought (Figure S3).
 
@@
 
-## Responses to drougth: tree-growth (BAI)
+## Growth trends
+The trees of the southern population were older and larger than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2).
+Competence indices were similar among sites (Table 1).
+The basal area increment (BAI) were different between sites (Figure 5).
+El sitio mas elevado (Ca-High) mostró mayores valores de BAI que los otros sitios, mostrando además una tendencia positiva a partir de 1970.
+Se observan diferencias en cuanto al crecimiento entre los sitios de las poblaciones del sur.
+Ca_Low se parece en crecimiento mas a SJ que a Ca_High.
 
-Although no significant differences were observed in the resistance of the oak growth to both drought events (Table 3; Figure 3b), the 2005 drought reduced growth more than 2012 did (Rt = 0.721 and 0.819 respectively) (Table S2).
-Northern population showed resistance values lower than southern ones, especially for 2005 drought event where the growth was reduced to 44.5 % of the preceding period (Figures 3b, 4b).
+Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5).
+La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona.
+Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.
+
+Los tres sitios mostraron valores de ring width indices bajos para el año 1995 (Figure 6).
+El sitio SJ mostró valores muy bajos de ring width indices en 1995, 2005, y 2012.
+Sin embargo esto no ocurrió igual en los sitios del sur (caL y CaH) donde si encontramos valores bajos de RWI para 1995, pero no para 2005 y 2012.
+
+Sorprende que en los sitios del sur (caH y caL), en los últimos años ha habido crecimientos muy grandes.
+Por ejemplo, en 2010 se observó uno de los mayores valores de RWI de toda la cronología tanto en caH como en caL.
+Algo similar ocurrió en sj para los años 2003 y 2013.
+
+## Climate and tree
+Tree-growth exhibited a significant greater response to water availability more than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
+Cumulative precipitation of the hydrological year was the climatic variable with a higher relationship with growth and influenced positively the tree-growth for the both northern and southern populations (Figure 7a).
+Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
+Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (r > 0.6) (Figure 7b), specially for northern population, which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
+Temperatures were much less related to growth than moisture variables.
+Yet spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 7c and 7d).
+Surprisingly minimum and maximum temperatures of current September positively influenced the tree-growth in the northern population.
+
+## Disturbances
+The analysis of growth changes revealed differences between sites (Figures 8 and 9).
+En el sitio del norte osbervamos dos eventos de liberación importantes (GC > 50 %), que ocurren en mas de la mitad de los árboles muestreados (Figures 8 and 9).
+El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
+Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
+Estos periodos se alternan con periodos de supresión.
+En los sitios del sur (caH y caL), sin embargo se observan algunos episodios débiles de supresión, aunque estos no se observan para los últimos años, de hecho incluso se observa una débil liberación (Figure 8) afectando a mas de la mitad de los árboles muestreados (Figure 9).
+
+## Responses to drougth: tree-growth
+
+Although no significant differences were observed in the resistance of oak growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
+The northern site, which is under a drier cliamte, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figures 3b, 4b).
 
@@
 Oak forests recovered their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1 (Table S2).
-Norhtern populations showed signigicantly higher values of recovery than southern ones (Tables 3, S2), which even not reach recovery values of Rc = 1 (Figure 3b, 4b).
+Northern populations showed signigicantly higher values of recovery than southern ones (Tables 3 and S2), which even not reach recovery values of Rc = 1 (Figure 3b, 4b).
 
@@
 We found no differences of resilience values for growth between northern and southern populations (Table 3).
-For 2005 drought event the resilience of northern population was lower than southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity (Figure 3b).
+For 2005 drought event the resilience of northern population was lower than that of southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity (Figure 3b).
 
-# Chronology
-
-The trees of the southern population were older and larger than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2).
-Competence indices were similar among sites (Table 1).
-The basal area increment (BAI) were different between sites (Figure 5).
-El sitio mas elevado (Ca-High) mostró mayores valores de BAI que los otros sitios, mostrando además una tendencia positiva a partir de 1970.
-Se observan diferencias en cuanto al crecimiento entre los sitios de las poblaciones del sur.
-Ca_Low se parece en crecimiento mas a SJ que a Ca_High.
-
-Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5).
-La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona.
-Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.
-
-and higher values of tree-growth (Table 2).
-
-El sitio caH mostró además los mayores valores de AR(1) y de MS
-
-El sitio del norte, mostró mayores valores de MS (Table 2). ¿esto sugiere que este sitio tiene una alta limitación climática?
-
-Different growth trends were observed for each site (Figure 5). Ca-High (the highest elevation site) had higher values of tree growth than the other sites.
-
-# Climate and tree
-
-Tree-growth exhibited a significant response to water availability more than to the temperatures.
-Cumulative precipitation of the hydrological year (September of the previous year to August of the growth year) influenced positively the tree-growth for the both northern and southern populations (FIG ).
-Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
-Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth, specially for northern population.
-Spring Maximum temperature was the most significant limitation of the tree-growth for the southern population while maximum temperatures of the previous August did for the northern one.
-Conversely, mean, minimum and maximum temperatures of current September positively influenced the tree-growth in the northern population.
-
-# Disturbance
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 
@@
 En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles
-
-notas para discusión
-
-En ambos casos (evi y bai) la sequía de 2005 y de 2012 redujeron el greenness y el crecimiento.
-Además en ambos casos la sequía de 2005 redujo mas la variable, sin embargo, para el greenness si existen diferencias entre ambos eventos de sequía, mientras que para el crecimiento no.
-Por tanto podemos decir que se ha reducido el greennes pero para el crecimiento esto está mas amortiguado.
-
-Al explorar la resistencia por sitio y drought se observa una curiosidad.
-La resistencia de los robledales de CaLow se mantiene casi igual.
-La de la cara norte (SJ) sufre mucho en 2005 y luego se recupera un poco.
-La de caH presenta menor resistencia en 2012 que en 2005 (y eso llama la atención)
-
-## References
+")
+```

### 2018-01-29 man/ms_3_discussion.Rmd
updating texts

@@
 
+and higher values of tree-growth (Table 2).
+
+El sitio caH mostró además los mayores valores de AR(1) y de MS
+
+El sitio del norte, mostró mayores valores de MS (Table 2). ¿esto sugiere que este sitio tiene una alta limitación climática?
+
+Different growth trends were observed for each site (Figure 5). Ca-High (the highest elevation site) had higher values of tree growth than the other sites.
+
+notas para discusión
+
+En ambos casos (evi y bai) la sequía de 2005 y de 2012 redujeron el greenness y el crecimiento.
+Además en ambos casos la sequía de 2005 redujo mas la variable, sin embargo, para el greenness si existen diferencias entre ambos eventos de sequía, mientras que para el crecimiento no.
+Por tanto podemos decir que se ha reducido el greennes pero para el crecimiento esto está mas amortiguado.
+
+Al explorar la resistencia por sitio y drought se observa una curiosidad.
+La resistencia de los robledales de CaLow se mantiene casi igual.
+La de la cara norte (SJ) sufre mucho en 2005 y luego se recupera un poco.
+La de caH presenta menor resistencia en 2012 que en 2005 (y eso llama la atención)
+
 ## References

### 2018-01-29 man/ms_figures.Rmd
updating texts

@@
 year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
+```
 
+```{r}
+# top10  
+top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
+top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
+top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
+top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
+```
+
+```{r}
 label_site <- c("0_sj" = "Northern (SJ)",

### 2018-01-29 man/ms_supp.Rmd
updating texts

@@
 library("ggcorrplot")
+library("stringr")
 ```
@@
 
-**Figure S1**.
+**Figure S1** Acumulated monthly precipitation during the hydrological year 2004-2005 (blue line) and 2011-2012 (red line).
+The boxplot representing the average from 1940-2015 period.
+Data from meteorological station Granada, Base Aérea (National Spanish Meteorological Services (AEMET)).
+
+```{r}
+ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
+
+source(here("/script/R/", "hydro_year.R"))
+
+ltdf <- ltdf %>% 
+  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
+  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
+         month = lubridate::month(fecha),
+         hmonth = ifelse(month <= 8, month + 4, month -8), 
+         hyear = hydro_year(fecha),
+         hyear_f = paste0(hyear-1,'-', hyear))
+
+# We used only the 5514 station (the most long term)
+# Filter by station and prec 
+# Filter from 1950 
+st5514 <- ltdf %>% 
+  filter(VARIABLE == 'PM1') %>% 
+  filter(cod == '5514') %>% 
+  filter(hyear >= 1950) %>% 
+  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
+
+st5514_pro <- st5514 %>% 
+  group_by(hyear_f, hmonth) %>% 
+  dplyr::summarise(pre = sum(pre)) %>%
+  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
+
+st5514_pro  %>% 
+  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
+  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
+  geom_boxplot(aes(x=as.factor(hmonth))) +  
+  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
+  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
+            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
+  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
+            aes(x=as.factor(hmonth), y=csum, group=1),
+            colour='red') +
+  theme_bw() + 
+  theme(strip.background = element_rect(fill = NA, color = "black")) + 
+  ylab('Precipitation (mm) acumulated') + xlab('') + 
+  ggtitle('Granada Base Aerea, 1950-2015') + 
+  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
+                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
+  annotate("text", label = "2004-2005", x = 2, y = 500, size = 6, colour = "blue") + 
+  annotate("text", label = "2011-2012", x = 2, y = 550, size = 6, colour = "red") + 
+  annotate("text", label = "mean 1950-2015", x = 3, y = 600, size = 6, colour = "black")
+```
+
+```{r, eval=FALSE, echo=FALSE}
+# Get file names
+myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
+
+mydf <- data.frame()
+
+# Loop to read files and prepare data
+for (j in myfiles){ 
+  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
+
+  # Remove csv and get name 
+  name_aux <- str_replace(j, ".csv", "")
+
+  # Get lat long
+  latlong <- str_replace(name_aux, "SPEI_", "")
+
+  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
+  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
+
+  aux$lat <- mylat
+  aux$long <- mylong
+
+  # Split time 
+  # Get year 
+  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
+
+  # Get month 
+  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
+
+  # Variable month in number 
+  aux$months <- match(aux$month, month.abb)
+
+  # assign(name_aux, aux)
+
+  mydf <- rbind(mydf, aux)
+}
+
+spei <- mydf %>% select(SPEI_12, year, months) %>% 
+  group_by(year,months) %>% 
+  summarise(spei6 = mean(SPEI_12),
+            spei6_sd = sd(SPEI_12)) %>% 
+  mutate(date = lubridate::ymd(paste(year, months, "01")))
+
+spei <- spei %>% 
+  mutate(sign = case_when(
+    spei6 > 0 ~ "pos",
+    TRUE ~ "neg"))
+
+limits <- as.Date(c("19549-01-01", "2015-12-01"))
+spei %>% 
+  ggplot(aes(x=date, y=spei6, fill=sign)) +  
+  geom_bar(stat = "identity") + 
+  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
+  theme_bw()  + 
+  labs(title= 'Drought evolution for Sierra Nevada',
+    x= '',
+    y='SPEI') +
+  theme(strip.background = element_rect(fill = "white"),
+        legend.position = "none") +
+  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
+```
+
+**Figure S2**.
 Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
@@
 
-**Figure S2.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
+**Figure S3.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
 

### 2018-01-30 man/notes/notes_history.Rmd
add notes usos

@@
 
+# Ganado
+## San Juan
+
+* Fuente: MIGRAME y FB.
+Entrevista a José Luis Pimentel (617851906) (guarda del P Nacional, hijo del guarda de la finca privada Dehesa de San Juan ...) Source: /Users/ajpelu/Desktop/2012_FB_recopilacion_historica/Obj_2_Usos_Suelo_Aprovechamientos/Act_212_Asociaciones profesionales/GuejarSierra
+
+Recopilación información ganadera (from MIGRAME)
+
+Respecto a los lugares concretos en los que, en principio, se va a desarrollar el proyecto se han recopilado los datos de detalle de la carga ganadera y su evolución a lo largo de la serie del 56 hasta la actualidad, así como su variación estacional.
+También se ha espacializado la información.
+Los lugares son: robledal del San Juan, el de Cañar y la Dehesa del Camarate.
+En Cañar ha sido a través de los pliegos y de entrevistas con pastores, técnicos del ayuntamiento  y al alcalde.
+En San Juan, a través de entrevista con el guarda cuando era privada y a través de los pliegos en el periodo en el que pasó a ser de Parques Nacionales.
+En el Camarate a través de entrevista con el guarda de las dos zonas (privada y la ahora pública), que antes era una sola.
+Los datos de la totalidad de montes públicos tanto de la junta de Andalucía, como de ayuntamientos y del ministerio de medio ambiente, los he conseguido pero de manera agregada.
+Estoy intentando por tres vías diferentes hacernos con ellos: que nos den de alta como usuarios en la aplicación SAGA, que gestiona los aprovechamientos, solicitarlos a la Rediam (que ya he cursado) y si no da resultado esto, buscar gente del parque que esté dada de alta en la aplicación y que nos deje pasarnos por allí a bichearla.
+
 # Links

### 2018-01-30 man/notes/notes_history.Rmd
add notes historic

@@
     variant: markdown_github
-bibliography: ../refs/references.bib
-csl: ../refs/ecology.csl
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecology.csl
 ---
@@
 ```
+
+# Actuaciones forestales
+
+Revision bbdd actuaciones forestales
+
+* PROYECTO DE ACTUACIONES EN LA DEHESA DE SAN JUAN. FINCA PÚBLICA DEL P.N. SIERRA NEVADA 
+	* exp. CMA. 10050119
+	* Director Javier Cano 
+	* cod. base de datos 36
+
+* MEJORA SELVÍCOLA EN LA DEHESA DE SAN JUAN
+	* Tragsa
+	* Director Juan Montoya
+	* exp. S/N 21/09 
+	* FECHA?? 
+	* cod. base de datos 105
+
+* GESTIÓN ADAPTATIVA DE MASAS DE QUERCUS PYRENAICA EN LA DEHESA DE SAN JUAN
+	* TRAGSA
+	* NO FECHA
+	* cod. base de datos 107 
+	* ??
+
+* REGENERACIÓN DE ECOSISTEMAS EN EL PARQUE NATURAL DE SIERRA NEVADA
+	* director ANDRÉS CASTILLO FERNAUD 
+	* Fecha fin ?? 
+	* cod. base de datos 70
+	* obs. En la base de datos aparece que la actuación es sobre Encinar, pero la espacialización es sobrel Robledal de Cáñar 
+	* En la bd de proyectos en obs. aparece en el campo OBSERV: Aclareos dejando en todo una densidad mínima de 1200 pies/ha. Resalveo no reduciendo más del 20% de FCC.
+
+* RESTAURACIÓN DE HÁBITATS DE ROBLEDAL Y ENCINAR EN EL ESPACIO NATURAL DE SIERRA NEVADA
+	* Fecha fin ?? 
+	* cod. base de datos 61
+	* No fecha en proyecto, pero si en actuación (2007) ?? 
+	* Es del que habla M. Valbuena??
+
+* RESTAURACION ZONAS INCENDIADAS EN MONTES MUNICIPALES DE SIERRA NEVADA
+	* JAVIER NAVARRO
+	* fecha fin? 
+	* cod. base de datos 53
+
+### datos 
+type, date_start, date_end, site, zona, id_proyecto, id_pols, obs
+resalveo 50 percentage, 01/10/2005, 31/08/2006, sj, hortichuela, 36, 1533|1534, no obs
+clara-clareo 100 percentage,,,sj, hortichuela ctjo. hoyo, 105, 3401, no fecha| densidad de poda 20 pies / ha
+resalveo 50 percentage, 01/06/1995, , ca, El Robledal | entre estacion y cortijo nevazo, 70, 2354, aclareos dejando una densidad mínima de 1200 pies / ha | Resalveo no reduciendo mas del 20 % de FCC 
+resalveo 50 percentage, 01/01/2007,,ca, El Robledal | entre estacion y cortijo nevazo, 61, 1813, 
+roza selectiva matorral 50 percentage, 01/01/2007,,ca, El Robledal | entre estacion y cortijo nevazo, 61, 1813, idem anterior
+resalveo 50 percentage,,,ca, buffer carril entre estacion y carril que sube (por donde subimos),, 1870, pone que es resalveo de encinar
+roza selectiva matorral 50 percentage,,,ca, buffer carril entre estacion y carril que sube (por donde subimos),, 1870, pone que es roza de matorral | idem anterior
+resalveo 50 percentage,,,ca, buffer carril entre estacion y carril treeline,, 1871, pone que es resalveo de encinar
+roza selectiva matorral 50 percentage,,,ca, buffer carril entre estacion y carril treeline,, 1871, pone que es roza de matorral | idem anterior
+resalveo 50 percentage,01/01/2008,,ca,zona derecha estacion,53, 1873, restauracion zonas incendiadas (incendio??)
+roza selectiva matorral 50 percentage,01/01/2008,,ca,zona derecha estacion,53, 1873, restauracion zonas incendiadas (incendio??)
+resalveo 50 percentage,,,ca,zona derecha estacion bajo hacia ca,, 1874, restauracion zonas incendiadas (incendio??)
+roza selectiva matorral 50 percentage,,,ca,zona derecha estacion bajo hacia ca,, 1874, restauracion zonas incendiadas (incendio??)
 

### 2018-01-30 man/ms_1_material_methods.Rmd
update content

@@
 The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3) and this length was used in other studies [@Gazol2017].
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S4) and this length was used in other studies [@Gazol2017].
 

### 2018-01-30 man/ms_2_results.Rmd
update content

@@
 cat("
-* Disminución del EVI --> Espacialización ver plots de trajectorias
-* Por ejemplo la disminución para 2005 fue heterogénea (
-
-* Durante 2005 la mayoría de los pixeles mostraron browning (99.36 % y 79.37 % para las poblaciones del norte y del sur respectivamente), sin embargo en 2012 la mayoría de los pixeles se clasificaron como no changes en las poblaciones del norte (89.60 %) y en las del sur (70.07 %) (Tabla supplementaria??)
-
-* Si atendemos a las sa (standardized anomalíes) y aplicamos el criterio de Gao, podemos decir que en 2005 se observó un bronwing en los bosques de Q.
-pyrenaica, sobre todo en las situadas en el northern slopes.
-
 > Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005.
@@
 
-> Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029)que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
-
-> * En 2005 y en 2012 las anomalías fueron negativas.
-> Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.
-
-#### Temporal trend EVI values 
-El 78.95 % de los pixeles mostraron una tendencia positiva en cuanto al EVI medio anual (siendo significativa para el 31.67 % de los pixeles).
-Esta tendencia positiva fué sobre todo mayor en algunas de las poblaciones del suroeste.
-La
-
---> referencia a las tendencias en EVI (lo hemos vuelto a calcular y además en el trabajo de ontologías también nos sale)
-
-Aproximadamete el x% de los robledales mostraron un incremento en los valores de productividad (greenees) ... (relacionar con tendencias EVI y ontologias).
-
-Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI.
-En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.
+> Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029) que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
 ")
@@
 
-## Climate and tree
-Tree-growth exhibited a significant greater response to water availability more than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
-Cumulative precipitation of the hydrological year was the climatic variable with a higher relationship with growth and influenced positively the tree-growth for the both northern and southern populations (Figure 7a).
-Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
-Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (r > 0.6) (Figure 7b), specially for northern population, which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
-Temperatures were much less related to growth than moisture variables.
-Yet spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 7c and 7d).
-Surprisingly minimum and maximum temperatures of current September positively influenced the tree-growth in the northern population.
-
-## Disturbances
-The analysis of growth changes revealed differences between sites (Figures 8 and 9).
-En el sitio del norte osbervamos dos eventos de liberación importantes (GC > 50 %), que ocurren en mas de la mitad de los árboles muestreados (Figures 8 and 9).
-El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
-Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
-Estos periodos se alternan con periodos de supresión.
-En los sitios del sur (caH y caL), sin embargo se observan algunos episodios débiles de supresión, aunque estos no se observan para los últimos años, de hecho incluso se observa una débil liberación (Figure 8) afectando a mas de la mitad de los árboles muestreados (Figure 9).
-
 ## Responses to drougth: tree-growth
@@
 For 2005 drought event the resilience of northern population was lower than that of southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity (Figure 3b).
+Es intersante que los valores de resiliencia para el tree-growth no alcanzaron el valor 1, excepto para 2005 en la población de SJ (Rs = 1.031)
+
+## Climate and tree-growth
+Tree-growth exhibited a significant greater response to water availability more than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
+Cumulative precipitation of the hydrological year was the climatic variable with a higher relationship with growth and influenced positively the tree-growth for the both northern and southern populations (Figure 7a).
+Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
+Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for northern population (r > 0.6), which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
+Temperatures were much less related to growth than moisture variables.
+Yet spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 7c and 7d).
+Surprisingly minimum and maximum temperatures of current September positively influenced the tree-growth in the northern population.
+
+## Disturbances
+The analysis of growth changes revealed differences between sites (Figures 8 and 9).
+En el sitio del norte osbervamos dos eventos de liberación importantes (GC > 50 %), que ocurren en mas de la mitad de los árboles muestreados (Figures 8 and 9).
+El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
+Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
+Estos periodos se alternan con periodos de supresión.
+En los sitios del sur (caH y caL), sin embargo se observan algunos episodios débiles de supresión, aunque estos no se observan para los últimos años, de hecho incluso se observa una débil liberación (Figure 8) afectando a mas de la mitad de los árboles muestreados (Figure 9).
 

### 2018-01-30 man/ms_3_discussion.Rmd
update content

--- 2018-01-18 man/ms_3_discussion.Rmd
@@
 
-# Discussions notes
+```{r, echo=FALSE}
+show_text <- FALSE
+```
 
-## Greenness
+# Algunas ideas para la discussion
 
-* Aproximadamete el x% de los robledales mostraron un incremento en los valores de productividad (greenees) ... (relacionar con tendencias EVI y ontologias).
-* tendencias ontologias
+### Efectos de la sequía en el greenneess y en el crecimiento.
 
+Hemos observado como las sequía, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento.
+Los datos de anomalías estandarizadas de EVI mostraron un browning para la mayoría de los robledales de Sierra Nevada durante la sequía de 2005.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 * Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI.
-En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.
+En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor
+")
+```
 
-## Dos cronos en Cañar 
-* @GeaIzquierdo2014 muestrea solo en CA-Low (QUPY10) y obtiene resultados similares a CA-Low. 
-* En tan poco espacio hay diferencias enormes entre las dos cronos
-* *Las tendencias de BAI de Cáñar baja altitud son diferentes a las del norte, las cuales se parecen mucho a las que suelen encontrarse en la mayoría de sitios.
-Puedes mirar por ejemplo las tendencias de las líneas negras (de más de 100 años, sobre todo) en la Fig 2 del artículo de Ecosystems… fíjate que QUPY3 se parece más a QUPY10… y QUPY3 era la única dehesa (en parte) y uno de los sitios más secos (por otro lado, aunque frío).
-En fin, todo esto es algo especulativo, pero sirva como exploración de tus cronos, y para mostrar que me parece muy interesante la diferencia entre Cáñar alto y bajo (y nuevo).*
+La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
+Esto se ha observado también en otras especies en el sur de la P.
+Ibérica (p.ej.
+P.
+nigra en Andalusia [@SanchezSalguero2013], ...
+INCLUIR otras especies y citas).
 
-## Reducción crecimiento y sequías
+Algunas notas sobre la sequía de 2005:
 
+* 2004/2005 hydrological year is considered one of the worst drooughts ever recorded in the Iberian Peninsula, particularly in the central and southern sectors [@GarciaHerrera2007].
+* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].
+* The hydrological year from October 2004 to September 2005 was the driest on record at several locations throughout Iberia [@GarciaHerrera2007].
+* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 Se observó una gran reducción de crecimiento en 1994-1995, 1999 y 2005 para Pinus nigra en el SE de Andalusia [@SanchezSalguero2013] coincidiendo con años de sequía para esa zona.
 La máxima reducción en crecimiento fue observada para 2005, siendo significativamente mayor que para el resto de años [@SanchezSalguero2013].
+")
+```
 
-## Relaciones crecimiento-clima en otras especies
+### La resiliencia de las poblaciones de roble de Sierra Nevada.
+
+El greennness de los robledades en Sierra Nevada ha mostrado una tendencia positiva hacia mas verdor en los ultimos años, que coincide con lo que ya observamos con datos de NDVI [@PerezLuque2015onto], sobre todo para las poblaciones del sur.
+Esta tendencia positiva, obtenida de variables derivadas de remote sensing (EVI y NDVI), parece que también se observa en el crecimiento.
+Por ejemplo en las poblaciones del sur (CaHigh y CaLow) observamos en los últimos años una ligera tendencia de crecimiento.
+
+Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia.
+Por ejemplo para las poblaciones del norte, los valores de EVI durante la sequía de 2005 descendieron hasta el 81 %, mientras que el BAI (sitio SJ) lo hizo hasta el 45 %.
+Sin embargo, la recuperación fue rápida, así los valores de Recovery (Rc) para el EVI tras la sequía de 2005 (en las poblaciones del norte) fue de 1.17, mientras que para el BAI el Rc fue de 1.112.
+En definitiva, estamos observando altos valores de resiliencia en estos robledales.
+
+### Diferencias entre sitios
+
+Llama la atención que en el rear-edge de la distribución de la especie, en el sitio mas meridional dentro de este rear edge, y en la parte mas alta (estas parcelas están en el treeline de la especie en SN (en torno a 1900)), es donde encontramos mayores crecimientos.
+Parece que los robles estuvieran mejor en esta zona.
+Algunas reflexiones sobre esto:
+
+* Quizá están creciendo donde les están dejando (menor impacto antrópico en las zonas mas altas, al menos en los últimos años??)
+* ¿Diferentes niveles de compentencia?. Parece que la competencia es similar en los tres sitios: no diferencias sig. para valores de Stand density ySize ratio proportional to distance (ver table 1). 
+* ¿Diferencias entre suelos??
+En principio no.
+[@CoboDiaz2017] en un trabajo sobre microbiota del suelo, explora diferencias en un gradiente altitudinal en Cáñar.
+Analiza 3 sitios: por encima del treline (XZF, el piornal), low altitudinal oak forest (LAF, en la zona baja del robledal) and high altitudinal oak forest (HAF).
+En su trabajo, LAF está mucho mas bajo, pero HAF está cerca de las parcelas caHigh.
+Restulados:
+    * tipo de suelo: LAF es Sandy-loam, HAF es loam
+    * HAF tuvo los valores mayores de disponibilidad hídrica (% available water tabla s1 en [@CoboDiaz2017]) --> Esto es importante, creo
+    * Los tres sitios eran pobres en materia orgánica, pero el HAF dobló los valores de los otros dos sitios
+    * No diferencias en contenido de N, C/N ratio similares en HAF, y LAF
+
+* ¿Menor estrés hídrico en zonas mas altas?
+Puede ser que tengamos menor estrés hídrico en esta zona??
+Este robledal, está en la cara sur de SN, y en una ladera con bastante insolación.
+Algunos trabajos antiguos (de fitosociólogos) hablan de que su presencia aquí se debe a que reciben un aporte extra de humedad procedente de las brisas del mediterráneo, para suplir el mínimo de humedad que necesitan en verano.
+Quizá también tendríamos que incluir el papel de las acequias.
+La zona caHigh tiene una acequia muy cerca (y por encima) de donde muestreamos (recordad el roble mas grande, y el mas alto, esta justo en una acequia).
+No se si es interesante que lo comentemos.
+
+* Otro punto de interés a introducir aquí es que frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], sin embargo algunos estudios están demostrando esto no es siempre así (ver por ejemplo [@Cavin2017; @Granda2017]).
+Nuestros resultados creo que van en esta línea creo.
+Esto, como apunta [@MartinezVilalta2018]m tiene que ver con la que consideramos como habitat marginal de la especie (...
+When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+## Rear-edge
+
+* Se asume una alta vulnerabilidad de las especies a la sequía en su límite inferior, sin embargo esta vulnerabilidad no necesariamente se mantiene: The results of Granda et al.
+2018 are consistent with recent studies showing that the frequently assumed higher vulnerability of dry, rear-edge populations to drought does not necessarily hold (e.g.
+Cavin and Jump 2017) (from: Martinez-Vilalta 2018)
+
+* Definición de *marginal* populations.
+[@MartinezVilalta2018] indica que a veces no está claro la definición de marginal populations.
+When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).
+
+* Quizá estamos asumiendo que es una población marginal, pero sin embargo lo es solamente geográficamente
+")
+```
+
+### ¿Que factor es mas limitante para el crecimiento en el rear-edge de Q. pyrenaica? 
+Aquí pueder ser interesante comentar algo de el peso de las variables climáticas en el crecimiento para poblaciones situadas en el borde de distribución (pesa mas la temperatura o la disponibilidad de agua?).
+
+Para *Q.
+pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q.
+pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al.
+2015 European Journal of Forest Research).
+Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea.
+
+Podríamos complementarlo con lo que le pasa a otras especies en su borde de distribución: por ejemplo en Baza, Herrero et al.
+2013, encontraron para Pinus nigra y sylvestris que la temperatura tenía mas peso que la disponibilidad de agua).
+O también ver algunos de los trabajos de Camarero et al 2013 para el P.
+nigra en su borde de distribución u otros similares (el de Sanchez-Salguero et al.
+2013, 2015) …
+
+## historia forestal de ambos sitios 
+Incluimos lo que conocemos de la historia forestal de los sitios??
+Me explico, tenemos datos de manejo y uso antrópico de las dos zonas que proceden de varias fuentes.
+En resumen, mas o menos, tenemos:
+
+* Minería: datos de minería, que afectan sobre todo a la población de SJ (los tenemos localizados temporalmente)
+* Actuaciones forestales: Tenemos una bd con actuaciones forestales, y he contactado con varios de los responsables de proyectos de actuaciones forestales en ambas zonas, y tenemos con bastante detalle la información sobre actuaciones forestales (al menos espacial y temporalmente)
+* Incendios, Carboneo, Ganadería, etc --> Tenemos información menos estructurada sobre estos ámbitos, procedentes de varias fuentes, que tienen una incertidumbre mayor espacial y temporalmente.
+
+Todo esto lo comento, porque quizá podemos utilizar dicha información para la discussión, o a lo mejor mete mas ruido (¿que opináis?)
+
+Por otro lado, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia.
+Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética).
+
+## refugio??
+
+Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático.
+Hemos observado un aumento en el greenness en los últimos años.
+Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness.
+Además, estos robledales tienen una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
+¿Sierra Nevada (regiones de montaña) como refugio??
+quizá este rear-edge esté actuando como refugio??
+(esto es muy especulativo)
 

### 2018-01-30 man/ms_supp.Rmd
update content

@@
 
-**Figure S3.** Correlation
+**Figure S4.** Correlation between indices of resilience (Rt, resistance; Rc, recovery; Rs, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
+Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
+ws2, ws3 and ws4 indicate periods of 2, 3 and 4 years after a drought.
 

### 2018-01-30 man/notes/notas_resultados.Rmd
update content

---
title: "Notas Resultados"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017"
output: 
  word_document
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# EVI 
## Tendencias

* Cerca del 79 % de los pixeles han mostrado una tendencia positiva en el EVI durante 2000-2016 (RELACIONAR con ontologías y citar)
* Los taus mayores sobre todo en las zonas del SW
* Plot de trajectorias para ver la heterogeneidad. Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI. En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor.

## EVI y sequias

* Disminución del EVI durante las sequias, sobre todo la de 2005. Heterogeneidad de la disminución del EVI (ver plot de trajectorias)
* Anomalías estandarizadas

    * En 2005 y en 2012 las anomalías fueron negativas. 
    * 2005 la mayoría de los pixeles sobrepasas $EVI~sa < -2$. 
    * Durante 2005 la mayoría de los pixeles mostraron browning (99.36 % y 79.37 % para las poblaciones del norte y del sur respectivamente), sin embargo en 2012 la mayoría de los pixeles se clasificaron como no changes en las poblaciones del norte (89.60 %) y en las del sur (70.07 %). En 2005 se observó un bronwing en los bosques de Q. pyrenaica, sobre todo en las situadas en el northern slopes. 
    * Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought.

Pero si miramos las anomalías estandarizadas, vemos que es en 2005 donde los valores sobrepasan EVIsa < -2 (browing < -1). Es decir, atendiendo a las anomalias sa, 2005 fue muy severa, y sobre todo para las poblaciones del N, ya que: 
* Valores promedios de EVIsa mas negativos (sobrepasan -2)
* El 99 % de los pixeles de las poblaciones del N, con pixels browing

## Resiliencia EVI

* La sequía de 2005 redujo el EVI hasta un 85 %, mientras que la de 2012 solo lo hizo hasta el 94 % (es decir mayor resistencia a la sequía de 2012, lógico porque parece que fue menos severa)

    * Las poblaciones del S resistieron mejor que las del N en 2005 (el EVI se redujo hasta 90% frente al 81 %)

* Sorprendentemente, los robles recuperaron el greenness mas rápido tras 2005 que tras 2012 (Rc2005 = 1.12 vs. Rc2012 = 1.057).

    * N se recuperó mas rápidamente tras 2005 que tras 2012
    * No diff en la recuperación tras 2005 y tras 2012 para el S

* Los robles recuperan su valor previo de greenness (Rs ~ 1).
* Para 2005 la Rs fue similar entre N y S

Ante eventos de sequía fuerte (como el de 2005), aunque afecte de forma diferencial, parece que la recuperación ha sido similar

# Tree-growth 
## Trends

* Vemos que el crecimiento ha sido mayor en Shigh que en Slow and N (en estos es mas parecido). De hecho podemos ver una tendencia positiva en los últimos años para Shigh.

## Tree-growth y sequía 
Las sequías afecataron al BAI y al RWI. 
* Para el BAI, se vio una reducción en los años 2005, 1995, pero dudosa para 2012.
* Para RWI, claramente una reducción en 1995, y 
    * N si se observó una reducción en 2005 y 2012
    * S no se observó ni en 2005 ni en 2012 (si para años mas tarde, 2006, 2007 ver top tables).

* Destaca que en los últimos años en la población de S (Sh y Slow) encontrammos de los valores mas altos de RWI

## Tree-growht y Resilience

* 2005 redujo el BAI hasta 72 %
* 2012 redujo el BAI hasta 82 % (sorprende differencias entre reducción del BAI y del EVI --> aunque no parezca una gran reducción del EVI, hay que tener cuidado)
* Para 2005, en el N, se redujo el BAI hasta 45 %, mientras que en el S solo hasta ~ 90 %

* La recuperación fue mejor tras 2012 (1.16) que tras 2005 (0.96). Sorprende que N se recupera muy bien, incluso en 2005, la Rc fue de 1.112. 
* Valores de Resiliencia mas bajos

# Sistema Vascular de *Q. pyrenaica*

* Ring-porous tress are assumed to be particularly sensitive to climate warming, because they operate with narrow hydraulic safety margins (see Choat et al. )[@PerezdeLis2016]

* Vessel size chronologies were related to previous summer rainfall (from González-González et al. 2014)[@PerezdeLis2016]

* Ring-porous trees yearly restore their photosynthetic apparatus and hydraulic architecture using exclusively carbon reserves from the previous season. This is particularly true for *Q. pyrenaica* whose leaves were not fully expanded until the end of earlywood formation.

* Climate signals in multi-decadal tree rings series indicate that vessel traits in ring-porous species are mostly influenced by climatic conditions in two physiological distinc periods: during the previous year and at the onset of cambial activity. `$TODO$`: Relation with lag effects of the drought

* **Previous summer rainfall** has been stated as the most relevant climatic factor determining vessel size rather than temperature in winter and spring (see González-González et al. 2014; 2015 in [@PerezdeLis2016]). Moisture availability was reported to be the most limiting factor driving radial growth in Iberian *Q. pyrenaica* populations (see Gea-Izquierdo and Cañellas 2014). :arrow_right: Vessel size could be more dependent on environmental signals operating at time windows that are different from those of vessel enlargement, such as precipitation during the previous growing season.

* Formation of narrower vessels is considered an adaptive response of trees to water stress in ring-porous species. It could be somewhat triggered by declining carbon gain after severe droughts, due to reliance of ring-porous oaks on the reserves of the previous year for earlywood formation and canopy development

# Sequía

## Summer of 2005 
* 2004/2005 hydrological year is considered one of the worst drooughts ever recorded in the Iberian Peninsula, particularly in the central and southern sectors [@GarciaHerrera2007].

* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].

* The hydrological year from October 2004 to September 2005 was the driest on record at several locations throughout Iberia [@GarciaHerrera2007].

* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].

# References

### 2018-02-02 man/ms_0_intro.Rmd
update discussion

@@
 # Intro
+
+* Analysing the impact of drought on tree growth is therefore a priority under the current climatic conditions, in which global warming is probably increasing the severity of droughts (Williams et al., 2013).
+
+En este estudio combinamos el uso de datos de dendrocronologia con información de productividad procedente de satélite.
+Por un lado, sabemos que el annual radial growth increment se utiliza como proxy para estudiar la forest vitality [@Fritts1976; @Dobbertin2005]; por otro lado el EVI para productividad ...
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+* Tree ring width or annual radial growth increment is a widely used proxy for tree vitality (ver  [@Bhuyan2017] [@Dobbertin2005]
+
+* Ver en Gea-Izquierdo 2014: El análisis del inter-annual growth variability can be used to study changes in productiviyt and forest vitality to detect vulnerability to stress (Fritts 1976; Doberti)
+
+")
+```
+
+* Diferentes estudios sugieren que, a consecuencia del cambio climático, se incrementará la severidad, la extensión geográfica y la frecuencia de las sequías [@Dai2011; @Trenberth2014]
+
+* Proyeccciones de crecimeinto de para *Q.
+pyrenaica* sugieren un declive en el crecimiento en las proximas decadas a lo largo de su rango de distribución en la P.
+Iberica, en diferentes escenarios climáticos [@GeaIzquierdo2013]: Predictions suggest that QUPY productivity would decline in the next decades all along its distributional range in the Iberian Peninsula for all the climate scenarios studied.
+Este declive debería ser mas dramático en las baja altitudes de las zonas mas calientes del sur de su distribución
+
+--> vulnerabilidad local (related to rear-edge)
+La tendencia de reducción drástica sugerida por el modelo para el dry-edge de la distribución de la especie podría expresar una aumentada vulnerabilidad de los árboles al incremento del estrés hídrico forzado por un clima mas cálido [@GeaIzquierdo2013].
+
+Negative trends over recent years and the greater response to moisture availability found at warmer and drier low elevations in the south suggest vulnerability to warming at the local low elevation dry edge of the species's range (e.g.
+QUPY9).
+Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()
+
+...
+Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
 

### 2018-02-02 man/ms_3_discussion.Rmd
update discussion

@@
 
+En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations).
+The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone.
+
+Otra cosa.
+@GeaIzquierdo2014 encontró que a lo largo del gradiente altitudinal el BAI se estabilizó o disminuyó a partir de 1970.
+DUDA: Nosotros encontramos un ligero (y débil) aumento, sobre todo en caHigh.
+
 ```{r echo=FALSE, results='asis', eval=show_text}
@@
 
+* Vicente-Serrano: 
+    * We found that some forests from cold and humid areas respond to shorter drought time-scales than forests from dry areas, which usually respond to longer time-scales
+    * Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011).
+    * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.
+
+duda --> las proyecciones de crecimiento que hiciste [@GeaIzquierdo2013] sugerían un declive en el crecimiento. Nosotros estamos obteniendo algo diferente no? DUDA
+
 ## historia forestal de ambos sitios 

### 2018-02-07 man/ms_0_intro.Rmd
update biblio

@@
 In the Iberian Peninsula, major drought episodies were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
-The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013].
+The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Pascoa2017].
 These events were extreme in both its magnitude and spatial extent [@Gouveia2014].

### 2018-02-07 man/ms_0_intro.Rmd
update

@@
 
+-- Sequías --
+
+Drought is a natural phenomenon that occurs when water availability is significantly below-normal levels [@Dai2011].
+Drought occurs over most part of the world and affects ecological systems across every climatic zone worldwide [@Mishra2010], being one of the most economically and ecologically disruptive extreme events affecting millions of people globally [@Dai2011].
+Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
+In addition, climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013] particularly for the Mediterranean area [@Spinoni2017a].
+
+In the Iberian Peninsula, major drought episodies were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
+The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Pascoa2017].
+These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
+
+Drought can cause significant changes in ecosystem productivity and water dynamics, and extreme drought can severely affect forest through changes in plant physiology, phenology and carbon allocation (see @Ummenhofer2017 for a revision).
+
+La vulnerabilidad de los bosques a los eventos climáticos extremos ...
+
+no obstante los ecosistemas situados en zonas áridas responden mas rápidamente a la sequía.. ver VicenteSerrano2013
+
+Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
+(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...
+
+...
+mejorar--> y esto puede ser especialmente importante en zonas como la región mediterránea,con un alto uso antrópico, donde la sequía presenta no solo efectos aislados, sino también la interacción con otros factores [@DoblasMiranda2017].
+(ver abstract trabajo)
+
+No solo efectos aislados, sino también en conjunción con otros factores, sobre todo en el mediterráneo donde convergen muchos de los factores que pueden interaccionar:
+
 ## 
@@
 ### Sequías 
-
-Although summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
-Además las proyecciones apuntan a que este incremento en la frecuencia continuará [@Spinoni2017a].
-
-In the Iberian Peninsula, major drought episodies were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
-The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Pascoa2017].
-These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
-
 > notas sequia
@@
 ### Extreme sequias y ecologia
-
-* ojo hablar de que la sequía es un proceso natural, pero que se está viendo aumentado en los últimos años:
-
-Drought is a natural phenomenon that occurs when water availability is significantly below normal levels over a long period and the supply cannot meet the existing demand.
-
-> ver intro Ecosphere:
-> Drought affects ecological systems across every climatic zone worldwide and is exacerbated by climate change and increasing anthropogenic water demands (Mishra and Singh 2010).
-Characterized by below-normal precipitation (Dai 2011), meteorological drought results from complex interactions between the atmosphere and hydrologic processes within the biosphere.
-Unlike aridity, which is a permanent feature of climate (Wilhite 1992), drought is a temporary extreme event (Palmer 1965, Mishra and Singh 2010) that can persist for extended time periods (months to years; Mishra and Singh 2010).
-Drought can cause significant changes in ecosystem productivity and water dynamics, and it is one of the most economically and ecologically disruptive extreme events affecting millions of people globally (Dai 2011).
-
-Extreme climate events (e.g.
-droughts) severely affect forests and grasslands throuhg changes in plant physiology, phenology and carbon allocation [@Ummenhofer2017] (incluir citas para los efectos de la sequía sobre la vegetación).
 
@@
 
-Climate change projections indicate that extreme events will become more common in the future (IPCC 2013), making it important that we understand how ecosystems respond to these events and the potential feedbacks to radiative forcing.
-
 ### 
@@
 ```
+
+.. ver [@VicenteSerrano2013] impacto de las sequías en diferentes zonas (i.e montañas)
+
+Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @AghaKouchak2015; @Norman2016]
+
+... ojo los remote sensing datos (aun con alta resolucion temporal y/o espacial) son insuficientes por si solos para demostrar 
+ Remote sensing data, even high temporal or spatial resolution data, are insufficient alone to unequivocally demonstrate that drought is the causal agent of a particular change in reflectance values
 

### 2018-02-12 man/ms_1_material_methods.Rmd
add review of GGI

@@
 
-The Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula [@Franco1990] reaching its southern limit in north of Morocco.
-In the Iberian Peninsula these forests live under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002] living on siliceous soils, or soils poor in basic ions [@Vilches2014].
+Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in northern Morocco [@Franco1990].
+In the Iberian Peninsula these forests live on siliceous soils, or soils poor in basic ions [@Vilches2014]
+under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002].
 *Q.
-pyrenaica* requires between 650 and 1200 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species [@Gavilan2007; @delRio2007].
+pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species [@Gavilan2007; @delRio2007].
 
-This species reaches its southernmost European limit at Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*.
+This species reaches its southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*.
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
-There are eight oak patches (2400 Has) identified (Figure 1) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and generally associated to major river valleys.
-Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species during glaciation [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].
+There are eight oak patches (2400 Has) identified (Figure 1) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
+Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].
 
-The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], undergoing intensive anthropic use in the last few decades [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
+The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], undergoing intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
 In fact, the status of conservation of this species for southern Spain is "Vulnerable" [@Vivero2000].
-The relict presence of this species in Sierra Nevada is related both to its genetic resilience as well as to its high intraspecific genetic diversity [@Valbuena2010].
-However, they are also expected to suffer the impact of climate change, due to their climate requirements (wet summers).
+The relict presence of this species in Sierra Nevada is related both to its genetic resilience and high intraspecific genetic diversity [@Valbuena2010].
+However, they are also expected to suffer the impact of climate change, due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species.
 Thus, simulations of the climate change effects on this habitat forecast a reduction in suitable habitats for Sierra Nevada [@Benito2011].
@@
 
-The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years particularly for the southern of the Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and the were charecterized by extreme drogught (Figure S1).
+The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and they were charecterized as extreme drought in our climatic data (Figure S1).
 Aunque sabemos que existen otros años de sequía extrema (e.g.
@@
 
-To characterize the vegetation greenness of *Q.
+As a proxy to primary growth, to characterize the vegetation greenness of *Q.
 pyrenaica* we used the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
@@
 We explored the temporal trends of EVI at a pixel scale.
-We computed the nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945), using the `trend` package [@Thorsten2017].
+We computed the nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945), using the `trend` package in R [@Thorsten2017].
 We set 0.05 the alpha level for the test.
@@
 $$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
-where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for the year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for the year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
+where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
 Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2010; @Samanta2012].
@@
 ### Tree sampling 
-Sampling was carried during autumn of 2016.
+Tree sampling was carried during autumn of 2016.
 Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern aspect) and Cáñar (CA; southern aspect) (Figure 1; Table 1).
-For the southern site two elevations were sampled: CA-Low and CA-High.
+-- GGI: Diferencias microclimáticas Decir algo aquí sobre difrencias microclimáticas entre el norte (más seco, más continental?) y el sur (más húmedo pero más oceánico/cálido?).
+-- For the southern site two elevations were sampled: CA-Low and CA-High.
 All the sites were oak monospecific and representatives of two of the three population clusters identified for the species in this mountain range [@PerezLuque2015].
@@
 We computed two distance independent indices: *density* in $\mathrm{trees \cdot ha^{-1}}$, and *basal area* (BA) in $\mathrm{m^{2} \cdot ha^{-1}}$; and one distance dependent indices: size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
-We used non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] for search differences between sites for height, DBH and competence indices.
+We used non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] to analyse differences between sites for height, DBH and competence indices.
 When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values.
-These analysis were performed uisng the `PMCMR` package [@Pohlert2014].
+These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014].
 
@@
 Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
-Validation of the cross-dating was done using COFECHA software [@Holmes1983].
+Cross-dating validation was finally done using COFECHA software [@Holmes1983].
 
@@
 Residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990].
-The statistical quality of each chronology was checked via the expressed population signal [@Wigley1984].
-A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.
+The statistical quality of each chronology was checked via the expressed population signal.
+A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable [@Wigley1984].
 
 The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
-BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008].
+Theoretically, BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008].
 We used measured dbh and raw ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
@@
 
-Both RWI and BAI chronologies were compared to different climatic indices to analyse the tree low and the high-frequency response to climate.
-Statistical descriptive parameters [@Fritts1976] including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology.
+Both RWI and BAI chronologies were compared to different climatic indices to analyse the low and the high-frequency response to climate of trees.
+Statistical descriptive parameters, including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology [@Fritts1976].
 
@@
 
-The relationships between residual site chronologies and the climatic variables were assessed by a bootstrapped Pearson's correlation estimate using the `treeclim` package [@Zang2015].
-The bootstrapped confidence intervals were used to estimate the significance (p < 0.05) of the correlation coefficients.
+Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using the `treeclim` package [@Zang2015].
+The bootstrapped confidence intervals were used to estimate significance (p < 0.05) of the correlation coefficients.
 
@@
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
-To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % (DUDA) of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
+To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
 
@@
 We computed the values of these indices for tree growth (BAI) and greenness during each drought event.
-We considered 2005 and 2012 as singles drought events.
+We considered 2005 and 2012 as single drought events.
 The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
@@
 
-We used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and oak population (northern and southern slopes) for each of the three resilience indices studied.
-These test were used beacuse original and log-transformed data both did not match the assumptions of normality and homogeneity of variance [@Wilcox2012].
+We used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects) for each of the three resilience indices studied.
+These test were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012].
 Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012].

### 2018-02-12 man/ms_2_results.Rmd
add review comments of GGI

@@
 
-## Vegetation Greenness
+## Time trends in vegetation greenness
 
-Trends analysis showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
-The strongest trends were observed in southwestern populations (Figure S2).
-Vegetation greenness of *Q.
+The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
+The strongest trends were observed in southwestern populations (Figure S2), in which the Cañar site are included Vegetation greenness of *Q.
 pyrenaica* forests were lower during both the 2005 and 2012 years than the greenness observed for the reference period (Figure 2a).
-The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b).
+The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b).
 Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought (Figure 2b).
@@
 According to the standardized anomalies, *Q.
-pyrenaica* forests suffered a browning during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), however no changes were observed for the 2012 drought (Figure S3).
+pyrenaica* forests suffered a "browning episode" during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S3).
 
@@
 
-## Responses to drougth: greenness
+## Greenness resilience to drought events
 
 *Q.
-pyrenaica* forest showed significantly lower resistance to 2005 drought event than to 2012 one (Table 3; Figure 3a).
-The 2005 drought reduced the greenness of oak to 85.8 % while the 2012 reduced 94.3 % (Table S1).
+pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3a).
+The 2005 drought reduced the greenness of oaks to a 85.8 % respect the reference period while the 2012 reduced it 94.3 % (Table S1).
 Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1.
@@
 
-The oak forests recovered their greenness significantly more rapidly after the 2005 drought than after 2012 (Tables 3, S1).
-In the period after 2005 drought, greeness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
+The oak forests recovered (*Rc*) their greenness significantly more rapidly after the 2005 drought than after 2012 (Tables 3, S1).
+In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
 A similar recovery after the 2005 and 2012 drought event was observed for southern populations (p = 0.2453; Figure 4a), whilst the northern populations showed a significantly greater recovery after the 2005 drought than after the 2012 drought (Figure 4a).
@@
 
-Resilience values were significantly higher for the 2012 drought event than for 2005 (Tables 3, S1), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
+Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables 3, S1), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
 The southern populations showed higher resilience values than the northern ones, although they were not significantly different for 2005 drought (p = 0.036; Figures 3a, 4a).
@@
 
-## Growth trends
-The trees of the southern population were older and larger than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2).
-Competence indices were similar among sites (Table 1).
-The basal area increment (BAI) were different between sites (Figure 5).
-El sitio mas elevado (Ca-High) mostró mayores valores de BAI que los otros sitios, mostrando además una tendencia positiva a partir de 1970.
-Se observan diferencias en cuanto al crecimiento entre los sitios de las poblaciones del sur.
-Ca_Low se parece en crecimiento mas a SJ que a Ca_High.
+## Radial growth trends and climate
+
+Likely as a consequence of their different management origin (i.e.
+land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2).
+The competition index was similar among sites but plot basal area was greatest in Ca-High (Table 1).
+Yet the basal area increment (BAI) growth in the highest plot in altitude (Ca-High) was the greatest  (Figure 5).
+Trees from this location showed in addition a positive growth trend since the late 1970s.
+This positive trend was not found in any of the other two locations and the two altitudes sampled in the Southern populations showed differences in growth, being more similar Ca_Low with growth in the Northern population (SJ) than with Ca_High (Figure 5).
+`VAS POR AQUI CARPINTO (BREAK)`
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+> Relative Resilience (De esto no he dicho nada)
+
 Sorprende que en los sitios del sur (caH y caL), en los últimos años ha habido crecimientos muy grandes.
@@
 Algo similar ocurrió en sj para los años 2003 y 2013.
+")
+```
 
-## Responses to drougth: tree-growth
+## Resilience of radial tree-growth to drougth events
 
-Although no significant differences were observed in the resistance of oak growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
-The northern site, which is under a drier cliamte, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figures 3b, 4b).
+Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
+Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figures 3b, 4b).
 
@@
 
-Oak forests recovered their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1 (Table S2).
+Oak forests recovered (*Rc*) their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1 (Table S2).
 Northern populations showed signigicantly higher values of recovery than southern ones (Tables 3 and S2), which even not reach recovery values of Rc = 1 (Figure 3b, 4b).
@@
 ## Climate and tree-growth
-Tree-growth exhibited a significant greater response to water availability more than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
-Cumulative precipitation of the hydrological year was the climatic variable with a higher relationship with growth and influenced positively the tree-growth for the both northern and southern populations (Figure 7a).
+Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
+Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 7a).
 Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
-Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for northern population (r > 0.6), which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
-Temperatures were much less related to growth than moisture variables.
-Yet spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 7c and 7d).
-Surprisingly minimum and maximum temperatures of current September positively influenced the tree-growth in the northern population.
+Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for the northern population (r > 0.6).
+Relationships with temperatures were lower than those with moisture variables.
+Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 7c and 7d).
+Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population.
+No positive relationship with temperature was found for southern trees.
 
@@
 El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
-Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
 Estos periodos se alternan con periodos de supresión.

### 2018-02-12 man/ms_figures.Rmd
include suggestions by GGI

@@
 **Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain range, where three cluster of oak populations have been identified [@PerezLuque2015] (showed in different colour) (b).
+Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain range, where three clusters of oak populations have been identified [@PerezLuque2015] (showed in different colour) (b).
 A grid of with the MODIS pixels for each population is shown (see material and methods).
-Detailed location of the sampling sites: northern (San Juan, SJ) (c) and southern ones (Cáñar: CALow and CAHigh)(d).
+Detailed location of the sampling sites: northern (San Juan, SJ) (c) and southern ones (Cáñar: CALow and CAHigh) (d).
 Colour Orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
@@
 
-**Figure 2.** Comparison of EVI profile for the reference period (*gray*) and during the 2005 (*green*) and 2012 (*blue*) drought events (a).
+**Figure 2.** Comparison of EVI profile for the reference period (*gray*) and during the 2005 (*green*) and 2012 (*blue*) drought events (a) (mean of all analyzed pixels).
 EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations (b).
@@
 **Figure 3.** Response of northern (*black*) and southern (*blue*) populations of *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; a) and tree growth (BAI; b) for the years 2005 and 2012.
+pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) for the years 2005 and 2012.
 Different letters above bars indicate significant post hoc differences between groups (see material and methods).
@@
 **Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree growth (BAI; b) by site.
+pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
 Different letters above bars indicate significant post hoc differences between groups (see material and methods).
@@
 
-pgc_median <- gc_treesMed %>% ggplot() +
+pgc_median50 <- gc_treesMed50 %>% ggplot() +
   geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
@@
                                         fill='white')) 
-pgc_median
+pgc_median50
 
@@
 
-**Figure 9**. Percentage of *Q. pyrenaica* trees affected by GC > 25 % by site. *Black* line shows number of trees (rigth-axis).
+**Figure 9**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).

### 2018-02-12 analysis/drought/explore_spei_longterm.Rmd
update

---
title: "SPEI_longterm"
author: "Antonio J. Perez-Luque"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("here")
library("dplyr")
library("tidyverse")
library("stringr")
library("pander")
library("magrittr")

# machine <- 'ajpeluLap'
# machine <- 'ajpelu'
```

## Read and prepare data 
```{r}

# Get file names
myfiles <- list.files(path = here("data_raw/spei/spei_longterm"), pattern = "\\.csv$")

mydf <- data.frame()

for (j in myfiles){ 
  aux <- read.csv(file=here("data_raw/spei/spei_longterm", j), header = TRUE, sep = ',')

  # Remove csv and get name 
  name_aux <- str_replace(j, ".csv", "")

  # Get lat long
  latlong <- str_replace(name_aux, "spei_", "")
  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[2]

  # Store lat and long 
  aux$lat <- mylat
  aux$long <- mylong

  # year and month 
  aux$month <- rep(1:12, 115)

  auxy <- c()
  years <- c() 
  for (j in 1901:2015) { 
    auxy <- as.data.frame(rep(j,12))
    years <- rbind(years, auxy)
    } 
  names(years) <- 'year'

  aux <- cbind(aux,years)

  # join all df 
  mydf <- rbind(mydf, aux)
}

# Convert to date  
mydf <- mydf %>% mutate(date = sprintf("%d-%02d", year, month),
                        date = as.Date(paste0(date, '-01'), format="%Y-%m-%d"))
```

```{r}
speidf <- mydf %>% dplyr::select(SPEI06,SPEI12,SPEI24,month, year, date) %>% 
  group_by(year,month, date) %>%
  summarise_at(vars(SPEI06,SPEI12,SPEI24), funs(mean,sd,se=sd(.)/sqrt(n()))) %>% 
  as.data.frame() 
```

# Plot 
```{r}
speiPlot <- function(df, vi, mytitle, year_scaleX){
  require(dplyr)
  require(scales)

  # Select variable of interest 
  dfs <- df %>% dplyr::select(year, month, date, starts_with(vi))

  # Rename variables 
  dfs <- dfs %>% rename_all(~sub(paste0(vi,'_'), '',.x))
  # Create variable pos and neg 
  dfs <- dfs %>% mutate(signo = ifelse(mean > 0, "pos", "neg"))

  # Plot  
  p <- ggplot(dfs, aes(x = date, y = mean, fill = signo)) + 
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
    theme_bw()  + 
    labs(title= mytitle, x= '', y = paste0('z-scored (',vi,')')) +
    theme(legend.position = "none")

  # Date Scale in X 
  if (missing(year_scaleX)){
    p
    } else { 
    year_interval <- paste0(year_scaleX," year")
    p <- p + scale_x_date(date_breaks = year_interval, date_labels = "%Y")
    }

  return(p)
}
```

```{r}
speiPlot(speidf, vi = 'SPEI12', mytitle = "SPEI for SN", year_scaleX = 20)
```

# Analysis

## Computo de indicadores

Vamos a calcular para el promedio espacial de la serie (valores medios de los pixeles), los eventos de sequía. El sesgo de la agregación espacial de estos datos puede ser un problema a la hora de comparar entre regiones, pero puede corregirse ([@Spinoni2015b]). No obstante en nuestro caso no vamos a realizar comparaciones entre regiones, por lo que utilizaremos la agregación espacial (un promedio) para dar una visión general de la sequía en Sierra Nevada.

* Hemos utilizado una categorización mixta, así para valores de sequía hemos seguido las categorías propuestas por [@Agnew2000] y para los valores positivos, la clasificación de [@Mckee1993]
* Un evento de sequía comienza cuando el valor del índice está por debajo de un umbral [@Pascoa2017] (en este caso < -0.84 moderate, < -1.28 severe y < -1.65 extreme)
* Utilizaremos la escala de 6 y 12 de SPEI
* Computaremos:

    * Duración del evento de sequía (*drought duration*, *dd*): número de meses consecutivos con el valor de spei por debajo de un umbral. 
    * Intensidad (*drought intensity*, *di*): valor medio del spei durante el periodo de sequía.
    * Severidad (*drought severity*, *ds*): suma de los valores de spei (en valores absolutos) durante el periodo de sequía 
    * Inicio (*start*, *dstart*): mes de inicio del periodo de sequía
    * Fin (*end*, *dend*): mes de fin del periodo de sequía 
    * Frecuencia de sequías (*drought frequency*, *df*): número de eventos de sequía por década.

Consideramos un evento cuando se dan las condiciones en al menos dos meses consecutivos (ver [@Spinoni2014; @Pascoa2017]). Para mas info ver [@Pascoa2017; @Spinoni2014; @Spinoni2015]

```{r}
droughtIndicators <- function(df, vname, threshold){
  require(dplyr)
  require(magrittr)

outlist <- c()

# https://github.com/tidyverse/dplyr/issues/1661

# Compute months under threshold 
out <- df %>% mutate(
  is_drought = ifelse(
    .[[vname]] < threshold & lead(.[[vname]] < threshold), 1, 0))

# Identified the different events and numbered them
# Compute lenght of drought events

out2 <- out %>% 
  group_by(index_events = rle(is_drought) %>% extract2("lengths") %>% rep(seq_along(.), .)) %>% 
  mutate(drought_duration = sum(is_drought)) %>% 
  as.data.frame()

# Filter events with drought duration > 1 
out3 <- out2 %>% filter(drought_duration > 1) %>% as.data.frame()

# Compute several indicators (drought assessments) 
da <- out3 %>% group_by(index_events) %>% 
  summarise(d_duration = unique(drought_duration),
            d_intensity = mean(spei, na.rm=TRUE),
            d_severity = sum(abs(spei), na.rm=TRUE),
            lowest_spei = min(spei),
            month_peak = month[which.min(spei)], 
            minyear = min(year),
            maxyear = max(year)) %>%
  as.data.frame()

outlist$data <- out2
outlist$drought_events <- out3
outlist$drought_assessment <- da

return(outlist) 
} 
```

```{r}
# Compute for all pixels (mean values of SN) only for scales 6 and 12

spei6avg <- speidf %>% dplyr::select(year, month, date, spei = SPEI06_mean) 
spei12avg <- speidf %>% dplyr::select(year, month, date, spei = SPEI12_mean)

da6_mod <- droughtIndicators(spei6avg, threshold = -0.84, vname = 'spei') 
da12_mod <- droughtIndicators(spei12avg, threshold = -0.84, vname = 'spei')

da6_sev <- droughtIndicators(spei6avg, threshold = -1.28, vname = 'spei') 
da12_sev <- droughtIndicators(spei12avg, threshold = -1.28, vname = 'spei')

da6_ext <- droughtIndicators(spei6avg, threshold = -1.65, vname = 'spei') 
da12_ext <- droughtIndicators(spei12avg, threshold = -1.65, vname = 'spei')

# Join all moderate and compute decade (# https://stackoverflow.com/questions/35352914/floor-a-year-to-the-decade-in-r

mod <- rbind(da6_mod$drought_assessment %>% mutate(variable= 'spei6'),
      da12_mod$drought_assessment %>% mutate(variable= 'spei12')) %>% 
  mutate(cat = 'moderate', decade = (minyear %/% 10) * 10)

sev<- rbind(da6_sev$drought_assessment %>% mutate(variable= 'spei6'),
      da12_sev$drought_assessment %>% mutate(variable= 'spei12')) %>% 
  mutate(cat = 'severe', decade = (minyear %/% 10) * 10)

ext <- rbind(da6_ext$drought_assessment %>% mutate(variable= 'spei6'),
      da12_ext$drought_assessment %>% mutate(variable= 'spei12')) %>% 
  mutate(cat = 'extreme', decade = (minyear %/% 10) * 10)

# Compute DF (*drought frequency*)

dfmod <- mod %>% group_by(variable, decade, cat) %>% summarize(df = n())
dfsev <- sev %>% group_by(variable, decade, cat) %>% summarize(df = n())
dfext <- ext %>% group_by(variable, decade, cat) %>% summarize(df = n())

df <- dfmod %>% bind_rows(dfsev, dfext)
```

Utilizar SPEI 12

```{r}
sev12 <- sev %>% filter(variable == "spei12") %>% arrange(desc(d_duration)) %>% 
  dplyr::select(-index_events, -cat, -decade, -variable)

pander(sev12)

```

### 2018-02-12 man/ms_3_discussion.Rmd
update

@@
 
+### OTRA COSA 
+#### Climate and tree relationship 
+Frase de resultados que tenemos que poner en discussion: 
+Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
+Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for the northern population (r > 0.6), ...
+**which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]**.
+#### Disturbance
+Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
+
 ### ¿Que factor es mas limitante para el crecimiento en el rear-edge de Q. pyrenaica? 

### 2018-02-12 man/ms_figures.Rmd
update

@@
 
+cro_rwi <- cro_rwi %>% 
+  mutate(siteF = case_when(
+    site_sorted == "0_sj" ~ "Northern (SJ)",
+    site_sorted == "1_caH" ~ "Southern High (Ca-High)",
+    site_sorted == "2_caL" ~ "Southern Low (Ca-Low)"))
+
 # 
@@
         strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        axis.text = element_text(face="bold", size=10), 
+        # strip.text = element_text(face="bold", size=12), 
+        strip.text = element_blank(), 
+        axis.text = element_text(face="bold", size=8), 
         axis.title = element_text(face="bold", size=12)) + 
@@
   geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "fixed", ncol = 1,
+  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
              labeller = as_labeller(label_site)) + 
@@
   geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
+  # geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
   geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
   scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
+                     breaks= seq(1820, 2020, by=20)) +
+  annotation_custom(x=1980, y = 1.6)
 

### 2018-02-12 man/ms_supp.Rmd
update

@@
 The boxplot representing the average from 1940-2015 period.
-Data from meteorological station Granada, Base Aérea (National Spanish Meteorological Services (AEMET)).
+Data from meteorological station Granada, Base Aérea (National Spanish Meteorological Services, AEMET).
 
@@
 ```{r}
-df <- read.csv(file='../out/anomalies/evi/percen_browing_greening.csv', header=TRUE)
+df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
 
@@
   geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), vjust= - 0.75) + 
+  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
   facet_wrap(~clu_pop2) + 
@@
   theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + theme_minimal() + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B"))
+        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
+  scale_y_continuous(expand=c(.1,0))
 

### 2018-02-12 analysis/bai/trends_resilience_bai.Rmd
include spei severe droughts and resilience trends

--- 2017-09-15 analysis/analysis_resilience_bai.Rmd
@@
 ---
-title: "Growth Resilience"
+title: "BAI resiliences ten droughts"
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 June"
 output: 
@@
 ```{r global_options, include=FALSE}
-knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, error=TRUE, message=FALSE)
+knitr::opts_chunk$set(fig.width=12, fig.height=8, warning=FALSE, error=TRUE, message=FALSE)
 ```
 
-```{r, warning=FALSE, message=FALSE}
+```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
+library("here")
 library("tidyverse")
 library("dplR")
-library("stringr")
 library("knitr")
@@
 library('devtools')
-# devtools::install_github("ajpelu/auxiliar")
-library('car')
-library('auxiliar')
-library('WRS2')
-library('MASS')
-library('rcompanion')
 ```
@@
 
-```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
+```{r}
+# machine <- 'ajpelu'
+
+machine <- 'ajpeluLap'
+
+di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
 
@@
 # Drought years
-dyears <- c(1995, 2005, 2012)
 
-# SJ 
-res_4_sj <- baiResilience(bai_sj, event_years = dyears, window = 4)
-res_3_sj <- baiResilience(bai_sj, event_years = dyears, window = 3)
-res_2_sj <- baiResilience(bai_sj, event_years = dyears, window = 2)
+dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
 
-# caL
-res_4_caL <- baiResilience(bai_caL, event_years = dyears, window = 4)
-res_3_caL <- baiResilience(bai_caL, event_years = dyears, window = 3)
-res_2_caL <- baiResilience(bai_caL, event_years = dyears, window = 2)
+dy <- dy %>% arrange(desc(maxyear))
 
-# caH
-res_4_caH <- baiResilience(bai_caH, event_years = dyears, window = 4)
-res_3_caH <- baiResilience(bai_caH, event_years = dyears, window = 3)
-res_2_caH <- baiResilience(bai_caH, event_years = dyears, window = 2)
-```
+dys <- dy %>% filter(maxyear > 1940) %>% as.data.frame()
 
-### Computar correlaciones ventanas temporales 
-```{r}
-# Vector with objects name
-obj <- c('res_2_sj', 'res_3_sj', 'res_4_sj',
-         'res_2_caL', 'res_3_caL', 'res_4_caL',
-         'res_2_caH', 'res_3_caH', 'res_4_caH')
+dys <- dys[,'maxyear']
 
-correla_ws <- c()
-
-for (i in obj){ 
-  x <- get(i)
-  xres <- x$resilience
-  out <- xres %>% 
-    mutate(ws = paste0('ws_', as.character(str_extract(i, "([0-9])"))),
-           site = str_replace(i, "res_[0-9]_", '')) %>%
-    dplyr::select(-disturb_year, -tree)
-
-  correla_ws <- bind_rows(correla_ws, out)
-
-}
-
-# Split by window size 
-correla <- correla_ws %>% split(.$ws)
-
-# Change names 
-names(correla[["ws_2"]])[1:4] <- paste0(names(correla[["ws_2"]])[1:4], '2')
-names(correla[["ws_3"]])[1:4] <- paste0(names(correla[["ws_3"]])[1:4], '3')
-names(correla[["ws_4"]])[1:4] <- paste0(names(correla[["ws_4"]])[1:4], '4')
-
-cor2 <- correla[["ws_2"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-cor3 <- correla[["ws_3"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-cor4 <- correla[["ws_4"]] %>% dplyr::select(-ws) %>% mutate(ind = row_number())
-
-correlations <- inner_join(cor2, cor3, by='ind') %>% inner_join(cor4, by='ind')
-
-# Resistance
-aux_coefs <- c()
-
-model <- lm(rt2~rt3, data=correlations)
-p_rt23 <- correlations %>% ggplot(aes(rt2, rt3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rt','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rt2~rt4, data=correlations)
-p_rt24 <- correlations %>% ggplot(aes(rt2, rt4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rt','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rt3~rt4, data=correlations)
-p_rt34 <- correlations %>% ggplot(aes(rt3, rt4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rt (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rt','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rt23, p_rt24, p_rt34,ncol=3)
-
-# Recovery 
-model <- lm(rc2~rc3, data=correlations)
-p_rc23 <- correlations %>% ggplot(aes(rc2, rc3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rc','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rc2~rc4, data=correlations)
-p_rc24 <- correlations %>% ggplot(aes(rc2, rc4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rc','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rc3~rc4, data=correlations)
-p_rc34 <- correlations %>% ggplot(aes(rc3, rc4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rc (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rc','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rc23, p_rc24, p_rc34,ncol=3)
-
-# Resilience
-model <- lm(rs2~rs3, data=correlations)
-p_rs23 <- correlations %>% ggplot(aes(rs2, rs3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rs','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rs2~rs4, data=correlations)
-p_rs24 <- correlations %>% ggplot(aes(rs2, rs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rs','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rs3~rs4, data=correlations)
-p_rs34 <- correlations %>% ggplot(aes(rs3, rs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rs','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rs23, p_rs24, p_rs34,ncol=3)
-
-# Relative Resilience
-model <- lm(rrs2~rrs3, data=correlations)
-p_rrs23 <- correlations %>% ggplot(aes(rrs2, rrs3)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) + 
-  theme(legend.position = c(.2, .75))
-aux <- as.data.frame(cbind('rrs','2-3', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rrs2~rrs4, data=correlations)
-p_rrs24 <- correlations %>% ggplot(aes(rrs2, rrs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rrs','2-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-model <- lm(rrs3~rrs4, data=correlations)
-p_rrs34 <- correlations %>% ggplot(aes(rrs3, rrs4)) + 
-  geom_point(aes(colour=site.x)) + theme_bw() + geom_smooth(method = 'lm', se=FALSE) + 
-  ggtitle(paste('rrs (R2) = ', round(summary(model)$r.squared, 3))) +
-  theme(legend.position = 'none')
-aux <- as.data.frame(cbind('rrs','3-4', as.numeric(summary(model)$r.squared)))
-aux_coefs <- rbind(aux_coefs, aux)
-
-grid.arrange(p_rrs23, p_rrs24, p_rrs34,ncol=3)
-
-names(aux_coefs) <- c('var', 'window_size', 'r2')
-
-write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), row.names = F)
+# ojo quito 1985 
+dys <- dys[which(dys!=1985)] 
 ```
@@
 ```{r}
-aux_coefs %>% pander()
+# SJ 
+res_3_sj <- baiResilience(bai_sj, event_years = dys, window = 3)
+# caL
+res_3_caL <- baiResilience(bai_caL, event_years = dys, window = 3)
+# caH
+res_3_caH <- baiResilience(bai_caH, event_years = dys, window = 3)
 ```
-
-Nos quedamos con 3 años de ventana temporal.
-
-# Plots Crecimiento
-## Boxplot with outliers 
-```{r}
-gsj <- res_3_sj$growth %>% mutate(site='SJ')
-gcaL <- res_3_caL$growth %>% mutate(site='caL')
-gcaH <- res_3_caH$growth %>% mutate(site='caH')
-
-g <- bind_rows(gsj, gcaL, gcaH)
-
-# Export csv 
-write.csv(g, file=paste0(di, 'data/resilience/crecimientos_drought.csv'), row.names = FALSE)
-
-g %>% mutate(disturb = dplyr::recode(disturb, 
-                              dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  ggplot(aes(y=mean_period/100, x=disturb)) + 
-  geom_boxplot() + 
-  facet_grid(site~disturb_year, scales='free_y') + 
-  theme_bw()
-```
-
-## Mean + se 
-```{r}
-g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  group_by(disturb, disturb_year, site) %>% 
-  summarise(mean = mean(mean_period),
-            sd = sd(mean_period),
-            se = sd/sqrt(length(mean_period))) %>%
-  ggplot(aes(y=mean/100, x=disturb)) + 
-  geom_errorbar(aes(ymin=mean/100 - se/100, 
-                    ymax=mean/100 + se/100),
-                width = 0.2) +
-  geom_point(size=2, shape=21, fill='white') + 
-  facet_grid(site~disturb_year, scales='free_y') + 
-  theme_bw() + ylab('BAI (cm2/year)') + 
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-pgrowht <- g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  group_by(disturb, disturb_year, site) %>% 
-  summarise(mean = mean(mean_period),
-            sd = sd(mean_period),
-            se = sd/sqrt(length(mean_period))) %>%
-  ggplot(aes(y=mean/100, x=disturb, colour=site)) + 
-  geom_errorbar(aes(ymin=mean/100 - se/100, 
-                    ymax=mean/100 + se/100),
-                width = 0.15) +
-  geom_point(size=3, aes(shape=site), fill='white') + 
-  geom_line(aes(group=site))+
-  facet_wrap(~disturb_year, scales='free_y') + 
-  theme_bw() + ylab('BAI (cm2/year)') + 
-  theme(panel.grid = element_blank())
-
-pgrowht
-ggsave(plot=pgrowht, width=8, height = 4,
-       filename=paste0(di, 'out/fig/resilience/bai_events.pdf')) 
-```
-
-```{r}
-g %>% 
-  mutate(disturb = dplyr::recode(disturb, dr = '1_dr', pre = '0_pre', post = '2_post')) %>%
-  dplyr::group_by(site, disturb_year, disturb) %>% 
-  dplyr::summarise(mean = mean(mean_period/100), 
-                   sd = sd(mean_period/100),
-                   se = sd/sqrt(length(mean_period/100))) %>% as.data.frame() %>% pander() 
-```
-
-# Anovas Resiliencia
 
@@
 re <- bind_rows(rsj, rcaL, rcaH)
-re$disturb_year <- as.factor(re$disturb_year)
+# re$disturb_year <- as.factor(re$disturb_year)
 re$site <- as.factor(re$site)
 
-# Export csv 
-write.csv(re, file=paste0(di, 'data/resilience/resilience_bai.csv'), row.names = FALSE)
-```
+re %>% ggplot(aes(x=disturb_year, ))
 
-## Asumptions
+s <- re %>% 
+  group_by(site, disturb_year) %>% 
+  summarise_at(c("rt", "rs", "rc", "rrs"), funs(mean,sd,se=sd(.)/sqrt(n())))
 
-* Explorar si se cumplen los supuestos de normalidad y homocedasticidad. Tenemos que comprobar que cada uno de los grupos son normales (1995,2005,2012; site: SJ, CaH, CaL; e interactions)
-
-```{r}
-shapirosNormal <- function(df, resp_var, factor_vars) {
-  rv <- enquo(resp_var)
-  out <- df %>%
-    group_by_(.dots=factor_vars) %>%
-    summarise(statistic = round(shapiro.test(!!rv)$statistic,5),
-              p_value = round(shapiro.test(!!rv)$p.value,5)) %>%
-    data.frame()
-
-  return(out)
-}
-
-homogetest <- function(resp_var, factores, df){
-
-  out_factores <- c()
-
-  for (f in factores){
-    hv <- c()
-    myformula <- as.formula(paste0(resp_var, "~", f))
-    #tests
-    fk <- fligner.test(myformula, data = df)
-    lv <- leveneTest(myformula, data = df)
-    # out
-    hv$fk_stat <- fk$statistic
-    hv$fk_pvalue <- fk$p.value
-    hv$lev_stat <- lv$`F value`[1]
-    hv$lev_pvalue <- lv$`Pr(>F)`[1]
-    hv$factor <- f
-    hv <- as.data.frame(hv)
-    row.names(hv) <- NULL
-
-    out_factores <- rbind(out_factores, hv)}
-  return(out_factores)
-
-}
-```
-## Normalidad
-
-```{r}
-# See auxiliar::shapirosNormal
-### Resilience
-
-nrsA <- shapirosNormal(re, resp_var = rs, 'disturb_year')
-nrsA$var <- 'rs'
-nrsB <- shapirosNormal(re, resp_var = rs, 'site')
-nrsB$var <- 'rs'
-nrsAB <- shapirosNormal(re, resp_var = rs, c('disturb_year','site'))
-nrsAB$var <- 'rs'
-
-### Recovery
-nrcA <- shapirosNormal(re, resp_var = rc, 'disturb_year')
-nrcA$var <- 'rc'
-nrcB <- shapirosNormal(re, resp_var = rc, 'site') 
-nrcB$var <- 'rc'
-nrcAB <- shapirosNormal(re, resp_var = rc, c('disturb_year','site'))
-nrcAB$var <- 'rc'
-
-### Resistance
-nrtA <- shapirosNormal(re, resp_var = rt, 'disturb_year')
-nrtA$var <- 'rt'
-nrtB <- shapirosNormal(re, resp_var = rt, 'site') 
-nrtB$var <- 'rt'
-nrtAB <- shapirosNormal(re, resp_var = rt, c('disturb_year','site'))
-nrtAB$var <- 'rt'
-
-### Relative Resilience 
-nrrsA <- shapirosNormal(re, resp_var = rrs, 'disturb_year')
-nrrsA$var <- 'rrs'
-nrrsB <- shapirosNormal(re, resp_var = rrs, 'site') 
-nrrsB$var <- 'rrs'
-nrrsAB <- shapirosNormal(re, resp_var = rrs, c('disturb_year','site'))
-nrrsAB$var <- 'rrs'
-```
-
-```{r}
-normtestAB <- rbind(nrcAB, nrtAB, nrsAB, nrrsAB) 
-normtestAB %>% pander()
-write.csv(normtestAB, 
-          file=paste0(di, '/out/anovas_resilience/normo_disturb_year_site.csv'), row.names = F)
-```
-
-```{r}
-rm(nrcA, nrcB, nrcAB,
-   nrsA, nrsB, nrsAB,
-   nrrsA, nrrsB, nrrsAB,
-   nrtA, nrtB, nrtAB)
-```
-
-* No se cumplen los requisitos de normalidad
-
-## Heterocedasticidad
-
-```{r}
-## See auxiliar::homogetest
-factores <- c('disturb_year', 'site', 'interaction(disturb_year, site)') 
-responses <- c('rs', 'rc', 'rt', 'rrs')
-homo <- c()
-
-for (i in responses){ 
-  ht <- homogetest(resp_var = i, factores = factores, df = re)
-  ht <- ht %>% mutate(response = i)
-  homo <- rbind(homo, ht)
-}
-
-homo %>% pander()
-write.csv(homo, 
-          file=paste0(di, '/out/anovas_resilience/homocedasticidad.csv'), row.names = F)
-```
-
-* Tampoco se cumplen los requisitos de homogeneidad de varianzas entre grupos
-
-* Probamos a transformar los datos con log y reanalizar los supuestos de homocedasticidad
-
-```{r}
-factores <- c('disturb_year', 'site', 'interaction(disturb_year, site)') 
-responses <- c('logrs', 'logrc', 'logrt', 'logrrs')
-homo_log <- c()
-
-re <- re %>% 
-  mutate(
-    logrs = log(rs),
-    logrc = log(rc),
-    logrt = log(rc),
-    logrrs = log(rrs)
-)
-
-for (i in responses){ 
-  ht <- homogetest(resp_var = i, factores = factores, df = re)
-  ht <- ht %>% mutate(response = i)
-  homo_log <- rbind(homo_log, ht)
-}
-
-homo_log %>% pander()
-write.csv(homo_log, 
-          file=paste0(di, '/out/anovas_resilience/homocedasticidad_log.csv'), row.names = F)
-```
-
-* Tampoco se cumplen
-
-## Custom functions
-
-```{r}
-# Custom Function to compute ANOVAS
-aovas <- function(df, vars, resp_var){ 
-  require('dplyr')
-  require('broom')
-
-  # Create subset 
-  dfsel <- df %>% dplyr::select_(.dots=c(vars, resp_var))
-
-  # Model 
-  myformula <- as.formula(paste0(resp_var,  " ~ ",
-                                 paste(vars, collapse = '*')))
-
-  mymodel <- aov(myformula, data=dfsel)
-
-  # Output model Summary http://my.ilstu.edu/~wjschne/444/ANOVA.html#(1)
-  model_coeff <- broom::tidy(mymodel)
-  model_summary <- broom::glance(mymodel)
-
-  out <- c() 
-  out$model_coeff <- model_coeff
-  out$model_summary <- model_summary
-  out$mymodel <- mymodel
-
-  return(out)
-}
-
-# Post-Hoc comparison
-phc <- function(mymodel, resp_var){
-  require(lsmeans)
-
-  # Disturb Event 
-  ph_event <- lsmeans(mymodel, pairwise ~ disturb_year, adjust = "bon")
-
-  # differences letters 
-  cld_event <- cld(ph_event, alpha   = 0.01, 
-                   Letters = letters, 
-                   adjust  = "bon")
-
-  # Site  
-  ph_site <- lsmeans(mymodel, pairwise ~ site, adjust = "bon")
-  cld_site <- cld(ph_site, alpha   = 0.01, 
-                 Letters = letters, 
-                 adjust  = "bon")
-
-  # interaction 
-  ph_i <- lsmeans(mymodel, pairwise ~ disturb_year:site, adjust = "bon")
-  cld_i <- cld(ph_i, alpha   = 0.01, 
-                 Letters = letters, 
-                 adjust  = "bon")
-
-  # Objets for plot
-  aux_ph_site <- as.data.frame(summary(ph_site$lsmeans)) 
-  aux_ph_site <- aux_ph_site %>% mutate(var = resp_var)
-  aux_ph_event <- as.data.frame(summary(ph_event$lsmeans)) 
-  aux_ph_event <- aux_ph_event %>% mutate(var = resp_var)
-  aux_ph_i <- as.data.frame(summary(ph_i$lsmeans)) 
-  aux_ph_i <- aux_ph_i %>% mutate(var = resp_var)
-
-  # Return objects
-  cat('\n### Event ###\n')
-  print(ph_event)
-  print(cld_event)
-  cat('\n### Clu pop ###\n')
-  print(ph_site)
-  print(cld_site)
-  cat('\n### Event:Clu pop ###\n')
-  print(ph_i)
-  return(list(aux_ph_site, aux_ph_event, aux_ph_i, cld_site, cld_event, cld_i))
-}
-```
-
-**OJO SOLO 2005 y 2012**
-
-```{r}
-# Only 2005 and 2012
-re <- re %>% filter(disturb_year != 1995) %>% as.data.frame()
-vars <- c('disturb_year','site')
-re$disturb_year <- factor(re$disturb_year)
-```
-
-## Recovery 
-```{r}
-# Variable
-resp_var <- 'rc'
-
-# AOV
-aov_rc <- aovas(re, vars=vars, resp_var = resp_var)
-
-mc <- aov_rc$model_coeff
-
-pander(mc, round=5,
-       caption = paste0("ANOVA table: ", resp_var), missing = '', 
-       emphasize.strong.cells = 
-         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-
-gm <- aov_rc$model_summary
-
-gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
-colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
-rownames(gm) <- "Statistic"
-pander(t(gm)) 
-```
-
-```{r}
-# Post hoc Define model
-mymodel <- aov_rc$mymodel
-postH_rc <- phc(mymodel = mymodel, resp_var = resp_var)
-```
-
-```{r, echo=FALSE}
-#### ~ Site
-ps <- plot(effect("site",mymodel))
-#### ~ Disturb Year
-pd <- plot(effect('disturb_year', mymodel))
-#### Disturb Year:Site
-picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
+s %>% 
+  ggplot(aes(y=rs_mean, x=disturb_year, colour=site)) + 
+  geom_errorbar(aes(ymin=rs_mean - rs_sd, 
+                    ymax=rs_mean + rs_sd),
+                width = 0.2, 
+                position=position_dodge(2)) +
+  geom_point(size=2, shape=21, fill='white', position=position_dodge(2)) + 
+  #facet_wrap(~site, scales='fixed', ncol=1) + 
+  theme_bw() # + geom_smooth(aes(colour=site), se=FALSE)
 
 ```
-
-```{r, fig.width = 4, fig.height=4}
-ps
-```
-
-```{r, fig.width = 4, fig.height=4}
-pd
-```
-
-```{r, fig.width = 4, fig.height=4}
-picollapse
-```
-
-```{r, fig.width = 6}
-pi
-```
-
-## Resistance 
-```{r, echo =FALSE}
-# Variable
-resp_var <- 'rt'
-vars <- c('disturb_year','site')
-
-# AOV
-aov_rt <- aovas(re, vars=vars, resp_var = resp_var)
-
-mc <- aov_rt$model_coeff
-
-pander(mc, round=5,
-       caption = paste0("ANOVA table: ", resp_var), missing = '', 
-       emphasize.strong.cells = 
-         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-
-gm <- aov_rt$model_summary
-
-gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
-colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
-rownames(gm) <- "Statistic"
-pander(t(gm)) 
-```
-
-```{r}
-# Post hoc Define model
-mymodel <- aov_rt$mymodel
-postH_rt <- phc(mymodel = mymodel, resp_var = resp_var)
-```
-
-```{r, echo=FALSE}
-#### ~ Site
-ps <- plot(effect("site",mymodel))
-#### ~ Disturb Year
-pd <- plot(effect('disturb_year', mymodel))
-#### Disturb Year:Site
-picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
-
-```
-
-```{r, fig.width = 4, fig.height=4}
-ps
-```
-
-```{r, fig.width = 4, fig.height=4}
-pd
-```
-
-```{r, fig.width = 4, fig.height=4}
-picollapse
-```
-
-```{r, fig.width = 6}
-pi
-```
-
-## Relative Resilience 
-```{r, echo=FALSE}
-# Variable
-resp_var <- 'rrs'
-
-# AOV
-aov_rrs <- aovas(re, vars=vars, resp_var = resp_var)
-
-mc <- aov_rrs$model_coeff
-
-pander(mc, round=5,
-       caption = paste0("ANOVA table: ", resp_var), missing = '', 
-       emphasize.strong.cells = 
-         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-
-gm <- aov_rrs$model_summary
-
-gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
-colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
-rownames(gm) <- "Statistic"
-pander(t(gm)) 
-```
-
-```{r}
-# Post hoc Define model
-mymodel <- aov_rrs$mymodel
-postH_rrs <- phc(mymodel = mymodel, resp_var = resp_var)
-```
-
-```{r, echo=FALSE}
-#### ~ Site
-ps <- plot(effect("site",mymodel))
-#### ~ Disturb Year
-pd <- plot(effect('disturb_year', mymodel))
-#### Disturb Year:Site
-picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
-
-```
-
-```{r, fig.width = 4, fig.height=4}
-ps
-```
-
-```{r, fig.width = 4, fig.height=4}
-pd
-```
-
-```{r, fig.width = 4, fig.height=4}
-picollapse
-```
-
-```{r, fig.width = 6}
-pi
-```
-
-## Resilience 
-```{r, echo=FALSE}
-# Variable
-resp_var <- 'rs'
-
-# AOV
-aov_rs <- aovas(re, vars=vars, resp_var = resp_var)
-
-mc <- aov_rs$model_coeff
-
-pander(mc, round=5,
-       caption = paste0("ANOVA table: ", resp_var), missing = '', 
-       emphasize.strong.cells = 
-         which(mc < 0.1 & mc == mc$p.value, arr.ind = T))
-
-gm <- aov_rs$model_summary
-
-gm <- apply(gm, 1, formatC, digits = 2, format = "f") %>% t()
-colnames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e"),"$")
-rownames(gm) <- "Statistic"
-pander(t(gm)) 
-```
-
-```{r}
-# Post hoc Define model
-mymodel <- aov_rs$mymodel
-postH_rs <- phc(mymodel = mymodel, resp_var = resp_var)
-```
-
-```{r, echo=FALSE}
-#### ~ Site
-ps <- plot(effect("site",mymodel))
-#### ~ Disturb Year
-pd <- plot(effect('disturb_year', mymodel))
-#### Disturb Year:Site
-picollapse <- plot(effect("disturb_year:site",mymodel), multiline = TRUE, ci.style = 'bars')
-pi <- plot(effect("disturb_year:site",mymodel), layout=c(2,1))
-
-```
-
-```{r, fig.width = 4, fig.height=4}
-ps
-```
-
-```{r, fig.width = 4, fig.height=4}
-pd
-```
-
-```{r, fig.width = 4, fig.height=4}
-picollapse
-```
-
-```{r, fig.width = 6}
-pi
-```
-
-```{r, echo=FALSE}
-# Interactions plot all  
-means_site <- postH_rc[[4]] %>% mutate(var = 'rc') %>% 
-  bind_rows(postH_rt[[4]] %>% mutate(var = 'rt')) %>% 
-  bind_rows(postH_rs[[4]] %>% mutate(var = 'rs')) %>% 
-  bind_rows(postH_rrs[[4]] %>% mutate(var = 'rrs')) %>% 
-  rename(letras = .group)
-
-means_disturb <- postH_rc[[5]] %>% mutate(var = 'rc') %>% 
-  bind_rows(postH_rt[[5]] %>% mutate(var = 'rt')) %>% 
-  bind_rows(postH_rs[[5]] %>% mutate(var = 'rs')) %>% 
-  bind_rows(postH_rrs[[5]] %>% mutate(var = 'rrs')) %>% 
-  rename(letras = .group)
-
-means_distub_site <- postH_rc[[6]] %>% mutate(var = 'rc') %>% 
-  bind_rows(postH_rt[[6]] %>% mutate(var = 'rt')) %>% 
-  bind_rows(postH_rs[[6]] %>% mutate(var = 'rs')) %>% 
-  bind_rows(postH_rrs[[6]] %>% mutate(var = 'rrs')) %>% 
-  rename(letras = .group)
-
-```
-
-```{r, echo=FALSE}
-dodge <- position_dodge(width = 0.3)
-micolor <- '#455883'
-mierrorbarSE <- aes(ymin=lsmean - SE, ymax=lsmean + SE)
-mierrorbar <- aes(ymin=lower.CL, ymax=upper.CL)
-```
-
-```{r, echo=FALSE}
-plot_ms <- means_site %>%  
-  ggplot(aes(x=site, y=lsmean)) + 
-  geom_point(colour=micolor, 
-             size=3, position = dodge) +
-  theme_bw() + xlab('') + ylab('') + 
-  facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15) +
-  theme(strip.background = element_rect(colour = "black", fill = "white"))
-
-plot_msSE <- plot_ms + geom_errorbar(mierrorbarSE,color=micolor, 
-                                     size=.5, width=.15, position = dodge)
-
-plot_msCI <- plot_ms + geom_errorbar(mierrorbar,color=micolor, 
-                                     size=.5, width=.15, position = dodge)
-```
-
-```{r, echo=FALSE}
-plot_md <- means_disturb %>%  
-  ggplot(aes(x=disturb_year, y=lsmean)) + 
-  geom_point(colour=micolor, 
-             size=3, position = dodge) +
-  theme_bw() + xlab('') + ylab('') + 
-  facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean, label=letras), nudge_x = 0.15) +
-  theme(strip.background = element_rect(colour = "black", fill = "white"))
-
-plot_mdSE <- plot_md + geom_errorbar(mierrorbarSE,color=micolor, 
-                                     size=.5, width=.15, position = dodge)
-
-plot_mdCI <- plot_md + geom_errorbar(mierrorbar,color=micolor, 
-                                     size=.5, width=.15, position = dodge)
-```
-
-```{r}
-plot_mds <- means_distub_site %>%  
-  ggplot(aes(x=site, y=lsmean, group=disturb_year, colour=disturb_year)) + 
-  geom_point(aes(shape=disturb_year), size=3) + 
-  geom_line() +
-  theme_bw() + xlab('') + ylab('') + 
-  facet_wrap(~var, scales='free_y', ncol = 1) +
-  geom_text(aes(y=lsmean+SE, label=letras), nudge_x = 0.15)+
-  theme(strip.background = element_rect(colour = "black", fill = "white"),
-        legend.position = c(0.8, 0.93),
-        legend.background = element_blank()) +
-  scale_colour_manual(values = c(micolor, "red"))
-
-plot_mdsSE <- plot_mds + geom_errorbar(mierrorbarSE, size=.5, width=.15)
-plot_mdsCI <- plot_mds + geom_errorbar(mierrorbar, size=.5, width=.15)
-```
-
-```{r}
-pdf(paste0(di, 'out/fig/resilience/interaction_plotsSE.pdf'), width=9, height = 9)
-grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
-dev.off()
-
-pdf(paste0(di, 'out/fig/resilience/interaction_plotsCI.pdf'), width=9, height = 9)
-grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
-dev.off()
-
-```
-
-```{r}
-aovas_coeff <- aov_rc$model_coeff %>% mutate(var = 'rc') %>% 
-  bind_rows(aov_rt$model_coeff %>% mutate(var = 'rt')) %>% 
-  bind_rows(aov_rs$model_coeff %>% mutate(var = 'rs')) %>% 
-  bind_rows(aov_rrs$model_coeff%>% mutate(var = 'rrs'))
-
-write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/anovas_statistics.csv'), row.names = F)
-
-aovas_coeff %>% pander()
-```
-
-```{r}
-aovas_model_summary <- aov_rc$model_summary %>% mutate(var = 'rc') %>% 
-  bind_rows(aov_rt$model_summary %>% mutate(var = 'rt')) %>% 
-  bind_rows(aov_rs$model_summary %>% mutate(var = 'rs')) %>% 
-  bind_rows(aov_rrs$model_summary%>% mutate(var = 'rrs'))
-
-write.csv(aovas_model_summary, 
-          file=paste0(di, '/out/anovas_resilience/anovas_summary_modelos.csv'), row.names = F)
-
-gm <- apply(aovas_model_summary, 1, formatC, digits = 2, format = "f") 
-rownames(gm) <- paste0("$",c("R^2","\\mathrm{adj}R^2","\\sigma_e","F","p","df_m","\\mathrm{logLik}","AIC","BIC","\\mathrm{dev}","df_e", "variable"),"$")
-colnames(gm) <- c("rc", "rt", "rs", "rrs")
-
-pander(gm)
-```
-
-# ROBUST ANOVA
-
-* Ver Wilcox (2005, 2012)
-* Vamos a realizar un Robust factorial ANOVA. En concreto:
-
-  * Two-way robust factorial ANOVA on M-estimator
-  * pkg WRS2
-
-```{r}
-robustANOVA <- function(df, resp_var, factores, 
-         alpha, nboot, treshold) {
-  # alpha: alpha ci for huber m-estimation 
-  # nboot: numer of iterations 
-  # treshoold for letter (posthoc)
-  # See http://rcompanion.org/rcompanion/d_08a.html
-
-  # Create interaction 
-  df$interaction <- interaction(df$disturb_year, df$site)
-
-  # Formulas 
-  formulaFull <- as.formula(paste0(resp_var,  " ~ ",
-                                 paste(factores, collapse = '+')))
-
-  formula_A <- as.formula(paste0(resp_var,  " ~ ", factores[1]))
-  formula_B <- as.formula(paste0(resp_var,  " ~ ", factores[2]))
-  formula_AB <- as.formula(paste0(resp_var,  " ~ interaction"))
-
-  # Produce Huber M-estimators and confidence intervals by group
-  mest <- groupwiseHuber(formulaFull, data = df, ci.type = 'wald', conf.level = alpha)
-  mest_a <- groupwiseHuber(formula_A, data = df, ci.type = 'wald', conf.level = alpha)
-  mest_b <- groupwiseHuber(formula_B, data = df, ci.type = 'wald', conf.level = alpha)
-
-  # Two-way robust analysis 
-  x <- pbad2way(formulaFull, data = df, est = "mom", nboot = nboot)
-
-  pbad2way(rs ~ disturb_year + site + disturb_year:site, data = df, 
-           est = "mom", nboot = nboot)
-
-  out_ra <- data.frame(
-    term = c(x$varnames[2], 
-             x$varnames[3], 
-             paste0(x$varnames[2], ':', x$varnames[3])),
-    p_value = c(x$A.p.value, x$B.p.value, x$AB.p.value))
-
-  # post-hoc 
-  ## factor A
-  pha <- pairwiseRobustTest(formula_A, data = df, est = "mom", 
-                            nboot = nboot, method="bonferroni")
-
-  ## factor B
-  phb <- pairwiseRobustTest(formula_B, data = df, est = "mom", 
-                            nboot = nboot, method="bonferroni")
-  ## interaction effect (AB)
-  phab <- pairwiseRobustTest(formula_AB, data = df, est = "mom", 
-                            nboot = nboot, method="bonferroni")
-
-  ph <- rbind(pha, phb, phab)
-
-  phRWS2 <- mcp2a(formulaFull, data=df, est = "mom", nboot = nboot)
-
-  out <- list()  
-  out$mest <- mest # Huber M-estimators and Confidence Intervals
-  out$mest_a <- mest_a
-  out$mest_b <- mest_b
-  out$ra <- out_ra # Output for Two-way robust analysis (M-estimators)
-  out$ph <- ph # posthoc comparison usinng pairwiseRobustTest 
-  out$pha <- pha
-  out$phb <- phb
-  out$phab <- phab
-
-  print(out_ra)
-  print(phRWS2)
-
-  return(out)
-}
-
-  # if (exists('letters_phb')) {
-  # letters_phb <- letters_phb} else {
-  # myerror <- evaluate('cldList(comparison = phb$Comparison, 
-  #                     p.value = phb$p.adjust, threshold  = treshold)')
-  # letters_phb <- as.character(myerror[[2]]$message)}
-  # 
-  # if (exists('letters_pha')) {
-  # letters_pha <- letters_pha} else {
-  # myerror <- evaluate('cldList(comparison = pha$Comparison, 
-  #                     p.value = pha$p.adjust, threshold  = treshold)')
-  # letters_pha <- as.character(myerror[[2]]$message)}
-  # 
-  # if (exists('letters_phab')) {
-  # letters_phab <- letters_phab} else {
-  # myerror <- evaluate('cldList(comparison = phab$Comparison, 
-  #                     p.value = phab$p.adjust, threshold  = treshold)')
-  # letters_phab <- as.character(myerror[[2]]$message)}
-
-```
-
-```{r}
-factores = c('disturb_year', 'site', 'disturb_year:site')
-```
-
-## Resilience 
-```{r}
-rars <- robustANOVA(df=re, resp_var='rs', factores=factores,
-              alpha = 0.95, nboot = 3000, treshold = 0.01)
-```
-
-#### Rs Letters 
-```{r}
-x <-rars
-
-letraArs <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rs')
-letraArs
-
-letraBrs <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rs')
-letraBrs # Manual (IMPROVE IT)
-letraBrs <- data.frame(Group = c('SJ', 'caH', 'caL'),
-                     Letter = c('a','a','a'),
-                     MonoLetter = c('a','a','a')) %>% mutate(var ='rs')
-letraBrs
-
-letraABrs  <- cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01) %>% mutate(var ='rs')
-letraABrs
-
-```
-
-## Recovery 
-```{r}
-rarc <- robustANOVA(df=re, resp_var='rc', factores=factores,
-              alpha = 0.95, nboot = 3000, treshold = 0.01)
-```
-
-#### Rc Letters 
-```{r}
-x <-rarc
-
-letraArc <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rc')
-letraArc
-letraBrc <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rc')
-letraBrc
-letraABrc <- cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01) %>% mutate(var ='rc')
-letraABrc
-
-```
-
-## Resistance 
-```{r}
-rart <- robustANOVA(df=re, resp_var='rt', factores=factores,
-              alpha = 0.95, nboot = 3000, treshold = 0.01)
-```
-
-#### Rt Letters 
-```{r}
-x <-rart
-
-letraArt <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rt')
-letraArt #Manual (IMPROVE IT)
-letraArt <- data.frame(Group = as.factor(c('25', '212')),
-                     Letter = as.factor(c('a','a')),
-                     MonoLetter = as.factor(c('a','a'))) %>% mutate(var ='rt')
-letraArt 
-letraBrt <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rt')
-letraBrt
-letraABrt <- cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01) %>% mutate(var ='rt')
-letraABrt 
-```
-
-## Relative Resilience
-```{r}
-rarrs <- robustANOVA(df=re, resp_var='rrs', factores=factores,
-              alpha = 0.95, nboot = 3000, treshold = 0.01)
-```
-
-#### Rt Letters 
-```{r}
-x <-rarrs
-
-letraArrs <- cldList(comparison = x$pha$Comparison, p.value = x$pha$p.adjust, threshold  = 0.01) %>% mutate(var ='rrs')
-letraArrs
-letraBrrs <- cldList(comparison = x$phb$Comparison, p.value = x$phb$p.adjust, threshold  = 0.01) %>% mutate(var ='rrs')
-letraBrrs
-letraABrrs <- cldList(comparison = x$phab$Comparison, p.value = x$phab$p.adjust, threshold  = 0.01) %>% mutate(var ='rrs')
-letraABrrs 
-```
-
-```{r}
-letrasA <- rbind(letraArs, letraArc, letraArt, letraArrs) %>% 
-  mutate(disturb_year = 
-           case_when(Group == "25" ~ "2005",
-                     Group == "212" ~ "2012")) %>% 
-  dplyr::select(-Group)
-
-letrasB <- rbind(letraBrs, letraBrc, letraBrt, letraBrrs) %>% rename(site = Group)
-
-letrasAB <- rbind(letraABrs, letraABrc, letraABrt, letraABrrs) %>% 
-  separate(Group, into=c('disturb_year', 'site')) %>% 
-  mutate(disturb_year = 
-           case_when(disturb_year == "25" ~ "2005",
-                     disturb_year == "212" ~ "2012"))
-```
-
-### Estimadores de huber 
-```{r} 
-rars$mest$var <- 'rs'
-rarc$mest$var <- 'rc'
-rart$mest$var <- 'rt'
-rarrs$mest$var <- 'rrs'
-
-mhuber <- rbind(rarc$mest, rart$mest, rars$mest, rarrs$mest)
-
-# add letras 
-mhuber <- mhuber %>% inner_join(letrasAB, by=c('var','disturb_year', 'site'))
-
-mhuber_agg <- mhuber %>% 
-  mutate(M.Huber = round(M.Huber, 4),
-         lower.ci = round(lower.ci, 4),
-         upper.ci = round(upper.ci, 4)) %>% 
-  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
-  mutate(ci = paste0('(', ci, ')')) %>% 
-  dplyr::select(-MonoLetter) %>% 
-  dplyr::select(var, disturb_year:ci, Letter) %>% as.data.frame()
-
-mhuber_agg %>% pander() 
-```
-
-```{r}
-rars$mest_a$var <- 'rs'
-rarc$mest_a$var <- 'rc'
-rart$mest_a$var <- 'rt'
-rarrs$mest_a$var <- 'rrs'
-
-mhuber_a <- rbind(rarc$mest_a, rart$mest_a, rars$mest_a, rarrs$mest_a)
-mhuber_a <- mhuber_a %>% inner_join(letrasA, by=c('var','disturb_year'))
-
-mhuber_agg_a <- mhuber_a %>% 
-  mutate(M.Huber = round(M.Huber, 4),
-         lower.ci = round(lower.ci, 4),
-         upper.ci = round(upper.ci, 4)) %>% 
-  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
-  mutate(ci = paste0('(', ci, ')')) %>% 
-  dplyr::select(-MonoLetter) %>% 
-  dplyr::select(var, disturb_year:ci, Letter) %>% as.data.frame()
-
-mhuber_agg_a %>% pander()
-
-```
-
-```{r}
-rars$mest_b$var <- 'rs'
-rarc$mest_b$var <- 'rc'
-rart$mest_b$var <- 'rt'
-rarrs$mest_b$var <- 'rrs'
-
-mhuber_b <- rbind(rarc$mest_b, rart$mest_b, rars$mest_b, rarrs$mest_b)
-mhuber_b <- mhuber_b %>% inner_join(letrasB, by=c('var','site'))
-
-mhuber_agg_b <- mhuber_b %>% 
-  mutate(M.Huber = round(M.Huber, 4),
-         lower.ci = round(lower.ci, 4),
-         upper.ci = round(upper.ci, 4)) %>% 
-  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
-  mutate(ci = paste0('(', ci, ')')) %>% 
-  dplyr::select(-MonoLetter) %>% 
-  dplyr::select(var, site:ci, Letter) %>% as.data.frame()
-
-mhuber_agg_b %>% pander() 
-```
-
-### Pairwise comparison
-```{r}
-
-rars$ph$var <- 'rs'
-rarc$ph$var <- 'rc'
-rart$ph$var <- 'rt'
-rarrs$ph$var <- 'rrs'
-
-pairwise <- rbind(rarc$ph, rart$ph, rars$ph, rarrs$ph) 
-pairwise %>% pander()
-```
-
-## Interaction plot 
-### Response ~ (x=Drought)
-
-```{r}
-mhuber<- mhuber %>% 
-  mutate(var_sorted = case_when(var == "rc" ~ "1_rc",
-                                var == "rt" ~ "0_rt",
-                                var == "rs" ~ "2_rs",
-                                var == "rrs" ~ "3_rrs"))
-
-pd <- position_dodge(.2)
-
-robust_plot_evi_drought <- ggplot(mhuber, aes(x=disturb_year, y=M.Huber, color = site, group=site, fill=site)) + 
-  geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), 
-                width=.1, size=0.7, position=pd) + 
-  geom_line(aes(group=site,color=site, linetype=site), position=pd) + 
-  geom_point(shape=21, size=3.5, position=pd) +
-  facet_wrap(~var_sorted, nrow = 2, scales = 'free_y',
-             labeller=as_labeller(c('0_rt' = 'Resistance', 
-                                 '1_rc' = 'Recovery',
-                                 '2_rs' = 'Resilience',
-                                 '3_rrs' = 'Relative Resilience'))) +
-  scale_color_manual(values=c('black','blue','red')) + 
-  scale_fill_manual(values=c('black','blue','red')) + theme_bw() +
-  scale_linetype_manual(values=c("solid", "dashed", 'dashed')) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'),
-        legend.position="bottom") +
-  ylab('') + xlab('Drought event')
-
-robust_plot_evi_drought
-
-pdf(paste0(di, 'out/fig/resilience/robust_plot_evi_drought.pdf'), width=9, height = 9)
-robust_plot_evi_drought 
-dev.off()
-```
-
-### Response ~ (x=site)
-```{r}
-pd <- position_dodge(.2)
-
-robust_plot_evi_site <- ggplot(mhuber, aes(x=site, y=M.Huber, color = disturb_year, group=disturb_year, fill=disturb_year)) + 
-  geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), 
-                width=.1, size=0.7, position=pd) + 
-  geom_line(aes(group=disturb_year,color=disturb_year, linetype=disturb_year), position=pd) + 
-  geom_point(shape=21, size=3.5, position=pd) +
-  facet_wrap(~var_sorted, nrow = 2, scales = 'free_y',
-             labeller=as_labeller(c('0_rt' = 'Resistance', 
-                                 '1_rc' = 'Recovery',
-                                 '2_rs' = 'Resilience',
-                                 '3_rrs' = 'Relative Resilience'))) +
-  scale_color_manual(values=c('black','blue')) + 
-  scale_fill_manual(values=c('black','blue')) + theme_bw() +
-  scale_linetype_manual(values=c("solid", "dashed")) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white'),
-        legend.position="bottom") +
-  ylab('') + xlab('Drought event')
-
-robust_plot_evi_site
-
-pdf(paste0(di, 'out/fig/resilience/robust_plot_evi_site.pdf'), width=9, height = 9)
-robust_plot_evi_site 
-dev.off()
-```
-
-```{r}
-# Export data 
-write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/robust_mhuber.csv'), row.names = F)
-write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg.csv'), row.names = F)
-write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_a.csv'), row.names = F)
-write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg_a.csv'), row.names = F)
-write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_b.csv'), row.names = F)
-write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/robust_mhuber_agg_b.csv'), row.names = F)
-
-write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/robust_pairwise.csv'), row.names = F)
-```
-
-# References

### 2018-02-12 analysis/drought/explore_spei_longterm.Rmd
include spei severe droughts and resilience trends

@@
 sev12 <- sev %>% filter(variable == "spei12") %>% arrange(desc(d_duration)) %>% 
-  dplyr::select(-index_events, -cat, -decade, -variable)
+  dplyr::select(-index_events, -decade, -variable)
+
+write.csv(sev12, file=here::here('data/sequias', 'severe_spei12.csv'), row.names=FALSE)
 
 pander(sev12)
-
 ```

### 2018-02-13 analysis/drought/explore_prec_longterm.Rmd
compute 2sd precipitation

---
title: "LongTerm Hydrological Prec"
author: "AJ Perez-Luque (@ajpelu)"
output: 
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, warning=FALSE, error=TRUE, message=FALSE)
```

```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
library("here")
library("tidyverse")
library("lubridate")
library("ggplot2")
library("ggrepel")
```

```{r}
source(here::here('script/R', 'hydro_year.R'))
```

```{r}
ltdf <- read.csv(file=here::here('data_raw/meteo/', 'long_term_data_rediam_data.txt'), header=T, sep=';', dec=',')

ltdf <- ltdf %>% 
  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
         month = lubridate::month(fecha),
         hmonth = ifelse(month <= 8, month + 4, month -8), 
         hyear = hydro_year(fecha),
         hyear_f = paste0(hyear-1,'-', hyear))

st5514 <- ltdf %>% 
  filter(VARIABLE == 'PM1') %>% 
  filter(cod == '5514') %>% 
  filter(hyear >= 1950) %>% 
  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)

df <- st5514 %>% 
  group_by(hyear_f) %>% 
  dplyr::mutate(csum = cumsum(pre)) %>% 
  filter(hmonth == 12) %>% 
  dplyr::select(-month, -pre, -fecha) %>% as.data.frame()

meandf <- mean(df$csum)
sddf <- sd(df$csum)

prec_hydrol <- ggplot(df, aes(x=hyear, y=csum)) + 
  geom_point() + theme_bw() + 
  geom_hline(yintercept = meandf) + 
  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
    geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
  geom_text_repel(
    data = subset(df, csum < (meandf - 1*sddf)),
    aes(label = hyear)) +
  ylab('Cumulative prec (hydrological year)')

pdf(here::here('/out/precipitation_2sd', 'prec_hydrol.pdf'), width=12, height = 9)
prec_hydrol
dev.off()

df2sd <- df %>% filter(csum < (meandf - 2*sddf)) %>% as.data.frame()
pander(df2sd)
write.csv(df2sd, here::here("/data/sequias", "prec19502012_2sd.csv"), row.names = FALSE)

df1sd <- df %>% filter(csum < (meandf - 1*sddf)) %>% as.data.frame()
pander(df1sd)
write.csv(df1sd, here::here("/data/sequias", "prec19502012_1sd.csv"), row.names = FALSE)

```

### 2018-02-13 analysis/bai/trends_resilience_bai.Rmd
resiliences bais longterm

@@
 library('devtools')
+library("ggrepel")
 ```
@@
 
-machine <- 'ajpeluLap'
+# machine <- 'ajpeluLap'
 
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
+# di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
 
 # sj 
-sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
+sj <- read.rwl(fname=here::here('data_raw/dendro_ring/sn_sanjuan', 'sn_sanjuan.rwl'), format="tucson")
 
 # canar 
-ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
+ca <- read.rwl(fname=here::here('data_raw/dendro_ring/sn_canar', 'sn_canar.rwl'), format="tucson")
 
 # Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+compete <- read.csv(file=here::here('data_raw', 'dendro_competence.csv'), header=TRUE, sep=',')
 ```
@@
 ```{r}
-source(paste0(di, 'script/R/rw_byTree.R'))
-source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/baiResilience.R'))
+source(here::here('script/R','rw_byTree.R'))
+source(here::here('script/R', 'bai_piovesan.R'))
+source(here::here('script/R', 'baiResilience.R'))
 ```
@@
 # Drought years 
-
+# Utilizamos spei12 
 dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-
 dy <- dy %>% arrange(desc(maxyear))
 
-dys <- dy %>% filter(maxyear > 1940) %>% as.data.frame()
-
+# solamente aquellos mayor de 2 meses 
+dys <- dy %>% filter(d_duration > 2) %>% as.data.frame() 
 dys <- dys[,'maxyear']
 
-# ojo quito 1985 
-dys <- dys[which(dys!=1985)] 
 ```
@@
 ```{r}
+# Computar resiliencia para otros years 
 # SJ 
 res_3_sj <- baiResilience(bai_sj, event_years = dys, window = 3)
+# Remove 1914 
+dyssj <- dys[which(dys!=1914)]
+res10sj <- baiResilience(bai_sj, event_years = dyssj, window = 10)
+
 # caL
 res_3_caL <- baiResilience(bai_caL, event_years = dys, window = 3)
+res10caL <- baiResilience(bai_caL, event_years = dys, window = 10)
 # caH
 res_3_caH <- baiResilience(bai_caH, event_years = dys, window = 3)
+res10caH <- baiResilience(bai_caH, event_years = dys, window = 10)
 ```
@@
 
+rsj10 <- res10sj$resilience %>% mutate(site='SJ')
+rcaL10<- res10caL$resilience %>% mutate(site='caL')
+rcaH10 <- res10caH$resilience %>% mutate(site='caH')
+
 re <- bind_rows(rsj, rcaL, rcaH)
+re10 <- bind_rows(rsj10, rcaL10, rcaH10)
 # re$disturb_year <- as.factor(re$disturb_year)
 re$site <- as.factor(re$site)
+re10$site <- as.factor(re10$site)
 
-re %>% ggplot(aes(x=disturb_year, ))
+## Sample depth (#trees by site)
+sampleDepth <- function(x, site){ 
+  x$samps <- rowSums(!is.na(x))
+  x$year <- as.numeric(row.names(x))
+  x$site <- site
+  out <- x[,c("samps", "year", "site")]
+  return(out)
+}
 
+d_sj <- sampleDepth(sj_tree, site='SJ')
+d_caL <- sampleDepth(caL_tree, site='caL')
+d_caH <- sampleDepth(caH_tree, site='caH')
+
+d <- rbind(d_sj, d_caL, d_caH)
+
+# Get mean values by site and add sample deepth
 s <- re %>% 
   group_by(site, disturb_year) %>% 
-  summarise_at(c("rt", "rs", "rc", "rrs"), funs(mean,sd,se=sd(.)/sqrt(n())))
+  summarise_at(c("rt", "rs", "rc", "rrs"), funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(d, by=c("disturb_year" = "year", "site"="site")) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
 
-s %>% 
+s10 <- re10 %>% 
+  group_by(site, disturb_year) %>% 
+  summarise_at(c("rt", "rs", "rc", "rrs"), funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(d, by=c("disturb_year" = "year", "site"="site")) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+```
+
+```{r}
+# ojo he ploteado los que tiene mas de 10 árboles 
+resi_temporal <- s %>% filter(samps >= 10) %>% 
   ggplot(aes(y=rs_mean, x=disturb_year, colour=site)) + 
@@
                 position=position_dodge(2)) +
-  geom_point(size=2, shape=21, fill='white', position=position_dodge(2)) + 
-  #facet_wrap(~site, scales='fixed', ncol=1) + 
-  theme_bw() # + geom_smooth(aes(colour=site), se=FALSE)
+  geom_point(aes(size=d_duration), shape=21, fill='white', position=position_dodge(2)) + 
+  facet_wrap(~site, scales='free_y', ncol=1) + 
+  theme_bw() + 
+  theme(strip.background = element_rect(fill='transparent')) + 
+  ylab('Resilience (Rs)') + xlab('years') + 
+  scale_x_continuous(breaks=seq(1910,2020, by=5)) # + geom_smooth(aes(colour=site), se=FALSE)
+
+resi_temporal
+
+pdf(here::here('/out/bai_resiliences_severes', 'res_bai_longterm.pdf'), width=12, height = 9)
+resi_temporal
+dev.off()
+
+resi_temporal10 <- s10 %>% filter(samps >= 10) %>% 
+  ggplot(aes(y=rs_mean, x=disturb_year, colour=site)) + 
+  geom_errorbar(aes(ymin=rs_mean - rs_sd, 
+                    ymax=rs_mean + rs_sd),
+                width = 0.2, 
+                position=position_dodge(2)) +
+  geom_point(aes(size=d_duration), shape=21, fill='white', position=position_dodge(2)) + 
+  facet_wrap(~site, scales='free_y', ncol=1) + 
+  theme_bw() + 
+  theme(strip.background = element_rect(fill='transparent')) + 
+  ylab('Resilience (Rs)') + xlab('years') + 
+  scale_x_continuous(breaks=seq(1910,2020, by=5)) # + geom_smooth(aes(colour=site), se=FALSE)
+
+pdf(here::here('/out/bai_resiliences_severes', 'res_bai_longterm10.pdf'), width=12, height = 9)
+resi_temporal10
+dev.off()
 
 ```
+
+```{r}
+
+#https://gist.github.com/ottadini/6882677 
+lm_eqn = function(m) {
+  # Displays regression line equation and R^2 value on plot
+  # Usage:
+  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
+
+  l <- list(#a = format(coef(m)[1], digits = 2),
+            # b = format(abs(coef(m)[2]), digits = 2),
+            r2 = format(summary(m)$r.squared, digits = 3));
+
+  if (coef(m)[2] >= 0)  {
+    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
+    eq <- substitute(~~italic(r)^2~"="~r2,l)
+  } else {
+    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
+    eq <- substitute(~~italic(r)^2~"="~r2,l)
+  }
+
+  as.character(as.expression(eq));                 
+}
+
+sdf <- s %>% filter(samps >= 10) %>% group_by(disturb_year) %>% 
+  summarise(rs_avg = mean(rs_mean, na.rm=TRUE), 
+            rt_avg = mean(rt_mean, na.rm=TRUE), 
+            rc_avg = mean(rc_mean, na.rm=TRUE)) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+tR <- list(geom_smooth(method='lm', se=FALSE),
+           theme_bw(), xlab("Drought Severity"), 
+          # scale_size_continuous("Drought Intensity", breaks = c(-2,-1.8,-1.6, -1.4, -1.2), labels= c(-2,-1.8,-1.6, -1.4, -1.2)),
+          geom_text_repel(),
+          scale_size_continuous("Drought Intensity (*-1)"),
+          geom_point(),
+          theme(legend.position ='top'))
+
+sevRs <- sdf %>% 
+  ggplot(aes(y=rs_avg,  x=d_severity, label=disturb_year)) +  
+  geom_point(aes(size=-1*d_intensity)) + 
+  ylab("Resilience (Rs) (3year window size)") + tR +
+  annotate("text", x=7, y = 1.4, label=lm_eqn(lm(rs_avg~d_severity, sdf)), parse=TRUE) 
+sevRs
+
+sevRt <- sdf %>% 
+  ggplot(aes(y=rt_avg,  x=d_severity, label=disturb_year)) + 
+  geom_point(aes(size=-1*d_intensity)) + 
+  ylab("Resistance (Rt) (3year window size)") + tR +
+  annotate("text", x=7, y = 1.3, label=lm_eqn(lm(rt_avg~d_severity, sdf)), parse=TRUE) 
+sevRt
+
+sevRc <- sdf %>% 
+  ggplot(aes(y=rc_avg,  x=d_severity, label=disturb_year)) + 
+  geom_point(aes(size=-1*d_intensity)) + 
+  ylab("Recovery (Rc) (3year window size)") + tR +
+  annotate("text", x=7, y = 2, label=lm_eqn(lm(rc_avg~d_severity, sdf)), parse=TRUE) 
+sevRc
+
+pdf(here::here('/out/bai_resiliences_severes', 'rs_vs_drought_3years.pdf'), width=18, height = 7)
+sevRt + sevRc + sevRs 
+dev.off()
+
+sdf10 <- s10 %>% filter(samps >= 10) %>% group_by(disturb_year) %>% 
+  summarise(rs_avg = mean(rs_mean, na.rm=TRUE), 
+            rt_avg = mean(rt_mean, na.rm=TRUE), 
+            rc_avg = mean(rc_mean, na.rm=TRUE)) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+sevRs10 <- sdf10 %>% 
+  ggplot(aes(y=rs_avg,  x=d_severity, label=disturb_year)) +  
+  geom_point(aes(size=-1*d_intensity)) + 
+  ylab("Resilience (Rs) (10year window size)") + tR +
+  annotate("text", x=7, y = 1.4, label=lm_eqn(lm(rs_avg~d_severity, sdf10)), parse=TRUE) 
+sevRs10
+
+sevRt10 <- sdf10 %>% 
+  ggplot(aes(y=rt_avg,  x=d_severity, label=disturb_year)) + 
+  geom_point(aes(size=-1*d_intensity)) + 
+  ylab("Resistance (Rt) (10year window size)") + tR +
+  annotate("text", x=7, y = 1.3, label=lm_eqn(lm(rt_avg~d_severity, sdf10)), parse=TRUE) 
+sevRt10
+
+sevRc10 <- sdf10 %>% 
+  ggplot(aes(y=rc_avg,  x=d_severity, label=disturb_year)) + 
+  geom_point(aes(size=-1*d_intensity)) + 
+  ylab("Recovery (Rc) (10year window size)") + tR +
+  annotate("text", x=7, y = 2, label=lm_eqn(lm(rc_avg~d_severity, sdf10)), parse=TRUE) 
+sevRc10
+
+pdf(here::here('/out/bai_resiliences_severes', 'rs_vs_drought_10years.pdf'), width=18, height = 7)
+sevRt10 + sevRc10 + sevRs10 
+dev.off()
+```

### 2018-02-14 analysis/bai/trends_resilience_bai.Rmd
organizar estos

@@
 
+!!!!!!!! ----- ESTO QUE esl lo que es!!!! compaeeee 
+# With error bar 
+sdfbe <- s %>% filter(samps >= 10) %>% group_by(disturb_year) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
 tR <- list(geom_smooth(method='lm', se=FALSE),

### 2018-02-15 analysis/drought/explore_prec_longterm.Rmd
update

@@
 ```
+
+## PREC + 2sd for more series
+
+```{r}
+
+prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
+
+dfpre <- c()
+
+for (i in prec_files){
+
+  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
+  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
+
+  dfpre <- rbind(dfpre, auxdf)
+}
+
+dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
+
+dfpre <- dfpre %>% 
+  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
+         month = lubridate::month(fecha),
+         hmonth = ifelse(month <= 8, month + 4, month -8), 
+         hyear = hydro_year(fecha),
+         hyear_f = paste0(hyear-1,'-', hyear))
+
+# sql to get lat/lon en Linaria
+# SELECT 
+#   cli_estaciones.id, 
+#   cli_estaciones.nombre, 
+#   cli_estaciones.codigo, 
+#   cli_estaciones.punto_x, 
+#   cli_estaciones.punto_y, 
+#   cli_estaciones.geometria, 
+#   ST_X(ST_Transform (geometria, 4326)) AS "Longitude",
+#   ST_Y(ST_Transform (geometria, 4326)) AS "Latitude"
+# FROM 
+#   public.cli_estaciones 
+# WHERE cli_estaciones.codigo IN ('5501','5502','5504','5504A','5505','5509','5509A','5511E',
+# '5517','5518','6224','6234','6234E','6236','6242U','6243','6243E','6248','6251','6252','6253',
+# '6257','6257I','6280U','6281E','6282','6307','6307X','6308');
+
+stations  <- read.csv(here::here('data_raw/meteo/prec_subclima/mdestaciones', 'estaciones_latlong.csv'), header=TRUE, sep=";")
+
+stations <- stations %>% dplyr::select(nombre, codigo, Longitude, Latitude)
+
+aux <- dfpre %>% 
+  group_by(hyear_f, code) %>% 
+  dplyr::mutate(csum = cumsum(prec)) %>% 
+  filter(hmonth == 12) %>% 
+  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
+
+presn <- aux %>% group_by(hyear) %>% 
+     dplyr::summarise(
+     csum_mean = mean(csum, na.rm = TRUE),
+     csum_sd = sd(csum, na.rm = TRUE), 
+     csum_se = csum_sd/sqrt(length(csum_sd)), 
+     n = n())
+
+meandf <- mean(presn$csum_mean)
+sddf <- sd(presn$csum_mean)
+
+prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
+  geom_hline(yintercept = meandf) + 
+  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
+  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
+  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
+  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
+  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
+  geom_point(size=2) + theme_bw() +
+  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
+  scale_y_continuous(breaks=seq(0,1500, by=250)) +
+  geom_text_repel(
+    data = subset(presn, csum_mean < (meandf - 1*sddf)),
+    aes(label = hyear)) +
+  theme(panel.grid.minor = element_blank()) +
+  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
+prec_hydrol_sn
+
+pdf(here::here('/out/precipitation_2sd', 'prec_hydrol_sn.pdf'), width=12, height = 7)
+prec_hydrol_sn
+dev.off()
+
+```

### 2018-02-15 man/ms_3_discussion.Rmd
update

@@
 # Algunas ideas para la discussion
+
+## FORTALEZA METODOLÓGICA
+
+* Algunos trabajos han analizado la relación ente RWI y remote sensing [@VicenteSerrano2013]; [(Bhuyan et al.
+Remote Sensing 9(6):526)]; [Vicente Serrano et al.
+https://www.sciencedirect.com/science/article/pii/S003442571630373X] ...
+
+* Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q.
+pyrenaica situadas en su reaar edge frente a los eventos de sequía.
+
+--- notas para esto 
+Se trata de expresar la idea de mezcla de aproximacioes: remote-sensing con dendro.
+Aunque existen algunos trabajos que utilizado RWI y remote sensing nuestra aproximación es importante, sobre todo para estudiar poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
+De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
+
+Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
+(ver esto y completar)
+
+Aunque el NDVI o similar ha sido utilizado para monitorizar la respuesta de bosques frente a sequía, su escala espacial limita su uso para estudiar cambios en poblacioens situadas en su rear edge.
+
+ Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability.
+In such regions, targeted collection of high-resolution data is necessary in order to identify currently occurring changes and predict the magnitude and spatial distribution of future decline.
+[@Jump2010]
+
+comentario de GGI
+
+Gracias Antonio.
+Desde luego que debes resaltar a lo largo del artículo la combinación de sensores y dendro como fortaleza de tu trabajo al analizar el tema de resiliencia desde dos escalas espacio-temporales complementarias.
+Y eso independientemente que lo diga o no Alistair Jump (que las citas vienen bien, y seguro que el trabajo es bueno), combinar ambas escalas/datos es beneficioso no sólo en poblaciones rear-edge, sino en general.
+Es decir, desde mi punto de vista, aunque decidas no centrarte tanto en el aspecto metodológico como sugieren Regino y Curro, eso no quiere decir que este punto no lo presentes de forma explícita en tu trabajo (en abstract debe aparecer, luego en intro, discusión y tal vez una frase en conclusiones) como uno de los elementos fuertes (un término medio, vamos).
+Creo que ya hemos discutido esto cuando hemos hablado.
+No es el primer trabajo que lo hace, como bien señalas, pero tampoco hay que pensar que no tiene valor, pues no hay tantos trabajos que combinen ambas aproximaciones (habrá cada vez más de hecho).
+Así que ánimo que queda poco y el trabajo puede quedar interesante si somos capaces de escribir e interpretar bien los resultados que nos has enviado.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+# Notas Jump 2010 
+Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)
+
+NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.
+
+Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!
+
+Otra idea --> Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)
+
+NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.
+
+Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!
+
+El NDVI se ha utilizado para monitorizar la respuesta de bosques frente a sequía (ej.: 24,26,32, 44 en Jump2009 )
+
+Necesidad de Ground-based assessment (p 1974 Jump 2009) -> Aunque utilizemos datos de satélite, siempre es necesario utilizar aproximaciones combinda
+
+Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
+(ver esto y completar)
+
+El problema de los datos ground-based assessment, existe un trade-off entre el área total cubierta por una red de monitoreo y su resolución local.
+--Por tanto también existen limitaciones de las ground-based para estudiar poblaciones en su rear edge.
+")
+```
 

### 2018-02-15 man/ms_3_discussion.Rmd
update

@@
 
+## MAs cosas diferentes 
+Otras especies: rear edge
+
+Evaluar el crecimiento en el rear edge de F. sylvatica
+
+https://link.springer.com/article/10.1007/s10342-016-0982-7?wt_mc=Internal.Event.1.SEM.ArticleAuthorOnlineFirst
+
+¿Que limita al crecimiento en el borde sur de su distribucion? --> Ver esto 
+http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2486.2006.01250.x/abstract
+
+Pinus sylvestris (Baza) (Matias et al 2017)
+Radial growth was maximal at medium altitude and treeline of the southernmost populations.
+Temperature was the main factor controlling growth variability along the gradients, although the timing and strength of climatic variables affecting growth shifted with latitude and altitude.
+
+http://onlinelibrary.wiley.com/doi/10.1111/gcb.13627/full
+
+Idea --> algunos autores comentan que puede existir una alteración en el balance competitivo entre especies en mixed stands.
+Por ejemplo en Montseny, borde equatorial para F.
+sylvatica, se está viendo que el F.
+sylvatica está siendo reemplazado por Q.
+ilex.
+...
+En SN, el artículo de B.
+Benito (Climatic Change) habla del remplazo que existirá de Q.
+pyrenaica por Q.
+ilex, sin embargo, estamos viendo que los crecimientos son muy grandes y que Q.
+pyrenaica tiene mucha resiliencia --> entonces que pasa con las predicciones de dichos modelos ???
+
 ## References

### 2018-02-16 analysis/drought/explore_spei_longterm.Rmd
computation drought

@@
 ```
+
+# Datos sequia antes y despues
+* Ver los valores de spei antes y despues
+
+```{r}
+
+dys <- sev12 %>% filter(d_duration > 2) %>% as.data.frame() 
+dys <- dys[,'maxyear']
+
+f <- c()
+
+for (y in dys){
+
+  aux <- speidf %>% 
+    dplyr::select(year, month, SPEI12_mean)
+
+  aux1 <- aux %>% 
+    filter(year == y + 1) %>% 
+    summarise(speiyear_mean = mean(SPEI12_mean), 
+              speiyear_sum = sum(SPEI12_mean)) %>% 
+    mutate(droughtyear = y, t = 1)
+
+    aux2 <- aux %>% 
+    filter(year == y + 2) %>% 
+    summarise(speiyear_mean = mean(SPEI12_mean), 
+              speiyear_sum = sum(SPEI12_mean)) %>% 
+    mutate(droughtyear = y, t = 2)
+
+      aux3 <- aux %>% 
+    filter(year == y + 3) %>% 
+    summarise(speiyear_mean = mean(SPEI12_mean), 
+              speiyear_sum = sum(SPEI12_mean)) %>% 
+    mutate(droughtyear = y, t = 3)
+
+  f <- rbind(f, aux1)
+
+  }
+
+f <- speidf %>% filter(year
+
+  sev12
+
+```

### 2018-02-16 analysis/bai/trends_resilience_bai.Rmd
add post event

--- 2018-02-12 analysis/bai/trends_resilience_bai.Rmd
@@
 library('devtools')
+library("ggrepel")
+library("patchwork")
 ```
@@
 * Computar métricas de resiliencia BAI para los tres sitios. 
-* Computar tres eventos climáticos: 1995, 2005, 2012
-* Computar ventanas temporales: 2, 3 y 4
+* Computar varios eventos climáticos: 1995, 2005, 2012 ... (ver dys)
+* Computar ventanas temporales: 3
 
@@
 ```{r}
+# Computar resiliencia para otros years (ventana temporal 3y) 
 # SJ 
@@
 ```{r}
+# Computar resiliencia para otros years (ventana temporal 10y) 
+# SJ 
+# Remove 1914 
+dyssj <- dys[which(dys!=1914)]
+res10sj <- baiResilience(bai_sj, event_years = dyssj, window = 10)
+# caL
+res10caL <- baiResilience(bai_caL, event_years = dys, window = 10)
+# caH
+res10caH <- baiResilience(bai_caH, event_years = dys, window = 10)
+```
+
+* Sample depth?
+```{r}
+## Sample depth (#trees by site)
+sampleDepth <- function(x, site){ 
+  x$samps <- rowSums(!is.na(x))
+  x$year <- as.numeric(row.names(x))
+  x$site <- site
+  out <- x[,c("samps", "year", "site")]
+  return(out)
+}
+
+d_sj <- sampleDepth(sj_tree, site='SJ')
+d_caL <- sampleDepth(caL_tree, site='caL')
+d_caH <- sampleDepth(caH_tree, site='caH')
+
+d <- rbind(d_sj, d_caL, d_caH)
+```
+
+```{r}
 # Prepara data 
@@
 re <- bind_rows(rsj, rcaL, rcaH)
-# re$disturb_year <- as.factor(re$disturb_year)
 re$site <- as.factor(re$site)
 
-re %>% ggplot(aes(x=disturb_year, ))
-
+# Get mean values by site and add sample deepth
 s <- re %>% 
   group_by(site, disturb_year) %>% 
-  summarise_at(c("rt", "rs", "rc", "rrs"), funs(mean,sd,se=sd(.)/sqrt(n())))
+  summarise_at(c("rt", "rs", "rc", "rrs"), funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(d, by=c("disturb_year" = "year", "site"="site")) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear")) 
+```
 
-s %>% 
+```{r}
+# Prepara data 
+rsj10 <- res10sj$resilience %>% mutate(site='SJ')
+rcaL10<- res10caL$resilience %>% mutate(site='caL')
+rcaH10 <- res10caH$resilience %>% mutate(site='caH')
+
+re10 <- bind_rows(rsj10, rcaL10, rcaH10)
+re10$site <- as.factor(re10$site)
+
+# Get mean values by site and add sample deepth
+s10 <- re10 %>% 
+  group_by(site, disturb_year) %>% 
+  summarise_at(c("rt", "rs", "rc", "rrs"), funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(d, by=c("disturb_year" = "year", "site"="site")) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear")) 
+```
+
+### Plot Resilience temporal 
+* Resilience of several drought years. 
+* Point size = duration of the drought event 
+* point = mean value of resilience (para eventos con mas de 10 árboles, samps > 10). 
+* error bar = standard_deviation of resilience.
+
+Para cada árbol se calcula, la resiliencia (wsize = 3 years) de BAI en cada uno de los años de sequía.
+Si para un año de sequía no tenemos un mínimo de 10 años, pues entonces no lo ploteamos.
+Las barras de error indican la sd de los valores de resiliencia de los arboles para ese evento de sequía.
+
+```{r}
+# ojo he ploteado los que tiene mas de 10 árboles 
+resi_temporal <- s %>% filter(samps >= 10) %>% 
   ggplot(aes(y=rs_mean, x=disturb_year, colour=site)) + 
@@
                 position=position_dodge(2)) +
-  geom_point(size=2, shape=21, fill='white', position=position_dodge(2)) + 
-  #facet_wrap(~site, scales='fixed', ncol=1) + 
-  theme_bw() # + geom_smooth(aes(colour=site), se=FALSE)
+  geom_point(aes(size=d_duration), shape=21, fill='white', position=position_dodge(2)) + 
+  facet_wrap(~site, scales='free_y', ncol=1) + 
+  theme_bw() + 
+  theme(strip.background = element_rect(fill='transparent')) + 
+  ylab('Resilience (Rs)') + xlab('years') + 
+  scale_x_continuous(breaks=seq(1910,2020, by=5)) # + geom_smooth(aes(colour=site), se=FALSE)
+
+resi_temporal
+
+pdf(here::here('/out/bai_resiliences_severes', 'res_bai_longterm.pdf'), width=12, height = 9)
+resi_temporal
+dev.off()
+```
+
+```{r} 
+resi_temporal10 <- s10 %>% filter(samps >= 10) %>% 
+  ggplot(aes(y=rs_mean, x=disturb_year, colour=site)) + 
+  geom_errorbar(aes(ymin=rs_mean - rs_sd, 
+                    ymax=rs_mean + rs_sd),
+                width = 0.2, 
+                position=position_dodge(2)) +
+  geom_point(aes(size=d_duration), shape=21, fill='white', position=position_dodge(2)) + 
+  facet_wrap(~site, scales='free_y', ncol=1) + 
+  theme_bw() + 
+  theme(strip.background = element_rect(fill='transparent')) + 
+  ylab('Resilience (Rs)') + xlab('years') + 
+  scale_x_continuous(breaks=seq(1910,2020, by=5)) # + geom_smooth(aes(colour=site), se=FALSE)
+
+pdf(here::here('/out/bai_resiliences_severes', 'res_bai_longterm10.pdf'), width=12, height = 9)
+resi_temporal10
+dev.off()
+```
+
+### Relation resiliencia y sequias (intensidad, etc)
+
+```{r}
+#https://gist.github.com/ottadini/6882677 
+lm_eqn = function(m) {
+  # Displays regression line equation and R^2 value on plot
+  # Usage:
+  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
+
+  l <- list(#a = format(coef(m)[1], digits = 2),
+            # b = format(abs(coef(m)[2]), digits = 2),
+            r2 = format(summary(m)$r.squared, digits = 3));
+
+  if (coef(m)[2] >= 0)  {
+    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
+    eq <- substitute(~~italic(r)^2~"="~r2,l)
+  } else {
+    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
+    eq <- substitute(~~italic(r)^2~"="~r2,l)
+  }
+
+  as.character(as.expression(eq));                 
+}
 
 ```
+
+```{r}
+sdf <- s %>% filter(samps >= 10) %>% 
+  group_by(disturb_year) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+plotrel <- function(df, v, x, label, XLab, YLab){
+  require(ggplot2)
+  require(ggrepel)
+
+  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
+
+  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
+                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
+
+  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
+    geom_errorbar(mapping = yerrbar) + 
+    geom_point(size=4, pch = 21, colour="black", fill="white") + 
+    theme_bw() + 
+    geom_smooth(method='lm', se=FALSE) +
+    xlab(XLab) + ylab(YLab) + geom_smooth(method='lm', se=FALSE, colour="black") +
+    geom_text_repel() + 
+    annotate("text", x=7, y = 0.5, 
+             label=lm_eqn(lm(formulita, df)), parse=TRUE)
+
+  }
+
+sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)")
+
+sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Resistance (Rt) (3year window size)")
+
+sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Recovery (Rc) (3year window size)")
+
+pdf(here::here('/out/bai_resiliences_severes', 'rs_vs_drought_3years.pdf'), width=18, height = 7)
+sevRt + sevRc + sevRs 
+dev.off()
+```
+
+```{r}
+sdf10 <- s10 %>% filter(samps >= 10) %>% group_by(disturb_year) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+sevRs10 <- plotrel(sdf10, v = "rs", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Resilience (Rs) (10year window size)")
+
+sevRt10 <- plotrel(sdf10, v = "rt", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Resistance (Rt) (10year window size)")
+
+sevRc10 <- plotrel(sdf10, v = "rc", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Recovery (Rc) (10year window size)")
+
+pdf(here::here('/out/bai_resiliences_severes', 'rs_vs_drought_10years.pdf'), width=18, height = 7)
+sevRt10 + sevRc10 + sevRs10 
+dev.off()
+```
+
+# Plots de Resiliencies con lo que ocurre tras el evento 
+```{r}
+
+post <- read.csv(file=here::here('data/sequias', 'post_severe.csv'), header=TRUE)
+names(post) <- c('spei_avg', 'spei_sum', 'dyear', 'tlag')
+
+u <- post %>% 
+  gather(variable, value, -(dyear:tlag)) %>%
+  unite(temp, variable, tlag) %>%
+  spread(temp, value)
+
+postdf <- s %>% filter(samps >= 10) %>% 
+  group_by(disturb_year) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(u, by=c("disturb_year" = "dyear"))
+
+rsdf <- postdf %>% 
+  dplyr::select(c(rs_mean_mean, rs_mean_se, spei_avg_1:spei_sum_30)) %>% 
+  mutate(v = 'rs') %>% 
+  rename(mean = rs_mean_mean, se = rs_mean_se)
+
+rcdf <- postdf %>% 
+  dplyr::select(c(rc_mean_mean, rc_mean_se, spei_avg_1:spei_sum_30)) %>% 
+  mutate(v = 'rc') %>% 
+  rename(mean = rc_mean_mean, se = rc_mean_se)
+
+rtdf <- postdf %>% 
+  dplyr::select(c(rt_mean_mean, rt_mean_se, spei_avg_1:spei_sum_30)) %>% 
+  mutate(v = 'rt') %>% 
+  rename(mean = rt_mean_mean, se = rt_mean_se)
+
+auxdf <- rbind(rsdf, rcdf, rtdf) %>% 
+  gather(-mean, -se, -v, key="var", value = "value")
+
+# Get r2 
+variables <- c('rs', 'rc', 'rt')
+
+auxdf_eq <- c()
+
+for (vari in variables){ 
+  varidf  <- auxdf %>% filter(v == vari)
+
+  speis <- unique(varidf$var)
+
+  o <- c()
+  out <- data.frame()
+
+  for (s in speis){ 
+    varidf_spei <- varidf %>% filter(var == s)
+
+    m <- lm(mean ~ value, varidf_spei)
+
+    o$var <- as.character(s)
+    o$v <- as.character(vari)
+    o$a <- as.numeric(format(coef(m)[1], digits = 5))
+    o$b <- as.numeric(format(coef(m)[2], digits = 5))
+    o$r2 <- as.numeric(format(summary(m)$r.squared, digits = 3))
+
+    o <- as.data.frame(o)
+
+    out <- rbind(out, o)
+  }
+
+  auxdf_eq <- rbind(auxdf_eq, out)
+}
+
+# Resiliences vs SPEI avg 
+auxdf_avg <- auxdf %>% filter(var %in% c('spei_avg_1', 'spei_avg_2', 'spei_avg_3', 'spei_avg_20', 'spei_avg_30')) 
+auxdf_avgeq <- auxdf_eq %>% filter(var %in% c('spei_avg_1', 'spei_avg_2', 'spei_avg_3', 'spei_avg_20', 'spei_avg_30'))
+
+rs_vs_post_avg <- auxdf_avg %>% 
+  ggplot(aes(x = value, y = mean)) + 
+  facet_grid(v ~ var, scales = "free_y") +
+  geom_point() + theme_bw() + 
+  geom_errorbar(aes(ymin = mean - se, ymax =mean +se)) +
+  geom_smooth(method='lm', se=FALSE, colour="black") +
+  geom_text(data=auxdf_avgeq, aes(x =-.5, y = 0.6, label=r2), parse = TRUE, inherit.aes=FALSE)
+
+pdf(here::here('/out/bai_resiliences_severes', 'rs_vs_post_avg.pdf'), width=14, height =10)
+rs_vs_post_avg 
+dev.off()
+
+# Resiliences vs SPEI sum
+auxdf_sum <- auxdf %>% filter(var %in% c('spei_sum_1', 'spei_sum_2', 'spei_sum_3', 'spei_sum_20', 'spei_sum_30')) 
+auxdf_sumeq <- auxdf_eq %>% filter(var %in% c('spei_sum_1', 'spei_sum_2', 'spei_sum_3', 'spei_sum_20', 'spei_sum_30'))
+
+rs_vs_post_sum <- auxdf_sum %>% 
+  ggplot(aes(x = value, y = mean)) + 
+  facet_grid(v ~ var, scales = "free_y") +
+  geom_point() + theme_bw() + 
+  geom_errorbar(aes(ymin = mean - se, ymax =mean +se)) +
+  geom_smooth(method='lm', se=FALSE, colour="black") +
+  geom_text(data=auxdf_sumeq, aes(x =-.5, y = 0.6, label=r2), parse = TRUE, inherit.aes=FALSE)
+
+pdf(here::here('/out/bai_resiliences_severes', 'rs_vs_post_sum.pdf'), width=14, height =10)
+rs_vs_post_sum
+dev.off()
+
+```

### 2018-02-16 analysis/drought/explore_spei_longterm.Rmd
add post event

--- 2018-02-12 analysis/drought/explore_spei_longterm.Rmd
@@
 ```
+
+# Datos sequia antes y despues
+* Ver los valores de spei antes y despues.
+Computa para el t+1, t+2 y t+3 el valor de spei medio (media de todos los valores de spei del año t+1 o t+2 o t+3) y el valor de spei acumulado (lo mismo pero sumando, ojo sin poner en valor absoluto), es decir, me
+)media de to
+```{r}
+
+dys <- sev12 %>% filter(d_duration > 2) %>% as.data.frame() 
+dys <- dys[,'maxyear']
+
+f <- c()
+
+for (y in dys){
+
+  aux <- speidf %>% 
+    dplyr::select(year, month, SPEI12_mean)
+
+  aux1 <- aux %>% filter(year == y + 1) %>% 
+    summarise(speiyear_mean = mean(SPEI12_mean), 
+              speiyear_sum = sum(SPEI12_mean)) %>% 
+    mutate(droughtyear = y, tlag = 1)
+
+  aux2 <- aux %>% filter(year == y + 2) %>% 
+    summarise(speiyear_mean = mean(SPEI12_mean), 
+              speiyear_sum = sum(SPEI12_mean)) %>% 
+    mutate(droughtyear = y, tlag = 2)
+
+  aux3 <- aux %>% filter(year == y + 3) %>% 
+    summarise(speiyear_mean = mean(SPEI12_mean), 
+              speiyear_sum = sum(SPEI12_mean)) %>% 
+    mutate(droughtyear = y, tlag = 3)
+
+  aux2ev <- aux %>% 
+    filter(year >= y + 1) %>% 
+    filter(year <= y + 2) %>% 
+    summarise(speiyear_mean = mean(SPEI12_mean), 
+              speiyear_sum = sum(SPEI12_mean)) %>% 
+    mutate(droughtyear = y, tlag = 20)
+
+  aux3ev <- aux %>% 
+    filter(year >= y + 1) %>% 
+    filter(year <= y + 3) %>% 
+    summarise(speiyear_mean = mean(SPEI12_mean), 
+              speiyear_sum = sum(SPEI12_mean)) %>% 
+    mutate(droughtyear = y, tlag = 30)
+
+  f <- rbind(f, aux1, aux2, aux2ev, aux3, aux3ev)
+  }
+
+write.csv(f, file=here::here('data/sequias', 'post_severe.csv'), row.names=FALSE)
+```

### 2018-02-19 analysis/drought/explore_prec_longterm.Rmd
update

@@
 library("ggrepel")
+library("sp")
+library("sf")
+library("pander")
 ```
@@
 ```
+
+# Mapa de estaciones 
+```{r}
+es <- read.csv(here::here('data_raw/meteo/prec_subclima/mdestaciones', 'estaciones_latlong.csv'), header=TRUE, sep=";")
+
+es <- st_as_sf(x = es, 
+               coords = c("Longitude","Latitude"), 
+               crs = "+proj=longlat +datum=WGS84")
+
+# mapview(es)
+
+```

### 2018-02-19 man/ms_1_material_methods.Rmd
update

@@
 The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], undergoing intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
-In fact, the status of conservation of this species for southern Spain is "Vulnerable" [@Vivero2000].
+In fact, conservation status of this species for southern Spain is "Vulnerable" [@Vivero2000].
 The relict presence of this species in Sierra Nevada is related both to its genetic resilience and high intraspecific genetic diversity [@Valbuena2010].
@@
 1995, 1981) [@VicenteSerrano2014], no se han elegido por una limitación metodológica, esto es, nuestro objetivo es analizar el efecto de la sequía y la resiliencia a los eventos de sequía utilizando información de satélite y de dendrocronología, y la información de satélite utilizada (ver mas adelante) solamente está disponible desde el año 2000 en adelante.
+We compare only these previous drought years in radial growth-time series, to place older drought events with those in 2005 and 2012 ...
+DUDA GUILLERMO
+
+* ¿Decimos aquí algo así como?: Para los datos de radial-growth hemos computado la resiliencia en los 10 eventos mas severos (ver tabla S1)
 
@@
 The filtering was done using quality flags and VI Usefulness Indices accompanying the EVI data.
-We filter out those values affected by high content of aerosols, clouds, snow and shadows, following the filtering recomendations of EVI data for mountain regions [@Reyes2015].
+We filter out those values affected by high content of aerosols, clouds, snow and shadows, following the filtering recommendations of EVI data for mountain regions [@Reyes2015].
 

### 2018-02-19 man/ms_2_results.Rmd
update results

@@
 The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
+The lowest values for EVI standardized anomalies were recorded in 2005, being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2).
+According to the standardized anomalies, *Q.
+pyrenaica* forests suffered a "browning episode" during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S2).
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
 The strongest trends were observed in southwestern populations (Figure S2), in which the Cañar site are included Vegetation greenness of *Q.
@@
 According to the standardized anomalies, *Q.
-pyrenaica* forests suffered a "browning episode" during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S3).
+pyrenaica* forests suffered a browning episode during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S3).
+")
+```
 
@@
 pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3a).
-The 2005 drought reduced the greenness of oaks to a 85.8 % respect the reference period while the 2012 reduced it 94.3 % (Table S1).
-Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1.
-Figures 3a, 4a).
+Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figures 3a).
 
@@
 
-The oak forests recovered (*Rc*) their greenness significantly more rapidly after the 2005 drought than after 2012 (Tables 3, S1).
 In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
@@
 
-Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables 3, S1), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
+Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables S1, 3), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
 The southern populations showed higher resilience values than the northern ones, although they were not significantly different for 2005 drought (p = 0.036; Figures 3a, 4a).
@@
 
-Likely as a consequence of their different management origin (i.e.
-land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2).
+Likely as a consequence of their different management origin (*i.e.* land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2).
 The competition index was similar among sites but plot basal area was greatest in Ca-High (Table 1).
@@
 Trees from this location showed in addition a positive growth trend since the late 1970s.
-This positive trend was not found in any of the other two locations and the two altitudes sampled in the Southern populations showed differences in growth, being more similar Ca_Low with growth in the Northern population (SJ) than with Ca_High (Figure 5).
+This positive trend was not found in any of the other two locations and the two altitudes sampled in the Southern populations showed differences in growth, being more similar Ca_Low with growth in the northern population (SJ) than with Ca_High (Figure 5).
 `VAS POR AQUI CARPINTO (BREAK)`

### 2018-02-19 man/ms_figures.Rmd
update results

@@
 
-## Figure 2. Plot combinado con EVI profile (a) y anomalías (b)
+## Figure 2. Plot de anomalias
+- [X] Solamente Anomalias estandarizadas
+
+## Figure 3. 
 ")
@@
 
-```{r}
+```{r, eval=FALSE}
 iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
@@
 
-```{r, fig.height=7}
+```{r, fig.height=7, eval=FALSE}
 a_profile_compara <- profile_compara + labs(title = "a")
@@
 
-**Figure 2.** Comparison of EVI profile for the reference period (*gray*) and during the 2005 (*green*) and 2012 (*blue*) drought events (a) (mean of all analyzed pixels).
-EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations (b).
-Error bars show standard error.
+```{r, fig.height=7}
+plot_sa_clu 
+```
+
+**Figure 2.** EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations. Error bars show standard error.
 
@@
   ggtitle("b) BAI")
+
+```
+
+```{r}
+
+resilience_metrics <- c("0_rt" = "Resistance", "1_rc" = "Recovery", "2_rs" = "Resilience")
+variables <- c("0_EVI"="EVI", "1_BAI"="BAI")
+# width
+w <- .6
+pos <- position_dodge(width=w)
+
+line_lwd <- .8
+
+aux <- mhuber_filter %>% mutate(vari = '0_EVI')
+aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
+m <- bind_rows(aux2, aux)
+
+m %>% 
+  ggplot(aes(x = disturb_year, y = M.Huber, 
+                   color = site, group = site, 
+                   fill = site,
+                   ymin = lower.ci, ymax = upper.ci)) + 
+  geom_bar(stat = "identity", position="dodge", width=w) +
+  geom_errorbar(width=.2, position = pos) + 
+  facet_grid(var_sorted~vari, labeller = labeller(
+    var_sorted = resilience_metrics, 
+    vari = variables), 
+    scales = "free_y") + theme_bw() + 
+  theme_bw() + 
+  theme(
+    panel.grid.minor = element_blank(),
+    strip.background = element_rect(
+      colour = "black",
+      fill = "white"),
+    strip.text = element_text(face="bold", size = 14),
+    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1)))+  
+  ylab("Response value")+
+  xlab('Drought event') + 
+  geom_text(aes(label=Letter, y = (M.Huber + .05*lower.ci)), position=pos, vjust=-2) + 
+  scale_color_manual(values=c('gray', "#104E8B", 'gray', "#1E90FF", "#104E8B")) +
+  scale_fill_manual(values=c('gray', "#104E8B", 'gray', "#1E90FF", "#104E8B"))
 

### 2018-02-20 man/ms_figures.Rmd
add changes to fig and upload new template

--- 2018-02-12 man/ms_figures.Rmd
@@
     highlight: null
-    reference_docx: template_ms.docx
+    reference_docx: template_ms_new_page.docx
 bibliography: ../refs/references.bib
@@
 library("patchwork")
+library("ggcorrplot")
+library("stringr")
+library("pander")
+library("lemon")
 ```
@@
 
-### Figure 1
+#####
 
@@
 
-### Figure 2
+#####
 
-```{r}
+```{r, eval=FALSE}
 iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
@@
 
-**Figure 2.** Comparison of EVI profile for the reference period (*gray*) and during the 2005 (*green*) and 2012 (*blue*) drought events (a) (mean of all analyzed pixels).
-EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations (b).
-Error bars show standard error.
+```{r, fig.height=5}
+plot_sa_clu 
+```
 
-### Figure 3
+**Figure 2.** EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations. Error bars show standard error.
 
-```{r}
+##### 
+```{r, eval=FALSE, echo=FALSE}
 mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
@@
 
-```{r}
+```{r, eval=FALSE, echo=FALSE}
 # EVI_drought 
@@
 
-```{r, fig.height=7}
+```{r, eval=FALSE, fig.height=7}
 res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
@@
 
-**Figure 3.** Response of northern (*black*) and southern (*blue*) populations of *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) for the years 2005 and 2012.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
+```{r, fig.height=7}
+mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
+mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
 
+# NO plot rrs 
+mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
+mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
+
+aux <- mhuber_filter %>% mutate(vari = '0_EVI')
+aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
+m <- bind_rows(aux2, aux)
+
+pd <- position_dodge(width = 0.6)
+
+aes_disturb <- aes(x = disturb_year, y = M.Huber, 
+                   color = site,  
+                   fill = site, linetype=site,
+                   shape = site, 
+                   ymin = lower.ci, ymax = upper.ci)
+
+variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
+
+themeR <- list(
+  geom_errorbar(width=.1, position = pd),
+  geom_point(position = pd),
+  theme_bw(),
+  facet_wrap(~vari, labeller = labeller(vari = variables)),
+  xlab(''), 
+  theme(
+    panel.grid.minor = element_blank(),
+    strip.background = element_rect(
+      colour = "black",
+      fill = "white"),
+    strip.text = element_text(face="bold", size = 14),
+    axis.title.y = element_text(face="bold", size = 14),
+    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
+  scale_shape_manual(values=c(19, 21, 17, 22, 15)),
+  scale_color_manual(values=c('black', "blue", 'black','blue','blue')),
+  scale_fill_manual(values=c('black', 'white', 'black','white','blue')),
+  scale_linetype_manual(values=c("solid","solid","solid","solid", "solid")),
+  scale_size_manual(values=c(2,2,2.5, 2.5, 2.5))
+)
+
+rt <- m %>% filter(var == "rt") %>% 
+  ggplot(aes_disturb) + themeR + 
+  ylab("Resistance") + 
+  geom_text(aes(label=Letter, y = 1.1), position=pd, show.legend = FALSE)
+
+rc <- m %>% filter(var == "rc") %>% 
+  ggplot(aes_disturb) + themeR + 
+  ylab("Recovery")+
+  geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
+
+rs <- m %>% filter(var == "rs") %>% 
+  ggplot(aes_disturb) + themeR + 
+  ylab("Resilience") + xlab('Drought event') +
+  geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
+```
+
+```{r, fig.height=6}
+grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top')
+```
+
+**Figure 3.** Response *Q.
+pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
+For EVI we compared northern populations (black fill circle) with southern ones (blue empty circle).
+For BAI we compared northern population (San Juan, SJ; black triangle) with southerns populations: Cáñar-High (Ca-H; blue empty squares) and Cáñar-Low (Ca-L; blue fill sauares).
+Different letters above error bars indicate significant post hoc differences between groups (see material and methods).
+
+```{r, eval=FALSE}
 ### Figure 4
-```{r}
+
 theme_de <- list(
@@
 
-```{r}
+```{r, eval=FALSE}
 res_evi_site  + res_bai_site + plot_layout(ncol=1)
-```
 
@@
 Different letters above bars indicate significant post hoc differences between groups (see material and methods).
+```
 
-### Figure 5
+#####
 
@@
 
-cronos_bai <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+# cores 
+cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
+  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
   theme_bw() + ylab('BAI') + 
@@
   ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
+  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
+  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA_High'), 
+              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
   # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
   # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-
 samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
@@
         axis.title.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(), 
-        plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")) + 
+        axis.text.x=element_text(size=6),
+        # axis.text.x=element_blank(),
+        # axis.ticks.x=element_blank(), 
+        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
+        ) + 
   scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
@@
   scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-
   # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
@@
 
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.1, 0.9))
-
+plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
 plot_bai_s
@@
 
-**Figure 5.** Basal Area Increment (BAI) chronologies of *Q.
+```{r, eval=FALSE}
+
+# Lm CA-High
+df <- cro_sites %>% 
+  filter(year >= 1975) %>% 
+  filter(site == 'CA_High')
+
+m1 <- lm(bai_mean ~ year, df)
+summary(m1)
+r2 <- format(summary(m1)$r.squared, digits = 3)
+```
+
+**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
 pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA_Low; *pink*) and high-elevation (CA_High, *purple*) sites.
@@
 Number of series are displayed in the upper plot.
+We only show chronologies with # trees > 5.
 
-# Figure 6
+#####
 
@@
 year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
+
+# peores sequias 
+dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
+dy <- dy %>% arrange(desc(d_duration))
+# solamente aquellos mayor de 2 meses 
+dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
+dys <- dys[,'maxyear'] 
 ```
@@
 label_site <- c("0_sj" = "Northern (SJ)",
-                "1_caH" = "Southern High (CaH)",
-                "2_caL" = "Southern Low (CaL)")
+                "1_caH" = "Southern (Ca-High)",
+                "2_caL" = "Southern Low (Ca-Low)")
+
+cro_rwi <- cro_rwi %>% 
+  mutate(siteF = case_when(
+    site_sorted == "0_sj" ~ "Northern (SJ)",
+    site_sorted == "1_caH" ~ "Southern (Ca-High)",
+    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
 
@@
         strip.text = element_text(face="bold", size=12), 
-        axis.text = element_text(face="bold", size=10), 
+        # strip.text = element_blank(), 
+        axis.text = element_text(face="bold", size=8), 
         axis.title = element_text(face="bold", size=12)) + 
@@
   geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "fixed", ncol = 1,
+  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
              labeller = as_labeller(label_site)) + 
   ylab("Ring width index") + xlab("Year") + 
+  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
+  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
   geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
+  # geom_vline(xintercept = dys, linetype = "dotted") + 
   scale_x_continuous(limits = c(1819, 2016), 
@@
 plot_rwi_crono  
-
 ```
 
-**Figure 6.** Residual Tree-ring chronologies obtained for the *Q.
+**Figure 5.** Residual Tree-ring chronologies obtained for the *Q.
 pyrenaica* sites.
 Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three most recent severe drought years (1995, 2005 and 2012).
+Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
 
-# Figure 7
+##### 
 ```{r}
@@
 
-**Figure 7.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
+**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
 pyrenaica* and monthly climatic data (precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures) for northern site (*green* bars), low-elevation southern site (*light blue* bars) and high-elevation shouthern (*dark blue* bars) site.
@@
 
-# Figure 8
+#####
 
@@
   facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "Southern High (CaH)", "2_caL" = "Southern Low (CaL)"))) + 
+             labeller=as_labeller(c("1_caH" = "Southern (CA-High)", "2_caL" = "Southern(CA-Low)"))) + 
     ylab(custom_ylab) + custom_corrected_gc
@@
 
-**Figure 8.** Comparison of median growth change ($GC$) following [@Nowacki1997] for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
+**Figure 7.** Comparison of median growth change ($GC$) following [@Nowacki1997] for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
 
-# Figure 9
+#####
 
@@
 
-**Figure 9**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
+**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
+
+# Supplementary
+
+##### 
+**Appendix S1** Acumulated monthly precipitation during the hydrological year 2004-2005 (blue line) and 2011-2012 (red line).
+The boxplot representing the average from 1940-2015 period.
+Data from meteorological station Granada, Base Aérea (National Spanish Meteorological Services, AEMET).
+
+```{r, fig.height=5}
+ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
+
+source(here("/script/R/", "hydro_year.R"))
+
+ltdf <- ltdf %>% 
+  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
+  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
+         month = lubridate::month(fecha),
+         hmonth = ifelse(month <= 8, month + 4, month -8), 
+         hyear = hydro_year(fecha),
+         hyear_f = paste0(hyear-1,'-', hyear))
+
+# We used only the 5514 station (the most long term)
+# Filter by station and prec 
+# Filter from 1950 
+st5514 <- ltdf %>% 
+  filter(VARIABLE == 'PM1') %>% 
+  filter(cod == '5514') %>% 
+  filter(hyear >= 1950) %>% 
+  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
+
+st5514_pro <- st5514 %>% 
+  group_by(hyear_f, hmonth) %>% 
+  dplyr::summarise(pre = sum(pre)) %>%
+  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
+
+st5514_pro  %>% 
+  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
+  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
+  geom_boxplot(aes(x=as.factor(hmonth))) +  
+  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
+  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
+            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
+  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
+            aes(x=as.factor(hmonth), y=csum, group=1),
+            colour='red') +
+  theme_bw() + 
+  theme(strip.background = element_rect(fill = NA, color = "black")) + 
+  ylab('Precipitation (mm) acumulated') + xlab('') + 
+  ggtitle('Granada Base Aerea, 1950-2015') + 
+  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
+                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
+  annotate("text", label = "2004-2005", x = 2, y = 500, size = 6, colour = "blue") + 
+  annotate("text", label = "2011-2012", x = 2, y = 550, size = 6, colour = "red") + 
+  annotate("text", label = "mean 1950-2015", x = 3, y = 600, size = 6, colour = "black")
+```
+
+#####
+
+```{r, echo=FALSE}
+# Get file names
+myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
+
+mydf <- data.frame()
+
+# Loop to read files and prepare data
+for (j in myfiles){ 
+  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
+
+  # Remove csv and get name 
+  name_aux <- str_replace(j, ".csv", "")
+
+  # Get lat long
+  latlong <- str_replace(name_aux, "SPEI_", "")
+
+  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
+  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
+
+  aux$lat <- mylat
+  aux$long <- mylong
+
+  # Split time 
+  # Get year 
+  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
+
+  # Get month 
+  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
+
+  # Variable month in number 
+  aux$months <- match(aux$month, month.abb)
+
+  # assign(name_aux, aux)
+
+  mydf <- rbind(mydf, aux)
+}
+
+spei <- mydf %>% select(SPEI_12, year, months) %>% 
+  group_by(year,months) %>% 
+  summarise(spei6 = mean(SPEI_12),
+            spei6_sd = sd(SPEI_12)) %>% 
+  mutate(date = lubridate::ymd(paste(year, months, "01")))
+
+spei <- spei %>% 
+  mutate(sign = case_when(
+    spei6 > 0 ~ "pos",
+    TRUE ~ "neg"))
+```
+
+```{r, fig.height=5}
+limits <- as.Date(c("1949-01-01", "2015-12-01"))
+spei %>% 
+  ggplot(aes(x=date, y=spei6, fill=sign)) +  
+  geom_bar(stat = "identity") + 
+  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
+  theme_bw()  + 
+  labs(title= 'Drought evolution for Sierra Nevada',
+    x= '',
+    y='SPEI') +
+  theme(strip.background = element_rect(fill = "white"),
+        legend.position = "none") +
+  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
+```
+
+**Appendix S2**.
+Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+We used data from Global SPEI database.
+We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
+
+```{r, echo=FALSE, eval=FALSE}
+**Figure S2**.
+Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
+Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
+
+![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
+```
+
+##### 
+```{r}
+prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
+
+dfpre <- c()
+
+for (i in prec_files){
+
+  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
+  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
+
+  dfpre <- rbind(dfpre, auxdf)
+}
+
+dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
+
+dfpre <- dfpre %>% 
+  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
+         month = lubridate::month(fecha),
+         hmonth = ifelse(month <= 8, month + 4, month -8), 
+         hyear = hydro_year(fecha),
+         hyear_f = paste0(hyear-1,'-', hyear))
+
+aux <- dfpre %>% 
+  group_by(hyear_f, code) %>% 
+  dplyr::mutate(csum = cumsum(prec)) %>% 
+  filter(hmonth == 12) %>% 
+  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
+
+presn <- aux %>% group_by(hyear) %>% 
+     dplyr::summarise(
+     csum_mean = mean(csum, na.rm = TRUE),
+     csum_sd = sd(csum, na.rm = TRUE), 
+     csum_se = csum_sd/sqrt(length(csum_sd)), 
+     n = n())
+
+meandf <- mean(presn$csum_mean)
+sddf <- sd(presn$csum_mean)
+
+prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
+  geom_hline(yintercept = meandf) + 
+  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
+  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
+  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
+  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
+  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
+  geom_point(size=2) + theme_bw() +
+  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
+  scale_y_continuous(breaks=seq(0,1500, by=250)) +
+  geom_text_repel(
+    data = subset(presn, csum_mean < (meandf - 1*sddf)),
+    aes(label = hyear)) +
+  theme(panel.grid.minor = element_blank()) +
+  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
+prec_hydrol_sn
+```
+
+**Appendix S3.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017 for 28 meteorological stations (from National Spanish Meteorological Services, AEMET) distributed around the Sierra Nevada area.
+Points represent mean and errorbars standard error.
+*Black* line indicates mean for all period.
+*Red* lines represent -1 (*dotted* line) and -2 (*dashed* line) standard deviation.
+*Blue* lines represent +1 (*dotted* line) and +2 (*dashed* line) standard deviation.
+We indicate all years with average values below -1SD.
+
+##### 
+```{r}
+# peores sequias 
+dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
+dy <- dy %>% arrange(desc(d_duration))
+# solamente aquellos mayor de 2 meses 
+dys <- dy %>% filter(d_duration > 2) %>% as.data.frame()
+
+pander(dys)
+```
+
+**Appendix S4.** Tabla de drought severas.
+
+##### 
+**Appendix S5.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
+
+```{r}
+df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
+
+df <- df %>% add_row(clu_pop2 = 'Northern slope',
+                             type = 'greening',
+                             count_clu = 0, per = 0.00, y = 2005)
+
+per_pixels_green <- df %>% 
+  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
+  geom_bar(stat='identity', position = "dodge") + 
+  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
+  facet_wrap(~clu_pop2) + 
+  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
+  theme(strip.background = element_rect(colour = "black", fill = "white")
+        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
+  scale_y_continuous(expand=c(.1,0))
+
+per_pixels_green
+```
+
+#####
+
+**Appendix S6.** Reilience vs. droughts
+
+##### 
+**Appendix S7.** Correlation between indices of resilience (Rt, resistance; Rc, recovery; Rs, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
+Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
+ws2, ws3 and ws4 indicate periods of 2, 3 and 4 years after a drought.
+
+```{r}
+
+corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
+corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
+corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
+
+generaMatrix <- function(cordf, variable){ 
+  m <- cordf %>% filter(var == variable) %>% 
+    dplyr::select(-var, -window_size) %>% 
+    mutate(w1 = c(2, 2, 3), 
+           w2 = c(3,4,4)) 
+  mf <- m %>% 
+    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
+
+  df <- m %>% rbind(mf) %>% 
+    add_row(r2 = 1, w1=2, w2=2) %>% 
+    add_row(r2 = 1, w1=3, w2=3) %>% 
+    add_row(r2 = 1, w1=4, w2=4)
+
+    df <- df %>% 
+    mutate(w1 = case_when(
+      w1 == 2 ~ 'ws2',
+      w1 == 3 ~ 'ws3',
+      w1 == 4 ~ 'ws4'),
+      w2 = case_when(
+      w2 == 2 ~ 'ws2',
+      w2 == 3 ~ 'ws3',
+      w2 == 4 ~ 'ws4'))
+
+      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
+
+return(cor_df)
+}
+
+cor_rt <- generaMatrix(corBai, 'rt')
+cor_rc <- generaMatrix(corBai, 'rc')
+cor_rs <- generaMatrix(corBai, 'rs')
+
+cor_rte <- generaMatrix(corEvi, 'rt')
+cor_rce <- generaMatrix(corEvi, 'rc')
+cor_rse <- generaMatrix(corEvi, 'rs')
+
+g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
+g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
+g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
+g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
+g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
+g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
+
+(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
+```
+
+##### 
+**Appendix S8.** Tabla S1
+
+##### 
+**Appendix S9.** Tabla S2

### 2018-02-21 man/ms_figures.Rmd
add new figure

@@
 **Appendix S9.** Tabla S2
+
+#####
+**Appendix S10.** Comparison EVI previous and post
+
+```{r}
+df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
+
+ggplot(df, 
+       aes(x=disturb, y=mean, colour=site)) + 
+  geom_point(size=2) + 
+  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
+  geom_line(aes(group=site)) + 
+  facet_wrap(~disturb_year) + 
+  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
+  theme_bw() +
+  xlab("") + ylab("EVI") + 
+  theme(
+    strip.background = element_rect(fill = "white"),
+    strip.text = element_text(face="bold", size=12),
+    axis.title.y = element_text(face="bold", size=12)
+  ) +
+  scale_color_manual(values=c('black', 'blue', 'gray'))
+```

### 2018-02-21 analysis/evi/compute_resilience_evi.Rmd
create means value of EVI previous, post and during

@@
 library('pander')
+library('here')
 source(paste0(di, '/script/R/eviResilience.R'))
@@
 # Read data
-iv <- read.csv(file=paste(di, "/data/evi_mean.csv", sep=""), header = TRUE, sep = ',')
+iv <- read.csv(file=paste(di, "/data/evi/evi_mean.csv", sep=""), header = TRUE, sep = ',')
 
@@
 
-write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), row.names = F)
+write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size_evi.csv'), row.names = F)
 ```
@@
 write.csv(res3$evi, file=paste0(di, '/data/resilience/evi_period.csv'), row.names = FALSE)
-write.csv(res3$resilience, file=paste0(di, '/data/resilience/resiliences.csv'), row.names = FALSE)
+write.csv(res3$resilience, file=paste0(di, '/data/resilience/resilience_evi.csv'), row.names = FALSE)
 ```
+
+# Ver previos, during and post values of EVI 
+```{r}
+
+# add data of pop
+anomalias <- read.csv(file=paste(di, "/data/anomalies/anomalias_evimean.csv", sep=""), header = TRUE, sep = ',')
+
+attr_iv_malla_modis_id <- anomalias %>% dplyr::select(iv_malla_modi_id,long,lat,pop) %>% unique()
+
+df <- res3$evi %>% inner_join(attr_iv_malla_modis_id, by='iv_malla_modi_id')
+
+dfevi <- df %>% 
+  mutate(
+    clu_pop = as.factor(case_when(
+      pop == 1 ~ "Camarate",
+      pop %in% c(2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
+      pop == 9 ~ 'out')),
+    clu_pop2 = as.factor(case_when(
+      pop %in% c(1,2,3,4,5) ~ 'N',
+      pop %in% c(6,7,8) ~ 'S',
+      pop == 9 ~ 'out'))) %>% 
+  filter(clu_pop != 'out')
+
+# Change name of clu_pop2 and disturb_year 
+dfevi <- dfevi %>% dplyr::rename(site = clu_pop2) %>% 
+  mutate(disturb_year = as.factor(disturb_year)) %>% 
+  mutate(disturb = case_when(
+    disturb == "dr" ~ "1_dr",
+    disturb == "pre" ~ "0_pre",
+    disturb == "post" ~ "2_post"
+  ))
+
+# Get mean value of EVI for 2005 and 2012
+
+dfevi_globalmeans <- dfevi %>% 
+  group_by(disturb, disturb_year) %>% 
+  summarise_at(c("mean_period"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  mutate(site = "all") %>% as.data.frame()
+
+dfevi_globalmeans_site <- dfevi %>% 
+  group_by(disturb, disturb_year, site) %>% 
+  summarise_at(c("mean_period"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% as.data.frame()
+
+dfevi_all_globalmeans <- rbind(dfevi_globalmeans, dfevi_globalmeans_site)
+
+write.csv(dfevi_all_globalmeans, file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"),
+          row.names = FALSE)
+
+ggplot(dfevi_all_globalmeans, 
+       aes(x=disturb, y=mean, colour=site)) + 
+  geom_point(size=2) + 
+  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
+  geom_line(aes(group=site)) + 
+  facet_wrap(~disturb_year) + 
+  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
+  theme_bw() +
+  xlab("") + ylab("EVI") + 
+  theme(
+    strip.background = element_rect(fill = "white"),
+    strip.text = element_text(face="bold", size=12),
+    axis.title.y = element_text(face="bold", size=12)
+  ) +
+  scale_color_manual(values=c('black', 'blue', 'gray'))
+
+```

### 2018-02-21 man/ms_1_material_methods.Rmd
updating

@@
 The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S4) and this length was used in other studies [@Gazol2017].
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Appendix S7) and this length was used in other studies (e.g.
+@Gazol2017).
+In addition we calculated the same indices for BAI data for the most severe drought events (Appendix S4) since 1900.
 

### 2018-02-21 man/ms_2_results.Rmd
updating

@@
 
-The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
-The lowest values for EVI standardized anomalies were recorded in 2005, being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2).
-According to the standardized anomalies, *Q.
-pyrenaica* forests suffered a "browning episode" during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S2).
+The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant.
+The lowest values of EVI standardized anomalies were recorded in 2005, being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2).
+During 2005 drought, *Q.
+pyrenaica* forests suffered a "browning episode" ($EVI~sa < -1$) (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Appendix S5).
 
@@
 *Q.
-pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3a).
-Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figures 3a).
+pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3).
+Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figure 3) .
 
@@
 ```
-
+Recovery (*Rc*) of greenness was significnatly different bewteen drought events and sites (Tables 3).
 In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
-A similar recovery after the 2005 and 2012 drought event was observed for southern populations (p = 0.2453; Figure 4a), whilst the northern populations showed a significantly greater recovery after the 2005 drought than after the 2012 drought (Figure 4a).
+For southern populations, a similar recovery after the 2005 and 2012 drought event was observed (p = 0.2453; Figure 3; Table S1).
 
@@
 Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables S1, 3), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
-The southern populations showed higher resilience values than the northern ones, although they were not significantly different for 2005 drought (p = 0.036; Figures 3a, 4a).
+The southern populations showed higher resilience values than the northern ones, although these differences were not significant for the 2005 drought event (p = 0.036; Figure 3).
 
@@
 The competition index was similar among sites but plot basal area was greatest in Ca-High (Table 1).
-Yet the basal area increment (BAI) growth in the highest plot in altitude (Ca-High) was the greatest (Figure 5).
+Yet the basal area increment (BAI) growth in the highest plot in altitude (Ca-High) was the greatest (Figure 4).
 Trees from this location showed in addition a positive growth trend since the late 1970s.
-This positive trend was not found in any of the other two locations and the two altitudes sampled in the Southern populations showed differences in growth, being more similar Ca_Low with growth in the northern population (SJ) than with Ca_High (Figure 5).
-`VAS POR AQUI CARPINTO (BREAK)`
+This positive trend was not found in any of the other two locations and the two altitudes sampled in the southern populations showed differences in growth, being more similar Ca-Low with growth in the northern population (SJ) than with Ca-High (Figure 4; Appendix S11).
 
-Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5).
-La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona.
-Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.
+se
 

### 2018-02-21 man/ms_3_discussion.Rmd
updating

@@
 
+### Esto viene del apartado Radial growth trends and climate
+* Aqui me comentó GGI que distribuyera esto bien entre results y discussion 
+Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5).
+La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona.
+Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.
+
 ## References

### 2018-02-21 man/ms_figures.Rmd
updating

@@
 
-```{r, fig.height=7}
+```{r, fig.height=9}
 mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
@@
 aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site,  
+                   color = site, group = site,  
                    fill = site, linetype=site,
@@
   theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables)),
+  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free_y"),
   xlab(''), 
@@
   scale_linetype_manual(values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(values=c(2,2,2.5, 2.5, 2.5))
+  scale_size_manual(values=c(2,2,2.5, 2.5, 2.5)),
+  geom_text(aes(label=Letter, y=postext), 
+            position=pd, 
+            show.legend = FALSE, size = 3) 
+  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
+  # geom_text_repel( 
+  #                 nudge_x = - 0.5,
+  #                 nudge_y = 0.01,
+  #                 direction="both", 
+  #                 size = 3.5, show.legend = FALSE,
+  #                 segment.size = 0.5, 
+  #                 segment.alpha = 0.5) 
 )
+
+m <- m %>% mutate(postext = case_when(
+  vari == "0_EVI" ~
+    case_when(var_sorted == "0_rt" ~ 0.97,
+              var_sorted == "1_rc" ~ 1.19,
+              var_sorted == "2_rs" ~ 1.02),
+  vari == "1_BAI" ~
+     case_when(var_sorted == "0_rt" ~ 1.1,
+              var_sorted == "1_rc" ~ 1.7,
+              var_sorted == "2_rs" ~ 1.2)))
 
@@
   ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") + 
-  geom_text(aes(label=Letter, y = 1.1), position=pd, show.legend = FALSE)
+  ylab("Resistance") 
+# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
 
@@
   ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")+
-  geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
+  ylab("Recovery")
+# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
 
@@
   ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') +
-  geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
+  ylab("Resilience") + xlab('Drought event') 
+# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
 ```

### 2018-02-21 man/ms_figures.Rmd
change site names

@@
     plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(values=c(19, 21, 17, 22, 15)),
-  scale_color_manual(values=c('black', "blue", 'black','blue','blue')),
-  scale_fill_manual(values=c('black', 'white', 'black','white','blue')),
+  scale_shape_manual(
+    values=c("N" = 19, "S" = 21, "CA-High"=22, "CA-Low" = 15, "SJ" = 17)), 
+  scale_color_manual(
+    values=c("N" = 'black',"S" = "blue", "CA-High" = 'blue', "CA-Low" = 'blue', "SJ" = 'black')),
+  scale_fill_manual(
+    values=c("N" = 'black', "S" = "white", "CA-High" = 'white', "CA-Low" = 'blue', "SJ" = 'black')),
   scale_linetype_manual(values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(values=c(2,2,2.5, 2.5, 2.5)),
+  scale_size_manual(
+    values=c("N" = 2, "S" = 2, "CA-High"=2.5, "CA-Low" = 2.5, "SJ" = 2.5)),
   geom_text(aes(label=Letter, y=postext), 
@@
               var_sorted == "2_rs" ~ 1.2)))
+
+m <- m %>% mutate(site = as.factor(
+  case_when(
+  site == "caH" ~ "CA-High",
+  site == "caL" ~ "CA-Low",
+  TRUE ~ site
+)))
 

### 2018-02-22 analysis/notula/evi_resilience_fine_scale.Rmd
add analysis notula

---
title: "EVI-resilience fine-scale"
author: "AJ Perez-Luque (@ajpelu)"
output: 
  md_document:
    variant: markdown_github
---
```{r, echo=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(fig.align='center', message = FALSE, warning = FALSE) 
```

```{r wd, echo=FALSE}
#---------------------------------
machine <- 'ajpelu'
#machine <- 'ajpeluLap'
di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro', sep='')
#---------------------------------
```

```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
library("here")
library('dplyr')
library("sf")
library("mapview")
library('tidyverse')

library('broom')
library('pander')
library('lsmeans')
library('effects')
library('gridExtra')
library('WRS2')
library('car')
library('MASS')
library('rcompanion')
library('devtools')
# devtools::install_github("ajpelu/auxiliar")
```

# Resilience populations

```{r}
# Read data
raw_evires <- read.csv(file=here::here("/data/resilience", "resilience_evi.csv"), header = TRUE, sep = ',')

# add data of pop
anomalias <- read.csv(file=here::here("/data/anomalies", "anomalias_evimean.csv"), header = TRUE, sep = ',')

attr_iv_malla_modis_id <- anomalias %>% dplyr::select(iv_malla_modi_id,long,lat,pop) %>% unique()

raw_evires <- raw_evires %>% inner_join(attr_iv_malla_modis_id, by='iv_malla_modi_id')

# filter by pop and add new variable

# ¿Que poblaciones estamos evaluando? CA and SJ

evires <- raw_evires %>% 
  filter(pop %in% c(2,6)) %>% 
  mutate(
     clu_pop = as.factor(case_when(
      pop == 2 ~ "SJ",
      pop == 6 ~ "CA")),
     clu_pop2 = as.factor(case_when(
      pop == 2 ~ "N",
      pop == 6 ~ "S"))
     )

# Check spatial 
ev <- st_as_sf(x = evires, coords = c("long","lat"), 
               crs = "+proj=longlat +datum=WGS84")
mapview(ev)

# Change name of clu_pop2 and disturb_year para los analisis anovas
evires <- evires %>% dplyr::rename(site = clu_pop, site2 = clu_pop2) %>% 
  mutate(disturb_year = as.factor(disturb_year))
```

# Fine fine scale 
* Ver para la correspondencia para cada pixel de los árboles de muestreo (ver esto https://rawgit.com/ajpelu/qpyr_historic/master/analysis/historic_map.html)

```{r}
fine <- read.csv(file=here::here("/data_raw", "evi_dendro.csv"), header = TRUE, sep = ',')

# number of tree by pixels 
fineid <- fine %>% group_by(iv_malla_modi_id, site) %>% count() %>% as.data.frame()

pixel_selected <- fineid %>% dplyr::select(iv_malla_modi_id) %>% as.vector()

eviresf <- evires %>% filter(iv_malla_modi_id %in% pixel_selected$iv_malla_modi_id)
```

```{r}
# BAI resilience
baires <- read.csv(here::here("data/resilience", "resilience_bai.csv"), header = TRUE, sep=",")
```

# Prepare Data

```{r}
aux_bai <- baires %>% filter(disturb_year != 1995) %>% 
  mutate(vari = '2_BAI') %>% 
  dplyr::select(-tree) %>% 
  mutate(site = case_when(
    site == "caL" ~ "CA-L",
    site == "caH" ~ "CA-H",
    TRUE ~ as.character(site)))

aux_evi <- evires %>% 
  mutate(vari = '0_EVI_pop') %>% 
  dplyr::select(-iv_malla_modi_id, -long, -lat, -pop, -site2)

aux_evifine <- eviresf %>% 
  mutate(vari = '1_EVI_pix') %>%
  dplyr::select(-iv_malla_modi_id, -long, -lat, -pop, -site2)

df <- rbind(aux_bai, aux_evi, aux_evifine)

dfrs <- df %>% 
  dplyr::select(-rt, -rc, -rrs) %>% 
  group_by(vari, disturb_year, site) %>% 
  summarise(mean = mean(rs), 
                   sd = sd(rs),
                   se = sd/sqrt(length(rs)))

```

# Plot data

```{r}
pd <- position_dodge(width = 0.6)

aes_disturb <- aes(x = disturb_year, y = mean, 
                   color = site, group = site,  
                   fill = site, linetype=site,
                   shape = site, 
                   ymin = mean - se, ymax = mean + se)

variables  <- c("0_EVI_pop" = "EVI population scale", "1_EVI_pix" = "EVI fine scale","2_BAI" = "BAI")

rs <- ggplot(dfrs, aes_disturb) + 
  geom_errorbar(width=.1, position=pd) +
  geom_point(position = pd) + 
  facet_wrap(~vari, labeller = labeller(vari = variables)) + 
  theme_bw() + xlab('') + ylab('Resilience') + 
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(
      colour = "black",
      fill = "white"),
    strip.text = element_text(face="bold", size = 14),
    axis.title.y = element_text(face="bold", size = 14),
    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))) +
  scale_shape_manual(
    values=c("SJ" = 19, "CA" = 21, "CA-H"=22, "CA-L" = 15)) + 
  scale_color_manual(
    values=c("SJ" = 'black',"CA" = "blue", "CA-H" = 'blue', "CA-L" = 'blue')) +
  scale_fill_manual(
    values=c("SJ" = 'black', "CA" = "white", "CA-H" = 'white', "CA-L" = 'blue')) +
  scale_linetype_manual(values=c("solid","solid","solid","solid")) +
  scale_size_manual(
    values=c("SJ" = 2, "CA" = 2, "CA-H"=2.5, "CA-L" = 2.5))

grid_arrange_shared_legend(rs, ncol = 1, nrow = 1, position='top')
```

```{r}

# dfcorrela <-

df %>% 
  dplyr::select(-rt, -rc, -rrs) %>% 
  gather(vari, value, -disturb_year, -site) %>% 
  unite(temp, rs, site) %>% 
  spread(temp, value)

```

### 2018-02-22 man/ms_2_results.Rmd
update discussion and results

@@
 
-Likely as a consequence of their different management origin (*i.e.* land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (Ca-High) which had bigger and taller trees than the other sites (Tables 1, 2).
-The competition index was similar among sites but plot basal area was greatest in Ca-High (Table 1).
-Yet the basal area increment (BAI) growth in the highest plot in altitude (Ca-High) was the greatest (Figure 4).
+Likely as a consequence of their different management origin (*i.e.* land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2).
+The competition index was similar among sites but plot basal area was greatest in CA-High (Table 1).
+Yet the basal area increment (BAI) growth in the highest plot in altitude (CA-High) was the greatest (Figure 4).
 Trees from this location showed in addition a positive growth trend since the late 1970s.
-This positive trend was not found in any of the other two locations and the two altitudes sampled in the southern populations showed differences in growth, being more similar Ca-Low with growth in the northern population (SJ) than with Ca-High (Figure 4; Appendix S11).
+This positive trend was not found in any of the other two locations and the two altitudes sampled in the southern populations showed differences in growth, being more similar CA-Low with growth in the northern population (SJ) than with CA-High (Figure 4; Appendix S11).
 
-se
-
-Los tres sitios mostraron valores de ring width indices bajos para el año 1995 (Figure 6).
-El sitio SJ mostró valores muy bajos de ring width indices en 1995, 2005, y 2012.
-Sin embargo esto no ocurrió igual en los sitios del sur (caL y CaH) donde si encontramos valores bajos de RWI para 1995, pero no para 2005 y 2012.
+1995 was the lowest pointer year (*i.e.* the lowest growth) since 1950 in all chronologies (Figure 5).
+For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low).
 
@@
 
-## Resilience of radial tree-growth to drougth events
+## Resilience of radial tree-growth to drought events
 
 Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
-Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figures 3b, 4b).
+Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).
 
@@
 
-Oak forests recovered (*Rc*) their growth more rapidly after 2012 than after 2005 staying at levels of Rc < 1 (Table S2).
-Northern populations showed signigicantly higher values of recovery than southern ones (Tables 3 and S2), which even not reach recovery values of Rc = 1 (Figure 3b, 4b).
+Higher values of recovery (*Rc*) were observed after 2012 than after 2005 drougth events, staying at levels of *Rc* < 1 (Figure 3, Table S2).
+SJ site (northern aspect) showed significantly higher values of recovery than southern sites (CA-High and CA-Low) (Tables 3 and S2), which even not reach recovery values of Rc = 1 (Figure 3).
 
@@
 
-Significantly higher values of resilience of growth were observerd for the 2012 drought event than for the 2005 one (Table S2; Figure 4b).
-We found no differences of resilience values for growth between northern and southern populations (Table 3).
-For 2005 drought event the resilience of northern population was lower than that of southern ones, but opposite pattern was found for the 2012 drought event, with northern population showing values greater than unity (Figure 3b).
-Es intersante que los valores de resiliencia para el tree-growth no alcanzaron el valor 1, excepto para 2005 en la población de SJ (Rs = 1.031)
+Similar pattern of resilience (*Rs*) values was found for growth than for greenness respect to drought event: significantly higher values of *Rs* for the 2012 drought event than for the 2005 (Table S2; Figure 3).
+However, no differences were observed between sites (Table 3).
+For 2005 drought event, *Rs* value of SJ (northern site) was lower than that of southern ones (CA-High and CA-Low), but opposite pattern was found for the 2012 drought event.
+All values of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031).
+
+When we explored resilience values of growth (*Rs*) for other severe droughts, we found a relation between the resilience and drought severity ($r^2$ = 0.336) (Appendix S6).
 
@@
 Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
-Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 7a).
+Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 6a).
 Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
-Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for the northern population (r > 0.6).
+Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 6b), specially for the northern population (r > 0.6).
 Relationships with temperatures were lower than those with moisture variables.
-Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 7c and 7d).
+Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 6c and 6d).
 Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population.
@@
 
-## Disturbances
-The analysis of growth changes revealed differences between sites (Figures 8 and 9).
-En el sitio del norte osbervamos dos eventos de liberación importantes (GC > 50 %), que ocurren en mas de la mitad de los árboles muestreados (Figures 8 and 9).
-El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
-Estos periodos se alternan con periodos de supresión.
-En los sitios del sur (caH y caL), sin embargo se observan algunos episodios débiles de supresión, aunque estos no se observan para los últimos años, de hecho incluso se observa una débil liberación (Figure 8) afectando a mas de la mitad de los árboles muestreados (Figure 9).
+## Forest disturbances
+The analysis of growth changes revealed differences forest history between sites (Figures 7 and 8).
+Northern site (SJ) showed two release events ($GC$ > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001.
+These periods alternate with periods of supression.
+Southern sites (CA-High and CA-Low) showed some weak suppression episodes, but not in the last 50 years.
 

### 2018-02-22 man/ms_3_discussion.Rmd
update discussion and results

@@
 
-# Algunas ideas para la discussion
+## Oaks show high resilience in response to recent drought events
 
-## FORTALEZA METODOLÓGICA
+Hemos observado como las sequías severas, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento de *Q.
+pyrenaica*.
+Los datos de anomalías estandarizadas de EVI mostraron un browning para la mayoría de los robledales de Sierra Nevada durante la sequía de 2005.
+
+El efecto de la sequía de 2005 en el crecimiento también se ha visto en otras especies en el sur de la P. Ibérica 
+La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
+Esto se ha observado también en otras especies en el sur de la P.
+Ibérica (p.ej.
+P.
+nigra en Andalusia [@SanchezSalguero2013], ...
+INCLUIR otras especies y citas).
+
+* P.
+nigra: Se observó una gran reducción de crecimiento en 1994-1995, 1999 y 2005 para Pinus nigra en el SE de Andalusia [@SanchezSalguero2013] coincidiendo con años de sequía para esa zona.
+La máxima reducción en crecimiento fue observada para 2005, siendo significativamente mayor que para el resto de años [@SanchezSalguero2013].
+
+No obstante, a pesar de que 2005 y 2012 fueron dos de las sequías mas severas registradas para el sur de la P.
+Ibérica [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), nuestros resultados muestran que los robledales de *Q.
+pyrenaica* en su rear-edge presentan una tendencia positiva tanto en el greenness (78.9 % pixeles con greeness) como en el crecimiento (tendencias positivas del BAI).
+
+### La resiliencia de las poblaciones de roble de Sierra Nevada.
+
+El greennness de los robledades en Sierra Nevada ha mostrado una tendencia positiva hacia mas verdor en los ultimos años, que coincide con lo que ya observamos con datos de NDVI [@PerezLuque2015onto], sobre todo para las poblaciones del sur.
+Esta tendencia positiva, obtenida de variables derivadas de remote sensing (EVI y NDVI), parece que también se observa en el crecimiento.
+Por ejemplo en las poblaciones del sur (CaHigh y CaLow) observamos en los últimos años una ligera tendencia de crecimiento.
+
+Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia.
+Por ejemplo para las poblaciones del norte, los valores de EVI durante la sequía de 2005 descendieron hasta el 81 %, mientras que el BAI (sitio SJ) lo hizo hasta el 45 %.
+Sin embargo, la recuperación fue rápida, así los valores de Recovery (Rc) para el EVI tras la sequía de 2005 (en las poblaciones del norte) fue de 1.17, mientras que para el BAI el Rc fue de 1.112.
+En definitiva, estamos observando altos valores de resiliencia en estos robledales.
+
+Algunas notas sobre la sequía de 2005: 
+* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].
+
+Los valores de resiliencia observados para el greenness (cercanos a 1) muestran como esta especie, en su rear edge muestra una buena capacidad de recuperación frente a eventos extremos,
+
+¿A que se debe esa resiliencia?
+
+¿Casos parecidos?
+
+## Site environment shapes differential sensitivity to climate and drought of rear-edge oak populations
+
+Esta tendencia es importante sobre todo en las poblaciones situadas mas al sur, en donde observamos: menor br
+Encontramos diferencias entre poblaciones. Asi por ejemplo en las situadas mas al sur, vemos que las anomalias de EVI son menos negativas para el 2005; También vemos tendencias positivas en el BAI...
+
+## metodologico
+
+La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía se pueden ver tanto con información de satélite (hablar del browning), como con información mas de campo (hablar aquí de lo del BAI).
+Indicar también las diferencias entre las resiliencias detectadas por ambas metodologías.
+
+### Notas sobre metodologia
 
@@
 ```
-
-### Efectos de la sequía en el greenneess y en el crecimiento.
-
-Hemos observado como las sequía, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento.
-Los datos de anomalías estandarizadas de EVI mostraron un browning para la mayoría de los robledales de Sierra Nevada durante la sequía de 2005.
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-* Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI.
-En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor
-")
-```
-
-La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
-Esto se ha observado también en otras especies en el sur de la P.
-Ibérica (p.ej.
-P.
-nigra en Andalusia [@SanchezSalguero2013], ...
-INCLUIR otras especies y citas).
-
-Algunas notas sobre la sequía de 2005:
-
-* 2004/2005 hydrological year is considered one of the worst drooughts ever recorded in the Iberian Peninsula, particularly in the central and southern sectors [@GarciaHerrera2007].
-* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].
-* The hydrological year from October 2004 to September 2005 was the driest on record at several locations throughout Iberia [@GarciaHerrera2007].
-* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-Se observó una gran reducción de crecimiento en 1994-1995, 1999 y 2005 para Pinus nigra en el SE de Andalusia [@SanchezSalguero2013] coincidiendo con años de sequía para esa zona.
-La máxima reducción en crecimiento fue observada para 2005, siendo significativamente mayor que para el resto de años [@SanchezSalguero2013].
-")
-```
-
-### La resiliencia de las poblaciones de roble de Sierra Nevada.
-
-El greennness de los robledades en Sierra Nevada ha mostrado una tendencia positiva hacia mas verdor en los ultimos años, que coincide con lo que ya observamos con datos de NDVI [@PerezLuque2015onto], sobre todo para las poblaciones del sur.
-Esta tendencia positiva, obtenida de variables derivadas de remote sensing (EVI y NDVI), parece que también se observa en el crecimiento.
-Por ejemplo en las poblaciones del sur (CaHigh y CaLow) observamos en los últimos años una ligera tendencia de crecimiento.
-
-Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia.
-Por ejemplo para las poblaciones del norte, los valores de EVI durante la sequía de 2005 descendieron hasta el 81 %, mientras que el BAI (sitio SJ) lo hizo hasta el 45 %.
-Sin embargo, la recuperación fue rápida, así los valores de Recovery (Rc) para el EVI tras la sequía de 2005 (en las poblaciones del norte) fue de 1.17, mientras que para el BAI el Rc fue de 1.112.
-En definitiva, estamos observando altos valores de resiliencia en estos robledales.
 
@@
 
+---- old 
+### Efectos de la sequía en el greenneess y en el crecimiento.
+
+Hemos observado como las sequía, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento.
+Los datos de anomalías estandarizadas de EVI mostraron un browning para la mayoría de los robledales de Sierra Nevada durante la sequía de 2005.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+* Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI.
+En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor
+")
+```
+
+La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
+Esto se ha observado también en otras especies en el sur de la P.
+Ibérica (p.ej.
+P.
+nigra en Andalusia [@SanchezSalguero2013], ...
+INCLUIR otras especies y citas).
+
+Algunas notas sobre la sequía de 2005:
+
+* 2004/2005 hydrological year is considered one of the worst drooughts ever recorded in the Iberian Peninsula, particularly in the central and southern sectors [@GarciaHerrera2007].
+* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].
+* The hydrological year from October 2004 to September 2005 was the driest on record at several locations throughout Iberia [@GarciaHerrera2007].
+* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
+
 ## References

### 2018-02-26 analysis/notula/evi_resilience_fine_scale.Rmd
add plot

@@
 library('tidyverse')
-
-library('broom')
-library('pander')
-library('lsmeans')
-library('effects')
-library('gridExtra')
-library('WRS2')
-library('car')
-library('MASS')
-library('rcompanion')
-library('devtools')
-# devtools::install_github("ajpelu/auxiliar")
+library('lemon')
 ```
@@
 
+pdf(here::here('/analysis/notula/out', 'evi_vs_bai.pdf'), width=10, height = 6)
 grid_arrange_shared_legend(rs, ncol = 1, nrow = 1, position='top')
+dev.off()
+
 ```
@@
 ```{r}
+## VAS por aqui CARPINTO #############
 

### 2018-02-26 man/ms_0_intro.Rmd
update
update intro

@@
 
-* Analysing the impact of drought on tree growth is therefore a priority under the current climatic conditions, in which global warming is probably increasing the severity of droughts (Williams et al., 2013).
+Flujo:
 
-En este estudio combinamos el uso de datos de dendrocronologia con información de productividad procedente de satélite.
-Por un lado, sabemos que el annual radial growth increment se utiliza como proxy para estudiar la forest vitality [@Fritts1976; @Dobbertin2005]; por otro lado el EVI para productividad ...
+# 1) Sequías y Land use change
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-* Tree ring width or annual radial growth increment is a widely used proxy for tree vitality (ver  [@Bhuyan2017] [@Dobbertin2005]
+- importancia de analizar las sequías ya que estas están aumentando
 
-* Ver en Gea-Izquierdo 2014: El análisis del inter-annual growth variability can be used to study changes in productiviyt and forest vitality to detect vulnerability to stress (Fritts 1976; Doberti)
+- Esto cobra mayor enfasis en zonas mediterráneas, donde además de la existencia de sequías, hay una gran historia de impacto antrópico (land-use change driver, el otro driver del cambio global).
+Rececientemente Doblas-Miranda et al.
+han realizado una revisión donde han visto la importancia de la sequías, no solo como perturbación aislada, sino también como sus interacciones, siendo el land-use change el otra gran factor (motor) importante a tener en cuenta.
+Por tanto,
 
-")
-```
+En una revision [@DoblasMiranda2017] subraya la relevancia y el impacto de las interacciones entre los emerging global change factors in the Mediterranean area.
+De entre las interacciones revisadas, la sequía es un factor crucial ...
 
-* Diferentes estudios sugieren que, a consecuencia del cambio climático, se incrementará la severidad, la extensión geográfica y la frecuencia de las sequías [@Dai2011; @Trenberth2014]
+past land uses could be a key factor altering the effects of current global changes and thus differentiating the Basin from other Mediterranean regions of the world
 
-* Proyeccciones de crecimeinto de para *Q.
-pyrenaica* sugieren un declive en el crecimiento en las proximas decadas a lo largo de su rango de distribución en la P.
-Iberica, en diferentes escenarios climáticos [@GeaIzquierdo2013]: Predictions suggest that QUPY productivity would decline in the next decades all along its distributional range in the Iberian Peninsula for all the climate scenarios studied.
-Este declive debería ser mas dramático en las baja altitudes de las zonas mas calientes del sur de su distribución
-
---> vulnerabilidad local (related to rear-edge)
-La tendencia de reducción drástica sugerida por el modelo para el dry-edge de la distribución de la especie podría expresar una aumentada vulnerabilidad de los árboles al incremento del estrés hídrico forzado por un clima mas cálido [@GeaIzquierdo2013].
-
-Negative trends over recent years and the greater response to moisture availability found at warmer and drier low elevations in the south suggest vulnerability to warming at the local low elevation dry edge of the species's range (e.g.
-QUPY9).
-Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()
-
-...
-Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
-
--- Sequías --
-
-Drought is a natural phenomenon that occurs when water availability is significantly below-normal levels [@Dai2011].
-Drought occurs over most part of the world and affects ecological systems across every climatic zone worldwide [@Mishra2010], being one of the most economically and ecologically disruptive extreme events affecting millions of people globally [@Dai2011].
+Drought is a natural phenomenon occurs when water availability is significantly below-normal levels [@Dai2011].
+Drought is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Ummenhofer2017].
 Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
@@
 
-In the Iberian Peninsula, major drought episodies were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
-The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Pascoa2017].
-These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
+Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forest through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].
 
-Drought can cause significant changes in ecosystem productivity and water dynamics, and extreme drought can severely affect forest through changes in plant physiology, phenology and carbon allocation (see @Ummenhofer2017 for a revision).
+El estudio de como las sequías afectan a la vegetación es importante, sobre todo teneindo en cuenta que existen algunos estudios que sugieren un incremento de la severidad, extensión geográfica y frecuencia de las sequía a consecuencia del cambio climático [@Dai2011; @Trenberth2014].
+Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg.
+@Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al.
+https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI).
+Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).
 
-La vulnerabilidad de los bosques a los eventos climáticos extremos ...
-
-no obstante los ecosistemas situados en zonas áridas responden mas rápidamente a la sequía.. ver VicenteSerrano2013
-
-Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
-(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...
-
+Junto con las sequías el cambio en el uso del suelo es otro de los grandes motores del cambio global.. 
 ...
@@
 
+In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017].
+The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017].
+These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
+
+# 2) Combinación metodológica 
+Analizar resiliencia de poblaciones utilizando RS y dendro.
+
+Analizar el impacto de las sequías en los árboles es muy importante y prioriatrio...
+
+* Analysing the impact of drought on tree growth is therefore a priority under the current climatic conditions, in which global warming is probably increasing the severity of droughts (Williams et al., 2013).
+
+En este estudio combinamos el uso de datos de dendrocronologia con información de productividad procedente de satélite.
+Por un lado, sabemos que el annual radial growth increment se utiliza como proxy para estudiar la forest vitality [@Fritts1976; @Dobbertin2005]; por otro lado el EVI para productividad ...
+
+Una vez dicho esto...
+
+a) Como la dendro ayuda a evaluar las sequías (ver Gazol 2017)
+
+* Gea-Izquierdo 2014: El análisis del inter-annual growth variability can be used to study changes in productiviyt and forest vitality to detect vulnerability to stress (Fritts 1976; Doberti)
+
+* Tree ring width or annual radial growth increment is a widely used proxy for tree vitality (ver  [@Bhuyan2017] [@Dobbertin2005]
+
+b) Remote sensing:
+
+Remote sensing can be used for studying droughts impacts on ecosystems (*e.g.* @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016).
+Several studies have analyzed drought impacts on vegetation using remote sensing at different scales [@VicenteSerrano2013; @Panisset2018], and for Mediterranean area there were studies focused on local scales (*e.g* @Lloret2007; @Gouveia2012, @Zribi2016) and recently other covering all Mediterranean region [@Gouveia2017].
+
+sin embargo (aqui poner las limitaciones de una u otra metodología).
+
+Jump et al habla de utilizar combinaciones de ambos métodos para estudiar esto, y esto es importante para las poblaciones del rear edge.
+
+Existen algunos estudios que han combinado el uso de dendro y RS para estudiar el efecto de las sequías....
+
+* @Dorman2013 --> interesante compara NDVI y dendro 
+* Tree growth and vegetation activity at the ecosystem-scale in the eastern Mediterranean
+
+Aqui nosotros evaluamos el efecto de la sequía ... The vegetation response to water deficit was assessed using remote sensing information and dendrocronologial data.
+
+... ojo los remote sensing datos (aun con alta resolucion temporal y/o espacial) son insuficientes por si solos para demostrar 
+ Remote sensing data, even high temporal or spatial resolution data, are insufficient alone to unequivocally demonstrate that drought is the causal agent of a particular change in reflectance values
+
+-----------------------
+
+3) rear edge
+
+- Todo ello cobra especial importancia cuando las poblaciones de estudio están en su rear edge ...
+Por otro lado, es importante analizar las sequías en las poblaciones localizadas en el rear edge para ver su respuesta.
+(Jump et al)
+
+- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):
+
+* Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
+(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...
+
+...
+Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
+
 ## 
@@
 cat("
-
-* The vegetation response to water deficit was assessed using remote sensing information and dendrocronologial data.
 
@@
 
--- Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
-(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar)
-
-Una vez dicho esto...
-
-* Como la dendro ayuda a evaluar las sequías (ver Gazol 2017) 
-* Remote sensing:
-
-    * Uso de Remote sensing para estudiar la sequía --> Leer @Zhang2013 y escribir algo.
-Drought monitoring using remote-sensing approach was originally applied to agriculture.
-Several remote-sensing derived indices have been used to study the drought effects on vegetation ...
-    * Leer también a @AghaKouchak2015
-
-### 
-Uso de NDVI como estimador de la NPP:
-
-* The NDVI properties have allowed the use of this information for estimating the Net Primary Production (NPP) (Goward and Dye, 1987; Running et al., 2004; Hasenauer et al., 2012).
-Different studies have already found a strong relationship between NPP and radial growth (e.g., Granier et al., 2008; Babst et al., 2013, 2014a, 2014b; Vicente-Serrano et al., 2015), albeit with significant differences, particularly those related to species, sites and environmental conditions.
-
-* Ver Gilaber et al. 2017 http://www.mdpi.com/2072-4292/9/3/193
-
-`$NOTA$`: NDVI sirve para estimar la producción primaria neta. Existen diferentes estudio que han evaluado el efecto de la sequía sobre la producción primaria neta utilizando NDVI. 
 ")
@@
 .. ver [@VicenteSerrano2013] impacto de las sequías en diferentes zonas (i.e montañas)
-
-Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @AghaKouchak2015; @Norman2016]
-
-... ojo los remote sensing datos (aun con alta resolucion temporal y/o espacial) son insuficientes por si solos para demostrar 
- Remote sensing data, even high temporal or spatial resolution data, are insufficient alone to unequivocally demonstrate that drought is the causal agent of a particular change in reflectance values
 
@@
 * and to explore differences in the resilience metrics between populations located in contrasting slopes within the rear edge of the distribution of this species.Do northern and southern populations of Pyrenean oak forests differ in their resistance, resilience or recovery to extreme drought events?
+
+# Otras notas: 
+* To guide sustainable forest management, forest researchers are asked to provide concrete answers about forest resilience in response to expected climatic trends, and extreme climatic events (Lindner et al., 2014) https://www.sciencedirect.com/science/article/pii/S030147971400379X?via%3Dihub
+
+* Hablan de incorporar ademas de RS, el conocimiento ecológico (los principios ecológicos) para comprender las consecuencias de las disturbances analizadas con RS, y sobre todo para predecir futuro.
+(McDowell et al https://www.sciencedirect.com/science/article/pii/S1360138514002726)
+
+----------------------------
+
+* Proyeccciones de crecimeinto de para *Q.
+pyrenaica* sugieren un declive en el crecimiento en las proximas decadas a lo largo de su rango de distribución en la P.
+Iberica, en diferentes escenarios climáticos [@GeaIzquierdo2013]: Predictions suggest that QUPY productivity would decline in the next decades all along its distributional range in the Iberian Peninsula for all the climate scenarios studied.
+Este declive debería ser mas dramático en las baja altitudes de las zonas mas calientes del sur de su distribución
+
+--> vulnerabilidad local (related to rear-edge)
+La tendencia de reducción drástica sugerida por el modelo para el dry-edge de la distribución de la especie podría expresar una aumentada vulnerabilidad de los árboles al incremento del estrés hídrico forzado por un clima mas cálido [@GeaIzquierdo2013].
+
+Negative trends over recent years and the greater response to moisture availability found at warmer and drier low elevations in the south suggest vulnerability to warming at the local low elevation dry edge of the species's range (e.g.
+QUPY9).
+Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()

### 2018-02-27 man/ms_0_intro.Rmd
update intro
update texts

@@
 
-# Intro
-
-Flujo:
+# Introduction
 
@@
 
-- importancia de analizar las sequías ya que estas están aumentando
+Drought is a natural phenomenon occurs when water availability is significantly below-normal levels [@Dai2011].
+It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017].
+Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forest through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].
 
-- Esto cobra mayor enfasis en zonas mediterráneas, donde además de la existencia de sequías, hay una gran historia de impacto antrópico (land-use change driver, el otro driver del cambio global).
-Rececientemente Doblas-Miranda et al.
-han realizado una revisión donde han visto la importancia de la sequías, no solo como perturbación aislada, sino también como sus interacciones, siendo el land-use change el otra gran factor (motor) importante a tener en cuenta.
-Por tanto,
+In the last decades there is an crescent interest in assessing the drought impacts on ecosystem, especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014].
+There are studies analyzing the response of vegetation to drought from global (*e.g*.
+@Allen2010; @VicenteSerrano2013; @Norman2016) to local scales (*e.g* @MartinezVilalta2016 for a revision).
+Some studies have focusing on the drought effects on Mediterranean species which representing their southernmost limit [@Pasho2011; @Camarero2011; @Herrero2014; @Linares2014; @SanchezSalguero2017; @Caminero2018].
 
-En una revision [@DoblasMiranda2017] subraya la relevancia y el impacto de las interacciones entre los emerging global change factors in the Mediterranean area.
-De entre las interacciones revisadas, la sequía es un factor crucial ...
-
-past land uses could be a key factor altering the effects of current global changes and thus differentiating the Basin from other Mediterranean regions of the world
-
-Drought is a natural phenomenon occurs when water availability is significantly below-normal levels [@Dai2011].
-Drought is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Ummenhofer2017].
 Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
 In addition, climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013] particularly for the Mediterranean area [@Spinoni2017a].
-
-Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forest through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].
-
-El estudio de como las sequías afectan a la vegetación es importante, sobre todo teneindo en cuenta que existen algunos estudios que sugieren un incremento de la severidad, extensión geográfica y frecuencia de las sequía a consecuencia del cambio climático [@Dai2011; @Trenberth2014].
-Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg.
-@Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al.
-https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI).
-Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).
-
-Junto con las sequías el cambio en el uso del suelo es otro de los grandes motores del cambio global.. 
-...
-mejorar--> y esto puede ser especialmente importante en zonas como la región mediterránea,con un alto uso antrópico, donde la sequía presenta no solo efectos aislados, sino también la interacción con otros factores [@DoblasMiranda2017].
-(ver abstract trabajo)
-
-No solo efectos aislados, sino también en conjunción con otros factores, sobre todo en el mediterráneo donde convergen muchos de los factores que pueden interaccionar:
-
 In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017].
@@
 These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
+
+Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017].
+The impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017].
+This is specially relevant for areas with a long history of landscape modification such us Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016].
+In a recent review, @DoblasMiranda2017 found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.
+
+The impacts of drought are especially relevant for populations located in the retreating or rear edge of their distribution where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005].
+The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010].
+It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].
+
+There are several works que han analizado la respuesta de la vegetación a la sequía en su rear edge, y en este contexto es interesante analizar no solo el efecto de la sequía en estas poblaciones sino también su resiliencia.
+-- Resilience is the capacity of an ecosystem, community or individual to restore pre-disturbance structure and function after a disturbance (analogous to ‘engineering resilience’, see Holling, 1996; see also Lloret et al., 2011).
+Thus, the assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services.
+Furthermore, the analysis of resilience using multiple demographic rates allows us to detect key vulnerabilities to drought in tree population dynamics, such as reductions in recruitment or survival, which can prompt a decline in dominant species shaping forest structure.
+
+Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling..
+// dorado-liñan // sanchez salguero
+
+trabajos rear edge: 
+* sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
+* sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
+* dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
+* herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
+* dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
+buscar otros
+
+Sin embargo el estudio de las poblaciones situadas en su rear edge requiere de una aproximación combinada donde se estudie la respuesta de las poblaciones al cambio climático y su resiliencia utiliznado diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].
+
+Remote sensing
+
+These climatic alterations are likely to have important consequences for tree species dynamics at local and regional scales ( Peñuelas and Boada 2003 , Van Mantgem et al.
+2009 , Matías and Jump 2015 ).
 
@@
 
-3) rear edge
+### ojo un parrafito de resiliencia
 
-- Todo ello cobra especial importancia cuando las poblaciones de estudio están en su rear edge ...
-Por otro lado, es importante analizar las sequías en las poblaciones localizadas en el rear edge para ver su respuesta.
-(Jump et al)
+Shifts in forest structure and composition in response to drought could alter ecosystem functions profoundly, affecting hydrological processes (Allen, 2007; Adams et al., 2012), nutrient cycling (Hughes et al., 2006; Xiong et al., 2011) and land-surface properties (Royer et al., 2011; Anderegg et al., 2013).
 
-- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):
-
-* Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
-(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...
-
-...
-Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
+In this context, resilience is an important concept that is gaining unprecedented attention in ecology (e.g.
+Reyer et al., 2015).
+Resilience is the capacity of an ecosystem, community or individual to restore pre-disturbance structure and function after a disturbance (analogous to ‘engineering resilience’, see Holling, 1996; see also Lloret et al., 2011).
+Thus, the assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services.
+Furthermore, the analysis of resilience using multiple demographic rates allows us to detect key vulnerabilities to drought in tree population dynamics, such as reductions in recruitment or survival, which can prompt a decline in dominant species shaping forest structure.
 
@@
 Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()
+
+Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg.
+@Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al.
+https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI).
+Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).
+
+Case studies that focus on multiple scales - including local scales - and also valuables (Dale et al.
+2018 Frontiers) --> Se necesitan casos de estudio que se enfoquen en escalas multiples (incluyendo las locales) y que combinen varias metodologías (Jump et al.
+...)
+
+3) rear edge
+
+- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):
+
+* Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
+(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...
+
+...
+Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...

### 2018-02-27 man/ms_3_discussion.Rmd
update texts

@@
 ### OTRA COSA
+
+In fact, several authors have raise the need to consider land use y drought de forma conjunta en las environmental y forest management .... (ver Peñuelas et al 2017 y tb Doblas Miranda )
+
+Environmental and forest management policies should take into account all these characteristics
+of Mediterranean forests and the social, environmental and climatic conditions that are projected
+for the coming years and decades.
+
 #### Climate and tree relationship 

### 2018-02-27 man/ms_0_intro.Rmd
update

@@
 Remote sensing can be used for studying droughts impacts on ecosystems (*e.g.* @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016).
-Several studies have analyzed drought impacts on vegetation using remote sensing at different scales [@VicenteSerrano2013; @Panisset2018], and for Mediterranean area there were studies focused on local scales (*e.g* @Lloret2007; @Gouveia2012, @Zribi2016) and recently other covering all Mediterranean region [@Gouveia2017].
+Several studies have analyzed drought impacts on vegetation using remote sensing at different scales [@VicenteSerrano2013; @Panisset2018; @Assal2016], and for Mediterranean area there were studies focused on local scales (*e.g* @Lloret2007; @Gouveia2012, @Zribi2016) and recently other covering all Mediterranean region [@Gouveia2017].
 

### 2018-03-13 man/ms_0_intro.Rmd
update intro

--- 2018-02-27 man/ms_0_intro.Rmd
@@
 
-# 1) Sequías y Land use change
+Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011].
+It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017].
+Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
+Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013]  @Spinoni2017a], specially in the Mediterranean area [@Spinoni2017a].
 
-Drought is a natural phenomenon occurs when water availability is significantly below-normal levels [@Dai2011].
-It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017].
-Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forest through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].
+Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].
+In the last decades the assessment of the drought impacts on ecosystems has gained attention [@Allen2010; @Clark2016], especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014].
+The response of vegetation to drought has been analyzed at different scales, from global [*e.g*.
+@Allen2010; @VicenteSerrano2013; @Norman2016] to local scales [see @MartinezVilalta2016 for a revision].
 
-In the last decades there is an crescent interest in assessing the drought impacts on ecosystem, especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014].
-There are studies analyzing the response of vegetation to drought from global (*e.g*.
-@Allen2010; @VicenteSerrano2013; @Norman2016) to local scales (*e.g* @MartinezVilalta2016 for a revision).
-Some studies have focusing on the drought effects on Mediterranean species which representing their southernmost limit [@Pasho2011; @Camarero2011; @Herrero2014; @Linares2014; @SanchezSalguero2017; @Caminero2018].
+Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017], and some authors states that the impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017].
+This is specially relevant for areas with a long history of landscape modification as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016].
+For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.
 
-Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
-In addition, climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013] particularly for the Mediterranean area [@Spinoni2017a].
-In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017].
-The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017].
-These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
-
-Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017].
-The impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017].
-This is specially relevant for areas with a long history of landscape modification such us Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016].
-In a recent review, @DoblasMiranda2017 found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.
-
-The impacts of drought are especially relevant for populations located in the retreating or rear edge of their distribution where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005].
+The impacts of drought are also especially relevant for populations located in the rear edge of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005].
 The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010].
@@
 
-There are several works que han analizado la respuesta de la vegetación a la sequía en su rear edge, y en este contexto es interesante analizar no solo el efecto de la sequía en estas poblaciones sino también su resiliencia.
--- Resilience is the capacity of an ecosystem, community or individual to restore pre-disturbance structure and function after a disturbance (analogous to ‘engineering resilience’, see Holling, 1996; see also Lloret et al., 2011).
-Thus, the assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services.
-Furthermore, the analysis of resilience using multiple demographic rates allows us to detect key vulnerabilities to drought in tree population dynamics, such as reductions in recruitment or survival, which can prompt a decline in dominant species shaping forest structure.
+Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [*e.g*.
+@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018; ...], however only a few works have focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
+Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance (@Holling1973; @Lloret2011; @Hodgson2015).
+The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services.
+This is particularly relevant for populations located in their rear-edge of their distribution, especially when they are located in Mediterranean mountains.
+
+In this work we combined remote sensing information and dendroecological methods to evaluate the drought impacts in canopy greenness (as a proxy to primary growth) and radial growth of a Mediterranean tree species (*Quercus pyrenaica*) located in the rear-edge of their distribution.
+We also assessed the resilience of these forest to severe drought events.
+
+Sabemos que la sequía afecta al crecimiento y a la productividad (poner ejemplos de ambos trabajos) ..
+
+RS 
+Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
+Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Panisset2018; @Assal2016].
+For Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [*e.g* @Lloret2007; @Gouveia2012, @Zribi2016].
+
+hablar algo de las ventajas de la RS ...
+
+Jump et al habla de utilizar combinaciones de ambos métodos para estudiar esto, y esto es importante para las poblaciones del rear edge.
+
+Existen algunos estudios que han combinado el uso de dendro y RS para estudiar el efecto de las sequías.... 
+* @Dorman2013 --> interesante compara NDVI y dendro 
+* Tree growth and vegetation activity at the ecosystem-scale in the eastern Mediterranean
+
+e incluso otros trabajos han analizado RS y dendro para evaluar resiliencia, pero han utilizado datos de amplia escala (Gazol et al 2018) y no se han centrado en rear edge.
+
+Por otro lado
+
+Dendro
 
@@
 
+In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017].
+The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017].
+These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
+
 # 2) Combinación metodológica 
@@
 
-b) Remote sensing:
-
-Remote sensing can be used for studying droughts impacts on ecosystems (*e.g.* @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016).
-Several studies have analyzed drought impacts on vegetation using remote sensing at different scales [@VicenteSerrano2013; @Panisset2018], and for Mediterranean area there were studies focused on local scales (*e.g* @Lloret2007; @Gouveia2012, @Zribi2016) and recently other covering all Mediterranean region [@Gouveia2017].
-
-sin embargo (aqui poner las limitaciones de una u otra metodología).
-
-Jump et al habla de utilizar combinaciones de ambos métodos para estudiar esto, y esto es importante para las poblaciones del rear edge.
-
-Existen algunos estudios que han combinado el uso de dendro y RS para estudiar el efecto de las sequías....
-
-* @Dorman2013 --> interesante compara NDVI y dendro 
-* Tree growth and vegetation activity at the ecosystem-scale in the eastern Mediterranean
-
-Aqui nosotros evaluamos el efecto de la sequía ... The vegetation response to water deficit was assessed using remote sensing information and dendrocronologial data.
-
-... ojo los remote sensing datos (aun con alta resolucion temporal y/o espacial) son insuficientes por si solos para demostrar 
- Remote sensing data, even high temporal or spatial resolution data, are insufficient alone to unequivocally demonstrate that drought is the causal agent of a particular change in reflectance values
-
 -----------------------
-
-### ojo un parrafito de resiliencia
-
-Shifts in forest structure and composition in response to drought could alter ecosystem functions profoundly, affecting hydrological processes (Allen, 2007; Adams et al., 2012), nutrient cycling (Hughes et al., 2006; Xiong et al., 2011) and land-surface properties (Royer et al., 2011; Anderegg et al., 2013).
-
-In this context, resilience is an important concept that is gaining unprecedented attention in ecology (e.g.
-Reyer et al., 2015).
-Resilience is the capacity of an ecosystem, community or individual to restore pre-disturbance structure and function after a disturbance (analogous to ‘engineering resilience’, see Holling, 1996; see also Lloret et al., 2011).
-Thus, the assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services.
-Furthermore, the analysis of resilience using multiple demographic rates allows us to detect key vulnerabilities to drought in tree population dynamics, such as reductions in recruitment or survival, which can prompt a decline in dominant species shaping forest structure.
 
@@
 Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
+
+--> plantear la hipótesis... efecto de sequias recurrentes ...
+
+y luego los objetivos

### 2018-03-14 man/ms_0_intro.Rmd
update

@@
 Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
-Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013]  @Spinoni2017a], specially in the Mediterranean area [@Spinoni2017a].
+Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013], specially in the Mediterranean area [@Spinoni2017a].
 
@@
 
-Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [*e.g*.
-@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018; ...], however only a few works have focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
-Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance (@Holling1973; @Lloret2011; @Hodgson2015).
+Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], however only a few works have focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
+Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015].
 The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services.
@@
 
-In this work we combined remote sensing information and dendroecological methods to evaluate the drought impacts in canopy greenness (as a proxy to primary growth) and radial growth of a Mediterranean tree species (*Quercus pyrenaica*) located in the rear-edge of their distribution.
+In this work we combined remote sensing information and dendroecological methods to evaluate the drought impacts in canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica*) located in the rear-edge of their distribution.
+
+Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
+Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016], and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Gouveia2012; @Zribi2016].
+Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an idicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016].
+
+Tree-ring width is a widely used proxy for tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Gazol2018].
+
+Several studies han combindado
+
+Several studies have combined the use of satellite information with dendrochronological information to study the effect of droughts on vegetation
+
+Existen algunos estudios que han combinado el uso de dendro y RS para estudiar el efecto de las sequías.... 
+* @Dorman2013 --> interesante compara NDVI y dendro 
+* Tree growth and vegetation activity at the ecosystem-scale in the eastern Mediterranean
+
+e incluso otros trabajos han analizado RS y dendro para evaluar resiliencia, pero han utilizado datos de amplia escala (Gazol et al 2018) y no se han centrado en rear edge.
+
 We also assessed the resilience of these forest to severe drought events.
-
-Sabemos que la sequía afecta al crecimiento y a la productividad (poner ejemplos de ambos trabajos) ..
-
-RS 
-Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
-Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Panisset2018; @Assal2016].
-For Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [*e.g* @Lloret2007; @Gouveia2012, @Zribi2016].
-
-hablar algo de las ventajas de la RS ...
 

### 2018-03-14 man/ms_0_intro.Rmd
update

@@
 It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].
-
 Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], however only a few works have focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
@@
 
-In this work we combined remote sensing information and dendroecological methods to evaluate the drought impacts in canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica*) located in the rear-edge of their distribution.
-
 Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
-Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016], and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Gouveia2012; @Zribi2016].
-Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an idicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016].
+Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016].
+Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Gouveia2012; @Zribi2016].
 
 Tree-ring width is a widely used proxy for tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Gazol2018].
+Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.
 
-Several studies han combindado
+In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica*) located in the rear-edge of their distribution.
 
-Several studies have combined the use of satellite information with dendrochronological information to study the effect of droughts on vegetation
+Relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge can be particularly vulnerable to climate change.
+In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and radial growth of a Mediterranean tree species (*Quercus pyrenaica*) forests in Sierra Nevada, which represent a rear-edge of their distribution.
+Wealso evaluated the resilience of these forest to several extreme drought episodes.
+Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q.
+pyreancia* forests in their rear edge; (*ii*) to analyze the resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species.
+In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
 
-Existen algunos estudios que han combinado el uso de dendro y RS para estudiar el efecto de las sequías.... 
-* @Dorman2013 --> interesante compara NDVI y dendro 
-* Tree growth and vegetation activity at the ecosystem-scale in the eastern Mediterranean
-
-e incluso otros trabajos han analizado RS y dendro para evaluar resiliencia, pero han utilizado datos de amplia escala (Gazol et al 2018) y no se han centrado en rear edge.
-
-We also assessed the resilience of these forest to severe drought events.
-
-Jump et al habla de utilizar combinaciones de ambos métodos para estudiar esto, y esto es importante para las poblaciones del rear edge.
-
-Existen algunos estudios que han combinado el uso de dendro y RS para estudiar el efecto de las sequías.... 
-* @Dorman2013 --> interesante compara NDVI y dendro 
-* Tree growth and vegetation activity at the ecosystem-scale in the eastern Mediterranean
-
-e incluso otros trabajos han analizado RS y dendro para evaluar resiliencia, pero han utilizado datos de amplia escala (Gazol et al 2018) y no se han centrado en rear edge.
-
-Por otro lado
-
-Dendro
-
-Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling..
-// dorado-liñan // sanchez salguero
-
-trabajos rear edge: 
-* sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
-* sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
-* dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
-* herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
-* dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
-buscar otros
-
-Sin embargo el estudio de las poblaciones situadas en su rear edge requiere de una aproximación combinada donde se estudie la respuesta de las poblaciones al cambio climático y su resiliencia utiliznado diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].
-
-Remote sensing
-
-These climatic alterations are likely to have important consequences for tree species dynamics at local and regional scales ( Peñuelas and Boada 2003 , Van Mantgem et al.
-2009 , Matías and Jump 2015 ).
+## 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 
@@
 
-# 2) Combinación metodológica 
-Analizar resiliencia de poblaciones utilizando RS y dendro.
-
-Analizar el impacto de las sequías en los árboles es muy importante y prioriatrio...
-
-* Analysing the impact of drought on tree growth is therefore a priority under the current climatic conditions, in which global warming is probably increasing the severity of droughts (Williams et al., 2013).
-
-En este estudio combinamos el uso de datos de dendrocronologia con información de productividad procedente de satélite.
-Por un lado, sabemos que el annual radial growth increment se utiliza como proxy para estudiar la forest vitality [@Fritts1976; @Dobbertin2005]; por otro lado el EVI para productividad ...
-
-Una vez dicho esto...
-
-a) Como la dendro ayuda a evaluar las sequías (ver Gazol 2017)
-
-* Gea-Izquierdo 2014: El análisis del inter-annual growth variability can be used to study changes in productiviyt and forest vitality to detect vulnerability to stress (Fritts 1976; Doberti)
-
-* Tree ring width or annual radial growth increment is a widely used proxy for tree vitality (ver  [@Bhuyan2017] [@Dobbertin2005]
-
------------------------
-
-## 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
+These climatic alterations are likely to have important consequences for tree species dynamics at local and regional scales ( Peñuelas and Boada 2003 , Van Mantgem et al.
+2009 , Matías and Jump 2015 ).
 
@@
 ```
-
-.. ver [@VicenteSerrano2013] impacto de las sequías en diferentes zonas (i.e montañas)
-
-Objetivos del trabajo
-
-### Aims
-In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both greenees and growth of *Q. pyrenaica* forests in Sierra Nevada. Specifically,
-
-The aims of this work were:
-
-* To quantify how two extreme drought events influenced the greenness and radial growth of *Q. pyreancia* forests in their rear edge,
-
-* to analyze the resilience of these forests to successive extreme drought events,
-
-* and to explore differences in the resilience metrics between populations located in contrasting slopes within the rear edge of the distribution of this species.Do northern and southern populations of Pyrenean oak forests differ in their resistance, resilience or recovery to extreme drought events?
-
-# Otras notas: 
-* To guide sustainable forest management, forest researchers are asked to provide concrete answers about forest resilience in response to expected climatic trends, and extreme climatic events (Lindner et al., 2014) https://www.sciencedirect.com/science/article/pii/S030147971400379X?via%3Dihub
-
-* Hablan de incorporar ademas de RS, el conocimiento ecológico (los principios ecológicos) para comprender las consecuencias de las disturbances analizadas con RS, y sobre todo para predecir futuro.
-(McDowell et al https://www.sciencedirect.com/science/article/pii/S1360138514002726)
-
-----------------------------
-
-* Proyeccciones de crecimeinto de para *Q.
-pyrenaica* sugieren un declive en el crecimiento en las proximas decadas a lo largo de su rango de distribución en la P.
-Iberica, en diferentes escenarios climáticos [@GeaIzquierdo2013]: Predictions suggest that QUPY productivity would decline in the next decades all along its distributional range in the Iberian Peninsula for all the climate scenarios studied.
-Este declive debería ser mas dramático en las baja altitudes de las zonas mas calientes del sur de su distribución
-
---> vulnerabilidad local (related to rear-edge)
-La tendencia de reducción drástica sugerida por el modelo para el dry-edge de la distribución de la especie podría expresar una aumentada vulnerabilidad de los árboles al incremento del estrés hídrico forzado por un clima mas cálido [@GeaIzquierdo2013].
-
-Negative trends over recent years and the greater response to moisture availability found at warmer and drier low elevations in the south suggest vulnerability to warming at the local low elevation dry edge of the species's range (e.g.
-QUPY9).
-Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()
-
-Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg.
-@Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al.
-https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI).
-Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).
-
-Case studies that focus on multiple scales - including local scales - and also valuables (Dale et al.
-2018 Frontiers) --> Se necesitan casos de estudio que se enfoquen en escalas multiples (incluyendo las locales) y que combinen varias metodologías (Jump et al.
-...)
-
-3) rear edge
-
-- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):
-
-* Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
-(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...
-
-...
-Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
-
---> plantear la hipótesis... efecto de sequias recurrentes ...
-
-y luego los objetivos

### 2018-03-14 man/ms_3_discussion.Rmd
update

@@
 
+### Notas de comparación metodológica
+Babst et al. 2017 hablan tambien de combinar ambos métodos
+
+Jump et al habla de utilizar combinaciones de ambos métodos para estudiar esto, y esto es importante para las poblaciones del rear edge.
+
+### Otras 
+Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling..
+// dorado-liñan // sanchez salguero
+
+trabajos rear edge: 
+* sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
+* sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
+* dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
+* herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
+* dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
+buscar otros
+
+Sin embargo el estudio de las poblaciones situadas en su rear edge requiere de una aproximación combinada donde se estudie la respuesta de las poblaciones al cambio climático y su resiliencia utiliznado diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].
+
+.. ver [@VicenteSerrano2013] impacto de las sequías en diferentes zonas (i.e montañas)
+
+# Otras notas: 
+* To guide sustainable forest management, forest researchers are asked to provide concrete answers about forest resilience in response to expected climatic trends, and extreme climatic events (Lindner et al., 2014) https://www.sciencedirect.com/science/article/pii/S030147971400379X?via%3Dihub
+
+* Hablan de incorporar ademas de RS, el conocimiento ecológico (los principios ecológicos) para comprender las consecuencias de las disturbances analizadas con RS, y sobre todo para predecir futuro.
+(McDowell et al https://www.sciencedirect.com/science/article/pii/S1360138514002726)
+
+----------------------------
+
+* Proyeccciones de crecimeinto de para *Q.
+pyrenaica* sugieren un declive en el crecimiento en las proximas decadas a lo largo de su rango de distribución en la P.
+Iberica, en diferentes escenarios climáticos [@GeaIzquierdo2013]: Predictions suggest that QUPY productivity would decline in the next decades all along its distributional range in the Iberian Peninsula for all the climate scenarios studied.
+Este declive debería ser mas dramático en las baja altitudes de las zonas mas calientes del sur de su distribución
+
+--> vulnerabilidad local (related to rear-edge)
+La tendencia de reducción drástica sugerida por el modelo para el dry-edge de la distribución de la especie podría expresar una aumentada vulnerabilidad de los árboles al incremento del estrés hídrico forzado por un clima mas cálido [@GeaIzquierdo2013].
+
+Negative trends over recent years and the greater response to moisture availability found at warmer and drier low elevations in the south suggest vulnerability to warming at the local low elevation dry edge of the species's range (e.g.
+QUPY9).
+Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()
+
+Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg.
+@Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al.
+https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI).
+Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).
+
+Case studies that focus on multiple scales - including local scales - and also valuables (Dale et al.
+2018 Frontiers) --> Se necesitan casos de estudio que se enfoquen en escalas multiples (incluyendo las locales) y que combinen varias metodologías (Jump et al.
+...)
+
+3) rear edge
+
+- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):
+
+* Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
+(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...
+
+...
+Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
+
 ---- old 

### 2018-03-17 analysis/bai/trends_resilience_bai.Rmd
add

@@
 
-plotrel <- function(df, v, x, label, XLab, YLab){
+plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq){
   require(ggplot2)
@@
   ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
+    geom_smooth(method='lm', se=FALSE, colour="black") +
     geom_errorbar(mapping = yerrbar) + 
@@
     theme_bw() + 
-    geom_smooth(method='lm', se=FALSE) +
-    xlab(XLab) + ylab(YLab) + geom_smooth(method='lm', se=FALSE, colour="black") +
+    xlab(XLab) + ylab(YLab) + 
     geom_text_repel() + 
-    annotate("text", x=7, y = 0.5, 
+    annotate("text", x=coordX_eq, y = coordY_eq, 
              label=lm_eqn(lm(formulita, df)), parse=TRUE)
@@
 sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)")
+        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
+        coordX_eq = 7, coordY_eq = 0.5)
 
 sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resistance (Rt) (3year window size)")
+        XLab = "Drought Severity", YLab = "Resistance (Rt) (3year window size)",
+        coordX_eq = 7, coordY_eq = 0.5)
 
 sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Recovery (Rc) (3year window size)")
+        XLab = "Drought Severity", YLab = "Recovery (Rc) (3year window size)",
+        coordX_eq = 7, coordY_eq = 0.5)
 
@@
 sevRs10 <- plotrel(sdf10, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (10year window size)")
+        XLab = "Drought Severity", YLab = "Resilience (Rs) (10year window size)",
+        coordX_eq = 7, coordY_eq = 0.5)
 
 sevRt10 <- plotrel(sdf10, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resistance (Rt) (10year window size)")
+        XLab = "Drought Severity", YLab = "Resistance (Rt) (10year window size)",
+        coordX_eq = 7, coordY_eq = 0.5)
 
 sevRc10 <- plotrel(sdf10, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Recovery (Rc) (10year window size)")
+        XLab = "Drought Severity", YLab = "Recovery (Rc) (10year window size)",
+        coordX_eq = 7, coordY_eq = 0.5)
 
@@
 sevRt10 + sevRc10 + sevRs10 
+dev.off()
+```
+
+# relation con drought intensity 
+```{r}
+
+intensityRs <- plotrel(sdf, v = "rs", x = 'd_intensity', label = 'disturb_year',
+        XLab = "Drought Intensity", YLab = "Resilience (Rs) (3year window size)",
+        coordX_eq = -1.5, coordY_eq = 0.5)
+
+intensityRt <- plotrel(sdf, v = "rt", x = 'd_intensity', label = 'disturb_year',
+        XLab = "Drought Intensity", YLab = "Resistance (Rt) (3year window size)",
+        coordX_eq = -1.5, coordY_eq = 0.5)
+
+intensityRc <- plotrel(sdf, v = "rc", x = 'd_intensity', label = 'disturb_year',
+        XLab = "Drought Intensity", YLab = "Recovery (Rc) (3year window size)",
+        coordX_eq = -1.5, coordY_eq = 0.5) + ylim(c(0.5,3.5))
+
+pdf(here::here('/out/bai_resiliences_severes', 'rs_vs_drought_3years_intensity.pdf'), width=18, height = 7)
+intensityRt + intensityRc + intensityRs 
 dev.off()

### 2018-03-20 man/ms_3_discussion.Rmd
add notes sobre metodología

@@
 
+---------------------
+
+# Notas sobre metodología
+
+En este estudio se combinan ambas metodologías por varias razones: 
+* distinta sensibilidad
+
+@Gazol2018 analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR.
+Encontraron que los datos de TR son mas sensibles para la resiliencia del bosque a la sequía que los datos de RS
+
+* es importante utilizar varias metodologías en el rear-edge
+
+* La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017]
+
+* El uso de análisis combinados además nos permite distinguir la varibilidad en el crecimiento del bosque inducida por clima de aquella impulsada por otros procesos ecológicos a nivel de comunidad [@Babst2017]
+
+Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes.
+
+### Estudios que utilizan RS y TR
+
+* Wu et al. 2017. 
+	* En un estudio sobre el efecto de la sequía sobre el crecimiento en bosques, grass y shrub del hemisferio norte, encontraron que los bosques (Deep-rooted forest) exhibit a drought legacy response with reduced growth during up to 4 years despues de un evento extremo de sequía.
+
+	* Además vieron que los bosques mostraron una stronger drought resilience in forest (i.e. smaller growth reduction after severe drought) (Gazol el al. 2017; Wu et al. 2017).
+
+	* Esto se puede deber a que acceden a reservas de agua mas profundas (los pastos y los matorrales no pueden acceder a ese agua)
+
+### 
+Nosotros hemos obtenido una relación entre la severidad de la sequía y la resiliencia mostrada por los robledales en su rear edge. (ver gráfica )
+
+@Gazol2018 encontró La resiliencia está relacionada con la severidad de la sequía y con la composición del bosque
+
+Rt-NDVI and Rt-TR were positively related with drought intensity
+
+### Muchos estudios han evaluado la resiliencia de los bosques a un evento de sequía, pero pocos lo han hecho a
+
+## Peores años de sequía.
+1986, 1994-1995, 1999 y 2005 se consideran de los eventos de sequía mas severos en la región durante la segunda mitad del siglo XX. 
+También hablan de la sequía de 2012, pero tenían pocos datos
+
+## 
+@Gazol2018 Resistencia a la sequía mayor en la zona norte de españa que en las zonas secas del sur
+
+Qpyr 
+* Valores de Rt-TR aumentaron a lo largo del tiempo, sin embargo Rt-NDVI dismiuyó con el tiempo
+* Valores Rc-NDVI tendencia positiva
+
 ## References

### 2018-03-20 man/ms_3_discussion.Rmd
organizando la discusion

@@
 
-Algunas notas sobre la sequía de 2005: 
-* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].
-
 Los valores de resiliencia observados para el greenness (cercanos a 1) muestran como esta especie, en su rear edge muestra una buena capacidad de recuperación frente a eventos extremos,
@@
 Encontramos diferencias entre poblaciones. Asi por ejemplo en las situadas mas al sur, vemos que las anomalias de EVI son menos negativas para el 2005; También vemos tendencias positivas en el BAI...
-
-## metodologico
-
-La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía se pueden ver tanto con información de satélite (hablar del browning), como con información mas de campo (hablar aquí de lo del BAI).
-Indicar también las diferencias entre las resiliencias detectadas por ambas metodologías.
-
-### Notas sobre metodologia
-
-* Algunos trabajos han analizado la relación ente RWI y remote sensing [@VicenteSerrano2013]; [(Bhuyan et al.
-Remote Sensing 9(6):526)]; [Vicente Serrano et al.
-https://www.sciencedirect.com/science/article/pii/S003442571630373X] ...
-
-* Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q.
-pyrenaica situadas en su reaar edge frente a los eventos de sequía.
-
---- notas para esto 
-Se trata de expresar la idea de mezcla de aproximacioes: remote-sensing con dendro.
-Aunque existen algunos trabajos que utilizado RWI y remote sensing nuestra aproximación es importante, sobre todo para estudiar poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
-De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
-
-Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
-(ver esto y completar)
-
-Aunque el NDVI o similar ha sido utilizado para monitorizar la respuesta de bosques frente a sequía, su escala espacial limita su uso para estudiar cambios en poblacioens situadas en su rear edge.
-
- Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability.
-In such regions, targeted collection of high-resolution data is necessary in order to identify currently occurring changes and predict the magnitude and spatial distribution of future decline.
-[@Jump2010]
-
-comentario de GGI
-
-Gracias Antonio.
-Desde luego que debes resaltar a lo largo del artículo la combinación de sensores y dendro como fortaleza de tu trabajo al analizar el tema de resiliencia desde dos escalas espacio-temporales complementarias.
-Y eso independientemente que lo diga o no Alistair Jump (que las citas vienen bien, y seguro que el trabajo es bueno), combinar ambas escalas/datos es beneficioso no sólo en poblaciones rear-edge, sino en general.
-Es decir, desde mi punto de vista, aunque decidas no centrarte tanto en el aspecto metodológico como sugieren Regino y Curro, eso no quiere decir que este punto no lo presentes de forma explícita en tu trabajo (en abstract debe aparecer, luego en intro, discusión y tal vez una frase en conclusiones) como uno de los elementos fuertes (un término medio, vamos).
-Creo que ya hemos discutido esto cuando hemos hablado.
-No es el primer trabajo que lo hace, como bien señalas, pero tampoco hay que pensar que no tiene valor, pues no hay tantos trabajos que combinen ambas aproximaciones (habrá cada vez más de hecho).
-Así que ánimo que queda poco y el trabajo puede quedar interesante si somos capaces de escribir e interpretar bien los resultados que nos has enviado.
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Notas Jump 2010 
-Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)
-
-NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.
-
-Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!
-
-Otra idea --> Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)
-
-NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.
-
-Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!
-
-El NDVI se ha utilizado para monitorizar la respuesta de bosques frente a sequía (ej.: 24,26,32, 44 en Jump2009 )
-
-Necesidad de Ground-based assessment (p 1974 Jump 2009) -> Aunque utilizemos datos de satélite, siempre es necesario utilizar aproximaciones combinda
-
-Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
-(ver esto y completar)
-
-El problema de los datos ground-based assessment, existe un trade-off entre el área total cubierta por una red de monitoreo y su resolución local.
---Por tanto también existen limitaciones de las ground-based para estudiar poblaciones en su rear edge.
-")
-```
 
@@
 
-### Notas de comparación metodológica
-Babst et al. 2017 hablan tambien de combinar ambos métodos
-
-Jump et al habla de utilizar combinaciones de ambos métodos para estudiar esto, y esto es importante para las poblaciones del rear edge.
-
 ### Otras 
@@
 
-La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
-Esto se ha observado también en otras especies en el sur de la P.
-Ibérica (p.ej.
-P.
-nigra en Andalusia [@SanchezSalguero2013], ...
-INCLUIR otras especies y citas).
-
-Algunas notas sobre la sequía de 2005:
-
-* 2004/2005 hydrological year is considered one of the worst drooughts ever recorded in the Iberian Peninsula, particularly in the central and southern sectors [@GarciaHerrera2007].
-* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].
-* The hydrological year from October 2004 to September 2005 was the driest on record at several locations throughout Iberia [@GarciaHerrera2007].
-* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
-
 ---------------------
 
-# Notas sobre metodología
+# METODOLOGY
 
-En este estudio se combinan ambas metodologías por varias razones: 
-* distinta sensibilidad
+## metodologico
 
-@Gazol2018 analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR.
+La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía se pueden ver tanto con información de satélite (hablar del browning), como con información mas de campo (hablar aquí de lo del BAI).
+Indicar también las diferencias entre las resiliencias detectadas por ambas metodologías.
+
+Nuestro trabajo combinamos las dos aproximaciones metodológicas por varias razones:
+
+* Cada metodología muestra distinta sensibilidad
+    * @Gazol2018 analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR.
 Encontraron que los datos de TR son mas sensibles para la resiliencia del bosque a la sequía que los datos de RS
 
-* es importante utilizar varias metodologías en el rear-edge
+* Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares...
 
@@
 
-* El uso de análisis combinados además nos permite distinguir la varibilidad en el crecimiento del bosque inducida por clima de aquella impulsada por otros procesos ecológicos a nivel de comunidad [@Babst2017]
+* El uso de análisis combinados además nos permite distinguir la varibilidad en el crecimiento del bosque inducida por clima de aquella impulsada por otros procesos ecológicos a nivel de comunidad [@Babst2017].
+Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes.
 
-Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes.
+### Notas sobre metodologia
+
+* Algunos trabajos han analizado la relación ente RWI y remote sensing [@VicenteSerrano2013]; [(Bhuyan et al.
+Remote Sensing 9(6):526)]; [Vicente Serrano et al.
+https://www.sciencedirect.com/science/article/pii/S003442571630373X] ...
+
+* Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q.
+pyrenaica situadas en su reaar edge frente a los eventos de sequía.
+
+--- notas para esto 
+Se trata de expresar la idea de mezcla de aproximacioes: remote-sensing con dendro.
+Aunque existen algunos trabajos que utilizado RWI y remote sensing nuestra aproximación es importante, sobre todo para estudiar poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
+De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
+
+Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
+(ver esto y completar)
+
+Aunque el NDVI o similar ha sido utilizado para monitorizar la respuesta de bosques frente a sequía, su escala espacial limita su uso para estudiar cambios en poblacioens situadas en su rear edge.
+
+ Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability.
+In such regions, targeted collection of high-resolution data is necessary in order to identify currently occurring changes and predict the magnitude and spatial distribution of future decline.
+[@Jump2010]
+
+comentario de GGI
+
+Gracias Antonio.
+Desde luego que debes resaltar a lo largo del artículo la combinación de sensores y dendro como fortaleza de tu trabajo al analizar el tema de resiliencia desde dos escalas espacio-temporales complementarias.
+Y eso independientemente que lo diga o no Alistair Jump (que las citas vienen bien, y seguro que el trabajo es bueno), combinar ambas escalas/datos es beneficioso no sólo en poblaciones rear-edge, sino en general.
+Es decir, desde mi punto de vista, aunque decidas no centrarte tanto en el aspecto metodológico como sugieren Regino y Curro, eso no quiere decir que este punto no lo presentes de forma explícita en tu trabajo (en abstract debe aparecer, luego en intro, discusión y tal vez una frase en conclusiones) como uno de los elementos fuertes (un término medio, vamos).
+Creo que ya hemos discutido esto cuando hemos hablado.
+No es el primer trabajo que lo hace, como bien señalas, pero tampoco hay que pensar que no tiene valor, pues no hay tantos trabajos que combinen ambas aproximaciones (habrá cada vez más de hecho).
+Así que ánimo que queda poco y el trabajo puede quedar interesante si somos capaces de escribir e interpretar bien los resultados que nos has enviado.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+# Notas Jump 2010 
+Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)
+
+NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.
+
+Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!
+
+Otra idea --> Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)
+
+NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.
+
+Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!
+
+El NDVI se ha utilizado para monitorizar la respuesta de bosques frente a sequía (ej.: 24,26,32, 44 en Jump2009 )
+
+Necesidad de Ground-based assessment (p 1974 Jump 2009) -> Aunque utilizemos datos de satélite, siempre es necesario utilizar aproximaciones combinda
+
+Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
+(ver esto y completar)
+
+El problema de los datos ground-based assessment, existe un trade-off entre el área total cubierta por una red de monitoreo y su resolución local.
+--Por tanto también existen limitaciones de las ground-based para estudiar poblaciones en su rear edge.
+")
+```
 
@@
 
-## Peores años de sequía.
-1986, 1994-1995, 1999 y 2005 se consideran de los eventos de sequía mas severos en la región durante la segunda mitad del siglo XX. 
-También hablan de la sequía de 2012, pero tenían pocos datos
+------------------
+
+# Años de sequía
+
+### 2005 frente a 2012
+
+* Analizando los datos de precipitación acumulada de las estaciones alrededor de SN, observamos como 2005 fue el año mas seco desde 1950. 
+* Datos de SPEI (0.5 º Grid) 
+    * 2005 fue peor que 2012, pero no peor que 1995 o que 1999 (ver figura S2)
+    * Tabla S4. Severidad sequías, la de 2005 fue menos severa y menos intensa y de menor duración que la de 2005 (de donde vienen estos datos???)
+
+* La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
+Esto se ha observado también en otras especies en el sur de la P.
+Ibérica (p.ej.
+P.
+nigra en Andalusia [@SanchezSalguero2013], ...
+INCLUIR otras especies y citas).
+* 2004/2005 hydrological year is considered one of the worst drooughts ever recorded in the Iberian Peninsula, particularly in the central and southern sectors [@GarciaHerrera2007].
+* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].
+* The hydrological year from October 2004 to September 2005 was the driest on record at several locations throughout Iberia [@GarciaHerrera2007].
+
+### 2005 en contexto
+
+* Algunos trabajos hablan que los peores años de sequía en la P. Iberica durante la segunda mitad del siglo XX fueron 1986, 1994-1995, 1999, 2005 y 2012
+
+* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
 

### 2018-03-20 man/ms_figures.Rmd
fixed figure 1

@@
 **Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain range, where three clusters of oak populations have been identified [@PerezLuque2015] (showed in different colour) (b).
-A grid of with the MODIS pixels for each population is shown (see material and methods).
-Detailed location of the sampling sites: northern (San Juan, SJ) (c) and southern ones (Cáñar: CALow and CAHigh) (d).
+Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
+Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
+For each population, a grid with the MODIS pixels is shown (see material and methods).
+Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c) and southern ones (Cáñar: CA-Low and CA-High) (d).
 Colour Orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.

### 2018-03-20 man/ms_figures.Rmd
change order figure 2

@@
 
-**Figure 2.** EVI standardized anomaly ($EVI~sa$) during the period 2000-2016 for northern and southern populations. Error bars show standard error.
+**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
 
@@
 aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site,  
-                   fill = site, linetype=site,
-                   shape = site, 
+                   color = site_sorted, group = site_sorted,  
+                   fill = site_sorted, linetype=site_sorted,
+                   shape = site_sorted, 
                    ymin = lower.ci, ymax = upper.ci)
@@
 variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
+
+# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
+item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
 
@@
   theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free_y"),
+  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
   xlab(''), 
@@
     plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(
-    values=c("N" = 19, "S" = 21, "CA-High"=22, "CA-Low" = 15, "SJ" = 17)), 
-  scale_color_manual(
-    values=c("N" = 'black',"S" = "blue", "CA-High" = 'blue', "CA-Low" = 'blue', "SJ" = 'black')),
-  scale_fill_manual(
-    values=c("N" = 'black', "S" = "white", "CA-High" = 'white', "CA-Low" = 'blue', "SJ" = 'black')),
-  scale_linetype_manual(values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(
-    values=c("N" = 2, "S" = 2, "CA-High"=2.5, "CA-Low" = 2.5, "SJ" = 2.5)),
+  scale_shape_manual(labels = item_site, guide = guide_legend(title = NULL),
+    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 19)), 
+  scale_color_manual(labels = item_site, guide = guide_legend(title = NULL),
+    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
+  scale_fill_manual(labels = item_site, guide = guide_legend(title = NULL),
+    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
+  scale_linetype_manual(labels = item_site, guide = guide_legend(title = NULL),
+                        values=c("solid","solid","solid","solid", "solid")),
+  scale_size_manual(labels = item_site, guide = guide_legend(title = NULL),
+    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
+
   geom_text(aes(label=Letter, y=postext), 
@@
   TRUE ~ site
+)))
+
+m <- m %>% mutate(site_sorted = as.factor(
+  case_when(
+  site == "N" ~ "0_N",
+  site == "S" ~ "1_S",
+  site == "SJ" ~ "2_SJ",
+  site == "CA-High" ~ "3_CA-High",
+  site == "CA-Low" ~ "4_CA-Low"
 )))

### 2018-03-20 man/ms_figures.Rmd
include fix figures

@@
 
+tl <- guide_legend(title = "Site")
+
 themeR <- list(
@@
     plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = NULL),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 19)), 
-  scale_color_manual(labels = item_site, guide = guide_legend(title = NULL),
+  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
+    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
+  scale_color_manual(labels = item_site, guide = tl,
     values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = guide_legend(title = NULL),
+  scale_fill_manual(labels = item_site, guide = tl,
     values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = guide_legend(title = NULL),
+  scale_linetype_manual(labels = item_site, guide = tl,
                         values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = guide_legend(title = NULL),
+  scale_size_manual(labels = item_site, guide = tl,
     values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
@@
 pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (black fill circle) with southern ones (blue empty circle).
-For BAI we compared northern population (San Juan, SJ; black triangle) with southerns populations: Cáñar-High (Ca-H; blue empty squares) and Cáñar-Low (Ca-L; blue fill sauares).
-Different letters above error bars indicate significant post hoc differences between groups (see material and methods).
+For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
+For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
+Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
 

### 2018-03-20 man/ms_figures.Rmd
modify colors and legend

@@
 
+colores <- c('#7570b3','#e7298a','#1b9e77')
+# colores <- c('#33a02c', '#984ea3', '#a65628')
+
+cro_sites <- cro_sites %>% mutate(site = 
+  case_when(
+    site == "CA_Low" ~ "CA-Low", 
+    site == "CA_High" ~ "CA-High",
+    site == "SJ" ~ "SJ",
+  )
+)
+
 # cores 
@@
         strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_colour_manual(values = colores) +
+  scale_fill_manual(values = colores) +
   ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
   scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA_High'), 
+  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
               aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
@@
         ) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_colour_manual(values = colores) +
+  scale_fill_manual(values = colores) +
   scale_x_continuous(breaks = seq(1815, 2020, by=10))  
@@
   filter(year >= 1975) %>% 
-  filter(site == 'CA_High')
+  filter(site == 'CA-High')
 
@@
 **Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA_Low; *pink*) and high-elevation (CA_High, *purple*) sites.
+pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
 Shading areas coorespond to standard error of the mean.
@@
 We only show chronologies with # trees > 5.
+Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r2`).
 

### 2018-03-21 man/ms_figures.Rmd
update figures

@@
 ```{r}
-label_site <- c("0_sj" = "Northern (SJ)",
-                "1_caH" = "Southern (Ca-High)",
-                "2_caL" = "Southern Low (Ca-Low)")
+label_site <- c("0_sj" = "SJ",
+                "1_caH" = "CA-High",
+                "2_caL" = "CA-Low")
 
@@
 **Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data (precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures) for northern site (*green* bars), low-elevation southern site (*light blue* bars) and high-elevation shouthern (*dark blue* bars) site.
+pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
+*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site; and *dark blue* bars: high-elevation shouthern site.
 Asteriks indicate significant ($P < 0.05$) correlation coefficients.
@@
   facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "Southern (CA-High)", "2_caL" = "Southern(CA-Low)"))) + 
+             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
     ylab(custom_ylab) + custom_corrected_gc
@@
   facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "Northern (SJ)"))) + 
+             labeller=as_labeller(c("0_sj" = "SJ"))) + 
   custom_corrected_gc + ylab('') + 
@@
 
-**Figure 7.** Comparison of median growth change ($GC$) following [@Nowacki1997] for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
+**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
 

### 2018-03-21 man/ms_figures.Rmd
combine plots

@@
 #####
-**Appendix S1** Acumulated monthly precipitation during the hydrological year 2004-2005 (blue line) and 2011-2012 (red line).
-The boxplot representing the average from 1940-2015 period.
+
+**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017 for 28 meteorological stations (from National Spanish Meteorological Services, AEMET) distributed around the Sierra Nevada area.
+Points represent mean and errorbars standard error.
+*Black* line indicates mean for all period.
+*Red* lines represent -1 (*dotted* line) and -2 (*dashed* line) standard deviation.
+*Blue* lines represent +1 (*dotted* line) and +2 (*dashed* line) standard deviation.
+We labelled all years with average values below -1SD.
+***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
+The boxplot representing the average from 1950-2015 period.
 Data from meteorological station Granada, Base Aérea (National Spanish Meteorological Services, AEMET).
@@
 
-st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012'))) %>% 
-  ggplot(aes(x=as.factor(hmonth), y=csum)) + 
+df <- st5514_pro  %>% 
+  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
+
+cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
   geom_boxplot(aes(x=as.factor(hmonth))) +  
@@
   theme(strip.background = element_rect(fill = NA, color = "black")) + 
-  ylab('Precipitation (mm) acumulated') + xlab('') + 
-  ggtitle('Granada Base Aerea, 1950-2015') + 
+  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
+  # ggtitle('Granada Base Aerea, 1950-2015') + 
   scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                             "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 6, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 6, colour = "red") + 
-  annotate("text", label = "mean 1950-2015", x = 3, y = 600, size = 6, colour = "black")
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-We used data from Global SPEI database.
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
+  annotate("text", label = "2004-2005", x = 2, y = 500, size = 4, colour = "blue") + 
+  annotate("text", label = "2011-2012", x = 2, y = 550, size = 4, colour = "red") + 
+  annotate("text", label = "mean 1950-2015", x = 3, y = 600, size = 4, colour = "black") +
+  theme(panel.grid = element_blank())
+```
+
 ```{r}
@@
   ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-prec_hydrol_sn
-```
-
-**Appendix S3.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017 for 28 meteorological stations (from National Spanish Meteorological Services, AEMET) distributed around the Sierra Nevada area.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 (*dotted* line) and -2 (*dashed* line) standard deviation.
-*Blue* lines represent +1 (*dotted* line) and +2 (*dashed* line) standard deviation.
-We indicate all years with average values below -1SD.
+
+prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
+```
+
+#####
+
+```{r, echo=FALSE}
+# Get file names
+myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
+
+mydf <- data.frame()
+
+# Loop to read files and prepare data
+for (j in myfiles){ 
+  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
+
+  # Remove csv and get name 
+  name_aux <- str_replace(j, ".csv", "")
+
+  # Get lat long
+  latlong <- str_replace(name_aux, "SPEI_", "")
+
+  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
+  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
+
+  aux$lat <- mylat
+  aux$long <- mylong
+
+  # Split time 
+  # Get year 
+  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
+
+  # Get month 
+  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
+
+  # Variable month in number 
+  aux$months <- match(aux$month, month.abb)
+
+  # assign(name_aux, aux)
+
+  mydf <- rbind(mydf, aux)
+}
+
+spei <- mydf %>% select(SPEI_12, year, months) %>% 
+  group_by(year,months) %>% 
+  summarise(spei6 = mean(SPEI_12),
+            spei6_sd = sd(SPEI_12)) %>% 
+  mutate(date = lubridate::ymd(paste(year, months, "01")))
+
+spei <- spei %>% 
+  mutate(sign = case_when(
+    spei6 > 0 ~ "pos",
+    TRUE ~ "neg"))
+```
+
+```{r, fig.height=5}
+limits <- as.Date(c("1949-01-01", "2015-12-01"))
+spei %>% 
+  ggplot(aes(x=date, y=spei6, fill=sign)) +  
+  geom_bar(stat = "identity") + 
+  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
+  theme_bw()  + 
+  labs(title= 'Drought evolution for Sierra Nevada',
+    x= '',
+    y='SPEI') +
+  theme(strip.background = element_rect(fill = "white"),
+        legend.position = "none") +
+  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
+```
+
+**Appendix S2**.
+Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+We used data from Global SPEI database.
+We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
+
+```{r, echo=FALSE, eval=FALSE}
+**Figure S2**.
+Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
+Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
+
+![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
+```
 

### 2018-03-21 man/ms_1_material_methods.Rmd
change drouhgt and greenness description

@@
 
-The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and they were charecterized as extreme drought in our climatic data (Figure S1).
-Aunque sabemos que existen otros años de sequía extrema (e.g.
-1995, 1981) [@VicenteSerrano2014], no se han elegido por una limitación metodológica, esto es, nuestro objetivo es analizar el efecto de la sequía y la resiliencia a los eventos de sequía utilizando información de satélite y de dendrocronología, y la información de satélite utilizada (ver mas adelante) solamente está disponible desde el año 2000 en adelante.
-We compare only these previous drought years in radial growth-time series, to place older drought events with those in 2005 and 2012 ...
-DUDA GUILLERMO
-
-* ¿Decimos aquí algo así como?: Para los datos de radial-growth hemos computado la resiliencia en los 10 eventos mas severos (ver tabla S1)
+During the second half of the XX century, several extreme drought episodes were recorded for the Iberian Peninsula (*e.g*.
+1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014 @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were charecterized as extreme drought in our climatic data (Appendices S1-S3).
+We focused on these two drought events because limitations about temporal availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
+Nevertheless, for radial growth-time series, older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events.
+The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Pascoa2017 (see Appendix S3)
 
@@
 ## Greenness data 
+As a proxy of primary growth, we characterized the vegetation greenness of *Q.
+pyrenaica* by means of the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
+Vegetation indices derived from remote sensing information are linked with primary productivity [@Ruimy1994] and they have been been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009].
+These indices are linearly related to net primary productivity through the fraction of photosynthetically active radiation intercepted by green vegetation (*i.e.* vegetation greenness) [@Ruimy1994].
 
-As a proxy to primary growth, to characterize the vegetation greenness of *Q.
-pyrenaica* we used the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
-EVI and NDVI (*Normalized Difference Vegetation Index*) are the most common greenness vegetation indices.
-We used EVI instead of NDVI because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].
+We used EVI instead of NDVI (*Normalized Difference Vegetation Index*; another of the most common greenness vegetation indices) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].
 
@@
 
-After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2016.
+After the filter out process, we computed the annual mean EVI values as a surrogate of mean annual primary production, and the EVI anomaly for each pixel for the period 2000 - 2016.
 EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).

### 2018-03-21 man/ms_1_material_methods.Rmd
update

@@
 The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Appendix S7) and this length was used in other studies (e.g.
-@Gazol2017).
-In addition we calculated the same indices for BAI data for the most severe drought events (Appendix S4) since 1900.
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Appendix S6) and this length was used in other studies [*e.g.* @Gazol2017].
+In addition we calculated the same indices for BAI data for the most severe drought events (Appendix S3) since 1900.
 

### 2018-03-21 man/ms_figures.Rmd
update

@@
 
+**Appendix S2**.
+Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+Data from Global SPEI database (http://spei.csic.es/database.html).
+We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
+
 ```{r, fig.height=5}
@@
     x= '',
-    y='SPEI') +
+    y='SPEI (12-month scale)') +
   theme(strip.background = element_rect(fill = "white"),
@@
 ```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-We used data from Global SPEI database.
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
 
@@
 # solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% as.data.frame()
+dys <- dy %>% 
+  filter(d_duration > 2) %>%  
+  mutate(year = ifelse(minyear == maxyear, 
+                       minyear, 
+                       paste0(minyear, '-', maxyear))) %>% 
+  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
+  as.data.frame()
+
+colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
 
@@
 
-**Appendix S4.** Tabla de drought severas.
-
-##### 
-**Appendix S5.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
+**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
+A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
+A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g* @Spinoni2015; Spinoni2017a; @Pascoa2017].
+The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
+***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
+***Intensity*** and ***Lowest SPEI** refer to the mean and lowest value of SPEI respectively during the drought event duration.
+
+##### 
+**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
 
@@
 
-**Appendix S6.** Reilience vs. droughts
-
-##### 
-**Appendix S7.** Correlation between indices of resilience (Rt, resistance; Rc, recovery; Rs, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
+**Appendix S5.** Reilience vs. droughts
+
+##### 
+**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
 Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-ws2, ws3 and ws4 indicate periods of 2, 3 and 4 years after a drought.
 
@@
     mutate(w1 = case_when(
-      w1 == 2 ~ 'ws2',
-      w1 == 3 ~ 'ws3',
-      w1 == 4 ~ 'ws4'),
+      w1 == 2 ~ '2 years',
+      w1 == 3 ~ '3 years',
+      w1 == 4 ~ '4 years'),
       w2 = case_when(
-      w2 == 2 ~ 'ws2',
-      w2 == 3 ~ 'ws3',
-      w2 == 4 ~ 'ws4'))
+      w2 == 2 ~ '2 years',
+      w2 == 3 ~ '3 years',
+      w2 == 4 ~ '4 years'))
 
@@
 g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
+g_rse <- ggcorrplot(cor_rse, tpe='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
 

### 2018-03-21 man/ms_1_material_methods.Rmd
updae

@@
 Nevertheless, for radial growth-time series, older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events.
-The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Pascoa2017 (see Appendix S3)
+The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by Spinioni (see Appendix S3)
 

### 2018-03-26 man/ms_tables_portrait.Rmd
add temperature and prec data

@@
 t1raw <- t1raw %>% mutate(name = case_when(
-  site == 'caH' ~ 'Cáñar-High', 
-  site == 'caL' ~ 'Cáñar-Low',
-  site == 'sj' ~ 'San Juan'))
+  site == 'caH' ~ 'CA-High', 
+  site == 'caL' ~ 'CA-Low',
+  site == 'sj' ~ 'SJ'))
 
 t1 <- t1raw %>% dplyr::select(
-  site, lat_m, long_m, elev_mean, slope_mean, 
+  name, lat_m, long_m, elev_mean, slope_mean, 
   n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, 
@@
 
+# ADD data climate
+# clima obsnev (from repo clima SN; son datos de mapas pasados 100 x 100 1960 - 2010)
+
+clima <- read.csv(here("/data/climate_obsnev", "obsnev_qp_pop.csv"), header = TRUE)
+cli <- clima %>% filter(pop %in% c(2,6)) %>% mutate_if(is.numeric, funs(round(., 1))) %>% 
+  dplyr::select(pop, mean, variable) %>% 
+  spread(variable, mean) %>% 
+  mutate(name = case_when(
+    pop == 2 ~ 'SJ',
+    pop == 6 ~ 'CA-High'
+  )) %>% dplyr::select(-pop)
+
+# duplicate row for CA site
+x <- cli[2,] 
+x$name <- 'CA-Low'
+cli <- cli %>% bind_rows(x) %>% 
+  mutate_at(1, round, 0) %>% 
+  unite(temp, c(tmin, tmax), sep=" - ")
+
+t1 <- t1 %>% inner_join(cli, by='name') %>% dplyr::select(name, lat_m, long_m, elev_mean, slope_mean, prec, temp, n_trees:stand_density)
+
 colnames (t1) <- c("Name", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
+                   "Prec. (mm)", "Temp ((&deg; C)", 
                    "# trees", '# cores', "Dbh (cm)", "Height (m)", "Age (years)", 

### 2018-03-26 man/ms_1_material_methods.Rmd
add notes about climate

@@
 Tree sampling was carried during autumn of 2016.
-Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ; northern aspect) and Cáñar (CA; southern aspect) (Figure 1; Table 1).
--- GGI: Diferencias microclimáticas Decir algo aquí sobre difrencias microclimáticas entre el norte (más seco, más continental?) y el sur (más húmedo pero más oceánico/cálido?).
--- For the southern site two elevations were sampled: CA-Low and CA-High.
+Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ), xeric site located at northern aspect; and Cáñar (CA), humid but warm site located at southern aspect (Figure 1; Table 1).
+For the southern site two elevations were sampled: CA-Low and CA-High.
 All the sites were oak monospecific and representatives of two of the three population clusters identified for the species in this mountain range [@PerezLuque2015].

### 2018-03-26 man/ms_1_material_methods.Rmd
correct passive and active voice methods

@@
 ## Greenness data
-As a proxy of primary growth, we characterized the vegetation greenness of *Q.
-pyrenaica* by means of the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
-Vegetation indices derived from remote sensing information are linked with primary productivity [@Ruimy1994] and they have been been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009].
-These indices are linearly related to net primary productivity through the fraction of photosynthetically active radiation intercepted by green vegetation (*i.e.* vegetation greenness) [@Ruimy1994].
 
-We used EVI instead of NDVI (*Normalized Difference Vegetation Index*; another of the most common greenness vegetation indices) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].
-
-MODIS EVI Data from Collection 6 were obtanied using Google Earth Engine platform for the period 2000 - 2016.
+Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] and they are linked with primary productivity [@Ruimy1994].
+Here, vegetation greenness of *Q.
+pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
+MODIS EVI Data (Collection 6) were obtanied using Google Earth Engine platform for the period 2000 - 2016.
 EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
@@
 pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
-A data filtering was applied to select EVI valid values.
-The filtering was done using quality flags and VI Usefulness Indices accompanying the EVI data.
-We filter out those values affected by high content of aerosols, clouds, snow and shadows, following the filtering recommendations of EVI data for mountain regions [@Reyes2015].
+Then, a data filtering was applied to select EVI valid values.
+For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags).
+Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations of @Reyes2015 for mountain regions [@Reyes2015].
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+We used EVI instead of NDVI (*Normalized Difference Vegetation Index*; another of the most common greenness vegetation indices) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].
+
+")
+```
 

### 2018-03-26 man/ms_1_material_methods.Rmd
update

@@
 
-After the filter out process, we computed the annual mean EVI values as a surrogate of mean annual primary production, and the EVI anomaly for each pixel for the period 2000 - 2016.
-EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
-We explored the temporal trends of EVI at a pixel scale.
-We computed the nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945), using the `trend` package in R [@Thorsten2017].
-We set 0.05 the alpha level for the test.
+To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016].
+For each pixel, an annual EVI value was obtained by averaging EVI valid values.
+Then, the standardized anomaly was computed as:
+
+$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
+where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
+Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2010; @Samanta2012].
+
+As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016.
+$EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
+For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale.
+The nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945) was applied using the `trend` package in R [@Thorsten2017].
+The alpha level for the test was set to 0.05.
 
@@
 
-To explore the effect of drought events on greenness we calculated the EVI standardized anomaly ($EVI~sa$) pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016].
-For each pixel we annually averaged all the EVI valid values, and then the standardized anomaly was computed as:
-
-$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
-where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year), and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
-Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2010; @Samanta2012].
-
 ## Field sampling and dendrochronological methods
@@
 Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively.
-We assessed stand competition of target trees by recording distance, azimuth, dbh, species and total height of all neighboring living trees with dbh > 7.5 cm in a circular plot of 10 m radius.
-We computed two distance independent indices: *density* in $\mathrm{trees \cdot ha^{-1}}$, and *basal area* (BA) in $\mathrm{m^{2} \cdot ha^{-1}}$; and one distance dependent indices: size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
-We used non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] to analyse differences between sites for height, DBH and competence indices.
+Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius.
+Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
+Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices.
 When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values.
@@
 The growth trends were analyzed at different time scales.
-To study the response of growth to the inter-annual variability of climate (short-term response) we used pre-whitened residual chronologies (RWI) calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976].
+To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used.
+They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976].
 Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008; @Bunn2010] package.
@@
 Theoretically, BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008].
-We used measured dbh and raw ring-widths to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
+Raw ring-widths and measured DBH were used to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
 We used the following equation:
@@
 where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
-We calculated a mean BAI serie for each individual tree.
-Mean site BAI chronologies were obtained by averaging individual tree BAI time series.
+For each individual tree, a mean BAI serie was calculated.
+Then, mean site BAI chronologies were obtained by averaging individual tree BAI time series.
 
@@
 Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018].
-We also used the Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014].
-SPEI values for the period 1961-2014 were obtained with a spatial resolution of 1.1 km from the Drought indices dataset for Spain database (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017].
-We selected a temporal scale for the SPEI of 6 months.
+The Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014], was also used to characterize the drought conditions for the period 1961-2014.
+SPEI values with a temporal scale of 6 months and a spatial resolution of 1.1 km, were obtanied from the Drought indices dataset (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017].
 
@@
 where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median.
-We used medians since they are more robust estimator of central tendency than means [@Rubino2004; @Camarero2011].
+Medians were used since they are more robust estimator of central tendency than means [@Rubino2004; @Camarero2011].
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
@@
 
+Aqui hablar de recopilación de información histórica --RZ
+
+A detailed historical analysis was made on the use changes of the territory in the town of Cáñar
+(Alpujarra, Granada). Documentary sources were reviewed from 1568 to the present (Figure 1) to compile information concerning the distribution
+of vegetation and crops, land uses, and other variables that could be related. A series of variables were selected to describe
+the changes that occurred in the socio-economic dynamics of the town and their ecological consequences:
+
+population, land use (surface area of irrigated crop, rainfed crop, and forest surface area), tree density, sheep raising, beekeeping, and economic sectors (primary, secondary, and tertiary).
+All the variables were gathered at the municipal scale, except for those related to the economic sectors, which were established at provincial level.
+The temporal range of each variable differed according to the data availability.
+After the collection of data, the changes that occurred in the land use and the dynamics of the vegetation were inferred from a multidisciplinar approach (Figure 2).
+
+Historical revisions of
+
+[@MorenoLlorca2016]
+
+D
+
+Available historical statistics and land use maps demonstrate intensive use of this area, in which dry and irrigated crops were intermingled with patches of forest exploited for firewood and acorns (Catastro 1752).
+
 ## Resilience
 
-To evaluate the effects of drougth events on greeennes and tree growth (BAI) we used three resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
+To evaluate the effects of drougth events on greeennes and tree growth (BAI) resilience indices proposed by @Lloret2011 were used: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
 The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred.
@@
 
-We computed the values of these indices for tree growth (BAI) and greenness during each drought event.
-We considered 2005 and 2012 as single drought events.
+The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event.
+2005 and 2012 were considered as single drought events.
 The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
 A period of three years was chosen because we found similar results comparing periods of two, three and four years (Appendix S6) and this length was used in other studies [*e.g.* @Gazol2017].
-In addition we calculated the same indices for BAI data for the most severe drought events (Appendix S3) since 1900.
+To put in context the resilience observed for 2005 and 2012 drought events, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 (Appendix S3).
 
@@
 
-We used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects) for each of the three resilience indices studied.
-These test were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012].
+For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects).
+These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012].
 Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012].

### 2018-03-27 man/ms_1_material_methods.Rmd
update references and methods (historical data)

@@
 
-Aqui hablar de recopilación de información histórica --RZ
-
-A detailed historical analysis was made on the use changes of the territory in the town of Cáñar
-(Alpujarra, Granada). Documentary sources were reviewed from 1568 to the present (Figure 1) to compile information concerning the distribution
-of vegetation and crops, land uses, and other variables that could be related. A series of variables were selected to describe
-the changes that occurred in the socio-economic dynamics of the town and their ecological consequences:
-
-population, land use (surface area of irrigated crop, rainfed crop, and forest surface area), tree density, sheep raising, beekeeping, and economic sectors (primary, secondary, and tertiary).
-All the variables were gathered at the municipal scale, except for those related to the economic sectors, which were established at provincial level.
-The temporal range of each variable differed according to the data availability.
-After the collection of data, the changes that occurred in the land use and the dynamics of the vegetation were inferred from a multidisciplinar approach (Figure 2).
-
-Historical revisions of
-
-[@MorenoLlorca2016]
-
-D
-
-Available historical statistics and land use maps demonstrate intensive use of this area, in which dry and irrigated crops were intermingled with patches of forest exploited for firewood and acorns (Catastro 1752).
+In addition, the forest history of San Juan and Cáñar sites was inferred from a detailed analysis of historical land-use changes.
+For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc.
+Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].
 

### 2018-03-27 man/ms_1_material_methods.Rmd
mas update

@@
 For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc.
-Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].
+Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].
 

### 2018-03-27 man/ms_2_results.Rmd
update results

@@
 
-The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant.
-The lowest values of EVI standardized anomalies were recorded in 2005, being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2).
-During 2005 drought, *Q.
-pyrenaica* forests suffered a "browning episode" ($EVI~sa < -1$) (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Appendix S5).
+The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period.
+During the 2005 the lowest values of EVI standardized anomalies were recorded, particularly for northern populations (Figure 2).
+In fact, a "browning" episode ($EVI~sa < -1$) was observed for *Q.
+pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Appendix S4).
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 *Q.
 pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3).
-Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figure 3) .
+Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figure 3).
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
+Recovery (*Rc*) of greenness was significnatly different bewteen drought events and sites (Tables 3).
+In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
+For southern populations, a similar recovery after the 2005 and 2012 drought event was observed (p = 0.2453; Figure 3; Table S1).
+
 > Resistencia
@@
 ```
-Recovery (*Rc*) of greenness was significnatly different bewteen drought events and sites (Tables 3).
-In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
-For southern populations, a similar recovery after the 2005 and 2012 drought event was observed (p = 0.2453; Figure 3; Table S1).
 
@@
 Yet the basal area increment (BAI) growth in the highest plot in altitude (CA-High) was the greatest (Figure 4).
-Trees from this location showed in addition a positive growth trend since the late 1970s.
-This positive trend was not found in any of the other two locations and the two altitudes sampled in the southern populations showed differences in growth, being more similar CA-Low with growth in the northern population (SJ) than with CA-High (Figure 4; Appendix S11).
+Trees from this location showed in addition a positive growth trend since the late 1970s (Figure 4).
+This positive trend was not found in any of the other two locations and the two altitudes sampled in the southern populations showed differences in growth, being more similar CA-Low with growth in the northern population (SJ) than with CA-High (Figure 4; Appendix S7).
 
-1995 was the lowest pointer year (*i.e.* the lowest growth) since 1950 in all chronologies (Figure 5).
 For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low).
+Yet, the lowest pointer year (*i.e.* the lowest growth) since 1950 was 1995 in all chronologies (Figure 5).
 
@@
 
-## Resilience of radial tree-growth to drought events
+## Resilience of greenness and radial tree-growth to drought events. (VALIDO)
 
-Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
-Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).
+*Q.
+pyrenaica* forest showed significantly lower resistance values (*Rt*) to the 2005 drought event than to that in 2012 for greenness and for radial growth (Table 3; Figure 3).
+The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2).
+Resistance values to drought for greenness and tree-growth varied between sites (Table 3).
+Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2).
+It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).
+
+Recovery (*Rc*) of greenness and growth were significantly different bewteen drought events and sites (Tables 3).
+In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
+An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 3, Table S2).
+Northern populations showed significantly higher values of recovery than southern sites for greenness and tree-growth (Table S2).
+For southern populations, no significant differences were found for recovery of greenness after the 2005 and 2012 drought event (p = 0.2453; Figure 3; Table S1).
+Recovery values for tree-growth of southern populations were below or close to 1 (Figure 3, Table S2).
 
@@
 > Resistance:
+
+Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
+Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).
 
@@
 ```
-
-Higher values of recovery (*Rc*) were observed after 2012 than after 2005 drougth events, staying at levels of *Rc* < 1 (Figure 3, Table S2).
-SJ site (northern aspect) showed significantly higher values of recovery than southern sites (CA-High and CA-Low) (Tables 3 and S2), which even not reach recovery values of Rc = 1 (Figure 3).
 

### 2018-03-28 man/ms_2_results.Rmd
updating results

@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 ## Greenness resilience to drought events
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
 *Q.
@@
 For southern populations, a similar recovery after the 2005 and 2012 drought event was observed (p = 0.2453; Figure 3; Table S1).
+
+Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables S1, 3), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
+The southern populations showed higher resilience values than the northern ones, although these differences were not significant for the 2005 drought event (p = 0.036; Figure 3).
 
@@
 drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)
-")
-```
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
 > Recovery 
@@
 En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
-")
-```
 
-Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables S1, 3), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
-The southern populations showed higher resilience values than the northern ones, although these differences were not significant for the 2005 drought event (p = 0.036; Figure 3).
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-> Resilience 
 > * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
@@
 En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
-")
-```
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
 > Relative Resilience (De esto no he dicho nada)
@@
 
-## Radial growth trends and climate
+## Radial growth trends
 
@@
 cat("
-> Relative Resilience (De esto no he dicho nada)
-
 Sorprende que en los sitios del sur (caH y caL), en los últimos años ha habido crecimientos muy grandes.
@@
 
-## Resilience of greenness and radial tree-growth to drought events. (VALIDO)
+## Resilience of greenness and radial tree-growth to drought events
 
@@
 Recovery values for tree-growth of southern populations were below or close to 1 (Figure 3, Table S2).
+
+Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 3).
+Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3).
+Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2).
+For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 3).
+For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Tab
+
+The exploration of resilience metrics of tree-growth for other severe droughts, revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S6).
+A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155).
+Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S6) even not being the most severe drought events.
 
@@
 De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ???
-")
-```
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
 > Recovery 
@@
 > * Las poblacionees del sur (caH y caL; no diferencias entre ellas) mostraron una menor recuperación que las poblaciones del norte, es decir, las del norte se recuperaron mas rápidamente que las del sur que no llegaron incluso a valores de Rc = 1.
-") 
-```
 
@@
 
-When we explored resilience values of growth (*Rs*) for other severe droughts, we found a relation between the resilience and drought severity ($r^2$ = 0.336) (Appendix S6).
+") 
+```
 
@@
 ## Forest disturbances
-The analysis of growth changes revealed differences forest history between sites (Figures 7 and 8).
+The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8).
 Northern site (SJ) showed two release events ($GC$ > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001.

### 2018-03-28 analysis/bai/trends_resilience_bai.Rmd
new update

@@
   inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+s %>% write.csv(file = here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), row.names = FALSE) 
 ```

### 2018-03-28 man/ms_figures.Rmd
new update

@@
 
-#####
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
 **Figure 1**.
@@
 
-#####
+![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
+
+##### 
+**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
 
@@
 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-##### 
+##### 
+**Figure 3.** Response *Q.
+pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
+For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
+For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
+Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
+
 ```{r, eval=FALSE, echo=FALSE}
@@
 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
 ```{r, eval=FALSE}
@@
 ##### 
+**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
+pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
+Shading areas coorespond to standard error of the mean.
+Number of series are displayed in the upper plot.
+We only show chronologies with # trees > 5.
+Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r2`).
 
@@
 
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r2`).
-
-#####
+##### 
+**Figure 5.** Residual Tree-ring chronologies obtained for the *Q.
+pyrenaica* sites.
+Dashed red lines indicate the start of the reliable period (EPS > 0.85).
+Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
 
@@
 
-**Figure 5.** Residual Tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-##### 
+##### 
+**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
+pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
+*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site; and *dark blue* bars: high-elevation shouthern site.
+Asteriks indicate significant ($P < 0.05$) correlation coefficients.
+
 ```{r}
@@
 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site; and *dark blue* bars: high-elevation shouthern site.
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-#####
+#####
+**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
 
@@
 
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-#####
+##### 
+**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
 
@@
 ```
-
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-# Supplementary
-
-#####
-
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017 for 28 meteorological stations (from National Spanish Meteorological Services, AEMET) distributed around the Sierra Nevada area.
+##### 
+**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
 Points represent mean and errorbars standard error.
 *Black* line indicates mean for all period.
-*Red* lines represent -1 (*dotted* line) and -2 (*dashed* line) standard deviation.
-*Blue* lines represent +1 (*dotted* line) and +2 (*dashed* line) standard deviation.
-We labelled all years with average values below -1SD.
+*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
+*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
+Years with average values below -1SD are labelled.
+Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
 ***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
 The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea (National Spanish Meteorological Services, AEMET).
+Data from meteorological station Granada, Base Aérea.
 
@@
   theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black")) + 
+  theme(strip.background = element_rect(fill = NA, color = "black"),
+        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
+        axis.text.y = element_text(size=7),
+        axis.title.y = element_text(size = 7)) + 
   ylab('Cumulative Precipitation (mm)')  + xlab('') + 
@@
                             "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 4, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 4, colour = "red") + 
-  annotate("text", label = "mean 1950-2015", x = 3, y = 600, size = 4, colour = "black") +
+  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
+  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
+  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
   theme(panel.grid = element_blank())
@@
     data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear)) +
+    aes(label = hyear), 
+    size = 3) +
   theme(panel.grid.minor = element_blank()) +
@@
 ##### 
+**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
+A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
+A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g* @Spinoni2015; Spinoni2017a; @Pascoa2017].
+The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
+***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
+***Intensity*** and ***Lowest SPEI** refer to the mean and lowest value of SPEI respectively during the drought event duration.
+
 ```{r}
@@
 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g* @Spinoni2015; Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
 ##### 
@@
 ##### 
-
-**Appendix S5.** Reilience vs. droughts
+**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
+*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
+Points indicate average of resilience metrics for all populations.
+Error bar corresponds standard error.
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
+
+```{r, warning=FALSE, out.extra='angle=90', fig.height=5}
+# see trends-resilience_bai.Rmd 
+df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
+
+#https://gist.github.com/ottadini/6882677 
+lm_eqn = function(m) {
+  # Displays regression line equation and R^2 value on plot
+  # Usage:
+  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
+
+  l <- list(#a = format(coef(m)[1], digits = 2),
+            # b = format(abs(coef(m)[2]), digits = 2),
+            r2 = format(summary(m)$r.squared, digits = 3));
+
+  if (coef(m)[2] >= 0)  {
+    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
+    eq <- substitute(~~italic(r)^2~"="~r2,l)
+  } else {
+    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
+    eq <- substitute(~~italic(r)^2~"="~r2,l)
+  }
+
+  as.character(as.expression(eq));                 
+}
+
+sdf <- df %>% 
+  filter(samps >= 10) %>% 
+  group_by(disturb_year) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
+  require(ggplot2)
+  require(ggrepel)
+
+  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
+
+  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
+                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
+
+  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
+    geom_smooth(method='lm', se=FALSE, colour="black") +
+    geom_errorbar(mapping = yerrbar) + 
+    geom_point(size=4, pch = 21, colour="black", fill="white") + 
+    theme_bw() + 
+    xlab(XLab) + ylab(YLab) + 
+    geom_text_repel() + 
+    annotate("text", x=coordX_eq, y = coordY_eq, 
+             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
+    ggtitle(titulo) + 
+    theme(plot.title = element_text(size = 14, 
+                                    face = "bold.italic",
+                                    margin = margin(b = -15)), 
+          axis.title.y = element_blank())
+  }
+
+sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
+        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
+
+sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Resistance (Rt) (3year window size)",
+        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
+
+sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
+        XLab = "Drought Severity", YLab = "Recovery (Rc) (3year window size)",
+        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
+
+sevRt + sevRc + sevRs
+
+```
 
@@
 g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, tpe='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
+g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
 
@@
 ```
+
+##### 
+**Appendix S7.** Growth commparations
 

### 2018-04-02 man/ms_3_discussion.Rmd
udpate
update text discusion

@@
 
+1. 
+Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction in greenness and especially in the growth of *Q.
+pyrenaica*.
+During the 2005 drought, one of the worst drought events recorded in the Iberian Peninsula [@GarciaHerrera2007], we found a browning of the *Q.
+pyrenaica* forests.
+Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure 5).
+This common signal observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].
+
+Although 2005 and 2012 were two of the most severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), we found a positive trend for greenness of the forests of *Q.
+pyrenaica* located in their rear edge, as previously was reported using remote sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi].
+For tree-growth a similar pattern was observed in the last decades, particularly for the soouthern high-elevation site (CA-High, Figure 4).
+UNIR Ahora esto: Aqui hablar algo de las proyecciones de crecimiento de Gea-Izquierdo
+
+---> Hablar aqui de resiliencia en el rear edge.
+
+Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia.
+Por ejemplo para las poblaciones del norte, los valores de EVI durante la sequía de 2005 descendieron hasta el 81 %, mientras que el BAI (sitio SJ) lo hizo hasta el 45 %.
+Sin embargo, la recuperación fue rápida, así los valores de Recovery (Rc) para el EVI tras la sequía de 2005 (en las poblaciones del norte) fue de 1.17, mientras que para el BAI el Rc fue de 1.112.
+En definitiva, estamos observando altos valores de resiliencia en estos robledales.
+
+Los valores de resiliencia observados para el greenness (cercanos a 1) muestran como esta especie, en su rear edge muestra una buena capacidad de recuperación frente a eventos extremos,
+
+Comentar algo de resiliencia a largo plazo (otras valores de resiliencia)
+
+¿A que se debe esa resiliencia?
+
+¿Casos parecidos?
+
+Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling..
+// dorado-liñan // sanchez salguero
+
+trabajos rear edge: 
+* sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
+* sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
+* dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
+* herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
+* dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
+buscar otros
+
+pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.
+
+## refugio??
+
+Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático.
+Hemos observado un aumento en el greenness en los últimos años.
+Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness.
+Además, estos robledales tienen una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
+¿Sierra Nevada (regiones de montaña) como refugio??
+quizá este rear-edge esté actuando como refugio??
+(esto es muy especulativo)
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 Hemos observado como las sequías severas, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento de *Q.
@@
 
+La sequía de 2005 ha sido una de las mas severas de los últimos años, afectando significativamente al crecimiento
+
+#--------------------------------------------
+# Sequías
+
+### 2005 frente a 2012
+
+* Analizando los datos de precipitación acumulada de las estaciones alrededor de SN, observamos como 2005 fue el año mas seco desde 1950. 
+* Datos de SPEI (0.5 º Grid) 
+    * 2005 fue peor que 2012, pero no peor que 1995 o que 1999 (ver figura S2)
+    * Tabla S4. Severidad sequías, la de 2005 fue menos severa y menos intensa y de menor duración que la de 2005 (de donde vienen estos datos???)
+
+* La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
+Esto se ha observado también en otras especies en el sur de la P.
+Ibérica (p.ej.
+P.
+nigra en Andalusia [@SanchezSalguero2013], ...
+INCLUIR otras especies y citas).
+
+### 2005 en contexto
+
+* Algunos trabajos hablan que los peores años de sequía en la P. Iberica durante la segunda mitad del siglo XX fueron 1986, 1994-1995, 1999, 2005 y 2012
+
+* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
+
+### Reducción en el crecimiento de algunas especies 
 El efecto de la sequía de 2005 en el crecimiento también se ha visto en otras especies en el sur de la P. Ibérica 
@@
 
+### La resiliencia de las poblaciones de roble de Sierra Nevada.
+
+El greennness de los robledades en Sierra Nevada ha mostrado una tendencia positiva hacia mas verdor en los ultimos años, que coincide con lo que ya observamos con datos de NDVI [@PerezLuque2015onto], sobre todo para las poblaciones del sur.
+Esta tendencia positiva, obtenida de variables derivadas de remote sensing (EVI y NDVI), parece que también se observa en el crecimiento.
+Por ejemplo en las poblaciones del sur (CaHigh y CaLow) observamos en los últimos años una ligera tendencia de crecimiento.
+
 Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia.
@@
 En definitiva, estamos observando altos valores de resiliencia en estos robledales.
-
-Los valores de resiliencia observados para el greenness (cercanos a 1) muestran como esta especie, en su rear edge muestra una buena capacidad de recuperación frente a eventos extremos,
-
-¿A que se debe esa resiliencia?
-
-¿Casos parecidos?
+")
+```
 
@@
 
-Environmental and forest management policies should take into account all these characteristics
-of Mediterranean forests and the social, environmental and climatic conditions that are projected
-for the coming years and decades.
+Environmental and forest management policies should take into account all these characteristics of Mediterranean forests and the social, environmental and climatic conditions that are projected for the coming years and decades.
 
@@
 **which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]**.
+
 #### Disturbance
@@
 
-## refugio??
-
-Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático.
-Hemos observado un aumento en el greenness en los últimos años.
-Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness.
-Además, estos robledales tienen una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
-¿Sierra Nevada (regiones de montaña) como refugio??
-quizá este rear-edge esté actuando como refugio??
-(esto es muy especulativo)
-
 ## MAs cosas diferentes 
@@
 ### Otras
-Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling..
-// dorado-liñan // sanchez salguero
-
-trabajos rear edge: 
-* sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
-* sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
-* dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
-* herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
-* dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
-buscar otros
 
@@
 Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
-
----- old 
-### Efectos de la sequía en el greenneess y en el crecimiento.
-
-Hemos observado como las sequía, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento.
-Los datos de anomalías estandarizadas de EVI mostraron un browning para la mayoría de los robledales de Sierra Nevada durante la sequía de 2005.
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-* Si analizamos el plot de trajectorias (evi medio anual), vemos como en 2005 y 2012 se observó un browning para los valores medios de EVI.
-En ambos casos, se observó un patrón homogéneo de browing, aunque en 2012 el browning fue mucho menor
-")
-```
 
@@
 
-# Años de sequía
-
-### 2005 frente a 2012
-
-* Analizando los datos de precipitación acumulada de las estaciones alrededor de SN, observamos como 2005 fue el año mas seco desde 1950. 
-* Datos de SPEI (0.5 º Grid) 
-    * 2005 fue peor que 2012, pero no peor que 1995 o que 1999 (ver figura S2)
-    * Tabla S4. Severidad sequías, la de 2005 fue menos severa y menos intensa y de menor duración que la de 2005 (de donde vienen estos datos???)
-
-* La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
-Esto se ha observado también en otras especies en el sur de la P.
-Ibérica (p.ej.
-P.
-nigra en Andalusia [@SanchezSalguero2013], ...
-INCLUIR otras especies y citas).
-* 2004/2005 hydrological year is considered one of the worst drooughts ever recorded in the Iberian Peninsula, particularly in the central and southern sectors [@GarciaHerrera2007].
-* The southern half of Iberia received less than 45 % of the usual precipitation between October 2004 and June 2005 [@GarciaHerrera2007].
-* The hydrological year from October 2004 to September 2005 was the driest on record at several locations throughout Iberia [@GarciaHerrera2007].
-
-### 2005 en contexto
-
-* Algunos trabajos hablan que los peores años de sequía en la P. Iberica durante la segunda mitad del siglo XX fueron 1986, 1994-1995, 1999, 2005 y 2012
-
-* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
-
 ## 

### 2018-04-02 man/ms_3_discussion.Rmd
update discussion

@@
 
-Although 2005 and 2012 were two of the most severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), we found a positive trend for greenness of the forests of *Q.
+Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), we found a positive trend for greenness of the forests of *Q.
 pyrenaica* located in their rear edge, as previously was reported using remote sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi].
-For tree-growth a similar pattern was observed in the last decades, particularly for the soouthern high-elevation site (CA-High, Figure 4).
-UNIR Ahora esto: Aqui hablar algo de las proyecciones de crecimiento de Gea-Izquierdo
+For tree-growth, a similar pattern was observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
+These results differ from those of @GeaIzquierdo2014 that found a decrease or stabilization in BAI since 1970.
+Notwithstanding the vulnerability to warming suggested by @GeaIzquierdo2014 for *Q.
+pyrenaica* forests located at their rear-edge, our results for the same locations, point out a recovery in tree-growth in the last years (Figure 4).
+
+Despite the resilience values (*Rs*) of growth for 2005 and 2012 were close or below 1, we observed high resilience of these forests from a long-term perspective.
+As our results indicate, the highest resilience values for growth were recorded for two of the most severe drought events (1995 and 1999), with resilience values above 1.2 in both events (Appendix S5).
+
+Comentar algo de que los valores de resiliencia de 2012 fueron mayores que 2005.
+Ojo, para greenness no podemos comprobar la resiliencia a largo plazo, pero en el caso de growth si podemos hacerlo.
+y hemos visto que ante eventos de sequía muy severos, estas poblaciones están mostrando una alta resiliencia.
 
@@
 
-pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.
+Otros estudios que analizan la resiliencia de especies arbóreas frente a los eventos de sequía indican:
 
@@
 (esto es muy especulativo)
+
+pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.
 

### 2018-04-05 man/ms_3_discussion.Rmd
update discussion

@@
 
-Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), we found a positive trend for greenness of the forests of *Q.
-pyrenaica* located in their rear edge, as previously was reported using remote sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi].
-For tree-growth, a similar pattern was observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
-These results differ from those of @GeaIzquierdo2014 that found a decrease or stabilization in BAI since 1970.
-Notwithstanding the vulnerability to warming suggested by @GeaIzquierdo2014 for *Q.
-pyrenaica* forests located at their rear-edge, our results for the same locations, point out a recovery in tree-growth in the last years (Figure 4).
+Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), we found a positive trend for vegetation greenness of the forests of *Q.
+pyrenaica* located in their rear edge.
+Our result agrees with those obtained by previous works using other remote sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi], which suggests an increase in primary productivity during the last years for rear-edge populations of this species.
+
+For tree-growth, a similar positive pattern was observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
+This result differs from those previous reported for *Q.
+pyrenaica* along their distribution range [@GeaIzquierdo2013; @GeaIzquierdo2014].
+[@GeaIzquierdo2014] found a general decline in the growth of this species since the 1970s, particularly sharp for rear-edge populations.
+This decline trend in growth have also been oberved for other tree species located in their rear-edges [*e.g.* @SanchezSalguero2012; @SanchezSalguero2017; @Dorado2017].
+Furthermore, growth projections have forecasted a decrease in productivity for *Q.
+pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013].
+However, our results for similar locations have revealed a recovery in tree-growth for the last years (Figure 4).
+
+Las zonas rear edge son zonas que se supone son mas vulnerables al cambio climático por sus estrechos márgenes.
+De hecho como han visto algunos trabajos, las especies situadas en su rear edge son mas vulnerables.
+No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado).
+Además habría que introducir aqui lo que M.
+Villalta habla de la consideración de rear edge de.
+
+Cuando miramos a la resiliencia de estos bosques podemos ver como aunque en los últimos años ...
+
+Este aumento de BAI, y de greenness no se refleja tanto en la resiliencia observada en los ultimos años. No obstante, n
 
@@
 y hemos visto que ante eventos de sequía muy severos, estas poblaciones están mostrando una alta resiliencia.
+
+¿porque? clima + temperatura ...??
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 ¿Casos parecidos?
@@
 buscar otros 
-
-Otros estudios que analizan la resiliencia de especies arbóreas frente a los eventos de sequía indican:
+")
+```
 
@@
 Evaluar el crecimiento en el rear edge de F. sylvatica 
-
 https://link.springer.com/article/10.1007/s10342-016-0982-7?wt_mc=Internal.Event.1.SEM.ArticleAuthorOnlineFirst

### 2018-04-10 man/ms_3_discussion.Rmd
update discussion

@@
 
-1. 
 Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction in greenness and especially in the growth of *Q.
@@
 During the 2005 drought, one of the worst drought events recorded in the Iberian Peninsula [@GarciaHerrera2007], we found a browning of the *Q.
-pyrenaica* forests.
+pyrenaica* forests, but no changes in EVI standardized anomalies were recorded for 2012, which can be explained because 2012 drought event was a winter-drought [@Trigo2013].
 Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure 5).
@@
 
-Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), we found a positive trend for vegetation greenness of the forests of *Q.
+Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Appendix S3), we found a positive trend for vegetation greenness of the forests of *Q.
 pyrenaica* located in their rear edge.
@@
 
-For tree-growth, a similar positive pattern was observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
+For tree-growth, a similar positive trend was observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
 This result differs from those previous reported for *Q.
 pyrenaica* along their distribution range [@GeaIzquierdo2013; @GeaIzquierdo2014].
-[@GeaIzquierdo2014] found a general decline in the growth of this species since the 1970s, particularly sharp for rear-edge populations.
+[@GeaIzquierdo2014] found a general decline in the growth of this species since the 1970s, particularly sharp for populations located in their dry-edge.
 This decline trend in growth have also been oberved for other tree species located in their rear-edges [*e.g.* @SanchezSalguero2012; @SanchezSalguero2017; @Dorado2017].
@@
 
-Las zonas rear edge son zonas que se supone son mas vulnerables al cambio climático por sus estrechos márgenes.
-De hecho como han visto algunos trabajos, las especies situadas en su rear edge son mas vulnerables.
-No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado).
-Además habría que introducir aqui lo que M.
-Villalta habla de la consideración de rear edge de.
+Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 3).
+La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Appendix S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q.
+pyrenaica*.
+Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía, observamos como estas poblaciones muestran altos valores de resiliencia.
+Como muestrasn nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).
 
-Cuando miramos a la resiliencia de estos bosques podemos ver como aunque en los últimos años ...
+Desde una perspectiva a largo plazo, nuestros resultados muestran que las poblaciones de *Q.
+pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia.
+Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005].
+Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas).
+Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q.
+pyrenaica* situadas en su rear-edge.
 
-Este aumento de BAI, y de greenness no se refleja tanto en la resiliencia observada en los ultimos años. No obstante, n
-
-Despite the resilience values (*Rs*) of growth for 2005 and 2012 were close or below 1, we observed high resilience of these forests from a long-term perspective.
-As our results indicate, the highest resilience values for growth were recorded for two of the most severe drought events (1995 and 1999), with resilience values above 1.2 in both events (Appendix S5).
-
-Comentar algo de que los valores de resiliencia de 2012 fueron mayores que 2005.
-Ojo, para greenness no podemos comprobar la resiliencia a largo plazo, pero en el caso de growth si podemos hacerlo.
-y hemos visto que ante eventos de sequía muy severos, estas poblaciones están mostrando una alta resiliencia.
-
-¿porque? clima + temperatura ...??
-
----> Hablar aqui de resiliencia en el rear edge.
-
-Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia.
-Por ejemplo para las poblaciones del norte, los valores de EVI durante la sequía de 2005 descendieron hasta el 81 %, mientras que el BAI (sitio SJ) lo hizo hasta el 45 %.
-Sin embargo, la recuperación fue rápida, así los valores de Recovery (Rc) para el EVI tras la sequía de 2005 (en las poblaciones del norte) fue de 1.17, mientras que para el BAI el Rc fue de 1.112.
-En definitiva, estamos observando altos valores de resiliencia en estos robledales.
-
-Los valores de resiliencia observados para el greenness (cercanos a 1) muestran como esta especie, en su rear edge muestra una buena capacidad de recuperación frente a eventos extremos,
-
-Comentar algo de resiliencia a largo plazo (otras valores de resiliencia)
-
-¿A que se debe esa resiliencia?
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-¿Casos parecidos?
-
-Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling..
-// dorado-liñan // sanchez salguero
-
-trabajos rear edge: 
- * sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
- * sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
- * dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
- * herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
- * dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
-buscar otros 
-")
-```
-
-## refugio??
-
-Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático.
-Hemos observado un aumento en el greenness en los últimos años.
-Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness.
-Además, estos robledales tienen una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
-¿Sierra Nevada (regiones de montaña) como refugio??
-quizá este rear-edge esté actuando como refugio??
-(esto es muy especulativo)
-
-pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.
+Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018].
+En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q.
+pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.
+Esta idea se vería reforzada por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
+pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
 
@@
 
-### La resiliencia de las poblaciones de roble de Sierra Nevada.
-
-El greennness de los robledades en Sierra Nevada ha mostrado una tendencia positiva hacia mas verdor en los ultimos años, que coincide con lo que ya observamos con datos de NDVI [@PerezLuque2015onto], sobre todo para las poblaciones del sur.
-Esta tendencia positiva, obtenida de variables derivadas de remote sensing (EVI y NDVI), parece que también se observa en el crecimiento.
-Por ejemplo en las poblaciones del sur (CaHigh y CaLow) observamos en los últimos años una ligera tendencia de crecimiento.
-
 Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia.
@@
 En definitiva, estamos observando altos valores de resiliencia en estos robledales.
+
+Despite the resilience values (*Rs*) of growth for 2005 and 2012 were close or below 1, we observed high resilience of these forests from a long-term perspective.
+As our results indicate, the highest resilience values for growth were recorded for two of the most severe drought events (1995 and 1999), with resilience values above 1.2 in both events (Appendix S5).
+
+genetica
+
+Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático.
+Hemos observado un aumento en el greenness en los últimos años.
+Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness.
+Además, estos robledales tienen una alta resiliencia genética  ¿Sierra Nevada (regiones de montaña) como refugio??
+quizá este rear-edge esté actuando como refugio??
+(esto es muy especulativo)
+
+Las zonas rear edge son zonas que se supone son mas vulnerables al cambio climático por sus estrechos márgenes.
+De hecho como han visto algunos trabajos, las especies situadas en su rear edge son mas vulnerables.
+No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado).
+Además habría que introducir aqui lo que M.
+Villalta habla de la consideración de rear edge de.
+
+* Otro punto de interés a introducir aquí es que frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], sin embargo algunos estudios están demostrando esto no es siempre así (ver por ejemplo [@Cavin2017; @Granda2017]).
+Nuestros resultados creo que van en esta línea creo.
+Esto, como apunta [@MartinezVilalta2018]m tiene que ver con la que consideramos como habitat marginal de la especie (...
+When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).
+
+¿Casos parecidos?
+
+Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling..
+// dorado-liñan // sanchez salguero
+
+trabajos rear edge: 
+ * sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
+ * sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
+ * dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
+ * herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
+ * dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
+buscar otros
+
+...
+Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
+
+## refugio??
+
+Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático.
+Hemos observado un aumento en el greenness en los últimos años.
+Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness.
+Además, estos robledales tienen una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
+¿Sierra Nevada (regiones de montaña) como refugio??
+quizá este rear-edge esté actuando como refugio??
+(esto es muy especulativo)
+
+pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.
+
 ")
@@
 Encontramos diferencias entre poblaciones. Asi por ejemplo en las situadas mas al sur, vemos que las anomalias de EVI son menos negativas para el 2005; También vemos tendencias positivas en el BAI...
+
+No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado).
 
@@
 
-* Otro punto de interés a introducir aquí es que frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], sin embargo algunos estudios están demostrando esto no es siempre así (ver por ejemplo [@Cavin2017; @Granda2017]).
-Nuestros resultados creo que van en esta línea creo.
-Esto, como apunta [@MartinezVilalta2018]m tiene que ver con la que consideramos como habitat marginal de la especie (...
-When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).
-
 En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations).
 The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone.
-
-Otra cosa.
-@GeaIzquierdo2014 encontró que a lo largo del gradiente altitudinal el BAI se estabilizó o disminuyó a partir de 1970.
-DUDA: Nosotros encontramos un ligero (y débil) aumento, sobre todo en caHigh.
 
@@
 
-...
-Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
-
 ---------------------

### 2018-04-11 man/ms_3_discussion.Rmd
update

@@
 
-Sin embargo el estudio de las poblaciones situadas en su rear edge requiere de una aproximación combinada donde se estudie la respuesta de las poblaciones al cambio climático y su resiliencia utiliznado diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].
-
 .. ver [@VicenteSerrano2013] impacto de las sequías en diferentes zonas (i.e montañas)
@@
 * To guide sustainable forest management, forest researchers are asked to provide concrete answers about forest resilience in response to expected climatic trends, and extreme climatic events (Lindner et al., 2014) https://www.sciencedirect.com/science/article/pii/S030147971400379X?via%3Dihub
-
-* Hablan de incorporar ademas de RS, el conocimiento ecológico (los principios ecológicos) para comprender las consecuencias de las disturbances analizadas con RS, y sobre todo para predecir futuro.
-(McDowell et al https://www.sciencedirect.com/science/article/pii/S1360138514002726)
 
@@
 
-# METODOLOGY
+# 4) Punto fuerte metodológico (VALID)
 
-## metodologico
+La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía en la vegetación se pueden ver tanto con información de satélite (hablar del browning), como con información mas de campo (hablar aquí de lo del BAI).
+
+Nuestros resultados
+
+La combinación de metodologías que hemos empleado para evaluar la resiliencia de la vegetación frente a los eventos de sequía
+
+La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía en la vegetación se pueden ver tanto con información de satélite, como con información mas de campo (hablar aquí de lo del BAI).
+
+Los efectos de la sequía sobre la vegetación pueden observarse usando información procedente de diferentes aproximaciones metodológicas [see @Norman2016 for a review].
+
+[información procedente de remote sensing (RS)[@Zhang2013; @AghaKouchak2015], información dendrochronologica (),
+
+o combinando estas y otras metodologías [see @Norman2016 for a review]
+
+respuesta de
+
+The effects of drought on vegetation can be observed using information from several methodological approaches [see @Norman2016 for a review].
+Several works revealed the utility of remote-sensing or dendrochronology to quantify the dorught impacts on vegetation at different spatial and temporal scales ().
+
+Exiten trabajos que han utilizado remote-sensing o la dendrochronology para analizar los efectos de las sequías sobre la vegetación a diferentes escalas espaciales y temporales ( ).
+Asimismo, algunos otros trabajos han utilizado conjuntamente estas dos metodologias, y han visto que existe de forma general una relació
+a positive relationship between vegetation indices derived from remote sensing and annual tree growth [@VicenteSerrano2016; @Bhuyan2017b],
+
+In general, a positive relationship between NDVI in the growing season and RWI was seen for many forests in our study.
+
+Positive relationship between inter-annual NDVI variability and annual tree growth [@VicenteSerrano2016]
+
+Esta relación positiva también la encuentra
+
+>>>>
+* Hablan de incorporar ademas de RS, el conocimiento ecológico (los principios ecológicos) para comprender las consecuencias de las disturbances analizadas con RS, y sobre todo para predecir futuro.
+(McDowell et al https://www.sciencedirect.com/science/article/pii/S1360138514002726)
+
+Nuestro trabajo muestra una combinación de métodos
+
+Finalmente, destacamos la aproximación metodológica que hemos llevado a cabo.
 
 La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía se pueden ver tanto con información de satélite (hablar del browning), como con información mas de campo (hablar aquí de lo del BAI).
-Indicar también las diferencias entre las resiliencias detectadas por ambas metodologías.
 
-Nuestro trabajo combinamos las dos aproximaciones metodológicas por varias razones:
+# Relaciones RWI-RS 
+* Algunos trabajos han analizado la relación ente RWI y remote sensing [@VicenteSerrano2013]; [(Bhuyan et al.
+Remote Sensing 9(6):526)]; [Vicente Serrano et al.
+https://www.sciencedirect.com/science/article/pii/S003442571630373X] ...
 
+## Trabajos RWI-RS 
+*
+
+# Diferencias que hemos detectado usando RS y RWI (en resiliencias)
+* Relacionar con datos de Gazol et al. 2018
+
+# Jusfiticación de uso de ambos métodos
+Se trata de expresar la idea de mezcla de aproximacioes: remote-sensing con dendro.
+Aunque existen algunos trabajos que utilizado RWI y remote sensing nuestra aproximación es importante, sobre todo para estudiar poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
+De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
+
+Sin embargo el estudio de las poblaciones situadas en su rear edge requiere de una aproximación combinada donde se estudie la respuesta de las poblaciones al cambio climático y su resiliencia utiliznado diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].
+
+## De forma genérica y para el rear-edge 
+* Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares...
+
+* Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q.
+pyrenaica situadas en su reaar edge frente a los eventos de sequía.
+
+Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
+(ver esto y completar)
+
+## Sensibilidades diferentes de cada metodología 
 * Cada metodología muestra distinta sensibilidad
@@
 
-* Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares...
+* La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017]
 
-* La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017]
+end VALID 
+---------------------
 
@@
 Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes.
-
-### Notas sobre metodologia
-
-* Algunos trabajos han analizado la relación ente RWI y remote sensing [@VicenteSerrano2013]; [(Bhuyan et al.
-Remote Sensing 9(6):526)]; [Vicente Serrano et al.
-https://www.sciencedirect.com/science/article/pii/S003442571630373X] ...
-
-* Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q.
-pyrenaica situadas en su reaar edge frente a los eventos de sequía.
-
---- notas para esto 
-Se trata de expresar la idea de mezcla de aproximacioes: remote-sensing con dendro.
-Aunque existen algunos trabajos que utilizado RWI y remote sensing nuestra aproximación es importante, sobre todo para estudiar poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
-De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
-
-Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
-(ver esto y completar)
 

### 2018-04-16 analysis/analysis_age_elevation.Rmd
create plots maps elevation age

---
title: "Dendro review"
author: "AJ Perez-Luque (@ajpelu)"
output: 
  md_document:
    variant: markdown_github
bibliography: ../references.bib 
csl: ../ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, warning=FALSE, error=TRUE, message=FALSE)
```

```{r packages, warning=FALSE, message=FALSE, echo=FALSE}
library("here")
library("tidyverse")
library("maps")
library("mapdata")
library("ggmap")
library("ggrepel")
library("cowplot")
library("ggExtra")
```

Dendro review

```{r}
df <- read.csv(file=here::here("data_raw/dendro_review/dendro_qp_review.csv"), header=TRUE)

spain <- map_data('world', region=c("spain", "portugal"))

s <- ggplot() + 
  geom_polygon(data = spain, aes(x=long, y = lat, group = group),
                            fill=NA, colour="gray") + 
  coord_fixed(1.3) +
  geom_point(data = df, aes(x=long, y=lat, size=length), 
            shape = 21, colour = "black", fill="white") +
             #, size=length), alpha=0.3) + 
  # geom_text_repel(data = subset(df, ), aes(x=long, y=lat, label=refs)) + 
  # geom_label(data = df, aes(x=long, y=lat, size=length, label=length)) +
  scale_size(range=c(1,6), breaks=c(0,50,100,200,300,350),
             guide = guide_legend(title="Age (years)")) + 
  xlab('') + ylab('') + theme_void() + 
  theme(legend.position = "top")

  # theme(panel.grid = element_blank(),
  #       axis.line=element_blank())

elev_age <- ggplot(df, aes(x=elev, y=length)) + 
  # geom_smooth(data = subset(df, elev > 500), se=FALSE, method = 'lm', colour="gray") +
  geom_point(size = 3, colour = "black") + theme_bw() + theme(panel.grid = element_blank()) + 
  geom_text_repel(data = subset(df, refs %in% c("1","2","3","4","5","6","7","8")), size = 3, aes(label=refs)) +
  geom_label_repel(data = subset(df, refs %in% c("SJ","CA-High", "CA-Low")), 
                   size = 5, aes(label=refs),
                   box.padding = 0.25, point.padding = 0.3) +
  xlab('Elevation (m)') + ylab('Age (years)') +
  ylim(0,350)

elev_age + annotation_custom(ggplotGrob(s), xmin = 0, xmax = 1400, ymin = 160)
# elev_age <- ggMarginal(elev_age, margins="y")
```

```{r}
ggplot(df, aes(length)) + geom_density()
```

### 2018-04-17 analysis/analysis_age_elevation.Rmd
add new data

@@
 elev_age <- ggplot(df, aes(x=elev, y=length)) + 
-  # geom_smooth(data = subset(df, elev > 500), se=FALSE, method = 'lm', colour="gray") +
+  geom_smooth(data = subset(df, elev > 500), se=FALSE, method = 'lm', colour="gray") +
   geom_point(size = 3, colour = "black") + theme_bw() + theme(panel.grid = element_blank()) + 
-  geom_text_repel(data = subset(df, refs %in% c("1","2","3","4","5","6","7","8")), size = 3, aes(label=refs)) +
+  geom_text_repel(data = subset(df, !(refs %in% c("SJ","CA-High", "CA-Low"))), size = 3, aes(label=refs)) +
   geom_label_repel(data = subset(df, refs %in% c("SJ","CA-High", "CA-Low")), 
@@
 elev_age + annotation_custom(ggplotGrob(s), xmin = 0, xmax = 1400, ymin = 160)
-# elev_age <- ggMarginal(elev_age, margins="y")
+
 ```
@@
 ggplot(df, aes(length)) + geom_density()
+
+dff <- subset(df, elev > 500)
+
+summary(lm(elev~length, data=dff))
 ```

### 2018-04-17 man/ms_3_discussion.Rmd
update
update discussion and bib file

@@
 #### Climate and tree relationship
+
 Frase de resultados que tenemos que poner en discussion: 
@@
 
-#### Disturbance
-Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
+Moisture availability has been reported as the most limiting factor driving radial growth of *Q.
+pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014].
+We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 6a).
+Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]
+
+Para *Q.
+pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q.
+pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al.
+2015 European Journal of Forest Research).
+Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea.
 
@@
     * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.
+
+#### Disturbance
+Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
 

### 2018-04-23 man/ms_3_discussion.Rmd
update tables

@@
 ## Site environment shapes differential sensitivity to climate and drought of rear-edge oak populations
+
+Despite similar value of competence in all sites, we obtained higher values of growth for the highest elevation site (CA-High, Figure 4).
+
+El sitio del norte, ()
+
+Tanto en el greenness como en el growth, encotramos direfencias para el greenness y el crecimiento entre las poblaciones situadas al norte y al sur.
+
+Our results showed differences for greenness and tree-growth between northern and southern oak populations ().
 

### 2018-04-23 man/ms_tables/ms_tables.Rmd
update tables

---
title: "ms_tables" 
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 7.5
    highlight: null
    reference_docx: ../template_tables_portrait.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
library(dplyr)
```

**Table 1.** Characteristics of sampled plot. Lat = latitude; Long = longitude. Dbh and height of all trees, Basal Area (BA), Density and SRD (Size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see methods). Temp.: annual average of minimun and maximum temperatures. Values shown here correspond to site averages. Standard deviations are shown in parentheses. Different letters indicate statistically significant differences between sites (Kruskal-Wallis test followed by Dunn’s test, p<0.05).

```{r, echo=FALSE}
topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
ages <- read.csv(here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
dendro_summ <- read.csv(here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)
letritas <- read.csv(here("/out/kw_stand_features", "kw_letras.csv"), header = TRUE)

letritas <- letritas %>% 
  mutate(letter = as.character(letter)) %>% 
  mutate(mivar = paste0(mivariable, '_l')) %>%
  dplyr::select(-mivariable) %>% 
  spread(key=mivar, letter)

t1raw <- topo %>% inner_join(ages, by='site') %>% 
  inner_join(dendro_summ, by='site') %>% 
  inner_join(biotic, by='site') %>% 
  inner_join(letritas, by='site')

t1raw <- t1raw %>% mutate(name = case_when(
  site == 'caH' ~ 'CA-High', 
  site == 'caL' ~ 'CA-Low',
  site == 'sj' ~ 'SJ'))

# ADD data climate
# clima obsnev (from repo clima SN; son datos de mapas pasados 100 x 100 1960 - 2010)

clima <- read.csv(here("/data/climate_obsnev", "obsnev_qp_pop.csv"), header = TRUE)
cli <- clima %>% filter(pop %in% c(2,6)) %>% mutate_if(is.numeric, funs(round(., 1))) %>% 
  dplyr::select(pop, mean, variable) %>% 
  spread(variable, mean) %>% 
  mutate(name = case_when(
    pop == 2 ~ 'SJ',
    pop == 6 ~ 'CA-High'
  )) %>% dplyr::select(-pop)

# duplicate row for CA site
x <- cli[2,] 
x$name <- 'CA-Low'
cli <- cli %>% bind_rows(x) %>% 
  mutate_at(1, round, 0) %>% 
  unite(temp, c(tmin, tmax), sep=" - ")

t1raw <- t1raw %>% inner_join(cli, by='name')

t1 <- t1raw %>% dplyr::select(
  name, lat_m, long_m, elev_mean, slope_mean, prec, temp, 
  n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, 
  dn_mean_all, dn_sd_all, h_mean_all, h_sd_all,
  ba_mean, ba_sd, stand_density_mean, stand_density_sd, srd_mean, srd_sd,
  all_dn_l, all_h_l, ba_l, ft_dn_l, ft_h_l, std_l, srd_l
  ) %>% 
  mutate(pare = ")") %>% 
  mutate_at(c("lat_m","long_m","slope_mean","ba_mean","ba_sd","srd_mean","srd_sd"), 
            funs(formatC(., digits = 2,format='f'))) %>%  
  # mutate_at(c(10:19, 22:23), round, 1) %>% 
  mutate_at(c("dn_mean","dn_sd","h_mean","h_sd","age_mean","age_std","dn_mean_all","dn_sd_all","h_mean_all","h_sd_all","stand_density_mean","stand_density_sd"),
            funs(formatC(., digits = 1,format='f'))) %>%
  mutate_at(4, round, 0) %>% 
  unite(n_trees, c(n_trees, n_radii), sep=" (") %>% 
  unite(dn, c(dn_mean, dn_sd), sep=" (") %>% 
  unite(h, c(h_mean, h_sd), sep=" (") %>% 
  unite(age, c(age_mean, age_std), sep=" (") %>% 
  unite(dn_all, c(dn_mean_all, dn_sd_all), sep=" (") %>% 
  unite(h_all, c(h_mean_all, h_sd_all), sep=" (") %>% 
  unite(ba, c(ba_mean, ba_sd), sep=" (") %>% 
  unite(stand_density, c(stand_density_mean, stand_density_sd), sep=" (") %>% 
  unite(srd, c(srd_mean, srd_sd), sep=" (") %>% 
  mutate(
    n_trees = paste0(n_trees, pare),
    dn = paste(paste0(dn, pare), ft_dn_l, sep=" "),
    h = paste(paste0(h, pare), ft_h_l, sep = ' '),
    age = paste0(age, pare),
    dn_all = paste(paste0(dn_all, pare), all_dn_l, sep = ' '),
    h_all = paste(paste0(h_all, pare), all_h_l, sep = ' '),
    ba = paste(paste0(ba, pare), ba_l, sep = ' '),
    stand_density = paste(paste0(stand_density, pare), std_l, sep = ' '),
    srd = paste(paste0(srd, pare), srd_l, sep = ' ')) %>%
  dplyr::select(-pare, -all_dn_l, -all_h_l, -ba_l, -ft_dn_l, -ft_h_l, -std_l, -srd_l)

colnames (t1) <- c("Site", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
                   "Prec. (mm)", "Temp (&deg; C)",
                   "# trees (# cores)", "Dbh (cm)", "Height (m)", "Age (years)", 
                   "Dbh all (cm)", "Height all (m)", "BA (m2/ha)", "Density (trees/ha)", "SRD")

pander(t1, split.table = Inf)

```

**Table 2.** Characteristics of the mean tree ring chronologies. Values of the length year in parenthesis indicate years replicated more than five series. RW = mean annual ring width (standard deviation in parenthesis). MS = mean sensitivity. AR(1) = mean autocorrelation of raw series. Rbt = mean correlation between series. EPS = mean expressed population signal. EPS and Rbt are calculated for the mean residual chronologies of growth indices.

```{r, echo=FALSE}
topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
ages <- read.csv(here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
dendro_summ <- read.csv(here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)

t1raw <- topo %>% inner_join(ages, by='site') %>% 
  inner_join(dendro_summ, by='site') %>% 
  inner_join(biotic, by='site')

t1raw <- t1raw %>% mutate(name = case_when(
  site == 'caH' ~ 'CA-High', 
  site == 'caL' ~ 'CA-Low',
  site == 'sj' ~ 'SJ'))

t2 <- t1raw %>% dplyr::select(
  name, first_year, last_year, 
  length_year, length_year5, 
  n_trees, n_radii,
  rw_mean, rw_std, ms2, ar1, rbar.wt, eps)  %>% 
  mutate(pare = ")") %>%

  mutate_at(c("rw_mean", "rw_std", "ms2", "ar1", "rbar.wt", "eps"),
            funs(formatC(., digits = 3,format='f'))) %>%
  unite(length_year, c(length_year,length_year5), sep=" (") %>%
  unite(RW, c(rw_mean,rw_std), sep=" (") %>%
  mutate(
    length_year = paste0(length_year, pare),
    RW = paste0(RW, pare)) %>% 
  dplyr::select(-pare)

colnames(t2) <- c("Site", "First year", "Last year", "Length (years)",  
                   "# trees", '# cores', "RW (mm)", "MS", "AR(1)", "Rbt", "EPS")

pander(t2,
       split.table = Inf)

```

**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the factor drought events (2005 and 2012) and site.

```{r, echo=FALSE, include = FALSE}
### Table R1_evi
evi <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
names(evi) <- c('factores', 'F', 'p', 'variables')

dff_evi <- evi %>% gather(temp, numeros, F, p) %>% 
  unite(temp1, variables, temp) %>% 
  spread(temp1, numeros) %>% 
  mutate(factores = recode(factores, 
                           "disturb_year" = "Drougth event", 
                           "disturb_year:site" = "Drougth event X Site", 
                           "site" = "Site")) %>% 
  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site"))) %>% 
  mutate_at(c(2,4,6,8), round, 2) %>% 
  mutate_at(c(3,5,7,9), round, 3) %>% 
  mutate(variable = 'evi')

bai <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
names(bai) <- c('factores', 'F', 'p', 'variables')

dff_bai <- bai %>% gather(temp, numeros, F, p) %>% 
  unite(temp1, variables, temp) %>% 
  spread(temp1, numeros) %>% 
  mutate(factores = recode(factores, 
                           "disturb_year" = "Drougth event", 
                           "disturb_year:site" = "Drougth event X Site", 
                           "site" = "Site")) %>% 
  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site"))) %>% 
  mutate_at(c(2,4,6,8), round, 2) %>% 
  mutate_at(c(3,5,7,9), round, 3) %>% 
  mutate(variable = 'bai')

dff <- dff_evi %>% 
  bind_rows(dff_bai) %>% 
  dplyr::select(-rrs_F, -rrs_p) %>% 
  dplyr::select(variable, factores, rt_F, rt_p, rc_F, rc_p, rs_F, rs_p)
```

```{r, echo=FALSE}
colnames(dff) <- c("variable", "factores", "rt\nF", "rt\np", "rc\nF", "rc\np", "rs\nF", "rs\np")
pander(dff)
```

**Table S1.** Robust measures of central tendency of resilience indices for greenness (EVI) groupped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included in parentheses.

```{r, echo = FALSE, warning = FALSE}
mraw <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)

m <- mraw %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>% 
 unite("mh", c("M.Huber", "ci"), sep = "\n")

# Datos agrupados para todos los sitios 
mraw_a <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)

m_a <- mraw_a %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
mraw_b <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)

m_b <- mraw_b %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
  mutate(disturb_year = 2099)

mm <- m %>% 
  bind_rows(m_a) %>%
  bind_rows(m_b) %>% 
  gather(temp, numeros, mh) %>% 
  unite(temp1, c(variables, disturb_year), temp) %>% 
  spread(temp1, numeros, fill = "")

mm <- mm %>% 
  dplyr::select(site, 
                rt_2005_mh, rc_2005_mh, rs_2005_mh,
                rt_2012_mh, rc_2012_mh, rs_2012_mh,
                rt_2099_mh, rc_2099_mh, rs_2099_mh)

colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
pander(mm, keep.line.breaks = TRUE, split.table = Inf)
```

**Table S2.** Robust measures of central tendency of resilience indices for tree-growth (BAI) groupped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included in parentheses.

```{r, echo = FALSE, warning = FALSE}
mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)

m <- mraw %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>% 
 unite("mh", c("M.Huber", "ci"), sep = "\n")

# Datos agrupados para todos los sitios 
mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)

m_a <- mraw_a %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)

m_b <- mraw_b %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
  mutate(disturb_year = 2099)

mm <- m %>% 
  bind_rows(m_a) %>%
  bind_rows(m_b) %>% 
  gather(temp, numeros, mh) %>% 
  unite(temp1, c(variables, disturb_year), temp) %>% 
  spread(temp1, numeros, fill = "")

mm <- mm %>% 
  dplyr::select(site, 
                rt_2005_mh, rc_2005_mh, rs_2005_mh,
                rt_2012_mh, rc_2012_mh, rs_2012_mh,
                rt_2099_mh, rc_2099_mh, rs_2099_mh)

colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
pander(mm, keep.line.breaks = TRUE, split.table = Inf)
```

```{r, echo = FALSE, warning = FALSE, eval = FALSE}

# Este código es para print la tabla anterior en formato pdf latex (ver state_ms)c
### OJO REVISAR 
# ------------------------------

mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)

m <- mraw %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter))

# Datos agrupados para todos los sitios 
mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)

m_a <- mraw_a %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)

m_b <- mraw_b %>% 
  filter(var != 'rrs') %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  mutate(disturb_year = 2099)

mm <- m %>% 
  bind_rows(m_a) %>%
  bind_rows(m_b) %>% 
  gather(temp, numeros, c(M.Huber, ci)) %>% 
  unite(temp1, c(variables, disturb_year), temp) %>% 
  spread(temp1, numeros, fill = "")

mm <- mm %>% 
  dplyr::select(site, 
                rc_2005_M.Huber, rc_2005_ci, rt_2005_M.Huber, rt_2005_ci, rs_2005_M.Huber, rs_2005_ci,
                rc_2012_M.Huber, rc_2012_ci, rt_2012_M.Huber, rt_2012_ci, rs_2012_M.Huber, rs_2012_ci,
                rc_2099_M.Huber, rc_2099_ci, rt_2099_M.Huber, rt_2099_ci, rs_2099_M.Huber, rs_2099_ci)

mmt <- mm %>%
   rownames_to_column %>% 
   gather(var, value, -rowname) %>% 
   spread(rowname, value)

colnames(mmt) <- c("var", "SJ", "caH", "caL", "z_all") 
mmt <- mmt[-19,] 
mmt <- mmt %>% 
  mutate(vars = c(rep("Rc",6), rep("Rt",6), rep("Rs",6)),
         year = c(rep(
           c(rep("2005", 2), 
             rep("2012", 2),
             rep(" ", 2)), 3))) %>% 
  dplyr::select(vars, year, SJ, caH, caL, var)

kable(mmt, 
      format = "latex", booktabs = T) %>% 
  kable_styling(font_size = 6) %>% 
  add_header_above(c("", "", "Northern", "Southern" = 2, "")) 
```

### 2018-04-23 man/ms_tables/t_sampled_plots.Rmd
update tables

@@
   word_document:
-    reference_docx: "../../man/template_tables_portrait.docx"
+    fig_caption: yes
+    fig_height: 6
+    fig_width: 7.5
+    highlight: null
+    reference_docx: ../template_tables_portrait.docx
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecology.csl
 ---
@@
 Dbh and height of all trees, Basal Area (BA), Density and SRD (Size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see methods).
+Temp.: annual average of minimun and maximum temperatures.
 Values shown here correspond to site averages.
@@
 t1raw <- t1raw %>% mutate(name = case_when(
-  site == 'caH' ~ 'Cáñar-High', 
-  site == 'caL' ~ 'Cáñar-Low',
-  site == 'sj' ~ 'San Juan'))
+  site == 'caH' ~ 'CA-High', 
+  site == 'caL' ~ 'CA-Low',
+  site == 'sj' ~ 'SJ'))
+
+# ADD data climate
+# clima obsnev (from repo clima SN; son datos de mapas pasados 100 x 100 1960 - 2010)
+
+clima <- read.csv(here("/data/climate_obsnev", "obsnev_qp_pop.csv"), header = TRUE)
+cli <- clima %>% filter(pop %in% c(2,6)) %>% mutate_if(is.numeric, funs(round(., 1))) %>% 
+  dplyr::select(pop, mean, variable) %>% 
+  spread(variable, mean) %>% 
+  mutate(name = case_when(
+    pop == 2 ~ 'SJ',
+    pop == 6 ~ 'CA-High'
+  )) %>% dplyr::select(-pop)
+
+# duplicate row for CA site
+x <- cli[2,] 
+x$name <- 'CA-Low'
+cli <- cli %>% bind_rows(x) %>% 
+  mutate_at(1, round, 0) %>% 
+  unite(temp, c(tmin, tmax), sep=" - ")
+
+t1raw <- t1raw %>% inner_join(cli, by='name')
 
 t1 <- t1raw %>% dplyr::select(
-  site, lat_m, long_m, elev_mean, slope_mean, 
+  name, lat_m, long_m, elev_mean, slope_mean, prec, temp, 
   n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, 
@@
   mutate(pare = ")") %>% 
-  mutate_at(c(2,3,5,18,19, 22:23), round, 2) %>% 
+  mutate_at(c("lat_m","long_m","slope_mean","ba_mean","ba_sd","srd_mean","srd_sd"), 
+            funs(formatC(., digits = 2,format='f'))) %>%  
+  # mutate_at(c(10:19, 22:23), round, 1) %>% 
+  mutate_at(c("dn_mean","dn_sd","h_mean","h_sd","age_mean","age_std","dn_mean_all","dn_sd_all","h_mean_all","h_sd_all","stand_density_mean","stand_density_sd"),
+            funs(formatC(., digits = 1,format='f'))) %>%
   mutate_at(4, round, 0) %>% 
-  mutate_at(c(8:17, 20:21), round, 1) %>% 
   unite(n_trees, c(n_trees, n_radii), sep=" (") %>% 
@@
 colnames (t1) <- c("Site", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
+                   "Prec. (mm)", "Temp (&deg; C)",
                    "# trees (# cores)", "Dbh (cm)", "Height (m)", "Age (years)", 

### 2018-05-07 analysis/chronos/analysis_chronologies.Rmd
update analysis chrono
export data comparison bais chronos

@@
 
+write.csv(correla, file=paste(di, "data/correla_ventanas_temporales/correla.csv", sep=""), row.names = FALSE)
+
 correla %>% 
@@
   theme(strip.background = element_blank())
+
 ```

### 2018-05-07 man/ms_figures.Rmd
add data figures

@@
 ##### 
+```{r, echo=FALSE}
+cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
+
+# Lm CA-High
+df <- cro_sites %>% 
+  filter(year >= 1975) %>% 
+  filter(site == 'CA-High')
+
+m1 <- lm(bai_mean ~ year, df)
+summary(m1)
+r2 <- format(summary(m1)$r.squared, digits = 3)
+```
+
 **Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
@@
 
-```{r, eval=FALSE}
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-summary(m1)
-r2 <- format(summary(m1)$r.squared, digits = 3)
-```
-
 ##### 
@@
 ##### 
-**Appendix S7.** Growth commparations
+**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
+Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
+Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
+
+```{r}
+correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
+
+correla <- correla %>% 
+  mutate(
+    name_comparison = as.factor(case_when(
+      name_comparison == "caH-caL" ~ "CAHigh - CALow",
+      name_comparison == "sj-caL" ~ "SJ - CALow",
+      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
+
+correla_sitesSJCALH  <- correla %>% 
+  filter(metodo_correla == 'pearson') %>%
+  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
+  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
+  xlab('Smoothing (centred moving average window size)') + 
+  # geom_smooth(method = 'loess', se=FALSE) + 
+  theme(panel.grid = element_blank()) +
+  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
+  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
+  theme(legend.position = c(.8,.9), legend.title = element_blank())
+
+correla_sitesSJCALH
+```
 

### 2018-05-07 man/ms_2_results.Rmd
new update
update results and discussion

@@
 Likely as a consequence of their different management origin (*i.e.* land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2).
-The competition index was similar among sites but plot basal area was greatest in CA-High (Table 1).
+Competition was similar among sites but plot basal area was greatest in CA-High (Table 1).
 Yet the basal area increment (BAI) growth in the highest plot in altitude (CA-High) was the greatest (Figure 4).
-Trees from this location showed in addition a positive growth trend since the late 1970s (Figure 4).
-This positive trend was not found in any of the other two locations and the two altitudes sampled in the southern populations showed differences in growth, being more similar CA-Low with growth in the northern population (SJ) than with CA-High (Figure 4; Appendix S7).
+In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations.
+For southern population differences in growth were observed, with CA-Low site showing lower values than CA-High but similar to growth values of the northern population (SJ) (Figure 4; Appendix S7).
 
@@
 A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155).
-Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S6) even not being the most severe drought events.
+Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S5) even not being the most severe drought events.
 

### 2018-05-07 man/ms_3_discussion.Rmd
update results and discussion

@@
 ```
+# Discussion
 
@@
 pyrenaica* along their distribution range [@GeaIzquierdo2013; @GeaIzquierdo2014].
-[@GeaIzquierdo2014] found a general decline in the growth of this species since the 1970s, particularly sharp for populations located in their dry-edge.
+@GeaIzquierdo2014 found a general decline in the growth of this species since the 1970s, particularly sharp for populations located in their dry-edge.
 This decline trend in growth have also been oberved for other tree species located in their rear-edges [*e.g.* @SanchezSalguero2012; @SanchezSalguero2017; @Dorado2017].
@@
 
-Despite similar value of competence in all sites, we obtained higher values of growth for the highest elevation site (CA-High, Figure 4).
+Despite similar value of competence in our study sites, we obtained higher values of growth for the highest elevation site (CA-High, Figure 4).
 

### 2018-05-07 man/ms_figures.Rmd
updates figures

@@
 For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c) and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour Orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
+Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
+Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
 
@@
 ##### 
-```{r, echo=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-summary(m1)
-r2 <- format(summary(m1)$r.squared, digits = 3)
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r2`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
+```{r, warning=FALSE}
 cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
@@
 )
+
+# Lm CA-High
+df <- cro_sites %>% 
+  filter(year >= 1975) %>% 
+  filter(site == 'CA-High')
+
+m1 <- lm(bai_mean ~ year, df)
+# summary(m1)
+myr2 <- format(summary(m1)$r.squared, digits = 3)
+
+```
+
+**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
+pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
+Shading areas coorespond to standard error of the mean.
+Number of series are displayed in the upper plot.
+We only show chronologies with # trees > 5.
+Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
+
+```{r, warning=FALSE, out.extra='angle=90'}
 
@@
 ##### 
-**Figure 5.** Residual Tree-ring chronologies obtained for the *Q.
+**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
 pyrenaica* sites.
@@
 pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site; and *dark blue* bars: high-elevation shouthern site.
+*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
 Asteriks indicate significant ($P < 0.05$) correlation coefficients.
@@
                "T max" = "(c) T max", "T min" = "(d) T min")
+
+label_site <- c("SJ" = "SJ",
+                "caH" = "CA-High",
+                "caL" = "CA-Low")
 
@@
                      breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3")) + 
+  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
   ylab('Correlation')
@@
 A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g* @Spinoni2015; Spinoni2017a; @Pascoa2017].
+A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
 The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
 ***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI** refer to the mean and lowest value of SPEI respectively during the drought event duration.
+***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
 

### 2018-05-08 man/ms_3_discussion.Rmd
update

@@
 
+Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3).
+Las poblaciones del norte, que presentan condiciones mas secas (Tabla 1), se vieron mas afectadas por los eventos de sequía.
+Así por ejemplo vimos que las anomalías estandarizadas de EVI fueron mas negativas en la población del norte durante la sequía de 2005 (mayor intensidad de browing en 2005 para el norte).
+
+Con respecto al BAI, observamos 
+Llama la atención que en el rear-edge de la distribución de la especie, en el sitio mas meridional dentro de este rear edge, y en la parte mas alta (estas parcelas están en el treeline de la especie en SN (en torno a 1900)), es donde encontramos mayores crecimientos.
+Parece que los robles estuvieran mejor en esta zona.
+Algunas reflexiones sobre esto:
+
+A striking result is the difference for tree growth between southern sites (Figure 4).
+Despite these sites are very close to each other, we obtained higher values of BAI for the high-elevation site than for the low-elevation site.
+
+The trees of CA-High site, which are located around 1900 m.
+a.s.l.
+and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ).
+This is especially interesting for southern sites, which  are very close to each other.
+
+In addition, despite southern sites are very close to each other, we obtained higher values of BAI for the high-elevation site than for the low-elevation site.
+
+Algunos autores [@VicenteSerrano2014b] hablan de que las respuestas 
+Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011)
+
+-----------
+
 Despite similar value of competence in our study sites, we obtained higher values of growth for the highest elevation site (CA-High, Figure 4).
 
-El sitio del norte, ()
-
 Tanto en el greenness como en el growth, encotramos direfencias para el greenness y el crecimiento entre las poblaciones situadas al norte y al sur.
-
-Our results showed differences for greenness and tree-growth between northern and southern oak populations ().
 

### 2018-05-15 man/ms_3_discussion.Rmd
update text

--- 2018-05-07 man/ms_3_discussion.Rmd
@@
 
+Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3).
+Las poblaciones del norte, que presentan condiciones mas secas (Tabla 1), se vieron mas afectadas por los eventos de sequía.
+Así por ejemplo vimos que las anomalías estandarizadas de EVI fueron mas negativas en la población del norte durante la sequía de 2005 (mayor intensidad de browing en 2005 para el norte).
+
+A striking result is the difference for tree growth between sites (Figure 4).
+The trees of CA-High site, which are located around 1900 m.
+a.s.l.
+and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4).
+This is especially interesting for southern sites, which are very close to each other.
+Our results are consistent with previous findings that pointed out tree growth and tree responses to drought are site-dependent (e.g.
+soil features, tree competence, etc.) [e.g.
+@Babst2013, @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017, @Dorado2017c].
+
+When we explored the chracteristics of the southern sites, we firstly observed a difference in elevation.
+@Bhuyan2017, in a recent work across the European continent, found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.
+The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q.
+pyrenaica* [@GeaIzquierdo2014].
+In fact, the differences in available water of soils could explain our results.
+[@CoboDiaz2017], who studied microbiota of this location in an elevational gradient, found higher values of available water of the forest soils located on high elevation compared with those located at low elevations.
+In this sense, a remarkably observation supporting this idea, is the presence of a traditional irrigation ditches located uphill from the CA-High site.
+In most of the high-mountain watersheds of Sierra Nevada, there is a system of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014, @MartinCivantos2016].
+These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014, @Jodar2017].
+There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017].
+It brings water from snowmelt and from a nearby stream, soaking the *Q.
+pyrenaica* forests.
+Finally another idea to consider is the different degree of anthropic impact related to elevation.
+
+Aunque nuestros sitios presenten valores similares de competencia, la historia de manejo humano puede ser diferente, y en parte se relaciona con la altitud.
+Así en zonas altas encontramos un menor impacto humano que en las zonas bajas.
+Esto ocurre
+
+Another reason that could explain the higher values of BAI of CA-High site is el grado diferente de impacto antrópico.
+En nuestros resultados vemos como los bosques situados a mayor altitud presentan un menor impacto antrópico.
+Esto puede ser simplemente por el hecho de que estos bosques estén menos accesibles ---> ojo y la idea de que son bosque relativamente jóvenes
+---> Gráfica 
+Our sampled sited showed similar tree competence levels (Table 1), but ... impacto antrópico diferente
+
+* Quizá están creciendo donde les están dejando (menor impacto antrópico en las zonas mas altas, al menos en los últimos años??)
+
+The presence of *Q. pyrenaica* forests on the southern slopes of Sierra Nevada is partly explained by the additional contribution of humid air from the Mediterranean sea [@Prieto1975]
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+* Vicente-Serrano: 
+    * We found that some forests from cold and humid areas respond to shorter drought time-scales than forests from dry areas, which usually respond to longer time-scales
+    * Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011).
+    * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.
+
+-----------
+
 Despite similar value of competence in our study sites, we obtained higher values of growth for the highest elevation site (CA-High, Figure 4).
 
-El sitio del norte, ()
+------
+
+En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations).
+The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone.
+
+")
+
+```
 
@@
 
-Our results showed differences for greenness and tree-growth between northern and southern oak populations ().
-
 Esta tendencia es importante sobre todo en las poblaciones situadas mas al sur, en donde observamos: menor br
 Encontramos diferencias entre poblaciones. Asi por ejemplo en las situadas mas al sur, vemos que las anomalias de EVI son menos negativas para el 2005; También vemos tendencias positivas en el BAI...
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+## Rear-edge
+
+* Se asume una alta vulnerabilidad de las especies a la sequía en su límite inferior, sin embargo esta vulnerabilidad no necesariamente se mantiene: The results of Granda et al.
+2018 are consistent with recent studies showing that the frequently assumed higher vulnerability of dry, rear-edge populations to drought does not necessarily hold (e.g.
+Cavin and Jump 2017) (from: Martinez-Vilalta 2018)
+
+* Definición de *marginal* populations.
+[@MartinezVilalta2018] indica que a veces no está claro la definición de marginal populations.
+When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).
+
+* Quizá estamos asumiendo que es una población marginal, pero sin embargo lo es solamente geográficamente
+
+-----
 
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-## Rear-edge
+In addition, despite southern sites are very close to each other, we obtained higher values of BAI for the high-elevation site than for the low-elevation site.
 
-* Se asume una alta vulnerabilidad de las especies a la sequía en su límite inferior, sin embargo esta vulnerabilidad no necesariamente se mantiene: The results of Granda et al.
-2018 are consistent with recent studies showing that the frequently assumed higher vulnerability of dry, rear-edge populations to drought does not necessarily hold (e.g.
-Cavin and Jump 2017) (from: Martinez-Vilalta 2018)
-
-* Definición de *marginal* populations.
-[@MartinezVilalta2018] indica que a veces no está claro la definición de marginal populations.
-When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).
-
-* Quizá estamos asumiendo que es una población marginal, pero sin embargo lo es solamente geográficamente
 ")
@@
 
-Para *Q.
-pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q.
-pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al.
-2015 European Journal of Forest Research).
-Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea.
-
 ### ¿Que factor es mas limitante para el crecimiento en el rear-edge de Q. pyrenaica? 
@@
 2013, 2015) …
-
-* Vicente-Serrano: 
-    * We found that some forests from cold and humid areas respond to shorter drought time-scales than forests from dry areas, which usually respond to longer time-scales
-    * Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011).
-    * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.
 

### 2018-05-15 man/ms_3_discussion.Rmd
update

@@
 soil features, tree competence, etc.) [e.g.
-@Babst2013, @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017, @Dorado2017c].
+@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017c].
 
@@
 In this sense, a remarkably observation supporting this idea, is the presence of a traditional irrigation ditches located uphill from the CA-High site.
-In most of the high-mountain watersheds of Sierra Nevada, there is a system of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014, @MartinCivantos2016].
-These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014, @Jodar2017].
+In most of the high-mountain watersheds of Sierra Nevada, there is a system of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016].
+These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017].
 There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017].

### 2018-05-16 man/ms_3_discussion.Rmd
update

@@
 In fact, the differences in available water of soils could explain our results.
-[@CoboDiaz2017], who studied microbiota of this location in an elevational gradient, found higher values of available water of the forest soils located on high elevation compared with those located at low elevations.
-In this sense, a remarkably observation supporting this idea, is the presence of a traditional irrigation ditches located uphill from the CA-High site.
+@CoboDiaz2017, in a work analyzing the variation of the microbiota along an elevational gradient within this location, found higher values of available water of the forest soils located on high elevation compared with those located at low elevations.
+In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply.
 In most of the high-mountain watersheds of Sierra Nevada, there is a system of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016].
 These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017].
+The network of traditional ditches present in one of our sampling site (Cáñar) was recentlty described, with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015].
 There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017].
@@
 pyrenaica* forests.
+This traditional system could supply an extra of water that could be used by the trees located downstream of these channels.
+This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site.
+
 Finally another idea to consider is the different degree of anthropic impact related to elevation.

### 2018-05-16 man/ms_1_material_methods.Rmd
update

--- 2018-03-27 man/ms_1_material_methods.Rmd
@@
 For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc.
-Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].
+Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@MartinMontanes2015; @RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].
 

### 2018-05-16 man/ms_3_discussion.Rmd
update

@@
 ---------------------
+Methodological approach
 
-# 4) Punto fuerte metodológico (VALID)
+Vegetation reflects the environmental conditions, and the effects of drought on vegetation can be observed using information from several methodological approaches [see @Norman2016 for a review].
+Several works revealed the utility of remote-sensing [*e.g*.
+@Zhang2013; @AghaKouchak2015] and of the dendrochronology [*e.g.* @Eilmann2012; @Bhuyan2017] to quantify the drought impacts on vegetation at different spatial and temporal scales, but to our knowledge the combination of both approaches is scarce [@VicenteSerrano2013; @VicenteSerrano2016; @Bhuyan2017b; @Wu2018].
+Our approach combining both methodologies to explore the resilience metrics have
+
+The combination of these two approach to explore the resilience of forests to drought have been recently used in a study across Iberian Peninsulsa, but
+
+Recently, Gazol et al. 2018 combine these two approaches to explore resilience of several iberian tree species, and encontraron ...
 
@@
 
-Nuestros resultados
-
 La combinación de metodologías que hemos empleado para evaluar la resiliencia de la vegetación frente a los eventos de sequía
 
-La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía en la vegetación se pueden ver tanto con información de satélite, como con información mas de campo (hablar aquí de lo del BAI).
-
-Los efectos de la sequía sobre la vegetación pueden observarse usando información procedente de diferentes aproximaciones metodológicas [see @Norman2016 for a review].
-
-[información procedente de remote sensing (RS)[@Zhang2013; @AghaKouchak2015], información dendrochronologica (),
-
-o combinando estas y otras metodologías [see @Norman2016 for a review]
-
-respuesta de
-
-The effects of drought on vegetation can be observed using information from several methodological approaches [see @Norman2016 for a review].
-Several works revealed the utility of remote-sensing or dendrochronology to quantify the dorught impacts on vegetation at different spatial and temporal scales ().
+Se trata de expresar la idea de mezcla de aproximacioes: remote-sensing con dendro.
+Aunque existen algunos trabajos que utilizado RWI y remote sensing nuestra aproximación es importante, sobre todo para estudiar poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
+De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
 
 Exiten trabajos que han utilizado remote-sensing o la dendrochronology para analizar los efectos de las sequías sobre la vegetación a diferentes escalas espaciales y temporales ( ).
-Asimismo, algunos otros trabajos han utilizado conjuntamente estas dos metodologias, y han visto que existe de forma general una relació
+Asimismo, algunos otros trabajos han utilizado conjuntamente estas dos metodologias, y han visto que existe de forma general una relación
 a positive relationship between vegetation indices derived from remote sensing and annual tree growth [@VicenteSerrano2016; @Bhuyan2017b],
@@
 
->>>>
-* Hablan de incorporar ademas de RS, el conocimiento ecológico (los principios ecológicos) para comprender las consecuencias de las disturbances analizadas con RS, y sobre todo para predecir futuro.
-(McDowell et al https://www.sciencedirect.com/science/article/pii/S1360138514002726)
-
-Nuestro trabajo muestra una combinación de métodos
-
-Finalmente, destacamos la aproximación metodológica que hemos llevado a cabo.
-
-La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía se pueden ver tanto con información de satélite (hablar del browning), como con información mas de campo (hablar aquí de lo del BAI).
-
-# Relaciones RWI-RS 
-* Algunos trabajos han analizado la relación ente RWI y remote sensing [@VicenteSerrano2013]; [(Bhuyan et al.
-Remote Sensing 9(6):526)]; [Vicente Serrano et al.
-https://www.sciencedirect.com/science/article/pii/S003442571630373X] ...
-
-## Trabajos RWI-RS 
-*
-
-# Diferencias que hemos detectado usando RS y RWI (en resiliencias)
+* Diferencias que hemos detectado usando RS y RWI (en resiliencias)
 * Relacionar con datos de Gazol et al. 2018
-
-# Jusfiticación de uso de ambos métodos
-Se trata de expresar la idea de mezcla de aproximacioes: remote-sensing con dendro.
-Aunque existen algunos trabajos que utilizado RWI y remote sensing nuestra aproximación es importante, sobre todo para estudiar poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
-De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
 
@@
 
-Aunque el NDVI o similar ha sido utilizado para monitorizar la respuesta de bosques frente a sequía, su escala espacial limita su uso para estudiar cambios en poblacioens situadas en su rear edge.
-
  Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability.
@@
 [@Jump2010]
-
-comentario de GGI
-
-Gracias Antonio.
-Desde luego que debes resaltar a lo largo del artículo la combinación de sensores y dendro como fortaleza de tu trabajo al analizar el tema de resiliencia desde dos escalas espacio-temporales complementarias.
-Y eso independientemente que lo diga o no Alistair Jump (que las citas vienen bien, y seguro que el trabajo es bueno), combinar ambas escalas/datos es beneficioso no sólo en poblaciones rear-edge, sino en general.
-Es decir, desde mi punto de vista, aunque decidas no centrarte tanto en el aspecto metodológico como sugieren Regino y Curro, eso no quiere decir que este punto no lo presentes de forma explícita en tu trabajo (en abstract debe aparecer, luego en intro, discusión y tal vez una frase en conclusiones) como uno de los elementos fuertes (un término medio, vamos).
-Creo que ya hemos discutido esto cuando hemos hablado.
-No es el primer trabajo que lo hace, como bien señalas, pero tampoco hay que pensar que no tiene valor, pues no hay tantos trabajos que combinen ambas aproximaciones (habrá cada vez más de hecho).
-Así que ánimo que queda poco y el trabajo puede quedar interesante si somos capaces de escribir e interpretar bien los resultados que nos has enviado.
 
@@
 
-### Muchos estudios han evaluado la resiliencia de los bosques a un evento de sequía, pero pocos lo han hecho a
-
-------------------
-
 ## 

### 2018-07-09 man/ms_3_discussion.Rmd
fix typo refs
update

@@
 Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure 5).
-This common signal observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].
+The decline in growth observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].
 
@@
 
-Desde una perspectiva a largo plazo, nuestros resultados muestran que las poblaciones de *Q.
-pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia.
+Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q.
+pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Appendix S5).
 Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005].
@@
 Las poblaciones del norte, que presentan condiciones mas secas (Tabla 1), se vieron mas afectadas por los eventos de sequía.
-Así por ejemplo vimos que las anomalías estandarizadas de EVI fueron mas negativas en la población del norte durante la sequía de 2005 (mayor intensidad de browing en 2005 para el norte).
+Por ejemplo las anomalías estandarizadas de EVI fueron mas negativas (mayor intensidad en el browning) en la población del norte durante la sequía de 2005.
+SJ site showed stronger correlations of tree-growth with SPEI (Hydrological and summer) (Figure 6), which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
+
+Moisture availability has been reported as the most limiting factor driving radial growth of *Q.
+pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014].
+We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 6a).
+Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]
 
@@
 
-----------------------------
-
 * Proyeccciones de crecimeinto de para *Q.
@@
 end VALID
----------------------
 

### 2018-09-07 analysis/analysis_age_elevation.Rmd
update analysis age elevation

@@
   geom_point(data = df, aes(x=long, y=lat, size=length), 
-            shape = 21, colour = "black", fill="white") +
+            shape = 21, colour = "black", fill="white", alpha=0.4) +
              #, size=length), alpha=0.3) + 
@@
   #       axis.line=element_blank())
-
-elev_age <- ggplot(df, aes(x=elev, y=length)) + 
-  geom_smooth(data = subset(df, elev > 500), se=FALSE, method = 'lm', colour="gray") +
-  geom_point(size = 3, colour = "black") + theme_bw() + theme(panel.grid = element_blank()) + 
-  geom_text_repel(data = subset(df, !(refs %in% c("SJ","CA-High", "CA-Low"))), size = 3, aes(label=refs)) +
-  geom_label_repel(data = subset(df, refs %in% c("SJ","CA-High", "CA-Low")), 
-                   size = 5, aes(label=refs),
-                   box.padding = 0.25, point.padding = 0.3) +
-  xlab('Elevation (m)') + ylab('Age (years)') +
-  ylim(0,350)
-
-elev_age + annotation_custom(ggplotGrob(s), xmin = 0, xmax = 1400, ymin = 160)
-
 ```
@@
 
-dff <- subset(df, elev > 500)
+dfall <- subset(df, elev > 500)
+lm_all <- lm(elev~length, data=dfall)
 
-summary(lm(elev~length, data=dff))
+dfsn <- subset(df, sn == 'yes')
+lm_sn <- lm(elev~length, data=dfsn)
+
+# Labels r2 
+label_sn <- paste("R[sn]^2 == ", round(summary(lm_sn)$r.squared,4))
+
+label_all <- paste("R[all]^2 == ", round(summary(lm_all)$r.squared,4))
+
+label_both <- paste0('atop(R[all]^2 ==',round(summary(lm_all)$r.squared,4),
+            ', R[sn]^2 ==',round(summary(lm_sn)$r.squared,4), ')')
 ```
+
+```{r}
+
+elev_age <- ggplot(df, aes(x=elev, y=length, colour=sn)) + 
+  geom_smooth(data = subset(df, elev > 500), se=FALSE, method = 'lm', colour="black") +
+  geom_smooth(data = subset(df, sn == 'yes'),se=FALSE, method = 'lm', colour='grey71') +
+  geom_point(size = 3) + theme_bw() + 
+  theme(panel.grid = element_blank(), legend.position="none") + 
+  geom_text_repel(data = subset(df, !(refs %in% c("SJ","CA-High", "CA-Low"))), 
+                  size = 3, aes(label=refs)) +
+  geom_label_repel(data = subset(df, refs %in% c("SJ","CA-High", "CA-Low")), 
+                   size = 5, aes(label=refs),
+                   box.padding = 0.25, point.padding = 0.3) +
+  xlab('Elevation (m)') + ylab('Age (years)') +
+  scale_colour_manual(values = c("black", "grey71")) + 
+  ylim(0,350)
+
+elev_age + annotation_custom(ggplotGrob(s), xmin = 0, xmax = 1400, ymin = 160) +
+  annotate("text", label = label_both, x = 150, y = 50, size = 5, parse = TRUE)
+
+```

### 2018-09-18 man/ms_3_discussion.Rmd
update discussion

@@
 Las poblaciones del norte, que presentan condiciones mas secas (Tabla 1), se vieron mas afectadas por los eventos de sequía.
-Por ejemplo las anomalías estandarizadas de EVI fueron mas negativas (mayor intensidad en el browning) en la población del norte durante la sequía de 2005.
-SJ site showed stronger correlations of tree-growth with SPEI (Hydrological and summer) (Figure 6), which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
+For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2).
+In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for SJ site (Figure 6), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
 
@@
 a.s.l.
-and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4).
+and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4).
 This is especially interesting for southern sites, which are very close to each other.
@@
 
+***Reescribir esto del impacto antrópico***
+
 Finally another idea to consider is the different degree of anthropic impact related to elevation.
@@
 Esto puede ser simplemente por el hecho de que estos bosques estén menos accesibles ---> ojo y la idea de que son bosque relativamente jóvenes
----> Gráfica 
-Our sampled sited showed similar tree competence levels (Table 1), but ... impacto antrópico diferente
+
+* ---> Gráfica. Our sampled sited showed similar tree competence levels (Table 1), but ... impacto antrópico diferente
 
@@
 ```
-
-Tanto en el greenness como en el growth, encotramos direfencias para el greenness y el crecimiento entre las poblaciones situadas al norte y al sur.
-
-Esta tendencia es importante sobre todo en las poblaciones situadas mas al sur, en donde observamos: menor br
-Encontramos diferencias entre poblaciones. Asi por ejemplo en las situadas mas al sur, vemos que las anomalias de EVI son menos negativas para el 2005; También vemos tendencias positivas en el BAI...
 
@@
 
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 #### Climate and tree relationship
@@
 **which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]**.
-
-Moisture availability has been reported as the most limiting factor driving radial growth of *Q.
-pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014].
-We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 6a).
-Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]
+")
+```
 
@@
 Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
-
-duda --> las proyecciones de crecimiento que hiciste [@GeaIzquierdo2013] sugerían un declive en el crecimiento. Nosotros estamos obteniendo algo diferente no? DUDA
 

### 2018-10-05 man/ms_figures.Rmd
change plot orientation S5
update

@@
 
-```{r, warning=FALSE, out.extra='angle=90', fig.height=5}
+```{r, warning=FALSE, dpi = 100, fig.height=5}
+# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
 # see trends-resilience_bai.Rmd 
@@
     geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=4, pch = 21, colour="black", fill="white") + 
+    geom_point(size=2, pch = 21, colour="black", fill="white") + 
     theme_bw() + 
@@
     ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 14, 
+    theme(plot.title = element_text(size = 8, 
                                     face = "bold.italic",
@@
 sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resistance (Rt) (3year window size)",
+        XLab = "", YLab = "Resistance (Rt) (3year window size)",
         coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
@@
 sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Recovery (Rc) (3year window size)",
+        XLab = "", YLab = "Recovery (Rc) (3year window size)",
         coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
 
-sevRt + sevRc + sevRs
+sevRt + sevRc + sevRs + plot_layout(nrow = 1)
 

### 2018-10-08 man/ms_1_material_methods.Rmd
update

@@
 During the second half of the XX century, several extreme drought episodes were recorded for the Iberian Peninsula (*e.g*.
-1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014 @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were charecterized as extreme drought in our climatic data (Appendices S1-S3).
+1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3).
 We focused on these two drought events because limitations about temporal availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
@@
 For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags).
-Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations of @Reyes2015 for mountain regions [@Reyes2015].
+Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].
 
@@
 As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016.
-$EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
+$EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
 For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale.
@@
 For the southern site two elevations were sampled: CA-Low and CA-High.
-All the sites were oak monospecific and representatives of two of the three population clusters identified for the species in this mountain range [@PerezLuque2015].
+All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015].
 In each site between 15 and 20 dominant trees were randomly selected.

### 2018-10-08 man/ms_2_results.Rmd
update

@@
 For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 3).
-For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Tab
+For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2).
 
-The exploration of resilience metrics of tree-growth for other severe droughts, revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S6).
+The exploration of resilience metrics of tree-growth for other severe droughts, revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S5).
 A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155).

### 2018-10-08 man/ms_3_discussion.Rmd
update

@@
 pyrenaica*.
-Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía, observamos como estas poblaciones muestran altos valores de resiliencia.
+Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía, observamos como estas poblaciones presentan altos valores de resiliencia.
 Como muestrasn nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).

### 2018-10-09 man/report.Rmd
update
remove supp figures from repor rmd
prepare v2

--- 2017-09-15 man/report.Rmd
@@
 ---
-title: "ms_borrador"
-author: "AJ Perez-Luque; G. Gea-Izquierdo; R. Zamora; FJ. Bonet"
-date: "2017"
-output:  word_document
-bibliography: ../references.bib
-csl: ../ecology.csl
+title: "Relict populations of a deciduous oak show high resilience to drought despite land-use legacies"
+author: "AJ Perez-Luque; G. Gea-Izquierdo; ..."
+output:
+  word_document:
+    fig_caption: yes
+    fig_height: 6
+    fig_width: 6
+    highlight: null
+    reference_docx: template_ms.docx
+bibliography: ../refs/references.bib
+csl: ../refs/ecology.csl
 ---
+
+# Abstract 
+* Trees showed resilience to drought events, despite the ‘a priori’ vulnerability that they can express as a result of the impact of climate change on relict stands resulting from intensive land-use in history.
+
+* Oak resilience to drought followed a dryness gradient: Northern populations, which thrive under a drier climate, were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.
 
@@
 ```
+
+```{r child = 'ms_3_discussion.Rmd'}
+```

### 2018-10-15 man/ms_figures.Rmd
add lm for cro_sites

@@
 # Lm CA-High
-df <- cro_sites %>% 
+dfcaH <- cro_sites %>% 
   filter(year >= 1975) %>% 
@@
 
-m1 <- lm(bai_mean ~ year, df)
+mCaH <- lm(bai_mean ~ year, dfCaH)
 # summary(m1)
-myr2 <- format(summary(m1)$r.squared, digits = 3)
+r2caH <- format(summary(mCaH)$r.squared, digits = 3)
+
+# Lm CA-Low 
+dfCaL <- cro_sites %>% 
+  filter(year >= 1975) %>% 
+  filter(site == 'CA-Low')
+
+mCaL <- lm(bai_mean ~ year, dfCaL)
+r2caL <- format(summary(mCaL)$r.squared, digits = 3)
+
+# Lm SJ
+dfSJ <- cro_sites %>% 
+  filter(year >= 1975) %>% 
+  filter(site == 'SJ')
+
+mSJ <- lm(bai_mean ~ year, dfSJ)
+r2SJ <- format(summary(mSJ)$r.squared, digits = 3)
 
@@
 We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
+Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r r2caH`).
 
@@
   geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
+              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) +
+  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-Low'), 
+              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) +
+  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'SJ'), 
               aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE)
+
   # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +

### 2018-10-16 man/ms_figures.Rmd
update

@@
 
-## Figure 1 Location map 
-- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset.
+- [] Figure 1. Location map
 
     - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf
-    - Ojo con los nombres SJ (Que hacemos?)
-
-## Figure 2. Plot de anomalias
-- [X] Solamente Anomalias estandarizadas
-
-## Figure 3. 
+
+- [] Figure 2. Plot de anomalias
+- [] Figure 3. Resilience metrics
+- [] Figure 4. BAI chronologies
+
 ")
@@
 **Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
+Distribution of *Quercus pyrenaica* forests in the Iberian Peninsula (a) and in Sierra Nevada mountain range (b).
 Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
@@
 Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
+Colour orthophotography of 2009 from Regional Ministry of the Environment [@IECA2009].
 
@@
 
+# To get p.value of the model 
+# https://stackoverflow.com/questions/5587676/pull-out-p-values-and-r-squared-from-a-linear-regression
+lmp <- function(modelobject) {
+    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
+    f <- summary(modelobject)$fstatistic
+    p <- pf(f[1],f[2],f[3],lower.tail=F)
+    attributes(p) <- NULL
+    return(p)
+}
+
 # Lm CA-High
-dfcaH <- cro_sites %>% 
+dfCaH <- cro_sites %>% 
   filter(year >= 1975) %>% 
@@
 mCaH <- lm(bai_mean ~ year, dfCaH)
-# summary(m1)
+summary(mCaH)
 r2caH <- format(summary(mCaH)$r.squared, digits = 3)
+pCaH <- lmp(mCaH)
 
@@
 mCaL <- lm(bai_mean ~ year, dfCaL)
+summary(mCaL)
 r2caL <- format(summary(mCaL)$r.squared, digits = 3)
+pCaL <- lmp(mCaL)
 
@@
 mSJ <- lm(bai_mean ~ year, dfSJ)
+summary(mSJ)
 r2SJ <- format(summary(mSJ)$r.squared, digits = 3)
-
+pSJ <- lmp(mSJ)
 ```
@@
 pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
+Shading areas correspond to standard error of the mean.
 Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
+We only show years replicated with # series > 5.
 Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r r2caH`).

### 2018-10-16 analysis/dendro_summary.Rmd
add na.rm TRUE

@@
   group_by(site) %>% 
-  dplyr::summarise(age_mean = mean(year), 
+  dplyr::summarise(age_mean = mean(year, na.rm = TRUE), 
          age_std = sd(year)) %>% as.data.frame()
@@
                    year_length = max(year),
-                   rw_mean = mean(mean), 
-                   rw_std = mean(stdev),
-                   sens1 = mean(sens1), 
-                   sens2 = mean(sens2), 
-                   ar1 = mean(ar1)) %>% 
+                   rw_mean = mean(mean, na.rm = TRUE), 
+                   rw_std = mean(stdev, na.rm = TRUE),
+                   sens1 = mean(sens1, na.rm = TRUE), 
+                   sens2 = mean(sens2, na.rm = TRUE), 
+                   ar1 = mean(ar1, na.rm = TRUE)) %>% 
   as.data.frame()

### 2018-10-16 analysis/chronos/get_chronos.Rmd
create analysis chronos

--- 2018-05-07 analysis/chronos/analysis_chronologies.Rmd
@@
 ---
-title: "chronologies"
+title: "get_chronologies"
 author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 Feb"
+date: "2018"
 output:  
@@
 knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
+```
+
+## Prepare data
+
+* Leer datos `rwl` de SJ y CA
+* Leer datos de diametros de los focal tree
+* Utilizaremos tres sitios: SJ, CAH y CAL (ver [./analysis/analysis_chronologies.md]('./analysis/analysis_chronologies.md))
+
+```{r}
+machine <- 'ajpelu'
+# machine <- 'ajpeluLap'
+di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
+
+# sj 
+sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
+
+# canar 
+ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
+
+# Read diameters data
+compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+```
+
+```{r}
+source(paste0(di, 'script/R/rw_byTree.R'))
+source(paste0(di, 'script/R/bai_piovesan.R'))
+source(paste0(di, 'script/R/baiResilience.R'))
+```
+
+* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low
+
+```{r, echo=FALSE}
+# Replace SNA by SJ and SNB by CA
+names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
+names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")
+
+# Remove g in name of some cores of CA. 
+names(ca) <- stringr::str_replace(names(ca), "g", "")
+```
+
+```{r, echo=FALSE}
+# Create subset to compare between sites 
+caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
+             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
+             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
+caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
+              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
+              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
+
+```
+
+* Lectura y preparación de datos de diámetro
+
+```{r} 
+# Prepare Diameter data
+
+# Compute diameter (mm)
+compete <- compete %>% 
+  mutate(dn_mm = (perim_mm / pi))
+
+# Change name focal according to loc
+compete <- compete %>% 
+  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))
+
+# Get only focal trees, and only selected variables 
+ft <- compete %>% 
+  filter(sp=='Focal') %>% 
+  filter(id_focal!='Fresno') %>% 
+  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm)
+
+# Set levels of eleveation 
+ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
+            paste0('CA', 26:30))
+ca_highcode <- paste0('CA', 11:25)
+
+ft <- ft %>% 
+  mutate(site = as.factor(
+    ifelse(id_focalLoc %in% ca_lowcode, 'CAL', 
+           ifelse(id_focalLoc %in% ca_highcode, 'CAH', 'SJ'))))
 ```
@@
 
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
+* Crear dataframes `rwl` por cada CA_High, CA_Low, SJ 
     * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
@@
               "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
-
-sjL <- sj[, c("SJ0101","SJ0102","SJ0201","SJ0202","SJ0301","SJ0302","SJ0401","SJ0402","SJ0501","SJ0502",
-              "SJ0601","SJ0602","SJ0603","SJ0701","SJ0702","SJ0801","SJ0802","SJ0901","SJ0902","SJ0903",
-              "SJ1001","SJ1002","SJ1003")]
-sjH <- sj[, c("SJ1101","SJ1102","SJ1103","SJ1201","SJ1202","SJ1203","SJ1301","SJ1302","SJ1303","SJ1401",
-              "SJ1402","SJ1501","SJ1502","SJ1601","SJ1602","SJ1701","SJ1702","SJ1703","SJ1801","SJ1802",
-              "SJ1901","SJ1902","SJ2001","SJ2003","SJ2002")]
 ```
@@
   mutate(site = as.factor(paste0(loc, '_', elev)))
+
+# SJ only site 
+ft <- ft %>% 
+  mutate(site = case_when(
+    loc == 'SJ' ~ 'SJ',
+    TRUE ~ as.character(site))
+  )
 ```
@@
 bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
+
+bai_sj1 <-bai_piovesan(rwdf = sj, diam_df = d_sj)
+
+str(sj_tree)
+str(d_sj)
+
 bai_ca <- bai_piovesan(rwdf = ca_tree, diam_df = d_ca)
@@
 
-write.csv(correla, file=paste(di, "data/correla_ventanas_temporales/correla.csv", sep=""), row.names = FALSE)
+write.csv(correla, file=paste(di, "data/correla_ventanas_temporales/correla_chronos.csv", sep=""), row.names = FALSE)
 

### 2018-10-16 script/R/chrono_bai.R
add na.rm TRUe

@@ -21,7 +21,7 @@ chrono_bai <- function(bai_object){
   # Compute mean, sd, se and sample deepth (n_trees)
   out <- b_aux %>% 
     group_by(year) %>% 
-    summarise(bai_mean = mean(bai), 
+    summarise(bai_mean = mean(bai, na.rm = TRUE), 
               bai_sd = sd(bai), 
               bai_se = bai_sd / sqrt(n()), 
               n_trees = n()) %>% 

### 2018-10-16 man/figures/location_map.Rmd
update figure location map

--- 2018-10-16 man/ms_figures.Rmd
@@
     highlight: null
-    reference_docx: template_ms_new_page.docx
-bibliography: ../refs/references.bib
-csl: ../refs/ecology.csl
+    reference_docx: ../../templates/template_pagebreak.docx
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecology.csl
 ---
@@
 
+- [] Figure 5. Residual chronologies --> Supplementary
+
 ")
@@
 
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
+![](../../out/fig/geomap/mapa_situacion.jpg){height=600px}
 
@@
 ```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# To get p.value of the model 
-# https://stackoverflow.com/questions/5587676/pull-out-p-values-and-r-squared-from-a-linear-regression
-lmp <- function(modelobject) {
-    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
-    f <- summary(modelobject)$fstatistic
-    p <- pf(f[1],f[2],f[3],lower.tail=F)
-    attributes(p) <- NULL
-    return(p)
-}
-
-# Lm CA-High
-dfCaH <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-mCaH <- lm(bai_mean ~ year, dfCaH)
-summary(mCaH)
-r2caH <- format(summary(mCaH)$r.squared, digits = 3)
-pCaH <- lmp(mCaH)
-
-# Lm CA-Low 
-dfCaL <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-Low')
-
-mCaL <- lm(bai_mean ~ year, dfCaL)
-summary(mCaL)
-r2caL <- format(summary(mCaL)$r.squared, digits = 3)
-pCaL <- lmp(mCaL)
-
-# Lm SJ
-dfSJ <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'SJ')
-
-mSJ <- lm(bai_mean ~ year, dfSJ)
-summary(mSJ)
-r2SJ <- format(summary(mSJ)$r.squared, digits = 3)
-pSJ <- lmp(mSJ)
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas correspond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show years replicated with # series > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r r2caH`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-Low'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'SJ'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE)
-
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-10-16 man/ms.Rmd
update

---
title: "Relict populations of a deciduous oak show high resilience to drought despite land-use legacies"
author: "AJ Perez-Luque; G. Gea-Izquierdo; ..."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
```

```{r setup, echo=FALSE}
show_text <- FALSE
```

# Abstract 
* Trees showed resilience to drought events, despite the ‘a priori’ vulnerability that they can express as a result of the impact of climate change on relict stands resulting from intensive land-use in history.

* Oak resilience to drought followed a dryness gradient: Northern populations, which thrive under a drier climate, were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.

# Introduction

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017]. Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b]. Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013], specially in the Mediterranean area [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017]. In the last decades the assessment of the drought impacts on ecosystems has gained attention [@Allen2010; @Clark2016], especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014]. The response of vegetation to drought has been analyzed at different scales, from global [*e.g*. @Allen2010; @VicenteSerrano2013; @Norman2016] to local scales [see @MartinezVilalta2016 for a revision].

Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017], and some authors states that the impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017]. This is specially relevant for areas with a long history of landscape modification as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016]. For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.

The impacts of drought are also especially relevant for populations located in the rear edge of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015]. The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in their rear-edge of their distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and tree-ring [@Jump2010].

Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Zribi2016].

Tree-ring width is a widely used proxy for tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Gazol2018]. Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, considered a rear-edge of their distribution. We are also interested in assessing resilience of these forest to several extreme drought episodes. We hypothesized that, relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyreancia* forests in their rear edge; (*ii*) to analyze the resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

```{r echo=FALSE, results='asis', eval=show_text}
cat("

In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017]. The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017]. These events were extreme in both its magnitude and spatial extent [@Gouveia2014].

These climatic alterations are likely to have important consequences for tree species dynamics at local and regional scales ( Peñuelas and Boada 2003 , Van Mantgem et al. 2009 , Matías and Jump 2015 ).

### Sequías 
> notas sequia
> 
> * the drought events have been longest and most severe in the period 1991-2010 for mediterranean are of Southern Europe  
> * Seasonally, drought frequency is projected to increase everywhere in Europe for both scenarios in spring and summer, especially over southern Europe, and less intensely in autumn [@Spinoni2017a]
> Aumento de la frecuencia de sequías en spring and summer desde 1950 hasta 2014 [@Spinoni2017b]
> Both for frequency and severity, the evolution towards drier conditions is more relevant in the last three decades over Mediterranean area in summer, 
> an increase in the drought severity in the Iberian Peninsula has been observed in the last decades [@VicenteSerrano2014].

Althought several works have reported these two years as some of the worst drought events, we characterised the drought at several spatio-temporal scales in the study area. From a long-term perspective, we compare the accumulated monthly precipitation at a meteorological station (Granada, Base Aérea) during the hydrological years 2004-2005 and 2011-2012 with the average of accumulated monthly precipitation for the period 1950-2015.

### Extreme sequias y ecologia

No solo efectos aislados, sino también en conjunción con otros factores, sobre todo en el mediterráneo donde convergen muchos de los factores que pueden interaccionar:

* La sequía es un factor crucial a tener en cuenta, ya que es además de los efectos que puede tener de forma aislada, se ha visto que además presenta muchas interacciones con otros factores, siendo por tanto un factor crucial (`$IMPROVE$` ver [@DoblasMiranda2017]):
* Some interactions alter the effects of a single factor, as drought enhances or decreases the effects of atmospheric components on plant ecophysiology
* Drought and land use changes, among others, alter water resources and lead to land degradation, vegetation regeneration decline, and expansion of forest diseases.
* Climate change, and especially drought, emerges as a crucial factor in most of the reviewed interactions and therefore it should be considered when it comes to designing and applying international management policies
* Drought should be considered when designing and applying management policies.

Además la sequía se espera que tengan ...Droughts are most likely to have the largest and most long-lasting impacts globally due to large indirect and lagged impacts and long recovery especially for forest ecosystems (ver  18 en @Ummenhofer2017). Así por ejemplo se ha visto que todos los biomas presentan una vulnerabilidad similar ...
En una revision sobre la vulnerabilidad del sistema de trasnporte en plantas al embolismo inducido por sequía, ha mostrado que una convergencia de la vulnerabilidad de los bosques a la sequía, mostrando que todos los biomas son igualmente vulnerables a los fallos hidráulicos independientemente del régimen de precipitación (Choat et al. 2012, doi:10.1038/nature11688)

")
```

# Materials and methods

## Species and study site

Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils, or soils poor in basic ions [@Vilches2014]
under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002]. *Q. pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species [@Gavilan2007; @delRio2007].

This species reaches its southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight oak patches (2400 Has) identified (Figure 1) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].

The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], undergoing intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. In fact, conservation status of this species for southern Spain is "Vulnerable" [@Vivero2000]. The relict presence of this species in Sierra Nevada is related both to its genetic resilience and high intraspecific genetic diversity [@Valbuena2010]. However, they are also expected to suffer the impact of climate change, due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species. Thus, simulations of the climate change effects on this habitat forecast a reduction in suitable habitats for Sierra Nevada [@Benito2011].

## Drought episodes

During the second half of the XX century, several extreme drought episodes were recorded for the Iberian Peninsula (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3). We focused on these two drought events because limitations about temporal availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events. The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by Spinioni (see Appendix S3)

```{r echo=FALSE, results='asis', eval=show_text}
cat("
* Ver ejemplo Gazol et al 2017 Plan Ecology
* un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
")
```

## Greenness data

Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] and they are linked with primary productivity [@Ruimy1994]. Here, vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. MODIS EVI Data (Collection 6) were obtanied using Google Earth Engine platform for the period 2000 - 2016. EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Then, a data filtering was applied to select EVI valid values. For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

```{r echo=FALSE, results='asis', eval=show_text}
cat("
We used EVI instead of NDVI (*Normalized Difference Vegetation Index*; another of the most common greenness vegetation indices) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].

")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2). Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked. Besides, VI usefulness indices between 0 to 11 essentially include all EVI data. Thus, these two conditions serve as additional checks.

and then a quality assessment was carried out to filter the ... [@Reyes2015]

    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds. On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]

The presence of clouds (adjacent clouds, mixed clouds and shadows) 'obscures' the surface in a radiometric sense, thus corrupting inferred EVI values. In addition, two types of aerosol loadings typically corrupt EVI—climatology and high aerosols. Use of aerosol climatology

The EVI data are geometrically and atmospherically corrected and include information about the quality ass....

These data are geometrically and atmospherically corrected, and include an index of data quality (reliability, which range from 0 – good quality data – to 4 – raw data or absent for different reasons) based on the environmental conditions in which the data was recorded

We first used the Quality Assesment (QA band) information of this product to filter out those values affected by high content of aerosols, clouds, shadows, snow or water; and then a quality assessment was carried out to filter the ... [@Reyes2015]

:red_circle: reescribir esto de la calidad. Ver Samanta et al 2012 y como describe el proceso de calidad

")
```

To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as:

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2010; @Samanta2012].

As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016. $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale. The nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945) was applied using the `trend` package in R [@Thorsten2017]. The alpha level for the test was set to 0.05.

```{r echo=FALSE, results='asis', eval=show_text}
cat("
(:red_circle: Hemos seleccionado EVI medio, además de por los consejos que me ha dicho Domingo, porque he comprobado que existe una correlación entre el evi medio y el evi estacional, sobre todo el de verano. Ver esto: [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md). Además presenta alta correlaciones significativas con el EVI de verano: 0.88; de primavera: 0.76 y anual: 0.81)

> Procedimiento de Filtrado de datos (ver [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md))

* Información contenida en banda QA.

    * Nos quedamos con pixeles marcados como Good Data (57.89 %)
    * Filter out los marcados como Snow/Ice y/o Cloudy (2.57 + 7.08 = 9.65 %)
    * Pixeles marcados como Marginal Data (32.33 %) (ver siguiente paso)

* Explorar distribución temporal y analizar banda QA Detailed y llevar a cabo un filtrado siguiendo las especificaciones de @Reyes2015.

    * Vemos los composites marcados con Aerosoles, Adjacent cluods, y Shadow.  
    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds. On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]

* Finalmente nos hemos quedado con las siguientes cifras. De un total de 360064 images composites for the study zone were downloaded (928 x 20 x 1 + 928 x 23 x 16 = 360064), tras el filtrado, nos quedamos con 286825 (79.65 %)

")
```

## Field sampling and dendrochronological methods

### Tree sampling 
Tree sampling was carried during autumn of 2016. Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ), xeric site located at northern aspect; and Cáñar (CA), humid but warm site located at southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High. All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014].

### Dendroecological analyses 
Tree cores were air dried, glued onto wooden mounts and sanded. Annual radial growth (ring width, RW) was measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally done using COFECHA software [@Holmes1983].

The growth trends were analyzed at different time scales. To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008; @Bunn2010] package. Residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990]. The statistical quality of each chronology was checked via the expressed population signal. A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable [@Wigley1984].

The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). Theoretically, BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008]. Raw ring-widths and measured DBH were used to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008]. We used the following equation:

$$\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$$
where $r$ is the radius of the tree and $t$ is the year of tree-ring formation. For each individual tree, a mean BAI serie was calculated. Then, mean site BAI chronologies were obtained by averaging individual tree BAI time series.

Both RWI and BAI chronologies were compared to different climatic indices to analyse the low and the high-frequency response to climate of trees. Statistical descriptive parameters, including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology [@Fritts1976].

### Climate and growth

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018]. The Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014], was also used to characterize the drought conditions for the period 1961-2014. SPEI values with a temporal scale of 6 months and a spatial resolution of 1.1 km, were obtanied from the Drought indices dataset (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017].

```{r echo=FALSE, results='asis', eval=show_text}
cat("
SPEI values between 1 and −1 are considered normal, whereas values <−1 indicate drought and values <−1.5 indicate severe drought47. We defined a drought as a period where SPEI12 decreased to ≤−1.5.
")
```

Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using the `treeclim` package [@Zang2015]. The bootstrapped confidence intervals were used to estimate significance (p < 0.05) of the correlation coefficients.

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g. logging) in the past. By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].

```{r echo=FALSE, results='asis', eval=show_text}
cat("
This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
")
```

Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:

$$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median. Medians were used since they are more robust estimator of central tendency than means [@Rubino2004; @Camarero2011]. Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Disturbance chronologies were built using tree-ring width to identify historical abrupt positive (releases) or negative (suppressions) changes in growth in each species (Nowacki and Abrams, 1997). This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
* Para obtener los datos de SPEI, hemos ido bajado los datos desde http://sac.csic.es/spei/map/maps.html 
* sanjuan: 37.133, -3.382 = spei_-3.25_37.25.csv 
* canar: 36.957, -3.4284 = spei_-3.25_36.75.csv

OJO ver http://monitordesequia.csic.es/
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
To obtain ring-width series of each individual and species (species chronology) the cross-dated ring-width series were standardized and detrended using the dplR library (Bunn 2010) in the R statistical language (R Development Core Team 2015). First, we fitted a spline through the ring-width series of each measured radius. We used a spline where the frequency response is 0.5 at a wavelength of 0.67 multiplied by the length of the series in years (Bunn 2010). Then, the ring-width data were divided by the fitted lines to obtain residual ring-width indices (RWI) that were subjected to autoregressive modelling (pre-whitening) and then averaged for each year using a biweight robust estimation of the mean (Fritts 1976). Calculating RWIs using subtraction instead of division rendered similar results as those here presented. Thus, we obtained mean residual chronologies for each species of prewhitened growth indices.

 * BAI represents a more accurate indicator of growth than ring-width, since it reduces the variation caused by adding volume to a circular stem 
 * BAI is a more meaningful indicator of growth than ring width because it removes variation in growth attributale to increasing circumference. 
 * BAI is more closely related to biomass increment 
 * BAI removes variation in growth attributatle to increasing stem circumference and captures changes in growth better than linear measures as tree-ring width [@Biondi2008] 
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
 To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
")
```

In addition, the forest history of San Juan and Cáñar sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc. Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@MartinMontanes2015; @RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].

## Resilience

To evaluate the effects of drougth events on greeennes and tree growth (BAI) resilience indices proposed by @Lloret2011 were used: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and before the disturbance (drought) event:

Resistance (*Rt*) = Drought / Predrought

The Recovery index (*Rc*) is the ability to recover from disturbance relative to its severity, and it is estimated as the ratio between performance after and during disturbance:

Recovery (*Rc*) = Postdrought / Drought

The Resilience index (*Rs*) is the capacity to reach pre-disturbance performance levels, and it is estimated as the ratio between the performance after and before disturbance:

Resilience (*Rs*) = Postdrought / Predrought

```{r echo=FALSE, results='asis', eval=show_text}
cat("
The Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance, and it is estimated as:

Relative Resilience (*RRs*) = (Postdrought - Drought) / Predrought
")
```

The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Appendix S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context the resilience observed for 2005 and 2012 drought events, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 (Appendix S3).

## Statistical analysis

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages. The level of significance was set to 0.05 and adjusted for multiple comparisons. All analyses were carried out in R software [@R2017] using several packages (see Supplementary S?)

# Results

## Time trends in vegetation greenness

The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period. During the 2005 the lowest values of EVI standardized anomalies were recorded, particularly for northern populations (Figure 2). In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q. pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Appendix S4).

```{r echo=FALSE, results='asis', eval=show_text}
cat("
The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends. The strongest trends were observed in southwestern populations (Figure S2), in which the Cañar site are included Vegetation greenness of *Q. pyrenaica* forests were lower during both the 2005 and 2012 years than the greenness observed for the reference period (Figure 2a). The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b). Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought (Figure 2b).

According to the standardized anomalies, *Q. pyrenaica* forests suffered a browning episode during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S3). 
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005. Sin embargo vemos valores menos negativos para 2005. Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years

> Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029) que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
## Greenness resilience to drought events

*Q. pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3). Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figure 3).

Recovery (*Rc*) of greenness was significnatly different bewteen drought events and sites (Tables 3). In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1). For southern populations, a similar recovery after the 2005 and 2012 drought event was observed (p = 0.2453; Figure 3; Table S1).

Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables S1, 3), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1). The southern populations showed higher resilience values than the northern ones, although these differences were not significant for the 2005 drought event (p = 0.036; Figure 3).

> Resistencia
> 
> * Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001]
> * Menor resistencia de las poblaciones del Norte a los eventos de sequía que las del Sur [N: 0.883 (0.877-0.889); S: 0.921 (0.918 - 0.925); p <0.0001]
> * La resistencia varió en función de la sequía y de la población. Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
> * notas:  La sequía de 2005 redujo el EVI medio hasta el 85.8%, mientras que la de 2012 la redujo hasta el 94.3%. La sequía de 2005 redujo el evi medio hasta el 81.9 % en las poblaciones del N, y hasta el 90.2 % en las del S. ... drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)

> Recovery 
> 
> * La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
> * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
> * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001]. En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.

> * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
> * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152). En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)

> Relative Resilience (De esto no he dicho nada)

  * Los robledales mostraron mayor resiliencia relativa a la sequía de 2005 que a la de 2012 [2005: 0.099 (0.095-0.105); 2012: 0.053 (0.050 - 0.056); p <0.0001]  
  * Las poblaciones del sur mostraron menor resiliencia relativa que las del norte [N: 0.086 (0.082-0.092); S: 0.063 (0.060 - 0.066); p <0.0001], debido sobre todo a la diferencia en la resiliencia relativa para las poblaciones del norte entre los dos eventos de sequía (mucha mayor resiliencia relativa en 2012 que en 2005 para las poblaciones del N)
  * Las poblaciones del sur no mostraron diferencias en cuanto a la resiliencia relativa entre los dos eventos de sequía (padj = 0.152)
")
```

## Radial growth trends

Likely as a consequence of their different management origin (*i.e.* land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2). Competition was similar among sites but plot basal area was greatest in CA-High (Table 1). Yet the basal area increment (BAI) growth in the highest plot in altitude (CA-High) was the greatest (Figure 4). In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations. For southern population differences in growth were observed, with CA-Low site showing lower values than CA-High but similar to growth values of the northern population (SJ) (Figure 4; Appendix S7).

For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low). Yet, the lowest pointer year (*i.e.* the lowest growth) since 1950 was 1995 in all chronologies (Figure 5).

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Sorprende que en los sitios del sur (caH y caL), en los últimos años ha habido crecimientos muy grandes. Por ejemplo, en 2010 se observó uno de los mayores valores de RWI de toda la cronología tanto en caH como en caL. Algo similar ocurrió en sj para los años 2003 y 2013. 
")
```

## Resilience of greenness and radial tree-growth to drought events

*Q. pyrenaica* forest showed significantly lower resistance values (*Rt*) to the 2005 drought event than to that in 2012 for greenness and for radial growth (Table 3; Figure 3). The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2). Resistance values to drought for greenness and tree-growth varied between sites (Table 3). Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2). It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).

Recovery (*Rc*) of greenness and growth were significantly different bewteen drought events and sites (Tables 3). In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1). An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 3, Table S2). Northern populations showed significantly higher values of recovery than southern sites for greenness and tree-growth (Table S2). For southern populations, no significant differences were found for recovery of greenness after the 2005 and 2012 drought event (p = 0.2453; Figure 3; Table S1). Recovery values for tree-growth of southern populations were below or close to 1 (Figure 3, Table S2).

Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 3). Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3). Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2). For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 3). For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2).

The exploration of resilience metrics of tree-growth for other severe droughts, revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S5). A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155). Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S5) even not being the most severe drought events.

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Resistance:

Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2). Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).

* No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012 (F = 6.0189; p = 0.019). Valores de resistencia menor al evento de 2005. [2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.025; trimmed-p = 0.019]

* Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001). Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].

* La interacción también fue significativa. De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ???

> Recovery 
> * Mayor recuperación del crecimiento de los robledales tras la sequía de 2005 que tras la de 2012 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
> * Las poblacionees del sur (caH y caL; no diferencias entre ellas) mostraron una menor recuperación que las poblaciones del norte, es decir, las del norte se recuperaron mas rápidamente que las del sur que no llegaron incluso a valores de Rc = 1.

Similar pattern of resilience (*Rs*) values was found for growth than for greenness respect to drought event: significantly higher values of *Rs* for the 2012 drought event than for the 2005 (Table S2; Figure 3). However, no differences were observed between sites (Table 3). For 2005 drought event, *Rs* value of SJ (northern site) was lower than that of southern ones (CA-High and CA-Low), but opposite pattern was found for the 2012 drought event. All values of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031).

") 
```

## Climate and tree-growth
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 6a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 6b), specially for the northern population (r > 0.6). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 6c and 6d). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees.

## Forest disturbances
The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8). Northern site (SJ) showed two release events ($GC$ > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed some weak suppression episodes, but not in the last 50 years.

```{r echo=FALSE, results='asis', eval=show_text}
cat("

En las poblaciones del norte osbervamos dos eventos de PGC por encima del 25 e incluso el 50 %. El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001. Estos periodos se alternan con periodos de NGC: previo 1940, 1960 - 1991, y quizá aunque no lo vemos en al actualidad (?? duda. Esto es especular un poco). Esto responde creo que a:

(i) actividad minera y tratamientos forestales. La actividad minera en la zona fue intensa (aunque a veces intermitente, entre 1890 y 1957 (cuando se clausuró); y previamente desde 1858 tras los trabajos de Amalio Maestre). Hay costancia de la existencia de hornos de fundición de mineral en alguna de las minas (La Probadora).
(ii) actividades forestales:
* Se conoce la existencia de un proyecto redactado en 1943 de repoblación forestal de la cabecera del río Genil (he pedido los mapas).
* Ver actividades forestales (bd proyectos de gestión)

En el trabajo de [@GeaIzquierdo2014] se observa una liberación posterior a 1950 (QPUY9) (entre 1950 y 1960 aprox.) y luego una supresión hasta mediados de 1990. Esto coincide con lo que hemos observado en nuestra chrono.

Respecto a las poblaciones del sur, destaca en la caH una primera liberación en torno a 1829-1830 y 1837-1840; y luego predominan los eventos de supresión que se alternan con algunos de liberación (muy débiles) hastas 1950. Posteriormente se alternan liberaciones-supresiones débiles. Destaca una liberación en torno a 2000-2003 (relación con clima??? posterior a sequía 1999???)

En caL, observamos un periodod de limitación hasta 1885 aprox. Luego se alternan al igual que antes periodos de liberación con otos de supresión con señales más fuertes que en caL, y al igual que antes destaca con mas intensidad si cabe, una liberación entre 2000 y 2003 (antrópico o climática ??)

Duda: ocurre en los dos sitios caL y caH, entonces, es aumento de crecimiento debido a actuación forestal? o debido a clima??

[@GeaIzquierdo2014] encontró un pico de disturbance en torno a 1860 en QUPY10 (cañar) sugiriendo que la baja densidad en esos sitios fue presumible el resultado de un thinning (aclareo). El patrón de supresión que observa para este sitio va en la linea de lo que nosotros encontramos (supresiones que alcanzan a veces el 25 % y otras no)

En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles. En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles 
")
```

# Discussion
## Impacts of severe droughts on rear-edge populations

Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction in greenness and especially in the growth of *Q. pyrenaica*. During the 2005 drought, one of the worst drought events recorded in the Iberian Peninsula [@GarciaHerrera2007], we found a browning of the *Q. pyrenaica* forests, but no changes in EVI standardized anomalies were recorded for 2012, which can be explained because 2012 drought event was a winter-drought [@Trigo2013]. 
Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure 5). The decline in growth observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].

## Oaks show high resilience in response to recent drought events
Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Appendix S3), we found a positive trend for vegetation greenness of the forests of *Q. pyrenaica* located in their rear edge. Our results agree with those obtained by previous works using other remote-sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi], which suggests an increase in primary productivity during the last years for rear-edge populations of this species.

For tree-growth, a similar positive trend was observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). This result differs from those previous reported for *Q. pyrenaica* along their distribution range [@GeaIzquierdo2013; @GeaIzquierdo2014]. @GeaIzquierdo2014 found a general decline in the growth of this species since the 1970s, particularly sharp for populations located in their dry-edge. This decline trend in growth have also been oberved for other tree species located in their rear-edges [*e.g.* @SanchezSalguero2012; @SanchezSalguero2017; @Dorado2017]. Furthermore, growth projections have forecasted a decrease in productivity for *Q. pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013]. However, our results for similar locations have revealed a recovery in tree-growth for the last years (Figure 4). In addition, a similar positive trend in BAI for the last years, has been reported for another *Q. pyrenaica* population of the Sierra Nevada [@RubioCuadrado2018].

Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 3). La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Appendix S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q. pyrenaica*. Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía (long-term perspective), observamos como estas poblaciones presentan altos valores de resiliencia. Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).

Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q. pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Appendix S5). Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005]. Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas). Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q. pyrenaica* situadas en su rear-edge.

Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018]. En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q. pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático. Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q. pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017].

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Hemos observado como las sequías severas, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento de *Q. pyrenaica*. Los datos de anomalías estandarizadas de EVI mostraron un browning para la mayoría de los robledales de Sierra Nevada durante la sequía de 2005.

La sequía de 2005 ha sido una de las mas severas de los últimos años, afectando significativamente al crecimiento

#--------------------------------------------
# Sequías

### 2005 frente a 2012

* Analizando los datos de precipitación acumulada de las estaciones alrededor de SN, observamos como 2005 fue el año mas seco desde 1950. 
* Datos de SPEI (0.5 º Grid) 
    * 2005 fue peor que 2012, pero no peor que 1995 o que 1999 (ver figura S2)
    * Tabla S4. Severidad sequías, la de 2005 fue menos severa y menos intensa y de menor duración que la de 2005 (de donde vienen estos datos???)

* La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento. Esto se ha observado también en otras especies en el sur de la P. Ibérica (p.ej. P. nigra en Andalusia [@SanchezSalguero2013], ... INCLUIR otras especies y citas).

### 2005 en contexto

* Algunos trabajos hablan que los peores años de sequía en la P. Iberica durante la segunda mitad del siglo XX fueron 1986, 1994-1995, 1999, 2005 y 2012

* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].

### Reducción en el crecimiento de algunas especies 
El efecto de la sequía de 2005 en el crecimiento también se ha visto en otras especies en el sur de la P. Ibérica 
La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento. Esto se ha observado también en otras especies en el sur de la P. Ibérica (p.ej. P. nigra en Andalusia [@SanchezSalguero2013], ... INCLUIR otras especies y citas).

* P. nigra: Se observó una gran reducción de crecimiento en 1994-1995, 1999 y 2005 para Pinus nigra en el SE de Andalusia [@SanchezSalguero2013] coincidiendo con años de sequía para esa zona. La máxima reducción en crecimiento fue observada para 2005, siendo significativamente mayor que para el resto de años [@SanchezSalguero2013].

No obstante, a pesar de que 2005 y 2012 fueron dos de las sequías mas severas registradas para el sur de la P. Ibérica [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), nuestros resultados muestran que los robledales de *Q. pyrenaica* en su rear-edge presentan una tendencia positiva tanto en el greenness (78.9 % pixeles con greeness) como en el crecimiento (tendencias positivas del BAI).

### La resiliencia de las poblaciones de roble de Sierra Nevada.

El greennness de los robledades en Sierra Nevada ha mostrado una tendencia positiva hacia mas verdor en los ultimos años, que coincide con lo que ya observamos con datos de NDVI [@PerezLuque2015onto], sobre todo para las poblaciones del sur. Esta tendencia positiva, obtenida de variables derivadas de remote sensing (EVI y NDVI), parece que también se observa en el crecimiento. Por ejemplo en las poblaciones del sur (CaHigh y CaLow) observamos en los últimos años una ligera tendencia de crecimiento.

Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia. Por ejemplo para las poblaciones del norte, los valores de EVI durante la sequía de 2005 descendieron hasta el 81 %, mientras que el BAI (sitio SJ) lo hizo hasta el 45 %. Sin embargo, la recuperación fue rápida, así los valores de Recovery (Rc) para el EVI tras la sequía de 2005 (en las poblaciones del norte) fue de 1.17, mientras que para el BAI el Rc fue de 1.112. En definitiva, estamos observando altos valores de resiliencia en estos robledales.

Despite the resilience values (*Rs*) of growth for 2005 and 2012 were close or below 1, we observed high resilience of these forests from a long-term perspective. As our results indicate, the highest resilience values for growth were recorded for two of the most severe drought events (1995 and 1999), with resilience values above 1.2 in both events (Appendix S5).

genetica

Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático. Hemos observado un aumento en el greenness en los últimos años. Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness. Además, estos robledales tienen una alta resiliencia genética  ¿Sierra Nevada (regiones de montaña) como refugio?? quizá este rear-edge esté actuando como refugio?? (esto es muy especulativo)

Las zonas rear edge son zonas que se supone son mas vulnerables al cambio climático por sus estrechos márgenes. De hecho como han visto algunos trabajos, las especies situadas en su rear edge son mas vulnerables. No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado). Además habría que introducir aqui lo que M. Villalta habla de la consideración de rear edge de.

* Otro punto de interés a introducir aquí es que frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], sin embargo algunos estudios están demostrando esto no es siempre así (ver por ejemplo [@Cavin2017; @Granda2017]). Nuestros resultados creo que van en esta línea creo. Esto, como apunta [@MartinezVilalta2018]m tiene que ver con la que consideramos como habitat marginal de la especie (... When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).

¿Casos parecidos?

Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling.. // dorado-liñan // sanchez salguero

trabajos rear edge: 
 * sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
 * sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
 * dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
 * herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
 * dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
buscar otros

... Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...

## refugio??

Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático. Hemos observado un aumento en el greenness en los últimos años. Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness. Además, estos robledales tienen una alta resiliencia genética [@Valbuena2013; @Valbuena2017]. ¿Sierra Nevada (regiones de montaña) como refugio?? quizá este rear-edge esté actuando como refugio?? (esto es muy especulativo)

pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.

")
```

## Site environment shapes differential sensitivity to climate and drought of rear-edge oak populations

Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3). Las poblaciones del norte, que presentan condiciones mas secas (Tabla 1), se vieron mas afectadas por los eventos de sequía. For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2). In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 6), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].

Moisture availability has been reported as the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014]. We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 6a). Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our results are consistent with previous findings that pointed out tree growth and tree responses to drought are site-dependent (e.g. soil features, tree competence, etc.) [e.g. @Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017c].

When we explored the chracteristics of the southern sites, we firstly observed a difference in elevation. @Bhuyan2017, in a recent work across the European continent, found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. In fact, the differences in available water of soils could explain our results. @CoboDiaz2017, in a work analyzing the variation of the microbiota along an elevational gradient within this location, found higher values of available water of the forest soils located on high elevation compared with those located at low elevations. In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply. In most of the high-mountain watersheds of Sierra Nevada, there is a system of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016]. These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017]. The network of traditional ditches present in one of our sampling site (Cáñar) was recentlty described, with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015]. There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017]. It brings water from snowmelt and from a nearby stream, soaking the *Q. pyrenaica* forests. This traditional system could supply an extra of water that could be used by the trees located downstream of these channels. This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site.

## Land-use legacy effects shape distribution and sensitivity to climate change of read-edge oak populations

Although our sampled sited showed similar tree competence levels (Table 1), the review of historical documents showed that la historia de manejo humano ha sido diferente para cada sitio. Different values of tree-growth were observed between populations and for different elevations within southern location (Figure 4). Low-elevation sites showed lower BAI values than higher-elevation site. A reason that could explain these results is the different degree of anthropic impact related to elevation. Los bosques situados a mas elevación presentan menor presión antrópica que los localizados a menor altitud, por el simple hecho de que estos últimos están mas accesibles. During the last decades, the rural abandonment has lead an decrease of anthropic pressures on the high-elevations zones of Mediterranean mountains (INLCUIR CITA).

comentario --> Aqui pretendo incluir la idea que hablamos en su momento. Quizá están creciendo donde les están dejando crecer (menor impacto antrópico en las zonas mas altas, al menos en los últimos años -abandono de zonas rurales?-)
He ralizado una revisión de trabajos que utilicen dendro de Q. pyrenaica para ver si existe una relación entre la edad de las cronos con la elevación.

* Son bosques relativamente jóvenes 
comentaio --> 
Aquí me falta desarrollar este apartado algo más incluyendo cosas de los documentos históricos (la minería, etc, etc) 
Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5). La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona. Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.

```{r echo=FALSE, results='asis', eval=show_text}
cat("

#### Disturbance
Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).

## historia forestal de ambos sitios 
Incluimos lo que conocemos de la historia forestal de los sitios?? Me explico, tenemos datos de manejo y uso antrópico de las dos zonas que proceden de varias fuentes. En resumen, mas o menos, tenemos:

* Minería: datos de minería, que afectan sobre todo a la población de SJ (los tenemos localizados temporalmente)
* Actuaciones forestales: 
* Incendios, Carboneo, Ganadería, etc --> Tenemos información menos estructurada sobre estos ámbitos, procedentes de varias fuentes, que tienen una incertidumbre mayor espacial y temporalmente. 
")

```

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution. Para  *Q. pyrenaica* se ha visto que moisture availability es el factor limitante que determina el crecimiento en las poblaciones situadas en su rear edge [@GeaIzquierdo2014], pero para otras especies (*i.e.* *Pinus nigra*, *P. sylvestris*) la temperatura tiene mas peso que la disponibilidad de agua  [@Herrero2013; @Matias2017]. Pero es importante considerar además la historia de manejo que han tenido esos bosques a la hora de la forest management [@Penuelas2017; @DoblasMiranda2017], ya que, como observamos en nuestros resultados puede condicionar el crecimiento y la resiliencia de las especies.

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Podríamos complementarlo con lo que le pasa a otras especies en su borde de distribución: por ejemplo en Baza, Herrero et al. 2013, encontraron para Pinus nigra y sylvestris que la temperatura tenía mas peso que la disponibilidad de agua). O también ver algunos de los trabajos de Camarero et al 2013 para el P. nigra en su borde de distribución u otros similares (el de Sanchez-Salguero et al. 2013, 2015)

Para *Q. pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q. pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al. 2015 European Journal of Forest Research). Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea.

")

```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
* Vicente-Serrano: 
    * We found that some forests from cold and humid areas respond to shorter drought time-scales than forests from dry areas, which usually respond to longer time-scales
    * Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011).
    * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.

-----------

Despite similar value of competence in our study sites, we obtained higher values of growth for the highest elevation site (CA-High, Figure 4).

------

En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations). 
The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone.

")

```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
## Rear-edge

* Se asume una alta vulnerabilidad de las especies a la sequía en su límite inferior, sin embargo esta vulnerabilidad no necesariamente se mantiene: The results of Granda et al. 2018 are consistent with recent studies showing that the frequently assumed higher vulnerability of dry, rear-edge populations to drought does not necessarily hold (e.g. Cavin and Jump 2017) (from: Martinez-Vilalta 2018)

* Definición de *marginal* populations. [@MartinezVilalta2018] indica que a veces no está claro la definición de marginal populations. When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).

* Quizá estamos asumiendo que es una población marginal, pero sin embargo lo es solamente geográficamente

-----

No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado).

### Diferencias entre sitios

Llama la atención que en el rear-edge de la distribución de la especie, en el sitio mas meridional dentro de este rear edge, y en la parte mas alta (estas parcelas están en el treeline de la especie en SN (en torno a 1900)), es donde encontramos mayores crecimientos. Parece que los robles estuvieran mejor en esta zona. Algunas reflexiones sobre esto:

* Quizá están creciendo donde les están dejando (menor impacto antrópico en las zonas mas altas, al menos en los últimos años??)
* ¿Diferentes niveles de compentencia?. Parece que la competencia es similar en los tres sitios: no diferencias sig. para valores de Stand density ySize ratio proportional to distance (ver table 1). 
* ¿Diferencias entre suelos?? En principio no. [@CoboDiaz2017] en un trabajo sobre microbiota del suelo, explora diferencias en un gradiente altitudinal en Cáñar. Analiza 3 sitios: por encima del treline (XZF, el piornal), low altitudinal oak forest (LAF, en la zona baja del robledal) and high altitudinal oak forest (HAF). En su trabajo, LAF está mucho mas bajo, pero HAF está cerca de las parcelas caHigh. Restulados:  
    * tipo de suelo: LAF es Sandy-loam, HAF es loam
    * HAF tuvo los valores mayores de disponibilidad hídrica (% available water tabla s1 en [@CoboDiaz2017]) --> Esto es importante, creo
    * Los tres sitios eran pobres en materia orgánica, pero el HAF dobló los valores de los otros dos sitios
    * No diferencias en contenido de N, C/N ratio similares en HAF, y LAF

* ¿Menor estrés hídrico en zonas mas altas? Puede ser que tengamos menor estrés hídrico en esta zona?? Este robledal, está en la cara sur de SN, y en una ladera con bastante insolación. Algunos trabajos antiguos (de fitosociólogos) hablan de que su presencia aquí se debe a que reciben un aporte extra de humedad procedente de las brisas del mediterráneo, para suplir el mínimo de humedad que necesitan en verano. Quizá también tendríamos que incluir el papel de las acequias. La zona caHigh tiene una acequia muy cerca (y por encima) de donde muestreamos (recordad el roble mas grande, y el mas alto, esta justo en una acequia). No se si es interesante que lo comentemos.

En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations). 
The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone.

In addition, despite southern sites are very close to each other, we obtained higher values of BAI for the high-elevation site than for the low-elevation site.

")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
### OTRA COSA

In fact, several authors have raise the need to consider land use y drought de forma conjunta en las environmental y forest management .... (ver Peñuelas et al 2017 y tb Doblas Miranda )

Environmental and forest management policies should take into account all these characteristics of Mediterranean forests and the social, environmental and climatic conditions that are projected for the coming years and decades.
")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
#### Climate and tree relationship

Frase de resultados que tenemos que poner en discussion: 
Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for the northern population (r > 0.6), ... **which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]**.  
")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
### ¿Que factor es mas limitante para el crecimiento en el rear-edge de Q. pyrenaica? 
Aquí pueder ser interesante comentar algo de el peso de las variables climáticas en el crecimiento para poblaciones situadas en el borde de distribución (pesa mas la temperatura o la disponibilidad de agua?).

Para *Q. pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q. pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al. 2015 European Journal of Forest Research). Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea.

Podríamos complementarlo con lo que le pasa a otras especies en su borde de distribución: por ejemplo en Baza, Herrero et al. 2013, encontraron para Pinus nigra y sylvestris que la temperatura tenía mas peso que la disponibilidad de agua). O también ver algunos de los trabajos de Camarero et al 2013 para el P. nigra en su borde de distribución u otros similares (el de Sanchez-Salguero et al. 2013, 2015) …

## MAs cosas diferentes 
Otras especies: rear edge

Evaluar el crecimiento en el rear edge de F. sylvatica 
https://link.springer.com/article/10.1007/s10342-016-0982-7?wt_mc=Internal.Event.1.SEM.ArticleAuthorOnlineFirst

¿Que limita al crecimiento en el borde sur de su distribucion? --> Ver esto 
http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2486.2006.01250.x/abstract

Pinus sylvestris (Baza) (Matias et al 2017)
Radial growth was maximal at medium altitude and treeline of the southernmost populations. Temperature was the main factor controlling growth variability along the gradients, although the timing and strength of climatic variables affecting growth shifted with latitude and altitude.

http://onlinelibrary.wiley.com/doi/10.1111/gcb.13627/full

Idea --> algunos autores comentan que puede existir una alteración en el balance competitivo entre especies en mixed stands. Por ejemplo en Montseny, borde equatorial para F. sylvatica, se está viendo que el F. sylvatica está siendo reemplazado por Q. ilex. ... En SN, el artículo de B. Benito (Climatic Change) habla del remplazo que existirá de Q. pyrenaica por Q. ilex, sin embargo, estamos viendo que los crecimientos son muy grandes y que Q. pyrenaica tiene mucha resiliencia --> entonces que pasa con las predicciones de dichos modelos ???

The presence of *Q. pyrenaica* forests on the southern slopes of Sierra Nevada is partly explained by the additional contribution of humid air from the Mediterranean sea [@Prieto1975]

")
```

En este sentido, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia. Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética).

```{r echo=FALSE, results='asis', eval=show_text}
cat("

### ORDENAR ???

.. ver [@VicenteSerrano2013] impacto de las sequías en diferentes zonas (i.e montañas)

# Otras notas: 
* To guide sustainable forest management, forest researchers are asked to provide concrete answers about forest resilience in response to expected climatic trends, and extreme climatic events (Lindner et al., 2014) https://www.sciencedirect.com/science/article/pii/S030147971400379X?via%3Dihub

* Proyeccciones de crecimeinto de para *Q. pyrenaica* sugieren un declive en el crecimiento en las proximas decadas a lo largo de su rango de distribución en la P. Iberica, en diferentes escenarios climáticos [@GeaIzquierdo2013]: Predictions suggest that QUPY productivity would decline in the next decades all along its distributional range in the Iberian Peninsula for all the climate scenarios studied. Este declive debería ser mas dramático en las baja altitudes de las zonas mas calientes del sur de su distribución

--> vulnerabilidad local (related to rear-edge)
La tendencia de reducción drástica sugerida por el modelo para el dry-edge de la distribución de la especie podría expresar una aumentada vulnerabilidad de los árboles al incremento del estrés hídrico forzado por un clima mas cálido [@GeaIzquierdo2013].

Negative trends over recent years and the greater response to moisture availability found at warmer and drier low elevations in the south suggest vulnerability to warming at the local low elevation dry edge of the species's range (e.g. QUPY9). Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()

Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg. @Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al. https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI). Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).

Case studies that focus on multiple scales - including local scales - and also valuables (Dale et al. 2018 Frontiers) --> Se necesitan casos de estudio que se enfoquen en escalas multiples (incluyendo las locales) y que combinen varias metodologías (Jump et al. ...)

3) rear edge

- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):

* Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ... (Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...

### Estudios que utilizan RS y TR

* Wu et al. 2017. 
	* En un estudio sobre el efecto de la sequía sobre el crecimiento en bosques, grass y shrub del hemisferio norte, encontraron que los bosques (Deep-rooted forest) exhibit a drought legacy response with reduced growth during up to 4 years despues de un evento extremo de sequía.

	* Además vieron que los bosques mostraron una stronger drought resilience in forest (i.e. smaller growth reduction after severe drought) (Gazol el al. 2017; Wu et al. 2017).

	* Esto se puede deber a que acceden a reservas de agua mas profundas (los pastos y los matorrales no pueden acceder a ese agua)

### 
Nosotros hemos obtenido una relación entre la severidad de la sequía y la resiliencia mostrada por los robledales en su rear edge. (ver gráfica )

@Gazol2018 encontró La resiliencia está relacionada con la severidad de la sequía y con la composición del bosque

Rt-NDVI and Rt-TR were positively related with drought intensity

## 
@Gazol2018 Resistencia a la sequía mayor en la zona norte de españa que en las zonas secas del sur

Qpyr 
* Valores de Rt-TR aumentaron a lo largo del tiempo, sin embargo Rt-NDVI dismiuyó con el tiempo
* Valores Rc-NDVI tendencia positiva

")
```

Methodological approach

comentario: tengo que rematar este apartado.

Vegetation reflects the environmental conditions, and the effects of drought on vegetation can be observed using information from several methodological approaches [see @Norman2016 for a review]. Several works revealed the utility of remote-sensing [*e.g*. @Zhang2013; @AghaKouchak2015] and of the dendrochronology [*e.g.* @Eilmann2012; @Bhuyan2017] to quantify the drought impacts on vegetation at different spatial and temporal scales, but to our knowledge the combination of both approaches is scarce [@VicenteSerrano2013; @VicenteSerrano2016; @Bhuyan2017b; @Wu2018]. Estos trabajos han observado que existe de forma general a positive relationship between vegetation indices derived from remote sensing and annual tree growth [@VicenteSerrano2016; @Bhuyan2017b; @Gazol2018].

Aunque existen algunos trabajos que han utilizado RWI y remote sensing nuestra aproximación (creo) es interesante, sobre todo porque la aplicamos al estudio de poblaciones que están en su límite de distribución (rear-edge) [@Jump2010]. Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares que analicen la respuesta de las poblaciones al cambio climático y su resiliencia utilizando diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].

- Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q. pyrenaica situadas en su reaar edge frente a los eventos de sequía.

- Hemos encontrado diferencias en cuanto a la resiliencia utilizando RS y RWI. Diferente sensibilidad de satélite frente a los datos de campo. Esto también se ha observado en @Gazol2018, que analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR. Encontraron que los datos de TR son mas sensibles para la resiliencia del bosque a la sequía que los datos de RS

- De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.

- La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017] (... Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes. )

- Nota from @Jump2010: Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability. In such regions, targeted collection of high-resolution data is necessary in order to identify currently occurring changes and predict the magnitude and spatial distribution of future decline. [@Jump2010]

```{r echo=FALSE, results='asis', eval=show_text}
cat("
# Notas Jump 2010 
Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)

NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.

Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!

Otra idea --> Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)

NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.

Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!

El NDVI se ha utilizado para monitorizar la respuesta de bosques frente a sequía (ej.: 24,26,32, 44 en Jump2009 )

Necesidad de Ground-based assessment (p 1974 Jump 2009) -> Aunque utilizemos datos de satélite, siempre es necesario utilizar aproximaciones combinda

Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ... (ver esto y completar)

El problema de los datos ground-based assessment, existe un trade-off entre el área total cubierta por una red de monitoreo y su resolución local. --Por tanto también existen limitaciones de las ground-based para estudiar poblaciones en su rear edge. 
")
```

## References

### 2018-10-16 man/figures/0_figures.Rmd
last update

---
output:
  word_document:
    fig_caption: yes
    highlight: null
    reference_docx: ../../templates/template_pagebreak.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

```{r child = 'location_map.Rmd'}
```

#####

```{r child = 'std_anomalies.Rmd'}
```

### 2018-10-16 man/figures/location_map.Rmd
last update

@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-- [] Figure 1. Location map
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf
-
-- [] Figure 2. Plot de anomalias
-- [] Figure 3. Resilience metrics
-- [] Figure 4. BAI chronologies
-
-- [] Figure 5. Residual chronologies --> Supplementary
-
-")
-```
-
 **Figure 1**.
@@
 
-![](../../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
-##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
-```
-
-```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
-```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
-```
+![](../../out/fig/geomap/mapa_situacion_v2.jpg){height=600px}

### 2018-10-16 man/figures/std_anomalies.Rmd
last update

--- 2018-10-05 man/ms_figures.Rmd
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-## Figure 1 Location map 
-- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset.
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf 
-    - Ojo con los nombres SJ (Que hacemos?)
-
-## Figure 2. Plot de anomalias
-- [X] Solamente Anomalias estandarizadas
-
-## Figure 3. 
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
-##### 
 **Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
@@
 
-# profile_compara 
+profile_compara 
 ```
@@
   theme_bw() +
-  theme(text = element_text(size=16),
+  theme(text = element_text(size=12),
         legend.position = "none",
@@
         strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
+        plot.title = element_text(hjust = -0.05, size = 12, 
                                     margin = margin(t=-5, b=1)))
@@
 
-```{r, fig.height=5}
+```{r, fig.width=3.5, fig.height=5, fig.asp = .8}
 plot_sa_clu 
 ```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-# summary(m1)
-myr2 <- format(summary(m1)$r.squared, digits = 3)
-
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-10-16 man/ms.Rmd
add comments of review

@@
 ---
-title: "Relict populations of a deciduous oak show high resilience to drought despite land-use legacies"
+title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
 author: "AJ Perez-Luque; G. Gea-Izquierdo; ..."
@@
 # Abstract 
-* Trees showed resilience to drought events, despite the ‘a priori’ vulnerability that they can express as a result of the impact of climate change on relict stands resulting from intensive land-use in history.
-
-* Oak resilience to drought followed a dryness gradient: Northern populations, which thrive under a drier climate, were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.
+* Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified species composition and the forest structure as a result of intensive human pressure in the past.
+We analyzed forest resilience of adult trees to drought and climate change of relict Mediterranean *Quercus pyrenaica* Willd.
+populations in the Southern Iberian Peninsula.
+Forest resilience was analyzed using both dendroecological estimates of growth and MODIS satellite images as proxies of secondary and primary growth respectively.
+
+* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure.
+Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (i.e.
+warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge.
+Furthermore, the coldest stands show a positive growth trend since the end of the 1970s matching the simultaneous increase in temperature observed.
+
+* Resilience indices showed ecological coherence along a gradient of drought severity in time.
+In addition, oak resilience to drought followed a dryness gradient among the three populations studied: Northern populations thriving under a drier climate were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.
+
+* Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought.
+Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e.
+not only adult performance but also regeneration) of those stands located at the lowest elevations, hence the altitudinal rear-edge, need to be monitored under future expected drier and warmer climatic scenarios and, particularly, if high management pressure remains.
 
@@
 It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017].
-Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b].
+Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for at least the last three decades [@Spinoni2017b].
 Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013], specially in the Mediterranean area [@Spinoni2017a].
@@
 In the last decades the assessment of the drought impacts on ecosystems has gained attention [@Allen2010; @Clark2016], especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014].
-The response of vegetation to drought has been analyzed at different scales, from global [*e.g*.
+The functional response of vegetation to drought has been analyzed at different scales, from global [*e.g*.
 @Allen2010; @VicenteSerrano2013; @Norman2016] to local scales [see @MartinezVilalta2016 for a revision].
@@
 Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017], and some authors states that the impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017].
-This is specially relevant for areas with a long history of landscape modification as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016].
-For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.
-
-The impacts of drought are also especially relevant for populations located in the rear edge of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005].
+This is especially relevant for areas with a long history of landscape modification such as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016].
+`REVIEW` For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.
+
+The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005].
 The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010].
@@
 
-Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
+Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution `REVIEW`[@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
 Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015].
 The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services.
-This is particularly relevant for populations located in their rear-edge of their distribution, especially when they are located in Mediterranean mountains.
-
-Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and tree-ring [@Jump2010].
-
+This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.
+
+Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth.
 Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
@@
 
-Tree-ring width is a widely used proxy for tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Gazol2018].
+Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
 Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.
 
-In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, considered a rear-edge of their distribution.
-We are also interested in assessing resilience of these forest to several extreme drought episodes.
+In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened.
+We are also interested in assessing resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.
 We hypothesized that, relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts.
 Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q.
-pyreancia* forests in their rear edge; (*ii*) to analyze the resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species.
+pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change.
 In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
@@
 
-Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in northern Morocco [@Franco1990].
-In the Iberian Peninsula these forests live on siliceous soils, or soils poor in basic ions [@Vilches2014]
-under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002].
+Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
+In the Iberian Peninsula these forests live on siliceous soils, or soils poor in basic ions [@Vilches2014] under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002].
 *Q.
-pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species [@Gavilan2007; @delRio2007].
-
-This species reaches its southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations of between 860 m and 3482 m *a.s.l.*.
+pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species, which can hence be considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@Gavilan2007; @delRio2007].
+
+The forests of this species reaches their `REVIEW` southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*.
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
-There are eight oak patches (2400 Has) identified (Figure 1) in this mountain range, ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
+There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
 Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].
-
-The populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], undergoing intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
-In fact, conservation status of this species for southern Spain is "Vulnerable" [@Vivero2000].
-The relict presence of this species in Sierra Nevada is related both to its genetic resilience and high intraspecific genetic diversity [@Valbuena2010].
-However, they are also expected to suffer the impact of climate change, due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species.
-Thus, simulations of the climate change effects on this habitat forecast a reduction in suitable habitats for Sierra Nevada [@Benito2011].
+For this reason, the populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], having these forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
+Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].`REVIEW CITA`
 
@@
 
-During the second half of the XX century, several extreme drought episodes were recorded for the Iberian Peninsula (*e.g*.
-1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3).
-We focused on these two drought events because limitations about temporal availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
-Nevertheless, for radial growth-time series, older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events.
-The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by Spinioni (see Appendix S3)
+During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*.
+1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3).
+We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
+Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term.
+The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015 (see Appendix S3).
 
@@
 
-Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] and they are linked with primary productivity [@Ruimy1994].
+Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] because they are linked with primary productivity [@Ruimy1994].
 Here, vegetation greenness of *Q.
@@
 
+As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016.
 To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016].
@@
 
-As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016.
-$EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
+Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
 For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale.
@@
 ### Tree sampling 
-Tree sampling was carried during autumn of 2016.
-Trees were sampled at two locations in contrasting slopes of Sierra Nevada: San Juan (SJ), xeric site located at northern aspect; and Cáñar (CA), humid but warm site located at southern aspect (Figure 1; Table 1).
-For the southern site two elevations were sampled: CA-Low and CA-High.
+Tree sampling was carried out during autumn 2016.
+Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warm site located at southern aspect (Figure 1; Table 1).
+For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal tree-line in the sampled slope.
 All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015].

### 2018-10-17 man/ms.Rmd
update

@@
 
-To evaluate the effects of drougth events on greeennes and tree growth (BAI) resilience indices proposed by @Lloret2011 were used: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
+To evaluate the effects of drougth events on greeennes and tree growth (BAI) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
 The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred.
@@
 A period of three years was chosen because we found similar results comparing periods of two, three and four years (Appendix S6) and this length was used in other studies [*e.g.* @Gazol2017].
-To put in context the resilience observed for 2005 and 2012 drought events, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 (Appendix S3).
+To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure S5).
 
@@
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
-All analyses were carried out in R software [@R2017] using several packages (see Supplementary S?)
+All analyses were carried out in R software [@R2017] using several packages (see Supplementary `REVIEW` S?)
 
@@
 
-The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period.
+The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period.
 During the 2005 the lowest values of EVI standardized anomalies were recorded, particularly for northern populations (Figure 2).
@@
 
-## Radial growth trends
-
-Likely as a consequence of their different management origin (*i.e.* land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2).
-Competition was similar among sites but plot basal area was greatest in CA-High (Table 1).
-Yet the basal area increment (BAI) growth in the highest plot in altitude (CA-High) was the greatest (Figure 4).
+## Radial growth trends and growth disturbances
+
+The trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2).
+`REVIEW meter en Discussion` This can be likely explained as a consequence of their different management origin (*i.e.* land-use legacy), as shown by the release event in SJ suggesting their coppice origin and (references).
+`END REVIEW` Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4).
 In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations.

### 2018-10-17 man/ms.Rmd
add more review

@@
 Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g.
-logging) in the past.
+logging, neighbor mortality) in the past.
 By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].
@@
 `END REVIEW` Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4).
-In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations.
-For southern population differences in growth were observed, with CA-Low site showing lower values than CA-High but similar to growth values of the northern population (SJ) (Figure 4; Appendix S7).
+`REVIEW` In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations.
+`END REVIEW` For southern population differences in growth were observed.
+CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figure 4; Appendix S7).
 
@@
 Algo similar ocurrió en sj para los años 2003 y 2013.
+")
+```
+
+The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8).
+Northern site (SJ) showed two release events ($GC$ > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001.
+These periods alternate with periods of supression.
+Southern sites (CA-High and CA-Low) showed some weak suppression episodes, but not in the last 50 years.
+
+The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8).
+Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001.
+These periods alternate with periods of supression.
+Southern sites (CA-High and CA-Low) showed no release event except one for CA-High at the beginning of the `REVIEW`1840s?
+and no suppression event in the last 50 years.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+
+En las poblaciones del norte osbervamos dos eventos de PGC por encima del 25 e incluso el 50 %.
+El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
+Estos periodos se alternan con periodos de NGC: previo 1940, 1960 - 1991, y quizá aunque no lo vemos en al actualidad (??
+duda.
+Esto es especular un poco).
+Esto responde creo que a:
+
+(i) actividad minera y tratamientos forestales.
+La actividad minera en la zona fue intensa (aunque a veces intermitente, entre 1890 y 1957 (cuando se clausuró); y previamente desde 1858 tras los trabajos de Amalio Maestre).
+Hay costancia de la existencia de hornos de fundición de mineral en alguna de las minas (La Probadora).
+(ii) actividades forestales:
+* Se conoce la existencia de un proyecto redactado en 1943 de repoblación forestal de la cabecera del río Genil (he pedido los mapas).
+* Ver actividades forestales (bd proyectos de gestión)
+
+En el trabajo de [@GeaIzquierdo2014] se observa una liberación posterior a 1950 (QPUY9) (entre 1950 y 1960 aprox.) y luego una supresión hasta mediados de 1990.
+Esto coincide con lo que hemos observado en nuestra chrono.
+
+Respecto a las poblaciones del sur, destaca en la caH una primera liberación en torno a 1829-1830 y 1837-1840; y luego predominan los eventos de supresión que se alternan con algunos de liberación (muy débiles) hastas 1950.
+Posteriormente se alternan liberaciones-supresiones débiles.
+Destaca una liberación en torno a 2000-2003 (relación con clima???
+posterior a sequía 1999???)
+
+En caL, observamos un periodod de limitación hasta 1885 aprox.
+Luego se alternan al igual que antes periodos de liberación con otos de supresión con señales más fuertes que en caL, y al igual que antes destaca con mas intensidad si cabe, una liberación entre 2000 y 2003 (antrópico o climática ??)
+
+Duda: ocurre en los dos sitios caL y caH, entonces, es aumento de crecimiento debido a actuación forestal? o debido a clima??
+
+[@GeaIzquierdo2014] encontró un pico de disturbance en torno a 1860 en QUPY10 (cañar) sugiriendo que la baja densidad en esos sitios fue presumible el resultado de un thinning (aclareo).
+El patrón de supresión que observa para este sitio va en la linea de lo que nosotros encontramos (supresiones que alcanzan a veces el 25 % y otras no)
+
+En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles.
+En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles
 ")
@@
 
-The exploration of resilience metrics of tree-growth for other severe droughts, revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S5).
+The strength of the response of the resilience indices was related to drought severity.
+The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data `REVIEW`(1900 - 2017), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S5).
 A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155).
 Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S5) even not being the most severe drought events.
+
+The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (period 1900-2017?),
 
@@
 No positive relationship with temperature was found for southern trees.
-
-## Forest disturbances
-The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8).
-Northern site (SJ) showed two release events ($GC$ > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001.
-These periods alternate with periods of supression.
-Southern sites (CA-High and CA-Low) showed some weak suppression episodes, but not in the last 50 years.
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-
-En las poblaciones del norte osbervamos dos eventos de PGC por encima del 25 e incluso el 50 %.
-El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
-Estos periodos se alternan con periodos de NGC: previo 1940, 1960 - 1991, y quizá aunque no lo vemos en al actualidad (??
-duda.
-Esto es especular un poco).
-Esto responde creo que a:
-
-(i) actividad minera y tratamientos forestales.
-La actividad minera en la zona fue intensa (aunque a veces intermitente, entre 1890 y 1957 (cuando se clausuró); y previamente desde 1858 tras los trabajos de Amalio Maestre).
-Hay costancia de la existencia de hornos de fundición de mineral en alguna de las minas (La Probadora).
-(ii) actividades forestales:
-* Se conoce la existencia de un proyecto redactado en 1943 de repoblación forestal de la cabecera del río Genil (he pedido los mapas).
-* Ver actividades forestales (bd proyectos de gestión)
-
-En el trabajo de [@GeaIzquierdo2014] se observa una liberación posterior a 1950 (QPUY9) (entre 1950 y 1960 aprox.) y luego una supresión hasta mediados de 1990.
-Esto coincide con lo que hemos observado en nuestra chrono.
-
-Respecto a las poblaciones del sur, destaca en la caH una primera liberación en torno a 1829-1830 y 1837-1840; y luego predominan los eventos de supresión que se alternan con algunos de liberación (muy débiles) hastas 1950.
-Posteriormente se alternan liberaciones-supresiones débiles.
-Destaca una liberación en torno a 2000-2003 (relación con clima???
-posterior a sequía 1999???)
-
-En caL, observamos un periodod de limitación hasta 1885 aprox.
-Luego se alternan al igual que antes periodos de liberación con otos de supresión con señales más fuertes que en caL, y al igual que antes destaca con mas intensidad si cabe, una liberación entre 2000 y 2003 (antrópico o climática ??)
-
-Duda: ocurre en los dos sitios caL y caH, entonces, es aumento de crecimiento debido a actuación forestal? o debido a clima??
-
-[@GeaIzquierdo2014] encontró un pico de disturbance en torno a 1860 en QUPY10 (cañar) sugiriendo que la baja densidad en esos sitios fue presumible el resultado de un thinning (aclareo).
-El patrón de supresión que observa para este sitio va en la linea de lo que nosotros encontramos (supresiones que alcanzan a veces el 25 % y otras no)
-
-En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles.
-En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles
-")
-```
 

### 2018-10-25 man/figures/cumulative_prec.Rmd
add lines to cumulative prec and create file

--- 2018-10-05 man/ms_figures.Rmd
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-## Figure 1 Location map 
-- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset.
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf 
-    - Ojo con los nombres SJ (Que hacemos?)
-
-## Figure 2. Plot de anomalias
-- [X] Solamente Anomalias estandarizadas
-
-## Figure 3. 
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
-##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
-```
-
-```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
-```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
-```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-# summary(m1)
-myr2 <- format(summary(m1)$r.squared, digits = 3)
-
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
 ##### 
@@
 
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
+```{r}
+ltdf <- read.table(file=here::here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
 
-source(here("/script/R/", "hydro_year.R"))
+source(here::here("/script/R/", "hydro_year.R"))
 
@@
   geom_point(size=2) + theme_bw() + 
+  geom_line(size=.3) + 
   scale_x_continuous(breaks=seq(1950,2020, by=5)) +
@@
 ```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-10-25 analysis/drought/explore_spei_longterm.Rmd
modify drought severity to include months

@@
             minyear = min(year),
-            maxyear = max(year)) %>%
+            maxyear = max(year),
+            rangeDate = paste(month(min(date), label=TRUE), '-',
+                             (month(max(date), label=TRUE)))) %>%
   as.data.frame()
@@
 write.csv(sev12, file=here::here('data/sequias', 'severe_spei12.csv'), row.names=FALSE)
+pander(sev12)
 
-pander(sev12)
+```
+
+```{r}
+sev6 <- sev %>% filter(variable == "spei6") %>% arrange(desc(d_duration)) %>% 
+  dplyr::select(-index_events, -decade, -variable)
+
+write.csv(sev6, file=here::here('data/sequias', 'severe_spei6.csv'), row.names=FALSE)
+pander(sev6)
+```
+
+## Duration vs. Lowest SPEI (SPEI06, SPEI12)
+```{r}
+require(ggrepel)
+# Duracion vs
+sev %>% ggplot(aes(d_duration, y=lowest_spei, label=maxyear, colour=variable)) + geom_point() + 
+  geom_text_repel() 
+```
+
+## Duration vs. Severity (SPEI06, SPEI12)
+```{r}
+sev %>% ggplot(aes(d_duration, y=d_severity, label=maxyear, colour=variable)) + geom_point() + 
+  geom_text_repel() 
+```
+
+## Duration vs. Intensity (SPEI06, SPEI12)
+```{r}
+sev %>% ggplot(aes(d_duration, y=d_intensity, label=maxyear, colour=variable)) + geom_point() + 
+  geom_text_repel() 
+```
+
+## Severity vs. Intensity (SPEI06, SPEI12)
+```{r}
+sev %>% ggplot(aes(d_severity, y=d_intensity, label=maxyear, colour=variable)) + geom_point() + 
+  geom_text_repel() 
 ```

### 2018-10-26 analysis/rwi/climate_rwi_relate.Rmd
add periods for climate-tree growht relation (prec)

@@
          method = "correlation",  boot = "std",
-        .range(-3:9) + .sum(-9:8))
+        .range(-3:9) + .sum(-9:8) + .sum(-12:2) + .sum(3:5) + .sum(6:8) + .sum(9:11))
 
@@
       month == "sep...AUG" ~ "Hydrol", 
+      month == "dec...FEB" ~ "Winter",
+      month == "MAR...MAY" ~ "Spring",
+      month == "JUN...AUG" ~ "Summer",
+      month == "SEP...NOV" ~ "Autumn",
       TRUE ~ as.character(.$month)))
@@
 * quitar tmean 
+* De prec incluir también periodos
+* De SPEI como salen casi todos correlacionados, mostrar solo periodos
 

### 2018-10-26 analysis/rwi/climate_rwi_relate.Rmd
update

@@
 
-write.csv(rprec_all, here("/data/rwi_climate", "rprec_all.csv"), row.names = FALSE)
+write.csv(rprec_all, here::here("/data/rwi_climate", "rprec_all.csv"), row.names = FALSE)
 ```
@@
          method = "correlation",  boot = "std",
-        .range(-9:10) + .sum(-9:8))
+        .range(-9:10) + .sum(-9:8) + .sum(-12:2) + .sum(3:5) + .sum(6:8) + .sum(9:11))
 

### 2018-10-26 man/figures/residual_chronos.Rmd
update

---
output: 
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_pagebreak.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
```

```{r}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")

library("grid")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
library("stringr")
library("pander")
library("lemon")
```

```{r, echo=FALSE}
show_text <- FALSE
```

##### 
**Figure 5.** Residual tree-ring chronologies obtained for the *Q. pyrenaica* sites. Dashed red lines indicate the start of the reliable period (EPS > 0.85). Dotted black lines showing the severe drought years identified in our climatic data (see Table S3).

```{r}
cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')

# peores sequias 
dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
dy <- dy %>% arrange(desc(d_duration))
# solamente aquellos mayor de 2 meses 
dys <- dy %>% filter(d_duration > 2) %>% as.data.frame() 
dys <- dys[,'maxyear'] 
```

```{r}
# Mayores y menores crecimientos 
# top10  
top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
```

```{r}
label_site <- c("0_sj" = "SJ",
                "1_caH" = "CA-High",
                "2_caL" = "CA-Low")

cro_rwi <- cro_rwi %>% 
  mutate(siteF = case_when(
    site_sorted == "0_sj" ~ "Northern (SJ)",
    site_sorted == "1_caH" ~ "Southern (Ca-High)",
    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))

# 
plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(face="bold", size=12), 
        # strip.text = element_blank(), 
        axis.text = element_text(face="bold", size=8), 
        axis.title = element_text(face="bold", size=12)) + 
  geom_hline(yintercept = 1, colour = "grey") + 
  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
             labeller = as_labeller(label_site)) + 
  ylab("Ring width index") + xlab("Year") + 
  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
  geom_vline(xintercept = dys, linetype = "dotted") + 
  # geom_vline(xintercept = dys, linetype = "dotted") + 
  scale_x_continuous(limits = c(1819, 2016), 
                     breaks= seq(1820, 2020, by=20))

plot_rwi_crono  
```

### 2018-10-26 man/figures/resilience_metrics.Rmd
update

--- 2018-10-05 man/ms_figures.Rmd
@@
 show_text <- FALSE
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-## Figure 1 Location map 
-- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset.
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf 
-    - Ojo con los nombres SJ (Que hacemos?)
-
-## Figure 2. Plot de anomalias
-- [X] Solamente Anomalias estandarizadas
-
-## Figure 3. 
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
-##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
-```
-
-```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
-```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
 ```
@@
 ```{r, eval=FALSE}
-### Figure 4
 
@@
 
-**Figure 4.** Response of *Q.
+**Figure .** Response of *Q.
 pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
@@
 ```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-# summary(m1)
-myr2 <- format(summary(m1)$r.squared, digits = 3)
-
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-10-26 man/figures/ring_climate.Rmd
update

--- 2018-10-05 man/ms_figures.Rmd
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-## Figure 1 Location map 
-- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset.
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf 
-    - Ojo con los nombres SJ (Que hacemos?)
-
-## Figure 2. Plot de anomalias
-- [X] Solamente Anomalias estandarizadas
-
-## Figure 3. 
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
-##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
-```
-
-```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
-```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
-```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-# summary(m1)
-myr2 <- format(summary(m1)$r.squared, digits = 3)
-
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
 ##### 
 **Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
+pyrenaica* and monthly climatic data: precipitation (a), 6-month SPEI (b), maximun (c) and minimun (d) temperatures.
 *green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
@@
 # cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
+# exdf <- data.frame(
+#   id = c(16:19),
+#   varname = rep("prec", 4), 
+#   month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
+#   coef = rep(0, 4), 
+#   significant = rep('FALSE', 4),
+#   ci_lower = rep(0, 4),
+#   ci_upper = rep(0, 4), 
+#   sig = rep("", 4), 
+#   month_name = c("Winter", "Spring", "Summer", "Autumn"), 
+#   v_name = rep("Prec", 4))
 
@@
 
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
+# clima_ring <- ddf %>% 
+  df %>% 
+  ggplot(aes(x=id, y=coef, group = site)) + 
   geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
+  geom_bar(aes(fill=site), stat="identity", position = position_dodge2(padding = 0.1)) +
+ #  facet_wrap(~v_name, labeller = as_labeller(label_var), ncol = 1) +
+  facet_rep_wrap(~v_name, repeat.tick.labels = TRUE, ncol = 1) + 
   geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
@@
         panel.grid.minor.y = element_blank(),
+        panel.grid.minor.x = element_line(size=1),
         axis.text.x = element_text(angle = 90, vjust = 0.3),
@@
         strip.background = element_rect(fill=NA, colour=NA),
+        strip.placement = 'inside',
         strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
+  scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name)) +
   scale_y_continuous(limits = c(-0.4, 0.8), 
@@
 ```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-10-26 man/figures/ring_climate.Rmd
add review GGI

@@
 csl: ../../refs/ecology.csl
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-# cutre solucion para el problema de PREC 
-# exdf <- data.frame(
-#   id = c(16:19),
-#   varname = rep("prec", 4), 
-#   month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-#   coef = rep(0, 4), 
-#   significant = rep('FALSE', 4),
-#   ci_lower = rep(0, 4),
-#   ci_upper = rep(0, 4), 
-#   sig = rep("", 4), 
-#   month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-#   v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
 label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
@@
 
-# clima_ring <- ddf %>% 
-  df %>% 
-  ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = position_dodge2(padding = 0.1)) +
- #  facet_wrap(~v_name, labeller = as_labeller(label_var), ncol = 1) +
-  facet_rep_wrap(~v_name, repeat.tick.labels = TRUE, ncol = 1) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
+theme_ring_climate <- list(
+  geom_hline(yintercept = 0, colour="gray"), 
+  geom_bar(aes(fill=site), stat="identity", position = position_dodge2(padding = 0.1)),
+  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.06))), 
+            position = position_dodge(.9), size = 6), 
   theme_bw() +
@@
         panel.grid.minor.y = element_blank(),
-        panel.grid.minor.x = element_line(size=1),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.placement = 'inside',
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
+        panel.grid.minor.x = element_line(size=.5),
+        axis.text.x = element_text(angle = 90, vjust = 0.3)),
+  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site)
+)
 
-clima_ring 
+p_tmax <- df %>% filter(v_name == 'T max') %>% 
+  ggplot(aes(x=id, y=coef, group = site)) + 
+  theme_ring_climate + 
+  scale_y_continuous(limits = c(-0.4, 0.4), 
+                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4)) +
+  scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'bottom') + 
+  ylab('') + 
+  annotate("text", x = 2, y = 0.35, label = "T max", fontface =2, size = 5)
+
+p_tmin <- df %>% filter(v_name == 'T min') %>% 
+  ggplot(aes(x=id, y=coef, group = site)) + 
+  theme_ring_climate + 
+  scale_y_continuous(limits = c(-0.4, 0.6), 
+                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4)) +
+  scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'top') + 
+  ylab('') +
+  annotate("text", x = 2, y = 0.35, label = "T min", fontface =2, size = 5)
+
+# prec-spei
+
+dfspei <- df %>% 
+  filter(v_name == 'SPEI') %>% 
+  filter(month_name %in% c('Hydrol', 'Winter','Spring','Summer','Autumn')) %>% 
+  # truquillo para poner juntas las dos graficas 
+  mutate(id = id + 6)
+
+dfprec <- df %>% filter(v_name == 'Prec')
+
+# Joins df
+dfprec_spei <- bind_rows(dfspei, dfprec)
+
+# solucion para poner juntas prec + spei 
+labelitas <- unique(dfprec_spei[c('id','month_name','v_name')])
+
+p_prec_spei <- dfprec_spei %>% 
+  ggplot(aes(x=id, y=coef, group = site)) + 
+  geom_vline(xintercept=20) +
+  theme_ring_climate + 
+  scale_y_continuous(limits = c(-0.2, 0.8), 
+                     breaks = c(-0.2, 0, 0.2, 0.4, 0.6, 0.8)) +
+  scale_x_continuous(breaks = labelitas$id, labels = labelitas$month_name, position = 'top') + 
+  ylab('') +
+  annotate("text", x = 2, y = 0.75, label = "Prec", fontface =2, size = 5) + 
+  annotate("text", x = 23, y = 0.75, label = "SPEI", fontface =2, size = 5)
+
+ptmin
+
+ptemperatura <- p_tmin + theme(legend.position="none") + p_tmax + theme(legend.position="bottom") + plot_layout(ncol=1)
+
+p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b'))
+
 ```

### 2018-10-27 man/figures/growth_changes.Rmd
add new figure

---
output: 
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_pagebreak.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
```

```{r}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")

library("grid")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
library("stringr")
library("pander")
library("lemon")
```

```{r, echo=FALSE}
show_text <- FALSE
```

**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).

```{r}
gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')

# customize plot (corrected scale)
pos_neg_log <- scales::trans_new('signed_log', 
                                 transform = function(x) sign(x)*log(abs(x)),
                                 inverse = function(x) sign(x)*exp(abs(x)))

custom_ylab <- '% GC'
micolor <- c("#838B8B")

# solution for arrange
# http://felixfan.github.io/stacking-plots-same-x/

custom_corrected_gc <- list(
  geom_bar(stat='identity', fill=micolor, colour=micolor),
  theme_bw(), 
  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),

        # strip.background = element_rect(colour='black',fill='white'))
  )

# Function to return a desired sequence breaks along axes
# https://gist.github.com/valentinitnelav/ec7c253c80f4e045f4bd1c9b2920d7bd 
my.breaks <- function(my.vector, step){
    my.min <- floor(min(my.vector))
    my.max <- ceiling(max(my.vector))
    my.seq <- seq(from = round(my.min/step)*step, 
                  to   = round(my.max/step)*step,
                  by   = step)
    return(my.seq)
}

gc_todos %>% 
  filter(site == 'gc_caH') %>% 
  ggplot(aes(x=year, y=gc_md)) + 
  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
  ylab(custom_ylab) +
  GCtheme + 
  ylim(-100, 150) +
  scale_x_continuous(name = "Year", breaks = my.breaks(gc_todos$year, step = 25))

GCtheme <- list(
  # BAR
  geom_bar(stat='identity', fill=micolor, colour=micolor),
  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
  theme_bw(),
  theme(# AXIS ticks
        axis.ticks.length = unit(-1.4, "mm"),
        axis.text.x = element_text(size = 8, margin = unit(c(t = 2.5, r = 0, b = 0, l = 0), "mm")),
        panel.grid.minor = element_blank(),
        # Remove strip 
        strip.background = element_blank(),
        strip.text = element_blank())
)

d <-

  gc_todos %>% 
  # mutate(year = lubridate::ymd(year, truncated = 2L)) %>% 
  filter(site != 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_md)) + 
  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
  # scale_y_continuous(trans = pos_neg_log) +
  facet_wrap(~site_sorted, nrow = 2) +
             # labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
    ylab(custom_ylab) + custom_corrected_gc +
  scale_x_continuous(name = "Year", breaks = my.breaks(gc_todos$year, step = 25))

  scale_x_date(date_breaks = "25 year", date_labels = "%Y")

d <- gc_todos %>% 
  filter(site == 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_md)) + 
  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
  facet_wrap(~site_sorted, nrow = 2,
             labeller=as_labeller(c("0_sj" = "SJ"))) + 
  custom_corrected_gc + ylab('') + 
  scale_y_continuous(trans = pos_neg_log, 
                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
plot_medians_corregido

```

### 2018-10-27 man/figures/ring_climate.Rmd
add new figure

@@
 **Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), 6-month SPEI (b), maximun (c) and minimun (d) temperatures.
+pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures.
 *green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
@@
 ```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
+rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'T max')
+rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'T min')
+rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name ='SPEI')
+rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'Prec')
 
@@
   scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'bottom') + 
-  ylab('') + 
+  ylab('Correlation') + 
   annotate("text", x = 2, y = 0.35, label = "T max", fontface =2, size = 5)
@@
   scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'top') + 
-  ylab('') +
+  ylab('Correlation') +
   annotate("text", x = 2, y = 0.35, label = "T min", fontface =2, size = 5)
@@
   scale_x_continuous(breaks = labelitas$id, labels = labelitas$month_name, position = 'top') + 
-  ylab('') +
+  ylab('Correlation') +
   annotate("text", x = 2, y = 0.75, label = "Prec", fontface =2, size = 5) + 
@@
 
-ptmin
-
 ptemperatura <- p_tmin + theme(legend.position="none") + p_tmax + theme(legend.position="bottom") + plot_layout(ncol=1)
 
-p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b'))
+ring_climate <- p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b'))
 
+ring_climate
 ```

### 2018-10-30 analysis/legacy_effects.Rmd
legacy effects sensu Anderegg 2015

--- 2018-10-16 analysis/chronos/analysis_chronologies.Rmd
@@
 ---
-title: "chronologies"
+title: "legacy effects"
 author: "AJ Perez-Luque (@ajpelu)"
@@
 library("knitr")
-# library("detrendeR")
+library("here")
 library("pander")
-library('boot')
-```
+```
+
+* Observed post-drougth growth --> detrended-only chronologies
+* Predicted post-drought growth --> prewithened tree growth
+
+* utilizamos datos procedentes de analysis_splines.Rmd
+* Los datos de la chrono standard corresponden con la detrended (obs) mientras que la residual es la prewithened
+
+```{r}
+# Datos de chronos 
+pre <- read.csv(here::here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
+
+# Datos de sequia 
+# peores sequias 
+dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
+dy <- dy %>% arrange(desc(d_duration))
+# solamente aquellos mayor de 2 meses 
+dys <- dy %>% 
+  filter(d_duration > 2) %>%  
+  mutate(year = ifelse(minyear == maxyear, 
+                       minyear, 
+                       paste0(minyear, '-', maxyear))) %>% 
+  dplyr::select(-minyear, -cat, -month_peak) %>% 
+  as.data.frame()
+
+# LEGACY EFFECTs
+
+# drougth_years 
+event <- dys$maxyear
+out <- c()
+
+for (i in event){
+
+  # Years post Drought 
+  lag_years <- seq(from = i, to = i + 6, by =1)
+
+  # Filter data 
+  aux <- pre %>% 
+    filter(year %in% lag_years) %>% 
+    mutate(ypd = year - i)
+
+  # Combined data 
+  out <- rbind(out, aux)
+  }
+
+# Computed OBS - PRED and filter by year 
+legacy <- out %>% 
+  filter(ypd != 0) %>% 
+  mutate(lag = std - res)
+
+mlegacy <- legacy %>% group_by(site, ypd) %>% 
+  summarise(mean=mean(lag, na.rm=T),
+            sd = sd(lag, na.rm=T),
+            se = sd/sqrt(length(lag)))
+
+legacy %>% ggplot(aes(x = ypd, y = lag, colour = site)) + geom_point() + geom_line()
+
+mlegacy %>% ggplot(aes(x = ypd, y = mean, colour = site)) + 
+  geom_point() + 
+  geom_errorbar(aes(ymin=mean - se, ymax=mean + se))
+
+ggplot(aes(x=ypd, y=lag, colour=site, group=site)) + geom_boxplot()
+
+# http://www.stat.ucla.edu/~rgould/110as02/bsci
+
+```
+
+# Leer datos de RWI 
+* Son datos de las series detrended ([./analysis/chronos/analysis_splines.md]('./analysis/chronos/analysis_splines.md)),
 

### 2018-11-02 analysis/legacy_effects.Rmd
update legacy

@@
 csl: ../ecology.csl
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library("pander")
+library("rms")
+library("magrittr")
+library("lemon")
 ```
@@
   mutate(lag = std - res) 
+```
 
-mlegacy <- legacy %>% group_by(site, ypd) %>% 
+```{r}
+# Get summary statistics and  plot
+
+mlegacy <- legacy %>% group_by(site_sorted, ypd) %>%
   summarise(mean=mean(lag, na.rm=T),
@@
 
-legacy %>% ggplot(aes(x = ypd, y = lag, colour = site)) + geom_point() + geom_line()
-
-mlegacy %>% ggplot(aes(x = ypd, y = mean, colour = site)) + 
-  geom_point() + 
-  geom_errorbar(aes(ymin=mean - se, ymax=mean + se))
-
-ggplot(aes(x=ypd, y=lag, colour=site, group=site)) + geom_boxplot()
-
-# http://www.stat.ucla.edu/~rgould/110as02/bsci
-
-```
-
-# Leer datos de RWI 
-* Son datos de las series detrended ([./analysis/chronos/analysis_splines.md]('./analysis/chronos/analysis_splines.md)),
-
-# Read y Prepare data 
-* Leer datos `rwl` de SJ y CA
-* Leer datos de diametros de los focal tree
-
-```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
-
-# sj 
-sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
-
-# canar 
-ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
-
-# Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+po <- position_dodge(0.9)
+mlegacy %>% ggplot(aes(x = ypd, y = mean, colour = site_sorted)) + 
+  geom_line(position = po, alpha=0.8) + 
+  geom_point(position = po) + 
+  geom_errorbar(aes(ymin=mean - se, ymax=mean + se), position = po) +
+  theme_bw() + 
+  geom_hline(yintercept = 0, colour = 'black') + 
+  theme(panel.grid = element_blank()) + 
+  scale_x_continuous("Years post drougth event", breaks=c(1:6)) + 
+  scale_y_continuous("Observed - Predicted Growth")
 ```
@@
 ```{r}
-source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/rw_byTree.R'))
-source(paste0(di, 'script/R/chrono_bai.R'))
-source(paste0(di, 'script/R/suaviza_cronos.R'))
-source(paste0(di, 'script/R/correlaBootIC.R'))
-```
+# Boostrap Confidence Intervals
 
-* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low (:red_circle: mejorar la forma de hacerlo)
-    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
+# Nest the data 
+legacy_nest <- legacy %>% 
+  group_by(site_sorted, ypd) %>% 
+  nest()
 
-```{r}
-# Replace SNA by SJ and SNB by CA
-names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
-names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")
+# Boostrap the mean 
+## boot_mean
+boot_mean <- function(d, i) {
+    mean(d[i])
+}
 
-# Remove g in name of some cores of CA. 
-names(ca) <- stringr::str_replace(names(ca), "g", "")
+## Use purrr::map to add another column to legacy_nest, in which each row contains the object returned by mapping the boot::boot function (which is responsible for applying the bootstrap_mean function) over the tibbles stored in each row of the “data” list-column.
+
+legacy_nest %<>% 
+  dplyr::mutate(booted = purrr::map(.x = data, # The list-column containing <S3: tibble>
+                                      ~ boot::boot(data = .x$lag, # The <S3 tibble> column being sampled
+                                                 statistic = boot_mean, # The user-defined function
+                                                 R = 10000, # The number of replicates
+                                                 stype = "i")))
+
+# compute the Confidence interval 
+legacy_nest %<>%
+    dplyr::mutate(booted_ci = purrr::map(.x = booted, # The list-column containing <S3: boot> objects
+                                         ~ boot::boot.ci(.x,
+                                                         conf = 0.95, # Interval width
+                                                         type = "bca")))  # Calculate a BCa interval
+
+# Tidy the CI 
+legacy_CI <- legacy_nest %>% 
+  dplyr::mutate(statistic = purrr::map(.x = booted_ci, # The list-column containing <S3 bootci> objects
+                                         ~ .x$t0), # The point estimate
+                  lower_ci = purrr::map(.x = booted_ci,
+                                        ~ .x$bca[[4]]), # The value of the lower 2.5% limit
+                  upper_ci = purrr::map(.x = booted_ci,
+                                        ~ .x$bca[[5]])) %>% # The value of teh upper 97.5% limit
+    # Drop the list-columns (no longer needed)
+    dplyr::select(-data, -booted, -booted_ci) %>%
+    # Unnest the dataframe
+    tidyr::unnest()
+
+# Join to mean values
+mlegacy %<>% inner_join(legacy_CI, by=c("site_sorted", "ypd")) %>% 
+  rename(site = site_sorted)
+
+label_site <- c("0_sj" = "SJ",
+                "1_caH" = "CA-High",
+                "2_caL" = "CA-Low")
+
+p <- mlegacy %>% ggplot(aes(x = ypd, y = mean)) + 
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill=site), alpha=.5) + 
+  geom_line(aes(colour=site)) + 
+  geom_point(aes(colour=site), size=3) + 
+  facet_wrap(~site, labeller = as_labeller(label_site)) + 
+  # geom_errorbar(aes(ymin=mean - sd, ymax=mean + sd), position = po) +
+  theme_bw() + 
+  geom_hline(yintercept = 0, colour = 'black') + 
+  theme(panel.grid = element_blank(), 
+        strip.background = element_blank(), 
+        legend.position = "none") + 
+  scale_x_continuous("Years post drougth event", breaks=c(1:6)) + 
+  scale_y_continuous("Observed - Predicted Growth", breaks=seq(-0.2, 0.2, by=0.05))
+p
+
+pdf(here::here('/out/legacy', 'legacy_all_droughts.pdf'), width=8, height = 5)
+p
+dev.off()
 ```
@@
 ```{r}
-# Create subset to compare between sites 
-caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
-             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
-             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
-caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
-              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
-              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
+# Filter 
+l2005 <- legacy %>%
+  filter(year %in% seq(2005+1, 2005+6, by=1)) %>% 
+  dplyr::select(-site) %>% rename(site = site_sorted)
+l2012 <- legacy %>% 
+  filter(year %in% seq(2012+1, 2012+6, by=1)) %>% 
+  dplyr::select(-site) %>% rename(site = site_sorted)
 
-sjL <- sj[, c("SJ0101","SJ0102","SJ0201","SJ0202","SJ0301","SJ0302","SJ0401","SJ0402","SJ0501","SJ0502",
-              "SJ0601","SJ0602","SJ0603","SJ0701","SJ0702","SJ0801","SJ0802","SJ0901","SJ0902","SJ0903",
-              "SJ1001","SJ1002","SJ1003")]
-sjH <- sj[, c("SJ1101","SJ1102","SJ1103","SJ1201","SJ1202","SJ1203","SJ1301","SJ1302","SJ1303","SJ1401",
-              "SJ1402","SJ1501","SJ1502","SJ1601","SJ1602","SJ1701","SJ1702","SJ1703","SJ1801","SJ1802",
-              "SJ1901","SJ1902","SJ2001","SJ2003","SJ2002")]
-```
+p1 <- mlegacy %>% ggplot(aes(x = ypd, y = mean)) + 
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill=site), alpha=.5) + 
+  geom_line(aes(colour=site)) + 
+  geom_point(aes(colour=site), size=3) + 
+  geom_line(data = l2005, aes(x = ypd, y =  lag)) + 
+  geom_line(data = l2012, aes(x = ypd, y =  lag), linetype = "dashed") + 
+  facet_wrap(~site, labeller = as_labeller(label_site)) + 
+  theme_bw() + 
+  geom_hline(yintercept = 0, colour = 'black') + 
+  theme(panel.grid = element_blank(), 
+        strip.background = element_blank(), 
+        legend.position = "none") + 
+  scale_x_continuous("Years post drougth event", breaks=c(1:6)) + 
+  scale_y_continuous("Observed - Predicted Growth", breaks=seq(-0.4, 0.2, by=0.05))
 
-* Lectura y preparación de datos de diámetro
+p1
 
-```{r} 
-# Prepare Diameter data
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Change name focal according to loc
-compete <- compete %>% 
-  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm)
-
-# Set levels of eleveation 
-sj_lowcode  <- paste0('SJ', str_pad(1:10, 2, pad='0'))
-sj_highcode <- paste0('SJ', 11:20)
-ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
-            paste0('CA', 26:30))
-ca_highcode <- paste0('CA', 11:25)
-
-ft <- ft %>% 
-  mutate(elev = as.factor(
-    ifelse(id_focalLoc %in% sj_lowcode, 'Low',
-                       ifelse(id_focalLoc %in% sj_highcode, 'High',
-                              ifelse(id_focalLoc %in% ca_lowcode, 'Low', 'High'))))) %>%
-  mutate(site = as.factor(paste0(loc, '_', elev)))
-```
-
-# Summary dendrochronology statistics
-
-* Por cada loc (ca/sj) o site (ca_high, ca_low, sj_high, sj_low) calculamos los siguientes datos: 
-    * age: mean + sd (average of all series)
-    * year: min, max of all series
-    * year range
-    * autocorrelation order 1 (ar1): mean + sd 
-    * seires intercorrelation (icor): mean + sd
-
-## By site 
-```{r}
-objects_rwl <- c('caL','caH','sjL','sjH')
-
-out <- c()
-
-for (i in objects_rwl){
-
-  aux <- get(i)
-  # Stats of singles RWL
-  aux_stats <- rwl.stats(aux)
-  # Add site 
-  aux_stats$site <- as.factor(i)
-
-  # Compute interseries correlation 
-  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
-  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
-
-  out <- rbind(out, aux_stats)
-  }
-
-rwl_by_site <- out %>% group_by(site) %>% 
-  summarize(age_mean = round(mean(year),2), 
-            age_sd = round(sd(year), 2),
-            year_min = min(first),
-            year_max = max(last), 
-            year_range = year_max - year_min + 1, 
-            ar1_mean = round(mean(ar1), 3),
-            ar1_sd = round(sd(ar1), 3),
-            icor_mean = round(mean(interser_cor_val), 3),
-            icor_sd = round(sd(interser_cor_val), 3)) %>% 
-  as.data.frame()
-
-rwl_by_site %>% 
-  write.csv(file=paste(di, "data/dendro_summary/site_dendro.csv", sep=""), row.names = FALSE)
-
-pander(rwl_by_site, caption='Dendrochronological summary by sites') 
-```
-
-## By Loc
-```{r}
-objects_rwl <- c('ca','sj')
-
-out <- c()
-
-for (i in objects_rwl){
-
-  aux <- get(i)
-  # Stats of singles RWL
-  aux_stats <- rwl.stats(aux)
-  # Add site 
-  aux_stats$loc <- as.factor(i)
-
-  # Compute interseries correlation 
-  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
-  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
-
-  out <- rbind(out, aux_stats)
-  }
-
-rwl_by_loc <- out %>% group_by(loc) %>% 
-  summarize(age_mean = round(mean(year),2), 
-            age_sd = round(sd(year), 2),
-            year_min = min(first),
-            year_max = max(last), 
-            year_range = year_max - year_min + 1, 
-            ar1_mean = round(mean(ar1), 3),
-            ar1_sd = round(sd(ar1), 3),
-            icor_mean = round(mean(interser_cor_val), 3),
-            icor_sd = round(sd(interser_cor_val), 3)) %>% 
-  as.data.frame()
-
-rwl_by_loc %>% 
-  write.csv(file=paste(di, "data/dendro_summary/loc_dendro.csv", sep=""), row.names = FALSE)
-
-pander(rwl_by_loc, caption='Dendrochronological summary by sites') 
-```
-
-# Aggregate RW by tree
-
-* Agregar valores medios de RW por loc (obtenemos sj_tree / ca_tree) o por site (sjL_tree, sjH_tree / caL_tree, caH_tree) 
-* ver fun rw_byTree o utilizar treeMean (dplR)
-
-```{r}
-# Remember snc = structure of core name SJ0101 (site | tree | core)
-sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
-ca_tree <- rw_byTree(ca, snc =c(2,2,2), locname = 'CA')
-
-sjL_tree <- rw_byTree(sjL, snc =c(2,2,2), locname = 'SJ')
-sjH_tree <- rw_byTree(sjH, snc =c(2,2,2), locname = 'SJ')
-caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
-caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')
-
-# Si decido usar treeMean debería ser (solo un ejemplo)
-# ids_sj <- read.ids(sj, stc = c(2, 2, 2))
-# sj_meantree <- treeMean(sj, ids = ids_sj,  na.rm=TRUE)
-```
-
-* Crear diferentes dataset de diametro por sitio 
-```{r}
-# # function to subset averaged RW 
-# subsetRW <- function(rwdf, codes){ 
-#   require(dplyr)
-#   aux_codes <- paste0('rw_', codes)
-#   mysubset <- rwdf %>% dplyr::select(one_of(aux_codes), year)
-#   mysubset
-#   }
-
-# diam <- ft %>% group_by(site) %>% do(data = (.)) %>% 
-#              dplyr::select(data) %>% 
-#              lapply(function(x) {(x)})
-
-diam <- ft %>%
-  mutate(diameter = dn_mm, 
-         id = id_focalLoc) %>%
-  dplyr::select(id, diameter, site) %>% 
-  split(.$site)
-
-d_caH <- diam$CA_High[,c('id','diameter')]
-d_caL <- diam$CA_Low[,c('id','diameter')]
-d_sjH <- diam$SJ_High[,c('id','diameter')]
-d_sjL <- diam$SJ_Low[,c('id','diameter')]
-
-diamloc <- ft %>% 
-  mutate(diameter = dn_mm, 
-         id = id_focalLoc) %>%
-  dplyr::select(id, diameter, loc) %>% split(.$loc)
-
-d_ca <- diamloc$CA[,c('id','diameter')]
-d_sj <- diamloc$SJ[,c('id','diameter')]
-```
-
-# Cómputo del BAI por loc y/o site
-
-* He construido una funcion para el computo del BAI, teniendo en cuenta la aproximación de [@Piovesan2008]. Es similar a `bai.out`
-
-```{r}
-bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
-
-bai_sj1 <-bai_piovesan(rwdf = sj, diam_df = d_sj)
-
-str(sj_tree)
-str(d_sj)
-
-bai_ca <- bai_piovesan(rwdf = ca_tree, diam_df = d_ca)
-
-bai_sjH <- bai_piovesan(rwdf = sjH_tree, diam_df = d_sjH)
-bai_sjL <- bai_piovesan(rwdf = sjL_tree, diam_df = d_sjL)
-bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
-bai_caL <- bai_piovesan(rwdf = caL_tree, diam_df = d_caL)
-
-# bai_test <- round(bai.out(rwl = sj_tree, diam = d_sj), 4)
-# test bai.out and bai_piovesan  
-# unique(round(bai_sj$SJ03,4) == round(bai_test$SJ03,4))
-
-# Set class to bai object 
-# Esto es para que funcionen algunas otras funciones de dplR 
-bais <- c('bai_sj', 'bai_ca', 'bai_sjH', 'bai_sjL', 'bai_caH', 'bai_caL')
-
-for (i in bais){
-  aux <- get(i)
-
-  class(aux) <- c('rwl', 'data.frame')
-
-  assign(i, aux)
-}
+pdf(here::here('/out/legacy', 'legacy_all_droughts_2005_2012.pdf'), width=8, height = 5)
+p1
+dev.off()
 
 ```
-
-# Cronologías by site / loc
-
-### Mean vs. Tukey’s Biweight Robust Mean
-
-* Computo la cronologia usando la función `chron`. 
-* Para SJ hag una comparación de la cronologia construida con biweight (*treatment = biweigth*) o construida con la media aritmética (*treatment = No biweigth*) (calculada con la función `chron`)
-* Idem para CA, pero en CA si separo entre CA high y CA low
-
-```{r chro_sj_compare_method}
-# Robust mean biweith or not
-cro_sjb <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
-cro_sjnotb <- chron(bai_sj, prefix = "SJ", biweight = FALSE)
-
-cro_b <- cro_sjb %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = SJstd, samp.depth, year, treatment)
-
-cro_notb <- cro_sjnotb %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = SJstd, samp.depth, year, treatment)
-
-aux_cro <- rbind(cro_b, cro_notb)
-
-aux_cro %>% ggplot(aes(x = year, y = bai_mean / 100, colour=treatment)) + 
-  geom_point() + geom_line() + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
-  theme_bw() + theme(panel.grid.major = element_blank()) +
-  ggtitle('SJ chronology built using Biweight or Mean')
-```
-
-```{r chro_ca_compare_method}
-# Robust mean biweith or not
-cro_cab_h <- chron(bai_caH, prefix = "CA", biweight = TRUE)
-cro_canotb_h <- chron(bai_caH, prefix = "CA", biweight = FALSE)
-
-cro_cab_h_aux <- cro_cab_h %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_canotb_h_aux <- cro_canotb_h  %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-aux_croca_h <- rbind(cro_cab_h_aux, cro_canotb_h_aux)
-aux_croca_h$site <- 'CA High'
-
-cro_cab_l <- chron(bai_caL, prefix = "CA", biweight = TRUE)
-cro_canotb_l <- chron(bai_caL, prefix = "CA", biweight = FALSE)
-
-cro_cab_l_aux <- cro_cab_l %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_canotb_l_aux <- cro_canotb_l  %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-aux_croca_l <- rbind(cro_cab_l_aux, cro_canotb_l_aux)
-aux_croca_l$site <- 'CA Low'
-
-aux_croca <- rbind(aux_croca_h, aux_croca_l)
-
-aux_croca %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
-  geom_point() + geom_line() + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
-  theme_bw() + theme(panel.grid.major = element_blank()) + 
-  facet_wrap(~site, nrow = 2) +
-    ggtitle('CA chronology built using Biweight or Mean')
-```
-
-## Compute Chrono 
-### San Juan chronologies by site
-
-```{r chrono_sj_two_sites}
-# San Juan 
-cro_sjH <- chrono_bai(bai_sjH)
-cro_sjH$site <- 'SJ_High'
-
-cro_sjL <- chrono_bai(bai_sjL)
-cro_sjL$site <- 'SJ_Low'
-
-cro_sj<- rbind(cro_sjH, cro_sjL)
-cro_sj$loc <- 'Norte (San Juan)'
-
-plot_cro_sj <- cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                    ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + geom_point(size=1) + 
-  # scale_colour_manual(values = c("red", "blue")) + 
-  theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1920, 2020, by=10))
-
-plot_cro_sj
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sj_HL.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sj, width=12, height = 4) 
-```
-
-### Cañar chronologies by site
-
-```{r chrono_ca_two_sites}
-# Canar 
-cro_caH <- chrono_bai(bai_caH)
-cro_caH$site <- 'CA_High'
-
-cro_caL <- chrono_bai(bai_caL)
-cro_caL$site <- 'CA_Low'
-
-cro_ca <- rbind(cro_caH,cro_caL)
-cro_ca$loc <- 'Sur (Canar)'
-
-plot_cro_ca <- cro_ca %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=1) + 
-  # scale_colour_manual(values = c("red", "blue")) + 
-  theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep='')))+
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-
-plot_cro_ca
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_ca_HL.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_ca, width=12, height = 4) 
-```
-
-```{r chronos_4_sites}
-# Combine and plot 
-cro_sites <- rbind(cro_ca, cro_sj)
-
-plot_cro_sites <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + facet_wrap(~loc, nrow = 2) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-
-plot_cro_sites
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites, width=12, height = 8)
-```
-
-```{r chronos_4_sites_collapsed}
-plot_cro_sites_collapsed <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  # geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-plot_cro_sites_collapsed
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed, width=12, height = 4)
-plot_cro_sites_collapsed
-```
-
-```{r chronos_4_sites_collapsed50}
-plot_cro_sites_collapsed50  <- cro_sites %>% 
-  filter(year > 1949) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  # geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1950, 2020, by=5)) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-
-plot_cro_sites_collapsed50
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed50.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed50, width=12, height = 4) 
-plot_cro_sites_collapsed50
-```
-
-```{r chronos_4_sites_collapsed90}
-plot_cro_sites_collapsed90  <- cro_sites %>% 
-  filter(year > 1989) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  # geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1990, 2016, by=2))
-
-plot_cro_sites_collapsed90
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed90.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed90, width=12, height = 4) 
-plot_cro_sites_collapsed90
-```
-
-# Correlation between chronos
-
-To compute similarity between chronos we use @Dorado2017 approach.
-
-* Para cada chrono se realiza un suavizado, utilizando diferente tamaño de ventana (1 a 40 years). Así evaluamos la sincronía entre dos series, tanto a nivel interanual como a largo plazo. 
-* Luego calculamos la correlación entre series para cada tamaño de ventana.
-Usamos correlaciones de Pearson o Sperman entre las series (es decir entre las zonas comunes de las cronologías).
-Ojo con la autocorrelación.
-Para ello puedo utilizar `interseries.cor()` del pkg `dplR`; o utilizar `cor.test()` (`stats`).
-El primero está basado en el segundo, así que utilizo el segundo.
-* Obtener IC utilizando bootstrap. Varias aproximaciones:
-
-    * Utilizar el pkg `boot` y conseguir IC. 
-    * Aproximación de @Mudelsee2014.
-En su paper habla de correlacion entre dos series y utiliza un programa en Fortran-90.
-He contactado con el autor y no tiene pensado hacerlo en R.
-He intentado utilizar Fortran-90, pero no tengo ni idea.
-Asi que desisto de utilizar esta computación de los IC.
-
-```{r}
-# centred moved average smooth
-suaviza_caH <- suaviza_cronos(cro_caH, win_vector=seq(1,40, by=1), crononame = 'cro_caH')
-suaviza_caL <- suaviza_cronos(cro_caL, win_vector=seq(1,40, by=1), crononame = 'cro_caL')
-suaviza_sjH <- suaviza_cronos(cro_sjH, win_vector=seq(1,40, by=1), crononame = 'cro_sjH')
-suaviza_sjL <- suaviza_cronos(cro_sjL, win_vector=seq(1,40, by=1), crononame = 'cro_sjL')
-```
-
-## between sites similarity (sjH *vs* sjL; caH *vs* caL)
-
-This note is for the plot:
-
-* Correlation among chronologies for the common period of each site in different time-domains after pre-filtering the time-series with increasing size of the moving-average window.
-
-```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-# set seed
-set.seed(333)
-
-# Metodo correlation
-tc <- c('spearman', 'pearson')
-
-# SJ 
-suaviza_sj <- rbind(suaviza_sjH, suaviza_sjL)
-correla_sj <- correlaBootIC(suaviza_sj, tc, years_suavizado = 40, 
-                            name_comparison = 'sjH-sjL')
-
-# CA 
-suaviza_ca <- rbind(suaviza_caH, suaviza_caL)
-correla_ca <- correlaBootIC(suaviza_ca, tc, years_suavizado = 40, 
-                            name_comparison = 'caH-caL')
-
-correla <- rbind(correla_ca, correla_sj)
-```
-
-```{r correla_chronos_sites, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-correla %>% 
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  facet_wrap(~metodo_correla, ncol=2) +
-  theme(strip.background = element_blank())
-```
-
-```{r correla_chronos_sites_pearson}
-correla_sites_pearson <- correla %>% filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  theme(legend.position = c(.8,.7))
-correla_sites_pearson
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/correla_boot_sitesHL.pdf') 
-ggsave(filename=fileplot, plot=correla_sites_pearson, width=5, height = 5) 
-```
-
-## between final sites similarity (sj *vs* caH; sj *vs* caL; caH *vs* caL)
-
-```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-# Metodo correlation
-tc <- c('spearman', 'pearson')
-
-set.seed(333)
-
-# Compute the suaviza_sj desde el principio, sin diferenciar h and l. Para ello la crono la computo de nuevo 
-cro_sjloc <- chrono_bai(bai_sj)
-cro_sjloc$site <- 'SJ'
-suaviza_sjloc <- suaviza_cronos(cro_sjloc, win_vector=seq(1,40, by=1), crononame = 'cro_sj')
-
-# sjcaH 
-suaviza_sjcaH <- rbind(suaviza_sjloc, suaviza_caH)
-correla_sjcaH <- correlaBootIC(suaviza_sjcaH, tc, years_suavizado = 40, 
-                            name_comparison = 'sj-caH')
-
-# sjcaL 
-suaviza_sjcaL <- rbind(suaviza_sjloc, suaviza_caL)
-correla_sjcaL <- correlaBootIC(suaviza_sjcaL, tc, years_suavizado = 40, 
-                            name_comparison = 'sj-caL')
-
-# Join all correlations 
-correla <- rbind(correla_ca, correla_sjcaL, correla_sjcaH)
-
-write.csv(correla, file=paste(di, "data/correla_ventanas_temporales/correla_chronos.csv", sep=""), row.names = FALSE)
-
-correla %>% 
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  facet_wrap(~metodo_correla, ncol=2) +
-  theme(strip.background = element_blank())
-
-```
-
-```{r correla_sitesSJCALH}
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/correla_boot_sitesSJCALH.pdf') 
-ggsave(filename=fileplot, plot=correla_sitesSJCALH, width=5, height = 5) 
-```
-
-# Chronologies by site (SJ, CAL, CAH)
-
-```{r}
-# Compute la crono de SJ de nuevo y llamar al site SJ
-cro_sjf <- chrono_bai(bai_sj)
-cro_sjf$site <- 'SJ'
-```
-
-```{r crono_compara_sitesSJCALH}
-# Combine and plot 
-cro_sites <- rbind(cro_caL, cro_caH, cro_sjf)
-
-## Export data of bai (chro)
-write.csv(cro_sites, file=paste(di, "data/cronos_medias/cronos_sites.csv", sep=""), row.names = FALSE)
-
-plot_cro_sitesSJCALH <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-
-plot_cro_sitesSJCALH
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH, width=12, height = 5)
-plot_cro_sitesSJCALH
-```
-
-```{r crono_compara_sitesSJCALH50}
-plot_cro_sitesSJCALH50 <- cro_sites %>% 
-  filter(year > 1949) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
-plot_cro_sitesSJCALH50
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH50.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH50, width=12, height = 5)
-```
-
-```{r crono_compara_sitesSJCALH90}
-plot_cro_sitesSJCALH90 <- cro_sites %>% 
-  filter(year > 1989) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-plot_cro_sitesSJCALH90
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH90.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH90, width=12, height = 5)
-```
-
-```{r, echo=FALSE, eval=FALSE}
-
-# CODE SNIPPETS 
-  # Loess 
-  l <- loess(b_aux$bai ~ b_aux$year, span = alpha)
-  years_v <- seq(min(b_aux$year), max(b_aux$year), 1)
-
-  l_pred_aux <- predict(l, years_v, se=TRUE)
-  l_pred <- as.data.frame(cbind(loess_fit = l_pred_aux$fit, 
-                loess_se = l_pred_aux$se.fit,
-                year = years_v))
-
-  out <- out %>% inner_join(l_pred, by='year')
-
-  out
-}
-
-# SJ
-
-cro_sj <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
-cro_sj %>% plot(add.spline=TRUE, nyrs=30)
-
-cro_ca <- chron(bai_ca, prefix = "CA", biweight = TRUE)
-cro_ca %>% plot(add.spline=TRUE, nyrs=30)
-
-cro_ca2 <- cro_ca %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('CA')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_sj2 <- cro_sj %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('SJ')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
-
-cro2 <- rbind(cro_ca2, cro_sj2)
-
-cro2 %>% ggplot(aes(x=year, y=bai_mean, colour=treatment)) + 
-  geom_line() + geom_smooth(method= 'loess', span = 0.30, se=FALSE) + theme_bw() + 
-  geom_bar(aes(x=year, stat = (samp.depth*10), colour=treatment))
-
-## San Juan Data
-
-Algunos árboles tienen 3 cores. Sin embargo, no los tres cores llegan hasta la corteza, por lo tanto, no podemos utilizar el diametro para estimar bai. Vamos a realizar lo siguiente:
-
-* Crear dos datasets: 
-    * Dataset con series de datos que llegan hasta corteza (`sj_cor`)
-    * Dataset con series de datos que no llegan hasta corteza (`sj_sin`): se trata de cores que no llegan hasta la corteza.
-
-* Para el cálculo del BAI, en el dataset `sj` utilizamos el diámetro medido en campo. En el caso del dataset `sj_sin` utilizamos la suma de todos los diámetros (:red_circle: `$TODO$ ASK to Guillermo`)
-
-# Get summary ring-width series
-sj_summ <- summary(sj)
-
-# Get names cores with last year different to 2016 
-id_cores_no_bark <- sj_summ %>% 
-  filter(last != 2016) %>% select(series) %>% mutate(series = factor(series))
-
-id_cores_no_bark <- unname(unlist(id_cores_no_bark))
-
-# Subbet datasets 
-sj_cor <- sj[which(! colnames(sj) %in% (id_cores_no_bark))]
-sj_sin <- sj[which(colnames(sj) %in% (id_cores_no_bark))]
-
-Lectura y preparación de datos de diámetro
-
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  select(id_focal, loc, dn_mm, height_cm)
-
-# Select only SJ trees and 
-ft_sj <- ft %>% filter(loc=='SJ')
-
-# get Create a dataframe with core ID
-cores <- data.frame(id_cores=colnames(sj))
-
-# Extract replicate and Tree ID from core ID 
-cores <- cores %>% 
-  mutate(id_focal = as.factor(stringr::str_sub(id_cores, 3,5)),
-         id_replica = stringr::str_sub(id_cores, 6,8))
-
-# Create df with diameter and height for each core ID 
-diam_cores <- cores %>% inner_join(ft_sj, by='id_focal') %>% 
-  select(id_cores, dn_mm)
-
-# remove diamm of cores without bark 
-diam_cores_sj <- diam_cores %>% filter(! id_cores %in% id_cores_no_bark)
-
-rwl.report(sj_cor)
-
-bai_p <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
-bai_d <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
-
-diam_cores_sjsin <- data.frame(id_cores = colnames(sj_sin),
-                               dn_mm = colSums(sj_sin, na.rm = TRUE))
-rownames(diam_cores_sjsin) <- NULL
-
-bai_psin <- bai_piovesan(rwdf = sj_sin, diam_df = diam_cores_sjsin)
-bai_dsinsin <- bai.out(rwl = sj_sin)
-bai_dsin <- bai.out(rwl = sj_sin, diam = diam_cores_sjsin)
-
-set.seed(1234)
-serie <- sample(colnames(b_test_dplR),1)
-
-require(reshape2)
-
-bp <- melt(t(b_test_piovesan))
-names(bp) <- c('cores', 'years', 'bai_p')
-
-bd <- melt(t(b_test_dplR))
-names(bd) <- c('cores', 'years', 'bai_d')
-
-diff <- bp %>% 
-  inner_join(bd, by=c('cores','years')) %>% 
-  mutate(diff = bai_p - bai_d)
-
-ggplot(x, aes(y=bai, x=years, colour=test)) + 
-  geom_point() + 
-  facet_wrap(~cores)
-
-# BAI dplR
-baisj <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
-
-baisjt <- bai.out(rwl = sjt)
-
-# Function to compute BAI (Piovesan et al)
-
-# rwl_file
-myrwl <- sj_cor
-# Diameter df  
-diam_df <- diam_cores_sj
-
-# Core
-id_core <- 'SNA0101'
-
-# Ring-width serie
-myrwl_serie <- myrwl[[id_core]]
-
-# Diameter for core id_core 
-dbh <- subset(diam_df, id_cores == id_core)
-dbh <- dbh[,2]
-
-myrwl_serie2 <- na.omit(myrwl_serie)
-cum
-
-r0 <- dbh/2 - c(0,cumsum(rev(myrwl_serie2)))
-
-bai <- (pi/4)*(dbh^2 - ((dbh - 2*myrwl_serie)^2))
-
-s <- myrwl_serie[94:96]
-
-d <- sum(s)*2
-
-########## test Piovesan
-# rwl_file
-myrwl <- baisj
-
-#id_core
-
-# Extract rwl serie 
-myrwl_i <- myrwl[[1]]
-
-dbh <- 372.4226
-
-# Year
-n.vec <- seq_len(nrow(myrwl))
-
-# Series
-seq_len(ncol(myrwl))
-
-# Select a ring-width series
-dat <- myrwl[[1]] 
-# Remove NA
-dat2 <- na.omit(dat)
-
-ggplot(s1, aes(y=SNA0101, x=row.names(s1))) + 
-  geom_point() + 
-  geom_line(b1, aes(y=SNA0101, x=row.names(b1)))
-
-ggplot(s1, aes(y=SNA0101, x=row.names(b1), group=1)) + geom_line()
-
-x <- c(1,2,2,3,1,3,4,5)
-
-xx <- rollmean(x=x, k=3, align = 'center', na.pad = T)
-
-# tes plots
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-carwi <- detrend(rwl = ca_tree, method = "ModNegExp")
-ca_prwi <- data.frame(rw = rowMeans(carwi, na.rm=TRUE))
-ca_prwi <- ca_prwi %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-xx <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-plot(xx$rw~rownames(xx))
-
-sjcrn <- chron(sjrwi, prefix = "rw_")
-plot(sjcrn, add.spline=TRUE, nyrs=20)
-plot()
-
-sj_tree <- rw_byTree(sj)
-ca_tree <- rw_byTree(ca)
-
-rownames(ca_tree) <- ca_tree$year
-ca_tree <- ca_tree[,-1]
-ca_p <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-ca_p <- ca_p %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-sj_p <- data.frame(rw = rowMeans(sj_tree, na.rm=TRUE))
-
-sj_p <- sj_p %>% mutate(year = rownames(.),
-                        site = as.factor('sj'),
-                        rw_log = log(rw))
-
-xx <- rbind(ca_p, sj_p)
-
-xx %>% ggplot(aes(x=as.Date(year, format='%Y'), y=rw_log)) + 
-  geom_line(aes(group=1)) + 
-  scale_x_date(date_labels = "%Y") + 
-  facet_wrap(~site, ncol=1, scales = 'free_y') + 
-  theme_bw() + 
-  geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-xx %>% filter(site =='ca') %>% 
-  ggplot(aes(x=as.Date(year, format='%Y'), y=rw)) + geom_point(colour='blue') +
-  geom_line(aes(group=1), colour='blue') + 
-  scale_x_date(date_labels = "%Y") + 
-  theme_bw() + 
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-ca_prwi %>%
-  ggplot(aes(x = as.Date(year, format = '%Y'), y = rw)) + geom_point() +
-  geom_line(aes(group = 1)) +
-  scale_x_date(date_labels = "%Y") +
-  theme_bw() +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width Index')
-
-```

### 2018-11-08 man/figures/growth_changes.Rmd
add improvement to gc

@@
 library("lemon")
+library("magrittr")
 ```
@@
 gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
+
+# gc_todos %<>% mutate(yearF = as.Date(as.character(year), format="%Y"))
 
@@
   geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black',fill='white'))
+  )
+
+gccaL <- gc_todos %>% 
+  filter(site == 'gc_caL') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
+  # scale_y_continuous(trans = pos_neg_log) +
+  # facet_wrap(~site_sorted, nrow = 2,
+  #            labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
+    ylab(custom_ylab) + custom_corrected_gc
+
+gccaH <- gc_todos %>% 
+  filter(site == 'gc_caH') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
+  # scale_y_continuous(trans = pos_neg_log) +
+  # facet_wrap(~site_sorted, nrow = 2,
+  #            labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
+    ylab(custom_ylab) + custom_corrected_gc
+
+ggplot_build(gccaH)$layout$panel_scales_y[[1]]$range$range
+ggplot_build(gccaL)$layout$panel_scales_y[[1]]$range$range
+
+##### -------------------------
+# VALIDO 
+# NEW
+th <- list(
+
+)
+
+custom_corrected_gc <- list(
+  geom_bar(stat='identity', fill=micolor, colour=micolor),
+  theme_bw(), 
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
+  theme(panel.grid.minor = element_blank(), 
+        strip.background = element_rect(colour='black',fill='white'))
+  )
+
+# Function for plot
+
+plotGC <- function(df, y, yse, x, my_ylab, micolor, xrange, ...){ 
+  # Get error bar aesthetic
+  yerrorbar <- aes_string(ymin = paste(y, yse, sep = '-'),
+                         ymax = paste(y, yse, sep = '+'))
+
+  p <- ggplot(df, aes_string(x = x, y = y)) +
+    geom_errorbar(mapping = yerrorbar, colour='gray', width=.35, size=.3) + 
+    geom_bar(stat='identity', fill=micolor, colour=micolor) +
+    geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
+    ylab(my_ylab) +
+    # Add line to %GC = 50 
+    geom_hline(yintercept=c(50, -50), linetype=2, colour='black') +
+    theme_bw() +
+    theme(panel.grid = element_blank()) +
+    # Customize limits and breaks to min of year data 
+    scale_x_continuous(limits = c(min(xrange), max(xrange)), 
+                       breaks = xrange, 
+                       sec.axis = dup_axis())
+  p
+}
+
+# Get min and max of year range and create vector of 10 years 
+xrange <- seq(round(min(gc_todos$year) / 10) * 10, 
+    round(max(gc_todos$year) / 10) * 10, by = 10)
+
+gc_todos %>% dplyr::filter(site == 'gc_caH') %>% 
+  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
+       my_ylab = '% GC', micolor = "#838B8B", xrange=xrange)
+
+gc_todos %>% dplyr::filter(site == 'gc_caL') %>% 
+  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
+       my_ylab = NULL, micolor = "#838B8B", xrange=xrange)
+
+gc_todos %>% dplyr::filter(site != 'gc_sj') %>% 
+  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
+       my_ylab = NULL, micolor = "#838B8B", xrange=xrange) + 
+  facet_rep_wrap(~site, ncol = 1) + 
+  theme(panel.spacing = unit(5, "lines"),
+        strip.placement = "inside", 
+        strip.background = element_blank())
+
+##### -------------------------
+# END VALIDO
+
+# Round to the nearest decade 
+# https://stackoverflow.com/questions/35352914/floor-a-year-to-the-decade-in-r 
+# options (lubridate::round_date(min(gc_todos$yearF), 'years')
+
+custom_ylab <- '% GC'
+micolor <- c("#838B8B")
+
+gccaL <- gc_todos %>% 
+  filter(site == 'gc_caL') %>% 
+  ggplot(aes(x=year, y=gc_md)) +
+
+gccaL <- gc_todos %>% 
+  filter(site == 'gc_caL') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
+  # scale_y_continuous(trans = pos_neg_log) +
+  # facet_wrap(~site_sorted, nrow = 2,
+  #            labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
+    ylab(custom_ylab) + custom_corrected_gc
+
+c <- gc_todos %>% 
+  filter(site != 'gc_sj') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
+  # scale_y_continuous(trans = pos_neg_log) +
+  facet_wrap(~site_sorted, nrow = 2,
+             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
+    ylab(custom_ylab) + custom_corrected_gc
+
+d <- gc_todos %>% 
+  filter(site == 'gc_sj') %>% 
+  ggplot(aes(x=year, y=gc_md)) + 
+  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
+  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
+  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
+  facet_wrap(~site_sorted, nrow = 2,
+             labeller=as_labeller(c("0_sj" = "SJ"))) + 
+  custom_corrected_gc + ylab('') + 
+  scale_y_continuous(trans = pos_neg_log, 
+                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
+  theme(axis.title.x = element_blank(),
+        axis.text.x = element_blank())
+
+plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
+plot_medians_corregido
+```
+
+gc_todos <- read.csv(here::here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
+
+# customize plot (corrected scale)
+pos_neg_log <- scales::trans_new('signed_log', 
+                                 transform = function(x) sign(x)*log(abs(x)),
+                                 inverse = function(x) sign(x)*exp(abs(x)))
+
+custom_ylab <- '% GC'
+micolor <- c("#838B8B")
+
+# solution for arrange
+# http://felixfan.github.io/stacking-plots-same-x/
+
+custom_corrected_gc <- list(
+  geom_bar(stat='identity', fill=micolor, colour=micolor),
+  theme_bw(), 
+  geom_hline(yintercept=c(50, -50), linetype=2, colour='black')
 

### 2018-11-08 man/figures/ring_climate.Rmd
add improvement to gc

@@
 
-ring_climate <- p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b'))
+ring_climate <- p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b')) & theme(plot.tag = element_text(size = 20, face = 2))
 

### 2018-11-08 man/figures/growth_changes.Rmd
ya casi lo tengo

@@
 
-plotGC <- function(df, y, yse, x, my_ylab, micolor, xrange, ...){ 
+plotGC <- function(df, y, yse, x, my_ylab, 
+                   micolor, ...){ 
   # Get error bar aesthetic
@@
   p <- ggplot(df, aes_string(x = x, y = y)) +
-    geom_errorbar(mapping = yerrorbar, colour='gray', width=.35, size=.3) + 
-    geom_bar(stat='identity', fill=micolor, colour=micolor) +
+    geom_errorbar(mapping = yerrorbar, colour = 'gray', width = .35, size = .3) + 
+    geom_bar(stat = 'identity', fill = micolor, colour = micolor) +
     geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
@@
     # Add line to %GC = 50 
-    geom_hline(yintercept=c(50, -50), linetype=2, colour='black') +
+    geom_hline(yintercept = c(50, -50), linetype=2, colour='black') +
     theme_bw() +
-    theme(panel.grid = element_blank()) +
+    theme(panel.grid = element_blank()) 
     # Customize limits and breaks to min of year data 
-    scale_x_continuous(limits = c(min(xrange), max(xrange)), 
-                       breaks = xrange, 
-                       sec.axis = dup_axis())
   p
@@
 
-gc_todos %>% dplyr::filter(site == 'gc_caH') %>% 
+    scale_x_continuous(limits = limitesx, 
+                       breaks = xrange, 
+                       position = pos) 
+                       #sec.axis = dup_axis(name = NULL))
+
+# Get automatically limits of y for caH anc caL
+minimoY <- min((gc_todos %>% 
+      dplyr::filter(site != 'gc_sj' & gc_md < 0) %>% 
+      mutate(error = gc_md - gc_md_se))$error, na.rm = TRUE)
+
+maximoY <- max((gc_todos %>% 
+      dplyr::filter(site != 'gc_sj' & gc_md > 0) %>% 
+      mutate(error = gc_md + gc_md_se))$error, na.rm = TRUE)
+
+# Ojo ver limites de ylim 
+# ggplot_build(pcaL)$layout$panel_params[[1]]$y.range
+# ggplot_build(pcaH)$layout$panel_params[[1]]$y.range
+
+# Establece rango de datos para y 
+mismo_y <- scale_y_continuous(limits=c(minimoY, maximoY), 
+                             breaks = c(-100, -50, 0, 50, 50, 100, 150))
+
+mismo_x <- scale_x_continuous(limits=c(min(gc_todos$year), max(gc_todos$year)), 
+                             breaks = xrange, position = 'bottom')
+
+#### --> Hay un error con la localizacion de las barras de error en el eje x. REVISAR
+
+pcaH <- gc_todos %>% dplyr::filter(site == 'gc_caH') %>% 
   plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
-       my_ylab = '% GC', micolor = "#838B8B", xrange=xrange)
+         my_ylab = '% GC', micolor = "#838B8B") + mismo_y + mismo_x +
+  annotate("text", x = 1990, y = 120, label = "CA-High", fontface = 2, size = 5)
+
+pcaL <- gc_todos %>% dplyr::filter(site == 'gc_caL') %>% 
+  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year',
+         my_ylab = '% GC', micolor = "#838B8B") +  
+         # xrange=xrange, pos = 'bottom', limitesx = limitesx) + 
+  ylab('') + mismo_y + mismo_x + 
+  annotate("text", x = 1990, y = 120, label = "CA-Low", fontface = 2, size = 5)
+
+psj <- gc_todos %>% dplyr::filter(site == 'gc_sj') %>% 
+  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
+         my_ylab = '% GC', micolor = "#838B8B") + 
+         # xrange=xrange, pos = 'top', limitesx = limitesx) + 
+  ylab('') + xlab('') + 
+  scale_y_continuous(trans = pos_neg_log, 
+                     breaks=c(-1000, -500, -100, 0, 100, 500)) + mismo_x + 
+  geom_hline(yintercept = 1) +
+  annotate("text", x = 1990, y = 500, label = "SJ", fontface = 2, size = 5)
+
+c(-116, 150)
+
+ggplot_build(pcaH)$layout$panel_scales_y[[1]]$range$range
+ggplot_build(gccaL)$layout$panel_scales_y[[1]]$range$range
+
+layer_scales(pcaH)$y$range$range
+
+layer_scales(c)$x$range$range[2]) +
 
@@
        my_ylab = NULL, micolor = "#838B8B", xrange=xrange) + 
-  facet_rep_wrap(~site, ncol = 1) + 
-  theme(panel.spacing = unit(5, "lines"),
-        strip.placement = "inside", 
+  facet_rep_wrap(~site, ncol = 1, strip.position = 'bottom') + 
+  theme(panel.spacing = unit(0, "lines"),
+        strip.placement = "outside", 
         strip.background = element_blank())

### 2018-11-09 man/figures/growth_changes.Rmd
update plot gc

--- 2018-10-27 man/figures/growth_changes.Rmd
@@
 
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
+**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q.
+pyrenaica* sites.
+Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
+Note that y-axes do not correspond in all of the three panels for the sake of clarity.
 
@@
 
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
+# Function for plot
 
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
+plotGC <- function(df, y, yse, x, my_ylab, 
+                   micolor, limitesX, ...){ 
+  # Get error bar aesthetic
+  yerrorbar <- aes_string(ymin = paste(y, yse, sep = '-'),
+                         ymax = paste(y, yse, sep = '+'))
 
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-
-        # strip.background = element_rect(colour='black',fill='white'))
-  )
-
-# Function to return a desired sequence breaks along axes
-# https://gist.github.com/valentinitnelav/ec7c253c80f4e045f4bd1c9b2920d7bd 
-my.breaks <- function(my.vector, step){
-    my.min <- floor(min(my.vector))
-    my.max <- ceiling(max(my.vector))
-    my.seq <- seq(from = round(my.min/step)*step, 
-                  to   = round(my.max/step)*step,
-                  by   = step)
-    return(my.seq)
+  p <- ggplot(df, aes_string(x = x, y = y)) +
+    geom_errorbar(mapping = yerrorbar, colour = 'gray', width = .35, size = .3, 
+                  na.rm = TRUE) + 
+    geom_bar(stat = 'identity', fill = micolor, colour = micolor) +
+    geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
+    ylab(my_ylab) +
+    # Add line to %GC = 50 
+    geom_hline(yintercept = c(50, -50), linetype=2, colour='black') +
+    theme_bw() +
+    theme(panel.grid = element_blank()) +
+    coord_cartesian(xlim=limitesX)
+    # Customize limits and breaks to min of year data 
+  p
 }
 
-gc_todos %>% 
-  filter(site == 'gc_caH') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
-  ylab(custom_ylab) +
-  GCtheme + 
-  ylim(-100, 150) +
-  scale_x_continuous(name = "Year", breaks = my.breaks(gc_todos$year, step = 25))
+# ---------- Min and Max axis        
+# Get min and max of year range and create vector of 10 years for labelling
+xrange <- seq(round(min(gc_todos$year) / 10) * 10, 
+    round(max(gc_todos$year) / 10) * 10, by = 10)
 
-GCtheme <- list(
-  # BAR
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme_bw(),
-  theme(# AXIS ticks
-        axis.ticks.length = unit(-1.4, "mm"),
-        axis.text.x = element_text(size = 8, margin = unit(c(t = 2.5, r = 0, b = 0, l = 0), "mm")),
-        panel.grid.minor = element_blank(),
-        # Remove strip 
-        strip.background = element_blank(),
-        strip.text = element_blank())
-)
+limitesX <- c(min(gc_todos$year), max(gc_todos$year))
 
-d <-
+# Get automatically limits of y for caH anc caL
+minimoY <- min((gc_todos %>% 
+      dplyr::filter(site != 'gc_sj' & gc_md < 0) %>% 
+      mutate(error = gc_md - gc_md_se))$error, na.rm = TRUE)
 
-  gc_todos %>% 
-  # mutate(year = lubridate::ymd(year, truncated = 2L)) %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2) +
-             # labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc +
-  scale_x_continuous(name = "Year", breaks = my.breaks(gc_todos$year, step = 25))
+maximoY <- max((gc_todos %>% 
+      dplyr::filter(site != 'gc_sj' & gc_md > 0) %>% 
+      mutate(error = gc_md + gc_md_se))$error, na.rm = TRUE)
 
-  scale_x_date(date_breaks = "25 year", date_labels = "%Y")
+# Establece rango de datos para y 
+mismo_y <- scale_y_continuous(limits=c(minimoY, maximoY), 
+                             breaks = c(-100, -50, 0, 50, 50, 100, 150))
 
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
+# plots
+
+pcaH <- gc_todos %>% dplyr::filter(site == 'gc_caH') %>% 
+  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
+         my_ylab = '% GC', micolor = "#838B8B", limitesX = limitesX) + mismo_y +
+  scale_x_continuous(breaks = xrange, position = 'bottom') + xlab('') + 
+  annotate("text", x = 1990, y = 120, label = "CA-High", fontface = 2, size = 7)
+
+pcaL <- gc_todos %>% dplyr::filter(site == 'gc_caL') %>% 
+  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year',
+         my_ylab = NULL, micolor = "#838B8B", limitesX = limitesX) +  mismo_y +
+  scale_x_continuous(breaks = xrange, position = 'bottom') + 
+  annotate("text", x = 1990, y = 120, label = "CA-Low", fontface = 2, size = 7)
+
+psj <- gc_todos %>% dplyr::filter(site == 'gc_sj') %>% 
+  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
+         my_ylab = NULL, micolor = "#838B8B", limitesX = limitesX) + 
+  xlab('') + 
+  scale_x_continuous(breaks = xrange, position = 'bottom') + 
   scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
+                     breaks=c(-1000, -500, -100, 1, 100, 500)) + 
+  geom_hline(yintercept = 1) +
+  annotate("text", x = 1990, y = 450, label = "SJ", fontface = 2, size = 7)
 
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
+gg <- psj + pcaH + pcaL + 
+  plot_layout(ncol=1) & 
+  theme(plot.tag = element_text(size = 20, face = 2)) & 
+  theme(plot.margin = margin(t=-1,r=3,b=-1,l=3, "pt"))
+
+gg
+
+# gg + plot_annotation(tag_levels = c('a','b', 'c'))
+
+# Ojo ver limites de ylim 
+# ggplot_build(pcaL)$layout$panel_params[[1]]$y.range
+# ggplot_build(pcaH)$layout$panel_params[[1]]$y.range
 

### 2018-11-09 man/figures/correla_boot_sites.Rmd
create correla figure

--- 2018-11-08 man/figures/ring_climate.Rmd
@@
 ##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
+**Figure S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
+Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
+Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
+
+```{r}
+correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
+
+correla <- correla %>% 
+  mutate(
+    name_comparison = as.factor(case_when(
+      name_comparison == "caH-caL" ~ "CAHigh - CALow",
+      name_comparison == "sj-caL" ~ "SJ - CALow",
+      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
+
+correla_sitesSJCALH  <- correla %>% 
+  filter(metodo_correla == 'pearson') %>%
+  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
+  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
+  xlab('Smoothing (centred moving average window size)') + 
+  # geom_smooth(method = 'loess', se=FALSE) + 
+  theme(panel.grid = element_blank()) +
+  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
+  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
+  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
+  theme(legend.position = c(.8,.9), legend.title = element_blank())
+
+correla_sitesSJCALH
+```
 

### 2018-11-09 man/figures/correla_boot_sites.Rmd
update

@@
 library("lemon")
+library("viridis")
+library("ggthemes")
 ```
@@
   filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
+  ggplot(aes(x=size, y=estimate)) + 
+  geom_line(aes(colour=name_comparison), size = 1.4) + 
+  theme_bw() + ylab('Correlation coefficient') +
   xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
   geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
+  geom_ribbon(aes(ymax=ci_upper, ymin=ci_lower, fill=name_comparison), alpha=0.3) + 
+  scale_fill_colorblind() + 
+  scale_color_colorblind() +
+  theme(legend.position = c(.8,.9), 
+        legend.title = element_blank(), 
+        legend.text = element_text(size = 14),
+        panel.grid = element_blank(), 
+        axis.title = element_text(size = 14), 
+        axis.text = element_text(size = 11))
 
 correla_sitesSJCALH
+
 ```
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'T max')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'T min')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name ='SPEI')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-theme_ring_climate <- list(
-  geom_hline(yintercept = 0, colour="gray"), 
-  geom_bar(aes(fill=site), stat="identity", position = position_dodge2(padding = 0.1)),
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.06))), 
-            position = position_dodge(.9), size = 6), 
-  theme_bw() +
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        panel.grid.minor.x = element_line(size=.5),
-        axis.text.x = element_text(angle = 90, vjust = 0.3)),
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site)
-)
-
-p_tmax <- df %>% filter(v_name == 'T max') %>% 
-  ggplot(aes(x=id, y=coef, group = site)) + 
-  theme_ring_climate + 
-  scale_y_continuous(limits = c(-0.4, 0.4), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4)) +
-  scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'bottom') + 
-  ylab('Correlation') + 
-  annotate("text", x = 2, y = 0.35, label = "T max", fontface =2, size = 5)
-
-p_tmin <- df %>% filter(v_name == 'T min') %>% 
-  ggplot(aes(x=id, y=coef, group = site)) + 
-  theme_ring_climate + 
-  scale_y_continuous(limits = c(-0.4, 0.6), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4)) +
-  scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'top') + 
-  ylab('Correlation') +
-  annotate("text", x = 2, y = 0.35, label = "T min", fontface =2, size = 5)
-
-# prec-spei
-
-dfspei <- df %>% 
-  filter(v_name == 'SPEI') %>% 
-  filter(month_name %in% c('Hydrol', 'Winter','Spring','Summer','Autumn')) %>% 
-  # truquillo para poner juntas las dos graficas 
-  mutate(id = id + 6)
-
-dfprec <- df %>% filter(v_name == 'Prec')
-
-# Joins df
-dfprec_spei <- bind_rows(dfspei, dfprec)
-
-# solucion para poner juntas prec + spei 
-labelitas <- unique(dfprec_spei[c('id','month_name','v_name')])
-
-p_prec_spei <- dfprec_spei %>% 
-  ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_vline(xintercept=20) +
-  theme_ring_climate + 
-  scale_y_continuous(limits = c(-0.2, 0.8), 
-                     breaks = c(-0.2, 0, 0.2, 0.4, 0.6, 0.8)) +
-  scale_x_continuous(breaks = labelitas$id, labels = labelitas$month_name, position = 'top') + 
-  ylab('Correlation') +
-  annotate("text", x = 2, y = 0.75, label = "Prec", fontface =2, size = 5) + 
-  annotate("text", x = 23, y = 0.75, label = "SPEI", fontface =2, size = 5)
-
-ptemperatura <- p_tmin + theme(legend.position="none") + p_tmax + theme(legend.position="bottom") + plot_layout(ncol=1)
-
-ring_climate <- p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b')) & theme(plot.tag = element_text(size = 20, face = 2))
-
-ring_climate
-```

### 2018-11-09 man/figures/browning.Rmd
update

---
output: 
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_pagebreak.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
```

```{r}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")

library("grid")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
library("stringr")
library("pander")
library("lemon")
library("viridis")
library("ggthemes")
```

```{r, echo=FALSE}
show_text <- FALSE
```
##### 
**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.

```{r}
df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)

df <- df %>% add_row(clu_pop2 = 'Northern slope',
                             type = 'greening',
                             count_clu = 0, per = 0.00, y = 2005)

df <- df %>% mutate(type = stringr::str_to_title(type))

per_pixels_green <- df %>% 
  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
  geom_bar(stat='identity', position = "dodge") + 
  # geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
  facet_wrap(~clu_pop2) + 
  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
  theme(strip.background = element_rect(colour = "black", fill = "white")
        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
  scale_y_continuous(expand=c(.1,0)) + 
  geom_text_repel(aes(label = paste0(per, " %")), 
                  position=position_dodge(width=0.9))

per_pixels_green
```

### 2018-11-09 man/figures/corr_plot_years.Rmd
update

--- 2018-10-05 man/ms_figures.Rmd
@@
     highlight: null
-    reference_docx: template_ms_new_page.docx
-bibliography: ../refs/references.bib
-csl: ../refs/ecology.csl
+    reference_docx: ../../templates/template_pagebreak.docx
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecology.csl
+editor_options: 
+  chunk_output_type: console
 ---
@@
 ```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-## Figure 1 Location map 
-- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset.
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf 
-    - Ojo con los nombres SJ (Que hacemos?)
-
-## Figure 2. Plot de anomalias
-- [X] Solamente Anomalias estandarizadas
-
-## Figure 3. 
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
-##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
-```
-
-```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
-```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
-```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-# summary(m1)
-myr2 <- format(summary(m1)$r.squared, digits = 3)
-
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
 ##### 
@@
 
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
+upper <- cor_rt 
+upper[upper.tri(cor_rt, diag = TRUE)]<- ""
+upper <- as.data.frame(upper)
+upper
+
+g_rt <- ggcorrplot(cor_rt, method = "circle", 
+                   type='upper', ggtheme = theme_minimal,
+                   lab=TRUE, title='BAI Rt', show.legend = FALSE, tl.cex = 14)
+
+# https://briatte.github.io/ggcorr/
+
+ggcorr(data = NULL, cor_matrix = cor_rt, geom = 'text', 
+       label_size = 7, label_round = 2, legend.position = 'none')
+
 g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
@@
 ```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-11-09 man/tables/table_droughts.Rmd
update table droughts long term

--- 2018-10-05 man/ms_figures.Rmd
@@
 library("dplyr")
-library("ggrepel")
-library("scales")
 
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-## Figure 1 Location map 
-- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset.
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf 
-    - Ojo con los nombres SJ (Que hacemos?)
-
-## Figure 2. Plot de anomalias
-- [X] Solamente Anomalias estandarizadas
-
-## Figure 3. 
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
 ##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
-```
-
-```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
-```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
-```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-# summary(m1)
-myr2 <- format(summary(m1)$r.squared, digits = 3)
-
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
+**Appendix S3.** Drought events for Sierra Nevada based on SPEI index (12 months scale).
 A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
@@
 ***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
+SPEI data were obtained for all 0.5º grid cells covering Sierra Nevada.
 
@@
 
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
+colnames(dys) <- c("Duration \n(months)", "Intensity", "Severity", "Lowest SPEI", "Months", "Year")
 
@@
 ```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-11-09 man/figures/browning.Rmd
update figure browing

@@
 ##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
+**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
 
@@
   geom_bar(stat='identity', position = "dodge") + 
-  # geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
+  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=4, angle = 90) + 
   facet_wrap(~clu_pop2) + 
   theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0)) + 
-  geom_text_repel(aes(label = paste0(per, " %")), 
-                  position=position_dodge(width=0.9))
+  theme(strip.background = element_rect(colour = "white", fill = "white"),
+        strip.text = element_text(size = 12), 
+        axis.title.y = element_text(size = 14, face = 'bold'),
+        axis.text = element_text(size = 11),
+        panel.grid = element_blank(),
+        legend.position = 'bottom') + 
+  labs(fill = 'Drought event') + 
+  scale_fill_colorblind() + 
+  scale_y_continuous(limits = c(0, 110), breaks = c(0, 25, 50, 75, 100))
 

### 2018-11-09 man/figures/plot_spei12.Rmd
update figure appendix s2

---
output: 
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_pagebreak.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
```

```{r}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")
library("stringr")
library("viridis")
library("ggthemes")
```

```{r, echo=FALSE}
show_text <- FALSE
```

##### 
**Appendix S2**. Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI). Data from Global SPEI database (http://spei.csic.es/database.html). We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.

```{r, echo=FALSE}
# Get file names
myfiles <- list.files(path = here::here("/data_raw/spei_all/"), pattern = "\\.csv$")

mydf <- data.frame()

# Loop to read files and prepare data
for (j in myfiles){ 
  aux <- read.csv(file=paste0(here::here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')

  # Remove csv and get name 
  name_aux <- str_replace(j, ".csv", "")

  # Get lat long
  latlong <- str_replace(name_aux, "SPEI_", "")

  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]

  aux$lat <- mylat
  aux$long <- mylong

  # Split time 
  # Get year 
  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))

  # Get month 
  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")

  # Variable month in number 
  aux$months <- match(aux$month, month.abb)

  # assign(name_aux, aux)

  mydf <- rbind(mydf, aux)
}

# Select SPEI 12
spei <- mydf %>% select(SPEI_12, year, months) %>% 
  group_by(year,months) %>% 
  summarise(spei = mean(SPEI_12),
            spei_sd = sd(SPEI_12)) %>% 
  mutate(date = lubridate::ymd(paste(year, months, "01")))

spei <- spei %>% 
  mutate(sign = case_when(
    spei > 0 ~ "pos",
    TRUE ~ "neg"))
```

```{r, fig.height=5}
p <- spei %>% 
  ggplot(aes(x=date, y=spei, fill=sign)) +  
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = c("pos" = "#696969", "neg" = "red")) +
  theme_bw() + 
  geom_hline(yintercept = 0, colour = 'gray') +
  labs(x = '', y ='SPEI (12-month scale)') +
  scale_x_date(date_breaks = "5 year", date_labels = "%Y", 
               limits=as.Date(c("1949-01-01", "2015-12-01"))) +
  theme(strip.background = element_rect(fill = "white"),
        legend.position = "none", 
        panel.grid = element_blank(), 
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=9),
        axis.title.y = element_text(size = 12)) 
```

### 2018-11-10 man/figures/plot_spei12.Rmd
change color

@@
   geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "#696969", "neg" = "red")) +
+  scale_fill_manual(values = c("pos" = "#696969", "neg" = "brown4")) +
   theme_bw() + 
@@
         axis.title.y = element_text(size = 12))
+
+p
 ```

### 2018-11-10 man/figures/gc_perc_trees.Rmd
add figure corrected

---
output: 
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_pagebreak.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
```

```{r}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")
library("grid")
library("gridExtra")
library("patchwork")
library("stringr")
library("pander")
library("lemon")
library("magrittr")
```

```{r, echo=FALSE}
show_text <- FALSE
```

##### 
**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis). Data for number of trees > 2 is shown.

```{r}
gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')

colorpositivo <- c("#104E8B")
colornegativo <- c("#CD3333")

# Get min and max year and round to nearest decade
getBreaks <- function(df, v, by){ 
  x <- seq(round(min(df[v]) / 10)*10, 
           round(max(df[v]) / 10)*10, by= by) 
  x
  }

xrange <- getBreaks(gc_treesMed50, 'year', by=10)

label_site <- c("0_sj" = "SJ",
                "1_caH" = "CA-High",
                "2_caL" = "CA-Low")

pgc_median50 <- gc_treesMed50 %>% 
  filter(total_tree > 2) %>% 
  ggplot() +
  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, 
           stat = "identity", position = "dodge") + 
  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, 
           stat = "identity", position = "dodge") +
  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
  theme_bw() + ylab('% trees') +
  geom_hline(yintercept=0, alpha=.4) +
  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
  scale_x_continuous(breaks = xrange) +
  scale_y_continuous(limits = c(-100, 100), breaks = seq(-100, 100, by = 50),
                     sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
  facet_rep_wrap(~site_sorted, nrow = 3, 
                 repeat.tick.labels = TRUE, labeller = as_labeller(label_site)) +
  theme(panel.grid = element_blank(), 
        strip.background = element_blank(), 
        strip.text = element_blank(),
        axis.text.y = element_text(size=10),
        axis.text.x = element_text(size=9),
        axis.title = element_text(size = 12))

pgc_median50
```

### 2018-11-12 man/figures/resilience_long_term.Rmd
figure resilience long_term

--- 2018-10-05 man/ms_figures.Rmd
@@
 library("patchwork")
-library("ggcorrplot")
 library("stringr")
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-## Figure 1 Location map 
-- [X] La figura 1 puede tener un mapa de localización de SN, otro de las poblaciones de roble (clasificadas por colores: cluster; y señalando las dos poblaciones muestradas en dendro).
-Ver MIGRAME dataset.
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf 
-    - Ojo con los nombres SJ (Que hacemos?)
-
-## Figure 2. Plot de anomalias
-- [X] Solamente Anomalias estandarizadas
-
-## Figure 3. 
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in Iberian Peninsula (a) and in Sierra Nevada mountain (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment, Regional Government of Andalusia.
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
 ##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
-```
+**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events (as from Appendix S3).
+*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
+Points indicate average of resilience metrics for all populations.
+Error bar corresponds standard error.
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
 
 ```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
-```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
-```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# Lm CA-High
-df <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-m1 <- lm(bai_mean ~ year, df)
-# summary(m1)
-myr2 <- format(summary(m1)$r.squared, digits = 3)
-
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas coorespond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show chronologies with # trees > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r myr2`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) 
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
+df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
 
@@
 dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
+# dys <- dys[,'maxyear']
 
@@
 
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
+plotrel <- function(df, v, x, label, coordX_eq, coordY_eq, titulo, ...){
   require(ggplot2)
@@
     geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
+    geom_point(size=1.5, pch = 21, colour="black", fill="white") + 
     theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
+    geom_text_repel(nudge_x = 0.5) + 
     annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
+             label=lm_eqn(lm(formulita, df)), parse=TRUE, 
+             size = 5, fontface = "bold") +
     ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
+    theme(plot.title = element_text(size = 14, face = "bold.italic"),
+          panel.grid = element_blank(), 
+          axis.text.y = element_text(size=10),
+          axis.text.x = element_text(size=9),
+          axis.title = element_text(size = 12))
   }
 
+sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
+                 coordX_eq = 10, coordY_eq = 0.5, titulo = "Rt") +
+  xlab('') + ylab('')
+
 sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
+                 coordX_eq = 10, coordY_eq = 0.5, titulo = "Rs") + 
+  xlab('') + ylab('')
 
 sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
+                 coordX_eq = 10, coordY_eq = 0.5, titulo = "Rc") +
+  xlab("Drought Severity") + ylab('')
 
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
+res_long_term <- sevRt + sevRc + sevRs + plot_layout(nrow = 1)
 
+res_long_term
 ```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-11-12 man/figures/corr_plot_years.Rmd
correlation plots

@@
 library("grid")
+library("GGally")
 library("gridExtra")
@@
 
-upper <- cor_rt 
-upper[upper.tri(cor_rt, diag = TRUE)]<- ""
-upper <- as.data.frame(upper)
-upper
+correP <- function(df, mytitle, ...){
+  require(ggplot2)
+  require(ggcorrplot)
 
-g_rt <- ggcorrplot(cor_rt, method = "circle", 
-                   type='upper', ggtheme = theme_minimal,
-                   lab=TRUE, title='BAI Rt', show.legend = FALSE, tl.cex = 14)
+  p <- ggcorrplot(df, method = "square", type ='upper', 
+                  outline.color = "black", colors = "white",
+                  ggtheme = theme_minimal, 
+                  lab = TRUE, lab_size = 12,
+                  title = mytitle, show.legend = FALSE, 
+                  tl.cex = 10)
+  p <- p + theme(plot.title = element_text(size = 14, face = "bold.italic"))
+  p      
+}
 
-# https://briatte.github.io/ggcorr/
-
-ggcorr(data = NULL, cor_matrix = cor_rt, geom = 'text', 
-       label_size = 7, label_round = 2, legend.position = 'none')
-
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
+g_rt <- correP(cor_rt, 'BAI Rt')
+g_rc <- correP(cor_rc, 'BAI Rc')
+g_rs <- correP(cor_rs, 'BAI Rs')
+g_rte <- correP(cor_rte, 'EVI Rt')
+g_rce <- correP(cor_rce, 'EVI Rc')
+g_rse <- correP(cor_rse, 'EVI Rs')
 

### 2018-11-12 analysis/analysis_stand_features.Rmd
update analisis stand features

@@
 date: "2018 Jan"
-output:  pdf_document
-bibliography: references.bib 
-csl: ecology.csl
+output:  
+  md_document:
+    toc: true
+    variant: markdown_github
+bibliography: ../refs/references.bib
+csl: ../refs/ecology.csl
 ---
@@
 ```{r, warning=FALSE, message=FALSE, echo=FALSE}
+library("here")
 library("tidyverse")
@@
 ```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
+# machine <- 'ajpelu'
+# # machine <- 'ajpeluLap'
+# di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
 
 # Read tree data
-tree <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
+tree <- read.csv(file=here::here('/data_raw/', 'dendro_competence.csv'), header=TRUE, sep=',')
 
@@
 # Centroid of focal tree 
-## filename 
-mygpx_ca <- paste0(di, 'data_raw/geoinfo/dendro_ca.GPX')
-mygpx_sj <- paste0(di, 'data_raw/geoinfo/dendro_sj.GPX')
-
-## reading  
-field_work_ca <- readOGR(mygpx_ca, layer="waypoints", verbose = FALSE)
-field_work_sj <- readOGR(mygpx_sj, layer="waypoints", verbose = FALSE)
+field_work_ca <- readOGR(here::here("/data_raw/geoinfo","dendro_ca.GPX" ), layer="waypoints", verbose = FALSE)
+field_work_sj <- readOGR(here::here("/data_raw/geoinfo","dendro_sj.GPX" ), layer="waypoints", verbose = FALSE)
 ```
@@
 # Read competition
-compe <- read.csv(file=paste(di, "/data/competence/competence_indices.csv", sep=""), header=TRUE, sep=',')
+compe <- read.csv(here::here("/data/competence", "competence_indices.csv"), header=TRUE, sep=',')
 
@@
   out <- get(paste0('aov_', i))
-  file_out <- file(paste0(di,'out/anovas_competition/aov_', i, '.txt'), "w")
+  file_out <- file(here::here("out/anovas_competition", paste0("aov_", i, ".txt")), "w")
   sink(file_out)
@@
 
-tm <- s$tm
+tm <- s$tm %>% as.data.frame()
 
@@
 # Read topo data 
-topo <- read.csv(file=paste(di, "/data/topo/topo.csv", sep=""), header=TRUE, sep=',')
+topo <- read.csv(file=here::here("/data/topo", "topo.csv"), header=TRUE, sep=',')
 
@@
 
-  file_out <- file(paste0(di,'out/anovas_topo/aov_', i, '.txt'), "w")
+  file_out <- file(here::here("/out/anovas_topo", paste0("aov_", i, ".txt")), "w")
   sink(file_out)
@@
 
-  file_out <- file(paste0(di,'out/anovas_ft/aov_', i, '.txt'), "w")
+  file_out <- file(here::here("out/anovas_ft", paste0("aov_", i, ".txt")), "w")
   sink(file_out)
@@
 # Export data 
-write.csv(stand_biotic, file=paste(di, "data/competence/biotic_stand_features.csv", sep=""), row.names = FALSE)
+write.csv(stand_biotic, file=here::here("data/competence", "biotic_stand_features.csv"), row.names = FALSE)
 
@@
 # write.csv(out4, file=paste(di, "data/proto_tables/site_features.csv", sep=""), row.names = FALSE)
-write.csv(out4, file=paste(di, "data/competence/resumen_medias_indices_competencia.csv", sep=""), row.names = FALSE)
+write.csv(out4, file=here::here("data/competence","resumen_medias_indices_competencia.csv"), row.names = FALSE)
 
@@
   elev_max = max(mde), 
-  slope_mean = mean(slope)
+  elev_sd = sd(mde), 
+  elev_se = sd(mde)/sqrt(n()),
+  slope_mean = mean(slope), 
+  slope_sd = sd(slope), 
+  slope_se = sd(slope)/sqrt(n()),
 )
@@
 
-write.csv(site_general, file=paste(di, "data/sites_summary/site_general.csv", sep=""), row.names = FALSE)
+write.csv(site_general, file=here::here("data/sites_summary", "site_general.csv"), row.names = FALSE)
 ```

### 2018-11-12 man/tables/t_sampled_plots.Rmd
update table sampled plots

--- 2018-04-23 man/ms_tables/t_sampled_plots.Rmd
@@
 ```{r, echo=FALSE}
-topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
-ages <- read.csv(here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
-dendro_summ <- read.csv(here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
-biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)
-letritas <- read.csv(here("/out/kw_stand_features", "kw_letras.csv"), header = TRUE)
+topo <- read.csv(here::here("/data/sites_summary", "site_general.csv"), header = TRUE)
+ages <- read.csv(here::here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
+dendro_summ <- read.csv(here::here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
+biotic <- read.csv(here::here("/data/competence", "biotic_stand_features.csv"), header = TRUE)
+letritas <- read.csv(here::here("/out/kw_stand_features", "kw_letras.csv"), header = TRUE)
 
@@
 
-clima <- read.csv(here("/data/climate_obsnev", "obsnev_qp_pop.csv"), header = TRUE)
+clima <- read.csv(here::here("/data/climate_obsnev", "obsnev_qp_pop.csv"), header = TRUE)
 cli <- clima %>% filter(pop %in% c(2,6)) %>% mutate_if(is.numeric, funs(round(., 1))) %>% 
@@
 t1 <- t1raw %>% dplyr::select(
-  name, lat_m, long_m, elev_mean, slope_mean, prec, temp, 
+  name, lat_m, long_m, elev_min, elev_max, slope_mean, slope_sd, prec, temp, 
   n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, 
@@
   mutate(pare = ")") %>% 
-  mutate_at(c("lat_m","long_m","slope_mean","ba_mean","ba_sd","srd_mean","srd_sd"), 
+  mutate_at(c("lat_m","long_m","slope_mean","slope_sd","ba_mean","ba_sd","srd_mean","srd_sd"), 
             funs(formatC(., digits = 2,format='f'))) %>%  
@@
             funs(formatC(., digits = 1,format='f'))) %>%
-  mutate_at(4, round, 0) %>% 
+  mutate_at(c("elev_min", "elev_max"), funs(round(., 0))) %>% 
+  unite(elev, c(elev_min, elev_max), sep= " - ") %>% 
+  unite(slope, c(slope_mean, slope_sd), sep=" (") %>% 
   unite(n_trees, c(n_trees, n_radii), sep=" (") %>% 
@@
   mutate(
+    slope = paste0(slope, pare),
     n_trees = paste0(n_trees, pare),

### 2018-11-14 man/tables/t_sampled_plots.Rmd
update table

@@
     highlight: null
-    reference_docx: ../template_tables_portrait.docx
+    reference_docx: ../../templates/template_tables_portrait.docx
 bibliography: ../../refs/references.bib
@@
 colnames (t1) <- c("Site", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
-                   "Prec. (mm)", "Temp (&deg; C)",
+                   "Prec. (mm)", "Temp. (&deg; C)",
                    "# trees (# cores)", "Dbh (cm)", "Height (m)", "Age (years)", 

### 2018-11-14 man/old_tables/ms_tables.Rmd
update table 2

--- 2018-01-22 man/ms_tables_portrait.Rmd
@@
 ---
-title: "ms_tables_portrait"
+title: "ms_tables"
 output:  
-  word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 7.5
-    highlight: null
-    reference_docx: template_tables_portrait.docx
-bibliography: ../refs/references.bib
-csl: ../refs/ecology.csl
+  md_document:
+    toc: true
+    variant: markdown_github
+bibliography: ../references.bib
+csl: ../ecology.csl
 ---
@@
 
+```{r}
+panderOptions('table.split.table', Inf)
+```
+
 ```{r echo=FALSE, results='asis', eval=show_text}
@@
 
-# Table 3 ANOVAS resilience BAI and EVI
+# Table 2 ANOVAS resilience BAI and EVI
 
@@
 ## Table 1. Characteristics of the sampled plots 
-
 ```{r, echo=FALSE}
-# https://rdrr.io/cran/pander/man/panderOptions.html
-
 topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
@@
 t1 <- t1raw %>% dplyr::select(
-  site, lat_m, long_m, elev_mean, slope_mean, 
+  site, lat_m, long_m, elev_mean, elev_min, elev_max, slope_mean, 
   n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, 
   dn_mean_all, dn_sd_all, h_mean_all, h_sd_all,
-  ba_mean, ba_sd, stand_density_mean, stand_density_sd) %>% 
-  mutate(pare = ")") %>% 
-  mutate_at(c(2,3,5,18,19), round, 2) %>% 
-  mutate_at(4, round, 0) %>% 
-  mutate_at(c(8:17, 20:21), round, 1) %>% 
-  unite(dn, c(dn_mean, dn_sd), sep=" (") %>% 
-  unite(h, c(h_mean, h_sd), sep=" (") %>% 
-  unite(age, c(age_mean, age_std), sep=" (") %>% 
-  unite(dn_all, c(dn_mean_all, dn_sd_all), sep=" (") %>% 
-  unite(h_all, c(h_mean_all, h_sd_all), sep=" (") %>% 
-  unite(ba, c(ba_mean, ba_sd), sep=" (") %>% 
-  unite(stand_density, c(stand_density_mean, stand_density_sd), sep=" (") %>% 
-  mutate(
-    dn = paste0(dn, pare),
-    h = paste0(h, pare), 
-    age = paste0(age, pare),
-    dn_all = paste0(dn_all, pare),
-    h_all = paste0(h_all, pare),
-    ba = paste0(ba, pare),
-    stand_density = paste0(stand_density, pare)) %>% 
-  dplyr::select(-pare)
+  ba_mean, ba_sd, stand_density_mean, stand_density_sd)
 
-colnames (t1) <- c("Name", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
-                   "# trees", '# cores', "Dbh (cm)", "Height (m)", "Age (years)", 
-                   "Dbh all (cm)", "Height all (m)", "BA (m2/ha)", "Density (trees/ha)")
+colnames (t1) <- c("Name", "Lat (º)", "Long (º)", "Elevation (m)", "", "", "Slope (º)", 
+                   "# trees", '# cores', "Dbh (cm)", "", "Height (m)", "", "Age (years)", "", 
+                   "Dbh all (cm)", "", "Height all (m)", "", "BA (m2/ha)", "", "Density (trees/ha)", "")
 
-pander(t1, split.table = Inf)
+pander(t1,
+       round=c(0, 2, 2, 0, 0, 0, 2, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 1, 1),
+             keep.line.breaks = TRUE)
 ```
 
-## Table 2. Dendro summary
-
-```{r, echo=FALSE}
-t2 <- t1raw %>% dplyr::select(
-  site, first_year, last_year, 
-  length_year, length_year5, 
-  n_trees, n_radii,
-  rw_mean, rw_std, ms2, ar1, rbar.wt, eps)  %>% 
-  mutate(pare = ")") %>% 
-  mutate_at(c(8:13), round, 3) %>% 
-  unite(length_year, c(length_year,length_year5), sep=" (") %>%
-  unite(RW, c(rw_mean,rw_std), sep=" (") %>%
-  mutate(
-    length_year = paste0(length_year, pare),
-    RW = paste0(RW, pare)) %>% 
-  dplyr::select(-pare)
-
-colnames(t2) <- c("Name", "First year", "Last year", "Length (years)",  
-                   "# trees", '# cores', "RW (mm)", "MS", "AR(1)", "Rbt", "EPS")
-
-pander(t2,
-       split.table = Inf)
-
-```
-
-## Table 3. Robust ANOVAS (F-value)
+## Table 2. Robust ANOVAS (F-value)
 ```{r, echo=FALSE, include = FALSE}
@@
 ```
-
-## Table RS1
-
-* Estimadores huber resiliences EVI
-
-```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)
-
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
-
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
-
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
-```
-
-# Table RS2
-
-* Estimadores huber resiliences Bai
-
-```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
-
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
-
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
-
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
-```
-
-```{r, echo = FALSE, warning = FALSE, eval = FALSE}
-
-# Este código es para print la tabla anterior en formato pdf latex (ver state_ms)c
-### OJO REVISAR 
-# ------------------------------
-
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
-
-m <- mraw %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter))
-
-# Datos agrupados para todos los sitios 
-mraw_a <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
-
-m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  mutate(site = "z_all")
-
-# Datos agrupados para todos los sitios 
-mraw_b <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
-
-m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
-  rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>%  
-  mutate(disturb_year = 2099)
-
-mm <- m %>% 
-  bind_rows(m_a) %>%
-  bind_rows(m_b) %>% 
-  gather(temp, numeros, c(M.Huber, ci)) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
-
-mm <- mm %>% 
-  dplyr::select(site, 
-                rc_2005_M.Huber, rc_2005_ci, rt_2005_M.Huber, rt_2005_ci, rs_2005_M.Huber, rs_2005_ci,
-                rc_2012_M.Huber, rc_2012_ci, rt_2012_M.Huber, rt_2012_ci, rs_2012_M.Huber, rs_2012_ci,
-                rc_2099_M.Huber, rc_2099_ci, rt_2099_M.Huber, rt_2099_ci, rs_2099_M.Huber, rs_2099_ci)
-
-mmt <- mm %>%
-   rownames_to_column %>% 
-   gather(var, value, -rowname) %>% 
-   spread(rowname, value)
-
-colnames(mmt) <- c("var", "SJ", "caH", "caL", "z_all") 
-mmt <- mmt[-19,] 
-mmt <- mmt %>% 
-  mutate(vars = c(rep("Rc",6), rep("Rt",6), rep("Rs",6)),
-         year = c(rep(
-           c(rep("2005", 2), 
-             rep("2012", 2),
-             rep(" ", 2)), 3))) %>% 
-  dplyr::select(vars, year, SJ, caH, caL, var)
-
-kable(mmt, 
-      format = "latex", booktabs = T) %>% 
-  kable_styling(font_size = 6) %>% 
-  add_header_above(c("", "", "Northern", "Southern" = 2, "")) 
-```

### 2018-11-14 man/tables/t_dendro_summary.Rmd
update table 2

--- 2018-01-22 man/ms_tables/t_dendro_summary.Rmd
@@
 **Table 2.** Characteristics of the mean tree ring chronologies.
-Values of the length year in parenthesis indicate years replicated more than five series.
+Values of the length year in parenthesis indicate years replicated more with than five series.
 RW = mean annual ring width (standard deviation in parenthesis).
@@
 ```{r, echo=FALSE}
-topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
-ages <- read.csv(here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
-dendro_summ <- read.csv(here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
-biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)
+dendro_summ <- read.csv(here::here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
 
-t1raw <- topo %>% inner_join(ages, by='site') %>% 
-  inner_join(dendro_summ, by='site') %>% 
-  inner_join(biotic, by='site')
+t1raw <- dendro_summ %>% mutate(name = case_when(
+  site == 'caH' ~ 'CA-High', 
+  site == 'caL' ~ 'CA-Low',
+  site == 'sj' ~ 'SJ'))
 
-t1raw <- t1raw %>% mutate(name = case_when(
-  site == 'caH' ~ 'Cáñar-High', 
-  site == 'caL' ~ 'Cáñar-Low',
-  site == 'sj' ~ 'San Juan'))
-
-t2 <- t1raw %>% dplyr::select(
-  site, first_year, last_year, 
-  length_year, length_year5, 
-  n_trees, n_radii,
-  rw_mean, rw_std, ms2, ar1, rbar.wt, eps)  %>% 
+t2 <- t1raw %>% 
   mutate(pare = ")") %>% 
@@
     RW = paste0(RW, pare)) %>% 
-  dplyr::select(-pare)
+  dplyr::select(name, first_year, last_year, 
+  length_year, n_trees, n_radii,
+  RW, ms2, ar1, rbar.wt, eps, -pare, -site)
 

### 2018-11-14 man/ms_tables/t_robust_anovas.Rmd
update tables

@@
 output:  
-  word_document:
-    reference_docx: "../../man/template_tables_portrait.docx"
+  pdf_document
 ---
@@
   mutate(factores = recode(factores, 
-                           "disturb_year" = "Drougth event", 
-                           "disturb_year:site" = "Drougth event X Site", 
+                           "disturb_year" = "Drought event", 
+                           "disturb_year:site" = "Drought event X Site", 
                            "site" = "Site")) %>% 
-  arrange(match(factores, c("Drougth event", "Site", "Drougth event X Site"))) %>% 
+  arrange(match(factores, c("Drought event", "Site", "Drought event X Site"))) %>% 
   mutate_at(c(2,4,6,8), round, 2) %>% 
@@
 ```{r, echo=FALSE}
-colnames(dff) <- c("variable", "factores", "rt\nF", "rt\np", "rc\nF", "rc\np", "rs\nF", "rs\np")
-pander(dff)
+# https://haozhu233.github.io/kableExtra/use_kableExtra_with_formattable.html
+colnames(dff) <- c("variable", "factors", "F", "p", "F", "p", "F", "p")
+kable(dff, format = "latex") %>% add_header_above(c("", "", "Resistance" = 2, "Recovery" = 2, "Resilience" = 2))
+
+# https://davidgohel.github.io/flextable/articles/examples.html
+
 ```

### 2018-11-16 analysis/evi/explore_resilience.Rmd
update

@@
     variant: markdown_github
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library('dplyr')
+library('here')
 library('tidyverse')
@@
 # Read data
-raw_evires <- read.csv(file=paste(di, "/data/resilience/resilience_evi.csv", sep=""), header = TRUE, sep = ',')
+raw_evires <- read.csv(here::here("/data/resilience","resilience_evi.csv"), header = TRUE, sep = ',')
 
 # add data of pop
-anomalias <- read.csv(file=paste(di, "/data/anomalies/anomalias_evimean.csv", sep=""), header = TRUE, sep = ',')
+anomalias <- read.csv(here::here("/data/anomalies", "anomalias_evimean.csv"), header = TRUE, sep = ',')
 
@@
 ```{r, echo=FALSE}
-pdf(paste0(di, '/out/mean_resilience/evi/interaction_plotsSE.pdf'), width=9, height = 9)
+pdf(here::here('/out/mean_resilience/evi', 'interaction_plotsSE.pdf'), width=9, height = 9)
 grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
@@
 
-pdf(paste0(di, '/out/mean_resilience/evi/interaction_plotsCI.pdf'), width=9, height = 9)
+pdf(here::here('/out/mean_resilience/evi','interaction_plotsCI.pdf'), width=9, height = 9)
 grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
@@
 
-write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/evi/anovas_statistics.csv'), row.names = F)
+write.csv(aovas_coeff, file=here::here('/out/anovas_resilience/evi', 'anovas_statistics.csv'), row.names = F)
 
@@
 write.csv(aovas_model_summary, 
-          file=paste0(di, '/out/anovas_resilience/evi/anovas_summary_modelos.csv'), row.names = F)
+          file=file=here::here('/out/anovas_resilience/evi','anovas_summary_modelos.csv'), row.names = F)
 
@@
 write.csv(normtestA, 
-          file=paste0(di, '/out/anovas_resilience/evi/normo_disturb_year.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/evi','normo_disturb_year.csv'), row.names = F)
 ```
@@
 write.csv(normtestB, 
-          file=paste0(di, '/out/anovas_resilience/evi/normo_site.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/evi','normo_site.csv'), row.names = F)
 ```
@@
 write.csv(normtestAB, 
-          file=paste0(di, '/out/anovas_resilience/evi/normo_disturb_year_site.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/evi','/normo_disturb_year_site.csv'), row.names = F)
 ```
@@
 write.csv(homo, 
-          file=paste0(di, '/out/anovas_resilience/evi/homocedasticidad.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/evi','homocedasticidad.csv'), row.names = F)
 ```
@@
 write.csv(homo_log, 
-          file=paste0(di, '/out/anovas_resilience/evi/homocedasticidad_log.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/evi','homocedasticidad_log.csv'), row.names = F)
 ```
@@
 write.csv(homo_log, 
-          file=paste0(di, '/out/anovas_resilience/evi/homocedasticidad_log_plus_1.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/evi','homocedasticidad_log_plus_1.csv'), row.names = F)
 ```
@@
 
-pdf(paste0(di, '/out/resilience_robust/evi/robust_plot_evi_drought.pdf'), width=9, height = 9)
+pdf(here::here('/out/resilience_robust/evi/','robust_plot_evi_drought.pdf'), width=9, height = 9)
 robust_plot_evi_drought 
@@
 
-pdf(paste0(di, '/out/resilience_robust/evi/robust_plot_evi_site.pdf'), width=9, height = 9)
+pdf(here::here('/out/resilience_robust/evi/', 'robust_plot_evi_site.pdf'), width=9, height = 9)
 robust_plot_evi_site 
@@
 # Export data 
-write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber.csv'), row.names = F)
-write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_agg.csv'), row.names = F)
-write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_a.csv'), row.names = F)
-write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_agg_a.csv'), row.names = F)
-write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_b.csv'), row.names = F)
-write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_mhuber_agg_b.csv'), row.names = F)
-
-write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/huber_evi/robust_pairwise.csv'), row.names = F)
-write.csv(trimmedanovas, file=paste0(di, '/out/anovas_resilience/huber_evi/trimmed_anovas.csv'), row.names = F)
-```
+write.csv(mhuber, file=here::here('/out/anovas_resilience/huber_evi','robust_mhuber.csv'), row.names = F)
+write.csv(mhuber_agg, file=here::here('/out/anovas_resilience/huber_evi','robust_mhuber_agg.csv'), row.names = F)
+write.csv(mhuber_a, file=here::here('/out/anovas_resilience/huber_evi','robust_mhuber_a.csv'), row.names = F)
+write.csv(mhuber_agg_a, file=here::here('/out/anovas_resilience/huber_evi','robust_mhuber_agg_a.csv'), row.names = F)
+write.csv(mhuber_b, file=here::here('/out/anovas_resilience/huber_evi','robust_mhuber_b.csv'), row.names = F)
+write.csv(mhuber_agg_b, file=here::here('/out/anovas_resilience/huber_evi','robust_mhuber_agg_b.csv'), row.names = F)
+
+write.csv(pairwise, file=here::here('/out/anovas_resilience/huber_evi','/robust_pairwise.csv'), row.names = F)
+write.csv(trimmedanovas, file=here::here('/out/anovas_resilience/huber_evi','trimmed_anovas.csv'), row.names = F)
+```

### 2018-11-16 man/ms_tables/t_huber_res_bai.Rmd
update

@@
     reference_docx: "../../man/template_tables_portrait.docx"
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(dplyr)
+library(flextable)
 ```
@@
 ```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg.csv"), header = TRUE)
+mraw <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
 
@@
   rename(variables = var)  %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
+  dplyr::select(c(-n, -Letter, -MonoLetter, -var_sorted)) %>% 
+  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., 3)))
+
  unite("mh", c("M.Huber", "ci"), sep = "\n")

### 2018-11-16 man/tables/t_robust_anovas.Rmd
update

---
output:  
  pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
library(dplyr)
```

**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the two factors drought events (in 2005 and 2012) and site.

```{r, echo=FALSE, include = FALSE}
### Table R1_evi
evi <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
names(evi) <- c('factores', 'F', 'p', 'variables')

dff_evi <- evi %>% gather(temp, numeros, F, p) %>% 
  unite(temp1, variables, temp) %>% 
  spread(temp1, numeros) %>% 
  mutate(factores = recode(factores, 
                           "disturb_year" = "Drought event", 
                           "disturb_year:site" = "Drought event:Site", 
                           "site" = "Site")) %>% 
  arrange(match(factores, c("Drought event", "Site", "Drought event:Site"))) %>% 
  mutate_at(c(2,4,6,8), round, 1) %>% 
  mutate_at(c(3,5,7,9), round, 3) %>% 
  mutate(variable = 'EVI')

bai <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
names(bai) <- c('factores', 'F', 'p', 'variables')

dff_bai <- bai %>% gather(temp, numeros, F, p) %>% 
  unite(temp1, variables, temp) %>% 
  spread(temp1, numeros) %>% 
  mutate(factores = recode(factores, 
                           "disturb_year" = "Drought event", 
                           "disturb_year:site" = "Drought event:Site", 
                           "site" = "Site")) %>% 
  arrange(match(factores, c("Drought event", "Site", "Drought event:Site"))) %>% 
  mutate_at(c(2,4,6,8), round, 1) %>% 
  mutate_at(c(3,5,7,9), round, 3) %>% 
  mutate(variable = 'BAI')

dff <- dff_evi %>% 
  bind_rows(dff_bai) %>% 
  dplyr::select(-rrs_F, -rrs_p) %>% 
  dplyr::select(variable, factores, rt_F, rt_p, rc_F, rc_p, rs_F, rs_p)
```

```{r, echo=FALSE}
# # https://haozhu233.github.io/kableExtra/use_kableExtra_with_formattable.html
# colnames(dff) <- c("variable", "factors", "F", "p", "F", "p", "F", "p")
# colnames(dff) <- c("variable", "factors", "rt\nF", "rt\np","rc\nF", "rc\np", "rs\nF", "rs\np")
# kable(dff, format = "latex") %>% add_header_above(c("", "", "Resistance" = 2, "Recovery" = 2, "Resilience" = 2))
#

# https://davidgohel.github.io/flextable/articles/examples.html

mt <- regulartable(
  data = dff, 
  col_keys = c("variable", "factores", "c3","rt_F", "rt_p", "c1", "rc_F", "rc_p","c2", "rs_F", "rs_p"),
)

mt <- theme_vanilla(mt)
mt <- autofit(mt)
mt <- empty_blanks(mt)
mt <- align(mt, j = 1,  align = "center")

mt <- merge_v(mt, ~variable)

mt <- set_header_labels(mt, variable = "", factores = "", rt_F = "Resistance", rt_p = "Resistance", rc_F = "Recovery", rc_p = "Recovery", rs_F = "Resilience", rs_p = "Resilience")

# # Add row of labels 
mt <- add_header(mt, variable = "Variable", factores = "Factor", rt_F = "F", rt_p = "p", rc_F = "F", rc_p = "p", rs_F = "F", rs_p = "p", top = FALSE)

# Format 
mt <- bold(mt, part = 'header') 
mt <- fontsize(mt, part = "header", size = 12)
mt <- fontsize(mt, part = "body", size = 11)
mt <- width(mt, j = 2, width = 2)

# mt <- align(mt, i = 1, j=2:3, align = "left")

mt <- italic(mt, i = ~ rt_p < 0.01, j = ~ rt_p)
mt <- italic(mt, i = ~ rc_p < 0.01, j = ~ rc_p)
mt <- italic(mt, i = ~ rs_p < 0.01, j = ~ rs_p)
mt

mt <- set_formatter(mt, 
                    rt_F = function(x) format(round(x, 1)),
                    rc_F = function(x) format(round(x, 1)),
                    rs_F = function(x) format(round(x, 1))
                    )

mt

```

### 2018-11-16 analysis/evi/explore_resilience.Rmd
update mhuber export data

@@
 
-```{r wd, echo=FALSE}
-#---------------------------------
-machine <- 'ajpelu'
-#machine <- 'ajpeluLap'
-di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro', sep='')
-#---------------------------------
-```
+# ```{r wd, echo=FALSE}
+# #---------------------------------
+# machine <- 'ajpelu'
+# #machine <- 'ajpeluLap'
+# di <- paste('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro', sep='')
+# #---------------------------------
+# ```
 
@@
 mhuber_agg <- mhuber %>% 
-  mutate(M.Huber = round(M.Huber, 4),
-         lower.ci = round(lower.ci, 4),
-         upper.ci = round(upper.ci, 4)) %>% 
-  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
-  mutate(ci = paste0('(', ci, ')')) %>% 
-  dplyr::select(-MonoLetter) %>% 
-  dplyr::select(var, disturb_year:ci, Letter) %>% as.data.frame()
+  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., 4))) %>% 
+  dplyr::select(var, disturb_year:Letter) %>% as.data.frame()
 
@@
 mhuber_agg_a <- mhuber_a %>% 
-  mutate(M.Huber = round(M.Huber, 4),
-         lower.ci = round(lower.ci, 4),
-         upper.ci = round(upper.ci, 4)) %>% 
-  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
-  mutate(ci = paste0('(', ci, ')')) %>% 
-  dplyr::select(-MonoLetter) %>% 
-  dplyr::select(var, disturb_year:ci, Letter) %>% as.data.frame()
+  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., 4))) %>% 
+  dplyr::select(var, disturb_year:Letter) %>% as.data.frame()
 
@@
 mhuber_agg_b <- mhuber_b %>% 
-  mutate(M.Huber = round(M.Huber, 4),
-         lower.ci = round(lower.ci, 4),
-         upper.ci = round(upper.ci, 4)) %>% 
-  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
-  mutate(ci = paste0('(', ci, ')')) %>% 
-  dplyr::select(-MonoLetter) %>% 
-  dplyr::select(var, site:ci, Letter) %>% as.data.frame()
+  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., 4))) %>% 
+  dplyr::select(var, site:Letter) %>% as.data.frame()
 

### 2018-11-16 man/ms_tables/t_huber_res_evi.Rmd
update table

@@
 ```{r, echo = FALSE, warning = FALSE}
-mraw <- df <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)
 
+# Number of digits 
+d <- 3
+
+mraw <- df <- read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)
 m <- mraw %>% 
-  filter(var != 'rrs') %>% 
   rename(variables = var) %>% 
-  dplyr::select(c(-n, -Letter)) %>% 
- unite("mh", c("M.Huber", "ci"), sep = "\n")
+  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
+  dplyr::select(c(-n, -Letter))
 
 # Datos agrupados para todos los sitios 
-mraw_a <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
-
+mraw_a <- read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
 m_a <- mraw_a %>% 
-  filter(var != 'rrs') %>% 
   rename(variables = var)  %>% 
+  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
   dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>% 
   mutate(site = "z_all")
@@
 # Datos agrupados para todos los sitios 
-mraw_b <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)
+mraw_b <- read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)
 
 m_b <- mraw_b %>% 
-  filter(var != 'rrs') %>% 
   rename(variables = var)  %>% 
   dplyr::select(c(-n, -Letter)) %>%  
-  unite("mh", c("M.Huber", "ci"), sep = "\n") %>%
+  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>%
   mutate(disturb_year = 2099)
@@
   bind_rows(m_b) %>% 
-  gather(temp, numeros, mh) %>% 
-  unite(temp1, c(variables, disturb_year), temp) %>% 
-  spread(temp1, numeros, fill = "")
+  unite("v", c("disturb_year","variables"), sep="")
 
-mm <- mm %>% 
-  dplyr::select(site, 
-                rt_2005_mh, rc_2005_mh, rs_2005_mh,
-                rt_2012_mh, rc_2012_mh, rs_2012_mh,
-                rt_2099_mh, rc_2099_mh, rs_2099_mh)
+mm_MH <- mm %>% 
+  dplyr::select(v, site, M.Huber) %>% 
+  mutate(M.Huber = as.character(format(M.Huber, digits = d))) %>% 
+  spread(v, M.Huber) %>% 
+  mutate(what = 'mh')
 
-colnames(mm) <- c(" \nsite", "2005\nRt", "2005\nRc", "2005\nRs", "2012\nRt", "2012\nRc", "2012\nRs", "Rt", "Rc", "Rs")
-pander(mm, keep.line.breaks = TRUE, split.table = Inf)
+mm_CI <- mm %>% 
+  dplyr::select(-M.Huber) %>% 
+  mutate_at(.vars = c("lower.ci","upper.ci"), funs(format(., digits = d))) %>%  
+  unite("ci", c("lower.ci","upper.ci"), sep=" - ") %>% 
+  mutate(ci = paste0('(', ci, ')')) %>% 
+  spread(v,ci) %>% mutate(what = 'ci')
+
+mi <- bind_rows(mm_MH, mm_CI) %>% arrange(site)
+
+mt <- regulartable(mi,
+                   col_keys = c("site","c1", "2005rt","2005rc","2005rs", "c2",
+                                "2012rt","2012rc","2012rs","c3","2099rt","2099rc", "2099rs"))
+
+mt <- autofit(mt)
+mt <- empty_blanks(mt)
+mt <- align(mt, align = "center")
+mt <- merge_v(mt, ~site)
+
+mt <- set_header_labels(mt, site = "Site", 
+                        '2005rt' = "Resistance", '2005rc' = "Recovery", '2005rs' = "Resilience",
+                        '2012rt' = "Resistance", '2012rc' = "Recovery", '2012rs' = "Resilience",
+                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience")
+
+# Add row of labels 
+mt <- add_header(mt, site="", '2005rt' = "", '2005rc' = "2005", '2005rs' = "",
+                        '2012rt' = "", '2012rc' = "2012", '2012rs' = "",
+                        '2099rt' = "", '2099rc' = "", '2099rs' = "")
+
+mt
+
 ```

### 2018-11-16 analysis/bai/analysis_resilience_bai.Rmd
change output figure resilience BAI
update bai hubers

@@
 library('rcompanion')
+library('here')
 ```
@@
 ```{r}
-machine <- 'ajpelu'
+# machine <- 'ajpelu'
 # machine <- 'ajpeluLap'
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
+# di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
 
 # sj 
-sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
+sj <- read.rwl(fname=here::here('/data_raw/dendro_ring/sn_sanjuan','sn_sanjuan.rwl'), format="tucson")
 
 # canar 
-ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
+ca <- read.rwl(fname=here::here('/data_raw/dendro_ring/sn_canar', 'sn_canar.rwl'), format="tucson")
 
 # Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-```
-
-```{r}
-source(paste0(di, 'script/R/rw_byTree.R'))
-source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/baiResilience.R'))
+compete <- read.csv(file=here::here('/data_raw', 'dendro_competence.csv'), header=TRUE, sep=',')
+```
+
+```{r}
+source(here::here('/script/R/', 'rw_byTree.R'))
+source(here::here('/script/R/', 'bai_piovesan.R'))
+source(here::here('/script/R/', 'baiResilience.R'))
 ```
@@
 
-write.csv(aux_coefs, file=paste0(di, '/out/correla_resilience/correla_window_size.csv'), row.names = F)
+write.csv(aux_coefs, file=here::here('/out/correla_resilience', 'correla_window_size.csv'), row.names = F)
 ```
@@
 # Export csv 
-write.csv(g, file=paste0(di, 'data/resilience/crecimientos_drought.csv'), row.names = FALSE)
+write.csv(g, file=here::here('/data/resilience', 'crecimientos_drought.csv'), row.names = FALSE)
 
@@
 ggsave(plot=pgrowht, width=8, height = 4,
-       filename=paste0(di, 'out/fig/resilience/bai_events.pdf')) 
+       filename=here::here('/out/mean_resilience/','bai_events.pdf')) 
 ```
@@
 # Export csv 
-write.csv(re, file=paste0(di, 'data/resilience/resilience_bai.csv'), row.names = FALSE)
+write.csv(re, file=here::here('data/resilience/','resilience_bai.csv'), row.names = FALSE)
 ```
@@
 write.csv(normtestA, 
-          file=paste0(di, '/out/anovas_resilience/bai/normo_disturb_year.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/bai', 'normo_disturb_year.csv'), row.names = F)
 ```
@@
 write.csv(normtestB, 
-          file=paste0(di, '/out/anovas_resilience/bai/normo_site.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/bai','normo_site.csv'), row.names = F)
 ```
@@
 write.csv(normtestAB, 
-          file=paste0(di, '/out/anovas_resilience/bai/normo_disturb_year_site.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/bai', 'normo_disturb_year_site.csv'), row.names = F)
 ```
@@
 write.csv(homo, 
-          file=paste0(di, '/out/anovas_resilience/bai/homocedasticidad.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/bai','homocedasticidad.csv'), row.names = F)
 ```
@@
 write.csv(homo_log, 
-          file=paste0(di, '/out/anovas_resilience/bai/homocedasticidad_log.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/bai/','homocedasticidad_log.csv'), row.names = F)
 ```
@@
 ```{r, echo=FALSE}
-pdf(paste0(di, '/out/mean_resilience/bai/interaction_plotsSE.pdf'), width=9, height = 9)
+pdf(here::here('/out/mean_resilience/bai','interaction_plotsSE.pdf'), width=9, height = 9)
 grid.arrange(plot_mdSE, plot_msSE, plot_mdsSE, ncol=3)
@@
 
-pdf(paste0(di, '/out/mean_resilience/bai/interaction_plotsCI.pdf'), width=9, height = 9)
+pdf(here::here('/out/mean_resilience/bai','interaction_plotsCI.pdf'), width=9, height = 9)
 grid.arrange(plot_mdCI, plot_msCI, plot_mdsCI, ncol=3)
@@
 
-write.csv(aovas_coeff, file=paste0(di, '/out/anovas_resilience/bai/anovas_statistics.csv'), row.names = F)
+write.csv(aovas_coeff, file=here::here('/out/anovas_resilience/bai','anovas_statistics.csv'), row.names = F)
 
@@
 write.csv(aovas_model_summary, 
-          file=paste0(di, '/out/anovas_resilience/bai/anovas_summary_modelos.csv'), row.names = F)
+          file=here::here('/out/anovas_resilience/bai','anovas_summary_modelos.csv'), row.names = F)
 
@@
 mhuber_agg <- mhuber %>% 
-  mutate(M.Huber = round(M.Huber, 4),
-         lower.ci = round(lower.ci, 4),
-         upper.ci = round(upper.ci, 4)) %>% 
-  unite_("ci", c("lower.ci", "upper.ci"), sep=",") %>% 
-  mutate(ci = paste0('(', ci, ')')) %>% 
-  dplyr::select(-MonoLetter) %>% 
-  dplyr::select(var, disturb_year:ci, Letter) %>% as.data.frame()
+  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., 4))) %>%
+  dplyr::select(var, disturb_year:Letter) %>% as.data.frame()
 
@@
 
-pdf(paste0(di, 'out/resilience_robust/bai/robust_plot_bai_drought.pdf'), width=9, height = 9)
+pdf(here::here('out/resilience_robust/bai','robust_plot_bai_drought.pdf'), width=9, height = 9)
 robust_plot_bai_drought 
@@
 
-pdf(paste0(di, 'out/resilience_robust/bai/robust_plot_bai_site.pdf'), width=9, height = 9)
+pdf(here::here('out/resilience_robust/bai','robust_plot_bai_site.pdf'), width=9, height = 9)
 robust_plot_bai_site 
@@
 # Export data 
-write.csv(mhuber, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber.csv'), row.names = F)
-write.csv(mhuber_agg, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg.csv'), row.names = F)
-write.csv(mhuber_a, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_a.csv'), row.names = F)
-write.csv(mhuber_agg_a, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg_a.csv'), row.names = F)
-write.csv(mhuber_b, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_b.csv'), row.names = F)
-write.csv(mhuber_agg_b, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_mhuber_agg_b.csv'), row.names = F)
-
-write.csv(pairwise, file=paste0(di, '/out/anovas_resilience/huber_bai/robust_pairwise.csv'), row.names = F)
-
-write.csv(trimmedanovas, file=paste0(di, '/out/anovas_resilience/huber_bai/trimmed_anovas.csv'), row.names = F)
+write.csv(mhuber, file=here::here('/out/anovas_resilience/huber_bai/','robust_mhuber.csv'), row.names = F)
+write.csv(mhuber_agg, file=here::here('/out/anovas_resilience/huber_bai/','robust_mhuber_agg.csv'), row.names = F)
+write.csv(mhuber_a, file=here::here('/out/anovas_resilience/huber_bai/','robust_mhuber_a.csv'), row.names = F)
+write.csv(mhuber_agg_a, file=here::here('/out/anovas_resilience/huber_bai/','robust_mhuber_agg_a.csv'), row.names = F)
+write.csv(mhuber_b, file=here::here('/out/anovas_resilience/huber_bai/','robust_mhuber_b.csv'), row.names = F)
+write.csv(mhuber_agg_b, file=here::here('/out/anovas_resilience/huber_bai/','robust_mhuber_agg_b.csv'), row.names = F)
+
+write.csv(pairwise, file=here::here('/out/anovas_resilience/huber_bai/','robust_pairwise.csv'), row.names = F)
+
+write.csv(trimmedanovas, file=here::here('/out/anovas_resilience/huber_bai/','trimmed_anovas.csv'), row.names = F)
 ```

### 2018-11-16 man/tables/t_huber_res_bai.Rmd
update table and move

---
output:  
  word_document:
    reference_docx: "../../man/template_tables_portrait.docx"
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
library(dplyr)
library(flextable)
library(officer)
```

**Table S2.** Robust measures of central tendency of resilience indices for tree-growth (BAI) groupped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included in parentheses.

```{r, echo = FALSE, warning = FALSE}

# Number of digits 
d <- 3

mraw <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)

m <- mraw %>% 
  rename(variables = var) %>% 
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
  dplyr::select(c(-n, -MonoLetter, -Letter, -var_sorted))

# Datos agrupados para todos los sitios 
mraw_a <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)

m_a <- mraw_a %>% 
  rename(variables = var)  %>% 
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
  dplyr::select(-n, -Letter) %>%  
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
mraw_b <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)

m_b <- mraw_b %>% 
  rename(variables = var)  %>% 
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
  dplyr::select(-n, -Letter) %>% 
  mutate(disturb_year = 2099)

mm <- m %>% 
  bind_rows(m_a) %>%
  bind_rows(m_b) %>% 
  unite("v", c("disturb_year","variables"), sep="")

mm_MH <- mm %>% 
  dplyr::select(v, site, M.Huber) %>% 
  mutate(M.Huber = as.character(format(M.Huber, digits = d))) %>% 
  spread(v, M.Huber) %>% 
  mutate(what = 'mh')

mm_CI <- mm %>% 
  dplyr::select(-M.Huber) %>% 
  mutate_at(.vars = c("lower.ci","upper.ci"), funs(format(., digits = d))) %>%  
  unite("ci", c("lower.ci","upper.ci"), sep=" - ") %>% 
  mutate(ci = paste0('(', ci, ')')) %>% 
  spread(v,ci) %>% mutate(what = 'ci')

mi <- bind_rows(mm_MH, mm_CI) %>% arrange(site) %>% 
  mutate(site = case_when(
    site == 'z_all' ~ "All",
    site == 'caH' ~ 'CA-High', 
    site == 'caL' ~ 'CA-Low',
    site == 'SJ' ~ 'SJ'))

# CREATE TABLE

# Define borders
big_border = fp_border(width = 2)
std_border = fp_border(width = 1)

mt <- regulartable(mi, col_keys = c("site","c1", "2005rt","2005rc","2005rs", "c2",
                                "2012rt","2012rc","2012rs","c3","2099rt","2099rc", "2099rs"))
mt <- mt %>% 
  # Header 
  set_header_labels(site = "Sites", 
                        '2005rt' = "Resistance", '2005rc' = "Recovery", '2005rs' = "Resilience",
                        '2012rt' = "Resistance", '2012rc' = "Recovery", '2012rs' = "Resilience",
                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience") %>% 
  add_header(site="", '2005rt' = "2005", '2005rc' = "2005", '2005rs' = "2005",'2012rt' = "2012", '2012rc' = "2012", '2012rs' = "2012",
                        '2099rt' = "", '2099rc' = "", '2099rs' = "", top = TRUE) %>% 
  merge_at(i = 1, j =c(3:5) , part = "header") %>% 
  merge_at(i = 1, j =c(7:9) , part = "header") %>% 
  align(align = "center", part = "header") %>% 
  bold(part = 'header') %>% 
  fontsize(part = "header", size = 12) %>%

  # General format 
  fontsize(part = "body", size = 11) %>% 
  autofit() %>% 
  empty_blanks() %>% 
  align(align = "center") %>%

  # Borders 
  border_remove() %>% 
  hline(i = 1:2, border = big_border, part = "header") %>% 
  hline(i = 6, border = big_border, part = "body") %>% 
  vline(j = c(2,6,10), border = big_border, part = "all") %>% 
  hline(i = c(2,4), border = std_border, part = "body") %>%

  # First column 
  merge_v(~site) %>% 
  italic(j=1, part = "body")

mt 
```

### 2018-11-16 man/tables/t_huber_res_evi.Rmd
update table and move

--- 2018-11-16 man/ms_tables/t_huber_res_evi.Rmd
@@
 library(flextable)
+library('officer')
 ```
@@
 
-mi <- bind_rows(mm_MH, mm_CI) %>% arrange(site)
+mi <- bind_rows(mm_MH, mm_CI) %>% arrange(site) %>% 
+  mutate(site = case_when(
+    site == 'z_all' ~ "All",
+    site == 'N' ~ "Northern slope", 
+    site == 'S' ~ "Southern slope"))
 
-mt <- regulartable(mi,
-                   col_keys = c("site","c1", "2005rt","2005rc","2005rs", "c2",
+# CREATE TABLE
+
+# Define borders
+big_border = fp_border(width = 2)
+std_border = fp_border(width = 1)
+
+mt <- regulartable(mi, col_keys = c("site","c1", "2005rt","2005rc","2005rs", "c2",
                                 "2012rt","2012rc","2012rs","c3","2099rt","2099rc", "2099rs"))
-
-mt <- autofit(mt)
-mt <- empty_blanks(mt)
-mt <- align(mt, align = "center")
-mt <- merge_v(mt, ~site)
-
-mt <- set_header_labels(mt, site = "Site", 
+mt <- mt %>% 
+  # Header 
+  set_header_labels(site = "Oak populations", 
                         '2005rt' = "Resistance", '2005rc' = "Recovery", '2005rs' = "Resilience",
                         '2012rt' = "Resistance", '2012rc' = "Recovery", '2012rs' = "Resilience",
-                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience")
+                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience") %>% 
+  add_header(site="", '2005rt' = "2005", '2005rc' = "2005", '2005rs' = "2005",'2012rt' = "2012", '2012rc' = "2012", '2012rs' = "2012",
+                        '2099rt' = "", '2099rc' = "", '2099rs' = "", top = TRUE) %>% 
+  merge_at(i = 1, j =c(3:5) , part = "header") %>% 
+  merge_at(i = 1, j =c(7:9) , part = "header") %>% 
+  align(align = "center", part = "header") %>% 
+  bold(part = 'header') %>% 
+  fontsize(part = "header", size = 12) %>%
 
-# Add row of labels 
-mt <- add_header(mt, site="", '2005rt' = "", '2005rc' = "2005", '2005rs' = "",
-                        '2012rt' = "", '2012rc' = "2012", '2012rs' = "",
-                        '2099rt' = "", '2099rc' = "", '2099rs' = "")
+  # General format 
+  fontsize(part = "body", size = 11) %>% 
+  autofit() %>% 
+  empty_blanks() %>% 
+  align(align = "center") %>%
+
+  # Borders 
+  border_remove() %>% 
+  hline(i = 1:2, border = big_border, part = "header") %>% 
+  hline(i = 4, border = big_border, part = "body") %>% 
+  vline(j = c(2,6,10), border = big_border, part = "all") %>% 
+  hline(i = 2, border = std_border, part = "body") %>%
+
+  # First column 
+  merge_v(~site) %>% 
+  italic(j=1, part = "body")
 
 mt 
-
 ```

### 2018-11-16 man/tables/t_dendro_summary.Rmd
update tables

@@
   word_document:
-    reference_docx: "../../man/template_tables_portrait.docx"
+    fig_caption: yes
+    highlight: null
+    reference_docx: ../../templates/template_tables_portrait.docx
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecology.csl
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r, message=FALSE, echo=FALSE}
+```{r}
 library(tidyverse)
@@
 
-```{r, echo=FALSE}
+```{r}
 dendro_summ <- read.csv(here::here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
@@
   unite(length_year, c(length_year,length_year5), sep=" (") %>%
+  mutate_at(.vars = c("rw_mean","rw_std"), funs(formatC(., digits = 3, format = 'f'))) %>% 
   unite(RW, c(rw_mean,rw_std), sep=" (") %>%
@@
 
-colnames(t2) <- c("Site", "First year", "Last year", "Length (years)",  
-                   "# trees", '# cores', "RW (mm)", "MS", "AR(1)", "Rbt", "EPS")
+ta2 <- regulartable(t2, col_keys =c("name", "first_year", "last_year", "n_trees", "n_radii", "RW", "ms2",
+                                   "ar1","rbar.wt","eps"))
+ta2 <- ta2 %>% set_header_labels(name = "Site", first_year = "First year", last_year = "Last year",
+                                 length_year = "Length (years)", n_trees ="# trees", n_radii = '# cores',
+                                 RW ="RW (mm)", ms2="MS", ar1="AR(1)", rbar.wt = "Rbt", eps = "EPS") %>%
+  align(align = "center", part = "header") %>% 
+  bold(part = 'header') %>% 
+  fontsize(part = "header", size = 10) %>% 
+  fontsize(part = "body", size = 9) %>% 
+  autofit(add_w = 0, add_h = 0) %>% 
+  align(align = "center")
 
-pander(t2,
-       split.table = Inf)
+ta2
 

### 2018-11-16 man/tables/t_droughts.Rmd
update tables

--- 2018-11-09 man/tables/table_droughts.Rmd
@@
     fig_caption: yes
-    fig_height: 6
-    fig_width: 6
     highlight: null
@@
 csl: ../../refs/ecology.csl
-editor_options: 
-  chunk_output_type: console
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
@@
 library("dplyr")
+library("flextable")
 
-library("grid")
-library("gridExtra")
-library("patchwork")
-library("ggcorrplot")
-library("stringr")
-library("pander")
-library("lemon")
-library("viridis")
-library("ggthemes")
-```
-
-```{r, echo=FALSE}
-show_text <- FALSE
 ```

### 2018-11-16 man/tables/t_huber_res_bai.Rmd
update tables

@@
     reference_docx: "../../man/template_tables_portrait.docx"
-editor_options: 
-  chunk_output_type: console
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
@@
 
-```{r, echo = FALSE, warning = FALSE}
-
+```{r}
 # Number of digits 
@@
   bold(part = 'header') %>% 
-  fontsize(part = "header", size = 12) %>%
+  fontsize(part = "header", size = 9) %>%
 
   # General format 
-  fontsize(part = "body", size = 11) %>% 
-  autofit() %>% 
+  fontsize(part = "body", size = 8) %>% 
+  autofit(add_w = 0, add_h = 0) %>% 
   empty_blanks() %>% 

### 2018-11-16 man/tables/t_huber_res_evi.Rmd
update tables

@@
 
-```{r, echo = FALSE, warning = FALSE}
+```{r}
 

### 2018-11-16 man/tables/t_robust_anovas.Rmd
update tables

@@
 output:
-  pdf_document
+  word_document:
+    fig_caption: yes
+    highlight: null
+    reference_docx: ../../templates/template_tables_portrait.docx
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecology.csl
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(dplyr)
+library(flextable)
+library(officer)
 ```
@@
 
-```{r, echo=FALSE, include = FALSE}
+```{r}
 ### Table R1_evi
@@
 
-```{r, echo=FALSE}
-# # https://haozhu233.github.io/kableExtra/use_kableExtra_with_formattable.html
-# colnames(dff) <- c("variable", "factors", "F", "p", "F", "p", "F", "p")
-# colnames(dff) <- c("variable", "factors", "rt\nF", "rt\np","rc\nF", "rc\np", "rs\nF", "rs\np")
-# kable(dff, format = "latex") %>% add_header_above(c("", "", "Resistance" = 2, "Recovery" = 2, "Resilience" = 2))
-#
-
-# https://davidgohel.github.io/flextable/articles/examples.html
-
+```{r}
 mt <- regulartable(
@@
 
-mt <- theme_vanilla(mt)
-mt <- autofit(mt)
-mt <- empty_blanks(mt)
-mt <- align(mt, j = 1,  align = "center")
+mti <- mt %>% 
+  autofit() %>% 
+  empty_blanks() %>% 
+  align(j = 1,  align = "center") %>% 
+  merge_v(~variable) %>% 
+  set_header_labels(variable = "", factores = "", rt_F = "Resistance", rt_p = "Resistance", rc_F = "Recovery", rc_p = "Recovery", rs_F = "Resilience", rs_p = "Resilience") %>% 
+  add_header(variable = "Variable", factores = "Factor", rt_F = "F", rt_p = "p", rc_F = "F", rc_p = "p", rs_F = "F", rs_p = "p", top = FALSE) %>% 
+  bold(part = 'header') %>% 
+  fontsize(part = "header", size = 12) %>%
+  fontsize(part = "body", size = 11) %>% 
+  width(j = 2, width = 2) %>% 
+  italic(i = ~ rt_p < 0.01, j = ~ rt_p) %>% 
+  italic(i = ~ rc_p < 0.01, j = ~ rc_p) %>% 
+  italic(i = ~ rs_p < 0.01, j = ~ rs_p)
+mti
 
-mt <- merge_v(mt, ~variable)
-
-mt <- set_header_labels(mt, variable = "", factores = "", rt_F = "Resistance", rt_p = "Resistance", rc_F = "Recovery", rc_p = "Recovery", rs_F = "Resilience", rs_p = "Resilience")
-
-# # Add row of labels 
-mt <- add_header(mt, variable = "Variable", factores = "Factor", rt_F = "F", rt_p = "p", rc_F = "F", rc_p = "p", rs_F = "F", rs_p = "p", top = FALSE)
-
-# Format 
-mt <- bold(mt, part = 'header') 
-mt <- fontsize(mt, part = "header", size = 12)
-mt <- fontsize(mt, part = "body", size = 11)
-mt <- width(mt, j = 2, width = 2)
-
-# mt <- align(mt, i = 1, j=2:3, align = "left")
-
-mt <- italic(mt, i = ~ rt_p < 0.01, j = ~ rt_p)
-mt <- italic(mt, i = ~ rc_p < 0.01, j = ~ rc_p)
-mt <- italic(mt, i = ~ rs_p < 0.01, j = ~ rs_p)
-mt
-
-mt <- set_formatter(mt, 
-                    rt_F = function(x) format(round(x, 1)),
-                    rc_F = function(x) format(round(x, 1)),
-                    rs_F = function(x) format(round(x, 1))
-                    )
-
-mt
+# mt <- set_formatter(mt, 
+#                     rt_F = function(x) format(round(x, 1)),
+#                     rc_F = function(x) format(round(x, 1)),
+#                     rs_F = function(x) format(round(x, 1))
+#                     ) 
+# 
+# mt
 

### 2018-11-16 man/tables/t_sampled_plots.Rmd
update tables

--- 2018-11-12 man/tables/t_sampled_plots.Rmd
@@
     fig_caption: yes
-    fig_height: 6
-    fig_width: 7.5
     highlight: null
-    reference_docx: ../template_tables_portrait.docx
+    reference_docx: ../../templates/template_tables_portrait.docx
 bibliography: ../../refs/references.bib
@@
 
-```{r, echo=FALSE}
+```{r}
 topo <- read.csv(here::here("/data/sites_summary", "site_general.csv"), header = TRUE)
@@
             funs(formatC(., digits = 2,format='f'))) %>%  
+  mutate_at(.vars = c("prec"), funs(format(., digits = 0))) %>% 
   # mutate_at(c(10:19, 22:23), round, 1) %>% 
@@
 
-colnames (t1) <- c("Site", "Lat (&deg;)", "Long (&deg;)", "Elevation (m)", "Slope (&deg;)", 
-                   "Prec. (mm)", "Temp (&deg; C)",
-                   "# trees (# cores)", "Dbh (cm)", "Height (m)", "Age (years)", 
-                   "Dbh all (cm)", "Height all (m)", "BA (m2/ha)", "Density (trees/ha)", "SRD")
+ta <- regulartable(t1, col_keys =c("name", "lat_m", "long_m", "elev", "slope", "prec", "temp",
+                                   "n_trees","dn","h","age","dn_all","h_all","ba","stand_density","srd"))
 
-pander(t1, split.table = Inf)
+ta <- ta %>% set_header_labels(name = "Site", lat_m = "Lat (\u00B0)", long_m = "Long (\u00B0)", 
+                               elev = "Elevation\n(m)", slope = "Slope (\u00B0)",  prec = "Prec.\n(mm)",
+                               temp = "Temp.\n(\u00B0 C)", n_trees = "# trees\n(# cores)", dn = "Dbh (cm)",
+                               h = "Height\n(m)", age = "Age\n(years)", dn_all = "Dbh all\n(cm)", 
+                               h_all = "Height all\n(m)", ba = "BA\n(m2/ha)", 
+                               stand_density = "Density\n(trees/ha)", srd = "SRD") %>% 
+  add_header(name="", lat_m ="", long_m="", elev="", slope ="", prec = "", temp = "",
+             n_trees = "Cored trees", dn = "Cored trees", h = "Cored trees", age = "Cored trees", 
+             dn_all = "Competence", h_all = "Competence", ba = "Competence", 
+             stand_density = "Competence", srd = "Competence", top = TRUE) %>% 
+  merge_at(i = 1, j =c(8:11) , part = "header") %>% 
+  merge_at(i = 1, j =c(12:16) , part = "header") %>% 
+  align(align = "center", part = "header") %>% 
+  bold(part = 'header') %>% 
+  fontsize(i= 1, part = "header", size = 8) %>% 
+  fontsize(i= 2, part = "header", size = 8) %>% 
+  italic(i=1, part = "header") %>% 
+    # General format 
+  fontsize(part = "body", size = 7) %>% 
+  autofit(add_w = 0) %>% 
+  align(align = "center") %>% 
+  width(j = c(2,3), width = 0.35) %>% 
+  width(j = c(6,7,8), width = 0.6)
+
+ta
 

### 2018-11-19 man/tables/t_dendro_summary.Rmd
update tables

@@
   word_document:
-    fig_caption: yes
-    highlight: null
     reference_docx: ../../templates/template_tables_portrait.docx

### 2018-11-19 man/tables/t_droughts.Rmd
update tables

@@
   word_document:
-    fig_caption: yes
-    highlight: null
-    reference_docx: ../../templates/template_pagebreak.docx
+    reference_docx: ../../templates/template_tables_portrait.docx
 bibliography: ../../refs/references.bib
@@
 
-colnames(dys) <- c("Duration \n(months)", "Intensity", "Severity", "Lowest SPEI", "Months", "Year")
-
-pander(dys)
+td <- regulartable(dys, col_keys = names(dys))
+td <- td %>% 
+  set_header_labels(d_duration = "Duration \n(months)", d_intensity = "Intensity", 
+                   d_severity = "Severity", lowest_spei = "Lowest \nSPEI", rangeDate = "Months", year = "Year") %>% 
+  align(align = "center", part ="all") %>% 
+  bold(part = 'header') %>% 
+  fontsize(part = "header", size = 10) %>%
+  fontsize(part = "body", size = 9) 
+td
 ```

### 2018-11-19 man/tables/t_robust_anovas.Rmd
update tables

@@
   bold(part = 'header') %>% 
-  fontsize(part = "header", size = 12) %>%
-  fontsize(part = "body", size = 11) %>% 
+  fontsize(part = "header", size = 10) %>%
+  fontsize(part = "body", size = 9) %>% 
   width(j = 2, width = 2) %>% 
@@
   italic(i = ~ rc_p < 0.01, j = ~ rc_p) %>% 
-  italic(i = ~ rs_p < 0.01, j = ~ rs_p)
+  italic(i = ~ rs_p < 0.01, j = ~ rs_p) %>% 
+  merge_at(i = 1, j=4:5, part="header") %>% 
+  merge_at(i = 1, j=7:8, part="header") %>% 
+  merge_at(i = 1, j=10:11, part="header")
+
 mti
-
 # mt <- set_formatter(mt, 

### 2018-11-19 man/tables/t_sampled_plots.Rmd
update tables

@@
 library(flextable)
+library(officer)
 ```
@@
 
+std_border = fp_border(width = 1)
+
 ta <- regulartable(t1, col_keys =c("name", "lat_m", "long_m", "elev", "slope", "prec", "temp",
@@
     # General format 
-  fontsize(part = "body", size = 7) %>% 
+  fontsize(part = "body", size = 8) %>% 
   autofit(add_w = 0) %>% 
@@
 
-ta
+ta %>% 
+   vline(j = c(7, 11), border = std_border, part="all")
 

### 2018-11-19 man/figures/0_figures.Rmd
modify drought severity to include months
add new figure
update

@@
 output:
-  word_document:
-    fig_caption: yes
-    highlight: null
-    reference_docx: ../../templates/template_pagebreak.docx
+  pdf_document
+
 bibliography: ../../refs/references.bib
@@
 ---
+
+```{r eval = FALSE}
+  # word_document:
+  #   fig_caption: yes
+  #   highlight: null
+  #   reference_docx: ../../templates/template_pagebreak.docx
+```
 
@@
 ```
+
+#####
+
+```{r child = 'browning.Rmd'}
+```
+
+#####
+
+```{r child = 'bai_cronos.Rmd'}
+```
+#####
+
+```{r child = 'growth_changes.Rmd'}
+```
+
+#####
+
+```{r child = 'resilience_long_term.Rmd'}
+```
+
+#####
+
+```{r child = 'resilience_metrics.Rmd'}
+```
+
+#####
+
+```{r child = 'ring_climate.Rmd'}
+```
+
+#####
+
+```{r child = 'gc_perc_trees.Rmd'}
+```
+
+#####
+
+```{r child = 'cumulative_prec.Rmd'}
+```
+
+#####
+
+```{r child = 'plot_spei12.Rmd'}
+```
+
+#####
+
+#####
+
+```{r child = 'gc_perc_trees.Rmd'}
+```
+
+#####
+
+```{r child = 'residual_chronos.Rmd'}
+```
+
+#####
+
+```{r child = 'cumulative_prec.Rmd'}
+```

### 2018-11-19 man/figures/bai_cronos.Rmd
update

--- 2018-10-16 man/ms_figures.Rmd
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-- [] Figure 1. Location map
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf
-
-- [] Figure 2. Plot de anomalias
-- [] Figure 3. Resilience metrics
-- [] Figure 4. BAI chronologies
-
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in the Iberian Peninsula (a) and in Sierra Nevada mountain range (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment [@IECA2009].
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
 ##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
-```
-
-```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
-```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
-```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```
-
+**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
+pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
+Shading areas correspond to standard error of the mean.
+Number of series are displayed in the upper plot.
+We only show years replicated with # series > 5.
+Linear trend since 1975 is shown for southern high-elevation site (CA-High) .AREGLAR ESTO
 ##### 
@@
 
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas correspond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show years replicated with # series > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r r2caH`).
-
 ```{r, warning=FALSE, out.extra='angle=90'}
@@
 ```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-11-19 man/ms_figures.Rmd
last update
update

@@
 - [] Figure 4. BAI chronologies
+
+- [] Figure 5. Residual chronologies --> Supplementary
 
@@
 res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure 4.** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
 ```

### 2018-11-23 man/figures/plot_spei12.Rmd
update figures std anomalies and EVI profile

@@
   word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 6
-    highlight: null
     reference_docx: ../../templates/template_pagebreak.docx

### 2018-11-23 man/figures/profile_EVI.Rmd
update figures std anomalies and EVI profile

--- 2018-11-19 man/ms_figures.Rmd
@@
 output:
-  word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 6
-    highlight: null
-    reference_docx: template_ms_new_page.docx
-bibliography: ../refs/references.bib
-csl: ../refs/ecology.csl
+  word_document
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
-library("ggrepel")
 library("scales")
-
-library("grid")
-library("gridExtra")
-library("patchwork")
-library("ggcorrplot")
-library("stringr")
-library("pander")
-library("lemon")
 ```
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Lista de figuras
-
-- [] Figure 1. Location map
-
-    - [] Ver https://github.com/ajpelu/qpyr_dendro/blob/master/out/fig/geomap/mapa_situacion.pdf
-
-- [] Figure 2. Plot de anomalias
-- [] Figure 3. Resilience metrics
-- [] Figure 4. BAI chronologies
-
-- [] Figure 5. Residual chronologies --> Supplementary
-
-")
-```
-
-**Figure 1**.
-Distribution of *Quercus pyrenaica* forests in the Iberian Peninsula (a) and in Sierra Nevada mountain range (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
-Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment [@IECA2009].
-
-![](../out/fig/geomap/mapa_situacion.jpg){height=600px}
-
-##### 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
-
-```{r, eval=FALSE}
+```{r data, eval=FALSE}
 iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
@@
 
+```
+
+```{r plot_profile,  }
 # colref <- '#455883'
@@
   scale_color_manual(values=c(col2005, col2012, colref))
-
-# profile_compara 
 ```
 
-```{r}
-# From explore_anomalies.Rmd 
-# Read data 
-anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')
-
-anomalias_evimean <- anomalias_evimean %>% 
-  mutate(
-    clu_pop = as.factor(case_when(
-      pop == 1 ~ "Camarate",
-      pop %in% c(2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out')),
-    clu_pop2 = as.factor(case_when(
-      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
-      pop %in% c(6,7,8) ~ 'Southern slope',
-      pop == 9 ~ 'out'))) %>% 
-  filter(clu_pop != 'out')
-
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
-# Standardized Anomalies 
-avg_sa_clu <- anomalias_evimean %>% 
-  group_by(clu_pop2, y) %>% 
-  dplyr::summarise(mean=mean(sa, na.rm=T),
-            sd = sd(sa, na.rm=T),
-            se = sd/sqrt(length(sa))) %>% 
- mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
-
-myylab <- 'EVI Standardized Anomaly'
-
-plot_sa_clu  <- avg_sa_clu %>%  
-  ggplot(aes(x=y, y=mean, fill=signo)) + 
-  geom_bar(stat = "identity") + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour=signo), width=.5) +
-  facet_wrap(~clu_pop2, nrow = 2) + 
-  scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
-  theme_bw() +
-  theme(text = element_text(size=16),
-        legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
+```{r out_plot, fig.width= 4, dpi = 300, out.width = 0.8}
+profile_compara 
 ```
-
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
-```
-
-```{r, fig.height=7, eval=FALSE}
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
-```
-
-```{r, fig.height=5}
-plot_sa_clu 
-```
-
-##### 
-**Figure 3.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
-
-```{r, eval=FALSE, echo=FALSE}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-pd <- position_dodge(.2)
-line_lwd <- .8
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site, group = site, 
-                   fill = site, linetype=site,
-                   shape = site, 
-                   ymin = lower.ci, ymax = upper.ci)
-aes_site <- aes(x = site, y = M.Huber, 
-                color = disturb_year, group=disturb_year, 
-                fill=disturb_year, linetype=disturb_year,
-                shape = disturb_year, 
-                ymin = lower.ci, ymax = upper.ci)
-
-theme_res <- list(
-  geom_errorbar(width=.2, position=pd, lwd=line_lwd), 
-  geom_line(position = pd, lwd=line_lwd),
-  geom_point(size = 2.5, position = pd), 
-  facet_wrap(
-    ~var_sorted, nrow = 1, scales="free_y",
-    labeller = as_labeller(c(
-      "0_rt" = "Resistance",
-      "1_rc" = "Recovery",
-      "2_rs" = "Resilience"))),
-  geom_text_repel(aes(label=Letter), 
-                  nudge_x = - 0.5,
-                  nudge_y = 0.01,
-                  direction="both", 
-                  size = 3.5, show.legend = FALSE,
-                  segment.size = 0.5, 
-                  segment.alpha = 0.5),
-  theme_bw(), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),  
-  ylab("Response value"), 
-  xlab("")
-)
-
-theme_evi <-  list(
-  scale_color_manual(values = c("black", "blue")),
-  scale_shape_manual(values=c(19, 22)),
-  scale_fill_manual(values = c("black", "blue")),
-  scale_linetype_manual(values = c("solid", "solid"))
-)
-
-theme_bai <- list(
-  scale_color_manual(values=c('black','blue','blue')), 
-  scale_shape_manual(values=c(19, 22, 15)),
-  scale_fill_manual(values=c('black','white','blue')),  
-  scale_linetype_manual(values=c("solid", "solid", 'solid'))
-)
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# EVI_drought 
-res_evi_drought <-
-  ggplot(mhuber_filter, aes_disturb) + 
-  theme_res + theme_evi + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_drought 
-res_bai_drought <-
-  ggplot(mhuberbai_filter, aes_disturb) + 
-  theme_res + theme_bai + theme(legend.position='bottom') + 
-  xlab('Drought event') + 
-  ggtitle("b) BAI")
-
-```
-
-```{r, eval=FALSE, fig.height=7}
-res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
-```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-### Figure 4
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-```
-
-##### 
-```{r, warning=FALSE}
-cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
-
-cro_sites <- cro_sites %>% mutate(site = 
-  case_when(
-    site == "CA_Low" ~ "CA-Low", 
-    site == "CA_High" ~ "CA-High",
-    site == "SJ" ~ "SJ",
-  )
-)
-
-# To get p.value of the model 
-# https://stackoverflow.com/questions/5587676/pull-out-p-values-and-r-squared-from-a-linear-regression
-lmp <- function(modelobject) {
-    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
-    f <- summary(modelobject)$fstatistic
-    p <- pf(f[1],f[2],f[3],lower.tail=F)
-    attributes(p) <- NULL
-    return(p)
-}
-
-# Lm CA-High
-dfCaH <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-High')
-
-mCaH <- lm(bai_mean ~ year, dfCaH)
-summary(mCaH)
-r2caH <- format(summary(mCaH)$r.squared, digits = 3)
-pCaH <- lmp(mCaH)
-
-# Lm CA-Low 
-dfCaL <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'CA-Low')
-
-mCaL <- lm(bai_mean ~ year, dfCaL)
-summary(mCaL)
-r2caL <- format(summary(mCaL)$r.squared, digits = 3)
-pCaL <- lmp(mCaL)
-
-# Lm SJ
-dfSJ <- cro_sites %>% 
-  filter(year >= 1975) %>% 
-  filter(site == 'SJ')
-
-mSJ <- lm(bai_mean ~ year, dfSJ)
-summary(mSJ)
-r2SJ <- format(summary(mSJ)$r.squared, digits = 3)
-pSJ <- lmp(mSJ)
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas correspond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show years replicated with # series > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) ($r_2$ = `r r2caH`).
-
-```{r, warning=FALSE, out.extra='angle=90'}
-
-# cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-Low'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'SJ'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE)
-
-  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
-samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
-  theme_bw() + ylab('# trees') + 
-  geom_line() + 
-  theme(panel.grid = element_blank(),
-        legend.position = "none", 
-        axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
-        # axis.text.x=element_blank(),
-        # axis.ticks.x=element_blank(), 
-        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
-  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
-  #       axis.title.y = element_text(vjust =0.25))
-
-plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
-plot_bai_s
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the three of most recent severe drought years (1995, 2005 and 2012).
-
-```{r}
-cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
-year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
-year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
-
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
-dys <- dys[,'maxyear'] 
-```
-
-```{r}
-# top10  
-top10pos50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg50 <- cro_rwi %>% filter(year > 1950) %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-top10pos <- cro_rwi %>% arrange_(~desc(res)) %>% group_by_(~site) %>%  slice(1:10)
-top10neg <- cro_rwi %>% arrange_(~res) %>% group_by_(~site) %>%  slice(1:10)
-```
-
-```{r}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
-
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
-
-# 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
-  theme_bw() + 
-  theme(panel.grid = element_blank(), 
-        strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
-        # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
-  geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
-  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
-             labeller = as_labeller(label_site)) + 
-  ylab("Ring width index") + xlab("Year") + 
-  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
-  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
-  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
-  # geom_vline(xintercept = dys, linetype = "dotted") + 
-  scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
-
-plot_rwi_crono  
-```
-
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation (a), SPEI (b), maximun (c) and minimun (d) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
-rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',')
-rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',')
-rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',')
-rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',')
-
-rtmax <- rtmax %>% mutate(v_name = 'T max')
-rtmin <- rtmin %>% mutate(v_name = 'T min')
-rsequia <- rsequia %>% mutate(v_name ='SPEI')
-rprec <- rprec %>% mutate(v_name = 'Prec')
-
-df <- rbind(rtmax, rtmin, rsequia, rprec)
-
-# cutre solucion para el problema de PREC 
-exdf <- data.frame(
-  id = c(16:19),
-  varname = rep("prec", 4), 
-  month = c('dec...FEB', 'MAR...MAY', 'JUN...AUG', 'SEP...NOV'),
-  coef = rep(0, 4), 
-  significant = rep('FALSE', 4),
-  ci_lower = rep(0, 4),
-  ci_upper = rep(0, 4), 
-  sig = rep("", 4), 
-  month_name = c("Winter", "Spring", "Summer", "Autumn"), 
-  v_name = rep("Prec", 4))
-
-exdf <- rbind(exdf, exdf, exdf)
-exdf$X <- c(46:57)
-exdf$site <- c(rep("SJ", 4), rep("caL", 4), rep("caH", 4))
-
-ddf <- rbind(df, exdf)
-
-label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
-               "T max" = "(c) T max", "T min" = "(d) T min")
-
-label_site <- c("SJ" = "SJ",
-                "caH" = "CA-High",
-                "caL" = "CA-Low")
-
-clima_ring <- ddf %>% ggplot(aes(x=id, y=coef, group = site)) + 
-  geom_hline(yintercept = 0, colour="gray") +
-  geom_bar(aes(fill=site), stat="identity", position = "dodge") +
-  facet_wrap(~v_name, labeller = as_labeller(label_var)) + 
-  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.05))), 
-            position = position_dodge(.9), size = 6) + 
-  theme_bw() + 
-  theme(axis.title.x = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor.y = element_blank(),
-        axis.text.x = element_text(angle = 90, vjust = 0.3),
-        legend.position = "bottom",
-        strip.background = element_rect(fill=NA, colour=NA),
-        strip.text.x = element_text(size = 10, face = "bold")) + 
-  scale_x_continuous(breaks = unique(ddf$id), labels = unique(ddf$month_name)) +
-  scale_y_continuous(limits = c(-0.4, 0.8), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6,0.8)) + 
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site) + 
-  ylab('Correlation')
-
-clima_ring 
-```
-
-#####
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
-
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
-
-# customize plot (corrected scale)
-pos_neg_log <- scales::trans_new('signed_log', 
-                                 transform = function(x) sign(x)*log(abs(x)),
-                                 inverse = function(x) sign(x)*exp(abs(x)))
-
-custom_ylab <- '% GC'
-micolor <- c("#838B8B")
-
-# solution for arrange
-# http://felixfan.github.io/stacking-plots-same-x/
-
-custom_corrected_gc <- list(
-  geom_bar(stat='identity', fill=micolor, colour=micolor),
-  theme_bw(), 
-  geom_hline(yintercept=c(50, -50), linetype=2, colour='black'),
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black',fill='white'))
-  )
-
-c <- gc_todos %>% 
-  filter(site != 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) + 
-  # scale_y_continuous(trans = pos_neg_log) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("1_caH" = "CA-High", "2_caL" = "CA-Low"))) + 
-    ylab(custom_ylab) + custom_corrected_gc
-
-d <- gc_todos %>% 
-  filter(site == 'gc_sj') %>% 
-  ggplot(aes(x=year, y=gc_md)) + 
-  xlim(layer_scales(c)$x$range$range[1],layer_scales(c)$x$range$range[2]) +
-  geom_hline(yintercept = 1, colour = 'gray', lwd = 0.3) + 
-  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour=micolor, width=.3, size=.2) +
-  facet_wrap(~site_sorted, nrow = 2,
-             labeller=as_labeller(c("0_sj" = "SJ"))) + 
-  custom_corrected_gc + ylab('') + 
-  scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-1000, -500, -100, 0, 100, 500)) +
-  theme(axis.title.x = element_blank(),
-        axis.text.x = element_blank())
-
-plot_medians_corregido <- d + c + plot_layout(ncol=1, heights = c(1, 2))
-plot_medians_corregido
-
-```
-
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis).
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
-
-colorpositivo <- c("#104E8B")
-colornegativo <- c("#CD3333")
-
-pgc_median50 <- gc_treesMed50 %>% ggplot() +
-  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, stat = "identity", position = "dodge") + 
-  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, stat = "identity", position = "dodge") +
-  facet_wrap(~site_sorted, nrow = 3, labeller = as_labeller(label_site)) + 
-  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
-  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
-  theme(panel.grid.minor = element_blank(), 
-        strip.background = element_rect(colour='black', 
-                                        fill='white')) 
-pgc_median50
-
-```
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r, fig.height=5}
-ltdf <- read.table(file=here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
-
-source(here("/script/R/", "hydro_year.R"))
-
-ltdf <- ltdf %>% 
-  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
-  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-# We used only the 5514 station (the most long term)
-# Filter by station and prec 
-# Filter from 1950 
-st5514 <- ltdf %>% 
-  filter(VARIABLE == 'PM1') %>% 
-  filter(cod == '5514') %>% 
-  filter(hyear >= 1950) %>% 
-  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)
-
-st5514_pro <- st5514 %>% 
-  group_by(hyear_f, hmonth) %>% 
-  dplyr::summarise(pre = sum(pre)) %>%
-  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()
-
-df <- st5514_pro  %>% 
-  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))
-
-cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
-  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
-  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'),
-            aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
-  theme_bw() + 
-  theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
-        axis.text.y = element_text(size=7),
-        axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
-  # ggtitle('Granada Base Aerea, 1950-2015') + 
-  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
-                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
-```
-
-```{r}
-prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
-
-dfpre <- c()
-
-for (i in prec_files){
-
-  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
-  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
-
-  dfpre <- rbind(dfpre, auxdf)
-}
-
-dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")
-
-dfpre <- dfpre %>% 
-  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
-         month = lubridate::month(fecha),
-         hmonth = ifelse(month <= 8, month + 4, month -8), 
-         hyear = hydro_year(fecha),
-         hyear_f = paste0(hyear-1,'-', hyear))
-
-aux <- dfpre %>% 
-  group_by(hyear_f, code) %>% 
-  dplyr::mutate(csum = cumsum(prec)) %>% 
-  filter(hmonth == 12) %>% 
-  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()
-
-presn <- aux %>% group_by(hyear) %>% 
-     dplyr::summarise(
-     csum_mean = mean(csum, na.rm = TRUE),
-     csum_sd = sd(csum, na.rm = TRUE), 
-     csum_se = csum_sd/sqrt(length(csum_sd)), 
-     n = n())
-
-meandf <- mean(presn$csum_mean)
-sddf <- sd(presn$csum_mean)
-
-prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
-  geom_hline(yintercept = meandf) + 
-  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
-  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
-  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
-  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
-  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
-  geom_point(size=2) + theme_bw() +
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
-  scale_y_continuous(breaks=seq(0,1500, by=250)) +
-  geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
-    aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
-
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
-```
-
-#####
-
-```{r, echo=FALSE}
-# Get file names
-myfiles <- list.files(path = here("/data_raw/spei_all/"), pattern = "\\.csv$")
-
-mydf <- data.frame()
-
-# Loop to read files and prepare data
-for (j in myfiles){ 
-  aux <- read.csv(file=paste0(here("/data_raw/spei_all/"), j), header = TRUE, sep = ',')
-
-  # Remove csv and get name 
-  name_aux <- str_replace(j, ".csv", "")
-
-  # Get lat long
-  latlong <- str_replace(name_aux, "SPEI_", "")
-
-  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
-  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
-
-  aux$lat <- mylat
-  aux$long <- mylong
-
-  # Split time 
-  # Get year 
-  aux$year <- as.numeric(str_extract(aux$DATA, "[0-9]+"))
-
-  # Get month 
-  aux$month <- str_extract(aux$DATA, "[aA-zZ]+")
-
-  # Variable month in number 
-  aux$months <- match(aux$month, month.abb)
-
-  # assign(name_aux, aux)
-
-  mydf <- rbind(mydf, aux)
-}
-
-spei <- mydf %>% select(SPEI_12, year, months) %>% 
-  group_by(year,months) %>% 
-  summarise(spei6 = mean(SPEI_12),
-            spei6_sd = sd(SPEI_12)) %>% 
-  mutate(date = lubridate::ymd(paste(year, months, "01")))
-
-spei <- spei %>% 
-  mutate(sign = case_when(
-    spei6 > 0 ~ "pos",
-    TRUE ~ "neg"))
-```
-
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, fig.height=5}
-limits <- as.Date(c("1949-01-01", "2015-12-01"))
-spei %>% 
-  ggplot(aes(x=date, y=spei6, fill=sign)) +  
-  geom_bar(stat = "identity") + 
-  scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
-  theme_bw()  + 
-  labs(title= 'Drought evolution for Sierra Nevada',
-    x= '',
-    y='SPEI (12-month scale)') +
-  theme(strip.background = element_rect(fill = "white"),
-        legend.position = "none") +
-  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limits=limits) 
-```
-
-```{r, echo=FALSE, eval=FALSE}
-**Figure S2**.
-Spatial distribution of the EVI-annual trends ($\tau$) for the period 2000-2016.
-Negative values (*red* values) indicate negative trends and positive values (*green* values) indicate positive trends.
-
-![](../analysis/evi/compute_MannKendall_evimean_files/figure-markdown_github/rater_map_tau_EVI-1.png){width=150%}
-```
-
-##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index.
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
-```{r}
-# peores sequias 
-dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
-# solamente aquellos mayor de 2 meses 
-dys <- dy %>% 
-  filter(d_duration > 2) %>%  
-  mutate(year = ifelse(minyear == maxyear, 
-                       minyear, 
-                       paste0(minyear, '-', maxyear))) %>% 
-  dplyr::select(-minyear, -maxyear, -cat, -month_peak) %>% 
-  as.data.frame()
-
-colnames(dys) <- c("Duration (months)", "Intensity", "Severity", "Lowest SPEI", "Year")
-
-pander(dys)
-```
-
-##### 
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
-
-```{r}
-df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
-
-df <- df %>% add_row(clu_pop2 = 'Northern slope',
-                             type = 'greening',
-                             count_clu = 0, per = 0.00, y = 2005)
-
-per_pixels_green <- df %>% 
-  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
-  facet_wrap(~clu_pop2) + 
-  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0))
-
-per_pixels_green
-```
-
-##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
-```{r, warning=FALSE, dpi = 100, fig.height=5}
-# ```{r, warning=FALSE, out.extra='angle=90', fig.height=4, fig.width=2}
-# see trends-resilience_bai.Rmd 
-df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
-
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
-plotrel <- function(df, v, x, label, XLab, YLab, coordX_eq, coordY_eq, titulo){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=2, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    xlab(XLab) + ylab(YLab) + 
-    geom_text_repel() + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE) +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 8, 
-                                    face = "bold.italic",
-                                    margin = margin(b = -15)), 
-          axis.title.y = element_blank())
-  }
-
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-        XLab = "Drought Severity", YLab = "Resilience (Rs) (3year window size)", 
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rs")
-
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Resistance (Rt) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rt")
-
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-        XLab = "", YLab = "Recovery (Rc) (3year window size)",
-        coordX_eq = 7, coordY_eq = 0.5, titulo = "Rc")
-
-sevRt + sevRc + sevRs + plot_layout(nrow = 1)
-
-```
-
-##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
-corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
-corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
-corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)
-
-generaMatrix <- function(cordf, variable){ 
-  m <- cordf %>% filter(var == variable) %>% 
-    dplyr::select(-var, -window_size) %>% 
-    mutate(w1 = c(2, 2, 3), 
-           w2 = c(3,4,4)) 
-  mf <- m %>% 
-    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)
-
-  df <- m %>% rbind(mf) %>% 
-    add_row(r2 = 1, w1=2, w2=2) %>% 
-    add_row(r2 = 1, w1=3, w2=3) %>% 
-    add_row(r2 = 1, w1=4, w2=4)
-
-    df <- df %>% 
-    mutate(w1 = case_when(
-      w1 == 2 ~ '2 years',
-      w1 == 3 ~ '3 years',
-      w1 == 4 ~ '4 years'),
-      w2 = case_when(
-      w2 == 2 ~ '2 years',
-      w2 == 3 ~ '3 years',
-      w2 == 4 ~ '4 years'))
-
-      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))
-
-return(cor_df)
-}
-
-cor_rt <- generaMatrix(corBai, 'rt')
-cor_rc <- generaMatrix(corBai, 'rc')
-cor_rs <- generaMatrix(corBai, 'rs')
-
-cor_rte <- generaMatrix(corEvi, 'rt')
-cor_rce <- generaMatrix(corEvi, 'rc')
-cor_rse <- generaMatrix(corEvi, 'rs')
-
-g_rt <- ggcorrplot(cor_rt, type='upper', lab=TRUE, title='BAI Rt', show.legend = FALSE)
-g_rc <- ggcorrplot(cor_rc, type='upper', lab=TRUE, title='BAI Rc', show.legend = FALSE)
-g_rs <- ggcorrplot(cor_rs, type='upper', lab=TRUE, title='BAI Rs', show.legend = FALSE)
-g_rte <- ggcorrplot(cor_rte, type='upper', lab=TRUE, title='EVI Rt', show.legend = FALSE)
-g_rce <- ggcorrplot(cor_rce, type='upper', lab=TRUE, title='EVI Rc', show.legend = FALSE)
-g_rse <- ggcorrplot(cor_rse, type='upper', lab=TRUE, title='EVI Rs', show.legend = TRUE)
-
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
-```
-
-##### 
-**Appendix S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
-
-```{r}
-correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
-
-correla <- correla %>% 
-  mutate(
-    name_comparison = as.factor(case_when(
-      name_comparison == "caH-caL" ~ "CAHigh - CALow",
-      name_comparison == "sj-caL" ~ "SJ - CALow",
-      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
-
-correla_sitesSJCALH  <- correla %>% 
-  filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) +
-  theme(legend.position = c(.8,.9), legend.title = element_blank())
-
-correla_sitesSJCALH
-```
-
-##### 
-**Appendix S8.** Tabla S1
-
-##### 
-**Appendix S9.** Tabla S2
-
-#####
-**Appendix S10.** Comparison EVI previous and post
-
-```{r}
-df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')
-
-ggplot(df, 
-       aes(x=disturb, y=mean, colour=site)) + 
-  geom_point(size=2) + 
-  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
-  geom_line(aes(group=site)) + 
-  facet_wrap(~disturb_year) + 
-  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
-  theme_bw() +
-  xlab("") + ylab("EVI") + 
-  theme(
-    strip.background = element_rect(fill = "white"),
-    strip.text = element_text(face="bold", size=12),
-    axis.title.y = element_text(face="bold", size=12)
-  ) +
-  scale_color_manual(values=c('black', 'blue', 'gray'))
-```

### 2018-11-23 man/figures/std_anomalies.Rmd
update figures std anomalies and EVI profile

@@
 output:
-  word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 6
-    highlight: null
-    reference_docx: ../../templates/template_pagebreak.docx
-bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
+  word_document
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
-library("ggrepel")
 library("scales")
-
-library("grid")
-library("gridExtra")
 library("patchwork")
-library("ggcorrplot")
-library("stringr")
-library("pander")
-library("lemon")
-```
-
-```{r, echo=FALSE}
-show_text <- FALSE
 ```
@@
 
-```{r, eval=FALSE}
-iv <- read.csv(here("data/evi", "iv_composite.csv"), header = TRUE, sep = ',')
-
-drought_years <- c(2005, 2012)
-
-evi_profile_dat_ref <- iv %>% 
-  filter(!year %in% drought_years) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = 'reference')
-
-evi_profile_dat_2005 <- iv %>% 
-  filter(year == 2005) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2005')
-
-evi_profile_dat_2012 <- iv %>% 
-  filter(year == 2012) %>% 
-  group_by(composite) %>% 
-  dplyr::summarise(mean=mean(evi, na.rm=TRUE),
-            sd = sd(evi, na.rm=TRUE),
-            se = sd/sqrt(length(evi))) %>% 
-  mutate(composite_dates = 
-           plyr::mapvalues(composite,
-                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
-                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
-                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
-                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
-  mutate(cd = as.Date(composite_dates, format = '%m-%d')) %>% 
-  mutate(period = '2012')
-
-# Join the three dataframes
-evi_profile_compara_dat <- rbind(evi_profile_dat_ref, evi_profile_dat_2005, evi_profile_dat_2012)
-
-# colref <- '#455883'
-# colref2 <- '#10253F'
-# col2005 <- '#00BA38'
-# col2012 <- '#F8766D'
-
-col2012 <- '#0700fe'
-col2005 <- '#19e00b'
-colref <- '#7b7b7b'
-
-profile_compara <- ggplot(evi_profile_compara_dat, aes(cd, y=mean, color=period)) + 
-  geom_errorbar(aes(ymin = mean - 2*se, ymax= mean + 2*se), width=3, size=.5) + 
-  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') + 
-  geom_line(size=.9) + 
-  geom_point(size=1.5, fill='white', shape=21) +
-  scale_x_date(labels = function(x) format(x, "%b"),
-               breaks = date_breaks('month')) + 
-  ylab('EVI') + xlab('Date') + 
-  theme_bw() +
-  theme_classic() + 
-  theme(panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),
-        legend.position = c(0.1,.9),
-        plot.title = element_text(hjust = -0.05, size = 16, 
-                                    margin = margin(t=-5, b=1))) +
-  scale_color_manual(values=c(col2005, col2012, colref))
-
-profile_compara 
-```
-
-```{r}
+```{r data}
 # From explore_anomalies.Rmd 
@@
 
-# colours 
-color_neg <- '#a63603'
-color_pos <- '#006d2c'
-
 # Standardized Anomalies 
@@
  mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
+```
+
+```{r plot}
+# colours 
+color_neg <- '#a63603'
+color_pos <- '#006d2c'
 
@@
                                     margin = margin(t=-5, b=1)))
-
-# plot_sa_clu
 ```
 
-```{r, fig.height=9} 
-# grid.arrange(profile_compara, plot_sa_clu)
+```{r out_plot, fig.width= 4, dpi = 300, out.width = 0.8}
+plot_sa_clu 
 ```
 
-```{r, fig.height=7, eval=FALSE}
+```{r oldFigure, eval=FALSE, fig.height=7}
+## See profile_EVI.Rmd 
+# grid.arrange(profile_compara, plot_sa_clu)
+
 a_profile_compara <- profile_compara + labs(title = "a")
@@
 ```
-
-```{r, fig.width=3.5, fig.height=5, fig.asp = .8}
-plot_sa_clu 
-```

### 2018-11-23 man/figures/browning.Rmd
adjust size

--- 2018-11-09 man/figures/browning.Rmd
@@
 output:
-  word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 6
-    highlight: null
-    reference_docx: ../../templates/template_pagebreak.docx
-bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
-editor_options: 
-  chunk_output_type: console
+  word_document
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
-library("ggrepel")
 library("scales")
-
-library("grid")
-library("gridExtra")
-library("patchwork")
-library("ggcorrplot")
-library("stringr")
-library("pander")
-library("lemon")
-library("viridis")
-library("ggthemes")
+library("viridis") # colours 
+library("ggthemes") # colours 
 ```
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
 #####
-**Appendix S4.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 droguht events according to EVI standardized anomalies.
 
-```{r}
+**Figure 3.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
+
+```{r data}
 df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
@@
 df <- df %>% mutate(type = stringr::str_to_title(type))
+```
 
+```{r plot}
 per_pixels_green <- df %>% 
   ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
-  geom_bar(stat='identity', position = "dodge") + 
-  # geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
+  geom_bar(stat = 'identity', position = "dodge") + 
+  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=2, angle = 90) + 
   facet_wrap(~clu_pop2) + 
   theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
-  theme(strip.background = element_rect(colour = "black", fill = "white")
-        ) + labs(fill = 'Drought event') + scale_fill_manual(values = c("#DAA520", "#104E8B")) +
-  scale_y_continuous(expand=c(.1,0)) + 
-  geom_text_repel(aes(label = paste0(per, " %")), 
-                  position=position_dodge(width=0.9))
+  theme(strip.background = element_rect(colour = "white", fill = "white"),
+        strip.text = element_text(size = 8), 
+        axis.title.y = element_text(size = 10, face = 'bold'),
+        axis.text = element_text(size = 8),
+        panel.grid = element_blank(),
+        legend.position = 'bottom') + 
+  labs(fill = 'Drought event') + 
+  scale_fill_colorblind() + 
+  scale_y_continuous(limits = c(0, 110), breaks = c(0, 25, 50, 75, 100))
+```
 
+```{r out_plot, fig.width= 4, dpi = 300, out.width = 0.8}
 per_pixels_green

### 2018-11-23 man/figures/growth_changes.Rmd
adjust sizes

@@
 output:
-  word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 6
-    highlight: null
-    reference_docx: ../../templates/template_pagebreak.docx
+  word_document
 bibliography: ../../refs/references.bib
 csl: ../../refs/ecology.csl
-editor_options: 
-  chunk_output_type: console
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
-library("ggrepel")
 library("scales")
-library("grid")
-library("gridExtra")
 library("patchwork")
-library("ggcorrplot")
-library("stringr")
-library("pander")
-library("lemon")
-library("magrittr")
 ```
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
-
-**Figure 7.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q.
+**Figure 5.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q.
 pyrenaica* sites.
@@
 
-```{r}
-gc_todos <- read.csv(here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
+```{r data}
+gc_todos <- read.csv(here::here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')
 
@@
                                  inverse = function(x) sign(x)*exp(abs(x)))
+```
 
+```{r plot_function}
 # Function for plot
@@
     theme_bw() +
-    theme(panel.grid = element_blank()) +
+    theme(panel.grid = element_blank(),
+          axis.text.y = element_text(size = 5.5),
+          axis.text.x = element_text(size = 6)) +
     coord_cartesian(xlim=limitesX)
@@
 
+```
+
+```{r plot_configure}
 # ---------- Min and Max axis        
@@
                             breaks = c(-100, -50, 0, 50, 50, 100, 150))
+```
 
-# plots
-
+```{r plot_caH}
 pcaH <- gc_todos %>% dplyr::filter(site == 'gc_caH') %>% 
@@
   scale_x_continuous(breaks = xrange, position = 'bottom') + xlab('') + 
-  annotate("text", x = 1990, y = 120, label = "CA-High", fontface = 2, size = 7)
+  annotate("text", x = 1990, y = 120, label = "CA-High", fontface = 2, size = 3)
+```
 
+```{r plot_caL}
 pcaL <- gc_todos %>% dplyr::filter(site == 'gc_caL') %>% 
@@
          my_ylab = NULL, micolor = "#838B8B", limitesX = limitesX) +  mismo_y +
-  scale_x_continuous(breaks = xrange, position = 'bottom') + 
-  annotate("text", x = 1990, y = 120, label = "CA-Low", fontface = 2, size = 7)
+  scale_x_continuous(breaks = xrange, position = 'bottom') + xlab('Year') + 
+  annotate("text", x = 1990, y = 120, label = "CA-Low", fontface = 2, size = 3) 
+```
 
+```{r plot_SJ}
 psj <- gc_todos %>% dplyr::filter(site == 'gc_sj') %>% 
@@
                      breaks=c(-1000, -500, -100, 1, 100, 500)) + 
-  geom_hline(yintercept = 1) +
-  annotate("text", x = 1990, y = 450, label = "SJ", fontface = 2, size = 7)
+  geom_hline(yintercept = 1)+
+  annotate("text", x = 1990, y = 440, label = "SJ", fontface = 2, size = 3)
 
+```
+
+```{r plot_join }
 gg <- psj + pcaH + pcaL + 
   plot_layout(ncol=1) & 
-  theme(plot.tag = element_text(size = 20, face = 2)) & 
+  # theme(plot.tag = element_text(size = 12, face = 2)) & 
   theme(plot.margin = margin(t=-1,r=3,b=-1,l=3, "pt"))
+```
 
+```{r out_plot, fig.height=5, dpi = 300, out.width = 0.8}
 gg
@@
 # gg + plot_annotation(tag_levels = c('a','b', 'c'))  
-
-# Ojo ver limites de ylim 
-# ggplot_build(pcaL)$layout$panel_params[[1]]$y.range
-# ggplot_build(pcaH)$layout$panel_params[[1]]$y.range
-
 ```

### 2018-11-23 man/figures/resilience_long_term.Rmd
adjust sizes

@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
+library("scales")
+library("patchwork")
 library("ggrepel")
-library("scales")
-library("grid")
-library("gridExtra")
-library("patchwork")
-library("stringr")
-library("pander")
-library("lemon")
-library("magrittr")
-```
-
-```{r, echo=FALSE}
-show_text <- FALSE
 ```
@@
 ##### 
-**Appendix S5.** Resilience metrics of the tree-growth for the most severe drought events (as from Appendix S3).
+**Figure 6.** Resilience metrics of the tree-growth for the most severe drought events (as from Appendix S3).
 *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
@@
 
-```{r}
+```{r data}
 df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)
@@
 
+sdf <- df %>% 
+  filter(samps >= 10) %>% 
+  group_by(disturb_year) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+```
+
+```{r plot_functions}
 #https://gist.github.com/ottadini/6882677 
@@
 
-sdf <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
-
 plotrel <- function(df, v, x, label, coordX_eq, coordY_eq, titulo, ...){
@@
   ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="black") +
+    geom_smooth(method='lm', se=FALSE, colour="gray") +
     geom_errorbar(mapping = yerrbar) + 
@@
     theme_bw() + 
-    geom_text_repel(nudge_x = 0.5) + 
+    geom_text_repel(nudge_x = 0.5, size = 3) + 
     annotate("text", x=coordX_eq, y = coordY_eq, 
              label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-             size = 5, fontface = "bold") +
+             size = 4, fontface = "bold") +
     ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 14, face = "bold.italic"),
+    theme(plot.title = element_text(size = 11, face = "bold.italic"),
           panel.grid = element_blank(), 
-          axis.text.y = element_text(size=10),
+          axis.text.y = element_text(size=9),
           axis.text.x = element_text(size=9),
-          axis.title = element_text(size = 12))
+          axis.title = element_text(size = 10))
   }
 
+```
+
+```{r plot_rt}
 sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
                  coordX_eq = 10, coordY_eq = 0.5, titulo = "Rt") +
-  xlab('') + ylab('')
+  xlab('') + ylab('') +
+  theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
+```
 
+```{r plot_rs}
 sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
                  coordX_eq = 10, coordY_eq = 0.5, titulo = "Rs") + 
-  xlab('') + ylab('')
+  xlab('') + ylab('') +
+  theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt"))
+```
 
+```{r plot_rc}
 sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
                  coordX_eq = 10, coordY_eq = 0.5, titulo = "Rc") +
-  xlab("Drought Severity") + ylab('')
+  xlab("Drought Severity") + ylab('') +
+  theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
+```
 
-res_long_term <- sevRt + sevRc + sevRs + plot_layout(nrow = 1)
+```{r plot_join}
+res_long_term <- sevRt + sevRc + sevRs + 
+  plot_layout(nrow = 1) 
+```
 
+```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8}
 res_long_term

### 2018-11-23 man/figures/cumulative_prec.Rmd
fix error size figures

@@
 output: 
-  word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 6
-    highlight: null
-    reference_docx: ../../templates/template_pagebreak.docx
-bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
+  word_document
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
 library("ggrepel")
@@
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
+#####
 
-##### 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-
-```{r}
+```{r data}
 ltdf <- read.table(file=here::here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')
@@
 
+```
+
+```{r plot_cummulatedPrec}
+
+size_line <- .25
+
 cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
-  geom_boxplot(aes(x=as.factor(hmonth))) +  
-  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black') + 
+  geom_boxplot(aes(x=as.factor(hmonth)), size = 0.25, outlier.size = 0.5) +  
+  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black', size = size_line) + 
   geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
-            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue') +
+            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue', size = size_line) +
   geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'), 
             aes(x=as.factor(hmonth), y=csum, group=1),
-            colour='red') +
+            colour='red', size = size_line) +
   theme_bw() + 
   theme(strip.background = element_rect(fill = NA, color = "black"),
-        axis.text.x = element_text(angle = 90, hjust = 1, size=7),
+        axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5, size=7),
         axis.text.y = element_text(size=7),
         axis.title.y = element_text(size = 7)) + 
-  ylab('Cumulative Precipitation (mm)')  + xlab('') + 
+  ylab("Cumulative \n Precipitation (mm)") + xlab('') + 
   # ggtitle('Granada Base Aerea, 1950-2015') + 
@@
                             "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
-  annotate("text", label = "2004-2005", x = 2, y = 500, size = 2, colour = "blue") + 
-  annotate("text", label = "2011-2012", x = 2, y = 550, size = 2, colour = "red") + 
-  annotate("text", label = "1950-2015", x = 3, y = 600, size = 2, colour = "black") +
-  theme(panel.grid = element_blank())
+  annotate("text", label = "2004-2005", x = 2, y = 420, size = 2, colour = "blue") + 
+  annotate("text", label = "2011-2012", x = 2, y = 520, size = 2, colour = "red") + 
+  annotate("text", label = "1950-2015", x = 2, y = 620, size = 2, colour = "black") +
+  theme(panel.grid = element_blank(), 
+        axis.title.y = element_text(hjust = 0.5),
+        plot.background = element_blank())
 ```
 
-```{r}
+```{r data_prec}
 prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")
@@
 sddf <- sd(presn$csum_mean)
+```
+
+```{r}
+
+yeah <- presn %>% filter(csum_mean < (meandf - 1*sddf)) %>% 
+  mutate(pos = csum_mean - 50) %>%  as.data.frame()
 
@@
   geom_point(size=2) + theme_bw() + 
-  geom_line(size=.3) + 
-  scale_x_continuous(breaks=seq(1950,2020, by=5)) +
+  geom_line(size=.25, colour = 'black') + 
+  scale_x_continuous(breaks=seq(1950,2015, by=5)) +
   scale_y_continuous(breaks=seq(0,1500, by=250)) + 
   geom_text_repel(
-    data = subset(presn, csum_mean < (meandf - 1*sddf)),
+    data = yeah,
     aes(label = hyear), 
-    size = 3) +
-  theme(panel.grid.minor = element_blank()) +
-  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
+    size = 2.5, 
+    # position = position_dodge(1), 
+    alpha = .5,
+    colour = "black",
+    segment.size = 0.1, 
+    segment.alpha = 0.5,
+    point.padding=unit(.7,'lines'),
+    nudge_y = -yeah$csum_mean,
+    direction = 'x'
+    ) +
+  theme(panel.grid = element_blank(),
+        axis.title = element_text(size = 9),
+        axis.text.y = element_text(size = 8)) +
+  ylab('Cumulative Precipitation (mm)') + xlab('Year') +
+  expand_limits(y=c(0,2000)) 
+```
 
-prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = 2020, ymin = 1050)
+```{r}
+p <- prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = Inf, ymin=1050)
 ```
+
+**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
+Points represent mean and errorbars standard error.
+*Black* line indicates mean for all period.
+*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
+*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
+Years with average values below -1SD are labelled.
+Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
+***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
+The boxplot representing the average from 1950-2015 period.
+Data from meteorological station Granada, Base Aérea.
+
+```{r plot_out, fig.width=7, dpi = 300, out.width = 0.8} 
+p
+```

### 2018-11-23 man/figures/growth_changes.Rmd
fix error size figures

@@
 
-```{r plot_join }
+```{r plot_join}
 gg <- psj + pcaH + pcaL + 
@@
 
-```{r out_plot, fig.height=5, dpi = 300, out.width = 0.8}
+```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8}
 gg

### 2018-11-23 man/figures/resilience_metrics.Rmd
fix error size figures

@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
 library("ggrepel")
-library("scales")
-
 library("grid")
@@
 library("patchwork")
-library("ggcorrplot")
-library("stringr")
-library("pander")
 library("lemon")
-```
-
-```{r, echo=FALSE}
-show_text <- FALSE
+library("bookdown")
 ```
@@
 #####
-**Figure 3.** Response *Q.
+
+```{r data}
+mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
+mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
+
+# NO rrs
+mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
+mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
+
+aux <- mhuber_filter %>% mutate(vari = '0_EVI')
+aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
+m <- bind_rows(aux2, aux)
+```
+
+```{r plot}
+pd <- position_dodge(width = 0.6)
+
+aes_disturb <- aes(x = disturb_year, y = M.Huber, 
+                   color = site_sorted, group = site_sorted,  
+                   fill = site_sorted, linetype=site_sorted,
+                   shape = site_sorted, 
+                   ymin = lower.ci, ymax = upper.ci)
+
+variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
+
+# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
+item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
+
+tl <- guide_legend(title = "Site")
+
+themeR <- list(
+  geom_errorbar(width=.1, position = pd),
+  geom_point(position = pd),
+  theme_bw(),
+  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
+  xlab(''), 
+  theme(
+    panel.grid.minor = element_blank(),
+    strip.background = element_rect(
+      colour = "black",
+      fill = "white"),
+    strip.text = element_text(face="bold", size = 10),
+    axis.title.y = element_text(face="bold", size = 10),
+    plot.title = element_text(hjust = -0.05, size = 10, margin = margin(t=-5, b=1)),
+    axis.text = element_text(size = 9),
+    panel.spacing.x = unit(1, "lines")),
+  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
+    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
+  scale_color_manual(labels = item_site, guide = tl,
+    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
+  scale_fill_manual(labels = item_site, guide = tl,
+    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
+  scale_linetype_manual(labels = item_site, guide = tl,
+                        values=c("solid","solid","solid","solid", "solid")),
+  scale_size_manual(labels = item_site, guide = tl,
+    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
+
+  geom_text(aes(label=Letter, y=postext), 
+            position=pd, 
+            show.legend = FALSE, size = 3)
+  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
+  # geom_text_repel( 
+  #                 nudge_x = - 0.5,
+  #                 nudge_y = 0.01,
+  #                 direction="both", 
+  #                 size = 3.5, show.legend = FALSE,
+  #                 segment.size = 0.5, 
+  #                 segment.alpha = 0.5) 
+)
+
+m <- m %>% mutate(postext = case_when(
+  vari == "0_EVI" ~
+    case_when(var_sorted == "0_rt" ~ 0.97,
+              var_sorted == "1_rc" ~ 1.19,
+              var_sorted == "2_rs" ~ 1.02),
+  vari == "1_BAI" ~
+     case_when(var_sorted == "0_rt" ~ 1.1,
+              var_sorted == "1_rc" ~ 1.7,
+              var_sorted == "2_rs" ~ 1.2)))
+
+m <- m %>% mutate(site = as.factor(
+  case_when(
+  site == "caH" ~ "CA-High",
+  site == "caL" ~ "CA-Low",
+  TRUE ~ site
+)))
+
+m <- m %>% mutate(site_sorted = as.factor(
+  case_when(
+  site == "N" ~ "0_N",
+  site == "S" ~ "1_S",
+  site == "SJ" ~ "2_SJ",
+  site == "CA-High" ~ "3_CA-High",
+  site == "CA-Low" ~ "4_CA-Low"
+)))
+
+rt <- m %>% filter(var == "rt") %>% 
+  ggplot(aes_disturb) + themeR + 
+  ylab("Resistance") +
+  theme(plot.margin = margin(r=0, l=0, t=0, b=-2, "pt"))
+# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
+
+rc <- m %>% filter(var == "rc") %>% 
+  ggplot(aes_disturb) + themeR + 
+  ylab("Recovery") +
+  theme(plot.margin = margin(r=0, l=0, t=0, b=-2, "pt"))
+# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
+
+rs <- m %>% filter(var == "rs") %>% 
+  ggplot(aes_disturb) + themeR + 
+  ylab("Resilience") + xlab('Drought event') +
+    theme(plot.margin = margin(r=0, l=0, t=0, b=-2, "pt"))
+# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
+```
+
+**Figure 7.** Response *Q.
 pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
@@
 
-```{r, eval=FALSE, echo=FALSE}
+```{r plot_out, fig.height=6.5, dpi = 300, out.width = 0.8} 
+grid_arrange_shared_legend(rt, rc, rs, ncol = 1, nrow = 3, position ='top')
+
+```
+
+```{r old1, eval=FALSE}
+
+theme_de <- list(
+  scale_color_manual(values=c('black','red')), 
+  scale_shape_manual(values=c(19, 15)),
+  scale_fill_manual(values=c('black','red')),  
+  scale_linetype_manual(values=c("solid", "solid")),
+  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
+
+)
+
+# EVI_site
+res_evi_site <-
+  ggplot(mhuber_filter, aes_site) + 
+  theme_res + theme_de + theme(legend.position='top') + 
+  ggtitle("a) EVI")
+
+# BAI_site
+res_bai_site <-
+  ggplot(mhuberbai_filter, aes_site) + 
+  theme_res + theme_de + theme(legend.position='bottom') + 
+  xlab('Site') + 
+  ggtitle("b) BAI")
+
+res_evi_site  + res_bai_site + plot_layout(ncol=1)
+
+# **Figure .** Response of *Q.
+pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
+Different letters above bars indicate significant post hoc differences between groups (see material and methods).
+```
+
+```{r old2, eval=FALSE, echo=FALSE, fig.height=7}
 mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
@@
 )
-```
 
-```{r, eval=FALSE, echo=FALSE}
 # EVI_drought 
@@
 
-```
-
-```{r, eval=FALSE, fig.height=7}
 res_evi_drought  + res_bai_drought + plot_layout(ncol=1)
 ```
-
-```{r, fig.height=9}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
-
-# NO plot rrs 
-mhuber_filter <- mhuber %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-mhuberbai_filter <- mhuberbai %>% filter(var != 'rrs') %>% mutate(disturb_year = as.factor(disturb_year))
-
-aux <- mhuber_filter %>% mutate(vari = '0_EVI')
-aux2 <- mhuberbai_filter %>% mutate(vari = '1_BAI')
-m <- bind_rows(aux2, aux)
-
-pd <- position_dodge(width = 0.6)
-
-aes_disturb <- aes(x = disturb_year, y = M.Huber, 
-                   color = site_sorted, group = site_sorted,  
-                   fill = site_sorted, linetype=site_sorted,
-                   shape = site_sorted, 
-                   ymin = lower.ci, ymax = upper.ci)
-
-variables  <- c("0_EVI" = "EVI", "1_BAI" = "BAI")
-
-# orden <- c("N", "S", "SJ", "CA-High","CA-Low")
-item_site <- c("0_N" = "N", "1_S" = "S", "2_SJ" = "SJ", "3_CA-High" = "CA-High", "4_CA-Low" = "CA-Low")
-
-tl <- guide_legend(title = "Site")
-
-themeR <- list(
-  geom_errorbar(width=.1, position = pd),
-  geom_point(position = pd),
-  theme_bw(),
-  facet_wrap(~vari, labeller = labeller(vari = variables), scales="free"),
-  xlab(''), 
-  theme(
-    panel.grid.minor = element_blank(),
-    strip.background = element_rect(
-      colour = "black",
-      fill = "white"),
-    strip.text = element_text(face="bold", size = 14),
-    axis.title.y = element_text(face="bold", size = 14),
-    plot.title = element_text(hjust = -0.05, size = 16, margin = margin(t=-5, b=1))),
-  scale_shape_manual(labels = item_site, guide = guide_legend(title = "Site"),
-    values=c("0_N" = 19, "1_S" = 21, "3_CA-High" = 22, "4_CA-Low" = 15, "2_SJ" = 17)), 
-  scale_color_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "blue", "3_CA-High" = 'blue', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_fill_manual(labels = item_site, guide = tl,
-    values=c("0_N" = 'black', "1_S" = "white", "3_CA-High" = 'white', "4_CA-Low" = 'blue', "2_SJ" = 'black')),
-  scale_linetype_manual(labels = item_site, guide = tl,
-                        values=c("solid","solid","solid","solid", "solid")),
-  scale_size_manual(labels = item_site, guide = tl,
-    values=c("0_N" = "2", "1_S" = "2", "3_CA-High" = 2.5, "4_CA-Low" = 2.5, "2_SJ" = 2.5)),
-
-  geom_text(aes(label=Letter, y=postext), 
-            position=pd, 
-            show.legend = FALSE, size = 3)
-  # , position_jitterdodge(dodge.width = 0.6, jitter.width = .2)
-  # geom_text_repel( 
-  #                 nudge_x = - 0.5,
-  #                 nudge_y = 0.01,
-  #                 direction="both", 
-  #                 size = 3.5, show.legend = FALSE,
-  #                 segment.size = 0.5, 
-  #                 segment.alpha = 0.5) 
-)
-
-m <- m %>% mutate(postext = case_when(
-  vari == "0_EVI" ~
-    case_when(var_sorted == "0_rt" ~ 0.97,
-              var_sorted == "1_rc" ~ 1.19,
-              var_sorted == "2_rs" ~ 1.02),
-  vari == "1_BAI" ~
-     case_when(var_sorted == "0_rt" ~ 1.1,
-              var_sorted == "1_rc" ~ 1.7,
-              var_sorted == "2_rs" ~ 1.2)))
-
-m <- m %>% mutate(site = as.factor(
-  case_when(
-  site == "caH" ~ "CA-High",
-  site == "caL" ~ "CA-Low",
-  TRUE ~ site
-)))
-
-m <- m %>% mutate(site_sorted = as.factor(
-  case_when(
-  site == "N" ~ "0_N",
-  site == "S" ~ "1_S",
-  site == "SJ" ~ "2_SJ",
-  site == "CA-High" ~ "3_CA-High",
-  site == "CA-Low" ~ "4_CA-Low"
-)))
-
-rt <- m %>% filter(var == "rt") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resistance") 
-# + geom_text(aes(label=Letter), position=position_jitterdodge(jitter.width = .2, dodge.width = 0.6), show.legend = FALSE)
-
-rc <- m %>% filter(var == "rc") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Recovery")
-# + geom_text(aes(label=Letter, y = 1.65), position=pd, show.legend = FALSE)
-
-rs <- m %>% filter(var == "rs") %>% 
-  ggplot(aes_disturb) + themeR + 
-  ylab("Resilience") + xlab('Drought event') 
-# + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE) 
-```
-
-```{r, fig.height=6}
-grid_arrange_shared_legend(rt, rc, rs,ncol = 1, nrow = 3, position='top') 
-```
-
-```{r, eval=FALSE}
-
-theme_de <- list(
-  scale_color_manual(values=c('black','red')), 
-  scale_shape_manual(values=c(19, 15)),
-  scale_fill_manual(values=c('black','red')),  
-  scale_linetype_manual(values=c("solid", "solid")),
-  labs(color = 'Drougth event', fill='Drougth event', linetype='Drougth event', shape='Drougth event')
-
-)
-
-# EVI_site
-res_evi_site <-
-  ggplot(mhuber_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='top') + 
-  ggtitle("a) EVI")
-
-# BAI_site
-res_bai_site <-
-  ggplot(mhuberbai_filter, aes_site) + 
-  theme_res + theme_de + theme(legend.position='bottom') + 
-  xlab('Site') + 
-  ggtitle("b) BAI") 
-```
-
-```{r, eval=FALSE}
-res_evi_site  + res_bai_site + plot_layout(ncol=1)
-
-**Figure .** Response of *Q.
-pyrenaica* forests to 2005 (*black*) and 2012 (*red*) drought events in terms of resistance, recovery and resilience of greenness (EVI; a) and tree radial growth (BAI; b) by site.
-Different letters above bars indicate significant post hoc differences between groups (see material and methods).
-```

### 2018-11-23 man/figures/ring_climate.Rmd
fix error size figures

--- 2018-10-27 man/figures/ring_climate.Rmd
@@
 output: 
-  word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 6
-    highlight: null
-    reference_docx: ../../templates/template_pagebreak.docx
-bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
-editor_options: 
-  chunk_output_type: console
+  word_document
 ---
@@
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
+#####
 
-##### 
-**Figure 6.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
-
-```{r}
+```{r data}
 rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'T max')
@@
                 "caL" = "CA-Low")
+```
 
+```{r plot_options}
 theme_ring_climate <- list(
@@
         panel.grid.minor.x = element_line(size=.5),
-        axis.text.x = element_text(angle = 90, vjust = 0.3)),
+        axis.text.x.top = element_text(angle = 90, hjust = 0, vjust = 0.5),
+        axis.text.x.bottom = element_text(angle = 90, hjust = 1, vjust = 0.5)),
   scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site)
 ) 
+```
 
+```{r plot_tmax}
 p_tmax <- df %>% filter(v_name == 'T max') %>% 
@@
   ylab('Correlation') + 
-  annotate("text", x = 2, y = 0.35, label = "T max", fontface =2, size = 5)
+  annotate("text", x = 2, y = 0.35, label = "T max", fontface =2, size = 3.5)
+```
 
+```{r plot_tmin}
 p_tmin <- df %>% filter(v_name == 'T min') %>% 
@@
   theme_ring_climate + 
-  scale_y_continuous(limits = c(-0.4, 0.6), 
+  scale_y_continuous(limits = c(-0.4, 0.4), 
                      breaks = c(-0.4, -0.2, 0, 0.2, 0.4)) +
@@
   ylab('Correlation') +
-  annotate("text", x = 2, y = 0.35, label = "T min", fontface =2, size = 5)
+  annotate("text", x = 2, y = 0.35, label = "T min", fontface =2, size = 3.5)
+```
 
+```{r plot_prec_spei}
 # prec-spei
@@
   ylab('Correlation') +
-  annotate("text", x = 2, y = 0.75, label = "Prec", fontface =2, size = 5) + 
-  annotate("text", x = 23, y = 0.75, label = "SPEI", fontface =2, size = 5)
+  annotate("text", x = 2, y = 0.75, label = "Prec", fontface =2, size = 3.5) + 
+  annotate("text", x = 23, y = 0.75, label = "SPEI", fontface =2, size = 3.5)
+```
 
+```{r plot_all}
 ptemperatura <- p_tmin + theme(legend.position="none") + p_tmax + theme(legend.position="bottom") + plot_layout(ncol=1)
 
-ring_climate <- p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b'))
+ring_climate <- p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b')) & theme(plot.tag = element_text(size = 12, face = 2)) &
+theme(plot.margin = margin(t=0,r=0,b=-2,l=0, "pt"))
+```
 
+**Figure 8.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
+pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures.
+*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
+Asteriks indicate significant ($P < 0.05$) correlation coefficients.
+
+```{r plot_out, fig.height=7, dpi = 300, out.width = 0.8} 
 ring_climate

### 2018-11-23 man/figures/bai_cronos.Rmd
update figures

@@
 
-##### 
 **Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
@@
 ##### 
-```{r, warning=FALSE}
+```{r, warning=FALSE, echo=FALSE}
 cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')

### 2018-11-23 man/figures/browning.Rmd
update figures

@@
 
-```{r load_pkg}
+```{r pkg}
 library("tidyverse")
@@
 
-#####
-
 **Figure 3.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
 
-```{r data}
+```{r}
 df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)

### 2018-11-23 man/figures/corr_plot_years.Rmd
update figures

@@
   word_document:
-    fig_caption: yes
-    fig_height: 6
-    fig_width: 6
-    highlight: null
-    reference_docx: ../../templates/template_pagebreak.docx
-bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
-editor_options: 
-  chunk_output_type: console
 ---
@@
 ##### 
-**Appendix S6.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
+**Appendix S5.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
 Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.

### 2018-11-23 man/figures/correla_boot_sites.Rmd
update figures

@@
 
-##### 
-**Figure S7.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
+**Figure S6.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
 Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.

### 2018-11-23 man/figures/cumulative_prec.Rmd
update figures

@@
 ```
-
-#####
 
@@
 
-```{r}
-
+```{r plot_prec}
 yeah <- presn %>% filter(csum_mean < (meandf - 1*sddf)) %>% 
@@
     size = 2.5, 
-    # position = position_dodge(1), 
     alpha = .5,
@@
 
-```{r}
+```{r plot_all}
 p <- prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = Inf, ymin=1050)

### 2018-11-23 man/figures/gc_perc_trees.Rmd
update figures

@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
-library("ggrepel")
 library("scales")
+
 library("grid")
@@
 library("patchwork")
+library("ggcorrplot")
 library("stringr")
@@
 library("lemon")
-library("magrittr")
 ```
 
-```{r, echo=FALSE}
-show_text <- FALSE
+**Figure S4.** Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis). Data for number of trees > 2 is shown.
+
+```{r data}
+gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
+gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
 ```
 
-##### 
-**Figure 8**. Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis). Data for number of trees > 2 is shown.
-
-```{r}
-gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
-gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
+```{r plotGC}
 
@@
            stat = "identity", position = "dodge") +
-  geom_line(aes(x=year, y=total_tree), colour = 'black') + 
-  theme_bw() + ylab('% trees') +
+  geom_line(aes(x=year, y=total_tree), colour = 'black', size = 0.4) + 
+  theme_bw() + ylab('% of trees') + xlab("Year") + 
   geom_hline(yintercept=0, alpha=.4) +
@@
         strip.text = element_blank(),
-        axis.text.y = element_text(size=10),
-        axis.text.x = element_text(size=9),
-        axis.title = element_text(size = 12))
+        axis.text.y.left = element_text(size=6),
+        axis.text.y.right  = element_text(size=5),
+        axis.text.x = element_text(size=6),
+        axis.title = element_text(size = 9))
 
+```
+
+```{r plot_out, fig.height=4.5, dpi = 300, out.width = 0.8} 
 pgc_median50

### 2018-11-23 man/figures/plot_spei12.Rmd
update figures

@@
 output: 
-  word_document:
-    reference_docx: ../../templates/template_pagebreak.docx
-bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
-editor_options: 
-  chunk_output_type: console
+  word_document
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
-library("ggrepel")
-library("scales")
-library("stringr")
-library("viridis")
-library("ggthemes")
 ```
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
-
-##### 
-**Appendix S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-
-```{r, echo=FALSE}
+```{r data}
 # Get file names
@@
 
-```{r, fig.height=5}
-p <- spei %>% 
-  ggplot(aes(x=date, y=spei, fill=sign)) +  
+```{r plot_spei}
+plot_spei <- spei %>%   
+  ggplot(aes(x=date, y=spei, fill = sign)) + 
   geom_bar(stat = "identity") + 
@@
 
-p
 ```
+
+**Appendix S2**.
+Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+Data from Global SPEI database (http://spei.csic.es/database.html).
+We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
+
+```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8} 
+plot_spei
+```

### 2018-11-23 man/figures/residual_chronos.Rmd
update figures

@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
 library("here")
-library("dplyr")
-library("ggrepel")
 library("scales")
@@
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
-
-##### 
-**Figure 5.** Residual tree-ring chronologies obtained for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the severe drought years identified in our climatic data (see Table S3).
-
-```{r}
+```{r data}
 cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
@@
 
-```{r}
+```{r data_aux}
 # Mayores y menores crecimientos 
@@
 
-```{r}
+```{r plot_rwi}
 label_site <- c("0_sj" = "SJ",
@@
         strip.background = element_rect(fill = "white"),
-        strip.text = element_text(face="bold", size=12), 
+        strip.text = element_text(face="bold", size=9), 
         # strip.text = element_blank(), 
-        axis.text = element_text(face="bold", size=8), 
-        axis.title = element_text(face="bold", size=12)) + 
+        axis.text = element_text(size=8), 
+        axis.title = element_text(face="bold", size=9)) + 
   geom_hline(yintercept = 1, colour = "grey") + 
@@
 
+```
+
+**Figure S3.** Residual tree-ring chronologies obtained for the *Q.
+pyrenaica* sites.
+Dashed red lines indicate the start of the reliable period (EPS > 0.85).
+Dotted black lines showing the severe drought years identified in our climatic data (see Table S3).
+
+```{r plot_out, fig.height=4.5, dpi = 300, out.width = 0.8} 
 plot_rwi_crono  

### 2018-11-23 man/figures/location_map.Rmd
update figures std anomalies and EVI profile
increase size

@@
 
-![](../../out/fig/geomap/mapa_situacion_v2.jpg){height=600px}
+![](../../out/fig/geomap/mapa_situacion_v2.jpg){height=750px}

### 2018-11-24 man/figures/browning.Rmd
fix sizes

@@
   theme(strip.background = element_rect(colour = "white", fill = "white"),
-        strip.text = element_text(size = 8), 
-        axis.title.y = element_text(size = 10, face = 'bold'),
-        axis.text = element_text(size = 8),
+        strip.text = element_text(size = 7), 
+        axis.title.y = element_text(size = 7, face = 'bold'),
+        axis.text = element_text(size = 7),
         panel.grid = element_blank(),
-        legend.position = 'bottom') + 
+        legend.position = 'bottom', 
+        legend.text = element_text(size = 7), 
+        legend.box.spacing = unit(0.01, "cm"))+
   labs(fill = 'Drought event') + 
   scale_fill_colorblind() + 
-  scale_y_continuous(limits = c(0, 110), breaks = c(0, 25, 50, 75, 100))
+  scale_y_continuous(limits = c(0, 110), breaks = c(0, 25, 50, 75, 100)) +
+  guides(fill = guide_legend(title.theme = element_text(size = 7)))
 ```

### 2018-11-24 man/figures/std_anomalies.Rmd
fix sizes

@@
   ggplot(aes(x=y, y=mean, fill=signo)) + 
+  geom_hline(yintercept = 0, size=0.3, colour = "gray") + 
   geom_bar(stat = "identity") + 
@@
   scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
+  ylab(myylab) + xlab('Year') +
   theme_bw() +
-  theme(text = element_text(size=12),
+  theme(axis.text = element_text(size=8),
         legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 12, 
+        panel.grid = element_blank(),
+        strip.background = element_rect(fill = "transparent", colour = "black", size = 0.2),
+        plot.title = element_text(hjust = -0.05, size = 9, 
                                     margin = margin(t=-5, b=1)))

### 2018-11-24 man/figures/growth_changes.Rmd
line size

@@
   p <- ggplot(df, aes_string(x = x, y = y)) +
-    geom_errorbar(mapping = yerrorbar, colour = 'gray', width = .35, size = .3, 
+    geom_errorbar(mapping = yerrorbar, colour = 'gray', width = .35, size = .2, 
                   na.rm = TRUE) + 
     geom_bar(stat = 'identity', fill = micolor, colour = micolor) +
-    geom_hline(yintercept = 0, colour = 'black', lwd = 0.3) + 
+    geom_hline(yintercept = 0, colour = 'black', lwd = 0.2) + 
     ylab(my_ylab) +
     # Add line to %GC = 50 
-    geom_hline(yintercept = c(50, -50), linetype=2, colour='black') +
+    geom_hline(yintercept = c(50, -50), linetype=2, colour='black', lwd=0.2) +
     theme_bw() +
@@
                      breaks=c(-1000, -500, -100, 1, 100, 500)) + 
-  geom_hline(yintercept = 1)+
+  geom_hline(yintercept = 1,  lwd=0.2)+
   annotate("text", x = 1990, y = 440, label = "SJ", fontface = 2, size = 3)

### 2018-11-24 man/figures/resilience_long_term.Rmd
update figures
fix font size

@@
 ```
-
-##### 
-**Figure 6.** Resilience metrics of the tree-growth for the most severe drought events (as from Appendix S3).
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right* Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
 
@@
              label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-             size = 4, fontface = "bold") +
+             size = 4.5, fontface = "bold") +
     ggtitle(titulo) + 
@@
 
+**Figure 6.** Resilience metrics of the tree-growth for the most severe drought events (as from Table S3).
+*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
+Points indicate average of resilience metrics for all populations.
+Error bar corresponds standard error.
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
+
 ```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8}

### 2018-11-24 man/figures/plot_spei12.Rmd
fix errors

@@
   ggplot(aes(x=date, y=spei, fill = sign)) + 
-  geom_bar(stat = "identity") + 
+  geom_bar(stat = "identity", colour = "transparent") + 
   scale_fill_manual(values = c("pos" = "#696969", "neg" = "brown4")) +
@@
   geom_hline(yintercept = 0, colour = 'gray') +
-  labs(x = '', y ='SPEI (12-month scale)') +
+  labs(x = 'Year', y ='SPEI (12-month scale)') +
   scale_x_date(date_breaks = "5 year", date_labels = "%Y", 
@@
 
-**Appendix S2**.
+**Figure S2**.
 Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).

### 2018-11-24 man/figures/ring_climate.Rmd
update figures
fix errors

@@
         panel.grid.minor.x = element_line(size=.5),
-        axis.text.x.top = element_text(angle = 90, hjust = 0, vjust = 0.5),
-        axis.text.x.bottom = element_text(angle = 90, hjust = 1, vjust = 0.5)),
+        axis.text.x.top = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 7),
+        axis.text.x.bottom = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
+        axis.text.y = element_text(size = 7), 
+        axis.title.y = element_text(size = 8),
+        legend.text = element_text(size = 7), 
+        legend.box.spacing = unit(0.01, "cm")),
   scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site)
@@
 
-```{r plot_out, fig.height=7, dpi = 300, out.width = 0.8} 
+```{r plot_out, fig.height=6, dpi = 300, out.width = 0.8} 
 ring_climate

### 2018-11-26 analysis/chronos/get_chronos.Rmd
update figures and chronos

@@
 
-```{r global_options, include=FALSE}
-knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
+```{r load_pkg, warning=FALSE, message=FALSE}
+library('here')
+library("tidyverse")
+library("stringr")
+library("dplR")
+library("knitr")
+# library("detrendeR")
+library("pander")
+library('boot')
+```
+
+```{r load_custom_fx}
+source(here::here('/script/R','rw_byTree.R'))
+source(here::here('/script/R','bai_piovesan.R'))
+source(here::here('/script/R','baiResilience.R'))
+source(here::here('/script/R','chrono_bai.R'))
+source(here::here('/script/R','suaviza_cronos.R'))
+source(here::here('/script/R','correlaBootIC.R'))
 ```
@@
 
-```{r}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep = '')
-
+```{r load_data}
 # sj 
-sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
+sj <- read.rwl(fname=here::here('/data_raw/dendro_ring/sn_sanjuan', 'sn_sanjuan.rwl'), format="tucson")
 
 # canar 
-ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
+ca <- read.rwl(fname=here::here('/data_raw/dendro_ring/sn_canar','sn_canar.rwl'), format="tucson")
 
 # Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-```
-
-```{r}
-source(paste0(di, 'script/R/rw_byTree.R'))
-source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/baiResilience.R'))
+compete <- read.csv(file=here::here('/data_raw', 'dendro_competence.csv'), header=TRUE, sep=',')
 ```
@@
 
-```{r, warning=FALSE, message=FALSE}
-library("tidyverse")
-library("stringr")
-library("dplR")
-library("knitr")
-# library("detrendeR")
-library("pander")
-library('boot')
-```
-
-# Read y Prepare data 
-* Leer datos `rwl` de SJ y CA
-* Leer datos de diametros de los focal tree
-
-```{r, echo=FALSE}
-machine <- 'ajpelu'
-# machine <- 'ajpeluLap'
-di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')
-
-# sj 
-sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")
-
-# canar 
-ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")
-
-# Read diameters data
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-```
-
-```{r}
-source(paste0(di, 'script/R/bai_piovesan.R'))
-source(paste0(di, 'script/R/rw_byTree.R'))
-source(paste0(di, 'script/R/chrono_bai.R'))
-source(paste0(di, 'script/R/suaviza_cronos.R'))
-source(paste0(di, 'script/R/correlaBootIC.R'))
-```
-
-* Crear dataframes `rwl` por cada CA_High, CA_Low, SJ 
-    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
-
-```{r}
-# Replace SNA by SJ and SNB by CA
-names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
-names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")
-
-# Remove g in name of some cores of CA. 
-names(ca) <- stringr::str_replace(names(ca), "g", "")
-```
-
-```{r}
-# Create subset to compare between sites 
-caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
-             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
-             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
-caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
-              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
-              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]
-```
-
-* Lectura y preparación de datos de diámetro
-
-```{r} 
-# Prepare Diameter data
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Change name focal according to loc
-compete <- compete %>% 
-  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm)
-
-# Set levels of eleveation 
-sj_lowcode  <- paste0('SJ', str_pad(1:10, 2, pad='0'))
-sj_highcode <- paste0('SJ', 11:20)
-ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
-            paste0('CA', 26:30))
-ca_highcode <- paste0('CA', 11:25)
-
-ft <- ft %>% 
-  mutate(elev = as.factor(
-    ifelse(id_focalLoc %in% sj_lowcode, 'Low',
-                       ifelse(id_focalLoc %in% sj_highcode, 'High',
-                              ifelse(id_focalLoc %in% ca_lowcode, 'Low', 'High'))))) %>%
-  mutate(site = as.factor(paste0(loc, '_', elev)))
-
-# SJ only site 
-ft <- ft %>% 
-  mutate(site = case_when(
-    loc == 'SJ' ~ 'SJ',
-    TRUE ~ as.character(site))
-  )
-```
-
 # Summary dendrochronology statistics
 
-* Por cada loc (ca/sj) o site (ca_high, ca_low, sj_high, sj_low) calculamos los siguientes datos: 
+* Por cada site (ca_high, ca_low, sj) calculamos los siguientes datos: 
     * age: mean + sd (average of all series)
@@
 rwl_by_site %>% 
-  write.csv(file=paste(di, "data/dendro_summary/site_dendro.csv", sep=""), row.names = FALSE)
+  write.csv(file=here::here("/data/dendro_summary", "site_dendro.csv"), row.names = FALSE)
 
@@
 
-## By Loc
+# Reports 
 ```{r}
-objects_rwl <- c('ca','sj')
-
-out <- c()
-
-for (i in objects_rwl){
-
-  aux <- get(i)
-  # Stats of singles RWL
-  aux_stats <- rwl.stats(aux)
-  # Add site 
-  aux_stats$loc <- as.factor(i)
-
-  # Compute interseries correlation 
-  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
-  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
-
-  out <- rbind(out, aux_stats)
-  }
-
-rwl_by_loc <- out %>% group_by(loc) %>% 
-  summarize(age_mean = round(mean(year),2), 
-            age_sd = round(sd(year), 2),
-            year_min = min(first),
-            year_max = max(last), 
-            year_range = year_max - year_min + 1, 
-            ar1_mean = round(mean(ar1), 3),
-            ar1_sd = round(sd(ar1), 3),
-            icor_mean = round(mean(interser_cor_val), 3),
-            icor_sd = round(sd(interser_cor_val), 3)) %>% 
-  as.data.frame()
-
-rwl_by_loc %>% 
-  write.csv(file=paste(di, "data/dendro_summary/loc_dendro.csv", sep=""), row.names = FALSE)
-
-pander(rwl_by_loc, caption='Dendrochronological summary by sites') 
+rwl.report(sj)
+rwl.report(caL)
+rwl.report(caH)
 ```
@@
 
-* Agregar valores medios de RW por loc (obtenemos sj_tree / ca_tree) o por site (sjL_tree, sjH_tree / caL_tree, caH_tree) 
+* Agregar valores medios de RW por site (sj / caL_tree, caH_tree) 
 * ver fun rw_byTree o utilizar treeMean (dplR)
@@
 # Remember snc = structure of core name SJ0101 (site | tree | core)
-sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
-ca_tree <- rw_byTree(ca, snc =c(2,2,2), locname = 'CA')
+sj_treeRaw <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
+sj_tree <- subset(sj_treeRaw, select = -nseries)
 
-sjL_tree <- rw_byTree(sjL, snc =c(2,2,2), locname = 'SJ')
-sjH_tree <- rw_byTree(sjH, snc =c(2,2,2), locname = 'SJ')
-caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
-caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')
+caL_treeRaw <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
+caL_tree <- subset(caL_treeRaw , select = -nseries)
+
+caH_treeRaw <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')
+caH_tree <- subset(caH_treeRaw , select = -nseries)
 
@@
 # sj_meantree <- treeMean(sj, ids = ids_sj,  na.rm=TRUE)
+
 ```
@@
 bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
-
-bai_sj1 <-bai_piovesan(rwdf = sj, diam_df = d_sj)
-
-str(sj_tree)
-str(d_sj)
-
-bai_ca <- bai_piovesan(rwdf = ca_tree, diam_df = d_ca)
-
-bai_sjH <- bai_piovesan(rwdf = sjH_tree, diam_df = d_sjH)
-bai_sjL <- bai_piovesan(rwdf = sjL_tree, diam_df = d_sjL)
 bai_caH <- bai_piovesan(rwdf = caH_tree, diam_df = d_caH)
@@
 # Esto es para que funcionen algunas otras funciones de dplR 
-bais <- c('bai_sj', 'bai_ca', 'bai_sjH', 'bai_sjL', 'bai_caH', 'bai_caL')
+bais <- c('bai_sj', 'bai_caH', 'bai_caL')
 
@@
 }
+```
+
+# Cronologías by site
+
+```{r}
+cro_sj <- chrono_bai(bai_sj)
+cro_sj$site <- 'SJ'
+# Add nseries 
+sj_treeRaw$year <- as.numeric(row.names(sj_treeRaw))
+cro_sj <- cro_sj %>% inner_join(sj_treeRaw[,c('year','nseries')], by='year')
+
+cro_caH <- chrono_bai(bai_caH)
+cro_caH$site <- 'CA_High'
+# Add nseries 
+caH_treeRaw$year <- as.numeric(row.names(caH_treeRaw))
+cro_caH <- cro_caH %>% inner_join(caH_treeRaw[,c('year','nseries')], by='year')
+
+cro_caL <- chrono_bai(bai_caL)
+cro_caL$site <- 'CA_Low'
+# Add nseries 
+caL_treeRaw$year <- as.numeric(row.names(caL_treeRaw))
+cro_caL <- cro_caL %>% inner_join(caL_treeRaw[,c('year','nseries')], by='year')
 
@@
 
-# Cronologías by site / loc
+```{r crono_compara_sitesSJCALH}
+# Combine and plot 
+cro_sites <- rbind(cro_caL, cro_caH, cro_sj)
 
-### Mean vs. Tukey’s Biweight Robust Mean
+## Export data of bai (chro)
+write.csv(cro_sites, file=here::here("/data/cronos_medias", "cronos_sites.csv"), row.names = FALSE)
 
-* Computo la cronologia usando la función `chron`. 
-* Para SJ hag una comparación de la cronologia construida con biweight (*treatment = biweigth*) o construida con la media aritmética (*treatment = No biweigth*) (calculada con la función `chron`)
-* Idem para CA, pero en CA si separo entre CA high y CA low
-
-```{r chro_sj_compare_method}
-# Robust mean biweith or not
-cro_sjb <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
-cro_sjnotb <- chron(bai_sj, prefix = "SJ", biweight = FALSE)
-
-cro_b <- cro_sjb %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = SJstd, samp.depth, year, treatment)
-
-cro_notb <- cro_sjnotb %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = SJstd, samp.depth, year, treatment)
-
-aux_cro <- rbind(cro_b, cro_notb)
-
-aux_cro %>% ggplot(aes(x = year, y = bai_mean / 100, colour=treatment)) + 
-  geom_point() + geom_line() + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
-  theme_bw() + theme(panel.grid.major = element_blank()) +
-  ggtitle('SJ chronology built using Biweight or Mean')
-```
-
-```{r chro_ca_compare_method}
-# Robust mean biweith or not
-cro_cab_h <- chron(bai_caH, prefix = "CA", biweight = TRUE)
-cro_canotb_h <- chron(bai_caH, prefix = "CA", biweight = FALSE)
-
-cro_cab_h_aux <- cro_cab_h %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_canotb_h_aux <- cro_canotb_h  %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-aux_croca_h <- rbind(cro_cab_h_aux, cro_canotb_h_aux)
-aux_croca_h$site <- 'CA High'
-
-cro_cab_l <- chron(bai_caL, prefix = "CA", biweight = TRUE)
-cro_canotb_l <- chron(bai_caL, prefix = "CA", biweight = FALSE)
-
-cro_cab_l_aux <- cro_cab_l %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_canotb_l_aux <- cro_canotb_l  %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('No biweight')) %>% 
-  dplyr::select(bai_mean = CAstd, samp.depth, year, treatment)
-
-aux_croca_l <- rbind(cro_cab_l_aux, cro_canotb_l_aux)
-aux_croca_l$site <- 'CA Low'
-
-aux_croca <- rbind(aux_croca_h, aux_croca_l)
-
-aux_croca %>% ggplot(aes(x = year, y = bai_mean/100, colour=treatment)) + 
-  geom_point() + geom_line() + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) + 
-  theme_bw() + theme(panel.grid.major = element_blank()) + 
-  facet_wrap(~site, nrow = 2) +
-    ggtitle('CA chronology built using Biweight or Mean')
-```
-
-## Compute Chrono 
-### San Juan chronologies by site
-
-```{r chrono_sj_two_sites}
-# San Juan 
-cro_sjH <- chrono_bai(bai_sjH)
-cro_sjH$site <- 'SJ_High'
-
-cro_sjL <- chrono_bai(bai_sjL)
-cro_sjL$site <- 'SJ_Low'
-
-cro_sj<- rbind(cro_sjH, cro_sjL)
-cro_sj$loc <- 'Norte (San Juan)'
-
-plot_cro_sj <- cro_sj %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+plot_cro_sitesSJCALH <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
   theme_bw() + ylab('BAI') + 
@@
                   ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + geom_point(size=1) + 
-  # scale_colour_manual(values = c("red", "blue")) + 
-  theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1920, 2020, by=10))
-
-plot_cro_sj
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sj_HL.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sj, width=12, height = 4) 
-```
-
-### Cañar chronologies by site
-
-```{r chrono_ca_two_sites}
-# Canar 
-cro_caH <- chrono_bai(bai_caH)
-cro_caH$site <- 'CA_High'
-
-cro_caL <- chrono_bai(bai_caL)
-cro_caL$site <- 'CA_Low'
-
-cro_ca <- rbind(cro_caH,cro_caL)
-cro_ca$loc <- 'Sur (Canar)'
-
-plot_cro_ca <- cro_ca %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
+                  fill=site), alpha=.4, colour=NA) +
   geom_line() + 
-  geom_point(size=1) + 
-  # scale_colour_manual(values = c("red", "blue")) + 
-  theme(panel.grid = element_blank()) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep='')))+
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-
-plot_cro_ca
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_ca_HL.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_ca, width=12, height = 4) 
-```
-
-```{r chronos_4_sites}
-# Combine and plot 
-cro_sites <- rbind(cro_ca, cro_sj)
-
-plot_cro_sites <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + facet_wrap(~loc, nrow = 2) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-
-plot_cro_sites
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites, width=12, height = 8)
-```
-
-```{r chronos_4_sites_collapsed}
-plot_cro_sites_collapsed <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  # geom_point(size=1) +
+  geom_point(size=.8) + 
   theme(panel.grid = element_blank(),
@@
         strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
   ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
   scale_x_continuous(breaks = seq(1815, 2020, by=10))
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
-plot_cro_sites_collapsed
-```
 
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed, width=12, height = 4)
-plot_cro_sites_collapsed
-```
+plot_cro_sitesSJCALH
 
-```{r chronos_4_sites_collapsed50}
-plot_cro_sites_collapsed50  <- cro_sites %>% 
-  filter(year > 1949) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
+# test_bai SJ error std
+plot_cro_sitesSJ_testerror <- cro_sites %>% filter(site == "SJ") %>% 
+  ggplot(aes(x=year, y=bai_mean/100)) + 
   theme_bw() + ylab('BAI') + 
   geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
+                  ymax = (bai_mean + bai_se)/100), alpha=.4) +
   geom_line() + 
-  # geom_point(size=1) +
+  geom_line(aes(y = (bai_mean - bai_se)/100), linetype = "dotted") +  
+  geom_line(aes(y = (bai_mean + bai_se)/100), linetype = "dotted") +
+
+  geom_point(size=.8) + 
   theme(panel.grid = element_blank(),
@@
         strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
+  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
+  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
   ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1950, 2020, by=5)) 
-#  geom_smooth(method= 'loess', span = 0.75, se=FALSE)
+  scale_x_continuous(breaks = seq(1815, 2020, by=10))
 
-plot_cro_sites_collapsed50
-```
+plot_cro_sitesSJ_testerror
 
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed50.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed50, width=12, height = 4) 
-plot_cro_sites_collapsed50
-```
-
-```{r chronos_4_sites_collapsed90}
-plot_cro_sites_collapsed90  <- cro_sites %>% 
-  filter(year > 1989) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  # geom_point(size=1) +
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  scale_fill_manual(values = c('#1b9e77','#d95f02','#7570b3','#e7298a')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1990, 2016, by=2))
-
-plot_cro_sites_collapsed90
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sites_collapsed90.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sites_collapsed90, width=12, height = 4) 
-plot_cro_sites_collapsed90
 ```
@@
 
-```{r}
-# centred moved average smooth
-suaviza_caH <- suaviza_cronos(cro_caH, win_vector=seq(1,40, by=1), crononame = 'cro_caH')
-suaviza_caL <- suaviza_cronos(cro_caL, win_vector=seq(1,40, by=1), crononame = 'cro_caL')
-suaviza_sjH <- suaviza_cronos(cro_sjH, win_vector=seq(1,40, by=1), crononame = 'cro_sjH')
-suaviza_sjL <- suaviza_cronos(cro_sjL, win_vector=seq(1,40, by=1), crononame = 'cro_sjL')
-```
-
 ## between sites similarity (sjH *vs* sjL; caH *vs* caL)
@@
 # set seed
+# centred moved average smooth
+suaviza_caH <- suaviza_cronos(cro_caH, win_vector=seq(1,40, by=1), crononame = 'cro_caH')
+suaviza_caL <- suaviza_cronos(cro_caL, win_vector=seq(1,40, by=1), crononame = 'cro_caL')
+suaviza_sj <- suaviza_cronos(cro_sj, win_vector=seq(1,40, by=1), crononame = 'cro_sj')
+
 set.seed(333)
@@
 
-# SJ 
-suaviza_sj <- rbind(suaviza_sjH, suaviza_sjL)
-correla_sj <- correlaBootIC(suaviza_sj, tc, years_suavizado = 40, 
-                            name_comparison = 'sjH-sjL')
-
-# CA 
-suaviza_ca <- rbind(suaviza_caH, suaviza_caL)
-correla_ca <- correlaBootIC(suaviza_ca, tc, years_suavizado = 40, 
+# caHcaL
+suaviza_caHcaL <- rbind(suaviza_caH, suaviza_caL)
+correla_caHcaL <- correlaBootIC(suaviza_caHcaL, tc, years_suavizado = 40, 
                             name_comparison = 'caH-caL')
 
-correla <- rbind(correla_ca, correla_sj)
-```
-
-```{r correla_chronos_sites, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-correla %>% 
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  # geom_smooth(method = 'loess', se=FALSE) + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2)  +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  facet_wrap(~metodo_correla, ncol=2) +
-  theme(strip.background = element_blank())
-```
-
-```{r correla_chronos_sites_pearson}
-correla_sites_pearson <- correla %>% filter(metodo_correla == 'pearson') %>%
-  ggplot(aes(x=size, y=estimate, colour=name_comparison)) + 
-  geom_line()+ theme_bw() + ylab('Correlation coefficient') +
-  xlab('Smoothing (centred moving average window size)') + 
-  theme(panel.grid = element_blank()) +
-  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
-  geom_line(aes(x=size, y=ci_lower, colour=name_comparison), linetype = 2) +
-  geom_line(aes(x=size, y=ci_upper, colour=name_comparison), linetype = 2) + 
-  theme(legend.position = c(.8,.7))
-correla_sites_pearson
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/correla_boot_sitesHL.pdf') 
-ggsave(filename=fileplot, plot=correla_sites_pearson, width=5, height = 5) 
-```
-
-## between final sites similarity (sj *vs* caH; sj *vs* caL; caH *vs* caL)
-
-```{r, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
-# Metodo correlation
-tc <- c('spearman', 'pearson')
-
-set.seed(333)
-
-# Compute the suaviza_sj desde el principio, sin diferenciar h and l. Para ello la crono la computo de nuevo 
-cro_sjloc <- chrono_bai(bai_sj)
-cro_sjloc$site <- 'SJ'
-suaviza_sjloc <- suaviza_cronos(cro_sjloc, win_vector=seq(1,40, by=1), crononame = 'cro_sj')
-
 # sjcaH 
-suaviza_sjcaH <- rbind(suaviza_sjloc, suaviza_caH)
+suaviza_sjcaH <- rbind(suaviza_sj, suaviza_caH)
 correla_sjcaH <- correlaBootIC(suaviza_sjcaH, tc, years_suavizado = 40, 
                             name_comparison = 'sj-caH') 
-
 # sjcaL 
-suaviza_sjcaL <- rbind(suaviza_sjloc, suaviza_caL)
+suaviza_sjcaL <- rbind(suaviza_sj, suaviza_caL)
 correla_sjcaL <- correlaBootIC(suaviza_sjcaL, tc, years_suavizado = 40, 
@@
 # Join all correlations 
-correla <- rbind(correla_ca, correla_sjcaL, correla_sjcaH)
+correla <- rbind(correla_caHcaL, correla_sjcaL, correla_sjcaH)
 
-write.csv(correla, file=paste(di, "data/correla_ventanas_temporales/correla_chronos.csv", sep=""), row.names = FALSE)
+write.csv(correla, file=here::here("/data/correla_ventanas_temporales", "correla_chronos.csv"), row.names = FALSE)
 
@@
 
-```
-
-```{r correla_sitesSJCALH}
 correla_sitesSJCALH  <- correla %>% 
@@
 ```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/correla_boot_sitesSJCALH.pdf') 
-ggsave(filename=fileplot, plot=correla_sitesSJCALH, width=5, height = 5) 
-```
-
-# Chronologies by site (SJ, CAL, CAH)
-
-```{r}
-# Compute la crono de SJ de nuevo y llamar al site SJ
-cro_sjf <- chrono_bai(bai_sj)
-cro_sjf$site <- 'SJ'
-```
-
-```{r crono_compara_sitesSJCALH}
-# Combine and plot 
-cro_sites <- rbind(cro_caL, cro_caH, cro_sjf)
-
-## Export data of bai (chro)
-write.csv(cro_sites, file=paste(di, "data/cronos_medias/cronos_sites.csv", sep=""), row.names = FALSE)
-
-plot_cro_sitesSJCALH <- cro_sites %>% ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-
-plot_cro_sitesSJCALH
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH, width=12, height = 5)
-plot_cro_sitesSJCALH
-```
-
-```{r crono_compara_sitesSJCALH50}
-plot_cro_sitesSJCALH50 <- cro_sites %>% 
-  filter(year > 1949) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10)) 
-plot_cro_sitesSJCALH50
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH50.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH50, width=12, height = 5)
-```
-
-```{r crono_compara_sitesSJCALH90}
-plot_cro_sitesSJCALH90 <- cro_sites %>% 
-  filter(year > 1989) %>% 
-  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + 
-  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
-                  ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
-  geom_line() + 
-  geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
-  scale_colour_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  scale_fill_manual(values = c('#7570b3','#e7298a','#1b9e77')) +
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))
-plot_cro_sitesSJCALH90
-```
-
-```{r}
-fileplot <- paste0(di, '/out/chronos/crono_compara_sitesSJCALH90.pdf') 
-ggsave(filename=fileplot, plot=plot_cro_sitesSJCALH90, width=12, height = 5)
-```
-
-```{r, echo=FALSE, eval=FALSE}
-
-# CODE SNIPPETS 
-  # Loess 
-  l <- loess(b_aux$bai ~ b_aux$year, span = alpha)
-  years_v <- seq(min(b_aux$year), max(b_aux$year), 1)
-
-  l_pred_aux <- predict(l, years_v, se=TRUE)
-  l_pred <- as.data.frame(cbind(loess_fit = l_pred_aux$fit, 
-                loess_se = l_pred_aux$se.fit,
-                year = years_v))
-
-  out <- out %>% inner_join(l_pred, by='year')
-
-  out
-}
-
-# SJ
-
-cro_sj <- chron(bai_sj, prefix = "SJ", biweight = TRUE)
-cro_sj %>% plot(add.spline=TRUE, nyrs=30)
-
-cro_ca <- chron(bai_ca, prefix = "CA", biweight = TRUE)
-cro_ca %>% plot(add.spline=TRUE, nyrs=30)
-
-cro_ca2 <- cro_ca %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('CA')) %>% 
-  select(bai_mean = CAstd, samp.depth, year, treatment)
-
-cro_sj2 <- cro_sj %>% 
-  mutate(year = as.numeric(rownames(.))) %>% 
-  mutate(treatment = as.factor('SJ')) %>% 
-  select(bai_mean = SJstd, samp.depth, year, treatment)
-
-cro2 <- rbind(cro_ca2, cro_sj2)
-
-cro2 %>% ggplot(aes(x=year, y=bai_mean, colour=treatment)) + 
-  geom_line() + geom_smooth(method= 'loess', span = 0.30, se=FALSE) + theme_bw() + 
-  geom_bar(aes(x=year, stat = (samp.depth*10), colour=treatment))
-
-## San Juan Data
-
-Algunos árboles tienen 3 cores. Sin embargo, no los tres cores llegan hasta la corteza, por lo tanto, no podemos utilizar el diametro para estimar bai. Vamos a realizar lo siguiente:
-
-* Crear dos datasets: 
-    * Dataset con series de datos que llegan hasta corteza (`sj_cor`)
-    * Dataset con series de datos que no llegan hasta corteza (`sj_sin`): se trata de cores que no llegan hasta la corteza.
-
-* Para el cálculo del BAI, en el dataset `sj` utilizamos el diámetro medido en campo. En el caso del dataset `sj_sin` utilizamos la suma de todos los diámetros (:red_circle: `$TODO$ ASK to Guillermo`)
-
-# Get summary ring-width series
-sj_summ <- summary(sj)
-
-# Get names cores with last year different to 2016 
-id_cores_no_bark <- sj_summ %>% 
-  filter(last != 2016) %>% select(series) %>% mutate(series = factor(series))
-
-id_cores_no_bark <- unname(unlist(id_cores_no_bark))
-
-# Subbet datasets 
-sj_cor <- sj[which(! colnames(sj) %in% (id_cores_no_bark))]
-sj_sin <- sj[which(colnames(sj) %in% (id_cores_no_bark))]
-
-Lectura y preparación de datos de diámetro
-
-compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
-
-# Compute diameter (mm)
-compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
-
-# Get only focal trees, and only selected variables 
-ft <- compete %>% 
-  filter(sp=='Focal') %>% 
-  filter(id_focal!='Fresno') %>% 
-  select(id_focal, loc, dn_mm, height_cm)
-
-# Select only SJ trees and 
-ft_sj <- ft %>% filter(loc=='SJ')
-
-# get Create a dataframe with core ID
-cores <- data.frame(id_cores=colnames(sj))
-
-# Extract replicate and Tree ID from core ID 
-cores <- cores %>% 
-  mutate(id_focal = as.factor(stringr::str_sub(id_cores, 3,5)),
-         id_replica = stringr::str_sub(id_cores, 6,8))
-
-# Create df with diameter and height for each core ID 
-diam_cores <- cores %>% inner_join(ft_sj, by='id_focal') %>% 
-  select(id_cores, dn_mm)
-
-# remove diamm of cores without bark 
-diam_cores_sj <- diam_cores %>% filter(! id_cores %in% id_cores_no_bark)
-
-rwl.report(sj_cor)
-
-bai_p <- bai_piovesan(rwdf = sj_cor, diam_df = diam_cores_sj)
-bai_d <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
-
-diam_cores_sjsin <- data.frame(id_cores = colnames(sj_sin),
-                               dn_mm = colSums(sj_sin, na.rm = TRUE))
-rownames(diam_cores_sjsin) <- NULL
-
-bai_psin <- bai_piovesan(rwdf = sj_sin, diam_df = diam_cores_sjsin)
-bai_dsinsin <- bai.out(rwl = sj_sin)
-bai_dsin <- bai.out(rwl = sj_sin, diam = diam_cores_sjsin)
-
-set.seed(1234)
-serie <- sample(colnames(b_test_dplR),1)
-
-require(reshape2)
-
-bp <- melt(t(b_test_piovesan))
-names(bp) <- c('cores', 'years', 'bai_p')
-
-bd <- melt(t(b_test_dplR))
-names(bd) <- c('cores', 'years', 'bai_d')
-
-diff <- bp %>% 
-  inner_join(bd, by=c('cores','years')) %>% 
-  mutate(diff = bai_p - bai_d)
-
-ggplot(x, aes(y=bai, x=years, colour=test)) + 
-  geom_point() + 
-  facet_wrap(~cores)
-
-# BAI dplR
-baisj <- bai.out(rwl = sj_cor, diam = diam_cores_sj)
-
-baisjt <- bai.out(rwl = sjt)
-
-# Function to compute BAI (Piovesan et al)
-
-# rwl_file
-myrwl <- sj_cor
-# Diameter df  
-diam_df <- diam_cores_sj
-
-# Core
-id_core <- 'SNA0101'
-
-# Ring-width serie
-myrwl_serie <- myrwl[[id_core]]
-
-# Diameter for core id_core 
-dbh <- subset(diam_df, id_cores == id_core)
-dbh <- dbh[,2]
-
-myrwl_serie2 <- na.omit(myrwl_serie)
-cum
-
-r0 <- dbh/2 - c(0,cumsum(rev(myrwl_serie2)))
-
-bai <- (pi/4)*(dbh^2 - ((dbh - 2*myrwl_serie)^2))
-
-s <- myrwl_serie[94:96]
-
-d <- sum(s)*2
-
-########## test Piovesan
-# rwl_file
-myrwl <- baisj
-
-#id_core
-
-# Extract rwl serie 
-myrwl_i <- myrwl[[1]]
-
-dbh <- 372.4226
-
-# Year
-n.vec <- seq_len(nrow(myrwl))
-
-# Series
-seq_len(ncol(myrwl))
-
-# Select a ring-width series
-dat <- myrwl[[1]] 
-# Remove NA
-dat2 <- na.omit(dat)
-
-ggplot(s1, aes(y=SNA0101, x=row.names(s1))) + 
-  geom_point() + 
-  geom_line(b1, aes(y=SNA0101, x=row.names(b1)))
-
-ggplot(s1, aes(y=SNA0101, x=row.names(b1), group=1)) + geom_line()
-
-x <- c(1,2,2,3,1,3,4,5)
-
-xx <- rollmean(x=x, k=3, align = 'center', na.pad = T)
-
-# tes plots
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-carwi <- detrend(rwl = ca_tree, method = "ModNegExp")
-ca_prwi <- data.frame(rw = rowMeans(carwi, na.rm=TRUE))
-ca_prwi <- ca_prwi %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-xx <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-plot(xx$rw~rownames(xx))
-
-sjcrn <- chron(sjrwi, prefix = "rw_")
-plot(sjcrn, add.spline=TRUE, nyrs=20)
-plot()
-
-sj_tree <- rw_byTree(sj)
-ca_tree <- rw_byTree(ca)
-
-rownames(ca_tree) <- ca_tree$year
-ca_tree <- ca_tree[,-1]
-ca_p <- data.frame(rw = rowMeans(ca_tree, na.rm=TRUE))
-
-ca_p <- ca_p %>% mutate(year = rownames(.),
-                        site = as.factor('ca'),
-                        rw_log = log(rw))
-
-rownames(sj_tree) <- sj_tree$year
-sj_tree <- sj_tree[,-1]
-sj_p <- data.frame(rw = rowMeans(sj_tree, na.rm=TRUE))
-
-sj_p <- sj_p %>% mutate(year = rownames(.),
-                        site = as.factor('sj'),
-                        rw_log = log(rw))
-
-xx <- rbind(ca_p, sj_p)
-
-xx %>% ggplot(aes(x=as.Date(year, format='%Y'), y=rw_log)) + 
-  geom_line(aes(group=1)) + 
-  scale_x_date(date_labels = "%Y") + 
-  facet_wrap(~site, ncol=1, scales = 'free_y') + 
-  theme_bw() + 
-  geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-xx %>% filter(site =='ca') %>% 
-  ggplot(aes(x=as.Date(year, format='%Y'), y=rw)) + geom_point(colour='blue') +
-  geom_line(aes(group=1), colour='blue') + 
-  scale_x_date(date_labels = "%Y") + 
-  theme_bw() + 
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width (1/100 mm)')
-
-ca_prwi %>%
-  ggplot(aes(x = as.Date(year, format = '%Y'), y = rw)) + geom_point() +
-  geom_line(aes(group = 1)) +
-  scale_x_date(date_labels = "%Y") +
-  theme_bw() +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2005-06-01')))) +
-  # geom_vline(xintercept = as.numeric(as.Date(c('2012-06-01')))) +
-  xlab('Year') + ylab('Ring width Index')
-
-```

### 2018-11-26 man/figures/bai_cronos.Rmd
update figures and chronos

@@
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *green*) and southern ones: low-elevation (CA-Low; *pink*) and high-elevation (CA-High, *purple*) sites.
-Shading areas correspond to standard error of the mean.
-Number of series are displayed in the upper plot.
-We only show years replicated with # series > 5.
-Linear trend since 1975 is shown for southern high-elevation site (CA-High) .AREGLAR ESTO
-##### 
-```{r, warning=FALSE, echo=FALSE}
+```{r data}
 cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')
-
-colores <- c('#7570b3','#e7298a','#1b9e77')
-# colores <- c('#33a02c', '#984ea3', '#a65628')
 
@@
 mCaH <- lm(bai_mean ~ year, dfCaH)
-summary(mCaH)
 r2caH <- format(summary(mCaH)$r.squared, digits = 3)
@@
 mCaL <- lm(bai_mean ~ year, dfCaL)
-summary(mCaL)
 r2caL <- format(summary(mCaL)$r.squared, digits = 3)
@@
 mSJ <- lm(bai_mean ~ year, dfSJ)
-summary(mSJ)
 r2SJ <- format(summary(mSJ)$r.squared, digits = 3)
 pSJ <- lmp(mSJ)
+
+lms <- data.frame(year = as.numeric(c(2018, 2018, 2018)),
+                     site = c("CA-Low", "CA-High", "SJ"), 
+                     r2 = as.numeric(c(r2caL, r2caH, r2SJ)),
+                     pvalue = as.numeric(round(c(pCaL, pCaH, pSJ), 4)),
+                     y = as.numeric(c(as.numeric(predict(mCaL)[40]), 
+                           as.numeric(predict(mCaH)[40]), 
+                           as.numeric(predict(mSJ)[40]))))
+lms$y <- lms$y / 100
+lms <- lms %>% mutate(
+  pc = ifelse(pvalue < 0.001, "p < 0.001", as.character(paste0("p = ", pvalue))),
+  label = round(r2,3))
+
 ```
 
-```{r, warning=FALSE, out.extra='angle=90'}
+```{r plot_configure}
+# colores <- c("#B2DF8A","#1F78B4","#A6CEE3")
+# colores <- c('#7570b3','#e7298a','#1b9e77')
+# colores <- c('#33a02c', '#984ea3', '#a65628')
+colores <- c("#000000", "#0072B2", "#D55E00") 
+```
+
+```{r plot_cronobai}
+colores <- c("#B2DF8A","#1F78B4","#A6CEE3")
+colores <- c("#000000", "#0072B2", "#D55E00")
 
 # cores 
-cronos_bai <- cro_sites %>% filter(n_trees >5) %>% 
+cronos_bai <- cro_sites %>% filter(nseries > 5) %>% 
   ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
@@
                   ymax = (bai_mean + bai_se)/100,
-                  fill=site), alpha=.2, colour=NA) +
+                  fill=site), alpha=.4, colour=NA) +
   geom_line() + 
   geom_point(size=.8) + 
-  theme(panel.grid = element_blank(),
-        axis.text.x=element_text(size=6),
-        legend.position = c(0.1,.8),
-        strip.background = element_blank()) + 
   scale_colour_manual(values = colores) +
@@
   ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020)) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020), sec.axis = dup_axis()) +
   geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) +
+              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5) +
   geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-Low'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE) +
+              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5) +
   geom_smooth(data = subset(cro_sites, year > 1975 & site == 'SJ'), 
-              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE)
+              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5)  +
+  geom_text_repel(data = lms, 
+                  aes(x=year, y = y, colour = site, label = label),
+                  nudge_x = Inf, 
+                  direction = "x", show.legend = FALSE, 
+                  size = 3) + 
+  theme(panel.grid = element_blank(),
+        axis.text.x=element_text(size=6),
+        legend.position = c(0.2,.8),
+        strip.background = element_blank(),
+        axis.title.x.top = element_blank(), 
+        axis.text.x.top = element_blank())  
+```
 
+```{r plot_samp}
   # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
@@
   geom_line() + 
+  scale_colour_manual(values = colores) +
+  scale_fill_manual(values = colores) +
+  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020), 
+                     sec.axis = dup_axis(), position = "top") +
   theme(panel.grid = element_blank(),
@@
         axis.title.x=element_blank(),
-        axis.text.x=element_text(size=6),
+        axis.text.x.top = element_text(size=6), 
+        axis.text.x.bottom = element_blank(), 
         # axis.text.x=element_blank(),
@@
         # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
-        ) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
-  scale_x_continuous(breaks = seq(1815, 2020, by=10))  
+        ) 
   # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
   #       axis.title.y = element_text(vjust =0.25))
+```
 
+```{r plot_all}
 plot_bai_s <- samp + cronos_bai + plot_layout(ncol=1, heights = c(0.15, 0.85))
+
+```
+
+**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
+pyrenaica* for northern population (SJ; *orange*) and southern ones: low-elevation (CA-Low; *blue*) and high-elevation (CA-High, *black*) sites.
+Shading areas correspond to standard error of the mean.
+Number of series are displayed in the upper plot.
+We only show years replicated with # series > 5.
+Linear trends since 1975 are shown all sites (numbers indicate $r^2$ values).
+
+```{r out_plot, fig.height = 5, fig.width=7, dpi = 300, out.width = 0.8}
 plot_bai_s

### 2018-11-26 man/figures/corr_plot_years.Rmd
update figures and chronos

@@
 output: 
-  word_document:
+  word_document
 ---
@@
 ```{r, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE, dpi = 200)
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r}
+```{r load_pkg}
 library("tidyverse")
@@
 library("grid")
-library("GGally")
+# library("GGally")
 library("gridExtra")
@@
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
-##### 
-**Appendix S5.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
-
-```{r}
-
+```{r data}
 corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
@@
 cor_rse <- generaMatrix(corEvi, 'rs')
+```
 
+```{r plot_function}
 correP <- function(df, mytitle, ...){
@@
                   ggtheme = theme_minimal, 
-                  lab = TRUE, lab_size = 12,
+                  lab = TRUE, lab_size = 5,
                   title = mytitle, show.legend = FALSE, 
-                  tl.cex = 10)
-  p <- p + theme(plot.title = element_text(size = 14, face = "bold.italic"))
+                  tl.cex = 7,
+                  tl.srt = 1)
+  p <- p + theme(plot.title = element_text(size = 10, face = "bold.italic"),
+                 panel.grid = element_blank(),
+                 axis.text.x = element_text(hjust = 0.5)) + scale_x_discrete(position = "top") 
   p      
 }
+```
 
+```{r plot}
 g_rt <- correP(cor_rt, 'BAI Rt')
@@
 
-(g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
+g <- (g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
 ```
+
+**Figure S5.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
+Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
+
+```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8} 
+g
+```

### 2018-11-26 man/figures/correla_boot_sites.Rmd
update figures and chronos

@@
 
-```{r, echo=FALSE}
-show_text <- FALSE
-```
-
 **Figure S6.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
@@
 
-```{r}
+```{r data}
 correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)
@@
       name_comparison == "sj-caH" ~ "SJ - CAHigh")))
+```
 
+```{r plot}
 correla_sitesSJCALH  <- correla %>% 
@@
   ggplot(aes(x=size, y=estimate)) + 
-  geom_line(aes(colour=name_comparison), size = 1.4) + 
+  geom_line(aes(colour=name_comparison), size = 1) + 
   theme_bw() + ylab('Correlation coefficient') +
@@
   scale_color_colorblind() +
-  theme(legend.position = c(.8,.9), 
+  theme(legend.position = c(.8,.87), 
         legend.title = element_blank(), 
-        legend.text = element_text(size = 14),
+        legend.text = element_text(size = 6),
         panel.grid = element_blank(), 
-        axis.title = element_text(size = 14), 
-        axis.text = element_text(size = 11))
+        axis.title = element_text(size = 10), 
+        axis.text = element_text(size = 7),
+        legend.background = element_blank())
+```
 
+```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8} 
 correla_sitesSJCALH
-
 ```

### 2018-11-26 man/figures/gc_perc_trees.Rmd
update figures and chronos

@@
            stat = "identity", position = "dodge") +
-  geom_line(aes(x=year, y=total_tree), colour = 'black', size = 0.4) + 
+  geom_line(aes(x=year, y=total_tree*2), colour = 'black', size = 0.2) + 
   theme_bw() + ylab('% of trees') + xlab("Year") + 
@@
   scale_y_continuous(limits = c(-100, 100), breaks = seq(-100, 100, by = 50),
-                     sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
+                     sec.axis = sec_axis(~., name = '# trees', breaks = c(0,15,30))) +
   facet_rep_wrap(~site_sorted, nrow = 3, 

### 2018-11-26 script/R/chrono_bai.R

@@ -22,7 +22,7 @@ chrono_bai <- function(bai_object){
   out <- b_aux %>% 
     group_by(year) %>% 
     summarise(bai_mean = mean(bai, na.rm = TRUE), 
-              bai_sd = sd(bai), 
+              bai_sd = sd(bai, na.rm = TRUE), 
               bai_se = bai_sd / sqrt(n()), 
               n_trees = n()) %>% 
     as.data.frame()

### 2018-11-26 script/R/rw_byTree.R

@@ -29,16 +29,18 @@ rw_byTree <- function(rwdf, snc, locname){
     # Create a variable name for the average by tree 
     name_avg <- paste0(stringr::str_sub(i, star=s_site+1, end=s_site+s_tree))
     
+
     # Create aux dataframe with rwl mean values and the year
     aux <- rwdf %>% 
       dplyr::select(starts_with(i)) %>% 
-      mutate_(.dots = setNames(list(~rowMeans(., na.rm=TRUE),
-                                    ~as.numeric(rownames(.))),
+      mutate_(.dots = setNames(list(~rowMeans(., na.rm=TRUE), # rw mean
+                                    ~as.numeric(rownames(.))), # year
                                nm=c(name_avg, "year"))) %>% 
       dplyr::select(-starts_with(i))
     
     # Inner join with output by year  
     output_avg <- output_avg %>% inner_join(aux, by='year')
+    
   }
   
   # Change NaN by NA
@@ -51,6 +53,10 @@ rw_byTree <- function(rwdf, snc, locname){
   
   # Add locname 
   names(output_avg) <- paste0(locname, names(output_avg))
+  
+  # Add Nseries 
+  output_avg$nseries <- rowSums(!is.na(rwdf))
+  
   output_avg
   
   

### 2018-12-10 analysis/legacy_effects.Rmd
add legacy years

@@
 # Boostrap Confidence Intervals 
+set.seed(1234)
 
@@
 ```
+
+```{r}
+ylegacy <- legacy %>% 
+  filter(!(year %in% seq(2005+1, 2005+6, by=1))) %>% 
+  filter(!(year %in% seq(2012+1, 2012+6, by=1))) %>% 
+  group_by(ypd) %>%
+  summarise(mean=mean(lag, na.rm=T),
+            sd = sd(lag, na.rm=T),
+            se = sd/sqrt(length(lag))) %>% 
+  mutate(d = "all")
+
+ly2005 <- legacy %>%
+  filter(year %in% seq(2005+1, 2005+6, by=1)) %>% 
+  dplyr::select(-site) %>% rename(site = site_sorted) %>% 
+  group_by(ypd) %>%
+  summarise(mean=mean(lag, na.rm=T),
+            sd = sd(lag, na.rm=T),
+            se = sd/sqrt(length(lag))) %>% 
+  mutate(d = "2005")
+
+ly2012 <- legacy %>%
+  filter(year %in% seq(2012+1, 2012+6, by=1)) %>% 
+  dplyr::select(-site) %>% rename(site = site_sorted) %>% 
+  group_by(ypd) %>%
+  summarise(mean=mean(lag, na.rm=T),
+            sd = sd(lag, na.rm=T),
+            se = sd/sqrt(length(lag))) %>% 
+  mutate(d = "2012")
+
+ly <- rbind(ylegacy, ly2005, ly2012)
+
+p2 <- ly %>% ggplot(aes(x = ypd, y = mean)) + 
+  # geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci, fill=site), alpha=.5) + 
+  geom_line(aes(colour=d)) + 
+  geom_point(aes(colour=d), size=3) + 
+  geom_errorbar(data = (dplyr::filter(ly, d == "all") %>% as.data.frame()), 
+                aes(ymin=mean - se, ymax = mean + se, colour = d), width = 0.2) + 
+  theme_bw() + 
+  geom_hline(yintercept = 0, colour = 'black') + 
+  theme(panel.grid = element_blank(), 
+        strip.background = element_blank()) +  
+  scale_x_continuous("Years post drougth event", breaks=c(1:6)) + 
+  scale_y_continuous("Observed - Predicted Growth", breaks=seq(-0.4, 0.2, by=0.05))
+
+pdf(here::here('/out/legacy', 'legacy_droughts_2005_2012.pdf'), width=8, height = 5)
+p2
+dev.off()
+
+```

### 2018-12-27 analysis/legacy_effects.Rmd
add legacy effects

@@
   md_document:
-    variant: markdown_github
-bibliography: ../references.bib
-csl: ../ecology.csl
+    variant: gfm
+bibliography: ../refs/references.bib
+csl: ../refs/ecology.csl
 editor_options: 
@@
 library("pander")
-library("rms")
+# library("rms")
 library("magrittr")
@@
 ```
+
+## Legacy effects for all Q. pyrenaica forests
+
+```{r}
+all_legacy <- legacy %>% group_by(ypd) %>%
+  summarise(mean=mean(lag, na.rm=T),
+            sd = sd(lag, na.rm=T),
+            se = sd/sqrt(length(lag)))
+
+po <- position_dodge(0.9)
+all_legacy %>% 
+  ggplot(aes(x = ypd, y = mean)) + 
+  geom_line(position = po, alpha=0.8) + 
+  geom_point(position = po) + 
+  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), position = po) +
+  theme_bw() + 
+  geom_hline(yintercept = 0, colour = 'black') + 
+  theme(panel.grid = element_blank()) + 
+  scale_x_continuous("Years post drougth event", breaks=c(1:6)) + 
+  scale_y_continuous("Observed - Predicted Growth")
+
+# Boostrap Confidence Intervals 
+set.seed(1234)
+
+all_legacy_nest <- legacy %>% 
+  group_by(ypd) %>% 
+  nest()
+
+# Boostrap the mean 
+## boot_mean
+boot_mean <- function(d, i) {
+    mean(d[i])
+}
+
+## Use purrr::map to add another column to legacy_nest, in which each row contains the object returned by mapping the boot::boot function (which is responsible for applying the bootstrap_mean function) over the tibbles stored in each row of the “data” list-column.
+
+all_legacy_nest %<>% 
+  dplyr::mutate(booted = purrr::map(.x = data, # The list-column containing <S3: tibble>
+                                      ~ boot::boot(data = .x$lag, # The <S3 tibble> column being sampled
+                                                 statistic = boot_mean, # The user-defined function
+                                                 R = 10000, # The number of replicates
+                                                 stype = "i")))
+
+# compute the Confidence interval 
+all_legacy_nest %<>%
+    dplyr::mutate(booted_ci = purrr::map(.x = booted, # The list-column containing <S3: boot> objects
+                                         ~ boot::boot.ci(.x,
+                                                         conf = 0.95, # Interval width
+                                                         type = "bca")))  # Calculate a BCa interval
+# Tidy the CI 
+all_legacy_CI <- all_legacy_nest %>% 
+  dplyr::mutate(statistic = purrr::map(.x = booted_ci, # The list-column containing <S3 bootci> objects
+                                         ~ .x$t0), # The point estimate
+                  lower_ci = purrr::map(.x = booted_ci,
+                                        ~ .x$bca[[4]]), # The value of the lower 2.5% limit
+                  upper_ci = purrr::map(.x = booted_ci,
+                                        ~ .x$bca[[5]])) %>% # The value of teh upper 97.5% limit
+    # Drop the list-columns (no longer needed)
+    dplyr::select(-data, -booted, -booted_ci) %>%
+    # Unnest the dataframe
+    tidyr::unnest()
+
+# Join to mean values
+all_legacy %<>% inner_join(all_legacy_CI, by="ypd")
+
+g <- all_legacy %>% ggplot(aes(x = ypd, y = mean)) + 
+  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha=.5) + 
+  geom_line() + 
+  geom_point(size=3) + 
+  theme_bw() + 
+  geom_hline(yintercept = 0, colour = 'black') + 
+  geom_line(data = (all_legacy %>% mutate(ac = cumsum(mean))), aes(y = ac, x = ypd), lty = 2) +
+  theme(panel.grid = element_blank(), 
+        strip.background = element_blank(), 
+        legend.position = "none") + 
+  scale_x_continuous("Years post drougth event", breaks=c(1:6)) + 
+  scale_y_continuous("Observed - Predicted Growth", breaks=seq(-0.2, 0.2, by=0.05))
+g
+
+# dashed line indicates integrated legacy effects (sensu Anderegg et al. 2015; i.e: sum of legacy effects)
+
+# Integrated legacy effects (sensu Anderegg et al. 2015)
+sum(all_legacy$mean)
+```
+
+## Legacy effects for site
 
@@
   geom_line(data = l2012, aes(x = ypd, y =  lag), linetype = "dashed") + 
+  geom_line(data = (mlegacy %>% group_by(site) %>% mutate(ac = cumsum(mean))), 
+            aes(y = ac, x = ypd, colour = site), lty = "dotdash") + 
   facet_wrap(~site, labeller = as_labeller(label_site)) + 

### 2018-12-31 man/ms.Rmd
update figure numbers

@@
     highlight: null
-    reference_docx: ../../templates/template_ms.docx
-bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
+    reference_docx: ../templates/template_ms.docx
+bibliography: ../refs/references.bib
+csl: ../refs/ecology.csl
 ---
@@
 
-## Species and study site
+## Tree species and study site
 
@@
 During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*.
-1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Appendices S1-S3).
+1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
 We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
 Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term.
-The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015 (see Appendix S3).
+The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015 (see Table S3).
 
@@
 In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q.
-pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Appendix S4).
+pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Figure 3).
 
@@
 `END REVIEW` For southern population differences in growth were observed.
-CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figure 4; Appendix S7).
+CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figures 4 and S6).
 
 For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low).
-Yet, the lowest pointer year (*i.e.* the lowest growth) since 1950 was 1995 in all chronologies (Figure 5).
+Yet, the lowest pointer year (*i.e.* the lowest growth) since 1950 was 1995 in all chronologies (Figure S3).
 
@@
 
-The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8).
-Northern site (SJ) showed two release events ($GC$ > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001.
-These periods alternate with periods of supression.
-Southern sites (CA-High and CA-Low) showed some weak suppression episodes, but not in the last 50 years.
-
-The analysis of growth changes revealed differences in forest history between sites (Figures 7 and 8).
+The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4).
 Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001.
@@
 
+The strength of the response of the resilience indices was related to drought severity.
+The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data `REVIEW`(1900 - 2017), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Figure 6).
+A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155).
+Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Figure 6) even not being the most severe drought events.
+`REVIEW` (CONEXION con el siguiente parrafo)
+
 *Q.
-pyrenaica* forest showed significantly lower resistance values (*Rt*) to the 2005 drought event than to that in 2012 for greenness and for radial growth (Table 3; Figure 3).
+pyrenaica* forest showed significantly lower resistance values (*Rt*) to the 2005 drought event than to that in 2012 for greenness and for radial growth (Table 3; Figure 7).
 The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2).
@@
 Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2).
-It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).
-
-Recovery (*Rc*) of greenness and growth were significantly different bewteen drought events and sites (Tables 3).
-In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
-An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 3, Table S2).
+It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 7).
+
+Recovery (*Rc*) of greenness and growth were significantly different bewteen drought events and sites (Table 3).
+In the 3-year period after the 2005 drought, greenness was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
+An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 7, Table S2).
 Northern populations showed significantly higher values of recovery than southern sites for greenness and tree-growth (Table S2).
-For southern populations, no significant differences were found for recovery of greenness after the 2005 and 2012 drought event (p = 0.2453; Figure 3; Table S1).
-Recovery values for tree-growth of southern populations were below or close to 1 (Figure 3, Table S2).
-
-Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 3).
+For southern populations, no significant differences were found for recovery of greenness after the 2005 and 2012 drought event (p = 0.2453; Figure 7; Table S1).
+Recovery values for tree-growth of southern populations were below or close to 1 (Figure 7, Table S2).
+
+Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 7).
 Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3).
 Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2).
-For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 3).
+For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 7).
 For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2).
-
-The strength of the response of the resilience indices was related to drought severity.
-The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data `REVIEW`(1900 - 2017), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Appendix S5).
-A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155).
-Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Appendix S5) even not being the most severe drought events.
 
@@
 Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
-Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 6a).
+Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 8a).
 Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
-Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 6b), specially for the northern population (r > 0.6).
+Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), specially for the northern population (r > 0.6).
 Relationships with temperatures were lower than those with moisture variables.
-Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 6c and 6d).
+Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figure 8b and 8c).
 Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population.
@@
 pyrenaica* forests, but no changes in EVI standardized anomalies were recorded for 2012, which can be explained because 2012 drought event was a winter-drought [@Trigo2013].
-Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure 5).
+Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure S3).
 The decline in growth observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].
@@
 ## Oaks show high resilience in response to recent drought events
-Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Appendix S3), we found a positive trend for vegetation greenness of the forests of *Q.
+Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of the forests of *Q.
 pyrenaica* located in their rear edge.
@@
 
-Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 3).
-La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Appendix S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q.
+Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 7).
+La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Table S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q.
 pyrenaica*.
 Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía (long-term perspective), observamos como estas poblaciones presentan altos valores de resiliencia.
-Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).
+Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Figure 6).
 
 Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q.
-pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Appendix S5).
+pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Figure 6).
 Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005].
@@
 Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3).
-Las poblaciones del norte, que presentan condiciones mas secas (Tabla 1), se vieron mas afectadas por los eventos de sequía.
+Las poblaciones del norte, que presentan condiciones mas secas (Table 1), se vieron mas afectadas por los eventos de sequía.
 For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2).
-In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 6), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
+In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 8), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
 
@@
 pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014].
-We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 6a).
+We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a).
 Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]

### 2019-01-28 man/ms.Rmd
update changes intro

@@
 Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011].
-It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017].
-Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for at least the last three decades [@Spinoni2017b].
-Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013], specially in the Mediterranean area [@Spinoni2017a].
+It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Dai2011; @Ummenhofer2017].
+Despite drought is considered a feature of Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events has been recorded in the last decades [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported [@Spinoni2017b].
+This is especially important, given the climate change projections for the Mediterranean area that forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Trenberth2014; @Spinoni2017a].
 
 Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].
-In the last decades the assessment of the drought impacts on ecosystems has gained attention [@Allen2010; @Clark2016], especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014].
-The functional response of vegetation to drought has been analyzed at different scales, from global [*e.g*.
-@Allen2010; @VicenteSerrano2013; @Norman2016] to local scales [see @MartinezVilalta2016 for a revision].
-
-Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017], and some authors states that the impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017].
-This is especially relevant for areas with a long history of landscape modification such as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016].
-`REVIEW` For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.
-
-The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005].
+The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global [*e.g.*, @VicenteSerrano2013; @Norman2016] and at local scales [see @MartinezVilalta2016 for a revision].
+
+Ecosystem responses to drought are influenced by other drivers of environmental changes [@Fischer2006; @Oliver2014; @Penuelas2017].
+In a global-change context, the impacts of drought on vegetation need to be evaluate considering the interactions with other drivers of change [@Clavero2011; @DoblasMiranda2017].
+This is especially relevant for areas with a long history of landscape modification, such us Mediterranean region, where land-use change is a key driver [@Navarro2013; @Ameztegui2016], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].
+
+The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate [@Hampe2005].
 The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010].
@@
 
-Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution `REVIEW`[@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
-Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015].
-The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services.
+Several studies analyzed the drought effects on Mediterranean tree species representing the southernmost limit of their distribution
+[*e.g.*, @SanchezSalguero2012; @Linares2014; @Dorado2017b; @Caminero2018].
+Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
+The assessment of resilience --*i.e.* the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Hodgson2015]-- to climate disturbances provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services.
 This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.
@@
 Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth.
-Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
-Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016].
-Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Zribi2016].
+Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.*, @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
+Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016].
+Drought impacts on vegetation using remote sensing have been analyzed at different scales.
+For instance, a global-scale study showed a stronger influence of the drought for drier biomes [@VicenteSerrano2013].
+For Mediterranean, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].
 
 Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
+
+The combination of both approaches, namely remote sensing and dendroecology, has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]
+
 Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.
@@
 In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened.
-We are also interested in assessing resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.
-We hypothesized that, relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts.
-Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q.
-pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to successive extreme drought events, both in recent times and in the long-term using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change.
+We also assess the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.
+Our hypothesis is that these relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts.
+Our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of Q.
+pyrenaica forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change.
 In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

### 2019-02-05 analysis/evi_vs_severe_droughts.Rmd
compute seasonality of droughts

---
title: "Growing season vs. Drought"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Feb"
output:  
  md_document:
    variant: gfm
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
```

```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("here")
library("lubridate")
library("anytime")
library("rdryad")
```

```{r}
# Datos de iv 
iv <- read.csv(here::here("data/evi/iv_composite.csv"), header = TRUE, sep = ',')

evi_profile_dat <- iv %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=T),
            sd = sd(evi, na.rm=T),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d'))

```

```{r}
# datos de Drought
dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)

# Convertir en datos de fecha inicio y fecha fin y convertir en falsa fecha

dy <- dy %>% 
  filter(d_duration > 2) %>%  
  separate(rangeDate, c("init", "end"), sep=" - ", remove = FALSE) %>% 
  mutate(miny = ifelse(minyear %in% c(1913,1998), 2018, 2019), 
         maxy = ifelse(maxyear == 1946, 2020, 2019),
         iD = anydate(paste(miny, init, sep="-")),
         eD = anydate(paste(maxy, end, sep="-"))) %>% 
  mutate(r = rank(d_severity, ties.method = "first")) %>% 
  mutate(mean = rescale(r, to =c(0.2, 0.33))) %>% 
  mutate(year = ifelse(minyear == maxyear, minyear, paste(minyear,maxyear, sep="-")))

```

```{r}
micolor <- '#455883'

evi_profile <- ggplot(evi_profile_dat, aes(cd, y=mean)) + 
  geom_errorbar(aes(ymin = mean - 10*se, ymax= mean + 10*se), width=4, colour=micolor, size=.8) + 
  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') +
  geom_line(colour=micolor, size=.8) + 
  geom_point(size=3,colour=micolor) +
  geom_point(size=1.5, colour='white')+
  scale_x_date(labels = function(x) format(x, "%b"),
               breaks = date_breaks('month')) + 
  ylab('EVI') + xlab('Date') + 
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank()) +
  theme_classic()
#format(x, "%d-%b")

combinado <- evi_profile + 
  geom_segment(data = dy, aes(y = mean, x= iD, xend =eD, yend = mean, colour = d_severity), 
               size = 8, alpha = 0.7) +
  geom_text(data = dy, aes(label = year, x = (iD + floor(eD-iD)/2))) +
  scale_colour_gradient(low = "orange", high = "red") + 
  labs(colour = "Drought Severity")
```

```{r}
# PLOT EVI + DROUGHT 
combinado2 <- ggplot(evi_profile_dat, aes(cd, y=mean)) + 
  scale_x_date(labels = function(x) format(x, "%b"), breaks = date_breaks('month')) + 
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank()) +
  theme_classic() +
  geom_segment(data = dy, aes(y = mean, x= iD, xend =eD, yend = mean, colour = d_severity), 
               size = 8, alpha = 0.7) +
  geom_text(data = dy, aes(label = year, x = (iD + floor(eD-iD)/2))) +
  scale_colour_gradient(low = "orange", high = "brown") + 
  labs(colour = "Drought Severity") +
  geom_errorbar(aes(ymin = mean - 10*se, ymax= mean + 10*se), width=4, colour=micolor, size=.8) + 
  geom_line(colour=micolor, size=.8) + 
  geom_point(size=3,colour=micolor) +
  geom_point(size=1.5, colour='white') + 
  ylab('EVI') + xlab('Date')

pdf(here::here('/out/drought_evi', 'evi_droughts.pdf'), width=8, height = 5)
combinado2
dev.off()

# Get data from Dryad 
# Peres de Lis 2016 et al. https://datadryad.org/resource/doi:10.5061/dryad.1cn19
# x <- dryad_files('10.5061/dryad.1cn19')

```

### 2019-02-05 man/figures/std_anomalies.Rmd
add more ticks

--- 2018-11-23 man/figures/std_anomalies.Rmd
@@
   word_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error.
+**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error. See main text for details on EVI calculation.
 
@@
   scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  ylab(myylab) + xlab('') +
+  scale_x_continuous(breaks = seq(2000, 2016, by=2)) +
+  ylab(myylab) + xlab('Year') +
   theme_bw() +
-  theme(text = element_text(size=12),
+  theme(axis.text = element_text(size=8),
         legend.position = "none",
-        panel.grid.major.x = element_blank(),
-        panel.grid.minor = element_blank(),
-        strip.background = element_rect(fill = "white"),
-        plot.title = element_text(hjust = -0.05, size = 12, 
+        panel.grid = element_blank(),
+        strip.background = element_rect(fill = "transparent", colour = "black", size = 0.2),
+        plot.title = element_text(hjust = -0.05, size = 9, 
                                     margin = margin(t=-5, b=1)))

### 2019-02-05 man/figures/ring_climate.Rmd
update figures and chronos
add tick marks

@@
   scale_y_continuous(limits = c(-0.4, 0.4), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4)) +
+                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4), 
+                     sec.axis = dup_axis(name = NULL)) +
   scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'bottom') + 
@@
   scale_y_continuous(limits = c(-0.4, 0.4), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4)) +
+                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4), 
+                     sec.axis = dup_axis(name = NULL)) +
   scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'top') + 
@@
   scale_y_continuous(limits = c(-0.2, 0.8), 
-                     breaks = c(-0.2, 0, 0.2, 0.4, 0.6, 0.8)) +
+                     breaks = c(-0.2, 0, 0.2, 0.4, 0.6, 0.8), 
+                     sec.axis = dup_axis(name = NULL)) +
   scale_x_continuous(breaks = labelitas$id, labels = labelitas$month_name, position = 'top') + 
@@
 
-```{r plot_out, fig.height=6, dpi = 300, out.width = 0.8} 
+```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
 ring_climate

### 2019-02-05 man/ms.Rmd
update

@@
 title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
-author: "AJ Perez-Luque; G. Gea-Izquierdo; ..."
+author: "AJ Perez-Luque; G. Gea-Izquierdo; Zamora, R. ..."
 output:
@@
 The trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2).
-`REVIEW meter en Discussion` This can be likely explained as a consequence of their different management origin (*i.e.* land-use legacy), as shown by the release event in SJ suggesting their coppice origin and (references).
-`END REVIEW` Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4).
+Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4).
 `REVIEW` In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations.
@@
 For 2005 drought event, *Rs* value of SJ (northern site) was lower than that of southern ones (CA-High and CA-Low), but opposite pattern was found for the 2012 drought event.
-All values of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031).
+All valu`es of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031).
 
@@
 # Discussion
-## Impacts of severe droughts on rear-edge populations
-
-Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction in greenness and especially in the growth of *Q.
-pyrenaica*.
+## Rear-edge oaks show high resilience in response both to recent drought events and long-term climatic variability.
+
+Our findings show that severe droughts, such us 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q.
+pyrenaica* (Figures 2 and 4).
+These results are consistent with previous studies that reported reductions of tree-growth of this species during extreme drought events, particularly during 1995 drought [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018].
+At this pointer-year, we observed the greatest reduction of tree-growth in our study sites, in concordance with findings reported for several Mediterranean-tree species [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].
+
+Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
+pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period).
+This confirms previous findings which suggests an increase in primary productivity for this forests [@PerezLuque2015onto], considered the most productive ecosystems in this mountain region [@Alcaraz2016obsnev_ndvi].
+For tree-growth, positive trends were also observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
+Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014].
+This pattern contrasts with the decline trends observed for other mediterranean [*e.g.*, *Pinus nigra*; *P.
+sylvestris*; @SanchezSalguero2012; @Camarero2015] and temperate [*e.g*, *Fagus sylvatica*; @Dorado2017] tree-species located in their rear-edges.
+
+Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2).
+Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005.
+These results could be explained by the different seasonality of the two droughts, *i.e.*, the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought.
+The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q.
+pyrenaica* forests [@PeresLis2017].
+Algunos autores han apuntado la importancia de los legacies (i.e.
+cuanto tiempo tardan en recuperarse los individuos tras una sequía) (Anderrbegg… CITAS).
+Nuestros resultados sugieren que otro aspecto adicional a considerar al analizar los impactos de las sequías sobre las especies de árboles es el momento en el que ocurre la sequía.
+
+A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6).
+A plausible explanation could be the water availability after the drought event.
+For instance, the months with the highest cumulative rainfall of the past century occured during the first months of 1996 in our southern site (Cáñar) [Torres-MESA]; and years after 1995 were within the wetter of the past 60 years.
+
+In addition, from a long-term perspective, we obtained a positive relation between tree-growth resilience and drought severity (Figure 6).
+Surprisingly, this result contradicts our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge.
+
+--> Aquí iría la idea de lo de refugio
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q. pyrenaica*.
+
 During the 2005 drought, one of the worst drought events recorded in the Iberian Peninsula [@GarciaHerrera2007], we found a browning of the *Q.
 pyrenaica* forests, but no changes in EVI standardized anomalies were recorded for 2012, which can be explained because 2012 drought event was a winter-drought [@Trigo2013].
+
 Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure S3).
@@
 
-## Oaks show high resilience in response to recent drought events
 Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of the forests of *Q.
@@
 Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Figure 6).
+") 
+```
 

### 2019-02-06 man/figures/resilience_metrics.Rmd
update figures
reduce fig size
update figures and chronos
update legend

@@
 **Figure 7.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree radial growth (BAI; right-plots) for the years 2005 and 2012.
+pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree growth (BAI; right-plots) for the years 2005 and 2012.
 For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared northern population (San Juan, SJ; *black triangle*) with southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
+For BAI we compared the northern population (San Juan, SJ; *black triangle*) with the two southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
 Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
 
-```{r plot_out, fig.height=6.5, dpi = 300, out.width = 0.8} 
+```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
 grid_arrange_shared_legend(rt, rc, rs, ncol = 1, nrow = 3, position ='top')

### 2019-02-06 man/figures/cumulative_prec.Rmd
fix errors
update S1 legend

@@
 
-**Appendix S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
+**Figure S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
 Points represent mean and errorbars standard error.
-*Black* line indicates mean for all period.
+*Black* line indicates mean for the whole period (`r meandf` mm).
 *Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).

### 2019-02-06 man/figures/plot_spei12.Rmd
update figures and chronos
add vertical lines

@@
         axis.text.x = element_text(size=9),
-        axis.title.y = element_text(size = 12))
+        axis.title.y = element_text(size = 12)) +
+  geom_vline(xintercept = as.Date(c("2005-01-01", "2012-01-01")), colour = "grey", alpha = 0.8)
 

### 2019-02-06 man/figures/gc_perc_trees.Rmd
add legend labels

--- 2018-11-23 man/figures/gc_perc_trees.Rmd
@@
         strip.background = element_blank(), 
-        strip.text = element_blank(),
+        strip.text = element_text(face = "bold", size = 12),
+        strip.placement = "inside", 
         axis.text.y.left = element_text(size=6),

### 2019-02-06 man/figures/evi_vs_severe_droughts.Rmd
update figures

--- 2019-02-05 analysis/evi_vs_severe_droughts.Rmd
@@
 ---
-title: "Growing season vs. Drought"
-author: "AJ Perez-Luque (@ajpelu)"
-date: "2017 Feb"
 output:
-  md_document:
-    variant: gfm
-bibliography: ../refs/references.bib
-csl: ../refs/ecology.csl
-editor_options: 
-  chunk_output_type: console
+  word_document
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecology.csl
 ---
 
-```{r global_options, include=FALSE}
-knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
+```{r, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
 ```
 
-```{r, warning=FALSE, message=FALSE}
+```{r load_pkg}
 library("tidyverse")
@@
 library("rdryad")
+library("scales")
 ```
 
-```{r}
+**Figure S7.** EVI annual profile (average of the period 2000-2016) for *Q.
+pyrenaica* forests in Sierra Nevada and drought events.
+Horizontal bars correspond to the most severe droughts for Sierra Nevada since 1900 (computed as in Table S3).
+Their position indicate the start and end months of each drought event.
+Bars length show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017]).
+
+```{r preparedata}
 # Datos de iv 
@@
 
-```
-
-```{r}
 # datos de Drought
@@
 
-```{r}
+```{r plotEVI, eval = FALSE, echo = FALSE}
 micolor <- '#455883'
@@
   geom_point(size=1.5, colour='white') + 
-  ylab('EVI') + xlab('Date')
-
-pdf(here::here('/out/drought_evi', 'evi_droughts.pdf'), width=8, height = 5)
-combinado2
-dev.off()
+  ylab('EVI') + xlab('')
 
@@
 # x <- dryad_files('10.5061/dryad.1cn19')
+```
 
+```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8}
+combinado2
 ```

### 2019-02-06 man/tables/t_droughts.Rmd
update figures

@@
 ##### 
-**Appendix S3.** Drought events for Sierra Nevada based on SPEI index (12 months scale).
+**Table S3.** Drought events for Sierra Nevada based on SPEI index (12 months scale).
 A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI value are below threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
+A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
 The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
@@
 dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
-dy <- dy %>% arrange(desc(d_duration))
+dy <- dy %>% arrange(desc(d_duration)) %>% 
+  arrange(desc(d_severity))
 # solamente aquellos mayor de 2 meses 

### 2019-02-06 man/tables/t_robust_anovas.Rmd
update figures

@@
 
-**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the two factors drought events (in 2005 and 2012) and site.
+**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the two drought events (in 2005 and 2012) and site.
 
@@
                            "disturb_year" = "Drought event", 
-                           "disturb_year:site" = "Drought event:Site", 
+                           "disturb_year:site" = "Drought event*Site", 
                            "site" = "Site")) %>% 
-  arrange(match(factores, c("Drought event", "Site", "Drought event:Site"))) %>% 
+  arrange(match(factores, c("Drought event", "Site", "Drought event*Site"))) %>% 
   mutate_at(c(2,4,6,8), round, 1) %>% 

### 2019-02-06 man/tables/t_sampled_plots.Rmd
update figures

@@
 Dbh and height of all trees, Basal Area (BA), Density and SRD (Size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see methods).
-Temp.: annual average of minimun and maximum temperatures.
+Temp.: annual average of mean monthly minimun and maximum temperatures.
 Values shown here correspond to site averages.
@@
              n_trees = "Cored trees", dn = "Cored trees", h = "Cored trees", age = "Cored trees", 
-             dn_all = "Competence", h_all = "Competence", ba = "Competence", 
-             stand_density = "Competence", srd = "Competence", top = TRUE) %>% 
+             dn_all = "Competition", h_all = "Competition", ba = "Competition", 
+             stand_density = "Competition", srd = "Competition", top = TRUE) %>% 
   merge_at(i = 1, j =c(8:11) , part = "header") %>% 

### 2019-02-06 man/ms.Rmd
update

@@
 Despite drought is considered a feature of Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events has been recorded in the last decades [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported [@Spinoni2017b].
-This is especially important, given the climate change projections for the Mediterranean area that forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Trenberth2014; @Spinoni2017a].
+This is especially important, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Trenberth2014; @Spinoni2017a].
 
@@
 In a global-change context, the impacts of drought on vegetation need to be evaluate considering the interactions with other drivers of change [@Clavero2011; @DoblasMiranda2017].
-This is especially relevant for areas with a long history of landscape modification, such us Mediterranean region, where land-use change is a key driver [@Navarro2013; @Ameztegui2016], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].
+This is especially relevant for areas with a long history of landscape modification, such us Mediterranean region, where land-use change plays a significant role [@Navarro2013; @Ameztegui2016], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].
 
@@
 Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
-The assessment of resilience --*i.e.* the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Hodgson2015]-- to climate disturbances provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services.
+Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Hodgson2015].
+The assessment of resilience to climate disturbances provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services.
 This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.
@@
 For instance, a global-scale study showed a stronger influence of the drought for drier biomes [@VicenteSerrano2013].
-For Mediterranean, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].
+For Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].
 
 Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
-
-The combination of both approaches, namely remote sensing and dendroecology, has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]
-
-Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.
+The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]
 
@@
 
-The forests of this species reaches their `REVIEW` southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*.
+The forests of this species reaches their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*.
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
@@
 For this reason, the populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], having these forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
-Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].`REVIEW CITA`
+Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].
 

### 2019-02-08 man/ms.Rmd
change

@@
 2005 and 2012 were considered as single drought events.
-The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drougth event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Appendix S6) and this length was used in other studies [*e.g.* @Gazol2017].
+The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively.
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017].
 To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure S5).
@@
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
-All analyses were carried out in R software [@R2017] using several packages (see Supplementary `REVIEW` S?)
+All analyses were carried out in R software [@R2017] using several packages (see Appendix S7).
 
@@
 pyrenaica* (Figures 2 and 4).
-These results are consistent with previous studies that reported reductions of tree-growth of this species during extreme drought events, particularly during 1995 drought [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018].
-At this pointer-year, we observed the greatest reduction of tree-growth in our study sites, in concordance with findings reported for several Mediterranean-tree species [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].
-
-Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
+These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018].
+The greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018].
+
+During the last decades, the south of Iberian Peninsula suffered two of the worst recorded drought [2005 and 2012; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3).
+However, we found a positive trend for vegetation greenness of *Q.
 pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period).
-This confirms previous findings which suggests an increase in primary productivity for this forests [@PerezLuque2015onto], considered the most productive ecosystems in this mountain region [@Alcaraz2016obsnev_ndvi].
-For tree-growth, positive trends were also observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
+This confirms previous findings findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi].
+For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
 Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014].
-This pattern contrasts with the decline trends observed for other mediterranean [*e.g.*, *Pinus nigra*; *P.
-sylvestris*; @SanchezSalguero2012; @Camarero2015] and temperate [*e.g*, *Fagus sylvatica*; @Dorado2017] tree-species located in their rear-edges.
+Our pattern differs from the decline trends observed for several mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
+sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].
 
@@
 Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005.
-These results could be explained by the different seasonality of the two droughts, *i.e.*, the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought.
+These results could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought.
 The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q.
 pyrenaica* forests [@PeresLis2017].
-Algunos autores han apuntado la importancia de los legacies (i.e.
-cuanto tiempo tardan en recuperarse los individuos tras una sequía) (Anderrbegg… CITAS).
-Nuestros resultados sugieren que otro aspecto adicional a considerar al analizar los impactos de las sequías sobre las especies de árboles es el momento en el que ocurre la sequía.
+Our results highlight the importance of the drought timing on the tree-growth responses, as have been indicated for other Mediterranean-tree species [*e.g.* *Q.
+ilex*, @Camarero2015b].
+
+As we mentioned above, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3).
+This pattern was also observed for several *Pinus* species in their rear edge [@SanchezSalguero2012; @Camarero2015].
+But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6).
+In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6).
+Surprisingly this result contradicts our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge.
+
+Several considerations have to be taken into account to explain this result.
+First of all we need to explore the cumulative precipitation of the previous and posterior years of these drought events.
+For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern.
+In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009].
+In addition, this argument could also explain the higher values of recovery found for 1995 (Figure 6).
 

### 2019-02-21 man/ms.Rmd
update

@@
 The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4).
-Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001.
+Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1995-2000.
 These periods alternate with periods of supression.
-Southern sites (CA-High and CA-Low) showed no release event except one for CA-High at the beginning of the `REVIEW`1840s?
-and no suppression event in the last 50 years.
+Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression event in the last 50 years.
 
@@
 pyrenaica* forests [@PeresLis2017].
-Our results highlight the importance of the drought timing on the tree-growth responses, as have been indicated for other Mediterranean-tree species [*e.g.* *Q.
-ilex*, @Camarero2015b].
+Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018],  that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].
 
@@
 In addition, this argument could also explain the higher values of recovery found for 1995 (Figure 6).
+
+--> Ver y resumir páginas 18-19 Notas Naturkunde
 

### 2019-03-14 man/ms.Rmd
update bib and ms

@@
 
-Although our sampled sited showed similar tree competence levels (Table 1), the review of historical documents showed that la historia de manejo humano ha sido diferente para cada sitio.
-Different values of tree-growth were observed between populations and for different elevations within southern location (Figure 4).
-Low-elevation sites showed lower BAI values than higher-elevation site.
-A reason that could explain these results is the different degree of anthropic impact related to elevation.
-Los bosques situados a mas elevación presentan menor presión antrópica que los localizados a menor altitud, por el simple hecho de que estos últimos están mas accesibles.
-During the last decades, the rural abandonment has lead an decrease of anthropic pressures on the high-elevations zones of Mediterranean mountains (INLCUIR CITA).
-
-comentario --> Aqui pretendo incluir la idea que hablamos en su momento.
-Quizá están creciendo donde les están dejando crecer (menor impacto antrópico en las zonas mas altas, al menos en los últimos años -abandono de zonas rurales?-)
-He ralizado una revisión de trabajos que utilicen dendro de Q. pyrenaica para ver si existe una relación entre la edad de las cronos con la elevación.
-
-* Son bosques relativamente jóvenes 
-comentaio --> 
-Aquí me falta desarrollar este apartado algo más incluyendo cosas de los documentos históricos (la minería, etc, etc) 
-Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5).
-La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona.
-Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.
+The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008].
+Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017, @MesaFernandez2018].
+From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018].
+Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada.
+As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia1991].
+
+Our study sites have been subjected to an intense exploitation of forest resources.
+Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before.
+For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015].
+For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings.
+Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@GomezCruz1992].
+These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
+@GeaIzquierdo2014; GeaIzquierdo2015; @Dorado2017].
+
+We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy).
+Our study sites had different land-use trajectories driven by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014].
+
+In Northern sites (*e.g.* San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO].
+Grasslands and shrublands for cattle farming are located at high elevations.
+Then forests formation with some croplands; and then, at lowlands, irrigated terraces with tree crops.
+Another activities could conditioning the natural resources uses.
+For instance, San Juan oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas), with two periods of intense activity, the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita].
+The mining activity in San Juan site have strongly conditioned the forest structure.
+
+On the other hand, woodland areas of the southern slopes (*e.g.* Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes)[@Calatrava2019].
+Forest resources (firewood, charcoal, acorns) were used in a continuously way, as some historical sources have pointed out [@ValbuenaCarabana2014, @JimenezSerrano2004].
+In fact, in Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] to late twenty century [@MesaTorres2009; @Bonet2014_conama].
+
+The different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5).
+We observed a tree-growth release event around 1940 in the SJ oak woodland (Figures 5, S4) which concurs with one period of mining activity for this area.
+Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections (CITA).
+During this period there were an increase on the use of timbers for the tunnels of the mines (CITE) and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area (cite).
+This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4).
+Our results also coincide with a concurrent but less intense tree-growth release event reported for a closed oak woodland [@GeaIzquierdo2014].
+In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017].
+
+The release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4).
+The revision of forests practices carried out in this area in the last 30 years did not reveal any clearing or cutting event for those years.
+Two non-mutually exclusive ways could explain the tree-growth release observed.
+Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004; @Gentilesca2017].
+On the other hand, the strong positive correlations of SPEI with tree-growth obtained for this site (Figure 8) indicates a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014].
+Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006].
+Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].
 

### 2019-03-28 man/ms.Rmd
update ms

@@
 Our hypothesis is that these relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts.
-Our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of Q.
-pyrenaica forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change.
+Our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q.
+pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change.
 In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
@@
 pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period).
-This confirms previous findings findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi].
+This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi].
 For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
@@
 
-As we mentioned above, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3).
-This pattern was also observed for several *Pinus* species in their rear edge [@SanchezSalguero2012; @Camarero2015].
+As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3).
+This pattern was also observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015].
 But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6).
 In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6).
-Surprisingly this result contradicts our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge.
+Surprisingly this result seems contradict our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge after extreme drought events.
+
+Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005].
+For instance, higher vulnerability to drought is ofted assumed for rear-edge populations [@MartinezVilalta2018], however some studies showed different findings [@Cavin2017; @Granda2017], as our results suggest, with high values of resilience for populations of * Q.
+pyrenaica * located in its rear-edge.
+
+This would be related with the definition of the marginal population.
+Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018].
+En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q.
+pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.
+Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
+pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017]
 
 Several considerations have to be taken into account to explain this result.
-First of all we need to explore the cumulative precipitation of the previous and posterior years of these drought events.
+Firstly it would be interesting to explore the cumulative precipitation of the previous and posterior years of these drought events.
 For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern.
@@
 
+A plausible explanation could be
+
 --> Ver y resumir páginas 18-19 Notas Naturkunde
@@
 For instance, the months with the highest cumulative rainfall of the past century occured during the first months of 1996 in our southern site (Cáñar) [Torres-MESA]; and years after 1995 were within the wetter of the past 60 years.
-
-In addition, from a long-term perspective, we obtained a positive relation between tree-growth resilience and drought severity (Figure 6).
-Surprisingly, this result contradicts our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge.
 
@@
 pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Figure 6).
-Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005].
-Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas).
-Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q.
-pyrenaica* situadas en su rear-edge.
-
-Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018].
-En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q.
-pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.
-Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
-pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
 
@@
 pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.
+
+Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005].
+Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas).
+Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q.
+pyrenaica* situadas en su rear-edge.
+
+Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018].
+En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q.
+pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.
+Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
+pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
 
@@
 The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008].
-Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017, @MesaFernandez2018].
+Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017; @MesaFernandez2018].
 From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018].
@@
 
-In Northern sites (*e.g.* San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO].
+In Northern sites (e.g.
+San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO].
 Grasslands and shrublands for cattle farming are located at high elevations.
@@
 Another activities could conditioning the natural resources uses.
-For instance, San Juan oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas), with two periods of intense activity, the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita].
-The mining activity in San Juan site have strongly conditioned the forest structure.
-
-On the other hand, woodland areas of the southern slopes (*e.g.* Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes)[@Calatrava2019].
-Forest resources (firewood, charcoal, acorns) were used in a continuously way, as some historical sources have pointed out [@ValbuenaCarabana2014, @JimenezSerrano2004].
-In fact, in Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] to late twenty century [@MesaTorres2009; @Bonet2014_conama].
+For instance, the mining activity in San Juan site have strongly conditioned the forest structure.
+This oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas).
+Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita].
+
+On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019].
+Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014, @JimenezSerrano2004].
+In Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century.
+From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].
 
@@
 
-The release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4).
-The revision of forests practices carried out in this area in the last 30 years did not reveal any clearing or cutting event for those years.
+The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4).
+We revised the forest practices carried out in this area in the last 30 years [CITA], but we did not find any clearing or cutting event during 1995 - 2000 period.
 Two non-mutually exclusive ways could explain the tree-growth release observed.
 Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004; @Gentilesca2017].
-On the other hand, the strong positive correlations of SPEI with tree-growth obtained for this site (Figure 8) indicates a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014].
+On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014].
 Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006].

### 2019-04-02 man/figures/bai_cronos.Rmd
udate nls

@@
                      pvalue = as.numeric(round(c(pCaL, pCaH, pSJ), 4)),
-                     y = as.numeric(c(as.numeric(predict(mCaL)[40]), 
+                     y = as.numeric(
+                       c(as.numeric(predict(mCaL)[40]), 
                            as.numeric(predict(mCaH)[40]), 
                            as.numeric(predict(mSJ)[40]))))
-lms$y <- lms$y / 100
+# lms$y <- lms$y / 100
 lms <- lms %>% mutate(
+  y = y/100,
   pc = ifelse(pvalue < 0.001, "p < 0.001", as.character(paste0("p = ", pvalue))),
-  label = round(r2,3))
+  label = round(r2,3), 
+  labelp = ifelse(pvalue < 0.001, paste0(label, "**"), label))
 
@@
   geom_text_repel(data = lms, 
-                  aes(x=year, y = y, colour = site, label = label),
-                  nudge_x = Inf, 
-                  direction = "x", show.legend = FALSE, 
+                  aes(x=year, y = y, colour = site, label = labelp),
+                  direction = "x", nudge_x = Inf, nudge_y = 1, show.legend = FALSE, 
                   size = 3) + 
@@
 We only show years replicated with # series > 5.
-Linear trends since 1975 are shown all sites (numbers indicate $r^2$ values).
+Linear trends since 1975 are shown all sites (numbers indicate $r^2$ values; asterisks indicate significant linear trend, *p < 0.001*).
 

### 2019-04-02 man/figures/resilience_long_term.Rmd
udate nls

@@
 
+```{r nls}
+# subset de datos para Non linear squares
+d <- sdf %>% 
+  dplyr::select(d_severity, rs_mean_mean) %>% 
+  rename(sev = d_severity, rs = rs_mean_mean) %>% as.data.frame()
+
+library("nlstools")
+
+f <- as.formula(rs ~ a-b*exp(-c*sev))
+
+preview(f, data=d, start=list(a=-1.6, b=-0.2, c=0.4))
+
+# install nls2
+devtools::install_github("ggrothendieck/nls2")
+
+fo <- rs ~ Const + B * (sev ^ A)
+st1 <- expand.grid(Const = seq(-2, 2, len = 10),
+B = seq(-2,2, len = 10), A = seq(-1, 1, len = 10))
+mod1 <- nls2(fo, start = st1, data=d, algorithm = "brute-force")
+mod1
+# use nls object mod1 just calculated as starting value for
+# nls optimization. Same as: nls(fo, start = coef(mod1))
+nls2(fo, start = mod1)
+
+summary(mod1)
+
+plot(sdf$d_severity, sdf$rs_mean_mean, pch=19)
+lines(d$sev, predict(mod, list(sev = d$sev)))
+
+```
+
 ```{r plot_rt}

### 2019-04-02 man/ms.Rmd
udate nls

@@
 
-The strength of the response of the resilience indices was related to drought severity.
-The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data `REVIEW`(1900 - 2017), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Figure 6).
+At long-term perspective, the strength of the response of the resilience indices was related to drought severity.
+The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (1900 - 2017) (Table S3), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Figure 6).
 A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155).
-Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Figure 6) even not being the most severe drought events.
-`REVIEW` (CONEXION con el siguiente parrafo)
+
+Similar patterns were obtanied Resilience values (*Rs*) varied for the two last severe drought events (2005, 2012; Table 3).
+We obtained significantly higher values for the 2012, the most severe event (Table S3), than for 2005 in both variables, namely greenness and tree-growth (Tables S1-S2; Figure 7).
+Differences were more pronunced for BAI between drought event than for EVI (Tables S1-S2)
+
+Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Figure 6) even not being the most severe drought events. `REVIEW` (CONEXION con el siguiente parrafo)
 
@@
 For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2).
-
-The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (period 1900-2017?),
 

### 2019-04-02 man/figures/resilience_long_term.Rmd
add gam
get nls

@@
 
+library("mgcv")
+library("rcompanion")
+mg <- gam(rs~s(sev), data=d)
+summary(mg)
+
+po <- nls(rs ~ exp.eq(sev, a, b), data = d, start = list(a = .02, b = -.07))
+
+p <- nls(rs ~ a*exp(b*sev), data=d, start = list(a = 0.81, b = 0.021))
+
+plot(d$sev, d$rs, pch=19)
+x <- seq(5,25, 0.1)
+y <- predict(p, list(sev=x))
+lines(x, y)
+abline(lm(d$rs~d$sev), col="red")
+
 library("nlstools")
+
+# exploring
+plot(d$sev, d$rs, pch=19, xlim=c(0,25), ylim=c(0.2, 1.6))
+
+# try exponential 
+f <- as.formula(rs ~ a*exp(b*sev))
+preview(f, data = d, start = list(a = 0.8, b=0.019))
+
+nls(f, data = d, start = list(a = 0.75, b=-0.019))
+
+summary(mo)
+nd <- expand.grid(rs=seq(0.4,1.6, by=0.1),
+                  sev=seq(5,25, by=1))
+
+lines(d$sev, predict(m, nd))
+
+lines(nd$sev, predict(m, nd))
+
+library("FlexParamCurve")
+
+#modpar(posneg.data$age,posneg.data$mass,pn.options = "myoptions.1")
+
+modpar(d$sev, d$rs, pn.options = "myoptions.1")
+
+modcompare <- pn.mod.compare(d$sev, d$rs, d$id, pn.options = "myoptions.1")
+modstep <-  pn.modselect.step(d$sev, d$rs, d$id,existing = TRUE, pn.options = "myoptions.1")
+
+posneg.modstep <- pn.modselect.step(posneg.data$age, posneg.data$mass, posneg.data$id, existing = TRUE, pn.options = "myoptions.1")
+  #existing = TRUE used to avoid refitting existing nlsList models
 
@@
 
-plot(sdf$d_severity, sdf$rs_mean_mean, pch=19)
-lines(d$sev, predict(mod, list(sev = d$sev)))
+g <- gam(rs~s(sev), data=d)
+
+plot(g)
+
+library
 

### 2019-04-03 man/figures/nls_long_term_resilience.Rmd
update nls

--- 2019-04-02 man/figures/resilience_long_term.Rmd
@@
 
-```{r plot_functions}
-#https://gist.github.com/ottadini/6882677 
-lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
-
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
-
-  as.character(as.expression(eq));                 
-}
-
-plotrel <- function(df, v, x, label, coordX_eq, coordY_eq, titulo, ...){
-  require(ggplot2)
-  require(ggrepel)
-
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="gray") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=1.5, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    geom_text_repel(nudge_x = 0.5, size = 3) + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-             size = 4.5, fontface = "bold") +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 11, face = "bold.italic"),
-          panel.grid = element_blank(), 
-          axis.text.y = element_text(size=9),
-          axis.text.x = element_text(size=9),
-          axis.title = element_text(size = 10))
-  }
-
-```
-
 ```{r nls}
@@
 
+library("minpack.lm")
+library("nls2")
+library("nlshelper")
+library("nlstools")
+
+# Catalogo de funciones en Catalog of Curves for Curve Fitting
+# V. Sit and M. Poulin-Costello 1994 Biometrics Information Handbook 
+# https://www.for.gov.bc.ca/hfd/pubs/docs/bio/bio04.htm
+
+set.seed(1467)
+# Exponential type I
+
+fe1 <- as.formula(rs ~ I(a*sev^b + c))
+preview(fe1, data = d, start = list(a= 0.0015, b = 2, c=0.8))
+
+e1 <- nls(fe1, data = d, start = list(a = 0.0015, b = 2, c=.8),
+          trace = TRUE, control = list(maxiter = 100))
+summary(e1)
+
+set.seed(1467)
+
+te <- nls(rs ~ I(exp(1)^(a+b*sev)), data = d, start = list(a=0.8, b = 0.03), trace = TRUE)
+summary(te)
+
+f <- as.formula(d$rs ~ d$sev + d$sev^2)
+lf <- lm(f, data=d)
+
+alpha <- 0.05
+df.new <- data.frame(rs=seq(2,24, 2))
+conf.dist <- predict(lf, newdata = df.new, interval="confidence", level=1-alpha) 
+pred.dist <- predict(lf, newdata = df.new, interval="prediction", level=1-alpha) 
+df.new[c("fit","lwr.conf", "upr.conf")] <- conf.dist
+df.new[c("lwr.pred", "upr.pred")] <- pred.dist[,2:3]
+p +   
+  geom_ribbon(data=df.new, aes(x=rs, ymin=lwr.pred, ymax=upr.pred), alpha=0.1, inherit.aes=F, fill="blue") +
+    geom_line(data=df.new, aes(x=rs, y=fit), colour="#339900", size=1)
+
++ 
+  geom_ribbon(data=df.new, aes(x=speed, ymin=lwr.conf, ymax=upr.conf), alpha=0.2, inherit.aes=F, fill="#339900") +
+
+p <- ggplot(d) + geom_point(aes(x=sev, y=rs))
+
+pl <- ggplot(cars) + geom_point(aes(x=speed, y=dist), size=2, colour="#993399") + 
+  xlab("Speed (mph)") + ylab("Stopping distance (ft)")
+
+offset <- round(min(d$rs),2)
+
+d$lrs <- log(d$rs + 0.5)
+d$lsev <- log(d$sev)
+
+l1 <- lm(d$rs~d$lsev, data=d)
+plot(d$rs~d$lsev) 
+abline(l1)
+summary(l1)
+
+R2 
+RSSp <- sum(residuals(e1)^2)
+
+TSS <- sum((d$rs - mean(d$rs))^2)
+
+1 - RSSp/TSS
+
+# R2
+
+r2s <- function(modelo, dataframe, y){
+
+  RSS <- sum(residuals(modelo)^2)
+  TSS <- sum(dataframe[,y] - mean())
+
+  out <- data.frame(RSS, TSS)
+  return(out)
+
+}
+
+r2s(e1, d, "rs")
+
+RSS
+
+# Adjusted-R2
+
+n <- dim(d)[1]
+p <- length(coefficients(e1))
+
+r2adj_pi <- 1 - ((n -1))/(n-p)) * (1- r2.pi)
+
 library("mgcv")
 library("rcompanion")
-mg <- gam(rs~s(sev), data=d)
+library("gam")
+
+# Modelo 1 gam 
+mg <- mgcv::gam(rs~s(sev), data=d, trace = TRUE)
 summary(mg)
@@
 
+```{r plot_functions}
+#https://gist.github.com/ottadini/6882677 
+lm_eqn = function(m) {
+  # Displays regression line equation and R^2 value on plot
+  # Usage:
+  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
+
+  l <- list(#a = format(coef(m)[1], digits = 2),
+            # b = format(abs(coef(m)[2]), digits = 2),
+            r2 = format(summary(m)$r.squared, digits = 3));
+
+  if (coef(m)[2] >= 0)  {
+    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
+    eq <- substitute(~~italic(r)^2~"="~r2,l)
+  } else {
+    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
+    eq <- substitute(~~italic(r)^2~"="~r2,l)
+  }
+
+  as.character(as.expression(eq));                 
+}
+
+plotrel <- function(df, v, x, label, coordX_eq, coordY_eq, titulo, ...){
+  require(ggplot2)
+  require(ggrepel)
+
+  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
+
+  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
+                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
+
+  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
+    geom_smooth(method='lm', se=FALSE, colour="gray") +
+    geom_errorbar(mapping = yerrbar) + 
+    geom_point(size=1.5, pch = 21, colour="black", fill="white") + 
+    theme_bw() + 
+    geom_text_repel(nudge_x = 0.5, size = 3) + 
+    annotate("text", x=coordX_eq, y = coordY_eq, 
+             label=lm_eqn(lm(formulita, df)), parse=TRUE, 
+             size = 4.5, fontface = "bold") +
+    ggtitle(titulo) + 
+    theme(plot.title = element_text(size = 11, face = "bold.italic"),
+          panel.grid = element_blank(), 
+          axis.text.y = element_text(size=9),
+          axis.text.x = element_text(size=9),
+          axis.title = element_text(size = 10))
+  }
+
+```
+
 ```{r plot_rt}

### 2019-04-04 man/figures/nls_long_term_resilience.Rmd
add pvalue and rsquared

@@
 
+TE Quedaste por aquí. No acabaste de encontrar un NLS mejor que la LM
+
 ```{r nls}
@@
 
+d$ly <- log(d$rs)
+d$lx <- log(d$sev)
+
+summary(lm(ly~lx, data=d))
+
+d$y <- log(d$rs-0.6)
+
 # Catalogo de funciones en Catalog of Curves for Curve Fitting
@@
 
+pF <- function(x,a,b,c) {
+  I(a*x^b + c)
+}
+
 fe1 <- as.formula(rs ~ I(a*sev^b + c))
-preview(fe1, data = d, start = list(a= 0.0015, b = 2, c=0.8))
+m1 <- nls(fe1, data = d, start = list(a=2.755, b=.00014, c=.85), 
+          control = list(maxiter = 500), trace = TRUE)
+summary(m1)
+abline(lm(rs~sev, data=d), col="blue")
 
-e1 <- nls(fe1, data = d, start = list(a = 0.0015, b = 2, c=.8),
+summary(lm(rs~sev, data=d))
+
+grid.start <- expand.grid(list(
+  a = seq(0.001, 0.03, by=0.001),
+  b = seq(1.8, 2.2, by=0.01),
+  c= c(0.8)))
+
+m2 <- nls2(rs~pF(sev, a,b,c), data = d, 
+          start = grid.start,
+          algorithm = "brute-force")
+
+preview(fe1, data = d, start = list(a= 0.5, b = 0.23, c=0.8))
+
+e1 <- nls(fe1, data = d, start = list(a= 0.5, b = 0.18, c=0.8),
           trace = TRUE, control = list(maxiter = 100))

### 2019-04-04 man/figures/resilience_long_term.Rmd
add pvalue and rsquared

--- 2018-11-24 man/figures/resilience_long_term.Rmd
@@
           axis.title = element_text(size = 10))
+
   }
@@
   theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
+
+summary(lm(rt_mean_mean~d_severity, data=sdf))
 ```
@@
   theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt"))
+
+summary(lm(rs_mean_mean~d_severity, data=sdf))
 ```
@@
   theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
+
+glance(lm(rc_mean_mean~d_severity, data=sdf))
 ```
@@
 ```
+
+```{r}
+df <- sdf %>% 
+  dplyr::select(d_severity, rt_mean_mean, rs_mean_mean, rc_mean_mean) %>% 
+  gather(key = res_metric, value,c(rt_mean_mean,rs_mean_mean,rc_mean_mean))
+
+nested_df <- group_by(df, res_metric) %>% nest()
+
+fit_model <- function(df){
+  lm(value ~ d_severity, data = df)
+}
+
+nested_df <- nested_df %>% 
+  mutate(model = map(data, fit_model))
+
+# Get R2 and p.value 
+get_rsq <- function(mod) glance(mod)$r.squared
+get_pvalue <- function(mod) glance(mod)$p.value
+
+nested_df <- nested_df %>% 
+  mutate(r.squared = map_dbl(model, get_rsq),
+         pvalue = map_dbl(model, get_pvalue))
+
+```

### 2019-04-08 man/figures/resilience_long_term.Rmd
add pvalues regression

@@
 library("ggrepel")
+library("broom")
 ```
@@
 
-  l <- list(#a = format(coef(m)[1], digits = 2),
-            # b = format(abs(coef(m)[2]), digits = 2),
-            r2 = format(summary(m)$r.squared, digits = 3));
+  # l <- list(#a = format(coef(m)[1], digits = 2),
+  #           # b = format(abs(coef(m)[2]), digits = 2),
+  #           rvalue = format(summary(m)$r.squared, digits = 3),
+  #           pvalue = format(summary(m)$coefficients[2,'Pr(>|t|)'],3));
 
-  if (coef(m)[2] >= 0)  {
-    # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  } else {
-    # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-    eq <- substitute(~~italic(r)^2~"="~r2,l)
-  }
+  # eq <- substitute(~~italic(r)^2~"="~r2,l[1], "p=",l[2])
 
-  as.character(as.expression(eq));                 
+  eq2 <- substitute(~~italic(r)^2~"="~rvalue*","~italic(p)~"="~pvalue,  
+                    list(rvalue = format(summary(m)$r.squared, digits = 3), 
+                         pvalue = format(summary(m)$coefficients[2,4], digits = 3)))
+
+  # if (coef(m)[2] >= 0)  {
+  #   # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
+  #   eq <- substitute(~~italic(r)^2~"="~r2,l[1], "p=",l[2])
+  # } else {
+  #   # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
+  #   eq <- substitute(~~italic(r)^2~"="~r2,l)
+  # }
+
+  as.character(as.expression(eq2));                 
 }
@@
 
-summary(lm(rt_mean_mean~d_severity, data=sdf))
+i <- summary(lm(rt_mean_mean~d_severity, data=sdf))
 ```
@@
   theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
-
-glance(lm(rc_mean_mean~d_severity, data=sdf))
 ```
@@
 
-```{r}
+```{r, eval=FALSE, echo=FALSE}
 df <- sdf %>% 

### 2019-04-08 man/figures/evi_vs_severe_droughts.Rmd
fix figure typo

@@
 ```{r}
+micolor <- '#455883'
+
 # PLOT EVI + DROUGHT 

### 2019-04-08 man/figures/resilience_long_term.Rmd
fix figure typo

@@
     geom_text_repel(nudge_x = 0.5, size = 3) + 
-    annotate("text", x=coordX_eq, y = coordY_eq, 
+    annotate("text", x = coordX_eq, y = coordY_eq, 
              label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-             size = 4.5, fontface = "bold") +
+             size = 3, fontface = "bold", hjust="left") +
     ggtitle(titulo) + 
@@
 sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-                 coordX_eq = 10, coordY_eq = 0.5, titulo = "Rt") +
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rt") +
   xlab('') + ylab('') +
@@
 
-i <- summary(lm(rt_mean_mean~d_severity, data=sdf))
 ```
@@
 sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-                 coordX_eq = 10, coordY_eq = 0.5, titulo = "Rs") + 
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rs") + 
   xlab('') + ylab('') +
   theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt"))
-
-summary(lm(rs_mean_mean~d_severity, data=sdf))
 ```
@@
 sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-                 coordX_eq = 10, coordY_eq = 0.5, titulo = "Rc") +
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rc") +
   xlab("Drought Severity") + ylab('') +

### 2019-04-08 man/ms.Rmd
update resilience results

--- 2019-03-28 man/ms.Rmd
@@
 ## Resilience of greenness and radial tree-growth to drought events
-
-The strength of the response of the resilience indices was related to drought severity.
-The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data `REVIEW`(1900 - 2017), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336) (Figure 6).
-A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438) but negative for resistance (*Rt*, $r^2$ = 0.155).
-Resilience values of tree-growth for 2005 was the lowest of the drought events analyzed (Figure 6) even not being the most severe drought events.
-`REVIEW` (CONEXION con el siguiente parrafo)
-
+At long-term perspective, the strength of the response of the resilience indices was related to drought severity.
+The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (1900 - 2017) (Table S3), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336, p = 0.048) (Figure 6).
+A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438, p = 0.019) but negative for resistance (*Rt*, $r^2$ = 0.155, p = 0.205).
+
+For greenness and tree-growth, resilience metrics were significantly different bewteen drought events at short-term (Table 3).
+We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7).
+Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth.
+Recovery values for greenness was higher for 2005 than for 2012 drought event (Table S1, Figure 7).
+
+Resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3).
+Southern populations showed lower values for Recovery (*Rc*) than northern site for greenness and tree-growth (Tables S1 and S2).
+In opposite, Resistance and Resilience values were higher for southern sites than for northern ones (Tables S1 and S2).
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+> Resistance:
+
+Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
+Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).
+
+* No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012 (F = 6.0189; p = 0.019).
+Valores de resistencia menor al evento de 2005.
+[2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.025; trimmed-p = 0.019]
+
+* Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001).
+Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].
+
+* La interacción también fue significativa.
+De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ???
+
+> Recovery 
+> * Mayor recuperación del crecimiento de los robledales tras la sequía de 2005 que tras la de 2012 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
+> * Las poblacionees del sur (caH y caL; no diferencias entre ellas) mostraron una menor recuperación que las poblaciones del norte, es decir, las del norte se recuperaron mas rápidamente que las del sur que no llegaron incluso a valores de Rc = 1.
+
+Similar pattern of resilience (*Rs*) values was found for growth than for greenness respect to drought event: significantly higher values of *Rs* for the 2012 drought event than for the 2005 (Table S2; Figure 3).
+However, no differences were observed between sites (Table 3).
+For 2005 drought event, *Rs* value of SJ (northern site) was lower than that of southern ones (CA-High and CA-Low), but opposite pattern was found for the 2012 drought event.
+All valu`es of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031).
+
+YA ESTOY DESESPERADO CON ESTE PAPER
 *Q.
@@
 An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 7, Table S2).
+
 Northern populations showed significantly higher values of recovery than southern sites for greenness and tree-growth (Table S2).
@@
 
-The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (period 1900-2017?),
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-> Resistance:
-
-Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
-Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).
-
-* No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012 (F = 6.0189; p = 0.019).
-Valores de resistencia menor al evento de 2005.
-[2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.025; trimmed-p = 0.019]
-
-* Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001).
-Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].
-
-* La interacción también fue significativa.
-De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ???
-
-> Recovery 
-> * Mayor recuperación del crecimiento de los robledales tras la sequía de 2005 que tras la de 2012 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
-> * Las poblacionees del sur (caH y caL; no diferencias entre ellas) mostraron una menor recuperación que las poblaciones del norte, es decir, las del norte se recuperaron mas rápidamente que las del sur que no llegaron incluso a valores de Rc = 1.
-
-Similar pattern of resilience (*Rs*) values was found for growth than for greenness respect to drought event: significantly higher values of *Rs* for the 2012 drought event than for the 2005 (Table S2; Figure 3).
-However, no differences were observed between sites (Table 3).
-For 2005 drought event, *Rs* value of SJ (northern site) was lower than that of southern ones (CA-High and CA-Low), but opposite pattern was found for the 2012 drought event.
-All valu`es of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031).
+Similar to the long-term observed pattern,
+
+was greater after more severe drought event (2012). Opposite pattern was observed for greenness with
+
+Lower resistance values (*Rt*) were observed to the 2005 drought event than to that in 2012 for both variables.
+The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2).
+Resistance values to drought for greenness and tree-growth varied between sites (Table 3).
+Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2).
+It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 7).
 
@@
 --> Aquí iría la idea de lo de refugio
+
+---> Nota
+
+Para la discussion: 
+- Resilience values of RW for 2005 was the lowest of the drought events analized even not being the moste severe drought event 
+- 2005 reduced greenness and bai more than 2012
 

### 2019-04-08 man/ms.Rmd
add drought event as method and remove from table

@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-SPEI values between 1 and −1 are considered normal, whereas values <−1 indicate drought and values <−1.5 indicate severe drought47. We defined a drought as a period where SPEI12 decreased to ≤−1.5.
-")
-```
-
 Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using the `treeclim` package [@Zang2015].
 The bootstrapped confidence intervals were used to estimate significance (p < 0.05) of the correlation coefficients.
+
+To identify drought events at long-term perspective we used SPEI index (12 months) for all 0.5º grid cells covering Sierra Nevada obtainde from Global SPEI Database (http://spei.csic.es/database.html).
+A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
+A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
+Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+SPEI values between 1 and −1 are considered normal, whereas values <−1 indicate drought and values <−1.5 indicate severe drought47. We defined a drought as a period where SPEI12 decreased to ≤−1.5.
+")
+```
 
@@
 A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017].
-To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure S5).
+To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure 6, Table S3).
 

### 2019-04-08 man/tables/t_droughts.Rmd
add drought event as method and remove from table

@@
 ##### 
-**Table S3.** Drought events for Sierra Nevada based on SPEI index (12 months scale).
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-The ***duration*** of a drought event is the number of consecutive months with the SPEI lower than a certain threshold.
-***Severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event.
-***Intensity*** and ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-SPEI data were obtained for all 0.5º grid cells covering Sierra Nevada.
+**Table S3.** Drought events for Sierra Nevada based on SPEI index (12 months scale). See methods for details.
 

### 2019-04-10 man/figures/plot_spei12.Rmd
update plot spei12

@@
 
+```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8} 
+# plot_spei
+```
+
+```{r}
+
+# DATOS desde 1901
+
+# Get file names
+myfiles <- list.files(path = here("data_raw/spei/spei_longterm"), pattern = "\\.csv$")
+
+mydf <- data.frame()
+
+for (j in myfiles){ 
+  aux <- read.csv(file=here("data_raw/spei/spei_longterm", j), header = TRUE, sep = ',')
+
+  # Remove csv and get name 
+  name_aux <- str_replace(j, ".csv", "")
+
+  # Get lat long
+  latlong <- str_replace(name_aux, "spei_", "")
+  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
+  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
+
+  # Store lat and long 
+  aux$lat <- mylat
+  aux$long <- mylong
+
+  # year and month 
+  aux$month <- rep(1:12, 115)
+
+  auxy <- c()
+  years <- c() 
+  for (j in 1901:2015) { 
+    auxy <- as.data.frame(rep(j,12))
+    years <- rbind(years, auxy)
+    } 
+  names(years) <- 'year'
+
+  aux <- cbind(aux,years)
+
+  # join all df 
+  mydf <- rbind(mydf, aux)
+}
+
+# Convert to date  
+mydf <- mydf %>% mutate(date = sprintf("%d-%02d", year, month),
+                        date = as.Date(paste0(date, '-01'), format="%Y-%m-%d"))
+
+speidf <- mydf %>% dplyr::select(SPEI06,SPEI12,SPEI24,month, year, date) %>% 
+  group_by(year,month, date) %>%
+  summarise_at(vars(SPEI06,SPEI12,SPEI24), funs(mean,sd,se=sd(.)/sqrt(n()))) %>% 
+  as.data.frame() 
+```
+
+```{r}
+speiPlot <- function(df, vi, mytitle, year_scaleX){
+  require(dplyr)
+  require(scales)
+
+  # Select variable of interest 
+  dfs <- df %>% dplyr::select(year, month, date, starts_with(vi))
+
+  # Rename variables 
+  dfs <- dfs %>% rename_all(~sub(paste0(vi,'_'), '',.x))
+  # Create variable pos and neg 
+  dfs <- dfs %>% mutate(signo = ifelse(mean > 0, "pos", "neg"))
+
+  # Plot  
+  p <- ggplot(dfs, aes(x = date, y = mean, fill = signo)) + 
+    geom_rect(aes(xmin=as.Date("2005-01-01"), xmax = as.Date("2005-12-01"), ymin=-Inf, ymax=Inf),
+              fill="grey", alpha=.9) + 
+    geom_rect(aes(xmin=as.Date("2012-01-01"), xmax = as.Date("2012-12-01"), ymin=-Inf, ymax=Inf),
+              fill="grey", alpha=.9) + 
+    geom_bar(stat = "identity") +
+    scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
+    theme_bw()  + 
+    labs(title= mytitle, x= '', 
+         y = paste0(substr(vi, 1, 4), ' (', gsub("[^0-9]","",vi),'-month scale)')) +
+    theme(strip.background = element_rect(fill = "white"),
+        legend.position = "none", 
+        panel.grid = element_blank(), 
+        axis.text.y = element_text(size=10),
+        axis.text.x = element_text(size=9),
+        axis.title.y = element_text(size = 12))
+
+  # Date Scale in X 
+  if (missing(year_scaleX)){
+    p
+    } else { 
+    year_interval <- paste0(year_scaleX," year")
+    p <- p + scale_x_date(date_breaks = year_interval, date_labels = "%Y",
+                          limits=as.Date(c("1901-01-01", "2015-12-01")))
+    }
+
+  return(p)
+}
+
+s <- speiPlot(speidf, vi = 'SPEI12', mytitle = "", year_scaleX = 10)
+```
+
 **Figure S2**.
-Drought severity in the Sierra Nevada for the 1950-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+Drought severity in the Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
 Data from Global SPEI database (http://spei.csic.es/database.html).
 We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
+Horizontal gray bars indicate 2005 and 2012 year.
 
-```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8} 
-plot_spei
+```{r plot_out1900, fig.width=6, dpi = 300, out.width = 0.8} 
+s
 ```

### 2019-04-10 man/ms.Rmd
update figures and texto

@@
 We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
-Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term.
-The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015 (see Table S3).
+Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3).
+The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015.
+We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html).
+A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017].
+A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a].
+Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
 
@@
 
-To identify drought events at long-term perspective we used SPEI index (12 months) for all 0.5º grid cells covering Sierra Nevada obtainde from Global SPEI Database (http://spei.csic.es/database.html).
-A drought event starts in the month when SPEI falls below the threshold of -1.28 [@Pascoa2017].
-A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a; @Pascoa2017].
-Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
-
 ```{r echo=FALSE, results='asis', eval=show_text}
@@
 Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4).
-`REVIEW` In addition, trees from this location showed a positive growth trend since the late 1970s (Figure 4), which was not found for any of the other two locations.
-`END REVIEW` For southern population differences in growth were observed.
+Trees for southern sites showed significant positive growth trend since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High).
+We also observed differences in growth for southern sites.
 CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figures 4 and S6).
 
+Las sequías produjeron reducciones de crecimiento en los tres sitios. De los eventos de sequía analizados, destaca por un lado que 1995 produjo la mayor reducción de crecimiento. 
+VAS POR AQUI 
 For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low).
@@
 Relationships with temperatures were lower than those with moisture variables.
-Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figure 8b and 8c).
+Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while no significant limitations were recorded for northen site regarding the temperatures (Figure 8b and 8c).
 Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population.

### 2019-04-11 man/tables/t_dendro_summary.Rmd
tables

@@
 
-ta2 <- regulartable(t2, col_keys =c("name", "first_year", "last_year", "n_trees", "n_radii", "RW", "ms2",
+ta2 <- regulartable(t2, col_keys =c("name", "first_year", "last_year", "length_year", "n_trees", "n_radii", "RW", "ms2",
                                    "ar1","rbar.wt","eps"))

### 2019-04-11 man/figures/evi_vs_severe_droughts.Rmd
fix sizes figures

@@
   geom_segment(data = dy, aes(y = mean, x= iD, xend =eD, yend = mean, colour = d_severity), 
-               size = 8, alpha = 0.7) +
-  geom_text(data = dy, aes(label = year, x = (iD + floor(eD-iD)/2))) +
-  scale_colour_gradient(low = "orange", high = "brown") + 
-  labs(colour = "Drought Severity") +
-  geom_errorbar(aes(ymin = mean - 10*se, ymax= mean + 10*se), width=4, colour=micolor, size=.8) + 
-  geom_line(colour=micolor, size=.8) + 
+               size = 6, alpha = 0.7) +
+  geom_text(data = dy, aes(label = year, x = (iD + floor(eD-iD)/2)), size = 3) +
+  scale_colour_gradient(low = "orange", high = "brown", 
+                        guide = guide_colourbar(direction = "horizontal")) + 
+  labs(colour = "Drought \n Severity") +
+  geom_errorbar(aes(ymin = mean - 10*se, ymax= mean + 10*se), width=4, colour=micolor, size=.6) + 
+  geom_line(colour=micolor, size=.6) + 
   geom_point(size=3,colour=micolor) +
@@
   ylab('EVI') + xlab('')
+
+combinado2 <- combinado2 + 
+  theme(legend.position = "bottom",
+        legend.text=element_text(size=rel(0.8)),
+        legend.title =element_text(size=rel(0.8)))
 

### 2019-04-11 man/figures/ring_climate.Rmd
fix sizes figures

@@
   geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.06))), 
-            position = position_dodge(.9), size = 6), 
+            position = position_dodge(.9), size = 3), 
   theme_bw() +

### 2019-04-11 man/ms.Rmd
update text

@@
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
-All analyses were carried out in R software [@R2017] using several packages (see Appendix S7).
+All analyses were carried out in R software [@R2017] using several packages (see Appendix S8).
 
@@
 Las sequías produjeron reducciones de crecimiento en los tres sitios. De los eventos de sequía analizados, destaca por un lado que 1995 produjo la mayor reducción de crecimiento.
+
 VAS POR AQUI 
@@
 A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438, p = 0.019) but negative for resistance (*Rt*, $r^2$ = 0.155, p = 0.205).
-
-For greenness and tree-growth, resilience metrics were significantly different bewteen drought events at short-term (Table 3).
+Trees showed the highest value of tree-growth resilience for the worst drought event in our study area (1995, see Table S3).
+
+During the last two drought events, resilience metrics for greenness and tree-growth were significantly different bewteen drought events (Table 3).
 We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7).

### 2019-04-12 man/ms.Rmd
comentario radial growth

@@
 
-Las sequías produjeron reducciones de crecimiento en los tres sitios. De los eventos de sequía analizados, destaca por un lado que 1995 produjo la mayor reducción de crecimiento.
-
-VAS POR AQUI 
-For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low).
-Yet, the lowest pointer year (*i.e.* the lowest growth) since 1950 was 1995 in all chronologies (Figure S3).
+From a long-term perspective, drought events produced reduction of RWI for all sites (Figure S3), particularly in 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies.
+Southern sites (CA-High and CA-Low) showed weaker reductions than northern site (SJ), especially for 2005 and 2012 (Figure S3).
+Trees from northern site also suffered a great reduction during 1945-1946 drought event.
 

### 2019-04-22 man/ms.Rmd
update

@@
 * The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure.
-Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (i.e.
-warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge.
+Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge.
 Furthermore, the coldest stands show a positive growth trend since the end of the 1970s matching the simultaneous increase in temperature observed.
@@
 2005 and 2012 were considered as single drought events.
-The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively.
+The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively.
 A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017].
@@
 During the last two drought events, resilience metrics for greenness and tree-growth were significantly different bewteen drought events (Table 3).
+The 2005 drought event reduced greennes and growth more than that of 2012 (Tables S1 and S2).
 We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7).
@@
 
-Our findings show that severe droughts, such us 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q.
-pyrenaica* (Figures 2 and 4).
+Severe drought events negatively affected to primary and secondary growth of *Q.
+pyrenaica* forests in the rear edge of their distribution.
+Our findings show that 2005 and 2012 drought events, provoked reductions in greenness and tree-growth of these forests.
 These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018].
-The greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018].
-
-During the last decades, the south of Iberian Peninsula suffered two of the worst recorded drought [2005 and 2012; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3).
-However, we found a positive trend for vegetation greenness of *Q.
-pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period).
+In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018].
+
+Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
+pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period).
 This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi].
@@
 Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014].
-Our pattern differs from the decline trends observed for several mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
+Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
 sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].
@@
 The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q.
-pyrenaica* forests [@PeresLis2017].
+pyrenaica* forests [@PeresLis2017] (Figure S7).
 Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018],  that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].
@@
 Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005].
-For instance, higher vulnerability to drought is ofted assumed for rear-edge populations [@MartinezVilalta2018], however some studies showed different findings [@Cavin2017; @Granda2017], as our results suggest, with high values of resilience for populations of * Q.
-pyrenaica * located in its rear-edge.
-
+For instance, higher vulnerability to drought is often assumed for rear-edge populations [@MartinezVilalta2018], however some studies showed different findings [@Cavin2017; @Granda2017], as our results suggest, with high values of resilience for populations of *Q.
+pyrenaica* located in its rear-edge.
 This would be related with the definition of the marginal population.
@@
 
-Several considerations have to be taken into account to explain this result.
+Several considerations have to be taken into account to explain the high resilience values found.
 Firstly it would be interesting to explore the cumulative precipitation of the previous and posterior years of these drought events.
@@
 In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009].
-In addition, this argument could also explain the higher values of recovery found for 1995 (Figure 6).
+This could explain the higher values of recovery found for 1995 (Figure 6).
+
+VAS POR AQUI
+
+De forma general: 
+* dependencia del agua
+* sitaución en Sierra Nevada
+* Mayor resiliencia (mayor crecimiento) los robles mas altos
+
+* Existencia de aportes extra de humedad. Las acequias
+
+- Por un lado una alta disponibilidad de agua. Conocemos que esta especies es mas sensitive a la humedad del suelo que otras especies arbóreas mediterráneas (GeaFEM)
+
+Por un lado la dependencia del agua. Sabemos que esta especie es mas sensitive a la humedad del suelo que otras especies arbóreas mediterráneas (GeaFEM). En Sie
 

### 2019-04-23 man/ms.Rmd
update discussion

@@
 ## Rear-edge oaks show high resilience in response both to recent drought events and long-term climatic variability. 
-
 Severe drought events negatively affected to primary and secondary growth of *Q.
@@
 Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005].
-For instance, higher vulnerability to drought is often assumed for rear-edge populations [@MartinezVilalta2018], however some studies showed different findings [@Cavin2017; @Granda2017], as our results suggest, with high values of resilience for populations of *Q.
-pyrenaica* located in its rear-edge.
-This would be related with the definition of the marginal population.
+A high vulnerability to drought is often assumed for populatios located at their rear-edge [@MartinezVilalta2018].
+However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q.
+pyrenaica* and high values of resilience to severe droughts were also reported.
+Our findings are in agreenment with those studies that have showed that the assumed higher vulnerability of dry edges does not necesarily hold [@Cavin2017; @Granda2017].
+@MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations.
+The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada, seem to be indicative of high local adaptation of this oak to this mountain range.
+In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
 Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018].
@@
 pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.
+
+A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6).
+A plausible explanation could be the water availability previously and after the drought event.
+For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern.
+In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009].
+This could explain the higher values of recovery found for 1995 (Figure 6).
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+
+OOOOOOOOOOOJOOOOOOOOOOO CON ESTO
+
 Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
@@
 
-Several considerations have to be taken into account to explain the high resilience values found.
-Firstly it would be interesting to explore the cumulative precipitation of the previous and posterior years of these drought events.
-For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern.
-In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009].
-This could explain the higher values of recovery found for 1995 (Figure 6).
-
-VAS POR AQUI
-
-De forma general: 
-* dependencia del agua
-* sitaución en Sierra Nevada
-* Mayor resiliencia (mayor crecimiento) los robles mas altos
-
-* Existencia de aportes extra de humedad. Las acequias
-
-- Por un lado una alta disponibilidad de agua. Conocemos que esta especies es mas sensitive a la humedad del suelo que otras especies arbóreas mediterráneas (GeaFEM)
-
 Por un lado la dependencia del agua. Sabemos que esta especie es mas sensitive a la humedad del suelo que otras especies arbóreas mediterráneas (GeaFEM). En Sie
@@
 --> Ver y resumir páginas 18-19 Notas Naturkunde
-
-A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6).
-A plausible explanation could be the water availability after the drought event.
-For instance, the months with the highest cumulative rainfall of the past century occured during the first months of 1996 in our southern site (Cáñar) [Torres-MESA]; and years after 1995 were within the wetter of the past 60 years.
-
---> Aquí iría la idea de lo de refugio
 
@@
 - 2005 reduced greenness and bai more than 2012 
+")
+```
 
@@
 ```
-
-Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q.
-pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Figure 6).
 

### 2019-05-02 man/ms.Rmd
update

@@
 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
-We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
+We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
 Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3).
@@
 These periods alternate with periods of supression.
-Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression event in the last 50 years.
+Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression events in the last 50 years.
 
@@
 For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014].
+Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970 [@BladeCastroDiez2010CeEppyf; @VicenteSerranoetal2017CIG].
 Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
@@
 ```
+
+Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. The exploration of the average reduction for the 2005 and 2012 droguht events
+revaled that EVI was reduced to 90.25 % (Table S1), whereas BAI was reduced to 70.83 % (Table S2).
+This suggests a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate.
+For instance, during 2005 drought event, the BAI were reduced to 44.5 % respect to that of the preceding period in northern site (SJ), whereas the EVI was reduced to 81.9 % for the same site during the same drought event (Tables S1 and S2).
+
+The different sensitivity
+
+Causas: 
+Gazol
+EVI of the understory (Dioniosio) 
+Different scales
 
@@
 
-En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations).
-The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone.
-
 In addition, despite southern sites are very close to each other, we obtained higher values of BAI for the high-elevation site than for the low-elevation site.
 
-")
-```
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-### OTRA COSA
-
-In fact, several authors have raise the need to consider land use y drought de forma conjunta en las environmental y forest management .... (ver Peñuelas et al 2017 y tb Doblas Miranda )
-
-Environmental and forest management policies should take into account all these characteristics of Mediterranean forests and the social, environmental and climatic conditions that are projected for the coming years and decades.
 ")
@@
 
-- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):
-
 * Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
@@
 ###
-Nosotros hemos obtenido una relación entre la severidad de la sequía y la resiliencia mostrada por los robledales en su rear edge. (ver gráfica )
 
@@
 
-Methodological approach
-
-comentario: tengo que rematar este apartado.
-
-Vegetation reflects the environmental conditions, and the effects of drought on vegetation can be observed using information from several methodological approaches [see @Norman2016 for a review].
-Several works revealed the utility of remote-sensing [*e.g*.
-@Zhang2013; @AghaKouchak2015] and of the dendrochronology [*e.g.* @Eilmann2012; @Bhuyan2017] to quantify the drought impacts on vegetation at different spatial and temporal scales, but to our knowledge the combination of both approaches is scarce [@VicenteSerrano2013; @VicenteSerrano2016; @Bhuyan2017b; @Wu2018].
-Estos trabajos han observado que existe de forma general a positive relationship between vegetation indices derived from remote sensing and annual tree growth [@VicenteSerrano2016; @Bhuyan2017b; @Gazol2018].
-
-Aunque existen algunos trabajos que han utilizado RWI y remote sensing nuestra aproximación (creo) es interesante, sobre todo porque la aplicamos al estudio de poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
-Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares que analicen la respuesta de las poblaciones al cambio climático y su resiliencia utilizando diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].
-
-- Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q.
-pyrenaica situadas en su reaar edge frente a los eventos de sequía.
-
-- Hemos encontrado diferencias en cuanto a la resiliencia utilizando RS y RWI.
-Diferente sensibilidad de satélite frente a los datos de campo.
-Esto también se ha observado en @Gazol2018, que analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR.
-Encontraron que los datos de TR son mas sensibles para la resiliencia del bosque a la sequía que los datos de RS
-
-- De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
-
-- La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017] (...
-Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes.
-)
-
-- Nota from @Jump2010: Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability.
-In such regions, targeted collection of high-resolution data is necessary in order to identify currently occurring changes and predict the magnitude and spatial distribution of future decline.
-[@Jump2010]
-
 ```{r echo=FALSE, results='asis', eval=show_text}

### 2019-05-06 man/ms.Rmd
update

@@
 
-Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. The exploration of the average reduction for the 2005 and 2012 droguht events
-revaled that EVI was reduced to 90.25 % (Table S1), whereas BAI was reduced to 70.83 % (Table S2).
+Greenness and tree-growth of *Q.
+pyrenaica* showed different sensitivity to severe droughts.
+During 2005 and 2012 drought events EVI was reduced to 90.25 % (Tabñe 1), whereas BAI was reduced to 70.83 % (Table S2).
 This suggests a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate.
 For instance, during 2005 drought event, the BAI were reduced to 44.5 % respect to that of the preceding period in northern site (SJ), whereas the EVI was reduced to 81.9 % for the same site during the same drought event (Tables S1 and S2).
-
-The different sensitivity
-
-Causas: 
-Gazol
-EVI of the understory (Dioniosio) 
-Different scales
+These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018].
+Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink thant by source limitations [@Gazol2018].
+Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisio2012].
+On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Gazol2018].
 

### 2019-05-14 man/ms.Rmd
update

@@
 
-The forests of this species reaches their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*.
+The forests of this species reaches one of their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*.
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
 There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
-Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005].
-For this reason, the populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], having these forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
+Sierra Nevada mountain is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010].
+In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005].
+They are considered relictic forests [@Melendo2000; @Vivero2000], and the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blanca1998; @Lorite2009].
+These forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
 Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].
@@
 
-As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3).
-This pattern was also observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015].
+As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015].
 But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6).
@@
 pyrenaica* showed different sensitivity to severe droughts.
-During 2005 and 2012 drought events EVI was reduced to 90.25 % (Tabñe 1), whereas BAI was reduced to 70.83 % (Table S2).
-This suggests a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate.
-For instance, during 2005 drought event, the BAI were reduced to 44.5 % respect to that of the preceding period in northern site (SJ), whereas the EVI was reduced to 81.9 % for the same site during the same drought event (Tables S1 and S2).
+During the last drought events EVI was reduced to 90.25 % (Table 1), whereas BAI was reduced to 70.83 % (Table S2), suggesting a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate.
+For our driest site (SJ, the northern one), BAI were reduced to 44.5 % respect to that of the preceding period during 2005 drought event, whereas the EVI was reduced to 81.9 % (Tables S1 and S2).
 These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018].
-Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink thant by source limitations [@Gazol2018].
+Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018].
 Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisio2012].
-On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Gazol2018].
+On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
 
@@
 Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3).
-Las poblaciones del norte, que presentan condiciones mas secas (Table 1), se vieron mas afectadas por los eventos de sequía.
+The former, which have drier conditions (Table 1), were more impacted by the drought events than the latter.
 For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2).
@@
 This is especially interesting for southern sites, which are very close to each other.
-Our results are consistent with previous findings that pointed out tree growth and tree responses to drought are site-dependent (e.g.
+Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (e.g.
 soil features, tree competence, etc.) [e.g.
@@
 pyrenaica* [@GeaIzquierdo2014].
-In fact, the differences in available water of soils could explain our results.
-@CoboDiaz2017, in a work analyzing the variation of the microbiota along an elevational gradient within this location, found higher values of available water of the forest soils located on high elevation compared with those located at low elevations.
+Soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017].
 In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply.
-In most of the high-mountain watersheds of Sierra Nevada, there is a system of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016].
+In most of the high-mountain watersheds of Sierra Nevada, there are systems of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016].
 These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017].
-The network of traditional ditches present in one of our sampling site (Cáñar) was recentlty described, with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015].
+Several works have recently described the network of traditional ditches for our southern sites (Cáñar), with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015].
 There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017].
@@
 This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site.
+
+In addition to the lower water stress, it is likely that the sites at higher elevation had a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources.
 
@@
 These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
-@GeaIzquierdo2014; GeaIzquierdo2015; @Dorado2017].
+@GeaIzquierdo2014; @GeaIzquierdo2015; @Dorado2017].
 
@@
 On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019].
-Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014, @JimenezSerrano2004].
+Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014; @JimenezSerrano2004].
 In Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century.
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-
-#### Disturbance
-Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
-
-## historia forestal de ambos sitios 
-Incluimos lo que conocemos de la historia forestal de los sitios??
-Me explico, tenemos datos de manejo y uso antrópico de las dos zonas que proceden de varias fuentes.
-En resumen, mas o menos, tenemos:
-
-* Minería: datos de minería, que afectan sobre todo a la población de SJ (los tenemos localizados temporalmente)
-* Actuaciones forestales: 
-* Incendios, Carboneo, Ganadería, etc --> Tenemos información menos estructurada sobre estos ámbitos, procedentes de varias fuentes, que tienen una incertidumbre mayor espacial y temporalmente. 
-")
-
-```
-
-Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution.
-Para  *Q.
-pyrenaica* se ha visto que moisture availability es el factor limitante que determina el crecimiento en las poblaciones situadas en su rear edge [@GeaIzquierdo2014], pero para otras especies (*i.e.* *Pinus nigra*, *P.
-sylvestris*) la temperatura tiene mas peso que la disponibilidad de agua  [@Herrero2013; @Matias2017].
-Pero es importante considerar además la historia de manejo que han tenido esos bosques a la hora de la forest management [@Penuelas2017; @DoblasMiranda2017], ya que, como observamos en nuestros resultados puede condicionar el crecimiento y la resiliencia de las especies.
+Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013], since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.
+
+En este sentido, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia.
+Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética).
 
@@
 
-")
-
-```
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
+.
+Para  *Q.
+pyrenaica* se ha visto que moisture availability es el factor limitante que determina el crecimiento en las poblaciones situadas en su rear edge [@GeaIzquierdo2014], pero para otras especies (*i.e.* *Pinus nigra*, *P.
+sylvestris*) la temperatura tiene mas peso que la disponibilidad de agua  [@Herrero2013; @Matias2017].
+Pero es importante considerar además la historia de manejo que han tenido esos bosques a la hora de la forest management [@Penuelas2017; @DoblasMiranda2017], ya que, como observamos en nuestros resultados puede condicionar el crecimiento y la resiliencia de las especies.
+
 * Vicente-Serrano: 

### 2019-05-16 man/ms.Rmd
organizar notas

@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-
-In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017].
-The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017].
-These events were extreme in both its magnitude and spatial extent [@Gouveia2014].
-
-These climatic alterations are likely to have important consequences for tree species dynamics at local and regional scales ( Peñuelas and Boada 2003 , Van Mantgem et al.
-2009 , Matías and Jump 2015 ).
-
-### Sequías 
-> notas sequia
-> 
-> * the drought events have been longest and most severe in the period 1991-2010 for mediterranean are of Southern Europe  
-> * Seasonally, drought frequency is projected to increase everywhere in Europe for both scenarios in spring and summer, especially over southern Europe, and less intensely in autumn [@Spinoni2017a]
-> Aumento de la frecuencia de sequías en spring and summer desde 1950 hasta 2014 [@Spinoni2017b]
-> Both for frequency and severity, the evolution towards drier conditions is more relevant in the last three decades over Mediterranean area in summer, 
-> an increase in the drought severity in the Iberian Peninsula has been observed in the last decades [@VicenteSerrano2014].
-
-Althought several works have reported these two years as some of the worst drought events, we characterised the drought at several spatio-temporal scales in the study area.
-From a long-term perspective, we compare the accumulated monthly precipitation at a meteorological station (Granada, Base Aérea) during the hydrological years 2004-2005 and 2011-2012 with the average of accumulated monthly precipitation for the period 1950-2015.
-
-### Extreme sequias y ecologia
-
-No solo efectos aislados, sino también en conjunción con otros factores, sobre todo en el mediterráneo donde convergen muchos de los factores que pueden interaccionar:
-
-* La sequía es un factor crucial a tener en cuenta, ya que es además de los efectos que puede tener de forma aislada, se ha visto que además presenta muchas interacciones con otros factores, siendo por tanto un factor crucial (`$IMPROVE$` ver [@DoblasMiranda2017]):
-* Some interactions alter the effects of a single factor, as drought enhances or decreases the effects of atmospheric components on plant ecophysiology
-* Drought and land use changes, among others, alter water resources and lead to land degradation, vegetation regeneration decline, and expansion of forest diseases.
-* Climate change, and especially drought, emerges as a crucial factor in most of the reviewed interactions and therefore it should be considered when it comes to designing and applying international management policies
-* Drought should be considered when designing and applying management policies.
-
-Además la sequía se espera que tengan ...Droughts are most likely to have the largest and most long-lasting impacts globally due to large indirect and lagged impacts and long recovery especially for forest ecosystems (ver  18 en @Ummenhofer2017).
-Así por ejemplo se ha visto que todos los biomas presentan una vulnerabilidad similar ...
-En una revision sobre la vulnerabilidad del sistema de trasnporte en plantas al embolismo inducido por sequía, ha mostrado que una convergencia de la vulnerabilidad de los bosques a la sequía, mostrando que todos los biomas son igualmente vulnerables a los fallos hidráulicos independientemente del régimen de precipitación (Choat et al.
-2012, doi:10.1038/nature11688)
-
-")
-```
-
 # Materials and methods
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-* Ver ejemplo Gazol et al 2017 Plan Ecology
-* un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
-")
-```
-
 ## Greenness data
@@
 Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-We used EVI instead of NDVI (*Normalized Difference Vegetation Index*; another of the most common greenness vegetation indices) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].
-
-")
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2).
-Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked.
-Besides, VI usefulness indices between 0 to 11 essentially include all EVI data.
-Thus, these two conditions serve as additional checks.
-
-and then a quality assessment was carried out to filter the ... [@Reyes2015]
-
-    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds.
-On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]
-
-The presence of clouds (adjacent clouds, mixed clouds and shadows) 'obscures' the surface in a radiometric sense, thus corrupting inferred EVI values.
-In addition, two types of aerosol loadings typically corrupt EVI—climatology and high aerosols.
-Use of aerosol climatology
-
-The EVI data are geometrically and atmospherically corrected and include information about the quality ass....
-
-These data are geometrically and atmospherically corrected, and include an index of data quality (reliability, which range from 0 – good quality data – to 4 – raw data or absent for different reasons) based on the environmental conditions in which the data was recorded
-
-We first used the Quality Assesment (QA band) information of this product to filter out those values affected by high content of aerosols, clouds, shadows, snow or water; and then a quality assessment was carried out to filter the ...
-[@Reyes2015]
-
-:red_circle: reescribir esto de la calidad. Ver Samanta et al 2012 y como describe el proceso de calidad
-
-")
-```
 
@@
 The alpha level for the test was set to 0.05.
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-(:red_circle: Hemos seleccionado EVI medio, además de por los consejos que me ha dicho Domingo, porque he comprobado que existe una correlación entre el evi medio y el evi estacional, sobre todo el de verano.
-Ver esto: [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md).
-Además presenta alta correlaciones significativas con el EVI de verano: 0.88; de primavera: 0.76 y anual: 0.81)
-
-> Procedimiento de Filtrado de datos (ver [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md))
-
-* Información contenida en banda QA.
-
-    * Nos quedamos con pixeles marcados como Good Data (57.89 %)
-    * Filter out los marcados como Snow/Ice y/o Cloudy (2.57 + 7.08 = 9.65 %)
-    * Pixeles marcados como Marginal Data (32.33 %) (ver siguiente paso)
-
-* Explorar distribución temporal y analizar banda QA Detailed y llevar a cabo un filtrado siguiendo las especificaciones de @Reyes2015.
-
-    * Vemos los composites marcados con Aerosoles, Adjacent cluods, y Shadow.  
-    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds.
-On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]
-
-* Finalmente nos hemos quedado con las siguientes cifras.
-De un total de 360064 images composites for the study zone were downloaded (928 x 20 x 1 + 928 x 23 x 16 = 360064), tras el filtrado, nos quedamos con 286825 (79.65 %)
-
-")
-```
 
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-SPEI values between 1 and −1 are considered normal, whereas values <−1 indicate drought and values <−1.5 indicate severe drought47. We defined a drought as a period where SPEI12 decreased to ≤−1.5.
-")
-```
-
 ### Disturbance analyses 
@@
 By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
-")
-```
 
@@
 To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-Disturbance chronologies were built using tree-ring width to identify historical abrupt positive (releases) or negative (suppressions) changes in growth in each species (Nowacki and Abrams, 1997).
-This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
-")
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-* Para obtener los datos de SPEI, hemos ido bajado los datos desde http://sac.csic.es/spei/map/maps.html 
-* sanjuan: 37.133, -3.382 = spei_-3.25_37.25.csv 
-* canar: 36.957, -3.4284 = spei_-3.25_36.75.csv
-
-OJO ver http://monitordesequia.csic.es/
-")
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-To obtain ring-width series of each individual and species (species chronology) the cross-dated ring-width series were standardized and detrended using the dplR library (Bunn 2010) in the R statistical language (R Development Core Team 2015).
-First, we fitted a spline through the ring-width series of each measured radius.
-We used a spline where the frequency response is 0.5 at a wavelength of 0.67 multiplied by the length of the series in years (Bunn 2010).
-Then, the ring-width data were divided by the fitted lines to obtain residual ring-width indices (RWI) that were subjected to autoregressive modelling (pre-whitening) and then averaged for each year using a biweight robust estimation of the mean (Fritts 1976).
-Calculating RWIs using subtraction instead of division rendered similar results as those here presented.
-Thus, we obtained mean residual chronologies for each species of prewhitened growth indices.
-
- * BAI represents a more accurate indicator of growth than ring-width, since it reduces the variation caused by adding volume to a circular stem 
- * BAI is a more meaningful indicator of growth than ring width because it removes variation in growth attributale to increasing circumference. 
- * BAI is more closely related to biomass increment 
- * BAI removes variation in growth attributatle to increasing stem circumference and captures changes in growth better than linear measures as tree-ring width [@Biondi2008] 
-")
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
- To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
-")
-```
 
@@
 Resilience (*Rs*) = Postdrought / Predrought
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-The Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance, and it is estimated as:
-
-Relative Resilience (*RRs*) = (Postdrought - Drought) / Predrought
-")
-```
 
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends.
-The strongest trends were observed in southwestern populations (Figure S2), in which the Cañar site are included Vegetation greenness of *Q.
-pyrenaica* forests were lower during both the 2005 and 2012 years than the greenness observed for the reference period (Figure 2a).
-The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b).
-Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought (Figure 2b).
-
-According to the standardized anomalies, *Q.
-pyrenaica* forests suffered a browning episode during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S3).
-")
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-> Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005.
-Sin embargo vemos valores menos negativos para 2005.
-Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years
-
-> Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029) que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
-")
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-## Greenness resilience to drought events
-
-*Q.
-pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3).
-Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figure 3).
-
-Recovery (*Rc*) of greenness was significnatly different bewteen drought events and sites (Tables 3).
-In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
-For southern populations, a similar recovery after the 2005 and 2012 drought event was observed (p = 0.2453; Figure 3; Table S1).
-
-Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables S1, 3), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1).
-The southern populations showed higher resilience values than the northern ones, although these differences were not significant for the 2005 drought event (p = 0.036; Figure 3).
-
-> Resistencia
-> 
-> * Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001]
-> * Menor resistencia de las poblaciones del Norte a los eventos de sequía que las del Sur [N: 0.883 (0.877-0.889); S: 0.921 (0.918 - 0.925); p <0.0001]
-> * La resistencia varió en función de la sequía y de la población.
-Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
-> * notas:  La sequía de 2005 redujo el EVI medio hasta el 85.8%, mientras que la de 2012 la redujo hasta el 94.3%.
-La sequía de 2005 redujo el evi medio hasta el 81.9 % en las poblaciones del N, y hasta el 90.2 % en las del S.
-...
-drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)
-
-> Recovery 
-> 
-> * La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
-> * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
-> * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001].
-En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte.
-
-> * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
-> * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152).
-En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
-
-> Relative Resilience (De esto no he dicho nada)
-
-  * Los robledales mostraron mayor resiliencia relativa a la sequía de 2005 que a la de 2012 [2005: 0.099 (0.095-0.105); 2012: 0.053 (0.050 - 0.056); p <0.0001]  
-  * Las poblaciones del sur mostraron menor resiliencia relativa que las del norte [N: 0.086 (0.082-0.092); S: 0.063 (0.060 - 0.066); p <0.0001], debido sobre todo a la diferencia en la resiliencia relativa para las poblaciones del norte entre los dos eventos de sequía (mucha mayor resiliencia relativa en 2012 que en 2005 para las poblaciones del N)
-  * Las poblaciones del sur no mostraron diferencias en cuanto a la resiliencia relativa entre los dos eventos de sequía (padj = 0.152)
-")
-```
-
 ## Radial growth trends and growth disturbances
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-Sorprende que en los sitios del sur (caH y caL), en los últimos años ha habido crecimientos muy grandes.
-Por ejemplo, en 2010 se observó uno de los mayores valores de RWI de toda la cronología tanto en caH como en caL.
-Algo similar ocurrió en sj para los años 2003 y 2013.
-")
-```
-
 The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4).
@@
 Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression events in the last 50 years.
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-
-En las poblaciones del norte osbervamos dos eventos de PGC por encima del 25 e incluso el 50 %.
-El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001.
-Estos periodos se alternan con periodos de NGC: previo 1940, 1960 - 1991, y quizá aunque no lo vemos en al actualidad (??
-duda.
-Esto es especular un poco).
-Esto responde creo que a:
-
-(i) actividad minera y tratamientos forestales.
-La actividad minera en la zona fue intensa (aunque a veces intermitente, entre 1890 y 1957 (cuando se clausuró); y previamente desde 1858 tras los trabajos de Amalio Maestre).
-Hay costancia de la existencia de hornos de fundición de mineral en alguna de las minas (La Probadora).
-(ii) actividades forestales:
-* Se conoce la existencia de un proyecto redactado en 1943 de repoblación forestal de la cabecera del río Genil (he pedido los mapas).
-* Ver actividades forestales (bd proyectos de gestión)
-
-En el trabajo de [@GeaIzquierdo2014] se observa una liberación posterior a 1950 (QPUY9) (entre 1950 y 1960 aprox.) y luego una supresión hasta mediados de 1990.
-Esto coincide con lo que hemos observado en nuestra chrono.
-
-Respecto a las poblaciones del sur, destaca en la caH una primera liberación en torno a 1829-1830 y 1837-1840; y luego predominan los eventos de supresión que se alternan con algunos de liberación (muy débiles) hastas 1950.
-Posteriormente se alternan liberaciones-supresiones débiles.
-Destaca una liberación en torno a 2000-2003 (relación con clima???
-posterior a sequía 1999???)
-
-En caL, observamos un periodod de limitación hasta 1885 aprox.
-Luego se alternan al igual que antes periodos de liberación con otos de supresión con señales más fuertes que en caL, y al igual que antes destaca con mas intensidad si cabe, una liberación entre 2000 y 2003 (antrópico o climática ??)
-
-Duda: ocurre en los dos sitios caL y caH, entonces, es aumento de crecimiento debido a actuación forestal? o debido a clima??
-
-[@GeaIzquierdo2014] encontró un pico de disturbance en torno a 1860 en QUPY10 (cañar) sugiriendo que la baja densidad en esos sitios fue presumible el resultado de un thinning (aclareo).
-El patrón de supresión que observa para este sitio va en la linea de lo que nosotros encontramos (supresiones que alcanzan a veces el 25 % y otras no)
-
-En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles.
-En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles
-")
-```
 
@@
 In opposite, Resistance and Resilience values were higher for southern sites than for northern ones (Tables S1 and S2).
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-> Resistance:
-
-Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2).
-Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3).
-
-* No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012 (F = 6.0189; p = 0.019).
-Valores de resistencia menor al evento de 2005.
-[2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.025; trimmed-p = 0.019]
-
-* Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001).
-Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].
-
-* La interacción también fue significativa.
-De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ???
-
-> Recovery 
-> * Mayor recuperación del crecimiento de los robledales tras la sequía de 2005 que tras la de 2012 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
-> * Las poblacionees del sur (caH y caL; no diferencias entre ellas) mostraron una menor recuperación que las poblaciones del norte, es decir, las del norte se recuperaron mas rápidamente que las del sur que no llegaron incluso a valores de Rc = 1.
-
-Similar pattern of resilience (*Rs*) values was found for growth than for greenness respect to drought event: significantly higher values of *Rs* for the 2012 drought event than for the 2005 (Table S2; Figure 3).
-However, no differences were observed between sites (Table 3).
-For 2005 drought event, *Rs* value of SJ (northern site) was lower than that of southern ones (CA-High and CA-Low), but opposite pattern was found for the 2012 drought event.
-All valu`es of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031).
-
-YA ESTOY DESESPERADO CON ESTE PAPER
-*Q.
-pyrenaica* forest showed significantly lower resistance values (*Rt*) to the 2005 drought event than to that in 2012 for greenness and for radial growth (Table 3; Figure 7).
-The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2).
-Resistance values to drought for greenness and tree-growth varied between sites (Table 3).
-Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2).
-It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 7).
-
-Recovery (*Rc*) of greenness and growth were significantly different bewteen drought events and sites (Table 3).
-In the 3-year period after the 2005 drought, greenness was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1).
-An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 7, Table S2).
-
-Northern populations showed significantly higher values of recovery than southern sites for greenness and tree-growth (Table S2).
-For southern populations, no significant differences were found for recovery of greenness after the 2005 and 2012 drought event (p = 0.2453; Figure 7; Table S1).
-Recovery values for tree-growth of southern populations were below or close to 1 (Figure 7, Table S2).
-
-Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 7).
-Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3).
-Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2).
-For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 7).
-For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2).
-
-Similar to the long-term observed pattern,
-
-was greater after more severe drought event (2012). Opposite pattern was observed for greenness with
-
-Lower resistance values (*Rt*) were observed to the 2005 drought event than to that in 2012 for both variables.
-The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2).
-Resistance values to drought for greenness and tree-growth varied between sites (Table 3).
-Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2).
-It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 7).
-
-") 
-```
 
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-
-OOOOOOOOOOOJOOOOOOOOOOO CON ESTO
-
-Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
-pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017]
-
-Por un lado la dependencia del agua. Sabemos que esta especie es mas sensitive a la humedad del suelo que otras especies arbóreas mediterráneas (GeaFEM). En Sie
-
-A plausible explanation could be
-
---> Ver y resumir páginas 18-19 Notas Naturkunde
-
----> Nota
-
-Para la discussion: 
-- Resilience values of RW for 2005 was the lowest of the drought events analized even not being the moste severe drought event 
-- 2005 reduced greenness and bai more than 2012 
-")
-```
-
 Greenness and tree-growth of *Q.
@@
 On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q. pyrenaica*.
-
-During the 2005 drought, one of the worst drought events recorded in the Iberian Peninsula [@GarciaHerrera2007], we found a browning of the *Q.
-pyrenaica* forests, but no changes in EVI standardized anomalies were recorded for 2012, which can be explained because 2012 drought event was a winter-drought [@Trigo2013].
-
-Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure S3).
-The decline in growth observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018].
-
-Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of the forests of *Q.
-pyrenaica* located in their rear edge.
-Our results agree with those obtained by previous works using other remote-sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi], which suggests an increase in primary productivity during the last years for rear-edge populations of this species.
-
-For tree-growth, a similar positive trend was observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
-This result differs from those previous reported for *Q.
-pyrenaica* along their distribution range [@GeaIzquierdo2013; @GeaIzquierdo2014].
-@GeaIzquierdo2014 found a general decline in the growth of this species since the 1970s, particularly sharp for populations located in their dry-edge.
-This decline trend in growth have also been oberved for other tree species located in their rear-edges [*e.g.* @SanchezSalguero2012; @SanchezSalguero2017; @Dorado2017].
-Furthermore, growth projections have forecasted a decrease in productivity for *Q.
-pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013].
-However, our results for similar locations have revealed a recovery in tree-growth for the last years (Figure 4).
-In addition, a similar positive trend in BAI for the last years, has been reported for another *Q.
-pyrenaica* population of the Sierra Nevada [@RubioCuadrado2018].
-
-Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 7).
-La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Table S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q.
-pyrenaica*.
-Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía (long-term perspective), observamos como estas poblaciones presentan altos valores de resiliencia.
-Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Figure 6).
-") 
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-Hemos observado como las sequías severas, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento de *Q.
-pyrenaica*.
-Los datos de anomalías estandarizadas de EVI mostraron un browning para la mayoría de los robledales de Sierra Nevada durante la sequía de 2005.
-
-La sequía de 2005 ha sido una de las mas severas de los últimos años, afectando significativamente al crecimiento
-
-#--------------------------------------------
-# Sequías
-
-### 2005 frente a 2012
-
-* Analizando los datos de precipitación acumulada de las estaciones alrededor de SN, observamos como 2005 fue el año mas seco desde 1950. 
-* Datos de SPEI (0.5 º Grid) 
-    * 2005 fue peor que 2012, pero no peor que 1995 o que 1999 (ver figura S2)
-    * Tabla S4. Severidad sequías, la de 2005 fue menos severa y menos intensa y de menor duración que la de 2005 (de donde vienen estos datos???)
-
-* La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
-Esto se ha observado también en otras especies en el sur de la P.
-Ibérica (p.ej.
-P.
-nigra en Andalusia [@SanchezSalguero2013], ...
-INCLUIR otras especies y citas).
-
-### 2005 en contexto
-
-* Algunos trabajos hablan que los peores años de sequía en la P. Iberica durante la segunda mitad del siglo XX fueron 1986, 1994-1995, 1999, 2005 y 2012
-
-* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].
-
-### Reducción en el crecimiento de algunas especies 
-El efecto de la sequía de 2005 en el crecimiento también se ha visto en otras especies en el sur de la P. Ibérica 
-La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento.
-Esto se ha observado también en otras especies en el sur de la P.
-Ibérica (p.ej.
-P.
-nigra en Andalusia [@SanchezSalguero2013], ...
-INCLUIR otras especies y citas).
-
-* P.
-nigra: Se observó una gran reducción de crecimiento en 1994-1995, 1999 y 2005 para Pinus nigra en el SE de Andalusia [@SanchezSalguero2013] coincidiendo con años de sequía para esa zona.
-La máxima reducción en crecimiento fue observada para 2005, siendo significativamente mayor que para el resto de años [@SanchezSalguero2013].
-
-No obstante, a pesar de que 2005 y 2012 fueron dos de las sequías mas severas registradas para el sur de la P.
-Ibérica [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), nuestros resultados muestran que los robledales de *Q.
-pyrenaica* en su rear-edge presentan una tendencia positiva tanto en el greenness (78.9 % pixeles con greeness) como en el crecimiento (tendencias positivas del BAI).
-
-### La resiliencia de las poblaciones de roble de Sierra Nevada.
-
-El greennness de los robledades en Sierra Nevada ha mostrado una tendencia positiva hacia mas verdor en los ultimos años, que coincide con lo que ya observamos con datos de NDVI [@PerezLuque2015onto], sobre todo para las poblaciones del sur.
-Esta tendencia positiva, obtenida de variables derivadas de remote sensing (EVI y NDVI), parece que también se observa en el crecimiento.
-Por ejemplo en las poblaciones del sur (CaHigh y CaLow) observamos en los últimos años una ligera tendencia de crecimiento.
-
-Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia.
-Por ejemplo para las poblaciones del norte, los valores de EVI durante la sequía de 2005 descendieron hasta el 81 %, mientras que el BAI (sitio SJ) lo hizo hasta el 45 %.
-Sin embargo, la recuperación fue rápida, así los valores de Recovery (Rc) para el EVI tras la sequía de 2005 (en las poblaciones del norte) fue de 1.17, mientras que para el BAI el Rc fue de 1.112.
-En definitiva, estamos observando altos valores de resiliencia en estos robledales.
-
-Despite the resilience values (*Rs*) of growth for 2005 and 2012 were close or below 1, we observed high resilience of these forests from a long-term perspective.
-As our results indicate, the highest resilience values for growth were recorded for two of the most severe drought events (1995 and 1999), with resilience values above 1.2 in both events (Appendix S5).
-
-genetica
-
-Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático.
-Hemos observado un aumento en el greenness en los últimos años.
-Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness.
-Además, estos robledales tienen una alta resiliencia genética  ¿Sierra Nevada (regiones de montaña) como refugio??
-quizá este rear-edge esté actuando como refugio??
-(esto es muy especulativo)
-
-Las zonas rear edge son zonas que se supone son mas vulnerables al cambio climático por sus estrechos márgenes.
-De hecho como han visto algunos trabajos, las especies situadas en su rear edge son mas vulnerables.
-No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado).
-Además habría que introducir aqui lo que M.
-Villalta habla de la consideración de rear edge de.
-
-* Otro punto de interés a introducir aquí es que frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], sin embargo algunos estudios están demostrando esto no es siempre así (ver por ejemplo [@Cavin2017; @Granda2017]).
-Nuestros resultados creo que van en esta línea creo.
-Esto, como apunta [@MartinezVilalta2018]m tiene que ver con la que consideramos como habitat marginal de la especie (...
-When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).
-
-¿Casos parecidos?
-
-Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling..
-// dorado-liñan // sanchez salguero
-
-trabajos rear edge: 
- * sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
- * sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
- * dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
- * herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
- * dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
-buscar otros
-
-...
-Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia...
-
-## refugio??
-
-Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático.
-Hemos observado un aumento en el greenness en los últimos años.
-Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness.
-Además, estos robledales tienen una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
-¿Sierra Nevada (regiones de montaña) como refugio??
-quizá este rear-edge esté actuando como refugio??
-(esto es muy especulativo)
-
-pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.
-
-Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005].
-Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas).
-Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q.
-pyrenaica* situadas en su rear-edge.
-
-Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018].
-En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q.
-pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.
-Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
-pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
-
-")
-```
 
@@
 
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-Podríamos complementarlo con lo que le pasa a otras especies en su borde de distribución: por ejemplo en Baza, Herrero et al.
-2013, encontraron para Pinus nigra y sylvestris que la temperatura tenía mas peso que la disponibilidad de agua).
-O también ver algunos de los trabajos de Camarero et al 2013 para el P.
-nigra en su borde de distribución u otros similares (el de Sanchez-Salguero et al.
-2013, 2015)
-
-Para *Q.
-pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q.
-pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al.
-2015 European Journal of Forest Research).
-Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea.
-
-.
-Para  *Q.
-pyrenaica* se ha visto que moisture availability es el factor limitante que determina el crecimiento en las poblaciones situadas en su rear edge [@GeaIzquierdo2014], pero para otras especies (*i.e.* *Pinus nigra*, *P.
-sylvestris*) la temperatura tiene mas peso que la disponibilidad de agua  [@Herrero2013; @Matias2017].
-Pero es importante considerar además la historia de manejo que han tenido esos bosques a la hora de la forest management [@Penuelas2017; @DoblasMiranda2017], ya que, como observamos en nuestros resultados puede condicionar el crecimiento y la resiliencia de las especies.
-
-* Vicente-Serrano: 
-    * We found that some forests from cold and humid areas respond to shorter drought time-scales than forests from dry areas, which usually respond to longer time-scales
-    * Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011).
-    * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.
-
------------
-
-Despite similar value of competence in our study sites, we obtained higher values of growth for the highest elevation site (CA-High, Figure 4).
-
-------
-
-En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations).
-The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone.
-
-")
-
-```
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-## Rear-edge
-
-* Se asume una alta vulnerabilidad de las especies a la sequía en su límite inferior, sin embargo esta vulnerabilidad no necesariamente se mantiene: The results of Granda et al.
-2018 are consistent with recent studies showing that the frequently assumed higher vulnerability of dry, rear-edge populations to drought does not necessarily hold (e.g.
-Cavin and Jump 2017) (from: Martinez-Vilalta 2018)
-
-* Definición de *marginal* populations.
-[@MartinezVilalta2018] indica que a veces no está claro la definición de marginal populations.
-When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors).
-
-* Quizá estamos asumiendo que es una población marginal, pero sin embargo lo es solamente geográficamente
-
------
-
-No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado).
-
-### Diferencias entre sitios
-
-Llama la atención que en el rear-edge de la distribución de la especie, en el sitio mas meridional dentro de este rear edge, y en la parte mas alta (estas parcelas están en el treeline de la especie en SN (en torno a 1900)), es donde encontramos mayores crecimientos.
-Parece que los robles estuvieran mejor en esta zona.
-Algunas reflexiones sobre esto:
-
-* Quizá están creciendo donde les están dejando (menor impacto antrópico en las zonas mas altas, al menos en los últimos años??)
-* ¿Diferentes niveles de compentencia?. Parece que la competencia es similar en los tres sitios: no diferencias sig. para valores de Stand density ySize ratio proportional to distance (ver table 1). 
-* ¿Diferencias entre suelos??
-En principio no.
-[@CoboDiaz2017] en un trabajo sobre microbiota del suelo, explora diferencias en un gradiente altitudinal en Cáñar.
-Analiza 3 sitios: por encima del treline (XZF, el piornal), low altitudinal oak forest (LAF, en la zona baja del robledal) and high altitudinal oak forest (HAF).
-En su trabajo, LAF está mucho mas bajo, pero HAF está cerca de las parcelas caHigh.
-Restulados:
-    * tipo de suelo: LAF es Sandy-loam, HAF es loam
-    * HAF tuvo los valores mayores de disponibilidad hídrica (% available water tabla s1 en [@CoboDiaz2017]) --> Esto es importante, creo
-    * Los tres sitios eran pobres en materia orgánica, pero el HAF dobló los valores de los otros dos sitios
-    * No diferencias en contenido de N, C/N ratio similares en HAF, y LAF
-
-* ¿Menor estrés hídrico en zonas mas altas?
-Puede ser que tengamos menor estrés hídrico en esta zona??
-Este robledal, está en la cara sur de SN, y en una ladera con bastante insolación.
-Algunos trabajos antiguos (de fitosociólogos) hablan de que su presencia aquí se debe a que reciben un aporte extra de humedad procedente de las brisas del mediterráneo, para suplir el mínimo de humedad que necesitan en verano.
-Quizá también tendríamos que incluir el papel de las acequias.
-La zona caHigh tiene una acequia muy cerca (y por encima) de donde muestreamos (recordad el roble mas grande, y el mas alto, esta justo en una acequia).
-No se si es interesante que lo comentemos.
-
-In addition, despite southern sites are very close to each other, we obtained higher values of BAI for the high-elevation site than for the low-elevation site.
-
-")
-```
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-#### Climate and tree relationship
-
-Frase de resultados que tenemos que poner en discussion: 
-Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
-Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for the northern population (r > 0.6), ...
-**which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]**.
-")
-```
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-### ¿Que factor es mas limitante para el crecimiento en el rear-edge de Q. pyrenaica? 
-Aquí pueder ser interesante comentar algo de el peso de las variables climáticas en el crecimiento para poblaciones situadas en el borde de distribución (pesa mas la temperatura o la disponibilidad de agua?).
-
-Para *Q.
-pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q.
-pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al.
-2015 European Journal of Forest Research).
-Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea.
-
-Podríamos complementarlo con lo que le pasa a otras especies en su borde de distribución: por ejemplo en Baza, Herrero et al.
-2013, encontraron para Pinus nigra y sylvestris que la temperatura tenía mas peso que la disponibilidad de agua).
-O también ver algunos de los trabajos de Camarero et al 2013 para el P.
-nigra en su borde de distribución u otros similares (el de Sanchez-Salguero et al.
-2013, 2015) …
-
-## MAs cosas diferentes 
-Otras especies: rear edge
-
-Evaluar el crecimiento en el rear edge de F. sylvatica 
-https://link.springer.com/article/10.1007/s10342-016-0982-7?wt_mc=Internal.Event.1.SEM.ArticleAuthorOnlineFirst
-
-¿Que limita al crecimiento en el borde sur de su distribucion? --> Ver esto 
-http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2486.2006.01250.x/abstract
-
-Pinus sylvestris (Baza) (Matias et al 2017)
-Radial growth was maximal at medium altitude and treeline of the southernmost populations.
-Temperature was the main factor controlling growth variability along the gradients, although the timing and strength of climatic variables affecting growth shifted with latitude and altitude.
-
-http://onlinelibrary.wiley.com/doi/10.1111/gcb.13627/full
-
-Idea --> algunos autores comentan que puede existir una alteración en el balance competitivo entre especies en mixed stands.
-Por ejemplo en Montseny, borde equatorial para F.
-sylvatica, se está viendo que el F.
-sylvatica está siendo reemplazado por Q.
-ilex.
-...
-En SN, el artículo de B.
-Benito (Climatic Change) habla del remplazo que existirá de Q.
-pyrenaica por Q.
-ilex, sin embargo, estamos viendo que los crecimientos son muy grandes y que Q.
-pyrenaica tiene mucha resiliencia --> entonces que pasa con las predicciones de dichos modelos ???
-
-The presence of *Q. pyrenaica* forests on the southern slopes of Sierra Nevada is partly explained by the additional contribution of humid air from the Mediterranean sea [@Prieto1975]
-
-")
-```
-
-En este sentido, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia.
-Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética).
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-
-### ORDENAR ???
-
-.. ver [@VicenteSerrano2013] impacto de las sequías en diferentes zonas (i.e montañas)
-
-# Otras notas: 
-* To guide sustainable forest management, forest researchers are asked to provide concrete answers about forest resilience in response to expected climatic trends, and extreme climatic events (Lindner et al., 2014) https://www.sciencedirect.com/science/article/pii/S030147971400379X?via%3Dihub
-
-* Proyeccciones de crecimeinto de para *Q.
-pyrenaica* sugieren un declive en el crecimiento en las proximas decadas a lo largo de su rango de distribución en la P.
-Iberica, en diferentes escenarios climáticos [@GeaIzquierdo2013]: Predictions suggest that QUPY productivity would decline in the next decades all along its distributional range in the Iberian Peninsula for all the climate scenarios studied.
-Este declive debería ser mas dramático en las baja altitudes de las zonas mas calientes del sur de su distribución
-
---> vulnerabilidad local (related to rear-edge)
-La tendencia de reducción drástica sugerida por el modelo para el dry-edge de la distribución de la especie podría expresar una aumentada vulnerabilidad de los árboles al incremento del estrés hídrico forzado por un clima mas cálido [@GeaIzquierdo2013].
-
-Negative trends over recent years and the greater response to moisture availability found at warmer and drier low elevations in the south suggest vulnerability to warming at the local low elevation dry edge of the species's range (e.g.
-QUPY9).
-Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()
-
-Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg.
-@Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al.
-https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI).
-Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).
-
-Case studies that focus on multiple scales - including local scales - and also valuables (Dale et al.
-2018 Frontiers) --> Se necesitan casos de estudio que se enfoquen en escalas multiples (incluyendo las locales) y que combinen varias metodologías (Jump et al.
-...)
-
-3) rear edge
-
-* Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ...
-(Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ...
-
-### Estudios que utilizan RS y TR
-
-* Wu et al. 2017. 
-	* En un estudio sobre el efecto de la sequía sobre el crecimiento en bosques, grass y shrub del hemisferio norte, encontraron que los bosques (Deep-rooted forest) exhibit a drought legacy response with reduced growth during up to 4 years despues de un evento extremo de sequía.
-
-	* Además vieron que los bosques mostraron una stronger drought resilience in forest (i.e. smaller growth reduction after severe drought) (Gazol el al. 2017; Wu et al. 2017).
-
-	* Esto se puede deber a que acceden a reservas de agua mas profundas (los pastos y los matorrales no pueden acceder a ese agua)
-
-###
-
-@Gazol2018 encontró La resiliencia está relacionada con la severidad de la sequía y con la composición del bosque
-
-Rt-NDVI and Rt-TR were positively related with drought intensity
-
-## 
-@Gazol2018 Resistencia a la sequía mayor en la zona norte de españa que en las zonas secas del sur
-
-Qpyr 
-* Valores de Rt-TR aumentaron a lo largo del tiempo, sin embargo Rt-NDVI dismiuyó con el tiempo
-* Valores Rc-NDVI tendencia positiva
-
-")
-```
-
-```{r echo=FALSE, results='asis', eval=show_text}
-cat("
-# Notas Jump 2010 
-Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)
-
-NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.
-
-Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!
-
-Otra idea --> Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)
-
-NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales.
-
-Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!
-
-El NDVI se ha utilizado para monitorizar la respuesta de bosques frente a sequía (ej.: 24,26,32, 44 en Jump2009 )
-
-Necesidad de Ground-based assessment (p 1974 Jump 2009) -> Aunque utilizemos datos de satélite, siempre es necesario utilizar aproximaciones combinda
-
-Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
-(ver esto y completar)
-
-El problema de los datos ground-based assessment, existe un trade-off entre el área total cubierta por una red de monitoreo y su resolución local.
---Por tanto también existen limitaciones de las ground-based para estudiar poblaciones en su rear edge.
-")
-```
-
 ## References

### 2019-05-16 man/ms.Rmd
update

@@
 Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge.
-Furthermore, the coldest stands show a positive growth trend since the end of the 1970s matching the simultaneous increase in temperature observed.
+Furthermore, all stands show a positive growth trend (stronger in high elevation stand) since the end of the 1970s matching the simultaneous increase in temperature observed.
 
-* Resilience indices showed ecological coherence along a gradient of drought severity in time.
-In addition, oak resilience to drought followed a dryness gradient among the three populations studied: Northern populations thriving under a drier climate were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.
+* Resilience to drought followed a dryness gradient among the three populations studied: Northern populations thriving under a drier climate were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.
 
 * Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought.
+
 Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e.
@@
 
-Several studies analyzed the drought effects on Mediterranean tree species representing the southernmost limit of their distribution
-[*e.g.*, @SanchezSalguero2012; @Linares2014; @Dorado2017b; @Caminero2018].
+Several studies analyzed the drought effects on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Linares2014; @Dorado2017b; @Caminero2018].
 Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
@@
 We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy).
-Our study sites had different land-use trajectories driven by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014].
-
-In Northern sites (e.g.
-San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO].
-Grasslands and shrublands for cattle farming are located at high elevations.
-Then forests formation with some croplands; and then, at lowlands, irrigated terraces with tree crops.
-Another activities could conditioning the natural resources uses.
-For instance, the mining activity in San Juan site have strongly conditioned the forest structure.
-This oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas).
-Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita].
-
+The review of historical documents revealed that our study sites had different land-use trajectories driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014].
+On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [Catalogo Paisajes]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops.
+Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure.
+For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history [quarries].
+Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [mineria1950].
 On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019].
-Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014; @JimenezSerrano2004].
-In Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century.
+Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@ValbuenaCarabana2014; @JimenezSerrano2004].
+In Cáñar site, there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century.
 From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].
 
-The different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5).
-We observed a tree-growth release event around 1940 in the SJ oak woodland (Figures 5, S4) which concurs with one period of mining activity for this area.
-Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections (CITA).
+Those different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5).
+We observed a tree-growth release event at 1940s in the SJ oak woodland (Figures 5, S4) which concurs with one period of maximum mining activity for this area.
+Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections [tranvia].
 During this period there were an increase on the use of timbers for the tunnels of the mines (CITE) and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area (cite).
@@
 
-Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013], since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.
+Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013; @Ameztegui2016] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.
+Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource.
+
+Conclussions
+
+Severe drought events have provoked reductions in primary and secondary growth *Q. pyrenaica* populations located in the rear-edge of their distribution.
+
+Despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q.
+pyrenaica* in Sierra Nevada mountain range.
+Surprisingly we also observed
+
+Those trends could have relation with the
+
+A corto plazo, la tendencia positiva observada tanto para el crecimiento primario como para el secundario podría tener relación con varios factores. Por un lado
+
+La tendencia creciente del crecimiento en el crecimiento observada en los últimos años podría relacionarse con un abandono de las practicas tradicionales en esta region de montaña, así como 
+el abandono de las prácticas tradicionales así como por el caracter de re
+
+El abandono de las masas de Q. pyrenaica ha provocado un estado general de degradación, que puede tener consecu
+
+La reducción de la presión antrópica durante las últimas décadas ha provocado que estas bosques presenten un avanzado estado de degradación
+
+After at least five decades of abandonment, old Cs of Q. pyrenaica in Spain present a general state of degradation
+
+In this work, molecular tools have been successfully applied to assess levels of genetic diversity in Cs, rejecting one of the most widespread and preconceived idea in Spanish silviculture: the decline in genetic diversity promoted by historical coppicing in Q.
+pyrenaica.
 

### 2019-05-17 man/ms.Rmd
update ms

@@
 For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings.
-Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@GomezCruz1992].
+Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@Cruz1991].
 These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
@@
 The review of historical documents revealed that our study sites had different land-use trajectories driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014].
-On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [Catalogo Paisajes]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops.
+On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [@2015Zoido]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops.
 Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure.
-For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history [quarries].
-Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [mineria1950].
+For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history [@Navarro2014].
+Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@Maestre1858]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [@2009MartinMartin].
 On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019].
-Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@ValbuenaCarabana2014; @JimenezSerrano2004].
+Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@Valbuena2013; @JimenezSerrano2004].
 In Cáñar site, there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century.
 From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].
+Acorn resources have been tradiotionally exploited in this location up to 1950s [@MesaTorres2009; @Bonet2014_conama].
 
@@
 We observed a tree-growth release event at 1940s in the SJ oak woodland (Figures 5, S4) which concurs with one period of maximum mining activity for this area.
-Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections [tranvia].
-During this period there were an increase on the use of timbers for the tunnels of the mines (CITE) and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area (cite).
+Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections [@Titos1990].
+During this period there were an increase on the use of timbers for the tunnels of the mines and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area [@Titos1990].
 This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4).
@@
 The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4).
-We revised the forest practices carried out in this area in the last 30 years [CITA], but we did not find any clearing or cutting event during 1995 - 2000 period.
+We revised the forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest], but we did not find any clearing or cutting event during 1995 - 2000 period.
 Two non-mutually exclusive ways could explain the tree-growth release observed.
@@
 
+The only positive release events showed for CA-High site could be related with converiosn from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013; @2015VilaTraver].
+
 Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013; @Ameztegui2016] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.
 Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource.
+This pattern concurs with a forest expansion into marginal abandonded croplands and an increase in the tree-cover [@CamachoOlmedo2002].
 

### 2019-05-21 man/ms.Rmd
update

@@
 Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge.
-Furthermore, all stands show a positive growth trend (stronger in high elevation stand) since the end of the 1970s matching the simultaneous increase in temperature observed.
+Furthermore, we found a positive growth trend (stronger in high elevation stand) since the end of the 1970s matching the simultaneous increase in temperature observed.
 
@@
 In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005].
-They are considered relictic forests [@Melendo2000; @Vivero2000], and the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blanca1998; @Lorite2009].
+They are considered relictic forests [@Melendo2000; @Vivero2000], and the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blancaetal1998BC; @Loriteetal2008ABG].
 These forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
@@
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
-All analyses were carried out in R software [@R2017] using several packages (see Appendix S8).
+All analyses were carried out in R software [@R2017].
 
@@
 Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018].
-Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisio2012].
+Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisioetal2012IPoGEC].
 On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
@@
 These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
-@GeaIzquierdo2014; @GeaIzquierdo2015; @Dorado2017].
+@GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].
 
@@
 
-Conclussions
+# Conclussions
 
-Severe drought events have provoked reductions in primary and secondary growth *Q. pyrenaica* populations located in the rear-edge of their distribution.
+Using both dendroecological estimates of growth and MODIS satellite images as proxies of secondary and primary growth respectively, we analyzed forest resilience to severe droughts of *Q.
+pyrenaica* forests located at rear edge of their distribution.
+Severe drought events have provoked important reductions in primary and secondary growth of *Q.
+pyrenaica* forests.
+However, despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q.
+pyrenaica* in Sierra Nevada mountain range.
+Surprisingly we observed high values of resilience to drought, particularly at long-term scales.
+This results are consistent with research showing high genetic resilience values [@Valbuena2013; @Valbuena2017], and overall suggest this mountain region still acting as refuge for deciduous *Quercus* to climate changes [@Brewer2002; @Olalde2002; @RodriguezSanchez2010].
 
-Despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q.
-pyrenaica* in Sierra Nevada mountain range.
-Surprisingly we also observed
+The differences observed in tree-growth and resilience to drought between very close sites seem to confirm that responses to drought are site-dependent [@Babst2013; Cavin2017; @Dorado2017c].
+This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia [@VilaCabrera2019].
 
-Those trends could have relation with the
+The analysis of tree-growth dynamics revealed suppression and release events that are consistent with socieconomic dynamic, derived from the review of historical documents.
+Recently researchs have combined information from both tree-ring and remote-sensing to analyze the resilience of forest to drought [*e.g.* @Gazol2018].
+However our approach, which includes also an exhaustive review of historical documents, helps to disentangle the impacts of climatic and land use for forest systems.
+In this sense our results highligth the importance of landuse legacies for highly transformed Mediterranean systemas [@Navarro2013].
 
-A corto plazo, la tendencia positiva observada tanto para el crecimiento primario como para el secundario podría tener relación con varios factores. Por un lado
+Growth projections have forecasted a decrease in productivity for *Q.
+pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013].
+However, the positive trens for both primary (*i.e.* greenness) and secondary growth (BAI) observed ar our study site and the high resilience values to drought showed at long term scale, seem to suggest that Sierra Nevada constitutes still a refuge for some species.
+This is relevant for tree species with higher sensitivity to climate change, such us *Q.
+pyrenaica* (García-Valdes et al.
+2013; Benito-Garzón et al.
+2008), no solo por la conservación de la especie en sí, sino por los servicios ecosistémicos que estos bosques tan sigulares ofrecen.
 
+En este sentido, los futuros trabajos deberan ir en ...
+
+VAS POR AQUI
+
+En estas zonas refugio en particular parece tener mucha importancia el abandono de las actividades tradicionales, que por un lado y de acuerdo a nuestros resultados parecen estar condicionando el crecimiento
+
+Future work:
+
+C5 Servicios ecosistémicos
+
+Aumento de crecimiento 
+Refugio para biodivesidad 
+Pero ojo ---> estancamiento y/o mortalidad?? Necesidad de prácticas de gestión que aseguren un crecimiento viable, promoviendo la regeneración sexual (evitar el estancamiento) para asi mantener los 
 La tendencia creciente del crecimiento en el crecimiento observada en los últimos años podría relacionarse con un abandono de las practicas tradicionales en esta region de montaña, así como 
@@
 
+Además las diferencias en el crecimiento entre los dos sitios mas cercanos así como lo de la resiliencia, van en la línea de lo expresado por Vilá referente a lo de microrefugios, por lo que la gestión de estas masas, situadas en su rear edge, además ha de analizarse teniendo en cuenta este aspecto.
+
 El abandono de las masas de Q. pyrenaica ha provocado un estado general de degradación, que puede tener consecu
-
 La reducción de la presión antrópica durante las últimas décadas ha provocado que estas bosques presenten un avanzado estado de degradación

### 2019-05-21 man/ms.Rmd
update

@@
 title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
-author: "AJ Perez-Luque; G. Gea-Izquierdo; Zamora, R. ..."
+author: "A.J. Perez-Luque; G. Gea-Izquierdo and Zamora, R."
 output:
@@
 * Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought.
-
 Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e.
@@
 The analysis of tree-growth dynamics revealed suppression and release events that are consistent with socieconomic dynamic, derived from the review of historical documents.
-Recently researchs have combined information from both tree-ring and remote-sensing to analyze the resilience of forest to drought [*e.g.* @Gazol2018].
-However our approach, which includes also an exhaustive review of historical documents, helps to disentangle the impacts of climatic and land use for forest systems.
+Recently works have combined information from both tree-ring and remote-sensing to analyze the resilience of forest to drought [*e.g.* @Gazol2018].
+The approach presented here, which also includes an exhaustive review of historical documents, help to identify the response of vegetation to climate and to land use changes.
 In this sense our results highligth the importance of landuse legacies for highly transformed Mediterranean systemas [@Navarro2013].
@@
 pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013].
-However, the positive trens for both primary (*i.e.* greenness) and secondary growth (BAI) observed ar our study site and the high resilience values to drought showed at long term scale, seem to suggest that Sierra Nevada constitutes still a refuge for some species.
+However, the positive trends for both primary (*i.e.* greenness) and secondary growth (BAI) observed ar our study site and the high resilience values to drought showed at long term scale, seem to suggest that Sierra Nevada constitutes still a refuge for some species.
 This is relevant for tree species with higher sensitivity to climate change, such us *Q.
@@
 2013; Benito-Garzón et al.
-2008), no solo por la conservación de la especie en sí, sino por los servicios ecosistémicos que estos bosques tan sigulares ofrecen.
+2008), not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.
 
-En este sentido, los futuros trabajos deberan ir en ...
+The abandonment of traditional activities along with the increase in temperature since the 70s, could be explaining the positive growth trend observed.
+This trend and the increase of tree-cover observed in the last decades in this mountain region [@CamachoOlmedo2002], would enhance the carbon sequestration of this forest promoting its role as carbon sink.
+However, the abandonment of traditional management has provoked a general state of degradation of this forest formation [@Canellas2004; @Corcuera2006].
+In this sense, es necesario analizar con mas detalle como se verán afectados los servicios ecosistémicos que ofrecen estos robledales en su rear edge en un contexto de cambio global, identificando claramente los impactos del clima y del land-use change en la provisión de servicios ecosistémicos.
 
-VAS POR AQUI
+# Acknowledgments
 
-En estas zonas refugio en particular parece tener mucha importancia el abandono de las actividades tradicionales, que por un lado y de acuerdo a nuestros resultados parecen estar condicionando el crecimiento
+We would like to thank Sammy Lucy Behle, M.
+Suarez-Muñoz and F.J.
+Bonet-García for their assistence in the field monitoring.
+A.
+Reyes and J.
+Blanco helped in the filtering of satellite data.
+We also are very thankful to F.J.
+Bonet-García for its valuable comments in an earlier version of the manuscript.
+Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work.
+This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project).
+We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.
 
-Future work:
-
-C5 Servicios ecosistémicos
-
-Aumento de crecimiento 
-Refugio para biodivesidad 
-Pero ojo ---> estancamiento y/o mortalidad?? Necesidad de prácticas de gestión que aseguren un crecimiento viable, promoviendo la regeneración sexual (evitar el estancamiento) para asi mantener los 
-La tendencia creciente del crecimiento en el crecimiento observada en los últimos años podría relacionarse con un abandono de las practicas tradicionales en esta region de montaña, así como 
-el abandono de las prácticas tradicionales así como por el caracter de re
-
-Además las diferencias en el crecimiento entre los dos sitios mas cercanos así como lo de la resiliencia, van en la línea de lo expresado por Vilá referente a lo de microrefugios, por lo que la gestión de estas masas, situadas en su rear edge, además ha de analizarse teniendo en cuenta este aspecto.
-
-El abandono de las masas de Q. pyrenaica ha provocado un estado general de degradación, que puede tener consecu
-La reducción de la presión antrópica durante las últimas décadas ha provocado que estas bosques presenten un avanzado estado de degradación
-
-After at least five decades of abandonment, old Cs of Q. pyrenaica in Spain present a general state of degradation
-
-In this work, molecular tools have been successfully applied to assess levels of genetic diversity in Cs, rejecting one of the most widespread and preconceived idea in Spanish silviculture: the decline in genetic diversity promoted by historical coppicing in Q.
-pyrenaica.
-
-En este sentido, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia.
-Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética).
-
-## References
+# References

### 2019-05-27 analysis/bai/trends_resilience_bai.Rmd
update

@@
 ```{r}
+sj_tree <- subset(sj_tree, select=-c(nseries))
+caH_tree <- subset(caH_tree, select=-c(nseries))
+caL_tree <- subset(caL_tree, select=-c(nseries))
+
 bai_sj <- bai_piovesan(rwdf = sj_tree, diam_df = d_sj)
@@
 resi_temporal
+dev.off()
+```
+
+```{r}
+# Variante todos juntos 
+resi_temporal_one <- s %>% filter(samps >= 10) %>% 
+  ggplot(aes(y=rs_mean, x=disturb_year, colour=site)) + 
+  geom_errorbar(aes(ymin=rs_mean - rs_sd, 
+                    ymax=rs_mean + rs_sd),
+                width = 0.75, 
+                size = .7,
+                position=position_dodge(2)) +
+  geom_point(aes(size=d_severity), shape=21, fill='white', position=position_dodge(2)) +
+
+  # facet_wrap(~site, scales='free_y', ncol=1) + 
+  theme_bw() + 
+  theme(strip.background = element_rect(fill='transparent'), 
+        panel.grid.minor = element_blank(),
+        panel.grid.major.x = element_blank()) + 
+  ylab('Resilience (Rs)') + xlab('years') + 
+  scale_x_continuous(breaks=seq(1910,2020, by=5)) # + geom_smooth(aes(colour=site), se=FALSE)
+
+resi_temporal_one
+
+pdf(here::here('/out/bai_resiliences_severes', 'res_bai_longterm_one.pdf'), width=12, height = 9)
+resi_temporal_one
 dev.off()
@@
 
+```{r sin_95}
+sdf95 <- sdf %>% filter(disturb_year != 1995)
+
+sevRt95 <- plotrel(sdf95, v = "rt", x = 'd_severity', label = 'disturb_year',
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rt") +
+  xlab('') + ylab('') +
+  theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
+
+sevRs95 <- plotrel(sdf95, v = "rs", x = 'd_severity', label = 'disturb_year',
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rs") + 
+  xlab('') + ylab('') +
+  theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt"))
+
+sevRc95 <- plotrel(sdf95, v = "rc", x = 'd_severity', label = 'disturb_year',
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rc") +
+  xlab("Drought Severity") + ylab('') +
+  theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
+
+res_long_term95 <- sevRt95 + sevRc95 + sevRs95 + 
+  plot_layout(nrow = 1)
+
+res_long_term95
+
+pdf(here::here('/out/bai_resiliences_severes', 'rs_vs_drought_3years_95.pdf'), width=18, height = 7)
+res_long_term95
+dev.off()
+
+```
+
 # Plots de Resiliencies con lo que ocurre tras el evento 

### 2019-05-30 man/ecosystems/ms_ecosystems.Rmd
update comments from GGI and format according to ECOSYSTEMS style

--- 2019-05-21 man/ms.Rmd
@@
 ---
-title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
-author: "A.J. Perez-Luque; G. Gea-Izquierdo and Zamora, R."
+title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies along a climatic gradient in the rear-edge"
+author: "A.J. Perez-Luque^1,2^; G. Gea-Izquierdo^3^ and Zamora, R.^1,2^"
 output:
@@
     highlight: null
-    reference_docx: ../templates/template_ms.docx
-bibliography: ../refs/references.bib
-csl: ../refs/ecology.csl
+    reference_docx: ../../templates/template_ms.docx
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecosystems.csl
 ---
@@
 
-# Abstract 
+## **Short Title**: Resilience of relict oaks in the rear-edge (máximo 45 caracteres)
+
+^1^ Instituto Interuniversitario de Investigación del Sistema Tierra en Andalucía (CEAMA), Universidad de Granada. Avda. del Mediterráneo s/n, E-18006 Granada, Spain
+
+^2^ Grupo de Ecología Terrestre, Departamento de Ecología, Facultad de Ciencias, Universidad de Granada, Avda. Fuentenueva s/n, E-18071 Granada, Spain
+
+^3^ INIA-CIFOR. Ctra. La Coruña km 7.5. E-28040 Madrid, Spain
+
+## **Author Contributions**
+AJPL conceived and desing the study. AJPL and GGI conducted lab work and led article preparation. 
+AJPL, GGI and RZ conducted field work, collect the data and contribute to writing and editing the manuscript. All authors contribute to the drafts and gave final approval for publication.
+
+# Abstract (369)
 * Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified species composition and the forest structure as a result of intensive human pressure in the past.
@@
 populations in the Southern Iberian Peninsula.
-Forest resilience was analyzed using both dendroecological estimates of growth and MODIS satellite images as proxies of secondary and primary growth respectively.
+Forest resilience was analyzed using both dendroecological growth estimates and MODIS satellite images as proxies for secondary and primary growth respectively.
 
-* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure.
-Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge.
-Furthermore, we found a positive growth trend (stronger in high elevation stand) since the end of the 1970s matching the simultaneous increase in temperature observed.
+* Past intense land-use in the studied populations were likey expressed by disturbance chronologiess, hence 
+partly reflecting the origin of the current forest structure and land-use legacies.
+Trees expressed a high sensitivity to drought.
+Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming).
+This would contradict the *a priori* vulnerability that we expected in respone to climate change of the studied relict stands at the species current rear-edge.
+Yet, there was variation in the response to climate and resilience following a water-stress gradiente along our three studied forests.
+We found a positive growth trend that was strongest in the high elevation stand and non-significant in the most xeric site (or the most transformed???) since the end of the 1970s matching the simultaneous increase in temperature observed.
 
@@
 
-* Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought.
-Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e.
-not only adult performance but also regeneration) of those stands located at the lowest elevations, hence the altitudinal rear-edge, need to be monitored under future expected drier and warmer climatic scenarios and, particularly, if high management pressure remains.
+* Despite being considered relict stands at the dry rear-edge for the species, the trees showed high resilience in response to drought.
+Yet, there were a divergent response to a changing climate along very short climatic gradients.
+The oaks were more sensitive to water availability expressed at the most xeric sites whereas they exhibited positive growth trends only in the coldest and more humid stands.
+These results suggest that stand dynamics, particularly at the lowest elevations (*i.e.* the altitudinal rear-edge), need to be monitored under future expected drier and warmer climatic scenarios bearing in mind that management legacies likely play a key role in current stand dynamics.
+
+## keywords (6 to 10)
+
+## Manuscript highlights
+* H1
+* H2
+* H3
+
+Provide three bullets (up to 85 characters for each, including spaces) that highlight the important contributions of your study for a general reader.
 
@@
 
-Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011].
-It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Dai2011; @Ummenhofer2017].
-Despite drought is considered a feature of Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events has been recorded in the last decades [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported [@Spinoni2017b].
-This is especially important, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Trenberth2014; @Spinoni2017a].
+Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of Mediterranean climate [@Dai2011; @Lionello2012].
+There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017].
+This is especially critical, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Spinoni2017a].
 
-Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].
+Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation.
 The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global [*e.g.*, @VicenteSerrano2013; @Norman2016] and at local scales [see @MartinezVilalta2016 for a revision].
 
-Ecosystem responses to drought are influenced by other drivers of environmental changes [@Fischer2006; @Oliver2014; @Penuelas2017].
-In a global-change context, the impacts of drought on vegetation need to be evaluate considering the interactions with other drivers of change [@Clavero2011; @DoblasMiranda2017].
-This is especially relevant for areas with a long history of landscape modification, such us Mediterranean region, where land-use change plays a significant role [@Navarro2013; @Ameztegui2016], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].
+Ecosystem responses to drought are influenced by other drivers of environmental changes [@Penuelas2017].
+In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017].
+This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].
 
@@
 
-Several studies analyzed the drought effects on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Linares2014; @Dorado2017b; @Caminero2018].
+Many studies have analyzed the effects of drought on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Dorado2017b; @Caminero2018].
 Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
-Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Hodgson2015].
-The assessment of resilience to climate disturbances provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services.
+Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973].
+The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services.
 This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.
 
-Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth.
-Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.*, @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
+The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously.
+Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
+The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
+Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.*, @Zhang2013; @AghaKouchak2015; @Norman2016].
 Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016].
 Drought impacts on vegetation using remote sensing have been analyzed at different scales.
-For instance, a global-scale study showed a stronger influence of the drought for drier biomes [@VicenteSerrano2013].
-For Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].
+For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].
 
-Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
-The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]
-
-In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened.
-We also assess the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.
-Our hypothesis is that these relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts.
-Our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q.
-pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change.
-In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
+In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) on southern relict forests at the rear-edge of species distribution and where species performance is considered to be threatened by climate change.
+We also assessed the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.
+Our main hypothesis is that these relict populations driven by historical land-use at their present climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show lower-values of resilience after several extreme drought only at the dry edge of climatic gradients.
+To address this hyphotesis our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q.
+pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change along climatic gradients.
+To achieve goal (*iii*) within the region of study we assessed whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
 
@@
 Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
-In the Iberian Peninsula these forests live on siliceous soils, or soils poor in basic ions [@Vilches2014] under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002].
+In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002].
 *Q.
-pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species, which can hence be considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@Gavilan2007; @delRio2007].
+pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982].
+Summer rainfall being is a key factor for the species, which is considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@Gavilan2007; @delRio2007].
 
@@
 In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005].
-They are considered relictic forests [@Melendo2000; @Vivero2000], and the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blancaetal1998BC; @Loriteetal2008ABG].
-These forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017].
-Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].
+They are the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blancaetal1998BC; @Loriteetal2008ABG].
+These relictic forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2017].
+Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge [@GeaIzquierdo2013].
 
@@
 During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*.
-1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
+1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
 We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
@@
 
-Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] because they are linked with primary productivity [@Ruimy1994].
+Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning because they are linked with primary productivity [*e.g* @AlcarazSegura2009].
 Here, vegetation greenness of *Q.
 pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
-MODIS EVI Data (Collection 6) were obtanied using Google Earth Engine platform for the period 2000 - 2016.
-EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
+MODIS EVI Data (Collection 6) were obtanied for the period 2000 - 2016.
+EVI data consits of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
 We selected the pixels covering the distribution of *Q.
 pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
-Then, a data filtering was applied to select EVI valid values.
-For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags).
+Valid EVI data were selected applying a filter.
 Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].
@@
 where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
-Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2010; @Samanta2012].
+Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2012].
 
@@
 For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale.
-The nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945) was applied using the `trend` package in R [@Thorsten2017].
-The alpha level for the test was set to 0.05.
+The nonparametric Mann–Kendall test was applied.
 
@@
 
-### Tree sampling 
 Tree sampling was carried out during autumn 2016.

### 2019-05-31 man/ecosystems/ms_ecosystems.Rmd
update

@@
 Tree sampling was carried out during autumn 2016.
-Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warm site located at southern aspect (Figure 1; Table 1).
-For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal tree-line in the sampled slope.
+Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warmer site located at the southern aspect (Figure 1; Table 1).
+For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope.
 All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015].
-In each site between 15 and 20 dominant trees were randomly selected.
+In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected.
 Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer.
-Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively.
-Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius.
+Diameter at breast height (DBH) and total height were measured for each tree.
+In addition, stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius.
 Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
-Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices.
-When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values.
-These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014].
+Then, non-parametric Kruskal-Wallis rank sum tests were used to analyse differences between sites for height, DBH and competition indices.
+When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct significance values.
 
-### Dendroecological analyses 
 Tree cores were air dried, glued onto wooden mounts and sanded.
@@
 They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976].
-Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008; @Bunn2010] package.
-Residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990].
+Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010].
+Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990].
 The statistical quality of each chronology was checked via the expressed population signal.
@@
 We selected grid cells covering each sampled sites.
-Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018].
-The Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014], was also used to characterize the drought conditions for the period 1961-2014.
-SPEI values with a temporal scale of 6 months and a spatial resolution of 1.1 km, were obtanied from the Drought indices dataset (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017].
+Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Frias2018].
+The Standardized Precipitation-Evapotranspiration Index (SPEI)  [@VicenteSerrano2010] was used to characterize the drought conditions for the period 1961-2014.
+SPEI values with a temporal scale of 6 months and a spatial resolution of 1.1 km, were obtanied from a high resolution database of drought indices for Spain  [@VicenteSerrano2017].
 
-Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using the `treeclim` package [@Zang2015].
-The bootstrapped confidence intervals were used to estimate significance (p < 0.05) of the correlation coefficients.
+Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using `treeclim` package [@Zang2015].
 
@@
 logging, neighbor mortality) in the past.
-By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005].
-
+By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997].
 Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:
@@
 For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc.
-Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@MartinMontanes2015; @RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].
+Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].
 
-## Resilience
+## Assessing ecosystem and tree individual resilience to drought
 
-To evaluate the effects of drougth events on greeennes and tree growth (BAI) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
+To evaluate the effects of drougth events on ecosystem resilience (using greeennes data) and tree individual resilience (using BAI growth data) we used indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
 The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred.
@@
 A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017].
-To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure 6, Table S3).
+To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared them with drought severity (Figure 6, Table S3).
 
@@
 Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012].
-When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012].
+When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012].
 Post-hoc differences were assessed pairwise using a similar boostrap test.
-All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages.
+All the robust ANOVA and post-hoc tests were carried out using the `WRS2` and `rcompanion`  packages.
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
-All analyses were carried out in R software [@R2017].
 
@@
 The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period.
-During the 2005 the lowest values of EVI standardized anomalies were recorded, particularly for northern populations (Figure 2).
+The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in northern (dry) populations (Figure 2).
 In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q.
-pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Figure 3).
+pyrenaica* forests during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).
 
-## Radial growth trends and growth disturbances
+## Analysis of radial growth trends and disturbances
 

### 2019-06-06 man/ecosystems/ms_ecosystems.Rmd
UPDATE TEXT

@@
 Post-hoc differences were assessed pairwise using a similar boostrap test.
-All the robust ANOVA and post-hoc tests were carried out using the `WRS2` and `rcompanion`  packages.
+All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package.
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
@@
 
-The trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2).
-Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4).
-Trees for southern sites showed significant positive growth trend since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High).
-We also observed differences in growth for southern sites.
-CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figures 4 and S6).
+The trees of the southern population were older than those from the northern one.
+In addition trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4).
+Both growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6).
+Competition was similar among all sites.
+Yet plot basal area was greatest in CA-High (Table 1), despite the highest individual BAI growth at that site (Figure 4).
+Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High).
 
-From a long-term perspective, drought events produced reduction of RWI for all sites (Figure S3), particularly in 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies.
-Southern sites (CA-High and CA-Low) showed weaker reductions than northern site (SJ), especially for 2005 and 2012 (Figure S3).
-Trees from northern site also suffered a great reduction during 1945-1946 drought event.
+Drought events produced reduction of RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies.
+Southern sites (CA-High and CA-Low) showed weaker reductions than the northern site (SJ), especially for 2005 and 2012 (Figure S3).
+Trees from northern site also suffered a great reduction during 1945-1946 drought event (Figure 5).
 
-The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4).
-Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1995-2000.
-These periods alternate with periods of supression.
-Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression events in the last 50 years.
+`$TODO$ PASAR A Discussion` The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4).
 
-## Resilience of greenness and radial tree-growth to drought events
-At long-term perspective, the strength of the response of the resilience indices was related to drought severity.
+The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940 decade (the most evindent) and the second in the period 1995-2000.
+These periods alternate with periods of supression for over 40 years.
+The two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.
+
+## Resilience to drought events at the stand and individual tree levels
+The strength of the response of the resilience indices was related to drought severity.
 The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (1900 - 2017) (Table S3), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336, p = 0.048) (Figure 6).
@@
 Resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3).
-Southern populations showed lower values for Recovery (*Rc*) than northern site for greenness and tree-growth (Tables S1 and S2).
-In opposite, Resistance and Resilience values were higher for southern sites than for northern ones (Tables S1 and S2).
+The two southern populations showed lower values for Recovery (*Rc*) than the northern site both for greenness and tree-growth (Tables S1 and S2).
+In contrast, Resistance and Resilience values were significantly higher for southern sites than for northern ones (Tables S1 and S2).
 
-## Climate and tree-growth
+## Tree-growth response to climate
 Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
@@
 Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
-Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), specially for the northern population (r > 0.6).
+Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), being this relationship higher in the northern population (r > 0.6 *vs.* r < 0.5).
 Relationships with temperatures were lower than those with moisture variables.
@@
 Our findings show that 2005 and 2012 drought events, provoked reductions in greenness and tree-growth of these forests.
-These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018].
-In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018].
+These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
+In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
 
@@
 For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970 [@BladeCastroDiez2010CeEppyf; @VicenteSerranoetal2017CIG].
+Similar long-term trends were described for this species along its distribution range [@GeaIzquierdo2014] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970 [@VicenteSerranoetal2017CIG].
 Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
@@
 pyrenaica* and high values of resilience to severe droughts were also reported.
-Our findings are in agreenment with those studies that have showed that the assumed higher vulnerability of dry edges does not necesarily hold [@Cavin2017; @Granda2017].
+Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necesarily hold [@Cavin2017].
 @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations.
@@
 
-## Site environment shapes differential sensitivity to climate and drought of rear-edge oak populations
+## Microenvironment shapes differential sensitivity to climate and drought of rear-edge oaks
 
@@
 We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a).
-Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]
+Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017]
 
@@
 soil features, tree competence, etc.) [e.g.
-@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017c].
+@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b].
 
@@
 
-## Land-use legacy effects shape distribution and sensitivity to climate change of read-edge oak populations
+## Land-use legacies shape distribution and sensitivity to climate change at the read-edge
 
@@
 Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada.
-As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia1991].
+As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].
 
-Our study sites have been subjected to an intense exploitation of forest resources.
-Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before.
-For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015].
-For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings.
-Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@Cruz1991].
+Our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation.
+Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before that date.
+For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991].
 These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
@@
 Two non-mutually exclusive ways could explain the tree-growth release observed.
-Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004; @Gentilesca2017].
+Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004].
 On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014].
@@
 
-The only positive release events showed for CA-High site could be related with converiosn from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013; @2015VilaTraver].
+The only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013; @2015VilaTraver].
 
-Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution [*e.g.* @GeaIzquierdo2014; @Herrero2013; @Matias2017], but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Penuelas2017; @DoblasMiranda2017; @Navarro2013; @Ameztegui2016] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.
+Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Navarro2013; @DoblasMiranda2017; @Penuelas2017] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.
 Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource.
@@
 # Conclussions 
+`$TODO$` las conclusiones
 
-Using both dendroecological estimates of growth and MODIS satellite images as proxies of secondary and primary growth respectively, we analyzed forest resilience to severe droughts of *Q.
-pyrenaica* forests located at rear edge of their distribution.
 Severe drought events have provoked important reductions in primary and secondary growth of *Q.
@@
 Surprisingly we observed high values of resilience to drought, particularly at long-term scales.
-This results are consistent with research showing high genetic resilience values [@Valbuena2013; @Valbuena2017], and overall suggest this mountain region still acting as refuge for deciduous *Quercus* to climate changes [@Brewer2002; @Olalde2002; @RodriguezSanchez2010].
-
-The differences observed in tree-growth and resilience to drought between very close sites seem to confirm that responses to drought are site-dependent [@Babst2013; Cavin2017; @Dorado2017c].
+These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*.
+The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent.
 This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia [@VilaCabrera2019].
@@
 We also are very thankful to F.J.
-Bonet-García for its valuable comments in an earlier version of the manuscript.
+Bonet-García for his valuable comments in an earlier version of the manuscript.
 Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work.

### 2019-06-10 man/ecosystems/ms_ecosystems.Rmd
update

@@
 ## keywords (6 to 10)
+extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing
 
@@
 
-Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of Mediterranean climate [@Dai2011; @Lionello2012].
+Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of Mediterranean climate [@Dai2011].
 There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017].
-This is especially critical, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Spinoni2017a].
+This is especially critical, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@Spinoni2017a].
 
 Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation.
-The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global [*e.g.*, @VicenteSerrano2013; @Norman2016] and at local scales [see @MartinezVilalta2016 for a revision].
+The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].
 
@@
 The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010].
-It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].
+It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2013].
 
@@
 The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously.
-Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
-The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
-Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.*, @Zhang2013; @AghaKouchak2015; @Norman2016].
-Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016].
+Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
+The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
+Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Zhang2013].
+Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013].
 Drought impacts on vegetation using remote sensing have been analyzed at different scales.
-For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].
+For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017].
+
+Aquí meter algo de lo que comentó RZ. Tree level response y stand response.
 
@@
 Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
-In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002].
+In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate.
 *Q.
-pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982].
-Summer rainfall being is a key factor for the species, which is considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@Gavilan2007; @delRio2007].
+pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm.
+Summer rainfall is a key factor for the species, which is considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@delRio2007].
 
@@
 There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
-Sierra Nevada mountain is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010].
+Sierra Nevada mountain is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002].
 In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005].
-They are the richest vascular-plant forest formation of Sierra Nevada [@PerezLuqueetal2014P], sheltering several endemic and endangered plant species [@Blancaetal1998BC; @Loriteetal2008ABG].
+They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG].
 These relictic forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2017].
@@
 During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*.
-1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
+1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
 We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
@@
 A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017].
-A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a].
+A drought event is considered only when SPEI values are below the threshold for at least two consecutive months.
 Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
@@
 Here, vegetation greenness of *Q.
-pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015].
+pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor.
 MODIS EVI Data (Collection 6) were obtanied for the period 2000 - 2016.
@@
 As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016.
-To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016].
+To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012].
 For each pixel, an annual EVI value was obtained by averaging EVI valid values.
@@
 The statistical quality of each chronology was checked via the expressed population signal.
-A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable [@Wigley1984].
+A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.
 
@@
 We selected grid cells covering each sampled sites.
-Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Frias2018].
-The Standardized Precipitation-Evapotranspiration Index (SPEI)  [@VicenteSerrano2010] was used to characterize the drought conditions for the period 1961-2014.
-SPEI values with a temporal scale of 6 months and a spatial resolution of 1.1 km, were obtanied from a high resolution database of drought indices for Spain  [@VicenteSerrano2017].
+Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R).
+The Standardized Precipitation-Evapotranspiration Index (SPEI) with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
 
@@
 where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median.
-Medians were used since they are more robust estimator of central tendency than means [@Rubino2004; @Camarero2011].
+Medians were used since they are more robust estimator of central tendency than means [@Rubino2004].
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
@@
 For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc.
-Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].
+Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
 
@@
 
-To evaluate the effects of drougth events on ecosystem resilience (using greeennes data) and tree individual resilience (using BAI growth data) we used indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
-The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred.
-It is estimated as the ratio between the performance during and before the disturbance (drought) event:
-
-Resistance (*Rt*) = Drought / Predrought
-
-The Recovery index (*Rc*) is the ability to recover from disturbance relative to its severity, and it is estimated as the ratio between performance after and during disturbance:
-
-Recovery (*Rc*) = Postdrought / Drought
-
-The Resilience index (*Rs*) is the capacity to reach pre-disturbance performance levels, and it is estimated as the ratio between the performance after and before disturbance:
-
-Resilience (*Rs*) = Postdrought / Predrought
-
+To evaluate the effects of drougth events on ecosystem resilience (using greeennes data) and tree individual resilience (using BAI growth data) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
+The Resistance index, estimated as the ratio between the performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred.
+The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity.
+Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels.
 The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event.
@@
 Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012].
-When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012].
+When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers.
 Post-hoc differences were assessed pairwise using a similar boostrap test.
@@
 For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends were described for this species along its distribution range [@GeaIzquierdo2014] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970 [@VicenteSerranoetal2017CIG].
+Similar long-term trends were described for this species along its distribution range [@GeaIzquierdo2014] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970.
 Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
@@
 The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q.
-pyrenaica* forests [@PeresLis2017] (Figure S7).
+pyrenaica* forests (Figure S7).
 Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].
@@
 The trees of CA-High site, which are located around 1900 m.
-*a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4).
+*a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4).
 This is especially interesting for southern sites, which are very close to each other.
@@
 In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply.
-In most of the high-mountain watersheds of Sierra Nevada, there are systems of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016].
+In most of the high-mountain watersheds of Sierra Nevada, there are systems of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014].
 These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017].
@@
 
-The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008].
-Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017; @MesaFernandez2018].
-From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018].
+The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history.
+Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017].
+From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017].
 Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada.
@@
 For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991].
-These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
+These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
 @GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].
@@
 The review of historical documents revealed that our study sites had different land-use trajectories driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014].
-On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [@2015Zoido]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops.
+On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [@JimenezOlivencia2015]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops.
 Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure.
-For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history [@Navarro2014].
-Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@Maestre1858]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [@2009MartinMartin].
+For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history.
+Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@Maestre1858]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity.
 On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019].
@@
 
-The only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013; @2015VilaTraver].
+The only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013].
 

### 2019-06-10 analysis/clima_real.Rmd
update

---
title: "climate_sites"
author: "Antonio J. Pérez-Luque"
date: "6/10/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("raster")
library("sp")
library("rgdal")
library("dplyr")
library("pander")
library("tidyverse")
library("here")
library("lubridate")
```

```{r}
prec <- stack("/Users/ajpelu/Google Drive/my_repos/clima/data_raw/gridded/worldclim/v2/30sec/prec/wc_prec_SN_30sec.grd")
prec_annual <- calc(prec, sum)

tavg <- stack("/Users/ajpelu/Google Drive/my_repos/clima/data_raw/gridded/worldclim/v2/30sec/tavg/wc_tavg_SN_30sec.grd")
plot(tavg)
tmean <- calc(tavg, mean)

# Read data dendro 
fw <- rgdal::readOGR(dsn = here::here("data_raw/geoinfo/shapes/"), layer = "field_work_4326", verbose = FALSE, encoding = "UTF-8")
```

```{r}
dendro_worldclim <- data.frame(
  coordinates(fw), 
  name = fw$name, 
  site = fw$site,
  loc = fw$loc, 
  prec = extract(prec_annual, fw),
  tavg = extract(tmean, fw)
)

dendro_worldclim <- dendro_worldclim %>% group_by(site) %>% summarise(p = mean(prec), t = mean(tavg))

pander(dendro_worldclim)
```

# Datos Estaciones

```{r}
## Datos bajados de Subsistema CLIMA 
yearly <- function(x){
  out <- x %>% dplyr::select(PM1, date) %>% 
    group_by(year=floor_date(date, "year")) %>%
    summarize(valor=sum(PM1, na.rm = TRUE), n = n())
  return(out)
}

## Datos de SubSistema clima metidos  en Linaria
d <- read_csv(file=here::here("/data_raw/climate/data-1559549919112.csv")) 
```

## Norte (SJ)
### 5501 GuejarSierra
```{r}
# 5501 GuejarSierra
s5501 <- read_delim(here::here("/data_raw/climate/Guejar-Sierra 5501 (1989-01-31 2005-10-31)_MENSUAL.txt"),
                    delim = ";")

s5501y <- s5501 %>% 
  dplyr::select(PM1, date) %>% 
  group_by(year=floor_date(date, "year")) %>%
  summarize(valor=sum(PM1, na.rm = TRUE), n = n())

max(s5501y$year)
min(s5501y$year)

s5501y %>% 
  filter(n == 12) %>% 
  ggplot(aes(y=valor, x=year)) +
  geom_point()

s5501y %>% 
  filter(n == 12) %>% 
  summarise(mean = mean(valor))
```

### 5518 Dilar Toril 
```{r}

s5518 <- read_delim(here::here("/data_raw/climate/Dílar, El Toril 5518 (1957-04-30 2017-11-30)_MENSUAL.txt"),
                    delim = ";")

s5518y <- yearly(s5518)
range(s5518y$year)

s5518y %>% 
  filter(n == 12) %>% 
  ggplot(aes(y=valor, x=year)) +
  geom_point()

mean(s5518y$valor)

s5518y %>% 
  filter(n == 12) %>% count()
```

### 5517 Dilar Central Eléctrica
```{r}
# Dilar Central Eléctrica
s5517 <- read_delim(here::here("/data_raw/climate/Dílar, Central Eléctrica 5517 (1946-06-30 2017-12-31)_MENSUAL.txt"),
                    delim = ";")

s5517y <- yearly(s5517)
range(s5517y$year)

s5517y %>% 
  filter(n == 12) %>% 
  ggplot(aes(y=valor, x=year)) +
  geom_point()
mean(s5517y$valor)

s5517y %>% 
  filter(n == 12) %>% count()

```

### 5506 Pantano de Quentar

```{r}
s5506lin <- d %>% 
  filter(codigo == 5506) %>% 
  filter(cli_variable_id == 130) %>% 
  group_by(year=floor_date(fecha, "year")) %>%
  summarize(valor=sum(valor, na.rm = TRUE), n = n()) %>% 
  filter(n > 360)

mean(s5506lin$valor)
```

### 5504 Tocon de Quentar 
```{r}

s5504lin <- d %>% 
  filter(codigo == 5504) %>% 
  filter(cli_variable_id == 130) %>% 
  group_by(year=floor_date(fecha, "year")) %>%
  summarize(valor=sum(valor, na.rm = TRUE), n = n()) %>% 
  filter(n > 360)

mean(s5504lin$valor)
```

### 5501O Canales 
```{r}
s5501o <- read_delim(here::here("/data_raw/climate/Presa de Canales 5501O (1992-01-31 2003-12-31)_MENSUAL.txt"),
                     delim = ";")

s5501oy <- yearly(s5501o)
range(s5501oy$year)

s5501oy %>% 
  filter(n == 12) %>% 
  ggplot(aes(y=valor, x=year)) +
  geom_point()
mean(s5517y$valor)
```

```{r}
norte <- s5501oy %>% 
  rbind(s5501y, s5518y, s5517y, s5504lin, s5506lin) %>% 
  dplyr::select(-n) %>% 
  group_by(year) %>% 
  summarise(prec_avg = mean(valor), 
            n = n())

norte %>% filter(year > 1980) %>% summarise(mean = mean(prec_avg))
norte %>% filter(year < 1980) %>% summarise(mean = mean(prec_avg))
mean(norte$prec_avg)

norte %>% ggplot(aes(y=prec_avg, x=year)) + geom_point(aes(size = n))

```

## Sur (Canar) 
### 6246Soportujar

```{r}
s6246 <- read_delim(here::here("/data_raw/climate/Soportújar, Casa Forestal 6246 (1946-01-31 2012-12-31)_MENSUAL.txt"),
                    delim = ";")

s6246y <- yearly(s6246)
range(s6246y$year)

s6246y %>% 
  filter(n == 12) %>% 
  ggplot(aes(y=valor, x=year)) +
  geom_point()
mean(s6246y$valor)
```

### Los JArales (6248)

```{r}
s6248 <- read_delim(here::here("/data_raw/climate/Canar_Los_Jarales_6248_(1945-01-31 1988-12-31)_MENSUAL.txt"),
           delim = ";")

s6248y <- yearly(s6248)
range(s6248y$year)

s6248y %>% 
  filter(n == 12) %>% 
  filter(valor > 350) %>% 
  ggplot(aes(y=valor, x=year)) +
  geom_point()

s6248y %>% 
  filter(n == 12) %>% 
  summarise(mean(valor), n())

```

```{r}
sur <- s6246y %>% rbind(s6248y) %>% 
  dplyr::select(-n) %>% 
  group_by(year) %>% 
  summarise(prec_avg = mean(valor), 
            n = n())

sur %>% filter(year > 1980) %>% summarise(mean = mean(prec_avg))
sur %>% filter(year < 1980) %>% summarise(mean = mean(prec_avg))
mean(sur$prec_avg)

sur %>% ggplot(aes(y=prec_avg, x=year)) + geom_point(aes(size = n)) + geom_smooth()

```

# Otras estaciones PSN

l <- list.files(path="/Users/ajpeluLap/Google Drive/CLIMA_SN/data/oapn/data_original/ca/", pattern = "*.dat", full.names = TRUE)

library(purrr)

psn07 <- l %>% map_dfr(read_delim, delim = ",", skip = 11)

pss <- function(file){
  d <- read_delim(file = file, delim = ",", skip = 11)
  d1 <- d %>% dplyr::select(FECHA, 
                      tmean = starts_with("Temperatura Media"),
                      prec = starts_with("Lluvia acumulada(mm)"))
  d2 <- d1 %>% group_by(year=floor_date(as.Date(FECHA, format="%d/%m/%Y"), "year")) %>%
    summarise(temp = mean(tmean), prec = sum(prec))
  return(d2)

}

p2010 <- pss("/Users/ajpeluLap/Google Drive/CLIMA_SN/data/oapn/data_original/ca/PSN07_mensual_2010.dat")

psn <- data.frame()

for (i in l){

  aux <- pss(i)
  psn <- rbind(psn, aux)

  }

p2008 <- pss(file = "/Users/ajpeluLap/Google Drive/CLIMA_SN/data/oapn/data_original/ca/PSN07_mensual_2008.dat")

p2008 <- read_delim(file="/Users/ajpeluLap/Google Drive/CLIMA_SN/data/oapn/data_original/ca/PSN07_mensual_2008.dat",
                    delim = ",", skip = 11) 
p2008 <- p2008 %>% dplyr::select(FECHA, 
                                  tmean = starts_with("Temperatura Media"),
                                  prec = starts_with("Lluvia acumulada(mm)"))
p2008 <- p2008 %>% group_by(year=floor_date(as.Date(FECHA, format="%d/%m/%Y"), "year")) %>%
  summarise(temp = mean(tmean), prec = sum(prec))

p2009 <- read_delim(file="/Users/ajpeluLap/Google Drive/CLIMA_SN/data/oapn/data_original/ca/PSN07_mensual_2009.dat",
                    delim = ",", skip = 11) 
p2009 <- p2009 %>% dplyr::select(FECHA, 
                                 tmean = starts_with("Temperatura Media"),
                                 prec = starts_with("Lluvia acumulada(mm)"))
p2009 <- p2009 %>% group_by(year=floor_date(as.Date(FECHA, format="%d/%m/%Y"), "year")) %>%
  summarise(temp = mean(tmean), prec = sum(prec))

psn072009 <- read_delim(file="/Users/ajpeluLap/Google Drive/CLIMA_SN/data/oapn/data_original/ca/PSN07_mensual_2008.dat",
                        delim = ",", skip = 11)
psn072010 <- read_delim(file="/Users/ajpeluLap/Google Drive/CLIMA_SN/data/oapn/data_original/ca/PSN07_mensual_2008.dat",
                        delim = ",", skip = 11)
psn072008 <- read_delim(file="/Users/ajpeluLap/Google Drive/CLIMA_SN/data/oapn/data_original/ca/PSN07_mensual_2008.dat",
                        delim = ",", skip = 11)

ajpeluLap/ ⁨Google Drive⁩ ▸ ⁨CLIMA_SN⁩ ▸ ⁨data⁩ ▸ ⁨oapn⁩ ▸ ⁨data_original⁩ ▸ ⁨ca

### 2019-06-11 analysis/clima_real.Rmd
update

@@
 
+tmin <- stack("/Users/ajpelu/Google Drive/my_repos/clima/data_raw/gridded/worldclim/v2/30sec/tmin/wc_tmin_SN_30sec.grd")
+tmin_avg <- calc(tmin, mean)
+
+tmax <- stack("/Users/ajpelu/Google Drive/my_repos/clima/data_raw/gridded/worldclim/v2/30sec/tmax/wc_tmax_SN_30sec.grd")
+
+tmax_avg <- calc(tmax, mean)
+
 # Read data dendro 
@@
   loc = fw$loc, 
-  prec = extract(prec_annual, fw),
-  tavg = extract(tmean, fw)
+  prec = raster::extract(prec_annual, fw),
+  tavg = raster::extract(tmean, fw), 
+  tmin = raster::extract(tmin_avg, fw),
+  tmax = raster::extract(tmax_avg, fw)
 )
 
-dendro_worldclim <- dendro_worldclim %>% group_by(site) %>% summarise(p = mean(prec), t = mean(tavg))
+dendro_worldclim <- dendro_worldclim %>% group_by(site) %>% summarise(p = mean(prec), t = mean(tavg), tmin = mean(tmin), tmax = mean(tmax))
 
@@
 ```{r}
-sur <- s6246y %>% rbind(s6248y) %>% 
-  dplyr::select(-n) %>% 
-  group_by(year) %>% 
+# 42 prec 
+# canar
+psn <- read_csv(here::here("/data_raw/climate/datos_psn07_03.csv"))
+
+psn07 <- psn %>% filter(cli_estacion_id == 11) %>% 
+  filter(cli_variable_id == 42) %>% 
+  group_by(year=floor_date(fecha, "year")) %>%
+  summarise(valor = sum(valor), n=n())
+
+```
+
+```{r}
+sur <- s6246y %>% rbind(s6248y, psn07) %>% 
+  dplyr::select(-n) %>% group_by(year) %>% 
   summarise(prec_avg = mean(valor), 
@@
 
-sur %>% ggplot(aes(y=prec_avg, x=year)) + geom_point(aes(size = n)) + geom_smooth()
+sur %>% ggplot(aes(y=prec_avg, x=year)) + geom_point(aes(size = n))
 
 ```
-
-# Otras estaciones PSN
 

### 2019-06-12 man/ecosystems/ms_ecosystems.Rmd
update

@@
 * H1
-* H2
+* H2 %>% %>% 
 * H3
@@
 Aquí meter algo de lo que comentó RZ. Tree level response y stand response.
+
+Mira la biblio de regino y esto https://www.researchgate.net/publication/327321775_Drought_Sensitiveness_on_Forest_Growth_in_Peninsular_Spain_and_the_Balearic_Islands
 
@@
 where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
-Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2012].
+Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI~sa > 1$), "browning" ($EVI~sa <- 1$) or "no changes" ($-1 > EVI~sa > 1$)[@Samanta2012].
 
 Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
-For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale.
-The nonparametric Mann–Kendall test was applied.
+For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale, using Mann–Kendall nonparametric test.
 
@@
 
-In addition, the forest history of San Juan and Cáñar sites was inferred from a detailed analysis of historical land-use changes.
+In addition, the forest and managment history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
 For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc.
-Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
+Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
 
@@
 The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017].
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2018].
 To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared them with drought severity (Figure 6, Table S3).
@@
 
-For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects).
+For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects).
 These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012].
@@
 For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern.
-In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009].
+In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar).
 This could explain the higher values of recovery found for 1995 (Figure 6).
@@
 Soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017].
-In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply.
+
+```{r eval=FALSE, echo=FALSE}
+# In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply.
 In most of the high-mountain watersheds of Sierra Nevada, there are systems of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014].
@@
 This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site.
+```
 

### 2019-06-12 man/figures/resilience_long_term.Rmd
update

@@
   inner_join(dy, by=c("disturb_year" = "maxyear")) 
-
 ```
@@
 ```
+
+```{r}
+library("purrr")
+
+sdfsite <- df %>% 
+  filter(samps >= 10) %>% 
+  group_by(disturb_year, site) %>% 
+  filter(disturb_year > 1940) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+plotrelsite <- function(df, v, x, label, site, coordX_eq, coordY_eq, titulo, ...){
+  require(ggplot2)
+  require(ggrepel)
+  require(purrr)
+
+  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
+
+  eqns <- df %>% split(.$site) %>% 
+    map(~ lm(formulita, data = .)) %>%
+    map(lm_eqn) %>% 
+    do.call(rbind, .) %>%
+    as.data.frame() %>%
+    set_names("equation") %>%
+    mutate(site = rownames(.))
+
+  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label, 
+                        colour = site, fill = site)) + 
+    geom_smooth(method='lm', se=FALSE) + 
+    geom_point(size=1.5, pch = 21) + 
+    theme_bw() + 
+    geom_text_repel(nudge_x = 0.5, size = 3) + 
+    annotate("text", x = coordX_eq, y = coordY_eq, 
+             label=lm_eqn(lm(formulita, df)), parse=TRUE, 
+             size = 3, fontface = "bold", hjust="left") +
+    ggtitle(titulo) + 
+    theme(plot.title = element_text(size = 11, face = "bold.italic"),
+          panel.grid = element_blank(), 
+          axis.text.y = element_text(size=9),
+          axis.text.x = element_text(size=9),
+          axis.title = element_text(size = 10)) +
+    geom_text_repel(data = eqns, aes(x = -Inf, y = Inf,label = equation), parse = TRUE, segment.size = 0)
+
+}
+
+plotrelsite(sdfsite, v = "rs", x = 'd_severity', label = 'disturb_year',
+                 site = "site",
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rt") +
+  xlab('') + ylab('') + 
+  theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
+
+```

### 2019-06-12 man/tables/t_sampled_plots.Rmd
update

@@
 
-clima <- read.csv(here::here("/data/climate_obsnev", "obsnev_qp_pop.csv"), header = TRUE)
-cli <- clima %>% filter(pop %in% c(2,6)) %>% mutate_if(is.numeric, funs(round(., 1))) %>% 
-  dplyr::select(pop, mean, variable) %>% 
-  spread(variable, mean) %>% 
-  mutate(name = case_when(
-    pop == 2 ~ 'SJ',
-    pop == 6 ~ 'CA-High'
-  )) %>% dplyr::select(-pop)
+# clima <- read.csv(here::here("/data/climate_obsnev", "obsnev_qp_pop.csv"), header = TRUE)
+# cli <- clima %>% filter(pop %in% c(2,6)) %>% mutate_if(is.numeric, funs(round(., 1))) %>% 
+#   dplyr::select(pop, mean, variable) %>% 
+#   spread(variable, mean) %>% 
+#   mutate(name = case_when(
+#     pop == 2 ~ 'SJ',
+#     pop == 6 ~ 'CA-High'
+#   )) %>% dplyr::select(-pop) 
+# 
+# # duplicate row for CA site
+# x <- cli[2,] 
+# x$name <- 'CA-Low'
+# cli <- cli %>% bind_rows(x) %>% 
+#   mutate_at(1, round, 0) %>% 
+#   unite(temp, c(tmin, tmax), sep=" - ")
 
-# duplicate row for CA site
-x <- cli[2,] 
-x$name <- 'CA-Low'
-cli <- cli %>% bind_rows(x) %>% 
-  mutate_at(1, round, 0) %>% 
-  unite(temp, c(tmin, tmax), sep=" - ")
+# Update use worlclim2 
+clima <- read.csv(here::here("/data/climate_worldclim", "dendro_worldclim.csv"), header = TRUE)
 
-t1raw <- t1raw %>% inner_join(cli, by='name')
+clima <- clima %>% 
+  mutate(site = case_when(
+    site == 'SJHigh' ~ 'SJ', 
+    site == 'SJLow' ~ 'SJ',  
+    site == 'CAHigh' ~ 'CA-High',
+    site == 'CALow' ~ 'CA-Low'
+  )) %>% 
+  mutate_at(2, round, 0) %>% 
+  mutate_at(3:5, round, 1) %>% 
+  group_by(site) %>% 
+  summarise_each(mean) %>% 
+  dplyr::select(name = site, prec = p, tmin, tmax) %>% 
+  unite(temp, c(tmin, tmax), sep="-")
+
+t1raw <- t1raw %>% inner_join(clima, by='name')
 

### 2019-06-13 analysis/dendro_summary.Rmd
update

@@
 library("pander")
+library("here")
+
+```
+
+```{r}
+# Compara 
+comparaKW <- function(df, mivariable){ 
+  require(PMCMR)
+  require(multcompView)
+
+  output <- list()
+
+  # Model formulation
+  myformula <- as.formula(paste0(mivariable, " ~ site"))
+
+  # Kruskal Wallis  
+  kt <- kruskal.test(myformula, data = df)
+
+  # Summary AOV (broom style)
+  tm <- broom::tidy(kt)
+  tm$mi_variable <- mivariable
+
+  # Dunn's test 
+  object <- posthoc.kruskal.dunn.test(myformula, data=df, p.adjust.method = "bonferroni")
+
+  # Get dataframe with letters and pvalues (#from summary.PMCMR)
+  pval <- as.numeric(object$p.value)
+  stat <- as.numeric(object$statistic)
+  grp1 <- as.numeric(c(col(object$p.value)))
+  cnam <- colnames(object$p.value)
+  grp2 <- as.numeric(c(row(object$p.value)))
+  rnam <- rownames(object$p.value)
+  H0 <- paste(cnam[grp1], " = ", rnam[grp2])
+  OK <- !is.na(pval)
+  xdf <- data.frame(H0 = H0[OK], statistic = stat[OK], p.value = format.pval(pval[OK], 5))
+
+  # Get letters (using multcompView) See viggnete PMCMR 
+  dt_letters  <- multcompView::multcompLetters(get.pvalues(object), threshold = 0.05)
+
+  output$KW <- tm
+  output$post_hoc <- xdf
+  output$letters <- dt_letters
+  return(output)
+  }
 
@@
          age_std = sd(year)) %>% as.data.frame()
+
+comparaAges <- comparaKW(rwl_stat_sites, 'year')
+
+letrasAges <- data.frame(site = names(comparaAges$letters$Letters),
+                            age_l = comparaAges$letters$Letters)
+
+age_sites <- age_sites %>% inner_join(letrasAges, by ='site')
 
@@
 ```
+
+```{r}
+
+cofecha <- read_delim(here::here("/data_raw/dendro_ring/cofechas.csv"), delim = ";") %>% 
+  mutate(site = as.factor(site))
+
+rbt <- cofecha %>% 
+  group_by(site) %>% 
+  summarise(rbt = mean(corr_w_master))
+
+```

### 2019-06-13 man/tables/t_dendro_summary.Rmd
update

--- 2018-11-19 man/tables/t_dendro_summary.Rmd
@@
 **Table 2.** Characteristics of the mean tree ring chronologies.
-Values of the length year in parenthesis indicate years replicated more with than five series.
+Values of the length year in parenthesis indicate years replicated with more than five series.
 RW = mean annual ring width (standard deviation in parenthesis).
@@
 dendro_summ <- read.csv(here::here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
+cofecha <- read_delim(here::here("/data_raw/dendro_ring/cofechas.csv"), delim = ";")
+
+rbt <- cofecha %>% 
+  group_by(name = site) %>% 
+  summarise(rbt = mean(corr_w_master))
 
@@
 t2 <- t1raw %>% 
+  inner_join(rbt, by="name") %>% 
   mutate(pare = ")") %>% 
@@
   length_year, n_trees, n_radii,
-  RW, ms2, ar1, rbar.wt, eps, -pare, -site)
+  RW, ms2, ar1, rbt, -rbar.wt, eps, -pare, -site)
 
-ta2 <- regulartable(t2, col_keys =c("name", "first_year", "last_year", "n_trees", "n_radii", "RW", "ms2",
-                                   "ar1","rbar.wt","eps"))
+ta2 <- regulartable(t2, col_keys =c("name", "first_year", "last_year", "length_year", "n_trees", "n_radii", "RW", "ms2",
+                                   "ar1","rbt","eps"))
 ta2 <- ta2 %>% set_header_labels(name = "Site", first_year = "First year", last_year = "Last year",
                                  length_year = "Length (years)", n_trees ="# trees", n_radii = '# cores',
-                                 RW ="RW (mm)", ms2="MS", ar1="AR(1)", rbar.wt = "Rbt", eps = "EPS") %>%
+                                 RW ="RW (mm)", ms2="MS", ar1="AR(1)", rbt = "Rbt", eps = "EPS") %>%
   align(align = "center", part = "header") %>% 

### 2019-06-13 man/tables/t_huber_res_bai.Rmd
update tables
update

@@
   word_document:
-    reference_docx: "../../man/template_tables_portrait.docx"
+    reference_docx: ../../templates/template_tables_portrait.docx
+bibliography: ../../refs/references.bib
+csl: ../../refs/ecology.csl
 ---
@@
 
-**Table S2.** Robust measures of central tendency of resilience indices for tree-growth (BAI) groupped by drought events, site and interaction.
+**Table S2.** Robust measures of central tendency of resilience indices for tree-growth (BAI) grouped by drought events, site and interaction.
 Measures of central tendency are M-estimators based on Huber's Psi (see material and methods).

### 2019-06-13 man/tables/t_huber_res_evi.Rmd
update tables
update

@@
 
-**Table S1.** Robust measures of central tendency of resilience indices for greenness (EVI) groupped by drought events, site and interaction.
+**Table S1.** Robust measures of central tendency of resilience indices for greenness (EVI) grouped by drought events, site and interaction.
 Measures of central tendency are M-estimators based on Huber's Psi (see material and methods).

### 2019-06-13 man/tables/t_sampled_plots.Rmd
update

@@
   name, lat_m, long_m, elev_min, elev_max, slope_mean, slope_sd, prec, temp, 
-  n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, 
+  n_trees, n_radii, dn_mean, dn_sd, h_mean, h_sd, age_mean, age_std, age_l, 
   dn_mean_all, dn_sd_all, h_mean_all, h_sd_all,
@@
     h = paste(paste0(h, pare), ft_h_l, sep = ' '),
-    age = paste0(age, pare),
+    age = paste(paste0(age, pare), age_l, sep = ' '),
     dn_all = paste(paste0(dn_all, pare), all_dn_l, sep = ' '),
@@
     srd = paste(paste0(srd, pare), srd_l, sep = ' ')) %>%
-  dplyr::select(-pare, -all_dn_l, -all_h_l, -ba_l, -ft_dn_l, -ft_h_l, -std_l, -srd_l)
+  dplyr::select(-pare, -all_dn_l, -all_h_l, -ba_l, -ft_dn_l, -ft_h_l, -std_l, -srd_l, -age_l)
 
@@
 
-ta %>% 
-   vline(j = c(7, 11), border = std_border, part="all")
+ta <- compose(
+  ta, j = "ba", part = "header",
+  value = as_paragraph(
+     "BA\n(m",  
+     as_chunk("2", props = fp_text(vertical.align = "superscript", font.size = 8)),
+     " ha", 
+     as_chunk("-1", props = fp_text(vertical.align = "superscript", font.size = 8)),
+     ")"
+
+  )
+)
+
+ta <- compose(
+  ta, j = "stand_density", part = "header",
+  value = as_paragraph(
+     "Density\n(trees ha",  
+     as_chunk("-1", props = fp_text(vertical.align = "superscript", font.size = 6)),
+     ")"
+
+  )
+)
+
+ta %>% vline(j = c(7, 11), border = std_border, part="all")
 

### 2019-06-13 man/figures/resilience_long_term.Rmd
update figures

@@
 library("broom")
+library("cowplot")
 ```
@@
 
-```{r plot_join}
+```{r plot_join, eval=FALSE, echo=FALSE}
 res_long_term <- sevRt + sevRc + sevRs + 
@@
 
-```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8}
+```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8, eval=FALSE, echo=FALSE}
 res_long_term
@@
 ```{r, eval=FALSE, echo=FALSE}
-df <- sdf %>% 
+dff <- sdf %>% 
   dplyr::select(d_severity, rt_mean_mean, rs_mean_mean, rc_mean_mean) %>% 
@@
 
-nested_df <- group_by(df, res_metric) %>% nest()
+nested_df <- group_by(dff, res_metric) %>% nest()
 
-fit_model <- function(df){
-  lm(value ~ d_severity, data = df)
+fit_model <- function(x){
+  lm(value ~ d_severity, data = x)
 }
@@
 
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label, 
-                        colour = site, fill = site)) + 
+  ggplot(df, 
+         aes_string(y=paste0(v, "_mean_mean"), x=x, label=label, colour=site, fill=site)) +
+    geom_smooth(method='lm', se=FALSE, aes(group=1), colour='black', linetype = "dashed", size = 1) +
     geom_smooth(method='lm', se=FALSE) + 
@@
     theme_bw() + 
-    geom_text_repel(nudge_x = 0.5, size = 3) + 
-    annotate("text", x = coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-             size = 3, fontface = "bold", hjust="left") +
+    geom_text_repel(nudge_x = 0.5, size = 3, show.legend = FALSE) + 
+    # annotate("text", x = coordX_eq, y = coordY_eq, 
+    #          label=lm_eqn(lm(formulita, df)), parse=TRUE, 
+    #          size = 3, fontface = "bold", hjust="left") +
     ggtitle(titulo) + 
@@
           axis.title = element_text(size = 10)) +
-    geom_text_repel(data = eqns, aes(x = -Inf, y = Inf,label = equation), parse = TRUE, segment.size = 0)
+    geom_text_repel(data = eqns, aes(x = -Inf, y = Inf, label = equation), 
+                    direction = "y", 
+                    parse = TRUE, segment.size = 0, show.legend = FALSE)
 
@@
 
-plotrelsite(sdfsite, v = "rs", x = 'd_severity', label = 'disturb_year',
-                 site = "site",
+```
+
+```{r}
+sevRt_site <- plotrelsite(sdfsite, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
                  coordX_eq = 6, coordY_eq = 0.5, titulo = "Rt") +
@@
   theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
+```
+
+```{r}
+sevRs_site <- plotrelsite(sdfsite, v = "rs", x = 'd_severity', label = 'disturb_year', site = "site",
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rs") + 
+  xlab('') + ylab('') + theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt")) 
+```
+
+```{r}
+sevRc_site <- plotrelsite(sdfsite, v = "rc", x = 'd_severity', label = 'disturb_year', site = "site",
+                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rc") +
+  xlab("Drought Severity") + ylab("") +
+  theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
+```
+
+```{r}
+res_long_term_sites <- sevRt_site + theme(legend.position="none") +
+  sevRc_site + theme(legend.position="bottom") +
+  sevRs_site + theme(legend.position = "none") + plot_layout(nrow = 1)
+
+ggsave(filename = here::here("./man/figures/svg/fig_6.jpg"), 
+       dpi = 300,  plot=res_long_term_sites, width=10, height=8)
 
 ```
+
+```{r, fig.width=6, dpi = 300, out.width = 0.8}
+res_long_term_sites
+```

### 2019-06-24 man/ecosystems/ms_ecosystems.Rmd
update ms

@@
 
-## keywords (6 to 10)
+## keywords
 extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing
@@
 A drought event is considered only when SPEI values are below the threshold for at least two consecutive months.
-Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.
+Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event.
 
@@
 
-Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning because they are linked with primary productivity [*e.g* @AlcarazSegura2009].
-Here, vegetation greenness of *Q.
-pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor.
+Vegetation greenness of *Q.
+pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS (*Moderate Resolution Imaging Spectroradiometer*) sensor.
 MODIS EVI Data (Collection 6) were obtanied for the period 2000 - 2016.
@@
 pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
-Valid EVI data were selected applying a filter.
 Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].
@@
 Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
-Then, non-parametric Kruskal-Wallis rank sum tests were used to analyse differences between sites for height, DBH and competition indices.
+Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests.
 When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct significance values.
@@
 Tree cores were air dried, glued onto wooden mounts and sanded.
-Annual radial growth (ring width, RW) was measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.001 mm.
+Annual radial growth (ring width, RW) was measured with a measuring device coupled to a stereomicroscope, with an accuracy of 0.001 mm.
 Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
@@
 Both RWI and BAI chronologies were compared to different climatic indices to analyse the low and the high-frequency response to climate of trees.
-Statistical descriptive parameters, including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology [@Fritts1976].
 
@@
 We selected grid cells covering each sampled sites.
-Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R).
 The Standardized Precipitation-Evapotranspiration Index (SPEI) with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
@@
 In addition, the forest and managment history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
-For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc.
+For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest Table XX
+
+$ convertir en tabla $ 
+: land-uses, forest managment actions, mining, wildfires, etc.
 Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].

### 2019-06-24 man/figures/resilience_long_term.Rmd
update ms

@@
   group_by(disturb_year, site) %>% 
-  filter(disturb_year > 1940) %>% 
+  filter(disturb_year > 1950) %>% 
   summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
@@
 
-plotrelsite <- function(df, v, x, label, site, coordX_eq, coordY_eq, titulo, ...){
+plotrelsite <- function(df, v, x, label, site, titulo, coord_eqX, ...){
   require(ggplot2)
@@
          aes_string(y=paste0(v, "_mean_mean"), x=x, label=label, colour=site, fill=site)) +
-    geom_smooth(method='lm', se=FALSE, aes(group=1), colour='black', linetype = "dashed", size = 1) +
+    geom_smooth(method='lm', se=FALSE, aes(group=1), colour='gray',  size = 1) + # linetype = "dotted",
     geom_smooth(method='lm', se=FALSE) + 
@@
     theme_bw() + 
-    geom_text_repel(nudge_x = 0.5, size = 3, show.legend = FALSE) + 
-    # annotate("text", x = coordX_eq, y = coordY_eq, 
-    #          label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-    #          size = 3, fontface = "bold", hjust="left") +
+    geom_text_repel(nudge_x = 0.5, size = 2, show.legend = FALSE) + 
     ggtitle(titulo) + 
@@
           axis.title = element_text(size = 10)) +
-    geom_text_repel(data = eqns, aes(x = -Inf, y = Inf, label = equation), 
+    geom_text_repel(data = eqns, aes(x = coord_eqX, y = Inf, label = equation), 
                     direction = "y", 
@@
 sevRt_site <- plotrelsite(sdfsite, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
-                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rt") +
-  xlab('') + ylab('') +
-  theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
+                 titulo = "Rt", coord_eqX = Inf) +
+  xlab('') + ylab('') + xlim(5,25)
+
+# + theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
 ```
@@
 sevRs_site <- plotrelsite(sdfsite, v = "rs", x = 'd_severity', label = 'disturb_year', site = "site",
-                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rs") + 
-  xlab('') + ylab('') + theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt")) 
+                 titulo = "Rs", coord_eqX = -Inf) + xlab('') + ylab('') + xlim(5,25)
+# theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt")) 
 ```
@@
 sevRc_site <- plotrelsite(sdfsite, v = "rc", x = 'd_severity', label = 'disturb_year', site = "site",
-                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rc") +
-  xlab("Drought Severity") + ylab("") +
-  theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
+                 titulo = "Rc",  coord_eqX = -Inf) +
+  xlab("Drought Severity") + ylab("") + xlim(5,25)
+#+ theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
 ```
@@
 ggsave(filename = here::here("./man/figures/svg/fig_6.jpg"), 
-       dpi = 300,  plot=res_long_term_sites, width=10, height=8)
+       dpi = 300,  plot=res_long_term_sites, width=9, height=7)
 

### 2019-06-24 man/tables/t_forest_history.Rmd
update ms

--- 2019-02-06 man/tables/t_robust_anovas.Rmd
@@
 
-**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the two drought events (in 2005 and 2012) and site.
+**Table S3.** Review of the foorest and management history of the sampling sites
 
-```{r}
+In addition, the forest and managment history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
+For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest Table XX
+
+$ convertir en tabla $ 
+: land-uses, forest managment actions, mining, wildfires, etc.
+Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
+
+```{r, eval=FALSE}
 ### Table R1_evi
@@
 
-```{r}
+```{r, eval = FALSE}
 mt <- regulartable(

### 2019-06-24 man/ecosystems/ms_ecosystems.Rmd
update

@@
 Drought events produced reduction of RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies.
-Southern sites (CA-High and CA-Low) showed weaker reductions than the northern site (SJ), especially for 2005 and 2012 (Figure S3).
+The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S3).
+Tree-growth reduction followed a moisture-gradient, being weaker for the wetter site (CA-High).
 Trees from northern site also suffered a great reduction during 1945-1946 drought event (Figure 5).
-
-`$TODO$ PASAR A Discussion` The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4).
 

### 2019-06-26 man/figures/resilience_long_term.Rmd
update figure 6

@@
   eq2 <- substitute(~~italic(r)^2~"="~rvalue*","~italic(p)~"="~pvalue,  
-                    list(rvalue = format(summary(m)$r.squared, digits = 3), 
-                         pvalue = format(summary(m)$coefficients[2,4], digits = 3)))
+                    list(rvalue = formatC(summary(m)$r.squared, digits = 3, format = "f"), 
+                         pvalue = formatC(summary(m)$coefficients[2,4], digits = 3, format = "f")))
 
@@
 ```{r}
+# Plot separately 
 library("purrr")
-
-sdfsite <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year, site) %>% 
-  filter(disturb_year > 1950) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
 
@@
     geom_text_repel(data = eqns, aes(x = coord_eqX, y = Inf, label = equation),
-                    direction = "y", 
-                    parse = TRUE, segment.size = 0, show.legend = FALSE)
-
+                    direction = "y", segment.size = 0, hjust = 0,
+                    parse = TRUE, show.legend = FALSE)
 }
@@
 ```{r}
+# Include 1995
+sdfsite <- df %>% 
+  filter(samps >= 10) %>% 
+  group_by(disturb_year, site) %>% 
+  filter(disturb_year > 1940) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+v = "rt"
+x = 'd_severity'
+formulita <- as.formula(paste0(v, "_mean_mean ~", x))
+
+  eqns <- sdfsite %>% split(.$site) %>% 
+    map(~ lm(formulita, data = .)) %>%
+    map(lm_eqn) %>% 
+    do.call(rbind, .) %>%
+    as.data.frame() %>%
+    set_names("equation") %>%
+    mutate(site = rownames(.))
+
 sevRt_site <- plotrelsite(sdfsite, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rt", coord_eqX = Inf) +
-  xlab('') + ylab('') + xlim(5,25)
+                 titulo = "Rt", coord_eqX = Inf) + xlab('') + ylab('') + xlim(5,25)
 
-# + theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
-```
-
-```{r}
 sevRs_site <- plotrelsite(sdfsite, v = "rs", x = 'd_severity', label = 'disturb_year', site = "site",
                  titulo = "Rs", coord_eqX = -Inf) + xlab('') + ylab('') + xlim(5,25)
-# theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt")) 
-```
 
-```{r}
 sevRc_site <- plotrelsite(sdfsite, v = "rc", x = 'd_severity', label = 'disturb_year', site = "site",
@@
   xlab("Drought Severity") + ylab("") + xlim(5,25)
-#+ theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
+
 ```
@@
 ggsave(filename = here::here("./man/figures/svg/fig_6.jpg"), 
-       dpi = 300,  plot=res_long_term_sites, width=9, height=7)
+       dpi = 300,  plot=res_long_term_sites, width=8.5, height=7)
 
@@
 ```
+
+```{r no95}
+sdfsite95 <- df %>% 
+  filter(samps >= 5) %>% 
+  group_by(disturb_year, site) %>% 
+  filter(disturb_year > 1940) %>% 
+  filter(disturb_year != 1995) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear"))
+
+sevRt_site95 <- plotrelsite(sdfsite95, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
+                 titulo = "Rt", coord_eqX = Inf) + xlab('') + ylab('')
+
+sevRs_site95 <- plotrelsite(sdfsite95, v = "rs", x = 'd_severity', label = 'disturb_year', site = "site",
+                 titulo = "Rs", coord_eqX = -Inf) + xlab('') + ylab('')
+
+sevRc_site95 <- plotrelsite(sdfsite95, v = "rc", x = 'd_severity', label = 'disturb_year', site = "site",
+                 titulo = "Rc",  coord_eqX = -Inf) + xlab("Drought Severity") + ylab("")
+
+res_long_term_sites95 <- sevRt_site95 + theme(legend.position="none") +
+  sevRc_site95 + theme(legend.position="bottom") +
+  sevRs_site95 + theme(legend.position = "none") + plot_layout(nrow = 1)
+
+ggsave(filename = here::here("./man/figures/svg/fig_6_95.jpg"), 
+       dpi = 300,  plot=res_long_term_sites95, width=9, height=7)
+
+```
+
+```{r, fig.width=6, dpi = 300, out.width = 0.8}
+res_long_term_sites95
+```

### 2019-06-26 man/figures/resilience_long_term.Rmd
update figure
add without 95

@@
 
-v = "rt"
-x = 'd_severity'
-formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  eqns <- sdfsite %>% split(.$site) %>% 
-    map(~ lm(formulita, data = .)) %>%
-    map(lm_eqn) %>% 
-    do.call(rbind, .) %>%
-    as.data.frame() %>%
-    set_names("equation") %>%
-    mutate(site = rownames(.))
-
 sevRt_site <- plotrelsite(sdfsite, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
@@
 sdfsite95 <- df %>% 
-  filter(samps >= 5) %>% 
+  filter(samps >= 10) %>% 
   group_by(disturb_year, site) %>% 
@@
 ggsave(filename = here::here("./man/figures/svg/fig_6_95.jpg"), 
-       dpi = 300,  plot=res_long_term_sites95, width=9, height=7)
+       dpi = 300,  plot=res_long_term_sites95, width=8.7, height=7)
 

### 2019-06-27 man/figures/evi_bai_ts.Rmd
update

---
output:
  word_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r load_pkg}
library("tidyverse")
library("here")
library("scales")
library("patchwork")
```

```{r data}
bai <- read.csv(file=here::here('data/cronos_medias/cronos_sites.csv'), header=TRUE)

df <- read.csv(file=here::here("data/evi", "evi_globalmeans_drouhgt.csv"), header = TRUE, sep = ',')

ggplot(df, 
       aes(x=disturb, y=mean, colour=site)) + 
  geom_point(size=2) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = .1) +
  geom_line(aes(group=site)) + 
  facet_wrap(~disturb_year) + 
  scale_x_discrete(labels = c("1_dr" = "Drought", "0_pre" = "Previous", "2_post" = "Posterior")) + 
  theme_bw() +
  xlab("") + ylab("EVI") + 
  theme(
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face="bold", size=12),
    axis.title.y = element_text(face="bold", size=12)
  ) +
  scale_color_manual(values=c('black', 'blue', 'gray'))

# BAI -> d
```

### 2019-06-28 man/ecosystems/ms_ecosystems.Rmd
update text

@@
 ## Resilience to drought events at the stand and individual tree levels
-The strength of the response of the resilience indices was related to drought severity.
-The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (1900 - 2017) (Table S3), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336, p = 0.048) (Figure 6).
-A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438, p = 0.019) but negative for resistance (*Rt*, $r^2$ = 0.155, p = 0.205).
-Trees showed the highest value of tree-growth resilience for the worst drought event in our study area (1995, see Table S3).
+The exploration of the resilience metrics of tree-growth for severe drought events since 1950 (*i.e.* for the share period among the three chronologies, excluding juvenile years, Table S3) revealed a positive relation between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6).
+A similar pattern was obtained for Resilience (*Rs*) but only significant for SJ site (Figure 6).
+Non-significant patterns were obtained if we exclude 1995 from the analysis (Figure S8).
+Trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), and particularly SJ, where our results suggest a major release event after 1995 (Figure 5).
 
 During the last two drought events, resilience metrics for greenness and tree-growth were significantly different bewteen drought events (Table 3).
-The 2005 drought event reduced greennes and growth more than that of 2012 (Tables S1 and S2).
+The 2005 drought event reduced greennes and growth more than that of 2012 (Tables S1 and S2) but the two metrics of resilience generally covaried in the same direction.
 We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7).
@@
 In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
-
 Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
@@
 pyrenaica* and high values of resilience to severe droughts were also reported.
-Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necesarily hold [@Cavin2017].
+Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017].
 @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations.
@@
 In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
-Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018].
-En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q.
-pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.
+A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@ MartinezVilalta2018].
+In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q.
+pyrenaica* in Sierra Nevada are located in a geographical but not climatic rear-edge.
 
@@
 A plausible explanation could be the water availability previously and after the drought event.
-For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern.
-In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar).
+For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern (*e.gfirst months of 1996 recorded the highest cumulative rainfall of the past century for our southern sites).
 This could explain the higher values of recovery found for 1995 (Figure 6).
+The high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality occurred in this area (*personal communications*).
 
@@
 We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a).
-Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017]
+Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in oaks [*e.g.* @GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017]
 
@@
 This is especially interesting for southern sites, which are very close to each other.
-Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (e.g.
-soil features, tree competence, etc.) [e.g.
-@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b].
-
-When we explored the chracteristics of the southern sites, we firstly observed a difference in elevation.
-@Bhuyan2017, in a recent work across the European continent, found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.
+Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b].
 The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q.
@@
 Soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017].
+Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive [@Bhuyan2017].
 
@@
 
-In addition to the lower water stress, it is likely that the sites at higher elevation had a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources.
+In addition to the lower water stress, it is likely that the sites at higher elevation had a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).
 

### 2019-07-01 man/ecosystems/ms_ecosystems.Rmd
update conclusions

@@
 
-## Resilience to drought events at the stand and individual tree levels
+## Resilience to drought events at the ecosystem and individual tree levels
 The exploration of the resilience metrics of tree-growth for severe drought events since 1950 (*i.e.* for the share period among the three chronologies, excluding juvenile years, Table S3) revealed a positive relation between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6).
@@
 # Conclussions 
-`$TODO$` las conclusiones
+Severe drought events have provoked important reductions in primary and secondary growth of *Q.
+pyrenaica* forests at Sierra Nevada mountain range.
+However, despite the expected vulnerability for those relict stands, we observed positive trends of both primary (*i.e.* greenness) at ecosystem scale and secondary growth at tree level in our study site, particularly at the most humid site along the climatic gradient analyzed.
+Surprisingly we observed high resilience values to drought, particularly at long-term scales.
+These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*.
 
-Severe drought events have provoked important reductions in primary and secondary growth of *Q.
-pyrenaica* forests.
-However, despite the expected vulnerability for those relict stands, we observed positive trends in primary growth (since 2000) and secondary growth (since 1970) of *Q.
-pyrenaica* in Sierra Nevada mountain range.
-Surprisingly we observed high values of resilience to drought, particularly at long-term scales.
-These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*.
 The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent.
-This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia [@VilaCabrera2019].
+This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019].
 
-The analysis of tree-growth dynamics revealed suppression and release events that are consistent with socieconomic dynamic, derived from the review of historical documents.
-Recently works have combined information from both tree-ring and remote-sensing to analyze the resilience of forest to drought [*e.g.* @Gazol2018].
-The approach presented here, which also includes an exhaustive review of historical documents, help to identify the response of vegetation to climate and to land use changes.
-In this sense our results highligth the importance of landuse legacies for highly transformed Mediterranean systemas [@Navarro2013].
+The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents.
+In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems.
+This is relevant for tree species with higher sensitivity to climate change, such us *Q.
+pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.
 
-Growth projections have forecasted a decrease in productivity for *Q.
-pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013].
-However, the positive trends for both primary (*i.e.* greenness) and secondary growth (BAI) observed ar our study site and the high resilience values to drought showed at long term scale, seem to suggest that Sierra Nevada constitutes still a refuge for some species.
-This is relevant for tree species with higher sensitivity to climate change, such us *Q.
-pyrenaica* (García-Valdes et al.
-2013; Benito-Garzón et al.
-2008), not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.
-
-The abandonment of traditional activities along with the increase in temperature since the 70s, could be explaining the positive growth trend observed.
-This trend and the increase of tree-cover observed in the last decades in this mountain region [@CamachoOlmedo2002], would enhance the carbon sequestration of this forest promoting its role as carbon sink.
-However, the abandonment of traditional management has provoked a general state of degradation of this forest formation [@Canellas2004; @Corcuera2006].
-In this sense, es necesario analizar con mas detalle como se verán afectados los servicios ecosistémicos que ofrecen estos robledales en su rear edge en un contexto de cambio global, identificando claramente los impactos del clima y del land-use change en la provisión de servicios ecosistémicos.
+Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.
 

### 2019-07-02 man/figures/resilience_long_term.Rmd
update

@@
 
-**Figure 6.** Resilience metrics of the tree-growth for the most severe drought events (as from Table S3).
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
-Points indicate average of resilience metrics for all populations.
-Error bar corresponds standard error.
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-
 ```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8, eval=FALSE, echo=FALSE}
@@
 
-```{r, eval=FALSE, echo=FALSE}
+**Figure 6.** Resilience metrics of the tree-growth for severe drought events since 1950.
+*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
+Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*).
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
+Gray line represents overall relationship for each Resilience metrics.
+
+```{r, echo=FALSE}
 dff <- sdf %>% 
@@
 
-plotrelsite <- function(df, v, x, label, site, titulo, coord_eqX, ...){
+plotrelsite <- function(df, v, x, label, site, titulo, coord_eqX, xLM, yLM, ...){
   require(ggplot2)
@@
                     direction = "y", segment.size = 0, hjust = 0,
-                    parse = TRUE, show.legend = FALSE)
+                    parse = TRUE, show.legend = FALSE) +
+    annotate("text", x = .0, y = 10, 
+             label=lm_eqn(lm(formulita, df)), parse=TRUE, 
+             size = 5, fontface = "bold", hjust="left") 
 }
@@
 sevRt_site <- plotrelsite(sdfsite, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rt", coord_eqX = Inf) + xlab('') + ylab('') + xlim(5,25)
+                 titulo = "Rt", coord_eqX = Inf) + xlab('') + ylab('')
+
++ 
+  xlim(5,25)
 

### 2019-07-08 man/ecosystems/ms_ecosystems.Rmd
update

@@
 Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Zhang2013].
-Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013].
+Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013].
 Drought impacts on vegetation using remote sensing have been analyzed at different scales.
@@
 
-Aquí meter algo de lo que comentó RZ. Tree level response y stand response.
-
-Mira la biblio de regino y esto https://www.researchgate.net/publication/327321775_Drought_Sensitiveness_on_Forest_Growth_in_Peninsular_Spain_and_the_Balearic_Islands
+Vegetation responses to drought requires of integrative and multidisciplinary strategies for assessing the impacts of disturbances on forest at several scales [@Hartman2018].
+Several works assessed vegetation responses in term of resistance and resilience to drought events highlithg a stronger reponse of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
+Nonetheless, is crucial to know if the responses at the tree level differ from responses at the ecosystem level, particularly in rear edge populations.
 
@@
 For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope.
+Sampling sites follow a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
 All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015].
@@
 In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
-A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@ MartinezVilalta2018].
+A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018].
 In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q.
@@
 A plausible explanation could be the water availability previously and after the drought event.
-For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern (*e.gfirst months of 1996 recorded the highest cumulative rainfall of the past century for our southern sites).
+For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern (*e.g.* first months of 1996 recorded the highest cumulative rainfall of the past century for our southern sites).
 This could explain the higher values of recovery found for 1995 (Figure 6).
@@
 pyrenaica* [@GeaIzquierdo2014].
-Soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017].
-Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive [@Bhuyan2017].
+In fact, soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017].
+Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.
+[@Bhuyan2017].
 
@@
 From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017].
-Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada.
+Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada (Table S4).
 As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].
@@
 These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*.
-
 The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent.
 This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019].
-
 The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents.
@@
 pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.
-
 Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.
@@
 Blanco helped in the filtering of satellite data.
+F.J.
+Cano-Manuel and F.J.
+Navarro provided worthwhile information about oak management projects in Sierra Nevada.
 We also are very thankful to F.J.

### 2019-07-10 man/ecosystems/ms_ecosystems.Rmd
update

@@
 Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure.
-For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history.
+For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history.
 Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@Maestre1858]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity.

### 2019-07-10 man/tables/t_dendro_summary.Rmd
table update

@@
 bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
+csl: ../../refs/ecosystems.csl
 editor_options: 

### 2019-07-10 man/tables/t_droughts.Rmd
update
table update

@@
 bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
+csl: ../../refs/ecosystems.csl
 ---
@@
 ##### 
-**Table S3.** Drought events for Sierra Nevada based on SPEI index (12 months scale). See methods for details.
+**Table S3.** Drought events for the 1901-2016 period for Sierra Nevada based on SPEI index (12 months scale). See methods for details.
 

### 2019-07-10 man/tables/t_forest_history.Rmd
table update

@@
 bibliography: ../../refs/references.bib
-csl: ../../refs/ecology.csl
+csl: ../../refs/ecosystems.csl
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(officer)
+library(tidyverse)
 ```
 
-**Table S3.** Review of the foorest and management history of the sampling sites
-
-In addition, the forest and managment history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
-For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest Table XX
-
-$ convertir en tabla $ 
-: land-uses, forest managment actions, mining, wildfires, etc.
+**Table S4.** Review of the forest and management history of the samplin site.
+An exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest.
 Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
 
-```{r, eval=FALSE}
-### Table R1_evi
-evi <- read.csv(here("/out/anovas_resilience/huber_evi", "trimmed_anovas.csv"), header = TRUE)
-names(evi) <- c('factores', 'F', 'p', 'variables')
+[@Catastro1752; @Lopez1776; @Madoz1846; @Maestre1852; @Maestre1858; @Cruz1991; @Titos1990; @Titos1997; @Ferrer1999; @MesaTorres2009; @Bonet2014_conama; @Navarro2014; @MartinCivantos2014; @MartinMontanes2015; @JimenezOlivencia2015; @2015Zoido; @Wing2015; @Bonet2016obsnev_forest; @MorenoLlorca2016; @RuizRuiz2017; @CMA2018; @Calatrava2019; J.
+Navarro and F.J.
+Cano-Manuel *personal communications*]
 
-dff_evi <- evi %>% gather(temp, numeros, F, p) %>% 
-  unite(temp1, variables, temp) %>% 
-  spread(temp1, numeros) %>% 
-  mutate(factores = recode(factores, 
-                           "disturb_year" = "Drought event", 
-                           "disturb_year:site" = "Drought event:Site", 
-                           "site" = "Site")) %>% 
-  arrange(match(factores, c("Drought event", "Site", "Drought event:Site"))) %>% 
-  mutate_at(c(2,4,6,8), round, 1) %>% 
-  mutate_at(c(3,5,7,9), round, 3) %>% 
-  mutate(variable = 'EVI')
+@Catastro1752; @MesaTorres2009; @Bonet2014_conama
+@Bonet2014_conama; @MorenoLlorca2016; @CMA2018
+@Bonet2016obsnev_forest; @MorenoLlorca2016; J. Navarro and F.J. Cano-Manuel *personal communications*
+@Cruz1991; @Wing2015
+@JimenezOlivencia2015; @2015Zoido; @MorenoLlorca2016; @Calatrava2019
+@Maestre1852; @Maestre1858; @Titos1990; @MesaTorres2009
+@Navarro2014
+@MartinCivantos2014; @MartinMontanes2015; @RuizRuiz2017
+@Catastro1752; @Lopez1776; @Madoz1846; @Titos1997; @Ferrer1999; @MesaTorres2009; @Bonet2014_conama
 
-bai <- read.csv(here("/out/anovas_resilience/huber_bai", "trimmed_anovas.csv"), header = TRUE)
-names(bai) <- c('factores', 'F', 'p', 'variables')
+```{r}
+u_raw <- read_delim(here::here("/data_raw/review_usos.tsv"), delim = "\t") %>% 
+  dplyr::select(-ref) %>% spread(Site, item)
 
-dff_bai <- bai %>% gather(temp, numeros, F, p) %>% 
-  unite(temp1, variables, temp) %>% 
-  spread(temp1, numeros) %>% 
-  mutate(factores = recode(factores, 
-                           "disturb_year" = "Drought event", 
-                           "disturb_year:site" = "Drought event*Site", 
-                           "site" = "Site")) %>% 
-  arrange(match(factores, c("Drought event", "Site", "Drought event*Site"))) %>% 
-  mutate_at(c(2,4,6,8), round, 1) %>% 
-  mutate_at(c(3,5,7,9), round, 3) %>% 
-  mutate(variable = 'BAI')
+u <- regulartable(u_raw, 
+                   col_keys = c("activity", "CA", "SJ", "refall"))
 
-dff <- dff_evi %>% 
-  bind_rows(dff_bai) %>% 
-  dplyr::select(-rrs_F, -rrs_p) %>% 
-  dplyr::select(variable, factores, rt_F, rt_p, rc_F, rc_p, rs_F, rs_p)
-```
-
-```{r, eval = FALSE}
-mt <- regulartable(
-  data = dff, 
-  col_keys = c("variable", "factores", "c3","rt_F", "rt_p", "c1", "rc_F", "rc_p","c2", "rs_F", "rs_p"),
-)
-
-mti <- mt %>% 
-  autofit() %>% 
-  empty_blanks() %>% 
-  align(j = 1,  align = "center") %>% 
-  merge_v(~variable) %>% 
-  set_header_labels(variable = "", factores = "", rt_F = "Resistance", rt_p = "Resistance", rc_F = "Recovery", rc_p = "Recovery", rs_F = "Resilience", rs_p = "Resilience") %>% 
-  add_header(variable = "Variable", factores = "Factor", rt_F = "F", rt_p = "p", rc_F = "F", rc_p = "p", rs_F = "F", rs_p = "p", top = FALSE) %>% 
-  bold(part = 'header') %>% 
+tabla <- u %>% 
+  set_header_labels(activity = "", CA = "Cáñar (CA sites)", SJ = "Güejar-Sierra (SJ site)", refall = "References") %>% 
+  bold(part = "header") %>% 
+  bold(j = 1) %>% 
+  border_outer(part = "all", border = fp_border(color = "black", width = 1.5)) %>% 
+  border_inner_h(part="all", border = fp_border(color = "gray")) %>% 
+  border_inner_v(part="all", border = fp_border(color = "gray")) %>% 
   fontsize(part = "header", size = 10) %>% 
   fontsize(part = "body", size = 9) %>% 
-  width(j = 2, width = 2) %>% 
-  italic(i = ~ rt_p < 0.01, j = ~ rt_p) %>% 
-  italic(i = ~ rc_p < 0.01, j = ~ rc_p) %>% 
-  italic(i = ~ rs_p < 0.01, j = ~ rs_p) %>% 
-  merge_at(i = 1, j=4:5, part="header") %>% 
-  merge_at(i = 1, j=7:8, part="header") %>% 
-  merge_at(i = 1, j=10:11, part="header")
+  autofit(add_w = 0, add_h = 0)
 
-mti
-# mt <- set_formatter(mt, 
-#                     rt_F = function(x) format(round(x, 1)),
-#                     rc_F = function(x) format(round(x, 1)),
-#                     rs_F = function(x) format(round(x, 1))
-#                     ) 
-# 
-# mt
-
+tabla 
 ```

### 2019-07-11 man/tables/t_forest_history.Rmd
update

@@
 @Catastro1752; @Lopez1776; @Madoz1846; @Titos1997; @Ferrer1999; @MesaTorres2009; @Bonet2014_conama
+@JimenezSerrano2004; @RomeroZurbano1909; @Arnedo2007
 

### 2019-07-13 man/ecosystems/ms_ecosystems.Rmd
update ms

@@
 ## **Author Contributions**
-AJPL conceived and desing the study. AJPL and GGI conducted lab work and led article preparation. 
+AJPL conceived and desing the study and wrote the manuscript. AJPL and GGI conducted lab work and led article preparation. 
 AJPL, GGI and RZ conducted field work, collect the data and contribute to writing and editing the manuscript. All authors contribute to the drafts and gave final approval for publication.
@@
 In addition, the forest and managment history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
-For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest Table XX
+For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest (Table S4).
 
-$ convertir en tabla $ 
-: land-uses, forest managment actions, mining, wildfires, etc.
+```{r eval=FALSE, echo=FALSE}
+# : land-uses, forest managment actions, mining, wildfires, etc.
 Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
+```
 
@@
 Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisioetal2012IPoGEC].
-On the other hand, tree-ring width data provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
+In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates.
+This is particularly important for *Q.
+pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014].
+On the other hand, tree-ring records allow to contextualize remote sensing data by reflecying climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
+Combined anlaysis, such the approach used in our study, may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [Babst2018].
 
@@
 From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017].
-Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada (Table S4).
+Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada, and on Pyrenean oak forest, as can be derived from the results of the review of historical documents (Table S4).
 As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].
@@
 We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy).
-The review of historical documents revealed that our study sites had different land-use trajectories driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014].
+The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4), driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015] (Table S4).
 On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [@JimenezOlivencia2015]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops.
 Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure.
-For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history.
-Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@Maestre1858]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity.
-On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019].
-Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@Valbuena2013; @JimenezSerrano2004].
+For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history (Table S4).
+On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation [@Calatrava2019].
+Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@Valbuena2013].
 In Cáñar site, there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century.
-From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].
-Acorn resources have been tradiotionally exploited in this location up to 1950s [@MesaTorres2009; @Bonet2014_conama].
+From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment.
+Acorn resources have been tradiotionally exploited in this location up to 1950s (Tabla S4).
 
 Those different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5).
-We observed a tree-growth release event at 1940s in the SJ oak woodland (Figures 5, S4) which concurs with one period of maximum mining activity for this area.
-Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections [@Titos1990].
+We observed a tree-growth release event at 1940s in the SJ oak woodland (Figures 5, S4) which concurs with one period of maximum mining activity for this area (1925 to 1957) (Table S4).
 During this period there were an increase on the use of timbers for the tunnels of the mines and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area [@Titos1990].
@@
 In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017].
-
 The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4).
@@
 Two non-mutually exclusive ways could explain the tree-growth release observed.
-Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004].
+Firstly it could be related with a natural drought-induced mortality event after 1995 (*personal communications*) as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004].
 On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014].
@@
 
-The only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013].
+For southern sites, the only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013].
 

### 2019-07-13 man/figures/resilience_long_term.Rmd
update ms

@@
 
-```{r plot_rt}
+```{r, eval=FALSE, echo=FALSE}}
 sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
@@
 
-```
-
-```{r plot_rs}
 sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
@@
   theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt"))
-```
 
-```{r plot_rc}
 sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
@@
   theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
-```
 
-```{r plot_join, eval=FALSE, echo=FALSE}
 res_long_term <- sevRt + sevRc + sevRs + 
@@
                     parse = TRUE, show.legend = FALSE) +
-    annotate("text", x = .0, y = 10, 
+    annotate("text", x = xLM, y = yLM, 
              label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-             size = 5, fontface = "bold", hjust="left") 
+             size = 4, fontface = "bold", hjust="left", col = "gray40") 
 }
@@
 sevRt_site <- plotrelsite(sdfsite, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rt", coord_eqX = Inf) + xlab('') + ylab('')
-
-+ 
-  xlim(5,25)
+             titulo = "Rt", coord_eqX = Inf, xLM = -Inf, yLM = 0.5) + xlab('') + ylab('') + xlim(5,25)
 
 sevRs_site <- plotrelsite(sdfsite, v = "rs", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rs", coord_eqX = -Inf) + xlab('') + ylab('') + xlim(5,25)
+                 titulo = "Rs", coord_eqX = -Inf, xLM = 10, yLM = 0.5) + xlab('') + ylab('') + xlim(5,25)
 
 sevRc_site <- plotrelsite(sdfsite, v = "rc", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rc",  coord_eqX = -Inf) +
+                 titulo = "Rc",  coord_eqX = -Inf, xLM = 10, yLM = 0.5) +
   xlab("Drought Severity") + ylab("") + xlim(5,25)
@@
 
-ggsave(filename = here::here("./man/figures/svg/fig_6.jpg"), 
-       dpi = 300,  plot=res_long_term_sites, width=8.5, height=7)
+ggsave(filename = here::here("./man/figures/svg/fig_6_95.jpg"), 
+       dpi = 300,  plot=res_long_term_sites, width=25, height=14, units = "cm")
+ggsave(filename = here::here("./man/figures/svg/fig_6_95.eps"), 
+       dpi = 300,  plot=res_long_term_sites, width=25, height=14, units = "cm")
+ggsave(filename = here::here("./man/figures/svg/fig_6_95.svg"), 
+       dpi = 300,  plot=res_long_term_sites, width=25, height=14, units = "cm")
 
@@
 sevRt_site95 <- plotrelsite(sdfsite95, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rt", coord_eqX = Inf) + xlab('') + ylab('')
+                 titulo = "Rt", coord_eqX = Inf, xLM = -Inf, yLM = 0.5) + xlab('') + ylab('')
 
 sevRs_site95 <- plotrelsite(sdfsite95, v = "rs", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rs", coord_eqX = -Inf) + xlab('') + ylab('')
+                 titulo = "Rs", coord_eqX = Inf, xLM = 10, yLM = 0.5) + xlab('') + ylab('')
 
 sevRc_site95 <- plotrelsite(sdfsite95, v = "rc", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rc",  coord_eqX = -Inf) + xlab("Drought Severity") + ylab("")
+                 titulo = "Rc",  coord_eqX = -Inf, xLM = 10, yLM = 0.5) + xlab("Drought Severity") + ylab("")
 
@@
 
-ggsave(filename = here::here("./man/figures/svg/fig_6_95.jpg"), 
-       dpi = 300,  plot=res_long_term_sites95, width=8.7, height=7)
-
+ggsave(filename = here::here("./man/figures/svg/fig_6_no95.jpg"), 
+       dpi = 300,  plot=res_long_term_sites95, width=25, height=14, units = "cm")
+ggsave(filename = here::here("./man/figures/svg/fig_6_no95.eps"), 
+       dpi = 300,  plot=res_long_term_sites95, width=25, height=14, units = "cm")
+ggsave(filename = here::here("./man/figures/svg/fig_6_no95.svg"), 
+       dpi = 300,  plot=res_long_term_sites95, width=25, height=14, units = "cm")
 ```
@@
 ```
+
+# Check lm partial
+
+```{r, eval= FALSE, echo = FALSE}
+library(ggpubr)
+
+ggscatter(
+  sd, x = "Sepal.Length", y = "Sepal.Width",
+  color = "Species", palette = "jco",
+  add = "reg.line"
+  ) +
+  facet_wrap(~site) +
+  stat_cor(label.y = 4.4) +
+  stat_regline_equation(label.y = 4.2)
+
+df1 <- sdfsite %>% dplyr::select(disturb_year, d_severity, rt_mean_mean, site)
+
+  ggscatter(df1, x = "d_severity", y = "rt_mean_mean", 
+            color = "site", palette = c("#00AFBB", "#E7B800", "#FC4E07"), 
+            label = "disturb_year", font.label = c(8, "plain"), repel = TRUE, 
+            add = "reg.line", ccor.coef = TRUE) +
+    ggpubr::stat_cor(label.x= 15, aes(color = site, label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
+    ggpubr::stat_cor(label.x= 15, aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")))
+
+df2 <-sdfsite %>% dplyr::select(disturb_year, site, rt_mean_mean, rs_mean_mean, rc_mean_mean, d_severity) %>%  gather(variable, value, c("rt_mean_mean", "rs_mean_mean", "rc_mean_mean"))
+
+ggscatter(df2, x = "d_severity", y = "value", 
+            color = "site", palette = c("#00AFBB", "#E7B800", "#FC4E07"), 
+            label = "disturb_year", font.label = c(8, "plain"), repel = TRUE, 
+            add = "reg.line", ccor.coef = TRUE) + 
+  facet_wrap(~variable, scales = "free") + 
+  ggpubr::stat_cor(label.x= 15, aes(color = site, label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
+    ggpubr::stat_cor(label.x= 15, aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")))
+
+```

### 2019-07-15 man/ecosystems/ms_ecosystems.Rmd
update ms

@@
 ## **Author Contributions**
-AJPL conceived and desing the study and wrote the manuscript. AJPL and GGI conducted lab work and led article preparation. 
-AJPL, GGI and RZ conducted field work, collect the data and contribute to writing and editing the manuscript. All authors contribute to the drafts and gave final approval for publication.
+AJPL, GGI and RZ conceived of or desing the study.
+AJPL, GGI and RZ conducted field work and collected the data.
+AJPL and GGI conducted lab work and led article preparation.
+AJPL analyzed data and wrote the paper.
+GGI and RZ contributed in the writing process.
+All authors contribute to the drafts and gave final approval for publication.
 
-# Abstract (369)
-* Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified species composition and the forest structure as a result of intensive human pressure in the past.
+# Abstract
+* Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified the forest structure as result of intensive human-pressures in the past.
 We analyzed forest resilience of adult trees to drought and climate change of relict Mediterranean *Quercus pyrenaica* Willd.
 populations in the Southern Iberian Peninsula.
-Forest resilience was analyzed using both dendroecological growth estimates and MODIS satellite images as proxies for secondary and primary growth respectively.
+Forest resilience was analyzed using both dendroecological growth estimates and greenness (NDVI) as proxies for secondary and primary growth respectively.
+Additionally forest history were inferred from an exhaustive review of historical documents.
 
-* Past intense land-use in the studied populations were likey expressed by disturbance chronologiess, hence 
-partly reflecting the origin of the current forest structure and land-use legacies.
+* Intense land-use were likely expressed by disturbance chronologies, reflecting the origin of the current forest structure and land-use legacies.
 Trees expressed a high sensitivity to drought.
-Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (*i.e.* warming).
-This would contradict the *a priori* vulnerability that we expected in respone to climate change of the studied relict stands at the species current rear-edge.
-Yet, there was variation in the response to climate and resilience following a water-stress gradiente along our three studied forests.
-We found a positive growth trend that was strongest in the high elevation stand and non-significant in the most xeric site (or the most transformed???) since the end of the 1970s matching the simultaneous increase in temperature observed.
-
-* Resilience to drought followed a dryness gradient among the three populations studied: Northern populations thriving under a drier climate were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.
+Resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and to long-term climatic changes.
+Response to climate and resilience following a water-stress gradient.
+We found a positive growth trend since the end of the 1970s, that was strongest in the high-elevation stand and non-significant in the most xeric site.
+Northern populations were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.
 
@@
 ## Manuscript highlights
-* H1
-* H2 %>% %>% 
-* H3
-
-Provide three bullets (up to 85 characters for each, including spaces) that highlight the important contributions of your study for a general reader.
+* Forest resilience of *Quercus pyrenaica* rear-edge forests were analyzed at tree and stand-level
+* Response to climate and resilience following a water-stress gradient
+* Trees and stand expressed a high sensitivity to drought and to land-use legacies
 

### 2019-07-15 man/ecosystems/ms_ecosystems.Rmd
update text and figures

@@
 title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies along a climatic gradient in the rear-edge"
-author: "A.J. Perez-Luque^1,2^; G. Gea-Izquierdo^3^ and Zamora, R.^1,2^"
 output:
@@
 
-## **Short Title**: Resilience of relict oaks in the rear-edge (máximo 45 caracteres)
+## **Short Title**
+Resilience of relict oaks in the rear-edge (max. 45 characters)
+
+## Authors [^+^] 
+A.J. Perez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^
 
@@
 
-## **Author Contributions**
-AJPL, GGI and RZ conceived of or desing the study.
+^a^ ajperez@ugr.es
+
+^b^ gea.guillermo@inia.es
+
+^c^ rzamora@ugr.es
+
+[^+^]: Author Contributions.
+AJPL, GGI and RZ conceived of or design the study.
 AJPL, GGI and RZ conducted field work and collected the data.
@@
 
-# Abstract
+## Manuscript highlights
+* Forest resilience of *Quercus pyrenaica* rear-edge forests were analyzed at tree and stand-level
+* Response to climate and resilience following a water-stress gradient
+* Trees and stand expressed a high sensitivity to drought and to land-use legacies
+
+## Abstract
 * Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified the forest structure as result of intensive human-pressures in the past.
@@
 
-## keywords
+## Keywords
 extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing
-
-## Manuscript highlights
-* Forest resilience of *Quercus pyrenaica* rear-edge forests were analyzed at tree and stand-level
-* Response to climate and resilience following a water-stress gradient
-* Trees and stand expressed a high sensitivity to drought and to land-use legacies
 

### 2019-07-15 man/figures/0_figures.Rmd
update figures std anomalies and EVI profile
update figures
update figures and chronos
update text and figures

--- 2018-10-16 man/figures/0_figures.Rmd
@@
 
+```{r, include=FALSE}
+options(knitr.duplicate.label = 'allow')
+```
+
 ```{r child = 'location_map.Rmd'}
@@
 ```
+
+#####
+
+```{r child = 'browning.Rmd'}
+```
+
+#####
+
+```{r child = 'bai_cronos.Rmd'}
+```
+
+#####
+
+```{r child = 'growth_changes.Rmd'}
+```
+
+#####
+
+**Figure 6.** Resilience metrics of the tree-growth for severe drought events since 1950.
+*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
+Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*).
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
+Gray line represents overall relationship for each Resilience metrics.
+
+![](../../man/figures/svg/fig_6_95.jpg){height=750px}
+
+#####
+
+```{r child = 'resilience_metrics.Rmd'}
+```
+
+#####
+
+```{r child = 'ring_climate.Rmd'}
+```
+
+#####
+
+```{r child = 'cumulative_prec.Rmd'}
+```
+
+#####
+
+```{r child = 'plot_spei12.Rmd'}
+```
+
+#####
+
+```{r child = 'residual_chronos.Rmd'}
+```
+
+#####
+
+```{r child = 'gc_perc_trees.Rmd'}
+```
+
+#####
+
+```{r child = 'corr_plot_years.Rmd'}
+```
+
+#####
+
+```{r child = 'correla_boot_sites.Rmd'}
+```
+
+#####
+
+```{r child = 'evi_vs_severe_droughts.Rmd'}
+```
+
+#####
+
+**Figure S8.** Resilience metrics of the tree-growth for severe drought events since 1950 (excluding 1995 drought event).
+*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
+Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*).
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
+Gray line represents overall relationship for each Resilience metrics.
+
+![](../../man/figures/svg/fig_6_no95.jpg){height=750px}

### 2019-08-05 analysis/bai/analysis_resilience_bai.Rmd
update review GGI

@@
 library('car')
-library('auxiliar')
+#library('auxiliar')
 library('WRS2')
@@
 compete <- compete %>% 
-  mutate(dn_mm = (perim_mm / pi))
+  dplyr::mutate(dn_mm = (perim_mm / base::pi))
 

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
update review GGI

@@
 ---
-title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies along a climatic gradient in the rear-edge"
+title: "Relict deciduous oaks show high resilience to drought along a climatic gradient in the rear-edge despite strong land-use legacies "
 output:
@@
 ## **Short Title**
-Resilience of relict oaks in the rear-edge (max. 45 characters)
+Relict oaks resilience at their rear-edge
 
-## Authors [^+^] 
+## Authors [^\*^] 
 A.J. Perez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^
@@
 
-[^+^]: Author Contributions.
-AJPL, GGI and RZ conceived of or design the study.
-AJPL, GGI and RZ conducted field work and collected the data.
+[^\*^]: Author Contributions.
+AJPL, GGI and RZ conceived the study, conducted field work and collected the data.
 AJPL and GGI conducted lab work and led article preparation.
-AJPL analyzed data and wrote the paper.
+AJPL analyzed data and led the writing of the paper.
 GGI and RZ contributed in the writing process.
-All authors contribute to the drafts and gave final approval for publication.
+All authors contributed to the drafts and gave final approval for publication.
 
 ## Manuscript highlights
-* Forest resilience of *Quercus pyrenaica* rear-edge forests were analyzed at tree and stand-level
-* Response to climate and resilience following a water-stress gradient
-* Trees and stand expressed a high sensitivity to drought and to land-use legacies
+* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
+* The responses to climate and resilience followed a water-stress gradient
+* Trees and stand expressed high sensitivity to drought and land-use legacies
 
 ## Abstract
-* Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified the forest structure as result of intensive human-pressures in the past.
-We analyzed forest resilience of adult trees to drought and climate change of relict Mediterranean *Quercus pyrenaica* Willd.
+
+* Global change challenges ecosystems in dry locations transformed by intensive human use.
+We analyzed resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd.
 populations in the Southern Iberian Peninsula.
-Forest resilience was analyzed using both dendroecological growth estimates and greenness (NDVI) as proxies for secondary and primary growth respectively.
-Additionally forest history were inferred from an exhaustive review of historical documents.
+Forest resilience to drought was analyzed in relation to historical records of land-use using dendroecological growth and greenness (EVI) as proxies for secondary and primary growth.
 
-* Intense land-use were likely expressed by disturbance chronologies, reflecting the origin of the current forest structure and land-use legacies.
-Trees expressed a high sensitivity to drought.
-Resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and to long-term climatic changes.
-Response to climate and resilience following a water-stress gradient.
-We found a positive growth trend since the end of the 1970s, that was strongest in the high-elevation stand and non-significant in the most xeric site.
-Northern populations were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.
+* The response to climate and resilience followed a water-stress gradient.
+Primary and secondary growth of more xeric populations (*i.e.* Northern aspect and low elevations) was less resilient to drought than that at more mesic sites.
+In addition, a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic high-elevation stands, but absent in the most xeric site.
+Trees were highly sensitivity to drought and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term.
+The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure.
+
+* The species showed high resilience to drought despite being relict and at its rear-edge.
+Yet, the trees exhibited a variable response to the changing climate along a very short climatic gradient.
+The trees were more sensitive to water availability at most xeric sites whereas only the coldest and more humid stands exhibited a recent positive growth trend.
+Stand dynamics at the most xeric locations (*i.e.* the altitudinal rear-edge) needs to be monitored under future expected climatic scenarios bearing in mind that land-use legacies play a key role in stand dynamics.
 

### 2019-08-05 man/tables/t_dendro_summary.Rmd
update review GGI

@@
 **Table 2.** Characteristics of the mean tree ring chronologies.
-Values of the length year in parenthesis indicate years replicated with more than five series.
+Values of length between parentheses indicate number of years replicated with more than five series.
 RW = mean annual ring width (standard deviation in parenthesis).
@@
 EPS = mean expressed population signal.
-EPS and Rbt are calculated for the mean residual chronologies of growth indices.
+EPS and Rbt were calculated for the mean residual chronologies of growth indices.
 

### 2019-08-05 man/tables/t_huber_res_bai.Rmd
update review GGI

@@
 Measures of central tendency are M-estimators based on Huber's Psi (see material and methods).
-95 % confidence intervals using 3000 bootstrap are included in parentheses.
+95 % confidence intervals using 3000 bootstrap are included between parentheses.
 
@@
   add_header(site="", '2005rt' = "2005", '2005rc' = "2005", '2005rs' = "2005",'2012rt' = "2012", '2012rc' = "2012", '2012rs' = "2012",
-                        '2099rt' = "", '2099rc' = "", '2099rs' = "", top = TRUE) %>% 
+                        '2099rt' = "Total", '2099rc' = "Total", '2099rs' = "Total", top = TRUE) %>% 
   merge_at(i = 1, j =c(3:5) , part = "header") %>% 

### 2019-08-05 man/tables/t_robust_anovas.Rmd
update review GGI

--- 2019-02-06 man/tables/t_robust_anovas.Rmd
@@
   italic(i = ~ rs_p < 0.01, j = ~ rs_p) %>% 
+  compose(i = ~rt_p == 0.001, j = ~rt_p,
+          value = as_paragraph("< ", as_chunk(rt_p))) %>% 
+  compose(i = ~rc_p == 0.001, j = ~rc_p,
+          value = as_paragraph("< ", as_chunk(rc_p))) %>% 
+  compose(i = ~rs_p == 0.001, j = ~rs_p,
+          value = as_paragraph("< ", as_chunk(rs_p))) %>% 
   merge_at(i = 1, j=4:5, part="header") %>% 

### 2019-08-05 man/tables/t_sampled_plots.Rmd
update review GGI

@@
 Different letters indicate statistically significant differences between sites (Kruskal-Wallis test followed by Dunn’s test, p<0.05).
+Stands were monospecific, hence all results correspond to oak data.
 
@@
              n_trees = "Cored trees", dn = "Cored trees", h = "Cored trees", age = "Cored trees", 
-             dn_all = "Competition", h_all = "Competition", ba = "Competition", 
-             stand_density = "Competition", srd = "Competition", top = TRUE) %>% 
+             dn_all = "Stand competition", h_all = "Stand competition", ba = "Stand competition", 
+             stand_density = "Stand competition", srd = "Stand competition", top = TRUE) %>% 
   merge_at(i = 1, j =c(8:11) , part = "header") %>% 

### 2019-08-05 script/R/bai_piovesan.R

@@ -13,11 +13,17 @@ bai_piovesan <- function(rwdf, diam_df){
   # Output 
   bai <- rwdf 
   
+  if("nseries" %in% colnames(rwdf)) { 
+    rwdf <- subset(rwdf, select=-nseries)
+    }
+
+
   # length years in Ring-width Data frame 
   vec_ly <- seq_len(nrow(rwdf))
   
   # ID cores 
   cores <- colnames(rwdf)
+  
   
   for (i in cores){ 
     # 1 # Select core 
@@ -48,7 +54,7 @@ bai_piovesan <- function(rwdf, diam_df){
     d <- d[-length(d)]
     
     # Bai
-    b <- rev((pi/4) * ((d)^2 - (d - 2*rws_rev)^2))
+    b <- rev((base::pi/4) * ((d)^2 - (d - 2*rws_rev)^2))
     
     # Store BAI and NA 
     ### Which elements are NA? 

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
incorporate introduction comments GGI RZ

@@
 
-Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of Mediterranean climate [@Dai2011].
+Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of the Mediterranean climate [@Dai2011].
 There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017].
-This is especially critical, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@Spinoni2017a].
+This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a].
 
-Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation.
-The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].
+Drought causes significant changes in ecosystem productivity and dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation.
+Thus, assessment of drought impacts on ecosystems has gained much attention in last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].
 
-Ecosystem responses to drought are influenced by other drivers of environmental changes [@Penuelas2017].
+Ecosystem responses to drought are influenced by other drivers of environmental changes.
 In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017].
@@
 
-The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate [@Hampe2005].
+The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005].
 The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010].
-It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2013].
+This becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].
 
-Many studies have analyzed the effects of drought on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Dorado2017b; @Caminero2018].
-Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
-Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973].
+Many studies have analyzed the effects of drought on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
+Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973].
 The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services.
-This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.
 
+Vegetation responses to drought requires of integrative and multidisciplinary strategies for assessing the impacts of disturbances on forest at several scales [@Hartman2018].
 The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously.
@@
 The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
-Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Zhang2013].
-Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013].
+Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013].
+Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us drought [*e.g.* @VicenteSerrano2013].
 Drought impacts on vegetation using remote sensing have been analyzed at different scales.
 For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017].
+Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
+Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level, particularly in rear edge populations.
 
-Vegetation responses to drought requires of integrative and multidisciplinary strategies for assessing the impacts of disturbances on forest at several scales [@Hartman2018].
-Several works assessed vegetation responses in term of resistance and resilience to drought events highlithg a stronger reponse of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
-Nonetheless, is crucial to know if the responses at the tree level differ from responses at the ecosystem level, particularly in rear edge populations.
-
-In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) on southern relict forests at the rear-edge of species distribution and where species performance is considered to be threatened by climate change.
-We also assessed the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.
-Our main hypothesis is that these relict populations driven by historical land-use at their present climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show lower-values of resilience after several extreme drought only at the dry edge of climatic gradients.
-To address this hyphotesis our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q.
-pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change along climatic gradients.
-To achieve goal (*iii*) within the region of study we assessed whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
+In this work we used remote sensing information and dendroecological methods to evaluate the drought impact in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of a Mediterranean tree species (*Quercus pyrenaica* Willd.) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be threatened by climate change.
+We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) in the long-term in the last decades.
+Our main hypothesis is that these relict populations driven by historical land-use are particularly vulnerable to climate change at their present climatic (either altitudinal or latitudinal) rear-edge, and will show low resilience after extreme drought particularly at the dry edge along climatic gradients at the small scale within the rear-edge.
+To address this hypothesis our specific aims were: (*i*) To quantify how recent extreme drought events influenced primary and secondary growth of *Q.
+pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of this species to detect vulnerability to climate change along climatic gradients.
+To achieve goal (*iii*) we assessed within the region of study whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
 
@@
 This is especially interesting for southern sites, which are very close to each other.
-Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b].
+Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
 The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q.

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
update comments MandM

@@
 
-The forests of this species reaches one of their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*.
+The forests of this species reache their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 m and 3482 m *a.s.l.*.
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
 There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
-Sierra Nevada mountain is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002].
-In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005].
+Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002].
+Today, *Q.
+pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005].
 They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG].
-These relictic forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2017].
-Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge [@GeaIzquierdo2013].
+These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2017].
+Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the rear-edge [@GeaIzquierdo2013].
 
@@
 
-During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*.
+The Iberian Peninsula suffered several extreme drought episodes in the last decades of the twentieth century (*e.g*.
 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
@@
 A drought event is considered only when SPEI values are below the threshold for at least two consecutive months.
-Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event.
+Several indicators were computed for each drought event: *duration* as the number of consecutive months with the SPEI lower than a certain threshold; *severity* of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
 
-## Greenness data
+## Greenness data to assess ecosystem resilience
 
@@
 pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS (*Moderate Resolution Imaging Spectroradiometer*) sensor.
-MODIS EVI Data (Collection 6) were obtanied for the period 2000 - 2016.
-EVI data consits of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
+MODIS EVI Data (Collection 6) were obtained for the period 2000 - 2016.
+EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
 We selected the pixels covering the distribution of *Q.
@@
 
-## Field sampling and dendrochronological methods
+## Field sampling and dendrochronological methods to assess individual resilience
 
 Tree sampling was carried out during autumn 2016.
-Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warmer site located at the southern aspect (Figure 1; Table 1).
-For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope.
+Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid but warmer site located at the southern aspect (Figure 1; Table 1).
+For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope.
 Sampling sites follow a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
@@
 In addition, stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius.
-Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$.
+Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].
 Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests.
@@
 Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
-Cross-dating validation was finally done using COFECHA software [@Holmes1983].
+Cross-dating validation was finally verified using COFECHA [@Holmes1983].
 
@@
 Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010].
-Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990].
+Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976].
 The statistical quality of each chronology was checked via the expressed population signal.
@@
 where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
-For each individual tree, a mean BAI serie was calculated.
+For each individual tree, a mean BAI series was calculated.
 Then, mean site BAI chronologies were obtained by averaging individual tree BAI time series.
@@
 
-Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using `treeclim` package [@Zang2015].
+Relationships between residual site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].
 
@@
 Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g.
-logging, neighbor mortality) in the past.
-By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997].
+logging, drought induced neighbor mortality) in the past.
 Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:
@@
 $$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
-where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median.
-Medians were used since they are more robust estimator of central tendency than means [@Rubino2004].
+where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
+Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
-To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance.
+To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance [*e.g.* @GeaIzquierdo2014].
 
-In addition, the forest and managment history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
-For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest (Table S4).
+In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
+For this purpose, an exhaustive review of existing abundant historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).
 
@@
 
-To evaluate the effects of drougth events on ecosystem resilience (using greeennes data) and tree individual resilience (using BAI growth data) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
-The Resistance index, estimated as the ratio between the performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred.
+To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
+The Resistance index, estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred.
 The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity.
@@
 The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event.
-2005 and 2012 were considered as single drought events.
 The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively.
@@
 Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012].
-When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers.
-Post-hoc differences were assessed pairwise using a similar boostrap test.
+When running the robust ANOVA test, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers.
+Post-hoc differences were assessed pairwise using a similar bootstrap test.
 All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package.

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
update Results

@@
 The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in northern (dry) populations (Figure 2).
-In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q.
-pyrenaica* forests during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).
+A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).
 
@@
 The trees of the southern population were older than those from the northern one.
-In addition trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4).
-Both growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6).
-Competition was similar among all sites.
-Yet plot basal area was greatest in CA-High (Table 1), despite the highest individual BAI growth at that site (Figure 4).
-Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High).
+In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4), despite competition measured as plot basal area was greatest in CA-High (Table 1).
+Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6).
+Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being much more pronounced for the highest, hence more humid, site (CA-High).
 
-Drought events produced reduction of RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies.
+Drought events produced reduction in RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
 The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S3).
-Tree-growth reduction followed a moisture-gradient, being weaker for the wetter site (CA-High).
-Trees from northern site also suffered a great reduction during 1945-1946 drought event (Figure 5).
+Tree-growth reduction followed a moisture-gradient, being weaker for the wettest site (CA-High, Figure 5).
 
-The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940 decade (the most evindent) and the second in the period 1995-2000.
-These periods alternate with periods of supression for over 40 years.
-The two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.
+The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940s (the most evident) and the second in 1995-2000.
+These periods alternated with periods of suppression.
+In contrast, the two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.
 
 ## Resilience to drought events at the ecosystem and individual tree levels
-The exploration of the resilience metrics of tree-growth for severe drought events since 1950 (*i.e.* for the share period among the three chronologies, excluding juvenile years, Table S3) revealed a positive relation between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6).
+Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6).
 A similar pattern was obtained for Resilience (*Rs*) but only significant for SJ site (Figure 6).
-Non-significant patterns were obtained if we exclude 1995 from the analysis (Figure S8).
+Importantly, non-significant patterns were obtained if we excluded 1995 from the analysis, except for *Rc* in SJ (Figure S8).
 Trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), and particularly SJ, where our results suggest a major release event after 1995 (Figure 5).
 
-During the last two drought events, resilience metrics for greenness and tree-growth were significantly different bewteen drought events (Table 3).
-The 2005 drought event reduced greennes and growth more than that of 2012 (Tables S1 and S2) but the two metrics of resilience generally covaried in the same direction.
+During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table 3).
+The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1 and S2) but the two metrics of resilience generally covaried in the same direction.
 We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7).
 Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth.
-Recovery values for greenness was higher for 2005 than for 2012 drought event (Table S1, Figure 7).
+Recovery values for greenness were higher for 2005 than for 2012 drought event (Table S1, Figure 7).
 
-Resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3).
+Overall, resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3), which was similar among the three sites.
 The two southern populations showed lower values for Recovery (*Rc*) than the northern site both for greenness and tree-growth (Tables S1 and S2).
@@
 ## Tree-growth response to climate
-Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge).
-Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 8a).
-Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population.
-Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), being this relationship higher in the northern population (r > 0.6 *vs.* r < 0.5).
-Relationships with temperatures were lower than those with moisture variables.
-Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while no significant limitations were recorded for northen site regarding the temperatures (Figure 8b and 8c).
-Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population.
-No positive relationship with temperature was found for southern trees.
+The response of tree-growth to water availability was greater than that to temperatures, as it can be expected for a Mediterranean site and particularly a species located closer to its xeric/dry distribution limit (rear-edge).
+Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 8a).
+Yet there were sone differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 8a).
+In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 8b and 8c).
+Minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
 

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
update review discussion

@@
 # Discussion
-## Rear-edge oaks show high resilience in response both to recent drought events and long-term climatic variability. 
-Severe drought events negatively affected to primary and secondary growth of *Q.
-pyrenaica* forests in the rear edge of their distribution.
-Our findings show that 2005 and 2012 drought events, provoked reductions in greenness and tree-growth of these forests.
-These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
-In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
-Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
-pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period).
-This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi].
-For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends were described for this species along its distribution range [@GeaIzquierdo2014] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970.
-Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
+## Relict oaks show high resilience to recent drought events and long-term climatic variability
+
+Severe drought negatively affects both primary and secondary growth of *Q.
+pyrenaica* forests.
+This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
+Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
+Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g.* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
+pyrenaica* for the last 16 years (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period).
+This confirms previous findings that pointed out a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi].
+For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4).
+Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014].
+However, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
 sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].
 
-Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2).
-Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005.
-These results could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought.
+Despite the 2012 drought event was more severe and intense than 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012.
+This could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought.
 The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q.
 pyrenaica* forests (Figure S7).
-Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].
-
-As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015].
-But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6).
-In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6).
-Surprisingly this result seems contradict our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge after extreme drought events.
+Thus, this highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], as also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].
+In addition, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species [@SanchezSalguero2012; @Camarero2015].
+But, interestingly, the highest values of resilience for tree-growth were obtained for these two drought events.
+In addition, we observed a positive relation between the tree-growth resilience and drought severity (Figures 6).
+This result seems contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events.
+Yet, the high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J.
+Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of individual resilience estimated from individual growth series.
 
 Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005].
-A high vulnerability to drought is often assumed for populatios located at their rear-edge [@MartinezVilalta2018].
+A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018].
 However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q.
-pyrenaica* and high values of resilience to severe droughts were also reported.
-Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017].
+pyrenaica* and high values of resilience to severe droughts.
+Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017].
 @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations.
-The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada, seem to be indicative of high local adaptation of this oak to this mountain range.
+The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013], seem to be indicative of high local adaptation of this oak.
 In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
@@
 In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q.
-pyrenaica* in Sierra Nevada are located in a geographical but not climatic rear-edge.
-
-A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6).
-A plausible explanation could be the water availability previously and after the drought event.
-For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern (*e.g.* first months of 1996 recorded the highest cumulative rainfall of the past century for our southern sites).
-This could explain the higher values of recovery found for 1995 (Figure 6).
-The high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality occurred in this area (*personal communications*).
+pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge.
+And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges.
 
@@
 pyrenaica* showed different sensitivity to severe droughts.
-During the last drought events EVI was reduced to 90.25 % (Table 1), whereas BAI was reduced to 70.83 % (Table S2), suggesting a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate.
-For our driest site (SJ, the northern one), BAI were reduced to 44.5 % respect to that of the preceding period during 2005 drought event, whereas the EVI was reduced to 81.9 % (Tables S1 and S2).
-These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018].
-Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018].
-Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisioetal2012IPoGEC].
+During the last drought events EVI was reduced to 90.3 % (Table 1), whereas BAI was reduced to 70.8 % (Table S2), suggesting a lower sensitivity to drought of greenness than of tree-growth, particularly for drier sites.
+These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018].
+Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based (*i.e.* ecosystem or stand based); the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018].
+The response of greenness to drought not only reflects the response of trees but from all vegetation within the pixel [@Dionisioetal2012IPoGEC].
 In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates.
@@
 pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014].
-On the other hand, tree-ring records allow to contextualize remote sensing data by reflecying climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
-Combined anlaysis, such the approach used in our study, may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [Babst2018].
+On the other hand, tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
+A combined analysis, such the approach used in our study, may allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018].
 
-## Microenvironment shapes differential sensitivity to climate and drought of rear-edge oaks
+## Small-scale environmental variability shapes the sensitivity to climate of trees within the rear-edge
 
-Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3).
-The former, which have drier conditions (Table 1), were more impacted by the drought events than the latter.
-For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2).
-In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 8), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
+Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event.
+In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site, can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014].
+Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
+pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
+Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017].
 
-Moisture availability has been reported as the most limiting factor driving radial growth of *Q.
-pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014].
-We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a).
-Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in oaks [*e.g.* @GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017]
+It has been described an effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive, thus high-elevations can be used as local refuge for species in response to climate change [@Bhuyan2017].
+It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
+Remarkably, the trees at CA-High, which are located around 1900 m.
+*a.s.l.* and represent today the upper altitudinal limit (treeline) for the species in this southernmost location, showed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4).
+This shows the high variability in the response to climate exhibited in a small gradient and is especially interesting for the southern sites, which are very close to each other, and overall both constitute the rear-edge for the species.
+The higher values of tree growth in taller trees that we observed in CA-High site strongly suggest a lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017].
+Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014].
 
-A striking result is the difference for tree growth between sites (Figure 4).
-The trees of CA-High site, which are located around 1900 m.
-*a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4).
-This is especially interesting for southern sites, which are very close to each other.
-Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
-The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q.
-pyrenaica* [@GeaIzquierdo2014].
-In fact, soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017].
-Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.
-[@Bhuyan2017].
+Finally and in a context of global change, in addition to the withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).
 
@@
 
-In addition to the lower water stress, it is likely that the sites at higher elevation had a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).
+## Land-use legacies shape forests and sensitivity to climate change at the rear-edge
 
-## Land-use legacies shape distribution and sensitivity to climate change at the read-edge
-
-The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history.
+The Mediterranean landscapes, even at high elevations, have suffered a strong transformation driven by human activities throughout history.
 Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017].
-From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017].
-Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada, and on Pyrenean oak forest, as can be derived from the results of the review of historical documents (Table S4).
+From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with a sharp intensification of the human activities in the last 150 years [@GarciaAlix2017].
+Mining, forest clearings, firewood extraction and charcoal exploitations (these were strongly suggested in the disturbance chronology in SJ), pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada, and on Pyrenean oak forest, as can be derived from the results of the review of historical documents.
 As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].
-
-Our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation.
-Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before that date.
+Therefore, our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation.
+Although the estimated age for our sampled trees was lower than 180 years, several documents reported abundant oak presence before that date.
 For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991].
@@
 
-We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy).
-The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4), driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015] (Table S4).
-On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [@JimenezOlivencia2015]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops.
-Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure.
-For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history (Table S4).
-On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation [@Calatrava2019].
-Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@Valbuena2013].
-In Cáñar site, there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century.
-From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment.
-Acorn resources have been tradiotionally exploited in this location up to 1950s (Tabla S4).
+Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between Northern (coppice) and Southern populations (high forest, open woodland) [@JimenezOlivencia2015].
+It has been described a sequentially distribution of the land uses along the elevational gradient in the northern site (*e.g.* San Juan): grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops [@JimenezOlivencia2015].
+Other activities (like mining) should have determined the forest structure at that northern location, sincr SJ is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history.
+On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands along the elevational gradient where oaks occurr [@Calatrava2019], being firewood, charcoal and acorns continuously exploited through history [@Valbuena2013].
+Charcoal extraction is weel known at least since the 15th century up to middle of the 20th century [@JimenezSerrano2004].
+From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment and use of gas and fosil fuels, similar to other Fagaceae stands across Europe and the Mediterranean [@Valbuena2013].
 
-Those different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5).
-We observed a tree-growth release event at 1940s in the SJ oak woodland (Figures 5, S4) which concurs with one period of maximum mining activity for this area (1925 to 1957) (Table S4).
-During this period there were an increase on the use of timbers for the tunnels of the mines and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area [@Titos1990].
-This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4).
-Our results also coincide with a concurrent but less intense tree-growth release event reported for a closed oak woodland [@GeaIzquierdo2014].
-In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017].
-The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4).
-We revised the forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest], but we did not find any clearing or cutting event during 1995 - 2000 period.
-Two non-mutually exclusive ways could explain the tree-growth release observed.
-Firstly it could be related with a natural drought-induced mortality event after 1995 (*personal communications*) as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004].
-On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014].
+Some of the previous historical records of land-use were expressed in our growth chronologies.
+The release event at 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990].
+This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth the remnant trees included in the northern site (Figure S4).
+In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in the northern site [@GarciaAlix2017].
+The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4).
+There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J.
+Cano-Manuel *personal communications*).
+Therefore, two non-mutually exclusive ways could explain the tree-growth release observed.
+Firstly it could be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
+On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014].
 Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006].
@@
 
-For southern sites, the only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013].
+For southern sites, the only positive release event showed at the earliest yers for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] documented in this site [@Valbuena2013].
 
-Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Navarro2013; @DoblasMiranda2017; @Penuelas2017] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species.
+... revisar 
 Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource.
-This pattern concurs with a forest expansion into marginal abandonded croplands and an increase in the tree-cover [@CamachoOlmedo2002].
+This pattern concurs with a forest expansion into marginal abandoned croplands and an increase in the tree-cover [@CamachoOlmedo2002].
+...
 
-# Conclussions 
+# Conclusions 
 Severe drought events have provoked important reductions in primary and secondary growth of *Q.

### 2019-08-05 man/ecosystems/ms_ecosystems.Rmd
update

@@
 # Conclusions 
-Severe drought events have provoked important reductions in primary and secondary growth of *Q.
-pyrenaica* forests at Sierra Nevada mountain range.
-However, despite the expected vulnerability for those relict stands, we observed positive trends of both primary (*i.e.* greenness) at ecosystem scale and secondary growth at tree level in our study site, particularly at the most humid site along the climatic gradient analyzed.
-Surprisingly we observed high resilience values to drought, particularly at long-term scales.
-These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*.
-The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent.
-This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019].
-The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents.
+Severe drought events provoke important reductions in primary and secondary growth of *Q.
+pyrenaica* forests in the studied relict, rear-edge forest.
+However, we observed no negative growth trends despite we had expected vulnerability to climate change for these relict stands.
+Furthermore, we observed positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a positive steep trend of secondary growth at the tree level at the most humid site along the climatic gradient analyzed.
+The trees exhibited high resilience values in response to drought, particularly in the long-term scales, and resilience was directly related to drought severity.
+These results would be consistent with other research showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*.
+The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent and can vary in very small gradients (*i.e.* following ecological thresholds).
+This is particularly relevant for rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
+The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents.
 In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems.
-This is relevant for tree species with higher sensitivity to climate change, such us *Q.
-pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.
-Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.
+This is relevant for tree species with a high sensitivity to climate change, such us *Q.
+pyrenaica*, not only for conservation per se of the species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.
+Following our study it is necessary to further disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.
 
-# Acknowledgments
+# Acknowledgements
 
@@
 Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work.
+AJPL wants to thank the invaluable support received from his family over the years.
 This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project).

### 2019-08-10 man/figures/browning.Rmd
change legend
fix typo error

@@
 
-**Figure 3.** Percentage of pixels showing browning, greenning or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
+**Figure 3.** Percentage of pixels showing browning, greening or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
+See main text for an explanation of greening and browning.
 

### 2019-08-10 man/figures/bai_cronos.Rmd
include color script

@@
 library("magrittr")
+source(here::here('/script/R/colores_den.R'))
+
 ```
@@
 
+year_init <- 1975
+
 # Lm CA-High
 dfCaH <- cro_sites %>% 
-  filter(year >= 1975) %>% 
+  filter(year >= year_init) %>% 
   filter(site == 'CA-High')
@@
 dfCaL <- cro_sites %>% 
-  filter(year >= 1975) %>% 
+  filter(year >= year_init) %>% 
   filter(site == 'CA-Low')
@@
 dfSJ <- cro_sites %>% 
-  filter(year >= 1975) %>% 
+  filter(year >= year_init) %>% 
   filter(site == 'SJ')
@@
 # colores <- c('#33a02c', '#984ea3', '#a65628')
-colores <- c("#000000", "#0072B2", "#D55E00") 
+# colores <- c("#000000", "#0072B2", "#D55E00") 
+# colores <- c("#B2DF8A","#1F78B4","#A6CEE3")
+# colores <- c("#000000", "#0072B2", "#D55E00")
+
 ```
@@
 ```{r plot_cronobai}
-colores <- c("#B2DF8A","#1F78B4","#A6CEE3")
-colores <- c("#000000", "#0072B2", "#D55E00")
 
@@
   geom_point(size=.8) + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
+  scale_colour_manual(values = colores_den) +
+  scale_fill_manual(values = colores_den) +
   ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
   scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020), sec.axis = dup_axis()) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-High'), 
+  geom_smooth(data = subset(cro_sites, year > year_init & site == 'CA-High'), 
               aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'CA-Low'), 
+  geom_smooth(data = subset(cro_sites, year > year_init & site == 'CA-Low'), 
               aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5) +
-  geom_smooth(data = subset(cro_sites, year > 1975 & site == 'SJ'), 
+  geom_smooth(data = subset(cro_sites, year > year_init & site == 'SJ'), 
               aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5)  +
@@
   geom_line() + 
-  scale_colour_manual(values = colores) +
-  scale_fill_manual(values = colores) +
+  scale_colour_manual(values = colores_den) +
+  scale_fill_manual(values = colores_den) +
   scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020), 
@@
 **Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
-pyrenaica* for northern population (SJ; *orange*) and southern ones: low-elevation (CA-Low; *blue*) and high-elevation (CA-High, *black*) sites.
+pyrenaica* for northern population (SJ; *red*) and southern ones: low-elevation (CA-Low; *green*) and high-elevation (CA-High, *blue*) sites.
 Shading areas correspond to standard error of the mean.
-Number of series are displayed in the upper plot.
+Number of series is displayed in the upper plot.
 We only show years replicated with # series > 5.

### 2019-08-10 script/R/colores_den.R

@@ -1,0 +1,5 @@
+# azul, verde, rojo 
+# CaH, CaL, SJ 
+colores_den <- c("#56B4E9","#009E73","#CD5C5C")
+
+

### 2019-08-12 man/figures/growth_changes.Rmd
update text and figures
cambiar colores figura 5

@@
 Note that y-axes do not correspond in all of the three panels for the sake of clarity.
+Error bars indicate standard error.
 
@@
 
+```{r}
+# azul, verde, rojo 
+# CaH, CaL, SJ 
+colores_den <- c("#56B4E9","#009E73","#CD5C5C")
+```
+
 ```{r plot_caH}
@@
   plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
-         my_ylab = '% GC', micolor = "#838B8B", limitesX = limitesX) + mismo_y +
+         my_ylab = NULL, micolor = colores_den[1], limitesX = limitesX) + mismo_y +
   scale_x_continuous(breaks = xrange, position = 'bottom') + xlab('') + 
@@
   plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year',
-         my_ylab = NULL, micolor = "#838B8B", limitesX = limitesX) +  mismo_y +
-  scale_x_continuous(breaks = xrange, position = 'bottom') + xlab('Year') + 
+         my_ylab = '% GC', micolor = colores_den[2], limitesX = limitesX) +  mismo_y +
+  scale_x_continuous(breaks = xrange, position = 'bottom') +  xlab('') +
   annotate("text", x = 1990, y = 120, label = "CA-Low", fontface = 2, size = 3) 
@@
   plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
-         my_ylab = NULL, micolor = "#838B8B", limitesX = limitesX) + 
-  xlab('') + 
+         my_ylab = NULL, micolor = colores_den[3], limitesX = limitesX) + 
+  xlab('') + xlab('Year') +
   scale_x_continuous(breaks = xrange, position = 'bottom') + 
@@
 ```{r plot_join}
-gg <- psj + pcaH + pcaL + 
+gg <- pcaH + pcaL + psj + 
   plot_layout(ncol=1) & 

### 2019-08-12 man/figures/ring_climate.Rmd
change colors and scale

@@
 library("lemon")
+source(here::here('/script/R/colores_den.R'))
 ```
@@
         legend.box.spacing = unit(0.01, "cm")),
-  scale_fill_manual(values = c("#B2DF8A","#1F78B4","#A6CEE3"), labels = label_site)
+  scale_fill_manual(values = colores_den, labels = label_site)
 ) 
+#c("#B2DF8A","#1F78B4","#A6CEE3")
 ```
@@
   theme_ring_climate + 
-  scale_y_continuous(limits = c(-0.4, 0.4), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4), 
+scale_y_continuous(limits = c(-0.4, 0.8), 
+                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8), 
                      sec.axis = dup_axis(name = NULL)) +
@@
   theme_ring_climate + 
-  scale_y_continuous(limits = c(-0.4, 0.4), 
-                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4), 
+  scale_y_continuous(limits = c(-0.4, 0.8), 
+                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8), 
                      sec.axis = dup_axis(name = NULL)) +
@@
   theme_ring_climate + 
-  scale_y_continuous(limits = c(-0.2, 0.8), 
-                     breaks = c(-0.2, 0, 0.2, 0.4, 0.6, 0.8), 
+  scale_y_continuous(limits = c(-0.4, 0.8), 
+                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8), 
                      sec.axis = dup_axis(name = NULL)) +

### 2019-08-12 man/figures/gc_perc_trees.Rmd
improve figures

@@
 gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')
+
+gc_treesMed50 <- gc_treesMed50 %>% mutate(site_sorted = 
+                                recode_factor(site,
+                                              "gc_caH" = "0_caH", 
+                                              "gc_caL" = "1_caL", "gc_sj" = "2_sj"))
+
+label_site <- c("2_sj" = "SJ",
+                "0_caH" = "CA-High",
+                "1_caL" = "CA-Low")
+
 ```
@@
 xrange <- getBreaks(gc_treesMed50, 'year', by=10)
-
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
 

### 2019-08-12 man/figures/residual_chronos.Rmd
update legend
improve figures

@@
 year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')
+
+# Reorder 
+cro_rwi <- cro_rwi %>% mutate(site_sorted = 
+                                recode_factor(site,
+                                              "caH" = "0_caH", 
+                                              "caL" = "1_caL", "sj" = "2_sj"))
+
+year_cutoffs <- year_cutoffs %>% mutate(site_sorted = 
+                                recode_factor(site,
+                                              "caH" = "0_caH", 
+                                              "caL" = "1_caL", "sj" = "2_sj"))
+
+year_sampledepth <- year_sampledepth %>% mutate(site_sorted = 
+                                   recode_factor(site,
+                                              "caH" = "0_caH", 
+                                              "caL" = "1_caL", "sj" = "2_sj"))
 
@@
 ```{r plot_rwi}
-label_site <- c("0_sj" = "SJ",
-                "1_caH" = "CA-High",
-                "2_caL" = "CA-Low")
 
-cro_rwi <- cro_rwi %>% 
-  mutate(siteF = case_when(
-    site_sorted == "0_sj" ~ "Northern (SJ)",
-    site_sorted == "1_caH" ~ "Southern (Ca-High)",
-    site_sorted == "2_caL" ~ "Southern (Ca-Low)"))
+label_site <- c("2_sj" = "SJ",
+                "0_caH" = "CA-High",
+                "1_caL" = "CA-Low")
 
 # 
-plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res)) +
+plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res, colour=site)) +
   theme_bw() + 
@@
         axis.text = element_text(size=8), 
-        axis.title = element_text(face="bold", size=9)) + 
+        axis.title = element_text(face="bold", size=9), 
+        legend.position = "none") + 
   geom_hline(yintercept = 1, colour = "grey") + 
@@
   scale_x_continuous(limits = c(1819, 2016), 
-                     breaks= seq(1820, 2020, by=20))
+                     breaks= seq(1820, 2020, by=20)) +
+  scale_colour_manual(values = colores_den)
 
@@
 Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the severe drought years identified in our climatic data (see Table S3).
+Dotted black lines showing the severe drought years identified in our climatic data (Table S3 and Figure S1).
 

### 2019-08-13 man/figures/0_figures.Rmd
figure 6 and s8

@@
 
-**Figure 6.** Resilience metrics of the tree-growth for severe drought events since 1950.
+**Figure 6.** Resilience metrics of tree-growth for eigth severe drought events since 1950 (see main text for details) as a function of drought severity.
 *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
-Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*).
+Points indicate resilience metrics for oak populations: SJ (*red*), CA-High (*blue*) and CA-Low (*green*).
 Resilience metrics were computed for each population (sample depth > 10) and drought event.
-Gray line represents overall relationship for each Resilience metrics.
+Gray lines represent overall relationships for each Resilience metrics.
 

### 2019-08-13 man/figures/resilience_long_term.Rmd
figure 6 and s8

@@
 library("cowplot")
+library("purrr")
+library("ggpubr")
+library("rstatix")
+library("ggpmisc")
+source(here::here('/script/R/colores_den.R'))
 ```
@@
 
-```{r plot_functions}
+```{r some_functions}
 #https://gist.github.com/ottadini/6882677 
 lm_eqn = function(m) {
-  # Displays regression line equation and R^2 value on plot
-  # Usage:
-  # p + annotate("text", x=25, y=300, label=lm_eqn(lm(y ~ x, df)), parse=TRUE)
-
-  # l <- list(#a = format(coef(m)[1], digits = 2),
-  #           # b = format(abs(coef(m)[2]), digits = 2),
-  #           rvalue = format(summary(m)$r.squared, digits = 3),
-  #           pvalue = format(summary(m)$coefficients[2,'Pr(>|t|)'],3));
-
-  # eq <- substitute(~~italic(r)^2~"="~r2,l[1], "p=",l[2])
-
   eq2 <- substitute(~~italic(r)^2~"="~rvalue*","~italic(p)~"="~pvalue,  
@@
 
-  # if (coef(m)[2] >= 0)  {
-  #   # eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
-  #   eq <- substitute(~~italic(r)^2~"="~r2,l[1], "p=",l[2])
-  # } else {
-  #   # eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l) 
-  #   eq <- substitute(~~italic(r)^2~"="~r2,l)
-  # }
-
   as.character(as.expression(eq2));                 
@@
 
-plotrel <- function(df, v, x, label, coordX_eq, coordY_eq, titulo, ...){
-  require(ggplot2)
-  require(ggrepel)
+fit_model <- function(x){
+  lm(value ~ d_severity, data = x)
+}
 
-  formulita <- as.formula(paste0(v, "_mean_mean ~", x))
-
-  yerrbar <- aes_string(ymin = paste0(v, "_mean_mean - ",  v, "_mean_se"), 
-                        ymax = paste0(v, "_mean_mean + ",  v, "_mean_se"))
-
-  ggplot(df, aes_string(y=paste0(v, "_mean_mean"), x=x, label=label)) + 
-    geom_smooth(method='lm', se=FALSE, colour="gray") +
-    geom_errorbar(mapping = yerrbar) + 
-    geom_point(size=1.5, pch = 21, colour="black", fill="white") + 
-    theme_bw() + 
-    geom_text_repel(nudge_x = 0.5, size = 3) + 
-    annotate("text", x = coordX_eq, y = coordY_eq, 
-             label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-             size = 3, fontface = "bold", hjust="left") +
-    ggtitle(titulo) + 
-    theme(plot.title = element_text(size = 11, face = "bold.italic"),
-          panel.grid = element_blank(), 
-          axis.text.y = element_text(size=9),
-          axis.text.x = element_text(size=9),
-          axis.title = element_text(size = 10))
-
-  }
+# Get R2 and p.value 
+get_rsq <- function(mod) broom::glance(mod)$r.squared
+get_pvalue <- function(mod) broom::glance(mod)$p.value
 
@@
 
-```{r, eval=FALSE, echo=FALSE}}
-sevRt <- plotrel(sdf, v = "rt", x = 'd_severity', label = 'disturb_year',
-                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rt") +
-  xlab('') + ylab('') +
-  theme(plot.margin = margin(r=-0.7, l=0, t=0, b=0, "pt"))
+**Figure 6.** Resilience metrics of tree-growth for eigth severe drought events since 1950 (see main text for details) as a function of drought severity.
+*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
+Points indicate resilience metrics for oak populations: SJ (*red*), CA-High (*blue*) and CA-Low (*green*).
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
+Gray lines represent overall relationships for each Resilience metrics.
 
-sevRs <- plotrel(sdf, v = "rs", x = 'd_severity', label = 'disturb_year',
-                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rs") + 
-  xlab('') + ylab('') +
-  theme(plot.margin = margin(r=0, l=-0.7, t=0, b=0, "pt"))
+```{r}
+# Include 1995
+dfall <- df %>% 
+  filter(samps >= 10) %>% 
+  group_by(disturb_year, site) %>% 
+  filter(disturb_year > 1940) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear")) %>% 
+  dplyr::select(disturb_year, site, rt_mean_mean, rs_mean_mean, rc_mean_mean, 
+               d_severity) %>% 
+  rename_at(vars(ends_with("_mean")), funs(str_remove(., "_mean_mean"))) %>% 
+  rename("a.rt" = "rt", "b.rc" = "rc", "c.rs" = "rs") %>% 
+  gather("res_index", "value", a.rt, b.rc, c.rs)
 
-sevRc <- plotrel(sdf, v = "rc", x = 'd_severity', label = 'disturb_year',
-                 coordX_eq = 6, coordY_eq = 0.5, titulo = "Rc") +
-  xlab("Drought Severity") + ylab('') +
-  theme(plot.margin = margin(r=-0.7, l=-0.7, t=0, b=0, "pt"))
+p <- ggplot(dfall, aes(x = d_severity, y = value, colour=site, 
+                    fill=site, label = disturb_year)) + 
+  geom_smooth(method='lm', se=FALSE) + 
+  geom_smooth(method='lm', se=FALSE, aes(group=1),  size = 1, colour = "grey") +
+  geom_point(size=1.5, pch = 21) +
+  geom_text_repel(nudge_x = 0.5, size = 2, show.legend = FALSE) +
+  theme_bw() +
+  theme(panel.grid = element_blank(), 
+        axis.text = element_text(size=9),
+        legend.position = "bottom", 
+        strip.background = element_rect(fill = "white"),
+        strip.text = element_text(size = 11, face = "bold.italic")) +
+  scale_colour_manual(values = colores_den) +
+  scale_fill_manual(values = colores_den) +
+  xlab("Drought Severity") + ylab("") + 
+  xlim(5,25) + ylim(0.3,3.8) +
+  facet_wrap(~res_index, labeller = as_labeller(c("a.rt" = "Resistance", 
+                                                  "b.rc" = "Recovery", 
+                                                  "c.rs" = "Resilience"))) +
+  stat_fit_glance(method = "lm",
+                  method.args = list(formula = y ~ x),
+                  aes(label = sprintf('r^2~"="~%.3f~~italic(p)~"="~%.3f',
+                      stat(r.squared), stat(p.value))),
+                  parse = TRUE) +
+  guides(fill = guide_legend(override.aes = list(linetype = 0)),
+        color = guide_legend(override.aes = list(size=3)))
 
-res_long_term <- sevRt + sevRc + sevRs + 
-  plot_layout(nrow = 1) 
+# Add global eq
+dfaux <- dfall %>%
+  group_by(res_index) %>% 
+  nest() %>%
+  mutate(model = map(data, fit_model)) %>% 
+  mutate(eq = map(model, lm_eqn))
+
+res_long_term_sites_all <- p + geom_text(data = dfaux, 
+              aes(x = 15, y = 0.3, label = eq), 
+              inherit.aes = FALSE, parse = TRUE, 
+              col = "gray40")
+
 ```
 
-```{r plot_out, fig.width=6, dpi = 300, out.width = 0.8, eval=FALSE, echo=FALSE}
-res_long_term
+```{r, eval=FALSE, echo=FALSE}
+## Esto no es necesario, es solamente para calcular lm y demás por tibles
+dfpub <- dfall %>% group_by(res_index, site) %>% nest()
+
+dfpub <- dfpub %>% 
+  mutate(model = map(data, fit_model)) %>% 
+  mutate(eq = map(model, lm_eqn))
+
+# mutate(r.squared = map_dbl(model, get_rsq),
+#                         pvalue = map_dbl(model, get_pvalue)) 
+# d <- dfpub %>% unnest(data, .drop = T) 
 ```
 
-**Figure 6.** Resilience metrics of the tree-growth for severe drought events since 1950.
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
-Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*).
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-Gray line represents overall relationship for each Resilience metrics.
+```{r, fig.width=6, dpi = 300, out.width = 0.8}
+res_long_term_sites_all
+```
 
-```{r, echo=FALSE}
+```{r}
+# SAVE as eps 
+ggsave(filename = here::here("./man/figures/svg/fig_6.jpg"), 
+       dpi = 300,  plot=res_long_term_sites_all, width=25, height=14, units = "cm")
+ggsave(filename = here::here("./man/figures/svg/fig_6.eps"), 
+       dpi = 300,  plot=res_long_term_sites_all, width=25, height=14, units = "cm")
+ggsave(filename = here::here("./man/figures/svg/fig_6.svg"), 
+       dpi = 300,  plot=res_long_term_sites_all, width=25, height=14, units = "cm")
+
+```
+
+```{r no95}
+# No 1995
+dfno95 <- df %>% 
+  filter(disturb_year != 1995) %>% 
+  filter(samps >= 10) %>% 
+  group_by(disturb_year, site) %>% 
+  filter(disturb_year > 1940) %>% 
+  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
+               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
+  inner_join(dy, by=c("disturb_year" = "maxyear")) %>% 
+  dplyr::select(disturb_year, site, rt_mean_mean, rs_mean_mean, rc_mean_mean, 
+               d_severity) %>% 
+  rename_at(vars(ends_with("_mean")), funs(str_remove(., "_mean_mean"))) %>% 
+  rename("a.rt" = "rt", "b.rc" = "rc", "c.rs" = "rs") %>% 
+  gather("res_index", "value", a.rt, b.rc, c.rs)
+
+g <- ggplot(dfno95, aes(x = d_severity, y = value, colour=site, 
+                    fill=site, label = disturb_year)) + 
+  geom_smooth(method='lm', se=FALSE) + 
+  geom_smooth(method='lm', se=FALSE, aes(group=1),  size = 1, colour = "grey") +
+  geom_point(size=1.5, pch = 21) +
+  geom_text_repel(nudge_x = 0.5, size = 2, show.legend = FALSE) +
+  theme_bw() +
+  theme(panel.grid = element_blank(), 
+        axis.text = element_text(size=9),
+        legend.position = "bottom", 
+        strip.background = element_rect(fill = "white"),
+        strip.text = element_text(size = 11, face = "bold.italic")) +
+  scale_colour_manual(values = colores_den) +
+  scale_fill_manual(values = colores_den) +
+  xlab("Drought Severity") + ylab("") + 
+  # xlim(5,20) + 
+  ylim(0.3,2) +
+  facet_wrap(~res_index, labeller = as_labeller(c("a.rt" = "Resistance", 
+                                                  "b.rc" = "Recovery", 
+                                                  "c.rs" = "Resilience"))) +
+  stat_fit_glance(method = "lm",
+                  method.args = list(formula = y ~ x),
+                  aes(label = sprintf('r^2~"="~%.3f~~italic(p)~"="~%.3f',
+                      stat(r.squared), stat(p.value))),
+                  parse = TRUE) +
+  guides(fill = guide_legend(override.aes = list(linetype = 0)),
+        color = guide_legend(override.aes = list(size=3)))
+
+# Add global eq
+dfauxno95 <- dfno95 %>%
+  group_by(res_index) %>% 
+  nest() %>%
+  mutate(model = map(data, fit_model)) %>% 
+  mutate(eq = map(model, lm_eqn))
+
+res_long_term_sites_no95 <-  g + geom_text(data = dfauxno95, 
+              aes(x = 12, y = 0.3, label = eq), 
+              inherit.aes = FALSE, parse = TRUE, 
+              col = "gray40")
+```
+
+```{r}
+# SAVE as eps 
+ggsave(filename = here::here("./man/figures/svg/fig_s8.jpg"), 
+       dpi = 300,  plot=res_long_term_sites_no95, width=25, height=14, units = "cm")
+ggsave(filename = here::here("./man/figures/svg/fig_s8.eps"), 
+       dpi = 300,  plot=res_long_term_sites_no95, width=25, height=14, units = "cm")
+ggsave(filename = here::here("./man/figures/svg/fig_s8.svg"), 
+       dpi = 300,  plot=res_long_term_sites_no95, width=25, height=14, units = "cm")
+
+```
+
+```{r, echo=FALSE, eval=FALSE}
+# Analisis global 
 dff <- sdf %>% 
@@
 
-fit_model <- function(x){
-  lm(value ~ d_severity, data = x)
-}
-
 nested_df <- nested_df %>% 
   mutate(model = map(data, fit_model))
-
-# Get R2 and p.value 
-get_rsq <- function(mod) glance(mod)$r.squared
-get_pvalue <- function(mod) glance(mod)$p.value
 
@@
 
-```{r}
+```{r, eval= FALSE, echo = FALSE}
+
 # Plot separately 
@@
              label=lm_eqn(lm(formulita, df)), parse=TRUE, 
-             size = 4, fontface = "bold", hjust="left", col = "gray40") 
+             size = 4, fontface = "bold", hjust="left", col = "gray40") +
+    scale_colour_manual(values = colores_den) +
+    scale_fill_manual(values = colores_den)
+
 }
 
-```
-
-```{r}
 # Include 1995
@@
 sevRt_site <- plotrelsite(sdfsite, v = "rt", x = 'd_severity', label = 'disturb_year', site = "site",
-             titulo = "Rt", coord_eqX = Inf, xLM = -Inf, yLM = 0.5) + xlab('') + ylab('') + xlim(5,25)
+             titulo = "Rt", coord_eqX = Inf, xLM = -Inf, yLM = 0.5) + xlab('') + ylab('') + xlim(5,25) + ylim(0,3.8)
 
 sevRs_site <- plotrelsite(sdfsite, v = "rs", x = 'd_severity', label = 'disturb_year', site = "site",
-                 titulo = "Rs", coord_eqX = -Inf, xLM = 10, yLM = 0.5) + xlab('') + ylab('') + xlim(5,25)
+                 titulo = "Rs", coord_eqX = -Inf, xLM = 10, yLM = 0.5) + xlab('') + ylab('') + xlim(5,25) +  ylim(0,3.8)
 
@@
                  titulo = "Rc",  coord_eqX = -Inf, xLM = 10, yLM = 0.5) +
-  xlab("Drought Severity") + ylab("") + xlim(5,25)
+  xlab("Drought Severity") + ylab("") + xlim(5,25) +  ylim(0,3.8)
 
-```
-
-```{r}
 res_long_term_sites <- sevRt_site + theme(legend.position="none") +
@@
   sevRs_site + theme(legend.position = "none") + plot_layout(nrow = 1)
-
-ggsave(filename = here::here("./man/figures/svg/fig_6_95.jpg"), 
-       dpi = 300,  plot=res_long_term_sites, width=25, height=14, units = "cm")
-ggsave(filename = here::here("./man/figures/svg/fig_6_95.eps"), 
-       dpi = 300,  plot=res_long_term_sites, width=25, height=14, units = "cm")
-ggsave(filename = here::here("./man/figures/svg/fig_6_95.svg"), 
-       dpi = 300,  plot=res_long_term_sites, width=25, height=14, units = "cm")
-
-```
-
-```{r, fig.width=6, dpi = 300, out.width = 0.8}
-res_long_term_sites
-```
-
-```{r no95}
-sdfsite95 <- df %>% 
-  filter(samps >= 10) %>% 
-  group_by(disturb_year, site) %>% 
-  filter(disturb_year > 1940) %>% 
-  filter(disturb_year != 1995) %>% 
-  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
-               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
-  inner_join(dy, by=c("disturb_year" = "maxyear"))
 
@@
        dpi = 300,  plot=res_long_term_sites95, width=25, height=14, units = "cm")
-```
 
-```{r, fig.width=6, dpi = 300, out.width = 0.8}
+# {r, fig.width=6, dpi = 300, out.width = 0.8}
 res_long_term_sites95
-```
 
-# Check lm partial
-
-```{r, eval= FALSE, echo = FALSE}
 library(ggpubr)
@@
 
+# another way
+
+# Create df with lm
+dflm <- df_test %>% 
+  group_by(res_index, site) %>% 
+  do(fit = lm(value ~ d_severity, data =.))
+
+dfeq <- dflm %>% mutate(r.squared = map_dbl(fit, get_rsq),
+         pvalue = map_dbl(fit, get_pvalue))
+
+dfeq <- dflm %>% broom::glance(fit)$r.squared
+
+# function to get expression p.value
+mieq = function(x) {
+      eq <- substitute(~~italic(r)^2~"="~rvalue*","~italic(p)~"="~pvalue,
+           list(rvalue = formatC(x$r.squared, digits = 3, format = "f"),
+                pvalue = formatC(x$p.value, digits = 3, format = "f")))
+
+  as.character(as.expression(eq));
+}
+
+equations <- dfeq %>% group_by(res_index, site) %>% summarise(eq = mieq(.data))
+
+df <- df_test %>% 
+  inner_join(dfeq, by = c("site", "res_index")) %>% 
+  inner_join(equations, by = c("site", "res_index"))
+
+# GGpubr 
+library(ggpubr)
+library(rstatix)
+
+format_pval <- function(pval){
+  pval <- scales::pvalue(pval, accuracy= 0.0001, add_p = TRUE)
+  gsub(pattern = "(=|<)", replacement = " \\1 ", x = pval)
+}
+#
+
+# trailing zeros
+
+ggscatter(df_test, x = "d_severity", y = "value", color="site", 
+          facet.by = "res_index", 
+          panel.labs = list("res_index" =c("Resistance", "Recovery", "Resilience")),
+          add = "reg.line", label = "disturb_year", repel = TRUE, font.label = c(8, "plain")) + 
+    scale_colour_manual(values = colores_den) +
+  scale_fill_manual(values = colores_den) +
+   stat_cor(
+     aes(color = site, 
+         label = #paste(
+           paste("italic(r)^2", round(..estimate..^2, digits=3), sep = "~`=`~"),
+           format_pval(..p..),
+          # paste("italic(p)", round(..p.value.., 3), sep = "~`=`~"),
+           sep = "~`,`~")))
+
 ```

### 2019-08-13 man/figures/bai_cronos.Rmd
modify figures and add eps

@@
 library("magrittr")
-source(here::here('/script/R/colores_den.R'))
+source(here::here('/script/R/custom_functions.R'))
 
@@
 ```
+
+```{r}
+# SAVE as eps
+
+saveImage(plot_bai_s, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_4", u = "cm") 
+```

### 2019-08-13 man/figures/browning.Rmd
modify figures and add eps

@@
 library("ggthemes") # colours 
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 ```
+
+```{r}
+saveImage(per_pixels_green, path="./man/figures/svg/", dpi=300, w=14, h=14, fig= "fig_3", u = "cm") 
+```

### 2019-08-13 man/figures/corr_plot_years.Rmd
modify figures and add eps

@@
 library("ggthemes")
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 ```
+
+```{r}
+# SAVE as eps 
+saveImage(g , path="./man/figures/svg/", dpi=300, w=20, h=10, fig= "fig_s5", u = "cm") 
+```

### 2019-08-13 man/figures/correla_boot_sites.Rmd
update text and figures
modify figures and add eps

@@
 Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016].
+Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot`.
 
@@
 ```
+
+```{r}
+# SAVE as eps 
+saveImage(correla_sitesSJCALH , path="./man/figures/svg/", dpi=300, w=20, h=14, fig= "fig_s6", u = "cm") 
+```

### 2019-08-13 man/figures/cumulative_prec.Rmd
modify figures and add eps

@@
 library("lemon")
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 ```
+
+```{r}
+# SAVE as eps 
+saveImage(p, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_s1", u = "cm") 
+```

### 2019-08-13 man/figures/evi_vs_severe_droughts.Rmd
modify figures and add eps

@@
 library("scales")
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 ```
+
+```{r}
+# SAVE as eps 
+saveImage(combinado2, path="./man/figures/svg/", dpi=300, w=20, h=14, fig= "fig_s7", u = "cm") 
+```

### 2019-08-13 man/figures/gc_perc_trees.Rmd
modify figures and add eps

@@
 ```
+
+```{r}
+# SAVE as eps 
+saveImage(pgc_median50 , path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_s4", u = "cm") 
+```

### 2019-08-13 man/figures/growth_changes.Rmd
modify figures and add eps

@@
 library("patchwork")
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
   p <- ggplot(df, aes_string(x = x, y = y)) +
-    geom_errorbar(mapping = yerrorbar, colour = 'gray', width = .35, size = .2, 
+    geom_errorbar(mapping = yerrorbar, colour = micolor, width = .35, size = .2, 
                   na.rm = TRUE) + 
@@
                             breaks = c(-100, -50, 0, 50, 50, 100, 150))
-```
-
-```{r}
-# azul, verde, rojo 
-# CaH, CaL, SJ 
-colores_den <- c("#56B4E9","#009E73","#CD5C5C")
 ```
@@
 ```
+
+```{r}
+saveImage(gg, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_5", u = "cm") 
+```

### 2019-08-13 man/figures/plot_spei12.Rmd
update text and figures
modify figures and add eps

@@
 library("here")
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 **Figure S2**.
-Drought severity in the Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
 Data from Global SPEI database (http://spei.csic.es/database.html).
@@
 ```
+
+```{r}
+# SAVE as eps 
+saveImage(s, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_s2", u = "cm") 
+```

### 2019-08-13 man/figures/residual_chronos.Rmd
modify figures and add eps

@@
 library("lemon")
-source(here::here('/script/R/colores_den.R'))
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 ```
+
+```{r}
+# SAVE as eps 
+saveImage(plot_rwi_crono , path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_s3", u = "cm") 
+```

### 2019-08-13 man/figures/resilience_long_term.Rmd
modify figures and add eps

@@
 library("ggpmisc")
-source(here::here('/script/R/colores_den.R'))
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 # SAVE as eps 
-ggsave(filename = here::here("./man/figures/svg/fig_6.jpg"), 
-       dpi = 300,  plot=res_long_term_sites_all, width=25, height=14, units = "cm")
-ggsave(filename = here::here("./man/figures/svg/fig_6.eps"), 
-       dpi = 300,  plot=res_long_term_sites_all, width=25, height=14, units = "cm")
-ggsave(filename = here::here("./man/figures/svg/fig_6.svg"), 
-       dpi = 300,  plot=res_long_term_sites_all, width=25, height=14, units = "cm")
-
+saveImage(res_long_term_sites_all, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_6", u = "cm") 
 ```
@@
 # SAVE as eps 
-ggsave(filename = here::here("./man/figures/svg/fig_s8.jpg"), 
-       dpi = 300,  plot=res_long_term_sites_no95, width=25, height=14, units = "cm")
-ggsave(filename = here::here("./man/figures/svg/fig_s8.eps"), 
-       dpi = 300,  plot=res_long_term_sites_no95, width=25, height=14, units = "cm")
-ggsave(filename = here::here("./man/figures/svg/fig_s8.svg"), 
-       dpi = 300,  plot=res_long_term_sites_no95, width=25, height=14, units = "cm")
+saveImage(res_long_term_sites_no95, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_s8", u = "cm")
 

### 2019-08-13 man/figures/ring_climate.Rmd
modify figures and add eps

@@
 ```
+
+```{r}
+# SAVE as eps 
+saveImage(ring_climate, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_8", u = "cm") 
+```

### 2019-08-13 man/figures/std_anomalies.Rmd
modify figures and add eps

@@
 
+```{r}
+saveImage(plot_sa_clu, path="./man/figures/svg/", dpi=300, w=14, h=14, fig= "fig_2", u = "cm") 
+```
+
 ```{r oldFigure, eval=FALSE, fig.height=7}

### 2019-08-13 script/R/custom_functions.R
include saveImage fx

@@ -1,0 +1,20 @@
+
+
+# SAVE IMAGE 
+# Function to save images as eps, jpg,  svb
+saveImage <- function(p, path, fig, dpi, w, h, u, ...){ 
+  ggsave(filename = here::here(path, paste0(fig, ".jpg")), 
+         dpi = dpi, plot = p, width = w, height = h, units = u)
+  ggsave(filename = here::here(path, paste0(fig, ".eps")), 
+         dpi = dpi, plot = p, width = w, height = h, units = u)
+  ggsave(filename = here::here(path, paste0(fig, ".svg")), 
+         dpi = dpi, plot = p, width = w, height = h, units = u)
+}
+
+
+# COLORES 
+# azul, verde, rojo 
+# CaH, CaL, SJ 
+colores_den <- c("#56B4E9","#009E73","#CD5C5C")
+
+

### 2019-08-14 man/ecosystems/ms_ecosystems.Rmd
update new version

@@
 ## Authors [^\*^] 
-A.J. Perez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^
+A.J. Pérez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^
 
@@
 AJPL, GGI and RZ conceived the study, conducted field work and collected the data.
-AJPL and GGI conducted lab work and led article preparation.
+AJPL and GGI conducted lab work.
 AJPL analyzed data and led the writing of the paper.
@@
 
-* The response to climate and resilience followed a water-stress gradient.
-Primary and secondary growth of more xeric populations (*i.e.* Northern aspect and low elevations) was less resilient to drought than that at more mesic sites.
+* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure.
+Trees were highly sensitivity to drought and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term.
+The response to climate and resilience followed a water-stress gradient.
+Primary and secondary growth of more xeric populations (*i.e.* northern aspect) was less resilient to drought than that at more mesic sites.
 In addition, a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic high-elevation stands, but absent in the most xeric site.
-Trees were highly sensitivity to drought and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term.
-The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure.
 
@@
 Stand dynamics at the most xeric locations (*i.e.* the altitudinal rear-edge) needs to be monitored under future expected climatic scenarios bearing in mind that land-use legacies play a key role in stand dynamics.
-
-* Despite being considered relict stands at the dry rear-edge for the species, the trees showed high resilience in response to drought.
-Yet, there were a divergent response to a changing climate along very short climatic gradients.
-The oaks were more sensitive to water availability expressed at the most xeric sites whereas they exhibited positive growth trends only in the coldest and more humid stands.
-These results suggest that stand dynamics, particularly at the lowest elevations (*i.e.* the altitudinal rear-edge), need to be monitored under future expected drier and warmer climatic scenarios bearing in mind that management legacies likely play a key role in current stand dynamics.
 
@@
 The Iberian Peninsula suffered several extreme drought episodes in the last decades of the twentieth century (*e.g*.
-1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
+1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
 We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
@@
 # Discussion
+## Land-use legacies shape forests and sensitivity to climate change at the rear-edge
+
+The Mediterranean landscapes, even at high elevations, have suffered a strong transformation driven by human activities throughout history [@GarciaAlix2017].
+Forest clearings, firewood extraction, charcoal exploitations and mining (these were strongly suggested in the disturbance chronology in SJ), have strongly impacted on the forest resources at Sierra Nevada as can be derived from the results of the review of historical documents (Table S4).
+Indeed a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation has been documented in this mountain region [@JimenezOlivencia2015].
+Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland) [@JimenezOlivencia2015].
+It has been described a sequentially distribution of the land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops [@JimenezOlivencia2015].
+Other activities (like mining) should have determined the forest structure at that northern location, since SJ is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history.
+On the other hand, woodland areas of the southern slopes (*e.g.* CA site) are mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015].
+Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013; @JimenezOlivencia2015].
+
+Some of the previous historical records of land-use were expressed in our growth chronologies.
+The release event at 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990].
+This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth the remnant trees included in the northern site (Figure S4).
+The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4).
+There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J.
+Cano-Manuel *personal communications*).
+We speculate that this might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
+On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014].
+For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] documented in this site [@Valbuena2013].
+
+As we inferred from our results, the management history can conditioned tree-growth and resilience of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017].
+Nevertheless, we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource.
+This pattern concurs with a forest expansion into marginal abandoned croplands and an increase in the tree-cover [@CamachoOlmedo2002].
+
+```{r eval=FALSE, echo=FALSE}
+~~Although the estimated age for our sampled trees was lower than 180 years, several documents reported abundant oak presence before that date.
+For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991].
+These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
+@GeaIzquierdo2014; @GeaIzquierdo2015EJFR; @Dorado2017].~~
+```
+
 ## Relict oaks show high resilience to recent drought events and long-term climatic variability
@@
 Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
-Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g.* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
-pyrenaica* for the last 16 years (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period).
-This confirms previous findings that pointed out a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi].
+Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g.* 2005 and 2012 were considered two of the worst recorded drought; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
+pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
 For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4).
 Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014].
-However, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
+However, the observed pattern for our sites (a rear-edge) differs from the declining trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
 sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].
@@
 Thus, this highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], as also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].
-In addition, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species [@SanchezSalguero2012; @Camarero2015].
-But, interestingly, the highest values of resilience for tree-growth were obtained for these two drought events.
+For tree-growth, the highest values of resilience were obtained for the most severe events, namely 1995 and 1999 (Table S3).
 In addition, we observed a positive relation between the tree-growth resilience and drought severity (Figures 6).
-This result seems contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events.
-Yet, the high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J.
+This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events.
+Yet, the high resilience values obtained for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J.
 Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of individual resilience estimated from individual growth series.
@@
 And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges.
+Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* Pinus sylvestris, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.
 
@@
 These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018].
-Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based (*i.e.* ecosystem or stand based); the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018].
+Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based (*i.e.* ecosystem or stand based); the growth reduction seems to be more mediated by sink that by source limitations [@Korner2013].
 The response of greenness to drought not only reflects the response of trees but from all vegetation within the pixel [@Dionisioetal2012IPoGEC].
@@
 It has been described an effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive, thus high-elevations can be used as local refuge for species in response to climate change [@Bhuyan2017].
-It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
+It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
 Remarkably, the trees at CA-High, which are located around 1900 m.
@@
 Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014].
-
 Finally and in a context of global change, in addition to the withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).
@@
 ```
-
-## Land-use legacies shape forests and sensitivity to climate change at the rear-edge
-
-The Mediterranean landscapes, even at high elevations, have suffered a strong transformation driven by human activities throughout history.
-Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017].
-From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with a sharp intensification of the human activities in the last 150 years [@GarciaAlix2017].
-Mining, forest clearings, firewood extraction and charcoal exploitations (these were strongly suggested in the disturbance chronology in SJ), pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada, and on Pyrenean oak forest, as can be derived from the results of the review of historical documents.
-As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015].
-Therefore, our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation.
-Although the estimated age for our sampled trees was lower than 180 years, several documents reported abundant oak presence before that date.
-For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991].
-These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
-@GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].
-
-Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between Northern (coppice) and Southern populations (high forest, open woodland) [@JimenezOlivencia2015].
-It has been described a sequentially distribution of the land uses along the elevational gradient in the northern site (*e.g.* San Juan): grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops [@JimenezOlivencia2015].
-Other activities (like mining) should have determined the forest structure at that northern location, sincr SJ is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history.
-On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands along the elevational gradient where oaks occurr [@Calatrava2019], being firewood, charcoal and acorns continuously exploited through history [@Valbuena2013].
-Charcoal extraction is weel known at least since the 15th century up to middle of the 20th century [@JimenezSerrano2004].
-From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment and use of gas and fosil fuels, similar to other Fagaceae stands across Europe and the Mediterranean [@Valbuena2013].
-
-Some of the previous historical records of land-use were expressed in our growth chronologies.
-The release event at 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990].
-This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth the remnant trees included in the northern site (Figure S4).
-In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in the northern site [@GarciaAlix2017].
-The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4).
-There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J.
-Cano-Manuel *personal communications*).
-Therefore, two non-mutually exclusive ways could explain the tree-growth release observed.
-Firstly it could be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
-On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014].
-Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006].
-Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].
-
-For southern sites, the only positive release event showed at the earliest yers for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] documented in this site [@Valbuena2013].
-
-... revisar 
-Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource.
-This pattern concurs with a forest expansion into marginal abandoned croplands and an increase in the tree-cover [@CamachoOlmedo2002].
-...
 

### 2019-08-14 man/figures/resilience_metrics.Rmd
update new version

@@
 library("bookdown")
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 ```{r data}
-mhuber <- read.csv(here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE)
-mhuberbai <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
+huber_evi <- 
+  read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE) %>% 
+  filter(var != "rrs") %>% 
+  mutate(site_sorted = as.factor(
+  case_when(
+  site == "N" ~ "0_N",
+  site == "S" ~ "1_S"
+)))
+
+huber_b_evi <- 
+  read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_b.csv"), header = TRUE) %>%
+  filter(var != "rrs") %>% 
+  mutate(var_sorted = as.factor(
+    case_when(
+      var == "rt" ~ "0_rt",
+      var == "rc" ~ "1_rc",
+      var == "rs" ~ "2_rs",
+))) %>% 
+mutate(site_sorted = as.factor(
+  case_when(
+  site == "N" ~ "0_N",
+  site == "S" ~ "1_S"
+)))
+
+huber_bai <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE) %>%
+  filter(var != "rrs") %>% 
+  mutate(site_sorted = as.factor(
+  case_when(
+  site == "SJ" ~ "0_SJ",
+  site == "caH" ~ "2_CA-High",
+  site == "caL" ~ "1_CA-Low"
+)))
+
+huber_b_bai <- 
+  read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber_b.csv"), header = TRUE) %>%
+  filter(var != "rrs") %>% 
+  mutate(var_sorted = as.factor(
+    case_when(
+      var == "rt" ~ "0_rt",
+      var == "rc" ~ "1_rc",
+      var == "rs" ~ "2_rs",
+))) %>% 
+  mutate(site_sorted = as.factor(
+  case_when(
+  site == "SJ" ~ "0_SJ",
+  site == "caH" ~ "2_CA-High",
+  site == "caL" ~ "1_CA-Low"
+)))
+```
+
+```{r}
+# plot_params 
+pd <- position_dodge(width = 1)
+aes_main <- aes(x=site_sorted, y = M.Huber, 
+                ymin = lower.ci, ymax = upper.ci,
+                fill = site_sorted, colour = site_sorted)
+aes_aux <- aes(shape = as.factor(disturb_year), 
+               label = Letter)
+labelita <- as_labeller(c("0_rt" = "Resistance",
+                                      "1_rc" = "Recovery", 
+                                      "2_rs" = "Resilience"))
+
+item_NS <- c("0_N" = "N", "1_S" = "S")
+item_site <- c("0_SJ" = "SJ", "2_CA-High" = "CA-High", "1_CA-Low" = "CA-Low")
+
+theme_res <- list(
+  geom_errorbar(width=.1, position = pd),
+  geom_point(size=7, shape=21),
+  geom_text(aes(label = str_to_upper(Letter)), fontface = "bold", colour = "white"), 
+  theme_bw(), xlab(""), ylab (""), 
+  theme(panel.grid = element_blank(), 
+        axis.text = element_text(size=9),
+        legend.title = element_blank(),
+        strip.background = element_rect(fill = "white"),
+        strip.text = element_text(size = 11, face = "bold.italic"), 
+        plot.title = element_text(size = 12, face = "bold")),
+  scale_shape_manual(values = c(24,22)),
+  facet_wrap(~var_sorted, scales = "free_y", labeller = labelita))
+
+# PLOT EVI 
+pevi <- ggplot(huber_b_evi, aes_main) + 
+  # by disturb year 
+  geom_errorbar(data=huber_evi, aes_aux, width=.1, position = pd) +
+  geom_point(data=huber_evi, aes_aux, position=pd, size = 4.5, fill="white") + 
+  geom_text(data = huber_evi, aes_aux, position = pd, size = 2.5) +
+  scale_x_discrete(labels = item_NS) +
+  scale_fill_manual(values = colores_den[3:2], labels = item_NS) +
+  scale_colour_manual(values = colores_den[3:2], labels = item_NS) + theme_res +
+  theme(legend.position = "top") + ggtitle ("EVI")
+
+# PLOT BAI
+pbai <- ggplot(huber_b_bai, aes_main) + 
+  # by disturb year 
+  geom_errorbar(data=huber_bai, aes_aux, width=.1, position = pd) +
+  geom_point(data=huber_bai, aes_aux, position=pd, size = 4.5, fill="white") + 
+  geom_text(data = huber_bai, aes_aux, position = pd, size = 2.5) +
+  scale_x_discrete(labels = item_site) + 
+  scale_fill_manual(values = rev(colores_den), labels = item_site) +
+  scale_colour_manual(values = rev(colores_den), labels = item_site) + 
+  theme_res + theme(legend.position = "bottom") + 
+  ggtitle("BAI")
+
+plot_res <- pevi + pbai + plot_layout(ncol=1, tag_level = "keep") + plot_annotation(tag_levels = "a") & theme(plot.tag = element_text(size = 12, face = 2)) & theme(plot.margin = margin(t=0,r=0,b=-2,l=0, "pt"))
+```
+
+**Figure 7.** Comparison of the response of  *Q.
+pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (a) and tree-growth (b).
+For EVI we compared northern populations (*red circle*) with southern ones (*green circle*).
+For BAI we compared the more xeric northern population (San Juan, SJ; *red circle*) with the two southerns populations: Cáñar-High (CA-High; *blue circle*) and Cáñar-Low (CA-Low; *green circle*).
+Different letters indicate significant *post hoc* differences between groups (see material and methods for details).
+
+```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
+plot_res 
+```
+
+```{r}
+# SAVE as eps 
+saveImage(plot_res, path="./man/figures/svg/", dpi=300, w=22, h=25, fig= "fig_7", u = "cm") 
+```
+
+```{r, eval=FALSE, echo = FALSE}
+# old plots
+
+mhuberbai <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)
 
@@
 m <- bind_rows(aux2, aux)
-```
 
-```{r plot}
 pd <- position_dodge(width = 0.6)
@@
 # + geom_text(aes(label=Letter, y = 1.2), position=pd, show.legend = FALSE)
-```
 
-**Figure 7.** Response *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (EVI; left-plots) and tree growth (BAI; right-plots) for the years 2005 and 2012.
-For EVI we compared northern populations (*black fill circle*) with southern ones (*blue empty circle*).
-For BAI we compared the northern population (San Juan, SJ; *black triangle*) with the two southerns populations: Cáñar-High (CA-High; *blue empty squares*) and Cáñar-Low (CA-Low; *blue fill squares*).
-Different letters above error bars indicate significant *post hoc* differences between groups (see material and methods).
+# ```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
+# grid_arrange_shared_legend(rt, rc, rs, ncol = 1, nrow = 3, position ='top')
 
-```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
-grid_arrange_shared_legend(rt, rc, rs, ncol = 1, nrow = 3, position ='top')
+# ```
 

### 2019-08-14 man/old_ms/ms.Rmd
update new version

--- 2019-05-21 man/ms.Rmd
@@
 bibliography: ../refs/references.bib
-csl: ../refs/ecology.csl
+csl: ../refs/ecosystems.csl
 ---

### 2019-08-14 man/old_ms/ms_0_intro.Rmd
update new version

--- 2018-03-14 man/ms_0_intro.Rmd
@@
 It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].
-Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], however only a few works have focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
+
+Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [@Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018], some focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014].
 Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Lloret2011; @Hodgson2015].
@@
 
+Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and tree-ring [@Jump2010].
+
 Remote sensing can be used for studying droughts impacts on ecosystems [@Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016].
 Vegetation indices obtained from satellite, such us EVI (*Enhanced Vegetation Index*) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016].
-Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Gouveia2012; @Zribi2016].
+Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Assal2016] and for Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [@Lloret2007; @VicenteSerrano2007; @Zribi2016].
 
@@
 
-In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica*) located in the rear-edge of their distribution.
-
-Relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge can be particularly vulnerable to climate change.
-In this study we combined remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and radial growth of a Mediterranean tree species (*Quercus pyrenaica*) forests in Sierra Nevada, which represent a rear-edge of their distribution.
-Wealso evaluated the resilience of these forest to several extreme drought episodes.
+In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, considered a rear-edge of their distribution.
+We are also interested in assessing resilience of these forest to several extreme drought episodes.
+We hypothesized that, relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts.
 Specifically, the aims of this work were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q.
@@
 
-## 
 ```{r echo=FALSE, results='asis', eval=show_text}

### 2019-08-14 man/old_ms/ms_3_discussion.Rmd
update new version

--- 2018-09-18 man/ms_3_discussion.Rmd
@@
 # Discussion
-
-## Oaks show high resilience in response to recent drought events
+## Impacts of severe droughts on rear-edge populations
 
@@
 
+## Oaks show high resilience in response to recent drought events
 Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Appendix S3), we found a positive trend for vegetation greenness of the forests of *Q.
 pyrenaica* located in their rear edge.
-Our result agrees with those obtained by previous works using other remote sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi], which suggests an increase in primary productivity during the last years for rear-edge populations of this species.
+Our results agree with those obtained by previous works using other remote-sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi], which suggests an increase in primary productivity during the last years for rear-edge populations of this species.
 
@@
 However, our results for similar locations have revealed a recovery in tree-growth for the last years (Figure 4).
+In addition, a similar positive trend in BAI for the last years, has been reported for another *Q.
+pyrenaica* population of the Sierra Nevada [@RubioCuadrado2018].
 
@@
 pyrenaica*.
-Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía, observamos como estas poblaciones muestran altos valores de resiliencia.
-Como muestrasn nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).
+Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía (long-term perspective), observamos como estas poblaciones presentan altos valores de resiliencia.
+Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Appendix S5).
 
@@
 pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático.
-Esta idea se vería reforzada por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
+Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q.
 pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017].
@@
 For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2).
-In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for SJ site (Figure 6), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
+In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 6), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014].
 
@@
 The trees of CA-High site, which are located around 1900 m.
-a.s.l.
-and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4).
+*a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4).
 This is especially interesting for southern sites, which are very close to each other.
@@
 
-***Reescribir esto del impacto antrópico***
+## Land-use legacy effects shape distribution and sensitivity to climate change of read-edge oak populations
 
-Finally another idea to consider is the different degree of anthropic impact related to elevation.
+Although our sampled sited showed similar tree competence levels (Table 1), the review of historical documents showed that la historia de manejo humano ha sido diferente para cada sitio.
+Different values of tree-growth were observed between populations and for different elevations within southern location (Figure 4).
+Low-elevation sites showed lower BAI values than higher-elevation site.
+A reason that could explain these results is the different degree of anthropic impact related to elevation.
+Los bosques situados a mas elevación presentan menor presión antrópica que los localizados a menor altitud, por el simple hecho de que estos últimos están mas accesibles.
+During the last decades, the rural abandonment has lead an decrease of anthropic pressures on the high-elevations zones of Mediterranean mountains (INLCUIR CITA).
 
-Aunque nuestros sitios presenten valores similares de competencia, la historia de manejo humano puede ser diferente, y en parte se relaciona con la altitud.
-Así en zonas altas encontramos un menor impacto humano que en las zonas bajas.
-Esto ocurre
+comentario --> Aqui pretendo incluir la idea que hablamos en su momento.
+Quizá están creciendo donde les están dejando crecer (menor impacto antrópico en las zonas mas altas, al menos en los últimos años -abandono de zonas rurales?-)
+He ralizado una revisión de trabajos que utilicen dendro de Q. pyrenaica para ver si existe una relación entre la edad de las cronos con la elevación.
 
-Another reason that could explain the higher values of BAI of CA-High site is el grado diferente de impacto antrópico.
-En nuestros resultados vemos como los bosques situados a mayor altitud presentan un menor impacto antrópico.
-Esto puede ser simplemente por el hecho de que estos bosques estén menos accesibles ---> ojo y la idea de que son bosque relativamente jóvenes
+* Son bosques relativamente jóvenes 
+comentaio --> 
+Aquí me falta desarrollar este apartado algo más incluyendo cosas de los documentos históricos (la minería, etc, etc) 
+Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5).
+La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona.
+Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.
 
-* ---> Gráfica. Our sampled sited showed similar tree competence levels (Table 1), but ... impacto antrópico diferente
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 
-* Quizá están creciendo donde les están dejando (menor impacto antrópico en las zonas mas altas, al menos en los últimos años??)
+#### Disturbance
+Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
 
-The presence of *Q. pyrenaica* forests on the southern slopes of Sierra Nevada is partly explained by the additional contribution of humid air from the Mediterranean sea [@Prieto1975]
+## historia forestal de ambos sitios 
+Incluimos lo que conocemos de la historia forestal de los sitios??
+Me explico, tenemos datos de manejo y uso antrópico de las dos zonas que proceden de varias fuentes.
+En resumen, mas o menos, tenemos:
 
+* Minería: datos de minería, que afectan sobre todo a la población de SJ (los tenemos localizados temporalmente)
+* Actuaciones forestales: 
+* Incendios, Carboneo, Ganadería, etc --> Tenemos información menos estructurada sobre estos ámbitos, procedentes de varias fuentes, que tienen una incertidumbre mayor espacial y temporalmente. 
+")
+
+```
+
+Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution.
+Para  *Q.
+pyrenaica* se ha visto que moisture availability es el factor limitante que determina el crecimiento en las poblaciones situadas en su rear edge [@GeaIzquierdo2014], pero para otras especies (*i.e.* *Pinus nigra*, *P.
+sylvestris*) la temperatura tiene mas peso que la disponibilidad de agua  [@Herrero2013; @Matias2017].
+Pero es importante considerar además la historia de manejo que han tenido esos bosques a la hora de la forest management [@Penuelas2017; @DoblasMiranda2017], ya que, como observamos en nuestros resultados puede condicionar el crecimiento y la resiliencia de las especies.
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+Podríamos complementarlo con lo que le pasa a otras especies en su borde de distribución: por ejemplo en Baza, Herrero et al.
+2013, encontraron para Pinus nigra y sylvestris que la temperatura tenía mas peso que la disponibilidad de agua).
+O también ver algunos de los trabajos de Camarero et al 2013 para el P.
+nigra en su borde de distribución u otros similares (el de Sanchez-Salguero et al.
+2013, 2015)
+
+Para *Q.
+pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q.
+pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al.
+2015 European Journal of Forest Research).
+Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea.
+
+")
+
+```
 ```{r echo=FALSE, results='asis', eval=show_text}
@@
 ```
-
 ```{r echo=FALSE, results='asis', eval=show_text}
@@
 ```
-
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 ### OTRA COSA
@@
 Environmental and forest management policies should take into account all these characteristics of Mediterranean forests and the social, environmental and climatic conditions that are projected for the coming years and decades.
-
+")
+```
 ```{r echo=FALSE, results='asis', eval=show_text}
@@
 ```
-
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
 ### ¿Que factor es mas limitante para el crecimiento en el rear-edge de Q. pyrenaica? 
@@
 2013, 2015) …
-
-#### Disturbance
-Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??).
-
-## historia forestal de ambos sitios 
-Incluimos lo que conocemos de la historia forestal de los sitios??
-Me explico, tenemos datos de manejo y uso antrópico de las dos zonas que proceden de varias fuentes.
-En resumen, mas o menos, tenemos:
-
-* Minería: datos de minería, que afectan sobre todo a la población de SJ (los tenemos localizados temporalmente)
-* Actuaciones forestales: Tenemos una bd con actuaciones forestales, y he contactado con varios de los responsables de proyectos de actuaciones forestales en ambas zonas, y tenemos con bastante detalle la información sobre actuaciones forestales (al menos espacial y temporalmente)
-* Incendios, Carboneo, Ganadería, etc --> Tenemos información menos estructurada sobre estos ámbitos, procedentes de varias fuentes, que tienen una incertidumbre mayor espacial y temporalmente.
-
-Todo esto lo comento, porque quizá podemos utilizar dicha información para la discussión, o a lo mejor mete mas ruido (¿que opináis?)
-
-Por otro lado, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia.
-Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética).
 
@@
 
-### Esto viene del apartado Radial growth trends and climate
-* Aqui me comentó GGI que distribuyera esto bien entre results y discussion 
-Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5).
-La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona.
-Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.
+The presence of *Q. pyrenaica* forests on the southern slopes of Sierra Nevada is partly explained by the additional contribution of humid air from the Mediterranean sea [@Prieto1975]
 
-### Otras
+")
+```
+
+En este sentido, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia.
+Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética).
+
+```{r echo=FALSE, results='asis', eval=show_text}
+cat("
+
+### ORDENAR ???
 
@@
 
----------------------
+### Estudios que utilizan RS y TR
+
+* Wu et al. 2017. 
+	* En un estudio sobre el efecto de la sequía sobre el crecimiento en bosques, grass y shrub del hemisferio norte, encontraron que los bosques (Deep-rooted forest) exhibit a drought legacy response with reduced growth during up to 4 years despues de un evento extremo de sequía.
+
+	* Además vieron que los bosques mostraron una stronger drought resilience in forest (i.e. smaller growth reduction after severe drought) (Gazol el al. 2017; Wu et al. 2017).
+
+	* Esto se puede deber a que acceden a reservas de agua mas profundas (los pastos y los matorrales no pueden acceder a ese agua)
+
+### 
+Nosotros hemos obtenido una relación entre la severidad de la sequía y la resiliencia mostrada por los robledales en su rear edge. (ver gráfica )
+
+@Gazol2018 encontró La resiliencia está relacionada con la severidad de la sequía y con la composición del bosque
+
+Rt-NDVI and Rt-TR were positively related with drought intensity
+
+## 
+@Gazol2018 Resistencia a la sequía mayor en la zona norte de españa que en las zonas secas del sur
+
+Qpyr 
+* Valores de Rt-TR aumentaron a lo largo del tiempo, sin embargo Rt-NDVI dismiuyó con el tiempo
+* Valores Rc-NDVI tendencia positiva
+
+")
+```
+
 Methodological approach
+
+comentario: tengo que rematar este apartado.
 
@@
 @Zhang2013; @AghaKouchak2015] and of the dendrochronology [*e.g.* @Eilmann2012; @Bhuyan2017] to quantify the drought impacts on vegetation at different spatial and temporal scales, but to our knowledge the combination of both approaches is scarce [@VicenteSerrano2013; @VicenteSerrano2016; @Bhuyan2017b; @Wu2018].
-Our approach combining both methodologies to explore the resilience metrics have
+Estos trabajos han observado que existe de forma general a positive relationship between vegetation indices derived from remote sensing and annual tree growth [@VicenteSerrano2016; @Bhuyan2017b; @Gazol2018].
 
-The combination of these two approach to explore the resilience of forests to drought have been recently used in a study across Iberian Peninsulsa, but
+Aunque existen algunos trabajos que han utilizado RWI y remote sensing nuestra aproximación (creo) es interesante, sobre todo porque la aplicamos al estudio de poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
+Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares que analicen la respuesta de las poblaciones al cambio climático y su resiliencia utilizando diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].
 
-Recently, Gazol et al. 2018 combine these two approaches to explore resilience of several iberian tree species, and encontraron ...
-
-La vegetación es un reflejo de las condiciones ambientales, y los efectos de la sequía en la vegetación se pueden ver tanto con información de satélite (hablar del browning), como con información mas de campo (hablar aquí de lo del BAI).
-
-La combinación de metodologías que hemos empleado para evaluar la resiliencia de la vegetación frente a los eventos de sequía
-
-Se trata de expresar la idea de mezcla de aproximacioes: remote-sensing con dendro.
-Aunque existen algunos trabajos que utilizado RWI y remote sensing nuestra aproximación es importante, sobre todo para estudiar poblaciones que están en su límite de distribución (rear-edge) [@Jump2010].
-De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
-
-Exiten trabajos que han utilizado remote-sensing o la dendrochronology para analizar los efectos de las sequías sobre la vegetación a diferentes escalas espaciales y temporales ( ).
-Asimismo, algunos otros trabajos han utilizado conjuntamente estas dos metodologias, y han visto que existe de forma general una relación
-a positive relationship between vegetation indices derived from remote sensing and annual tree growth [@VicenteSerrano2016; @Bhuyan2017b],
-
-In general, a positive relationship between NDVI in the growing season and RWI was seen for many forests in our study.
-
-Positive relationship between inter-annual NDVI variability and annual tree growth [@VicenteSerrano2016]
-
-Esta relación positiva también la encuentra
-
-* Diferencias que hemos detectado usando RS y RWI (en resiliencias)
-* Relacionar con datos de Gazol et al. 2018
-
-Sin embargo el estudio de las poblaciones situadas en su rear edge requiere de una aproximación combinada donde se estudie la respuesta de las poblaciones al cambio climático y su resiliencia utiliznado diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].
-
-## De forma genérica y para el rear-edge 
-* Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares...
-
-* Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q.
+- Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q.
 pyrenaica situadas en su reaar edge frente a los eventos de sequía.
 
-Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ...
-(ver esto y completar)
-
-## Sensibilidades diferentes de cada metodología 
-* Cada metodología muestra distinta sensibilidad
-    * @Gazol2018 analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR.
+- Hemos encontrado diferencias en cuanto a la resiliencia utilizando RS y RWI.
+Diferente sensibilidad de satélite frente a los datos de campo.
+Esto también se ha observado en @Gazol2018, que analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR.
 Encontraron que los datos de TR son mas sensibles para la resiliencia del bosque a la sequía que los datos de RS
 
-* La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017]
+- De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.
 
-end VALID
+- La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017] (...
+Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes.
+)
 
-* El uso de análisis combinados además nos permite distinguir la varibilidad en el crecimiento del bosque inducida por clima de aquella impulsada por otros procesos ecológicos a nivel de comunidad [@Babst2017].
-Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes.
-
- Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability.
+- Nota from @Jump2010: Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability.
 In such regions, targeted collection of high-resolution data is necessary in order to identify currently occurring changes and predict the magnitude and spatial distribution of future decline.
@@
 
-### Estudios que utilizan RS y TR
-
-* Wu et al. 2017. 
-	* En un estudio sobre el efecto de la sequía sobre el crecimiento en bosques, grass y shrub del hemisferio norte, encontraron que los bosques (Deep-rooted forest) exhibit a drought legacy response with reduced growth during up to 4 years despues de un evento extremo de sequía.
-
-	* Además vieron que los bosques mostraron una stronger drought resilience in forest (i.e. smaller growth reduction after severe drought) (Gazol el al. 2017; Wu et al. 2017).
-
-	* Esto se puede deber a que acceden a reservas de agua mas profundas (los pastos y los matorrales no pueden acceder a ese agua)
-
-### 
-Nosotros hemos obtenido una relación entre la severidad de la sequía y la resiliencia mostrada por los robledales en su rear edge. (ver gráfica )
-
-@Gazol2018 encontró La resiliencia está relacionada con la severidad de la sequía y con la composición del bosque
-
-Rt-NDVI and Rt-TR were positively related with drought intensity
-
-## 
-@Gazol2018 Resistencia a la sequía mayor en la zona norte de españa que en las zonas secas del sur
-
-Qpyr 
-* Valores de Rt-TR aumentaron a lo largo del tiempo, sin embargo Rt-NDVI dismiuyó con el tiempo
-* Valores Rc-NDVI tendencia positiva
-
 ## References

### 2019-08-14 man/tables/t_huber_res_bai.Rmd
update new version

@@
 csl: ../../refs/ecosystems.csl
+editor_options: 
+  chunk_output_type: console
 ---
@@
 95 % confidence intervals using 3000 bootstrap are included between parentheses.
+*Total* corresponds to the average of 2005 and 2012.
 
@@
 # Datos agrupados para todos los sitios 
-mraw_a <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
+mraw_a <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)
 
@@
 # Datos agrupados para todos los sitios 
-mraw_b <- read.csv(here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
+mraw_b <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)
 
@@
   merge_at(i = 1, j =c(7:9) , part = "header") %>%
+    merge_at(i=1, j =c(11:13), part = "header") %>% 
   align(align = "center", part = "header") %>% 

### 2019-08-14 man/tables/t_huber_res_evi.Rmd
update new version

@@
   merge_at(i = 1, j =c(7:9) , part = "header") %>% 
+  merge_at(i=1, j =c(11:13), part = "header") %>% 
   align(align = "center", part = "header") %>% 

### 2019-08-14 man/figures/0_figures_ecosystems.Rmd
update

---
output:
  word_document:
    fig_caption: yes
    highlight: null
    reference_docx: ../../templates/template_pagebreak.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

```{r, include=FALSE}
options(knitr.duplicate.label = 'allow')
```

**Figure 1**. Distribution of *Quercus pyrenaica* forests in the Iberian Peninsula (a) and in Sierra Nevada mountain range (b). Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015]. For each population, a grid with the MODIS pixels is shown (see material and methods). Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d). Colour orthophotography of 2009 from Regional Ministry of the Environment.

![](../../out/fig/geomap/mapa_situacion_v2.jpg){height=750px}

#####

**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error. See main text for details on EVI calculation.

![](../../man/figures/svg/fig_2.jpg)

#####

**Figure 3.** Percentage of pixels showing browning, greening or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies. See main text for an explanation of greening and browning.

![](../../man/figures/svg/fig_3.jpg)

#####

**Figure 4.** Basal Area Increment (BAI) chronologies of *Q. pyrenaica* for northern population (SJ; *red*) and southern ones: low-elevation (CA-Low; *green*) and high-elevation (CA-High, *blue*) sites. Shading areas correspond to standard error of the mean. Number of series is displayed in the upper plot. We only show years replicated with # series > 5. Linear trends since 1975 are shown all sites (numbers indicate $r^2$ values; asterisks indicate significant linear trend, *p < 0.001*).

![](../../man/figures/svg/fig_4.jpg)

#####

**Figure 5.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods). Note that y-axes do not correspond in all of the three panels for the sake of clarity. Error bars indicate standard error.

![](../../man/figures/svg/fig_5.jpg)

#####

**Figure 6.** Resilience metrics of tree-growth for eigth severe drought events since 1950 (see main text for details) as a function of drought severity. *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*). Points indicate resilience metrics for oak populations: SJ (*red*), CA-High (*blue*) and CA-Low (*green*). Resilience metrics were computed for each population (sample depth > 10) and drought event. Gray lines represent overall relationships for each Resilience metrics.

![](../../man/figures/svg/fig_6.jpg)

#####

**Figure 7.** Comparison of the response of  *Q. pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (a) and tree-growth (b). For EVI we compared northern populations (*red circle*) with southern ones (*green circle*). For BAI we compared the more xeric northern population (San Juan, SJ; *red circle*) with the two southerns populations: Cáñar-High (CA-High; *blue circle*) and Cáñar-Low (CA-Low; *green circle*). Different letters indicate significant *post hoc* differences between groups (see material and methods for details).

![](../../man/figures/svg/fig_7.jpg)

#####

**Figure 8.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q. pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures. *green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High). Asteriks indicate significant ($P < 0.05$) correlation coefficients.

![](../../man/figures/svg/fig_8.jpg)
#####

**Figure S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017. Points represent mean and errorbars standard error. *Black* line indicates mean for the whole period (585 mm). *Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively). *Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively). Years with average values below -1SD are labelled. Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET). ***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*). The boxplot representing the average from 1950-2015 period. Data from meteorological station Granada, Base Aérea.

![](../../man/figures/svg/fig_s1.jpg)

#####

**Figure S2**. Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI). Data from Global SPEI database (http://spei.csic.es/database.html). We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada. Horizontal gray bars indicate 2005 and 2012 year.

![](../../man/figures/svg/fig_s2.jpg)

#####

**Figure S3.** Residual tree-ring chronologies obtained for the *Q. pyrenaica* sites. Dashed red lines indicate the start of the reliable period (EPS > 0.85). Dotted black lines showing the severe drought years identified in our climatic data (Table S3 and Figure S1).

![](../../man/figures/svg/fig_s3.jpg)

#####

**Figure S4.** Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis). Data for number of trees > 2 is shown.

![](../../man/figures/svg/fig_s4.jpg)

#####

**Figure S5.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought). Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.

![](../../man/figures/svg/fig_s5.jpg)

#####

**Figure S6.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years). Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot`.

![](../../man/figures/svg/fig_s6.jpg)

##### 
**Figure S7.** EVI annual profile (average of the period 2000-2016) for *Q. pyrenaica* forests in Sierra Nevada and drought events. Horizontal bars correspond to the most severe droughts for Sierra Nevada since 1900 (computed as in Table S3). Their position indicate the start and end months of each drought event. Bars length show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017]).

![](../../man/figures/svg/fig_s7.jpg)

#####

**Figure S8.** Resilience metrics of the tree-growth for severe drought events since 1950 (excluding 1995 drought event). *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*). Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*). Resilience metrics were computed for each population (sample depth > 10) and drought event. Gray line represents overall relationship for each Resilience metrics.

![](../../man/figures/svg/fig_s8.jpg)

### 2019-08-14 man/tables/t_sampled_plots.Rmd
update

@@
 csl: ../../refs/ecosystems.csl
-editor_options: 
-  chunk_output_type: console
 ---

### 2019-08-22 man/ecosystems/ms_ecosystems.Rmd
update ms

@@
 ---
-title: "Relict deciduous oaks show high resilience to drought along a climatic gradient in the rear-edge despite strong land-use legacies "
+title: "Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today"
 output:
@@
 ## **Short Title**
-Relict oaks resilience at their rear-edge
+Relict oak resilience at the rear-edge
 
@@
 * Global change challenges ecosystems in dry locations transformed by intensive human use.
-We analyzed resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd.
-populations in the Southern Iberian Peninsula.
-Forest resilience to drought was analyzed in relation to historical records of land-use using dendroecological growth and greenness (EVI) as proxies for secondary and primary growth.
+Forest resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd.
+populations in the Southern Iberian Peninsula was analyzed in relation to historical records of land-use using dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
 
 * The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure.
-Trees were highly sensitivity to drought and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term.
+Trees were highly sensitive to moisture availability and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term.
 The response to climate and resilience followed a water-stress gradient.
-Primary and secondary growth of more xeric populations (*i.e.* northern aspect) was less resilient to drought than that at more mesic sites.
-In addition, a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic high-elevation stands, but absent in the most xeric site.
+Primary and secondary growth were generally less resilient to drought in xeric than in mesic sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and more humid) high-elevation stands, but absent in the most xeric site.
 
-* The species showed high resilience to drought despite being relict and at its rear-edge.
-Yet, the trees exhibited a variable response to the changing climate along a very short climatic gradient.
-The trees were more sensitive to water availability at most xeric sites whereas only the coldest and more humid stands exhibited a recent positive growth trend.
-Stand dynamics at the most xeric locations (*i.e.* the altitudinal rear-edge) needs to be monitored under future expected climatic scenarios bearing in mind that land-use legacies play a key role in stand dynamics.
+* Adult trees of the studied species showed high resilience to drought despite being considered relict and at its rear-edge.
+Yet, the trees exhibited a variable response to the changing climate along a very short climatic gradient and showed that the ecological and geographical rear-edges do not need to meet today in forests with strong land-uses legacies.
+Forest resilience including that of regeneration needs to be monitored at xeric locations (*i.e.* the low altitudinal rear-edge) under future expected climatic scenarios bearing in mind that land-use legacies play a key role in stand dynamics.
 
@@
 
-Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of the Mediterranean climate [@Dai2011].
+Resumir párrafos de Vilá-Cabrera (los que ha puesto RZ)
+
+Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].
 There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017].
 This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a].
-
-Drought causes significant changes in ecosystem productivity and dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation.
-Thus, assessment of drought impacts on ecosystems has gained much attention in last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].
-
-Ecosystem responses to drought are influenced by other drivers of environmental changes.
 In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017].
@@
 
-The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005].
+The impacts of drought are also especially relevant for populations considered to be located in the rear-edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005].
 The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010].
-This becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].
+Yet it is crucial to accurately characterize the rear-edge of species.
+This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabreraJump2019GreaterGrowth].
+Furthermore, it becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].
 
-Many studies have analyzed the effects of drought on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
+Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [*e.g.*, @SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
 Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973].
-The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services.
+The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and provide valuable ecosystem services.
 
-Vegetation responses to drought requires of integrative and multidisciplinary strategies for assessing the impacts of disturbances on forest at several scales [@Hartman2018].
-The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously.
-Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
-The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
+Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018].
+The resilience to drought of species is better assessed when several approaches like remote sensing and dendrocronology are combined simultaneously.
+Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
+The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
 Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013].
@@
 Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
-Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level, particularly in rear edge populations.
+Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.
 
-In this work we used remote sensing information and dendroecological methods to evaluate the drought impact in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of a Mediterranean tree species (*Quercus pyrenaica* Willd.) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be threatened by climate change.
+We used remote sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of *Quercus pyrenaica* Willd.
+on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
 We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) in the long-term in the last decades.
@@
 To address this hypothesis our specific aims were: (*i*) To quantify how recent extreme drought events influenced primary and secondary growth of *Q.
-pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of this species to detect vulnerability to climate change along climatic gradients.
-To achieve goal (*iii*) we assessed within the region of study whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
+pyrenaica* forests in its rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) within the rear edge to detect vulnerability to climate change along climatic gradients at the small spatial scale.
+To achieve goal (*iii*) and characterize variability in the forest response to drought within the rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
 
@@
 *Q.
-pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm.
-Summer rainfall is a key factor for the species, which is considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@delRio2007].
+pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation.
+It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].
 
-The forests of this species reache their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 m and 3482 m *a.s.l.*.
-The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August).
+The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 m and 3482 m *a.s.l.*.
+The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August) but with marked variability set by altitude.
 There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
@@
 Today, *Q.
-pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005].
+pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005].
 They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG].
@@
 
-The Iberian Peninsula suffered several extreme drought episodes in the last decades of the twentieth century (*e.g*.
-1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
-We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below).
+The Iberian Peninsula suffered several extreme drought episodes in the last three decades (*e.g*.
+1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].
+The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
+We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below).
 Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3).
@@
 A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017].
-A drought event is considered only when SPEI values are below the threshold for at least two consecutive months.
+A drought event is considered only when SPEI values are below that threshold for at least two consecutive months.
 Several indicators were computed for each drought event: *duration* as the number of consecutive months with the SPEI lower than a certain threshold; *severity* of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
@@
 
-As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016.
-To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012].
+Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016.
+The EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012].
 For each pixel, an annual EVI value was obtained by averaging EVI valid values.
@@
 $$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
-where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
+where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
 Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI~sa > 1$), "browning" ($EVI~sa <- 1$) or "no changes" ($-1 > EVI~sa > 1$)[@Samanta2012].
@@
 
-## Field sampling and dendrochronological methods to assess individual resilience
+## Field sampling and dendrochronological methods to assess tree individual resilience
 
 Tree sampling was carried out during autumn 2016.
-Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid but warmer site located at the southern aspect (Figure 1; Table 1).
-For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope.
-Sampling sites follow a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
-All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015].
+Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1).
+For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled aspect Sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
+All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015].
 In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected.
@@
 Diameter at breast height (DBH) and total height were measured for each tree.
-In addition, stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius.
+In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius.
 Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].
 Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests.
-When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct significance values.
+When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct for significance.
 
@@
 Theoretically, BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008].
-Raw ring-widths and measured DBH were used to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008].
-We used the following equation:
-
-$$\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$$
-where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
+Raw ring-widths and measured DBH were used to compute BAI [@Piovesan2008] using the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
 For each individual tree, a mean BAI series was calculated.
 Then, mean site BAI chronologies were obtained by averaging individual tree BAI time series.
-
-Both RWI and BAI chronologies were compared to different climatic indices to analyse the low and the high-frequency response to climate of trees.
 
@@
 The Standardized Precipitation-Evapotranspiration Index (SPEI) with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
-
-Relationships between residual site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].
+Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].
 
@@
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
-To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance [*e.g.* @GeaIzquierdo2014].
-
+To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014].
 In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
@@
 A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2018].
-To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared them with drought severity (Figure 6, Table S3).
+To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Figure 6, Table S3).
 
@@
 
-For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects).
+For each of the three resilience indices studied we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects).
 These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012].
@@
 
-The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period.
-The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in northern (dry) populations (Figure 2).
+The analysis of greenness time trends showed that EVI in 78.9 % of the pixels experienced a positive trend for the 2000-2016 period.
+The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2).
 A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).
@@
 The trees of the southern population were older than those from the northern one.
-In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4), despite competition measured as plot basal area was greatest in CA-High (Table 1).
+In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4) and despite competition measured as plot basal area was greatest in CA-High (Table 1).
 Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6).
-Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being much more pronounced for the highest, hence more humid, site (CA-High).
+Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being much more pronounced for the high-elevation, hence more humid, site (CA-High).
 
-Drought events produced reduction in RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
-The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S3).
-Tree-growth reduction followed a moisture-gradient, being weaker for the wettest site (CA-High, Figure 5).
+Drought events reduced radial growth for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
+Tree-growth reduction followed a moisture-gradient.
+The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S3), being the weakest for the wettest site (Figure 5).
 
@@
 ## Resilience to drought events at the ecosystem and individual tree levels
-Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6).
-A similar pattern was obtained for Resilience (*Rs*) but only significant for SJ site (Figure 6).
-Importantly, non-significant patterns were obtained if we excluded 1995 from the analysis, except for *Rc* in SJ (Figure S8).
-Trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), and particularly SJ, where our results suggest a major release event after 1995 (Figure 5).
+
+*Rs* and *Rt* varied in the same direction whereas *Rc* varied inversely to *Rs* and *Rt*.
+Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 6).
+A similar pattern was obtained for *Rs* but only significant for SJ site (Figure 6).
+Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S8).
+The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), particularly SJ where our results suggest a major release event also after 1995 (Figure 5).
 
 During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table 3).
-The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1 and S2) but the two metrics of resilience generally covaried in the same direction.
-We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7).
-Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth.
-Recovery values for greenness were higher for 2005 than for 2012 drought event (Table S1, Figure 7).
+The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1 and S2) but the metrics of resilience generally covaried in the same direction those two years.
+We obtained significantly higher Resilience and Resistance values for 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7).
+Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event (Table S1, Figure 7).
+Recovery showed a contrasting pattern for EVI and tree-growth.
 
-Overall, resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3), which was similar among the three sites.
-The two southern populations showed lower values for Recovery (*Rc*) than the northern site both for greenness and tree-growth (Tables S1 and S2).
-In contrast, Resistance and Resilience values were significantly higher for southern sites than for northern ones (Tables S1 and S2).
+The resilience metrics calculated significantly varied between sites, except for Resilience of tree-growth (p = 0.534; Table 3), which was similar among the three sites.
+The two southern populations showed lower values for Recovery than the northern site both for greenness and tree-growth (Tables S1 and S2) but Resistance and Resilience values were significantly higher for the southern site (Tables S1 and S2).
 
@@
 Yet there were sone differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 8a).
-In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 8b and 8c).
-Minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
+In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 8b and 8c), whereas minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
 

### 2019-08-26 man/ecosystems/ms_ecosystems.Rmd
update contents

@@
 # Discussion
-## Land-use legacies shape forests and sensitivity to climate change at the rear-edge
+## Land-use legacies shape sensitivity to climate change of forests and the present rear-edge
 
-The Mediterranean landscapes, even at high elevations, have suffered a strong transformation driven by human activities throughout history [@GarciaAlix2017].
-Forest clearings, firewood extraction, charcoal exploitations and mining (these were strongly suggested in the disturbance chronology in SJ), have strongly impacted on the forest resources at Sierra Nevada as can be derived from the results of the review of historical documents (Table S4).
-Indeed a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation has been documented in this mountain region [@JimenezOlivencia2015].
+The Mediterranean landscapes have suffered a strong transformation driven by human activities throughout history even at high elevations [@GarciaAlix2017].
+The review of historical documents shows how forest clearings, firewood extraction, charcoal exploitations and mining have strongly impacted the forests at Sierra Nevada (Table S4) where in addition it has been estimated a historical loss of about 90% of broadleaf *Quercus* species cover from medium and low elevations [@JimenezOlivencia2015].
+Therefore, it seems logic to think that the geographical limit of the current rear-edge and relict stands of the deciduous oak studied does not necessarily match the potential dsitribution limit related to edaphoclimatic constraints (*i.e.* the ecological rear-edge).
 Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland) [@JimenezOlivencia2015].
-It has been described a sequentially distribution of the land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops [@JimenezOlivencia2015].
-Other activities (like mining) should have determined the forest structure at that northern location, since SJ is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history.
-On the other hand, woodland areas of the southern slopes (*e.g.* CA site) are mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015].
+It has been described a sequential distribution of historical land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming were located at the highest elevations; then forest stands with some croplands; and at the lowest elevations, irrigated terraces with tree crops [@JimenezOlivencia2015].
+In addition, other activities like mining should have modified the forest structure at that northern location where there is with a high concentration of mines and quarries that were exploited intermittently throughout history.
+On the other hand, woodlands in the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015].
 Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013; @JimenezOlivencia2015].
 
-Some of the previous historical records of land-use were expressed in our growth chronologies.
-The release event at 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990].
-This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth the remnant trees included in the northern site (Figure S4).
+Some of the previous historical records of land-use were likely expressed by the growth chronologies.
+The release event in 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990].
+This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S4).
 The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4).
@@
 Cano-Manuel *personal communications*).
-We speculate that this might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
+Therefore this release might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
 On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014].
-For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] documented in this site [@Valbuena2013].
-
-As we inferred from our results, the management history can conditioned tree-growth and resilience of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017].
-Nevertheless, we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource.
-This pattern concurs with a forest expansion into marginal abandoned croplands and an increase in the tree-cover [@CamachoOlmedo2002].
+For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
+In this site we observed an increase in growth especially in high elevation trees, where moisture is less limiting, a pattern which concurs with a forest expansion into marginal abandonded croplads and an increase in tree-cover [@CamachoOlmedo2002].
 
@@
 This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
-Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
-Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g.* 2005 and 2012 were considered two of the worst recorded drought; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q.
+Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic precipitation negative anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
+Although the south of Iberian Peninsula suffered severe drought events in the last decades (Table S3), we found a positive trend for vegetation greenness of *Q.
 pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
@@
 Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014].
-However, the observed pattern for our sites (a rear-edge) differs from the declining trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P.
-sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].
+Importantly, for rear-edges where climate change is a threat we would have expected negative growth trends as shown for temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
+Therefore there is much small-scale variability in the response to climate within the rear-edge we had "a priori" considered in our study, where we even observed positive effects of climate change in certain stands.
+This suggests that the rear-edge needs to be redefined in consequence [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear-edge) and the potential ecological (limiting) rear-edge of species.
 
-Despite the 2012 drought event was more severe and intense than 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012.
-This could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought.
-The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q.
-pyrenaica* forests (Figure S7).
-Thus, this highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], as also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b].
-For tree-growth, the highest values of resilience were obtained for the most severe events, namely 1995 and 1999 (Table S3).
-In addition, we observed a positive relation between the tree-growth resilience and drought severity (Figures 6).
+Despite that 2012 drought event was more severe and intense than that in 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012.
+This could be explained by the different timing of the two droughts.
+The 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought and although the latter lasted less than the 2012 drought, it matched the period of maximum growth for *Q.
+pyrenaica* forests in late spring (Figure S7).
+Thus, as bserved for other species [*e.g.* Holm oak; @Camarero2015b] this highlights the importance of the timing of drought as a key factor determining the tree recovery after drought [@Huang2018].
+For tree-growth, the highest values of resilience were obtained for the two most severe events, namely 1995 and 1999 (Table S3) and we observed a positive relationship between the tree-growth resilience and drought severity (Figures 6).
 This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events.
 Yet, the high resilience values obtained for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J.
-Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of individual resilience estimated from individual growth series.
+Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of tree resilience estimated from individual growth series.
 
-Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005].
+Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of species [@Hampe2005].
 A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018].
@@
 Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017].
-@MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations.
-The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013], seem to be indicative of high local adaptation of this oak.
-In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
-A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018].
+@MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability showed by rear-edge populations.
+Our results would highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our studied forest.
+The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013] seem to be indicative of high local adaptation of this oak.
+In addition the local environmental conditions of the sites where this species inhabits at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
+It needs to be mentioned that we only studied adult individuals established decades or centuries ago, meaning that it needs to be assessed if the species expresses resilience or to the contrary vulnerability if we analyzed seedling regeneration, likewise many other Mediterranean species at their xeric limit (2-3 CITAS).
+A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018; @VilaCabrera2019].
 In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q.
@@
 
-Greenness and tree-growth of *Q.
-pyrenaica* showed different sensitivity to severe droughts.
-During the last drought events EVI was reduced to 90.3 % (Table 1), whereas BAI was reduced to 70.8 % (Table S2), suggesting a lower sensitivity to drought of greenness than of tree-growth, particularly for drier sites.
-These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018].
-Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based (*i.e.* ecosystem or stand based); the growth reduction seems to be more mediated by sink that by source limitations [@Korner2013].
-The response of greenness to drought not only reflects the response of trees but from all vegetation within the pixel [@Dionisioetal2012IPoGEC].
-In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates.
-This is particularly important for *Q.
-pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014].
+## Small-scale environmental variability shapes sensitivity to climate of trees within the rear-edge
+
+Since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018] a combined analysis as in our study allows to better disentangle the small and large-scale variability in growth induced by climate from that driven by community-level ecological processes [@Babst2017].
+Greenness exhibited a lower sensitivity to drought than tree-growth, particularly for drier sites.
+These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018] and would suggest that the growth reduction could more mediated by sink that by source limitations [@Korner2013; @Fatichi2014].
 On the other hand, tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
-A combined analysis, such the approach used in our study, may allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018].
 
-## Small-scale environmental variability shapes the sensitivity to climate of trees within the rear-edge
-
-Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event.
-In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site, can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014].
+Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event and the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014].
 Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
 pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
-Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017].
+Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @GeaIzquierdo2011; @GarciaGonzalez2017].
 
-It has been described an effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive, thus high-elevations can be used as local refuge for species in response to climate change [@Bhuyan2017].
+Higher elevations within warm and dry ecosystems can help species alleviate climatic stress and 
+increase resistance to drought and stands of several tree-species located at higher elevations were less drought sensitive.
+Thus high-elevations are generally considered as potential local refuge for species in response to climate change [@Bhuyan2017].
 It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
-Remarkably, the trees at CA-High, which are located around 1900 m.
-*a.s.l.* and represent today the upper altitudinal limit (treeline) for the species in this southernmost location, showed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4).
-This shows the high variability in the response to climate exhibited in a small gradient and is especially interesting for the southern sites, which are very close to each other, and overall both constitute the rear-edge for the species.
-The higher values of tree growth in taller trees that we observed in CA-High site strongly suggest a lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017].
+Trees at CA-High sowhed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4).
+This shows the high variability in the response to climate exhibited in a small gradient, which is especially interesting for southern sites, which are very close to each other, and overall both constitute the rear-edge for the species.
+The higher values of tree growth in taller trees that we observed in CA-High site strongly suggests lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017].
 Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014].
-Finally and in a context of global change, in addition to the withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).
+Finally and in a context of global change, in addition to withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).
+Overall, our results show how management history conditions tree-growth and resilience to climate change of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017] which, hence, will also strongly determine the current geographical distribution of marginal stands, regardless of the potential extent of ecological marginality of species.
 
@@
 # Conclusions 
+The ecological and geographical read-edges did not seem to fully match in our study.
 Severe drought events provoke important reductions in primary and secondary growth of *Q.
@@
 Furthermore, we observed positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a positive steep trend of secondary growth at the tree level at the most humid site along the climatic gradient analyzed.
-The trees exhibited high resilience values in response to drought, particularly in the long-term scales, and resilience was directly related to drought severity.
+The trees exhibited high resilience values in response to drought, particularly in the long-term scale, and resilience was directly related to drought severity.
 These results would be consistent with other research showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*.
@@
 Following our study it is necessary to further disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.
+Since we only studied resilience of already established adult trees, monitoring should be enlarged to analyze overall species resilience including that of seedling regeneration.
+Our results show how management history conditions tree-growth and resilience to climate of tree species.
+They highlight the importance of land-use legacies and how they affect the current geographical distribution of marginal stands, which therefore can be partly disconnected from actual ecologically marginal conditions.
 
@@
 We would like to thank Sammy Lucy Behle, M.
-Suarez-Muñoz and F.J.
+Suárez-Muñoz and F.J.
 Bonet-García for their assistence in the field monitoring.
@@
 Bonet-García for his valuable comments in an earlier version of the manuscript.
+AJPL wants to thank the invaluable support received from his family over the years.
 Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work.
-AJPL wants to thank the invaluable support received from his family over the years.
 This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project).

### 2019-09-23 man/ecosystems/ms_ecosystems.Rmd
update ms

@@
 
-Resumir párrafos de Vilá-Cabrera (los que ha puesto RZ)
+The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005].
+Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species’ range (*i.e.* centre-periphery hypotheses) [CITAS].
+It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance and thus higher risk of extinction than those populations at the core of the species’ range [@VilaCabrera2019; @Pironon2016].
+Nonetheless, recent reviews found species- and population- specific responses highlight that this hypothesis not always fulfill [*e.g.* Sexton2009; Abeli2014].
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginallity does not correspond with ecological marginallity [@VilaCabrera2019].
 

### 2019-09-23 man/figures/0_figures_ecosystems.Rmd
update

@@
 
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error. See main text for details on EVI calculation.
+**Figure 2.** ***a*** EVI standardized anomaly during the period 2000-2016 for northern and southern populations.
+Error bars show standard error.
+See main text for details on EVI calculation.
+***b*** Percentage of pixels showing browning, greening or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
+See main text for an explanation of greening and browning.
 
-![](../../man/figures/svg/fig_2.jpg)
+![](../../man/figures/svg/fig_2ab.jpg)
 
@@
 
-**Figure 3.** Percentage of pixels showing browning, greening or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
-See main text for an explanation of greening and browning.
-
-![](../../man/figures/svg/fig_3.jpg)
-
-#####
-
-**Figure 4.** Basal Area Increment (BAI) chronologies of *Q.
+**Figure 3.** Basal Area Increment (BAI) chronologies of *Q.
 pyrenaica* for northern population (SJ; *red*) and southern ones: low-elevation (CA-Low; *green*) and high-elevation (CA-High, *blue*) sites.
@@
 
-**Figure 5.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q.
+**Figure 4.** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q.
 pyrenaica* sites.
@@
 
-**Figure 6.** Resilience metrics of tree-growth for eigth severe drought events since 1950 (see main text for details) as a function of drought severity.
+**Figure 5.** Resilience metrics of tree-growth for eigth severe drought events since 1950 (see main text for details) as a function of drought severity.
 *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
@@
 
-**Figure 7.** Comparison of the response of  *Q.
+**Figure 6.** Comparison of the response of  *Q.
 pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (a) and tree-growth (b).
@@
 
-**Figure 8.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
+**Figure 7.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
 pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures.
@@
 ![](../../man/figures/svg/fig_8.jpg)
+
 #####
@@
 Data from meteorological station Granada, Base Aérea.
-
-![](../../man/figures/svg/fig_s1.jpg)
-
-#####
-
-**Figure S2**.
-Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+***B*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
 Data from Global SPEI database (http://spei.csic.es/database.html).
@@
 
-![](../../man/figures/svg/fig_s2.jpg)
+![](../../man/figures/svg/fig_s1ab.jpg)
 
@@
 
-**Figure S3.** Residual tree-ring chronologies obtained for the *Q.
+**Figure S2.** Residual tree-ring chronologies obtained for the *Q.
 pyrenaica* sites.
@@
 
-**Figure S4.** Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis). Data for number of trees > 2 is shown.
+**Figure S3.** Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis). Data for number of trees > 2 is shown.
 
@@
 
-**Figure S5.** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-Top plots (a, b and c) showing the resilience indices of greenness (EVI) to drought; and bottom plots (d, e, f) the resilience indices of tree-growth (BAI) to drought.
+**Figure S4.** ***A*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
+Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
+Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot`.
+***B*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
 
-![](../../man/figures/svg/fig_s5.jpg)
+![](../../man/figures/svg/fig_s4ab.jpg)
 
 ##### 
-
-**Figure S6.** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot`.
-
-![](../../man/figures/svg/fig_s6.jpg)
-
-##### 
-**Figure S7.** EVI annual profile (average of the period 2000-2016) for *Q.
+**Figure S5.** EVI annual profile (average of the period 2000-2016) for *Q.
 pyrenaica* forests in Sierra Nevada and drought events.
@@
 
-**Figure S8.** Resilience metrics of the tree-growth for severe drought events since 1950 (excluding 1995 drought event).
+**Figure S6.** Resilience metrics of the tree-growth for severe drought events since 1950 (excluding 1995 drought event).
 *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).

### 2019-09-23 man/figures/fig2_combined.Rmd
update

--- 2019-02-05 man/figures/std_anomalies.Rmd
@@
 library("here")
+library("ggrepel")
 library("scales")
+library("grid")
+library("gridExtra")
 library("patchwork")
+library("ggcorrplot")
+library("stringr")
+library("pander")
+library("lemon")
+library("ggpubr")
+library("viridis") # colours 
+library("ggthemes") # colours 
+source(here::here('/script/R/custom_functions.R'))
 ```
-
-**Figure 2.** EVI standardized anomaly during the period 2000-2016 for northern and southern populations. Error bars show standard error. See main text for details on EVI calculation.
 
@@
 
-```{r out_plot, fig.width= 4, dpi = 300, out.width = 0.8}
-plot_sa_clu 
+## Browing
+
+```{r}
+df <- read.csv(file=here::here('/out/anomalies/evi' ,'percen_browing_greening.csv'), header=TRUE)
+
+df <- df %>% add_row(clu_pop2 = 'Northern slope',
+                             type = 'greening',
+                             count_clu = 0, per = 0.00, y = 2005)
+
+df <- df %>% mutate(type = stringr::str_to_title(type))
 ```
 
-```{r oldFigure, eval=FALSE, fig.height=7}
-## See profile_EVI.Rmd 
-# grid.arrange(profile_compara, plot_sa_clu)
+```{r}
+per_pixels_green <- df %>% 
+  ggplot(aes(x=type, y=per, fill=as.factor(y))) + 
+  geom_bar(stat = 'identity', position = "dodge") + 
+  geom_text(aes(label = paste0(per, " %")), position=position_dodge(width=0.9), hjust=-.1, size=3, angle = 90) + 
+  facet_wrap(~clu_pop2) + 
+  theme_bw() + xlab('') + ylab('Frequency of pixels (%)') +
+  theme(strip.background = element_rect(colour = "white", fill = "white"),
+        strip.text = element_text(size = 7), 
+        axis.title.y = element_text(size = 7, face = 'bold'),
+        axis.text = element_text(size = 7),
+        panel.grid = element_blank(),
+        legend.position = 'bottom', 
+        legend.text = element_text(size = 7), 
+        legend.box.spacing = unit(0.01, "cm"))+
+  labs(fill = 'Drought event') + 
+  scale_fill_colorblind() + 
+  scale_y_continuous(limits = c(0, 110), breaks = c(0, 25, 50, 75, 100)) +
+  guides(fill = guide_legend(title.theme = element_text(size = 7)))
+```
 
-a_profile_compara <- profile_compara + labs(title = "a")
-b_plot_sa_clu <- plot_sa_clu + labs(title = "b")
-a_profile_compara + b_plot_sa_clu + plot_layout(ncol=1)
+```{r}
+f2 <- ggarrange(plot_sa_clu, per_pixels_green, 
+                    labels = c("a", "b"),
+          ncol = 2, nrow = 1)
 ```
+
+```{r}
+saveImage(f2, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_2ab", u = "cm") 
+```

### 2019-09-23 man/figures/s1_combined.Rmd
update

--- 2019-02-06 man/figures/cumulative_prec.Rmd
@@
 library("lemon")
+library("ggpubr")
+source(here::here('/script/R/custom_functions.R'))
 ```
@@
 
-```{r plot_out, fig.width=7, dpi = 300, out.width = 0.8} 
+##### SPEI
+
+```{r}
+# DATOS desde 1901
+
+# Get file names
+myfiles <- list.files(path = here("data_raw/spei/spei_longterm"), pattern = "\\.csv$")
+
+mydf <- data.frame()
+
+for (j in myfiles){ 
+  aux <- read.csv(file=here("data_raw/spei/spei_longterm", j), header = TRUE, sep = ',')
+
+  # Remove csv and get name 
+  name_aux <- str_replace(j, ".csv", "")
+
+  # Get lat long
+  latlong <- str_replace(name_aux, "spei_", "")
+  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
+  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[2]
+
+  # Store lat and long 
+  aux$lat <- mylat
+  aux$long <- mylong
+
+  # year and month 
+  aux$month <- rep(1:12, 115)
+
+  auxy <- c()
+  years <- c() 
+  for (j in 1901:2015) { 
+    auxy <- as.data.frame(rep(j,12))
+    years <- rbind(years, auxy)
+    } 
+  names(years) <- 'year'
+
+  aux <- cbind(aux,years)
+
+  # join all df 
+  mydf <- rbind(mydf, aux)
+}
+
+# Convert to date  
+mydf <- mydf %>% mutate(date = sprintf("%d-%02d", year, month),
+                        date = as.Date(paste0(date, '-01'), format="%Y-%m-%d"))
+
+speidf <- mydf %>% dplyr::select(SPEI06,SPEI12,SPEI24,month, year, date) %>% 
+  group_by(year,month, date) %>%
+  summarise_at(vars(SPEI06,SPEI12,SPEI24), funs(mean,sd,se=sd(.)/sqrt(n()))) %>% 
+  as.data.frame() 
+```
+
+```{r}
+speiPlot <- function(df, vi, mytitle, year_scaleX){
+  require(dplyr)
+  require(scales)
+
+  # Select variable of interest 
+  dfs <- df %>% dplyr::select(year, month, date, starts_with(vi))
+
+  # Rename variables 
+  dfs <- dfs %>% rename_all(~sub(paste0(vi,'_'), '',.x))
+  # Create variable pos and neg 
+  dfs <- dfs %>% mutate(signo = ifelse(mean > 0, "pos", "neg"))
+
+  # Plot  
+  p <- ggplot(dfs, aes(x = date, y = mean, fill = signo)) + 
+    geom_rect(aes(xmin=as.Date("2005-01-01"), xmax = as.Date("2005-12-01"), ymin=-Inf, ymax=Inf),
+              fill="grey", alpha=.9) + 
+    geom_rect(aes(xmin=as.Date("2012-01-01"), xmax = as.Date("2012-12-01"), ymin=-Inf, ymax=Inf),
+              fill="grey", alpha=.9) + 
+    geom_bar(stat = "identity") +
+    scale_fill_manual(values = c("pos" = "darkblue", "neg" = "red")) +
+    theme_bw()  + 
+    labs(title= mytitle, x= '', 
+         y = paste0(substr(vi, 1, 4), ' (', gsub("[^0-9]","",vi),'-month scale)')) +
+    theme(strip.background = element_rect(fill = "white"),
+        legend.position = "none", 
+        panel.grid = element_blank(), 
+        axis.text.y = element_text(size=10),
+        axis.text.x = element_text(size=9),
+        axis.title.y = element_text(size = 12))
+
+  # Date Scale in X 
+  if (missing(year_scaleX)){
     p
+    } else { 
+    year_interval <- paste0(year_scaleX," year")
+    p <- p + scale_x_date(date_breaks = year_interval, date_labels = "%Y",
+                          limits=as.Date(c("1901-01-01", "2015-12-01")))
+    }
+
+  return(p)
+}
+
+s <- speiPlot(speidf, vi = 'SPEI12', mytitle = "", year_scaleX = 10)
 ```
+
+```{r}
+s12 <- ggarrange(p, s,
+          labels = c("a", "b"),
+          ncol = 1, nrow = 2)
+```
+
+```{r}
+# SAVE as eps 
+saveImage(s12, path="./man/figures/svg/", dpi=300, w=25, h=20, fig= "fig_s1ab", u = "cm") 
+```
+
+**Figure S1.** ***A*** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
+Points represent mean and errorbars standard error.
+*Black* line indicates mean for the whole period (`r meandf` mm).
+*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
+*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
+Years with average values below -1SD are labelled.
+Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
+***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
+The boxplot representing the average from 1950-2015 period.
+Data from meteorological station Granada, Base Aérea.
+***B*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+Data from Global SPEI database (http://spei.csic.es/database.html).
+We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
+Horizontal gray bars indicate 2005 and 2012 year.

### 2019-09-23 man/figures/s4_combined.Rmd
update

---
output: 
  word_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r load_pkg}
library("tidyverse")
library("here")
library("dplyr")
library("ggrepel")
library("scales")
library("ggpubr")
library("grid")
# library("GGally")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
library("stringr")
library("pander")
library("lemon")
library("viridis")
library("ggthemes")
source(here::here('/script/R/custom_functions.R'))
```

```{r data}
corBai <- read.csv(file=here("/out/correla_resilience/", "correla_window_size.csv"), header = TRUE)
corEvi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_evi.csv"), header = TRUE)
corRwi <- read.csv(file=here("/out/correla_resilience/", "correla_window_size_rwi.csv"), header = TRUE)

generaMatrix <- function(cordf, variable){ 
  m <- cordf %>% filter(var == variable) %>% 
    dplyr::select(-var, -window_size) %>% 
    mutate(w1 = c(2, 2, 3), 
           w2 = c(3,4,4)) 
  mf <- m %>% 
    dplyr::select(r2, w2, w1) %>% rename(w1 = w2, w2 = w1)

  df <- m %>% rbind(mf) %>% 
    add_row(r2 = 1, w1=2, w2=2) %>% 
    add_row(r2 = 1, w1=3, w2=3) %>% 
    add_row(r2 = 1, w1=4, w2=4)

    df <- df %>% 
    mutate(w1 = case_when(
      w1 == 2 ~ '2 years',
      w1 == 3 ~ '3 years',
      w1 == 4 ~ '4 years'),
      w2 = case_when(
      w2 == 2 ~ '2 years',
      w2 == 3 ~ '3 years',
      w2 == 4 ~ '4 years'))

      cor_df <- as.data.frame.matrix(xtabs(r2 ~., df))

return(cor_df)
}

cor_rt <- generaMatrix(corBai, 'rt')
cor_rc <- generaMatrix(corBai, 'rc')
cor_rs <- generaMatrix(corBai, 'rs')

cor_rte <- generaMatrix(corEvi, 'rt')
cor_rce <- generaMatrix(corEvi, 'rc')
cor_rse <- generaMatrix(corEvi, 'rs')
```

```{r plot_function}
correP <- function(df, mytitle, ...){
  require(ggplot2)
  require(ggcorrplot)

  p <- ggcorrplot(df, method = "square", type ='upper', 
                  outline.color = "black", colors = "white",
                  ggtheme = theme_minimal, 
                  lab = TRUE, lab_size = 3.5,
                  title = mytitle, show.legend = FALSE, 
                  tl.cex = 7,
                  tl.srt = 1)
  p <- p + theme(plot.title = element_text(size = 10, face = "bold.italic"),
                 panel.grid = element_blank(),
                 axis.text.x = element_text(hjust = 0.5)) + scale_x_discrete(position = "top") 
  p      
}
```

```{r plot}
g_rt <- correP(cor_rt, 'BAI Rt')
g_rc <- correP(cor_rc, 'BAI Rc')
g_rs <- correP(cor_rs, 'BAI Rs')
g_rte <- correP(cor_rte, 'EVI Rt')
g_rce <- correP(cor_rce, 'EVI Rc')
g_rse <- correP(cor_rse, 'EVI Rs')

g <- (g_rte + g_rce + g_rse) / (g_rt + g_rc + g_rs) + plot_layout(nrow=2) + plot_annotation(tag_levels = c('a'))
```

```{r}
correla  <- read.csv(file=here("/data/correla_ventanas_temporales/", "correla_chronos.csv"), header = TRUE)

correla <- correla %>% 
  mutate(
    name_comparison = as.factor(case_when(
      name_comparison == "caH-caL" ~ "CAHigh - CALow",
      name_comparison == "sj-caL" ~ "SJ - CALow",
      name_comparison == "sj-caH" ~ "SJ - CAHigh")))
```

```{r}
correla_sitesSJCALH  <- correla %>% 
  filter(metodo_correla == 'pearson') %>%
  ggplot(aes(x=size, y=estimate)) + 
  geom_line(aes(colour=name_comparison), size = 1) + 
  theme_bw() + ylab('Correlation coefficient') +
  xlab('Smoothing (centred moving average window size)') + 
  geom_hline(yintercept = 0, colour='gray', linetype=2) + 
  geom_ribbon(aes(ymax=ci_upper, ymin=ci_lower, fill=name_comparison), alpha=0.3) + 
  scale_fill_colorblind() + 
  scale_color_colorblind() +
  theme(legend.position = c(.8,.87), 
        legend.title = element_blank(), 
        legend.text = element_text(size = 6),
        panel.grid = element_blank(), 
        axis.title = element_text(size = 10), 
        axis.text = element_text(size = 7),
        legend.background = element_blank())
```

```{r} 
correla_sitesSJCALH
```

```{r}
s4ab <- ggarrange(correla_sitesSJCALH,
          ggarrange(g_rte, g_rte, g_rce, g_rc, g_rse, g_rs, ncol=2, nrow=3), 
          ncol = 2,
          labels = c("a", "b"))
```

**Figure S4.** ***A*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years). Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot`. ***B*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).

```{r}
# SAVE as eps 
saveImage(s4ab, path="./man/figures/svg/", dpi=300, w=20, h=10, fig= "fig_s4ab", u = "cm")
```

### 2019-09-25 man/ecosystems/ms_ecosystems.Rmd
update

--- 2019-08-26 man/ecosystems/ms_ecosystems.Rmd
@@
     fig_height: 6
-    fig_width: 6
     highlight: null
@@
 
-## **Short Title**
-Relict oak resilience at the rear-edge
+### Short title: *Relict oak resilience at the rear-edge*
 
-## Authors [^\*^] 
+### Authors [^\*^] 
 A.J. Pérez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^
 
-^1^ Instituto Interuniversitario de Investigación del Sistema Tierra en Andalucía (CEAMA), Universidad de Granada. Avda. del Mediterráneo s/n, E-18006 Granada, Spain
+^1^Instituto Interuniversitario de Investigación del Sistema Tierra en Andalucía (CEAMA), Universidad de Granada.
+Avda.
+del Mediterráneo s/n, E-18006 Granada, Spain.
+^2^Grupo de Ecología Terrestre, Departamento de Ecología, Facultad de Ciencias, Universidad de Granada, Avda.
+Fuentenueva s/n, E-18071 Granada, Spain.
+^3^INIA-CIFOR.
+Ctra.
+La Coruña km 7.5.
+E-28040 Madrid, Spain
 
-^2^ Grupo de Ecología Terrestre, Departamento de Ecología, Facultad de Ciencias, Universidad de Granada, Avda. Fuentenueva s/n, E-18071 Granada, Spain
-
-^3^ INIA-CIFOR. Ctra. La Coruña km 7.5. E-28040 Madrid, Spain
-
-^a^ ajperez@ugr.es
-
-^b^ gea.guillermo@inia.es
-
-^c^ rzamora@ugr.es
+^a^ ajperez@ugr.es ^b^ gea.guillermo@inia.es ^c^ rzamora@ugr.es
 
@@
 
-## Manuscript highlights
+### Manuscript highlights
 * *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
@@
 
-## Keywords
+### Keywords 
 extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing
 
-# Introduction
+## Introduction
+The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005].
+Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range (*i.e.* centre-periphery hypothesis) [@Sagarin2002; @Pironon2016].
+It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance and thus higher risk of extinction than those populations at the core of the species' range [@VilaCabrera2019; @Pironon2016].
+Nonetheless, recent reviews found species- and population- specific responses highlight that this hypothesis not always fulfill [@Sexton2009; @Abeli2014].
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginallity does not correspond with ecological marginallity [@VilaCabreraJump2019GreaterGrowth].
 
-Resumir párrafos de Vilá-Cabrera (los que ha puesto RZ)
+Climate change is expected to cause important changes in the distribution and abundance of plant communities, and there are already obvious signs of processes associated with increased intensity and duration of droughts [@Allen2010].
+In this respect, several studies have warned on the rapid extinction of plant species in the near future, especially in mountain areas of southern Europe [*e.g.* @Engler2011].
+Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
 
-Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].
+Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; @MartinezVilalta2016].
 There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017].
@@
 In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017].
-This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].
+This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].
 
@@
 Yet it is crucial to accurately characterize the rear-edge of species.
-This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabreraJump2019GreaterGrowth].
+This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019].
 Furthermore, it becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].
 
-Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [*e.g.*, @SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
+Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
 Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973].
@@
 Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018].
-The resilience to drought of species is better assessed when several approaches like remote sensing and dendrocronology are combined simultaneously.
-Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
+The resilience to drought of species is better assessed when several approaches like dendrocronology and remote sensing are combined simultaneously.
+Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual tree-ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
 The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
@@
 Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us drought [*e.g.* @VicenteSerrano2013].
-Drought impacts on vegetation using remote sensing have been analyzed at different scales.
-For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017].
+Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013].
+For instance, a large spatial and seasonal variability of the effects of drought on vegetation has been documented for Mediterranean area [@Gouveia2017].
 Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
@@
 
-# Materials and methods
-
-## Tree species and study site
-
+## Materials and methods 
+### Tree species and study site 
 Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
@@
 
-The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 m and 3482 m *a.s.l.*.
-The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August) but with marked variability set by altitude.
+The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 and 3482 m *a.s.l.*.
+The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability set by altitude.
+Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002].
 There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
-Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002].
 Today, *Q.
@@
 They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG].
-These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2017].
+These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002].
 Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the rear-edge [@GeaIzquierdo2013].
 
-## Drought episodes
-
+### Drought episodes
 The Iberian Peninsula suffered several extreme drought episodes in the last three decades (*e.g*.
 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].
-The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3).
+The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figure S1; Table S3).
 We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below).
@@
 We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html).
-A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017].
+A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017].
 A drought event is considered only when SPEI values are below that threshold for at least two consecutive months.
-Several indicators were computed for each drought event: *duration* as the number of consecutive months with the SPEI lower than a certain threshold; *severity* of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
+For each drought event we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
 
-## Greenness data to assess ecosystem resilience
-
+### Greenness data to assess ecosystem resilience
 Vegetation greenness of *Q.
 pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS (*Moderate Resolution Imaging Spectroradiometer*) sensor.
+EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
 MODIS EVI Data (Collection 6) were obtained for the period 2000 - 2016.
-EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
 We selected the pixels covering the distribution of *Q.
@@
 Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016.
-The EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012].
+The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012].
 For each pixel, an annual EVI value was obtained by averaging EVI valid values.
@@
 where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
-Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI~sa > 1$), "browning" ($EVI~sa <- 1$) or "no changes" ($-1 > EVI~sa > 1$)[@Samanta2012].
+Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].
 
-Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88).
+Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88).
 For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale, using Mann–Kendall nonparametric test.
 
-## Field sampling and dendrochronological methods to assess tree individual resilience
-
+### Field sampling and dendrochronological methods to assess tree individual resilience
 Tree sampling was carried out during autumn 2016.
 Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1).
-For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled aspect Sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
+For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled aspect.
+Sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
 All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015].
@@
 In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius.
-Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].
+Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].
 Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests.
@@
 ### Climate and growth 
-
 Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008].
@@
 We selected grid cells covering each sampled sites.
-The Standardized Precipitation-Evapotranspiration Index (SPEI) with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
+SPEI index with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
 Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].
@@
 ### Disturbance analyses 
-Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g.
-logging, drought induced neighbor mortality) in the past.
-Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:
-
-$$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
-where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
-Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
+Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past.
+Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
+```{r, eval=FALSE, echo=FALSE}
+# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
+```
 Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
@@
 
-## Assessing ecosystem and tree individual resilience to drought
-
-To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*).
-The Resistance index, estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred.
+### Assessing ecosystem and tree individual resilience to drought
+To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011.
+The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred.
 The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity.
@@
 The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2018].
-To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Figure 6, Table S3).
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S4b) and this length was used in other studies [*e.g.* @Gazol2018].
+To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Figure 5, Table S3).
 
-## Statistical analysis
-
+### Statistical analysis 
 For each of the three resilience indices studied we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects).
@@
 
-# Results
+## Results
+### Time trends in vegetation greenness
+The analysis of greenness time trends showed that EVI in 78.9 % of the pixels experienced a positive trend for the 2000-2016 period.
+The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2a).
+A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 2b).
 
-## Time trends in vegetation greenness
+### Analysis of radial growth trends and disturbances 
+The trees of the southern population were older than those from the northern one.
+In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 3) and despite competition measured as plot basal area was greatest in CA-High (Table 1).
+Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S4a).
+Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid, site (CA-High).
 
-The analysis of greenness time trends showed that EVI in 78.9 % of the pixels experienced a positive trend for the 2000-2016 period.
-The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2).
-A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).
-
-## Analysis of radial growth trends and disturbances
-
-The trees of the southern population were older than those from the northern one.
-In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4) and despite competition measured as plot basal area was greatest in CA-High (Table 1).
-Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6).
-Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being much more pronounced for the high-elevation, hence more humid, site (CA-High).
-
-Drought events reduced radial growth for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
+Drought events reduced radial growth for all sites (Figure S2), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
 Tree-growth reduction followed a moisture-gradient.
-The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S3), being the weakest for the wettest site (Figure 5).
+The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2), being the weakest for the wettest site (Figure 4).
 
@@
 
-## Resilience to drought events at the ecosystem and individual tree levels
-
+### Resilience to drought events at the ecosystem and individual tree levels
 *Rs* and *Rt* varied in the same direction whereas *Rc* varied inversely to *Rs* and *Rt*.
-Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 6).
-A similar pattern was obtained for *Rs* but only significant for SJ site (Figure 6).
-Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S8).
-The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), particularly SJ where our results suggest a major release event also after 1995 (Figure 5).
+Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 5).
+A similar pattern was obtained for *Rs* but only significant for SJ site (Figure 5).
+Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S6).
+The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), particularly SJ where our results suggest a major release event also after 1995 (Figure 4).
 
@@
 The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1 and S2) but the metrics of resilience generally covaried in the same direction those two years.
-We obtained significantly higher Resilience and Resistance values for 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7).
-Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event (Table S1, Figure 7).
+We obtained significantly higher Resilience and Resistance values for 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 6).
+Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event (Table S1, Figure 6).
 Recovery showed a contrasting pattern for EVI and tree-growth.
@@
 
-## Tree-growth response to climate
+### Tree-growth response to climate
 The response of tree-growth to water availability was greater than that to temperatures, as it can be expected for a Mediterranean site and particularly a species located closer to its xeric/dry distribution limit (rear-edge).
-Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 8a).
-Yet there were sone differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 8a).
-In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 8b and 8c), whereas minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
+Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 7a).
+Yet there were sone differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
+In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
 
-# Discussion
-## Land-use legacies shape sensitivity to climate change of forests and the present rear-edge
-
+## Discussion
+### Land-use legacies shape sensitivity to climate change of forests and the present rear-edge
 The Mediterranean landscapes have suffered a strong transformation driven by human activities throughout history even at high elevations [@GarciaAlix2017].
@@
 Therefore, it seems logic to think that the geographical limit of the current rear-edge and relict stands of the deciduous oak studied does not necessarily match the potential dsitribution limit related to edaphoclimatic constraints (*i.e.* the ecological rear-edge).
-Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland) [@JimenezOlivencia2015].
+Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
 It has been described a sequential distribution of historical land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming were located at the highest elevations; then forest stands with some croplands; and at the lowest elevations, irrigated terraces with tree crops [@JimenezOlivencia2015].
@@
 On the other hand, woodlands in the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015].
-Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013; @JimenezOlivencia2015].
+Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013].
 
@@
 
-## Relict oaks show high resilience to recent drought events and long-term climatic variability
-
+### Relict oaks show high resilience to recent drought events and long-term climatic variability 
 Severe drought negatively affects both primary and secondary growth of *Q.
@@
 Our results would highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our studied forest.
-The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013] seem to be indicative of high local adaptation of this oak.
+The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013; @Valbuena2017] seem to be indicative of high local adaptation of this oak.
 In addition the local environmental conditions of the sites where this species inhabits at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
@@
 
-## Small-scale environmental variability shapes sensitivity to climate of trees within the rear-edge
-
+### Small-scale environmental variability shapes sensitivity to climate of trees within the rear-edge
 Since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018] a combined analysis as in our study allows to better disentangle the small and large-scale variability in growth induced by climate from that driven by community-level ecological processes [@Babst2017].
@@
 
-# Conclusions 
+## Conclusions 
 The ecological and geographical read-edges did not seem to fully match in our study.
@@
 
-# Acknowledgements
-
-We would like to thank Sammy Lucy Behle, M.
-Suárez-Muñoz and F.J.
-Bonet-García for their assistence in the field monitoring.
+## Acknowledgements
+We would like to thank to Sammy L.
+Behle and M.
+Suárez-Muñoz for their assistance in the field.
 A.
@@
 We also are very thankful to F.J.
-Bonet-García for his valuable comments in an earlier version of the manuscript.
+Bonet-García for his comments in earlier version of the manuscript and field assistance.
 AJPL wants to thank the invaluable support received from his family over the years.
-Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work.
-This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project).
+This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project).
 We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.
 
-# References
+## References

### 2019-10-28 man/ecosystems/ms_ecosystems.Rmd
update

@@
 
-Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014].
+Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [@Herrero2014].
 Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973].
@@
 Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual tree-ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
+Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013].
+Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such as drought.
 The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
-Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013].
-Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us drought [*e.g.* @VicenteSerrano2013].
-Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013].
-For instance, a large spatial and seasonal variability of the effects of drought on vegetation has been documented for Mediterranean area [@Gouveia2017].
 Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
@@
 We used remote sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of *Quercus pyrenaica* Willd.
-on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
+(*melojo oak*) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
 We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) in the long-term in the last decades.
@@
 pyrenaica* forests in its rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) within the rear edge to detect vulnerability to climate change along climatic gradients at the small spatial scale.
-To achieve goal (*iii*) and characterize variability in the forest response to drought within the rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
+To achieve goal (*iii*) and characterize variability in the forest response to drought within the rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.
 
@@
 ### Tree species and study site 
-Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
+*Melojo* oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
 In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate.
@@
 Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002].
-There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
+There are eight melojo oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
 Today, *Q.
@@
 We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below).
-Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3).
+Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and to analyze forest resilience to drought in a longer term (see Table S3).
 The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015.
@@
 Vegetation greenness of *Q.
-pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS (*Moderate Resolution Imaging Spectroradiometer*) sensor.
+pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor.
 EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
-MODIS EVI Data (Collection 6) were obtained for the period 2000 - 2016.
+MODIS EVI Data were obtained for the period 2000 - 2016.
 We selected the pixels covering the distribution of *Q.
@@
 For each pixel, an annual EVI value was obtained by averaging EVI valid values.
-Then, the standardized anomaly was computed as:
-
-$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
+Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$,
 where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
@@
 The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S4b) and this length was used in other studies [*e.g.* @Gazol2018].
+A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3b) and this length was used in other studies [*e.g.* @Gazol2018].
 To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Figure 5, Table S3).
@@
 In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 3) and despite competition measured as plot basal area was greatest in CA-High (Table 1).
-Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S4a).
+Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S3a).
 Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid, site (CA-High).
 
-Drought events reduced radial growth for all sites (Figure S2), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
+Drought events reduced radial growth for all sites (Figure S2a), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
 Tree-growth reduction followed a moisture-gradient.
-The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2), being the weakest for the wettest site (Figure 4).
+The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest for the wettest site (Figure 4).
 
@@
 The release event in 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990].
-This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S4).
-The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S4).
+This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S2b).
+The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S2b).
 There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J.

### 2019-10-28 man/figures/0_figures_ecosystems.Rmd
update

@@
 
-**Figure S2.** Residual tree-ring chronologies obtained for the *Q.
+**Figure S2.** ***A*** Residual tree-ring chronologies obtained for the *Q.
 pyrenaica* sites.
@@
 Dotted black lines showing the severe drought years identified in our climatic data (Table S3 and Figure S1).
+***B*** Percentage of *Q.
+pyrenaica* trees affected by GC > 50 % by site.
+*Black* line shows number of trees (rigth-axis).
+Data for number of trees > 2 is shown.
 
-![](../../man/figures/svg/fig_s3.jpg)
+![](../../man/figures/svg/fig_s2ab.jpg)
 
@@
 
-**Figure S3.** Percentage of *Q. pyrenaica* trees affected by GC > 50 % by site. *Black* line shows number of trees (rigth-axis). Data for number of trees > 2 is shown.
-
-![](../../man/figures/svg/fig_s4.jpg)
-
-#####
-
-**Figure S4.** ***A*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
+**Figure S3.** ***A*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
 Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
@@
 ##### 
-**Figure S5.** EVI annual profile (average of the period 2000-2016) for *Q.
+**Figure S4.** EVI annual profile (average of the period 2000-2016) for *Q.
 pyrenaica* forests in Sierra Nevada and drought events.
@@
 
-**Figure S6.** Resilience metrics of the tree-growth for severe drought events since 1950 (excluding 1995 drought event).
+**Figure S5.** Resilience metrics of the tree-growth for severe drought events since 1950 (excluding 1995 drought event).
 *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).

### 2019-10-28 man/figures/s2_combined.Rmd
update

---
output: 
  word_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r load_pkg}
library("tidyverse")
library("here")
library("ggrepel")
library("scales")
library("grid")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
library("stringr")
library("pander")
library("lemon")
library("ggpubr")
source(here::here('/script/R/custom_functions.R'))

```

```{r}
cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')

# Reorder 
cro_rwi <- cro_rwi %>% mutate(site_sorted = 
                                recode_factor(site,
                                              "caH" = "0_caH", 
                                              "caL" = "1_caL", "sj" = "2_sj"))

year_cutoffs <- year_cutoffs %>% mutate(site_sorted = 
                                recode_factor(site,
                                              "caH" = "0_caH", 
                                              "caL" = "1_caL", "sj" = "2_sj"))

year_sampledepth <- year_sampledepth %>% mutate(site_sorted = 
                                   recode_factor(site,
                                              "caH" = "0_caH", 
                                              "caL" = "1_caL", "sj" = "2_sj"))

# peores sequias 
dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
dy <- dy %>% arrange(desc(d_duration))
# solamente aquellos mayor de 2 meses 
dys <- dy %>% filter(d_duration > 2) %>% as.data.frame() 
dys <- dys[,'maxyear'] 
```

```{r plot_rwi}
label_site <- c("2_sj" = "SJ",
                "0_caH" = "CA-High",
                "1_caL" = "CA-Low")

# 
plot_rwi_crono <- ggplot(cro_rwi, aes(x=year, y=res, colour=site)) +
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(face="bold", size=9), 
        # strip.text = element_blank(), 
        axis.text = element_text(size=8), 
        axis.title = element_text(face="bold", size=9), 
        legend.position = "none") + 
  geom_hline(yintercept = 1, colour = "grey") + 
  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) +
  facet_wrap(~site_sorted, scales = "free_x", ncol = 1,
             labeller = as_labeller(label_site)) + 
  ylab("Ring width index") + xlab("Year") + 
  geom_vline(aes(xintercept = year), data = year_cutoffs,  linetype = "dashed", colour='red') + 
  # geom_vline(aes(xintercept = year), data = year_sampledepth,  linetype = "dashed", colour='blue') +
  geom_vline(xintercept = dys, linetype = "dotted") + 
  # geom_vline(xintercept = dys, linetype = "dotted") + 
  scale_x_continuous(limits = c(1819, 2016), 
                     breaks= seq(1820, 2020, by=20)) +
  scale_colour_manual(values = colores_den)

```

```{r}
gc_treesMed <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median.csv"), header = TRUE, sep = ',')
gc_treesMed50 <- read.csv(here("data/disturbance", "disturbance_trees_gc_treshold_median50.csv"), header = TRUE, sep = ',')

gc_treesMed50 <- gc_treesMed50 %>% mutate(site_sorted = 
                                recode_factor(site,
                                              "gc_caH" = "0_caH", 
                                              "gc_caL" = "1_caL", "gc_sj" = "2_sj"))

label_site <- c("2_sj" = "SJ",
                "0_caH" = "CA-High",
                "1_caL" = "CA-Low")

```

```{r plotGC}

colorpositivo <- c("#104E8B")
colornegativo <- c("#CD3333")

# Get min and max year and round to nearest decade
getBreaks <- function(df, v, by){ 
  x <- seq(round(min(df[v]) / 10)*10, 
           round(max(df[v]) / 10)*10, by= by) 
  x
  }

xrange <- getBreaks(gc_treesMed50, 'year', by=10)

dat_text <- data.frame(
  label_site = c("CA-Low","CA-High","SJ"),
  site_sorted = c("0_caH","1_caL","2_sj"),
  year = c(1835, 1835, 1835),
  y = c(90, 90, 90))

pgc_median50 <- gc_treesMed50 %>% 
  filter(total_tree > 2) %>% 
  ggplot() +
  geom_bar(aes(x=year, y=per_pos), fill=colorpositivo, colour=colorpositivo, 
           stat = "identity", position = "dodge") + 
  geom_bar(aes(x=year, y=(-1)*per_neg), fill=colornegativo, colour=colornegativo, 
           stat = "identity", position = "dodge") +
  geom_line(aes(x=year, y=total_tree*2), colour = 'black', size = 0.2) + 
  theme_bw() + ylab('% of trees') + xlab("Year") + 
  geom_hline(yintercept=0, alpha=.4) +
  geom_hline(yintercept=c(50, -50), linetype=2, alpha=.4) +
  scale_x_continuous(breaks = xrange) +
  scale_y_continuous(limits = c(-100, 100), breaks = seq(-100, 100, by = 50),
                     sec.axis = sec_axis(~., name = '# trees', breaks = c(0,15,30))) +
  facet_rep_wrap(~site_sorted, nrow = 3, 
                 repeat.tick.labels = TRUE, labeller = as_labeller(label_site)) +
  geom_text(data= dat_text, aes(label = label_site), 
            x = dat_text$year, y = dat_text$y,
            size = 3, fontface = "bold") + 
  theme(panel.grid = element_blank(), 
        strip.background = element_blank(), 
        strip.text = element_blank(),
        axis.text.y.left = element_text(size=6),
        axis.text.y.right  = element_text(size=5),
        axis.text.x = element_text(size=6),
        axis.title = element_text(size = 9))

```

```{r}
s34 <- ggarrange(plot_rwi_crono, pgc_median50,
          labels = c("a", "b"),
          ncol = 1, nrow = 2)
```

```{r}
# SAVE as eps 
saveImage(s34, path="./man/figures/svg/", dpi=300, w=25, h=20, fig= "fig_s2ab", u = "cm") 
```

### 2019-11-05 man/ecosystems/ms_ecosystems.Rmd
update text

@@
 ### Tree species and study site 
-*Melojo* oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
+*Quercus pyrenaica* Willd.
+(*melojo* oak) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
 In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate.
@@
 EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
-MODIS EVI Data were obtained for the period 2000 - 2016.
+MODIS EVI data were obtained for the period 2000 - 2016.
 We selected the pixels covering the distribution of *Q.
@@
 For each pixel, an annual EVI value was obtained by averaging EVI valid values.
-Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$,
+Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$,
 where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
@@
 Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past.
-Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
+Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
 ```{r, eval=FALSE, echo=FALSE}
@@
 A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3b) and this length was used in other studies [*e.g.* @Gazol2018].
-To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Figure 5, Table S3).
+To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Table S3).
 
@@
 The trees of the southern population were older than those from the northern one.
-In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 3) and despite competition measured as plot basal area was greatest in CA-High (Table 1).
+In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites and despite competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3).
 Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S3a).
@@
 Tree-growth reduction followed a moisture-gradient.
-The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest for the wettest site (Figure 4).
+The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest for the wettest site (Figure 3).
 
-The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940s (the most evident) and the second in 1995-2000.
+The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940s (the most evident) and the second in 1995-2000 (Figure 4).
 These periods alternated with periods of suppression.
@@
 Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 5).
-A similar pattern was obtained for *Rs* but only significant for SJ site (Figure 5).
-Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S6).
-The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), particularly SJ where our results suggest a major release event also after 1995 (Figure 4).
+A similar pattern was obtained for *Rs* but only significant for SJ site.
+Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S5).
+The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4).
 
 During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table 3).
-The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1 and S2) but the metrics of resilience generally covaried in the same direction those two years.
-We obtained significantly higher Resilience and Resistance values for 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 6).
+The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1-S2) but the metrics of resilience generally covaried in the same direction those two years.
+We obtained significantly higher resilience and resistance values for 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 6).
 Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event (Table S1, Figure 6).
@@
 
-The resilience metrics calculated significantly varied between sites, except for Resilience of tree-growth (p = 0.534; Table 3), which was similar among the three sites.
-The two southern populations showed lower values for Recovery than the northern site both for greenness and tree-growth (Tables S1 and S2) but Resistance and Resilience values were significantly higher for the southern site (Tables S1 and S2).
+The resilience metrics calculated significantly varied between sites, except for resilience of tree-growth (p = 0.534; Table 3), which was similar among the three sites.
+The two southern populations showed lower values for recovery than the northern site both for greenness and tree-growth but resistance and resilience values were significantly higher for the southern site (Tables S1-S2).
 
@@
 Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 7a).
-Yet there were sone differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
+Yet there were some differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
 In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
@@
 The review of historical documents shows how forest clearings, firewood extraction, charcoal exploitations and mining have strongly impacted the forests at Sierra Nevada (Table S4) where in addition it has been estimated a historical loss of about 90% of broadleaf *Quercus* species cover from medium and low elevations [@JimenezOlivencia2015].
-Therefore, it seems logic to think that the geographical limit of the current rear-edge and relict stands of the deciduous oak studied does not necessarily match the potential dsitribution limit related to edaphoclimatic constraints (*i.e.* the ecological rear-edge).
+Therefore, it seems logic to think that the geographical limit of the current rear-edge and relict stands of the deciduous oak studied does not necessarily match the potential distribution limit related to edaphoclimatic constraints (*i.e.* the ecological rear-edge).
 Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
@@
 Some of the previous historical records of land-use were likely expressed by the growth chronologies.
-The release event in 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral [@Titos1990].
+The release event in 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral.
 This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S2b).
-The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 5, S2b).
+The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 4, S2b).
 There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J.
@@
 Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic precipitation negative anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
-Although the south of Iberian Peninsula suffered severe drought events in the last decades (Table S3), we found a positive trend for vegetation greenness of *Q.
+Despite the severe drought events suffered in the last decades (Table S3), we found a positive trend for vegetation greenness of *Q.
 pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
@@
 This could be explained by the different timing of the two droughts.
-The 2012 drought was a winter drought [@Trigo2013] occurring earlier than the 2005 drought and although the latter lasted less than the 2012 drought, it matched the period of maximum growth for *Q.
-pyrenaica* forests in late spring (Figure S7).
-Thus, as bserved for other species [*e.g.* Holm oak; @Camarero2015b] this highlights the importance of the timing of drought as a key factor determining the tree recovery after drought [@Huang2018].
-For tree-growth, the highest values of resilience were obtained for the two most severe events, namely 1995 and 1999 (Table S3) and we observed a positive relationship between the tree-growth resilience and drought severity (Figures 6).
+The 2012 drought was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought.
+The latter matched the period of maximum growth for oak forests in late spring (Figure S4).
+This highlights the importance of the timing of drought as a key factor determining the tree recovery after drought [@Camarero2015b; @Huang2018].
+For tree-growth, the highest values of resilience were obtained for the two most severe events (1995 and 1999)(Table S3) and we observed a positive relationship between the tree-growth resilience and drought severity (Figure 5).
 This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events.
@@
 In addition the local environmental conditions of the sites where this species inhabits at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
+All this, together with the characteristic resprouting ability of the species from its stoloniferous root system, would suggest a long-term persistence of those populations.
 It needs to be mentioned that we only studied adult individuals established decades or centuries ago, meaning that it needs to be assessed if the species expresses resilience or to the contrary vulnerability if we analyzed seedling regeneration, likewise many other Mediterranean species at their xeric limit (2-3 CITAS).
 A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018; @VilaCabrera2019].
-In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q.
+In this sense, the high values of resilience observed, could suggest that the populations of *Q.
 pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge.
 And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges.
-Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* Pinus sylvestris, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.
+Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* *Pinus sylvestris*, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.
 
@@
 pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
-Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @GeaIzquierdo2011; @GarciaGonzalez2017].
+Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
 
@@
 We would like to thank to Sammy L.
-Behle and M.
-Suárez-Muñoz for their assistance in the field.
+Behle, F.J.
+Bonet and M.
+Suárez-Muñoz for field assistance.
 A.
@@
 Navarro provided worthwhile information about oak management projects in Sierra Nevada.
-We also are very thankful to F.J.
-Bonet-García for his comments in earlier version of the manuscript and field assistance.
+We are grateful for the comments made by F.J.
+Bonet in an earlier version of the manuscript.
 AJPL wants to thank the invaluable support received from his family over the years.
-This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project).
-We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support.
+This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” project.
+We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (No 641762), for their funding support.
 

### 2019-11-05 man/figures/0_figures_ecosystems.Rmd
update

@@
 
-**Figure 2.** ***a*** EVI standardized anomaly during the period 2000-2016 for northern and southern populations.
+**Figure 2.** ***a)*** EVI standardized anomaly during the period 2000-2016 for northern and southern populations.
 Error bars show standard error.
 See main text for details on EVI calculation.
-***b*** Percentage of pixels showing browning, greening or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
+***b)*** Percentage of pixels showing browning, greening or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
 See main text for an explanation of greening and browning.
@@
 
-**Figure S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
+**Figure S1.** ***a)*** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
 Points represent mean and errorbars standard error.
@@
 Data from meteorological station Granada, Base Aérea.
-***B*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+***b)*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
 Data from Global SPEI database (http://spei.csic.es/database.html).
@@
 
-**Figure S2.** ***A*** Residual tree-ring chronologies obtained for the *Q.
+**Figure S2.** ***a)*** Residual tree-ring chronologies obtained for the *Q.
 pyrenaica* sites.
@@
 Dotted black lines showing the severe drought years identified in our climatic data (Table S3 and Figure S1).
-***B*** Percentage of *Q.
+***b)*** Percentage of *Q.
 pyrenaica* trees affected by GC > 50 % by site.
@@
 
-**Figure S3.** ***A*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
+**Figure S3.** ***a)*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
 Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
 Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot`.
-***B*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
+***b)*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
 

### 2019-11-05 man/figures/resilience_long_term.Rmd
update

@@
 
-res_long_term_sites_no95 <-  g + geom_text(data = dfauxno95, 
-              aes(x = 12, y = 0.3, label = eq), 
+res_long_term_sites_no95 <-  g + 
+  geom_text(data = dfauxno95, 
+            aes(x = 12, y = 0.2, label = eq), 
             inherit.aes = FALSE, parse = TRUE, 
-              col = "gray40")
+            col = "gray40") + 
+  xlim(5,25) + ylim(0.1,3.5)
 ```

### 2019-11-25 man/ecosystems/ms_ecosystems.Rmd
update infinite version

@@
 * *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
-* The responses to climate and resilience followed a water-stress gradient
+* The growth response to climate and resilience followed a water-stress gradient
 * Trees and stand expressed high sensitivity to drought and land-use legacies
@@
 * The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure.
-Trees were highly sensitive to moisture availability and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term.
+Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events in the short- and the long-term.
 The response to climate and resilience followed a water-stress gradient.
-Primary and secondary growth were generally less resilient to drought in xeric than in mesic sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and more humid) high-elevation stands, but absent in the most xeric site.
+Primary and secondary growth were generally less resilient to drought in mesic than in xeric sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and more humid) high-elevation stands, but absent in the most xeric site.
 
-* Adult trees of the studied species showed high resilience to drought despite being considered relict and at its rear-edge.
-Yet, the trees exhibited a variable response to the changing climate along a very short climatic gradient and showed that the ecological and geographical rear-edges do not need to meet today in forests with strong land-uses legacies.
-Forest resilience including that of regeneration needs to be monitored at xeric locations (*i.e.* the low altitudinal rear-edge) under future expected climatic scenarios bearing in mind that land-use legacies play a key role in stand dynamics.
+* Adult oak trees showed high resilience to drought despite being considered relict and at its rear-edge.
+The trees exhibited a variable response to the changing climate along a very short climatic gradient and showed that the ecological and geographical rear-edges do not need to meet today in forests with strong land-uses legacies.
+Forest resilience including that of regeneration needs to be monitored at xeric of the species under future expected climatic scenarios bearing in mind that ecological boundaries for species distribution can be narrow and that land-use legacies can play a key role in stand dynamics.
 
@@
 ## Introduction
-The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005].
+The response of species to changing environments is likely to be largely determined by population responses at range margins [@Hampe2005].
 Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range (*i.e.* centre-periphery hypothesis) [@Sagarin2002; @Pironon2016].
@@
 Nonetheless, recent reviews found species- and population- specific responses highlight that this hypothesis not always fulfill [@Sexton2009; @Abeli2014].
-In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginallity does not correspond with ecological marginallity [@VilaCabreraJump2019GreaterGrowth].
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].
 
 Climate change is expected to cause important changes in the distribution and abundance of plant communities, and there are already obvious signs of processes associated with increased intensity and duration of droughts [@Allen2010].
-In this respect, several studies have warned on the rapid extinction of plant species in the near future, especially in mountain areas of southern Europe [*e.g.* @Engler2011].
-Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
-
-Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; @MartinezVilalta2016].
-There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017].
+In fact, there has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017].
 This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a].
+Under this climatic change context, it is reasonable to expect that population loss and range retractions in boreal, temperate and Mediterranean species should be seen in the lowest latitudes, altitudes and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
 In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017].
@@
 The impacts of drought are also especially relevant for populations considered to be located in the rear-edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005].
-The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010].
-Yet it is crucial to accurately characterize the rear-edge of species.
+The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of species to predicted climatic changes [@Jump2010].
+Yet it is crucial to accurately characterize the ecological rear-edge of species.
 This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019].
@@
 
-Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [@Herrero2014].
-Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973].
+Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018].
 The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and provide valuable ecosystem services.
-
-Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018].
 The resilience to drought of species is better assessed when several approaches like dendrocronology and remote sensing are combined simultaneously.
@@
 Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013].
-Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such as drought.
-The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018].
-Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
+The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].
 Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.
 
-We used remote sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of *Quercus pyrenaica* Willd.
+In this study we used remote sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of *Quercus pyrenaica* Willd.
 (*melojo oak*) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
@@
 pyrenaica* forests in its rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) within the rear edge to detect vulnerability to climate change along climatic gradients at the small spatial scale.
-To achieve goal (*iii*) and characterize variability in the forest response to drought within the rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.
+To achieve goal (*iii*) and characterize variability in the forest response to drought within the current geographical rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.
+
+```{r, eval=FALSE, echo = FALSE}
+# Retales de la intro 
+# Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
+
+# In this respect, several studies have warned on the rapid extinction of plant species in the near future, especially in mountain areas of southern Europe [*e.g.* @Engler2011].
+
+# Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
+
+# Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; @MartinezVilalta2016].
+
+# Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [@Herrero2014].
+Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973].
+```
 
@@
 ### Tree species and study site 
-*Quercus pyrenaica* Willd.
-(*melojo* oak) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
+*Quercus pyrenaica* forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
 In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate.
@@
 
-The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 and 3482 m *a.s.l.*.
+The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations up to 3482 m *a.s.l.*.
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability set by altitude.
 Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002].
-There are eight melojo oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
+There are eight melojo oak patches (2400 ha) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
 Today, *Q.
@@
 These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002].
-Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the rear-edge [@GeaIzquierdo2013].
+Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].
 
@@
 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].
-The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figure S1; Table S3).
+The 2005 and 2012 drought events have been documented among the worst in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figure S1; Table S3).
 We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below).
-Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and to analyze forest resilience to drought in a longer term (see Table S3).
+Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 and to analyze forest resilience to drought in a longer term (see Table S3).
 The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015.
@@
 Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016.
-The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012].
+The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about their magnitude [@Samanta2012].
 For each pixel, an annual EVI value was obtained by averaging EVI valid values.
@@
 Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1).
-For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled aspect.
-Sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
+For the southern site two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), which constitute the current low altitudinal limit (CA-Low) and the tree-line (CA-High), respectively in the sampled aspect.
+Despite the proximity of these two elevations (less than 200 m difference in altitude) the stands greatly differ in their structure and characteristics (Table 1).
+The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
 All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015].
-In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected.
+In each site between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly selected.
 Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer.
@@
 Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976].
-The statistical quality of each chronology was checked via the expressed population signal.
+The statistical quality of each chronology was checked via the expressed population signal (EPS).
 A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.
@@
 Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past.
-Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
+Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@Rubino2004].
 ```{r, eval=FALSE, echo=FALSE}
@@
 In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
-For this purpose, an exhaustive review of existing abundant historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).
+For this purpose, an exhaustive review of existing historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).
+We reviewd several documentary sources: historical documents and maps; detailed mining reports; official information about recent wildfires events and forest management practices; livestock farming; traditional irrigation ditches and other studies concerning the socioeconomic dynamics of forest of Sierra Nevada at different scales (see Table S4 for references).
 
@@
 A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3b) and this length was used in other studies [*e.g.* @Gazol2018].
-To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Table S3).
+Resilience metrics for BAI data were additionally computed for the most severe drought events since 1940 (*n* = 8; Table S3) and compared with drought severity.
 
@@
 Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S3a).
-Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid, site (CA-High).
+Only trees from the southern sites (*i.e.* the more humid aspect) showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid and colder, site (CA-High).
 
@@
 Tree-growth reduction followed a moisture-gradient.
-The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest for the wettest site (Figure 3).
+The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest growth reductions in trees from the wettest site (Figure 3).
 
@@
 Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 5).
-A similar pattern was obtained for *Rs* but only significant for SJ site.
+A similar pattern was obtained for *Rs* but only significant for SJ.
 Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S5).
@@
 ### Tree-growth response to climate
-The response of tree-growth to water availability was greater than that to temperatures, as it can be expected for a Mediterranean site and particularly a species located closer to its xeric/dry distribution limit (rear-edge).
+The response of tree-growth to water availability was greater than that to temperatures.
 Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 7a).
-Yet there were some differences between northern and southern populations: the positive relationship with SPEI was highest in the northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
+Yet there were some differences between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
 In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
@@
 ## Discussion
+
+By using a combined approach of remote sensing information and dendrocronology we have quantified the drought impacts on melojo oak forests on Sierra Nevada and its resilience to several severe drought events in the last decades.
+Our results clearly indicated that these relict oak populations driven by historical land-use are resilient to climate change at their present (either altitudinal or latitudinal) rear-edge.
+However, resistance, resilience and forest recovery to extreme drought events are strongly influenced by mountain aspect, local environmental conditions, and management legacies.
+
 ### Land-use legacies shape sensitivity to climate change of forests and the present rear-edge
-The Mediterranean landscapes have suffered a strong transformation driven by human activities throughout history even at high elevations [@GarciaAlix2017].
 The review of historical documents shows how forest clearings, firewood extraction, charcoal exploitations and mining have strongly impacted the forests at Sierra Nevada (Table S4) where in addition it has been estimated a historical loss of about 90% of broadleaf *Quercus* species cover from medium and low elevations [@JimenezOlivencia2015].
-Therefore, it seems logic to think that the geographical limit of the current rear-edge and relict stands of the deciduous oak studied does not necessarily match the potential distribution limit related to edaphoclimatic constraints (*i.e.* the ecological rear-edge).
 Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
-It has been described a sequential distribution of historical land uses along the elevational gradient in the northern slopes of Sierra Nevada (*e.g.* SJ site): grasslands and shrublands for cattle farming were located at the highest elevations; then forest stands with some croplands; and at the lowest elevations, irrigated terraces with tree crops [@JimenezOlivencia2015].
-In addition, other activities like mining should have modified the forest structure at that northern location where there is with a high concentration of mines and quarries that were exploited intermittently throughout history.
+In the northern slopes of Sierra Nevada (*e.g.* SJ site), land uses have been historically distributed along an altitudinal gradient: grasslands and shrublands for cattle farming at the highest elevations; then forest stands with some croplands; and irrigated terraces with tree crops at the lowest elevations [@JimenezOlivencia2015].
+In addition, other activities like mining should have modified the forest structure, *e.g.* SJ site contains a high concentration of small mines an quarries that were exploited intermittently throughout history.
+The release growth event observed in 1940s concurs with a period of maximum mining activity in this area (1925 to 1957) (Table S4), during wich there was an increase in the use of timbers for mine tunnels and furnaces, that also required large amounts of fuelwood to melt the mineral.
+This heavily exploitation of the neighbouring forest resources should have affected a significant part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S2b).
 On the other hand, woodlands in the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015].
 Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013].
+For CA-High site, the only positive release event showed at the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
 
-Some of the previous historical records of land-use were likely expressed by the growth chronologies.
-The release event in 1940s in the northern site concurs with one period of maximum mining activity in this area (1925 to 1957) (Table S4), with an increase on the use of timbers for the mine tunnels and furnaces that additionally required great quantities of fuelwood to melt the mineral.
-This heavily exploitation of the neighbouring forest resources should have affected a major part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S2b).
 The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 4, S2b).
-There are no records of forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest] and no record for any logging  during 1995 - 2000 period (F.J.
+No records of forest practices carried out in this area in the last 30 years were found [@Bonet2016obsnev_forest], and no logging was recorded during 1995 - 2000 (F.J.
 Cano-Manuel *personal communications*).
-Therefore this release might be related with a natural drought-induced mortality event after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
+Therefore this release might be related to natural drought-induced mortality after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
 On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014].
-For southern sites, the only positive release event showed at the earliest years for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
-In this site we observed an increase in growth especially in high elevation trees, where moisture is less limiting, a pattern which concurs with a forest expansion into marginal abandonded croplads and an increase in tree-cover [@CamachoOlmedo2002].
 
@@
 
-### Relict oaks show high resilience to recent drought events and long-term climatic variability 
+### Relict oaks sensitivity (resistance) to recent drought events
 Severe drought negatively affects both primary and secondary growth of *Q.
@@
 Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic precipitation negative anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
+Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
+pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
+Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
+
+Greenness exhibited a lower sensitivity to drought than tree-growth, particularly for drier sites.
+These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018] and would suggest that the growth reduction could more mediated by sink that by source limitations [@Korner2013; @Fatichi2014].
+The use of tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
+
+##### Small-scale environmental variability shapes sensitivity to climate of trees within the rear-edge
+Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event and the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014].
+It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competition, etc.) [@Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
+Trees at CA-High sowhed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4).
+This shows the high variability in the response to climate exhibited in a small gradient, which is especially interesting for southern sites, which are very close to each other, and overall both are considered to constitute the rear-edge for the species.
+
+```{r eval=FALSE, echo=FALSE}
+#The higher values of tree growth in taller trees that we observed in CA-High site strongly suggests lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017].
+Higher elevations within warm and dry ecosystems can help species alleviate climatic stress and increase resistance to drought and stands of several tree-species located at higher elevations were less drought sensitive.
+Thus high-elevations are generally considered as potential local refuge for species in response to climate change [@Bhuyan2017].
+Finally and in a context of global change, in addition to withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).
+```
+
+### Relict oaks show high resilience to recent drought events and long-term climatic variability 
 Despite the severe drought events suffered in the last decades (Table S3), we found a positive trend for vegetation greenness of *Q.
@@
 For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends were described for this species along its distribution range only at high elevation sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014].
-Importantly, for rear-edges where climate change is a threat we would have expected negative growth trends as shown for temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
-Therefore there is much small-scale variability in the response to climate within the rear-edge we had "a priori" considered in our study, where we even observed positive effects of climate change in certain stands.
-This suggests that the rear-edge needs to be redefined in consequence [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear-edge) and the potential ecological (limiting) rear-edge of species.
+Similar long-term trends were described for this species along its distribution range only at high elevation humid and cold sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014].
+Importantly, for rear-edges where climate change is a threat we would have expected negative growth trends as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
 
-Despite that 2012 drought event was more severe and intense than that in 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012.
+Albeit that 2012 drought event was more severe and intense than that in 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012.
 This could be explained by the different timing of the two droughts.
@@
 For tree-growth, the highest values of resilience were obtained for the two most severe events (1995 and 1999)(Table S3) and we observed a positive relationship between the tree-growth resilience and drought severity (Figure 5).
-This result seems to contradict our initial hypothesis in which we predicted low resilience values for this oak species in their rear-edge after extreme drought events.
-Yet, the high resilience values obtained for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality (F.J.
-Navarro *personal communications*), which would show the influence of stand dynamics partly shadowing interpretation of tree resilience estimated from individual growth series.
 
-Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of species [@Hampe2005].
+The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013; @Valbuena2017] seem to be indicative of high local adaptation of this oak.
+Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017].
+@MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations.
+Our results would highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our studied forest.
+All this, together with the characteristic high resprouting ability of the species, would suggest a long-term persistence of those populations [@Bellingham2000].
+It needs to be mentioned that we only studied adult individuals established decades or centuries ago, meaning that it needs to be assessed if the trees express resilience at the species level or to there is vulnerability if we analyzed seedling regeneration, likewise other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].
+
+```{r eval=FALSE, echo=FALSE}
+#Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of species [@Hampe2005].
 A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018].
@@
 pyrenaica* and high values of resilience to severe droughts.
-Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017].
-@MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability showed by rear-edge populations.
-Our results would highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our studied forest.
-The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013; @Valbuena2017] seem to be indicative of high local adaptation of this oak.
-In addition the local environmental conditions of the sites where this species inhabits at Sierra Nevada would explain the low vulnerability to drought showed by this oak.
-All this, together with the characteristic resprouting ability of the species from its stoloniferous root system, would suggest a long-term persistence of those populations.
-It needs to be mentioned that we only studied adult individuals established decades or centuries ago, meaning that it needs to be assessed if the species expresses resilience or to the contrary vulnerability if we analyzed seedling regeneration, likewise many other Mediterranean species at their xeric limit (2-3 CITAS).
-A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018; @VilaCabrera2019].
-In this sense, the high values of resilience observed, could suggest that the populations of *Q.
-pyrenaica* in Sierra Nevada are located in a geographical but not a climatic rear-edge.
-And the current niche is a result of land-use changes, which further complicates the definition of potential rear-edges.
-Similarly, high values of resilience showed by other Mediterranean species located in their rear-edge [*e.g.* *Pinus sylvestris*, @Herrero2014] seem to be related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events.
+```
 
-### Small-scale environmental variability shapes sensitivity to climate of trees within the rear-edge
-Since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018] a combined analysis as in our study allows to better disentangle the small and large-scale variability in growth induced by climate from that driven by community-level ecological processes [@Babst2017].
-Greenness exhibited a lower sensitivity to drought than tree-growth, particularly for drier sites.
-These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018] and would suggest that the growth reduction could more mediated by sink that by source limitations [@Korner2013; @Fatichi2014].
-On the other hand, tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
+Summarizing, two mains results stand out from this research.
+First, the high values of resilience observed in our study suggest that the populations of *Q.
+pyrenaica* in Sierra Nevada are located in a geographical but not a climatic, ecological rear-edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019].
+The current niche is a result of land-use changes, which further complicates the definition of potential rear-edges.
+The high resilience values observed could be also related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events, as it has been described for other species [*e.g.* *Pinus sylvestris*, @Herrero2014].
 
-Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event and the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014].
-Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
-pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
-Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
+Second, these resilience responses of oak forest to drought events are not spatially homogeneous around all mountain range, due to differences in ecological conditions and/or past management legacies.
+In fact, there was much small-scale variability in the response to climate within the rear-edge that we had "a priori" considered in our study.
+Furthermore we even observed positive effects of climate change in certain stands, as discussed, and which would be in disagreement with our hypothesis expecting oak vulnerability in the studied geographical rear-edge.
+This suggests that the rear-edge needs to be redefined in consequence [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear-edge) and the ecological (limiting) rear-edge of species.
 
-Higher elevations within warm and dry ecosystems can help species alleviate climatic stress and 
-increase resistance to drought and stands of several tree-species located at higher elevations were less drought sensitive.
-Thus high-elevations are generally considered as potential local refuge for species in response to climate change [@Bhuyan2017].
-It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
-Trees at CA-High sowhed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4).
-This shows the high variability in the response to climate exhibited in a small gradient, which is especially interesting for southern sites, which are very close to each other, and overall both constitute the rear-edge for the species.
-The higher values of tree growth in taller trees that we observed in CA-High site strongly suggests lower water stress at high-elevation sites [as shown by soil analyses along an elevational gradient in this location @CoboDiaz2017].
-Furthermore, as mentioned, the recent growth increase only at the most humid locations a likely concomitant positive effect of a warming climate [@GeaIzquierdo2014].
-Finally and in a context of global change, in addition to withstanding lower water stress, it is likely that the less accessible sites at higher elevations would suffer a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4).
 Overall, our results show how management history conditions tree-growth and resilience to climate change of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017] which, hence, will also strongly determine the current geographical distribution of marginal stands, regardless of the potential extent of ecological marginality of species.
@@
 ```{r eval=FALSE, echo=FALSE}
+
+# Since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [@GeaIzquierdo2015; @Babst2018] a combined analysis as in our study allows to better disentangle the small and large-scale variability in growth induced by climate from that driven by community-level ecological processes [@Babst2017].
+
 # In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply.
@@
 These results would be consistent with other research showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*.
-The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent and can vary in very small gradients (*i.e.* following ecological thresholds).
+The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent and can vary in very small spatial gradients (*i.e.* following ecological thresholds).
 This is particularly relevant for rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
@@
 This is relevant for tree species with a high sensitivity to climate change, such us *Q.
-pyrenaica*, not only for conservation per se of the species, but for all ecosystem services that these singular forests offer at Mediterranean mountain.
-Following our study it is necessary to further disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.
+pyrenaica*, not only for conservation per se of the species, but for all ecosystem services that these singular forests offer.
+Following our study it is necessary to disentangle how forest resistance and resilience are affected by the combination of climate change and land-use legacies in highly transformed landscapes such us Mediterranean mountain ecosystems.
+
+```{r eval=FALSE, echo=FALSE}
 Since we only studied resilience of already established adult trees, monitoring should be enlarged to analyze overall species resilience including that of seedling regeneration.
@@
 They highlight the importance of land-use legacies and how they affect the current geographical distribution of marginal stands, which therefore can be partly disconnected from actual ecologically marginal conditions.
+```
 

### 2019-11-26 man/ecosystems/ms_ecosystems.Rmd
update content

@@
 
-During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table 3).
-The 2005 drought event reduced greenness and growth more than that of 2012 (Tables S1-S2) but the metrics of resilience generally covaried in the same direction those two years.
-We obtained significantly higher resilience and resistance values for 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 6).
-Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event (Table S1, Figure 6).
+During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table S1).
+The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction those two years.
+We obtained significantly higher resilience and resistance values for 2012, the most severe event, than for 2005 in both variables (Table S2; Figure 6).
+Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event.
 Recovery showed a contrasting pattern for EVI and tree-growth.
 
-The resilience metrics calculated significantly varied between sites, except for resilience of tree-growth (p = 0.534; Table 3), which was similar among the three sites.
-The two southern populations showed lower values for recovery than the northern site both for greenness and tree-growth but resistance and resilience values were significantly higher for the southern site (Tables S1-S2).
+The resilience metrics calculated significantly varied between sites, except for resilience of tree-growth (p = 0.534; Table S1), which was similar among the three sites.
+The two southern populations showed lower values for recovery than the northern site both for greenness and tree-growth but resistance and resilience values were significantly higher for the southern site (Table S2).
 

### 2019-11-26 man/tables/t_huber_res.Rmd
update content

---
output:
  word_document: 
    reference_docx: ../../templates/template_tables_portrait.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
library(dplyr)
library(flextable)
library('officer')
```

**Table S2.** Robust measures of central tendency of resilience indices for greenness (EVI) and tree-growth (BAI), grouped by drought events, site and interaction. Measures of central tendency are M-estimators based on Huber's Psi (see material and methods). 95 % confidence intervals using 3000 bootstrap are included between parentheses. *Total* corresponds to the average of 2005 and 2012.

```{r}

# Number of digits 
d <- 3

eraw <- df <- read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg.csv"), header = TRUE)
e <- eraw %>% 
  rename(variables = var) %>% 
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
  dplyr::select(c(-n, -Letter))

# Datos agrupados para todos los sitios 
eraw_a <- read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_a.csv"), header = TRUE)
e_a <- eraw_a %>% 
  rename(variables = var)  %>% 
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
eraw_b <- read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_agg_b.csv"), header = TRUE)

e_b <- eraw_b %>% 
  rename(variables = var)  %>% 
  dplyr::select(c(-n, -Letter)) %>%  
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>%
  mutate(disturb_year = 2099)

emm <- e %>% 
  bind_rows(e_a) %>%
  bind_rows(e_b) %>% 
  unite("v", c("disturb_year","variables"), sep="")

emm_MH <- emm %>% 
  dplyr::select(v, site, M.Huber) %>% 
  mutate(M.Huber = as.character(format(M.Huber, digits = d))) %>% 
  spread(v, M.Huber) %>% 
  mutate(what = 'mh')

emm_CI <- emm %>% 
  dplyr::select(-M.Huber) %>% 
  mutate_at(.vars = c("lower.ci","upper.ci"), funs(format(., digits = d))) %>%  
  unite("ci", c("lower.ci","upper.ci"), sep=" - ") %>% 
  mutate(ci = paste0('(', ci, ')')) %>% 
  spread(v,ci) %>% mutate(what = 'ci')

emi <- bind_rows(emm_MH, emm_CI) %>% arrange(site) %>% 
  mutate(site = case_when(
    site == 'z_all' ~ "All",
    site == 'N' ~ "Northern slope", 
    site == 'S' ~ "Southern slope"),
    var = 'EVI')

# BAI
braw <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE)

b <- braw %>% 
  rename(variables = var) %>% 
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
  dplyr::select(c(-n, -MonoLetter, -Letter, -var_sorted))

# Datos agrupados para todos los sitios 
braw_a <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_a.csv"), header = TRUE)

b_a <- braw_a %>% 
  rename(variables = var)  %>% 
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
  dplyr::select(-n, -Letter) %>%  
  mutate(site = "z_all")

# Datos agrupados para todos los sitios 
braw_b <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber_agg_b.csv"), header = TRUE)

b_b <- braw_b %>% 
  rename(variables = var)  %>% 
  mutate_at(.vars = c("M.Huber", "lower.ci", "upper.ci"), funs(round(., d))) %>% 
  dplyr::select(-n, -Letter) %>% 
  mutate(disturb_year = 2099)

bmm <- b %>% 
  bind_rows(b_a) %>%
  bind_rows(b_b) %>% 
  unite("v", c("disturb_year","variables"), sep="")

bmm_MH <- bmm %>% 
  dplyr::select(v, site, M.Huber) %>% 
  mutate(M.Huber = as.character(format(M.Huber, digits = d))) %>% 
  spread(v, M.Huber) %>% 
  mutate(what = 'mh')

bmm_CI <- bmm %>% 
  dplyr::select(-M.Huber) %>% 
  mutate_at(.vars = c("lower.ci","upper.ci"), funs(format(., digits = d))) %>%  
  unite("ci", c("lower.ci","upper.ci"), sep=" - ") %>% 
  mutate(ci = paste0('(', ci, ')')) %>% 
  spread(v,ci) %>% mutate(what = 'ci')

bmi <- bind_rows(bmm_MH, bmm_CI) %>% arrange(site) %>% 
  mutate(site = case_when(
    site == 'z_all' ~ "All",
    site == 'caH' ~ 'CA-High', 
    site == 'caL' ~ 'CA-Low',
    site == 'SJ' ~ 'SJ'), 
    var = 'BAI')

latabla <- emi %>% bind_rows(bmi)

# CREATE TABLE

# Define borders
big_border = fp_border(width = 2)
std_border = fp_border(width = 1)

mt <- regulartable(latabla,
                   col_keys = c("var","site","c1", "2005rt","2005rc","2005rs", "c2",
                                "2012rt","2012rc","2012rs","c3","2099rt","2099rc", "2099rs"))
mt <- mt %>% 
  # Header 
  set_header_labels('var' = "Variable", site = "Sites", 
                        '2005rt' = "Resistance", '2005rc' = "Recovery", '2005rs' = "Resilience",
                        '2012rt' = "Resistance", '2012rc' = "Recovery", '2012rs' = "Resilience",
                        '2099rt' = "Resistance", '2099rc' = "Recovery", '2099rs' = "Resilience") %>% 
  add_header(var = "", site="", '2005rt' = "2005", '2005rc' = "2005", '2005rs' = "2005",'2012rt' = "2012", '2012rc' = "2012", '2012rs' = "2012",
                        '2099rt' = "Total", '2099rc' = "Total", '2099rs' = "Total", top = TRUE) %>% 
  merge_at(i = 1, j =c(4:7) , part = "header") %>% 
  merge_at(i = 1, j =c(8:10) , part = "header") %>%
  merge_at(i=1, j =c(12:14), part = "header") %>% 
  align(align = "center", part = "header") %>% 
  bold(part = 'header') %>% 
  fontsize(part = "header", size = 9) %>%

  # General format 
  fontsize(part = "body", size = 8) %>% 
  autofit(add_w = 0, add_h = 0) %>% 
  empty_blanks() %>% 
  align(align = "center") %>%

  # Borders 
  border_remove() %>% 
  hline(i = 1:2, border = big_border, part = "header") %>% 
  hline(i = 6, border = big_border, part = "body") %>% 
  vline(j = c(2,6,10), border = big_border, part = "all") %>% 
  hline(i = c(2,4,8,10,12), border = std_border, part = "body") %>%

  # First column 
  merge_v(~var) %>% 
  merge_v(~site) %>% 
  italic(j= c(1,2), part = "body") %>% 
  bold(j=1, part = "body")

mt 
```

### 2019-11-26 man/tables/t_robust_anovas.Rmd
update content

@@
 
-**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the two drought events (in 2005 and 2012) and site.
+**Table S1.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the two drought events (in 2005 and 2012) and site.
 

### 2019-12-07 man/ecosystems/ms_ecosystems.Rmd
update english review

@@
 
-### Short title: *Relict oak resilience at the rear-edge*
+### Short title: *Relict oak resilience at the rear edge*
 
 ### Authors [^\*^] 
-A.J. Pérez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^
+A.J. Pérez-Luque^1,2,a^; Gea-Izquierdo, G.^3,b^ and Zamora, R.^1,2,c^
 
@@
 [^\*^]: Author Contributions.
-AJPL, GGI and RZ conceived the study, conducted field work and collected the data.
-AJPL and GGI conducted lab work.
+AJPL, GGI and RZ conceived of the study, conducted field work, and collected the data.
+AJPL and GGI performed the lab work.
 AJPL analyzed data and led the writing of the paper.
@@
 ### Manuscript highlights
-* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
-* The growth response to climate and resilience followed a water-stress gradient
+* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand levels
+* Resilience and growth response to climate followed a water-stress gradient
 * Trees and stand expressed high sensitivity to drought and land-use legacies
@@
 Forest resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd.
-populations in the Southern Iberian Peninsula was analyzed in relation to historical records of land-use using dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
+populations in the southern Iberian Peninsula was analyzed in relation to historical records of land use employing dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
 
-* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure.
-Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events in the short- and the long-term.
-The response to climate and resilience followed a water-stress gradient.
-Primary and secondary growth were generally less resilient to drought in mesic than in xeric sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and more humid) high-elevation stands, but absent in the most xeric site.
+* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood removal) in the current forest structure.
+Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events over the short and the long term.
+Resilience and response to climate followed a water-stress gradient.
+Primary and secondary growth were generally less resilient to drought in mesic than in xeric sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and wetter) high-elevation stands, but absent in the most xeric (*i.e.* warmer and drier) site.
 
-* Adult oak trees showed high resilience to drought despite being considered relict and at its rear-edge.
-The trees exhibited a variable response to the changing climate along a very short climatic gradient and showed that the ecological and geographical rear-edges do not need to meet today in forests with strong land-uses legacies.
-Forest resilience including that of regeneration needs to be monitored at xeric of the species under future expected climatic scenarios bearing in mind that ecological boundaries for species distribution can be narrow and that land-use legacies can play a key role in stand dynamics.
+* Adult oak trees showed high resilience to drought despite being considered relict and at its rear edge.
+The trees varied in response to the changing climate along a narrow climatic gradient and showed that the ecological and geographical rear edges do not need to meet today in forests with strong land use legacies.
+Forest resilience including that of regeneration needs to be monitored at xeric of the species under future expected climatic scenarios given that ecological boundaries for species distribution can be narrow and that land-use legacies can play a key role in stand dynamics.
 
 ### Keywords 
-extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing
+extreme drought, resilience, rear edge, *Quercus pyrenaica*, tree growth, dendrochronology, remote sensing
 
 ## Introduction
-The response of species to changing environments is likely to be largely determined by population responses at range margins [@Hampe2005].
-Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range (*i.e.* centre-periphery hypothesis) [@Sagarin2002; @Pironon2016].
-It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance and thus higher risk of extinction than those populations at the core of the species' range [@VilaCabrera2019; @Pironon2016].
-Nonetheless, recent reviews found species- and population- specific responses highlight that this hypothesis not always fulfill [@Sexton2009; @Abeli2014].
-In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].
+The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005].
+Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016].
+It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance, higher vulnerability, and thus higher risk of extinction than do populations at the core of the species' range [@VilaCabrera2019; @Pironon2016].
+Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014].
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine
+under what environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].
 
-Climate change is expected to cause important changes in the distribution and abundance of plant communities, and there are already obvious signs of processes associated with increased intensity and duration of droughts [@Allen2010].
-In fact, there has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017].
-This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a].
-Under this climatic change context, it is reasonable to expect that population loss and range retractions in boreal, temperate and Mediterranean species should be seen in the lowest latitudes, altitudes and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
-In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017].
-This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].
+Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of processes associated with increased intensity and duration of droughts [@Allen2010].
+In fact, the frequency and severity of drought events
+have increased in recent decades, along with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
+This is especially critical, since climate-change projections for the Mediterranean area forecast more frequent and severe extreme climate events [@Spinoni2017a].
+In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
+In a global-change context, the impact of drought on vegetation needs to be evaluated regarding the interactions with other drivers of change [@DoblasMiranda2017].
+This is especially relevant for areas with a long history of landscape alteration, such as the Mediterranean region, where land-use change plays a major role in current ecosystem ecology [@Navarro2013], given that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].
 
-The impacts of drought are also especially relevant for populations considered to be located in the rear-edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005].
-The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of species to predicted climatic changes [@Jump2010].
+The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005].
+The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010].
 Yet it is crucial to accurately characterize the ecological rear-edge of species.
 This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019].
-Furthermore, it becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].
+Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].
 
-Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018].
-The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and provide valuable ecosystem services.
-The resilience to drought of species is better assessed when several approaches like dendrocronology and remote sensing are combined simultaneously.
-Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual tree-ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].
-Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013].
-The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].
-Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.
+Understanding the vegetation response to drought requires integrative and multidisciplinary strategies to assess the disturbances of forest at several scales [@Hartman2018].
+The assessment of resilience to climate disturbances such as droughts provides critical information concerning the capacity of the forests to maintain their structure and render valuable ecosystem services.
+The drought resilience of species is better assessed when several approaches, such as dendrochronology and remote sensing, are combined.
+Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and annual tree-ring widths can be used to study changes in growth as a response to drought at the individiual-tree scale [@Fritts1976].
+Remote sensing can be used to study the impact of drought on ecosystems and hence the stand level [*e.g.* @Zhang2013].
+The combination of both approaches (remote sensing and dendroecology) has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].
+Nonetheless, it is crucial to ascertain wether the responses at the tree level differ from those at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.
 
-In this study we used remote sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and tree-radial growth (as a proxy to secondary growth) of *Quercus pyrenaica* Willd.
-(*melojo oak*) on southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
-We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) in the long-term in the last decades.
-Our main hypothesis is that these relict populations driven by historical land-use are particularly vulnerable to climate change at their present climatic (either altitudinal or latitudinal) rear-edge, and will show low resilience after extreme drought particularly at the dry edge along climatic gradients at the small scale within the rear-edge.
-To address this hypothesis our specific aims were: (*i*) To quantify how recent extreme drought events influenced primary and secondary growth of *Q.
-pyrenaica* forests in its rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) within the rear edge to detect vulnerability to climate change along climatic gradients at the small spatial scale.
-To achieve goal (*iii*) and characterize variability in the forest response to drought within the current geographical rear-edge we assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.
+In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd.
+(*melojo oak*) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
+We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades).
+Our main hypothesis is that these relict populations driven by historical land use are vulnerable to climate change at their present climatic (either elevational or latitudinal) rear edge, and hence will show low resilience to extreme drought along climatic gradients at a small scale within the rear edge.
+To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q.
+pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
+For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.
 
@@
 ### Tree species and study site 
-*Quercus pyrenaica* forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990].
-In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate.
+*Quercus pyrenaica* forests extend throughout south-western France and the Iberian Peninsula reaching their southern limit in mountain areas of northern Morocco [@Franco1990].
+In the Iberian Peninsula, these forests occupy siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid, and hyperhumid ombroclimate.
 *Q.
 pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation.
-It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].
+As a submediterranean species, it has lower drought tolerance than evergreen Mediterranean taxa [@delRio2007].
 
-The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations up to 3482 m *a.s.l.*.
-The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability set by altitude.
-Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002].
-There are eight melojo oak patches (2400 ha) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys.
+The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*.
+The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability according to elevation.
+Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002].
+Eight melojo oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys.
 Today, *Q.
 pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005].
-They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG].
-These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002].
-Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].
+They are the richest forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG].
+These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002].
+Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].
 
-### Drought episodes
-The Iberian Peninsula suffered several extreme drought episodes in the last three decades (*e.g*.
-1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].
-The 2005 and 2012 drought events have been documented among the worst in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figure S1; Table S3).
-We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below).
-Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 and to analyze forest resilience to drought in a longer term (see Table S3).
-The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015.
-We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html).
+### Climatic data and drought episodes
+Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008].
+Monthly precipitation and minimum and maximum temperatures had a 0.25 x 0.25 º resolution for the 1950-2016 period.
+Grid cells were select to cover each sampled site.
+The SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] index with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
+
+The Iberian Peninsula underwent several extreme drought episodes in the last three decades (*e.g*.
+1995, 1999, 2005, 2012) [@VicenteSerrano2014].
+The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], and they were characterized as extreme drought in our climatic data (Figure S1; Table S3).
+We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below).
+Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3).
+A drought event was identified using the SPEI 12-months scale, following a procedure similar to the one proposed by @Spinoni2015.
+We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html).
 A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017].
-A drought event is considered only when SPEI values are below that threshold for at least two consecutive months.
-For each drought event we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
+A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months.
+For each drought event, we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity*, *i.e.* the sum of the SPEI values (absolute values) during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
 
@@
 EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
-MODIS EVI data were obtained for the period 2000 - 2016.
+MODIS EVI data were compiled for the period 2000 - 2016.
 We selected the pixels covering the distribution of *Q.
 pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
-Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].
+Any values affected by clouds, snow, shadows or high content of aerosols, were filtered out following recommendations for mountain regions [@Reyes2015].
 
-Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016.
-The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about their magnitude [@Samanta2012].
-For each pixel, an annual EVI value was obtained by averaging EVI valid values.
+The mean Annual EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016.
+The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, in order to minimize bias in the evaluation of anomalies and to provide more information concerning their magnitude [@Samanta2012].
+For each pixel, an annual EVI value was calculated by averaging EVI valid values.
 Then, the standardized anomaly was computed as: $\mathrm{EVI_{sa,\mathit{i}}}= (\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}})\Big/\sigma_{\mathrm{ref}}$,
@@
 
-Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88).
-For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale, using Mann–Kendall nonparametric test.
+Rather than other vegetation indices such as the NDVI, $EVI_{mean}$ was chosen because it is highly stable under the use of any filter [@Reyes2015] and because it showed highly significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88).
+For exploratory purposes, temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test.
 
-### Field sampling and dendrochronological methods to assess tree individual resilience
-Tree sampling was carried out during autumn 2016.
-Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1).
-For the southern site two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), which constitute the current low altitudinal limit (CA-Low) and the tree-line (CA-High), respectively in the sampled aspect.
-Despite the proximity of these two elevations (less than 200 m difference in altitude) the stands greatly differ in their structure and characteristics (Table 1).
+### Field sampling and dendrochronological methods to assess individual tree resilience
+Trees were sampled during autumn 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1).
+For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the sampled aspect.
+Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1).
 The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
 All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015].
-In each site between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly selected.
-Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer.
+At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled.
+Two cores of 5 mm of diameter were taken from each tree at breast height (1.3 m) using an increment borer.
 Diameter at breast height (DBH) and total height were measured for each tree.
-In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius.
+In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius.
 Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].
-Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests.
-When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct for significance.
+Differences between sites for height, DBH and competition indices were analyzed using non-parametric Kruskal-Wallis rank sum tests.
+When significant differences were detected, multiple comparisons were run using the Dunn's-test with Bonferroni adjustment to correct for significance.
 
-Tree cores were air dried, glued onto wooden mounts and sanded.
-Annual radial growth (ring width, RW) was measured with a measuring device coupled to a stereomicroscope, with an accuracy of 0.001 mm.
+Tree cores were air dried, glued onto wooden mounts, and sanded.
+Annual radial growth (ring width, RW) was determined with a measuring device coupled to a stereomicroscope, for an accuracy of 0.001 mm.
 Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
@@
 The growth trends were analyzed at different time scales.
-To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used.
-They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976].
+To study the growth response to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used.
+These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976].
 Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010].
-Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976].
+Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976].
 The statistical quality of each chronology was checked via the expressed population signal (EPS).
@@
 The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
-Theoretically, BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008].
-Raw ring-widths and measured DBH were used to compute BAI [@Piovesan2008] using the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
+In theory, BAI represents a more accurate indicator of growth than ring width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile growth [@Biondi2008].
+Raw ring widths and measured DBH were used to compute BAI [@Piovesan2008] with the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
 For each individual tree, a mean BAI series was calculated.
-Then, mean site BAI chronologies were obtained by averaging individual tree BAI time series.
-
-### Climate and growth 
-Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008].
-Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period.
-We selected grid cells covering each sampled sites.
-SPEI index with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
-Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].
+Then, mean site BAI chronologies were determined by averaging individual tree BAI time series.
 
 ### Disturbance analyses 
-Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past.
+Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past.
 Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@Rubino2004].
@@
 ```
-Site disturbance chronologies were constructed by averaging the individual disturbances series annually.
-To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014].
-In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes.
-For this purpose, an exhaustive review of existing historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).
-We reviewd several documentary sources: historical documents and maps; detailed mining reports; official information about recent wildfires events and forest management practices; livestock farming; traditional irrigation ditches and other studies concerning the socioeconomic dynamics of forest of Sierra Nevada at different scales (see Table S4 for references).
+Site-disturbance chronologies were constructed by annually averaging the individual disturbance series.
+To separate growth peaks caused by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50% of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014].
+In addition, the history of the forest and management of our sampling sites was inferred from a detailed analysis of historical land-use changes.
+For this, existing historical documents were exhaustively reviewed to compile information on socio-economical activities affecting the forests being studied (Table S4).
+We reviewed several documentary sources: historical documents and maps; detailed mining reports; official information on recent wildfires events and forest-management practices; livestock farming; traditional irrigation channels; and studies concerning the socioeconomic dynamics of forests on Sierra Nevada at different scales (see Table S4 for references).
 
@@
 ### Assessing ecosystem and tree individual resilience to drought
-To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011.
-The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred.
-The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity.
+To evaluate the effects of drought events on ecosystem resilience (using greenness data) and individual tree resilience (using BAI data), we used resilience indices proposed by @Lloret2011.
+The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$) quantifies the severity of the impact of the disturbance in the year it occurred.
+The Recovery index, computed as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity.
 Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels.
 The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event.
-The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively.
-A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S3b) and this length was used in other studies [*e.g.* @Gazol2018].
+The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value over a period of three years before and after the drought event, respectively.
+A period of three years was chosen because we found similar results on comparing periods of two, three, and four years (Figure S3b), and this time period has been used in other studies [*e.g.* @Gazol2018].
 Resilience metrics for BAI data were additionally computed for the most severe drought events since 1940 (*n* = 8; Table S3) and compared with drought severity.
@@
 ### Statistical analysis 
-For each of the three resilience indices studied we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects).
-These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012].
-Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012].
-When running the robust ANOVA test, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers.
-Post-hoc differences were assessed pairwise using a similar bootstrap test.
-All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package.
+Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].
+For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern aspects).
+These tests were used because original and log-transformed data did not follow the assumptions of normality or homogeneity of variance [@Wilcox2012].
+Robust measures of central tendency (M-estimator based on Huber's Psi) were used because they were close to the mean value in all cases [@Wilcox2012].
+When the robust ANOVA test was run, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers.
+*Post-hoc* differences were assessed pairwise using a similar bootstrap test.
+All the robust ANOVA and *post-hoc* tests were carried out using the `WRS2` package.
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
@@
 ### Time trends in vegetation greenness
-The analysis of greenness time trends showed that EVI in 78.9 % of the pixels experienced a positive trend for the 2000-2016 period.
-The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2a).
-A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 2b).
+The analysis of time trends in greenness showed that 78.9% of the EVI pixels followed a positive trend for the 2000-2016 period.
+The lowest values of EVI standardized anomalies for the study period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2a).
+A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was found during this drought event, whereas no changes in greenness in response to the 2012 drought were detected (Figure 2b).
 
-### Analysis of radial growth trends and disturbances 
+### Analysis of radial-growth trends and disturbances 
 The trees of the southern population were older than those from the northern one.
-In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites and despite competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3).
-Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S3a).
-Only trees from the southern sites (*i.e.* the more humid aspect) showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid and colder, site (CA-High).
+In addition, trees from the southern population at high elevation were taller and their growth was significantly faster than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3).
+The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 3 and S3a).
+Only trees from the southern sites (*i.e.* the wetter aspect) showed significant positive growth trends since the late 1970s (Figure 3), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).
 
-Drought events reduced radial growth for all sites (Figure S2a), being particularly evident the negative effect of the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
-Tree-growth reduction followed a moisture-gradient.
-The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S2a), being the weakest growth reductions in trees from the wettest site (Figure 3).
+Drought events reduced radial growth for all sites (Figure S2a), this negative effect being particularly evident for the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
+Tree-growth reduction followed a moisture gradient.
+The southern sites (CA-High and CA-Low) showed less tree-growth reductions than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site (Figure 3).
 
-The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940s (the most evident) and the second in 1995-2000 (Figure 4).
+The northern site (SJ) showed two major release events (GC > 50% occurring in more than 50% of trees sampled): the first during the 1940s (the most evident) and the second in 1995-2000 (Figure 4).
 These periods alternated with periods of suppression.
-In contrast, the two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.
+By contrast, the two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.
 
-### Resilience to drought events at the ecosystem and individual tree levels
+### Resilience to drought events at the ecosystem and individual-tree levels
 *Rs* and *Rt* varied in the same direction whereas *Rc* varied inversely to *Rs* and *Rt*.
 Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 5).
-A similar pattern was obtained for *Rs* but only significant for SJ.
-Importantly, non-significant patterns were obtained if we excluded 1995, except for *Rc* in SJ (Figure S5).
+A similar pattern was found for *Rs* but proved significant only for SJ.
+Importantly, non-significant patterns resulted when we excluded 1995, except for *Rc* in SJ (Figure S5).
 The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4).
@@
 During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table S1).
-The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction those two years.
-We obtained significantly higher resilience and resistance values for 2012, the most severe event, than for 2005 in both variables (Table S2; Figure 6).
+The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years.
+Resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 in both variables (Table S2; Figure 6).
 Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event.
-Recovery showed a contrasting pattern for EVI and tree-growth.
+Recovery showed a contrasting pattern for EVI and tree growth.
 
-The resilience metrics calculated significantly varied between sites, except for resilience of tree-growth (p = 0.534; Table S1), which was similar among the three sites.
-The two southern populations showed lower values for recovery than the northern site both for greenness and tree-growth but resistance and resilience values were significantly higher for the southern site (Table S2).
+The resilience metrics calculated varied significantly between sites, except for resilience of tree growth (p = 0.534; Table S1), which was similar among the three sites.
+The two southern populations showed lower recovery values than did the northern site both for greenness and tree growth, but resistance and resilience values were significantly higher for the southern site (Table S2).
 
 ### Tree-growth response to climate
-The response of tree-growth to water availability was greater than that to temperatures.
+The response of tree growth to water availability was greater than to temperatures.
 Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 7a).
-Yet there were some differences between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
-In addition, spring maximum temperature was the most significant limitation for tree-growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of current September positively influenced tree-growth only in the northern population.
+Nevertheless, differences appeared between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
+In addition, the spring maximum temperature was the most significant limitation for treemgrowth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of the current September positively influenced tree growth only in the northern population.
 
@@
 
-By using a combined approach of remote sensing information and dendrocronology we have quantified the drought impacts on melojo oak forests on Sierra Nevada and its resilience to several severe drought events in the last decades.
-Our results clearly indicated that these relict oak populations driven by historical land-use are resilient to climate change at their present (either altitudinal or latitudinal) rear-edge.
-However, resistance, resilience and forest recovery to extreme drought events are strongly influenced by mountain aspect, local environmental conditions, and management legacies.
+By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the melojo oak forests of Sierra Nevada and their resilience to several severe drought events in the recent decades.
+Our results indicated that these relict oak populations driven by historical land use are resilient to climate change at their present rear edge.
+However, resistance, resilience, and forest recovery to extreme drought events are strongly influenced by mountain aspect, local environmental conditions, and management legacies.
+This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance varies markedly along the rear edge under climate change.
 
-### Land-use legacies shape sensitivity to climate change of forests and the present rear-edge
-The review of historical documents shows how forest clearings, firewood extraction, charcoal exploitations and mining have strongly impacted the forests at Sierra Nevada (Table S4) where in addition it has been estimated a historical loss of about 90% of broadleaf *Quercus* species cover from medium and low elevations [@JimenezOlivencia2015].
-Together with the analysis of the disturbance chronologies, the observed strong differences in stand structures, tree size and age suggest different forest history and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
-In the northern slopes of Sierra Nevada (*e.g.* SJ site), land uses have been historically distributed along an altitudinal gradient: grasslands and shrublands for cattle farming at the highest elevations; then forest stands with some croplands; and irrigated terraces with tree crops at the lowest elevations [@JimenezOlivencia2015].
-In addition, other activities like mining should have modified the forest structure, *e.g.* SJ site contains a high concentration of small mines an quarries that were exploited intermittently throughout history.
-The release growth event observed in 1940s concurs with a period of maximum mining activity in this area (1925 to 1957) (Table S4), during wich there was an increase in the use of timbers for mine tunnels and furnaces, that also required large amounts of fuelwood to melt the mineral.
-This heavily exploitation of the neighbouring forest resources should have affected a significant part of this oak woodland as shown by growth of the remnant trees included in the northern site (Figure S2b).
-On the other hand, woodlands in the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks occur [@JimenezOlivencia2015].
-Firewood, charcoal extraction and acorns have been resources intensively exploited in southern sites, until at least the middle of the 20th century when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013].
-For CA-High site, the only positive release event showed at the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
+### Land-use legacies shape sensitivity to climate change of forests and the present rear edge
+The review of historical documents revealed that forest clearings, firewood removal, charcoal making, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species approaches 90% of the cover at medium and low elevations [@JimenezOlivencia2015].
+Together with the analysis of the disturbance chronologies, the observed notable differences in stand structure, tree size, and age suggest different forest histories and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
+On the northern slopes of Sierra Nevada (*e.g.* SJ site), land uses have been historically distributed along an elevational gradient: grasslands and shrublands for cattle farming at the highest elevations; then forest stands with some croplands; and irrigated terraces with tree crops at the lowest elevations [@JimenezOlivencia2015].
+In addition, other activities such as mining must have altered the forest structure, *e.g.* the SJ site has many small mines and quarries that were exploited intermittently throughout history.
+The release growth event discerned for the 1940s concurs with a period of maximum mining activity in this area (1925 to 1957) (Table S4), during wich timbers use increased for mine tunnels and furnaces, these also requiring large amounts of firewood to melt the mineral.
+This heavy exploitation of the neighboring forest resources must have affected a significant part of this oak woodland, as shown by growth of the remnant trees included in the northern site (Figure S2b).
+On the other hand, woodlands on the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks grow [@JimenezOlivencia2015].
+Firewood, charcoal, and acorns have been resources intensively exploited at southern sites, until at least the middle of the 20th-century, when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013].
+At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management type which has been applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
 
-The other release event observed for SJ site during 1995-2000 was lower than that occurred at 1940, but also affected most trees (Figures 4, S2b).
-No records of forest practices carried out in this area in the last 30 years were found [@Bonet2016obsnev_forest], and no logging was recorded during 1995 - 2000 (F.J.
-Cano-Manuel *personal communications*).
-Therefore this release might be related to natural drought-induced mortality after 1995 as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
-On the other hand, the strong positive correlations of SPEI with tree-growth for this site show a high sensitivity to water availability [@GeaIzquierdo2014].
+The other release event observed for the SJ site during 1995-2000 was lower than during 1940, but also affected most trees (Figures 4, S2b).
+No records of forest practices in this area over the last 30 years have been found [@Bonet2016obsnev_forest], and no logging has been recorded during the period 1995 - 2000 (F.J.
+Cano-Manuel *personal communication*).
+Therefore this release might be related to natural drought-induced mortality after 1995, as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
+On the other hand, the strong positive correlations of SPEI with tree growth for this site show high sensitivity to water availability [@GeaIzquierdo2014].
 
@@
 
-### Relict oaks sensitivity (resistance) to recent drought events
+### Sensitivity (resistance) of relict oaks to recent drought events
 Severe drought negatively affects both primary and secondary growth of *Q.
 pyrenaica* forests.
-This was expressed by the observed reduction in greenness and tree-growth in response to the 2005 and 2012 drought events, and the consistent radial growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
-Furthermore, the greatest reduction of tree-growth was observed during the 1995 drought, a characteristic precipitation negative anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
-Similar to many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
-pyrenaica* along their distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
+This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by the consistent radial-growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
+Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
+Similar to the situation with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
+pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
 Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
 
-Greenness exhibited a lower sensitivity to drought than tree-growth, particularly for drier sites.
-These findings are in accordance with previously works showing that tree-growth is a more sensitive metric of forest resilience than net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018] and would suggest that the growth reduction could more mediated by sink that by source limitations [@Korner2013; @Fatichi2014].
-The use of tree-ring records allow to complement remote sensing data in a longer time scale, by reflecting climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
+Greenness proved less sensitive to drought than did tree growth, particularly for drier sites.
+These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018], suggesting that the growth reduction could mediated by sink more than by source limitations [@Korner2013; @Fatichi2014].
+Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
 
-##### Small-scale environmental variability shapes sensitivity to climate of trees within the rear-edge
-Greenness and tree-growth were more affected by drought events in drier northern populations than in more humid southern oak populations, *e.g.* the northern site showed higher browning intensity than the southern sites during the 2005 drought event and the stronger correlations of tree-growth with SPEI (Hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought in drier sites [@GeaIzquierdo2014].
-It is well known that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competition, etc.) [@Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
-Trees at CA-High sowhed higher BAI than those located at lower-elevations (CA-Low and SJ) (Figure 4).
-This shows the high variability in the response to climate exhibited in a small gradient, which is especially interesting for southern sites, which are very close to each other, and overall both are considered to constitute the rear-edge for the species.
+Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations.
+For example, the northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and the stronger correlations of tree-growth with SPEI (hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014].
+It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition, etc.; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
+Trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4).
+This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie very close to each other and overall both are considered to constitute the rear edge for the species.
 
@@
 ### Relict oaks show high resilience to recent drought events and long-term climatic variability 
-Despite the severe drought events suffered in the last decades (Table S3), we found a positive trend for vegetation greenness of *Q.
-pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather humid decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
-For tree-growth, we also observed positive trends in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends were described for this species along its distribution range only at high elevation humid and cold sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold limited high elevation sites [@Salzer2009; @GeaIzquierdo2014].
-Importantly, for rear-edges where climate change is a threat we would have expected negative growth trends as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
+Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q.
+pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
+For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4).
+Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014].
+Importantly, for rear edges threatened by climate change, negative growth trends would have been expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
 
-Albeit that 2012 drought event was more severe and intense than that in 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012.
-This could be explained by the different timing of the two droughts.
-The 2012 drought was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought.
+Althought the 2012 drought event was more severe and intense than that of 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012.
+This could be due to the different timing of the two droughts.
+The 2012 event was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought.
 The latter matched the period of maximum growth for oak forests in late spring (Figure S4).
-This highlights the importance of the timing of drought as a key factor determining the tree recovery after drought [@Camarero2015b; @Huang2018].
-For tree-growth, the highest values of resilience were obtained for the two most severe events (1995 and 1999)(Table S3) and we observed a positive relationship between the tree-growth resilience and drought severity (Figure 5).
+This highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Camarero2015b; @Huang2018].
+For tree growth, the highest values of resilience were found for the two most severe events (1995 and 1999; Table S3) and tree-growth resilience was positively related to drought severity (Figure 5).
 
-The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada [@Valbuena2013; @Valbuena2017] seem to be indicative of high local adaptation of this oak.
-Our findings are in agreement with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017].
+The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests at Sierra Nevada reported elsewhere [@Valbuena2013; @Valbuena2017] appear to indicate the strong local adaptation of this oak.
+Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017].
 @MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations.
-Our results would highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our studied forest.
+Our results highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our forest.
 All this, together with the characteristic high resprouting ability of the species, would suggest a long-term persistence of those populations [@Bellingham2000].
-It needs to be mentioned that we only studied adult individuals established decades or centuries ago, meaning that it needs to be assessed if the trees express resilience at the species level or to there is vulnerability if we analyzed seedling regeneration, likewise other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].
+It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the trees express resilience at the species level or whether we would find vulnerability if we analyzed seedling regeneration, likewise other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].
 
@@
 
-Summarizing, two mains results stand out from this research.
-First, the high values of resilience observed in our study suggest that the populations of *Q.
-pyrenaica* in Sierra Nevada are located in a geographical but not a climatic, ecological rear-edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019].
-The current niche is a result of land-use changes, which further complicates the definition of potential rear-edges.
-The high resilience values observed could be also related with stabilizing mechanisms promoting community resilience that can buffer the impacts of extreme events, as it has been described for other species [*e.g.* *Pinus sylvestris*, @Herrero2014].
+In summary, two main results stand out from this research.
+First, the high values of resilience observed in our study suggest that *Q.
+pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019].
+The current niche is a result of land-use changes, which further complicate the definition of potential rear edges.
+The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @Herrero2014].
 
-Second, these resilience responses of oak forest to drought events are not spatially homogeneous around all mountain range, due to differences in ecological conditions and/or past management legacies.
-In fact, there was much small-scale variability in the response to climate within the rear-edge that we had "a priori" considered in our study.
-Furthermore we even observed positive effects of climate change in certain stands, as discussed, and which would be in disagreement with our hypothesis expecting oak vulnerability in the studied geographical rear-edge.
-This suggests that the rear-edge needs to be redefined in consequence [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear-edge) and the ecological (limiting) rear-edge of species.
+Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.
+In fact, there was much small-scale variability in the response to climate within the rear edge that we had *a priori* considered in our study.
+Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the geographical rear edge studied.
+This suggests that the rear edge therefore needs to be redefined in [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the ecological (limiting) rear edge of species.
 
-Overall, our results show how management history conditions tree-growth and resilience to climate change of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017] which, hence, will also strongly determine the current geographical distribution of marginal stands, regardless of the potential extent of ecological marginality of species.
+Overall, our results show that management history influences tree growth and resilience to climate change of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017] which, hence, will also strongly determine the current geographical distribution of marginal stands, regardless of the potential extent of ecological marginality of species.
 
@@
 ## Conclusions 
-The ecological and geographical read-edges did not seem to fully match in our study.
-Severe drought events provoke important reductions in primary and secondary growth of *Q.
-pyrenaica* forests in the studied relict, rear-edge forest.
-However, we observed no negative growth trends despite we had expected vulnerability to climate change for these relict stands.
-Furthermore, we observed positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a positive steep trend of secondary growth at the tree level at the most humid site along the climatic gradient analyzed.
+The ecological and geographical read edges did not appear to fully match in our study.
+Severe drought events provoke major reductions in primary and secondary growth of *Q.
+pyrenaica* forests in the relict, rear-edge forest studied.
+However, we found no negative growth trends despite our expectation of vulnerability to climate change for these relict stands.
+Furthermore, we detected positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a steep positive trend of secondary growth at the tree level at the wettest site along the climatic gradient analyzed.
 The trees exhibited high resilience values in response to drought, particularly in the long-term scale, and resilience was directly related to drought severity.
-These results would be consistent with other research showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*.
-The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent and can vary in very small spatial gradients (*i.e.* following ecological thresholds).
+These findings are consistent with other results showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*.
+The differences found in tree growth and resilience to drought between sites close together show that responses to drought were site dependent and can drastically vary even in very narrow spatial gradients (*i.e.* following ecological thresholds).
 This is particularly relevant for rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
-The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents.
-In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems.
-This is relevant for tree species with a high sensitivity to climate change, such us *Q.
-pyrenaica*, not only for conservation per se of the species, but for all ecosystem services that these singular forests offer.
-Following our study it is necessary to disentangle how forest resistance and resilience are affected by the combination of climate change and land-use legacies in highly transformed landscapes such us Mediterranean mountain ecosystems.
+The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land use in local forest dynamics, as inferred from an exhaustive review of historical documents.
+In this sense, our results highlight the importance of land-use legacies for highly transformed Mediterranean systems.
+This is relevant for tree species with a high sensitivity to climate change, such as *Q.
+pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these singular forests offer.
+As a follow-up to our study, research would be informative to disentangle how forest resistance and resilience might be affected by the combination of climate change and land-use legacies in highly transformed landscapes such as Mediterranean mountain ecosystems.
 
@@
 Behle, F.J.
-Bonet and M.
+Bonet, and M.
 Suárez-Muñoz for field assistance.
@@
 Cano-Manuel and F.J.
-Navarro provided worthwhile information about oak management projects in Sierra Nevada.
+Navarro provided worthwhile information about oak-management projects in Sierra Nevada.
 We are grateful for the comments made by F.J.
-Bonet in an earlier version of the manuscript.
-AJPL wants to thank the invaluable support received from his family over the years.
+Bonet on an earlier version of the manuscript.
+AJPL wishes to thank the invaluable support received from his family over the years.
 This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” project.

### 2019-12-10 man/ecosystems/ms_ecosystems.Rmd
update ms

@@
 ---
-title: "Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today"
+title: "Oak resilience to drought and land use show how the ecological and geographical rear edges do not necessarily meet today"
 output:
@@
 Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014].
-In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine
-under what environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine under what environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].
 
 Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of processes associated with increased intensity and duration of droughts [@Allen2010].
-In fact, the frequency and severity of drought events
-have increased in recent decades, along with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
+In fact, the frequency and severity of drought events have increased in recent decades, along with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
 This is especially critical, since climate-change projections for the Mediterranean area forecast more frequent and severe extreme climate events [@Spinoni2017a].
@@
 In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd.
-(*melojo oak*) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
+(Pyrenean oak, *melojo*) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
 We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades).
@@
 A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months.
-For each drought event, we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity*, *i.e.* the sum of the SPEI values (absolute values) during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
+For each drought event, we computed: the *duration* as the number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the absolute SPEI values during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
 
@@
 Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations.
-For example, the northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and the stronger correlations of tree-growth with SPEI (hydrological and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014].
+For example, the northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and the stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014].
 It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition, etc.; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
@@
 For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming at species-wise cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014].
+Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming for the species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014].
 Importantly, for rear edges threatened by climate change, negative growth trends would have been expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
@@
 Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the geographical rear edge studied.
-This suggests that the rear edge therefore needs to be redefined in [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the ecological (limiting) rear edge of species.
+This suggests that the rear edge therefore needs to be redefined [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the ecological (limiting) rear edge of species.
 

### 2019-12-10 man/figures/0_figures_ecosystems.Rmd
update ms

@@
 Distribution of *Quercus pyrenaica* forests in the Iberian Peninsula (a) and in Sierra Nevada mountain range (b).
-Different colours indicate oak population cluster's identified in Sierra Nevada [@PerezLuque2015].
-For each population, a grid with the MODIS pixels is shown (see material and methods).
+Different colors indicate oak-population clusters identified in Sierra Nevada [@PerezLuque2015].
+For each population, a grid with the MODIS pixels is shown (see Material and methods).
 Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
-Colour orthophotography of 2009 from Regional Ministry of the Environment.
+Color orthophotography of 2009 from Regional Ministry of the Environment.
 
@@
 See main text for details on EVI calculation.
-***b)*** Percentage of pixels showing browning, greening or no-changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
+***b)*** Percentage of pixels showing browning, greening or no changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
 See main text for an explanation of greening and browning.
@@
 Number of series is displayed in the upper plot.
-We only show years replicated with # series > 5.
-Linear trends since 1975 are shown all sites (numbers indicate $r^2$ values; asterisks indicate significant linear trend, *p < 0.001*).
+Only years replicated with # series > 5 are shown.
+Linear trends since 1975 are indicated for all sites (numbers indicate $r^2$ values; asterisks indicate significant linear trend, *p < 0.001*).
 
@@
 pyrenaica* sites.
-Dashed black lines indicate a threshold of 50 % of GC (see material and methods).
+Dashed black lines indicate a threshold of 50% of GC (see Material and methods).
 Note that y-axes do not correspond in all of the three panels for the sake of clarity.
@@
 
-**Figure 5.** Resilience metrics of tree-growth for eigth severe drought events since 1950 (see main text for details) as a function of drought severity.
+**Figure 5.** Resilience metrics of tree-growth for eight severe drought events since 1950 (see main text for details) as a function of drought severity.
 *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
@@
 **Figure 6.** Comparison of the response of  *Q.
-pyrenaica* forests to drought in terms of resistance, recovery and resilience of greenness (a) and tree-growth (b).
-For EVI we compared northern populations (*red circle*) with southern ones (*green circle*).
-For BAI we compared the more xeric northern population (San Juan, SJ; *red circle*) with the two southerns populations: Cáñar-High (CA-High; *blue circle*) and Cáñar-Low (CA-Low; *green circle*).
-Different letters indicate significant *post hoc* differences between groups (see material and methods for details).
+pyrenaica* forests to drought in terms of resistance, recovery, and resilience of greenness (a) and tree growth (b).
+For EVI, northern populations (*red circle*) were compared with southern ones (*green circle*).
+For BAI, the more xeric northern population (San Juan, SJ; *red circle*) was compared with the two southern populations, Cáñar-High (CA-High; *blue circle*) and Cáñar-Low (CA-Low; *green circle*).
+Different letters indicate significant *post hoc* differences between groups (see Material and methods for details).
 
@@
 
-**Figure 7.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High).
-Asteriks indicate significant ($P < 0.05$) correlation coefficients.
+**Figure 7.** Correlation coefficients found by relating tree-ring residual chronologies (RWI) of *Q.
+pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimum (b) and maximum (c) temperatures.
+*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation southern site (CA-High).
+Asterisks indicate significant ($P < 0.05$) correlation coefficients.
 
@@
 **Figure S1.** ***a)*** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent mean and errorbars standard error.
-*Black* line indicates mean for the whole period (585 mm).
-*Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively).
-*Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively).
-Years with average values below -1SD are labelled.
-Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*).
+Points represent the mean, and error bars the standard error.
+The black line indicates mean for the entire period (585 mm).
+The red lines represent -1 and -2 standard deviation (dotted and dashed lines, respectively).
+The blue lines represent +1 and +2 standard deviation (dotted and dashed lines, respectively).
+Years with average values below -1SD are labeled.
+Data from 28 meteorological stations distributed around the Sierra Nevada area (from the National Spanish Meteorological Services, AEMET).
+***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (blue line) and 2011-2012 (red line).
 The boxplot representing the average from 1950-2015 period.
 Data from meteorological station Granada, Base Aérea.
-***b)*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardised Precipitation-Evapotranspiration Index (SPEI).
+***b)*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardized Precipitation-Evapotranspiration Index (SPEI).
 Data from Global SPEI database (http://spei.csic.es/database.html).
-We obtanied the SPEI data for a 12 month scale and for all 0.5º grid cells covering Sierra Nevada.
-Horizontal gray bars indicate 2005 and 2012 year.
+We took the SPEI data for a 12-month scale and for all 0.5º grid cells covering Sierra Nevada.
+Horizontal gray bars indicate the years 2005 and 2012.
 
@@
 
-**Figure S2.** ***a)*** Residual tree-ring chronologies obtained for the *Q.
+**Figure S2.** ***a)*** Residual tree-ring chronologies determined for the *Q.
 pyrenaica* sites.
 Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines showing the severe drought years identified in our climatic data (Table S3 and Figure S1).
+Dotted black lines show the severe drought years identified in our climatic data (Table S3 and Figure S1).
 ***b)*** Percentage of *Q.
 pyrenaica* trees affected by GC > 50 % by site.
-*Black* line shows number of trees (rigth-axis).
+*Black* line shows number of trees (right-axis).
 Data for number of trees > 2 is shown.
@@
 
-**Figure S3.** ***a)*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time-domains after pre-filtering the time-series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centred moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair chronologies were calculated.
-Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot`.
+**Figure S3.** ***a)*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time domains after pre-filtering the time series with increasing size of the moving-average window (1 to 40 years).
+Each site chronology was smoothed using centered moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair of chronologies was calculated.
+Significance was tested using 1000 bootstrap replicates and with 95% confidence intervals built using the R package `boot`.
 ***b)*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
@@
 Horizontal bars correspond to the most severe droughts for Sierra Nevada since 1900 (computed as in Table S3).
-Their position indicate the start and end months of each drought event.
-Bars length show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017]).
+Their position indicates the start and end months of each drought event.
+Bars lengths show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017].
 
@@
 
-**Figure S5.** Resilience metrics of the tree-growth for severe drought events since 1950 (excluding 1995 drought event).
+**Figure S5.** Resilience metrics of the tree growth for severe drought events since 1950 (excluding 1995 drought event).
 *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
@@
 Resilience metrics were computed for each population (sample depth > 10) and drought event.
-Gray line represents overall relationship for each Resilience metrics.
+The gray line represents overall relationship for each Resilience metrics.
 

### 2019-12-10 man/tables/t_dendro_summary.Rmd
update ms

@@
 
-**Table 2.** Characteristics of the mean tree ring chronologies.
-Values of length between parentheses indicate number of years replicated with more than five series.
+**Table 2.** Characteristics of the mean tree-ring chronologies.
+Length values in parentheses indicate the number of years replicated with more than five series.
 RW = mean annual ring width (standard deviation in parenthesis).

### 2019-12-10 man/tables/t_droughts.Rmd
update review GGI
update ms

@@
 ##### 
-**Table S3.** Drought events for the 1901-2016 period for Sierra Nevada based on SPEI index (12 months scale). See methods for details.
+**Table S3.** Drought events for the 1901-2016 period for Sierra Nevada ranked according to drought severity calculated from the SPEI index (12 months scale). See Material and methods for details.
 

### 2019-12-10 man/tables/t_forest_history.Rmd
update ms

@@
 
-**Table S4.** Review of the forest and management history of the samplin site.
-An exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest.
-Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
+**Table S4.** Review of the forest and management history of the sampling sites.
+Historical documents were exhaustively reviewed to compile information on socio-economical activities affecting forests: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information on recent wildfires events and forest-management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
 

### 2019-12-10 man/tables/t_huber_res.Rmd
update ms

@@
 
-**Table S2.** Robust measures of central tendency of resilience indices for greenness (EVI) and tree-growth (BAI), grouped by drought events, site and interaction.
-Measures of central tendency are M-estimators based on Huber's Psi (see material and methods).
-95 % confidence intervals using 3000 bootstrap are included between parentheses.
+**Table S2.** Robust measures of central tendency of resilience indices for greenness (EVI) and tree growth (BAI), grouped by drought events, site, and interaction.
+Measures of central tendency are M-estimators based on Huber's Psi (see Material and methods).In parentheses are the 95% confidence intervals using 3000 bootstraps.
 *Total* corresponds to the average of 2005 and 2012.

### 2019-12-10 man/tables/t_robust_anovas.Rmd
update ms

--- 2019-08-05 man/tables/t_robust_anovas.Rmd
@@
 
-**Table 3.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree-growth (BAI) for the two drought events (in 2005 and 2012) and site.
+**Table S1.** Robust two-way ANOVAs of the resilience metrics of greenness (EVI) and tree growth (BAI) for the two drought events (in 2005 and 2012) and site.
 

### 2019-12-10 man/tables/t_sampled_plots.Rmd
update ms

@@
 Lat = latitude; Long = longitude.
-Dbh and height of all trees, Basal Area (BA), Density and SRD (Size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see methods).
+Dbh and height of all trees, Basal Area (BA), Density and SRD (size ratio proportional to distance) are computed for all trees within a 10-m radius of focal trees (see Material and methods).
 Temp.: annual average of mean monthly minimun and maximum temperatures.

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
update

@@
 ---
-title: "Oak resilience to drought and land use show how the ecological and geographical rear edges do not necessarily meet today"
+title: "Land-use legacies and high oak resilience to drought show how the ecological and geographical rear edges do not necessarily meet today"
 output:

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
Change title: Suggested by DN. Approved by coauthors

--- 2019-12-10 man/ecosystems/ms_ecosystems.Rmd
@@
 ---
-title: "Oak resilience to drought and land use show how the ecological and geographical rear edges do not necessarily meet today"
+title: "Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges"
 output:

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
change abstract. - incorporate suggestions by DN - remove two last items - include new item (RZ)

@@
 * Global change challenges ecosystems in dry locations transformed by intensive human use.
-Forest resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd.
-populations in the southern Iberian Peninsula was analyzed in relation to historical records of land use employing dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
+Resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd.
+populations in the southern Iberian Peninsula was analyzed in relation to historical records of land use, employing dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
 
@@
 Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events over the short and the long term.
-Resilience and response to climate followed a water-stress gradient.
-Primary and secondary growth were generally less resilient to drought in mesic than in xeric sites, whereas a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic (*i.e.* colder and wetter) high-elevation stands, but absent in the most xeric (*i.e.* warmer and drier) site.
+Resilience and the tree growth response to climate followed a water-stress gradient.
+A positive growth trend since the late 1970s was particularly evident in mesic (*i.e.* colder and wetter) high-elevation stands, but absent in the most xeric (*i.e.* warmer and drier) site.
 
-* Adult oak trees showed high resilience to drought despite being considered relict and at its rear edge.
-The trees varied in response to the changing climate along a narrow climatic gradient and showed that the ecological and geographical rear edges do not need to meet today in forests with strong land use legacies.
-Forest resilience including that of regeneration needs to be monitored at xeric of the species under future expected climatic scenarios given that ecological boundaries for species distribution can be narrow and that land-use legacies can play a key role in stand dynamics.
+* The high values of resilience observed in our study suggest that Q.
+pyrenaica populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge.
+These resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.
+This is particularly relevant for rear-edge populations where topographic and biophysical variability can facilitate the existence of refugia
 

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
changes at intro

@@
 
-Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of processes associated with increased intensity and duration of droughts [@Allen2010].
-In fact, the frequency and severity of drought events have increased in recent decades, along with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
+Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of more intense and longer droughts [@Allen2010].
+In fact, drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
 This is especially critical, since climate-change projections for the Mediterranean area forecast more frequent and severe extreme climate events [@Spinoni2017a].
@@
 
-The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005].
+The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution.
+In these cases, climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005].
 The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010].
-Yet it is crucial to accurately characterize the ecological rear-edge of species.
+Also, it is crucial to accurately characterize the ecological rear-edge of species.
 This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019].
@@
 
-Understanding the vegetation response to drought requires integrative and multidisciplinary strategies to assess the disturbances of forest at several scales [@Hartman2018].
 The assessment of resilience to climate disturbances such as droughts provides critical information concerning the capacity of the forests to maintain their structure and render valuable ecosystem services.
-The drought resilience of species is better assessed when several approaches, such as dendrochronology and remote sensing, are combined.
-Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and annual tree-ring widths can be used to study changes in growth as a response to drought at the individiual-tree scale [@Fritts1976].
-Remote sensing can be used to study the impact of drought on ecosystems and hence the stand level [*e.g.* @Zhang2013].
+The drought resilience of species is better assessed by a combination of several approaches, such as dendrochronology and remote sensing.
+Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and annual tree-ring widths can be used to study growth changes in response to drought among individiual trees [@Fritts1976].
+Remote sensing can be used to analyze the impact of drought on ecosystems and hence on the stand [*e.g.* @Zhang2013].
 The combination of both approaches (remote sensing and dendroecology) has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].
-Nonetheless, it is crucial to ascertain wether the responses at the tree level differ from those at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.
+Nonetheless, it is crucial to ascertain whether the responses at the tree level differ from those at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.
 
 In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd.
-(Pyrenean oak, *melojo*) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
+(Pyrenean oak) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
 We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades).

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
change melojo for Pyrenean oak

@@
 pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
-For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of *melojo* oak forests differ in their resistance, resilience and recovery to extreme drought events.
+For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
 
@@
 Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002].
-Eight melojo oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys.
+Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys.
 Today, *Q.
@@
 
-By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the melojo oak forests of Sierra Nevada and their resilience to several severe drought events in the recent decades.
+By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the *Q.
+pyrenaica* forests of Sierra Nevada and their resilience to several severe drought events in the recent decades.
 Our results indicated that these relict oak populations driven by historical land use are resilient to climate change at their present rear edge.

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
change hypothesis

@@
 We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades).
-Our main hypothesis is that these relict populations driven by historical land use are vulnerable to climate change at their present climatic (either elevational or latitudinal) rear edge, and hence will show low resilience to extreme drought along climatic gradients at a small scale within the rear edge.
+In the study area, the rear edge of relict populations is imposed by historical land use.
+Our main hypothesis is that, limited by their land-use legacy, these stands will show low resilience to extreme drought from climate change a small-scale gradient.
 To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q.
 pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
-For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.
+For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.
 

### 2020-03-09 man/ecosystems/ms_ecosystems.Rmd
update review DN

@@
 
-* The high values of resilience observed in our study suggest that Q.
-pyrenaica populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge.
+* The high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge.
 These resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.
@@
 Our main hypothesis is that, limited by their land-use legacy, these stands will show low resilience to extreme drought from climate change a small-scale gradient.
-To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q.
-pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
+To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Quercus pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
 For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.
@@
 ### Tree species and study site 
-*Quercus pyrenaica* forests extend throughout south-western France and the Iberian Peninsula reaching their southern limit in mountain areas of northern Morocco [@Franco1990].
+*Quercus pyrenaica* forests extend throughout south-western France and the Iberian Peninsula, reaching their southern limit in mountain areas of northern Morocco [@Franco1990].
 In the Iberian Peninsula, these forests occupy siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid, and hyperhumid ombroclimate.
-*Q.
-pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation.
+Pyrenean oak is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation.
 As a submediterranean species, it has lower drought tolerance than evergreen Mediterranean taxa [@delRio2007].
@@
 Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys.
-Today, *Q.
-pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005].
-They are the richest forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG].
+Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005].
+They are the richest forest formation of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG].
 These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002].
@@
 Monthly precipitation and minimum and maximum temperatures had a 0.25 x 0.25 º resolution for the 1950-2016 period.
-Grid cells were select to cover each sampled site.
+Grid cells were selected to cover each sampled site.
 The SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] index with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
 
-The Iberian Peninsula underwent several extreme drought episodes in the last three decades (*e.g*.
-1995, 1999, 2005, 2012) [@VicenteSerrano2014].
-The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], and they were characterized as extreme drought in our climatic data (Figure S1; Table S3).
+The Iberian Peninsula underwent several extreme drought episodes in the last three decades [*e.g.* 1995, 1999, 2005, 2012;  @VicenteSerrano2014].
+The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], appearing as extreme drought in our climatic data (Figure S1; Table S3).
 We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below).
@@
 ### Greenness data to assess ecosystem resilience
-Vegetation greenness of *Q.
-pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor.
+Vegetation greenness of *Quercus pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS sensor.
 EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m.
 MODIS EVI data were compiled for the period 2000 - 2016.
-We selected the pixels covering the distribution of *Q.
-pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
-Any values affected by clouds, snow, shadows or high content of aerosols, were filtered out following recommendations for mountain regions [@Reyes2015].
+We selected the pixels covering the distribution of *Quercus pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
+Any values affected by clouds, snow, shadows or high content aerosols, were filtered out following recommendations for mountain regions [@Reyes2015].
 
@@
 ### Field sampling and dendrochronological methods to assess individual tree resilience
-Trees were sampled during autumn 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1).
+Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1).
 For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the sampled aspect.
@@
 At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled.
-Two cores of 5 mm of diameter were taken from each tree at breast height (1.3 m) using an increment borer.
+Two cores of 5 mm in diameter were taken from each tree at breast height (1.3 m) using an increment borer.
 Diameter at breast height (DBH) and total height were measured for each tree.
@@
 Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].
-Differences between sites for height, DBH and competition indices were analyzed using non-parametric Kruskal-Wallis rank sum tests.
+Differences between sites for height, DBH, and competition indices were analyzed using non-parametric Kruskal-Wallis rank sum tests.
 When significant differences were detected, multiple comparisons were run using the Dunn's-test with Bonferroni adjustment to correct for significance.
@@
 ### Disturbance analyses 
-Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past.
+Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (suppressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past.
 Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@Rubino2004].
@@
 Site-disturbance chronologies were constructed by annually averaging the individual disturbance series.
-To separate growth peaks caused by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50% of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014].
+To separate growth peaks caused by disturbance events and expressing stand-wise disturbances from those caused by climate, we considered a threshold of 50% of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014].
 In addition, the history of the forest and management of our sampling sites was inferred from a detailed analysis of historical land-use changes.
@@
 To evaluate the effects of drought events on ecosystem resilience (using greenness data) and individual tree resilience (using BAI data), we used resilience indices proposed by @Lloret2011.
-The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$) quantifies the severity of the impact of the disturbance in the year it occurred.
-The Recovery index, computed as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity.
-Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels.
+The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought:PreDrought$) quantifies the severity of the impact of the disturbance in the year it occurred.
+The Recovery index, computed as the ratio between performance after and during disturbance ($Rc = PostDrought:Drought$), represents the ability to recover from disturbance relative to its severity.
+Finally, the Resilience index ($Rs = PostDrought:PreDrought$) is the capacity to reach pre-disturbance performance levels.
 The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event.
@@
 Tree-growth reduction followed a moisture gradient.
-The southern sites (CA-High and CA-Low) showed less tree-growth reductions than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site (Figure 3).
+The southern sites (CA-High and CA-Low) showed less tree-growth reduction than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site.
 
@@
 
-During the last two drought events, resilience metrics for greenness and tree-growth were significantly different between drought events (Table S1).
+During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1).
 The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years.
@@
 Nevertheless, differences appeared between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
-In addition, the spring maximum temperature was the most significant limitation for treemgrowth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of the current September positively influenced tree growth only in the northern population.
+In addition, the spring maximum temperature was the most significant limitation for tree growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of the current September positively influenced tree growth only in the northern population.
 
@@
 
-By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the *Q.
-pyrenaica* forests of Sierra Nevada and their resilience to several severe drought events in the recent decades.
+By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the *Quercus pyrenaica* forests of Sierra Nevada and their resilience to several severe drought events in the recent decades.
 Our results indicated that these relict oak populations driven by historical land use are resilient to climate change at their present rear edge.
@@
 In addition, other activities such as mining must have altered the forest structure, *e.g.* the SJ site has many small mines and quarries that were exploited intermittently throughout history.
-The release growth event discerned for the 1940s concurs with a period of maximum mining activity in this area (1925 to 1957) (Table S4), during wich timbers use increased for mine tunnels and furnaces, these also requiring large amounts of firewood to melt the mineral.
-This heavy exploitation of the neighboring forest resources must have affected a significant part of this oak woodland, as shown by growth of the remnant trees included in the northern site (Figure S2b).
+The release growth event discerned for the 1940s concurs with a period of maximum mining activity in this area (1925 to 1957), during which timber use increased for mine tunnels and furnaces, these also requiring large amounts of firewood to melt the mineral (Table S4).
+This heavy exploitation of the neighboring forest resources must have affected a significant part of this oak woodland, as shown by growth of the remnant trees at the northern site (Figure S2b).
+
 On the other hand, woodlands on the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks grow [@JimenezOlivencia2015].
-Firewood, charcoal, and acorns have been resources intensively exploited at southern sites, until at least the middle of the 20th-century, when these activities sharply decreased mainly due to rural abandonment and use of gas and fossil fuels [@Valbuena2013].
+Firewood, charcoal, and acorns have been intensively exploited at the southern sites, until at least the mid-20th century, when these activities sharply declined due mainly to rural abandonment and the use of gas and fossil fuels [@Valbuena2013].
 At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management type which has been applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
 
-The other release event observed for the SJ site during 1995-2000 was lower than during 1940, but also affected most trees (Figures 4, S2b).
+The other release event observed for the SJ site during the period 1995-2000 was lower than during 1940, but also affected most trees (Figures 4, S2b).
 No records of forest practices in this area over the last 30 years have been found [@Bonet2016obsnev_forest], and no logging has been recorded during the period 1995 - 2000 (F.J.
@@
 ### Sensitivity (resistance) of relict oaks to recent drought events
-Severe drought negatively affects both primary and secondary growth of *Q.
-pyrenaica* forests.
+Severe drought negatively affects both primary and secondary growth of *Quercus pyrenaica* forests.
 This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by the consistent radial-growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
 Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
-Similar to the situation with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
-pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
-Thus, our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
+As with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Quercus pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
+Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
 
 Greenness proved less sensitive to drought than did tree growth, particularly for drier sites.
-These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018], suggesting that the growth reduction could mediated by sink more than by source limitations [@Korner2013; @Fatichi2014].
+These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013; @Fatichi2014].
 Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
@@
 For example, the northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and the stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014].
-It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition, etc.; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
+It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
 Trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4).
-This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie very close to each other and overall both are considered to constitute the rear edge for the species.
+This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and overall both are considered to constitute the rear edge for the species.
 
@@
 ### Relict oaks show high resilience to recent drought events and long-term climatic variability 
-Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q.
-pyrenaica* for the last 16 years, which is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
+Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Quercus pyrenaica* for the last 16 years.
+This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
 For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014], which could be related to a non-linear positive effect of warming for the species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014].
+Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014].
+This could be related to a non-linear positive effect of warming for the species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014].
 Importantly, for rear edges threatened by climate change, negative growth trends would have been expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
 
-Althought the 2012 drought event was more severe and intense than that of 2005 (Table S3), resilience values for greenness and tree-growth were greater for 2012.
+Although the 2012 drought event was more severe and intense than that of 2005 (Table S3), resilience values for greenness and tree growth were greater for 2012.
 This could be due to the different timing of the two droughts.
@@
 
-The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests at Sierra Nevada reported elsewhere [@Valbuena2013; @Valbuena2017] appear to indicate the strong local adaptation of this oak.
+The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@Valbuena2013; @Valbuena2017] appear to indicate the strong local adaptation of this oak.
 Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017].
@@
 All this, together with the characteristic high resprouting ability of the species, would suggest a long-term persistence of those populations [@Bellingham2000].
-It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the trees express resilience at the species level or whether we would find vulnerability if we analyzed seedling regeneration, likewise other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].
+It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the trees express resilience at the species level or whether we would find vulnerability if we analyzed seedling regeneration, as in other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].
 
@@
 In summary, two main results stand out from this research.
-First, the high values of resilience observed in our study suggest that *Q.
-pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019].
+First, the high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019].
 The current niche is a result of land-use changes, which further complicate the definition of potential rear edges.
@@
 
-## Conclusions 
-The ecological and geographical read edges did not appear to fully match in our study.
-Severe drought events provoke major reductions in primary and secondary growth of *Q.
-pyrenaica* forests in the relict, rear-edge forest studied.
+## Concluding comments
+The ecological and geographical rear edges did not appear to fully match in our study.
+Severe drought events provoke major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the relict, rear-edge forest studied.
 However, we found no negative growth trends despite our expectation of vulnerability to climate change for these relict stands.
 Furthermore, we detected positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a steep positive trend of secondary growth at the tree level at the wettest site along the climatic gradient analyzed.
-The trees exhibited high resilience values in response to drought, particularly in the long-term scale, and resilience was directly related to drought severity.
+The trees exhibited high resilience values in response to drought, particularly in the long-term scale.
 These findings are consistent with other results showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*.
@@
 In this sense, our results highlight the importance of land-use legacies for highly transformed Mediterranean systems.
-This is relevant for tree species with a high sensitivity to climate change, such as *Q.
-pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these singular forests offer.
-As a follow-up to our study, research would be informative to disentangle how forest resistance and resilience might be affected by the combination of climate change and land-use legacies in highly transformed landscapes such as Mediterranean mountain ecosystems.
+This is relevant for tree species with a high sensitivity to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these singular forests offer.
 

### 2020-03-10 man/ecosystems/ms_ecosystems.Rmd
last changes

@@
 Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016].
-It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance, higher vulnerability, and thus higher risk of extinction than do populations at the core of the species' range [@VilaCabrera2019; @Pironon2016].
+It has been assumed that geographical and ecological rear-edge populations match.
+This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@VilaCabrera2019; @Pironon2016].
 Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014].
-In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine under what environmental conditions geographical marginality does not correspond with ecological marginality [@VilaCabreraJump2019GreaterGrowth].
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine under what environmental conditions the geographical rear edge does not correspond with the ecological one [@VilaCabreraJump2019GreaterGrowth].
 
@@
 Also, it is crucial to accurately characterize the ecological rear-edge of species.
-This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019].
+This can be particularly misleading in ecosystems where land-use legacies have strongly modified species distribution [@VilaCabrera2019].
 Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].
@@
 Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1).
-For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the sampled aspect.
+For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the site sampled.
 Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1).
@@
 Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].
-For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern aspects).
+For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern exposures).
 These tests were used because original and log-transformed data did not follow the assumptions of normality or homogeneity of variance [@Wilcox2012].
@@
 The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 3 and S3a).
-Only trees from the southern sites (*i.e.* the wetter aspect) showed significant positive growth trends since the late 1970s (Figure 3), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).
+Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 3), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).
 
-Drought events reduced radial growth for all sites (Figure S2a), this negative effect being particularly evident for the 1995 drought (the worst drought spell in our climatic record, Table S3), when the greatest growth reduction was recorded for all chronologies.
+Drought events reduced radial growth for all sites (Figure S2a).
+The strongest reduction in radial growth occurred in the 1995 drought (the worst drought spell in our climatic record, Table S3) for all sites.
 Tree-growth reduction followed a moisture gradient.
-The southern sites (CA-High and CA-Low) showed less tree-growth reduction than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site.
+The southern sites (CA-High and CA-Low) showed less tree-growth reduction than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site (CA-High).
 

### 2020-03-10 man/ecosystems/ms_ecosystems.Rmd
aspect / exposure

@@
 Our results indicated that these relict oak populations driven by historical land use are resilient to climate change at their present rear edge.
-However, resistance, resilience, and forest recovery to extreme drought events are strongly influenced by mountain aspect, local environmental conditions, and management legacies.
+However, resistance, resilience, and forest recovery to extreme drought events are strongly influenced by mountain exposure, local environmental conditions, and management legacies.
 This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance varies markedly along the rear edge under climate change.

### 2020-05-24 man/ecosystems/2_revision2/fig/aux/bai_vs_rwi.Rmd
update figures

---
output: 
  word_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r load_pkg}
library("tidyverse")
library("here")
library("ggrepel")
library("grid")
library("gridExtra")
library("patchwork")
library("lemon")
library("bookdown")
source(here::here('/script/R/custom_functions.R'))
```

```{r data}
huber_bai <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "SJ" ~ "0_SJ",
  site == "caH" ~ "2_CA-High",
  site == "caL" ~ "1_CA-Low"
)))

huber_b_bai <- 
  read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber_b.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(var_sorted = as.factor(
    case_when(
      var == "rt" ~ "0_rt",
      var == "rc" ~ "1_rc",
      var == "rs" ~ "2_rs",
))) %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "SJ" ~ "0_SJ",
  site == "caH" ~ "2_CA-High",
  site == "caL" ~ "1_CA-Low"
)))

huber_rwi <- read.csv(here::here("/out/anovas_resilience/huber_rwi", "robust_mhuber.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "SJ" ~ "0_SJ",
  site == "caH" ~ "2_CA-High",
  site == "caL" ~ "1_CA-Low"
)))

huber_b_rwi <- 
  read.csv(here::here("/out/anovas_resilience/huber_rwi", "robust_mhuber_b.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(var_sorted = as.factor(
    case_when(
      var == "rt" ~ "0_rt",
      var == "rc" ~ "1_rc",
      var == "rs" ~ "2_rs",
))) %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "SJ" ~ "0_SJ",
  site == "caH" ~ "2_CA-High",
  site == "caL" ~ "1_CA-Low"
)))
```

```{r}
df <- huber_bai %>% mutate(growth_var = "bai") %>% 
  bind_rows(huber_rwi %>% mutate(growth_var = "rwi"))
```

```{r}
# plot_params 
labelita <- c("0_rt" = "Resistance",
                                      "1_rc" = "Recovery", 
                                      "2_rs" = "Resilience")
item_site <- c("0_SJ" = "SJ", "2_CA-High" = "CA-High", "1_CA-Low" = "CA-Low")

pd <- position_dodge(width = .8)

ggplot(df, aes(x=site_sorted, y = M.Huber, 
               colour=as.factor(growth_var), 
               fill=as.factor(growth_var),
               ymin = lower.ci, 
               ymax = upper.ci)) + 
  geom_point(size=4, position = pd) + 
  facet_grid(disturb_year~var_sorted, 
             labeller = labeller(var_sorted = labelita)) +
  scale_fill_manual(values = c("black", "gray")) + 
  scale_color_manual(values = c("black", "gray")) + 
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size=9),
        legend.title = element_blank(),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 11, face = "bold.italic"), 
        plot.title = element_text(size = 12, face = "bold")) +
   theme_bw() + 
  geom_errorbar(width=.1, position = pd) + 
  xlab("") + ylab ("") + 
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size=9),
        legend.title = element_blank(),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 11, face = "bold.italic"), 
        plot.title = element_text(size = 12, face = "bold")) +
  scale_x_discrete(labels = item_site)

```

```{r}

dfres <- df %>% filter(var == "rs")

ggplot(dfres, aes(x=site_sorted, y = M.Huber, 
               colour=as.factor(growth_var), 
               fill=as.factor(growth_var),
               ymin = lower.ci, 
               ymax = upper.ci)) + 
  geom_point(size=4, position = pd) + 
  facet_wrap(~disturb_year) +
  scale_fill_manual(values = c("black", "gray")) + 
  scale_color_manual(values = c("black", "gray")) + 
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size=9),
        legend.title = element_blank(),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 11, face = "bold.italic"), 
        plot.title = element_text(size = 12, face = "bold")) +
   theme_bw() + 
  geom_errorbar(width=.1, position = pd) + 
  xlab("") + ylab ("Resilience") + 
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size=9),
        legend.title = element_blank(),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 11, face = "bold.italic"), 
        plot.title = element_text(size = 12, face = "bold")) +
  scale_x_discrete(labels = item_site)
```

### 2020-05-24 man/ecosystems/2_revision2/fig/aux/browing_corrected_comparison.R

@@ -1,0 +1,91 @@
+library("tidyverse")
+
+anomalias_evimean <- 
+  read.csv(file="./data/anomalies/anomalias_evimean.csv", header = TRUE, sep = ',')
+
+anomalias_evimean <- anomalias_evimean %>% 
+  mutate(
+    clu_pop = as.factor(case_when(
+      pop == 1 ~ "Camarate",
+      pop %in% c(2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
+      pop == 9 ~ 'out')),
+    clu_pop2 = as.factor(case_when(
+      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
+      pop == 9 ~ 'out'))) %>% 
+  filter(clu_pop != 'out')
+
+
+
+
+sa2005 <- anomalias_evimean %>% 
+  filter(y == 2005) %>% 
+  mutate(type = ifelse(sa < -1, "browning", 
+                       ifelse(sa > 1, "greening", "no change"))) %>% 
+  group_by(clu_pop2, type) %>% 
+  dplyr::summarise(count_clu = n()) %>% 
+  group_by(clu_pop2) %>% 
+  mutate(per = round(count_clu / sum(count_clu)*100, 2),
+         y = 2005) 
+
+sa2012 <- anomalias_evimean %>% filter(y == 2012) %>% 
+  mutate(type = ifelse(sa < -1, "browning", 
+                       ifelse(sa > 1, "greening", "no change"))) %>% 
+  group_by(clu_pop2, type) %>% 
+  dplyr::summarise(count_clu = n()) %>% 
+  group_by(clu_pop2) %>% 
+  mutate(per = round(count_clu / sum(count_clu)*100, 2),
+         y = 2012) 
+
+per_sa <- sa2005 %>% rbind(sa2012) %>% as.data.frame()
+
+
+## corrected 
+anomalias_evimeanc <- 
+  read.csv(file="./data/anomalies/anomalias_evimean_corrected.csv", header = TRUE, sep = ',')
+anomalias_evimeanc <- anomalias_evimeanc %>% 
+  mutate(
+    clu_pop = as.factor(case_when(
+      pop == 1 ~ "Camarate",
+      pop %in% c(2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
[…]

### 2020-05-24 analysis/evi/compute_anomalies.Rmd
update

@@
       iv_ref <- df_aux %>% 
-        filter(year != y) %>% 
+        # filter(year != y) %>% 
         summarise(mean(evi, na.rm=T))
@@
       sd_ref <- df_aux %>% 
-        filter(year != y) %>% 
+        # filter(year != y) %>% 
         summarise(sd(evi, na.rm=T))
@@
 
-write.csv(anomalias_evimean, file=paste(di, "/data/anomalies/anomalias_evimean.csv", sep=""), row.names = FALSE)
+write.csv(anomalias_evimean, file=paste(di, "/data/anomalies/anomalias_evimean_corrected.csv", sep=""), row.names = FALSE)
+
+# write.csv(anomalias_evimean, file=paste(di, "/data/anomalies/anomalias_evimean.csv", sep=""), row.names = FALSE)
+
 ```
@@
 write.csv(anomalias_composite, file=paste(di, "/data/anomalies/anomalias_composite.csv", sep=""), row.names = FALSE)
+
+a <- avg_sa_clu %>% 
+  mutate(method = "original") %>% 
+  bind_rows(avg_sa_cluc %>% mutate(method = "no remove"))
+
 ```
+
+```{r, eval=FALSE, echo=FALSE}
+df_aux <- evi[evi$iv_malla_modi_id == 144160, ]
+
+years <- unique(df$year)
+
+# Create empty df to store anomlay of composite j of all year 
+aux_anomaly <- data.frame()
+
+    for (y in years){
+      # get mean evi for the reference period
+      iv_ref <- df_aux %>% 
+        filter(year != y) %>% 
+        summarise(mean(evi, na.rm=T))
+
+      # Get sd ref for standardized anomaly
+      sd_ref <- df_aux %>% 
+        filter(year != y) %>% 
+        summarise(sd(evi, na.rm=T))
+
+      # get evi for the year
+      iv_year <- df_aux[df_aux$year == y, 'evi']
+      # To solve the problems with the year without composites (i.e 2000)
+      iv_year <- ifelse(length(iv_year) == 0, 0, iv_year)
+
+      # Compute anomaly
+      anomaly <- (iv_year - iv_ref)
+      names(anomaly) <- 'a'
+
+      # Compute normalized anomaly 
+      anomaly_nor <- ((iv_year - iv_ref) / (iv_year + iv_ref))*100
+      names(anomaly_nor) <- 'nora'
+
+      # Compute standardized anomaly
+      anomaly_std <- (iv_year - iv_ref) / sd_ref
+      names(anomaly_std) <- 'sa'
+
+      # Create dataframe 
+      aux <- cbind(y, iv_year, iv_ref, anomaly, anomaly_nor, anomaly_std)
+      aux_anomaly <- rbind(aux_anomaly, aux)
+      }
+
+  # Add name of the pixel 
+  aux_anomaly <- aux_anomaly %>% 
+    mutate(iv_malla_modi_id = i) %>% 
+    mutate(evi_mean_cte = mean(iv_year))
+
+```

### 2020-05-24 man/figures/0_figures_ecosystems.Rmd
update

@@
 ```
+
+**Figure 1**.
+Distribution of *Quercus pyrenaica* forests in the Iberian Peninsula (a) and in Sierra Nevada mountain range (b).
+Different colors indicate oak-population clusters identified in Sierra Nevada [@PerezLuque2015].
+For each population, a grid with the MODIS pixels is shown (see Material and methods).
+Detailed location of the dendroecological sampling sites: northern (San Juan, SJ) (c), and southern ones (Cáñar: CA-Low and CA-High) (d).
+Color orthophotography of 2009 from Regional Ministry of the Environment.
+
+**Figure 2.** ***a)*** EVI standardized anomaly during the period 2000-2016 for northern and southern populations.
+Error bars show standard error.
+See main text for details on EVI calculation.
+***b)*** Percentage of pixels showing browning, greening or no changes during the 2005 and 2012 drought events according to EVI standardized anomalies.
+See main text for an explanation of greening and browning.
+
+**Figure 3.** ***a)*** Basal Area Increment (BAI) chronologies of *Q.
+pyrenaica* for northern population (SJ; *red*) and southern ones: low-elevation (CA-Low; *green*) and high-elevation (CA-High, *blue*) sites.
+Shading areas correspond to standard error of the mean.
+Number of series is displayed in the upper plot.
+Only years replicated with # series > 5 are shown.
+Linear trends since 1975 are indicated for all sites (numbers indicate $r^2$ values; asterisks indicate significant linear trend, *p < 0.001*).
+***b)*** Comparison of median growth change ($GC$) following @Nowacki1997 for *Q.
+pyrenaica* sites.
+Dashed black lines indicate a threshold of 50% of GC (see Material and methods).
+Note that y-axes do not correspond in all of the three panels for the sake of clarity.
+Error bars indicate standard error.
+
+**Figure 4.** ***a)*** Resilience metrics of tree-growth for eight severe drought events since 1950 (see main text for details) as a function of drought severity.
+Points indicate resilience metrics for oak populations: SJ (*red*), CA-High (*blue*) and CA-Low (*green*).
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
+Gray lines represent overall relationships for each Resilience metrics.
+***b)*** Comparison of the response of  *Q.
+pyrenaica* forests to drought in terms of resistance, recovery, and resilience of greenness (b) and tree growth (c).
+For EVI, northern populations (*red circle*) were compared with southern ones (*green circle*).
+For BAI, the more xeric northern population (San Juan, SJ; *red circle*) was compared with the two southern populations, Cáñar-High (CA-High; *blue circle*) and Cáñar-Low (CA-Low; *green circle*).
+Different letters indicate significant *post hoc* differences between groups (see Material and methods for details).
+
+#####
+**Figure 1**
+![](../../out/fig/geomap/mapa_situacion_v2.jpg){height=750px} 
+#####
+**Figure 2.**
+![](../../man/figures/svg/fig_2ab.jpg)
+
+#####
+**Figure 3**
+
+![](../../man/figures/svg/cronos_bai_gc.jpg)
+
+#####
+**Figure 4**
+![](../../man/figures/svg/resiliences.jpg)
+
+#####
+
+**Figure S1.** ***a)*** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
+Points represent the mean, and error bars the standard error.
+The black line indicates mean for the entire period (585 mm).
+The red lines represent -1 and -2 standard deviation (dotted and dashed lines, respectively).
+The blue lines represent +1 and +2 standard deviation (dotted and dashed lines, respectively).
+Years with average values below -1SD are labeled.
+Data from 28 meteorological stations distributed around the Sierra Nevada area (from the National Spanish Meteorological Services, AEMET).
+***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (blue line) and 2011-2012 (red line).
+The boxplot representing the average from 1950-2015 period.
+Data from meteorological station Granada, Base Aérea.
+***b)*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardized Precipitation-Evapotranspiration Index (SPEI).
+Data from Global SPEI database (http://spei.csic.es/database.html).
+We took the SPEI data for a 12-month scale and for all 0.5º grid cells covering Sierra Nevada.
+Horizontal gray bars indicate the years 2005 and 2012.
+
+**Figure S2.** ***a)*** Residual tree-ring chronologies determined for the *Q.
+pyrenaica* sites.
+Dashed red lines indicate the start of the reliable period (EPS > 0.85).
+Dotted black lines show the severe drought years identified in our climatic data (Table S3 and Figure S1).
+***b)*** Percentage of *Q.
+pyrenaica* trees affected by GC > 50 % by site.
+*Black* line shows number of trees (right-axis).
+Data for number of trees > 2 is shown.
+
+**Figure S3.** ***a)*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time domains after pre-filtering the time series with increasing size of the moving-average window (1 to 40 years).
+Each site chronology was smoothed using centered moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair of chronologies was calculated.
+Significance was tested using 1000 bootstrap replicates and with 95% confidence intervals built using the R package `boot`.
+***b)*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
+
+**Figure S4.** EVI annual profile (average of the period 2000-2016) for *Q.
+pyrenaica* forests in Sierra Nevada and drought events.
+Horizontal bars correspond to the most severe droughts for Sierra Nevada since 1900 (computed as in Table S3).
+Their position indicates the start and end months of each drought event.
+Bars lengths show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017].
+
+**Figure S5.** Resilience metrics of the tree growth for severe drought events since 1950 (excluding 1995 drought event).
+*Left*: Resistance; *Center*: Recovery; *Right*: Resilience.
+Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*).
+Resilience metrics were computed for each population (sample depth > 10) and drought event.
+The gray line represents overall relationship for each Resilience metrics.
+
+**Figure S6.** Correlation coefficients found by relating tree-ring residual chronologies (RWI) of *Q.
+pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimum (b) and maximum (c) temperatures.
+*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation southern site (CA-High).
+Asterisks indicate significant ($P < 0.05$) correlation coefficients.
+
+**Figure S1**
+![](../../man/figures/svg/fig_s1ab.jpg)
+
+##### 
+**Figure S2**
+![](../../man/figures/svg/fig_s2ab.jpg)
+
+##### 
+**Figure S3**
+![](../../man/figures/svg/fig_s4ab.jpg)
+
+##### 
+**Figure S4**
+
+![](../../man/figures/svg/fig_s7.jpg)
+
+##### 
+**Figure S5**
+![](../../man/figures/svg/fig_s8.jpg)
+
+#####
+**Figure S6**
+![](../../man/figures/svg/fig_8.jpg)
+
+```{r, eval=FALSE, echo=FALSE}
+
+OLD
 
@@
 ![](../../man/figures/svg/fig_7.jpg)
-
-#####
-
-**Figure 7.** Correlation coefficients found by relating tree-ring residual chronologies (RWI) of *Q.
-pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimum (b) and maximum (c) temperatures.
-*green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation southern site (CA-High).
-Asterisks indicate significant ($P < 0.05$) correlation coefficients.
-
-![](../../man/figures/svg/fig_8.jpg)
-
-#####
-
-**Figure S1.** ***a)*** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017.
-Points represent the mean, and error bars the standard error.
-The black line indicates mean for the entire period (585 mm).
-The red lines represent -1 and -2 standard deviation (dotted and dashed lines, respectively).
-The blue lines represent +1 and +2 standard deviation (dotted and dashed lines, respectively).
-Years with average values below -1SD are labeled.
-Data from 28 meteorological stations distributed around the Sierra Nevada area (from the National Spanish Meteorological Services, AEMET).
-***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (blue line) and 2011-2012 (red line).
-The boxplot representing the average from 1950-2015 period.
-Data from meteorological station Granada, Base Aérea.
-***b)*** Drought severity in Sierra Nevada for the 1901-2016 period based on the Standardized Precipitation-Evapotranspiration Index (SPEI).
-Data from Global SPEI database (http://spei.csic.es/database.html).
-We took the SPEI data for a 12-month scale and for all 0.5º grid cells covering Sierra Nevada.
-Horizontal gray bars indicate the years 2005 and 2012.
-
-![](../../man/figures/svg/fig_s1ab.jpg)
-
-#####
-
-**Figure S2.** ***a)*** Residual tree-ring chronologies determined for the *Q.
-pyrenaica* sites.
-Dashed red lines indicate the start of the reliable period (EPS > 0.85).
-Dotted black lines show the severe drought years identified in our climatic data (Table S3 and Figure S1).
-***b)*** Percentage of *Q.
-pyrenaica* trees affected by GC > 50 % by site.
-*Black* line shows number of trees (right-axis).
-Data for number of trees > 2 is shown.
-
-![](../../man/figures/svg/fig_s2ab.jpg)
-
-#####
-
-**Figure S3.** ***a)*** Correlation among site chronologies (CA-High, CA-Low and SJ) in different time domains after pre-filtering the time series with increasing size of the moving-average window (1 to 40 years).
-Each site chronology was smoothed using centered moving averages with different window sizes (1 to 40 years), and then Pearson's correlation coefficient between the each pair of chronologies was calculated.
-Significance was tested using 1000 bootstrap replicates and with 95% confidence intervals built using the R package `boot`.
-***b)*** Correlation between indices of resilience (*Rt*, resistance; *Rc*, recovery; *Rs*, Resilience) using periods of several lengths (2, 3 and 4 years after a drought).
-
-![](../../man/figures/svg/fig_s4ab.jpg)
-
-##### 
-**Figure S4.** EVI annual profile (average of the period 2000-2016) for *Q.
-pyrenaica* forests in Sierra Nevada and drought events.
-Horizontal bars correspond to the most severe droughts for Sierra Nevada since 1900 (computed as in Table S3).
-Their position indicates the start and end months of each drought event.
-Bars lengths show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017].
-
-![](../../man/figures/svg/fig_s7.jpg)
-
-#####
-
-**Figure S5.** Resilience metrics of the tree growth for severe drought events since 1950 (excluding 1995 drought event).
-*Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*).
-Points indicate resilience metrics for oak populations: SJ (*blue*), CA-High (*red*) and CA-Low (*green*).
-Resilience metrics were computed for each population (sample depth > 10) and drought event.
-The gray line represents overall relationship for each Resilience metrics.
-
-![](../../man/figures/svg/fig_s8.jpg)
+```

### 2020-05-24 man/figures/cronos_bai_gc.Rmd
update

---
output:
  word_document
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r load_pkg}
library("tidyverse")
library("here")
library("ggrepel")
library("scales")
library("grid")
library("gridExtra")
library("patchwork")
library("stringr")
library("pander")
library("lemon")
library("magrittr")
source(here::here('/script/R/custom_functions.R'))
```

```{r}
cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')

cro_sites <- cro_sites %>% mutate(site = 
  case_when(
    site == "CA_Low" ~ "CA-Low", 
    site == "CA_High" ~ "CA-High",
    site == "SJ" ~ "SJ",
  )
)

# To get p.value of the model 
# https://stackoverflow.com/questions/5587676/pull-out-p-values-and-r-squared-from-a-linear-regression
lmp <- function(modelobject) {
    if (class(modelobject) != "lm") stop("Not an object of class 'lm' ")
    f <- summary(modelobject)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}

year_init <- 1975

# Lm CA-High
dfCaH <- cro_sites %>% 
  filter(year >= year_init) %>% 
  filter(site == 'CA-High')

mCaH <- lm(bai_mean ~ year, dfCaH)
r2caH <- format(summary(mCaH)$r.squared, digits = 3)
pCaH <- lmp(mCaH)

# Lm CA-Low 
dfCaL <- cro_sites %>% 
  filter(year >= year_init) %>% 
  filter(site == 'CA-Low')

mCaL <- lm(bai_mean ~ year, dfCaL)
r2caL <- format(summary(mCaL)$r.squared, digits = 3)
pCaL <- lmp(mCaL)

# Lm SJ
dfSJ <- cro_sites %>% 
  filter(year >= year_init) %>% 
  filter(site == 'SJ')

mSJ <- lm(bai_mean ~ year, dfSJ)
r2SJ <- format(summary(mSJ)$r.squared, digits = 3)
pSJ <- lmp(mSJ)

lms <- data.frame(year = as.numeric(c(2018, 2018, 2018)),
                     site = c("CA-Low", "CA-High", "SJ"), 
                     r2 = as.numeric(c(r2caL, r2caH, r2SJ)),
                     pvalue = as.numeric(round(c(pCaL, pCaH, pSJ), 4)),
                     y = as.numeric(
                       c(as.numeric(predict(mCaL)[40]), 
                           as.numeric(predict(mCaH)[40]), 
                           as.numeric(predict(mSJ)[40]))))
# lms$y <- lms$y / 100
lms <- lms %>% mutate(
  y = y/100,
  pc = ifelse(pvalue < 0.001, "p < 0.001", as.character(paste0("p = ", pvalue))),
  label = round(r2,3), 
  labelp = ifelse(pvalue < 0.001, paste0(label, "**"), label))

```

```{r plot_cronobai}
# cores 
cronos_bai <- cro_sites %>% filter(nseries > 5) %>% 
  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
  theme_bw() + ylab('BAI') + xlab("") + 
  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
                  ymax = (bai_mean + bai_se)/100,
                  fill=site), alpha=.4, colour=NA) +
  geom_line() + 
  geom_point(size=.8) + 
  scale_colour_manual(values = colores_den) +
  scale_fill_manual(values = colores_den) +
  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020), sec.axis = dup_axis()) +
  geom_smooth(data = subset(cro_sites, year > year_init & site == 'CA-High'), 
              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5) +
  geom_smooth(data = subset(cro_sites, year > year_init & site == 'CA-Low'), 
              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5) +
  geom_smooth(data = subset(cro_sites, year > year_init & site == 'SJ'), 
              aes(x=year, y=bai_mean/100, colour=site), method = 'lm', se=FALSE, size = .5)  +
  geom_text_repel(data = lms, 
                  aes(x=year, y = y, colour = site, label = labelp),
                  direction = "x", nudge_x = Inf, nudge_y = 1, show.legend = FALSE, 
                  size = 3) + 
  theme(panel.grid = element_blank(),
        axis.text.x=element_text(size=6),
        legend.position = c(0.2,.8),
        strip.background = element_blank(),
        axis.ticks.length.x.top = unit(-3.5, "pt"),
        axis.title.x.top = element_blank(), 
        axis.text.x.top = element_blank())  
```

```{r plot_samp}
  # scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
  # geom_line(aes(x=year, y=n_trees+40, colour=site)) 
samp <- cro_sites %>% ggplot(aes(x=year, y=n_trees, colour=site)) + 
  theme_bw() + ylab('# trees') + 
  geom_line() + 
  scale_colour_manual(values = colores_den) +
  scale_fill_manual(values = colores_den) +
  scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020), 
                     sec.axis = dup_axis(), position = "top") +
  theme(panel.grid = element_blank(),
        legend.position = "none", 
        axis.title.x=element_blank(),
        axis.text.x.top = element_text(size=6), 
        axis.text.x.bottom = element_blank(), 
        axis.ticks.x.bottom =element_blank(),
        axis.line.x.bottom = element_blank(), 
        #panel.border = element_blank(), 
        # axis.line.y.right = element_line(), 
        # axis.line.x.top = element_line(), 
        # axis.line.y.left = element_line(),
        plot.margin = unit(c(0,0,-4,0), "pt")
        # axis.ticks.x=element_blank(), 
        # plot.margin = margin(t = 5, r = 0, b = -5, l = 0, unit = "pt")
        ) 
  # theme(plot.margin = unit(c(1,5,-30,6),units="points"),
  #       axis.title.y = element_text(vjust =0.25))
```

```{r datagc}
gc_todos <- read.csv(here::here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')

# customize plot (corrected scale)
pos_neg_log <- scales::trans_new('signed_log', 
                                 transform = function(x) sign(x)*log(abs(x)),
                                 inverse = function(x) sign(x)*exp(abs(x)))

plotGC <- function(df, y, yse, x, my_ylab, 
                   micolor, limitesX, ...){ 
  # Get error bar aesthetic
  yerrorbar <- aes_string(ymin = paste(y, yse, sep = '-'),
                         ymax = paste(y, yse, sep = '+'))

  p <- ggplot(df, aes_string(x = x, y = y)) +
    geom_errorbar(mapping = yerrorbar, colour = micolor, width = .35, size = .2, 
                  na.rm = TRUE) + 
    geom_bar(stat = 'identity', fill = micolor, colour = micolor) +
    geom_hline(yintercept = 0, colour = 'black', lwd = 0.2) + 
    ylab(my_ylab) +
    # Add line to %GC = 50 
    geom_hline(yintercept = c(50, -50), linetype=2, colour='black', lwd=0.2) +
    theme_bw() +
    theme(panel.grid = element_blank(),
          axis.text.y = element_text(size = 5.5),
          axis.text.x = element_text(size = 6)) +
    coord_cartesian(xlim=limitesX)
    # Customize limits and breaks to min of year data 
  p
}

# ---------- Min and Max axis        
# Get min and max of year range and create vector of 10 years for labelling

# ajustar limites X para combinar con otra figura 
limitesX <- c(1815,2020)
# ajustar rango para combinar con otra figura 
mi_scalex <- scale_x_continuous(breaks = seq(1815, 2020, by=10), limits=c(1815,2020))

# xrange <- seq(round(min(gc_todos$year) / 10) * 10, 
#     round(max(gc_todos$year) / 10) * 10, by = 10)
# 
# limitesX <- c(min(gc_todos$year), max(gc_todos$year))

# Get automatically limits of y for caH anc caL
minimoY <- min((gc_todos %>% 
      dplyr::filter(site != 'gc_sj' & gc_md < 0) %>% 
      mutate(error = gc_md - gc_md_se))$error, na.rm = TRUE)

maximoY <- max((gc_todos %>% 
      dplyr::filter(site != 'gc_sj' & gc_md > 0) %>% 
      mutate(error = gc_md + gc_md_se))$error, na.rm = TRUE)

# Establece rango de datos para y 
mismo_y <- scale_y_continuous(limits=c(minimoY, maximoY), 
                            breaks = c(-100, -50, 0, 50, 50, 100, 150))

```

```{r plot_caH}
pcaH <- gc_todos %>% dplyr::filter(site == 'gc_caH') %>% 
  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
         my_ylab = NULL, micolor = colores_den[1], limitesX = limitesX) + 
  mismo_y + xlab('') + mi_scalex + 
  # scale_x_continuous(breaks = xrange, position = 'bottom') + 
  annotate("text", x = 1990, y = 120, label = "CA-High", fontface = 2, size = 3)
```

```{r plot_caL}
pcaL <- gc_todos %>% dplyr::filter(site == 'gc_caL') %>% 
  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year',
         my_ylab = '% GC', micolor = colores_den[2], limitesX = limitesX) + 
  mismo_y +  xlab('') + mi_scalex + 
  # scale_x_continuous(breaks = xrange, position = 'bottom') +  
  annotate("text", x = 1990, y = 120, label = "CA-Low", fontface = 2, size = 3) 
```

```{r plot_SJ}
psj <- gc_todos %>% dplyr::filter(site == 'gc_sj') %>% 
  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
         my_ylab = NULL, micolor = colores_den[3], limitesX = limitesX) + 
  xlab('') + xlab('Year') + mi_scalex + 
  # scale_x_continuous(breaks = xrange, position = 'bottom') + 
  scale_y_continuous(trans = pos_neg_log, 
                     breaks=c(-1000, -500, -100, 1, 100, 500)) + 
  geom_hline(yintercept = 1,  lwd=0.2)+
  annotate("text", x = 1990, y = 440, label = "SJ", fontface = 2, size = 3)

```

```{r}
samp <- samp + labs(tag='a')
cronos_bai <- cronos_bai + theme(plot.margin = unit(c(0,0,-4,0), "pt"))
pcaH <- pcaH + theme(plot.margin = unit(c(0,0,-5,0), "pt")) + labs(tag = 'b')
pcaL <- pcaL + theme(plot.margin = unit(c(0,0,-5,0), "pt"))
psj <- psj +  theme(plot.margin = unit(c(0,0,-4,0), "pt"))
```

```{r, fig.height=10, dpi = 300, out.width = 0.8}
cronos_bai_gc <- 
(samp / cronos_bai) + pcaH + pcaL + psj +
  plot_layout(ncol = 1, heights = c(0.13/2, 0.87/2, rep(1/3, 3)/2))

```

```{r, fig.height=10, dpi = 300, out.width = 0.8}
cronos_bai_gc 
```

```{r}
saveImage(cronos_bai_gc, path="./man/figures/svg/", dpi=300, w=25, h=27, fig= "cronos_bai_gc", u = "cm") 
```

**Figure 3.** a) Basal Area Increment (BAI) chronologies of *Q. pyrenaica* for northern population (SJ; *red*) and southern ones: low-elevation (CA-Low; *green*) and high-elevation (CA-High, *blue*) sites. Shading areas correspond to standard error of the mean. Number of series is displayed in the upper plot. We only show years replicated with # series > 5. Linear trends since 1975 are shown all sites (numbers indicate $r^2$ values; asterisks indicate significant linear trend, *p < 0.001*). b) Comparison of median growth change ($GC$) following @Nowacki1997 for *Q. pyrenaica* sites. Dashed black lines indicate a threshold of 50 % of GC (see material and methods). Note that y-axes do not correspond in all of the three panels for the sake of clarity. Error bars indicate standard error.

### 2020-05-24 man/figures/resiliences_figure_juntas.Rmd
update

---
output:
  word_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r load_pkg}
library("tidyverse")
library("here")
library("scales")
library("patchwork")
library("ggrepel")
library("broom")
library("cowplot")
library("purrr")
library("ggpubr")
library("rstatix")
library("ggpmisc")
source(here::here('/script/R/custom_functions.R'))
```

```{r data}
df <- read.csv(file=here::here('/out/bai_resiliences_severes' ,'resiliences_bai_droughts.csv'), header=TRUE)

# peores sequias 
dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
dy <- dy %>% arrange(desc(d_duration))
# solamente aquellos mayor de 2 meses 
dys <- dy %>% filter(d_duration > 2) %>% filter(maxyear > 1940 ) %>% as.data.frame() 
# dys <- dys[,'maxyear']

sdf <- df %>% 
  filter(samps >= 10) %>% 
  group_by(disturb_year) %>% 
  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
  inner_join(dy, by=c("disturb_year" = "maxyear")) 
```

```{r some_functions}
#https://gist.github.com/ottadini/6882677 
lm_eqn = function(m) {
  eq2 <- substitute(~~italic(r)^2~"="~rvalue*","~italic(p)~"="~pvalue,  
                    list(rvalue = formatC(summary(m)$r.squared, digits = 3, format = "f"), 
                         pvalue = formatC(summary(m)$coefficients[2,4], digits = 3, format = "f")))

  as.character(as.expression(eq2));                 
}

fit_model <- function(x){
  lm(value ~ d_severity, data = x)
}

# Get R2 and p.value 
get_rsq <- function(mod) broom::glance(mod)$r.squared
get_pvalue <- function(mod) broom::glance(mod)$p.value

```

**Figure 6.** Resilience metrics of tree-growth for eigth severe drought events since 1950 (see main text for details) as a function of drought severity. *Left*: Resistance (*Rt*); *Center*: Recovery (*Rc*); *Right*: Resilience (*Rs*). Points indicate resilience metrics for oak populations: SJ (*red*), CA-High (*blue*) and CA-Low (*green*). Resilience metrics were computed for each population (sample depth > 10) and drought event. Gray lines represent overall relationships for each Resilience metrics.

```{r}
# Include 1995
dfall <- df %>% 
  filter(samps >= 10) %>% 
  group_by(disturb_year, site) %>% 
  filter(disturb_year > 1940) %>% 
  summarise_at(c("rt_mean", "rs_mean", "rc_mean", "rrs_mean"), 
               funs(mean(., na.rm = TRUE), sd(., na.rm=TRUE),se=sd(., na.rm=TRUE)/sqrt(n()))) %>% 
  inner_join(dy, by=c("disturb_year" = "maxyear")) %>% 
  dplyr::select(disturb_year, site, rt_mean_mean, rs_mean_mean, rc_mean_mean, 
               d_severity) %>% 
  rename_at(vars(ends_with("_mean")), funs(str_remove(., "_mean_mean"))) %>% 
  rename("a.rt" = "rt", "b.rc" = "rc", "c.rs" = "rs") %>% 
  gather("res_index", "value", a.rt, b.rc, c.rs)

p <- ggplot(dfall, aes(x = d_severity, y = value, colour=site, 
                    fill=site, label = disturb_year)) + 
  geom_smooth(method='lm', se=FALSE) + 
  geom_smooth(method='lm', se=FALSE, aes(group=1),  size = 1, colour = "grey") +
  geom_point(size=1.5, pch = 21) +
  geom_text_repel(nudge_x = 0.5, size = 2, show.legend = FALSE) +
  theme_bw() +
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size=9),
        legend.position = "none",
         #legend.position = "bottom",
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 11, face = "bold.italic")) +
  scale_colour_manual(values = colores_den) +
  scale_fill_manual(values = colores_den) +
  xlab("Drought Severity") + ylab("") + 
  xlim(5,25) + ylim(0.3,3.8) +
  facet_wrap(~res_index, labeller = as_labeller(c("a.rt" = "Resistance", 
                                                  "b.rc" = "Recovery", 
                                                  "c.rs" = "Resilience"))) +
  stat_fit_glance(method = "lm",
                  method.args = list(formula = y ~ x),
                  aes(label = sprintf('r^2~"="~%.3f~~italic(p)~"="~%.3f',
                      stat(r.squared), stat(p.value))),
                  parse = TRUE) 
# +
#   guides(fill = guide_legend(override.aes = list(linetype = 0)),
#         color = guide_legend(override.aes = list(size=3)))

# Add global eq
dfaux <- dfall %>%
  group_by(res_index) %>% 
  nest() %>%
  mutate(model = map(data, fit_model)) %>% 
  mutate(eq = map(model, lm_eqn))

res_long_term_sites_all <- p + geom_text(data = dfaux, 
              aes(x = 20, y = 0.3, label = eq), 
              inherit.aes = FALSE, parse = TRUE, 
              col = "gray40") + labs(tag="a)") +
  theme(plot.tag = element_text(face = "bold"))

```

```{r load_pkg}
library("grid")
library("gridExtra")
library("patchwork")
library("lemon")
library("bookdown")
```

```{r}
huber_evi <- 
  read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber.csv"), header = TRUE) %>% 
  filter(var != "rrs") %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "N" ~ "0_N",
  site == "S" ~ "1_S"
)))

huber_b_evi <- 
  read.csv(here::here("/out/anovas_resilience/huber_evi", "robust_mhuber_b.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(var_sorted = as.factor(
    case_when(
      var == "rt" ~ "0_rt",
      var == "rc" ~ "1_rc",
      var == "rs" ~ "2_rs",
))) %>% 
mutate(site_sorted = as.factor(
  case_when(
  site == "N" ~ "0_N",
  site == "S" ~ "1_S"
)))

huber_bai <- read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "SJ" ~ "0_SJ",
  site == "caH" ~ "2_CA-High",
  site == "caL" ~ "1_CA-Low"
)))

huber_b_bai <- 
  read.csv(here::here("/out/anovas_resilience/huber_bai", "robust_mhuber_b.csv"), header = TRUE) %>%
  filter(var != "rrs") %>% 
  mutate(var_sorted = as.factor(
    case_when(
      var == "rt" ~ "0_rt",
      var == "rc" ~ "1_rc",
      var == "rs" ~ "2_rs",
))) %>% 
  mutate(site_sorted = as.factor(
  case_when(
  site == "SJ" ~ "0_SJ",
  site == "caH" ~ "2_CA-High",
  site == "caL" ~ "1_CA-Low"
)))
```

```{r}
# plot_params 
pd <- position_dodge(width = 1)
aes_main <- aes(x=site_sorted, y = M.Huber, 
                ymin = lower.ci, ymax = upper.ci,
                fill = site_sorted, colour = site_sorted)
aes_aux <- aes(shape = as.factor(disturb_year), 
               label = Letter)
labelita <- as_labeller(c("0_rt" = "Resistance",
                                      "1_rc" = "Recovery", 
                                      "2_rs" = "Resilience"))

item_NS <- c("0_N" = "N", "1_S" = "S")
item_site <- c("0_SJ" = "SJ", "2_CA-High" = "CA-High", "1_CA-Low" = "CA-Low")

theme_res <- list(
  geom_errorbar(width=.1, position = pd),
  geom_point(size=7, shape=21),
  geom_text(aes(label = str_to_upper(Letter)), fontface = "bold", colour = "white"), 
  theme_bw(), xlab(""), ylab (""), 
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size=9),
        legend.title = element_blank(),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 11, face = "bold.italic"), 
        plot.title = element_text(size = 12, face = "bold")),
  scale_shape_manual(values = c(24,22)),
  facet_wrap(~var_sorted, scales = "free_y", labeller = labelita))

# PLOT EVI 
pevi <- ggplot(huber_b_evi, aes_main) + 
  # by disturb year 
  geom_errorbar(data=huber_evi, aes_aux, width=.1, position = pd) +
  geom_point(data=huber_evi, aes_aux, position=pd, size = 4.5, fill="white") + 
  geom_text(data = huber_evi, aes_aux, position = pd, size = 2.5) +
  scale_x_discrete(labels = item_NS) +
  scale_fill_manual(values = colores_den[3:2], labels = item_NS) +
  scale_colour_manual(values = colores_den[3:2], labels = item_NS) + theme_res +
  labs(tag="b) EVI") +
  theme(legend.position = "top", 
        plot.tag = element_text(margin = margin(b = -20),
                                face = "bold")
  )

# PLOT BAI
pbai <- ggplot(huber_b_bai, aes_main) + 
  # by disturb year 
  geom_errorbar(data=huber_bai, aes_aux, width=.1, position = pd) +
  geom_point(data=huber_bai, aes_aux, position=pd, size = 4.5, fill="white") + 
  geom_text(data = huber_bai, aes_aux, position = pd, size = 2.5) +
  scale_x_discrete(labels = item_site) + 
  scale_fill_manual(values = rev(colores_den), labels = item_site) +
  scale_colour_manual(values = rev(colores_den), labels = item_site) + 
  theme_res +  
  labs(tag="c) BAI") +
  theme(legend.position = "bottom", 
        plot.tag = element_text(
                                face = "bold")
        )

```

```{r}
resilience <- res_long_term_sites_all + pevi + pbai + plot_layout(ncol=1) + theme(plot.margin = margin(t=-5,r=0,b=-2,l=0, "pt"))
```

```{r, fig.width=6, dpi = 300, out.width = 0.8}
resilience
```

```{r}
# SAVE as eps 
saveImage(resilience, path="./man/figures/svg/", dpi=300, w=25, h=29, fig= "resiliences", u = "cm") 
```

### 2020-05-24 man/figures/std_anomalies.Rmd
update

@@
 
+```{r}
+## Reviewer 1. Is it necessary/correct to remove year i from the calculation of EVImean, ref?
+
+# From explore_anomalies.Rmd 
+# Read data 
+anomalias_evimeanc <- read.csv(here("data/anomalies", "anomalias_evimean_corrected.csv"),header = TRUE, sep = ',')
+
+anomalias_evimeanc <- anomalias_evimeanc %>% 
+  mutate(
+    clu_pop = as.factor(case_when(
+      pop == 1 ~ "Camarate",
+      pop %in% c(2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
+      pop == 9 ~ 'out')),
+    clu_pop2 = as.factor(case_when(
+      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
+      pop %in% c(6,7,8) ~ 'Southern slope',
+      pop == 9 ~ 'out'))) %>% 
+  filter(clu_pop != 'out')
+
+# Standardized Anomalies 
+avg_sa_cluc <- anomalias_evimeanc %>% 
+  group_by(clu_pop2, y) %>% 
+  dplyr::summarise(mean=mean(sa, na.rm=T),
+            sd = sd(sa, na.rm=T),
+            se = sd/sqrt(length(sa))) %>% 
+ mutate(signo = ifelse(mean >= 0, 'pos', 'neg'))
+
+a <- avg_sa_clu %>% 
+  mutate(method = "original") %>% 
+  bind_rows(avg_sa_cluc %>% mutate(method = "no remove")) %>% 
+  unite("f", signo:method, remove = FALSE)
+
+comparative_anomalies <- a %>% ggplot(aes(x=y, y=mean, fill=f)) + 
+  geom_hline(yintercept = 0, size=0.3, colour = "gray") + 
+  geom_bar(position = "dodge", stat = "identity", width = 0.8) + geom_errorbar(aes(ymin = mean - se, ymax = mean + se, colour = f), 
+                width=.5, position = position_dodge(0.9)) +
+  facet_wrap(~clu_pop2, nrow = 2) + 
+  scale_fill_manual(values = c("neg_original" =  "#a63603",
+                               "neg_no remove" =  "#FFCCCC",
+                               "pos_original" = "#006d2c", 
+                               "pos_no remove" = "#CCFFCC"
+                                 )) +
+  scale_colour_manual(values = c("neg_original" =  "#a63603",
+                               "neg_no remove" =  "#FFCCCC",
+                               "pos_original" = "#006d2c", 
+                               "pos_no remove" = "#CCFFCC"
+                                 )) +
+  scale_x_continuous(breaks = seq(2000, 2016, by=2)) +
+  ylab('EVI Standardized Anomaly') + xlab('Year') +
+  theme_bw() +
+  theme(axis.text = element_text(size=8),
+        legend.position = "none",
+        panel.grid = element_blank(),
+        strip.background = element_rect(fill = "transparent", colour = "black", size = 0.2),
+        plot.title = element_text(hjust = -0.05, size = 9, 
+                                    margin = margin(t=-5, b=1)))
+```
+
 ```{r oldFigure, eval=FALSE, fig.height=7}

### 2020-06-07 analysis/drought/explore_prec_longterm.Rmd
new versions

@@
 ```
+
+```{r}
+
+aux_SJ <- aux %>% filter(code %in% c("5501","5502","5504","5504A","5505","5509","5509A")) %>% mutate(site = "SJ")
+aux_CA <- aux %>% filter(code %in% c("6242U","6243E","6248","6257")) %>% mutate(site = "CA")
+
+prec_sites <- aux_SJ %>% bind_rows(aux_CA)
+
+prec_sites_mean <- prec_sites %>% 
+  dplyr::select(-code) %>% 
+  group_by(hyear, site) %>% 
+     dplyr::summarise(
+     csum_mean = mean(csum, na.rm = TRUE),
+     csum_sd = sd(csum, na.rm = TRUE), 
+     csum_se = csum_sd/sqrt(length(csum_sd)), 
+     n = n())
+
+pos <- position_dodge(width = .9)
+
+prec_sites_mean %>% filter(hyear > 1990) %>% 
+  ggplot(aes(x=hyear, y=csum_mean, fill=site, colour = site)) + 
+  geom_line(position = pos) +
+  # geom_hline(yintercept = meandf) + 
+  # geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
+  # geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
+  # geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
+  # geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
+  scale_x_continuous(breaks=seq(1990,2020, by=5)) +
+  scale_y_continuous(breaks=seq(0,1500, by=250)) +
+  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, position = pos) +
+  geom_point(size=2, position = pos) + theme_bw() + 
+  theme(panel.grid.minor = element_blank()) +
+  ylab('Cumulative Precipitation (hydrological year)') + xlab('year')
+
+6257I
+6242U
+6236
+```

### 2020-06-07 man/ecosystems/2_revision2/fig/aux/comparison_sites.Rmd
new versions

---
title: "spei_locations"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(here)
library(SPEI)
library(lubridate)
source(here::here('script/R', 'hydro_year.R'))
source(here::here('/script/R/custom_functions.R'))
```

```{r}
# SJ

prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")

dfpre <- c()

for (i in prec_files){

  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")

  dfpre <- rbind(dfpre, auxdf)
}

dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")

dfpre <- dfpre %>% 
  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
         month = lubridate::month(fecha),
         hmonth = ifelse(month <= 8, month + 4, month -8), 
         hyear = hydro_year(fecha),
         hyear_f = paste0(hyear-1,'-', hyear))

aux <- dfpre %>% 
  group_by(hyear_f, code) %>% 
  dplyr::mutate(csum = cumsum(prec)) %>% 
  filter(hmonth == 12) %>% 
  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()

aux_sj <- aux %>% 
  filter(code %in% c("5501","5502","5504","5504A","5505","5509","5509A")) %>% 
  group_by(hyear) %>% 
     dplyr::summarise(
     csum_mean = mean(csum, na.rm = TRUE),
     csum_sd = sd(csum, na.rm = TRUE), 
     csum_se = csum_sd/sqrt(length(csum_sd)), 
     n = n())

## 
guejar <- read_csv("/Users/ajpelu/Downloads/cn_5501_72.csv") %>% 
  filter(variable_id == 29) %>% 
  group_by(mensual=floor_date(time, "month")) %>%
  summarize(prec=sum(value)) %>% 
    mutate(site = "Guejar")

quentar <- read_csv("/Users/ajpelu/Downloads/cn_E42_235.csv") %>% 
  filter(variable_id == 49) %>% 
  group_by(mensual=floor_date(time, "month")) %>%
  summarize(prec=sum(value)) %>% 
  mutate(site = "Quentar")

quentar_pantano <- read_csv("/Users/ajpelu/Downloads/cn_5506_100.csv") %>% 
  filter(variable_id == 29) %>% 
  group_by(mensual=floor_date(time, "month")) %>%
  summarize(prec=sum(value)) %>% 
  mutate(site = "Quentar Pantano")

df <- bind_rows(guejar, quentar, quentar_pantano) %>% 
  filter(prec >= 0) %>% 
  mutate(month = lubridate::month(mensual),
         hmonth = ifelse(month <= 8, month + 4, month -8), 
         hyear = hydro_year(mensual),
         hyear_f = paste0(hyear-1,'-', hyear)) %>% 
  group_by(hyear_f, site) %>% 
  dplyr::mutate(csum = cumsum(prec)) %>% 
  filter(hmonth == 12) %>% 
  dplyr::select(-month, -prec, -mensual) %>% 
  group_by(hyear) %>% 
     dplyr::summarise(
     csum_mean = mean(csum, na.rm = TRUE),
     csum_sd = sd(csum, na.rm = TRUE), 
     csum_se = csum_sd/sqrt(length(csum_sd)), 
     n = n())

# Compare df con aux_sj

# aux_sj <- aux_sj %>% mutate(tipo = "old") 
# df <- df %>% mutate(tipo = "new")
# 
# s <- bind_rows(aux_sj, df)
# 
# 
# s %>% filter(hyear > 1990) %>% 
#   ggplot(aes(x=hyear, y=csum_mean, fill=tipo, colour = tipo)) + 
#   geom_line(position = pos) +
#   scale_x_continuous(breaks=seq(1990,2020, by=5)) +
#   scale_y_continuous(breaks=seq(0,1500, by=250)) +
#   geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, position = pos) + geom_point(size=2, position = pos) + theme_bw()

# utilizamos el old 
```

# Pra CA sites usamos

```{r}

#6236
ca <- aux %>% 
  filter(code %in% c("6257I","6242U","6236")) %>% 
  group_by(hyear) %>% 
     dplyr::summarise(
     csum_mean = mean(csum, na.rm = TRUE),
     csum_sd = sd(csum, na.rm = TRUE), 
     csum_se = csum_sd/sqrt(length(csum_sd)), 
     n = n()) %>% 
  mutate(site = "CA")

sj <- aux %>% 
  filter(code %in% c("5501", "5502","5504", "5504A", "5505")) %>% 
  group_by(hyear) %>% 
     dplyr::summarise(
     csum_mean = mean(csum, na.rm = TRUE),
     csum_sd = sd(csum, na.rm = TRUE), 
     csum_se = csum_sd/sqrt(length(csum_sd)), 
     n = n()) %>% 
    mutate(site = "SJ")

s <- bind_rows(sj, ca)

# Get mean and sd 
ca_mean <- mean(ca$csum_mean)
ca_sd <- sd(ca$csum_mean)
sj_mean <- mean(sj$csum_mean)
sj_sd <- sd(sj$csum_mean)

prec_sites_plot <- s %>% filter(hyear > 1990) %>% 
  ggplot(aes(x=hyear, y=csum_mean, fill=site, colour = site)) + 
  geom_line(position = pos) +
  scale_x_continuous(breaks=seq(1990,2020, by=2)) +
  scale_y_continuous(breaks=seq(0,1500, by=250)) +
  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, position = pos) + geom_point(size=2, position = pos) + theme_bw() +
  ylab('Cumulative Precipitation (mm)') + xlab('Year') +
  scale_fill_manual(values = colores_den[3:2]) +
  scale_colour_manual(values = colores_den[3:2]) + theme(panel.grid = element_blank())

```

Temporal evolution of cumulative precipitation (hydrological year) during the period 1990-2017. Points represent the mean, and error bars the standard error. Data from meteorological stations distributed around northern (SJ, n=4) and southern (CA, n=4) sites respectively.

### 2020-06-07 man/figures/0_figures_ecosystems.Rmd
new versions

@@
 
+**Figure S7.** Schema of the metrics (*a*) and analysis (*b*) used in the manuscript (see Material and methods for details) and their relation with the aims (numbers in grey circles) of our study (see Introduction).
+The severe drought events since 1901 were identified using SPEI-12 and regional climatic data.
+They were characterized in terms of duration, severity, intensity.
+Resilience metrics of BAI were computed for the most severe drought events since 1950, and the relationship of them to the severity of the drought were explored.
+For 2005 and 2012 drought events we also compute resilience metrics.
+A comparison of EVI and BAI resilience metrics for the 2005 and 2012 drought events was conducted among Q.
+pyrenaica populations.
+
 **Figure S1**
@@
 ![](../../man/figures/svg/fig_8.jpg)
+
+#####
+**Figure S7**
+![](../../man/figures/schema_methods/propuesta_esquema2.jpg)
 

### 2020-06-07 man/figures/schematic_plots.Rmd
new versions

---
output:
  word_document
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r load_pkg}
library("tidyverse")
library("here")
library("ggrepel")
library("scales")
library("grid")
library("gridExtra")
library("patchwork")
library("stringr")
library("pander")
library("lemon")
library("magrittr")
source(here::here('/script/R/custom_functions.R'))
```

```{r}
# elementos comunes

# co <- "#678cab"
co <- "#6e7561"
# co <- colores_den[3]

# co_evi <- "#6e7561"
co_evi <- "#678cab"

limitesX <- c(1925, 2020)

eje_x <- scale_x_continuous(breaks = seq(1925, 2020, by=10), limits=c(1925,2020))
```

```{r}
# BAI 
cro_sites <- read.csv(here("data/cronos_medias", "cronos_sites.csv"), header = TRUE, sep = ',')

cro_sites <- cro_sites %>% mutate(site = 
  case_when(
    site == "CA_Low" ~ "CA-Low", 
    site == "CA_High" ~ "CA-High",
    site == "SJ" ~ "SJ",
  )
)

pbai <- cro_sites %>% 
  filter(nseries > 5) %>% 
  filter(site == "SJ") %>% 
  ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
  theme_bw() + ylab('BAI') + xlab("") + 
  geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
                  ymax = (bai_mean + bai_se)/100,
                  fill=site), alpha=.4, colour=NA) +
  geom_line() + 
  geom_point(size=.8) + 
  scale_colour_manual(values = co) +
  scale_fill_manual(values = co) +
  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
  theme(panel.grid = element_blank(),
        legend.position = "none", #c(0.2,.8),
        strip.background = element_blank(),
        axis.ticks.length.x.top = unit(-3.5, "pt"),
        axis.title.x.top = element_blank(), 
        axis.text.x.top = element_blank(), 
        panel.border = element_blank(),
        axis.line.y = element_line(), 
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank()) + 
  eje_x

```

```{r}
# GC
gc_todos <- read.csv(here::here("data/disturbance", "disturbance_gc.csv"), header = TRUE, sep = ',')

# customize plot (corrected scale)
pos_neg_log <- scales::trans_new('signed_log', 
                                 transform = function(x) sign(x)*log(abs(x)),
                                 inverse = function(x) sign(x)*exp(abs(x)))

plotGC <- function(df, y, yse, x, my_ylab, 
                   micolor, limitesX, ...){ 
  # Get error bar aesthetic
  yerrorbar <- aes_string(ymin = paste(y, yse, sep = '-'),
                         ymax = paste(y, yse, sep = '+'))

  p <- ggplot(df, aes_string(x = x, y = y)) +
    geom_errorbar(mapping = yerrorbar, colour = micolor, width = .35, size = .2, 
                  na.rm = TRUE) + 
    geom_bar(stat = 'identity', fill = micolor, colour = micolor) +
    geom_hline(yintercept = 0, colour = 'black', lwd = 0.2) + 
    ylab(my_ylab) +
    # Add line to %GC = 50 
    geom_hline(yintercept = c(50, -50), linetype=2, colour='black', lwd=0.2) +
    theme_bw() +
    theme(panel.grid = element_blank(),
          axis.text.y = element_text(size = 5.5),
          axis.text.x = element_text(size = 6)) +
    coord_cartesian(xlim=limitesX)
    # Customize limits and breaks to min of year data 
  p
}

```

```{r plot_SJ}
gcsj <- gc_todos %>% dplyr::filter(site == 'gc_sj') %>% 
  plotGC(y = 'gc_md', yse = 'gc_md_se', x = 'year', 
         my_ylab = '% GC', micolor = co, limitesX = limitesX) + 
  xlab('') + xlab('Time') + 
  eje_x + 
  # scale_x_continuous(breaks = xrange, position = 'bottom') + 
  scale_y_continuous(trans = pos_neg_log, 
                     breaks=c(-50, 1, 50)) + 
  geom_hline(yintercept = 1,  lwd=0.2)  +
  theme(
    panel.border = element_blank(),
        axis.line.y = element_line(), 
        axis.ticks.x = element_blank(),
    axis.line.x = element_line(colour = "grey")
  )
```

```{r}
cro_rwi <- read.csv(here("data/cronos_medias", "cronos_sites_rwi.csv"), header = TRUE, sep = ',')
year_cutoffs <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_cutoffs.csv"), header = TRUE, sep = ',')
year_sampledepth <- read.csv(here("data/cronos_medias", "cronos_sites_rwi_year_sampledepth.csv"), header = TRUE, sep = ',')

# Reorder 
cro_rwi <- cro_rwi %>% mutate(site_sorted = 
                                recode_factor(site,
                                              "caH" = "0_caH", 
                                              "caL" = "1_caL", "sj" = "2_sj"))

year_cutoffs <- year_cutoffs %>% mutate(site_sorted = 
                                recode_factor(site,
                                              "caH" = "0_caH", 
                                              "caL" = "1_caL", "sj" = "2_sj"))

year_sampledepth <- year_sampledepth %>% mutate(site_sorted = 
                                   recode_factor(site,
                                              "caH" = "0_caH", 
                                              "caL" = "1_caL", "sj" = "2_sj"))

# peores sequias 
dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)
dy <- dy %>% arrange(desc(d_duration))
# solamente aquellos mayor de 2 meses 
dys <- dy %>% filter(d_duration > 2) %>% as.data.frame() 
dys <- dys[,'maxyear']

rwi_sj <- cro_rwi %>% filter(site =="sj") %>% 
  ggplot(aes(x=year, y=res, colour=site)) +
  theme_bw() +
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size=8), 
        legend.position = "none", 
        panel.border = element_blank(),
        axis.line.y = element_line(), 
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank()) + 
  geom_hline(yintercept = 1, colour = "grey") + 
  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) + 
  ylab("RWI") + xlab("") + 
  geom_vline(xintercept = c(1995, 2005, 2012), linetype = "dotted") + 
  eje_x +
  scale_colour_manual(values = co)

```

```{r}
# Datos de iv 
iv <- read.csv(here::here("data/evi/iv_composite.csv"), header = TRUE, sep = ',')

ivv_mean <- iv %>% 
   mutate(
    clu_pop = as.factor(case_when(
      pop %in% c(1,2,3,4,5) ~ 'N',
      pop %in% c(6,7,8) ~ 'S',
      pop == 9 ~ 'out'))) %>% 
  filter(clu_pop == 'N') %>% 
  mutate(date = as.Date(date, format = "%Y-%m-%d")) %>% 
  group_by(date, clu_pop) %>% 
  summarise(iv_avg = mean(evi, na.rm = TRUE), 
            iv_sd = sd(evi, na.rm = TRUE), 
            iv_se = iv_sd/sqrt(length(evi)))

# iv_plot <- ivv_mean %>% ggplot(aes(x=date, y=iv_avg)) +
#   scale_x_date(limits= as.Date(c("1920-01-01","2020-01-01"))) + 
#   # geom_rect(aes(xmin=as.Date("2012-01-01"), 
#   #               xmax = as.Date("2012-12-31"), 
#   #               ymin=-Inf, ymax=Inf),
#   #           fill="grey", alpha=.9) + 
#   # geom_rect(aes(xmin=as.Date("2005-01-01"), 
#   #               xmax = as.Date("2005-12-31"), 
#   #               ymin=-Inf, ymax=Inf),
#   #           fill="grey", alpha=.9) + 
#   geom_line(colour=co_evi) +
#   theme_bw() + ylab("EVI") + xlab("") + 
#   theme(panel.grid = element_blank(), 
#         panel.border = element_blank(),
#         axis.line.y = element_line(), 
#         axis.ticks.x = element_blank(), 
#         axis.text.x = element_blank())

ivv_mean_year <- iv %>% 
   mutate(
    clu_pop = as.factor(case_when(
      pop %in% c(1,2,3,4,5) ~ 'N',
      pop %in% c(6,7,8) ~ 'S',
      pop == 9 ~ 'out'))) %>% 
  filter(clu_pop == 'S') %>% 
  mutate(date = as.Date(date, format = "%Y-%m-%d"), 
         year = lubridate::year(date)) %>% 
  group_by(year, clu_pop) %>% 
  summarise(iv_avg = mean(evi, na.rm = TRUE), 
            iv_sd = sd(evi, na.rm = TRUE), 
            iv_se = iv_sd/sqrt(length(evi))) %>% 
  mutate(yr = as.Date(as.character(year), format = "%Y"))

iv_plot <- ivv_mean_year %>% 
  ggplot(aes(x=yr, y=iv_avg)) + 
  geom_line(data=ivv_mean, aes(x=date, y=iv_avg), colour = "gray", alpha=.4) +
  geom_errorbar(aes(ymin = iv_avg - 5*iv_se, ymax = iv_avg + 5*iv_se), width=.3, colour = co_evi) +
  geom_point(size=1, shape=21, fill=co_evi, colour = co_evi) + 
  theme_bw() + ylab("EVI") + xlab("") + 
  theme(panel.grid = element_blank(), 
        panel.border = element_blank(),
        axis.line.y = element_line(), 
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank()) + 
  scale_x_date(limits = c(as.Date("1920-01-01"), as.Date("2020-12-31")))

```

```{r}
# From explore_anomalies.Rmd 
# Read data 
anomalias_evimean <- read.csv(here("data/anomalies", "anomalias_evimean.csv"),header = TRUE, sep = ',')

anomalias_evimean <- anomalias_evimean %>% 
  mutate(
    clu_pop = as.factor(case_when(
      pop == 1 ~ "Camarate",
      pop %in% c(2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out')),
    clu_pop2 = as.factor(case_when(
      pop %in% c(1,2,3,4,5) ~ 'Northern slope',
      pop %in% c(6,7,8) ~ 'Southern slope',
      pop == 9 ~ 'out'))) %>% 
  filter(clu_pop != 'out')

# Standardized Anomalies 
avg_sa_clu <- anomalias_evimean %>% 
  group_by(clu_pop2, y) %>% 
  dplyr::summarise(mean=mean(sa, na.rm=T),
            sd = sd(sa, na.rm=T),
            se = sd/sqrt(length(sa))) %>% 
 mutate(signo = ifelse(mean >= 0, 'pos', 'neg'),
        yr = as.Date(as.character(y), format = "%Y"))

sa <- avg_sa_clu %>%
  filter(clu_pop2 =="Northern slope") %>% 
  ggplot(aes(x=yr, y=mean)) + 
  geom_hline(yintercept = 0, size=0.3, colour = "gray") + 
  geom_bar(stat = "identity", fill=co_evi, colour="white", size=.1) + 
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width=.5, colour=co_evi) +
  # facet_wrap(~clu_pop2, nrow = 2) + 
  # scale_fill_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  # scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
  scale_x_date(limits = c(as.Date("1920-01-01"), as.Date("2020-12-31"))) +  
  ylab('EVI std. Anomaly') + xlab("") +
  theme_bw() +
  theme(panel.grid = element_blank(), 
        panel.border = element_blank(),
        axis.line.y = element_line(), 
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank())
```

# SPEI

```{r}

# DATOS desde 1901

# Get file names
myfiles <- list.files(path = here("data_raw/spei/spei_longterm"), pattern = "\\.csv$")

mydf <- data.frame()

for (j in myfiles){ 
  aux <- read.csv(file=here("data_raw/spei/spei_longterm", j), header = TRUE, sep = ',')

  # Remove csv and get name 
  name_aux <- str_replace(j, ".csv", "")

  # Get lat long
  latlong <- str_replace(name_aux, "spei_", "")
  mylong <- as.numeric(str_split_fixed(latlong, "_", 2))[1]
  mylat <- as.numeric(str_split_fixed(latlong, "_", 2))[2]

  # Store lat and long 
  aux$lat <- mylat
  aux$long <- mylong

  # year and month 
  aux$month <- rep(1:12, 115)

  auxy <- c()
  years <- c() 
  for (j in 1901:2015) { 
    auxy <- as.data.frame(rep(j,12))
    years <- rbind(years, auxy)
    } 
  names(years) <- 'year'

  aux <- cbind(aux,years)

  # join all df 
  mydf <- rbind(mydf, aux)
}

# Convert to date  
mydf <- mydf %>% mutate(date = sprintf("%d-%02d", year, month),
                        date = as.Date(paste0(date, '-01'), format="%Y-%m-%d"))

speidf <- mydf %>% dplyr::select(SPEI06,SPEI12,SPEI24,month, year, date) %>% 
  group_by(year,month, date) %>%
  summarise_at(vars(SPEI06,SPEI12,SPEI24), funs(mean,sd,se=sd(.)/sqrt(n()))) %>% 
  as.data.frame() 
```

```{r}

vi <- "SPEI12"

df <- speidf %>% dplyr::select(date, starts_with("SPEI12")) %>% 
  select_all(~gsub("SPEI12_", "", .)) %>% 
  mutate(signo = ifelse(mean > 0, "pos", "neg"))

p_spei <- df %>% 
  ggplot(aes(x = date, y = mean)) + 
    geom_bar(stat = "identity", colour="grey", fill="grey") + 
  ylab("SPEI") + xlab("") + 
  theme_bw() +
  theme(panel.grid = element_blank(), 
        panel.border = element_blank(),
        axis.line = element_line()) +
  scale_x_date(limits = c(as.Date("1901-01-01"), as.Date("2016-12-31")), date_labels = "%Y", date_breaks ="5 years")
```

```{r}

a <- rwi_sj + theme(plot.margin = unit(c(0,0,-5,0), "pt")) 
b <- pbai + theme(plot.margin = unit(c(0,0,-5,0), "pt"))
c <- gcsj +  theme(plot.margin = unit(c(0,0,-4,0), "pt"))

m <- p_spei + iv_plot + sa + a+b+c+plot_layout(ncol = 1)
```

```{r}
saveImage(m, path="./man/figures/svg/", dpi=300, w=25, h=15, fig= "plot_metrics", u = "cm") 
```

### 2020-06-07 man/ecosystems/2_revision2/fig/aux/elev.R

@@ -1,0 +1,61 @@
+
+library("tidyverse")
+library("here")
+library("ggpubr")
+
+source(here::here('/script/R/custom_functions.R'))
+
+
+elev <- read_csv(here::here("./data/elev.csv"))
+
+sj_lowcode  <- paste0('A', str_pad(1:10, 2, pad='0'))
+sj_highcode <- paste0('A', 11:20)
+ca_lowcode <- c(paste0('B', str_pad(1:10, 2, pad='0')),
+                paste0('B', 26:30))
+ca_highcode <- paste0('B', 11:25)
+
+
+elev <- elev %>%
+  mutate(site = ifelse(name %in% sj_lowcode, 'SJ-Low',
+                       ifelse(name %in% sj_highcode, 'SJ-High',
+                              ifelse(name %in% ca_lowcode, 'CA-Low', 'CA-High')))) 
+
+elev %>% ggplot(aes(y=elev_mde, x= site, colour = site)) + 
+  geom_boxplot() + 
+  geom_jitter(width = 0.2) + 
+  theme_bw() + 
+  theme(legend.position = "none")
+ylab("Elevation (m)")
+
+
+sj <- elev %>% filter(site %in% c("SJ-Low", "SJ-High")) 
+
+
+ggboxplot(elev, x = "site", y = "elev_mde", 
+          color = "site",
+          ylab = "Elevation", xlab = "Site") + 
+  scale_y_continuous(
+    breaks = seq(1150, 2000, by=50))
+
+
+t.test(elev_mde ~ site, data = sj) 
+
+ca <- elev %>% filter(site %in% c("CA-Low", "CA-High")) 
+t.test(elev_mde ~ site, data = ca) 
+
+
+<!-- # Create subset to compare between sites  -->
+  <!-- caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502", -->
+                      <!--              "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002", -->
+                      <!--              "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")]  -->
+  <!-- caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502", -->
[…]

### 2020-06-07 man/ecosystems/ms_ecosystems.Rmd
fix highlights

@@
 ### Manuscript highlights
-* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand levels
-* Resilience and growth response to climate followed a water-stress gradient
-* Trees and stand expressed high sensitivity to drought and land-use legacies
+* *Quercus pyrenaica* rear-edge forests showed high resilience to drought
+* Resilience and the growth response to climate followed a water-stress gradient
+* Tree growth and vegetation indices were highly sensitive to drought and land-use
 

### 2020-06-07 man/ecosystems/ms_ecosystems.Rmd
change abstract

@@
 
-* Global change challenges ecosystems in dry locations transformed by intensive human use.
+* Global change challenges ecosystems in xeric locations transformed by intensive human use.
 Resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd.
-populations in the southern Iberian Peninsula was analyzed in relation to historical records of land use, employing dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
+populations in the southern Iberian Peninsula was analyzed in relation to historical records of land-use, combining dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
 
-* The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood removal) in the current forest structure.
+* The growth trends reflected a strong influence of old land-use legacies (*e.g.* firewood removal) in the current forest structure.
 Trees were highly sensitive to moisture availability but both primary and secondary growth expressed high resilience to drought events over the short and the long term.
 Resilience and the tree growth response to climate followed a water-stress gradient.
-A positive growth trend since the late 1970s was particularly evident in mesic (*i.e.* colder and wetter) high-elevation stands, but absent in the most xeric (*i.e.* warmer and drier) site.
+A positive growth trend since the late 1970s was particularly evident in mesic (*i.e.* colder and wetter) high-elevation stands, but absent in the most xeric (*i.e.* warmer and drier) stands.
 
-* The high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge.
-These resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.
-This is particularly relevant for rear-edge populations where topographic and biophysical variability can facilitate the existence of refugia
+* The high values of resilience observed in our study suggest that *Q.
+pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic and ecological rear edge.
+Resilience of oak stands to drought events was not spatially homogeneous across the mountain range, due to differences in ecological conditions and/or past-management legacies.
+This is particularly relevant for rear-edge populations where topographic and biophysical variability can facilitate the existence of refugia.
 

### 2020-06-07 man/ecosystems/ms_ecosystems.Rmd
update intro

@@
 ## Introduction
-The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005].
+
+The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@Hampe2005].
 Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016].
-It has been assumed that geographical and ecological rear-edge populations match.
-This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@VilaCabrera2019; @Pironon2016].
+It has been often assumed that geographically marginal populations represent ecological marginal populations.
+This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pironon2016; @VilaCabrera2019; @Oldfatheretal2020RangeEdges].
 Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014].
-In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, we need to determine under what environmental conditions the geographical rear edge does not correspond with the ecological one [@VilaCabreraJump2019GreaterGrowth].
+This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and an genetic components [@VilaCabrera2019], but also an anthropogenic (*i.e.* land-use) dimension.
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to determine the factors that cause mismatches between the geographical and the ecological rear-edge [@VilaCabreraJump2019GreaterGrowth].
 
-Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs of more intense and longer droughts [@Allen2010].
-In fact, drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
-This is especially critical, since climate-change projections for the Mediterranean area forecast more frequent and severe extreme climate events [@Spinoni2017a].
+Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sexton2009].
+Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allen2010].
+Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
 In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
-In a global-change context, the impact of drought on vegetation needs to be evaluated regarding the interactions with other drivers of change [@DoblasMiranda2017].
-This is especially relevant for areas with a long history of landscape alteration, such as the Mediterranean region, where land-use change plays a major role in current ecosystem ecology [@Navarro2013], given that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].
+The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@Hampe2005].
 
-The effects of drought are also especially relevant for populations considered to be located along the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution.
-In these cases, climatic conditions mark the species limits, and the populations are likely to be more sensitive to minor climatic and microtopographic variations [@Hampe2005].
-The response of rear-edge populations to climate variations could be useful in estimating the adaptation of species to predicted climatic changes [@Jump2010].
-Also, it is crucial to accurately characterize the ecological rear-edge of species.
-This can be particularly misleading in ecosystems where land-use legacies have strongly modified species distribution [@VilaCabrera2019].
-Furthermore, this characterization becomes even more critical for relict populations driven by historical land uses when they harbor high levels of intraspecific genetic diversity [@Valbuena2013].
+It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Sala2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @Navarro2013; @DoblasMiranda2017].
+In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects].
+The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLandUse].
+For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests].
+Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate‐driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth].
+In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forest at the southern geographical range.
+Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLandUse], and particularly that of marginal rear edge populations [@VilaCabrera2019].
 
-The assessment of resilience to climate disturbances such as droughts provides critical information concerning the capacity of the forests to maintain their structure and render valuable ecosystem services.
-The drought resilience of species is better assessed by a combination of several approaches, such as dendrochronology and remote sensing.
-Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and annual tree-ring widths can be used to study growth changes in response to drought among individiual trees [@Fritts1976].
-Remote sensing can be used to analyze the impact of drought on ecosystems and hence on the stand [*e.g.* @Zhang2013].
-The combination of both approaches (remote sensing and dendroecology) has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].
-Nonetheless, it is crucial to ascertain whether the responses at the tree level differ from those at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.
+The assessment of resilience to climate and human disturbances provides critical information concerning the capacity of forests to maintain their structure and render valuable ecosystem services.
+Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYoua].
+@Lloret2011 proposed an approach, which decomposes resilience to drought in three components: resistance to drought, recovery after drought and resilience.
+Forest resilience is determined by the capacity to reduce the impact (resistance) and the capacity to recover from the impact of disturbance (recovery) [@IngrischBahn2018ComparableQuantification].
+This conceptual approach has been very popular to assess forest resilience recently, because it allows a simple, yet highly efficient assessment of short-term responses of trees to drought.
+Nevertheless, it is not exempted from some criticism and needs to be applied carefully to avoid over interpretation and potential bias associated at different levels [@Schwarzetal2020QuantifyingGrowth].
+In this sense we assessed forest resilience both in the long-term to several recent extreme drought episodes and to climate change (*i.e.* warming) in the long-term (the last few decades) using two different proxies to characterize resilience.
+Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and, generally annual tree-ring widths have been used to study growth changes in response to drought at the individual tree level [@Fritts1976; @Dobbertin2005TreeGrowth].
+Remote sensing can be used to analyze the impact of drought on ecosystems and hence at the stand level [*e.g.* @Zhang2013].
+The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].
 
-In the present study, we used remote-sensing information and dendroecological methods to evaluate the impact of drought in both canopy greenness (as a proxy to primary growth) and radial tree growth (as a proxy for secondary growth) of *Quercus pyrenaica* Willd.
-(Pyrenean oak) in southern relict forests at the rear-edge of the species distribution and where species performance is considered to be severely threatened by climate change.
-We also assessed the resilience of these forests both to several extreme drought episodes and to climate change (*i.e.* warming) over the long term (in the last few decades).
-In the study area, the rear edge of relict populations is imposed by historical land use.
-Our main hypothesis is that, limited by their land-use legacy, these stands will show low resilience to extreme drought from climate change a small-scale gradient.
-To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Quercus pyrenaica* forests at their  present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
-For the latter task, we characterized the variability in the forest response to drought within the current geographical rear edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.
+In the present study, we assessed resilience of *Quercus pyrenaica* Willd.
+(Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be severely threatened by climate change.
+We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth).
+The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests.
+To do so we conducted a review of historical documents to reconstruct the forest history of our study sites, and to determine how anthropogenic drivers have shaped the current forest structure.
+Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local‐scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations.
+Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient.
+To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q.
+pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
+For the latter task, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.
 
@@
 They are the richest forest formation of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG].
+They also harbor high levels of intraspecific genetic diversity [@Valbuena2013].
 These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002].

### 2020-06-07 man/ecosystems/ms_ecosystems.Rmd
update methods

@@
 Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005].
-They are the richest forest formation of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG].
+They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG].
 They also harbor high levels of intraspecific genetic diversity [@Valbuena2013].
@@
 ### Climatic data and drought episodes
-Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008].
-Monthly precipitation and minimum and maximum temperatures had a 0.25 x 0.25 º resolution for the 1950-2016 period.
-Grid cells were selected to cover each sampled site.
-The SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] index with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014.
 
@@
 Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3).
-A drought event was identified using the SPEI 12-months scale, following a procedure similar to the one proposed by @Spinoni2015.
+A drought event was identified using the SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] (SPEI 12-months scale), following a procedure similar to the one proposed by @Spinoni2015.
 We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html).
@@
 For each drought event, we computed: the *duration* as the number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the absolute SPEI values during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.
+
+To explore the relationships between climatic variables and tree-growth variables we used climate data obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008].
+Monthly precipitation and minimum and maximum temperatures had a 0.25 x 0.25 º resolution for the 1950-2016 period.
+Grid cells were selected to cover each sampled site.
+The SPEI 6-months scale index was used to characterize the drought conditions for the period 1961-2014.
 
@@
 Rather than other vegetation indices such as the NDVI, $EVI_{mean}$ was chosen because it is highly stable under the use of any filter [@Reyes2015] and because it showed highly significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88).
-For exploratory purposes, temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test.
 
-### Field sampling and dendrochronological methods to assess individual tree resilience
+### Field sampling and dendroecological methods to assess individual tree resilience
 Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1).
@@
 Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].
-Differences between sites for height, DBH, and competition indices were analyzed using non-parametric Kruskal-Wallis rank sum tests.
-When significant differences were detected, multiple comparisons were run using the Dunn's-test with Bonferroni adjustment to correct for significance.
 
@@
 To evaluate the effects of drought events on ecosystem resilience (using greenness data) and individual tree resilience (using BAI data), we used resilience indices proposed by @Lloret2011.
-The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought:PreDrought$) quantifies the severity of the impact of the disturbance in the year it occurred.
-The Recovery index, computed as the ratio between performance after and during disturbance ($Rc = PostDrought:Drought$), represents the ability to recover from disturbance relative to its severity.
-Finally, the Resilience index ($Rs = PostDrought:PreDrought$) is the capacity to reach pre-disturbance performance levels.
+The Resistance index estimated as the ratio between performance during and before the disturbance ($Resistance=Drought:PreDrought$) quantifies the severity of the impact of the disturbance in the year it occurred.
+The Recovery index, computed as the ratio between performance after and during disturbance ($Recovery= PostDrought:Drought$), represents the ability to recover from disturbance relative to its severity.
+Finally, the Resilience index ($Resilience = PostDrought:PreDrought$) is the capacity to reach pre-disturbance performance levels.
 The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event.
@@
 ### Statistical analysis
-Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].
+
+The severe drought events since 1901 were identified using SPEI-12 and regional climatic data.
+They were characterized in terms of duration, severity, intensity [see @Spinoni2015].
+In a first step we explored temporal trends of EVI and BAI variables.
+Temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test (Figure S7a).
+Temporal trends for BAI were assessed using mean BAI chronologies.
+The impact of drought in greenness and growth was exploring using the $EVI_{sa}$ and the mean RWI site chronologies (Figures S7a).
+Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim` [@Zang2015].
+The non-climatic disturbance impacts on tree-growth were evaluated using the site disturbance chronologies.
+
+Differences between sites for height, DBH, and competition indices were analyzed using non-parametric Kruskal-Wallis rank sum tests.
+When significant differences were detected, multiple comparisons were run using the Dunn's-test with Bonferroni adjustment to correct for significance.
+
 For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern exposures).
@@
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
+Resilience metrics of BAI were also computed for the most severe drought events since 1950, and the relationship of them to the severity of the drought were explored.
 

### 2020-06-07 man/ecosystems/ms_ecosystems.Rmd
update results

@@
 ### Climatic data and drought episodes
-
 The Iberian Peninsula underwent several extreme drought episodes in the last three decades [*e.g.* 1995, 1999, 2005, 2012;  @VicenteSerrano2014].
@@
 ### Statistical analysis 
-
-The severe drought events since 1901 were identified using SPEI-12 and regional climatic data.
+The severe drought events since 1901 were identified using SPEI-12 and regional climatic data (Figure 2).
 They were characterized in terms of duration, severity, intensity [see @Spinoni2015].
 In a first step we explored temporal trends of EVI and BAI variables.
-Temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test (Figure S7a).
+Temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test.
 Temporal trends for BAI were assessed using mean BAI chronologies.
-The impact of drought in greenness and growth was exploring using the $EVI_{sa}$ and the mean RWI site chronologies (Figures S7a).
+The impact of drought in greenness and growth was exploring using the $EVI_{sa}$ and the mean RWI site chronologies.
 Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim` [@Zang2015].
@@
 The analysis of time trends in greenness showed that 78.9% of the EVI pixels followed a positive trend for the 2000-2016 period.
-The lowest values of EVI standardized anomalies for the study period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2a).
-A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was found during this drought event, whereas no changes in greenness in response to the 2012 drought were detected (Figure 2b).
+The lowest values of EVI standardized anomalies for the study period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 3a).
+A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was found during this drought event, whereas no changes in greenness in response to the 2012 drought were detected (Figure 3b).
 
@@
 The trees of the southern population were older than those from the northern one.
-In addition, trees from the southern population at high elevation were taller and their growth was significantly faster than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3).
-The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 3 and S3a).
-Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 3), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).
+In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a).
+The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 4a and S3a).
+Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 4a), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).
 
@@
 
-The northern site (SJ) showed two major release events (GC > 50% occurring in more than 50% of trees sampled): the first during the 1940s (the most evident) and the second in 1995-2000 (Figure 4).
+The response of tree growth to water availability was greater than to temperatures.
+Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure S6a).
+Nevertheless, differences appeared between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure S6a).
+
+The northern site (SJ) showed two major release events (GC > 50% occurring in more than 50% of trees sampled): the first during the 1940s (the most evident) and the second in 1995-2000 (Figure 4b).
 These periods alternated with periods of suppression.
@@
 ### Resilience to drought events at the ecosystem and individual-tree levels
-*Rs* and *Rt* varied in the same direction whereas *Rc* varied inversely to *Rs* and *Rt*.
-Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 5).
-A similar pattern was found for *Rs* but proved significant only for SJ.
-Importantly, non-significant patterns resulted when we excluded 1995, except for *Rc* in SJ (Figure S5).
-The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4).
+Resilience and resistance varied in the same direction whereas recovery varied inversely to resilience and resistance.
+Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and recovery, significant for all oak populations (Figure 5a).
+A similar pattern was found for resilience but proved significant only for SJ.
+Importantly, non-significant patterns resulted when we excluded 1995, except for recovery in SJ (Figure S5).
+The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4b).
 
@@
 The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years.
-Resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 in both variables (Table S2; Figure 6).
-Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event.
-Recovery showed a contrasting pattern for EVI and tree growth.
+For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event.
+For BAI, the resilience, resistance and recovery values were higher for 2012 than for 2005 (Table S2, Figure 5c).
 
@@
 The two southern populations showed lower recovery values than did the northern site both for greenness and tree growth, but resistance and resilience values were significantly higher for the southern site (Table S2).
-
-### Tree-growth response to climate
-The response of tree growth to water availability was greater than to temperatures.
-Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 7a).
-Nevertheless, differences appeared between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure 7a).
-In addition, the spring maximum temperature was the most significant limitation for tree growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of the current September positively influenced tree growth only in the northern population.
 

### 2020-06-07 man/figures/0_figures_ecosystems.Rmd
update figures

@@
 
-**Figure 2.** ***a)*** EVI standardized anomaly during the period 2000-2016 for northern and southern populations.
+**Figure 2**.
+Schema of the metrics and analysis used in the manuscript (see Material and methods for details) and their relation with the aims (numbers in grey circles) of our study (see Introduction).
+The severe drought events since 1901 were identified using SPEI-12 and regional climatic data.
+They were characterized in terms of duration, severity, intensity.
+Resilience metrics of BAI were computed for the most severe drought events since 1950, and the relationship of them to the severity of the drought were explored.
+For 2005 and 2012 drought events we also compute resilience metrics.
+A comparison of EVI and BAI resilience metrics for the 2005 and 2012 drought events was conducted among Q.
+pyrenaica populations.
+
+**Figure 3.** ***a)*** EVI standardized anomaly during the period 2000-2016 for northern and southern populations.
 Error bars show standard error.
@@
 
-**Figure 3.** ***a)*** Basal Area Increment (BAI) chronologies of *Q.
+**Figure 4.** ***a)*** Basal Area Increment (BAI) chronologies of *Q.
 pyrenaica* for northern population (SJ; *red*) and southern ones: low-elevation (CA-Low; *green*) and high-elevation (CA-High, *blue*) sites.
@@
 
-**Figure 4.** ***a)*** Resilience metrics of tree-growth for eight severe drought events since 1950 (see main text for details) as a function of drought severity.
+**Figure 5.** ***a)*** Resilience metrics of tree-growth for eight severe drought events since 1950 (see main text for details) as a function of drought severity.
 Points indicate resilience metrics for oak populations: SJ (*red*), CA-High (*blue*) and CA-Low (*green*).
@@
 **Figure 2.**
+![](../../man/figures/schema_methods/propuesta_esquema2.jpg)
+
+#####
+**Figure 3.**
 ![](../../man/figures/svg/fig_2ab.jpg)
@@
 #####
-**Figure 3**
+**Figure 4**
 
@@
 #####
-**Figure 4**
+**Figure 5**
 ![](../../man/figures/svg/resiliences.jpg)
@@
 
-**Figure S7.** Schema of the metrics (*a*) and analysis (*b*) used in the manuscript (see Material and methods for details) and their relation with the aims (numbers in grey circles) of our study (see Introduction).
-The severe drought events since 1901 were identified using SPEI-12 and regional climatic data.
-They were characterized in terms of duration, severity, intensity.
-Resilience metrics of BAI were computed for the most severe drought events since 1950, and the relationship of them to the severity of the drought were explored.
-For 2005 and 2012 drought events we also compute resilience metrics.
-A comparison of EVI and BAI resilience metrics for the 2005 and 2012 drought events was conducted among Q.
-pyrenaica populations.
-
 **Figure S1**
@@
 ![](../../man/figures/svg/fig_8.jpg)
-
-#####
-**Figure S7**
-![](../../man/figures/schema_methods/propuesta_esquema2.jpg)
 

### 2020-06-10 man/figures/schematic_plots.Rmd
figure

@@
 
-limitesX <- c(1925, 2020)
+limitesX <- c(1901, 2016)
+limitesD <- c(as.Date("1901-01-01"), as.Date("2016-01-01"))
 
-eje_x <- scale_x_continuous(breaks = seq(1925, 2020, by=10), limits=c(1925,2020))
+eje_x <- scale_x_continuous(breaks = seq(1900, 2015, by=5), limits = limitesX)
+eje_xdate <- scale_x_date
+
+texto_ejes <- element_text(size = 12) 
+titulo_eje_y <- element_text(size=14, face = "bold")
+
+custom_tema <- 
+  theme(panel.background = element_blank(), 
+        panel.border = element_blank(),
+        panel.grid = element_blank(),
+        # ejes 
+        axis.title.y = element_text(size=14, face = "bold"), 
+        axis.text.y = element_text(size = 12),
+        # x 
+        axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()
+)
 ```
@@
   ggplot(aes(x=year, y=bai_mean/100, colour=site)) + 
-  theme_bw() + ylab('BAI') + xlab("") + 
+  ylab('BAI') + xlab("") + 
   geom_ribbon(aes(ymin = (bai_mean - bai_se)/100,
@@
   scale_colour_manual(values = co) + 
+  scale_y_continuous(breaks = c(10,20), position="right") + 
   scale_fill_manual(values = co) + 
-  ylab(expression(paste("BAI (", cm^2, year^-1, ")", sep=''))) +
-  theme(panel.grid = element_blank(),
-        legend.position = "none", #c(0.2,.8),
-        strip.background = element_blank(),
-        axis.ticks.length.x.top = unit(-3.5, "pt"),
-        axis.title.x.top = element_blank(), 
-        axis.text.x.top = element_blank(), 
-        panel.border = element_blank(),
-        axis.line.y = element_line(), 
-        axis.ticks.x = element_blank(), 
-        axis.text.x = element_blank()) + 
+  ylab("BAI") + 
+  custom_tema + 
+  theme(legend.position = "none", 
+        axis.line.y = element_line()) +
+
   eje_x
@@
     geom_hline(yintercept = c(50, -50), linetype=2, colour='black', lwd=0.2) + 
-    theme_bw() +
-    theme(panel.grid = element_blank(),
-          axis.text.y = element_text(size = 5.5),
-          axis.text.x = element_text(size = 6)) +
     coord_cartesian(xlim=limitesX)
-    # Customize limits and breaks to min of year data 
   p
@@
   scale_y_continuous(trans = pos_neg_log, 
-                     breaks=c(-50, 1, 50)) + 
+                     breaks=c(-50, 1, 50), position="right") + 
   geom_hline(yintercept = 1,  lwd=0.2) + 
-  theme(
-    panel.border = element_blank(),
-        axis.line.y = element_line(), 
-        axis.ticks.x = element_blank(),
-    axis.line.x = element_line(colour = "grey")
-  )
+  custom_tema + 
+  theme(axis.line.y = element_line(),
+        axis.line.x = element_line(colour = "grey"), 
+        axis.text.x = element_text())
 ```
@@
   ggplot(aes(x=year, y=res, colour=site)) +
-  theme_bw() +
-  theme(panel.grid = element_blank(), 
-        axis.text = element_text(size=8), 
-        legend.position = "none", 
-        panel.border = element_blank(),
-        axis.line.y = element_line(), 
-        axis.ticks.x = element_blank(), 
-        axis.text.x = element_blank()) + 
+  custom_tema + 
+  theme(legend.position = "none", 
+        axis.line.y = element_line()) +
   geom_hline(yintercept = 1, colour = "grey") + 
-  geom_line(size=.8) + scale_y_continuous(breaks = seq(0.4,1.8, 0.2)) + 
+  geom_line(size=.8) + scale_y_continuous(breaks = c(0.5, 1, 1.5), position="right") +
+                                          # seq(0.4,1.8, 0.2)) + 
   ylab("RWI") + xlab("") + 
@@
   geom_point(size=1, shape=21, fill=co_evi, colour = co_evi) + 
-  theme_bw() + ylab("EVI") + xlab("") + 
-  theme(panel.grid = element_blank(), 
-        panel.border = element_blank(),
-        axis.line.y = element_line(), 
-        axis.ticks.x = element_blank(), 
-        axis.text.x = element_blank()) + 
-  scale_x_date(limits = c(as.Date("1920-01-01"), as.Date("2020-12-31")))
+  ylab("EVI") + xlab("") + custom_tema + 
+  theme(axis.line.y = element_line()) + scale_y_continuous(position="right") + 
+  scale_x_date(limits = limitesD)
 
@@
   # scale_color_manual(values = c("pos" = color_pos, "neg" =  color_neg)) +
-  scale_x_date(limits = c(as.Date("1920-01-01"), as.Date("2020-12-31"))) +  
-  ylab('EVI std. Anomaly') + xlab("") +
-  theme_bw() +
-  theme(panel.grid = element_blank(), 
-        panel.border = element_blank(),
-        axis.line.y = element_line(), 
-        axis.ticks.x = element_blank(), 
-        axis.text.x = element_blank())
+  scale_x_date(limits = limitesD) + 
+  ylab(expression(bold("EVI"["sa"]))) + xlab("") + 
+  scale_y_continuous(position="right") +
+   custom_tema + 
+  theme(axis.line.y = element_line())
+
 ```
@@
     geom_bar(stat = "identity", colour="grey", fill="grey") + 
-  ylab("SPEI") + xlab("") + 
-  theme_bw() +
-  theme(panel.grid = element_blank(), 
-        panel.border = element_blank(),
-        axis.line = element_line()) +
-  scale_x_date(limits = c(as.Date("1901-01-01"), as.Date("2016-12-31")), date_labels = "%Y", date_breaks ="5 years")
+  ylab("SPEI") + xlab("") + custom_tema + 
+  theme(axis.line = element_line(), 
+        axis.text.x = element_text(size=10), 
+        axis.ticks.x = element_line()) +
+  scale_x_date(limits = limitesD, date_labels = "%Y", date_breaks ="10 years", position = "top")
+
 ```
@@
 
-m <- p_spei + iv_plot + sa + a+b+c+plot_layout(ncol = 1)
+ev <- iv_plot + theme(plot.margin = unit(c(-5,0,-5,0), "pt"))
+evsa <- sa +  theme(plot.margin = unit(c(-5,0,5,0), "pt"))
+
+m <- p_spei + ev + sa + a+b+c+plot_layout(ncol = 1)
 ```
@@
 ```{r}
-saveImage(m, path="./man/figures/svg/", dpi=300, w=25, h=15, fig= "plot_metrics", u = "cm") 
+saveImage(m, path="./man/figures/svg/", dpi=300, w=25, h=17, fig= "plot_metrics", u = "cm") 
 ```

### 2020-06-10 man/figures/0_figures_ecosystems.Rmd
update legend figures

@@
 **Figure 2**.
-Schema of the metrics and analysis used in the manuscript (see Material and methods for details) and their relation with the aims (numbers in grey circles) of our study (see Introduction).
-The severe drought events since 1901 were identified using SPEI-12 and regional climatic data.
-They were characterized in terms of duration, severity, intensity.
-Resilience metrics of BAI were computed for the most severe drought events since 1950, and the relationship of them to the severity of the drought were explored.
-For 2005 and 2012 drought events we also compute resilience metrics.
-A comparison of EVI and BAI resilience metrics for the 2005 and 2012 drought events was conducted among Q.
-pyrenaica populations.
+Schema of the different metrics (a) and analyses (b) used in the manuscript (see Material and methods for details).
+The severe drought events since 1901 were identified using SPEI-12 and characterized in terms of duration, severity and intensity.
+Climate impacts on vegetation were assessed for greenness and tree-growth.
+Non climatic disturbances on vegetation were quantified using growth changes on tree growth (% GC) and were also related with anthropogenic alterations inferred from review of historical documents.
+Responses of vegetation to disturbances were explored in the short- and the long-term using resilience metrics and temporal trends respectively for both EVI and BAI.
+Resilience metrics of BAI were computed for the eight most severe drought events since 1950, and their relationship with drought severity were explored.
+For the 2005 and 2012 drought events we also compared EVI and BAI resilience metrics among the three *Q.
+pyrenaica* populations.
 
@@
 **Figure 2.**
-![](../../man/figures/schema_methods/propuesta_esquema2.jpg)
+![](../../man/figures/schema_methods/schema_analysisv3@1,5x.jpg)0
 

### 2020-06-10 man/ecosystems/ms_ecosystems.Rmd
update abstract

@@
 AJPL analyzed data and led the writing of the paper.
-GGI and RZ contributed in the writing process.
-All authors contributed to the drafts and gave final approval for publication.
+All authors contributed in the writing process to the drafts and gave final approval for publication.
 
@@
 
-* The high values of resilience observed in our study suggest that *Q.
-pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic and ecological rear edge.
+* The high values of resilience observed suggest that the studied *Q.
+pyrenaica* populations are located in a geographical but not a climatic or ecological rear edge.
 Resilience of oak stands to drought events was not spatially homogeneous across the mountain range, due to differences in ecological conditions and/or past-management legacies.
@@
 Temporal trends for BAI were assessed using mean BAI chronologies.
-The impact of drought in greenness and growth was exploring using the $EVI_{sa}$ and the mean RWI site chronologies.
+The impact of drought in greenness and growth was explored using the $EVI_{sa}$ and the mean RWI site chronologies.
 Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim` [@Zang2015].

### 2020-06-10 man/ecosystems/ms_ecosystems.Rmd
update intro

@@
 Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014].
-This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and an genetic components [@VilaCabrera2019], but also an anthropogenic (*i.e.* land-use) dimension.
-In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to determine the factors that cause mismatches between the geographical and the ecological rear-edge [@VilaCabreraJump2019GreaterGrowth].
+This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and a genetic component [@VilaCabrera2019], but also an anthropogenic (*i.e.* land-use) dimension.
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to identify and understand mismatches between the geographical and the ecological rear-edges [@VilaCabreraJump2019GreaterGrowth].
 
@@
 Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
-In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
-The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@Hampe2005].
+In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge.
+The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@Hampe2005; @VilaCabrera2019].
 
@@
 For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests].
-Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate‐driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth].
+Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica].
 In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forest at the southern geographical range.
@@
 In the present study, we assessed resilience of *Quercus pyrenaica* Willd.
-(Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be severely threatened by climate change.
+(Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjectionsa; @GeaIzquierdoetal2017RiskyFuture].
 We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth).
 The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests.
-To do so we conducted a review of historical documents to reconstruct the forest history of our study sites, and to determine how anthropogenic drivers have shaped the current forest structure.
-Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local‐scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations.
+To do so we conducted a review of historical documents to reconstruct forest history in our study sites, and to determine how anthropogenic drivers have shaped the current forest structure.
+Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations.
 Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient.
@@
 pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
-For the latter task, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.
+For the latter, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.
 

### 2020-06-10 man/ecosystems/ms_ecosystems.Rmd
add statisticall methods

@@
 In the present study, we assessed resilience of *Quercus pyrenaica* Willd.
-(Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjectionsa; @GeaIzquierdoetal2017RiskyFuture].
+(Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdo2013; @GeaIzquierdoetal2017RiskyFuture].
 We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth).
@@
 The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*.
-The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability according to elevation.
+The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability according to elevation and aspect.
 Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002].
@@
 These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002].
-Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013].
+Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013; @GeaIzquierdoetal2017RiskyFuture].
 
@@
 ### Field sampling and dendroecological methods to assess individual tree resilience
-Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1).
-For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit (CA-Low) and the tree-line (CA-High), respectively, in the site sampled.
+Trees were sampled during the autumn of 2016 at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect (around 1400 m); and Cáñar (CA), a wetter site located at the southern aspect (Figure 1; Table 1).
+For the southern site, two elevations were sampled: CA-Low (around 1700 m) and CA-High (around 1860 m), constituting the current low-elevational limit for the species (CA-Low) and the maximum altitude currently reached by trees (CA-High), respectively, in the site sampled.
 Despite the proximity of these two elevations (less than a 200-m difference) the stands differ markedly in their structure and characteristics (Table 1).
@@
 These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976].
-Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010].
+Tree-ring width series were standardized and detrended using dplR [@Bunn2010].
 Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976].
@@
 The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
-In theory, BAI represents a more accurate indicator of growth than ring width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile growth [@Biondi2008].
+In theory, BAI represents a more accurate indicator of growth than ring width, because it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile growth [@Biondi2008].
 Raw ring widths and measured DBH were used to compute BAI [@Piovesan2008] with the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
@@
 
-### Disturbance analyses 
+### Disturbance analyses and land-use history review 
 Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (suppressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past.
@@
 For this, existing historical documents were exhaustively reviewed to compile information on socio-economical activities affecting the forests being studied (Table S4).
-We reviewed several documentary sources: historical documents and maps; detailed mining reports; official information on recent wildfires events and forest-management practices; livestock farming; traditional irrigation channels; and studies concerning the socioeconomic dynamics of forests on Sierra Nevada at different scales (see Table S4 for references).
+We exhaustively reviewed existing documentary sources: historical documents and maps; detailed mining reports; official information on recent wildfires events and forest-management practices; livestock farming; traditional irrigation channels; and studies concerning the socioeconomic dynamics of forests on Sierra Nevada at different spatio-temporal scales (see Table S4 for references).
 
@@
 ### Statistical analysis
-The severe drought events since 1901 were identified using SPEI-12 and regional climatic data (Figure 2).
-They were characterized in terms of duration, severity, intensity [see @Spinoni2015].
-In a first step we explored temporal trends of EVI and BAI variables.
-Temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test.
-Temporal trends for BAI were assessed using mean BAI chronologies.
-The impact of drought in greenness and growth was explored using the $EVI_{sa}$ and the mean RWI site chronologies.
-Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim` [@Zang2015].
-The non-climatic disturbance impacts on tree-growth were evaluated using the site disturbance chronologies.
 
@@
 When significant differences were detected, multiple comparisons were run using the Dunn's-test with Bonferroni adjustment to correct for significance.
+
+The severe drought events since 1901 were identified using SPEI-12.
+They were characterized in terms of duration, severity and intensity [see @Spinoni2015].
+In a first step, the impact of drought in greenness and growth was explored using the $EVI_{sa}$ and the mean RWI site chronologies (Figure 2).
+Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim [@Zang2015].
+The non-climatic disturbance impacts on tree-growth were evaluated using site disturbance chronologies (buit using growth changes on tree growth).
+
+Responses of vegetation to disturbances were explored in the short- and the long-term using resilience metrics (resilience, resistance and recovery) and temporal trends respectively for both EVI and BAI (Figure 2).
+Resilience metrics of BAI were computed for the eight most severe drought events since 1950 (including 2005 and 2012), and their relationship with drought severity were explored.
+Resilience metrics of EVI were computed only for 2005 and 2012 drought events.
+Temporal trends of $EVI_{mean}$ (pixel scale) and BAI (mean BAI site chronologies) were examined using non-parametric (Mann-Kendall) and parametric test (Pearson) respectively.
 
@@
 *Post-hoc* differences were assessed pairwise using a similar bootstrap test.
-All the robust ANOVA and *post-hoc* tests were carried out using the `WRS2` package.
+All the robust ANOVA and *post-hoc* tests were carried out using the WRS2 R package [@Mair2017].
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
-Resilience metrics of BAI were also computed for the most severe drought events since 1950, and the relationship of them to the severity of the drought were explored.
 

### 2020-06-10 man/ecosystems/ms_ecosystems.Rmd
update text

@@
 The trees of the southern population were older than those from the northern one.
-In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a).
-The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 4a and S3a).
-Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 4a), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).
+In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites, despite stand competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a).
+The growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4a and S3a).
+Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 4a), this trend being far more pronounced for the wetter and colder high elevation site (CA-High).
 
 Drought events reduced radial growth for all sites (Figure S2a).
-The strongest reduction in radial growth occurred in the 1995 drought (the worst drought spell in our climatic record, Table S3) for all sites.
-Tree-growth reduction followed a moisture gradient.
-The southern sites (CA-High and CA-Low) showed less tree-growth reduction than did the northern site (SJ), especially for 2005 and 2012 (Figure S2a), with the weakest growth reductions being in trees from the wettest site (CA-High).
+The strongest reduction in radial growth occurred in response to the 1995 drought (the worst drought spell in our climatic record, Table S3) for all sites.
+Tree-growth reductions in response to drought followed a moisture gradient.
+Tree-growth reductions in response to the studied drought events were lower in the southern sites (CA-High and CA-Low) than in the northern site (SJ), especially for 2005 and 2012 (Figure S2a).
+The weakest growth reductions were found in trees from the wettest site (CA-High).
 
@@
 Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure S6a).
-Nevertheless, differences appeared between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure S6a).
+Nevertheless, there were differences between populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure S6a).
 
@@
 
-### Resilience to drought events at the ecosystem and individual-tree levels
+### Resilience to drought events at the forest stand and individual-tree levels
 Resilience and resistance varied in the same direction whereas recovery varied inversely to resilience and resistance.
+During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1).
+The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years.
+For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event.
+For BAI, the resilience, resistance and recovery values were higher for 2012 than for 2005 (Table S2, Figure 5c).
+
+The recovery and resistance for greenness and growth varied significantly between sites.
+Resilience calculated for greenness also differed between sites but not for tree growth (p = 0.534; Table S1).
+The two southern populations showed lower recovery values than did the northern site both for greenness and tree growth, but resistance and resilience values were significantly higher for the southern site (Table S2).
+
 Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and recovery, significant for all oak populations (Figure 5a).
@@
 
-During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1).
-The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years.
-For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event.
-For BAI, the resilience, resistance and recovery values were higher for 2012 than for 2005 (Table S2, Figure 5c).
+## Discussion
+///////
+By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the *Quercus pyrenaica* forests of Sierra Nevada and their resilience to several severe drought events in the recent decades.
+//////
 
-The resilience metrics calculated varied significantly between sites, except for resilience of tree growth (p = 0.534; Table S1), which was similar among the three sites.
-The two southern populations showed lower recovery values than did the northern site both for greenness and tree growth, but resistance and resilience values were significantly higher for the southern site (Table S2).
+Our results indicate that these relict oak populations driven by historical land use have been resilient to climate change at their present rear edge.
+However, resistance, resilience, and forest recovery to extreme drought events were strongly influenced by mountain exposure, local environmental conditions, and management legacies.
+This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance can vary markedly along the rear edge under climate change
 
-## Discussion
+### High sensitivity of oaks to recent drought events
+Severe drought negatively affects both primary and secondary growth of *Q.
+pyrenaica* forests.
+This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by the consistent radial-growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
+Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
 
-By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the *Quercus pyrenaica* forests of Sierra Nevada and their resilience to several severe drought events in the recent decades.
-Our results indicated that these relict oak populations driven by historical land use are resilient to climate change at their present rear edge.
-However, resistance, resilience, and forest recovery to extreme drought events are strongly influenced by mountain exposure, local environmental conditions, and management legacies.
-This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance varies markedly along the rear edge under climate change.
+Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations of Sierra Nevada.
+The northern site showed higher browning intensity than the southern sites during the 2005 drought event, and stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014].
+Greenness proved less sensitive to drought than did tree growth, particularly for drier sites.
+These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013; @Fatichi2014].
+Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
+~~It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]~~.
+On the other hand, trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4).
+This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and overall both are considered to constitute the rear edge for the species.
 
-### Land-use legacies shape sensitivity to climate change of forests and the present rear edge
-The review of historical documents revealed that forest clearings, firewood removal, charcoal making, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species approaches 90% of the cover at medium and low elevations [@JimenezOlivencia2015].
+As with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
+pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
+Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
+In fact, growth of some tree species in their rear edge (*e.g.* Abies alba) have also shown to be more sensitive to moisture-related variables [@MartinezSanchoGutierrez2019EvidenceThat]; others species respond simultaneously to both temperature and moisture related variables [*e.g.* *Fagus sylvatica*, @Dorado2017; @Dorado2017c; *Pinus nigra* subsp.
+*salzmanii*, @SanchezSalguero2012].
+Even more, other species be more dependent to temperatures [*e.g.* *Pinus sylvestris*, @Herrero2013].
+This diversity of responses by tree species to the two main components of climate change (*i.e.* increase of temperature and rainfall reduction) highlihts that the vulnerability to climate is not consistently expressed within the rear edge, evidencing that geographically marginal forests are not necessarily climatically marginal [see @DoradoLinanetal2019GeographicalAdaptation and references therein].
+~~A negative effect of temperature expressing a limiting effect of high vapor pressure deficit and high potential evapotranspiration is to be expected at drought-limited rear-edges.
+In any case, this proves that the vulnerability to climate is not consistently expressed within the rear edge, evidencing that geographically marginal forests are not necessarily climatically marginal~~
+
+~~Esta diversidad de respuestas de las diferentes especies forestales a los dos componentes principales del cambio climático (incremento de temperatura vs reducción de precipitación) dificultan aún mas la posibilidad de encontrar un patrón general de ajustes finos entre limite ecológico y límite geográfico (see Dorado-Liñan and others 2019 and references therein).~~
+
+### Land-use legacies shape forest sensitivity to climate change of forests at the present rear edge
+The review of historical documents revealed that forest clearings, firewood removal, charcoal making, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species would approach 90% in tree cover at medium and low elevations [@JimenezOlivencia2015].
 Together with the analysis of the disturbance chronologies, the observed notable differences in stand structure, tree size, and age suggest different forest histories and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
@@
 Firewood, charcoal, and acorns have been intensively exploited at the southern sites, until at least the mid-20th century, when these activities sharply declined due mainly to rural abandonment and the use of gas and fossil fuels [@Valbuena2013].
-At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management type which has been applied in the past in Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
+At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management practice often applied in the past in many Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
 
-The other release event observed for the SJ site during the period 1995-2000 was lower than during 1940, but also affected most trees (Figures 4, S2b).
+The other release event observed for the SJ site during the period 1995-2000 was lower than during 1940, but also affected most trees (Figures 4b, S2b).
 No records of forest practices in this area over the last 30 years have been found [@Bonet2016obsnev_forest], and no logging has been recorded during the period 1995 - 2000 (F.J.
@@
 Therefore this release might be related to natural drought-induced mortality after 1995, as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
-On the other hand, the strong positive correlations of SPEI with tree growth for this site show high sensitivity to water availability [@GeaIzquierdo2014].
+
+### Relict oaks show high resilience to recent drought events and long-term climatic variability 
+Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q.
+pyrenaica* for the last 16 years.
+This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
+For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4a).
+Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014].
+This could be related to a non-linear positive effect of warming for species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014].
+Importantly, for rear edges threatened by climate change, negative growth trends were expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
+
+Although the 2012 drought event was more severe and intense than that of 2005 (Table S3), resilience values for greenness and tree growth were greater for 2012.
+This could be due to the different timing of the two droughts.
+The 2012 event was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought.
+The latter matched the period of maximum growth for oak forests in late spring (Figure S4).
+This would highlight the importance of the timing of the drought as a key factor determining tree recovery after drought [@Camarero2015b; @Huang2018].
+For tree growth, the highest values of resilience were found for the two most severe events (1995 and 1999; Table S3) and tree-growth resilience was positively related to drought severity (Figure 5a).
+
+The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@Valbuena2013; @Valbuena2017] appear to indicate the strong local adaptation of this oak species.
+Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017].
+In our case, this can be explained by the fact that the current geographical rear-edge does not match with the potential ecologically rear-edge for the species because this has been determined mostly by human use.
+@MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations.
+Our results highlight the ample small-scale variability at the ecological boundary and need to better define and delimit the rear-edges.
+All this, together with the characteristic high resprouting ability of the species, show the long-term persistence of these populations  [@Bellingham2000].
+It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the high resilience observed is expressed at the species level or only in adult trees.
+The rear-edge might be different for different ontogenic stages.
+It is important to assess whether there is vulnerability if we analyzed seedling regeneration and recruitment, as in other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].
+
+In summary, two main results stand out from this research.
+First, the high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019].
+The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @Herrero2014].
+Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.
+In fact, there was much small-scale variability in the response to climate within the rear edge that we had *a priori* considered in our study.
+Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the whole geographical rear edge studied.
+This suggests that the rear edge therefore needs to be redefined in space but also in time [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the potential ecological (limiting) rear-edge of species.
+The rear-edge concept should also consider historical aspects in addition to the geographic, climatic and genetic dimensions [@VilaCabrera2019], particularly in areas with a long-history of anthropic management, such as Mediterranean mountains.
+Therefore, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate‐driven population decline at rear edges [@VilaCabrera2019].
+Furthermore, even the geographical dimension of the rear edge is not a continuous, uniform space of ecological conditions in mountain landscapes.
+In other words: in mountains, heterogeneity of ecological conditions at fine scales is the rule, enabling the existence of microrefugia and enhancing species persistence [@Olaldeetal2002WhiteOaksa; @SerraDiazetal2015DisturbanceClimate].
+In this respect, our results showed that resilience of oak stands to drought events was not spatially homogeneous across the mountain range, due to differences in ecological conditions and/or past-management legacies.
+This is particularly relevant for rear-edge populations where topographic and biophysical variability can facilitate the existence of refugia.
+
+## Concluding comments
+The ecological and geographical rear edges did not appear to fully match in our study sistem.
+Severe drought events provoked major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the a priori relict, rear-edge forest studied.
+Nevertheless, contrary to our expectations, the trees exhibited high resilience in response to drought, particularly in the long-term.
+The differences found in tree growth and resilience to drought between close neighboring sites show that responses to drought are site dependent and can drastically vary in very narrow spatial gradients.
+This is particularly relevant to refine the real extension of rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
+The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land use in local forest dynamics, as inferred from an exhaustive review of historical documents.
+In consequence, the geographical and ecological rear-edges for specific species do not necessarily meet after centuries or millennia of human impacts on the real world landscapes.
+This is relevant for tree species with a high sensitivity to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these forests offer.
+
+9999999999999999
+OLD
 
@@
 
-### Sensitivity (resistance) of relict oaks to recent drought events
-Severe drought negatively affects both primary and secondary growth of *Quercus pyrenaica* forests.
-This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by the consistent radial-growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
-Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
-As with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Quercus pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
-Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
-
-Greenness proved less sensitive to drought than did tree growth, particularly for drier sites.
-These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013; @Fatichi2014].
-Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
-
-Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations.
-For example, the northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and the stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014].
-It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM].
-Trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4).
-This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and overall both are considered to constitute the rear edge for the species.
-
 ```{r eval=FALSE, echo=FALSE}
@@
 
-### Relict oaks show high resilience to recent drought events and long-term climatic variability 
-Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Quercus pyrenaica* for the last 16 years.
-This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
-For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4).
-Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014].
-This could be related to a non-linear positive effect of warming for the species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014].
-Importantly, for rear edges threatened by climate change, negative growth trends would have been expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
-
-Although the 2012 drought event was more severe and intense than that of 2005 (Table S3), resilience values for greenness and tree growth were greater for 2012.
-This could be due to the different timing of the two droughts.
-The 2012 event was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought.
-The latter matched the period of maximum growth for oak forests in late spring (Figure S4).
-This highlights the importance of the timing of the drought as a key factor determining tree recovery after drought [@Camarero2015b; @Huang2018].
-For tree growth, the highest values of resilience were found for the two most severe events (1995 and 1999; Table S3) and tree-growth resilience was positively related to drought severity (Figure 5).
-
-The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@Valbuena2013; @Valbuena2017] appear to indicate the strong local adaptation of this oak.
-Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017].
-@MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations.
-Our results highlight the ample small-scale variability at the ecological boundary and need to better confine the rear-edge limit in our forest.
-All this, together with the characteristic high resprouting ability of the species, would suggest a long-term persistence of those populations [@Bellingham2000].
-It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the trees express resilience at the species level or whether we would find vulnerability if we analyzed seedling regeneration, as in other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].
-
 ```{r eval=FALSE, echo=FALSE}
@@
 ```
-
-In summary, two main results stand out from this research.
-First, the high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019].
-The current niche is a result of land-use changes, which further complicate the definition of potential rear edges.
-The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @Herrero2014].
-
-Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.
-In fact, there was much small-scale variability in the response to climate within the rear edge that we had *a priori* considered in our study.
-Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the geographical rear edge studied.
-This suggests that the rear edge therefore needs to be redefined [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the ecological (limiting) rear edge of species.
-
-Overall, our results show that management history influences tree growth and resilience to climate change of tree species, highlighting the importance of land-use legacies in Mediterranean forests [@Navarro2013; @DoblasMiranda2017] which, hence, will also strongly determine the current geographical distribution of marginal stands, regardless of the potential extent of ecological marginality of species.
 
@@
 ```
-
-## Concluding comments
-The ecological and geographical rear edges did not appear to fully match in our study.
-Severe drought events provoke major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the relict, rear-edge forest studied.
-However, we found no negative growth trends despite our expectation of vulnerability to climate change for these relict stands.
-Furthermore, we detected positive trends for primary growth (*i.e.* greenness) at the ecosystem scale and a steep positive trend of secondary growth at the tree level at the wettest site along the climatic gradient analyzed.
-The trees exhibited high resilience values in response to drought, particularly in the long-term scale.
-These findings are consistent with other results showing that this mountain region is still acting as a refuge for deciduous species, including *Quercus*.
-The differences found in tree growth and resilience to drought between sites close together show that responses to drought were site dependent and can drastically vary even in very narrow spatial gradients (*i.e.* following ecological thresholds).
-This is particularly relevant for rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
-The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land use in local forest dynamics, as inferred from an exhaustive review of historical documents.
-In this sense, our results highlight the importance of land-use legacies for highly transformed Mediterranean systems.
-This is relevant for tree species with a high sensitivity to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these singular forests offer.
 

### 2020-06-16 man/ecosystems/ms_ecosystems.Rmd
update biblio and figures

@@
     reference_docx: ../../templates/template_ms.docx
-bibliography: ../../refs/references.bib
+bibliography: ../../refs/refs_phd_dendro.bib
 csl: ../../refs/ecosystems.csl
@@
 * Global change challenges ecosystems in xeric locations transformed by intensive human use.
-Resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd.
-populations in the southern Iberian Peninsula was analyzed in relation to historical records of land-use, combining dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
+Resilience to drought of relict Mediterranean *Quercus pyrenaica* populations in the southern Iberian Peninsula was analyzed in relation to historical records of land-use, combining dendroecological growth of adult trees and greenness (EVI) as proxies for secondary and primary growth.
 
@@
 ### Keywords 
-extreme drought, resilience, rear edge, *Quercus pyrenaica*, tree growth, dendrochronology, remote sensing
+extreme drought, resilience, rear edge, *Quercus pyrenaica*, tree growth, dendroecology, remote sensing
 
@@
 
-The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@Hampe2005].
-Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @Sagarin2002; @Pironon2016].
+The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@HampePetit2005ConservingBiodiversity].
+Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @SagarinGaines2002AbundantCentre; @Pirononetal2017GeographicVariation].
 It has been often assumed that geographically marginal populations represent ecological marginal populations.
-This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pironon2016; @VilaCabrera2019; @Oldfatheretal2020RangeEdges].
-Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sexton2009; @Abeli2014].
-This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and a genetic component [@VilaCabrera2019], but also an anthropogenic (*i.e.* land-use) dimension.
+This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pirononetal2017GeographicVariation; @VilaCabreraetal2019RefiningPredictions].
+Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sextonetal2009EvolutionEcology; @Abelietal2014EffectsMarginality; @Oldfatheretal2020RangeEdges].
+This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and a genetic component [@VilaCabreraetal2019RefiningPredictions], but also an anthropogenic (*i.e.* land-use) dimension.
 In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to identify and understand mismatches between the geographical and the ecological rear-edges [@VilaCabreraJump2019GreaterGrowth].
 
-Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sexton2009].
-Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allen2010].
-Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerrano2014; @Stagge2017].
+Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sextonetal2009EvolutionEcology].
+Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allenetal2010GlobalOverview].
+Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerranoetal2014EvidenceIncreasing; @Staggeetal2017ObservedDrought].
 In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge.
-The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@Hampe2005; @VilaCabrera2019].
+The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@HampePetit2005ConservingBiodiversity; @VilaCabreraetal2019RefiningPredictions].
 
-It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Sala2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @Navarro2013; @DoblasMiranda2017].
+It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Salaetal2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @NavarroGonzalezetal2013WeightLanduse; @DoblasMirandaetal2017ReviewCombination].
 In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects].
-The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLandUse].
+The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLanduse].
 For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests].
 Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica].
-In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forest at the southern geographical range.
-Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLandUse], and particularly that of marginal rear edge populations [@VilaCabrera2019].
+In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forests at their southern geographical range.
+Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLanduse], and particularly that of marginal rear edge populations [@VilaCabreraetal2019RefiningPredictions].
 
 The assessment of resilience to climate and human disturbances provides critical information concerning the capacity of forests to maintain their structure and render valuable ecosystem services.
-Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYoua].
-@Lloret2011 proposed an approach, which decomposes resilience to drought in three components: resistance to drought, recovery after drought and resilience.
+Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYou].
+@Lloretetal2011ComponentsTree proposed an approach, which decomposes resilience to drought in three components: resistance to drought, recovery after drought and resilience.
 Forest resilience is determined by the capacity to reduce the impact (resistance) and the capacity to recover from the impact of disturbance (recovery) [@IngrischBahn2018ComparableQuantification].
@@
 In this sense we assessed forest resilience both in the long-term to several recent extreme drought episodes and to climate change (*i.e.* warming) in the long-term (the last few decades) using two different proxies to characterize resilience.
-Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and, generally annual tree-ring widths have been used to study growth changes in response to drought at the individual tree level [@Fritts1976; @Dobbertin2005TreeGrowth].
-Remote sensing can be used to analyze the impact of drought on ecosystems and hence at the stand level [*e.g.* @Zhang2013].
-The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazol2018; @PenaGallardo2018].
+Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and, generally annual tree-ring widths have been used to study growth changes in response to drought at the individual tree level [@Fritts1976TreeRings; @Dobbertin2005TreeGrowth].
+Remote sensing can be used to analyze the impact of drought on ecosystems and hence at the stand level [*e.g.* @Zhangetal2013MonitoringEstimating].
+The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerranoetal2013ResponseVegetation; @Coulthardetal2017TreeGrowth], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness].
 
 In the present study, we assessed resilience of *Quercus pyrenaica* Willd.
-(Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdo2013; @GeaIzquierdoetal2017RiskyFuture].
+(Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].
 We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth).
@@
 # Retales de la intro 
-# Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018].
+# Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardoetal2018DroughtSensitiveness; @Gazoletal2018ForestResilience].
 
@@
 
-# Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].
+# Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@HampePetit2005ConservingBiodiversity].
 
-# Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; @MartinezVilalta2016].
+# Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allenetal2010GlobalOverview], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerranoetal2013ResponseVegetation; @MartinezVilalta2016].
 
-# Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [@Herrero2014].
+# Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @DoradoLinanetal2017LargescaleAtmospheric], and some focused on the resilience to drought of rear-edge populations [@Herrero2014].
 Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973].
@@
 ### Tree species and study site 
-*Quercus pyrenaica* forests extend throughout south-western France and the Iberian Peninsula, reaching their southern limit in mountain areas of northern Morocco [@Franco1990].
+*Quercus pyrenaica* forests extend throughout south-western France and the Iberian Peninsula, reaching their southern limit in mountain areas of northern Morocco [@Franco1990Quercus].
 In the Iberian Peninsula, these forests occupy siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid, and hyperhumid ombroclimate.
 Pyrenean oak is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation.
-As a submediterranean species, it has lower drought tolerance than evergreen Mediterranean taxa [@delRio2007].
+As a submediterranean species, it has lower drought tolerance than evergreen Mediterranean taxa [@delRioetal2007BioclimaticAnalysis].
 
@@
 The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability according to elevation and aspect.
-Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olalde2002].
+Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olaldeetal2002WhiteOaks].
 Eight Pyrenean oak patches (2400 ha) have been identified in this mountain range (Figure 1), from 1100 to 2000 m *a.s.l.* and often associated with major river valleys.
-Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005].
-They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008ABG].
-They also harbor high levels of intraspecific genetic diversity [@Valbuena2013].
-These relict forests have undergone intensive human use throughout history [@CamachoOlmedo2002].
-Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdo2013; @GeaIzquierdoetal2017RiskyFuture].
+Today, *Quercus pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@HampePetit2005ConservingBiodiversity].
+They are the richest forest formation in vascular plant species of Sierra Nevada, containing several endemic and endangered plant species [@Loriteetal2008PhytosociologicalReview].
+They also harbor high levels of intraspecific genetic diversity [@ValbuenaCarabanaGil2013GeneticResilience].
+These relict forests have undergone intensive human use throughout history [@CamachoOlmedoetal2002DinamicaEvolutiva].
+Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].
 
 ### Climatic data and drought episodes
-The Iberian Peninsula underwent several extreme drought episodes in the last three decades [*e.g.* 1995, 1999, 2005, 2012;  @VicenteSerrano2014].
-The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], appearing as extreme drought in our climatic data (Figure S1; Table S3).
+The Iberian Peninsula underwent several extreme drought episodes in the last three decades [*e.g.* 1995, 1999, 2005, 2012;  @VicenteSerranoetal2014EvidenceIncreasing].
+The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoaetal2017DroughtTrends], appearing as extreme drought in our climatic data (Figure S1; Table S3).
 We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below).
 Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3).
-A drought event was identified using the SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] (SPEI 12-months scale), following a procedure similar to the one proposed by @Spinoni2015.
+A drought event was identified using the SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerranoetal2010MultiscalarDrought] (SPEI 12-months scale), following a procedure similar to the one proposed by @Spinonietal2015EuropeanDrought.
 We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html).
-A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017].
+A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinonietal2018WillDrought; @Pascoaetal2017DroughtTrends].
 A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months.
@@
 
-To explore the relationships between climatic variables and tree-growth variables we used climate data obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008].
+To explore the relationships between climatic variables and tree-growth variables we used climate data obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylocketal2008EuropeanDaily].
 Monthly precipitation and minimum and maximum temperatures had a 0.25 x 0.25 º resolution for the 1950-2016 period.
@@
 We selected the pixels covering the distribution of *Quercus pyrenaica* forests in Sierra Nevada (*n* = 928 pixels).
-Any values affected by clouds, snow, shadows or high content aerosols, were filtered out following recommendations for mountain regions [@Reyes2015].
+Any values affected by clouds, snow, shadows or high content aerosols, were filtered out following recommendations for mountain regions [@ReyesDiezetal2015ImplicacionesFiltrado].
 
 The mean Annual EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016.
-The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, in order to minimize bias in the evaluation of anomalies and to provide more information concerning their magnitude [@Samanta2012].
+The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, in order to minimize bias in the evaluation of anomalies and to provide more information concerning their magnitude [@Samantaetal2012InterpretationVariations].
 For each pixel, an annual EVI value was calculated by averaging EVI valid values.
@@
 where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period.
-Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].
+Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samantaetal2012InterpretationVariations].
 
-Rather than other vegetation indices such as the NDVI, $EVI_{mean}$ was chosen because it is highly stable under the use of any filter [@Reyes2015] and because it showed highly significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88).
+Rather than other vegetation indices such as the NDVI, $EVI_{mean}$ was chosen because it is highly stable under the use of any filter [@ReyesDiezetal2015ImplicacionesFiltrado] and because it showed highly significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88).
 
@@
 The three sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1).
-All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015].
+All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuqueetal2015DatasetMIGRAME].
 At each site, between 15 and 20 trees from either the single dominant-codominant layer in CA or the open canopy in SJ were randomly sampled.
@@
 In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species, and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot with a 10-m radius.
-Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details].
+Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdoCanellas2009AnalysisHolm for more details].
 
@@
 Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI).
-Cross-dating validation was finally verified using COFECHA [@Holmes1983].
+Cross-dating validation was finally verified using COFECHA [@Holmes1983ComputerassistedQuality].
 
@@
 To study the growth response to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used.
-These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976].
-Tree-ring width series were standardized and detrended using dplR [@Bunn2010].
-Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976].
+These were calculated from ratios between raw growth measurements and individual cubic splines with a 50% frequency cutoff at 30 years [@Fritts1976TreeRings].
+Tree-ring width series were standardized and detrended using dplR [@Bunn2010StatisticalVisual].
+Mean residual site chronologies were established by computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Fritts1976TreeRings].
 The statistical quality of each chronology was checked via the expressed population signal (EPS).
@@
 The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$).
-In theory, BAI represents a more accurate indicator of growth than ring width, because it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile growth [@Biondi2008].
-Raw ring widths and measured DBH were used to compute BAI [@Piovesan2008] with the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
+In theory, BAI represents a more accurate indicator of growth than ring width, because it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile growth [@BiondiQeadan2008TheorydrivenApproach].
+Raw ring widths and measured DBH were used to compute BAI [@Piovesanetal2008DroughtdrivenGrowth] with the following equation: $\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$ where $r$ is the radius of the tree and $t$ is the year of tree-ring formation.
 For each individual tree, a mean BAI series was calculated.
@@
 ### Disturbance analyses and land-use history review 
-Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (suppressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past.
-Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@Rubino2004].
+Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (suppressions) in radial growth [@NowackiAbrams1997RadialgrowthAveraging] as indirect estimates of possible disturbance events (*e.g.* logging, drought-induced neighbor mortality) in the past.
+Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left[(M1 - M2)\Big/ M2\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median [@RubinoMcCarthy2004ComparativeAnalysis].
 ```{r, eval=FALSE, echo=FALSE}
@@
 Site-disturbance chronologies were constructed by annually averaging the individual disturbance series.
-To separate growth peaks caused by disturbance events and expressing stand-wise disturbances from those caused by climate, we considered a threshold of 50% of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014].
+To separate growth peaks caused by disturbance events and expressing stand-wise disturbances from those caused by climate, we considered a threshold of 50% of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdoCanellas2014LocalClimate].
 In addition, the history of the forest and management of our sampling sites was inferred from a detailed analysis of historical land-use changes.
@@
 
-### Assessing ecosystem and tree individual resilience to drought
-To evaluate the effects of drought events on ecosystem resilience (using greenness data) and individual tree resilience (using BAI data), we used resilience indices proposed by @Lloret2011.
+### Assessing resilience to drought at the forest stand and individual-tree levels
+To evaluate the effects of drought events on ecosystem resilience (using greenness data) and individual tree resilience (using BAI data), we used resilience indices proposed by @Lloretetal2011ComponentsTree.
 The Resistance index estimated as the ratio between performance during and before the disturbance ($Resistance=Drought:PreDrought$) quantifies the severity of the impact of the disturbance in the year it occurred.
@@
 The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value over a period of three years before and after the drought event, respectively.
-A period of three years was chosen because we found similar results on comparing periods of two, three, and four years (Figure S3b), and this time period has been used in other studies [*e.g.* @Gazol2018].
+A period of three years was chosen because we found similar results on comparing periods of two, three, and four years (Figure S3b), and this time period has been used in other studies [*e.g.* @Gazoletal2018ForestResilience].
 Resilience metrics for BAI data were additionally computed for the most severe drought events since 1940 (*n* = 8; Table S3) and compared with drought severity.
@@
 The severe drought events since 1901 were identified using SPEI-12.
-They were characterized in terms of duration, severity and intensity [see @Spinoni2015].
+They were characterized in terms of duration, severity and intensity [see @Spinonietal2015EuropeanDrought].
 In a first step, the impact of drought in greenness and growth was explored using the $EVI_{sa}$ and the mean RWI site chronologies (Figure 2).
-Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim [@Zang2015].
+Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim [@ZangBiondi2015TreeclimPackage].
 The non-climatic disturbance impacts on tree-growth were evaluated using site disturbance chronologies (buit using growth changes on tree growth).
@@
 For each of the three resilience indices studied, we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the oak populations studied (northern and southern exposures).
-These tests were used because original and log-transformed data did not follow the assumptions of normality or homogeneity of variance [@Wilcox2012].
-Robust measures of central tendency (M-estimator based on Huber's Psi) were used because they were close to the mean value in all cases [@Wilcox2012].
+These tests were used because original and log-transformed data did not follow the assumptions of normality or homogeneity of variance [@Wilcox2012IntroductionRobust].
+Robust measures of central tendency (M-estimator based on Huber's Psi) were used because they were close to the mean value in all cases [@Wilcox2012IntroductionRobust].
 When the robust ANOVA test was run, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers.
 *Post-hoc* differences were assessed pairwise using a similar bootstrap test.
-All the robust ANOVA and *post-hoc* tests were carried out using the WRS2 R package [@Mair2017].
+All the robust ANOVA and *post-hoc* tests were carried out using the WRS2 R package [@Mairetal2017WRS2Wilcox].
 The level of significance was set to 0.05 and adjusted for multiple comparisons.
@@
 ## Discussion
-///////
-By using a combined approach of remote-sensing information and dendrochronology, we have quantified the drought impact on the *Quercus pyrenaica* forests of Sierra Nevada and their resilience to several severe drought events in the recent decades.
-//////
+By using a combined approach of remote-sensing information and dendroecology, we have quantified growth of adult trees and greenness (EVI) as proxies for secondary and primary growth of relict Mediterranean *Quercus pyrenaica* populations in the southern Iberian Peninsula.
+These relict oak populations driven by historical land-use have been resilient to climate change at their present rear edge.
+However, resistance, resilience, and forest recovery to extreme drought events were strongly influenced by mountain exposure, local environmental conditions, and management legacies.
+This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance can vary markedly along the rear edge under climate change.
 
-Our results indicate that these relict oak populations driven by historical land use have been resilient to climate change at their present rear edge.
-However, resistance, resilience, and forest recovery to extreme drought events were strongly influenced by mountain exposure, local environmental conditions, and management legacies.
-This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance can vary markedly along the rear edge under climate change
-
-### High sensitivity of oaks to recent drought events
+### High sensitivity and variability in the oak sensitivity to climate at the rear-edge
 Severe drought negatively affects both primary and secondary growth of *Q.
 pyrenaica* forests.
-This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by the consistent radial-growth suppression for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014].
-Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018].
+This was expressed by the observed reduction in greenness and tree growth in response to the 2005 and 2012 drought events as well as by radial-growth suppression during extreme drought events [@Corcueraetal2006RadialgrowthWoodanatomical; @GeaIzquierdoCanellas2014LocalClimate].
+Furthermore, the greatest reduction of tree growth was detected during the 1995 drought, a characteristic negative precipitation anomaly that caused severe and extensive damage in the Mediterranean Iberian Peninsula [@Penuelasetal2001SevereDrought; @Gazoletal2018ForestResilience].
 
+The tree responses to drought are site-dependent [@Babstetal2013SiteSpecies], particularly for rear-edge populations [@CavinJump2017HighestDrought; @DoradoLinanetal2017LargescaleAtmospheric].
 Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations of Sierra Nevada.
-The northern site showed higher browning intensity than the southern sites during the 2005 drought event, and stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdo2014].
-Greenness proved less sensitive to drought than did tree growth, particularly for drier sites.
-These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018; @PenaGallardo2018], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013; @Fatichi2014].
-Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018].
-~~It is well known that tree growth and tree responses to drought are site-dependent [*e.g.* soil features, tree competition; @Babst2013], particularly for rear-edge populations [@Cavin2017; @Dorado2017AFM]~~.
+The northern site showed higher browning intensity than the southern sites during the 2005 drought event, and stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdoCanellas2014LocalClimate].
+Greenness was less sensitive to drought than tree growth, particularly for drier sites.
+These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babstetal2013SiteSpecies; @Coulthardetal2017TreeGrowth; @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013GrowthControls; @Fatichietal2014MovingPhotosynthesis].
+Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babstetal2017ImprovedTreering] and provide an accurate measure of growth responses to droughts [@Bhuyanetal2017DifferentResponses; @Gazoletal2018ForestResilience].
 On the other hand, trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4).
-This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and overall both are considered to constitute the rear edge for the species.
+This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and both are considered to constitute the rear edge for the species.
 
 As with many other forest species under Mediterranean climates, moisture availability is generally the most limiting factor driving radial growth of *Q.
-pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdo2014].
-Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippo2010; @GeaIzquierdo2011; @GarciaGonzalez2017].
-In fact, growth of some tree species in their rear edge (*e.g.* Abies alba) have also shown to be more sensitive to moisture-related variables [@MartinezSanchoGutierrez2019EvidenceThat]; others species respond simultaneously to both temperature and moisture related variables [*e.g.* *Fagus sylvatica*, @Dorado2017; @Dorado2017c; *Pinus nigra* subsp.
-*salzmanii*, @SanchezSalguero2012].
-Even more, other species be more dependent to temperatures [*e.g.* *Pinus sylvestris*, @Herrero2013].
-This diversity of responses by tree species to the two main components of climate change (*i.e.* increase of temperature and rainfall reduction) highlihts that the vulnerability to climate is not consistently expressed within the rear edge, evidencing that geographically marginal forests are not necessarily climatically marginal [see @DoradoLinanetal2019GeographicalAdaptation and references therein].
-~~A negative effect of temperature expressing a limiting effect of high vapor pressure deficit and high potential evapotranspiration is to be expected at drought-limited rear-edges.
-In any case, this proves that the vulnerability to climate is not consistently expressed within the rear edge, evidencing that geographically marginal forests are not necessarily climatically marginal~~
+pyrenaica* along its distribution range in the Iberian Peninsula [@GeaIzquierdoCanellas2014LocalClimate].
+Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippoetal2010ClimateChange; @GeaIzquierdoetal2011TreeringsReflect; @GarciaGonzalezSoutoHerrero2017EarlywoodVessel].
+A positive effect of moisture availability and negative of temperature expressing a limiting effect of high vapor pressure deficit and potential evapotranspiration can be expected at drought-limited rear-edges.
+Yet, at the rear-edge, growth of some tree species (*e.g.* *Abies alba*) have been shown to be more sensitive to moisture-related variables [@MartinezSanchoGutierrezMerino2019EvidenceThat], others species were more sensitive to temperatures [*e.g.* *Pinus sylvestris*, @Herreroetal2013VaryingClimate], and growth of other species responded simultaneously to both temperature and moisture related variables [*e.g.* *Fagus sylvatica*, @DoradoLinanetal2017CoexistenceMediterraneanTemperate; @DoradoLinanetal2017ClimateThreats; *Pinus nigra* subsp.
+*salzmanii*, @SanchezSalgueroetal2012DroughtMain].
+This diversity in the response of tree species to precipitation and temperature suggests that vulnerability to climate change is not consistently expressed within the rear edge, therefore evidencing that geographically marginal forests are not necessarily climatically or ecologically marginal [see @DoradoLinanetal2019GeographicalAdaptation and references therein].
 
-~~Esta diversidad de respuestas de las diferentes especies forestales a los dos componentes principales del cambio climático (incremento de temperatura vs reducción de precipitación) dificultan aún mas la posibilidad de encontrar un patrón general de ajustes finos entre limite ecológico y límite geográfico (see Dorado-Liñan and others 2019 and references therein).~~
-
-### Land-use legacies shape forest sensitivity to climate change of forests at the present rear edge
-The review of historical documents revealed that forest clearings, firewood removal, charcoal making, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species would approach 90% in tree cover at medium and low elevations [@JimenezOlivencia2015].
-Together with the analysis of the disturbance chronologies, the observed notable differences in stand structure, tree size, and age suggest different forest histories and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
-On the northern slopes of Sierra Nevada (*e.g.* SJ site), land uses have been historically distributed along an elevational gradient: grasslands and shrublands for cattle farming at the highest elevations; then forest stands with some croplands; and irrigated terraces with tree crops at the lowest elevations [@JimenezOlivencia2015].
-In addition, other activities such as mining must have altered the forest structure, *e.g.* the SJ site has many small mines and quarries that were exploited intermittently throughout history.
-The release growth event discerned for the 1940s concurs with a period of maximum mining activity in this area (1925 to 1957), during which timber use increased for mine tunnels and furnaces, these also requiring large amounts of firewood to melt the mineral (Table S4).
-This heavy exploitation of the neighboring forest resources must have affected a significant part of this oak woodland, as shown by growth of the remnant trees at the northern site (Figure S2b).
-
-On the other hand, woodlands on the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks grow [@JimenezOlivencia2015].
-Firewood, charcoal, and acorns have been intensively exploited at the southern sites, until at least the mid-20th century, when these activities sharply declined due mainly to rural abandonment and the use of gas and fossil fuels [@Valbuena2013].
-At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management practice often applied in the past in many Iberian oak woodlands [@Canellas2004; @GeaIzquierdo2011] and which has been documented for this site [@Valbuena2013].
-
-The other release event observed for the SJ site during the period 1995-2000 was lower than during 1940, but also affected most trees (Figures 4b, S2b).
-No records of forest practices in this area over the last 30 years have been found [@Bonet2016obsnev_forest], and no logging has been recorded during the period 1995 - 2000 (F.J.
-Cano-Manuel *personal communication*).
-Therefore this release might be related to natural drought-induced mortality after 1995, as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelas2001; @Lloret2004].
-
-### Relict oaks show high resilience to recent drought events and long-term climatic variability 
+### Relict oaks show high resilience to drought at different spatio-temporal scales: do the geographical an ecological rear-edges match?
 Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q.
 pyrenaica* for the last 16 years.
-This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuque2015onto].
+This is consistent with previous findings stressing a recent short-term increase in primary productivity for these forests coinciding with a rather wet decade in the 2000s after a dry decade in the 1990s [@PerezLuqueetal2015OntologicalSystem].
 For tree growth, positive trends also appeared in the last decade, particularly for the southern high-elevation site (CA-High, Figure 4a).
-Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdo2014].
-This could be related to a non-linear positive effect of warming for species at cold-limited high-elevation sites [@Salzer2009; @GeaIzquierdo2014].
-Importantly, for rear edges threatened by climate change, negative growth trends were expected, as shown for some temperate and Mediterranean species [@SanchezSalguero2012; @Camarero2015; @Dorado2017].
+Similar long-term trends have been described for this species along its distribution range only at high-elevation wet and cold sites [@GeaIzquierdoCanellas2014LocalClimate].
+This could be related to a non-linear positive effect of warming for species at cold-limited high-elevation sites [@Salzeretal2009RecentUnprecedented; @GeaIzquierdoCanellas2014LocalClimate].
+Importantly, for rear edges threatened by climate change, negative growth trends were expected, as shown for some temperate and Mediterranean species [@SanchezSalgueroetal2012DroughtMain; @Camareroetal2015NotEarly; @DoradoLinanetal2017CoexistenceMediterraneanTemperate].
 
@@
 This could be due to the different timing of the two droughts.
-The 2012 event was a winter drought [@Trigo2013] occurring earlier than the shorter 2005 drought.
+The 2012 event was a winter drought [@Trigoetal2013RecordWinter] occurring earlier than the shorter 2005 drought.
 The latter matched the period of maximum growth for oak forests in late spring (Figure S4).
-This would highlight the importance of the timing of the drought as a key factor determining tree recovery after drought [@Camarero2015b; @Huang2018].
+This would highlight the importance of the timing of the drought as a key factor determining tree recovery after drought [@Camareroetal2015TimingDrought; @Huangetal2018DroughtTiming].
 For tree growth, the highest values of resilience were found for the two most severe events (1995 and 1999; Table S3) and tree-growth resilience was positively related to drought severity (Figure 5a).
 
-The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@Valbuena2013; @Valbuena2017] appear to indicate the strong local adaptation of this oak species.
-Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @Cavin2017].
+The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@ValbuenaCarabanaGil2013GeneticResilience; @ValbuenaCarabanaGil2017CentenaryCoppicing] appear to indicate the strong local adaptation of this oak species.
+Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @CavinJump2017HighestDrought].
 In our case, this can be explained by the fact that the current geographical rear-edge does not match with the potential ecologically rear-edge for the species because this has been determined mostly by human use.
-@MartinezVilalta2018 pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations.
+@MartinezVilalta2018RearWindow pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations.
 Our results highlight the ample small-scale variability at the ecological boundary and need to better define and delimit the rear-edges.
-All this, together with the characteristic high resprouting ability of the species, show the long-term persistence of these populations  [@Bellingham2000].
+All this, together with the characteristic high resprouting ability of the species, show the long-term persistence of these populations  [@BellinghamSparrow2000ResproutingLife].
 It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the high resilience observed is expressed at the species level or only in adult trees.
 The rear-edge might be different for different ontogenic stages.
-It is important to assess whether there is vulnerability if we analyzed seedling regeneration and recruitment, as in other Mediterranean species at their xeric limit [@Castro2004; @VilaCabrera2011; @GeaIzquierdo2015EJFR].
+It is important to assess whether there is vulnerability if we analyzed seedling regeneration and recruitment, as in other Mediterranean species at their xeric limit [@Castroetal2004SeedlingEstablishment; @VilaCabreraetal2011StructuralClimatic; @GeaIzquierdoetal2015ThisEnd].
+
+### Land-use legacies shape forest sensitivity to climate change of forests and the present rear-edge
+The review of historical documents revealed that forest clearings, firewood removal, charcoal production, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species would approach 90% in tree cover at medium and low elevations [@JimenezOlivenciaetal2015MedioSiglo].
+Together with the analysis of the disturbance chronologies, the observed notable differences in stand structure, tree size, and age suggest different forest histories and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
+On the northern slopes of Sierra Nevada (*e.g.* SJ site), land uses have been historically distributed along an elevational gradient: grasslands and shrublands for cattle farming at the highest elevations; then forest stands with some croplands; and irrigated terraces with tree crops at the lowest elevations [@JimenezOlivenciaetal2015MedioSiglo].
+In addition, other activities such as mining must have altered the forest structure, *e.g.* the SJ site has many small mines and quarries that were exploited intermittently throughout history.
+The release growth event expressed in the 1940s concurs with a period of maximum mining activity in this area (1925 to 1957), during which timber use increased for mine tunnels and furnaces, these also requiring large amounts of firewood to melt the mineral (Table S4).
+This heavy exploitation of the neighboring forest resources must have affected a significant part of this oak woodland, as shown by growth of the remnant trees at the northern site (Figure S2b).
+
+On the other hand, woodlands on the southern slopes (*e.g.* CA site) were mixed with a greater percentage of croplands along the elevational gradient where oaks grow [@JimenezOlivenciaetal2015MedioSiglo].
+Firewood, charcoal, and acorns were intensively exploited at the southern sites, until at least the mid-20th century, when these activities sharply declined due mainly to rural abandonment and the use of gas and fossil fuels [@ValbuenaCarabanaGil2013GeneticResilience].
+At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management practice often applied in the past in many Iberian oak woodlands [@Canellasetal2004GrowthResponse; @GeaIzquierdoetal2011TreeringsReflect] and which is documented for this site [@ValbuenaCarabanaGil2013GeneticResilience].
+
+The other release event observed for the SJ site during the period 1995-2000 was lower than during 1940, but also affected most trees (Figures 4b, S2b).
+No records of forest practices in this area over the last 30 years have been found [@Bonetetal2016HistorySierra], and no logging was recorded during the period 1995 - 2000 (F.J.
+Cano-Manuel *personal communication*).
+Therefore this release might be related to natural drought-induced mortality after 1995, as has been reported for other Mediterranean tree species after severe drought [*e.g.* @Penuelasetal2001SevereDrought; @Lloretetal2004CanopyRecovery].
 
 In summary, two main results stand out from this research.
-First, the high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018; @VilaCabrera2019].
-The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @Herrero2014].
+First, the high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018RearWindow; @VilaCabreraetal2019RefiningPredictions].
+The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @HerreroZamora2014PlantResponses].
 Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.
@@
 Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the whole geographical rear edge studied.
-This suggests that the rear edge therefore needs to be redefined in space but also in time [@VilaCabrera2019], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the potential ecological (limiting) rear-edge of species.
-The rear-edge concept should also consider historical aspects in addition to the geographic, climatic and genetic dimensions [@VilaCabrera2019], particularly in areas with a long-history of anthropic management, such as Mediterranean mountains.
-Therefore, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate‐driven population decline at rear edges [@VilaCabrera2019].
+This suggests that the rear edge therefore needs to be redefined in space but also in time [@VilaCabreraetal2019RefiningPredictions], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the potential ecological (limiting) rear-edge of species.
+The rear-edge concept should also consider historical aspects in addition to the geographic, climatic and genetic dimensions [@VilaCabreraetal2019RefiningPredictions], particularly in areas with a long-history of anthropic management, such as Mediterranean mountains.
+Therefore, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraetal2019RefiningPredictions].
 Furthermore, even the geographical dimension of the rear edge is not a continuous, uniform space of ecological conditions in mountain landscapes.
-In other words: in mountains, heterogeneity of ecological conditions at fine scales is the rule, enabling the existence of microrefugia and enhancing species persistence [@Olaldeetal2002WhiteOaksa; @SerraDiazetal2015DisturbanceClimate].
+In other words: in mountains, heterogeneity of ecological conditions at fine scales is the rule, enabling the existence of microrefugia and enhancing species persistence [@Olaldeetal2002WhiteOaks; @SerraDiazetal2015DisturbanceClimate].
 In this respect, our results showed that resilience of oak stands to drought events was not spatially homogeneous across the mountain range, due to differences in ecological conditions and/or past-management legacies.
@@
 ## Concluding comments
-The ecological and geographical rear edges did not appear to fully match in our study sistem.
+The ecological and geographical rear edges did not appear to fully match in the studied forest ecosystem.
 Severe drought events provoked major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the a priori relict, rear-edge forest studied.
-Nevertheless, contrary to our expectations, the trees exhibited high resilience in response to drought, particularly in the long-term.
-The differences found in tree growth and resilience to drought between close neighboring sites show that responses to drought are site dependent and can drastically vary in very narrow spatial gradients.
-This is particularly relevant to refine the real extension of rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
+Nevertheless, contrary to our expectations, the trees exhibited high resilience in the response to drought, particularly in the long-term.
+The differences found in tree growth, climatic sensitivity and tree resilience between close neighboring sites show that responses to drought are site dependent and can drastically vary in very narrow spatial gradients.
+This is particularly relevant to refine the real extension and nature (*i.e.* geographical and/or ecological) of rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
 The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land use in local forest dynamics, as inferred from an exhaustive review of historical documents.
@@
 This is relevant for tree species with a high sensitivity to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these forests offer.
-
-9999999999999999
-OLD
 
@@
 These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g.
-@GeaIzquierdo2014; @GeaIzquierdo2015EJFR; @Dorado2017].~~
+@GeaIzquierdoCanellas2014LocalClimate; @GeaIzquierdo2015EJFR; @Dorado2017].~~
 ```
@@
 ```{r eval=FALSE, echo=FALSE}
-#Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of species [@Hampe2005].
+#Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of species [@HampePetit2005ConservingBiodiversity].
 A high vulnerability to drought is often assumed for populations located at their rear-edge [@MartinezVilalta2018].
@@
 A.
-Reyes and J.
+Reyes (*in memoriam*) and J.
 Blanco helped in the filtering of satellite data.

### 2020-06-16 man/figures/0_figures_ecosystems.Rmd
update
updating

@@
 pyrenaica* populations.
+Numbers (*grey circles*) indicate the study aims to which the analyses are related.
 
@@
 
-#####
 **Figure 1**
 ![](../../out/fig/geomap/mapa_situacion_v2.jpg){height=750px}
+
 ##### 
 **Figure 2.**
-![](../../man/figures/schema_methods/schema_analysisv3@1,5x.jpg)0
+![](../../man/figures/schema_methods/schema_analysisv4.jpg)
 

### 2020-08-05 man/ecosystems/ms_ecosystems.Rmd
update version 3

@@
 
-The response of species to changing environments (*e.g.* distributional shifts) can be largely determined by population responses at range margins [@HampePetit2005ConservingBiodiversity].
-Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range [*i.e.* center-periphery hypothesis; @SagarinGaines2002AbundantCentre; @Pirononetal2017GeographicVariation].
-It has been often assumed that geographically marginal populations represent ecological marginal populations.
+The response of species to changing environments (*e.g.* distributional shifts) can be determined largely by population responses at range margins [@HampePetit2005ConservingBiodiversity].
+Peripheral populations are usually considered more vulnerable compared with populations at the center of a species' range [*i.e.* center-periphery hypothesis; @SagarinGaines2002AbundantCentre; @Pirononetal2017GeographicVariation].
+Geographically marginal populations have often been assumed to represent ecologically marginal populations.
 This means lower performance, higher vulnerability, and thus higher risk of extinction than for populations at the core of the species' range [@Rehmetal2015LosingYour; @Pirononetal2017GeographicVariation; @VilaCabreraetal2019RefiningPredictions].
 Nonetheless, recent reviews report that species- and population-specific responses do not always support this hypothesis [@Sextonetal2009EvolutionEcology; @Abelietal2014EffectsMarginality; @Oldfatheretal2020RangeEdges].
-This is partly explained because rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical and a genetic component [@VilaCabreraetal2019RefiningPredictions], but also an anthropogenic (*i.e.* land-use) dimension.
-In this respect, to fully understand changes in distribution and abundance of species as a consequence of global-change, it is crucial to identify and understand mismatches between the geographical and the ecological rear-edges [@VilaCabreraJump2019GreaterGrowth].
+This is partly because a rear-edge is a multidimensional concept including an ecological (*i.e.* climatic and edaphic), a geographical, and a genetic component [@VilaCabreraetal2019RefiningPredictions], but also an anthropogenic dimension (*i.e.* land-use).
+In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change, it is crucial to identify and understand mismatches between the geographical and the ecological rear edges [@VilaCabreraJump2019GreaterGrowth].
 
 Limits of species distribution are strongly determined by climatic factors and biotic interactions [@Gaston2009GeographicRange; @Sextonetal2009EvolutionEcology].
-Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and there are already signs that more intense and longer droughts are altering forest dynamics [@Allenetal2010GlobalOverview].
+Climate change is expected to cause major shifts in the distribution and abundance of plant communities, and signs already indicate that more intense and longer droughts are altering forest dynamics [@Allenetal2010GlobalOverview].
 Drought frequency and severity have increased in recent decades, with a trend towards drier summers, particularly for Southern Europe [@VicenteSerranoetal2014EvidenceIncreasing; @Staggeetal2017ObservedDrought].
-In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes, elevations, and drought-prone areas of a species' distribution, *i.e.* the rear edge.
-The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly relevant [@HampePetit2005ConservingBiodiversity; @VilaCabreraetal2019RefiningPredictions].
+In this climatic-change context, population loss and range retractions are expected in boreal, temperate, and Mediterranean species at the lowest latitudes and elevations, as well as in drought-prone areas of a species' distribution, *i.e.* the rear edge.
+The rear-edge populations are likely to be more sensitive to minor climatic and microtopographic variations and therefore the effects of droughts are expected to be particularly noteworthy [@HampePetit2005ConservingBiodiversity; @VilaCabreraetal2019RefiningPredictions].
 
-It is often overlooked that anthropic-actions are a driver of change as relevant or more as natural drivers like climate [@Salaetal2000GlobalBiodiversity], particularly for regions with long land-use history like the Mediterranean Region [*e.g.* @NavarroGonzalezetal2013WeightLanduse; @DoblasMirandaetal2017ReviewCombination].
-In these areas, the susceptibility and response of ecosystems to natural disturbances are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects].
-The past land-use legacies interact with recent natural disturbances and may confound their interpretation [@Fosteretal2003ImportanceLanduse].
-For example, recent works showed that a quarter of current forests in the Iberian Peninsula, the rear edge of several temperate and boreal tree species, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests].
-Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica].
-In this context, the main focus of our work is to identify the impacts and responses to natural (severe drought) and anthropogenic (*e.g.* logging) disturbances on oak forests at their southern geographical range.
-Thus, the incorporation of an historical perspective should aid to better interpret the responses of ecosystems to natural disturbances [@Fosteretal2003ImportanceLanduse], and particularly that of marginal rear edge populations [@VilaCabreraetal2019RefiningPredictions].
+It is often overlooked that human activity constitutes a driver of change as powerful or even more powerful as natural drivers, *i.e.* natural variation in climate, particularly for regions with long land-use history such as the Mediterranean Region [*e.g.* @NavarroGonzalezetal2013WeightLanduse; @DoblasMirandaetal2017ReviewCombination].
+In these areas, the susceptibility and response of ecosystems to climate change are conditioned by legacies of historical land-use activity [*e.g.* @Munteanuetal2015Legacies19th; @Mausolfetal2018LegacyEffects].
+The past land-use legacies interact with recent human-caused climate disturbances and may confound their interpretation [@Fosteretal2003ImportanceLanduse].
+For example, recent works showed that a quarter of current forests in the Iberian Peninsula, are growing on former agricultural and grazing land abandoned after the 1950s [@VilaCabreraetal2017NewForests].
+Consequently, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate-driven population decline at rear edges [@VilaCabreraJump2019GreaterGrowth; @SanchezdeDiosetal2020FagusSylvatica].
+In this context, our work seeks to identify the impacts and responses to natural (*e.g.* severe drought) and human disturbances (*e.g.* logging) on oak forests at their southern geographical range.
+A historical perspective should help us to interpret the responses of ecosystems to disturbances [@Fosteretal2003ImportanceLanduse], particularly regarding marginal rear-edge populations [@VilaCabreraetal2019RefiningPredictions].
 
@@
 Resilience is the capacity of an ecosystem to persist and maintain its state and functions in the face of disturbance [@Holling1973ResilienceStability; @Hodgsonetal2015WhatYou].
-@Lloretetal2011ComponentsTree proposed an approach, which decomposes resilience to drought in three components: resistance to drought, recovery after drought and resilience.
-Forest resilience is determined by the capacity to reduce the impact (resistance) and the capacity to recover from the impact of disturbance (recovery) [@IngrischBahn2018ComparableQuantification].
-This conceptual approach has been very popular to assess forest resilience recently, because it allows a simple, yet highly efficient assessment of short-term responses of trees to drought.
-Nevertheless, it is not exempted from some criticism and needs to be applied carefully to avoid over interpretation and potential bias associated at different levels [@Schwarzetal2020QuantifyingGrowth].
-In this sense we assessed forest resilience both in the long-term to several recent extreme drought episodes and to climate change (*i.e.* warming) in the long-term (the last few decades) using two different proxies to characterize resilience.
-Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality, and, generally annual tree-ring widths have been used to study growth changes in response to drought at the individual tree level [@Fritts1976TreeRings; @Dobbertin2005TreeGrowth].
-Remote sensing can be used to analyze the impact of drought on ecosystems and hence at the stand level [*e.g.* @Zhangetal2013MonitoringEstimating].
+@Lloretetal2011ComponentsTree proposed an approach, which decomposes resilience into three components: resistance to drought, recovery after drought and resilience.
+This resilience is determined by the forest's ability to mitigate the disturbance (resistance) and the capacity to recover from the impact (recovery) [@IngrischBahn2018ComparableQuantification].
+This conceptual approach has recently become widely used to assess forest resilience, because it allows a simple, yet highly efficient assessment of short-term responses of trees to drought.
+Nevertheless, not exempt from criticism, this approach needs to be applied carefully to avoid potential bias at different levels [@Schwarzetal2020QuantifyingGrowth].
+In this sense we assessed forest resilience both over the short-term to several recent extreme drought episodes, as well as over the long-term to climate change (*i.e.* warming on the last few decades), using two different proxies to characterize resilience.
+Dendroecological estimates of growth (*i.e.* tree-ring width) are commonly used proxies to characterize tree vitality and have commonly been used to study growth changes in response to drought at the individual tree level [@Fritts1976TreeRings; @Dobbertin2005TreeGrowth].
+Remote sensing can be used to analyze the impact of drought on ecosystems at the stand level [*e.g.* @Zhangetal2013MonitoringEstimating].
+Tree-ring records complement remote-sensing data with a longer time scale.
+Tree rings can reflect tree-growth anomalies induced by climate or other disturbance over decades to centuries [@Babstetal2017ImprovedTreering] and provide an accurate measure of growth responses to droughts [@Bhuyanetal2017DifferentResponses].
 The combination of remote sensing and dendroecology has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerranoetal2013ResponseVegetation; @Coulthardetal2017TreeGrowth], and also to evaluate growth resilience to drought in several tree species [*e.g.* @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness].
 
-In the present study, we assessed resilience of *Quercus pyrenaica* Willd.
-(Pyrenean oak) from southern relict forests at the rear-edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].
-We combined remote-sensing information and dendroecological methods to evaluate the impact of drought both in canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth).
-The analysis of resilience to climate was discussed taken into account the land-use history of these transformed forests.
-To do so we conducted a review of historical documents to reconstruct forest history in our study sites, and to determine how anthropogenic drivers have shaped the current forest structure.
-Based on this discussion of forest resilience, we specifically developed a rationale that integrates the ecological and anthropogenic dimensions of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of oak rear edge populations.
-Our main hypothesis is that these stands will show low resilience to extreme droughts along a small-scale gradient.
+In the present study, we assess resilience of *Quercus pyrenaica* Willd.
+(Pyrenean oak) from southern relict forests at the rear edge of the species distribution, where species performance is considered to be threatened by climate change [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].
+For this, we combined remote-sensing information and dendroecological methods to evaluate the impact of drought both on canopy greenness (as a proxy for primary growth) and radial tree growth (as a proxy for secondary growth).
+For the analysis of forest resilience to climate, we took into account the land-use history of these transformed forests, thoroughly reviewing historical documents to reconstruct forest history at the study sites, and analyzing how anthropogenic drivers have shaped the current forest structure.
+Based on this analysis, we developed a rationale that integrates the ecological and anthropogenic components of marginality to determine the regional and local scale mechanisms shaping the probability of persistence (or extinction) of rear-edge oak populations.
+Our main hypothesis is that range edge stands will show low resilience to extreme droughts, but that the vulnerability to drought will be reduced quickly across a fine-scale topographic gradient of decreasing aridity.
 To test this hypothesis, we: (*i*) quantified how recent extreme drought events influenced primary and secondary growth of *Q.
-pyrenaica* forests at their present geographical rear edge; (*ii*) conducted and explored a review of historical documents to reconstruct forest management history and to infer how it impacted tree-growth in time; (*iii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iv*) and examined differences in the resilience metrics between populations located in contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
-For the latter, we characterized variability in the forest response to drought within the current geographical rear-edge and assessed whether the effect of aspect and environmental conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience, and recovery to extreme drought events.
+pyrenaica* forests at their present geographical rear edge; (*ii*) analyzed the long-term resilience of these forests to extreme drought events, using time-series of radial growth; (*iii*) reviewed historical documents to reconstruct forest-management history and to infer how it impacted tree growth and stand dynamics over time; (*iv*) and examined differences in the resilience metrics between populations under contrasting ecological conditions (*i.e.* xeric *vs.* mesic) along environmental gradients within the rear edge in order to detect vulnerability to climate change at the small spatial scale.
 
@@
 The forests of this species reach their southernmost European limit in Andalusian mountains such as Sierra Nevada (37°N, 3°W), a high-mountain range with elevations of up to 3482 m *a.s.l.*.
-The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability according to elevation and aspect.
+The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought and increasing aridity with decreasing altitude, and marked variability in annual rainfall according to elevation and aspect.
 Sierra Nevada is considered a glacial refuge for deciduous *Quercus* species [@Olaldeetal2002WhiteOaks].
@@
 These relict forests have undergone intensive human use throughout history [@CamachoOlmedoetal2002DinamicaEvolutiva].
-Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, reducing its suitable habitats in the near future [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].
+Furthermore, the conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer from climate change, potentially reducing its suitable habitats in the near future [@GeaIzquierdoetal2013GrowthProjections; @GeaIzquierdoetal2017RiskyFuture].
 
@@
 Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim [@ZangBiondi2015TreeclimPackage].
-The non-climatic disturbance impacts on tree-growth were evaluated using site disturbance chronologies (buit using growth changes on tree growth).
+The non-climatic disturbance impacts on tree-growth were evaluated using site disturbance chronologies (built using growth changes on tree growth).
 
@@
 ## Results
-### Time trends in vegetation greenness
-The analysis of time trends in greenness showed that 78.9% of the EVI pixels followed a positive trend for the 2000-2016 period.
+### Temporal trends in vegetation greenness
+The analysis of temporal trends in greenness showed that 78.9% of the EVI pixels followed a positive trend for the 2000-2016 period.
 The lowest values of EVI standardized anomalies for the study period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 3a).
@@
 The trees of the southern population were older than those from the northern one.
-In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites, despite stand competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a).
+In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites.
+Stand competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a).
 The growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4a and S3a).
@@
 
-### Resilience to drought events at the forest stand and individual-tree levels
+### Resilience, resistance and recovery to drought events at the stand and individual-tree levels
 Resilience and resistance varied in the same direction whereas recovery varied inversely to resilience and resistance.
 During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1).
-The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years.
+The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two drought events.
 For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event.
@@
 ## Discussion
-By using a combined approach of remote-sensing information and dendroecology, we have quantified growth of adult trees and greenness (EVI) as proxies for secondary and primary growth of relict Mediterranean *Quercus pyrenaica* populations in the southern Iberian Peninsula.
-These relict oak populations driven by historical land-use have been resilient to climate change at their present rear edge.
-However, resistance, resilience, and forest recovery to extreme drought events were strongly influenced by mountain exposure, local environmental conditions, and management legacies.
+By using a combined approach of remote-sensing information and dendroecology, we quantified the growth of adult trees and greenness (EVI) as proxies for secondary and primary growth of relict Mediterranean *Quercus pyrenaica* populations in the southern Iberian Peninsula.
+These relict oak populations, driven by historical land-use, have been resilient to climate change at their present rear edge.
+However, resistance, resilience, and forest recovery after extreme drought events were strongly influenced by mountain exposure, local environmental conditions, and management legacies.
 This means that the geographical and the ecological rear edges do not necessarily match and, at a small spatial scale, tree performance can vary markedly along the rear edge under climate change.
 
-### High sensitivity and variability in the oak sensitivity to climate at the rear-edge
+### High sensitivity and variability in the oak sensitivity to climate at the rear edge
 Severe drought negatively affects both primary and secondary growth of *Q.
@@
 Greenness and tree growth were more affected by drought events in drier northern populations than in wetter southern oak populations of Sierra Nevada.
-The northern site showed higher browning intensity than the southern sites during the 2005 drought event, and stronger correlations of tree-growth with SPEI (hydrological year and summer) in the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdoCanellas2014LocalClimate].
+The northern site showed higher browning intensity than did the southern sites during the 2005 drought event, and stronger correlations of tree-growth with SPEI (hydrological year and summer) at the northern site can be interpreted as higher sensitivity to drought at drier sites [@GeaIzquierdoCanellas2014LocalClimate].
 Greenness was less sensitive to drought than tree growth, particularly for drier sites.
 These findings agree with previous works showing tree growth to be a more sensitive metric of forest resilience than is net primary productivity [*e.g.* @Babstetal2013SiteSpecies; @Coulthardetal2017TreeGrowth; @Gazoletal2018ForestResilience; @PenaGallardoetal2018DroughtSensitiveness], suggesting that the growth reduction could be mediated by sink more than by source limitations [@Korner2013GrowthControls; @Fatichietal2014MovingPhotosynthesis].
-Tree-ring records complement remote-sensing data in longer time scales by reflecting tree-growth anomalies induced by climate or disturbance over decades to centuries [@Babstetal2017ImprovedTreering] and provide an accurate measure of growth responses to droughts [@Bhuyanetal2017DifferentResponses; @Gazoletal2018ForestResilience].
-On the other hand, trees at CA-High registered higher BAI than those located at lower elevations (CA-Low and SJ; Figure 4).
-This shows the high variability in the response to climate exhibited along a narrow gradient, which is especially noteworthy for southern sites, as these lie close to each other and both are considered to constitute the rear edge for the species.
+On the other hand, trees at CA-High registered higher BAI than did those located at lower elevations (CA-Low and SJ; Figure 4).
+This shows the high variability in the response to climate exhibited along a narrow gradient, which was especially noteworthy for southern sites, as these lie close to each other and both are considered to constitute the rear edge for the species.
 
@@
 Thus, our results are consistent with those of previous studies highlighting the influence of precipitation on tree-ring growth in different oak species [*e.g.* @Tessieretal1994DeciduousQuercus; @DiFilippoetal2010ClimateChange; @GeaIzquierdoetal2011TreeringsReflect; @GarciaGonzalezSoutoHerrero2017EarlywoodVessel].
-A positive effect of moisture availability and negative of temperature expressing a limiting effect of high vapor pressure deficit and potential evapotranspiration can be expected at drought-limited rear-edges.
-Yet, at the rear-edge, growth of some tree species (*e.g.* *Abies alba*) have been shown to be more sensitive to moisture-related variables [@MartinezSanchoGutierrezMerino2019EvidenceThat], others species were more sensitive to temperatures [*e.g.* *Pinus sylvestris*, @Herreroetal2013VaryingClimate], and growth of other species responded simultaneously to both temperature and moisture related variables [*e.g.* *Fagus sylvatica*, @DoradoLinanetal2017CoexistenceMediterraneanTemperate; @DoradoLinanetal2017ClimateThreats; *Pinus nigra* subsp.
+A positive effect of moisture availability and negative impact of temperature expressing a limiting effect of high vapor-pressure deficit and potential evapotranspiration can be expected at drought-limited rear-edges.
+Yet, at the rear edge, the growth of some tree species (*e.g.* *Abies alba*) has been shown to be more sensitive to moisture-related variables [@MartinezSanchoGutierrezMerino2019EvidenceThat], while others species were more sensitive to temperatures [*e.g.* *Pinus sylvestris*, @Herreroetal2013VaryingClimate], and still other species responded simultaneously to both temperature and moisture-related variables [*e.g.* *Fagus sylvatica*, @DoradoLinanetal2017CoexistenceMediterraneanTemperate; @DoradoLinanetal2017ClimateThreats; *Pinus nigra* subsp.
 *salzmanii*, @SanchezSalgueroetal2012DroughtMain].
@@
 
-### Relict oaks show high resilience to drought at different spatio-temporal scales: do the geographical an ecological rear-edges match?
+### Relict oaks show high resilience to drought at different spatio-temporal scales: do the geographical and ecological rear-edges match?
 Despite the severe drought events in recent decades (Table S3), we found a positive trend for vegetation greenness of *Q.
@@
 
-The high drought-resilience values reported here, coinciding with high values of genetic resilience for those forests on Sierra Nevada reported elsewhere [@ValbuenaCarabanaGil2013GeneticResilience; @ValbuenaCarabanaGil2017CentenaryCoppicing] appear to indicate the strong local adaptation of this oak species.
-Our findings agree with those of studies showing that the assumed higher vulnerability of dry edges does not necessarily hold [*e.g.* @CavinJump2017HighestDrought].
-In our case, this can be explained by the fact that the current geographical rear-edge does not match with the potential ecologically rear-edge for the species because this has been determined mostly by human use.
+The high drought-resilience values reported here, in addition to the potential role of local adaptation [*i.e.* high values of genetic resilience for oak forests on Sierra Nevada; @ValbuenaCarabanaGil2013GeneticResilience; @ValbuenaCarabanaGil2017CentenaryCoppicing], suggest that land-use also has a key role to determine tree resilience to drought and the range edge of species.
+Our findings agree with those of studies showing that the assumed higher vulnerability of current geographical dry edges does not necessarily hold [*e.g.* @CavinJump2017HighestDrought].
+In our case, this can be explained by the fact that the current geographical rear-edge does not match with the potential ecological rear edge for the species because this has been modified and determined mostly by human use.
 @MartinezVilalta2018RearWindow pointed out the importance of local adaptation and plasticity, and also of local environmental factors on the vulnerability shown by rear-edge populations.
-Our results highlight the ample small-scale variability at the ecological boundary and need to better define and delimit the rear-edges.
-All this, together with the characteristic high resprouting ability of the species, show the long-term persistence of these populations  [@BellinghamSparrow2000ResproutingLife].
-It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the high resilience observed is expressed at the species level or only in adult trees.
-The rear-edge might be different for different ontogenic stages.
-It is important to assess whether there is vulnerability if we analyzed seedling regeneration and recruitment, as in other Mediterranean species at their xeric limit [@Castroetal2004SeedlingEstablishment; @VilaCabreraetal2011StructuralClimatic; @GeaIzquierdoetal2015ThisEnd].
+Our results highlight the ample small-scale variability at the ecological boundary and thus the rear edges need to be more clearly defined and delineated.
+All the above points, together with the characteristic high resprouting ability of the species, show the long-term persistence of these populations  [@BellinghamSparrow2000ResproutingLife].
+It should be mentioned that we studied only adult individuals established decades or centuries ago, meaning that it needs to be assessed whether the high resilience found is expressed at the species level (*i.e.* also including regeneration) or only in adult trees.
+The rear-edge might differ for different ontogenic stages.
+It is important to assess whether seedling regeneration and recruitment are vulnerable, as in other Mediterranean species at some locations including their xeric limit [@Castroetal2004SeedlingEstablishment; @VilaCabreraetal2011StructuralClimatic; @GeaIzquierdoetal2015ThisEnd].
 
-### Land-use legacies shape forest sensitivity to climate change of forests and the present rear-edge
-The review of historical documents revealed that forest clearings, firewood removal, charcoal production, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species would approach 90% in tree cover at medium and low elevations [@JimenezOlivenciaetal2015MedioSiglo].
+### Land-use legacies in relation to forest response under climate change and to the present rear edge 
+The review of historical documents revealed that forest clearings, firewood removal, charcoal production, and mining have strongly affected the forests on Sierra Nevada (Table S4), where an estimated historical loss of broadleaf *Quercus* species has approached 90% in tree cover at medium and low elevations [@JimenezOlivenciaetal2015MedioSiglo].
 Together with the analysis of the disturbance chronologies, the observed notable differences in stand structure, tree size, and age suggest different forest histories and a different management origin (*i.e.* land-use legacy) between northern (coppice) and southern populations (high forest, open woodland).
-On the northern slopes of Sierra Nevada (*e.g.* SJ site), land uses have been historically distributed along an elevational gradient: grasslands and shrublands for cattle farming at the highest elevations; then forest stands with some croplands; and irrigated terraces with tree crops at the lowest elevations [@JimenezOlivenciaetal2015MedioSiglo].
+On the northern slopes of Sierra Nevada (*e.g.* the SJ site), land uses have been historically distributed along an elevational gradient: grasslands and shrublands for cattle farming at the highest elevations; next forest stands with some croplands; and, finally, irrigated terraces with tree crops at the lowest elevations [@JimenezOlivenciaetal2015MedioSiglo].
 In addition, other activities such as mining must have altered the forest structure, *e.g.* the SJ site has many small mines and quarries that were exploited intermittently throughout history.
@@
 Firewood, charcoal, and acorns were intensively exploited at the southern sites, until at least the mid-20th century, when these activities sharply declined due mainly to rural abandonment and the use of gas and fossil fuels [@ValbuenaCarabanaGil2013GeneticResilience].
-At the CA-High site, the only positive release event found for the earliest years could be related with conversion from closed forest to an open silvopastoral system, a common management practice often applied in the past in many Iberian oak woodlands [@Canellasetal2004GrowthResponse; @GeaIzquierdoetal2011TreeringsReflect] and which is documented for this site [@ValbuenaCarabanaGil2013GeneticResilience].
+At the CA-High site, the only positive release event found for the earliest years could be related to the conversion from closed forest to an open silvopastoral system, a common management practice often applied in the past in many Iberian oak woodlands [@Canellasetal2004GrowthResponse; @GeaIzquierdoetal2011TreeringsReflect] and which has been documented for this site [@ValbuenaCarabanaGil2013GeneticResilience].
 
@@
 
-In summary, two main results stand out from this research.
-First, the high values of resilience observed in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018RearWindow; @VilaCabreraetal2019RefiningPredictions].
-The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @HerreroZamora2014PlantResponses].
-Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past-management legacies.
-In fact, there was much small-scale variability in the response to climate within the rear edge that we had *a priori* considered in our study.
-Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability in the whole geographical rear edge studied.
-This suggests that the rear edge therefore needs to be redefined in space but also in time [@VilaCabreraetal2019RefiningPredictions], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the potential ecological (limiting) rear-edge of species.
-The rear-edge concept should also consider historical aspects in addition to the geographic, climatic and genetic dimensions [@VilaCabreraetal2019RefiningPredictions], particularly in areas with a long-history of anthropic management, such as Mediterranean mountains.
-Therefore, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate driven population decline at rear edges [@VilaCabreraetal2019RefiningPredictions].
+## Conclusions 
+```{r eval=FALSE, echo=FALSE}
+The ecological and geographical rear edges did not appear to fully match each other in the forest ecosystem studied.
+Severe drought events provoked major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the *a priori* relict, rear-edge forest studied.
+```
+Two main results could be highligthened from our research.
+First, the high values of resilience in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical, but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018RearWindow; @VilaCabreraetal2019RefiningPredictions].
+Contrary to our expectations, the trees exhibited high resilience in the response to drought, particularly over the long-term.
+The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience or enhancing resilience of already established adult individuals [*e.g.* stress tolerance capacity linked to local adaptation; @Lloretetal2012ExtremeClimatic], that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @HerreroZamora2014PlantResponses].
+Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past management legacies.
+In fact, there was much small-scale variability in the response to climate along the rear edge that we had not *a priori* considered in our study.
+The differences found in tree growth, climatic sensitivity and tree resilience between close neighboring sites showed that responses to drought were site dependent and could drastically vary in extremely narrow spatial gradients.
+In other words: in mountains, heterogeneity of ecological conditions at fine scales is the rule, enabling the existence of microrefugia and lengthening species persistence [@Olaldeetal2002WhiteOaks; @SerraDiazetal2015DisturbanceClimate].
+This is particularly relevant to define the real extent and nature (*i.e.* geographical and/or ecological) of rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
+
+The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land use in local forest dynamics, as inferred from an exhaustive review of historical documents.
+This suggest that the rear edge therefore needs to be redefined in space but also in time [@VilaCabreraetal2019RefiningPredictions], partly because of land-use legacies and their effect on the possible mismatch between the current distribution of species (*i.e.* determining the "available" geographical rear edge) and the potential ecological (limiting) rear edge of species.
+The rear-edge concept should also consider historical aspects in addition to the geographic, climatic, and genetic ones [@VilaCabreraetal2019RefiningPredictions], particularly in areas with a long history of human management, such as Mediterranean mountains.
+Therefore, anthropogenic habitat modification and its legacies represent a critical dimension of marginality as they may intensify, confound or delay climate-driven population decline at the rear edges [@VilaCabreraetal2019RefiningPredictions].
+This is relevant for tree species that are highly sensitive to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these forests offer.
+In this sense, it needs to be analyzed the resilience of all demographic stages of species, to assure that the observed resilience in adult trees it is also manifested in its demographic recruitment dynamic expressed by the natural regeneration.
+The rear-edge could also differ for different age cohorts or in seedlings compared to resprouts.
+
+```{r eval=FALSE, echo=FALSE}
+### OLDDd
+## Conclusions AJ
+In summary, this research provides two main results.
+First, the high values of resilience in our study suggest that *Quercus pyrenaica* populations in Sierra Nevada are located in a geographical, but not a climatic, ecological rear edge [*sensu* @MartinezVilalta2018RearWindow; @VilaCabreraetal2019RefiningPredictions].
+Contrary to our expectations, the trees exhibited high resilience in the response to drought, particularly over the long-term.
+The high resilience values observed could also be related to stabilizing mechanisms promoting community resilience or enhancing resilience of already established adult individuals [*e.g.* stress tolerance capacity linked to local adaptation; @Lloretetal2012ExtremeClimatic], that can buffer the impact of extreme events, as has been described for other species [*e.g.* *Pinus sylvestris*, @HerreroZamora2014PlantResponses].
+Second, these resilience responses of oak forest to drought events are not spatially homogeneous throughout the mountain range, due to differences in ecological conditions and/or past management legacies.
+In fact, there was much small-scale variability in the response to climate along the rear edge that we had not *a priori* considered in our study.
+
+Furthermore, we even found positive effects of climate change in certain stands, as discussed, in disagreement with our hypothesis of expecting oak vulnerability along the entire geographical rear edge studied.
+
  Furthermore, even the geographical dimension of the rear edge is not a continuous, uniform space of ecological conditions in mountain landscapes.
-In other words: in mountains, heterogeneity of ecological conditions at fine scales is the rule, enabling the existence of microrefugia and enhancing species persistence [@Olaldeetal2002WhiteOaks; @SerraDiazetal2015DisturbanceClimate].
-In this respect, our results showed that resilience of oak stands to drought events was not spatially homogeneous across the mountain range, due to differences in ecological conditions and/or past-management legacies.
+In other words: in mountains, heterogeneity of ecological conditions at fine scales is the rule, enabling the existence of microrefugia and lengthening species persistence [@Olaldeetal2002WhiteOaks; @SerraDiazetal2015DisturbanceClimate].
+In this respect, our results showed that the resilience of oak stands to drought events was not spatially homogeneous across the mountain range, due to differences in ecological conditions and/or past management legacies.
 This is particularly relevant for rear-edge populations where topographic and biophysical variability can facilitate the existence of refugia.
@@
 ## Concluding comments
-The ecological and geographical rear edges did not appear to fully match in the studied forest ecosystem.
-Severe drought events provoked major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the a priori relict, rear-edge forest studied.
-Nevertheless, contrary to our expectations, the trees exhibited high resilience in the response to drought, particularly in the long-term.
-The differences found in tree growth, climatic sensitivity and tree resilience between close neighboring sites show that responses to drought are site dependent and can drastically vary in very narrow spatial gradients.
-This is particularly relevant to refine the real extension and nature (*i.e.* geographical and/or ecological) of rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
+The ecological and geographical rear edges did not appear to fully match each other in the forest ecosystem studied.
+Severe drought events provoked major reductions in primary and secondary growth of *Quercus pyrenaica* forests in the *a priori* relict, rear-edge forest studied.
+Nevertheless, contrary to our expectations, the trees exhibited high resilience in the response to drought, particularly over the long-term.
+The differences found in tree growth, climatic sensitivity and tree resilience between close neighboring sites showed that responses to drought were site dependent and could drastically vary in extremely narrow spatial gradients.
+This is particularly relevant to define the real extent and nature (*i.e.* geographical and/or ecological) of rear-edge populations where topographic and biophysical variability facilitates the existence of microrefugia.
 The analysis of tree-growth dynamics revealed suppression and release events that were consistent with legacies left by land use in local forest dynamics, as inferred from an exhaustive review of historical documents.
-In consequence, the geographical and ecological rear-edges for specific species do not necessarily meet after centuries or millennia of human impacts on the real world landscapes.
-This is relevant for tree species with a high sensitivity to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these forests offer.
+Consequently, the geographical and ecological rear edges for specific species do not necessarily meet after centuries or millennia of human impact on the landscape.
+This is relevant for tree species that are highly sensitive to climate change, such as *Quercus pyrenaica*, not only for conservation *per se* of the species, but for all ecosystem services that these forests offer.
+```
 
@@
 Bonet on an earlier version of the manuscript.
+We also thank two anonymous reviewers for their thoughtful comments on previous versions of the manuscript that have contributed to improve it significantly.
 AJPL wishes to thank the invaluable support received from his family over the years.
-This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” project.
-We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (No 641762), for their funding support.
+This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” monitoring program http://obsnev.es.
+We thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* for the funding support, and also to H2020 project *European Long-Term Ecosystem and socio-ecological Research Infrastructure (eLTER)* for partial funding.
 
